---
title: "Problem Set 1 Solutions"
subtitle: "QTM 220"
format: pdf
editor: visual
---

```{r}
#| echo: false
library(foreign)
library(ggplot2)
library(tinytex)
#students change working directory? 
#setwd("/GitHub/qtm220/Problem Sets")
```

## Mean and SD Review 

\begin{itemize}
\item Each of the following lists has a mean of 50. For which is the spread of the numbers around the mean biggest? Smallest? 
\begin{itemize}
\item (i) 0, 20, 40, 50, 60, 80, 100
\item (ii) 0, 48, 49, 50, 51, 52, 100  -- smallest
\item (iii) 0, 1, 2, 50, 98, 99, 100  -- biggest
\end{itemize}
\item For each pair of lists below, calculate the mean and SD of each list, then compare within pairs. 
\begin{itemize}
\item (i.a) 1, 3, 4, 5, 7 
\item mean = 4, SD= 2
\item (i.b) 6, 8, 9, 10, 12 
\item mean= 9, SD= 2
\item How is list (i.a) related to list (i.b)? How does this relationship carry over to the mean? the SD? 
\item List b is list a+5 on each entry. The mean increases by 5, but the standard deviation does not change. Adding a constant does not affect the standard deviation, but does change the mean. 

\item (ii.a) 1, 3, 4, 5, 7
\item mean = 4, SD= 2
\item (ii.b) 3, 9, 12, 15, 21 
\item mean= 12, SD= 6
\item How is list (ii.a) related to list (ii.b)? How does this relationship carry over to the mean? the SD? 
\item List b is list a*3. The mean is multiplied by 3 and the standard deviation is also multiplied by 3. Multiplying by a (positive) constant multiplies mean and SD by that number 

\item (iii.a) 5, -4, 3, -1, 7
\item mean = 2, SD= 4
\item (iii.b) -5, 4, -3, 1, -7 
\item mean= -2, SD= 4
\item How is list (iii.a) related to list (iii.b)? How does this relationship carry over to the mean? the SD? 
\item List b is list a with the signs flipped. The sign of the mean flips (and so do all individual deviations from the average), but does not affect the standard deviation 
\end{itemize}

\item Can the standard deviation ever be negative? Explain.

No

\item For set of positive numbers, can the standard deviation ever be larger than the average? Explain. 

Yes. Example: for a set 1,1,16, the mean is 6 and the standard deviation is about 7 

\item Consider the following three histograms. 
\begin{itemize}
\item Match each histogram with the following approximate means: 0.3, 0.5, 0.7. 
\item Match each histogram with the following descriptions: (a) the median about equal to the mean; (b) the median is larger than the mean; (c) the median is less than the mean. 
\item Is the standard deviation of Histogram 2 around 0.015, 0.15, or 1.5?
\item Ture or false (and explain): the standard deviation for Histogram 1 is a lot smaller than for Histogram 2. 
\end{itemize}
\end{itemize}

```{r}
#| echo: false
set.seed(220220)
hist(rbeta(10000,5,2), xlab="X", main="Histogram 1")
mean(rbeta(10000,5,2))
median(rbeta(10000,5,2))


hist(rbeta(10000,2,5), xlab="X", main="Histogram 2")
mean(rbeta(10000,2,5))
median(rbeta(10000,2,5))
sd(rbeta(10000,2,5))

hist(rbeta(10000,5,5), xlab="X", main="Histogram 3")
mean(rbeta(10000,5,5))
median(rbeta(10000,5,5))

```


## Supreme Court Justices

Load the provided dataset "EMdata.csv"

```{=tex}
\begin{itemize}
\item This data is on 27 justices from the Warren (`53 - `69), Burger (`69 - `86), and Rehnquist (`86 - `05) courts
\item The data can be interpreted as a census for the 1953 - 2005 era.  
\item Variables: 
\begin{itemize}
    \item Justice: last name of Supreme Court Justice 
    \item CLlib: Percentage of votes in liberal direction for each justice in civil liberties cases
  \item party: Political party that nominated the justice (Republican =0, Democrat=1)
  \item ur: Justice is a member of an under-represented group, such as a racial or gender minority (under-represented group=1, not in under-represented group=0)
\end{itemize}
\end{itemize}
```
```{r}
#| echo: false
EMdata <- read.csv("EMdata.csv")       
attach(EMdata)
```

(a) Create a histogram of the percentage of liberal votes, visually identifying the mean and median.

```{r}
#| echo: false
hist(CLlib,freq=FALSE,col="blue",ylim=c(-.002,.025),xlab="% Support for Liberal Position on Civil Liberties Cases (CLlib)")
lines(density(CLlib),col="red",lwd=2)
points(x=mean(CLlib),y=-.0015,pch=2,cex=3,lwd=2,col="red")
points(x=median(CLlib),y=-.0015,pch="|",cex=3,lwd=2,col="red")
```

(b) Describe what the mean and median represent in the context of the CLlib variable for these justices. Which is larger? What does that imply in this context?

SOLUTION

(c) Write your own function to calculate the standard deviation of CLlib (i.e. not using "sd") and report it. Add a visual identification of a standard deviation on each side of the mean.

```{r}
y.bar<- mean(CLlib)
s.resid<- rep(0,length(CLlib))
for(i in 1:length(CLlib)){
	s.resid[i] <- (CLlib[i] - y.bar)^2
}
sqrt(sum(s.resid)/(length(CLlib)-1))
sd(CLlib)
```

```{r}
#| echo: false
hist(CLlib,freq=FALSE,col="blue",ylim=c(-.002,.025),xlab="% Support for Liberal Position on Civil Liberties Cases (CLlib)")
lines(density(CLlib),col="red",lwd=2)
points(x=mean(CLlib),y=-.0015,pch=2,cex=3,lwd=2,col="red")
points(x=mean(CLlib)-sd(CLlib),y=-.0015,pch="[",cex=2,lwd=2,col="red")
points(x=mean(CLlib)+sd(CLlib),y=-.0015,pch="]",cex=2,lwd=2,col="red")
```

(d) Calculate the mean of 'ur', whether a justice is a member of an under-represented group. What is the substantive interpretation of this mean?

```{r}
mean(ur)
```

The proportion of justices who are 1s, the proportion of justices who are members of an under-represented group.

(e) Calculate the conditional means and medians of CLlib conditioning on nominating party. Create two histograms of CLlib, one for Republican-nominated justices, one for Democratic-nominated justices. What do the mean and median represent in the context of the CLib variable for the justices of each party? Which is larger, the mean among Republican justices or Democratic justices? Give a substantive interpretation of this. Which is larger, the median among Republican justices or Democratic justices? Give a substantive interpreation of this. Which difference is larger? What does that mean?

```{r}
mean(CLlib[party==0])
median(CLlib[party==0])
mean(CLlib[party==1])
median(CLlib[party==1])
```

```{r}
#| echo: false
hist(CLlib[party==0],freq=FALSE,col="blue",xlim=c(0,100),ylim=c(-.002,.025),xlab="% Support for Liberal Position on Civil Liberties Cases (CLlib)")
points(x=mean(CLlib[party==0]),y=-.0015,pch=2,cex=3,lwd=2,col="red")
points(x=median(CLlib[party==0]),y=-.0015,pch="|",cex=3,lwd=2,col="red")
hist(CLlib[party==1],freq=FALSE,col="blue",xlim=c(0,100),ylim=c(-.002,.025),xlab="% Support for Liberal Position on Civil Liberties Cases (CLlib)")
points(x=mean(CLlib[party==1]),y=-.0015,pch=2,cex=3,lwd=2,col="red")
points(x=median(CLlib[party==1]),y=-.0015,pch="|",cex=3,lwd=2,col="red")
```

SUBSTANTIVE INT.

(f) Create a scatter plot of CLlib with the nominating party as the x-axis. Visually indicate the conditional means (conditioning on party) on the plot.

```{r}
#| echo: false
plot(party,CLlib, xaxt="n",xlab="Party of Appointing President",ylab="% Support for Liberal Position on CL Cases")
axis(side=1, at=c(0,1), labels=c("Rep","Dem"))
points(c(0+.02,1+.02),tapply(CLlib,party,mean),pch=4,col="red",cex=1.25)
```

(g) Using R's lm() function, regress CLlib on party. Using abline(), add the regression line to your scatterplot. What is substantive interpretation of the intercept and slope of the line in terms of justices for each party?

```{r}
lm(CLlib~party)
```

```{r}
#| echo: false
plot(party,CLlib,xlab="Party of Appointing President",ylab="% Support for Liberal Position on CL Cases")
points(c(0+.02,1+.02),tapply(CLlib,party,mean),pch=4,col="red",cex=1.25)
abline(lm(CLlib~party),col="red",lwd=2)
```

SUBSTANTIVE

(h) Calculate the conditional standard deviations of CLlib, conditioning on party. Add a visual represenatation of the conditional standard deviations to your scatterplot.

```{r}
sd(CLlib[party==0])
sd(CLlib[party==1])
```

```{r}
#| echo: false
plot(party,CLlib,xlab="Party of Appointing President",ylab="% Support for Liberal Position on CL Cases")
points(c(0+.02,1+.02),tapply(CLlib,party,mean),pch=4,col="red",cex=1.25)
abline(lm(CLlib~party),col="red",lwd=2)
text(x=0,y=mean(CLlib[party==0])+sd(CLlib[party==0]),labels="]",cex=2,col="red",srt=90)
text(x=0,y=mean(CLlib[party==0])-sd(CLlib[party==0]),labels="]",cex=2,col="red",srt=-90)
text(x=1,y=mean(CLlib[party==1])+sd(CLlib[party==1]),labels="]",cex=2,col="red",srt=90)
text(x=1,y=mean(CLlib[party==1])-sd(CLlib[party==1]),labels="]",cex=2,col="red",srt=-90)
```

(i) Create a scatterplot of justices with under-represented group membership as the y-axis and party as the x-axis. Add the conditional means and standard deviations to the plot (conditioning on party). Using lm(), regress under-represented group membership on party and add the regression line to your plot using abline(). What is the substantive interpretation of the intercept and slope of the line in terms of justices for each party?

```{r}
lm(ur~party)
```

```{r}
#| echo: false
plot(jitter(party,.05),jitter(ur,.05),xlab="Party of Appointing President",ylab="Underrepresented Groups (Ur)")
points(c(0+.02,1+.02),tapply(ur,party,mean),pch=4,col="red",cex=1.25)
abline(lm(ur~party),col="red",lwd=2)
text(x=0,y=mean(ur[party==0])+sd(ur[party==0]),labels="]",cex=2,col="red",srt=90)
text(x=0,y=mean(ur[party==0])-sd(ur[party==0]),labels="]",cex=2,col="red",srt=-90)
text(x=1,y=mean(ur[party==1])+sd(ur[party==1]),labels="]",cex=2,col="red",srt=90)
text(x=1,y=mean(ur[party==1])-sd(ur[party==1]),labels="]",cex=2,col="red",srt=-90)
```

SUBSTANCE

## Matrix Algebra and Least Squares

(a) Let $Y$ be the $n$x1 vector of our outcome variable and $X$ be the $n$x2 matrix that concatenates our independent variable of interest with a row of 1s. $\hat{\beta}$ is a 2x1 vector.

$Y = \begin{bmatrix} y_1 \\ y_2 \\ y_3 \\  \vdots\\  y_n \end{bmatrix}$

$X = \begin{bmatrix} 1 & x_1 \\ 1 & x_2 \\ 1 & x_3 \\ \vdots & \vdots\\  1& x_n \end{bmatrix}$

$\hat{\beta} = \begin{bmatrix} \hat{\beta_0}\\ \hat{\beta_1} \end{bmatrix}$

Define $r$ as the $n$x1 vector of residuals as given by $Y - X \hat{\beta}$

Using matrix algebra, show that $(Y-X\hat{\beta})' (Y-X\hat{\beta})$ yields the sum of squared residuals.

\bigskip

\bigskip

$\hat{r} = \underbrace{(Y_{nx1} - \underbrace{X_{nx2} \hat{\beta}_{2x1}}_{\text{nx1}})}_{\text{nx1}}$ 

$\hat{r}' = \underbrace{(Y_{nx1} - \underbrace{X_{nx2} \hat{\beta}_{2x1}}_{\text{nx1}})'}_{\text{1xn}}$ 

$\hat{r}' * \hat{r} = \begin{bmatrix} 
\hat{r}_1 & \hat{r}_2 & \hat{r}_3 & \cdots & \hat{r}_n \\ 
\end{bmatrix} * \begin{bmatrix} 
\hat{r}_1 \\
\hat{r}_2 \\
\hat{r}_3 \\ 
\vdots \\ 
\hat{r}_n \\ 
\end{bmatrix}$

\ \ 

 $$ = \hat{r}_1 * \hat{r}_1 + \hat{r}_2 * \hat{r}_2  + \hat{r}_3  * \hat{r}_3 + \cdots + \hat{r}_n  * \hat{r}_n =  \sum_{i=1}^n \hat{r}_{i}^{2}   $$ 
 This is the sum of squared residuals. Therefore $(Y-X\hat{\beta})' (Y-X\hat{\beta})$ yields the sum of squared residuals.
 
\bigskip


(b) Using matrix algebra and $\frac{\partial}{\partial \hat{\beta}}$, show that $\hat{\beta} = (X'X)^{-1}X'Y$ minimizes the sum of squared residuals (assuming that $X'X$ is invertible.)

\bigskip
\bigskip

We want to choose $\hat{\beta}$ to minimize $(Y-X\hat{\beta})' (Y-X\hat{\beta})$, the sum of squared residuals. 

$(Y-X\hat{\beta})' (Y-X\hat{\beta}) = Y'Y - \hat{\beta}'X'Y' - Y'X\hat{\beta} + \hat{\beta}'X'X\hat{\beta}$

$Y'Y - 2Y'X\hat{\beta} + \hat{\beta}X'X\hat{\beta}$

First Order Condition $\frac{\partial}{\partial \hat{\beta}} = -2X'Y + 2X'X \hat{\beta} =0$

$-X'Y + X'X\hat{\beta} =0$ 

$X'X \hat{\beta} = X'Y$

assume $(X'X)^{-1}$ exists ($X'X$ is invertible) 

$\hat{\beta} = (X'X)^{-1} X'Y$



(c) Let $Y = \begin{bmatrix} 1 \\ 1 \\ 0 \\ 1 \end{bmatrix}$ and $X = \begin{bmatrix} 1 & 1 \\ 1 & 0 \\ 1 &0\\  1& 1 \end{bmatrix}$

Using matrix algebra, calculate $(X'X)^{-1}X'Y$.

\bigskip



$$X'X = \begin{bmatrix} 1 & 1 & 1 &1 \\ 1 & 0 & 0 &1 \end{bmatrix} \begin{bmatrix} 1 & 1 \\ 1 & 0 \\ 1 &0\\  1& 1 \end{bmatrix} = \begin{bmatrix} 4 & 2\\2 & 2 \end{bmatrix}$$

$$(X'X)^{-1}  = \frac{1}{8-4} \begin{bmatrix} 2 & -2\\-2 & 4 \end{bmatrix}$$

$$(X'X)^{-1} = \begin{bmatrix} \frac{1}{2} & -\frac{1}{2} \\ -\frac{1}{2} & 1 \end{bmatrix}$$

$$X'Y =  \begin{bmatrix} 1 & 1 & 1 &1 \\ 1 & 0 & 0 &1 \end{bmatrix} \begin{bmatrix} 1 \\ 1 \\ 0 \\ 1 \end{bmatrix} = \begin{bmatrix} 3 \\ 2 \end{bmatrix} $$

$$(X'X)^{-1}X'Y = \begin{bmatrix} \frac{1}{2} & -\frac{1}{2} \\ -\frac{1}{2} & 1 \end{bmatrix}  \begin{bmatrix} 3 \\ 2 \end{bmatrix} = \begin{bmatrix} \frac{3}{2} -1 \\ -\frac{3}{2} + 2  \end{bmatrix} = \begin{bmatrix}\frac{1}{2} \\ \frac{1}{2} \end{bmatrix}$$

\bigskip 

(d) In R, create a dataframe $y= c(1,1,0,1)$ and $x = c(1,0,0,1)$. Calculate the conditional means of $y$ when $x=1$ and when $x=0$

```{r}
y<- c(1,1,0,1)
x<- c(1,0,0,1)
df<- data.frame(y,x)
mean(df$y[x==0])
mean(df$y[x==1])
```

(e) Using R's lm() function, regress $y$ on $x$.

```{r}
lm(y~x)
```
