---
title: "Week 2 Homework"
format: html
---

```{r}
#| echo: false
library(ggplot2)
library(purrr)
theme_set(theme_minimal())
```

# NBA Data

A mutual friends gives us some data describing the performance of NBA players in the 2023 season. Actually, they give us different data---they give me 'Sample 1' and you 'Sample 2'. They say these are samples of size 100 drawn with replacement from the population of all players that played in the league that year.  

```{r}
#| echo: true
my.sample <- read.csv("nba_sample_1.csv")
your.sample <- read.csv("nba_sample_2.csv")
```

Theres a fair amount of information about the players in our samples.Here's a list of the variables we have.

- Player - The player's name.
- Team: The team whose this player is playing for this season in abbreviated term.
- Age:  The Age of the player.
- GP: The number of games that the player has played in this season.
- W: The number of games won in which the player has played.
- L: The numer of games lost in which the player has played.
- Min: The minutes the player has played for this season.
- PTS: The number of points made by the player.

## An Estimation Problem

What we want to know is how many points, on average, a player in the league scored over the course of the season. That is, if we call the number of points these players scored $y_1 \ldots y_m$, we want this.
$$
\underset{\text{estimation target}}{\theta} = \frac{1}{m}\sum_{j=1}^m y_j.
$$

We're going to estimate it exactly the same way, but using our sample $Y_1 \ldots Y_n$ of size $n=100$.

$$
\underset{\text{estimate}{\hat\theta} = \frac{1}{n}\sum_{i=1}^n Y_i
$$

Here we're following a convention in which we write our targets as some greek letter. Often $\theta$. And add a 'hat', the thing above the letter theta in $\hat\theta$, to indicate an estimator of that thing.   

My estimate, for what it's worth, is `r mean(my.sample$PTS)`.
```{r}
#| echo: true
Y <- my.sample$PTS
my.estimate <- mean(Y)
my.estimate
```

To put it in context, I plot my points around my estimate.

```{r}
#| echo: true
#|
n <- length(Y)
points.plot <- ggplot(data.frame(i = 1:n, y = Y)) +
    geom_point(aes(x = i, y = y)) +
    geom_hline(aes(yintercept = my.estimate))
points.plot
```

 Here the x-coordinate is irrelevant---they just reflect the
STUFF. Stuff leading to having a histogram.

```{r}
#| echo: true
points.histogram <- ggplot(data.frame(y = Y)) +
    geom_histogram(aes(x = Y))
points.histogram
```



To make this look really nice, we rotate our histogram and lay our two plots out side by side so that the bins of the histogram align with the individual players' points. 
```{r}
points.histogram + coord_flip()
points.plot
```

::: {#exr-your-estimate}

Replicate my analyis using your sample. Report your estimate as well as your two plots.

*Hint*. You should be able to do this with a one-line change to my code.

:::

You notice that my estimate and yours are fairly different. To try to understand what's going on, you want to run a simulation study. You call up our mutual friend and request he give you the data for the whole league. And you get it. Here it is.

```{r}
#| echo: true

our.pop = my.sample # there's some file we want. Call read.csv()
```

::: {#exr-sampling-distribution}
To get a sense of the sampling distribution of the estimator you and I are using, draw 10,000 samples of size $n=100$ with replacement from
this population. Plot it, histogram it, whatever. Then answer this: how weird is your estimate? Do you think it's likely that your sample was really drawn with replacement from this population?
:::


Text about calibrating interval estimate.

::: {#exr-interval-estimate}
To get interval estimates based on our two point estimates,
calculate the width of the middle 95% of our sampling distribution.
Report these two interval estimates and draw them on your plot from 
@exr-sampling-distribution.
:::

Here's some code that'll be useful. What it does is...
```{r}
#| label: width-code
#| echo: true

```

Hopefully this week's lecture has convinced you that this interval will have 95% coverage. But let's check. This'll be useful in the future when what we're doing is a little more complicated and error-prone.

::: {#exr-calculate-coverage}


:::


```{r}
#| echo: true
pops = rep(list(our.pop), 1000)
covers = pops |> map_vec(function(pop) {
  m = nrow(pop)
  n = 100
  
  J = sample(1:m, n, replace=TRUE )
  sam = pop[J,]

  interval.estimate = c(NA, NA)  #<1>
  this.one.covers = NA           #<2>
  this.one.covers
})
coverage.frequency = mean(covers)
```
1. You should have code from @exr-interval-estimate to calculate an interval estimate given a sample.
2. All you need to do here is check whether the estimation target, i.e. the population mean from ..., is between the lower and upper bound of your interval. If you want to check that `t` is between `a` and `b` in **R**, you can say `a <= t & t <= b`.