---
title: "Week 3 Homework"
format: html
---
```{r}
#| echo: false
library(ggplot2)
library(purrr)
library(tidyverse)
library(scales)
theme_set(theme_minimal())
```

::: hidden
$$
\DeclareMathOperator{\RMSE}{RMSE}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\Var}{V}
\DeclareMathOperator{\sd}{sd}
\DeclareMathOperator{\hVar}{\widehat{V}}
\DeclareMathOperator{\bias}{bias}
\newcommand{\thetaprior}{\theta^{\text{prior}}}
\newcommand{\nprior}{n^{\text{prior}}}
\newcommand{\yprior}{y^{\text{prior}}}
$$
:::

# Introduction

## Summary

In our previous homeworks, we've focused mostly on calibration.  We've talked about what our estimators' sampling distributions look like, how to use them to calibrate interval estimates, and what happens when we do. In this one, what we're really focusing on is the accuracy of our point estimators. We'll talk about different ways a point estimator can be untrustworthy, like being biased or highly variable, using our first example of a estimator that actually is biased. A simple one, but one that people really do use. This'll give us the opportunity to think a bit about how to choose between different estimators for the same estimation target. 

After thinking a bit about what we might actually want from an estimator, 
which does involve an element of individual preference, we'll take a moment
to understand *consistency*, the one property that almost everyone agrees 
they do want an estimator to have. Maybe people put a little too much emphasis on this, 
since it's really the bare minimum you could ask for. All that means is that if you're willing to collect any amount of data, you can get as close to the estimation target as anyone might want. Unless you're trying to estimate something really weird, you can afford to have higher standards.
The good news is that when you try to show an estimator is consistent, you usually wind up showing a lot more than that. 
You need to use some pretty weird arguments to show an estimator is consistent without actually getting some quantitative information about how close it gets to the estimation target at different sample sizes. 

The main idea we'll be using to think about all this is our estimators' root-mean-squared error---the square root of its mean-squared distance from the estimation target.^[I like to say 'typical distance' and 'typical error' instead of 'root-mean-squared distance' and 'root-mean-squared error' because it's shorter, gets the essential idea across, and doesn't have a technical meaning that might conflict with my use of it this way like 'average distance' would. I've written it out in full here, but you may find me saying 'typical distance' now and then when I speak.] This doesn't tell us *everything* that the whole sampling distribution does, but it's a lot easier to calculate, especially when we've got a complicated estimator or sampling scheme. As an example, we'll carry out the missing step from Lecture 5's [variance calculation](https://qtm285-1.github.io/assets/lectures/Lecture5.html#/the-standard-deviation-of-a-proportion-1) for a sample proportion when sampling without replacement. 

We'll conclude by returning to the familiar challenge of interval calibration. We'll see that information about root-mean-squared error
alone is enough to calibrate intervals so they get *at least* 95% coverage. We say this approach is *conservative* because we tend 
to have to sacrifice a bit of precision --- we get wider intervals than necessary --- to have greater faith that they do have the
coverage probability we're claiming. In this case, we get this 'extra faith' because we aren't relying on our sampling distribution
looking like the normal distribution or the bootstrap working. We're just relying on our knowledge of mean-squared-error which,
for an estimator we can prove to be unbiased, really just comes down to our ability to estimate our estimator's variance. We'll 
look into exactly how much precision this leaves on the table by comparing them to intervals based on normal approximation and 
the bootstrap using both a simple mathematical argument (@exr-markov-interval) and fake-data simulation (@exr-interval-comparison).

## The Point

Knowing this stuff will help you choose between estimators yourself, 
think about the choices others make, and communicate about these decisions and their implications. It'll help us think about calibration later on, too, because when we know our point estimators are close to the estimation target, we can get away with using approximations like Taylor series to understand their sampling distributions.

# Variance when Sampling Without Replacement

In [Lecture 3](https://qtm285-1.github.io/assets/lectures/Lecture3#/marginalization), 
we talked about the distribution of a sample proportion when we sample *without replacement*. This is, admittedly, a long read. But one thing we can take away from it---and I'll just tell here  rather than asking you to work it out --- is the probability distribution of $Y_1$ when we make a single call and the joint distribution of $Y_1, Y_2$ when we make two calls. Those are just special cases of the 'Partially Marginalized' distribution described in Lecture 3. Letting $m_0$ and $m_1$ be the number of zeros and ones in our binary population $y_1 \ldots y_m$, these are the distributions.

:::: columns
::: {.column width=50%}
$p$ | $Y_1$ 
--|--
$\frac{m_0}{m}$ | 0
$\frac{m_1}{m}$   | 1 

:::
::: {.column width=50%}

$p$ | $Y_1$ | $Y_2$
--|--|--
$\frac{m_0(m_0-1)}{m(m-1)}$  | 0 | 0
$\frac{m_0m_1}{m(m-1)}$      | 0 | 1
$\frac{m_0m_1}{m(m-1)}$      | 1 | 0
$\frac{m_1(m_1-1)}{m(m-1)}$  | 1 | 1

:::
::::

You can go ahead and cross out the $Y_1$ in the first table and write in $Y_i$,
because this is the marginal distribution of every individual observation $Y_i$ when we draw a sample
without replacement of *any size* $n$. You can cross out $Y_1$ and $Y_2$ in the second table and write in $Y_i$ and $Y_j$ for the same reason: it's the joint distribution of any pair $(Y_i, Y_j)$ for $i\neq j$ when we draw a sample without replacement of any size $n$. This makes them a lot more useful.
 
To justify this, we can do a simple two-step thought experiment. We'll think of drawing a sample without replacement as the process of shuffling a deck of cards labeled 1 ... m,  then drawing the first $n$ cards off the top. Letting $J_1$ be what the first card says, $J_2$ what the second one says, and so on, our sample $Y_1 \ldots Y_n$ is $y_{J_1} \ldots y_{J_n}$. Here's our argument.

1. No matter how many cards we're going to draw, the distribution of the first card is the same. Same with the first two cards. Thus, what we've written above is the marginal distribution of $Y_1$ and the joint distribution of  $(Y_1,Y_2)$ when we make any number of calls $n$.

2. If we'd pulled off the first $n$ cards and then counted from the bottom instead of the top, we'd get the same distribution. They're shuffled, after all. So the marginal distributions of $Y_1$ and $Y_n$ are the same and so are the joint distributions of $(Y_1,Y_2)$ and $(Y_n,Y_{n-1})$. The same thing would be true if we went through our top $n$ cards in any order. Thus, the marginal distributions of $Y_i$ are the same for all $i$ and the joint distributions of $(Y_i,Y_j)$ are the same for all $i \neq j$. 

Now let's use this to calculate the variance of our sample proportion $\hat\theta=\frac{1}{n}\sum_{i=1}^{n}Y_i$.
Most of the work is already done on [this slide](https://qtm285-1.github.io/assets/lectures/Lecture5.html#/the-standard-deviation-of-a-proportion-1) from Lecture 5.^[Look at the 'Sampling without Replacement' tab.]
In that calculation, I used the following identity. It's colored blue there to highlight it.

$$
\E\qty[ (Y_i - \E[Y_i])(Y_j - \E[Y_j]) ] = -\frac{\theta(1-\theta)}{m-1} \qfor i\neq j
$$

But I didn't actually show that it's true. That's what we're going to do now. Let's start by rewriting this expression 
so it's a little easier to work with. Like this.

$$ 
\E\qty[ (Y_i - \E[Y_i])(Y_j - \E[Y_j]) ] = \E[Y_i Y_j] - \E[Y_i]\E[Y_j]
$$

::: {.callout-tip collapse="true"}
### Proof of Equivalence
To show that these two expressions are equivalent, 
we 'multiply out'  $(Y_i - \E[Y_i])(Y_j - \E[Y_j])$ to
get four terms, distribute the expectation and pull out constants, 
and cancel some equal-and-opposite terms in the result.

$$
\begin{aligned}
\E\qty[ (Y_i - \E[Y_i])(Y_j - \E[Y_j]) ] 
&= \E\qty[ Y_i Y_j - Y_i\E[Y_j] - Y_j\E[Y_i] + \E[Y_i]\E[Y_j] ] \\
&= \E\qty[ Y_i Y_j ] - \E\qty[ Y_i\E[Y_j] ] - \E\qty[ Y_j\E[Y_i] ] + \E\qty[ \E[Y_i]\E[Y_j] ] \\
&= \E\qty[ Y_i Y_j ] - \E[Y_i]\E[Y_j] - \E[Y_j]\E[Y_i] + \E[Y_i]\E[Y_j] \\
&= \E\qty[ Y_i Y_j ] - \E[Y_i]\E[Y_j]
\end{aligned}
$$
:::

What's nice about this second form is that using our tables above to calculate these expectations is easy.
We actually have a row for $Y_i$ in our marginal table and can easily add a row to our joint table for $Y_i Y_j$.
I'd like to ask you to calculate the thing, but since it'd probably take a litte arithmetic to manipulate what you
get into the form $-\frac{\theta(1-\theta)}{m-1}$ we used in lecture, I'm going to ask you to verify my calculation
instead.

::: {#exr-covariance-with-replacement .callout-exercise}

Using this equivalent form, and the tables above, show that the following identity is true.
$$
\E\qty[ (Y_i - \E[Y_i])(Y_j - \E[Y_j]) ] = \frac{m_1(m_1-1)}{m(m-1)}-\frac{m_1^2}{m^2}
$$

:::

::: {.callout-solution}
$Y_iY_j$ is 1 if both $Y_i$ and $Y_j$ are 1 and otherwise zero, so $\E[Y_iY_j]$ is the probability that both $Y_i$ and $Y_j$ are 1.
Referring to our joint distribution table above, that's $\frac{m_1(m_1-1)}{m(m-1)}$. That's our first term. 
And because we've sampled uniformly-at-random, expected value of $Y_i$ (and $Y_j$) is $\theta=m_1/m$, the population mean,
so the product of the expectations is $\theta^2=(m_1/m)^2$. Subtracting these, as in our new 'equivalent formula' for the
cross term, gives us the result we're looking for.
:::

Simplifying this into the form $-\frac{\theta(1-\theta)}{m-1}$ is a little bit of work. As usual, the trick is to give your two terms a common denominator and see what cancels. 

I'll save you the trouble, but I'll 'fold' it like I usually do solutions, so if you'd like to try it on your own, you can without the solution staring you in the face. If you do just want to read it, expand the box by clicking on it.

::: {.callout-tip collapse="true"}
## Simplifying our Cross-Term
$$
\begin{aligned}
 \frac{m_1(m_1-1)}{m(m-1)}-\frac{m_1^2}{m^2} 
 &= \frac{m_1(m_1-1)m - m_1^2(m-1)}{m^2(m-1)} && \text{ getting a common denominator} \\
 &= \frac{ (m_1^2 - m_1m) - (m_1^2 m - m_1^2)}{m^2(m-1)} && \text{ expanding products} \\
 &= \frac{ m_1^2 - m_1m}{m^2(m-1)} && \text{ canceling equal and opposite terms} \\
 &= \frac{ m_1(m_1 - m)}{m^2(m-1)} && \text{ pulling out common factors} \\
 &= \frac{m_1}{m} \cdot \frac{m_1-m}{m} \times \frac{1}{m-1} && \text{ grouping factors in the numerator and denominator} \\
 &= \theta (\theta - 1) \times \frac{1}{m-1} && \text{ for } \  \theta = \frac{m_1}{m} \\
 &= -\frac{\theta(1-\theta)}{m-1} && \text{ as desired }
 \end{aligned}
$$
:::

# Bias and Consistency 

## Using Prior Information 
So far, we've exclusively talked about *unbiased estimators*. That is, estimators with the property that 
their expected value is equal to the estimation target. 
$$ 
\hat\theta \qqtext{ is called unbiased if } \E\qty[\hat{\theta}] = \theta. 
$$ 
We say an estimator is *biased* if this isn't true. To get a sense of what that means, let's consider a simple
example of a biased estimator. 
$$
\tilde{Y}_1 = \frac{1}{n+1}\cdot \qty{\frac{1}{2} + \sum_{i=1}^{n}Y_i}
$$

::: {#exr-bias-1 .callout-exercise}
Suppose that we've drawn a sample $Y_1 \ldots Y_n$ with replacement from a binary population $y_1 \ldots y_m$ in
which the proportion of ones is $\theta$. 
Calculate the expected value of the estimator $\hat\theta =\tilde{Y}_1$. 
Then calculate its bias $\E[\hat\theta] - \theta$ and its 
standard deviation $\sqrt{\Var[\hat\theta]}$. Report all three.
:::
::: {.callout-solution}
$$
\begin{aligned}
\E[\tilde Y_1] &= 
\E\qty[\frac{\frac{1}{2} + \sum_{i=1}^{n}Y_i}{n+1}]  
= \frac{\frac{1}{2} + \sum_{i=1}^{n}\E[Y_i]}{n+1} = \frac{\frac{1}{2} + n\theta}{n+1} \\
\text{bias} &= \E[\tilde Y_1] - \theta = \frac{\frac{1}{2} + n\theta}{n+1} - \theta = \frac{\frac{1}{2} + n\theta - (n+1)\theta}{n+1} = \frac{\frac{1}{2} - \theta}{n+1} \\
\Var[\tilde Y_1] 
&= \E\qty[ \qty{\frac{\qty(\frac{1}{2} + \sum_{i=1}^{n}Y_i)}{n+1} - \frac{\frac{1}{2} + n\theta}{n+1}}^2 ] \\
&= \E\qty[ \qty{\frac{\sum_{i=1}^{n}(Y_i - \theta)}{n+1}}^2  ] \\
&= \E\qty[ \qty{\frac{n}{n+1} \frac{1}{n}\sum_{i=1}^{n}(Y_i - \theta)}^2  ] \\
&= \qty(\frac{n}{n+1})^2 \E\qty[ \qty{\frac{1}{n}\sum_{i=1}^{n}(Y_i - \theta)}^2 ] \\
&= \qty(\frac{n}{n+1})^2 \Var[\bar Y] = \qty(\frac{n}{n+1})^2 \frac{\theta(1-\theta)}{n} = \frac{n\theta(1-\theta)}{(n+1)^2} \qqtext{ so } \\
\sd(\tilde Y_1) &= \sqrt{\frac{n\theta(1-\theta)}{(n+1)^2}}
\end{aligned}
$$
:::

This is a specific example of a more general estimator that integrates information from a prior study. Maybe a real study or maybe a study we're imagining. Suppose we have $\nprior$ observations $\yprior_1 \ldots \yprior_{\nprior}$ from this study.^[Sometimes we call these *pseudo-observations*. This particular interpretation, in which we think of them as coming from a prior study, can be used to derive this estimator from Bayesian principles.]
$$
\thetaprior = \frac{1}{\nprior}\sum_{i=1}^{\nprior}\yprior_i. 
$$ 

Then averaging all our observations --- from our current study and this prior one --- gives us the following estimator.

$$
\begin{aligned}
\tilde{Y}_{\nprior}
&= \frac{1}{\nprior + n} \qty{\sum_{i=1}^{\nprior}\yprior_i + \sum_{i=1}^{n} Y_i} \\
&= \frac{1}{\nprior + n}  \qty{ \nprior\thetaprior + n\bar Y }
\end{aligned}
$$

We treat these prior observations, and therefore their mean $\thetaprior$, as deterministic. 
The simple example we started with, $\tilde{Y}_1$, is a special case where we have a single prior observation $\yprior_1 = \frac{1}{2}$.

::: {#exr-bias-general .callout-exercise}
Repeat @exr-bias-1 for $\tilde{Y}_{\nprior}$. 
:::

::: {.callout-solution}
$$
\begin{aligned}
\E[\tilde Y_{\nprior}] &= 
\E\qty[\frac{\nprior\thetaprior + \sum_{i=1}^{n}Y_i}{n+\nprior}]  
= \frac{\nprior \thetaprior + n\theta}{\nprior + n} \\
\text{bias} &= \E[\tilde Y_{\nprior}] - \theta = \frac{\nprior \thetaprior + n\theta}{\nprior + n} - \theta = \frac{\nprior \thetaprior + n\theta - (n+\nprior)\theta}{n+\nprior} = \frac{\nprior( \thetaprior - \theta)}{\nprior + n} \\
\Var[\tilde Y_{\nprior}] 
&= \E\qty[ \qty{\frac{\nprior\thetaprior  + \sum_{i=1}^{n}Y_i}{\nprior + n} - \frac{\nprior\thetaprior + n\theta}{\nprior + n}}^2 ] \\ 
&= \E\qty[ \qty{\frac{\sum_{i=1}^{n}(Y_i - \theta)}{\nprior + n}}^2  ]
= \qty(\frac{n}{\nprior + n})^2 \Var[\bar Y] = \frac{n\theta(1-\theta)}{(\nprior + n)^2} \qqtext{ so } \\
\sd(\tilde Y_{\nprior}) &= \sqrt{\frac{n\theta(1-\theta)}{(\nprior + n)^2}}
\end{aligned}
$$
:::


## Visualizating the Impact of Prior Information

Lets's try to get a sense of how using prior information like this
impacts our inference. We'll do it visually. The plot below shows the sampling distributions of three estimators of a population mean $\theta$ at three sample sizes $n$: 10, 40, and 160. These are the estimators.

1. The estimator $\hat\theta_1 = \tilde{Y}_{\nprior}$ for $\thetaprior=3/4$ and $\nprior=10$ prior observations.
2. The estimator $\hat\theta_2 = \tilde{Y}_{n}$ for $\thetaprior=3/4$ and $\nprior=n$ prior observations. The bigger the sample, the more prior observations we use.
3. The sample mean, $\hat\theta_3 = \bar Y$.

As usual, the estimation target $\theta$ is indicated by a green line, the sampling distribution's mean by a solid blue line, and its mean plus and minus two standard deviations by dotted blue lines. In our first exercise, we'll match estimators to pictures.

::: {#exr-bias-visualize .callout-exercise}
What is the correspondence between $\hat\theta_1$, $\hat\theta_2$, and $\hat\theta_3$ described above 
and 'estimator a', 'estimator b', and 'estimator c' from the plot?
:::

::: {.callout-solution}
$\hat\theta_1$ is 'estimator b', $\hat\theta_2$ is 'estimator c', and $\hat\theta_3$ is 'estimator a'.
What gives this away is the bias of each estimator. The bias of estimator 'a' is zero (like the bias of the sample mean $\hat\theta_3$), the bias of estimator 'b' is decreasing as $n$ grows (like the bias $10(3/4-\theta)/(10+n)$ of $\hat\theta_1$), and the bias of estimator 'c' is constant (like the bias $n(3/4-\theta)/2n= (3/4-\theta)/2$ of $\hat\theta_2$).
:::

::: {#fig-3-sampling-distributions}
```{r}
#| echo: false
#| fig-width: 10 
#| fig-asp: 1
set.seed(1234)
n.prior = 10
theta.prior = .75
theta = .5

error = 2 * rnorm(100000)
bias = list(a = function(n) 0, 
            b = function(n) n.prior * (theta.prior - theta) / (n.prior + n), 
            c = function(n) n * (theta.prior-theta) / (n + n))                # n.prior = n
variance = list(a = function(n) theta*(1-theta)/n, 
                b = function(n) theta*(1-theta) * n / (n.prior + n)^2, 
                c = function(n) theta*(1-theta) * n / (n + n)^2)


plotdata = expand_grid(estimator = names(bias), n = c(10, 40, 160)) %>%
    pmap(function(estimator, n) {
        thebias = bias[[estimator]](n)
        thevar = variance[[estimator]](n)
        data.frame(estimator = estimator, n = n, muhat = thebias + error * sqrt(thevar))
    }) |> bind_rows()

summaries = plotdata |>
    group_by(n, estimator) |>
    summarize(mean = mean(muhat), sd = sd(muhat))

breaks = seq(-2/3, 2/3, by = 1/3) 
signs = case_when(
    breaks < 0 ~ "-",
    breaks == 0 ~ "",
    breaks > 0 ~ "+"
)
offsets = case_when(
    breaks != 0 ~ sprintf("%d/3", 3*abs(breaks)),
    breaks == 0 ~ ""
)
labels.abstract = sprintf("θ %s %s", signs, offsets)

ggplot(plotdata) +
    geom_vline(aes(xintercept = theta), data = data.frame(theta=0), color='green',alpha=.4, linewidth=1.5) +
    geom_histogram(aes(x = muhat, y = after_stat(density)), bins = 100, alpha = .2) +
    geom_vline(aes(xintercept = mean), data = summaries, alpha=.4, color='blue') +
    geom_vline(aes(xintercept = mean + 2 * sd), data = summaries, linetype = "dotted", color='blue',alpha=.4) +
    geom_vline(aes(xintercept = mean - 2 * sd), data = summaries, linetype = "dotted", color='blue',alpha=.4) +
    scale_x_continuous(breaks = breaks, labels = labels.abstract, limits = c(-3/4,3/4)) +
    theme(axis.text.x = element_text(angle = -90, vjust = 0.5, hjust = 1)) +
    xlab("") +
    ylab("") +
    facet_grid(n ~ estimator, labeller = labeller(n = label_both, estimator = label_both))
```

:::


## The Bias/Variance Tradeoff

As @fig-3-sampling-distributions shows, there are multiple ways to be a not-so-great estimator.
You can have no bias---or small bias---but so much variance that in many surveys (i.e. many draws from the sampling distribution) you're way off. That's what we see happening with 'estimator a' at sample size $n=10$. 
You can have low variance but comparatively high bias, like we see with 'estimator c' at sample 
sizes $n=40$ and $n=160$. We often use *root-mean-squared-error* to summarize the typical magnitude of
an estimator's error.

$$
\RMSE(\hat\theta) = \sqrt{ \E\qty[ \qty{\hat \theta - \theta}^2 ] }
$$

Famously, we can decompose its square into a sum of squared bias and variance.

$$
\begin{aligned}
\E\qty[ \qty{\hat\theta - \theta}^2 ] 
&= \E \qty[ \qty{ \qty(\hat\theta - \E[\hat\theta]) + \qty(\E[\hat\theta] - \theta) }^2 ] \\
&= \E \qty[ \qty{ \hat\theta - \E[\hat\theta]}^2 ] + 2\E \qty[ \qty{ \hat\theta - \E[\hat\theta]} \cdot \qty{\E[\hat\theta] - \theta} ] + \E \qty[ \qty{ \E[\hat\theta] - \theta }^2 ] \\
&= \Var[\hat\theta] + 2 \times 0 + \underset{\text{bias}(\hat\theta)^2}{\qty{ \E[\hat\theta] - \theta }^2} \\
\end{aligned}
$$

Just to check that you're following the math, do this quick exercise.

::: {#exr-linearity-of-expectation .callout-exercise}
Explain why $\E \qty[ \qty{ \hat\theta - \E[\hat\theta]} \cdot \qty{\E[\hat\theta] - \theta} ] = 0$.
:::

::: {.callout-solution}
It's the expected value of a mean-zero random variable $\qty{ \hat\theta - \E[\hat\theta]}$ times a constant $\qty{\E[\hat\theta] - \theta}$. 
Linearity of expectations justifies rewriting this as that constant times the expectation of the mean-zero random variable, i.e., times $0$.
:::

Now here's a real one. For the first time this semester, we're talking about 
different estimators for the same estimation target. This exercise is an opportunity 
to think about how you might choose between them. I've asked you to answer a few 
questions and sketch a few things to help you think through the issues, but I've tried to
leave this pretty open-ended because this really is a question without a single 
right answer. To some extent, it's about what you think is important. You don't
have to stand by this choice for the rest of your life, so it's okay if you miss
something important and wind up changing your mind later, e.g. when you talk with
your classmates or read the solution. It's really just to get you started thinking
about these kinds of choices and what you might want to consider when making them.

::: {#exr-bias-variance-rmse .callout-exercise}

Suppose we increase our number of prior observations, $\nprior$, 
without changing their mean $\thetaprior$. What happens to 
the bias of the estimator $\tilde{Y}_{\nprior}$? 
E.g. does it stay the same, increase, decrease, increase then decrease, decrease then increase, etc.?
What happens to its standard deviation? And what happens
to its root-mean-squared-error, $\RMSE(\tilde{Y}_{\nprior})$?
Is there a *bias/variance tradeoff* going on?

Thinking of bias, standard deviation, and root-mean-squared error 
as a functions of $\nprior$, sketch all three on the same axes. This sketch
doesn't have to be super precise, but try to convey the general shape
of each function and do identify precisely the location (value of $\nprior$)
and value (height of the function at the location) of any important features,
e.g. minima or maxima of one curve, points where two curves cross, etc.

Having done all this, propose a choice for $\nprior$ and explain why you'd make it.
Your choice can depend on $n$, $\thetaprior$, and $\nprior$, because you know all 
that stuff. But it can't depend on $\theta$. That's something you don't know.

:::

::: {.callout-solution}
Let's recall our formulas for the bias and standard deviation of $\tilde{Y}_{\nprior}$ and sum their squares to get 
its mean squared error.

$$
\begin{aligned}
\text{bias} &= \frac{\nprior(\thetaprior-\theta)}{\nprior+n} \\
\text{sd} &=\frac{\sqrt{n\theta(1-\theta)}}{\nprior+n} \\
\text{rmse}  &= \sqrt{\frac{ (\nprior)^2 (\thetaprior-\theta)^2 + n\theta(1-\theta)}{(\nprior+n)^2}}
\end{aligned}
$$

The bias of this estimator increases in magnitude as $\nprior$ grows.
In particular, it grows more or less linearly when $\nprior$ is small relative to $n$ and level off then slows, taking the value $(\thetaprior-\theta)/2$ for $\nprior=n$ and converging to $\thetaprior-\theta$ as $\nprior$ goes to infinity. The standard deviation of this estimator decreases as $\nprior$ grows, taking on half the value it has for $\nprior=0$ when $\nprior=n$ and converging to zero as $\nprior$ goes to infinity. The root-mean-squared error of this estimator, decreases then increases as $\nprior$ grows. 

I'm going to give you a long, detailed answer for the rest of this question. I wasn't expecting anything like it from you, 
but I thought it might interest you and maybe clarify a few things you might've thought about when you approached the question
yourself. I'll start with precise plots of the bias, sd, and rmse curves for $\thetaprior=.5$ and three values of $\theta$: $.6$, $.55,$ and $.525$. So that we can compare across different values of $\theta$, I'll plot each twice: once on a common set of axes (left) and once with the minimums of $\text{rmse}$ aligned (right).
```{r}
#| layout-ncol: 2
#| fig-width: 5
#| fig-cap:
#|  - bias (red), sd (blue), and rmse (green) as a function of nprior when thetaprior=.5, theta=.6, and n=100
thetaprior = .5
theta = .6
n = 100

bias   = function(np) { np*(thetaprior - theta)/(np + n) } 
stddev = function(np) { sqrt(n*theta*(1-theta)/(n + np)^2) }
rmse   = function(np) { sqrt( np^2*(thetaprior-theta)^2 + n*theta*(1-theta) ) / (n + np) }

x = seq(0,4*n,by=1)
p = ggplot() + 
    geom_line(aes(x=x, y=abs(bias(x))), color="red") + 
    geom_line(aes(x=x, y=stddev(x)), color="blue") + 
    geom_line(aes(x=x, y=rmse(x)), color="green") + labs(x='nprior', y='') + 
    geom_vline(aes(xintercept = theta*(1-theta)/(thetaprior-theta)^2), linetype='dotted')

p + xlim(0, 420)
p + xlim(0,2*theta*(1-theta)/((thetaprior-theta)^2))
```


```{r}
#| layout-ncol: 2
#| fig-width: 5
#| fig-cap:
#|  - bias (red), sd (blue), and rmse (green) as a function of nprior when thetaprior=.5, theta=.55, and n=100
#| 
thetaprior = .5
theta = .55
n = 100

bias   = function(np) { np*(thetaprior - theta)/(np + n) } 
stddev = function(np) { sqrt(n*theta*(1-theta)/(n + np)^2) }
rmse   = function(np) { sqrt( np^2*(thetaprior-theta)^2 + n*theta*(1-theta) ) / (n + np) }

x = seq(0,4*n,by=1)
p = ggplot() + 
    geom_line(aes(x=x, y=abs(bias(x))), color="red") + 
    geom_line(aes(x=x, y=stddev(x)), color="blue") + 
    geom_line(aes(x=x, y=rmse(x)), color="green") + labs(x='nprior', y='') + 
    geom_vline(aes(xintercept = theta*(1-theta)/(thetaprior-theta)^2), linetype='dotted')

p + xlim(0, 420)
p + xlim(0,2*theta*(1-theta)/((thetaprior-theta)^2))
```

```{r}
#| layout-ncol: 2
#| fig-width: 5
#| fig-cap:
#|  - bias (red), sd (blue), and rmse (green) as a function of nprior when thetaprior=.5, theta=.525, and n=100
#| 
thetaprior = .5
theta = .525
n = 100

bias   = function(np) { np*(thetaprior - theta)/(np + n) } 
stddev = function(np) { sqrt(n*theta*(1-theta)/(n + np)^2) }
rmse   = function(np) { sqrt( np^2*(thetaprior-theta)^2 + n*theta*(1-theta) ) / (n + np) }

x = seq(0,8*n,by=1)
p = ggplot() + 
    geom_line(aes(x=x, y=abs(bias(x))), color="red") + 
    geom_line(aes(x=x, y=stddev(x)), color="blue") + 
    geom_line(aes(x=x, y=rmse(x)), color="green") + labs(x='nprior', y='') + 
    geom_vline(aes(xintercept = theta*(1-theta)/(thetaprior-theta)^2), linetype='dotted')

p + xlim(0, 420)
p + xlim(0,2*theta*(1-theta)/((thetaprior-theta)^2))
```

One salient feature is the minimum of the rmse curve, which I'll indicate with a dotted line. We can find it by setting the mean squared error with respect to $\nprior$ to zero and solving for $\nprior$. We use the quotient rule $[f(x)/g(x)]' = [f'(x)g(x) - f(x)g'(x)]/g(x)^2$. Since we're looking for a minimum, we can ignore the denominator. We want $x=\nprior$ where $f'(x)g(x) - f(x)g'(x) = 0$ for $f(x) = x^2 (\thetaprior-\theta)^2 + n\theta(1-\theta)$  and $g(x) = (x + n)^2$.
$$
\begin{aligned}
0 &= f'(x)g(x) - f(x)g'(x) \\ 
&= 2x(\thetaprior-\theta)^2(x+n)^2 - 2x^2(\thetaprior-\theta)^2(x+n) - 2n\theta(1-\theta)(x+n) \\ 
&= 2x(\thetaprior-\theta)^2 (x+n)\qty{(x+n) - x} - 2n\theta(1-\theta)(x+n) \\ 
\end{aligned}
$$
which happens, cancelling the common factor of $2n(x+n)$, for $x=\theta(1-\theta)/(\thetaprior-\theta)^2$. 


Another salient feature is the point where bias and standard deviation are equal, which is indicated by a crossing of the red and blue lines in the plots. We can solve for this as follows.
$$
\frac{\nprior(\thetaprior-\theta)}{\nprior+n} = \frac{\sqrt{n\theta(1-\theta)}}{\nprior+n}
\qqtext { where }
\nprior = \sqrt{n\theta(1-\theta)} / (\thetaprior-\theta)
$$

One interesting thing to notice is that, while both of these points shift to the right as $\thetaprior$ gets closer to $\theta$, the point where rmse is minimized shifts to the right *faster*.  An implication of this that we can see in the plots is that choosing $\nprior$ to minimize rmse will result in an estimator with a larger $\text{bias}/\text{sd}$ ratio when $\thetaprior$ is close to $\theta$ than when it is far away. Why does this matter? Because the bigger the ratio of $\text{bias}/\text{sd}$, the worse our interval estimators' coverage probability will be. When this ratio is roughly 1.96, we'll get about 50% coverage. We see that [here in the slides](https://qtm285-1.github.io/assets/lectures/Lecture6.html#/implications-of-unbiasedness) when discussing implications of bias. So it tells us that choosing $\nprior$ to minimize $\rmse$ is not necessarily a good idea from an inferential perspective. When we do this, poor coverage isn't a price we pay for choosing $\thetaprior$ badly---it's our 'reward' for choosing it well. In light of this, I think choosing $\nprior$ to minimize $\rmse$ is a bad idea. We could try to do something a bit more complicated like choosing $\nprior$ to minimize $\rmse$ subject to a constraint on the bias/sd ratio, e.g., $\text{bias}/\text{sd} \le 1/2$, to ensure close-to-nominal coverage. But what I'd suggest is simpler: just set $\nprior=0$ and use the sample mean. 

All of this assumes we can actually choose $\nprior$ to minimize $\rmse$. It isn't obvious that we can given that our $\rmse$ formula involves the unknown population proportion $\theta$, but you can do something a lot like that by minimizing an *estimate* of $\text{rmse}$. People do this often, although typically in more complex estimation problems. The most popular technique for this is probably [Cross-Validation](https://en.wikipedia.org/wiki/Cross-validation_(statistics)), which we'll talk about later in the semester. [Stein's Unbiased Risk Estimate](https://en.wikipedia.org/wiki/Stein's_unbiased_risk_estimate) is an interesting alternative. We won't get there this semester, but it doesn't require all that much background to understand, so take a look at the wikipedia page if you're curious.

::: 




## Consistency

We say an estimator is *consistent* if it converges to the estimation target
as sample size increases to infinity. That's being vague. There are a couple
ways this is imprecise. 

First, what does it mean to be the same estimator 
at different sample sizes? E.g., the definition of $\hat\theta_2$ from @exr-bias-visualize
depends on $n$, so is that 'an estimator'?  The resolution for this one is easy---if we
really want to be precise, we explicitly specify an estimator for each sample size:
we say 'an estimator sequence' is consistent instead of 'an estimator'. When we say
'an estimator' is consistent, we expect the person we're talking about understands
what sequence of estimators we actually mean. This is just a language thing.

The second thing that's imprecise is what it actually means for a random variable,
like an estimator, to converge to something. It turns out that there are a lot of
different, useful ways to think about this happening. One of the simpler versions
is called *convergence in mean square*. A sequence of random variables $Z_1,Z_2,Z_3,\ldots$
converges in mean square to a a random variable $Z$, which is often but not necessarily a constant,
if the root mean square difference between them, $\sqrt{E[(Z_n - Z)^2]}$, gets arbitrarily small (i.e., converges to zero) 
as $n \to \infty$. And we say an estimator $\hat\theta$ is *consistent in mean-square* if it converges in mean square to 
the estimation target $\theta$ as *sample size* $n$ goes to infinity.

Here are a few exercises to get you thinking about what consistency can and can't look like.

::: {#exr-mean-square-convergence-and-rmse .callout-exercise}
If we know that an estimator $\hat\theta$ is mean-square consistent, do we know
that its root-mean-squared error goes to zero? What about 
its bias and standard deviation?

If we know that an estimator $\hat\theta$ is *not* consistent in mean square, what does that tell us about its root-mean-squared error, bias, and standard deviation?
:::

::: {.callout-solution}
As a matter of definition, an estimator is consistent if (and only if) its root-mean-squared error goes to zero. 
And because mean squared error is the sum of its squared bias and squared standard deviation, this implies that
both its bias and standard deviation go to zero. 

If an estimator is not consistent in mean square, then its root-mean-squared error does not go to zero, 
so *at least one of* its bias or its standard deviation does not go to zero.
::: 

::: {#exr-mean-square-convergence .callout-exercise}
Of the three estimators depicted in @fig-3-sampling-distributions --- estimators 'a', 'b', and 'c' ---
which are mean-square consistent?  Using the figure, explain your reasoning. 

Because consistency is about what happens as $n\to\infty$, it's not really possible to know
whether an estimator is consistent by looking at what happens at 3 sample sizes, like we have in the figure.
You'd have to guess what the next row would look like, and the next, etc. But you don't need to guess. 
You have the actual estimator definitions, albeit mismatched, in @exr-bias-visualize. That said, rather than phrasing your explanation in terms of any formulas you might have from @exr-bias-general, talk about what you see in the figure and what the formulas tell you that *you would see* if we were to add rows for $n$ increasing to $\infty$. 
:::
::: {.callout-solution}
Estimators 'a' and 'b' are mean-square consistent. The bias of 'a' is zero at all sample sizes and the bias of 'b' is decreasing to zero as sample size increases. And both have standard deviations that go to zero as sample size increases. Estimator 'c' is not mean-square consistent because its bias isn't improving with sample size.
:::

# Convergence in Probability

Now let's think about another notion of convergence. If an estimator's sampling distribution
ends up in the right location ($\text{bias} \to 0$) with arbitrarily little spread ($\text{standard deviation} \to 0$),
then it makes sense that *every draw from that sampling distribution* will be close to the estimation target. 
Or almost every draw, anyway.

We tend to visualize our estimate as a single dot, e.g. the black dot below. That's what it is. One number. But 
when we want to think about what estimates we could have plausibly gotten, we tend to think about the dots we'd
get if we were to repeat our survey a hundred times or a billion, each time calculating an estimate the same way.
The sampling distribution of our estimator. We can plot actual dots, like the purple ones below, or histogram them,
to get a sense of what this looks like. Hopefully all this is familiar-verging-on-boring to you by now. 

```{r}
#| echo: false
#| fig-width: 10
K = 100
plotdata$height = rep((3/100)*(1:100), nrow(plotdata)/100)
dotplotdata = plotdata |> filter(n == 40 & estimator == "b")
dotplotsummaries = summaries |> filter(n == 40 & estimator == "b")

ggplot(dotplotdata) +
    geom_histogram(aes(x = muhat, y = after_stat(density)), bins = 100, alpha = .3) +
    geom_vline(aes(xintercept = mean), data = dotplotsummaries) +
    geom_point(aes(x = muhat, y = height),
        alpha = .2,
        color = "purple", data = dotplotdata[c(1:49, 51:K), ]
    ) +
    geom_point(aes(x = muhat, y = height),
        alpha = 1,
        color = "black", data = dotplotdata[50, ]
    ) +
    scale_x_continuous(breaks = breaks, labels = labels.abstract, limits = range(breaks)) +
    theme(axis.text.x = element_text(angle = -90, vjust = 0.5, hjust = 1)) +
    xlab("") +
    ylab("")

```

Whereas convergence in mean square is about the *typical distance* between these dots and our estimation target, *convergence in probability* 
is about the fraction of these dots that falls within a small distance $\epsilon$. We say a random variable $\hat\theta$ converges in probability to $\theta$ if, for anyone's idea of 'sufficiently close', the probability that $\hat\theta$ is sufficiently close to $\theta$ goes to one. 
$$
P(\lvert\hat\theta - \theta\rvert \le \epsilon) \to 1 \qqtext{ as } n \to \infty \qqtext{ for any } \epsilon > 0
$$
Or equivalently, the probability that it isn't sufficiently close goes to zero. That is, if 
$$
P(\lvert\hat\theta - \theta\rvert > \epsilon) \to 0 \qqtext{ as } n \to \infty \qqtext{ for any } \epsilon > 0
$$

If $\hat \theta$ is an estimator and $\theta$ is our estimation target, we say $\hat\theta$
is *consistent in probability* if it converges in probability to $\theta$.^[This is nonstandard terminology. What I'm calling 'consistency in probability' is usually called *weak consistency* for reasons that are a little hard to explain without a fair amount of background in topology and measure-theoretic probability. If you say 'consistent in probability', people who usually say *weak consistency* will know what you mean and maybe even follow your lead, but it may take them a second. 'Convergence in probability' is standard terminology, for what it's worth.]

::: {.callout-note }
### Convergence in Probability and Interval Estimation
You can rephrase consistency in probability as a question about what happens to the coverage probability 
of a sequence of interval estimates $\hat\theta \pm \epsilon$ as sample size $n$ goes to infinity.
$\hat\theta$ is consistent in probability if, for every $\epsilon > 0$, the coverage probability
of the interval estimate $\hat\theta \pm \epsilon$ goes to one as $n$ goes to infinity. Why?
Because $\theta$ is in the interval $\hat\theta \pm \epsilon$ if and only if $\lvert\hat\theta - \theta\rvert \le \epsilon$, 
so the probability that interval covers is same as the probability that $\lvert\hat\theta - \theta\rvert \le \epsilon$.
:::

Convergence in mean square *implies* convergence in probability. Let's use *Markov's inequality* to see why.

## Markov's Inequality

Markov's inequality says that the probability that a non-negative random variable $X$
hits some threshold $\epsilon$ is bounded by the ratio of its expected value and $\epsilon$, i.e.,
$$ 
P(X \geq \epsilon) \leq \frac{\E[X]}{\epsilon} 
$$

Often, instead of using this to bound the random variable we're interested in directly,
e.g. $X=|\hat\theta - \theta|$, we use it to bound the random variable's square. 
$|X| \ge \epsilon$ if and only if $X^2 \ge \epsilon^2$, so the probability that $|X| \ge \epsilon$ is the same as the probability that $X^2 \ge \epsilon^2$. Applying Markov's inequality to the random variable $X^2$ gives us a bound in terms of $X$'s mean square which, in the specific case
that $X$ is $|\hat\theta-\theta|$, is the *mean squared error* of the estimator $\hat\theta$, $\RMSE(\hat\theta)^2=\E[(\hat\theta-\theta)^2]$

$$
P(X \ge \epsilon) = P(X^2 \ge \epsilon^2) \leq \frac{\E[X^2]}{\epsilon^2} \qqtext{ e.g.} 
P(|\hat\theta - \theta| \ge \epsilon) = P((\hat\theta - \theta)^2 \ge \epsilon^2) \leq \frac{\E[(\hat\theta - \theta)^2]}{\epsilon^2}
$$

This tells us that, if the root-mean-squared error of $\hat\theta$ goes to zero, then the probability that $\hat\theta$ is any distance $\epsilon$ away from $\theta$ goes to zero, i.e., consistency in mean-square implies consistency in probability.
 
*Plan*. We're going to conclude with a few exercises. First, we're going to prove Markov's inequality. 
It's good practice working with expectations and probabilities. Then, we're going to look 
at what Markov's inequality actually tells us about the sampling distribution of our estimator
$\hat\theta$ in the terms we're used to---we're going to use it to calibrate an interval estimate.
And we'll see how calibrating this way compares to calibration using the bootstrap and 
normal approximation. 

### Proving Markov's Inequality

The usual proof of Markov's inequality is based on a few simple observations.

1. The expectation of the indicator variable $1_{\ge t}(X)$ is the probability that $X$ exceeds $t$, $P(X \ge t)$. 
2. If we have some function of $u$ that's always larger than $1_{\ge t}$, i.e., 
   one satisfying $u_t(x) \ge 1_{\ge t}(x)$ for all $x$, we know that $\E[u_t(X)] \ge \E 1_{\ge t}(X)$
   for any random variable $X$. If it's always larger for non-negative $x$, then $\E[u_t(X)] \ge \E[1_{\ge t}(X)]$ for any non-negative random variable $X$.
3. The function $u_t(x)=x/t$ is such a function.^[If you're not convinced, sketch the two functions on the same axes. Sketching usually helps.]

::: { #exr-markov-proof  .callout-exercise}
Prove Markov's inequality. 
:::
::: {.callout-solution}
To see that $x/\epsilon \ge 1_{\ge \epsilon}(x)$ for all $x \ge 0$, consider separately the cases 
$x \ge \epsilon$ and $x < \epsilon$. If $x \ge \epsilon$, $1_{\ge \epsilon}(x)=1$ and $x/\epsilon \ge 1$. If $x < \epsilon$, $1_{\ge \epsilon}(x)=0$ and $x/\epsilon > 0$ (for $x > 0$). It follows that if $X$ is any non-negative random variable, then $X/\epsilon \ge 1_{\ge \epsilon}(X)$ and therefore that 
$$
\begin{aligned}
P(X \ge \epsilon) = \E[\int 1_{\ge \epsilon}(X)]  \le \E\qty[\frac{X}{\epsilon}] = \frac{\E[X]}{\epsilon}
\end{aligned}
$$
:::

## Markov's Inequality and Interval Estimation

So far, when we've calibrated interval estimates using our estimator's standard deviation, we've relied on normal approximation. In effect, we've been using a formula for $P(\lvert\hat\theta - \theta\rvert \le \epsilon)$ that's accurate when $\hat\theta$ has a normal distribution and close enough when its distribution is close enough to normal. In this problem, we're going to think about doing without this reliance on approximate normality.

Let's consider $\hat\theta$, an unbiased estimator of $\theta$ with standard deviation $\sigma$, so the 
normal approximation to the distribution of $\hat\theta-\theta$ has the density $f_{0,\sigma}(x)$ below.

$$
P\qty(|\hat\theta - \theta| \le \epsilon) 
\approx \int_{-\epsilon}^{\epsilon} f_{0,\sigma}(x) dx 
\qfor f_{0, \sigma}(x) = \frac{1}{\sqrt{2\pi}\sigma} e^{-x^2/2\sigma^2}
$$

The reason we've been talking about interval estimators of the form $\hat\theta \pm 1.96 \sigma$ is that, if this approximation were perfect, 
these interval estimators would have 95% coverage. That is, it'd be true that $P(|\hat\theta-\theta| \le 1.96 \sigma) = .95$.
And if the approximation is pretty good, we should still expect coverage close to that. But suppose we're not confident that it is.
Markov's inequality allows us to calibrate interval estimates in terms of our estimator's standard deviation 
*without any caveats* about its sampling distribution being approximately normal. Let's give it a shot.

::: {#exr-markov-interval  .callout-exercise}
Let $\hat\theta$ be an unbiased estimator of $\theta$ with standard deviation $\sigma$.
By applying Markov's inequality to $|\hat\theta-\theta|^2$, find a *lower bound* on $P(|\hat\theta-\theta| \le t\sigma)$ as a function of $t$.
For what choice of $t$ is this bound equal to $.95$?  Explain how you can use this to get an interval estimate $\hat\theta \pm t\sigma$
with a coverage probability of at least 95%. 
:::

::: {.callout-tip collapse="true"}
### From Upper Bounds to Lower Bounds.
If you have an upper bound on $P(X > \epsilon)$, then you have a lower bound on $P(X \le \epsilon)$. 
$P(X \le \epsilon)=1-P(X > \epsilon)$ because $X > \epsilon$ if and only if it is not the case that $X \le \epsilon$,
so if $P(X > \epsilon) \le u$, then $P(X \le \epsilon) = 1-P(X > \epsilon) \ge 1-u$. And if you're troubled that 
Markov's inequality says $\ge$ instead of $>$, don't be. You have a bound on $P(X > \epsilon)$ too. 
$P(X > \epsilon) \le  P(X\ge \epsilon)$ because $X \ge \epsilon$ whenever $X > \epsilon$.
:::

::: {.callout-solution}
Applying Markov's inequality to $|\hat\theta-\theta|^2$, which has mean-square $\sigma^2$ because $\hat\theta$ is unbiased, for$\epsilon = (t\sigma)^2$ gives

$$
P(|\hat\theta-\theta|^2 \ge (t\sigma)^2) \le \frac{\E[(\hat\theta-\theta)^2]}{(t\sigma)^2} = \frac{\sigma^2}{(t\sigma)^2} = \frac{1}{t^2}.
$$
The probability that $|\hat\theta-\theta| > t\sigma$ is also no larger than $1/t^2$ because this happens *only if* $|\hat\theta-\theta|^2 \ge (t\sigma)^2$. And it follows that the probability that $|\hat\theta-\theta| \le t\sigma$ is greater than $1-1/t^2$. 

To find the value of $t$ that makes this lower bound equal to $.95$, we solve $1-1/t^2=.95$ to find $t=\sqrt{1/(1-.95)}= \sqrt{20} \approx 4.5$.
And because $\hat\theta \in [\theta - t\sigma, \theta + t\sigma]$ if (and only if) $|\hat\theta-\theta| \le t\sigma$, it follows that this interval has a coverage probability of at least $.95$.

:::

Now let's try this out on the NBA data we've been working with recently.
```{r}
#| echo: true
prior.obs = read.csv("https://qtm285-1.github.io/assets/data/nba_sample_2.csv")
sam = read.csv("https://qtm285-1.github.io/assets/data/nba_sample_1.csv")
pop = read.csv("https://qtm285-1.github.io/assets/data/nba_population.csv") 

indicator = function(W,L,...) { W / (W+L) > 1/2 }

library(purrr)
Y.prior = prior.obs |> pmap_vec(indicator)
Y = sam |> pmap_vec(indicator)
y = pop |> pmap_vec(indicator)

n = length(Y)
m = length(y)
theta = mean(y)
```

The sample $Y_1 \ldots Y_{`r n`}$ we'll use is drawn with replacement from a population $y_1 \ldots y_{`r m`}$
of indicators. These indicators---one for each of the `r m` players who played in the NBA in 2023---are one if 
the player's team won more than half the games they played in and zero otherwise.

We'll consider two point estimators. The first is the sample mean, $\hat\theta=\bar{Y}$.
And the second is the mean-with-prior-observations estimator $\tilde{Y}_{100}$
we talked about in @exr-bias-general, using 100 prior observations from what 
we called 'your sample' in the Week 1 Homework. 

::: {#exr-interval-comparison .callout-exercise}
For each of these point estimators:

1. Plot the bootstrap sampling distribution and a 95% confidence interval calibrated using it.
2. Plot an estimate of the normal approximation to the estimator's sampling distribution and a second 95% confidence interval calibrated using that.
3. Plot a third 95% confidence interval, this time calibrated using Markov's inequality.
4. Taking advantage of your knowledge of the population, plot the estimator's actual sampling distribution and calculate the coverage probability of your three interval estimators.


Referring to your plots, comment on the behavior of your six^[$2\ \text{point estimators} \times 3\ \text{interval calibration methods}=6\ \text{interval estimators}$] interval estimators. If you had to do something like this again, e.g. using data from a different season, which would you choose?  Why?  Would it make a difference if your sample size were larger, e.g. $n=400$ instead of $n=100$? 

Style your plots as you see fit. Overlay the distributions or plot them side-by-side, use color, transparency, annotations, etc. Make it easy to see the stuff that comes up in your comments.

:::

::: callout-note
*Note.* 
Your interval estimates should not use any information about the population other than what's in the sample. If you need a standard deviation or something like that, estimate it using the sample.

:::


::: {.callout-solution}


I've made a separate plot for each of the two point estimators, showing the sampling distribution as a gray bar graph, the bootstrap sampling distribution as an orange bar graph, the normal approximation to the sampling distribution as a black line, and three interval estimates.  I've used the same axis scales and laid them out on top of one another so we can compare location and spread easily. To calibrate my bootstrap interval, I used the function `width` from previous homeworks.

```{r}
set.seed(12345)
```

```{r}
#| echo: false 
width = function(samples, center=mean(samples), alpha=.05) { 
    limits = c(0, 2*max(abs(samples-center)))
    covg_dif = function(w) { 
      actual = mean(center - w/2 <= samples & samples <= center + w/2)
      nominal = 1-alpha 
      actual - nominal  
  }
  uniroot(covg_dif, interval=limits)$root
}
```

```{r}
#| echo: true

library(tidyverse)

theta.prior = mean(Y.prior)
n.prior = length(Y.prior)
prior.obs.estimator = function(Y) { (theta.prior*n.prior + sum(Y)) / (n.prior + n) } 

sampling.distribution = function(estimator) {
    draws = 1:10000 |> map_vec(function(.) {
        J = sample(1:m, size=n, replace = TRUE)
        Y = y[J]
        estimator(Y)
    })
}

bootstrap.sampling.distribution = function(estimator, draws=10000) {
    1:draws |> map_vec(function(.) {
        I = sample(1:n, size=n, replace = TRUE)
        Y.star = Y[I] 
        estimator(Y.star)
    })
}

plot.estimated.sampling.distribution = function(point.estimate, sampling.draws, 
                                                bootstrap.draws, estimated.sd, binwidth) {
    xgrid = seq(point.estimate-5*estimated.sd, point.estimate+5*estimated.sd, 
                length.out = 1000)
    normal.approx = dnorm(xgrid, mean = point.estimate, sd = estimated.sd)
    intervals = data.frame(
        center = rep(point.estimate, 3),
        width = c(width(bootstrap.draws), 2*1.96*estimated.sd, 2*sqrt(20)*estimated.sd),
        method = c("bootstrap", "normal approx", "markov's inequality"),
        y=max(normal.approx)*c(.3,.4,.5))

    ggplot() + 
        geom_bar(aes(x = draws, y=after_stat(prop)/binwidth), 
            alpha = .2, data=data.frame(draws=sampling.draws)) +
        geom_bar(aes(x = draws, y=after_stat(prop)/binwidth), 
            alpha = .4, fill='orange', data=data.frame(draws=bootstrap.draws)) +
        geom_line(aes(x=x, y=density), data=data.frame(x=xgrid, density=normal.approx)) +
        geom_pointrange(aes(x=center, xmin=center-width/2, xmax=center+width/2, 
                            y=y, color=method), data=intervals)
}
scales = list(scale_x_continuous(limits = c(.2,.8)), scale_y_continuous(limits = c(0,16)))
```

```{r}
#| echo: true
#| fig-cap: 
#|  - "The sampling distribution of the sample mean and three interval estimates."
#|  - "The sampling distribution of the mean-with-prior-observations estimator and three interval estimates."

theta.hat = mean(Y)
plot.estimated.sampling.distribution(mean(Y), 
                                    sampling.distribution(mean), 
                                    bootstrap.sampling.distribution(mean), 
                                    sqrt(theta.hat * (1-theta.hat) / n), 1/n) + 
                                    scales + labs(x="", y="")

theta.hat = prior.obs.estimator(Y)
plot.estimated.sampling.distribution(prior.obs.estimator(Y), 
                                    sampling.distribution(prior.obs.estimator), 
                                    bootstrap.sampling.distribution(prior.obs.estimator), 
                                    sqrt(n*theta.hat * (1-theta.hat) / (n.prior+n)^2), 1/(n.prior+n)) + 
                                    scales + labs(x="", y="")
```

To calculate coverage, I ran this code. Because we have to bootstrap our estimator for each sample drawn from our population,
computation time can get out of control if we use 10k samples from the population and 10k draws from the bootstrap sampling distribution as usual. I used 1000 of each, which is about 100 times faster. 
 
```{r}
#| echo: true 
#| cache: true

library(tidyverse)

bootstrap.width = function(estimator) {
    bootstrap.draws = bootstrap.sampling.distribution(estimator, draws=1000)
    width(bootstrap.draws)
}
normal.width = function(estimator) {
    theta.hat = estimator(Y)
    if(identical(estimator, prior.obs.estimator)) {
        2*1.96*sqrt(n*theta.hat * (1-theta.hat) / (n.prior+n)^2)
    } else {
        2*1.96*sqrt(theta.hat * (1-theta.hat) / n)
    }
}

coverage = function(estimator) {
    1:1000 |> map(function(.) {
        J = sample(1:m, size=n, replace = TRUE)
        Y = y[J]
        point.estimate = estimator(Y)

        bootstrap.covers = abs(point.estimate - theta) <= bootstrap.width(estimator)/2 
        normal.covers = abs(point.estimate - theta) <= normal.width(estimator)/2  
        markov.covers = abs(point.estimate - theta) <= (sqrt(20)/1.96)*normal.width(estimator)/2 
        data.frame(bootstrap=bootstrap.covers, normal=normal.covers, markov=markov.covers)
    }) |> bind_rows() |> 
          summarize(across(everything(), mean))
}

mean.coverage = coverage(mean)
prior.obs.coverage = coverage(prior.obs.estimator)
```

  point estimator | bootstrap coverage | normal coverage | markov coverage |
 --- | -------- | -------- | -------- |
$\bar Y$         | `r mean.coverage$bootstrap` | `r mean.coverage$normal` | `r mean.coverage$markov` |
$\tilde Y_{100}$ | `r prior.obs.coverage$bootstrap` | `r prior.obs.coverage$normal` | `r prior.obs.coverage$markov` |


The clear takeway, from an interval estimation perspective, is that calibration via bootstrap or normal approximation is a better idea than calibration using Markov's inequality. As established in @exr-markov-interval, intervals calibrated using Markov's inequality are roughly $\sqrt{20}/1.96 \approx `r round(sqrt(20)/1.96,1)`$ times wider. The question of which point estimator to use is a little more subtle.  Here the prior-observations estimator looks pretty good. We get substantially narrower intervals and the coverage probability of our bootstrap and normal-approximation-calibrated intervals is close to the nominal level---it's 90%. But we know from our discussion in @exr-bias-visualize that using this estimator can be risky. 
We've managed to do well because our prior mean $\thetaprior \approx `r round(theta.prior, 2)`$ was close to the population mean $\theta \approx `r round(theta, 2)`$ and we haven't used it 'too much'. In fact, we're in a pretty similar situation to the I plotted in the last row in my solution to @exr-bias-visualize --- we have $n=100$, $\theta\approx .5$, and $\theta-\thetaprior\approx .025$ --- where using $\nprior=100$ as we have here is a substantial improvement relative to $\nprior=0$ in terms of root-mean-squared error and not-too-unreasonable inferentially because it gave us a $\text{bias}/\text{sd}$ ratio of roughly $1/2$. That said, I wouldn't have been comfortable using it because I wouldn't have known I was this situation without knowing the actual value of the estimation target $\theta$. If it had been the case that $\theta=.58$, which is consistent with what out 95% interval for the sample mean tells us is possible, then we'd have been a similar situation to the one I plotted in the first row in my solution to @exr-bias-visualize --- we'd have had a $\text{bias}/\text{sd}$ ratio of roughly $2$ --- and the result would be 50% coverage. 

Note that here what I'm doing when I talk about these plots is a doing-the-math-by-hand version of testing out an estimator on data I've simulated by drawing samples from a few different fake populations. What I'm saying is, in essence, that the estimator $\tilde Y_{\nprior}$ works well in some populations and not in others, and since I don't know which will happen in my actual population, I don't feel comfortable using it.


I'd have been more comfortable using it at the larger sample size $n=400$, 
since having more actual observations waters down the influence of the $100$ prior observations.
I still wouldn't have chosen it myself, as once the sample size gets large enough that the potential
for bias starts to erode, the interval-narrowness advantage erodes too. Here's a version of the plot I made in @exr-bias-visualize for $n=400$ with $\thetaprior=.5$ and $\theta=.6$, where you can see that at $\nprior=100$ we have a bias/sd ratio of $1$, which is ok but not great, but also a pretty minimal reduction in standard deviation
relative to using the sample mean (i.e. $\nprior=0$). 

```{r}
#| fig-width: 10
#| fig-cap:
#|  - bias (red), sd (blue), and rmse (green) as a function of nprior when thetaprior=.5, theta=.6, and n=400
thetaprior = .5
theta = .6
n = 400

bias   = function(np) { np*(thetaprior - theta)/(np + n) } 
stddev = function(np) { sqrt(n*theta*(1-theta)/(n + np)^2) }
rmse   = function(np) { sqrt( np^2*(thetaprior-theta)^2 + n*theta*(1-theta) ) / (n + np) }

x = seq(0,n,by=1)
p = ggplot() + 
    geom_line(aes(x=x, y=abs(bias(x))), color="red") + 
    geom_line(aes(x=x, y=stddev(x)), color="blue") + 
    geom_line(aes(x=x, y=rmse(x)), color="green") + labs(x='nprior', y='') + 
    geom_vline(aes(xintercept = theta*(1-theta)/(thetaprior-theta)^2), linetype='dotted')

p
```

The difference this larger sample size would make for me is that I'd be less inclined to disagree with someone else's choice to use $\tilde Y_{\nprior}$, as I'd be content knowing that there isn't much difference between it and the sample mean. Here the context does matter. I can be pretty confident that $\theta$ is close to $.5$ because,
while it's not exactly the average win/game ratio of *teams* in the NBA (which has to be $.5$ because one team wins and one loses in each game), it should be pretty close. It's not going to further out than $.4$ or $.6$. If it were plausible that $\thetaprior$ differed from $\theta$ more, we'd have to water down our prior observations with more real ones to ensure the prior observations weren't doing much.

:::
