---
title: "Week 1 Homework"
subtitle: "QTM 220"
---

```{r}
#| echo: false
library(ggplot2)
theme_set(theme_minimal())
```

# Mean and Standard Deviation Review

## Calculations

::: {#exr-calculations-1 .callout-exercise}
Each of the following lists has a mean of 50. For which is the standard deviation biggest? Smallest? 

  1. 0, 20, 40, 50, 60, 80, 1002.
  2. 0, 48, 49, 50, 51, 52, 100 
  3. 0, 1, 2, 50, 98, 99, 100  

:::

::: {#exr-calculations-2 .callout-exercise}
For the two lists below, calculate the mean and standard deviation of the numbers. Then compare the your answers for the two lists. Think about this comparison. How is the first list related to the second? How does this relationship carry over to the mean? The standard deviation? 

 
  1. 1, 3, 4, 5, 7 
  2. 6, 8, 9, 10, 12 

:::

::: {#exr-calculations-3 .callout-exercise}
Repeat this exercise for a new pair of lists. 

  1. 1, 3, 4, 5, 7 
  2. 3, 9, 12, 15, 21

:::

::: {#exr-calculations-4 .callout-exercise}
Repeat it again. 

  1. 5, -4, 3, -1, 7
  2. -5, 4, -3, 1, -7 

:::


## Properties

::: {#exr-properties-1 .callout-exercise}
Can a standard deviation ever be negative? Explain.
:::

::: {#exr-properties-2 .callout-exercise}
For set of positive numbers, can the standard deviation ever be larger than the average? Explain. 
:::


## Visualization

```{r}
#| echo: false
set.seed(220220)
sample.1 = rbeta(10000, 5, 2)
sample.2 = rbeta(10000, 2, 5)
sample.3 = rbeta(10000, 5, 5)
```

Consider the three histograms below.

::: {#exr-visual-1 .callout-exercise}
The means of the samples we've histogrammed are approximately `r round(mean(sample.2),1)`,
`r round(mean(sample.1),1)`, `r round(mean(sample.3),1)`. Which histogram corresponds to which mean?
:::

::: {#exr-visual-2 .callout-exercise}
True or false: the standard deviation of the sample summarized by Histogram 1 is a lot smaller than the one summarized by Histogram 2. Explain. 
:::

```{r}
#| echo: false
#| warning: false 
#| layout-ncol: 3
#| fig-cap:
#|  - "Histogram 1"
#|  - "Histogram 2"
#|  - "Histogram 3" 

ggplot() + geom_histogram(aes(x=sample.1, y=after_stat(density)), bins=100, alpha=.3, color='black') + xlim(0,1) 
ggplot() + geom_histogram(aes(x=sample.2, y=after_stat(density)), bins=100, alpha=.3, color='black') + xlim(0,1)
ggplot() + geom_histogram(aes(x=sample.3, y=after_stat(density)), bins=100, alpha=.3, color='black') + xlim(0,1)
```

# Supreme Court Justices

## The Data
Load the provided dataset "EMdata.csv". To do this, start an R session in the same folder as the csv file, then say 

```{r}
#| echo: true
EMdata = read.csv("EMdata.csv")
```

- This data is on 27 justices from the Warren ('53 - '69), Burger ('69 - '86), and Rehnquist ('86 - '05) courts
- The data can be interpreted as a census of justices for the 1953 - 2005 era.  Each row is a justice and each column is a variable. The column 'justice' is the name of the justice. We'll be looking at a few other variables.
   - CLlib: The percentage of votes in liberal direction for each justice in civil liberties cases
   - party: the political party that nominated the justice (Republican =0, Democrat=1)
   - ur:   the justice is a member of an under-represented group, such as a racial or gender minority (under-represented group=1, not in under-represented group=0)
- To get you started, I'm going to plot a histogram of the percentage of liberal votes, identifying the mean with a vertical line. You may want to edit this code to answer future questions.

```{r}
#| echo: true

CL.histogram = ggplot(EMdata) + 
           geom_histogram(aes(x=CLlib, y=after_stat(density)), 
                          bins=10, alpha=.3, color='black') +  
           geom_vline(aes(xintercept=mean(CLlib)), 
                          color="blue") +
           xlab("% Support for Liberal Position on Civil Liberties Cases (CLlib)")

CL.histogram
```

::: { #exr-justices-calc .callout-exercise}

Write your own function to calculate the standard deviation of CLlib (i.e. not using "sd") and report it. Use 'sd' to check your answer.^[You may be slightly off. In particular, you may be off by a factor of `r n=nrow(EMdata); n-1`/`r n`. That's
ok. There are two conventions for calculating the sample standard deviation---one involves division by $n$ and the other $n-1$.]
Add, to the plot above, vertical lines indicating the mean plus and minus 1 and 2 standard deviations. 
identification of a standard deviation on each side of the mean. What is the substantive interpretation of this mean?

:::


::: {#exr-justices-histogram-ur .callout-exercise}
Replicate the plot above, i.e. a histogram with lines for the mean plus or minus two standard deviations, for the variable 'ur'. What is this mean? And what is its substantive interpretation?
:::

::: {#exr-justices-histogram-clib-party .callout-exercise}
Draw two histograms of CLlib, one for Republican-nominated justices and one for Democratic-nominated justices. Calculate the mean of CLlib in each group. Which is larger, the mean among Republican justices or Democratic justices? Give a substantive interpretation of this difference.
:::

::: {#exr-justices-scatter .callout-exercise}

Below, I've drawn a scatter plot of CLlib with the nominating party as the x-axis. I've used `stat_summary` to draw in the mean plus and minus two standard deviations for each group. Calculate these means and standard deviations and report the mean and endpoints of the intervals that have been drawn in. Check your answer by visually for agreement with the plot. Can you give a substantive interpretation of these intervals?
:::

```{r}
#| echo: true

ggplot(EMdata) +  
  geom_point(aes(x=party, y=CLlib), 
             position=position_jitter(w=.1, h=0), alpha=.4) + 
  stat_summary(aes(x=party, y=CLlib), 
              fun=mean, 
              fun.min = function(x) mean(x) - 2*sd(x), 
              fun.max = function(x) mean(x) + 2*sd(x), 
              geom="pointrange") + 
  xlab("Party of Appointing President") + 
  ylab("%Support for Liberal Position on CL Cases")
```


::: {#exr-justices-scatter-ur .callout-exercise}
Repeat the exercise above, using the variable 'ur' instead of 'CLlib'. Draw your own plot.
:::
