---
title: "Week 5 Homework"
format: html
---
```{r}
#| echo: false
library(ggplot2)
library(purrr)
library(tidyverse)
library(scales)
theme_set(theme_minimal())
```

::: hidden
$$
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\Var}{V}
\DeclareMathOperator{\hVar}{\widehat{V}}
\DeclareMathOperator{\bias}{bias}
$$
:::

# Review of Linear Approximation and Gradient

In some questions of this homework we would rely on the idea of *Linear Approximation*, which helps us approximate the value of a function near some specific point. Given a function $f(x)$, the linear approximation $L(x)$ of the function at $x = c$ for some constant $c$ can be written as:
$$
L(x) = f(c) + f'(c)(x-c)
$$

In a higher dimension (let's say in the 2-d space), the multivariable linear approximation $L(x, y)$ of function $f(x)$ at $(p,q)$ can be written as:
$$
\begin{aligned}
L(x,y) & = f(p,q) + \nabla f(p,q)\begin{bmatrix}x-p \\ y-q\end{bmatrix}
\\
&= f(p,q)+[\frac{\partial f}{\partial x}(p,q)]\cdot(x-p)+[\frac{\partial f}{\partial y}(p,q)]\cdot(y-q)
\end{aligned}
$$

And the gradient of a function can be calculated by taking *partial derivatives* with respect to each "direction" it possibly goes, that for some $p = (x_1, \ldots, x_n)$,
$$
\nabla f(p) = \nabla f\begin{pmatrix}x_1 \\ \vdots \\ x_n \end{pmatrix} = \begin{bmatrix}
\frac{\partial f}{x_1}(p) \\
\vdots\\
\frac{\partial f}{x_n}(p)
\end{bmatrix}
$$

*Linear Approximation* is basically the *First-Order Taylor Expansion*, both of which use the function value at a certain point and its first derivative to approximate the function value near that point. We will use that fact to derive more in the following exercises.

# More About Width 
In our previous homework, we have done analyzing the width of interval estimate both visually and computationally. Specifically, 
$$
\begin{aligned}
    |\hat{\theta} - \theta| 
    &\leq \frac{2\sqrt{\theta (1-\theta)}}{\sqrt{n}}\\
    &\leq \frac{1}{\sqrt{n}} \text{, where }\sqrt{\theta(1-\theta)}\leq\frac{1}{2}
\end{aligned}
$$

Now we know that the interval estimator centered around $\hat{\theta}$ and with length $\frac{2}{\sqrt{n}}\sqrt{\theta (1-\theta)}$ can be written as
$$
\hat{\theta} \pm \frac{2}{\sqrt{n}}\sqrt{\theta (1-\theta)}
$$

We can also visualize the relationship between $\theta$ and width to get a sense of how width would perform as $\theta$ changes: 
```{r}
#| echo: false
#| warning: false
#| layout-ncol: 3
#| fig-height: 3
#| fig-width: 3.5

width.binom <- function(theta, n, alpha = .05) {
    if (theta == 0 || theta == 1) {
        return(0)
    }
    covg.dif <- function(w) {
        actual <- pbinom(n * (theta + w / 2), n, theta) -
            pbinom(n * (theta - w / 2) - 1, n, theta)
        nominal <- 1 - alpha
        actual - nominal
    }
    limits <- c(0, 1)
    uniroot(covg.dif, interval = limits)$root
}

thetas <- seq(0, 1, by = 0.01)
widths.100 <- sapply(thetas, function(theta) width.binom(theta, n = 100, alpha = 0.05))
widths.500 <- sapply(thetas, function(theta) width.binom(theta, n = 500, alpha = 0.05))
widths.1000 <- sapply(thetas, function(theta) width.binom(theta, n = 1000, alpha = 0.05))

df.100 <- data.frame(theta = thetas, width = widths.100)
df.500 <- data.frame(theta = thetas, width = widths.500)
df.1000 <- data.frame(theta = thetas, width = widths.1000)

ggplot(df.100, aes(x = theta, y = width)) +
    geom_line() +
    labs(
        x = "theta", y = "width",
        title = "The width of a calibrated 95% CI as \na function of theta when n = 100."
    ) +
    theme_minimal()

ggplot(df.500, aes(x = theta, y = width)) +
    geom_line() +
    labs(
        x = "theta", y = "width",
        title = "The width of a calibrated 95% CI as \na function of theta when n = 500."
    ) +
    theme_minimal()

ggplot(df.1000, aes(x = theta, y = width)) +
    geom_line() +
    labs(
        x = "theta", y = "width",
        title = "The width of a calibrated 95% CI as \na function of theta when n = 1000."
    ) +
    theme_minimal()

```

::: {#exr-width-change  .callout-exercise}
Consider both computational and visual illustration of the relationship between $\theta$ and width. In both ways explain why our interval would not change too much near $\theta = \frac{1}{2}$.
:::

::: {#hint-width-change  .callout-tip collapse="true"}
In your calculation, use the First-Order Linear Approximation of $\sqrt{\theta(1-\theta)}$ at the point where $\theta = \frac{1}{2}$ to help you derive your conclusion. Refer to the first section for the computational details of the approach.

To help with, we can first consider the derivative of the width function with respect to $\theta$:
$$
\frac{\partial}{\partial \theta}\left(\frac{2}{\sqrt{n}}\sqrt{\theta(1-\theta)}\right) = \frac{2}{\sqrt{n}}\cdot\frac{1-2\theta}{2\sqrt{\theta(1-\theta)}}
$$

With that, we can then derive an expression that approximate the width function value at $\theta = \frac{1}{2}$
:::

:::{#exr-width-change-general  .callout-exercise}
Now, to put it in a more general sense, instead of expanding at $\theta = \frac{1}{2}$, derive a more general expansion around any $\theta$ that tells us its approximation of the width, and use the expression to explain how this affects the sensitivity of the width to changes of $\theta$ as $n$ increases, or how the sample size would impact the relationship between $\theta$ and the width.
:::

# More About Variance of Difference in Means

Remember in lecture that we've explained that for the unbiased estimators, $\hat{\mu}(1)$ and $\hat{\mu}(0)$,
$$
\Var[\hat{\mu}(1)-\hat{\mu}(0)] = \E[\frac{1}{N_1}\sigma^2(1)+\frac{1}{N_0}\sigma^2(0)] \text{ for } N_x = \sum_{i}1_{=x}(X_i)
$$

And here is how we derived such expression:
$$
\begin{aligned}
    \Var[\hat{\mu}(1)-\hat{\mu}(0)] 
    &= \E[\{\hat{\mu}(1) - \hat{\mu}(0)\} - \{\mu(1)-\mu(0)\}]^2 \\
    &= \E[\{\hat{\mu}(1) - \mu(1)\} - \{\hat{\mu}(0) -\mu(0)\}]^2 \\
    &= \E[\{\hat{\mu}(1) - \mu(1)\}]^2 + \E[\{\hat{\mu}(0) -\mu(0)\}]^2 - 2\E[\{\hat{\mu}(1) - \mu(1)\}\{\hat{\mu}(0) -\mu(0)\}]
\end{aligned}
$$

If we omit the cross-multiplying term at the end: $2\E[\{\hat{\mu}(1) - \mu(1)\}\{\hat{\mu}(0) -\mu(0)\}]$, we know what the leftover represents: this is just our *Variance of Difference in Means* we've lined out above:
$$
\E[\{\hat{\mu}(1) - \mu(1)\}]^2 + \E[\{\hat{\mu}(0) -\mu(0)\}]^2 = \E[\frac{1}{N_1}\sigma^2(1)+\frac{1}{N_0}\sigma^2(0)]
$$

But the only way to justify our argument above is to prove that the cross-multiplying term equals to zero, which becomes our next exercise.

:::{#exr-variance-crossterm  .callout-exercise}
Complete the proof argument by showing the cross term is zero:
$$
2\E[\{\hat{\mu}(1) - \mu(1)\}\{\hat{\mu}(0) -\mu(0)\}] = 0
$$
:::

::: {#hint-width-change  .callout-tip collapse="true"}
Try expressing $\hat{\mu}(1)$ and $\hat{\mu}(0)$ using the "indicator trick" we've talked about long time ago and consider applying their properties. What would happen when the cross item index $i=j$ and when $i \neq j$? Would the independence assumption among individual observations plays a role in any case?
:::

# How to Estimate *Ratio Between Means*

In the lecture, and the question above, we've investigated a mutant estimator from the pure $\hat{\mu}$, *Difference in Means* across two groups. But what if we would like to estimate the *Ratio* between means across groups? Specifically, we want to construct that 
$$
\begin{align}
\text{Estimate } \frac{\mu(1)}{\mu(0)}&& \text{with }\frac{\hat{\mu}(1)}{\hat{\mu}(0)}
\end{align}
$$

But would this estimator be biased? To see this, we can again use *Linear Approximation* or *First-Order Taylor Series Expansion*. 

Now we write our estimand to be a function $f$ such that $f(\mu(1), \mu(0)) = \frac{\mu(1)}{\mu(0)}$. Then the First-Order Taylor Expansion of $f$ around the point $(\mu(1), \mu(0))$ evaluated at $(\hat{\mu}(1), \hat{\mu}(0))$ is:
$$
f(\hat{\mu}(1), \hat{\mu}(0)) = f(\mu(1), \mu(0)) + \nabla f(\mu(1),\mu(0))\cdot \begin{pmatrix}
\hat{\mu}(1)-\mu(1) \\ \hat{\mu}(0)-\mu(0)
\end{pmatrix},
$$

where $\nabla f(\mu(1),\mu(0))$ is the gradient of $f$ evaluated at $(\mu(1), \mu(0))$. 

:::{#exr-ratio-estimate  .callout-exercise}
Now complete the calculation above by computing the gradient of $f$ and substitue it back to the expansion and unfold the function $f$. Then, examine whether the expanded term is unbiased.
:::

::: {#hint-ratio-estimate  .callout-tip collapse="true"}
1. Use the information provided at the beginning for your calculation.
2. What can we infer about this new estimator if we know $\hat{\mu}$ is unbiased estimator of $\mu$?
:::

