---
title: "Week 7 Homework"
format: html
---
::: hidden
$$
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\Var}{V}
\DeclareMathOperator{\hVar}{\widehat{V}}
\DeclareMathOperator{\bias}{bias}
$$
:::

# Summary
{{< include shared-code.qmd >}}
{{< include ca-income.qmd >}}
```{r}
set.seed(1)
theme_set(whitebw.theme())
```

In this one, we're going to review some stuff that people mixed up on the exam and work on visualization and computation for complex summaries of data with multiple and multivalued covariates. 


# Review 

```{r}
set.seed(123)
m = 10000000 
y = rnorm(m) + (6*rbinom(m, size=1, prob=.5)-3)
y = y - mean(y)

n = 10000
J = sample(1:m, size=n, replace=TRUE)
Y = y[J]
```
```{r}
theta.hat.star = 1:10000 |> map_vec(function(.) { 
  I = sample(1:n, size=n, replace=TRUE)
  Y.star = Y[I]
  mean(Y.star)
}) 
```

::: {#fig-histogram-data}
```{r}
#| layout-ncol: 2
#| fig-width: 5.5
#| fig-cap:
#|  - A histogram of the population, $y_1 \ldots y_{m}$ for $m=10,000,000$.
#|  - A histogram of the sample, $Y_1 \ldots Y_n$ for $n=10,000$.
ggplot() + geom_histogram(aes(x=y), bins=100, alpha=.3) + 
  geom_vline(xintercept=mean(y), color="blue") + 
  geom_vline(xintercept=mean(Y)-sd(Y), color="blue", linetype="dashed") + 
  geom_vline(xintercept=mean(Y)+sd(Y), color="blue", linetype="dashed") + 
  scale_x_continuous(limits=c(-8,8), breaks=seq(-8,8,by=2))
ggplot() + geom_histogram(aes(x=Y), bins=100, alpha=.3) + 
  geom_vline(xintercept=mean(Y), color="blue") + 
  geom_vline(xintercept=mean(Y)-sd(Y), color="blue", linetype="dashed") + 
  geom_vline(xintercept=mean(Y)+sd(Y), color="blue", linetype="dashed") + 
  scale_x_continuous(limits=c(-8,8), breaks=seq(-8,8,by=2))
```
A histogram of the sample $Y_1 \ldots Y_{10000}$. The mean of each distribution is marked with a blue line
and the mean plus and minus one standard deviation are marked with dashed blue lines.
:::

::: {#fig-histogram-theta-star}
```{r}
#| layout-ncol: 3
#| fig-cap:
#|  - The sampling distribution of estimator A.
#|  - The sampling distribution of estimator B.
#|  - The sampling distribution of estimator C.
ggplot() + geom_histogram(aes(x=theta.hat.star/2), bins=100, alpha=.3) + 
  geom_vline(xintercept=mean(theta.hat.star/2), color="blue")  
ggplot() + geom_histogram(aes(x=Y), bins=100, alpha=.3) + 
  geom_vline(xintercept=mean(Y), color="blue")  
ggplot() + geom_histogram(aes(x=1+theta.hat.star/2), bins=100, alpha=.3) + 
  geom_vline(xintercept=mean(1+theta.hat.star/2), color="blue")  
```
The bootstrap sampling distributions of our three estimators. 
The mean of each distribution is marked with a blue line.
:::

In @fig-histogram-data, I've drawn two histograms. The first shows a population $y_1 \ldots y_m$ with mean $\mu = \frac1m \sum_{j=1}^m y_j$ and variance $\sigma^2 = \frac1m \sum_{j=1}^m (y_j - \mu)^2$. The second shows a sample $Y_1 \ldots Y_n$ drawn with replacement from this population---the sample that was used in Problem 1 of Midterm 1.  

In @fig-histogram-theta-star, I've drawn the bootstrap sampling distributions of three estimators of the population mean. 

1. $\hat\mu = \bar{Y}/2$ for $\bar{Y} = \frac1n \sum_{i=1}^n Y_i$
2. $\hat\mu = 1 + \bar{Y}/2$.
3. $\hat\mu = Y_2$.

::: {#exr-match-theta-hat-star .callout-exercise}
Match each estimator to the plot of its bootstrap sampling distribution. 
That is, say something like `1:A`, `2:B`, `3:C`. But not exactly that.
:::

::: {#exr-mean .callout-exercise}
For each of these estimators, say whether it is *unbiased* and whether it is *consistent*.

*Tip*.
Each part of this is a Yes or No question. There are estimators that are unbiased or consistent for some populations and not for others, so if I were asking about a population I'd told you nothing concrete about, the answer might be *maybe*. But I'm asking about the specific population plotted above and the plot does give you enough information to determine the answer. 
:::

::: {#exr-mean-vis .callout-exercise}
For any estimator you described as *biased* or *inconsistent*, 
describe what you see in the plots above that suggests this is
the case.
:::

::: {#exr-variance .callout-exercise}
For each of the three estimators, do the following.

1. Write a formula for its standard deviation. Your formula may involve the sample size $n$ and some summaries of the population, e.g. $m$, $\mu$, or  $\sigma$. 
2. Using information shown in @fig-histogram-data, calculate it approximately and report the number. 
3. Check that your answer looks about right using @fig-histogram-theta-star and *describe how you checked*. 
:::


# Complex Summaries and Covariate Shift 

In this section, we'll work with a subset of the California income data considered in lecture. We'll look at incomes of CA residents who responded to the Current Population Survey in 2022, were in the age range 25-25, and graduated from high school. I've plotted this data below. As in lecture: 

 - [$W_i=0$ for male residents]{.groupa} and [$W_i=1$ for female residents]{.groupb}; 
 - $X_i$ is years of schooling.
 - $Y_i$ is income in dollars.

```{r}
ca = read.csv('../lectures/data/earnings.csv')
```

```{r}
xco = function(w,x, shift=.15) {
  w=as.numeric(w)
  x + shift*(2*w-1)
}

sam.dot = list(size=1,    alpha=.15)
pop.dot = list(size=.25,    alpha=.05)
scatter.by.group = function(data, dot=sam.dot, width=sd, 
                            highlight=NULL, 
                            highlight.color='yellow', highlight.alpha=.25,
                            line.alpha = NULL,
                            group.means.alpha=NULL, 
                            mean.shift=.05,
                            jitter.height=0)
                            { 
  p = ggplot(data)
  if(!is.null(highlight)) {
    p = p + annotate('rect', 
                      xmin=min(highlight)-.5,
                      xmax=min(highlight)+.5,
                      ymin=-Inf, ymax=Inf, 
                      fill=highlight.color, alpha=highlight.alpha)
  } 
  if(!is.null(group.means.alpha)) {
    summaries = data |> group_by(group) |> summarize(income=mean(income))
    p = p + geom_hline(aes(yintercept=income, color=group), alpha=group.means.alpha, data=summaries, linetype='dashed') 
  }
  if(!is.null(line.alpha)) { 
    p = p + stat_summary(aes(x=xco(group, education,shift=mean.shift), y=income, color=group), 
    geom='line', alpha=line.alpha, fun=mean, linetype='solid') 
  }

  p + geom_point(aes(x=xco(group, education),  y=income, color=group), 
               size=dot$size, position=position_jitter(seed=0,width=.15,height=jitter.height), alpha=dot$alpha) +
      stat_summary(aes(x=xco(group, education,shift=mean.shift), y=income, color=group), geom="pointrange", alpha=.7, fun.data=mean_sd)  +
    scale_x(c(11.5,20.5)) + scale_y() + scale_alpha(range=c(0,1)) + guides(color='none')
}

money = function(x) { sprintf('%1.0fk', round(x/1000, digits=0)) }  
```


```{r}
proportion.scale = 200000
scaled.interleaved.bars = list(geom_bar(aes(x=xco(group, education), y=proportion.scale*after_stat(prop), fill=group, color=group), 
                               position='identity', alpha=.1, linewidth=.1), 
                               scale_y(c(0,150000), sec.axis=sec_axis(~./proportion.scale, name='Proportion')),
                               guides(color='none', fill='none'))
scaled.purple.bars = geom_bar(aes(x=education, y=proportion.scale*after_stat(prop)), 
                       alpha=.1, linewidth=.025, fill='purple', color='purple', width=.2)
bars= list(geom_bar(aes(x=education, y=after_stat(prop), fill=group, color=group), 
                     position='identity', alpha=.05, linewidth=.4))
```

::: {#fig-greenred-scatter}
```{r}
#| fig-width: 10
ca$group = ca$sex == 'female'
ca = ca[ca$education >= 12,]
scatter.by.group(ca, group.means.alpha=.2, line.alpha=.5) + 
  scaled.interleaved.bars + 
  scaled.purple.bars 
```
Scatter plot of `education` vs `income` colored by `sex` with histograms of `education` overlaid. Note that, because incomes are on a very different scale than the proportions we're displaying in the histogram, two different scales for the y-axis are used. The scale on the left is for income. The scale on the right is for the proportions.  
:::

::: {#fig-greenred-histograms}
```{r}
#| fig-width: 5.5
#| layout-ncol: 2
facet.by.group = 
  list(facet_grid(row=vars(group)), theme(strip.background=element_blank(), strip.text=element_blank()))
ggplot(ca) + bars + facet.by.group + guides(color='none', fill='none')
ggplot(ca) + bars + guides(color='none', fill='none')
``` 
Histograms of `education` colored by `sex` in two layouts. On the left, they're plotted one above the other. On the right, they're overlaid.
:::


:::: columns 
::: {.column width="35%"}
```{r}
ca$W = as.numeric(ca$sex == 'female')
ca$X = ca$education
summaries = ca |> group_by(W,X) |> 
  summarize(mu.hat=money(mean(income)), N.wx=n())
```

```{r}
library(gt)
summaries |> ungroup() |> gt()
```
:::
::: {.column width="65%"}

We'll consider the following four estimators of the difference in income between the male and female CA residents in our sample.  
$$
\color{gray}
\begin{aligned}
\hat \Delta_{\text{raw}} &= \frac{1}{N_1}\sum_{i: W_i=1} Y_i - \frac{1}{N_0}\sum_{i: W_i=0} Y_i \\
\hat\Delta_0  &=\frac{1}{N_0}\sum_{i: W_i=0} \qty{\hat\mu(1,X_i) - \hat \mu(0, X_i)} \\ 
\hat \Delta_1 &= \frac{1}{N_1}\sum_{i: W_i=1} \qty{\hat\mu(1,X_i) - \hat \mu(0, X_i)} \\ 
\hat \Delta_{\text{all}} &= \frac{1}{n} \sum_{i=1}^n \qty{\hat\mu(1,X_i) - \hat \mu(0, X_i)} 
\end{aligned}
$$

::: {#exr-calc-ca-estimator .callout-exercise} 
Using the data in the table to the left, calculate each of these four estimators.
:::

:::
::::


In Lecture 9, we decomposed the raw difference $\hat\Delta_{\text{raw}}$ as a sum of the adjusted difference $\hat\Delta_1$ and a *covariate shift term*.  And we talked about how to use plots like the ones above to understand what the covariate shift term would look like. In this next exercise, we'll get in some practice interpreting plots like these. We'll work with three samples. Each sample is shown in a tab below. 

::: {.panel-tabset}

### Income in LA vs SFBay 
```{r}
#| fig-width: 10
groups = c('los angeles', 'san francisco', 'alameda')
this.sam = ca |> filter(county %in% groups)
this.sam$group = this.sam$county != groups[1]
scatter.by.group(this.sam, group.means.alpha=.2, line.alpha=.5) + 
  scaled.interleaved.bars + 
  scaled.purple.bars 
```
```{r}
#| fig-width: 5.5
#| layout-ncol: 2
ggplot(this.sam) + bars + facet.by.group + guides(color='none', fill='none')
ggplot(this.sam) + bars + guides(color='none', fill='none')
```

- $W_i$ is an indicator for county. 
  - [0 for residents of Los Angeles County]{.groupa}
  - [1 for residents of San Francisco and Alameda Counties]{.groupb}
- $X_i$ is education in years of schooling 
- $Y_i$ is income in dollars. 


### Unemployment Rates in Southern California   

```{r}
#| fig-width: 10
groups = c('san diego', 'orange')
this.sam = ca |> filter(county %in% groups)
this.sam$group = this.sam$county != groups[1]
this.sam$income = as.numeric(!this.sam$employed)

proportion.scale = 2
scatter.by.group(this.sam, group.means.alpha=.2, line.alpha=.5, jitter.height=.2) + 
  scale_y_continuous(breaks=seq(0,1), 
    sec.axis = sec_axis(~./proportion.scale, name='Proportion')) +    
  coord_cartesian(ylim=c(-.2,1.2)) 
```
```{r}
#| fig-width: 5.5
#| layout-ncol: 2
ggplot(this.sam) + bars + facet.by.group + guides(color='none', fill='none')
ggplot(this.sam) + bars + guides(color='none', fill='none')
```


- $W_i$ is an indicator for county.
  - [0 for residents of San Diego]{.groupa}
  - [1 for residents of Orange County]{.groupb} 
- $X_i$ is education in years of schooling 
- $Y_i$ is unemployment status: 0 for employed and 1 for unemployed. 


### Primary Turnout in MI 

```{r}
#| fig-width: 10
this.sam = read.csv('data/social-pressure-data.csv') |>
  mutate(income = as.numeric(voted=='Yes'),
         education = yob, 
         group = p2004=='Yes') |>
  filter(yob >= 1950) 

set.seed(1)
this.sam = this.sam[sample(1:nrow(this.sam), size=100000),]

pop.dot = list(size=.2,    alpha=.05)
proportion.scale = 1
p=scatter.by.group(this.sam, group.means.alpha=.2, line.alpha=.5, jitter.height=.2, dot=pop.dot) +
  scale_y_continuous(breaks=seq(0,1), 
                      sec.axis = sec_axis(~./proportion.scale, name='Proportion')) +    
  scale_x_continuous(breaks=seq(1900,2000, by=10)) +
  coord_cartesian(ylim=c(-.2,1.2), xlim=c(min(this.sam$yob)-1, max(this.sam$yob)+1)) 

p
```

```{r}
#| fig-width: 5.5
#| layout-ncol: 2
ggplot(this.sam) + bars + facet.by.group + guides(color='none', fill='none')
ggplot(this.sam) + bars + guides(color='none', fill='none')
```

- $W_i$ is an indicator for voting in the 2004 primary. 
  - [1 if they voted]{.groupb} in the last primary, 
  - [0 if they didn't]{.groupa}. 
- $X_i$ is birth-year.
- $Y_i$ is an indicator for voting in the 2006 primary.
  - 1 if they voted
  - 0 if they didn't

:::

::: {#exr-shift-interp-practice .callout-exercise} 
For each of the three samples, answer these Yes or No questions. No need to explain. 

1. Is the raw difference $\hat\Delta_{\text{raw}}$ larger or smaller than the adjusted difference $\hat\Delta_1$? 
2. Is the magnitude of the raw difference, $| \hat\Delta_{\text{raw}}|$, larger or smaller than the magnitude of the adjusted difference, $|\hat\Delta_1|$?
   
*Tip*. The second question is a bit hard. You should be able to use the plot to know the sign (+/-) of the adjusted difference and covariate shift term. If the signs are the same, you know the magnitude of the raw difference is larger. But if the signs are opposite, it's a bit more subtle. What happens depends on the relative magnitudes of the adjusted difference and covariate shift term. Usually, the magnitude of the covariate shift is less than twice the magnitude of the adjusted difference, which makes the raw difference smaller in magnitude than the adjusted one.  But if the magnitude of the covariate shift term is more than twice the magnitude of the adjusted difference, we have an extreme case of Simpson's paradox: the raw difference is larger in magnitude but opposite in sign to the adjusted difference. For the purpose of this assignment, don't worry about this happening. 
:::

We can derive similar decompositions for our other adjusted differences $\hat\Delta_0$ and $\hat\Delta_{\text{all}}$. That is, we can decompose $\hat\Delta_{\text{raw}}$ as the sum of $\hat\Delta_0$ or $\hat\Delta_{\text{all}}$ and a different, but conceptually similar covariate shift term. 

::: {#exr-decompose-ca-estimator .callout-exercise}
Derive a formula for $\hat\Delta_{\text{raw}}$ as the 
sum of $\hat\Delta_0$ and a covariate shift term.  Show your work.
:::

::: {#nte-decompose-ca-estimator .callout-note}
### A correction.
In a previously posted version of this assignment, @exr-decompose-ca-estimator asked you to do the same with $\hat\Delta_{\text{all}}$ in place of $\hat\Delta_0$. That's a bad exercise because there isn't a two-term decomposition of $\hat\Delta_{\text{raw}}$ as the sum of $\hat\Delta_{\text{all}}$ and a single covariate shift term. You can, however, get a useful *three-term* decomposition.


To do this, start with your formula for $\hat\Delta_{\text{raw}}$ in terms of $\hat\Delta_1$.
Then think of the difference $\hat\Delta_1 - \hat\Delta_{\text{all}}$ as *another* covariate shift term---only this time, the function we're averaging over our two histograms is $\hat\mu(1,x) - \hat\mu(0,x)$ instead of $\hat\mu(0,x)$.

$$
\begin{aligned}
\hat\Delta_{\text{raw}} 
&= \hat\Delta_1 + (\hat\Delta_{\text{raw}} - \hat\Delta_1) \\
&= \hat\Delta_{\text{all}} 
+ \underset{\text{new covariate shift term}}{\qty(\hat\Delta_{1} - \hat\Delta_{\text{all}})}
+ \underset{\text{$\hat\Delta_1$'s covariate shift term}}{(\hat\Delta_{\text{raw}} - \hat\Delta_1)}
\end{aligned}
$$

I've added an Extra Credit version of this exercise, @exr-decompose-ca-estimator-ec, on deriving and using this three-term decomposition.

:::


Now that we have these new decompositions, let's use them.

::: {#exr-using-decompositions .callout-exercise} 
Forget the calculations you did in @exr-calc-ca-estimator for a moment.
Explain how, using only @fig-greenred-scatter and @fig-greenred-histograms, 
you could predict whether $\hat\Delta_{\text{raw}}$ or $\hat\Delta_0$ would be larger. Do the same for $\hat\Delta_{\text{raw}}$ and $\hat\Delta_{\text{all}}$. 

Are your answers here are consistent with what you calculated in @exr-calc-ca-estimator? 
:::

Covariate shift isn't just the phenomenon behind the difference between raw and adjusted differences. It shows up wherever we're comparing averages over different groups. One example is comparing different adjusted differences, e.g. $\hat \Delta_1$ and $\hat \Delta_0$. The following exercises will explore that. These are extra credit exercises, but I encourage you to look them over and think about them at least for a minute or two. There's a good chance that, having made it this far, you'll find them pretty easy. 

::: {#exr-shift-between-adjusted .callout-exercise} 
*Extra Credit*. Explain how, using only @fig-greenred-scatter and @fig-greenred-histograms, you could predict whether $\hat\Delta_{1}$ or $\hat\Delta_0$ would be larger. 

*Hint*. Compare the 'histogram form' formulas for $\hat\Delta_1$ and $\hat\Delta_0$.
:::

::: {#exr-shift-between-adjusted .callout-exercise} 
*Extra Credit*. 

1. Write the simplest formula you can expressing 
$\hat\Delta_{\text{all}}$ in terms of $\hat\Delta_0$ and $\hat\Delta_1$ and other numbers you can compute from the sample. It should be very simple.
2. Referring to the formula, explain why, if you know which of $\hat\Delta_0$ or $\hat\Delta_1$ is larger, you also know whether $\hat\Delta_0$ or $\hat\Delta_{\text{all}}$ is larger and whether $\hat\Delta_1$ or $\hat\Delta_{\text{all}}$ is larger. 
3. Explain why in visual terms, too. 

*Hint*. How does the purple histogram in @fig-greenred-scatter relate to the green and red ones? This is not just relevant to the in-visual-terms part; if you don't already have a formula, what you see may guide you toward one. 
:::

::: {#exr-decompose-ca-estimator-ec .callout-exercise} 
*Extra Credit*. Write out the three-term decomposition 
for $\hat\Delta_{\text{raw}}$ described above in @nte-decompose-ca-estimator.
And, with reference to the California income data shown in @fig-greenred-scatter and @fig-greenred-histograms, explain how to use this to see whether $\hat\Delta_{\text{raw}}$ or $\hat\Delta_{\text{all}}$ is larger.
:::

