---
title: "Week 3 Homework"
format: html
---
```{r}
#| echo: false
library(ggplot2)
library(purrr)
library(tidyverse)
library(scales)
theme_set(theme_minimal())
```

```{r}
#| echo: false
if (knitr::is_latex_output()) {
    knitr::opts_chunk$set(dev = "cairo_pdf")
}
theme_set(theme_minimal())
```

::: hidden
$$
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\Var}{V}
\DeclareMathOperator{\hVar}{\widehat{V}}
\DeclareMathOperator{\bias}{bias}
\newcommand{\thetaprior}{\theta^{\text{prior}}}
\newcommand{\nprior}{n^{\text{prior}}}
\newcommand{\yprior}{y^{\text{prior}}}
$$
:::

# Variance when Sampling Without Replacement

In [Lecture 3](https://qtm285-1.github.io/assets/lectures/Lecture3#/marginalization), 
we talked about the distribution of a sample proportion when we sample *without replacement*. This is, admittedly, a long read. But one thing we can take away from it---and I'll just tell here  rather than asking you to work it out --- is the probability distribution of $Y_1$ when we make a single call and the joint distribution of $Y_1, Y_2$ when we make two calls. Those are just special cases of the 'Partially Marginalized' distribution described in Lecture 3. Letting $m_0$ and $m_1$ be the number of zeros and ones in our binary population $y_1 \ldots y_m$, these are the distributions.

:::: columns
::: {.column width=50%}
$p$ | $Y_1$ 
--|--
$\frac{m_0}{m}$ | 0
$\frac{m_1}{m}$   | 1 

:::
::: {.column width=50%}

$p$ | $Y_1$ | $Y_2$
--|--|--
$\frac{m_0(m_0-1)}{m(m-1)}$  | 0 | 0
$\frac{m_0m_1}{m(m-1)}$      | 0 | 1
$\frac{m_0m_1}{m(m-1)}$      | 1 | 0
$\frac{m_1(m_1-1)}{m(m-1)}$  | 1 | 1

:::
::::

You can go ahead and cross out the $Y_1$ in the first table and write in $Y_i$,
because this is the marginal distribution of every individual observation $Y_i$ when we draw a sample
without replacement of *any size* $n$. You can cross out $Y_1$ and $Y_2$ in the second table and write in $Y_i$ and $Y_j$ for the same reason: it's the joint distribution of any pair $(Y_i, Y_j)$ for $i\neq j$ when we draw a sample without replacement of any size $n$. This makes them a lot more useful.
 
To justify this, we can do a simple two-step thought experiment. We'll think of drawing a sample without replacement as the process of shuffling a deck of cards labeled 1 ... m,  then drawing the first $n$ cards off the top. Letting $J_1$ be what the first card says, $J_2$ what the second one says, and so out, our sample $Y_1 \ldots Y_n$ is $y_{J_1}, y_{J_2}, \ldots y_{J_n}$. Here's our argument.

1. No matter how many cards we're going to draw, the distribution of the first card is the same. Same with the first two cards. Thus, what we've written above is the marginal distribution of $Y_1$ and the joint distribution of  $(Y_1,Y_2)$ when we make any number of calls $n$.

2. If we'd pulled off the first $n$ cards and then counted from the bottom instead of the top, we'd get the same distribution. They're shuffled, after all. So the marginal distributions of $Y_1$ and $Y_n$ are the same and so are the joint distributions of $(Y_1,Y_2)$ and $(Y_n,Y_{n-1})$. The same thing would be true if we went through our top $n$ cards in any order. Thus, the marginal distributions of $Y_i$ are the same for all $i$ and the joint distributions of $(Y_i,Y_j)$ are the same for all $i \neq j $. 

Now let's use this to calculate the variance of our sample proportion $\hat\theta=\frac{1}{n}\sum_{i=1}^{n}Y_i$.
Most of the work is already done on [this slide](https://qtm285-1.github.io/assets/lectures/Lecture5.html#/the-standard-deviation-of-a-proportion-1) from Lecture 5.^[Look at the 'Sampling without Replacement' tab.]
In that calculation, I used the following identity. It's colored blue there to highlight it.

$$
\E\qty[ (Y_i - \E[Y_i])(Y_j - \E[Y_j]) ] = -\frac{\theta(1-\theta)}{m-1} \qfor i\neq j
$$

But I didn't actually show that it's true. That's what we're going to do now. Let's start by rewriting this expression 
so it's a little easier to work with. Like this.

$$ 
\E\qty[ (Y_i - \E[Y_i])(Y_j - \E[Y_j]) ] = \E[Y_i Y_j] - \E[Y_i]\E[Y_j]
$$

::: {.callout-tip collapse="true"}
### Proof of Equivalence
To show that these two expressions are equivalent, 
we 'multiply out'  $(Y_i - \E[Y_i])(Y_j - \E[Y_j])$ to
get four terms, distribute the expectation and pull out constants, 
and cancel some equal-and-opposite terms in the result.

$$
\begin{aligned}
\E\qty[ (Y_i - \E[Y_i])(Y_j - \E[Y_j]) ] 
&= \E\qty[ Y_i Y_j - Y_i\E[Y_j] - Y_j\E[Y_i] + \E[Y_i]\E[Y_j] ] \\
&= \E\qty[ Y_i Y_j ] - \E\qty[ Y_i\E[Y_j] ] - \E\qty[ Y_j\E[Y_i] ] + \E\qty[ \E[Y_i]\E[Y_j] ] \\
&= \E\qty[ Y_i Y_j ] - \E[Y_i]\E[Y_j] - \E[Y_j]\E[Y_i] + \E[Y_i]\E[Y_j] \\
&= \E\qty[ Y_i Y_j ] - \E[Y_i]\E[Y_j]
\end{aligned}
$$
:::

What's nice about this second form is that using our tables above to calculate these expectations is easy.
We actually have a row for $Y_i$ in our marginal table and can easily add a row to our joint table for $Y_i Y_j$.
I'd like to ask you to calculate the thing, but since it'd probably take a litte arithmetic to manipulate what you
get into the form $-\frac{\theta(1-\theta)}{m-1}$ we used in lecture, I'm going to ask you to verify my calculation
instead.

::: {#exr-covariance-with-replacement}

Using this equivalent form, and the tables above, show that the following identity is true.
$$
\E\qty[ (Y_i - \E[Y_i])(Y_j - \E[Y_j]) ] = \frac{m_1(m_1-1)}{m(m-1)}-\frac{m_1^2}{m^2}
$$

:::

Simplifying this into the form $-\frac{\theta(1-\theta)}{m-1}$ is a little bit of work, but not too bad.
As usual, the trick is to give your two terms a common denominator and see what cancels. I'll 'fold' this
so if you'd like to try it on your own, you can do so. If you just want to read it, expand the box by clicking on it

::: {.callout-tip collapse="true"}
## Simplifying our Cross-Term
$$
\begin{aligned}
 \frac{m_1(m_1-1)}{m(m-1)}-\frac{m_1^2}{m^2} 
 &= \frac{m_1(m_1-1)m - m_1^2(m-1)}{m^2(m-1)} && \text{ getting a common denominator} \\
 &= \frac{ (m_1^2 - m_1m) - (m_1^2 m - m_1^2)}{m^2(m-1)} && \text{ expanding products} \\
 &= \frac{ m_1^2 - m_1m}{m^2(m-1)} && \text{ canceling equal and opposite terms} \\
 &= \frac{ m_1(m_1 - m)}{m^2(m-1)} && \text{ pulling out common factors} \\
 &= \frac{m_1}{m} \cdot \frac{m_1-m}{m} \times \frac{1}{m-1} && \text{ grouping factors in the numerator and denominator} \\
 &= \theta (\theta - 1) \times \frac{1}{m-1} && \text{ for } \  \theta = \frac{m_1}{m} \\
 &= -\frac{\theta(1-\theta)}{m-1} && \text{ as desired }
 \end{aligned}
$$
:::

# Bias and Consistency 

## Using Prior Information 
So far, we've exclusively talked about *unbiased estimators*. That is, estimators with the property that 
their expected value is equal to the estimation target. 
$$ 
\hat\theta \qqtext{ is called unbiased if } \E[\hat{\theta}] = \theta. 
$$ 
We say an estimator is *biased* if this isn't true. To get a sense of what that means, let's consider a simple
example of a biased estimator. 
$$
\tilde{Y}_1 = \frac{1}{n+1}\cdot \{\frac{1}{2} + \sum_{i=1}^{n}Y_i\}
$$

::: {#exr-bias-1 .callout-exercise}
Calculate the expected value of the estimator $\hat\theta =\tilde{Y}_1$. 
Then calculate its bias $\E[\hat\theta] - \theta$ and its 
standard deviation $\sqrt{\Var[\hat\theta]}$. Report all three.
:::

This is a specific example of a more general estimator that integrates information from a prior study. Maybe a real study or maybe a study we're imagining. Suppose we have $\nprior$ observations $\yprior_1 \ldots \yprior_{\nprior}$ from this study^[Sometimes we call these *pseudo-observations*. This particular interpretation, in which we think of them as coming from a prior study, can be used to derive this estimator from Bayesian principles.]
$$
\thetaprior = \frac{1}{\nprior}\sum_{i=1}^{\nprior}\yprior_i. 
$$ 

Then averaging all our observations --- from our current study and this prior one --- gives us the following estimator.

$$
\begin{aligned}
\tilde{Y}_{\nprior}
&= \frac{1}{\nprior + n} \qty{\sum_{i=1}^{\nprior}\yprior_i + \sum_{i=1}^{n} Y_i} \\
&= \frac{1}{\nprior + n}  \qty{ \nprior\thetaprior + n\bar Y }
\end{aligned}
$$

We treat these prior observations, and therefore their mean $\thetaprior$, as deterministic. 
The simple example we started with, $\tilde{Y}_1$, is a special case where we have a single prior observation $\yprior_1 = \frac{1}{2}$.

::: {#exr-bias-general .callout-exercise}
Repeat @exr-bias-1 for $\tilde{Y}_{\nprior}$. 
:::


## Visualizating the Impact of Prior Information

Lets's try to get a sense of how using prior information like this
impacts our inference. We'll do it visually. The plot below shows the sampling distributions of three estimators of a population mean $\theta$ at three sample sizes $n$: 10, 40, and 160. These are the estimators.

1. The estimator $\hat\theta_1 = \tilde{Y}_{\nprior}$ for $\thetaprior=3/4$ and $\nprior=10$ prior observations.
2. The estimator $\hat\theta_2 = \tilde{Y}_{n}$ for $\thetaprior=3/4$ and $\nprior=n$ prior observations---the bigger the sample, the more prior observations we incorporate.
3. The sample mean, $\hat\theta_3 = \bar Y$.

As usual, the estimation target $\theta$ is indicated by a green line, the sampling distribution's mean by a solid blue line, and its mean plus and minus two standard deviations by dotted blue lines. In our first exercise, we'll match estimators to pictures.

::: {#exr-bias-visualize .callout-exercise}
What is the correspondence between $\theta_1$, $\theta_2$, and $\theta_3$ described above 
and 'estimator a', 'estimator b', and 'estimator c' from the plot?
:::

::: {#fig-3-sampling-distributions}
```{r}
#| echo: false
#| out-height: 4in
set.seed(1234)
n.prior = 10
theta.prior = .75
theta = .5

error = 2 * rnorm(100000)
bias = list(a = function(n) 0, 
            b = function(n) n.prior * (theta.prior - theta) / (n.prior + n), 
            c = function(n) n * (theta.prior-theta) / (n + n))                # n.prior = n
variance = list(a = function(n) theta*(1-theta)/n, 
                b = function(n) theta*(1-theta) * n / (n.prior + n)^2, 
                c = function(n) theta*(1-theta) * n / (n + n)^2)


plotdata = expand_grid(estimator = names(bias), n = c(10, 40, 160)) %>%
    pmap(function(estimator, n) {
        thebias = bias[[estimator]](n)
        thevar = variance[[estimator]](n)
        data.frame(estimator = estimator, n = n, muhat = thebias + error * sqrt(thevar))
    }) |> bind_rows()

summaries = plotdata |>
    group_by(n, estimator) |>
    summarize(mean = mean(muhat), sd = sd(muhat))

breaks = seq(-2/3, 2/3, by = 1/3) 
signs = case_when(
    breaks < 0 ~ "-",
    breaks == 0 ~ "",
    breaks > 0 ~ "+"
)
offsets = case_when(
    breaks != 0 ~ sprintf("%d/3", 3*abs(breaks)),
    breaks == 0 ~ ""
)
labels.abstract = sprintf("θ %s %s", signs, offsets)

ggplot(plotdata) +
    geom_vline(aes(xintercept = theta), data = data.frame(theta=0), color='green',alpha=.4, linewidth=1.5) +
    geom_histogram(aes(x = muhat, y = after_stat(density)), bins = 100, alpha = .2) +
    geom_vline(aes(xintercept = mean), data = summaries, alpha=.4, color='blue') +
    geom_vline(aes(xintercept = mean + 2 * sd), data = summaries, linetype = "dotted", color='blue',alpha=.4) +
    geom_vline(aes(xintercept = mean - 2 * sd), data = summaries, linetype = "dotted", color='blue',alpha=.4) +
    scale_x_continuous(breaks = breaks, labels = labels.abstract, limits = c(-3/4,3/4)) +
    theme(axis.text.x = element_text(angle = -90, vjust = 0.5, hjust = 1)) +
    xlab("") +
    ylab("") +
    facet_grid(n ~ estimator, labeller = labeller(n = label_both, estimator = label_both))
```

:::




## The Bias/Variance Tradeoff







::: {#exr-bias-variance-rmse .callout-exercise}
Suppose we increase our number of prior observations, $\nprior$, 
without changing their mean $\thetaprior$. What happens to 
the bias of the estimator $\tilde{Y}_{\nprior}$? 
E.g. does it stay the same, increase, decrease, increase then decrease, decrease then increase, etc.?
What happens to its standard deviation? And what happens
to its *root-mean-squared-error*, $\RMSE(\tilde{Y}_{\nprior})$
where $\RMSE(\hat\theta) = \sqrt{\E\qty[(\hat\theta - \theta)^2]}$?
:::

Having a large bias or a large standard deviation are two ways of 




To illustrate your point, make a *polar plot* 

    
    \tilde{Y}_{\nprior}) = \sqrt{\Var(\tilde{Y}_{\nprior}) + \bias(\tilde{Y}_{\nprior})^2}$?




your
estimator's bias? What happens to its standard deviation.
are




What happens to the bias of $\tilde{

:::
Now suppose your 'prior observations' are made-up. 
You really just think the population proportion ought to be something like $\thetaprior$,



::: {#exr-bias-variance-tradeoff .callout-exercise}



What do you see about the relationship between your estimator's bias and variance?


:::

You 


## Visualizing 




## Implications for Consistency


```{r}
#| echo: false
#| out-height: 4in
set.seed(1234)
error <- 2 * rnorm(100000)
bias <- list(a = function(n) 0, b = function(n) sqrt(10) / sqrt(n), c = function(n) 1)

plotdata <- expand_grid(estimator = names(bias), n = c(10, 40, 160)) %>%
    pmap(function(estimator, n) {
        thebias <- bias[[estimator]](n)
        data.frame(estimator = estimator, n = n, muhat = thebias + error / sqrt(n))
    }) |>
    bind_rows()
summaries <- plotdata |>
    group_by(n, estimator) |>
    summarize(mean = mean(muhat))

breaks <- seq(-4, 4, by = 2) / 2
signs <- case_when(
    breaks < 0 ~ "-",
    breaks == 0 ~ "",
    breaks > 0 ~ "+"
)
offsets <- case_when(
    breaks != 0 ~ sprintf("%d", abs(breaks)),
    breaks == 0 ~ ""
)
labels.abstract <- sprintf("μ %s %s", signs, offsets)

ggplot(plotdata) +
    geom_histogram(aes(x = muhat, y = after_stat(density)), bins = 100, alpha = .4) +
    geom_vline(aes(xintercept = mean), data = summaries) +
    scale_x_continuous(breaks = breaks, labels = labels.abstract, limits = range(breaks)) +
    theme(axis.text.x = element_text(angle = -90, vjust = 0.5, hjust = 1)) +
    xlab("") +
    ylab("") +
    facet_grid(n ~ estimator, labeller = labeller(n = label_both, estimator = label_both))
```

::: { #exr-consistency_plot  .callout-exercise}
Consider another estimator of our sample mean $\bar{Y_i}$, when we observe a sample with replacement $Y_1 \ldots Y_n$ from a population with mean $\mu$ and standard deviation $\sigma$:
$$
\bar{Y_i} = \frac{1}{n}\sum_{i=1}^n Y_i
$$ 

Which of the visualizations corresponds to the more generalized estimator, $\underset{\text{general}}{\tilde{Y_i}}$, where we incorporate more information, and which of them correponds to the sample mean estimator, $\bar{Y_i}$? Are they consistent?
:::

# Convergence in Probability

Now let's think about another notion of convergence. If an estimator's sampling distribution
ends up in the right location ($\text{bias} \to 0$) with arbitrarily little spread ($\text{variance} \to 0$),
then it makes sense that \emph{any draw from it} will be close to the estimation target $\mu$. 

Recall that you can visualize your estimator as a single dot drawn
from its sampling distribution. Like the black one below. And if you want to think of what happens if you were to repeat your survey 100 times, you can visualize the results by plotting 100 dots. Like the purple ones.
```{r}
K <- 100
plotdata$height <- runif(nrow(plotdata), 0, 1.2)
dotplotdata <- plotdata |> filter(n == 40 & estimator == "b")
dotplotsummaries <- summaries |> filter(n == 40 & estimator == "b")

ggplot(dotplotdata) +
    geom_histogram(aes(x = muhat, y = after_stat(density)), bins = 100, alpha = .4) +
    geom_vline(aes(xintercept = mean), data = dotplotsummaries) +
    geom_point(aes(x = muhat, y = height),
        alpha = .2,
        color = "purple", data = dotplotdata[2:K, ]
    ) +
    geom_point(aes(x = muhat, y = height),
        alpha = 1,
        color = "black", data = dotplotdata[1, ]
    ) +
    scale_x_continuous(breaks = breaks, labels = labels.abstract, limits = range(breaks)) +
    theme(axis.text.x = element_text(angle = -90, vjust = 0.5, hjust = 1)) +
    xlab("") +
    ylab("")

```


We formalize this idea by talking about \emph{convergence in probability}. We say an estimator $\hat\mu$ converges in probability to $\mu$ if, for anyone's idea of `sufficiently close', the probability that $\hat\mu$ is sufficiently close to $\mu$ goes to one. Or equivalently, the probability that it isn't sufficiently close goes to zero.
$$
P(\lvert\hat\mu - \mu\rvert \ge \epsilon) \to 0 \qqtext{ as } n \to \infty
$$

Convergence in mean square *implies* convergence in probability. Let's see why. We'll use *Markov's inequality*.

## Markov's Inequality

Markov's inequality says that the probability the absolute value of a random variable $X$ 
exceeds some threshold $t$ is bounded by the ratio of its expectated absolute value and $t$, i.e.,
$$ P(X \geq t) \leq \frac{\E[X]}{t}$$
Often, instead of using this to bound the random variable we're interested in directly,
e.g. $X=\lvert\hat\mu - \mu\rvert$, we use it to bound its square. $X \ge t$ if and only if $X^2 \ge t^2$, so the probability that $X \ge t$ is the same as the probability that $X^2 \ge t^2$. Applying Markov's inequality to the random variable $X^2$ gives us a bound in terms of $X$'s mean square.
$$
P(X \ge t) = P(X^2 \ge t^2) \leq \frac{\E[X^2]}{t^2} \qqtext{ e.g.} 
P(\lvert\hat\mu - \mu\rvert \ge \epsilon) = P((\hat\mu - \mu)^2 \ge \epsilon^2) \leq \frac{\E[(\hat\mu - \mu)^2]}{\epsilon^2}
$$

This tells us, in terms of the mean squared error of $\hat\mu$, how likely it is that $\hat\mu$ is far from $\mu$. And, in particular, it tells us that if the mean squared error goes to zero, then the probability that $\hat\mu$ is any distance $\epsilon$ away from $\mu$ goes to zero. 

Now we're going to do two things. First, we're going to prove Markov's inequality. It's good practice working with expectations and probabilities. Second, we're going to look at what Markov's inequality tells us about
how close $\hat\mu$ is to $\mu$, i.e., we're going to use it to calculate the coverage of an interval estimate.

First we define the following two functions, $f$ and $g$, with $t, x\ge 0$.
$$
f(x,t) = \frac{x}{t}
$$
$$
g(x,t) = 1_{\ge t}(x) = \begin{cases}
1 & \text{ if } x \ge t \\
0 & \text{ otherwise }
\end{cases}
$$

::: { #exr-markov-plot  .callout-exercise}
We pick $t=3$, plot the two functions above on the same coordinate with $x$ on the x axis. Based on the graph, what can we infer about the relationship between $f(x,t)$ and $g(x,t)$?
:::

::: { #exr-markov-plot  .callout-exercise}
It turns out that your inference is correct! For all $t, x\ge 0$, $\frac{x}{t} \ge 1_{\ge t}(x)$. Using the fact to prove *Markov's inequality*.

*Hint 1*. $\E[Y]\ge \E[Z]$ if $Y\ge Z$

*Hint 2*. Knowing that $\frac{x}{t} \ge 1_{\ge t}(x)$, we can infer that $\frac{X}{t} \ge 1_{\ge t}(X)$ for $X$ being a non-negative random variable.
:::

## Markov's Inequality and Interval Estimation

So far, when we've calculated the coverage of an interval estimate in terms of its mean and variance, we've relied on normal approximation. That is, we've been using a different formula for 
$P(\lvert\hat\mu - \mu\rvert \ge \epsilon)$ based on the idea that $\hat\mu-\mu$ has approximately the same distribution as a normally-distributed random variable $Z$ with the same mean---zero---and variance $\sigma_{\hat\mu}^2=\Var[\hat\mu-\mu]$. We can use this approximation, in combination with our knowlege of the probability density of the normal approximation, to get a bound on the probability that $\lvert\hat\mu - \mu\rvert \ge \epsilon$.

$$
P(\abs{\hat\mu - \mu} \ge \epsilon) \approx P(\abs{Z} \ge \epsilon) = 1-P(\abs{Z} < \epsilon) = 1- \int_{-\epsilon}^{\epsilon} f_{\sigma_{\hat\mu}}(z) dz \qfor f_{\sigma}(z) = \frac{1}{\sqrt{2\pi}\sigma} e^{-z^2/2\sigma_{\hat\mu}^2}
$$

The reason we're always talking about intervals of the form $\hat\mu \pm 2 \sigma_{\hat\mu}$ is that, using this approximation, that yields 95% coverage. That is, $P(\abs{Z} < 2 \sigma_{\hat\mu}) \approx .95$. To see why we tend to use normal approximation instead of Markov's inequality, let's see what Markov's inequality tell us about how close an estimator is to it's estimation and about the coverage of this interval.

::: {#exr-markov-interval  .callout-exercise}
To see what Markov's inequality can tell us, first we can apply the property to the squared deviation from the mean:
$$
X = \{\hat{\theta}-\E[\hat{\theta}]\}^2
$$
$$
P(X\ge t)\le \frac{\E[\{\hat{\theta}-\E[\hat{\theta}]\}^2]}{t} = \frac{\Var(\hat{\theta})}{t}
$$

When we do this, what does Markov's inequality tell us about how close our estimator to its mean. What is the probability that it's within 1, 2, 3 standard deviations?
:::


:::{#exr-markov-coverage  .callout-exercise}
We can use this to calibrate an interval. What does it tell us about the probability of the interval, $\hat{\theta} \pm 2 \sigma_{\hat{\theta}}$ contains the $\E[\hat{\theta}]$. How wide do we need to make the interval width to make the interval width $w$ so that $\hat{\theta} \pm \frac{w}{2}$ has 95% coverage?

*Hint*. $\hat{\theta} \pm \frac{w}{2}$ contains $\theta$ if and only if $|\hat{\theta}-\theta| \le \frac{w}{2}$. Therefore, $P(\hat{\theta} \pm \frac{w}{2} \text{ contains } \theta) = P(|\hat{\theta}-\theta| \le \frac{w}{2})$.
:::

## Extra Credit (Markov's Inequality Continued)
:::{#exr-markov-compare  .callout-exercise}
You may have noticed that the interval we get from Markov's inequality is different from the one we get from using normal approximation.
:::