---
title: "QTM 220 Lecture 11"
subtitle: "Inferential Statistics with Non-Binary Covariates: Part 2"
format: 
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---

```{r load-libraries}
library(purrr)
library(tibble)
library(dplyr)
library(magrittr)
library(stringr)
library(ggplot2)
library(htmltools)
library(kableExtra)
library(scales)
library(cpsR)
library(mapdata)

options(knitr.kable.NA = '')

compact_number = label_number(accuracy=.1, scale_cut=cut_short_scale())
ellipsis <- function(df, nhead=5L, ntail=2) {
    rownames(df) = as.character(1:nrow(df))
    stopifnot("data.frame" %in% class(df))
    els <- rep(NA, ncol(df)) %>% 
        matrix(nrow=1, dimnames=list(NULL, names(df))) %>% 
        data.frame(stringsAsFactors=FALSE) %>% 
        tibble::as_tibble()
    rownames(els) = "⋮"
    dplyr::bind_rows(head(df, nhead), els, tail(df, ntail))
}

percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }
```

```{r ggplot-themes}
fontsize = 16
textcolor = "#888888"

class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
        panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x  = element_text(colour = textcolor, size=fontsize),
        axis.text.y  = element_text(colour = textcolor, size=fontsize),
        axis.title.x  = element_text(colour = textcolor),
        axis.title.y  = element_text(colour = textcolor, angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
              axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
           panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(simplified.theme)
```




```{r get-data}
#| cache: true
#| 
# for cps variable interpretation see https://www2.census.gov/programs-surveys/cps/datasets/2022/march/asec2022_ddl_pub_full.pdf
census.api.key = '8084f81e4f71e4d32638dac15f188b424f72d063'
cps.sampling.ratio = 1/25 # Not true. 1/2500 households are surveyed. But for the sake of plotting, it's ok.

california.state.code = 6
year = 2020
cps.data = results=get_asec(year=year, vars=c('pearnval', 'a_age', 'a_hga', 'gestfips', 'gtco', 'gtcbsasz'), key=census.api.key)
```

```{r transform-data}
#| cache: false
#| 
all.earnings = data.frame(state.code = cps.data$gestfips,
                          county.code = cps.data$gtco,
                          age=cps.data$a_age,
                          income=cps.data$pearnval,
                          education.code=cps.data$a_hga,
                          education=case_match(cps.data$a_hga,
                                           0  ~ 0,   # child
                                           31 ~ 0,   # < grade 1
                                           32 ~ 4,   #  grade 1-4
                                           33 ~ 6,   #  grade 5-6
                                           34 ~ 8,   #  grade 7-8 
                                           35 ~ 9,   #  grade 9
                                           36 ~ 10,  #  grade 10
                                           37 ~ 11,  #  grade 11
                                           38 ~ 11,  #  grade 12 no diploma
                                           39 ~ 12,  #  high school grad
                                           40 ~ 13,  #  some college
                                           41 ~ 14,  #  associate degree (vocational)
                                           42 ~ 14,  #  associate degree (academic)
                                           43 ~ 16,  #  bachelors degree
                                           44 ~ 18,  #  masters degree
                                           45 ~ 20,  #  professional school degree
                                           46 ~ 20)) #  doctorate

all.earnings$degree = factor(all.earnings$education >= 16, 
                     levels=c(FALSE, TRUE), 
                     labels=c('no 4-year degree','4-year degree')) 

coarsen = function(x, group, seed=0) {
  current.seed = .Random.seed
  set.seed(seed)
  x.coarsened = rep(NA, length(x))
  for(gg in unique(group)) {
    x.coarsened[group==gg] = runif(sum(group==gg), min(x[group==gg]), max(x[group==gg]))
  }
  set.seed(current.seed)
  x.coarsened
} 

in.interval = function(X, interval) { interval[1] <= X & X <= interval[2] }
earnings = filter(all.earnings, education >= 8 & 
                                state.code == california.state.code & 
                                age %>% in.interval(c(25,35)))
Y = earnings$income
X = earnings$education
```


```{r get-county-info}
# county info from https://en.wikipedia.org/wiki/List_of_counties_in_California
# converted to csv using https://wikitable2csv.ggor.de

to.number = function(s) { as.numeric(str_trim(gsub(",", "", s))) }
ca_county_info = read.csv('california_county_info.csv') %>%
  mutate(subregion = tolower(sub(' County', '', County))) %>%
  mutate(pop = to.number(Population..2022.)) %>%
  mutate(area = to.number(sub('sq.*', '', Area))) 

earnings.county = left_join(earnings, ca_county_info[,c('FIPS.code', 'subregion')], by=c('county.code' = 'FIPS.code'))
earnings$county = earnings.county$subregion
earnings$county[is.na(earnings$county)]='unknown'
```


```{r make-maps}
#| cache: true
#| 

# To give our survey recipients a bit more reality, we'll give them fake locations on a map
# And we'll make up a population of California residents that they're sampled from
# maps are based on code from https://eriqande.github.io/rep-res-eeb-2017/map-making-in-R.html

counties = map_data("county")
ca_county = counties %>%
  filter(region == "california")
ca_county_centers = ca_county %>%
  group_by(subregion) %>%
  summarize(county.lat = mean(lat), county.long=mean(long))

ca_county_more = left_join(ca_county_centers, ca_county_info, by="subregion")
ca_all = left_join(ca_county, ca_county_more, by="subregion")


ca.plot = ggplot(data=ca_all, mapping=aes(x=long, y=lat, group=group)) + 
  coord_quickmap() +
  geom_polygon(fill='gray', color = "white", alpha=.4) +
      scale_fill_gradient(trans="log10") + 
      theme_void()

ca.pop.plot = ggplot(data=ca_all, mapping=aes(x=long, y=lat, group=group)) + 
  coord_quickmap() +
  geom_polygon(aes(fill = pop/area), color = "white") +
      scale_fill_gradient(trans="log10") + 
      theme_void()

## we'll make up a population as follows
# 1. choose the number of people to be the number of people in our sample scaled up to account for the cps sampling ratio 1/2500
#    Because this would wind up being computationally and visually problematic, we'll scale up by 25 instead of 2500;
#    each population dot in the population stands for 100 people in the population.
# 2. for each, sample a county with probability proportional to its finite.population
# 3. for each, sample a latitude and longitude by taking a randomly weighted combination of the    
#    county's vertices. As long as the county is convex, these points will be inside the county.
# 4. we'll give them a degree too, chosen independently of all else from the empirical distribution on the sample
m = nrow(earnings) * 25 

# takes a list of subregions, possibly with duplicates
# for each element, chooses random lat/long coordinates within the sample.subregion
# by taking a random convex combination of the subregion's vertices
sample.within.subregion = function(subregions) {
  output = data.frame(subregion=subregions, lat=NA, long=NA)
  for(subregion in unique(subregions)) { 
      vertices = ca_county[ca_county$subregion==subregion, c('lat','long')]
      U = sum(subregions==subregion)
      V = nrow(vertices)
      weights = abs(rt(U*V, 1))
      dim(weights) = c(U,V)
      weights = (1/rowSums(weights)) * weights
      output[subregions==subregion, c('lat', 'long')] = 
                 cbind(weights %*% vertices$lat, weights %*% vertices$long)
    }
    output
}

finite.population.subregion = sample(ca_county_more$subregion, size=m, replace=TRUE, prob=ca_county_more$pop)
finite.population = sample.within.subregion(finite.population.subregion)
finite.population$degree = sample(earnings$degree, nrow(finite.population), replace=TRUE)
finite.population$group=1 # for plotting

display.population = finite.population
```


```{r make-sample}
# and we'll force the degree-distribution in the sample to match our survey responses
our.sample = sample_n(finite.population, nrow(earnings))
our.sample$degree=earnings$degree

pop.plot = ca.plot + geom_point(aes(x=long, y=lat), data=display.population, size=.02, alpha=.1) 
sample.plot = ca.plot + geom_point(aes(x=long, y=lat), data=our.sample, size=.2, alpha=.5)
pop.plot.inf = ca.pop.plot + geom_point(aes(x=long, y=lat), data=display.population, size=.01, alpha=.1)  
sample.plot.inf = ca.pop.plot + geom_point(aes(x=long, y=lat), data=our.sample, size=.2, alpha=.5)
```

```{r visualization-defaults}
scale_y = function(limits=c(-.3e5, 3e5)) { 
  scale_y_continuous(labels=label_dollar(), limits=limits)
}

years = 8:20
scale_x = function(limits=c(min(years)-.5, max(years)+.5)) { 
  scale_x_continuous(breaks=years, labels=sprintf('%dy', years), limits=limits)
}

dot.size = 2
```


```{r define-colors}
twocolor.red   = 'rgb(248,118,109)'
twocolor.green = 'rgb(0,191,196)'
```


```{r scatter-plots}

earnings.jitter = ggplot(earnings) + geom_point(aes(x=education, y=income), color='#F7766D',
                              size=dot.size, position=position_jitter(seed=0,w=.3), alpha=.1) +
                   scale_x() + scale_y() + guides(color=FALSE)

summaries = earnings %>% 
          group_by(education) %>% 
          summarize(mean=mean(income),
                    sd=sd(income))
                    
```

```{r bootstrapping}
#| cache: true

average.over.people = function(summaries) {
  muhat = function(x) { summaries$mean[summaries$education==x] }
  P = function(x) { summaries$N[summaries$education==x] / sum(summaries$N) }
  alphahat = function(x) { if (x==12) { P(12) } else if (x==8) { -P(9) } else { P(x) - P(x+1) } }  
  sapply(8:12, function(x) { alphahat(x) * muhat(x) }) %>% sum
}
run.boot = function(earnings, B=1000) { purrr::map(1:B, function(i) {
  earnings.boot = earnings[sample(1:nrow(earnings), replace=TRUE),]
  summaries = earnings.boot[earnings.boot$education <= 12, ] %>% group_by(education) %>% summarize(mean=mean(income), N=length(income))
  model = lm(income ~ education, data=earnings.boot[earnings.boot$education <= 12,])
  data.frame(people=average.over.people(summaries), 
             years=(summaries$mean[summaries$education==12] - summaries$mean[summaries$education==8])/4,
             line = coef(model)[2])
}) %>% bind_rows()
}
boots = run.boot(earnings)
summaries = earnings[earnings$education <= 12, ] %>% group_by(education) %>% summarize(mean=mean(income), N=length(income))
years.estimate = (summaries$mean[summaries$education==12] - summaries$mean[summaries$education==8])/4
```

```{r plotting-code}
width = function(samples, center=mean(samples), alpha=.05) {
  covg.dif = function(w) { 
    actual = mean( center - w/2 <= samples & samples <= center + w/2)
    nominal = 1-alpha 
    actual - nominal  
  }
  uniroot(covg.dif, interval=c(0, 2*max(abs(samples-center))))$root
}

distribution.plot = function(boot.samples, point.estimate, w=width(boot.samples), interval.replications=0, point.replications=0) { 
  drange = c(0, dnorm(0, mean=0, sd=sd(boot.samples)))
  p = ggplot(data.frame(samples=boot.samples)) + 
    geom_histogram(aes(x=samples, y=after_stat(density)), alpha=.3, bins=50)  +
    geom_pointrange(aes(y=dnorm(2*sd(samples), mean=0, sd=sd(samples)), 
      x=point.estimate, xmin=point.estimate - w/2, xmax=point.estimate + w/2)) + 
      geom_vline(xintercept=mean(boot.samples), color='blue') + 
      geom_vline(xintercept=quantile(boot.samples, c(.025,.975)), color='blue', linetype='dotted') 
  if(interval.replications > 0) { 
    K = interval.replications
    p = p + geom_pointrange(aes(y=y, x=x, xmin=xmin, xmax=xmax), 
          data=data.frame(y=seq(min(drange), max(drange), length.out=K), 
                  x=boot.samples[1:K], xmin=boot.samples[1:K] - w/2, xmax=boot.samples[1:K] + w/2), color='purple', alpha=.2)
  }
  if(point.replications > 0) { 
    K = point.replications
    p = p + geom_point(aes(y=y, x=x), 
          data=data.frame(y=seq(min(drange), max(drange), length.out=K), 
                  x=boot.samples[1:K], xmin=boot.samples[1:K] - w/2, xmax=boot.samples[1:K] + w/2), color='purple', alpha=.2)
  }
  p
}
```

## Last Time

::: {.hidden}
$$
% color defs aren't making it into MathJax
%\definecolor{twocolor-red}{RGB}{248, 118, 109}   % red   used by ggplot in plots w/ 2 colors 
%\definecolor{twocolor-green}{RGB}{0, 191, 196}   % green used by ggplot in plots w/ 2 colors
% so create commands for using those colors instead
\newcommand{\inred}[1]{\textcolor[RGB]{248,118,109}{#1}}
\newcommand{\ingreen}[1]{\textcolor[RGB]{0,191,196}{#1}}
\newcommand{\inblue}[1]{\textcolor{blue}{#1}}

\newcommand{\ind}[1]{1_{\{#1\}}}
\newcommand{\with}{\ingreen{\text{degree}}}
\newcommand{\without}{\inred{\text{no degree}}}

\DeclareMathOperator{\E}{E}degredegre
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\Cov}{Cov}
$$ 

:::


## Recently {.smaller}

```{r}
#| out-height: 250px
#| layout-ncol: 2
#| cache: true


earnings.jitter = ggplot(earnings) + geom_point(aes(x=education, y=income), color='#F7766D',
                              size=dot.size, position=position_jitter(seed=0,w=.3), alpha=.1) +
                   scale_x() + scale_y() + guides(color=FALSE)


# make a fake population by sampling with replacement from the sample then adding fuzz to incomes
# the actual sampled proportion in CPS is about 1/2500, so by doing 25x more dots,
# we're letting each dot  stand in for 100 people in the population
earnings.pop = earnings[sample(1:nrow(earnings), nrow(earnings)*25, replace=TRUE),]
earnings.pop$income = earnings.pop$income + rnorm(nrow(earnings.pop), sd=5000)
earnings.jitter.pop = ggplot(earnings.pop) + geom_point(aes(x=education, y=income), color='#F7766D',
                              size=dot.size/10, position=position_jitter(seed=0,w=.3), alpha=.05) +
                   scale_x() + scale_y() + guides(color=FALSE)

earnings.jitter
earnings.jitter.pop
```


:::: columns

::: column
### Sample
$$
\small{
\begin{array}{ccccccc}
X_1 & Y_1 & X_2 & Y_2 &  \ldots & X_n & Y_n \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\end{array}
}
$$

:::
::: column
### Population
$$
\small{
\begin{array}{ccccccc}
x_1 & y_1 & x_2 & y_2 &  \ldots & x_m & y_m \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\end{array}
}
$$

:::
::::

- We've been talking about ways of summarizing data when we have non-binary covariates.
  - Data that looks like the plot on the left.
  

- We talked about statistical inference based of summaries of a sample
- We'll talk about *statistical inference* based on these sample summaries.
- That is, we'll talk about what they tell us about the population our sample was drawn from.
- To keep us focused, we'll talk mostly about two different *aggregated comparisons*.
  - Both are summaries of the value of a year of high school. 
  - They're different averages of the year-on-year increments in mean income.


Qualitatively, the sample and the population are the same kind of thing. 
- They're both collections of people with incomes and education levels. 
- That means we can summarize them in the same ways.  
- It's natural to think of sample summaries as *estimates* of the corresponding population summaries.


$$
\small{
\begin{aligned}
&\hat \mu(x) = \frac{1}{N_x}\ \sum_{i:X_i=x} Y_i \qfor N_x = \sum_{i:X_i=x} 1 
&& \text{ estimates } \quad \mu(x) = \frac{1}{m_x}\ \sum_{j:x_j=x} y_j \qfor m_x = \sum_{j:x_j=x} 1 \\
&\frac{1}{4}\sum_{x=9}^{12} \qty{ \hat \mu(x) - \hat\mu(x-1) }                       
&& \text{ estimates } \quad \frac{1}{4}\sum_{x=9}^{12} \qty{ \mu(x) - \mu(x-1) }   \\
&\sum_{x=9}^{12} P_x \qty{ \hat \mu(x) - \hat\mu(x-1) } \qfor P_x = \frac{N_x}{\sum_{x=9}^{12} N_x}                       
&& \text{ estimates } \quad \sum_{x=9}^{12} p_x \qty{ \mu(x) - \mu(x-1) }   \qfor p_x = \frac{m_x}{\sum_{x=9}^{12} m_x} \\
\end{aligned}
}
$$


# Inference for Linear Combinations of Subsample Means: Random Coefficients



## Unbiasedness

- Consider the case where the coefficients in our linear combination are random---where they depend on covariates.
- Our estimator is unbiased *if the sample coefficients are unbiased estimates of the population coefficients*.
  - Sample proportions $P_x$ are unbiased for the population proportions $p_x$.
  - Differences in Proportions $P_x - P_{x-1}$ are unbiased for the population differences $p_x - p_{x-1}$.
  - That's enough to handle our example and many other cases.
- The proof is almost the same it is with non-random coefficients.
  - But now we use linearity of *conditional expectations*.
  - And *conditional unbiasedness* of the subsample means.


::: {.fragment}
$$
\small{
\begin{aligned}
&\E\qty[\sum_x \hat\alpha(x) \hat\mu(x)]  \\ 
&= \E\qty[ \E\qty{ \sum_x \hat\alpha(x) \hat\mu(x) \mid X_1 \ldots X_n } ] && \text{via law of iterated expectations} \\
&= \E\qty[ \sum_x \hat\alpha(x) \E\qty{ \hat\mu(x) \mid X_1 \ldots X_n } ] && \text{via linearity of conditional expectations} \\
&= \E\qty[ \sum_x \hat\alpha(x) \mu(x) ]                                   && \text{via conditional unbiasedness of the subsample means} \\
&= \sum_x \E[\hat\alpha(x)] \mu(x) = \sum_x \alpha(x) \mu(x)               && \text{via linearity of expectations and unbiasedness of coefficients}  
\end{aligned} 
}
$$
:::


## Variance Calculation Step 1: Total Variance


- As before, we'll use the law of total variance to break the variance of our estimator into two parts.
- But now, due to the randomness of the coefficients, neither part will be zero. 
- We'll have to deal with deal with both parts.
  - The one we're used to --- the $\E\Var[\ldots]$ part --- will be the almost the same as before.
  - The other one --- the $\Var\E[\ldots]$ part --- will be new.

$$
\small{
\begin{aligned}
&\Var\qty[ \sum_x \hat\alpha(x) \hat \mu(x) ] \\
&= \E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] + \Var\qty[ \E{\sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n} ] \\
&= \E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] + \Var\qty[ \sum_x \hat \alpha_x \E{ \hat \mu(x) \mid X_1 \ldots X_n} ] \\
&= \E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] + \Var\qty[ \sum_x \hat \alpha_x \mu(x) ] \\
\end{aligned}
}
$$


## Variance Calculation Step 2: The $\E[\Var[\ldots]]$ Part.

- What we get for this part is *almost* the same as with non-random coefficients.
- Now that the coefficients are random, their squares go inside the expectation with the inverse-sample-size $1/N_x$.
- That's all that changes. The calculation is essentially the same, too. It's below, but we'll skip it.

$$\small{
\begin{aligned}
&\E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] \\ 
&=\E\qty[ \E\qty{ \qty(\sum_x \hat\alpha(x) \hat\mu(x) - \E[\sum_x \hat\alpha(x) \hat\mu(x) \mid X_1 \ldots X_n])^2 \mid X_1 \ldots X_n }] \\
&=\E\qty[ \E\qty{ \qty(\sum_x \hat\alpha(x) \hat\mu(x) - \hat\alpha(x) \E\qty[\hat\mu(x) \mid X_1 \ldots X_n])^2 \mid X_1 \ldots X_n }] \\
&=\E\qty[ \E\qty{ \qty(\sum_x \hat\alpha(x) \qty{\hat\mu(x) - \E[\hat\mu(x) \mid X_1 \ldots X_n]})^2 \mid X_1 \ldots X_n }] \\
&=\E\qty[ \E\qty{ \sum_x\sum_x' \hat\alpha(x) \hat\alpha x' \qty{\hat\mu(x) - \E[\hat\mu(x) \mid X_1 \ldots X_n]}  \qty{\hat\mu(x) - \E[\hat\mu(x) \mid X_1 \ldots X_n]} \mid X_1 \ldots X_n }] \\ 
&=\E\qty[ \sum_x\sum_x' \hat\alpha(x) \hat\alpha(x') \E\qty{\qty{\hat\mu(x) - \E[\hat\mu(x) \mid X_1 \ldots X_n]}  \qty{\hat\mu(x) - \E[\hat\mu(x) \mid X_1 \ldots X_n]} \mid X_1 \ldots X_n }] \\ 
&=\E\qty[ \sum_x\sum_x' \hat\alpha(x) \hat\alpha(x') \begin{cases} \Var\qty{\hat\mu(x) \mid X_1 \ldots X_n} = \frac{\sigma^2(x)}{N_x} & \text{ if } x=x' \\ 0 & \text{ otherwise} \end{cases} ] \\
&=\E\qty[ \sum_x \hat \alpha^2(x) \frac{\sigma^2(x)}{N_x} ] = \sum_x \sigma^2(x) \times \E\qty[ \frac{\hat\alpha(x)^2}{N_x} ] \\
\end{aligned}
}
$$

## A Pause to Consider the Implications {.smaller}

```{r}
#| out-height: 400px
#| fig-asp: .25

N = function(x) { sum(earnings$education==x) }
P = function(x) { sum(earnings$education==x) / sum(earnings$education <= 12) }
alpha = function(x) { if (x==12) { P(12) } else if (x==8) { -P(9) } else { P(x) - P(x+1) } }
se.people.EV = sqrt(sum(sapply(8:12, function(x) { alpha(x)^2 * summaries$sd[summaries$education==x]^2 / sum(earnings$education==x) })))
se.people.EV.terms = sqrt(sapply(8:12, function(x) { alpha(x)^2 * summaries$sd[summaries$education==x]^2 / sum(earnings$education==x) }))
```

$$
\Var\qty[\sum_x \hat\alpha(x) \hat\mu(x)] \ge \E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] = \sum_x \sigma^2(x) \times \E\qty[ \frac{\hat\alpha(x)^2}{N_x} ]
$$

- Let's think about what this means for our two examples.
$$
\small{
\begin{aligned}
\hat\theta_{\text{years}}  &= \frac14\sum_{x=9}^{12} \qty{ \hat\mu(x) - \hat\mu(x-1)} = \sum_x \alpha(x) \hat \mu(x)  \qfor \alpha(x) = \begin{cases} \frac14 & \text{ if } x = 12 \\ -\frac14 & \text{ if } x= 8 \\ 0 & \text{ otherwise} \end{cases} \\
\hat\theta_{\text{people}} &= \sum_{x=9}^{12} P_x \qty{ \hat\mu(x) - \hat\mu(x-1)} = \sum_x \hat \alpha(x) \hat \mu(x) \qfor
\hat\alpha(x) = \begin{cases}
  P_{12} & \text{ if } x = 12 \\
  P_{x} - P_{x+1} & \text{ if } x \in \{9 \ldots 11\} \\
  -P_9 & \text{ if } x = 8 
\end{cases}
\end{aligned}
}
$$

- We've seen that our estimate of the average-over-years summary of the value of a year of high school has large variance.
- The problem was that it paired a relatively large weight--$\alpha(x)=-1/4$---with a very small sample size $N_8=`r sum(earnings$education==8)`$.
- The average-over-people sumary does better in that respect. It puts much less weight on means of very small subsamples. 
$$ \small{
\begin{array}{c|ccccc}
x & 8 & 9 & 10 & 11 & 12 \\
N_x & `r N(8)` & `r N(9)` & `r N(10)` & `r N(11)` & `r N(12)` \\
\hat\alpha(x) & `r approx(alpha(8))` & `r approx(alpha(9))` & `r approx(alpha(10))` & `r approx(alpha(11))` & `r approx(alpha(12))` \\
\end{array}
}
$$

## Instead ...
$$ \small{
\begin{array}{c|ccccc}
x & 8 & 9 & 10 & 11 & 12 \\
N_x & `r N(8)` & `r N(9)` & `r N(10)` & `r N(11)` & `r N(12)` \\
\hat\alpha(x) & `r approx(alpha(8))` & `r approx(alpha(9))` & `r approx(alpha(10))` & `r approx(alpha(11))` & `r approx(alpha(12))` \\
\end{array}
}
$$

- Much of its variance comes from putting *much more* weight on the mean of a small, but less small, sample: the one with $x=11$.
- This variance term, $\sigma^2(11) \times \E[\frac{\hat\alpha(11)^2}{N_{11}}] \approx \sigma^2(11) \times `r round(alpha(11)^2/N(11),3)`$,
  is bigger than the term $\sigma^2(8) \times (1/4)^2/N_8 \approx \sigma^2(8) \times `r round((1/16)/N(8),3)`$ we worried about before.
- Even based on this *part* of the variance, we can see that this summary will have a bigger variance than the average-over-years one.
- This part of the former's variance is $`r approx(se.people.EV)`^2$ vs. the latter's whole variance $`r approx(se)`^2$. Let's get a sense of the other part.


## Variance Calculation Step 3: The $\Var[\E[\ldots]]$ Part.

- Now we'll calculate the new part --- the $\Var[\E[\ldots]]$ part. Roughly speaking.
- This **additional variance** is what we pay for using *estimates* of the population coefficients $\alpha(x)$.
- We're going to use a formula for the variance of a linear combination of random variables that you may have seen before.
$$
\Var\qty[\sum_x a_x Z_x] = \sum_x \sum_x' a_x a_x' \Cov(Z_x,\ Z_y) \qfor \Cov(Z_x,\ Z_y) = \E\qty[ \qty{Z_x - \E[Z_x]} \qty{Z_y - \E[Z_y]} ]
$$
- It's easy enough to prove, but we'll skip that. Here's what it tells us.
$$
\Var\qty[ \E{\sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n} ] 
= \Var\qty[ \sum_x \mu(x) \hat \alpha(x)  ] \\
= \sum_x \sum_x' \mu(x) \mu(x') \Cov\qty[\hat\alpha(x),\ \hat\alpha(x')] 
$$
- Calculating this is do-able but pretty involved. We'll skip the details.

## Not Details

$$
\Var\qty[ \E{\sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n} ] 
= \Var\qty[ \sum_x \mu(x) \hat \alpha(x)  ] \\
= \sum_x \sum_x' \mu(x) \mu(x') \Cov\qty[\hat\alpha(x),\ \hat\alpha(x')] 
$$

- We'll just do enough to get a handle on the size of this additional variance is. 
- We'll do this by calculating one of these covariance factors---we'll do $\hat\alpha(9)$ and itself.
$$
\Cov(\hat\alpha(9), \alpha(9)) = \Cov(P_9-P_{10},\ P_{9}-P_{10}) = \Var(P_9) - 2\Cov(P_9,\ P_{10}) + \Var(P_{10}).
$$
- In short, it --- and presumably the other terms --- can't be much bigger than the variance of the individual sample proportions $P_{x}$.
- And those are roughly $p_x (1-p_x) / \sum_x N_x$, i.e., the variance of an average of Bernoulli random variables with probability $p_x$.
  - What's important is that *the denominator here is the sum of the sample sizes*. 
  - It's never small like the denominators in the other term can be.
  - As a result, this additional variance term is *often much smaller* than the other one.
- We'll see this in our average-over-people example.

## Application


```{r}
#| out-height: 250px
#| layout-ncol: 2
#| fig-asp: .4
#| fig-subcap: 
#|  - Our High School Data
#|  - Bootstrap sampling distributions for the average-over-people (red) and average-over-years (blue) summaries.


B=1000
boot = purrr::map(1:B, function(i) {
  earnings.boot = earnings[sample(1:nrow(earnings), replace=TRUE),]
  summaries = earnings.boot[earnings.boot$education <= 12, ] %>% group_by(education) %>% summarize(mean=mean(income), N=length(income))
  data.frame(people=average.over.people(summaries), years=(summaries$mean[summaries$education==12] - summaries$mean[summaries$education==8])/4)
}) %>% bind_rows()

summaries = earnings[earnings$education <= 12, ] %>% group_by(education) %>% summarize(mean=mean(income), N=length(income))
aggregate.comparison.plot + xlim(7.7,12.5)
ggplot(boot) + geom_histogram(aes(x=people,y=after_stat(density)), fill='red', alpha=.3, bins=100) + 
               geom_histogram(aes(x=years,y=after_stat(density)), fill='blue', alpha=.3, bins=100) 
```

- Now that we've made sense of the variance of the average-over-people summary in the sample, let's use it.
- We'll construct an interval estimate for value of a year of high school in our population. 
- Since we haven't done the variance calculation completely, we'll use the bootstrap to estimate the standard error.
- What we get is `r approx(sd(boot$people))`, very close to the standard error `r approx(se.people.EV)`
  we calculated for the 'old' $\E\Var[\ldots]$ part of the variance.
- And roughly twice as big as the standard error `r approx(se)` we calculated for the average-over-years summary.
$$
\begin{array}{l|l}
\text{Summary} & \text{Point Estimate} & \text{95\% Confidence Interval} \\
\theta_{\text{people}} & `r approx(estimate)` 
&  \qty[`r approx(estimate-1.96*se)`, \ `r approx(estimate+1.96*se)`] \\
\theta_{\text{years}}  & `r approx(average.over.people(summaries))` 
& \qty[`r approx(average.over.people(summaries)-1.96*sd(boot$people))`, \ `r approx(average.over.people(summaries)+1.96*sd(boot$people))`]
\end{array}
$$

```{r}

```

## This is Dissatisfying

$$
\begin{array}{l|ll}
\text{Summary} & \text{Point Estimate} & \text{95\% Confidence Interval} \\
\theta_{\text{years}} & `r approx(estimate)` 
&  \qty[`r approx(estimate-1.96*se)`, \ `r approx(estimate+1.96*se)`] \\
\theta_{\text{people}}  & `r approx(average.over.people(summaries))` 
& \qty[`r approx(average.over.people(summaries)-1.96*sd(boot$people))`, \ `r approx(average.over.people(summaries)+1.96*sd(boot$people))`]
\end{array}
$$

- No matter how we average, we don't have a very precise estimate of the 'value of a year of high school.'
- It's clear that the essential problem is that we're heavily weighting the mean of some very small subsample.
- Can we do better? Perhaps by estimating the subpopulation mean *without simply using the subsample mean*?
- Let's see what happens when we *fit a line* to the data we've been working with and use that instead. 
- This lets us use that larger sample, not just the subsample with $X_i=x$, to estimate the subpopulation mean $\mu(x)$.
$$ 
\small{
\hat \mu(x) = \hat b_1 x + \hat b_0 \qfor \hat b = \operatorname*{argmin}_{b} \sum_{i:X_i \in 8 \ldots 12}^n (b_1 X_i + b_0 - Y_i)^2 
\qqtext{ instead of } \hat \mu(x) = \operatorname*{argmin}_{m} \sum_{i:X_i=x} (m - Y_i)^2  
}
$$

## Estimating Subpopulation Means with a Fitted Line

```{r}
model = lm(income ~ education, data=earnings[earnings$education <= 12,])
muhat.line = function(x) { predict(model, newdata=data.frame(education=x)) }
x = seq(8,12,by=.01)

earnings.jitter.hs = ggplot(earnings) + geom_point(aes(x=education, y=income, alpha=education <= 12), color='#F7766D',
                              size=dot.size, position=position_jitter(seed=0,w=.3)) +
                   scale_x() + scale_y() + scale_alpha_manual(breaks=c(FALSE, TRUE), values=c(.01,.3)) + guides(color=FALSE, alpha=FALSE)
summaries = earnings %>% group_by(education) %>% summarize(mean=mean(income), sd=sd(income), N=length(income))
earnings.jitter.hs +
  geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries[summaries$education <= 12,]) +
  geom_line(aes(x=education, y=mean), data=summaries[summaries$education<=12,], color='black', linetype='dashed') +
  geom_point(aes(x=x,y=y), data=data.frame(x=8:12, y=muhat.line(8:12)), color='magenta') + 
  geom_line(aes(x=x, y=y), data=data.frame(x=x, y=muhat.line(x)), color='magenta', linetype='dashed') + xlim(7.7,12.5) + scale_y(c(0, .7e5)) 
```

- Here's our line. You may remember that, when our covariate is binary, the least squares line will go through the subsample means.
- That's not true more generally. It can't be---that'd require our subsample means to actually be on a line. They aren't.
- So our new estimates of the subpopulation means, $\hat\mu(x)$, differ from the subsample means.
- **Q**. What happens when we estimate our population summaries?

::: fragment

- **A**. They're the same---they're both the slope of the estimated line.
- Last class, we talked about that as *a different summary*. And it is.
- Using a model like this to estimate subpopulation means---one that can't really fit the data---is a way of *diguising* a different summary as the one we're interested in. 

$$
\begin{aligned}
\hat\theta_{\text{years}}  &= \frac{\hat\mu(12) - \hat \mu(8)}{4} = \frac{\qty(\hat b_1 \times 12 + \hat b_0) - \qty(\hat b_1 \times 8 + \hat b_0)}{4} = 
\frac{\hat b_1 \times (12-8)}{4} = \hat b_1 \\

\hat\theta_{\text{people}} &= 
\sum_{x \in 9 \ldots 12} P_x \qty{\hat\mu(x) - \hat \mu(x-1)} = \sum_{x \in 9 \ldots 12} P_x \qty{\qty(\hat b_1 \times x + \hat b_0) - \qty(\hat b_1 \times (x-1) + \hat b_0)} = 
\sum_{x \in 9 \ldots 12} P_x \hat b_1 = \hat b_1
\end{aligned}
$$

:::

## Result

```{r}
B=1000
boot.line = purrr::map(1:B, function(i) {
  earnings.boot = earnings[sample(1:nrow(earnings), replace=TRUE),]
  model = lm(income ~ education, data=earnings.boot[earnings.boot$education <= 12,])
  data.frame(line=coef(model)[2])
}) %>% bind_rows()
boots = bind_cols(boot[1:B,], boot.line)
```
$$
\begin{array}{l|ll}
\text{Summary} & \text{Point Estimate} & \text{95\% Confidence Interval} \\
\theta_{\text{years}} & `r approx(estimate)` 
&  \qty[`r approx(estimate-1.96*se)`, \ `r approx(estimate+1.96*se)`] \\
\theta_{\text{people}}  & `r approx(average.over.people(summaries))` 
& \qty[`r approx(average.over.people(summaries)-1.96*sd(boot$people))`, \ `r approx(average.over.people(summaries)+1.96*sd(boot$people))`] \\
\theta_{\text{slope}}  & `r approx(coef(model)[2])` 
& \qty[`r approx(coef(model)[2]-1.96*sd(boots$line))`, \ `r approx(coef(model)[2]+1.96*sd(boots$line))`]
\end{array}
$$

- In this instance, it's not a disaster---the point estimate is about the same as the other two.
- But it's not a huge win either---the standard error is more or less the same as the average-over-years summary's.
 
```{r}
#| out-height: 250px
#| fig-asp: .25
#| fig-cap: Bootstrap sampling distributions for the average-over-people (red), average-over-years (blue), and least squares slope (magenta) summaries.
ggplot(boots) + geom_histogram(aes(x=people,y=after_stat(density)), fill='red', alpha=.3, bins=100) + 
                geom_histogram(aes(x=years,y=after_stat(density)), fill='blue', alpha=.3, bins=100) +
                geom_histogram(aes(x=line,y=after_stat(density)), fill='magenta', alpha=.3, bins=100) 
```

## That Was Lucky {.smaller}

- Remember that last class, we looked at fake data in which these three summaries were very different.
- That can happen. And if you're pretending the least-squares slope is one of the other summaries, you'll be very wrong.
- What's worse is, you won't know it. Your confidence interval won't be wide enough to cover the summary you're interested in.
- That's because your slope-masquerading-as-something-else estimator is *biased* for the summary you're interested in.
- And our confidence intervals don't account for that. 

$$
$$

- Let's take a look at the three intervals we get using our fake data from last class. 
- We're not lucky in any of these cases.

## When We're Not Lucky

```{r}
mu = map_dbl(1:12, function(x) { mean(earnings$income[earnings$education==x]) })
epsilon = rnorm(nrow(earnings)) * .1*sd(earnings$income[earnings$education %in% 8:12])

earnings1 = earnings
earnings1$income = mu[ifelse(earnings$education==12, 12, 8)] + epsilon

earnings2 = earnings
earnings2$income = mu[ifelse(earnings$education >= 10, 12, 8)] + epsilon

mu3 = NA+mu
mu3[8]=mu[8]
mu3[9]=mu[10]
mu3[10]=(mu[11]+mu[12])/2
mu3[11]=mu[12]
mu3[12]=mu[12]
earnings3 = earnings
earnings3$income = mu3[earnings$education] + epsilon
```



- Story 1. All that matters is the diploma.

```{r}
boots=run.boot(earnings1)
```

$$
\small{
\begin{array}{l|ll}
\text{Summary} & \text{Point Estimate} & \text{95\% Confidence Interval} \\
\theta_{\text{years}} & `r approx(mean(boots$years))` 
&  \qty[`r approx(mean(boots$years) - 1.96*sd(boots$years))`, \ `r approx(mean(boots$years) + 1.96*sd(boots$years))`] \\
\theta_{\text{people}} & `r approx(mean(boots$people))` 
&  \qty[`r approx(mean(boots$people) - 1.96*sd(boots$people))`, \ `r approx(mean(boots$people) + 1.96*sd(boots$people))`] \\
\theta_{\text{line}} & `r approx(mean(boots$line))` 
&  \qty[`r approx(mean(boots$line) - 1.96*sd(boots$line))`, \ `r approx(mean(boots$line) + 1.96*sd(boots$line))`] \\
\end{array}
}
$$

- Story 2. All that matters is 10th grade algebra.
```{r}
boots=run.boot(earnings2)
```

$$
\small{
\begin{array}{l|ll}
\text{Summary} & \text{Point Estimate} & \text{95\% Confidence Interval} \\
\theta_{\text{years}} & `r approx(mean(boots$years))` 
&  \qty[`r approx(mean(boots$years) - 1.96*sd(boots$years))`, \ `r approx(mean(boots$years) + 1.96*sd(boots$years))`] \\
\theta_{\text{people}} & `r approx(mean(boots$people))` 
&  \qty[`r approx(mean(boots$people) - 1.96*sd(boots$people))`, \ `r approx(mean(boots$people) + 1.96*sd(boots$people))`] \\
\theta_{\text{line}} & `r approx(mean(boots$line))` 
&  \qty[`r approx(mean(boots$line) - 1.96*sd(boots$line))`, \ `r approx(mean(boots$line) + 1.96*sd(boots$line))`] \\
\end{array}
}
$$

- Story 3. They teach the most important stuff first. 
```{r}
boots=run.boot(earnings3)
```

$$
\small{
\begin{array}{l|ll}
\text{Summary} & \text{Point Estimate} & \text{95\% Confidence Interval} \\
\theta_{\text{years}} & `r approx(mean(boots$years))` 
&  \qty[`r approx(mean(boots$years) - 1.96*sd(boots$years))`, \ `r approx(mean(boots$years) + 1.96*sd(boots$years))`] \\
\theta_{\text{people}} & `r approx(mean(boots$people))` 
&  \qty[`r approx(mean(boots$people) - 1.96*sd(boots$people))`, \ `r approx(mean(boots$people) + 1.96*sd(boots$people))`] \\
\theta_{\text{line}} & `r approx(mean(boots$line))` 
&  \qty[`r approx(mean(boots$line) - 1.96*sd(boots$line))`, \ `r approx(mean(boots$line) + 1.96*sd(boots$line))`] \\
\end{array}
}
$$

## Next Class

- We'll talk about the implications of bias for confidence intervals. Something very interesting happens. 
- We'll see that using problematic models like our line can be ok in small samples, but not at all ok in large ones.
- That's part of what why fitting lines is such a big part of statistics education. It's anachronistic.
- Before we had computers, we had small samples. And using lines this way was --- sometimes --- justifiable.
- Now we, by and large, have large samples. And using lines this way is --- almost always --- not.
