---
title: "QTM 220 Lecture 11"
subtitle: "Inferential Statistics with Non-Binary Covariates: Part 2"
format: 
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---

```{r load-libraries}
library(purrr)
library(tibble)
library(dplyr)
library(magrittr)
library(stringr)
library(ggplot2)
library(htmltools)
library(kableExtra)
library(scales)
library(cpsR)
library(mapdata)

options(knitr.kable.NA = '')

compact_number = label_number(accuracy=.1, scale_cut=cut_short_scale())
ellipsis <- function(df, nhead=5L, ntail=2) {
    rownames(df) = as.character(1:nrow(df))
    stopifnot("data.frame" %in% class(df))
    els <- rep(NA, ncol(df)) %>% 
        matrix(nrow=1, dimnames=list(NULL, names(df))) %>% 
        data.frame(stringsAsFactors=FALSE) %>% 
        tibble::as_tibble()
    rownames(els) = "⋮"
    dplyr::bind_rows(head(df, nhead), els, tail(df, ntail))
}

percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }
```

```{r ggplot-themes}
fontsize = 16
textcolor = "#888888"

class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
        panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x  = element_text(colour = textcolor, size=fontsize),
        axis.text.y  = element_text(colour = textcolor, size=fontsize),
        axis.title.x  = element_text(colour = textcolor),
        axis.title.y  = element_text(colour = textcolor, angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
              axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
           panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(simplified.theme)
```




```{r get-data}
#| cache: true
#| 
# for cps variable interpretation see https://www2.census.gov/programs-surveys/cps/datasets/2022/march/asec2022_ddl_pub_full.pdf
census.api.key = '8084f81e4f71e4d32638dac15f188b424f72d063'
cps.sampling.ratio = 1/25 # Not true. 1/2500 households are surveyed. But for the sake of plotting, it's ok.

california.state.code = 6
year = 2020
cps.data = results=get_asec(year=year, vars=c('pearnval', 'a_age', 'a_hga', 'gestfips', 'gtco', 'gtcbsasz'), key=census.api.key)
```

```{r transform-data}
#| cache: false
#| 
all.earnings = data.frame(state.code = cps.data$gestfips,
                          county.code = cps.data$gtco,
                          age=cps.data$a_age,
                          income=cps.data$pearnval,
                          education.code=cps.data$a_hga,
                          education=case_match(cps.data$a_hga,
                                           0  ~ 0,   # child
                                           31 ~ 0,   # < grade 1
                                           32 ~ 4,   #  grade 1-4
                                           33 ~ 6,   #  grade 5-6
                                           34 ~ 8,   #  grade 7-8 
                                           35 ~ 9,   #  grade 9
                                           36 ~ 10,  #  grade 10
                                           37 ~ 11,  #  grade 11
                                           38 ~ 11,  #  grade 12 no diploma
                                           39 ~ 12,  #  high school grad
                                           40 ~ 13,  #  some college
                                           41 ~ 14,  #  associate degree (vocational)
                                           42 ~ 14,  #  associate degree (academic)
                                           43 ~ 16,  #  bachelors degree
                                           44 ~ 18,  #  masters degree
                                           45 ~ 20,  #  professional school degree
                                           46 ~ 20)) #  doctorate

all.earnings$degree = factor(all.earnings$education >= 16, 
                     levels=c(FALSE, TRUE), 
                     labels=c('no 4-year degree','4-year degree')) 

coarsen = function(x, group, seed=0) {
  current.seed = .Random.seed
  set.seed(seed)
  x.coarsened = rep(NA, length(x))
  for(gg in unique(group)) {
    x.coarsened[group==gg] = runif(sum(group==gg), min(x[group==gg]), max(x[group==gg]))
  }
  set.seed(current.seed)
  x.coarsened
} 

in.interval = function(X, interval) { interval[1] <= X & X <= interval[2] }
earnings = filter(all.earnings, education >= 8 & 
                                state.code == california.state.code & 
                                age %>% in.interval(c(25,35)))
Y = earnings$income
X = earnings$education
```


```{r get-county-info}
# county info from https://en.wikipedia.org/wiki/List_of_counties_in_California
# converted to csv using https://wikitable2csv.ggor.de

to.number = function(s) { as.numeric(str_trim(gsub(",", "", s))) }
ca_county_info = read.csv('california_county_info.csv') %>%
  mutate(subregion = tolower(sub(' County', '', County))) %>%
  mutate(pop = to.number(Population..2022.)) %>%
  mutate(area = to.number(sub('sq.*', '', Area))) 

earnings.county = left_join(earnings, ca_county_info[,c('FIPS.code', 'subregion')], by=c('county.code' = 'FIPS.code'))
earnings$county = earnings.county$subregion
earnings$county[is.na(earnings$county)]='unknown'
```


```{r make-maps}
#| cache: true
#| 

# To give our survey recipients a bit more reality, we'll give them fake locations on a map
# And we'll make up a population of California residents that they're sampled from
# maps are based on code from https://eriqande.github.io/rep-res-eeb-2017/map-making-in-R.html

counties = map_data("county")
ca_county = counties %>%
  filter(region == "california")
ca_county_centers = ca_county %>%
  group_by(subregion) %>%
  summarize(county.lat = mean(lat), county.long=mean(long))

ca_county_more = left_join(ca_county_centers, ca_county_info, by="subregion")
ca_all = left_join(ca_county, ca_county_more, by="subregion")


ca.plot = ggplot(data=ca_all, mapping=aes(x=long, y=lat, group=group)) + 
  coord_quickmap() +
  geom_polygon(fill='gray', color = "white", alpha=.4) +
      scale_fill_gradient(trans="log10") + 
      theme_void()

ca.pop.plot = ggplot(data=ca_all, mapping=aes(x=long, y=lat, group=group)) + 
  coord_quickmap() +
  geom_polygon(aes(fill = pop/area), color = "white") +
      scale_fill_gradient(trans="log10") + 
      theme_void()

## we'll make up a population as follows
# 1. choose the number of people to be the number of people in our sample scaled up to account for the cps sampling ratio 1/2500
#    Because this would wind up being computationally and visually problematic, we'll scale up by 25 instead of 2500;
#    each population dot in the population stands for 100 people in the population.
# 2. for each, sample a county with probability proportional to its finite.population
# 3. for each, sample a latitude and longitude by taking a randomly weighted combination of the    
#    county's vertices. As long as the county is convex, these points will be inside the county.
# 4. we'll give them a degree too, chosen independently of all else from the empirical distribution on the sample
m = nrow(earnings) * 25 

# takes a list of subregions, possibly with duplicates
# for each element, chooses random lat/long coordinates within the sample.subregion
# by taking a random convex combination of the subregion's vertices
sample.within.subregion = function(subregions) {
  output = data.frame(subregion=subregions, lat=NA, long=NA)
  for(subregion in unique(subregions)) { 
      vertices = ca_county[ca_county$subregion==subregion, c('lat','long')]
      U = sum(subregions==subregion)
      V = nrow(vertices)
      weights = abs(rt(U*V, 1))
      dim(weights) = c(U,V)
      weights = (1/rowSums(weights)) * weights
      output[subregions==subregion, c('lat', 'long')] = 
                 cbind(weights %*% vertices$lat, weights %*% vertices$long)
    }
    output
}

finite.population.subregion = sample(ca_county_more$subregion, size=m, replace=TRUE, prob=ca_county_more$pop)
finite.population = sample.within.subregion(finite.population.subregion)
finite.population$degree = sample(earnings$degree, nrow(finite.population), replace=TRUE)
finite.population$group=1 # for plotting

display.population = finite.population
```


```{r make-sample}
# and we'll force the degree-distribution in the sample to match our survey responses
our.sample = sample_n(finite.population, nrow(earnings))
our.sample$degree=earnings$degree

pop.plot = ca.plot + geom_point(aes(x=long, y=lat), data=display.population, size=.02, alpha=.1) 
sample.plot = ca.plot + geom_point(aes(x=long, y=lat), data=our.sample, size=.2, alpha=.5)
pop.plot.inf = ca.pop.plot + geom_point(aes(x=long, y=lat), data=display.population, size=.01, alpha=.1)  
sample.plot.inf = ca.pop.plot + geom_point(aes(x=long, y=lat), data=our.sample, size=.2, alpha=.5)
```

```{r visualization-defaults}
scale_y = function(limits=c(-.3e5, 3e5)) { 
  scale_y_continuous(labels=label_dollar(), limits=limits)
}

years = 8:20
scale_x = function(limits=c(min(years)-.5, max(years)+.5)) { 
  scale_x_continuous(breaks=years, labels=sprintf('%dy', years), limits=limits)
}

dot.size = 2
```


```{r define-colors}
twocolor.red   = 'rgb(248,118,109)'
twocolor.green = 'rgb(0,191,196)'
```


```{r scatter-plots}

earnings.jitter = ggplot(earnings) + geom_point(aes(x=education, y=income), color='#F7766D',
                              size=dot.size, position=position_jitter(seed=0,w=.3), alpha=.1) +
                   scale_x() + scale_y() + guides(color=FALSE)
```

```{r bootstrapping}
#| cache: true

average.over.people = function(summaries) {
  muhat = function(x) { summaries$mean[summaries$education==x] }
  P = function(x) { summaries$N[summaries$education==x] / sum(summaries$N) }
  alphahat = function(x) { if (x==12) { P(12) } else if (x==8) { -P(9) } else { P(x) - P(x+1) } }  
  sapply(8:12, function(x) { alphahat(x) * muhat(x) }) %>% sum
}
run.boot = function(earnings, B=1000) { purrr::map(1:B, function(i) {
  earnings.boot = earnings[sample(1:nrow(earnings), replace=TRUE),]
  summaries = earnings.boot[earnings.boot$education <= 12, ] %>% group_by(education) %>% summarize(mean=mean(income), N=length(income))
  model = lm(income ~ education, data=earnings.boot[earnings.boot$education <= 12,])
  data.frame(people=average.over.people(summaries), 
             years=(summaries$mean[summaries$education==12] - summaries$mean[summaries$education==8])/4,
             line = coef(model)[2])
}) %>% bind_rows()
}
boots = run.boot(earnings)
summaries = earnings[earnings$education <= 12, ] %>% group_by(education) %>% 
  summarize(mean=mean(income), sd=sd(income), N=length(income))
years.estimate = (summaries$mean[summaries$education==12] - summaries$mean[summaries$education==8])/4
```

```{r plotting-code}
width = function(samples, center=mean(samples), alpha=.05) {
  covg.dif = function(w) { 
    actual = mean( center - w/2 <= samples & samples <= center + w/2)
    nominal = 1-alpha 
    actual - nominal  
  }
  uniroot(covg.dif, interval=c(0, 2*max(abs(samples-center))))$root
}

distribution.plot = function(boot.samples, point.estimate, w=width(boot.samples), interval.replications=0, point.replications=0) { 
  drange = c(0, dnorm(0, mean=0, sd=sd(boot.samples)))
  p = ggplot(data.frame(samples=boot.samples)) + 
    geom_histogram(aes(x=samples, y=after_stat(density)), alpha=.3, bins=50)  +
    geom_pointrange(aes(y=dnorm(2*sd(samples), mean=0, sd=sd(samples)), 
      x=point.estimate, xmin=point.estimate - w/2, xmax=point.estimate + w/2)) + 
      geom_vline(xintercept=mean(boot.samples), color='blue') + 
      geom_vline(xintercept=quantile(boot.samples, c(.025,.975)), color='blue', linetype='dotted') 
  if(interval.replications > 0) { 
    K = interval.replications
    p = p + geom_pointrange(aes(y=y, x=x, xmin=xmin, xmax=xmax), 
          data=data.frame(y=seq(min(drange), max(drange), length.out=K), 
                  x=boot.samples[1:K], xmin=boot.samples[1:K] - w/2, xmax=boot.samples[1:K] + w/2), color='purple', alpha=.2)
  }
  if(point.replications > 0) { 
    K = point.replications
    p = p + geom_point(aes(y=y, x=x), 
          data=data.frame(y=seq(min(drange), max(drange), length.out=K), 
                  x=boot.samples[1:K], xmin=boot.samples[1:K] - w/2, xmax=boot.samples[1:K] + w/2), color='purple', alpha=.2)
  }
  p
}
```

## Recently {.smaller}

::: {.hidden}
$$
% color defs aren't making it into MathJax
%\definecolor{twocolor-red}{RGB}{248, 118, 109}   % red   used by ggplot in plots w/ 2 colors 
%\definecolor{twocolor-green}{RGB}{0, 191, 196}   % green used by ggplot in plots w/ 2 colors
% so create commands for using those colors instead
\newcommand{\inred}[1]{\textcolor[RGB]{248,118,109}{#1}}
\newcommand{\ingreen}[1]{\textcolor[RGB]{0,191,196}{#1}}
\newcommand{\inblue}[1]{\textcolor{blue}{#1}}

\newcommand{\ind}[1]{1_{\{#1\}}}
\newcommand{\with}{\ingreen{\text{degree}}}
\newcommand{\without}{\inred{\text{no degree}}}

\DeclareMathOperator{\E}{E}degredegre
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\Cov}{Cov}
$$ 

:::



```{r}
#| out-height: 200px
#| layout-ncol: 2
#| cache: true


earnings.jitter = ggplot(earnings) + geom_point(aes(x=education, y=income), color='#F7766D',
                              size=dot.size, position=position_jitter(seed=0,w=.3), alpha=.3) +
                   scale_x() + scale_y() + guides(color=FALSE)


# make a fake population by sampling with replacement from the sample then adding fuzz to incomes
# the actual sampled proportion in CPS is about 1/2500, so by doing 25x more dots,
# we're letting each dot  stand in for 100 people in the population
earnings.pop = earnings[sample(1:nrow(earnings), nrow(earnings)*25, replace=TRUE),]
earnings.pop$income = earnings.pop$income + rnorm(nrow(earnings.pop), sd=5000)
earnings.jitter.pop = ggplot(earnings.pop) + geom_point(aes(x=education, y=income), color='#F7766D',
                              size=dot.size/10, position=position_jitter(seed=0,w=.3), alpha=.15) +
                   scale_x() + scale_y() + guides(color=FALSE)

earnings.jitter
earnings.jitter.pop
```


:::: columns

::: column
### Sample
$$
\small{
\begin{array}{ccccccc}
X_1 & Y_1 & X_2 & Y_2 &  \ldots & X_n & Y_n \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\end{array}
}
$$

:::
::: column
### Population
$$
\small{
\begin{array}{ccccccc}
x_1 & y_1 & x_2 & y_2 &  \ldots & x_m & y_m \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\end{array}
}
$$

:::
::::

- We've been talking about ways of summarizing data when we have non-binary covariates.
  - Data that looks like the plot on the left.
- We've talked about how to use those summaries to estimate corresponding *population summaries*.
  - Assuming the data we've observed is sampled with replacement from that population.
  - A population that looks like the plot on the right.
- And we've talked about what makes some harder to estimate than others.
  - i.e., the *variance* of the corresponding sample summaries.


## Our Example {.smaller}

```{r}
#| out-height: 200px
#| layout-ncol: 2

pop.summaries = earnings.pop %>% 
          group_by(education) %>% 
          summarize(mean=mean(income), sd=sd(income), N=length(income))
earnings.jitter + geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries, size=1) +
                  geom_line(aes(x=education, y=mean), data=summaries, linetype='dotted') + xlim(7.5, 12.5) + scale_y(c(-10e3,80e3))
earnings.jitter.pop + geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), data=pop.summaries, size=1) +
                      geom_line(aes(x=education, y=mean), data=pop.summaries, linetype='dotted') + xlim(7.5,12.5) + scale_y(c(-10e3,80e3))
```

$$
\small{
\begin{aligned}
&\frac{1}{4}\sum_{x=9}^{12} \qty{ \hat \mu(x) - \hat\mu(x-1) }                       
&& \text{ estimates } \quad \frac{1}{4}\sum_{x=9}^{12} \qty{ \mu(x) - \mu(x-1) }   \\
&\sum_{x=9}^{12} P_x \qty{ \hat \mu(x) - \hat\mu(x-1) } \qfor P_x = \frac{N_x}{\sum_{x=9}^{12} N_x}                       
&& \text{ estimates } \quad \sum_{x=9}^{12} p_x \qty{ \mu(x) - \mu(x-1) }   \qfor p_x = \frac{m_x}{\sum_{x=9}^{12} m_x} \\
\end{aligned}
}
$$

- We've focused on two main examples. Both summarize the value of a year of high school. 
- The first is the mean increment in income associated with *the four years* of high school.
- The second is the mean increment in income associated with *students' last year* of high school ...
  - where these students are California residents who did not attend college.
    - In our sample summary, we're talking about *sampled California residents* who didn't attend college.
    - In our population summary, we're talking about *all California residents* who didn't attend college.
  - This summary speaks to the benefit to these residents of not stopping school a year earlier.



## What We See

```{r}
#| layout-ncol: 2
#| out-height: 300px
#| fig-asp: .3 
#| fig-subcap:
#|  - Interval Estimate for the Average over Years
#|  - Interval Estimate for the Average over People 

distribution.plot(boots$years, years.estimate) + xlim(-5000,10000)
distribution.plot(boots$people, average.over.people(summaries)) + xlim(-5000,10000)
```

- We use *confidence intervals* to characterize our uncertainty about the values of the population summaries.
- And we can calibrate these intervals using the *bootstrap sampling distribution* of our summaries.
- What we see is that we're less certain about the --- perhaps more meaningful --- average over people.
- And we've gotten about half-way to working out why that is.
  - We've worked out a formula for the variance of the average over years. 
   - This lets us see what's behind our uncertainty.
  - Today, we'll do the same for the average over people.

## Abstractions {.smaller }
```{r}
#| out-height: 200px
#| layout-ncol: 2

pop.summaries = earnings.pop %>% 
          group_by(education) %>% 
          summarize(mean=mean(income), sd=sd(income), N=length(income))
earnings.jitter + geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries, size=1) +
                  geom_line(aes(x=education, y=mean), data=summaries, linetype='dotted') + xlim(7.5, 12.5) + scale_y(c(-10e3,80e3))
earnings.jitter.pop + geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), data=pop.summaries, size=1) +
                      geom_line(aes(x=education, y=mean), data=pop.summaries, linetype='dotted') + xlim(7.5,12.5) + scale_y(c(-10e3,80e3))
```
$$
\hat \theta = \sum_x \hat\alpha(x) \hat \mu(x) \quad \text{ \textcolor{gray}{estimates }} \quad \theta = \sum_x \hat \alpha(x) \hat\mu(x)
$$

- Rather than tackling our two examples specifically, we'll work in more abstract terms.
- We'll talk about inference for *linear combinations of subsample means* --- coefficient-weighted sums of them..
- Working more abstractly is, in a way, easier. We can background some of the details.
- And it's also more useful. It means we can apply our results to other examples, too.
- All we have to do is rewrite our examples in these terms.

## Rewriting Our Examples {.smaller}

- Last time, we tackled combinations with *deterministic coefficients*, like the average over years. 
$$
\small{
\begin{aligned}
&\frac{1}{4}\sum_{x=9}^{12} \qty{ \hat \mu(x) - \hat\mu(x-1) }  
= \sum_x \alpha(x) \hat \mu(x) \qfor \alpha(x) = 
\begin{cases} 
-\frac14 & \text{ if } x=8 \\
\hphantom{-}\frac14 & \text{ if } x=12 \\
0 & \text{ otherwise}
\end{cases} && \text{ \textcolor{gray}{estimates}} \\
&\frac{1}{4}\sum_{x=9}^{12} \qty{ \mu(x) - \mu(x-1) }  = \sum_x \alpha(x) \mu(x)  \qfor \text{ the same coefficients } && 
\end{aligned}
}
$$

- Today, we'll look at combinations with *random coefficients*, like the average over people. 

$$
\small{
\begin{aligned}
&\sum_{x=9}^{12} P_x \qty{ \hat \mu(x) - \hat\mu(x-1) } =  \sum_x \hat\alpha(x) \hat\mu(x) \qfor
\hat\alpha(x) = \begin{cases}
  P_{12} & \text{ if } x = 12 \\
  P_{x} - P_{x+1} & \text{ if } x \in \{9 \ldots 11\} \\
  -P_9 & \text{ if } x = 8 
\end{cases}    && \text{ \textcolor{gray}{estimates} } \\
&\sum_{x=9}^{12} p_x \qty{ \hat \mu(x) - \hat\mu(x-1) } 
= \sum_{x} \alpha(x) \mu(x) \qfor
\alpha(x) = \begin{cases}
  p_{12} & \text{ if } x = 12 \\\
  p_{x} - p_{x+1} & \text{ if } x \in \{9 \ldots 11\} \\
  -p_9 & \text{ if } x = 8 
\end{cases} && 
\end{aligned}
}
$$

- In these combinations, we're in effect estimating the coefficients $\alpha(x)$.
- This is an additional source of variance. But usually not that much variance. We'll see. 

# Inference for Linear Combinations of Subsample Means: Random Coefficients

## Plan

```{r}
#| layout-ncol: 2
#| out-height: 250px
#| fig-asp: .3
#| fig-subcap:
#|  - Interval Estimate for the Average over Years
#|  - Interval Estimate for the Average over People 


x = seq(-2000,8000, by=10)
point.estimate = (summaries$mean[summaries$education==12] - summaries$mean[summaries$education==8]) / 4
w = 2*1.96*sd(boots$years) 
ggplot() + 
    geom_area(aes(x=x, y=dnorm(x, mean=mean(boots$years), sd=sd(boots$years))), color='purple', fill='purple', alpha=.08) +
    geom_vline(aes(xintercept=mean(boots$years)), color='purple', linetype='solid') +
    geom_vline(aes(xintercept=xi), data=data.frame(xi=qnorm(c(.025,.975), mean=mean(boots$years), sd=sd(boots$years))), color='purple', linetype='dotted') +
    geom_pointrange(aes(y=dnorm(2*sd(boots$years), mean=0, sd=sd(boots$years)), x=point.estimate, xmin=point.estimate - w/2, xmax=point.estimate + w/2)) + 
    xlim(-2000,8000)

x = seq(-2000,8000, by=10)
point.estimate = average.over.people(summaries)
w = 2*1.96*sd(boots$people) 
ggplot() + 
    geom_area(aes(x=x, y=dnorm(x, mean=mean(boots$people), sd=sd(boots$people))), color='purple', fill='purple', alpha=.08) +
    geom_vline(aes(xintercept=mean(boots$people)), color='purple', linetype='solid') +
    geom_vline(aes(xintercept=xi), data=data.frame(xi=qnorm(c(.025,.975), mean=mean(boots$people), sd=sd(boots$years))), color='purple', linetype='dotted') +
    geom_pointrange(aes(y=dnorm(2*sd(boots$people), mean=0, sd=sd(boots$people)), x=point.estimate, xmin=point.estimate - w/2, xmax=point.estimate + w/2)) + 
    xlim(-2000,8000)
```

1. We'll show that our estimator is unbiased. 

$$ 
\E[\hat\theta] = \theta
$$


2. We'll calculate --- and then estimate --- its standard error.
$$
\hat \sigma_{\hat\theta} \approx \sqrt{ \Var[\hat\theta] }
$$

3.  And we can use normal approximation to construct a *confidence interval*.
$$
\theta \in \qty[ \hat \theta - 1.96\hat\sigma_{\hat\theta}, \ \hat \theta + 1.96\hat\sigma_{\hat\theta}] 
$$

- Working out what contributes to the standard error will help us understand why our intervals are as wide as they are.

## Unbiasedness

- Our estimator is unbiased *if the sample coefficients are unbiased estimates of the population coefficients*.
  - Sample proportions $P_x$ are unbiased for the population proportions $p_x$.
  - Differences in Proportions $P_x - P_{x-1}$ are unbiased for the population differences $p_x - p_{x-1}$.
  - That's enough to handle our example and many other cases.
- The proof is almost the same it was with non-random coefficients.
  - But now we use linearity of *conditional expectations*.
  - And *conditional unbiasedness* of the subsample means.


::: {.fragment}
$$
\small{
\begin{aligned}
&\E\qty[\sum_x \hat\alpha(x) \hat\mu(x)]  \\ 
&= \E\qty[ \E\qty{ \sum_x \hat\alpha(x) \hat\mu(x) \mid X_1 \ldots X_n } ] && \text{via law of iterated expectations} \\
&= \E\qty[ \sum_x \hat\alpha(x) \E\qty{ \hat\mu(x) \mid X_1 \ldots X_n } ] && \text{via linearity of conditional expectations} \\
&= \E\qty[ \sum_x \hat\alpha(x) \mu(x) ]                                   && \text{via conditional unbiasedness of the subsample means} \\
&= \sum_x \E[\hat\alpha(x)] \mu(x) = \sum_x \alpha(x) \mu(x)               && \text{via linearity of expectations and unbiasedness of coefficients}  
\end{aligned} 
}
$$
:::




## Variance Calculation Step 1: Total Variance


- We'll use the law of total variance to break the variance of our estimator into two parts.
- One of these parts --- $\Var\E[\ldots]$ part --- has always been zero before.
  - It isn't now. This terms is how the *randomness of the coefficients* contributes.
  - It was zero before because the coefficients were constants.
- What we'll get for the remaining $\E[\Var[\ldots]]$ part is almost exactly what we did before.
  - We *sum* the conditional variances of the individual terms, then take the expectation of the result.
  - All that changes is that, the coefficients being random, we can't pull them out of the expectation

$$
\small{
\begin{aligned}
\Var\qty[ \sum_x \alpha(x) \hat \mu(x) ] 
&= \E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] + \Var\qty[ \E\qty{\sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n} ] \\
&= \E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] + \Var\qty[ \sum_x  \hat\alpha(x) \E\qty{ \hat \mu(x) \mid X_1 \ldots X_n} ] \\
&= \E\qty[ \Var\qty{ \sum_x \hat \alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] + \Var\qty[ \sum_x \hat\alpha(x) \mu(x) ] \\
&= \textcolor{blue}{\sum_x \E\qty[ \hat\alpha(x)^2 \Var\qty{\hat\mu(x) \mid X_1 \ldots X_n}]} \quad + \ldots \\
&=\textcolor{blue}{ \sum_x \sigma^2(x) \times \E\qty[ \frac{\hat\alpha(x)^2}{N_x} ]} \quad + \ldots
\end{aligned}
}
$$

## Variance Calculation Step 2a: $\E[\Var[\ldots]]$ 

$$\small{
\begin{aligned}
\E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] 
&=\E\qty[ \E\qty{ \qty(\sum_x \hat\alpha(x) \hat\mu(x) - \E\qty[\sum_x \alpha(x) \hat\mu(x) \mid X_1 \ldots X_n])^2 \mid X_1 \ldots X_n }] \\
&=\E\qty[ \E\qty{ \qty(\sum_x \hat\alpha(x) \hat\mu(x) - \alpha(x) \E\qty[\hat\mu(x) \mid X_1 \ldots X_n])^2 \mid X_1 \ldots X_n }] \\
&=\E\qty[ \E\qty{ \qty(\sum_x \hat\alpha(x) \qty{\hat\mu(x) - \E\qty[\hat\mu(x) \mid X_1 \ldots X_n]})^2 \mid X_1 \ldots X_n }] \\
&=\E\qty[ \E\qty{ \qty(\sum_x \hat\alpha(x) Z_x)^2  \mid X_1 \ldots X_n }] \quad \text{ for } \quad  Z_x=\hat\mu(x) - \E\qty[\hat\mu(x) \mid X_1 \ldots X_n] 
\end{aligned}
}
$$

- Here $Z_x$ is a *conditionally centered* version of our subsample mean $\hat\mu(x)$.
  - The *conditional variance* of $\hat\mu(x)$ is the conditional expectation of its square $Z_x^2$
  - And the conditional expectation of *products* $Z_x Z_{x'}$ for different subsamples is zero.

$$
\begin{aligned} 
\E[Z_x^2 \mid X_1 \ldots X_n] &= \Var[\hat\mu(x) \mid X_1 \ldots X_n] \\
\E[Z_x Z_{x'} \mid X_1 \ldots X_n] &= 0 \quad \text{ for } \quad x \neq x'
\end{aligned}
$$

## Variance Calculation Step 2b: $\E[\Var[\ldots]]$ and FOIL
- To make progress from here, we have to square out our sum. We get one term for each combination of $x$s.

$$
\small{
\qty(\sum_x \hat\alpha_x Z_x)^2 = \sum_x\sum_{x'} \alpha_x Z_x \ \alpha_{x'} Z_{x'} 
}
$$

- Making this substitution, we can boil things down to something we've calculated before.

$$\small{
\begin{aligned}
\E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] 
&=\E\qty[ \E\qty{ \qty(\sum_x \hat\alpha(x) Z_x)^2  \mid X_1 \ldots X_n }]  \\
&=\E\qty[ \E\qty{ \sum_x\sum_{x'} \hat\alpha(x) \alpha(x') Z_x Z_{x'} \mid X_1 \ldots X_n }] \\ 
&=\E\qty[ \sum_x\sum_{x'} \hat\alpha(x) \alpha(x') \E\qty{ Z_x Z_{x'} \mid X_1 \ldots X_n}] \\ 
&=\E\qty[ \sum_x\sum_{x'} \hat\alpha(x) \alpha(x') \begin{cases} \Var\qty{\hat\mu(x) \mid X_1 \ldots X_n} & \text{ if } x=x' \\ 0 & \text{ otherwise} \end{cases} ] \\
&=\E\qty[ \sum_x \hat\alpha(x)^2 \Var\qty{\hat\mu(x) \mid X_1 \ldots X_n}] \\
\end{aligned}
}
$$

## Concluding Part 2 

```{r}
#| layout-ncol: 2
#| out-height: 300px
#| fig-asp: .3
earnings.jitter
earnings.jitter.pop
```

$$ 
\begin{aligned}
\E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] 
&=  \E\qty[\sum_x  \hat\alpha(x)^2 \Var\qty{\hat\mu(x) \mid X_1 \ldots X_n}] \\
&=  \E\qty[\sum_x  \hat\alpha(x)^2  \frac{\sigma^2(x)}{N_x}] \\
&= \sum_x \sigma^2(x) \times  \E\qty[ \frac{\hat\alpha(x)^2}{N_x} ]
\end{aligned}
$$


## Variance Term Estimation

$$
\small{
\E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] 
\sum_x \sigma^2(x) times \E\qty[ \frac{\hat\alpha(x)^2}{N_x} ] \quad \text{ where } \quad \sigma^2(x) = \Var[Y_i \mid X_i=x]
}
$$

- This has two quantities we don't know.
  - The subpopulation variances $\sigma^2(x)$.
  - The expected coefficient-squared over subsample size $\hat\alpha^2(x) / N_x$.
- But the subsample variances are good esimates of the subpopulation variances
$$
\small{
\sigma^2(x) \approx \hat\sigma^2(x) := \frac{1}{N_x} \sum_{i:X_i=x} \qty{Y_i - \hat\mu(x)}^2
}
$$

- And the coefficient-squared over subsample size is usually a good estimate of its expectation.

$$
\small{
\frac{\hat\alpha(x)^2}{N_x} \approx \E[\frac{\hat\alpha(x)^2}{N_x}] 
}
$$

- Plugging them in gives us an estimate of *the first term* of our summary's variance.

$$\small{
\hat\sigma_{\hat\theta}^2 = \sum_x \frac{ \hat\alpha^2(x) \ \hat\sigma^2(x) }{ N_x } \quad + \quad ?
}
$$


## A Pause to Consider the Implications {.smaller}

```{r}
N = function(x) { sum(earnings$education==x) }
P = function(x) { sum(earnings$education==x) / sum(earnings$education <= 12) }
alphahat = function(x) { if (x==12) { P(12) } else if (x==8) { -P(9) } else { P(x) - P(x+1) } }
sigmahat = function(x) { sd(earnings$income[earnings$education==x]) }
alpha    = function(x) { case_match(x, 
  12 ~  1/4,
  8  ~ -1/4,
  .default = 0) 
}
se.years = sd(boots$years)
se.people.EV = sqrt(sum(sapply(8:12, function(x) { alpha(x)^2 * summaries$sd[summaries$education==x]^2 / sum(earnings$education==x) })))
se.people.EV.terms = sqrt(sapply(8:12, function(x) { alpha(x)^2 * summaries$sd[summaries$education==x]^2 / sum(earnings$education==x) }))
```

$$
\Var\qty[\sum_x \hat\alpha(x) \hat\mu(x)] \ge \E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] = \sum_x \sigma^2(x) \times \E\qty[ \frac{\hat\alpha(x)^2}{N_x} ]
$$

- Let's think about what this means for our two examples.
$$
\small{
\begin{aligned}
\hat\theta_{\text{years}}  &= \frac14\sum_{x=9}^{12} \qty{ \hat\mu(x) - \hat\mu(x-1)} = \sum_x \alpha(x) \hat \mu(x)  \qfor \alpha(x) = \begin{cases} \frac14 & \text{ if } x = 12 \\ -\frac14 & \text{ if } x= 8 \\ 0 & \text{ otherwise} \end{cases} \\
\hat\theta_{\text{people}} &= \sum_{x=9}^{12} P_x \qty{ \hat\mu(x) - \hat\mu(x-1)} = \sum_x \hat \alpha(x) \hat \mu(x) \qfor
\hat\alpha(x) = \begin{cases}
  P_{12} & \text{ if } x = 12 \\
  P_{x} - P_{x+1} & \text{ if } x \in \{9 \ldots 11\} \\
  -P_9 & \text{ if } x = 8 
\end{cases}
\end{aligned}
}
$$

- We've seen that our estimate of the average-over-years summary of the value of a year of high school has large variance.
- The problem was that it paired a relatively large weight--$\alpha(x)=-1/4$---with a very small sample size $N_8=`r sum(earnings$education==8)`$.
- And we know the average-over-people summary has an even larger variance. Let's compare.

## Comparison


```{r}
#| layout-ncol: 2
#| out-height: 250px
#| fig-asp: .3
#| fig-subcap:
#|  - Interval Estimate for the Average over Years
#|  - Interval Estimate for the Average over People 


x = seq(-2000,8000, by=10)
point.estimate = (summaries$mean[summaries$education==12] - summaries$mean[summaries$education==8]) / 4
w = 2*1.96*sd(boots$years) 
ggplot() + 
    geom_area(aes(x=x, y=dnorm(x, mean=mean(boots$years), sd=sd(boots$years))), color='purple', fill='purple', alpha=.08) +
    geom_vline(aes(xintercept=mean(boots$years)), color='purple', linetype='solid') +
    geom_vline(aes(xintercept=xi), data=data.frame(xi=qnorm(c(.025,.975), mean=mean(boots$years), sd=sd(boots$years))), color='purple', linetype='dotted') +
    geom_pointrange(aes(y=dnorm(2*sd(boots$years), mean=0, sd=sd(boots$years)), x=point.estimate, xmin=point.estimate - w/2, xmax=point.estimate + w/2)) + 
    xlim(-2000,8000)

x = seq(-2000,8000, by=10)
point.estimate = average.over.people(summaries)
w = 2*1.96*sd(boots$people) 
ggplot() + 
    geom_area(aes(x=x, y=dnorm(x, mean=mean(boots$people), sd=sd(boots$people))), color='purple', fill='purple', alpha=.08) +
    geom_vline(aes(xintercept=mean(boots$people)), color='purple', linetype='solid') +
    geom_vline(aes(xintercept=xi), data=data.frame(xi=qnorm(c(.025,.975), mean=mean(boots$people), sd=sd(boots$years))), color='purple', linetype='dotted') +
    geom_pointrange(aes(y=dnorm(2*sd(boots$people), mean=0, sd=sd(boots$people)), x=point.estimate, xmin=point.estimate - w/2, xmax=point.estimate + w/2)) + 
    xlim(-2000,8000)
```

$$ 
\small{
\begin{array}{c|ccccc}
x & 8 & 9 & 10 & 11 & 12 \\
N_x & `r N(8)` & `r N(9)` & `r N(10)` & `r N(11)` & `r N(12)` \\
\hat\sigma^2(x) & `r approx(sigmahat(8)^2)` & `r approx(sigmahat(9)^2)` & `r approx(sigmahat(10)^2)` & `r approx(sigmahat(11)^2)` & `r approx(sigmahat(12)^2)`  \\
 & & & & & \\
\hat\alpha^2(x) & `r approx(alphahat(8)^2)` & `r approx(alphahat(9)^2)` & `r approx(alphahat(10)^2)` & `r approx(alphahat(11)^2)` & `r approx(alphahat(12)^2)` \\
\frac{\hat\sigma^2(x)\alpha^2(x)}{N_x} & `r approx(sigmahat(8)^2*alphahat(8)^2/N(8))` & `r approx(sigmahat(9)^2*alphahat(9)^2/N(9))` & `r approx(sigmahat(10)^2*alphahat(10)^2/N(10))` & `r approx(sigmahat(11)^2*alphahat(11)^2/N(11))` & `r approx(sigmahat(12)^2*alphahat(12)^2/N(12))` \\`
& & & & & \\
\alpha^2(x) & `r approx(alpha(8)^2)` & `r approx(alpha(9)^2)` & `r approx(alpha(10)^2)` & `r approx(alpha(11)^2)` & `r approx(alpha(12)^2)` \\
\frac{\hat\sigma^2(x) \alpha^2(x)}{N_x} & `r approx(sigmahat(8)^2*alpha(8)^2/N(8))` & `r approx(sigmahat(9)^2*alpha(9)^2/N(9))` & `r approx(sigmahat(10)^2*alpha(10)^2/N(10))` & `r approx(sigmahat(11)^2*alpha(11)^2/N(11))` & `r approx(sigmahat(12)^2*alpha(12)^2/N(12))` \\`
\end{array}
}
$$

- Much of its variance comes from putting *much more* weight on the mean of a small, but less small, sample: the one with $x=11$.
- This variance term, $\sigma^2(11) \times \E[\frac{\hat\alpha(11)^2}{N_{11}}] \approx \sigma^2(11) \times `r round(alpha(11)^2/N(11),3)`$,
  is bigger than the term $\sigma^2(8) \times (1/4)^2/N_8 \approx \sigma^2(8) \times `r round((1/16)/N(8),3)`$ we worried about before.
- Even based on this *part* of the variance, we can see that this summary will have a bigger variance than the average-over-years one.
- This part of the former's variance is $`r approx(se.people.EV)`^2$ vs. the latter's whole variance $`r approx(se.years)`^2$. Let's get a sense of the other part.



## Variance Calculation Step 3: The $\Var[\E[\ldots]]$ Part.

- Now we'll calculate the new part --- the $\Var[\E[\ldots]]$ part. Roughly speaking.
- This **additional variance** is what we pay for using *estimates* of the population coefficients $\alpha(x)$.
- We're going to use a formula for the variance of a linear combination of random variables that you may have seen before.
$$
\Var\qty[\sum_x a_x Z_x] = \sum_x \sum_x' a_x a_x' \Cov(Z_x,\ Z_y) \qfor \Cov(Z_x,\ Z_y) = \E\qty[ \qty{Z_x - \E[Z_x]} \qty{Z_y - \E[Z_y]} ]
$$
- It's easy enough to prove, but we'll skip that. Here's what it tells us.
$$
\Var\qty[ \E{\sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n} ] 
= \Var\qty[ \sum_x \mu(x) \hat \alpha(x)  ] \\
= \sum_x \sum_x' \mu(x) \mu(x') \Cov\qty[\hat\alpha(x),\ \hat\alpha(x')] 
$$

- Calculating this is do-able but pretty involved. We'll skip the details.

## Not Details

$$
\Var\qty[ \E{\sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n} ] 
= \Var\qty[ \sum_x \mu(x) \hat \alpha(x)  ] \\
= \sum_x \sum_x' \mu(x) \mu(x') \Cov\qty[\hat\alpha(x),\ \hat\alpha(x')] 
$$

- We'll just do enough to get a handle on the size of this additional variance is. 
- We'll do this by calculating one of these covariance factors---we'll do $\hat\alpha(9)$ and itself.
$$
\Cov(\hat\alpha(9), \alpha(9)) = \Cov(P_9-P_{10},\ P_{9}-P_{10}) = \Var(P_9) - 2\Cov(P_9,\ P_{10}) + \Var(P_{10}).
$$
- In short, it --- and presumably the other terms --- can't be much bigger than the variance of the individual sample proportions $P_{x}$.
- And those are roughly $p_x (1-p_x) / \sum_x N_x$, i.e., the variance of an average of Bernoulli random variables with probability $p_x$.
  - What's important is that *the denominator here is the sum of the sample sizes*. 
  - It's never small like the denominators in the other term can be.
  - As a result, this additional variance term is *often much smaller* than the other one.
