---
title: "Activity"
subtitle: "Cross Validation Worksheet"
---

## 

Consider the following data with three observations.

```{r}
#| layout-ncol: 2
x <- c(1,2,3)
y <- c(1,3,1)
plot(x,y,pch=19, xlim=c(0,4), ylim=c(0,4))
```
Our goal is to use leave one out cross-validation to choose between a linear regression model and a constant model for this data. Below we have plotted these data three times. The hollow point is the data that is "left out" for the cross-validation, which we will use as our test data. 

- Draw a constant model (i.e., flat line) for each of the three plots based on the non-left-out data.

- Draw and calculate the residuals from the constant models to the left-out observation for each of the three plots.

- Draw a linear regression line (i.e., line that doesn't have to be flat) for each of the three plots based on the non-left-out data.

- Draw and calculate the residuals from the linear models to the left-out observation for each of the three plots.

##

```{r}
#| layout-ncol: 2
x <- c(1,2,3)
y <- c(1,3,1)
#par(mfrow=c(1,3))
plot(x,y,pch=c(1,19,19), xlim=c(0,4), ylim=c(0,4), main="A")
plot(x,y,pch=c(19,1,19), xlim=c(0,4), ylim=c(0,4), main="B")
plot(x,y,pch=c(19,19,1), xlim=c(0,4), ylim=c(0,4), main="C")
```


$$
\begin{array}{|c|p{3cm}|p{3cm}|p{3cm}|}
\hline
& A & B & C \\ 
\hline
 & & & \\
\text{Constant model  } \hat{y}_{test} & & & \\
 & & & \\
\hline 
 & & & \\
\text{Constant model  } y_{test} - \hat{y}_{test} \text{  (residual)} & & &  \\ 
 & & & \\
\hline 
 & & & \\
\text{Linear model  } \hat{y}_{test} & & &  \\
 & & & \\
\hline 
 & & & \\
\text{Linear model  } y_{test} - \hat{y}_{test} \text{  (residual)} & & & \\ 
 & & & \\
\hline
\end{array}
$$

Based on the residuals you calculated, which model is preferred? The constant model or the linear regression model? 