---
title: "Probability Background: Expectations"
---

## Setup

{{< include ../_extensions/r-wasm/live/_knitr.qmd >}}
{{< include shared-code.qmd >}}


```{webr}
#| label: ggplot-theme
#| include: false

lightgray = "#bbbbbb"
gridcolor = rgb(0,0,0,.1, maxColorValue=1)

lab.theme = theme(
        plot.background = element_rect(fill = "transparent", colour = NA),
        panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x  = element_text(colour = lightgray),
        axis.text.y  = element_text(colour = lightgray),
        axis.title.x  = element_blank(),
        axis.title.y  = element_blank(),
        panel.grid.major=element_line(colour=gridcolor),
        panel.grid.minor=element_line(colour=gridcolor, linewidth=0.25))

theme_set(lab.theme)
set.seed(1)
```

```{webr}

n = 100
m = 200
theta = 7/10

height.binom = max(dbinom(0:n, n, theta))
height.hyper = max(dhyper(0:n, m*theta, m*(1-theta), n))

S = 1000
probs = data.frame(x = (0:n)/n,
                   binom = dbinom(0:n, n, theta),
                   hyper = dhyper(0:n, m*theta, m*(1-theta), n))
samples = data.frame(binom = rbinom(S, n, theta)/n,
                     hyper = rhyper(S, m*theta, m*(1-theta), n)/n,
                     sample = (1:S)/S)
height = max(probs$hyper)

twodist.plot = ggplot() +
  geom_area(aes(x=x, y=prob, color=dist, fill=dist), alpha=.3, position="identity", 
    data=probs |> pivot_longer(cols=c(binom, hyper), names_to="dist", values_to="prob")) +
  geom_vline(xintercept = theta, color='green', alpha=.7, linewidth=1.5) +
  geom_point(aes(x=x, y=sample*height), color='purple',alpha=.2, size=.2, 
    data=samples |> pivot_longer(cols=c(binom, hyper), names_to="dist", values_to="x")) +
  scale_x_continuous(breaks=seq(0,1,by=.1), limits=c(.5,.9)) + 
 guides(color='none', fill='none') + labs(y="", x="")
```

## Comparing Distributions

```{webr}
#| warning: false
twodist.plot + facet_wrap(~dist)
```

- You're looking at the sampling distribution of the sample frequency, $\hat\theta$ when we've sampled 100 observations ...
- ... [with]{.groupa} and [without]{.groupb} replacement from a binary population of size 200 with [frequency $\theta = 7/10$]{.target}.
- **Q**. What do these distributions have in common? How do they differ? 

::: fragment

- What they have in common:
  1. Their essential shape. They're both 'single bumps'. 
  2. Their location. They're both centered on the population frequency $\theta$.
  
- What's different is their *spread*. The [binomial distribution]{.groupa} is wider than the [hypergeometric distribution]{.groupb}.

:::

## Expectation and Location

::: {.panel-tabset}
### Single Bumps 
```{webr}
#| warning: false
twodist.plot + facet_wrap(~dist)
```

### Flips
```{webr}
#| warning: false
flips = rbinom(S, 1, .7)
flips.plot =ggplot() + geom_point(aes(x=flips, y=1:S/S), color='purple', alpha=.2, size=.2) + 
           geom_bar(aes(x=flips, y=after_stat(prop)), alpha=.2) +
           geom_vline(xintercept = .7, color='green', alpha=.7, linewidth=1.5)
flips.plot
```

### Multiple Bumps 
```{webr}
#| warning: false
flips = rbinom(S, 1, .5)
B1 = rbinom(S, n, .3)/n
B2 = rbinom(S, n, .7)/n
B = ifelse(flips==1, B1, B2)
bumps.plot = ggplot() + geom_point(aes(x=B, y=(height/2)*(1:S/S)), color='purple', alpha=.2, size=.2) + 
           geom_bar(aes(x=B, y=after_stat(prop)),  alpha=.2) +
           geom_vline(xintercept = .5, color='green', alpha=.7, linewidth=1.5)
bumps.plot
```
:::


::: {.panel-tabset}
### Defining Expectation 
- The *expectation* of a random variable is its *probability-weighted average*.
  
$$
\E[X] = \sum_x \inpink{x}\inteal{P(X=x)}
$$

- That means summing, over the x axis, the product of ...
  - ... the [$x$, the position]{.pink} along the x-axis 
  - ... the [$P(X=x)$, the *bar height*]{.teal} at that position.

- Does that tell us where distribution is *located*, i.e., where draws are likely to be?
- If not, why is that useful?

### Quick Practice

- What is the expectation of ...
  1. $X$, a binary random variable that's one with probability $\theta$  
  2. $X+X=2X$
  3. $2X-1$
  3. $X + Y$ where $Y$ has the same distribution as $X$
  4. $X^2$
  5. $XY$ 

:::


## Spread

::: {.panel-tabset}
### Single Bumps 
```{webr}
#| warning: false
twodist.plot + facet_wrap(~dist)
```

### Flips
```{webr}
#| warning: false
flips.plot
```

### Multiple Bumps 
```{webr}
#| warning: false
bumps.plot
```
:::

- To summarize the spread of a distribution, we use the expectation of ...
  - some kind of *distance* 
  - from the center.
- The most common measure is the *variance*, which is the expectation of the squared distance from the mean.
  - The *standard deviation*, the square root of this, is the variance translated back into 'not-squared units'.
  - Other measures like this might be the *median absolute difference* from *the median*. 

::: {.panel-tabset}
### Definition
$$
\begin{aligned}
\Var[X] &= \E\qty[ (X - \E X)^2 ] \\
\sd[X]  &= \sqrt{\Var[X]}
\end{aligned}
$$

  
### Quick Practice
- What is the variance of ...
  1. $X$, a binary random variable that's one with probability $\theta$  
  2. $2X$
  3. $2X-1$
  3. $X + Y$ where $Y$ has the same distribution as $X$

:::

## Sampling, Means, and Variances

:::: columns
::: column
```{webr}
#| warning: false
twodist.plot + facet_wrap(~dist)
```
:::
::: column
```{webr}
#| warning: false
flips.plot 
```
:::
::::

:::: {.panel-tabset}
### Why Do We Care About These?

- Why do we care about the expectation and variance of *a random variable*?
- Criticism 1. The mean often isn't a great summary of location for a random variable.
  - Variance describes deviation from the mean. If the mean is bad, so is the variance.
- Criticism 2. Why do we care about a random variable anyway? We care about the *population* it's drawn from.
- Any comments in support of expectations and variances? Maybe responses to these criticisms?

### Re: Criticism 2

- If a random variable $Y$ is *equally likely* to be any member of a population $y_1 \ldots y_m$, ...
- ... then its expectation is the population mean.

$$
\E[Y] = \sum_{j=1}^m \underset{\text{response}}{y_j} \times \underset{\text{probability}}{\frac{1}{m}} = \frac{1}{m}\sum_{j=1}^m y_j = \bar y
$$

- Usually, we use sampling schemes like this. Even weird ones tend to have this property.
  - Sampling without replacement.
  - Sampling with replacement.
  - Randomized 'Circular' Convenience Sampling.
    - Line up your population and roll a die to decide who to survey first. 
    - Then survey them and the $n-1$ people to their right.
    - 'Wrap around', starting from the other side, if you run out of people.

### Re: Criticism 1

- It's very reasonable to say that it's nonsense to refer to any one location for a coin flip or a pair of bumps.
- But if you average a lot of coin flips, or a lot of independent draws from a pair of bumps...
  - what you'll get will start to look like a single bump.^[This is what the **central limit theorem** says.]
  - that bump will be centered on the expectation.^[This is due to the *linearity of expectation*.]
  - and it'll be *concentrated* around it---most draws will be close to that expectation.^[This is what the *law of large numbers* says.]

:::


# Properties of Expectations
These are repeated in the next lecture.

## Linearity of Expectations

::: {.panel-tabset}
### Linearity
$$ 
\begin{aligned}
E[ a Y + b Z ] 
&= E[aY] + E[bZ] \\
&= aE[Y] + bE[Z]  \\
& \text{ for random variables $Y, Z$ and numbers $a,b$ }
\end{aligned}
$$

  - There are two things going on here.
    - To average a sum of two things, we can take two averages and sum.
    - To average a constant times a random variable, we can multiply the random variable's average by the constant.
  - In other words, we can *distribute* expectations and can pull constants out of them.
  
### Proof

- In essence, it comes down to the fact that all we're doing is summing.
	- Expectations are probability-weighted sums.
	- And we're looking at the expectation of a sum.
- And we can change the order we sum in without changing what we get.

$$
\small{
\begin{aligned}
\E\qty[ a Y + b Z ]
&= \sum_{y}\sum_z (a y + b z) \ P(Y=y, Z=z) && \text{ by definition of expectation} \\
&= \sum_{y}\sum_z a y \ P(Y=y, Z=z) + \sum_{z}\sum_y b z \ P(Y=y, Z=z) && \text{changing the order in which we sum} \\
&= \sum_{y} a y \ \sum_z P(Y=y,Z=z) + \sum_{z} b z \ \sum_y P(Y=y,Z=z) && \text{pulling constants out of the inner sums} \\
&= \sum_{y} a y \ P(Y=y) + \sum_{z} b z \ P(Z=z)  && \text{summing to get marginal probabilities from our joint } \\
&= a\sum_{y} y \ P(Y=y) + b\sum_{z} z \ P(Z=z)  && \text{ pulling constants out of the remaining sum } \\
&= a\E Y + b \E Z  && \text{by definition}
\end{aligned}
}
$$

:::

## Factorization of Products of *Independent* Random Variables

The expectation of a product of **independent** random variables is the product of their expectations.
$$
\E[YZ] = \E[Y]\E[Z] \qqtext{when $Y$ and $Z$ are independent}
$$

This comes up a lot in variance calculations. Let's prove it. It'll be good practice.

::: fragment

$$
\begin{aligned}
\E[YZ] &= \sum_{yz} yz \ P(Y=y, Z=z) && \text{by definition of expectation} \\
&= \sum_y \sum_z yz \ P(Y=y) P(Z=z) && \text{factoring and ordering sums } \\
&= \inteal{\sum_y y \ P(Y=y)} \inpink{\sum_z z \ P(Z=z)} && \text{pulling factors that don't depend on $z$ out of the inner sum} \\
&  \inteal{\E[Y]} \inpink{\E[Z]} && \text{by definition of expectation}
\end{aligned}
$$
:::


# Implications for Sample Means

## Bias

```{r}
xx = seq(-6,6,by=.01)
K = 100
heights = seq(0,.4,length.out=K)
normals = rnorm(10000)

unbiased.plot = ggplot() + 
  geom_histogram(aes(x=normals, y=after_stat(density)), alpha=.2, bins=100) + 
  with_annotations(normals, estimand=0, dots=100, highlight.dots=0, interval.width=1.96) + xlim(min(xx),max(xx))

biased.plot = ggplot() + 
  geom_histogram(aes(x=normals+2, y=after_stat(density)), alpha=.2, bins=100) + 
  with_annotations(normals+2, estimand=0, dots=100, highlight.dots=0, interval.width=1.96) + xlim(min(xx),max(xx))
```

```{r}
#| layout-ncol: 2
unbiased.plot
biased.plot
```

::: {.panel-tabset}
### What Bias Is
- The **bias** of an estimator is the difference between its expected value and the value of the thing it's estimating.
- We haven't talked about this yet because there hasn't been any. We've been using **unbiased estimators**.
- You can see, looking at the sampling distribution on the right, that bias might cause some problems with coverage.
- If we calibrate interval estimates to cover **the estimator's mean** 95% of the time
  - and that's what we've really been doing
  how often will they cover the [thing we're estimating]{.fg style="color:green"}?
- This means biased estimators can cause big problems. We'll talk about this more later in the semester.
- But for now, let's just show the estimator we've been using is, in fact, unbiased.
  
### Unbiasedness of the Sample Mean

**Claim.** The sample mean is an unbiased estimator of the population mean when we use sampling with replacement, sampling without, randomized circular sampling, etc. 
$$ \E[\hat\mu] = \mu \qqtext{ for} \hat\mu = \frac1n\sum_{i=1}^n Y_i  \qand \mu = \frac1m\sum_{j=1}^m y_j $$

**Proof.**

::: fragment

$$
\begin{aligned}
\E\qty[\frac1n\sum_{i=1}^n Y_i] 
&= \frac1n\sum_{i=1}^n \E[Y_i] && \text{ via linearity } \\
&= \frac1n\sum_{i=1}^n \frac{1}{m}\sum_{j=1}^m y_j \times \frac{1}{m} && \text{ via equal-probability sampling } \\
&= \frac1{n} \mu &&\text{ by definition }
&= \frac1n \times n \times \mu = \mu.
\end{aligned}
$$

:::
:::

