---
title: "Probability Review: Conditional Expectations"
---

## A New Dataset

{{< include shared-code.qmd >}}
{{< include ca-income.qmd >}}

:::: columns

::: column
```{r map-sample}
#| fig-width: 5
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=earnings, size=.1, alpha=.5) + guides(color='none')
```
:::

::: column

```{r table}
style.tab = function(tab, cols=c('income', 'education', 'county'), nhead=3, ntail=1) {
  tab$income50k = 0+(tab$income > 50000)
  tab$income = sprintf('$%dk', round(as.numeric(tab$income) / 1000))
  tab$county = ifelse(tab$county=='los angeles', 'LA',  tab$county)
  ellipsis.tab = tab[, cols] %>% ellipsis(nhead=nhead,ntail=ntail)
  ellipsis.tab %>%
    kable %>% kable_styling() %>% 
    row_spec(which(ellipsis.tab$education >= 16),  background=twocolor.green) %>%
    row_spec(which(ellipsis.tab$education <  16),  background=twocolor.red) 
}

style.tab(sample.tab, nhead=4)
```
:::

::::

- We'll do the same thing we've been doing, but for **non-binary outcomes**.
- Using data from the Current Population Survey, we'll estimate ...
  - the mean income in our population.
  - the mean income among people in our population [with 4-year degrees]{.twocolor-green} 
  - the difference in mean income between people [with 4-year degrees]{.twocolor-green} and people [without them]{.twocolor-red}.
- Our population will be the set of California residents 
    - between the ages of 25 and 35
    - with at least an 8th-grade education


::: aside
The locations shown on the map are made up. They're not the actual locations of the people in the sample. \
The survey includes some location information for some people, but as you can see in the table, not for everyone.
:::


## Imagining a Population

```{r}
#| fig-width: 5 
#| layout-ncol: 2 
#| 
ca.plot + 
 geom_point(aes(x=long, y=lat, color=degree), data=pop.earnings, size=.02, alpha=.1) + 
 guides(color='none')
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=earnings, size=.1, alpha=.5) +
  guides(color='none')
```

:::: columns
::: column
```{r}
style.tab(pop.tab)
```
:::
::: column
```{r}
style.tab(sample.tab)
```
:::
::::

- We don't have much information about the people in our population. Except the ones in our sample. 
- But, for the sake of visualization, I've made some up. 
  - We'll put visualizations of this fake population on the left.
  - Visualizations of the (real) sample will be on the right.



## Sampling

```{r pop-sample-maps}
#| layout-ncol: 2 
#| fig-width: 7
ca.plot + 
 geom_point(aes(x=long, y=lat, color=degree), data=display.population, size=.02, alpha=.1) + 
 guides(color='none')
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=our.sample, size=.1, alpha=.5) 
```

- We'll act as if our data were sampled, with replacement, from this population.
  - On the left, I've illustrated the population.
  - On the right, I've illustrated our sample.
- Our working assumption: each dot on the right was chosen, from those on the left, by rolling a big die.
- That's not quite right, but we'll stick with it throughout the semester.

## Sample and Population Means

```{r}
#| fig-width: 7
#| layout-ncol: 2

ggplot(pop.earnings) + 
  geom_point(aes(x=0, y=income),  color='blue',
            size=.2, position=position_jitter(seed=0,w=.3), alpha=.05) +
  stat_summary(aes(x=0, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x))  + 
  scale_y() 

ggplot(earnings) + 
  geom_point(aes(x=0, y=income), color='blue', 
            size=2, position=position_jitter(seed=0,w=.3), alpha=.2) +
  stat_summary(aes(x=0, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x))  +
  scale_y() 
```

::: {.panel-tabset}
### Estimator and Target
- One thing we often want to estimate is the mean of our population.

$$
\mu = \frac{1}{m}\sum_{j=1}^m y_j 
$$

- We'll think of the mean in our sample as an estimate of it.

$$ 
\hat \mu = \frac{1}{n}\sum_{i=1}^n Y_i 
$$

### What We Know

- We have shown last time that this estimator is *unbiased*.

$$
\E[\hat \mu] = \mu \qfor \underset{\text{\faded{population mean}}}{\mu = \E[Y_i] = \frac{1}{m}\sum_{j=1}^m y_j}
$$

- We have almost calculated its standard deviation. Here it is.

$$
\sd[\hat \mu] = \frac{\sigma}{\sqrt{n}}  \qfor \underset{\text{\faded{population standard deviation}}}{\sigma=\sqrt{\Var[Y_i]}=\sqrt{\frac{1}{m}\sum_{j=1}^m (y_j - \mu)^2}}
$$

  - We did it in the special case of binary $Y_i$ in our last lecture.
  - We'll generalize next time. It's a one-line change.

:::

## Why Do We Care About Unbiasedness?

```{r}
xx = seq(-6,6,by=.01)
K = 100
heights = seq(0,.4,length.out=K)
normals = rnorm(10000)

unbiased.plot = ggplot() + geom_histogram(aes(x=normals, y=after_stat(density)), alpha=.3, bins=100) +
           geom_vline(xintercept=0, color='green', linewidth=3, alpha=.4) + 
           geom_vline(xintercept=0, color='blue') + 
           geom_vline(xintercept=c(-1,1), color='blue', linetype='dashed') +
           geom_vline(xintercept=c(-2,2), color='blue', linetype='dotted') +
           geom_pointrange(aes(y=heights, x=normals[1:K], xmin=normals[1:K]-1.96, xmax=normals[1:K]+1.96), 
                           color='purple', alpha=.15) + xlim(min(xx),max(xx))
           
biased.plot = ggplot() + geom_histogram(aes(x=normals+2, y=after_stat(density)), alpha=.3, bins=100) +
           geom_vline(xintercept=0, color='green', linewidth=3, alpha=.4) + 
           geom_vline(xintercept=2+0, color='blue') + 
           geom_vline(xintercept=2+c(-1,1), color='blue', linetype='dashed') +
           geom_vline(xintercept=2+c(-2,2), color='blue', linetype='dotted') +
           geom_pointrange(aes(y=heights, x=2+normals[1:K], xmin=2+normals[1:K]-1.96, xmax=2+normals[1:K]+1.96), 
                           color='purple', alpha=.15) + xlim(min(xx),max(xx))


```

```{r}
#| layout-ncol: 2
#| fig-width: 7   
unbiased.plot
biased.plot
```

::: {.panel-tabset}
### Unbiased and Biased Estimators
- *Unbiasedness* means that our sampling distribution is centered at our [estimation target]{.target}.
- On the left, we see the sampling distribution of an unbiased estimator. 
  - When it looks like that, we're in good shape.
- On the right, we see one for an estimator with substantial bias.
  - When it looks like that, we're in trouble.
  - You can see that a good number of our samples on the right are far off-target. 
  - Further away, e.g., than the width of the distribution.
  - You can see this might cause problems with coverage.

### Calibration
- If we calibrate interval estimates to cover [**the estimator's mean**]{.blue} 95% of the time, \
how often will they cover the [thing we actually intend to estimate]{.target}?
- Give me a rough estimate for the picture on the right. Is it about 90%? 80%? 50%? 20%?
- We'll see our first examples of biased estimators in this week's homework. 
- And we'll have to start thinking about this relationship between bias and coverage.

:::

## Subsample and Subpopulation Means

```{r}
#| layout-ncol: 2 
#| fig-width: 7

earnings.jitter.pop = ggplot(pop.earnings) + 
    geom_point(aes(x=degree, y=income, color=degree), 
               size=.2, position=position_jitter(seed=0,w=.3), alpha=.1) +
     scale_y() + guides(color='none')

earnings.jitter = ggplot(earnings) + 
    geom_point(aes(x=degree, y=income, color=degree), 
               size=dot.size, position=position_jitter(seed=0,w=.3), alpha=.1) +
     scale_y() + guides(color='none')

earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

- We'll often be interested in the mean income in subpopulations, too.
  - We'll think about the subsample with $X_i=x$ for some value $x$. 
  - e.g. $X_i=1$ for people [with a 4-year degree]{.twocolor-green} and $X_i=0$ for people [without]{.twocolor-red}.
$$
\mu(x) = \frac{1}{m_x}\sum_{j:x_j = x } y_j \quad \text{ where } \quad m_x = \sum_{j:x_j=x} 1.
$$

- We'll use the mean in the corresponding subsample to estimate it.

$$ 
\hat \mu(x) = \frac{1}{N_x}\sum_{i:X_i=x} Y_i \quad \text{ where } \quad N_x = \sum_{i:X_i=x} 1.
$$

- If we want to know a *difference of subpopulation means*, as we often do ...
- ... we can estimate it using a difference of subsample means. 

## Are These Good Estimators?

```{r}
#| layout-ncol: 2
#| fig-width: 7
unbiased.plot
biased.plot
```

- We want to know about bias, i.e. location of the sampling distribution relative to the [estimation target]{.target}.
  - That's where we're headed today. We'll show they're unbiased.
- And we want to know about precision, i.e., how wide our calibrated intervals would be.
  - We'll get to this next time.


# Random Variables and Conditioning
A lot of this is review.

::: footnotestyle
To make this a nice cohesive read, I've included all the unconditional probability stuff we've covered so far *and* the new conditional stuff.
:::


## Observations as Random Variables

```{r}
#| layout-ncol: 2 
#| fig-width: 7

summaries = earnings %>% group_by(degree) %>% summarize(mean=mean(income), sd=sd(income))

earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

- The notation $y_j$ refers to a number.
  - It's the income of the $j$th person in our population. That's just some person. 
- The notation $Y_i$ refers to something else.
  - It's the income of the $i$th person we call in our survey. That's not a person.
  - That's the result of a random process---the roll of a die.
- To summarize this result, we talk about the **probability distribution** of $Y_i$.

### Notation Conventions {#sec-notation}

- Random variables are written in **uppercase**: $X$, $Y$, $Z$, etc. Constants are written in **lowercase**: $x$, $y$, $z$, etc.
- We'll use the same letter for a random variable and the value it takes on. $x$ is a possible value of $X$, $y$ of $Y$, etc.
- Estimators are also random variables, but instead of uppercase we use a **hat**: $\hat\theta$,$\hat \mu$, $\hat \sigma$, etc.
- We'll use the same letter for the estimator and what it's meant to estimate. 
  - $\hat \mu$ is an estimator of $\mu$; $\hat\theta$ is an estimator of $\theta$.
- We'll use $\theta$ and $\mu$ in different but ocassionally overlapping ways.
  - $\mu$ will be our population mean and $\mu(x)$ the mean in the subpopulation with $X=x$.
  - $\theta$ will be our estimation target. 
    - So far, it's often been the population mean, so $\theta=\mu$.
    - Or a subpopulation mean, so $\theta=\mu(x)$.
    - Later in the semseter, it'll tend to be something a bit more complicated. 





## Probability Distributions: The Binary Case

```{r}
cols = c('income50k', 'income', 'education', 'county')
sample.tab$roll = sample.tab$J
```

:::: columns
::: column
```{r}
style.tab(pop.tab, cols=cols)
```
:::
::: column
```{r}
style.tab(sample.tab, cols=c('roll', cols))
```
:::
::::

- When our outcomes are binary, it's easy to describe this distribution.
  - All we need to know is the probability that $Y_i$ is 1. Why?
  - To do that, we **sum the probabilities** of the rolls that result in it being one.

$$
P(Y_i = 1) = \sum_{j: y_j = 1} P(\text{roll}_i = j) = \sum_{j: y_j = 1} \frac{1}{m} \times y_j = \mu.
$$

- This **collapses** out information about our random process that's irrelevant to $Y_i$.
- That is, it collapses the probability we roll each number in 1...m into our 'weighted coin flip'.


## Probability Distributions: The General Case

:::: columns
::: column
```{r}
style.tab(pop.tab, cols=cols)
```
:::
::: column
```{r}
style.tab(sample.tab, cols=c('roll', cols))
```
:::
::::


- When our outcomes are are nonbinary, it's a bit more complicated.
  - We need to know the probability that $Y_i$ takes on each possible value.But we calculate it the same way. 
  - We sum the probabilities of the rolls that result in it taking on those values.

$$
P(Y_i = y) = \sum_{j: y_j = y} P(\text{roll}_i = j) 
$$

- We're still collapsing out irrelevant information, but what we're left with is more complicated.
- It's a **weighted die roll**, with one face for each possible value of $Y_i$. 
- And If each person's income is different, then there's nothing to collapse out. Then, ...
$$ 
P(Y_i = y) = \begin{cases} \frac{1}{m} & \qqtext{for} y \in y_1 \ldots y_m \\ 
                            0 & \qqtext{otherwise}
             \end{cases}
$$

## Expectations

:::: columns
::: column
```{r}
style.tab(pop.tab, cols=cols)
```
:::
::: column
```{r}
style.tab(sample.tab, cols=c('roll', cols))
```
:::
::::


- That said, we often don't need to know the whole distribution.
- Often the only thing we care about is the expected value of $Y_i$. Or some related quantity.
  - The expected value of $Y_i$ is the **probability-weighted average** of the values it can take on.
  - What's nice about this is that we can think of this in 'uncollapsed' terms.
- When we sample as usual, this is just the population mean.
- The collapsed form is, in a sense, just summing in a different order.

$$
\begin{aligned}
\E[Y_i] = \sum_y P(Y_i = y) \times y
= \sum_y \qty(\sum_{j: y_j = y} \frac{1}{m}) \times y 
= \frac{1}{m}\sum_y\sum_{j:y_j = y} y 
= \frac{1}{m}\sum_{j=1}^m y_j
\end{aligned}
$$

## Expectations

:::: columns
::: column
```{r}
style.tab(pop.tab, cols=cols)
```
:::
::: column
```{r}
style.tab(sample.tab, cols=c('roll', cols))
```
:::
::::


- What's nice about the binary case is, in fact, that it's an expectation.
- When $Y_i$ is binary, the probability it's 1 is its expected value $Y_i$.

$$
\begin{aligned}
P(Y_i=1) = \sum_{j: y_j = 1} P(\text{roll}_i = j) 
         &= \sum_{j:y_j=1} \frac{1}{m} \\
         &= \sum_{j=1}^m \frac{1}{m} \times \begin{cases} 1 & \ \ \text{ if } y_j = 1 \\ 0 & \ \ \text{ otherwise} \\ \end{cases} \\
         &= \sum_{j=1}^m \frac{1}{m} \times y_j = \E[Y_i].
\end{aligned}
$$

- In fact, we like expectations so much that we often use them to work with probabilities.

$$
P(Z \in A) = \E[1_A(Z)] \qfor 1_A(z) = 
\begin{cases} 
1 & \qqtext{ for } z \in A \\ 0 & \qqtext{ otherwise} 
\end{cases}
$$


## Independence 


```{r}
#| layout-ncol: 2 
#| fig-width: 7

summaries = earnings %>% group_by(degree) %>% summarize(mean=mean(income), sd=sd(income))

earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

- Random variables are **independent** if ...
  - ... in intuitive terms, knowing the value of one doesn't tell us anything about the value of the other.
  - ... in mathematical terms, their joint probability distribution is the product of their individual marginals ones.

$$
P(Y_1 ... Y_k  = y_1 \ldots y_k) = P(Y_1=y_1) \times \ldots \times  P(Y_k=y_k).
$$

- That's what happens when the randomness in each $Y_i$ comes from a different roll of the die.
  - And since that's how we're doing our sampling in the Current Population Survey, it'll be true in our sample.
  - Because we draw each of our observations the same way, they also have the same probability distribution. 
  - We say they're **independent and indentically distributed.**
- At least, that's what we're pretending when we analyze CPS data in this class. Reality is more complicated.

## Conditioning

```{r}
#| layout-ncol: 2 
#| fig-width: 7

earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

- **Conditioning** is, in effect, a way of thinking about sampling as a two-stage process.
  - First, we choose the color of our dot, i.e., the value of $X_i$, according to it frequency in the population.
  - Then, we choose a specific one of those dots, i.e. $J_i$, from those with that color---with equal probability.
- Because this is just a way of thinking, each person still gets chosen with probability $1/m$.
$$
P(J_i=j) = 
\begin{cases}
   \frac{m_{green}}{m} \ \ \ \ \ \times \frac{1}{m_{green}} \ \ \ \ \ =\ \frac{1}{m} & \text{if the $j$th dot is green  ($x_j=1$) } \\
    \frac{m-m_{green}}{m} \times \frac{1}{m-m_{green}} \ = \ \frac{1}{m} & \text{otherwise}
\end{cases}
$$
  


## Conditioning

```{r}
#| layout-ncol: 2 
#| fig-width: 7

earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```


- The **Conditional Probability** of $Y_i$ is the probability resulting from the second stage.
- It's a function of the result of the first. 
  - $P(Y_i=y \mid X_i=1)$ is the probability distribution of $Y_i$ when we're rolling the 'green die'.
  - $P(Y_i=y \mid X_i=0)$ is the probability distribution of $Y_i$ when we're rolling the 'red die'.
- And the **Conditional Expectation** of $Y_i$ is the 'second stage expected value' in the same sense.
  - $\E[Y_i \mid X_i=1]$ is the expected value of $Y_i$ when we're rolling the 'green die'. 
  - That is, the mean value $\mu(1)$ of $y_j$ in the subpopulation drawn as little green dots.
  - $\E[Y_i \mid X_i=0]$ is the expected value of $Y_i$ when we're rolling the 'red die'.
  - That is, the mean value $\mu(0)$ of $y_j$ in the subpopulation drawn as little red dots.


# Working with Expectations 

## Conditioning

```{r}
#| layout-ncol: 2 
#| fig-width: 7
earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

### The law of iterated expectations
$$
E \{ E( Y \mid X ) \} \quad \text{ for any random variables $X, Y$} 
$$

To calculate the mean of $Y$, we can average within subpopulations, then across subpopulations.

### Irrelevance of independent conditioning variables
$$ E( Y \mid X, X' ) = E( Y \mid X ) \quad \text{ when $X'$ is independent of $X,Y$ } $$

If $X'$ is unrelated to $X$ and $Y$, holding it constant doesn't affect the relationship between them.

## Linearity of Expectations

```{r}
#| layout-ncol: 2 
#| fig-width: 7

earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

$$ 
\begin{aligned}
E ( a Y + b Z ) 
&= E (aY) + E (bZ) \\
&= aE(Y) + bE(Z)  \\
& \text{ for random variables $Y, Z$ and numbers $a,b$ }
\end{aligned}
$$

  - There are two things going on here.
    - To average a sum of two things, we can take two averages and sum.
    - To average a constant times a random variable, we multiply the random variable's average by the constant.
  - In other words, we can *distribute* expectations and can pull constants out of them.
  
## Proof

- In essence, it comes down to the fact that all we're doing is summing.
	- Expectations are probability-weighted sums.
	- And we're looking at the expectation of a sum.
- And we can change the order we sum in without changing what we get.

$$
\small{
\begin{aligned}
\E\qty( a Y + b Z ) 
&= \sum_{y}\sum_z (a y + b z) \ P(Y=y, Z=z) && \text{ by definition of expectation} \\
&= \sum_{y}\sum_z a y \ P(Y=y, Z=z) + \sum_{z}\sum_y b z \ P(Y=y, Z=z) && \text{changing the order in which we sum} \\
&= \sum_{y} a y \ \sum_z P(Y=y,Z=z) + \sum_{z} b z \ \sum_y P(Y=y,Z=z) && \text{pulling constants out of the inner sums} \\
&= \sum_{y} a y \ P(Y=y) + \sum_{z} b z \ P(Z=z)  && \text{summing to get marginal probabilities from our joint } \\
&= a\sum_{y} y \ P(Y=y) + b\sum_{z} z \ P(Z=z)  && \text{ pulling constants out of the remaining sum } \\
&= a\E Y + b \E Z  && \text{by definition}
\end{aligned}
}
$$

## Linearity of **Conditional** Expectations

```{r}
#| layout-ncol: 2 
#| fig-width: 7

earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

$$ 
\begin{aligned}
E\{ a(X) Y + b(X) Z \mid X \} 
&= E\{a(X)Y \mid X\} + E\{ b(X)Z \mid X\} \\
&= a(X)E(Y \mid X) + b(X)E(Z \mid X)
\end{aligned}
$$

- This is like linearity of expectations, but with a twist.
   - When we condition on $X$, we're working with subpopulations in which $X$ is constant.
   - This means we can act as if functions of $X$ **are** constants.
   - So we can pull them out of expectations that are conditional on $X$.
   
## It's important to distinguish between two things

```{r}
#| fig-asp: .25
Y = c(.75,1,1.25, 1,1.25,1.5)
X = 0 + (1:6 > 3)
Xdodge = X + rep(.02*c(-1,0,1),2)
mu = function(x) { mean(Y[X==x]) }

tinypop.plot = ggplot() + geom_point(aes(x=Xdodge, y=Y, color=factor(X)), alpha=.7, size=2) + 
  scale_x_continuous(breaks=c(0,1), labels=c('0','1')) + scale_y_continuous(breaks=seq(.25,1.75,by=.25)) +
  guides(color='none') 
tinypop.plot
```

  1.  The **conditional expectation function** $\mu(x)=E[Y \mid X=x]$.
      - $\mu$ is a function; evaluated at $x$, it's a number. It's not random.
      - It's the mean of the subpopulation of people with $X=x$.
   
  2.  The conditional expectation $\mu(X)=E[Y \mid X]$.
      - $\mu(X)$ is the mean of a **random subpopulation** of people.
      - It's the conditional expectation function **evaluated at** the random variable $X$.
      - This is the sort of thing that shows up when we use the law of iterated expectations.

## Check Your Understanding {.smaller}

```{r}
#| fig-asp: .25
tinypop.plot
```

- Suppose we sample a point $(X, Y)$  uniformly at random from the population of 6 points above.
  - What is the conditional expectation function $\mu(x)$ at [$x=0$]{.twocolor-red} and [$x=1$]{.twocolor-green}?
  - What is the conditional expectation $\mu(X)$?
  
::: fragment
  - [$\mu(0)=`r mu(0)`$]{.twocolor-red} and [$\mu(1)=`r mu(1)`$]{.twocolor-green}
:::
::: fragment
  - $\mu(X)$ is a random variable taking on these two values.
  $$ \mu(X) = \begin{cases}
      `r mu(0)` & \text{ when } X=0 \\
      `r mu(1)` & \text{ when } X=1
      \end{cases}
  $$
:::

## The Indicator Trick
```{r}
#| fig-asp: .25
tinypop.plot
```

- Suppose we sample a point $(X, Y)$  uniformly at random from the population of 6 points above.
  - What is $1_{=1}(X)\mu(X)$?

::: fragment

- $1_{=1}(X)\mu(X)$ is a random variable taking on these two values.
  $$ 1_{=1}(X)\mu(X) = \begin{cases}
      0 \times \mu(0) = 0 \times `r mu(0)` & \text{ when } X=0 \\
      1 \times \mu(1) = 1 \times `r mu(1)` & \text{ when } X=1
      \end{cases}
  $$

- We can write it equivalently as $1_{=1}(X)\mu(1)$ because either ...
  - $X=0$, so $1_{=1}(X)\mu(X) = 1_{=1}(X)\mu(1) = 0$ 
  - $X=1$, so $1_{=1}(X)\mu(X) = 1_{=1}(X)\mu(1) =\mu(1)$
- This comes up a lot working with subsample means.
- We'll swap $1_{=1}(X)\mu(X)$ for $1_{=1}(X)\mu(1)$ often, referring to the **indicator trick**.

:::


## A More Realistic Example

```{r}
#| fig-asp: .25
#| 
pop.summaries = pop.earnings %>% group_by(degree) %>% summarize(mean=mean(income), sd=sd(income))
mean.has.degree = round(pop.summaries$mean[pop.summaries$degree=='4-year degree'] / 10000) * 10
mean.no.degree  = round(pop.summaries$mean[pop.summaries$degree=='no 4-year degree'] / 10000) * 10
degree.proportion = round(mean(pop.earnings$degree=='4-year degree'),1)
earnings.jitter.pop + geom_pointrange(aes(x=degree, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries) 
```

 - Let's think about a random person drawn from this population
    - $Y$ is their income 
    - $X$ is an indicator for having a four-year degree. 
  - Suppose the subpopulation means are `r approx(mean.has.degree, 10)`k for people with degrees and `r approx(mean.no.degree, 10)`k for people without.
  - And that `r approx(degree.proportion*10,1)`/10 of our population has degrees.
  
- **Q.**  What is $E(Y \mid X)$? And what is $E(Y)$?

:::: columns
::: column
::: fragment

- $E(Y \mid X)$ is a random variable that is **either** `r mean.has.degree`k or `r mean.no.degree`k
  - It's `r mean.has.degree`k with probability `r degree.proportion*10`/10,       when $X=1$. 
  - It's `r mean.no.degree`k  with probability `r (1-degree.proportion)*10`/10,   when $X=0$. 

::: 
:::
::: column
::: fragment

- $\E(Y)$ is its expectation. A number.
- We'll calculate it using iterated expectations.

$$
\begin{aligned} 
E\{ E( Y \mid X ) \} 
&= E(Y|X=1)P(X=1) + E(Y|X=0)P(X=0) \\
&= `r mean.has.degree`k \cdot `r degree.proportion*10`/10 +
   `r mean.no.degree`k  \cdot `r (1-degree.proportion)*10`/10
=  `r mean.has.degree*degree.proportion`k +
   `r mean.no.degree*(1-degree.proportion)`k 
=  `r mean.has.degree*degree.proportion + mean.no.degree*(1-degree.proportion)`k
\end{aligned}
$$

:::
:::
::::

## Review Exercise 1
```{r}
#| fig-asp: .25
tinypop.plot
```

- Suppose we sample a point $(X, Y)$ in the plot above uniformly at random.
  - Ignore the jitter; just think of the points as being at $X=0$ and $X=1$.
  - What is $\E\{ \mu(X) \}$?

::: fragment
  - There are two ways of thinking about calculating $\E\{\mu(X)\}$.
      1. We just calculate the expectation of it, thinking of it as an arbitrary random variable.
      2. We use the law of iterated expectations to show it's the unconditional mean of $Y$.

$$
\small{
\begin{aligned}
\E\qty{ \mu(X) } &= \frac{1}{2}\mu(0) + \frac{1}{2}\mu(1)= \frac{1}{2} \cdot `r mu(0)` + \frac{1}{2} \cdot `r mu(1)` = `r (mu(0)+mu(1))/2` && \text{ the first way } \\
\E\qty{ \mu(X) } &= \E\qty{\E\qty(Y \mid X)} = \E\qty{Y} = \frac{1}{6} \cdot `r Y[1]` + \frac{1}{6} \cdot `r Y[2]` + \ldots && \text{ the second way }
\end{aligned}
}
$$

:::

## Review Exercise 2
```{r}
#| fig-asp: .25
tinypop.plot
```

- Suppose we sample a point $(X, Y)$ in the plot above uniformly at random.
  - Ignore the jitter; just think of the points as being at $X=0$ and $X=1$.
  - What is $\E\{ 1_{=1}(X)\mu(X) \}/\E(1_{=1}(X))$?

::: fragment

- It's $\mu(1)$. $1_{=1}(X)\mu(X)=1_{=1}(X)\mu(1)$, so ...
$$
\begin{aligned}
\E\{ 1_{=1}(X)\mu(X) \} 
&= \E\{ 1_{=1}(X)\mu(1) \} \\
&= \mu(1) \ \E\{ 1_{=1}(X) \} \\
\end{aligned}
$$

:::

## Review Exercise 3

```{r}
#| fig-asp: .25
tinypop.plot
```

- Suppose we sample a point $(X, Y)$ in the plot above uniformly at random.
  - Ignore the jitter; just think of the points as being at $X=0$ and $X=1$.
  - What is $\E\{ 1_{=0}(X)\mu(X) \}/\E(1_{=0}(X))$?

::: fragment

- It's $\mu(0)$. It's analogous to the last one. $1_{=0}(X)\mu(X)=1_{=0}(X)\mu(0)$, so ...
$$
\begin{aligned}
\E\{ 1_{=0}(X)\mu(X) \} 
&= \E\{ 1_{=0}(X)\mu(0) \} \\
&= \mu(0) \ \E\{ 1_{=0}(X) \} \\
\end{aligned}
$$

:::


# Unbiasedness of Means

## The Sample Mean

```{r}
#| layout-ncol: 2
#| fig-width: 7
unbiased.plot
biased.plot
```

**Claim.** The sample mean is an unbiased estimator of the population mean.
$$ \E[\hat\mu] = \mu $$

::: fragment

$$
\begin{aligned}
\E\qty[\frac1n\sum_{i=1}^n Y_i] 
&= \frac1n\sum_{i=1}^n \E[Y_i] && \text{ via linearity } \\
&= \frac1n\sum_{i=1}^n \mu     && \text{ via equal-probability sampling } \\
&= \frac1n \times n \times \mu = \mu.
\end{aligned}
$$

:::

## A Subsample Mean

**Claim.** The subsample mean is unbiased for the subpopulation mean.
$$ \E[\hat\mu(1)] = \mu(1) $$

::: {.panel-tabset}

### Proof Sketch

  - Use the Law of Iterated Expectations, conditioning on $X_1 \ldots X_n$.
  - Then the linearity of conditional expectations to push the conditional expectation into the sum.
  - Then irrelevance of independent conditioning variables to write things in terms of the *random variable* $\mu(X_i)$  
  - Then the **indicator trick**. What's $1_{=1}(X_i) \mu(X_i)$? How is it related to $\mu(1)$?

### How to Write Subsample Means

$$ 
\hat\mu(1) = \frac{\sum_{i:X_i=1} Y_i}{\sum_{i:X_i=1} 1} = \frac{\sum_{i=1}^{n} 1_{=1}(X_i) Y_{i}}{\sum_{i=1}^{n} 1_{=1}(X_i)} 
$$

  - It's easy to make mistakes using *linearity of expectations* when we're summing over a subsample.
    - Can we or can we not 'push' or 'pull' and expectation through a sum ...
    - ... when the terms in that sum depend on the value of a random variable? 
  - To make this a lot more obvious, we can rewrite these as sums over the whole sample.
    - Instead of 'excluding' the terms where $X_i=0$, we 'make them zero' by multiplying in the indicator $1_{=1}(X_i)$.
    - Then, of course we can distribute expectations through the sum. 
    - The question becomes whether we can 'pull out' the indicator.
    - And we have a rule for that. We can do it if we're conditioning on $X_i$.
  - When we write a subsample mean, we can do that ...
    - in the numerator, the sum of observations in the subsample.
    - in the denominator, the number of those terms, which is a *sum of ones* over the subsample.

### Proof

$$
\small{
\begin{aligned}
\E[\hat\mu(1)] 
&=\E\qty[ \E\qty{\frac{\sum_{i:X_i=1} Y_{i}}{\sum_{i:X_i=1} 1} \mid X_1 \ldots X_n}] \\
&=\E\qty[ \E\qty{\frac{\sum_{i=1}^{n} 1_{=1}(X_i) Y_{i}}{\sum_{i=1}^{n} 1_{=1}(X_i)} \mid X_1 \ldots X_n}] \\
&=\E\qty[\frac{\sum_{i=1}^{n} 1_{=1}(X_i) \E\qty{ Y_{i} \mid X_i}}{\sum_{i=1}^{n}1_{=1}(X_i)}] \\
&=\E\qty[\frac{\sum_{i=1}^{n}1_{=1}(X_i) \mu(X_i)}{\sum_{i=1}^{n}1_{=1}(X_i)}] \\
&=\E\qty[\frac{\sum_{i=1}^{n}1_{=1}(X_i) \mu(1)}{\sum_{i=1}^{n}1_{=1}(X_{i})}] \\
&=\E\qty[\frac{\mu(1)\sum_{i=1}^{n} 1_{=1}(X_i) }{\sum_{i=1}^{n}1_{=1}(X_{i})}] \\
&=\mu(1) \ \E\qty[\frac{\sum_{i=1}^{n} 1_{=1}(X_i) }{\sum_{i=1}^{n}1_{=1}(X_{i})}] = \mu(1) \E[1] = \mu(1).
\end{aligned}
}
$$

:::

## Differences in Subsample Means

**Claim.** The difference in subsample means is unbiased for the difference in subpopulation means.

$$ \E[\hat\mu(1) - \hat\mu(0)] = \mu(1) - \mu(0) $$

::: fragment
This follows from the linearity of expectations and unbiasedness of the subsample means.

$$
\E[\hat\mu(1) - \hat\mu(0)] = \E[\hat\mu(1)] - \E[\hat\mu(0)] =  \mu(1) - \mu(0) 
$$

:::

## Summary

```{r}
#| layout-ncol: 2
#| fig-width: 7
#| fig-cap: 
#|   - "Like this."
#|   - "Not like this."
xx = seq(-4,4,by=.01)
heights = seq(0,.4,length.out=100)
normals = rnorm(10000)

ggplot() + geom_histogram(aes(x=normals, y=after_stat(density)), alpha=.3, bins=100) +
           geom_vline(xintercept=0, color='green', linewidth=3, alpha=.4) + 
           geom_vline(xintercept=0, color='blue') + 
           geom_vline(xintercept=c(-1,1), color='blue', linetype='dashed') +
           geom_vline(xintercept=c(-2,2), color='blue', linetype='dotted') +
           geom_point(aes(y=heights, x=normals[1:length(heights)]), color='purple', alpha=.3)

ggplot() + geom_histogram(aes(x=normals+2, y=after_stat(density)), alpha=.3, bins=100) +
           geom_vline(xintercept=0, color='green', linewidth=3, alpha=.4) + 
           geom_vline(xintercept=2+0, color='blue') + 
           geom_vline(xintercept=2+c(-1,1), color='blue', linetype='dashed') +
           geom_vline(xintercept=2+c(-2,2), color='blue', linetype='dotted') +
           geom_point(aes(y=heights, x=normals[1:length(heights)]+2), color='purple', alpha=.3) +
           xlim(-4,4)
```

- Subsample means are *unbiased estimators* of the corresponding population means.
- And, expectation being linear, this extends to *differences in subsample means*.
- Now that we've worked out that the location is good, it's time to talk about spread.
- That's what we'll do next time.

