---
title: "Lab 5"
subtitle: "A Game of Telephone"
resources:
  - data
webr:
  resources:
    - data/tinypop.csv
    - data/mht-pop.csv
    - data/mht-sam.csv
    - data/earnings.csv
    - data/pop-earnings.csv
---


## Data {visibility="hidden"}

{{< include ../_extensions/r-wasm/live/_knitr.qmd >}}
{{< include shared-code.qmd >}}
{{< include ca-income.qmd >}}

```{r}
#| label: tiny-pop-data
pop = data.frame(
  names = c('Alice', 'Bob', 'Charlie', 'Dana', 'Ed', 'Frank', 'Grace', 'Heidi'),
  x = c(12,12,14,14,16,16,16,16),
  y = c(40,50,50,60,75,65,90,70),
  j = .03*c(-1,1,-1,1,-1,1,-1,1)) 
pop$jx = pop$x + pop$j
write.csv(pop, 'data/tinypop.csv')
```


```{r}
#| label: income-data
write.csv(pop.earnings, 'data/pop-earnings.csv')
write.csv(earnings, 'data/earnings.csv')
```

```{r}
#| label: mht-data
set.seed(1) 
m = 1e4 
pi = function(x) { 1/(1+exp(-(x-60)/8 )) }
mu = function(w, x) { 1 - (x-50)/10 + w*((x-50)/10)^2/2 }  

X = 50+2*round(5*runif(m))
W = 0 + (runif(m) <= pi(X))
epsilon0 = rnorm(m)
epsilon1 = rnorm(m)
Y0 = mu(0,X) + .25*epsilon0
Y1 = mu(1,X) + .25*epsilon1

pop = data.frame(X=X, W=W, Y=Y0*(1-W) + Y1*W, Y0=Y0, Y1=Y1)

n = 5000
sam = pop[sample(nrow(pop), n, replace=TRUE),]
write.csv(pop, 'data/mht-pop.csv')
write.csv(sam, 'data/mht-sam.csv')
```

```{r}
mean_sd = function(x) {
  data.frame(y=mean(x), ymin=mean(x)-sd(x), ymax=mean(x)+sd(x))
}

triangle =  'x25B2'
circle = 'x25CF'
square = 'x25A0'
diamond = 'x25C6'
```

## A Small Population 

```{=html}
<style>
.reveal { background: #ffffff; }
</style>
```

```{r}
#| echo: false
#| layout-ncol: 2
#| fig-asp: .4 

pop = read.csv('data/tinypop.csv')
scales =   list(scale_x_continuous(
                  breaks=c(12,14,16), 
                  labels=c('High School Diploma', '2-Year Degree', '4-Year Degree'), 
                  limits=c(11.5,16.5)),
                scale_y_continuous(
                  breaks=c(40,50,60,70,80,90), 
                  labels=c('40k', '50k', '60k', '70k', '80k', '90k')), 
                coord_cartesian(ylim=c(35,95)))  
plot.tiny.increments = function(pop) {
  summaries = pop |> group_by(x) |> summarize(mu=mean(y), sigma=sd(y), n=n())
  ggplot(pop) + 
    geom_point(aes(x=jx, y=y, color=factor(names)), size=3) + 
    geom_text(aes(x=jx + .03, y=y, label=names, color=factor(names)), hjust=0,vjust=0, size=5) +
    geom_pointrange(aes(x=x, y=mu, ymin=mu-sigma, ymax=mu+sigma, shape=factor(x)), data=summaries, alpha=.4, size=1.5) +
    geom_line(aes(x=x, y=mu), data=summaries, linetype='dashed', alpha=.3) + 
    scales + guides(color='none', shape='none')
}
theme_set(whitebw.theme())
```

```{r}
plot.tiny.increments(pop)
```

$$
\begin{aligned}
\theta_{\text{degrees}} 
&= \text{average in our population, over 2 and 4-year degrees, of the incremental value of the degree} \\
&=\frac{(\unicode{`r square`} - \unicode{`r triangle`}) + \class{fragment}{(\unicode{`r triangle`} - \unicode{`r circle`})}}{2} \\ 
&=\sum_x \alpha(x) \mu(x) \qfor  \alpha(x) = \begin{cases}  
\class{fragment}{\frac12}&\qqtext{if} x=\text{\unicode{`r square`} \ 4-year degree} \\
\class{fragment}{\frac{-1 + 1}{2}=0}&\qqtext{if} x=\text{\unicode{`r triangle`} \ 2-year degree} \\
\class{fragment}{-\frac12}&\qqtext{if} x=\text{\unicode{`r circle`} \ high school diploma} 
\end{cases} \\
& \\
& \\ 
\theta_{\text{people}}
&= \text{average in our population, over people with 2 and 4-year degrees, of the incremental value of their last degree} \\
&= \frac{4 \times (\unicode{`r square`} - \unicode{`r triangle`}) + \class{fragment}{2 \times (\unicode{`r triangle`} - \unicode{`r circle`})}}{6} \\ 
&=\sum_x \alpha(x) \mu(x) \qfor  \alpha(x) = \begin{cases}  
\class{fragment}{\frac46}&\qqtext{if} x=\text{\unicode{`r square`} \ 4-year degree} \\
\class{fragment}{\frac{-4 + 2}{6}=-\frac26}&\qqtext{if} x=\text{\unicode{`r triangle`} \ 2-year degree} \\
\class{fragment}{-\frac26}&\qqtext{if} x=\text{\unicode{`r circle`} \ high school diploma} 
\end{cases} \\
\end{aligned}
$$  
  

## A Small Population, Sampled

```{r}
J1 = c(1,3,5,7,8)
sam = pop[J1, ]
plot.tiny.increments(sam)
```

$$
\begin{aligned}
\hat\theta_{\text{degrees}} 
&= \text{average in our sample, over 2 and 4-year degrees, of the incremental value of the degree} \\
&=\frac{(\unicode{`r square`} - \unicode{`r triangle`}) + \class{fragment}{(\unicode{`r triangle`} - \unicode{`r circle`})}}{2} \\ 
&=\sum_x \hat\alpha(x) \hat\mu(x) \qfor  \hat\alpha(x) = \begin{cases}  
\class{fragment}{-\frac12}&\qqtext{if} x=\text{\unicode{`r square`} \ 4-year degree} \\
\class{fragment}{\frac{-1 + 1}{2}=0}&\qqtext{if} x=\text{\unicode{`r circle`} \ high school diploma} \\
\class{fragment}{\frac12}&\qqtext{if} x=\text{\unicode{`r triangle`} \ 2-year degree} 
\end{cases} \\
& \\
& \\ 
\hat\theta_{\text{people}}
&= \text{average in our sample, over people with 2 and 4-year degrees, of the incremental value of their last degree} \\
&= \frac{3 \times (\unicode{`r square`} - \unicode{`r triangle`}) + \class{fragment}{(\unicode{`r triangle`} - \unicode{`r circle`})}} {4} \\ 
&=\sum_x \hat\alpha(x) \hat\mu(x) \qfor  \hat\alpha(x) = \begin{cases}  
\class{fragment}{\frac34}&\qqtext{if} x=\text{\unicode{`r square`} \ 4-year degree} \\
\class{fragment}{\frac{-3 + 1}{4}=-\frac24}&\qqtext{if} x=\text{\unicode{`r circle`} \ high school diploma} \\
\class{fragment}{-\frac14}&\qqtext{if} x=\text{\unicode{`r triangle`} \ 2-year degree} 
\end{cases} \\
\end{aligned}
$$  
  

 
## A Small Population, Sampled Again

```{r}
#| fig-asp: .4
#| 
J2 = c(1,2,3,4,6,8)
sam = pop[J2, ]
plot.tiny.increments(sam)
```

$$
\begin{aligned}
\hat\theta_{\text{degrees}} 
&= \text{average in our sample, over 2 and 4-year degrees, of the incremental value of the degree} \\
&=\frac{\class{fragment}{(\unicode{`r square`} - \unicode{`r triangle`})} + \class{fragment}{(\unicode{`r triangle`} - \unicode{`r circle`})}}{2} \\ 
&=\sum_x \hat\alpha(x) \hat\mu(x) \qfor  \hat\alpha(x) = \begin{cases}  
\class{fragment}{\frac12}&\qqtext{if} x=\text{\unicode{`r square`} \ 4-year degree} \\ 
\class{fragment}{\frac{-1 + 1}{2}=0} &\qqtext{if} x=\text{\unicode{`r triangle`} \ 2-year degree} \\
\class{fragment}{-\frac12}&\qqtext{if} x=\text{\unicode{`r circle`} \ high school diploma} 
\end{cases} \\
& \\
& \\ 
\hat\theta_{\text{people}}
&= \text{average in our sample, over people with 2 and 4-year degrees, of the incremental value of their last degree} \\
&= \frac{2 \times (\unicode{`r square`} - \unicode{`r triangle`}) + \class{fragment}{2 \times (\unicode{`r triangle`} - \unicode{`r circle`})}}{4} \\ 
&=\sum_x \hat\alpha(x) \hat\mu(x) \qfor  \hat\alpha(x) = \begin{cases}  
\class{fragment}{\frac24}&\qqtext{if} x=\text{\unicode{`r square`} \ 4-year degree} \\
\class{fragment}{\frac{-2 + 2}{4}=0}&\qqtext{if} x=\text{\unicode{`r triangle`} \ 2-year degree} \\
\class{fragment}{-\frac24} &\qqtext{if} x=\text{\unicode{`r circle`} \ high school diploma} 
\end{cases} \\
\end{aligned}
$$  
  

## Setup
```{webr}
library(tidyverse)
pop = read.csv('data/tinypop.csv')

J1 = c(1,3,5,7,8)
sam = pop[J1,]

summary.functions = function(pop) { 
  summaries = pop |> group_by(x) |> summarize(mean=mean(y), sd=sd(y), n=n())
  list(mu = function(x) { summaries$mean[match(x, summaries$x)] }, 
       sigma = function(x) { summaries$sd[match(x, summaries$x)] },
       n = function(x) { summaries$n[match(x, summaries$x)] })
}

pop.functions = summary.functions(pop)
sam.functions = summary.functions(sam)

m  = pop.functions$n
mu = pop.functions$mu
sigma = pop.functions$sigma

N = sam.functions$n
muhat = sam.functions$mu
sigmahat = sam.functions$sigma

x = pop$x
X = sam$x 
```

## Writing the Code
```{webr}
degree.years = c(14,16)
theta.years = mean(mu(degree.years)) 
theta.people = mean(mu(X[X %in% degree.years]))
```

```{webr}
p = function(x) { m(x)/sum(m(degree.years)) }
theta.people.table = mean(mu(degree.years) * p(degree.years))
```

## A Bigger Version
```{r}
pop.dot.size = .75
n = nrow(earnings)
m = nrow(pop.earnings)

# reduce size of pop to 5x sample size to keep the browser happy
pop.idx = c(1:4*n, (m-n+1):m) 

jitter = function(earnings, idx=pop.idx, foreground=TRUE, color=default.color, background.alpha=.01, foreground.alpha=.2) { 
  earnings$foreground = foreground
  earnings$color = color
  ggplot(earnings[idx,]) + geom_point(aes(x=education, y=income, alpha=foreground, color=I(color)), 
                                size=pop.dot.size, position=position_jitter(seed=0,w=.3)) +
                                scale_x() + scale_y() + scale_alpha_manual(breaks=c(FALSE, TRUE), values=c(background.alpha,foreground.alpha)) + 
                                guides(color=FALSE, alpha=FALSE)
} 
degree.years = c(12, 14, 16, 18, 20)
plot.aggregate.comparison = function(earnings, idx=pop.idx) {
  summaries = earnings |> group_by(education) |> summarize(mean=mean(income), sd=sd(income), mean.X = mean(education))
  aggregate.comparison.plot= jitter(earnings, idx) + 
    geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), 
      data=summaries[summaries$education %in% degree.years,]) +
    geom_line(aes(x=education, y=mean), color='blue', alpha=.4,
      data=summaries[summaries$education %in% range(degree.years),]) + 
    geom_line(aes(x=education, y=mean), color='black', linetype='dashed',
      data=summaries[summaries$education %in% degree.years,]) 
  aggregate.comparison.plot + scale_x(c(11.5,20.5)) + scale_y(c(0, 2e5)) 
}

default.color = twocolor.red
epsilon = rnorm(m)*sd(pop.earnings$income)/2
mu = map_dbl(1:20, function(x) { mean(pop.earnings$income[pop.earnings$education==x]) })
earnings3 = pop.earnings 
earnings3$income = mu[ifelse(pop.earnings$education==16, 14, pop.earnings$education)] + epsilon

over.people = function(earnings) {
  degreeed = earnings$education[earnings$education %in% degree.years]
  mu = map_dbl(1:20, function(x) { mean(earnings$income[earnings$education==x]) })
  mean(mu[degreeed] - mu[degreeed-2])
}
over.years = function(earnings) {
  mu = map_dbl(1:20, function(x) { mean(earnings$income[earnings$education==x]) })
  mean(mu[degree.years[2:5]] - mu[degree.years[1:4]])
}
```

::: {.panel-tabset}
### Plots
```{r}
#| layout-ncol: 2
#| fig-width: 7

pe = earnings3
J = sample(1:nrow(pe), 2000)
sam = pe[J, ]
plot.aggregate.comparison(pe, idx=1:(10*nrow(sam)))
plot.aggregate.comparison(sam, idx=1:nrow(sam))
```

```{r}
mu = map_dbl(1:20, function(x) { mean(pe$income[pe$education==x]) })
mx = map_dbl(1:20, function(x) { sum(pe$education==x) })
px = mx/sum(mx[degree.years])

muhat = map_dbl(1:20, function(x) { mean(sam$income[sam$education==x]) })
mxhat = map_dbl(1:20, function(x) { sum(sam$education==x) })
pxhat = mxhat/sum(mxhat[degree.years[2:5]])
```

```{webr}
pop.earnings = read.csv('data/pop-earnings.csv')
earnings = read.csv()
```

### Summary Tables
:::: {.columns}
::: {.column}

$x$ | $m_x$ | $\mu(x)$  
---|--|--
`r degree.years[1]` | `r approx(mx[degree.years[1]])` | `r thousand(mu[degree.years[1]], 1)`
`r degree.years[2]` | `r approx(mx[degree.years[2]])` | `r thousand(mu[degree.years[2]], 1)`
`r degree.years[3]` | `r approx(mx[degree.years[3]])` | `r thousand(mu[degree.years[3]], 1)`
`r degree.years[4]` | `r approx(mx[degree.years[4]])` | `r thousand(mu[degree.years[4]], 1)`
`r degree.years[5]` | `r approx(mx[degree.years[5]])` | `r thousand(mu[degree.years[5]], 1)`

:::
::: {.column}
$x$ | $N_x$ | $\hat\mu(x)$  
---|--|--
`r degree.years[1]` | `r approx(mxhat[degree.years[1]])` | `r thousand(muhat[degree.years[1]], 1)`
`r degree.years[2]` | `r approx(mxhat[degree.years[2]])` | `r thousand(muhat[degree.years[2]], 1)`
`r degree.years[3]` | `r approx(mxhat[degree.years[3]])` | `r thousand(muhat[degree.years[3]], 1)`
`r degree.years[4]` | `r approx(mxhat[degree.years[4]])` | `r thousand(muhat[degree.years[4]], 1)`
`r degree.years[5]` | `r approx(mxhat[degree.years[5]])` | `r thousand(muhat[degree.years[5]], 1)`
:::
:::: 
:::

::: {.panel-tabset}
### R Setup
```{webr}
pop = read.csv('data/pop-earnings.csv')
sam = read.csv('data/earnings.csv')
pop$x = pop$education
sam$x = sam$education
pop$y = pop$income
sam$y = sam$income

pop.functions = summary.functions(pop)
sam.functions = summary.functions(sam)

m  = pop.functions$n
mu = pop.functions$mu
sigma = pop.functions$sigma

N = sam.functions$n
muhat = sam.functions$mu
sigmahat = sam.functions$sigma

x = pop$education
X = sam$education
```


### Exercise 1
- What is the average in our population, over the four degree types shown, of the incremental value of the degree? 
- How do we estimate it using our sample? 

```{webr}
degree.years = c(14,16,18,20)
mean(mu(degree.years) - mu(degree.years-2))
mean(muhat(degree.years) - muhat(degree.years-2))
```



### Exercise 2
- What is the average in our population, over people with these 4 degrees, of the incremental value of their last degree? 
- How do we estimate it using our sample?

:::

## A New Dataset

```{r}
set.seed(1) 
m = 1e4 
pi = function(x) { 1/(1+exp(-(x-60)/8 )) }
mu = function(w, x) { 1 - (x-50)/10 + w*((x-50)/10)^2/2 }  

X = 50+2*round(5*runif(m))
W = 0 + (runif(m) <= pi(X))
epsilon0 = rnorm(m)
epsilon1 = rnorm(m)
Y0 = mu(0,X) + .25*epsilon0
Y1 = mu(1,X) + .25*epsilon1

pop = data.frame(X=X, W=W, Y=Y0*(1-W) + Y1*W, Y0=Y0, Y1=Y1)

n = 5000
sam = pop[sample(nrow(pop), n, replace=TRUE),]
```

```{r}
scatter = ggplot(sam[sam$W==1,]) + 
  geom_point(aes(x=X, y=Y), alpha=.1, position=position_jitter(w=.15, h=0), color=twocolor.red) + 
  guides(color='none')

annotated.scatter = scatter +  
  stat_summary(aes(x=X, y=Y), geom="pointrange", alpha=.8,
                 fun=mean, 
                 fun.min = function(x) mean(x) - sd(x), 
                 fun.max = function(x) mean(x) + sd(x))
annotated.scatter + scale_x_continuous(breaks=seq(50,60,2)) 
```

```{r}
X = sam$X[sam$W==1]
Y = sam$Y[sam$W==1]
```

:::: columns 
::: {.column width="25%"}
x | $N_x$ | $\hat\mu(x)$
---|---|---
50 | `r sum(X==50)` | `r round(mean(Y[X==50]),2)`
52 | `r sum(X==52)` | `r round(mean(Y[X==52]),2)`
54 | `r sum(X==54)` | `r round(mean(Y[X==54]),2)`
56 | `r sum(X==56)` | `r round(mean(Y[X==56]),2)`
58 | `r sum(X==58)` | `r round(mean(Y[X==58]),2)`
60 | `r sum(X==60)` | `r round(mean(Y[X==60]),2)`

:::
::: {.column width="75%"}
::: {.panel-tabset}
### About the Data
- Menopausal Hormone Therapy, or MHT, is a treatment for the symptoms of menopause. 
  - It's said that it improves bone density. 
  - But its effect does seem to depend on when it's started.
- What we see here is a plot of $y$ vs $x$ in a sample of people who receive MHT.
  - y: bone density at age 65 
  - x: age at which MHT was started

### Exercise 
- You will be given something to estimate. 
 - You will describe it as precisely as possible and hand your description back.
 - Then you'll compute it, either by hand or using R in the **R** tab.
- Row 1. I'll give you an iPad. In your description, you may want to mark up the plot above.
- Row 2. Write your description in mathematical notation.
- Row 3. Write your description in plain English. And hold on to it. 

### R Setup (ignore)
```{webr}
set.seed(1) 
m = 1e4 
pi = function(x) { 1/(1+exp(-(x-60)/8 )) }
mu = function(w, x) { 1 - (x-50)/10 + w*((x-50)/10)^2/2 }  

X = 50+2*round(5*runif(m))
W = 0 + (runif(m) <= pi(X))
epsilon0 = rnorm(m)
epsilon1 = rnorm(m)
Y0 = mu(0,X) + .25*epsilon0
Y1 = mu(1,X) + .25*epsilon1

pop = data.frame(X=X, W=W, Y=Y0*(1-W) + Y1*W, Y0=Y0, Y1=Y1)

n = 5000
sam = pop[sample(nrow(pop), n, replace=TRUE),]
Y = sam$Y[sam$W==1]
X = sam$X[sam$W==1]
```

```{webr}
N = Vectorize(function(x) { sum(X==x) })
muhat = Vectorize(function(x) { mean(Y[X==x]) }) 
```
### R 
```{webr}
ages = c(50,52,54,56,58,60)
X[1:5]
Y[1:5]
N(ages)
muhat(ages)
```

:::
:::
::::
