---
title: "Lecture 12"
subtitle: "The Behavior of Least Squares Predictors"
---

# Introduction 

## Today's Topic

{{< include ../_extensions/r-wasm/live/_knitr.qmd >}}
{{< include shared-code.qmd >}}


```{r}
#| label: fake-data
x = rep(c(20,30,40), each=5000)

pi.pop = function(x) { case_when(x < 25 ~ .25, x < 35 ~ .5, TRUE ~ .75) } 
mu.pop = function(w,x) case_when(
  w==1 & x <  25  ~ 10, 
  w==1 & x <= 35  ~ 10,
  w==1            ~ 20, 
  w==0 & x <  25  ~ 5, 
  w==0 & x <= 35  ~ 10,
  w==0            ~ 10)

w = 0 + (runif(length(x)) < pi.pop(x))
y = rpois(length(x), mu.pop(w,x))
fake.pop = data.frame(w=w, x=x, y=y)
```

```{r}
pop = fake.pop 
m = nrow(pop)
n = 200
set.seed(5)
J = sample(1:m, size=n, replace=TRUE)
sam = pop[J,]
```

```{r}
prop.scale = 40
default.shift = .5
plotX = function(w,x, shift=default.shift) { ifelse(w==0, x-shift, x+shift) }
lsq.plot = function(sam, formula=NULL, dot.size=1, shift=default.shift, residual.alpha=0) {
  sam$jitterX = runif(nrow(sam), -1, 1)*shift
  base.plot = ggplot(sam) + 
    geom_point(aes(x=plotX(w,x)+jitterX,y=y,color=factor(w)), 
      alpha=.05, size=dot.size) + 
    stat_summary(aes(x=plotX(w,x), y=y), 
      fun.data=mean_sd, geom='pointrange', alpha=.2) +
    geom_bar(aes(x=x, y=prop.scale*after_stat(prop)), 
      fill='purple', alpha=.1, linewidth=.05) +
    scale_y_continuous(sec.axis = sec_axis(~./prop.scale, name='proportion')) +
    scale_x_continuous(breaks=c(20,30,40)) +
    coord_cartesian(xlim=c(15,45), ylim=c(-1,40)) +
    guides(color='none') 
    
    if(is.null(formula)) { base.plot } else { 
        model = lm(formula, data=sam)
        gridx = seq(-10, 50, length.out=1000)
        muhat = function(w,x) predict(model, newdata=data.frame(w=w,x=x))   
        the.plot = base.plot + 
          geom_point(aes(x=plotX(w,x), y=muhat(w,x), color=factor(w)), 
            alpha=.7, shape='square') +
          geom_line(aes(x=plotX(w,x), y=muhat(w,x), color=factor(w)), 
            data = expand_grid(w=0:1, x=gridx), alpha=.7, linetype='dotted') 
        if(residual.alpha > 0) {
          the.plot = the.plot + 
            geom_segment(aes(x=plotX(w,x)+jitterX, xend=plotX(w,x), y=y, yend=muhat(w,x)), 
              alpha=residual.alpha, color='orange', linewidth=.1)
        }
        the.plot 
    }
}
```

```{r}
lsq.plot.1 = function(sam, formula, dot.size=1, color=groupa, residual.alpha=0) { 
  sam$jitterX = runif(nrow(sam), -1, 1)
  base.plot = ggplot(sam) + 
    geom_point(aes(x=x+jitterX,y=y, color=color), 
      alpha=.05, size=dot.size) + 
    stat_summary(aes(x=x+jitterX, y=y), 
      fun.data=mean_sd, geom='pointrange', alpha=.2) +
    geom_bar(aes(x=x, y=prop.scale*after_stat(prop)), 
      fill='purple', alpha=.1, linewidth=.05) +
    scale_y_continuous(sec.axis = sec_axis(~./prop.scale, name='proportion')) +
    guides(color='none') 
    
    if(is.null(formula)) { base.plot } else { 
        model = lm(formula, data=sam)
        muhat = function(x) predict(model, newdata=data.frame(x=x))   
        the.plot = base.plot + 
          geom_point(aes(x=x, y=muhat(x)), color=color, alpha=.7, shape=muhat.shape) +
          geom_line(aes(x=x, y=muhat(x)),  color=color, alpha=.7, shape=muhat.linetype) 
        if(residual.alpha > 0) {
          the.plot = the.plot + 
            geom_segment(aes(x=x+jitterX, xend=x+jitterX, y=y, yend=muhat(x)), 
              alpha=residual.alpha, color='orange', linewidth=.05)
        }
        the.plot 
    }
}
```


```{r}
#| cache: true
estimate = function(sam, formula) {
  X = sam$x
  W = sam$w
  model = lm(formula, data=sam)
  muhat = function(w,x) predict(model, newdata=data.frame(w=w,x=x)) 
  mean(muhat(1, X) - muhat(0, X))
}

samples = 1:4000 |> map(function(.) { 
  J = sample(1:nrow(pop), size=n, replace=TRUE)
  pop[J,]
})

target  = pop |> estimate(y~w*factor(x))
estimates = data.frame( 
  all.functions = samples |> map_vec(estimate, y~w*factor(x)),
  lines = samples |> map_vec(estimate, y~w*x))
```

```{r}
#| fig.width: 7
#| layout-ncol: 2
#| fig-cap: 
#|  - 'A sample drawn with replacement from a population.'
#|  - 'That population.'
sam |> lsq.plot(y~w*x)
pop |> lsq.plot()
```

We're going to talk about statistical inference when we use least squares in linear models. For example ...
$$
\color{gray}
\begin{aligned}
\theta &= \frac{1}{m}\sum_{j=1}^m \qty{ \mu(1,x_j) -  \mu(0,x_j) } \qfor \mu(w,x)=\frac{1}{m_{w,x}}\sum_{j:w_j=w,x_j=x} y_j \\
\hat\theta &= \frac{1}{n}\sum_{i=1}^n \qty{ \hat\mu(1,X_i) -  \hat\mu(0,X_i) } \qfor \hat\mu = \argmin_{m \in \model} \sum_{i=1}^n \qty{ Y_i - m(X_i) }^2
\end{aligned}
$$

- Does it look like this is a good estimator?
- If not, do you think it's too big or too small?

## Summary

```{r}
#| fig.width: 7
#| layout-ncol: 2
#| fig-cap: 
#|  - 'The sampling distribution of $\hat\theta$ and a perfectly calibrated interval estimate.'
#|  - 'The population least squares predictor $\tilde\mu$ fit to the population.'
#| 
theta.hat = estimate(sam, y~w*x)
theta.tilde = estimate(pop, y~w*x)
theta.sampling.dist.plot =ggplot(estimates) + 
  geom_histogram(aes(x=all.functions, y=after_stat(density)), bins=50, alpha=.05, fill=pink, color='gray', linewidth=.05) + 
  geom_histogram(aes(x=lines, y=after_stat(density)), bins=50, alpha=.2, fill=teal, color='gray', linewidth=.05) + 
  with_annotations(c(theta.hat,estimates$lines), estimand=target, highlight.dots=1, interval.width=width(estimates$lines)) + 
  geom_vline(xintercept=theta.tilde, color=pink, alpha=.3, linewidth=2.5)

plot.pop = pop[sample(1:nrow(pop), size=10000),]
pop.lsq.plot = plot.pop |> lsq.plot(y~w*x)

theta.sampling.dist.plot
pop.lsq.plot
```

::: {.panel-tabset}
### What happened?
- In this case, $\hat\theta$ is too small. And that's not just a chance occurrence. 
  - When we look at its sampling distribution, we see that it's *systematically* too small.
  - It's *biased*. In roughly 95% of samples, $\hat\theta$ will be smaller than $\theta$.
- It boils down to a bad choice of regression model. 
  - To make sense of this, think about what happens when everything else is perfect.
  - i.e. if we fit this model to the whole population, then average over the population.
  - The result is *nonrandom*. And it's right in the middle of our estimator's sampling distribution.
$$
\inpink{\tilde\theta} = \frac{1}{m}\sum_{j=1}^m \qty{ \tilde\mu(1,x_j) -  \tilde\mu(0,x_j) } \qfor \tilde\mu = \argmin_{m \in \model} \frac{1}{m}\sum_{j=1}^m \qty{y_j - m(w_j,x_j)}^2 
$$

### What happens generally 

We can think of our estimator's error as a sum of two things.

$$
\hat\theta - \theta = 
\underset{\text{sampling error}}{\qty{\hat\theta - \inpink{\tilde\theta}}} + 
\underset{\text{modeling error}}{\qty{\inpink{\tilde\theta} - \theta}}
$$

1. *Sampling error* is the 'variance part'. It's random and has mean zero. 
  - It's the error we'd have had if what we'd really wanted to estimate was $\tilde\theta$.
  - We can get a good estimate of this term's distribution using the bootstrap. 
2. *Modeling error* is the 'bias part'. It's non-random---it's the bias of our estimator. 
  - It's the error we'd have had if we'd been given the whole population as our sample.
  - This is zero *if* your model contains the subpopulation mean function $\mu(w,x)$.
    - e.g. when you use the 'all functions' model.

  - That we have a nice, clean breakdown of our error like this isn't obvious.
    - You can, of course, add and subtract whatever you want. We chose $\inpink{\tilde\theta}$.
    - But it's not obvious that the sampling error term $\hat\theta - \inpink{\tilde\theta}$ has mean zero.

### Terminology
  
  - We say a model is *correctly specified* when it contains $\mu$ 
  - We say it is *misspecified* when it doesn't contain $\mu$.

:::
  
## Plan 

$$
\hat\theta - \theta = 
\underset{\text{sampling error}}{\qty{\hat\theta - \inpink{\tilde\theta}}} + 
\underset{\text{modeling error}}{\qty{\inpink{\tilde\theta} - \theta}}
$$

- Today, we'll do two things. 
 1. We'll prove that the sampling error term has mean zero. 
    - In fact, we'll prove that it's an average of mean-zero random variables.
    - That'll take a little calculus. It's not a hard proof, but it's a little clever.
 2. Then, we'll try to get a handle on what modeling error looks like for different models. 
  - This'll be a visualization exercise to help you develop some intuition.
  - This intuition comes in very handy because a lot of people don't have it.^[Blame us teachers. In a lot of classes, you do the math assuming you're using a correctly specified model. Then you learn some techniques for choosing a model that isn't obviously misspecified. The message people get is to put a little effort into choosing a model, then act as if it were correctly specified. That doesn't work.]
  - With even pretty basic intuition, you can catch a lot of mistakes and get them fixed.

# Modeling and Sampling Error

## Breaking Down the Problem

```{r}
#| fig.width: 7
#| layout-ncol: 2
#| fig-cap: 
#|  - 'A sample drawn with replacement from a population.'
#|  - 'That population.'
sam |> lsq.plot(y~w*x)
pop.lsq.plot
```


- The summaries we've been estimating are complex, but made up of simple pieces.
- In particular, they're all linear combinations of subpopulation means $\mu(w,x)$. 

$$
\theta = \frac{1}{m}\sum_{j=1}^m \qty{ \mu(1,x_j) -  \mu(0,x_j) } \qfor \mu(w,x)=\frac{1}{m_{w,x}}\sum_{j:w_j=w,x_j=x} y_j
$$

- Let's start by looking at the error we get when we estimate a single one of these. 
- We'll break it down into two pieces. 
  - One piece is the error of a *population least squares predictor* $\inpink{\tilde\mu}$.
  - The other is the difference between that and our actual predictor $\hat\mu$. 

$$
\hat\mu(w,x) - \mu(w,x) = \qty{\hat\mu(w,x) - \inpink{\tilde\mu(w,x)}} + \qty{\inpink{\tilde\mu(w,x)} - \mu(w,x)}
$$

## Estimating Subpopulation Means 

```{r}
#| cache: true
grid = expand.grid(w=0:1, x=c(20,30,40))
mutilde=function(w,x) { predict(lm(y~w*x, data=pop), newdata=data.frame(w=w,x=x)) }
mu=function(w,x) { predict(lm(y~w*factor(x), data=pop), newdata=data.frame(w=w,x=x)) }
mu.hat.grid = samples |> map(function(sam) { 
  newgrid=grid 
  newgrid$muhat=predict(lm(y~w*x, data=sam), newdata=grid)
  newgrid 
}) |> bind_rows()
```
```{r}
#| layout-ncol: 3
#| layout-nrow: 2
#| fig-width: 3.5

plothist = function(w,x, color=color) { 
  ggplot() + 
    geom_histogram(aes(x=muhat, y=after_stat(density)), 
      alpha=.2, bins=50, fill=color, data=mu.hat.grid[mu.hat.grid$w==w & mu.hat.grid$x==x,]) +
    geom_vline(aes(xintercept=mutilde(w,x)), alpha=.2, linewidth=2, color=pink) +
    geom_vline(aes(xintercept=mu(w,x)), alpha=.2, linewidth=2, color='green')
}

plothist(0,20, color=groupa)
plothist(0,30, color=groupa)
plothist(0,40, color=groupa)
plothist(1,20, color=groupb)
plothist(1,30, color=groupb)
plothist(1,40, color=groupb)
```

$$
\hat\mu(w,x) - \mu(w,x) = \qty{\hat\mu(w,x) - \inpink{\tilde\mu(w,x)}} + \qty{\inpink{\tilde\mu(w,x)} - \mu(w,x)}
$$

- What we see looks a lot like what we saw when we were estimating $\hat\mu$ using the 'all functions' model.
- But there's one significant difference: bias. 
  - The sampling distribution of $\hat\mu(w,x)$ isn't centered on the population mean $\mu(w,x)$.
  - It's centered on the population least squares predictor $\inpink{\tilde\mu(w,x)}$.
$$
\E[\hat\mu(w,x)] = \tilde\mu(w,x) \qfor \tilde\mu = \argmin_{m \in \model} \sum_{j=1}^m \qty{y_j - m(w_j,x_j)}^2
$$

- That's what we're going to prove now. 

**Q.** If the distribution of $\hat\mu$ is centered on the population least squares predictor $\tilde\mu$, \
why is $\hat\mu(w,x)$ unbiased when we use the 'all functions model'?


## Key Idea: Orthogonality of Residuals 
```{r}
#| layout-ncol: 2
#| fig-width: 7
sam |> lsq.plot(y~1, residual.alpha=.1)
sam |> lsq.plot(y~w, residual.alpha=.1)
```
$$
\textcolor{orange}{\hat\varepsilon_i} = Y_i - \hat\mu(W_i,X_i) \qquad \text{ are the \textbf{residuals}}
$$

::: {.panel-tabset}
### Properties in Familiar Cases
- When we use least squares to choose a ...
  - constant prediction, the residuals sum to zero. 
  - constant-within-groups prediction, the residuals sum to zero within groups. 
$$
\begin{aligned}
\hat\mu &= \argmin_{\text{constants} \ m} \sum_{i=1}^n \qty{ Y_i - m }^2 \qqtext{ satisfies } \\
0 &=\frac{d}{dm}\mid_{m=\hat\mu} \sum_{i=1}^n \qty{ Y_i - m }^2 = \sum_{i=1}^n 2 \times \qty{ Y_i - \hat\mu(X_i) } \times -1 = -2\sum_{i=1}^n \hat \varepsilon_i 
\end{aligned}
$$

### A Generalization

- The residuals are *orthogonal* to the predictions of every function in our model. 
$$
\begin{aligned}
0 &= \sum_{i=1}^n \hat\varepsilon_i m(W_i, X_i) && \qqtext{ for all } m \in \model
  &= \underset{\text{vector of residuals}}{\vec{\hat\varepsilon}} \cdot \underset{\text{vector of predictions}}{\vec{m(W,X)}} && \qfor \vec{m(W,X)} = [ m(W_1,X_1), \ldots, m(W_n,X_n) ]
\end{aligned}
$$

- For the constant model, that's the same as summing to zero. 
- For the constant-within-groups model, that's the same as summing to zero within groups.

### Why It's the Same

$$
\begin{aligned}
0 &= \sum_{i=1}^n \hat\varepsilon_i m && \qqtext{ for all constants } m \text{ if and only if } \\
0 &= \sum_{i=1}^n \hat\varepsilon_i.
\end{aligned}
$$

$$
\begin{aligned}
0 &= \sum_{i=1}^n \hat\varepsilon_i m(W_i) && \qqtext{ for all functions } m(w) 
\text{ if and only if }  \\
0 &= \sum_{i=1}^n \hat\varepsilon_i m(W_i) \\ 
  &= \sum_{w}\sum_{i:W_i=w} \hat\varepsilon_i m(w) \\
  &= \sum_{w} m(w) \inteal{\sum_{i:W_i=w} \hat\varepsilon_i}  && \text{ is a linear combination of \inteal{zeros} }.
\end{aligned}
$$

:::

## Orthogonality of Residuals in General
```{r}
#| fig.width: 7
#| layout-ncol: 2
sam |> lsq.plot(y~w*x, residual.alpha=.5)
```

- To prove this is true in general, we'll think about the optimality of $\hat\mu$ *one direction at a time*.
$$
\begin{aligned}
\hat\mu &\qqtext{minimizes} \sum_{i=1}^n \qty{ Y_i - m(W_i,X_i) }^2 &&\qqtext{ over } m \in \model \implies \\
\hat\mu &\qqtext{minimizes} \sum_{i=1}^n \qty{ Y_i - m(W_i,X_i) }^2 &&\qqtext{ over } m \in \model_1 \subseteq \model \\
        & \text{and in particular} && \qqtext{ over } m \in \model_1 = \qty{ \hat\mu + t m \ : \ t \in \mathbb{R} } \qqtext{ for any } m \in \model
\end{aligned}
$$

- Writing mean squared error as a function of $t$ in this model ... 
  - We can observe that it's minimized at $t=0$, since that's when $\hat\mu + tm=\hat\mu$
  - And we can set the derivative to zero there to get the orthogonality condition we want.
  
::: fragment

$$
\begin{aligned}
0 &= \frac{d}{dt}\mid_{t=0} \sum_{i=1}^n \qty{ Y_i - \hat\mu(W_i,X_i) - t m(W_i,X_i) }^2 \\
  &= \sum_{i=1}^n 2 \times \qty{ Y_i - \hat\mu(W_i,X_i) } \frac{d}{dt}\mid_{t=0} \qty{ Y_i - \hat\mu(W_i,X_i) - t m(W_i,X_i) }^2 \\
  &= \sum_{i=1}^n 2 \times \qty{ Y_i - \hat\mu(W_i,X_i) } \times -m(W_i,X_i) \\
  &= -2 \hat\varepsilon_i \dot m(W, X)
\end{aligned}
$$

:::

## Orthogonality of Population Residuals
```{r}
#| layout-ncol: 2
#| fig-width: 7
sam |> lsq.plot(y~w*x, residual.alpha=.5)
plot.pop |> lsq.plot(y~w*x, residual.alpha=.2)
```

- We can define *population residuals* $\tilde\varepsilon_j = y_j - \tilde\mu(w_j,x_j)$ analogously.
  - These are the residuals we get when we choose a predictor $\tilde\mu$ by least squares in the whole population.
  - And by the same argument, they're are orthogonal to every function in our model *in a population sense*.
$$
\begin{aligned}
0 &= \veryfaint{\frac{1}{n}}\sum_{i=1}^n \hat\varepsilon_i  m(W_i,X_i)  
&& \qqtext{ for all } m \in \model \qqtext{ where } \hat\varepsilon_i = Y_i - \hat\mu(W_i,X_i) \\
0 &= \veryfaint{\frac{1}{m}}\sum_{j=1}^m \tilde\varepsilon_j m(w_j,x_j) 
&& \qqtext{ for all } m \in \model \qqtext{ where } \tilde\varepsilon_j = y_j - \tilde\mu(w_j,x_j) \\
  &= \E \qty{ \frac{1}{n}\sum_{i=1}^n \qty{ Y_i - \tilde\mu(W_i,X_i) } m(W_i,X_i) } 
\end{aligned}
$$

- We've set out to prove that the population least squares predictor $\tilde\mu$ is the conditional expectation of the sample least squares predictor $\hat\mu$. 
- To do this, we'll combine these two orthogonality properties in a clever way.

## Relating $\hat\mu$ and $\tilde\mu$ (1/4)

**Claim**. 
$$
\begin{aligned}
\tilde\mu(w,x) &= \mu_E(w,x) 
&& \qfor \mu_E(w,x) = \E[\hat\mu(W_i,X_i) \mid W_i=w, X_i=x] \\
&&& \qand \tilde\mu = \argmin_{m \in \model} \sum_{j=1}^m \qty{ y_j - m(w_j,x_j) }^2
\end{aligned}
$$

**Sketch**.

1. We'll show that orthogonality of residuals implies a kind of orthogonality of the difference $\hat\mu - \tilde\mu$.
$$
0=\E\left[ \lcb{ \hat\mu(W_i,X_i) - \tilde\mu(W_i,X_i) } m(W_i,X_i) \right] \qfor \text{ all } m \in \model
$$
2. We'll plug in a clever choice of $m$: $m=\mu_E - \tilde \mu$.
$$
0=\E\left[ \lcb{ \hat\mu(W_i,X_i) - \tilde\mu(W_i,X_i) } \qty{ \mu_E(W_i,X_i) - \tilde\mu(W_i, X_i) } \right] 
$$
3. We'll use the law of iterated expectations to show that this is a mean-squared-difference. 
$$
0 = \E\left[  \lcb{ \mu_E(W_i,X_i) - \tilde\mu(W_i, X_i) }^2 \right]
$$
- That implies $\mu_E(W_i,X_i) - \tilde\mu(W_i, X_i)=0$. 
- That's how squares work. If something isn't zero, its mean-square is positive.

## Relating $\hat\mu$ and $\tilde\mu$ (2/4)

**Claim**. 
$$
\begin{aligned}
\tilde\mu(w,x) &= \mu_E(w,x) 
&& \qfor \mu_E(w,x) = \E[\hat\mu(W_i,X_i) \mid W_i=w, X_i=x] \\
&&& \qand \tilde\mu = \argmin_{m \in \model} \sum_{j=1}^m \qty{ y_j - m(w_j,x_j) }^2
\end{aligned}
$$

**Step 1**. 

- From the orthogonality of the sample residuals, we have 

$$ 
\begin{aligned}
0 &= \frac{1}{n}\sum_{i=1}^n \lcb{ Y_i -\hat\mu(W_i,X_i) } m(W_i,X_i)
\qqtext{ and therefore } \\
0 &= \E \left[ \frac{1}{n}\sum_{i=1}^n \lcb{ Y_i -\hat\mu(W_i,X_i) } m(W_i,X_i) \right] \\ 
  &=  \E\left[ \lcb{ Y_i - \hat\mu(W_i,X_i) } m(W_i,X_i) \right]. 
\end{aligned}
$$

- From the orthogonality of the population residuals, we have ...

$$
\begin{aligned}
0 &= \E\left[ \lcb{ Y_i - \tilde \mu(W_i,X_i)} m(W_i,X_i) \right] \\
\end{aligned}
$$

- Subtracting the first from the second, we get ...

$$
\begin{aligned}
0 &= \E\left[ \pcb{ \lcb{ Y_i- \tilde\mu(W_i,X_i) } - \lcb{ Y_i - \hat\mu(W_i,X_i) }} m(W_i,X_i) \right] \\
  &= \E\left[ \pcb{ \hat\mu(W_i,X_i) - \tilde\mu(W_i,X_i) } m(W_i,X_i) \right]
\end{aligned}
$$

## Relating $\hat\mu$ and $\tilde\mu$ (3/4)

**Claim**. 
$$
\begin{aligned}
\tilde\mu(w,x) &= \mu_E(w,x) 
&& \qfor \mu_E(w,x) = \E[\hat\mu(W_i,X_i) \mid W_i=w, X_i=x] \\
&&& \qand \tilde\mu = \argmin_{m \in \model} \sum_{j=1}^m \qty{ y_j - m(w_j,x_j) }^2
\end{aligned}
$$

**Step 2**.
$$
\begin{aligned}
0 &= \E\qty[ \lcb{ \hat\mu(W_i,X_i) - \tilde\mu(W_i,X_i) }m(W_i,X_i) ] \qqtext{ for all } m \in \model \implies \\
0 &= \E\qty[ \lcb{ \hat\mu(W_i,X_i) - \tilde\mu(W_i,X_i) } \lcb{ \mu_E(W_i,X_i) - \tilde\mu(W_i,X_i) } ]
\end{aligned}
$$

- This seems trivial, but to make sure this implication is valid, we have to check something.
  - We have to check that $m=\mu_E - \tilde\mu$ is in the model $\model$.
  - This ultimately boils down to linearity of the model and linearity of expectation.
  - $\mu_E$ is a (probability-weighted) linear combination of functions in the model, so it's in the model.
$$
\begin{aligned}
\mu_E &= \sum_{\substack{(w_1,x_1,y_1) \ldots (w_n,x_n,y_n) \\ w_i=w, x_i=x }} 
\hat\mu \times P\qty{ (W_1,X_1,Y_1) \ldots (W_n,X_n,Y_n) = (w_1,x_1,y_1) \ldots (w_n,x_n,y_n) \mid W_i=w, X_i=x} \\
&\qfor \hat\mu = \argmin_{m \in \model} \sum_{i=1}^n \qty{ y_i - m(w_i,x_i) }^2
\end{aligned}
$$

## Relating $\hat\mu$ and $\tilde\mu$ (4/4)
**Claim**. 
$$
\begin{aligned}
\tilde\mu(w,x) &= \mu_E(w,x) 
&& \qfor \mu_E(w,x) = \E[\hat\mu(W_i,X_i) \mid W_i=w, X_i=x] \\
&&& \qand \tilde\mu = \argmin_{m \in \model} \sum_{j=1}^m \qty{ y_j - m(w_j,x_j) }^2
\end{aligned}
$$

**Step 3**.
$$
\begin{aligned}
0 
&\explain{=}{Steps 1-2} 
\E\left[ \lcb{ \hat\mu(W_i,X_i) - \tilde\mu(W_i,X_i) } \lcb{ \mu_E(W_i,X_i) - \tilde\mu(W_i,X_i) } \right] \\
&\explain{=}{Law of Iterated Expectations} 
  \E\left[ \color{blue}\E\left[ \color{black}  \left\{ \hat\mu(W_i,X_i) - \tilde\mu(W_i,X_i) \right\} \left\{ \mu_E(W_i,X_i) - \tilde\mu(W_i,X_i) \right\} \color{blue} \mid W_i, X_i\right] \color{black} \right] \\
&\explain{=}{Linearity of Conditional Expectations} 
  \E\left[  \left\{ \color{blue}\E\left[ \color{black} \hat\mu(W_i,X_i) \color{blue} \mid W_i, X_i\right] \color{black} - \tilde\mu(W_i,X_i) \right\} \left\{ \mu_E(W_i,X_i) - \tilde\mu(W_i,X_i) \right\}  \right] \\
&\explain{=}{Recognizing $\mu_E$'s definition} 
\E\qty[ \qty{ \color{blue} \mu_E(W_i,X_i) \color{black} - \tilde\mu(W_i,X_i) }\qty{ \mu_E(W_i,X_i)  - \tilde\mu(W_i,X_i) } ] \
&\explain{=}{Recognizing a square} 
\E\left[ \left\{ \mu_E(W_i,X_i) - \tilde\mu(W_i,X_i) \right\}^2 \right] 
\end{aligned}
$$

- That concludes the proof. 
  - If $\mu_E(W_i,X_i) - \tilde\mu(W_i,X_i)$ weren't always zero ...
  - then the expected value of its square would be positive, not zero.


## Implications for Complex Targets

```{r}
#| fig.width: 7
#| layout-ncol: 2
#| fig-cap: 
#|  - 'A sample drawn with replacement from a population.'
#|  - 'That population.'
sam |> lsq.plot(y~w*x)
pop.lsq.plot
```


- Let's consider a more complex estimation target, e.g., the adjusted difference in means $\hat\Delta_{\text{all}}$.
  - This, like most of our targets, is a linear combination of column means.
  - To keep things simple, let's suppose that the weights are non-random.
  - e.g., if we were able to get the distribution of $w,x$ in the population from a census or voter file.
- This is how we've been estimating it. 
$$
\begin{aligned}
\hat\theta 
&= \frac{1}{m}\sum_{j=1}^m \qty{ \hat\mu(1,x_j) -  \hat\mu(0,x_j) } 
&&\qfor \hat\mu = \argmin_{m \in \model} \sum_{i=1}^n \qty{ Y_i - m(X_i) }^2 \\
&= \sum_{w,x} \alpha(w,x) \hat\mu(w,x) 
&&\qfor \alpha(w,x) = \begin{cases} p_{x} & \text{ for } w=1 \\
                                    -p_{x} & \text{ for } w=0. \end{cases}
\end{aligned}
$$

- We can decompose our estimator's error by comparing to $\tilde\mu$ term-by-term. 
$$
\begin{aligned}
\hat\theta-\theta 
&= \sum_{w,x} \alpha(w,x) \qty{ \hat\mu(w,x) - \mu(w,x) } \\ 
&= \underset{\faint{\text{sampling error}}}{\sum_{w,x} \alpha(w,x) \qty{ \hat\mu(w,x) - \tilde\mu(w,x) }} + \underset{\faint{\text{modeling error}}}{\sum_{w,x} \alpha(w,x) \qty{ \tilde\mu(w,x) - \mu(w,x) }} \\ 
\end{aligned}
$$

- This breaks our error down into two pieces. 
  - The *sampling error* has mean zero. Why?
  - The *modeling error* is deterministic. It's our estimator's bias. 

## What Do We Do With This?
```{r}
#| layout-ncol: 2
#| fig-width: 7
theta.sampling.dist.plot
pop.lsq.plot
```
$$
\begin{aligned}
\hat\theta-\theta 
&= \underset{\faint{\text{sampling error}}}{\sum_{w,x} \alpha(w,x) \qty{ \hat\mu(w,x) - \tilde\mu(w,x) }} + \underset{\faint{\text{modeling error}}}{\sum_{w,x} \alpha(w,x) \qty{ \tilde\mu(w,x) - \mu(w,x) }} \\ 
\end{aligned}
$$

- We can estimate the distribution of the sampling error as usual.
  - Meaning using the bootstrap.
  - There are variance formulas, but they're a little complicated.
- If there were no modeling error, this'd mean we could calibrate interval estimates for 95% coverage. With modeling error, we can't do that. 
- This leaves us with some bad options. 
  1. We act as if there's no modeling error. 
  2. We act as if what we really wanted to estimate was $\tilde\theta$.
- Because these choices are so popular, it's important to understand the implications of doing them.
- But we do have some good options.
  1. Use the all-functions model, so there really isn't any modeling error.
  2. Use *inverse probability weighting*, a special way estimation-target-specific way of estimating $\mu$ that can eliminate or at least significant reduce modeling error. That's what we'll discuss next class.
  
## Putting It All Together: An Example

```{r}
#| layout-ncol: 3
#| layout-nrow: 2
#| fig-width: 3.5

plothist(0,20, color=groupa)
plothist(0,30, color=groupa)
plothist(0,40, color=groupa)
plothist(1,20, color=groupb)
plothist(1,30, color=groupb)
plothist(1,40, color=groupb)
```

- Above, we see the sampling distribution of $\hat\mu(w,x)$ for $w=0,1$ and $x=20,30,40$.
- Roughly what's the bias of $\hat\theta$? 
  - Could you have guessed that from the plot of $\hat\mu$ on top of the sample? 
  - Or did you need all the population information displayed here?

$$
\hat\theta = \frac{1}{m}\sum_{j=1}^m \qty{ \hat\mu(1,x_j) -  \hat\mu(0,x_j) } 
$$

```{r}
#| fig.width: 7
#| layout-ncol: 2
#| fig-cap: 
#|  - 'A sample drawn with replacement from a population.'
#|  - 'That population.'
sam |> lsq.plot(y~w*x)
pop.lsq.plot
```

## Quiz 

[Here](https://qtm285-1.github.io/assets/homework/week3-homework.html#fig-3-sampling-distributions) we see the sampling distributions of three estimators of $\theta$ at three different sample sizes.

Which one *couldn't* be the sampling distribution of an estimator $\hat\theta$ like this?

$$
\hat\theta = \frac{1}{m}\sum_{j=1}^m \qty{ \hat\mu(1,x_j) -  \hat\mu(0,x_j) }  \qfor \hat\mu = \argmin_{m \in \model} \sum_{i=1}^n \qty{ Y_i - m(X_i) }^2
$$

# Modeling Error

## Setup

::: {.panel-tabset}
### Population
```{webr}
#| autorun: true
x = rep(c(20,30,40), each=1000)
pi.pop = function(x) { case_when(x < 25 ~ .25, x < 35 ~ .5, TRUE ~ .75) } 
mu.pop = function(w,x) case_when(
  w==1 & x <  25  ~ 10, 
  w==1 & x <= 35  ~ 10,
  w==1            ~ 20, 
  w==0 & x <  25  ~ 5, 
  w==0 & x <= 35  ~ 10,
  w==0            ~ 10)

w = 0 + (runif(length(x)) < pi.pop(x))
y = rpois(length(x), mu.pop(w,x))
pop = data.frame(w=w, x=x, y=y)
```

### Sample
```{webr}
#| autorun: true
m = nrow(pop)
n = 2000
set.seed(5)
J = sample(1:m, size=n, replace=TRUE)
sam = pop[J,]
```

### Plotting Code
```{webr}
#| autorun: true
prop.scale = 40
default.shift = .5
mean_sd = function(x, mult = 1) { mean_se(x, mult = mult*sqrt(length(stats::na.omit(x)))) }
plotX = function(w,x, shift=default.shift) { ifelse(w==0, x-shift, x+shift) }
lsq.plot = function(sam, formula=NULL, dot.size=1, shift=default.shift, residual.alpha=0) { 

  sam$jitterX = runif(nrow(sam), -1, 1)*shift
  base.plot = ggplot(sam) + 
    geom_point(aes(x=plotX(w,x)+jitterX,y=y,color=factor(w)), 
      alpha=.05, size=dot.size) + 
    stat_summary(aes(x=plotX(w,x), y=y), 
      fun.data=mean_sd, geom='pointrange', alpha=.2) +
    geom_bar(aes(x=x, y=prop.scale*after_stat(prop)), 
      fill='purple', alpha=.1, linewidth=.05) +
    scale_y_continuous(sec.axis = sec_axis(~./prop.scale, name='proportion')) +
    guides(color='none') 
    
    if(is.null(formula)) { base.plot } else { 
        model = lm(formula, data=sam)
        gridx = seq(-10, 50, length.out=1000)
        muhat = function(w,x) predict(model, newdata=data.frame(w=w,x=x))   
        the.plot = base.plot + 
          geom_point(aes(x=plotX(w,x), y=muhat(w,x), color=factor(w)), alpha=.7, shape='square') +
          geom_line(aes(x=plotX(w,gridx), y=muhat(w,gridx), color=factor(w)), alpha=.7) 
        if(residual.alpha > 0) {
          the.plot = the.plot + 
            geom_segment(aes(x=plotX(w,x)+jitterX, xend=plotX(w,x), y=y, yend=muhat(w,x)), 
              alpha=residual.alpha, color='orange', linewidth=.05)
        }
        the.plot 
    }
}
```

### Plot Styline

```{webr}
bgcolor = "#E7F0EC"
class.theme = theme(plot.background = element_rect(fill = bgcolor, colour = NA),
        panel.background = element_rect(fill = bgcolor, colour = NA),
        legend.background = element_rect(fill=bgcolor, colour = NA),
        legend.box.background = element_rect(fill=bgcolor, colour = NA),
        legend.key = element_rect(fill=bgcolor, colour = NA),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x  = element_text(colour = "#aaaaaa"),
        axis.text.y  = element_text(colour = "#aaaaaa"),
        axis.title.x  = element_text(colour = "#aaaaaa"),
        axis.title.y  = element_text(colour = "#aaaaaa", angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()
```

:::

## Example 

```{webr}
#| autorun: true
model.formula = y~w*x
sam |> lsq.plot(y~w*x)
```

```{webr}
estimate = function(sam, formula) {
  X = sam$x
  W = sam$w
  model = lm(formula, data=sam)
  muhat = function(w,x) predict(model, newdata=data.frame(w=w,x=x)) 
  mean(muhat(1, X) - muhat(0, X))
}
all.functions.formula = y~w*factor(x)

data.frame(
  thetahat   = sam |> estimate(model.formula),
  thetatilde = pop |> estimate(model.formula), 
  theta      = pop |> estimate(all.functions.formula)) 
```
