---
title: "QTM 220: Lecture 12"
subtitle: "Causal Inference with a Multivalued Treatment"
format: 
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---


```{r load-libraries}
library(purrr)
library(tibble)
library(dplyr)
library(magrittr)
library(stringr)
library(ggplot2)
library(htmltools)
library(kableExtra)
library(scales)

percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }
```

```{r ggplot-themes}
fontsize = 16
textcolor = "#888888"

class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
        panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x  = element_text(colour = textcolor, size=fontsize),
        axis.text.y  = element_text(colour = textcolor, size=fontsize),
        axis.title.x  = element_text(colour = textcolor),
        axis.title.y  = element_text(colour = textcolor, angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
              axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
           panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(simplified.theme)
```


```{r table-styling}
ellipsis <- function(df, nhead=5L, ntail=2) {
    rownames(df) = as.character(1:nrow(df))
    stopifnot("data.frame" %in% class(df))
    els <- rep(NA, ncol(df)) %>% 
        matrix(nrow=1, dimnames=list(NULL, names(df))) %>% 
        data.frame(stringsAsFactors=FALSE) %>% 
        tibble::as_tibble()
    rownames(els) = "⋮"
    dplyr::bind_rows(head(df, nhead), els, tail(df, ntail))
}
```


## Today

::: {.hidden}
$$
\newcommand{\alert}[1]{\textcolor{red}{#1}}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\V}{V}
\DeclareMathOperator{\Var}{\V}
\newcommand{\inred}[1]{\textcolor{red}{#1}}
\newcommand{\inblue}[1]{\textcolor{blue}{#1}}
$$ 
:::

- We're going to talk about causal inference with a multivalued treatment.
- Our context will be a randomized experiment with three treatments.
- A large employer is experimenting with a new benefit for their employees: subsized gym memberships.
- They're randomly selecting employees to participate in a trial program.
- Then randomizing these participants into three plans.
  - Plan 1: No gym membership. It costs them \$0/month.
  - Plan 2: A gym membership that covers entry but not classes. It costs \$50/month.
  - Plan 3: A gym membership that covers entry but and classes. It costs \$100/month.
- And surveying them to see how often they go to the gym, e.g. in days/month.
- They'll be able to use this information to decide what plan to offer everyone next year.

## How We Think

- We'll think about this using potential outcomes.
- We imagine, for each employee, that there's a number of times they'd go to the gym given each plan.
  - $y_j(0)$: The number of days/month they'd go to the gym if they had no membership.
  - $y_j(50)$: The number of days/month they'd go to the gym if they had a \$50 membership.
  - $y_j(100)$: The number of days/month they'd go to the gym if they had a \$100 membership.
- And we'll think about the effect of each plan on each employee.
  - $\tau_j(50)  = y_j(50)  - y_j(0)$: The effect of a \$50 gym membership vs none.
  - $\tau_j(100) = y_j(100) - y_j(50)$: The effect of a \$100 gym membership vs \$50.

$$
\small{
\begin{array}{c|ccc|cc}
j & y_j(0) & y_j(50) & y_j(100) & \tau_j(50) & \tau_j(100) \\
1 & 5  & 10 & 10 & 5 & 0 \\
2 & 10 & 10 & 10 & 0 & 0 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
327 & 0 & 0  &  20 & 0 & 20 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
358 & 15 & 18 & 20 & 3 & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
m & 10 & 10 & 5  & 0 & -5 \\
\end{array}
}
$$

## What We Want To Know

- What we want to know is the population-level average treatment effect.
  - i.e. the average increase in monthly gym visits if we offered these plans to every employee.

$$
\small{
\begin{aligned}
\tau(50)  &= \frac{1}{m} \sum_{j=1}^m \tau_j(50)  && \quad \text{this is the first increment---the effect of \$50 vs. \$0} \\
\tau(100) &= \frac{1}{m} \sum_{j=1}^m \tau_j(100) && \quad \text{this is the second increment---the effect of \$100 vs. \$50} \\
\end{aligned}
}
$$

- What's hard about this?
  - *Sampling*. We only observe information about a fraction of these people.
  - *Missing Data*. For that fraction, we only observe one of the potential outcomes.


$$
\small{
\begin{array}{c|ccc|cc}
j & y_j(0) & y_j(50) & y_j(100) & \tau_j(50) & \tau_j(100) \\
\color{gray}1 & \color{gray}5  & \color{gray}10 & \color{gray}10 & \color{gray}5 & \color{gray}0 \\
2 & 10 & \color{gray}10 & \color{gray}10 & \color{gray}0 & \color{gray}0 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
327 & \color{gray}0 & \color{gray}0  &  20 & \color{gray}0 & \color{gray}20 \\
\color{gray}\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
358 & \color{gray}15 & 18 & \color{gray}20 & \color{gray}3 & \color{gray}2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{gray}m & \color{gray}10 & \color{gray}10 & \color{gray}5  & \color{gray}0 & \color{gray}-5 \\
\end{array}
}
$$

## Sampling

:::: columns
::: column
### Our Population 
$$
\small{
\begin{array}{c|ccc|cc}
j & y_j(0) & y_j(50) & y_j(100) & \tau_j(50) & \tau_j(100) \\
\color{gray} 1 & \color{gray}5  & \color{gray}10 & \color{gray}10 & \color{gray}5 & \color{gray}0 \\ 
\color{blue}2 & \color{blue}10 & \color{blue}10 & \color{blue}10 & \color{blue}0 & \color{blue}0 \\
\vdots & \vdots  & \vdots & \vdots & \vdots & \vdots \\
\color{purple}327 & \color{purple}15 & \color{purple}18 & \color{purple}20 & \color{purple}3 & \color{purple}2 \\
\vdots & \vdots  & \vdots & \vdots & \vdots & \vdots \\
\color{green}358 & \color{green}0 & \color{green}0  &  \color{green}20 & \color{green}0 & \color{green}20 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{gray}m & \color{gray}10 & \color{gray}10 & \color{gray}5  & \color{gray}0 & \color{gray}-5 \\
\end{array}
}
$$

:::
::: column
### Our Sample
$$
\small{
\begin{array}{c|ccc|cc}
i &  Y_i(0) & Y_i(50) & Y_i(100) & \tau_i(50) & \tau_i(100) \\
\color{blue}1 & \color{blue}10 & \color{blue}10 & \color{blue}10 & \color{blue}0 & \color{blue}0 \\
\color{green}2 & \color{green}0 & \color{green}0  &  \color{green}20 & \color{green}0 & \color{green}20 \\
\vdots & \vdots  & \vdots & \vdots & \vdots \\
\color{purple}n & \color{purple}15 & \color{purple}18 & \color{purple}20 & \color{purple}3 & \color{purple}2 \\
\end{array}
}
$$

:::
::::

- We don't want to trouble all of our employees with a survey, so we sample a subset of them.
- Each employee in our sample is drawn, with replacement, from our population of employees---all $m$ of them.
- That means each row of our *sample table* is a random row from our *population table*.
  - I've colored them so we can track them from population into the sample.

## Missing Data

- We assign a treatment to each sampled employee.
  - That tells us which potential outcome we observe.
  - It's the one corresponding to the treatment that actually happens: $Y_i=Y_i(D_i)$.

$$
\small{
\begin{array}{c|cc|ccc|cc}
i & D_i & Y_i   & Y_i(0) & Y_i(50) & Y_i(100) & \tau_i(50) & \tau_i(100) \\
\color{blue}1   &  \color{blue}0 & \color{blue}10 & \color{blue}10 & \color{gray}10 &  \color{gray}10 &  \color{gray}0 &  \color{gray}0 \\
\color{green}2  &  \color{green}100& \color{green} 20 &  \color{gray}0  &  \color{gray}0  &  \color{green}20 &  \color{gray}0 &  \color{gray}20 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{purple}n &  \color{purple}50& \color{purple}18  & \color{gray}15 &  \color{purple}18 &  \color{gray}20 &  \color{gray}3 &  \color{gray}2 \\
\end{array}
}
$$

- We'll do this *randomly*, e.g. by rolling a die. In effect, this is just *another sampling step*.
- Think of the group that receives each treatment as being drawn from our population by a two-stage process.
  - First, roll an $m$-sided die to choose a row from our population table *to consider*.
  - Then, roll a 3-sided die to decide whether we're included in the subsample with $D_i=d$.


:::: columns
::: {.column width="30%"}
$$
\small{
\begin{array}{c|cc}
i & D_i & Y_i    \\
\color{blue}1   &  \color{blue}0 & \color{blue}10 \\
\vdots & 0 & \vdots  \\
\end{array}
}
$$
:::
::: {.column width="30%"}
$$
\small{
\begin{array}{c|cc}
i & D_i & Y_i    \\
\vdots & 50 & \vdots \\
\color{purple}n &  \color{purple}50& \color{purple}18 \\
\end{array}
}
$$
:::
::: {.column width="30%"}
$$
\small{
\begin{array}{c|cc}
i & D_i & Y_i    \\
\vdots & 100 & \vdots \\
\color{green}2  &  \color{green}100& \color{green} 20  \\
\end{array}
}
$$
:::
::::



## Getting By With What We Have


- If we had all the potential outcomes... 
- We could approximate a population-level average treatment effect by a sample average.

$$
\small{
\frac{1}{m} \sum_{j=1}^m y_j(50) - y_j(0) \approx \frac{1}{n} \sum_{i=1}^n Y_i(50) - Y_i(0) 
}
$$

- This works because of the law of large numbers. Sample averages approximate population averages.
- By the same logic, we can do the same using *each treatment group*. 
- When we randomize, these are random samples from the population, too.

$$
\small{
\frac{1}{m} \sum_{j=1}^m y_j(50) - y_j(0) \approx \frac{1}{N_d} \sum_{i:D_i=d}^n Y_i(50) - Y_i(0) = \frac{1}{N_d} \sum_{i:D_i=d} Y_i(50) - \frac{1}{N_d}\sum_{i:D_i=d} Y_i(0)
}
$$ 

- This still isn't something we can use. We don't have all the potential outcomes. But it just takes one tweak. 
- We use *different subsamples* to approximate the averages of diferent potential outcomes. 
- In particular, we use the subsample in which that potential outcome is observed.

$$
\small{
\frac{1}{m} \sum_{j=1}^m y_j(50) - y_j(0) \approx \frac{1}{N_d} \sum_{i:D_i=50} Y_i(50) - \frac{1}{N_0}\sum_{i:D_i=0} Y_i(0)  = \frac{1}{N_d} \sum_{i:D_i=50} Y_i - \frac{1}{N_0}\sum_{i:D_i=0} Y_i
}
$$


## Formal Identification

- Here, in formal language, is the crux of this argument. As a result of sampling and randomization, ...
- Subsample averages of realized outcomes are *unbiased and consistent* estimators of population averages of potential outcomes.
- That's because *randomized assignment* ensures that the treatment $D_i$ is *independent* of the potential outcomes.

$$
\small{
\begin{aligned}
\frac{1}{m}\sum_j y_j(d) 
  &= \E[Y_i(d)]            && \text{sampling} \\
  &= \E[Y_i(d) \mid D_i=d] && \text{irrelevance of independent conditioning variables} \\
  &= \E[Y_i \mid D_i=d]=\mu(d)    && \text{$Y_i(d)=Y_i$ \ when \ $D_i=d$ } \\
&\approx \hat\mu(d) = \frac{1}{N_d}\sum_{i:D_i=d} Y_i && \text{law of large numbers}
\end{aligned}
}
$$

- If we can express what we want to estimate as a linear combination of potential outcome means...
- This means we can *rewrite it* as a linear combination of *subpopulation means*.

$$
\small{
\begin{aligned}
\tau(50) &= \frac{1}{m}\sum_j y_j(50) - y_j(0) = \mu(50) - \mu(0) \\
\tau(100) &= \frac{1}{m}\sum_j y_j(100) - y_j(50) = \mu(100) - \mu(50) \\
&\frac{1}{m}\sum_j \alpha(100)y_j(100) + \alpha(50)y_j(50) + \alpha(0)y_j(0) = \alpha(100)\mu(100) + \alpha(50)\mu(50) + \alpha(0)\mu(0) \\
\end{aligned}
}
$$

## Where This Leaves Us

$$
\small{
\begin{aligned}
\tau(50) &= \frac{1}{m}\sum_j y_j(50) - y_j(0) = \mu(50) - \mu(0) \\
\tau(100) &= \frac{1}{m}\sum_j y_j(100) - y_j(50) = \mu(100) - \mu(50) \\
&\frac{1}{m}\sum_j \alpha(100)y_j(100) + \alpha(50)y_j(50) + \alpha(0)y_j(0) = \alpha(100)\mu(100) + \alpha(50)\mu(50) + \alpha(0)\mu(0) \\
\end{aligned}
}
$$

- Once we've done this, we can put away our causal thinking.
- We know, in terms of the distribution of our observations ($D_i,Y_i$), what we want to estimate.
- And there's nothing especially causal about that. We can just use the techniques we've already learned.


## Today's Focus
```{r}
#| out-height: 300px
#| fig-asp: .3 
set.seed(3)

# Sampling 250 employees. Our population is an even mix of ...
#   'blue guys', who go 1/4 days no matter what
#   'green guys', who go more if it's free but --- unlike the ones in our table earlier --- don't care about classes

n=250
is.blue = rbinom(n,1, 1/2)
Yblue   = rbinom(n, 30, 1/4) 
Ygreen0 = rbinom(n, 30, 1/10)
Ygreen50 = rbinom(n, 30, 1/5)

Y0   = ifelse(is.blue, Yblue, Ygreen0)
Y50  = ifelse(is.blue, Yblue, Ygreen50)
Y100 = Y50   ### <--- this tells us we've got no effect.

# Randomizing by rolling two dice. R is their sum
R = sample(c(1:6), size=n, replace=TRUE) + sample(c(1:6), size=n, replace=TRUE)
D = case_match(R, 1:6~0, 7:10~50, 11:12~100)
Y = case_match(D, 0~Y0, 50~Y50, 100~Y100)

# Randomizing by rolling two dice. R is their sum
R = sample(c(1:6), size=n, replace=TRUE) + sample(c(1:6), size=n, replace=TRUE)
D = case_match(R,
  1:6~0,
  7:10~50,
  11:12~100)
Y = case_match(D,
  0~Y0,
  50~Y50,
  100~Y100)

real.data = tibble(D=D, Y=Y)
summaries = tibble(D=D, Y=Y) %>% group_by(D) %>% summarize(mean=mean(Y), sd=sd(Y), n=length(Y))

scatter = ggplot() + geom_point(aes(x=D, y=Y), color="red", data=tibble(D=D,Y=Y),
                                     position=position_jitter(w=10,h=0), alpha=.3) +
           geom_pointrange(aes(x=D, y=mean, ymin=mean-2*sd, ymax=mean+2*sd), data=summaries) +
           scale_x_continuous(breaks=c(0,50,100)) + labs(x="D", y="Y") + ylim(-3,15)
scatter
``` 

- Let's talk about estimating $\tau(100)$, the effect of a \$100 gym membership vs. a \$50 membership.
  - We really don't want to spend \$100/month if we don't have to.
  - So we're hoping it's not much better than spending \$50/month.
- The statistical challenge is we've leaned into that preference in our experiment too.
- We've designed our experiment so not many people are in the \$100 group.
- In particular, for each employee in our sample, we're rolling two dice
  - If the sum is 11 or 12, they get the \$100 plan.
  - If the sum if 7,8,9, or 10, they get the \$50 plan.
  - If the sum is 1-6, they get the \$0 plan.

## We Know How To Do This
```{r}
#| out-height: 300px
#| fig-asp: .3 
point.estimate = summaries$mean[summaries$D==100] - summaries$mean[summaries$D==50]
scatter
``` 

$$
\tau(100) = \mu(100) - \mu(50)
$$

- We can use a simple difference in subsample means as a point estimate.

$$
\hat\tau(100) = \hat\mu(100) - \hat\mu(50) = `r approx(summaries$mean[summaries$D==100])` - `r approx(summaries$mean[summaries$D==50])` \approx `r approx(point.estimate)`
$$

- And we can use an estimate of its sampling distribution to calibrate an interval estimate.

```{r}
data = tibble(D=D, Y=Y)
boot.distn = function(data, B=1000) { 
  rows = purrr::map(1:B, function(i) {
    data.boot = data[sample(1:nrow(data), replace=TRUE),]
    summaries.boot = data.boot %>% group_by(D) %>% summarize(mean=mean(Y), sd=sd(Y), n=length(Y))
    model = lm(Y ~ D, data=data.boot)
    muhat = predict(model, newdata=data.frame(D=c(0,50,100)))

    tauhat.dif = mean(data.boot$Y[data.boot$D==100]) - mean(data.boot$Y[data.boot$D==50]) 
    tauhat.slope = muhat[3]-muhat[2]
    data.frame(dif=tauhat.dif, slope=tauhat.slope)
  })
  rows %>% bind_rows() 
}
boots = boot.distn(data)
```

```{r plotting-code}
width = function(samples, center=mean(samples), alpha=.05) {
  covg.dif = function(w) { 
    actual = mean( center - w/2 <= samples & samples <= center + w/2)
    nominal = 1-alpha 
    actual - nominal  
  }
  uniroot(covg.dif, interval=c(0, 2*max(abs(samples-center))))$root
}

distribution.plot = function(boot.samples, point.estimate, w=width(boot.samples), interval.replications=0, point.replications=0) { 
  drange = c(0, dnorm(0, mean=0, sd=sd(boot.samples)))
  p = ggplot(data.frame(samples=boot.samples)) + 
    geom_histogram(aes(x=samples, y=after_stat(density)), alpha=.3, bins=50)  +
    geom_pointrange(aes(y=dnorm(2*sd(samples), mean=0, sd=sd(samples)), 
      x=point.estimate, xmin=point.estimate - w/2, xmax=point.estimate + w/2)) + 
      geom_vline(xintercept=mean(boot.samples), color='blue') + 
      geom_vline(xintercept=quantile(boot.samples, c(.025,.975)), color='blue', linetype='dotted') 
  if(interval.replications > 0) { 
    K = interval.replications
    p = p + geom_pointrange(aes(y=y, x=x, xmin=xmin, xmax=xmax), 
          data=data.frame(y=seq(min(drange), max(drange), length.out=K), 
                  x=boot.samples[1:K], xmin=boot.samples[1:K] - w/2, xmax=boot.samples[1:K] + w/2), color='purple', alpha=.2)
  }
  if(point.replications > 0) { 
    K = point.replications
    p = p + geom_point(aes(y=y, x=x), 
          data=data.frame(y=seq(min(drange), max(drange), length.out=K), 
                  x=boot.samples[1:K], xmin=boot.samples[1:K] - w/2, xmax=boot.samples[1:K] + w/2), color='purple', alpha=.2)
  }
  p
}
```

```{r}
#| out-height: 250px
#| fig.asp: .25
#| layout-ncol: 2

se = sqrt( var(Y[D==100])/sum(D==100) + var(Y[D==50])/sum(D==50) )
x = seq(point.estimate-4*se, point.estimate+4*se, length.out=1000)
normal.distribution.plot = ggplot() + 
    geom_area(aes(x=x, y=dnorm(x, mean=point.estimate, sd=se)), color='purple', fill='purple', alpha=.08) +
    geom_vline(aes(xintercept=point.estimate), color='purple', linetype='solid') +
    geom_vline(aes(xintercept=point.estimate + c(-1,1)*1.96*se), color='purple', linetype='dotted') +
    geom_pointrange(aes(y=dnorm(2*se, mean=0, sd=se), x=point.estimate, xmin=point.estimate - 1.96*se, xmax=point.estimate + 1.96*se))

distribution.plot(boots$dif, point.estimate) + xlim(min(x), max(x))
normal.distribution.plot + xlim(min(x), max(x))
``` 

## But We're Not Happy With Our Precision
```{r}
#| out-height: 250px
#| fig.asp: .25
normal.distribution.plot + xlim(min(x), max(x))
``` 

- It's worth it to spend the extra money if it gets us at least 1 day/month for the average employee.
  - And we can't rule that out.
  - But we can't 'rule it in' either. 
- We just don't have compelling evidence in one direction or the other.
- So let's look at why that is. What did we do wrong when designing this experiment? 


## Variance
```{r}
#| out-height: 250px
#| fig.asp: .25
normal.distribution.plot + xlim(min(x), max(x))
``` 

- Let's calculate the variance of our point estimate to see where the problem is.
$$
\Var\qty[\sum_d \alpha(d) \hat\mu(d)]  = \sum_d \sigma^2(d) \alpha^2(d) \times \E\qty[ \frac{1}{N_d} ]
$$

- We can use this table.

$$
\begin{array}{c|ccc}
d & 0 & 50 & 100 \\
\hline
\hat\sigma^2(d) & `r approx(var(Y[D==0]))` & `r approx(var(Y[D==50]))` & `r approx(var(Y[D==100]))` \\
\alpha^2(d) &  &  &  \\
N_d & `r approx(sum(D==0))` & `r approx(sum(D==50))` & `r approx(sum(D==100))` \\
\frac{\hat\sigma^2(d) \alpha^2(d)}{N_d} &  &  &  \\
\end{array}
$$

- What do you see?

## Variance
```{r}
#| out-height: 250px
#| fig.asp: .25
normal.distribution.plot + xlim(min(x), max(x))
``` 


$$
\begin{array}{c|ccc}
d & 0 & 50 & 100 \\
\hline
\hat\sigma^2(d) & `r approx(var(Y[D==0]))` & `r approx(var(Y[D==50]))` & `r approx(var(Y[D==100]))` \\
\alpha^2(d) & 0 & 1 & 1 \\
N_d & `r approx(sum(D==0))` & `r approx(sum(D==50))` & `r approx(sum(D==100))` \\
\frac{\hat\sigma^2(d) \alpha^2(d)}{N_d} & 0 & `r approx(var(Y[D==50])/sum(D==50))` & `r approx(var(Y[D==100])/sum(D==100))` \\
\end{array}
$$

- What I see is that we don't have enough people randomized to the $D=100$ treatment.
  - The variance we get from just that one term is $`r approx(var(Y[D==100])/sum(D==100))`$.
  - Meaning the contribution to our standard error is 
  $\sqrt{`r approx(var(Y[D==100])/sum(D==100))`} \approx `r approx(sqrt(var(Y[D==100])/sum(D==100)))`$ 
  - And since we multiply standard error by 1.96 to get an 'arm' for our interval ...
  - This ensures our arms are at least $1.96 \times `r approx(sqrt(var(Y[D==100])/sum(D==100)))`  \approx `r approx(1.96*sqrt(var(Y[D==100])/sum(D==100)))`$ days wide.

## If We Could Do It Again
```{r}
#| out-height: 300px
#| fig-asp: .3 

n = 250
D = sample(c(50,100), size=n, replace=TRUE) 
Y = case_match(D,
  0~Y0,
  50~Y50,
  100~Y100)

summaries.new = tibble(D=D, Y=Y) %>% group_by(D) %>% summarize(mean=mean(Y), sd=sd(Y), n=length(Y))
point.estimate = summaries.new$mean[summaries.new$D==100] - summaries.new$mean[summaries.new$D==50]

scatter.new = ggplot() + geom_point(aes(x=D, y=Y), color="red", data=tibble(D=D,Y=Y),
                                     position=position_jitter(w=10,h=0), alpha=.3) +
           geom_pointrange(aes(x=D, y=mean, ymin=mean-2*sd, ymax=mean+2*sd), data=summaries) +
           scale_x_continuous(breaks=c(0,50,100)) + labs(x="D", y="Y") + ylim(-3,15)
scatter.new
```

- We'd assign more people to the $D=100$ treament to get a more precise estimate.
- Here's what we'd get if we assigned people to each of our 2 *relevant* treatments with probability $1/2$.
- This does what we need it to do: it gives us a confidence interval that to the left of $1$.

```{r}
boots.ifwecould = boot.distn(tibble(D=D, Y=Y))
```

```{r}
#| out-height: 250px
#| fig.asp: .25
#| layout-ncol: 2

se = sqrt( var(Y[D==100])/sum(D==100) + var(Y[D==50])/sum(D==50) )
x = seq(point.estimate-4*se, point.estimate+4*se, length.out=1000)
normal.distribution.plot = ggplot() + 
    geom_area(aes(x=x, y=dnorm(x, mean=point.estimate, sd=se)), color='purple', fill='purple', alpha=.08) +
    geom_vline(aes(xintercept=point.estimate), color='purple', linetype='solid') +
    geom_vline(aes(xintercept=point.estimate + c(-1,1)*1.96*se), color='purple', linetype='dotted') +
    geom_pointrange(aes(y=dnorm(2*se, mean=0, sd=se), x=point.estimate, xmin=point.estimate - 1.96*se, xmax=point.estimate + 1.96*se))

distribution.plot(boots.ifwecould$dif, point.estimate) + xlim(min(x), max(x))
normal.distribution.plot + xlim(min(x), max(x))
``` 

## But We Can't

```{r}
#| out-height: 300px
#| fig-asp: .3 

model=lm(Y ~ D, data=real.data)
mu.hat = predict(model, newdata=data.frame(D=c(0,50,100)))
slope.point.estimate = mu.hat[3]-mu.hat[2]

scatter + geom_line(aes(x=c(0,50,100), y=mu.hat), color='blue', linetype='dashed')
```

- This is our data. We have to do what we can. So we get clever.
- Instead of estimating each subpopulation mean separately, we *fit a line* to the data.
- We take $\hat\mu(d)$ to be the height of the line at $d$ and $\hat\tau(d)=\hat\mu(100)-\hat\mu(50)$ as before.
- And when we bootstrap it ... 
  - We get a much narrower interval than the one we had before.
  - But it's not in the same place at all.


```{r}
#| out-height: 250px
#| fig.asp: .25
#| layout-ncol: 2 
#| fig-subcap:
#|   - "Difference in Sample Means Sampling Dist"
#|   - "Fitted-Line-Based Sampling Dist" 

real.point.estimate = mean(real.data$Y[real.data$D==100]) - mean(real.data$Y[real.data$D==50])
x = seq(slope.point.estimate-7*se, slope.point.estimate+7*se, length.out=1000)
distribution.plot(boots$dif,   real.point.estimate) + xlim(min(x), max(x)) 
distribution.plot(boots$slope, slope.point.estimate) + xlim(min(x), max(x))
```

## What's Going On?

```{r}
#| out-height: 300px
#| fig-asp: .3 

scatter + geom_line(aes(x=c(0,50,100), y=mu.hat), color='blue', linetype='dashed') + 
  coord_cartesian(ylim=range(summaries$mean) + .5*c(-1,+1))
```

- Let's zoom in. It looks like our like our blue line ...
  - *overestimates* $\mu(100)$
  - *underestimates* $\mu(50)$
  - therefore *overestimates* $\tau(100)=\mu(100)-\mu(50)$.
- That's what it looks like when we compare to the subsample means, anyway.
- And that's why we get a bigger estimate than we did *using* the subsample means.
- It looks like---although we haven't done the calculation---this line-based estimator is *biased*.

## Implications
```{r}
#| out-height: 250px
#| fig.asp: .25
#| layout-ncol: 2 
#| fig-subcap:
#|   - "Difference in Sample Means Sampling Dist"
#|   - "Fitted-Line-Based Sampling Dist" 

distribution.plot(boots$dif,   real.point.estimate) + xlim(min(x), max(x))
distribution.plot(boots$slope, slope.point.estimate)  + xlim(min(x), max(x))
```

- This isn't a disaster if all we want to know is whether our effect $\tau(100)$ is &gt;1
  - i.e, if we want to know whether the \$100 plan gives us an additional gym day over the \$50 one.
  - No matter which approach we use, we get the same answer: we don't know.
  - Our interval estimates both include numbers above and below 1.
- But if what we wanted to know was whether $\tau(100) > 0$, we'd get different answers.
  - The difference in subsample means approach would say we don't know.
  - The fitted line approach would say we do know---in fact, that $\tau(100) > 1/2$.
- And this is a problem. This is fake data, so I can tell you it's wrong. The actual effect is exactly zero.
- In the data we're looking at, the potential outcomes $Y_i(50)$ and $Y_i(100)$ are *exactly the same*.

## The Data

```{r}
#| eval: false
#| echo: true

# Sampling 250 employees. Our population is an even mix of ...
#   'blue guys', who go 1/4 days no matter what
#   'green guys', who go more if it's free but --- unlike the ones in our table earlier --- don't care about classes

n=250
is.blue = rbinom(n,1, 1/2)
Yblue   = rbinom(n, 30, 1/4) 
Ygreen0 = rbinom(n, 30, 1/10)
Ygreen50 = rbinom(n, 30, 1/5)

Y0   = ifelse(is.blue, Yblue, Ygreen0)
Y50  = ifelse(is.blue, Yblue, Ygreen50)
Y100 = Y50   ### <--- this tells us we've got no effect.

# Randomizing by rolling two dice. R is their sum
R = sample(c(1:6), size=n, replace=TRUE) + sample(c(1:6), size=n, replace=TRUE)
D = case_match(R, 1:6~0, 7:10~50, 11:12~100)
Y = case_match(D, 0~Y0, 50~Y50, 100~Y100)
``` 

## Bias


```{r}

line.based.estimates = purrr::map(1:1000, function(b) {
  n=250
  is.blue = rbinom(n,1, 1/2)
  Yblue   = rbinom(n, 30, 1/4) 
  Ygreen0 = rbinom(n, 30, 1/10)
  Ygreen50 = rbinom(n, 30, 1/5)

  Y0   = ifelse(is.blue, Yblue, Ygreen0)
  Y50  = ifelse(is.blue, Yblue, Ygreen50)
  Y100 = Y50   ### <--- this tells us we've got no effect.

  # Randomizing by rolling two dice. R is their sum
  R = sample(c(1:6), size=n, replace=TRUE) + sample(c(1:6), size=n, replace=TRUE)
  D = case_match(R, 1:6~0, 7:10~50, 11:12~100)
  Y  = case_match(D, 0~Y0, 50~Y50, 100~Y100)

  model=lm(Y ~ D, data=tibble(D=D, Y=Y))
  mu.hat = predict(model, newdata=data.frame(D=c(0,50,100)))
  data.frame(estimate =  mu.hat[3]-mu.hat[2], se = sqrt(vcov(model)[2,2])*50)
}) %>% bind_rows()

```

```{r}
#| out-height: 500px

line.based.estimates$i = 1:nrow(line.based.estimates)
slope.se = sqrt(vcov(model)[2,2])*50
line.sampling.dist = ggplot() + 
  geom_histogram(aes(x=estimate, y=after_stat(density)), bins=50, alpha=.2, data=line.based.estimates) + 
  geom_vline(aes(xintercept=mean(estimate)), color='blue', data=line.based.estimates) +
  geom_pointrange(aes(x=slope.point.estimate, y=dnorm(2*slope.se, mean=0, sd=slope.se), 
                      xmin=slope.point.estimate - 1.96*slope.se, 
                      xmax=slope.point.estimate + 1.96*slope.se))
line.sampling.dist
``` 

- We can do a little simulation to calculate the bias of our line-based estimator.
- We'll generate 1000 fake datasets like the one we've been looking at.
  - And calculate our estimator for each one. This gives us draws from our estimator's *sampling distribution*. 
  - I've used them to draw a histogram that approximates this sampling sampling distribution.
- The mean of these point estimates is $`r approx(mean(line.based.estimates$estimate))`$. Since our actual effect is zero, that's our bias. The blue line.
- Our actual point estimate is off in the same direction---but a bit more than usual.
  - It's shown in black with a corresponding 95% interval.
  - An interval that fails to cover the actual effect of zero.

## Coverage

```{r}
#| out-height: 500px
covers =  abs(line.based.estimates$estimate) <= 1.96*line.based.estimates$se 
line.sampling.dist + geom_pointrange(aes(x=estimate, y=(8*i/100)*dnorm(2*slope.se, mean=0, sd=slope.se), 
                      xmin=estimate - 1.96*se, 
                      xmax=estimate + 1.96*se), data=line.based.estimates[1:100,], color='purple', alpha=.4)  
```

- We can construct 95% confidence intervals the same way in each fake dataset.
  - I've drawn 100 of these intervals in purple. 
  - You can see that they sometimes cover zero, but not as often as you'd like.
- `r sum(covers)` / `r nrow(line.based.estimates)` &asymp; `r percent(mean(covers))` of them cover zero. That's not good. 
- If that's typical of our analysis, people *really* shouldn't trust us. 

## Where We're Headed 

```{r}
#| out-height: 250px
#| fig.asp: .25
#| layout-ncol: 2 
#| fig-subcap:
#|   - "Difference in Sample Means Sampling Dist (n=250)"
#|   - "Fitted-Line-Based Sampling Dist (n=250)" 

xrange = c(-4,4)
scale_x = scale_x_continuous(breaks=seq(min(xrange), max(xrange), by=1), limits=xrange)
distribution.plot(boots$dif,   real.point.estimate) + scale_x
distribution.plot(boots$slope, slope.point.estimate)  + scale_x
```

:::: r-stack

::: {.fragment .fade-out data-fragment-index="0"}

- Next time, we're going to talk about why fitting lines **is** common practice and **isn't** a good idea.
- It has to do with sample size. When sample sizes are small, we get wide intervals.
  - And a small amount of bias may be negligible in comparison to it. 
  - That means we can often get away with stuff like fitting lines.
- You can see that in the sampling distribution plots below.
  - We've done the same analysis using only $50$ of our $250$ observations.
  - And the line-based estimate gives us a wider interval that covers the actual effect: zero.
-  We'll see exactly how much worse coverage gets as a function of sample size.
:::
::: {.fragment .current-visible data-fragment-index="0"}

- The reason fitting lines is common practice is historical.
- Common practice is rooted in tradition dating back to the 1930s or so. 
  - Nobody had computers. So samples were almost always small. 
  - And working with lines usually wasn't such a big problem.
- That changed in the 1970s. 
  - And what statisticians were thinking about changed too.
  - A lot of what are now called *machine learning methods* were developed at that time.
- It's taking some time to cover the basics so we'll be prepared, but that's where we're going.
:::
::::


```{r}
#| out-height: 250px
#| fig.asp: .25
#| layout-ncol: 2 
#| fig-subcap:
#|   - "Difference in Sample Means Sampling Dist (n=50)"
#|   - "Fitted-Line-Based Sampling Dist (n=50)" 

ind = c(51:100) 
# because 1:50 happened to give us an unusually large line-based estimate (of ~2)
# and that didn't help make the point we want

summaries.small = real.data[ind, ] %>% group_by(D) %>% summarize(mean=mean(Y), sd=sd(Y), n=length(Y))
small.point.estimate = summaries.small$mean[summaries.small$D==100] - summaries.small$mean[summaries.small$D==50]
smallboots = boot.distn(real.data[ind, ])

model=lm(Y ~ D, data=real.data[ind,])
mu.hat = predict(model, newdata=data.frame(D=c(0,50,100)))
small.slope.point.estimate = mu.hat[3]-mu.hat[2]

distribution.plot(smallboots$dif[is.finite(smallboots$dif)],   small.point.estimate) + scale_x
distribution.plot(smallboots$slope, small.slope.point.estimate)  + scale_x
# on the is.finite thing: 
#  - we get some NAs when our bootstrap sample includes no D=100 points. 
#  - It's pretty rare. We'll ignore it.
```

