---
title: "Lecture 8"
subtitle: "Working With Non-Binary Outcomes: Spread"
format: 
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---


```{r load-libraries}
library(purrr)
library(tibble)
library(dplyr)
library(magrittr)
library(stringr)
library(ggplot2)
library(htmltools)
library(kableExtra)
library(scales)
library(cpsR)
library(mapdata)

options(knitr.kable.NA = '')

percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }

library(cpsR)
census.api.key = '8084f81e4f71e4d32638dac15f188b424f72d063'
# for cps variable interpretation see https://www2.census.gov/programs-surveys/cps/datasets/2022/march/asec2022_ddl_pub_full.pdf
```

```{r ggplot-themes}
class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
        panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x  = element_text(colour = "#aaaaaa"),
        axis.text.y  = element_text(colour = "#aaaaaa"),
        axis.title.x  = element_text(colour = "#aaaaaa"),
        axis.title.y  = element_text(colour = "#aaaaaa", angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
              axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
           panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(simplified.theme)
```


```{r define-colors}
twocolor.red   = 'rgb(248,118,109)'
twocolor.green = 'rgb(0,191,196)'
```

```{r table-styling}
ellipsis <- function(df, nhead=5L, ntail=2) {
    rownames(df) = as.character(1:nrow(df))
    stopifnot("data.frame" %in% class(df))
    els <- rep(NA, ncol(df)) %>% 
        matrix(nrow=1, dimnames=list(NULL, names(df))) %>% 
        data.frame(stringsAsFactors=FALSE) %>% 
        tibble::as_tibble()
    rownames(els) = "⋮"
    dplyr::bind_rows(head(df, nhead), els, tail(df, ntail))
}
```


```{r get-california-data}
#| cache: true
cps.sampling.ratio = 1/2500
plotting.cps.sampling.ratio=1/25
california.state.code = 6
cps.data = cpsR::get_asec(year=2022, vars=c('pearnval', 'gestfips', 'h_numper', 'htotval', 'a_age', 'a_hga','gereg', 'a_maritl', 'a_sex','peafever','a_wkstat','gtco'), key=census.api.key)
```

```{r process-california-data}
all.earnings = data.frame(age=cps.data$a_age,
                          state.code = cps.data$gestfips,
                          county.code = cps.data$gtco,
                          income=cps.data$pearnval, 
                          education=case_match(cps.data$a_hga,
                                          0  ~ 0,   # child
                                          31 ~ 0,   # < grade 1
                                          32 ~ 4,   #  grade 1-4
                                          33 ~ 6,   #  grade 5-6
                                          34 ~ 8,   #  grade 7-8 
                                          35 ~ 9,   #  grade 9
                                          36 ~ 10,  #  grade 10
                                          37 ~ 11,  #  grade 11
                                          38 ~ 11,  #  grade 12 no diploma
                                          39 ~ 12,  #  high school grad
                                          40 ~ 13,  #  some college
                                          41 ~ 14,  #  associate degree(vocational)
                                          42 ~ 14,  #  associate degree (academic)
                                          43 ~ 16,  #  bachelors degree
                                          44 ~ 18,  #  masters degree
                                          45 ~ 20,  #  professional school degree
                                          46 ~ 20)) #  doctorate

in.interval = function(X, interval) { interval[1] <= X & X <= interval[2] }
all.earnings$employed<- ifelse(cps.data$a_wkstat >5, 0,ifelse(cps.data$a_wkstat<2,NA,1))
all.earnings$degree = factor(all.earnings$education >= 16,
                    levels=c(FALSE, TRUE), 
                    labels=c('no 4-year degree','4-year degree')) 

earnings = all.earnings[all.earnings$age %>% in.interval(c(25,35)) & 
                        all.earnings$state.code == california.state.code &
                        all.earnings$education > 8, ]

rownames(earnings)=NULL
earnings$person = 1:nrow(earnings)
n = nrow(earnings)
```

```{r fake-pop-and-sample-it}
set.seed(11)
pop.earnings = earnings[sample(nrow(earnings), nrow(display.population), replace=TRUE), ]
pop.earnings$income = pop.earnings$income + 
    ifelse(1:nrow(pop.earnings) <= nrow(earnings), 0, 3000*rnorm(nrow(pop.earnings)))
pop.j = sample(nrow(pop.earnings))
pop.earnings[pop.j, ] = pop.earnings

pop.tab = pop.earnings
J = pop.j[1:nrow(earnings)]
sample.tab = pop.tab[J, ]
sample.tab$J = J
```


```{r maps}
#| cache: true

## All of this is just to make maps showing the sample and population.
## And this code will be used later, so some of what's written here isn't 
## used in the lecture. I wouldn't read it unless you're interested in drawing maps in R.

# county info from https://en.wikipedia.org/wiki/List_of_counties_in_California
# converted to csv using https://wikitable2csv.ggor.de
to.number = function(s) { as.numeric(str_trim(gsub(",", "", s))) }
ca_county_info = read.csv('california_county_info.csv') %>%
  mutate(subregion = tolower(sub(' County', '', County))) %>%
  mutate(pop = to.number(Population..2022.)) %>%
  mutate(area = to.number(sub('sq.*', '', Area))) 

earnings.county = left_join(earnings, ca_county_info[,c('FIPS.code', 'subregion')], by=c('county.code' = 'FIPS.code'))
earnings$county = earnings.county$subregion
earnings$county[is.na(earnings$county)]='unknown'
  
# To give our survey recipients a bit more reality, we'll give them fake locations on a map
# And we'll make up a population of California residents that they're sampled from
# maps are based on code from https://eriqande.github.io/rep-res-eeb-2017/map-making-in-R.html

counties = map_data("county")
ca_county = counties %>%
  filter(region == "california")
ca_county_centers = ca_county %>%
  group_by(subregion) %>%
  summarize(county.lat = mean(lat), county.long=mean(long))

ca_county_more = left_join(ca_county_centers, ca_county_info, by="subregion")
ca_all = left_join(ca_county, ca_county_more, by="subregion")


ca.plot = ggplot(data=ca_all, mapping=aes(x=long, y=lat, group=group)) + 
  coord_quickmap() +
  geom_polygon(fill='gray', color = "white", alpha=.4) +
      scale_fill_gradient(trans="log10") + 
      theme_void()

ca.pop.plot = ggplot(data=ca_all, mapping=aes(x=long, y=lat, group=group)) + 
  coord_quickmap() +
  geom_polygon(aes(fill = pop/area), color = "white") +
      scale_fill_gradient(trans="log10") + 
      theme_void()

## we'll make up a population as follows
# 1. choose the number of people to be the number of people in our sample scaled up to account for the cps sampling ratio
# 2. for each, sample a county with probability proportional to its finite.population
# 3. for each, sample a latitude and longitude by taking a randomly weighted combination of the    
#    county's vertices. As long as the county is convex, these points will be inside the county.
# 4. we'll give them a degree too, chosen independently of all else from the empirical distribution on the sample
npop = nrow(earnings) / plotting.cps.sampling.ratio

# takes a list of subregions, possibly with duplicates
# for each element, chooses random lat/long coordinates within the sample.subregion
# by taking a random convex combination of the subregion's vertices
sample.within.subregion = function(subregions) {
  output = data.frame(subregion=subregions, lat=NA, long=NA)
  for(subregion in unique(subregions)) { 
      vertices = ca_county[ca_county$subregion==subregion, c('lat','long')]
      U = sum(subregions==subregion)
      V = nrow(vertices)
      weights = abs(rt(U*V, 1))
      dim(weights) = c(U,V)
      weights = (1/rowSums(weights)) * weights
      output[subregions==subregion, c('lat', 'long')] = 
                 cbind(weights %*% vertices$lat, weights %*% vertices$long)
    }
    output
}

finite.population.subregion = sample(ca_county_more$subregion, size=npop, replace=TRUE, prob=ca_county_more$pop)
finite.population = sample.within.subregion(finite.population.subregion)
finite.population$degree = sample(earnings$degree, nrow(finite.population), replace=TRUE)
finite.population$group=1 # for plotting

display.population = sample_n(finite.population, min(100000, nrow(finite.population)))

# and we'll force the degree-distribution in the sample to match our survey responses
our.sample = sample_n(finite.population, nrow(earnings))
our.sample$degree=earnings$degree

pop.plot = ca.plot + geom_point(aes(x=long, y=lat), data=display.population, size=.02, alpha=.1) 
sample.plot = ca.plot + geom_point(aes(x=long, y=lat), data=our.sample, size=.2, alpha=.5)
pop.plot.inf = ca.pop.plot + geom_point(aes(x=long, y=lat), data=display.population, size=.01, alpha=.1)  
sample.plot.inf = ca.pop.plot + geom_point(aes(x=long, y=lat), data=our.sample, size=.2, alpha=.5)
```



## Review

::: {.hidden}
$$
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\V}{V}
\DeclareMathOperator{\Var}{\V}
\newcommand{\inred}[1]{\textcolor{red}{#1}}
\newcommand{\inblue}[1]{\textcolor{blue}{#1}}
$$ 

:::

```{r}
our.seed = 0
width = function(samples, alpha=.05) {
  covg.dif = function(w) { 
    actual = mean( mean(samples) - w/2 <= samples & samples <= mean(samples) + w/2)
    nominal = 1-alpha 
    actual - nominal  
  }
  uniroot(covg.dif, interval=c(0, 2*max(abs(samples-mean(samples)))))$root
}

default.height = .04
scale_x = scale_x_continuous(breaks=seq(.6,.8, by=.025), limits=c(.6, .8)) 
scale_y = scale_y_continuous(breaks=seq(0,default.height,by=.01), limits=c(0, default.height))

annotate.histogram = function(histplot, samples, shade=FALSE, 
                              shade.width=width(samples, alpha=.05), shade.color='green', shade.alpha=.15) { 
    w66 = width(samples, alpha=1/3)
    w95 = width(samples, alpha=.05)
    p = histplot + 
           geom_vline(aes(xintercept=mean(samples)), color="blue") +
           geom_vline(aes(xintercept=mean(samples)+c(-1,1)*w66/2), color="blue", linetype='dashed') +
           geom_vline(aes(xintercept=mean(samples)+c(-1,1)*w95/2), color="blue", linetype='dotted') 
    if(shade) { 
      p + geom_rect(aes(xmin=mean(samples)-shade.width/2, 
                        xmax=mean(samples)+shade.width/2, 
                        ymin=0, ymax=Inf), alpha=shade.alpha, fill=shade.color)
    } else { p }
}

annotated.bar = function(samples, ...) { 
    p = ggplot() + geom_bar(aes(x=samples, y=after_stat(prop)), alpha=.3) +
        geom_vline(aes(xintercept=mean(y)), color='green', size=3, alpha=.7)
    annotate.histogram(p, samples, ...) + scale_x + scale_y
}

default.highlight.colors = c('black', 'magenta', 'cyan', 'gray')
with.points = function(histplot, samples, K=100, highlight=4, height=default.height) {
    set.seed(our.seed)
    highlight.colors = if(highlight) default.highlight.colors[1:highlight] else c()
    heights = sample(height*(1:K)/K)
    data = data.frame(y=heights, x=samples[1:K], 
                      color=c(highlight.colors, rep('purple', K-length(highlight.colors))),
                      alpha=c(rep(1, length(highlight.colors)), rep(.1, K-length(highlight.colors))))
    histplot + geom_point(aes(y=y, x=x, color=I(color), alpha=I(alpha)), data=data)
}

with.intervals = function(histplot, samples, w=width(samples, alpha=.05), K=100, 
                          highlight=4, highlight.alpha=1, height=default.height) { 
    set.seed(our.seed)
    highlight.colors = if(highlight) default.highlight.colors[1:highlight] else c()
    heights = sample(height*(1:K)/K)
    data = data.frame(y=heights, x=samples[1:K], w=w,
                      color=c(highlight.colors, rep('purple', K-length(highlight.colors))),
                      alpha=c(rep(highlight.alpha, length(highlight.colors)), rep(.1, K-length(highlight.colors))))
    histplot + 
        geom_pointrange(aes(y=y, x=x, xmin=x-w/2, xmax=x+w/2, color=I(color), alpha=I(alpha)), data=data)
}
```

```{r}
#| cache: true
social.pressure = read.csv('social-pressure-data.csv') 
keep = social.pressure$treatment %in% c(' Control', ' Neighbors')
D = as.numeric(social.pressure[keep,'treatment'] == ' Neighbors')
Y = as.numeric(social.pressure[keep,'voted'] == 'Yes')
n = length(Y)

difference = mean(Y[D==1]) - mean(Y[D==0])
difference.bootstrap.samples = array(dim=10000)
for(rr in 1:10000) {
    I = sample(1:n, n, replace=TRUE)
    Dstar = D[I]
    Ystar = Y[I]
    difference.bootstrap.samples[rr] = mean(Ystar[Dstar==1]) - mean(Ystar[Dstar==0])
}
```

## Where We Are

```{r visualization-defaults}
scale_y = function(limits=c(0, 4e5)) { 
  scale_y_continuous(labels=label_dollar(), limits=limits)
}

dot.size = 2
```

:::: columns

::: column
```{r map-sample}
#| fig-width: 5
#| out-height: 300px
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=our.sample, size=.1, alpha=.5) + guides(color='none')
```
:::

::: column

```{r table}
style.tab = function(tab, cols=c('income', 'education', 'county'), nhead=3, ntail=1) {
  tab$income50k = 0+(tab$income > 50000)
  tab$income = sprintf('$%dk', round(as.numeric(tab$income) / 1000))
  tab$county = ifelse(tab$county=='los angeles', 'LA',  tab$county)
  ellipsis.tab = tab[, cols] %>% ellipsis(nhead=nhead,ntail=ntail)
  ellipsis.tab %>%
    kable %>% kable_styling() %>% 
    row_spec(which(ellipsis.tab$education >= 16),  background=twocolor.green) %>%
    row_spec(which(ellipsis.tab$education <  16),  background=twocolor.red) 
}

style.tab(sample.tab, nhead=4)
```
:::

::::

- We've talked about **estimating the mean of a population**. 
- And means of subpopulations, too. And differences in those.
- We've been looking the incomes of California residents [with]{.twocolor-green} and [without]{.twocolor-red} 4-year degrees.
  - Our population is the set of all California residents meeting some age and education criteria.
  - And we'll think of our sample as being drawn with replacement from that population.
- To think carefully about the relationship between our sample and the population, we'll use **random variables**.

## Imagining a Population

```{r}
#| out-height: 300px
#| fig-width: 5 
#| layout-ncol: 2 
#| 
ca.plot + 
 geom_point(aes(x=long, y=lat, color=degree), data=display.population, size=.02, alpha=.1) + 
 guides(color='none')
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=our.sample, size=.1, alpha=.5) +
  guides(color='none')
```


:::: columns
::: column
```{r}
style.tab(pop.tab)
```
:::
::: column
```{r}
style.tab(sample.tab)
```
:::
::::


- We don't have much information about the people in our population. Except the ones in our sample. 
- But, for the sake of visualization, I've made some up. 
  - We'll put visualizations of this fake population on the left.
  - Visualizations of the (real) sample will be on the right.

## Unbiasedness

```{r}
xx = seq(-6,6,by=.01)
K = 100
heights = seq(0,.4,length.out=K)
normals = rnorm(10000)

unbiased.plot = ggplot() + geom_histogram(aes(x=normals, y=after_stat(density)), alpha=.3, bins=100) +
           geom_vline(xintercept=0, color='green', linewidth=3, alpha=.4) + 
           geom_vline(xintercept=0, color='blue') + 
           geom_vline(xintercept=c(-1,1), color='blue', linetype='dashed') +
           geom_vline(xintercept=c(-2,2), color='blue', linetype='dotted') +
           geom_pointrange(aes(y=heights, x=normals[1:K], xmin=normals[1:K]-1.96, xmax=normals[1:K]+1.96), 
                           color='purple', alpha=.15) + xlim(min(xx),max(xx))
           
biased.plot = ggplot() + geom_histogram(aes(x=normals+2, y=after_stat(density)), alpha=.3, bins=100) +
           geom_vline(xintercept=0, color='green', linewidth=3, alpha=.4) + 
           geom_vline(xintercept=2+0, color='blue') + 
           geom_vline(xintercept=2+c(-1,1), color='blue', linetype='dashed') +
           geom_vline(xintercept=2+c(-2,2), color='blue', linetype='dotted') +
           geom_pointrange(aes(y=heights, x=2+normals[1:K], xmin=2+normals[1:K]-1.96, xmax=2+normals[1:K]+1.96), 
                           color='purple', alpha=.15) + xlim(min(xx),max(xx))


```

```{r}
#| layout-ncol: 2
unbiased.plot
biased.plot
```

- We've shown that using sample means as estimators of population means gives us **unbiased estimators**.
  - That is, estimators whose expected values are the thing we want to estimate.
  - Intuitively, that means their sampling distributions are where we want them to be.
  - Like the plot on the left. Not the one on the right. 
- This is essentially necessary to get correctly calibrated interval estimates.
- It's more or less sufficient too. 
  - Calibration via bootstrap works for every unbiased estimator we'll talk about this semester. 
  - Things are more complicated for fancy machine learning stuff, but that's beyond the scope.

## So We Can Trust This Interval Estimate

```{r}
#| cache: true
X = as.numeric(earnings$education >= 16)
Y = earnings$income 
n = length(Y)

difference = mean(Y[X==1]) - mean(Y[X==0])
difference.bootstrap.samples = array(dim=10000)
for(rr in 1:10000) {
    I = sample(1:n, n, replace=TRUE)
    Xstar = X[I]
    Ystar = Y[I]
    difference.bootstrap.samples[rr] = mean(Ystar[Xstar==1]) - mean(Ystar[Xstar==0])
}
```

```{r}
#| out-height: 250px
#| fig-asp: .25 
 
xx = seq(min(difference.bootstrap.samples), max(difference.bootstrap.samples), length.out=1000)
wboot=width(difference.bootstrap.samples, alpha=.05)
difference.bootstrap.samples = difference.bootstrap.samples 
bootstrap.histogram=ggplot() + geom_histogram(aes(x=difference.bootstrap.samples, y=after_stat(density)), alpha=.3, bins=100) 
annotated.bootstrap.difference.histogram = annotate.histogram(bootstrap.histogram, difference.bootstrap.samples,
  shade=FALSE)
with.points(annotated.bootstrap.difference.histogram, difference.bootstrap.samples, height=.1, highlight=0) + 
  geom_pointrange(aes(x=x,y=y, xmin=x-w/2, xmax=x+wboot/2), 
                  data=data.frame(x=difference, y=.05, w=wboot), color='black')  +
  geom_line(aes(x=xx, y=dnorm(xx, mean(difference.bootstrap.samples), sd(difference.bootstrap.samples))), color='cyan') 
``` 

# Probability Stuff

## Review: Random Variables

```{r}
#| out-height: 300px
#| layout-ncol: 2 

earnings.jitter.pop = ggplot(pop.earnings) + 
    geom_point(aes(x=degree, y=income, color=degree), 
               size=.2, position=position_jitter(seed=0,w=.3), alpha=.1) +
     scale_y() + guides(color='none')

earnings.jitter = ggplot(earnings) + 
    geom_point(aes(x=degree, y=income, color=degree), 
               size=dot.size, position=position_jitter(seed=0,w=.3), alpha=.1) +
     scale_y() + guides(color='none')
```

## Review: Observations as Random Variables

```{r}
#| out-height: 300px
#| layout-ncol: 2 

summaries = earnings %>% group_by(degree) %>% summarize(mean=mean(income), sd=sd(income))

earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

- The notation $y_j$ refers to a number.
  - It's the income of the $j$th person in our population. That's just some person. 
- The notation $Y_i$ refers to something else.
  - It's the income of the $i$th person we call in our survey. That's not a person.
  - That's the result of a random process---the roll of a die.
- To summarize this result, we talk about the **probability distribution** of $Y_i$.

## Review: Expectations

:::: columns
::: column
```{r}
```
:::
::: column
```{r}
sample.tab
```
:::
::::

- Often the thing we care about is the expected value of $Y_i$. Or some related quantity.
  - The expected value of $Y_i$ is the **probability-weighted average** of the values it can take on.
  - What's nice about this is that we can think of this in 'uncollapsed' terms.
- When we sample as usual, this is just the population mean.
- The collapsed form is, in a sense, just summing in a different order.

$$
\begin{aligned}
\E[Y_i] = \sum_y P(Y_i = y) \times y
= \sum_y \qty(\sum_{j: y_j = y} \frac{1}{m}) \times y 
= \frac{1}{m}\sum_y\sum_{j:y_j = y} y 
= \frac{1}{m}\sum_{j=1}^m y_j
\end{aligned}
$$


## Spread

```{r}
#| out-height: 300px
#| layout-ncol: 2 
#| 
ggplot(earnings) + 
  geom_point(aes(x=0, y=income), color='blue', 
            size=2, position=position_jitter(seed=0,w=.3), alpha=.2) +
  stat_summary(aes(x=0, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x))  

earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

- To summarize the **spread** of a distribution, we often talk about variance.
- The variance is an expectation: it's the **expected squared difference** between a random variable and its expectation.
$$
\Var(Y) = \E \qty[ \qty{ Y - \E(Y) }^2 ]
$$

- To put that back on the same scale as $Y$---instead of the scale of $Y^2$---we take it's square root. That's the standard deviation.

$$
\text{standard deviation}(Y) = \sqrt{\Var(Y)}
$$

- We can talk about spread outright (left) --- that's (unconditional) variance.
- Or spread within groups (right) --- that's conditional variance.


## Variance as Excess {.smaller}

$$
\small{
\begin{aligned}
\Var(Y) &= \E \qty[ \qty{ Y - \E(Y) }^2 ] && \text{ the \emph{spread} idea: the mean square of a \emph{centered version} of $Y$ } \\
&= \E\qty(Y^2) - \qty{\E\qty(Y)}^2         && \text{ the \emph{excess} idea: the average amount $Y^2$ \emph{exceeds} the square of its mean}.
\end{aligned}
}
$$

```{r}
#| out-height: 250px
#| fig-asp: .25

Y = c(.75,1,1.25, 1,1.25,1.5)
X = 0 + (1:6 > 3)
Xdodge = X + rep(.02*c(-1,0,1),2)
mu = function(x) { mean(Y[X==x]) }

tinypop.square.plot = ggplot() + geom_point(aes(x=Xdodge, y=Y, color=factor(X)), alpha=.5, size=3*dot.size) +
                                 geom_point(aes(x=Xdodge, y=Y^2, color=factor(X)), shape=0, size=3*dot.size) + 
                                 geom_segment(aes(x=Xdodge, xend=Xdodge, y=Y, yend=Y^2), alpha=.3) + guides(color='none')
tinypop.square.plot + xlim(min(Xdodge[X==0])-sd(Xdodge[X==0]), max(Xdodge[X==0] + sd(Xdodge[X==0])))
```

Let's think, in terms of the plot, about why there is excess in the square.

- Suppose $Y$ takes on three values --- the dots ---with equal probability.
    - Its expectation is the average height of the black dots.
    - And its expectation, squared, is the square of this.
- Its square $Y^2$ takes on the the square of those values---the squares --- with equal probability.
    - Its expectation is the average height of the squares.
- Why is the second one bigger?

::: fragment

- Because squaring *increases big draws of Y* more than it *decreases small ones*.
- Visually, the line from $Y$ to $Y^2$ is longer when $Y$ is bigger than when it's smaller.

:::


## Equivalence Proof

::: fragment

$$
\small{
\begin{aligned}
\E \qty[ \qty{ Y - \E(Y) }^2 ] 
&= \E \qty[ Y^2 - 2 Y \E(Y) + \qty{\E(Y)}^2 ] && \text{ FOIL } \\
&= \E[ Y^2] - \E\qty[2 Y \E(Y)] + \E\qty[\qty{\E(Y)}^2] && \text{Distributing expectations (linearity)} \\
&= \E[ Y^2] - 2\E(Y)\E\qty[Y] + \qty{\E(Y)}^2\E[1] && \text{Pulling constants out of expectations (linearity)} \\
&= \E[ Y^2] - 2\qty{\E(Y)}^2 + \qty{\E(Y)}^2] && \text{Recognizing a product of something with itself as a square} \\
&= \E[ Y^2] - \cancel{2}\qty{\E(Y)}^2 + \cancel{\qty{\E(Y)}^2}] && \text{Subtracting} 
\end{aligned}
}
$$

:::

## Conditional Variance {.smaller}

- The **conditional variance** describes the spread of one random variable *within groups*. 
- Like the conditional expectation of the random variable, there are two related concepts.

1. The **conditional variance function**.
$$
\sigma^2(x) := \Var(Y \mid X=x) = \E\qty[ \textcolor{blue}{\qty{ Y-\E(Y \mid X) }^2} \mid X=x]
$$
    - This is the variance of $Y$, **within the subpopulation** of people with $X=x$.
    - It is the *conditional expectation function* of the random variable $\textcolor{blue}{\{Y-\E(Y \mid X)\}^2}$.
    - We call it 'the conditional variance of $Y$ given $X=x$'

2. The conditional variance $\sigma^2(X) = \Var(Y \mid X)$
$$
\sigma^2(X) := \Var(Y \mid X)   = \E\qty[ \textcolor{blue}{\qty{ Y-\E(Y \mid X) }^2} \mid X]
$$

    - This is the variance of $Y$ within **a random subpopulation** of people.
        - It's the conditional variance function **evaluated at** the random variable $X$.
        - And  the conditional expectation of the random variable $\textcolor{blue}{\{Y-\E(Y \mid X)\}^2}$.
    - We call it 'the conditional variance of $Y$ given $X$.
    - This is the sort of thing that shows up when you use the law of iterated expectations.



## Conditional Variance as Excess {.smaller}

$$
\small{
\begin{aligned}
\Var(Y \mid X) &= \E \qty[ \qty{ Y - \E(Y\mid X)}^2 \mid X ] && \text{ the \emph{spread} idea: the mean square of a conditionally centered version of $Y$ } \\
&= \E\qty(Y^2 \mid X) - \qty{\E\qty(Y \mid X)}^2              && \text{ the \emph{excess} idea: the average amount $Y^2$ exceeds the square of its cond. mean}
\end{aligned}
}
$$


```{r}
#| out-height: 500px
#| fig-asp: .5
tinypop.square.plot 
```

- We can think of the Conditional Variance as excess, too.
- It's the same as the unconditional case, but *within subpopulations*.


## The Law of Total Variance

$$
\Var[Y] = \E\qty{\Var( Y \mid X ) } + \Var\qty{\E( Y \mid X ) }  
$$

- The Law of Total Variance relates Conditional and Unconditional Variance
- It's like the Law of Total Expectation, but for Variance
- It's a useful way to decompose the variance a random variable.
- Think about where most of the variance in $Y$ is coming from in the two populations shown below.

```{r}
#| out-height: 500px
#| fig-ncol: 2

ggplot() + geom_point(aes(x=Xdodge, y=Y, color=factor(X)), alpha=.7, size=4) + 
  scale_y_continuous(breaks=seq(0,3,by=.5), limits=c(0,3)) + guides(color='none') 
ggplot() + geom_point(aes(x=Xdodge, y=Y+X, color=factor(X)), alpha=.7, size=4) + 
  scale_y_continuous(breaks=seq(0,3,by=.5), limits=c(0,3)) + guides(color='none') 

```


## An Activity 

```{r}
#| out-height: 250px
#| fig-asp: .25

mean.has.degree = round(summaries$mean[summaries$degree=='4-year degree'] / 10000) * 10
mean.no.degree  = round(summaries$mean[summaries$degree=='no 4-year degree'] / 10000) * 10
degree.proportion = round(mean(earnings$degree=='4-year degree'),1)
earnings.jitter + geom_pointrange(aes(x=degree, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries) + scale_y()
```

- Working with your neighbor, think about why we call it this.
- Try to put the three terms into words.
    - $\Var[Y]$ is the ...
    - $\E\qty{ \Var(Y \mid X) }$ is the ...
    - $\Var\qty{\E( Y \mid X)}$ is the ...
- Do this in two ways.
    1. Using 'Random Variables Language', 
        - e.g. 'the expected value of the conditional variance of Y given X'
    2. Using 'Population Language'
        - e.g. 'the variance of income (Y) within the random subpopulation with degree status X'


## Proof

**Claim.**
$$
\Var[Y] = \textcolor{blue}{\E\qty[\Var(Y \mid X)]} + \textcolor{green}{\Var\qty[\E(Y \mid X)]} 
$$

1. Rewrite the variances, conditional and unconditional, in *excess* form.

::: fragment
$$
\Var[Y]  = \textcolor{blue}{\E\qty[ \E(Y^2 \mid X) - \qty{\E\qty(Y \mid X)}^2]} 
                       + \textcolor{green}{\E\qty{ \E(Y \mid X)^2}  - \qty(\E\qty{\E\qty(Y \mid X)})^2} 
$$

:::

2. Distribute the outer expectations in the first term.

::: fragment
$$
\Var[Y] = \textcolor{blue}{\E\qty[ \E(Y^2 \mid X)] - \E\qty[\qty{\E\qty(Y \mid X)}^2]} 
                       + \textcolor{green}{\E \qty[ \qty{\E(Y \mid X)^2} ] - \qty(\E\qty{\E\qty(Y \mid X)})^2}
$$

:::

3. Cancel matching terms

::: fragment
$$
\Var[Y]  = \textcolor{blue}{\E\qty[ \E(Y^2 \mid X)] \cancel{- \E\qty[\qty{\E\qty(Y \mid X)}^2]}}  
                       + \textcolor{green}{\cancel{\E \qty[ \qty{\E(Y \mid X)^2} ]} -  \qty(\E\qty{\E\qty(Y \mid X)})^2}
$$

:::

4. Use the law of iterated expectations to simplify the remaining terms.

::: fragment
$$
\Var[Y]  = \textcolor{blue}{E[Y^2]  \cancel{- \E\qty[\qty{\E\qty(Y \mid X)}^2]}}
                       + \textcolor{green}{\cancel{\E \qty[ \qty{\E(Y \mid X)^2} ]} - (\E\qty{Y})^2}
$$

:::

## The Variance of The Mean

$$
\begin{aligned}
\Var[\bar Y] 
&= \E\qty[ \qty{ \frac{1}{n}\sum_i Y_i - \E \qty( \frac{1}{n}\sum_i Y_i ) }^2 ] && \text{definition}\\
&= \E\qty[ \qty{ \frac{1}{n}\sum_i (Y_i - \E Y_i) }^2 ]  && \text{Rule 2: Linearity} \\
&= \E\qty[ \qty{ \frac{1}{n}\sum_i Z_i }^2 ] && \text{for} \ \ Z_i = Y_i - \E Y_i \\
&= \E\qty[ \frac{1}{n^2}\sum_i \sum_j Z_i Z_j ]   && \text{via FOIL}
\end{aligned}
$$

### FOIL?

- Remember FOIL from high school algebra? 
- When we square a sum, we get *all products* of terms in the sum.
  
$$ 
(a+b)^2 = \inblue{(a+b)}\inred{(a+b)} 
= \inblue{a}\inred{a} + \inblue{a}\inred{b} + \inblue{b}\inred{a} + \inblue{b}\inred{b}
= a^2 + 2ab + b^2
$$

- Same deal for bigger sums.

$$ 
\qty{ \sum_i Z_i }^2 = \inblue{\sum_i Z_i}\inred{\sum_j Z_j} = \sum_i \sum_j \inblue{Z_i}\inred{Z_j}
$$

## Completing the Calculation

$$
\begin{aligned}
\Var[\bar Y] &= \E\qty[ \frac{1}{n^2}\sum_i \sum_j Z_i Z_j ]  \quad \text{ for } \quad Z_i = Y_i - \E Y_i \quad && \text{via FOIL}
\end{aligned}
$$

**Activity**. Complete the calculation to get a simple expression. 

:::: columns
::: column
 **Hints**.

  1. Use Rule 1 to push the expectation inside the sum.
  2. Think about what $\E Z_i Z_j$ is when $i=j$ and $i\neq j$.
  3. Is Rule 2 useful here?
  4. What is $\E Z_i^2$  
:::
::: column
**Rules**.
$$
\small{
\begin{aligned}
\text{1.} \quad & \E\qty[ \sum_i a_i Y_i] =  \sum_i a_i \E Y_i \quad \text{ for RVs $Y_i$ and constants $a_i$.} \\
\text{2.} \quad & \E\qty[ Z_i Z_j] = \E[Z_i] \E[ Z_j]  \quad \text{if $Z_i$ and $Z_j$ are independent.} \\
\end{aligned}
}
$$

:::
::::

## The Variance of The Mean: Answered
$$
\begin{aligned}
\Var[\bar Y] 
&= \E\qty[ \frac{1}{n^2}\sum_i \sum_j Z_i Z_j ]  && \text{via FOIL} \\
&= \frac{1}{n^2} \sum_i \sum_j \E Z_i Z_j    && \text{via Rule 1: Linearity } \\
&= \frac{1}{n^2} \sum_i \qty( \sum_{j : j\neq i} \E Z_i Z_j + \E Z_i^2  )  && \text{Taking out the term with $i=j$ term from the inner sum}  \\
&= \frac{1}{n^2} \sum_i \qty( \sum_{j : j\neq i} 0 + \E Z_i^2  )  && \text{via Rule 2} \\
&= \frac{1}{n^2} \sum_i \E\qty(Y_i-\E Y_i)^2  && \text{using the definition of $Z_i$} \\
&= \frac{1}{n^2} \sum_i \Var(Y_i)             && \text{using the definition of variance} \\
&= \frac{1}{n^2} \times n \times \Var(Y_i) = \frac{\Var(Y_i)}{n} = \frac{\sigma^2}{n}.   && \text{counting the copies of $\Var(Y_i)$ in the sum} \\
\end{aligned}
$$

## The Variance of the Subsample Mean

**Claim.** The variance of our point estimate is the **expected value** of the subpopulation variance divided by the number of people in the subsample.
$$ \Var[\hat\mu(1)] = \E\qty[ \frac{\sigma^2(1)}{N_1} ] \quad \text{ for } \quad N_1 = \sum_{i=1}^n X_i $$

- This is like what we saw in the unconditional case in Lecture 3, with a twist.
- The number of people in the subsample is random, so instead of $1/n$, we have $\E[ 1/N_1 ]$.
- Let's prove it.
  - It's a bit long. 
  - To keep organized, I'll remind you where we left off at the top of each slide.
  - Except the first one, where we're starting from the beginning.

## Proof, Part 1

- We'll start by using the Law of Total Variance, conditioning on $X_1 \ldots X_n$.
- We 'looked back' just now for a reason---we saw that one of these terms is zero.

$$
\small{
\begin{aligned}
\Var\qty[ \hat \mu(1) ] &= \E \qty[ \Var\qty{ \hat \mu(1) \mid X_1 \ldots X_n }] + \Var\qty[ \E{\hat \mu(1) \mid X_1 \ldots X_n} ] \\
&= \E \qty[ \Var\qty{ \hat \mu(1) \mid X_1 \ldots X_n }] + \Var[\mu(1)] \\
&= \E \qty[ \Var\qty{ \hat \mu(1) \mid X_1 \ldots X_n }] + 0
\end{aligned}
}
$$



## Proof, Part 2 {.smaller}

$$
\Var\qty[ \hat \mu(1) ] 
= \E\qty[ \Var\qty{ \hat \mu(1) \mid X_1 \ldots X_n }] 
$$

- Now we'll substitute in definitions and simplify.
- We'll end having to work with the conditional mean of a *squared sum*.

$$
\small{
\begin{aligned}
\Var\qty[ \hat \mu(1) ] 
&= \E\qty[ \Var\qty{ \hat \mu(1) \mid X_1 \ldots X_n }] \\
&= \E\qty[ \E\qty{  \qty(\hat \mu(1)-\mu(1))^2 \mid X_1 \ldots X_n }] \\
&= \E\qty[ \E\qty{  \qty(\frac{\sum_{i=1}^{n}X_i Y_{i}}{\sum_{i=1}^{n}X_{i}} - \mu(1))^2 \mid X_1 \ldots X_n }] \\
&= \E\qty[ \E\qty{  \qty(\frac{\sum_{i=1}^{n}X_i Y_{i}}{\sum_{i=1}^{n}X_{i}} - \frac{\mu(1)\sum_{i=1}^n X_i}{\sum_{i=1}^n X_i })^2 \mid X_1 \ldots X_n }] \\
&= \E\qty[ \E\qty{  \qty(\frac{\sum_{i=1}^{n}X_i \qty{Y_{i}-\mu(1)}}{\sum_{i=1}^{n}X_{i}})^2 \mid X_1 \ldots X_n }] \\
&= \E\qty[ \frac{\E\qty{ \qty(\sum_{i=1}^{n}X_i \qty{Y_{i}-\mu(1)})^2 \mid X_1 \ldots X_n}}{\qty(\sum_{i=1}^{n}X_{i})^2}]
\end{aligned}
}
$$

## Proof, Part 3 {.smaller}

$$
\Var\qty[ \hat \mu(1) ]  = \E\qty[ \frac{\E\qty{ \qty(\sum_{i=1}^{n}X_i \qty{Y_{i}-\mu(1)})^2 \mid X_1 \ldots X_n}}{\qty(\sum_{i=1}^{n}X_{i})^2}]
$$

- To work with this squared sum, we do our FOIL-like expansion. This is ...

$$ \small{ 
\begin{aligned}
\qty(\sum_{i}Z_i)^2 = \qty(\sum_{i}Z_i) \qty(\sum_{j}Z_i) = \sum_i\sum_j Z_i Z_j \quad && \text{ for } \quad Z_i = X_i \qty{Y_{i}-\mu(1)} \\
\text{or equivalently, using the \textbf{indicator trick}} && \text{ for } \quad Z_i = X_i\qty{Y_i - \mu(X_i)}
\end{aligned}
}
$$

- This lets us push the conditional mean *all the way into the sum*, so it meets the individual terms.
- And when we've done that, we can use the irrelevance of independent conditioning variables to simplify.

$$
\small{
\begin{aligned}
\Var\qty[ \hat \mu(1) ] 
&= \E\qty[ \frac{\E\qty{ \qty(\sum_{i=1}^{n}Z_i )^2 \mid X_1 \ldots X_n}}{\qty(\sum_{i=1}^{n}X_{i})^2}] \\
&= \E\qty[ \frac{\E\qty{ \sum_{i=1}^{n}\sum_{j=1}^n Z_i Z_j \mid X_1 \ldots X_n}}{\qty(\sum_{i=1}^{n}X_{i})^2}] \\
&= \E\qty[ \frac{\sum_{i=1}^{n}\sum_{j=1}^n \E\qty{Z_i Z_j \mid X_1 \ldots X_n}}{\qty(\sum_{i=1}^{n}X_{i})^2}] \\
&= \E\qty[ \frac{\sum_{i=1}^{n}\sum_{j=1}^n \E\qty{Z_i Z_j \mid X_i, X_j}}{\qty(\sum_{i=1}^{n}X_{i})^2}] \\
\end{aligned}
}
$$

## Proof, Part 4 {.smaller}

$$
\begin{aligned} 
\Var\qty[ \hat \mu(1) ] 
&= \E\qty[ \frac{\sum_{i=1}^{n}\sum_{j=1}^n \E\qty{Z_i Z_j \mid X_i, X_j}}{\qty(\sum_{i=1}^{n}X_{i})^2}] \quad \text{ for } \quad Z_i=Y_i-\mu(X_i)
\end{aligned}
$$


- Let's take a look at one of the terms in the numerator, $\E\qty{Z_i Z_j \mid X_i, X_j}$.
  - We're going to show that the cross-terms --- the ones for $i \neq j$ --- are zero.
  - This will let us drop them from the sum.
  - We'll use the fact that $Z_i$ and $Z_j$ are independent for $i \neq j$.
- Let's rewrite a cross-term using our usual stuff
  - the law of iterated expectations, conditioning on $Z_i$
  - linearity of expectations;
  - irrelevance of independent conditioning variables.

$$\small{
\begin{aligned}
\E\qty{Z_i Z_j \mid X_i, X_j} 
&= \E\qty{ \E(Z_i Z_j \mid Z_i,X_i,X_j) } \\
&= \E\qty{ Z_i \E(Z_j \mid X_i,X_j) } \\
&= \E\qty{ Z_i \textcolor{blue}{\E(Z_j \mid X_j)} } \quad \text{ where } \\
\textcolor{blue}{\E(Z_j \mid X_j)} 
&= \E\qty[X_j \qty{Y_j - \mu(X_j)}] \\ 
&= \E\qty{ \E\qty[X_j \qty{Y_j - \mu(X_j)} \mid X_j] } \\
&= \E\qty[X_j \qty{ \E\qty[Y_j \mid X_j] - \mu(X_j)}] \\
&= \E\qty[X_j \cross{\qty{ \mu(X_j) - \mu(X_j)}}] = 0. \\
\end{aligned}
}$$


## Proof, Part 5 {.smaller}

$$
\begin{aligned} 
\Var\qty[ \hat \mu(1) ] 
&= \E\qty[ \frac{\sum_{i=1}^{n}\sum_{j=1}^n \E\qty{Z_i Z_j \mid X_i, X_j}}{\qty(\sum_{i=1}^{n}X_{i})^2}] \quad \text{ for } \quad Z_i=Y_i-\mu(X_i)  \\
&= \E\qty[ \frac{\sum_{i=1}^{n} \E\qty{Z_i^2 \mid X_i}}{\qty(\sum_{i=1}^{n}X_{i})^2}] \quad \text{because cross-terms are zero}.
\end{aligned}
$$

- All we've got to do now is recognize what we have. 
- Let's look at a term in the numerator.
  - We'll use iterated expecations and linearity.
  - Then the indicator trick.

$$
\small{
\begin{aligned}
\E\qty[Z_i^2 \mid X_i] 
&= \E\qty[ X_i\qty{Y_i - \mu(X_i)}^2 \mid X_i] 
= X_i\E\qty[\qty{Y_i - \mu(X_i)}^2 \mid X_i] 
= X_i\Var\qty[Y_i \mid X_i] = X_i \sigma^2(X_i) 
= X_i\sigma^2(1) 
\end{aligned}
}
$$

- We finish up by substituting this back into what we started with.

$$
\small{ 
\begin{aligned}
\Var\qty[ \hat \mu(1) ] 
&= \E\qty[ \frac{\sum_{i=1}^{n} \E\qty{Z_i^2 \mid X_i}}{\qty(\sum_{i=1}^{n}X_{i})^2}]  
= \E\qty[ \frac{\sum_{i=1}^{n} X_i \sigma^2(1)}{\qty(\sum_{i=1}^{n}X_{i})^2}] 
= \E\qty[ \frac{\cancel{\sum_{i=1}^{n} X_i }\sigma^2(1)}{\qty(\sum_{i=1}^{n}X_{i})^{\cancel{2}}}] 
= \E\qty[ \frac{\sigma^2(1)}{\sum_{i=1}^{n}X_{i}}] = \sigma^2(1) \times \E\qty[ \frac{1}{N_1}]. 
\end{aligned}
}
$$
