---
title: "Potential Outcomes and Causality"
---

## A Campaign Finance Example

{{< include shared-code.qmd >}}

```{r}
#| label: ggw-donations-data 


set.seed(1) 
m = 12000 
pi = function(x) { rep(1/2,length(x)) }
pi2 = function(x) { 1/(1+exp(-(x-60)/8 )) }
mu = function(w, x) { 2 + 10*((x-50)/30 + w*((x-50)/30)^2/2) }  

X = 50+round(30*runif(m))
W = 0 + (runif(m) <= pi(X))
epsilon0 = runif(m, min=-1, max=1)*mu(0,X) 
epsilon1 = runif(m, min=-1, max=1)*mu(1,X)
Y0 = mu(0,X) + epsilon0
Y1 = mu(1,X) + epsilon1

pop = data.frame(X=X, W=W, Y=Y0*(1-W) + Y1*W, Y0=Y0, Y1=Y1)
counterfactual.pop = data.frame(X=X, W=W, Y=Y0*W + Y1*(1-W))  

n = 4000
data = pop[sample(nrow(pop), n, replace=TRUE),]
```

```{r}
#| label: ggw-dontations-scatter-plots
plotX = function(X,W,sep=.15) { X + sep*(2*W-1) }
pop.scatter = ggplot(pop) + 
  geom_point(aes(x=plotX(X,W), y=Y, color=factor(W)), alpha=.025, position=position_jitter(w=.1, h=0, seed=1), size=.2) + 
  guides(color='none')
counterfactual.pop.scatter = ggplot(counterfactual.pop) +
  geom_point(aes(x=plotX(X,W), y=Y, color=factor(W)), alpha=.025, position=position_jitter(w=.1, h=0, seed=1), size=.2) + 
  guides(color='none')
scatter = ggplot(data) + 
  geom_point(aes(x=plotX(X,W), y=Y, color=factor(W)), alpha=.3, size=.5, position=position_jitter(w=.1, h=0, seed=1)) + 
  guides(color='none')

po.scatter = ggplot(pop) + 
  geom_point(aes(x=plotX(X,0), y=Y0), color=groupa, alpha=.025, size=.2) + 
  geom_point(aes(x=plotX(X,1), y=Y1), color=groupb, alpha=.025, size=.2) + 
  stat_summary(aes(x=plotX(X,0), y=Y0), geom="pointrange", alpha=.8, fun.data=mean_sd, size=.5, color=groupa) + 
  stat_summary(aes(x=plotX(X,1), y=Y1), geom="pointrange", alpha=.8, fun.data=mean_sd, size=.5, color=groupb) +
  guides(color='none')

big.coords = coord_cartesian(xlim=c(50,80), ylim=c(-1.5,2.5))
narrow.coords = coord_cartesian(xlim=c(60,70), ylim=c(-1.5,2.5))
tight.coords = coord_cartesian(xlim=c(50,80), ylim=c(0,1.5))

annotated.pop.scatter = pop.scatter +  
  stat_summary(aes(x=plotX(X,W), y=Y, color=factor(W)), geom="pointrange", alpha=.8,
                 fun=mean, 
                 fun.min = function(x) mean(x) - sd(x), 
                 fun.max = function(x) mean(x) + sd(x))

annotated.counterfactual.pop.scatter = counterfactual.pop.scatter +  
  stat_summary(aes(x=plotX(X,W), y=Y, color=factor(W)), geom="pointrange", alpha=.8,
                 fun=mean, 
                 fun.min = function(x) mean(x) - sd(x), 
                 fun.max = function(x) mean(x) + sd(x))


proportion.scale = 200
scale = list(scale_y_continuous(breaks = c(0, 10, 20), 
                           labels=scales::dollar_format(),
                           sec.axis=sec_axis(~./proportion.scale, name='Proportion')),
               scale_x_continuous(breaks=seq(50,80,by=10)))

annotated.scatter = scatter + 
  stat_summary(aes(x=plotX(X,W), y=Y, color=factor(W)), 
               geom="pointrange", alpha=.5, size=.35, fun.data=mean_sd) + 
  stat_summary(aes(x=plotX(X,W), y=Y, color=factor(W)), 
               geom="line", alpha=.8, linewidth=.5, fun.data=mean_sd) +
  guides(color='none', fill='none')

annotated.scatter.with.bars = annotated.scatter + 
  geom_bar(aes(x=X, y=20000*after_stat(prop), fill=factor(W)), 
           alpha=.2, linewidth=.1, position="identity") 
```


```{r}
# plotting options
# the colors are given css definitions in class.scss. If you change them here, change them there.
default.dot.size = 3
default.mean.size  = 1.5
mean.shape = 23
filled.circ = 19
empty.circ = 1
x.circ = 13
gray = '#808080'
counterfactual.color = '#FF00FF'
sampled.color = '#0000FF'
randomized.color = '#000000'
aggregate.mean.shape = 24


# give each member of the population jittered coordinates
set.seed(1)
jitter.width = 2 
plotX2 = function(W,X) { X + 1.1*jitter.width*(2*W-1) }
default.shift = function(d,multiplier=1) { 
  jitter.width*multiplier
}

po.means = function(d, shift=default.shift(d, .5), color=counterfactual.color, realized=FALSE, size=default.mean.size) { 
  list(stat_summary(aes(x=plotX2(0,X)-shift, y=Y0), fill=groupa, 
        geom="pointrange", fun.data=mean_sd, size=size, shape=mean.shape, color=color, 
        data=if(realized) d[d$W==0,] else d),
      stat_summary(aes(x=plotX2(1,X)+shift, y=Y1), fill=groupb, 
        geom="pointrange", fun.data=mean_sd, size=size, shape=mean.shape, color=color, 
        data=if(realized) d[d$W==1,] else d))
}
sampled.po.means = function(d,shift=default.shift(d,.25), color=sampled.color, size=default.mean.size) { 
  po.means(d[d$sampled==1,], shift, color, realized=FALSE, size=size) 
}
randomized.means = function(d,shift=default.shift(d,0), color=randomized.color, size=default.mean.size) { 
  po.means(d[d$sampled==1,], shift, color, realized=TRUE, size=size) 
}

aggregate.po.means = function(d, shift=default.shift(d,.75), color=counterfactual.color, realized=FALSE, size=default.mean.size) { 
  d0 = if(realized) { d[d$W==0,] } else d
  d1 = if(realized) { d[d$W==1,] } else d
  list(geom_pointrange(aes(x=x, y=y, ymax=y+sd, ymin=y-sd), fill=groupa, 
        shape=aggregate.mean.shape, size=size, color=color,
        data=data.frame(x=mean(d0$meanX)-shift, y=mean(d0$Y0), sd=sd(d0$Y0))),
      geom_pointrange(aes(x=x, y=y, ymax=y+sd, ymin=y-sd), fill=groupb, 
        shape=aggregate.mean.shape, size=size, color=color,
        data=data.frame(x=mean(d1$meanX)+shift, y=mean(d1$Y1), sd=sd(d1$Y1))))
}
sampled.aggregate.po.means = function(d,shift=default.shift(d,.5), color=sampled.color, size=default.mean.size) { 
  aggregate.po.means(d[d$sampled==1,], shift, color, realized=FALSE, size=size) 
}
randomized.aggregate.means = function(d,shift=default.shift(d,.25), color=randomized.color, size=default.mean.size) { 
  aggregate.po.means(d[d$sampled==1,], shift, color, realized=TRUE, size=size) 
} 

potential.outcome.plot = function(pop, size=default.dot.size, shape=filled.circ, segment.alpha=.05) { 
  p = ggplot(pop) + 
    geom_point(aes(x=plotX2(0,jX), y=Y0), color=groupa, alpha=.7, size=size, shape=shape) + 
    geom_point(aes(x=plotX2(1,jX), y=Y1), color=groupb, alpha=.7, size=size, shape=shape) + 
    geom_segment(aes(x=plotX2(0,jX), xend=plotX2(1,jX), y=Y0, yend=Y1), 
    alpha=segment.alpha, linewidth=.5, color=gray) +
    guides(color='none', fill='none') 
  p 
}

sampled.potential.outcome.plot = function(pop, size=default.dot.size) {
  pop$shape     = ifelse(pop$sampled, filled.circ, empty.circ)
  pop$linewidth = ifelse(pop$sampled, .5, .1)
  pop$alpha = ifelse(pop$sampled, .7, .1)
  p = ggplot(pop) + 
    geom_point(aes(x=plotX2(0,jX), y=Y0, shape=I(shape), alpha=I(alpha)), color=groupa,size=size) + 
    geom_point(aes(x=plotX2(1,jX), y=Y1, shape=I(shape), alpha=I(alpha)), color=groupb, size=size) + 
    geom_segment(aes(x=plotX2(0,jX), xend=plotX2(1,jX), y=Y0, yend=Y1, alpha=I(alpha), linewidth=I(linewidth)), 
      alpha=.05, color=gray) +
    guides(color='none', fill='none', shape='none', linewidth='none', alpha='none')
  #+ scale_linewidth(range=c(0,.5)) 
  p 
}

sw.choice = function(pop, a, b, c) {
  ifelse(pop$sampled, 
    ifelse(pop$W, a, b), 
    c)
}
randomized.plot = function(pop, size=default.dot.size, segment.alpha=.05) {
  pop$shape0    = sw.choice(pop, x.circ, filled.circ, empty.circ)
  pop$alpha0    = sw.choice(pop, .3, .7, .1)
  pop$shape1    = sw.choice(pop, filled.circ, x.circ, empty.circ)
  pop$alpha1    = sw.choice(pop, .7, .3, .1)
  pop$color     = sw.choice(pop, twocolor.green, twocolor.red, gray)
  pop$linewidth = sw.choice(pop, .5, .5, .1)
  p = ggplot(pop) + 
    geom_point(aes(x=plotX2(0,jX), y=Y0, shape=I(shape0), alpha=I(alpha0)), color=groupa, size=size) + 
    geom_point(aes(x=plotX2(1,jX), y=Y1, shape=I(shape1), alpha=I(alpha1)), color=groupb, size=size) + 
    geom_segment(aes(x=plotX2(0,jX), xend=plotX2(1,jX), y=Y0, yend=Y1, 
    color=I(color), linewidth=I(linewidth)), alpha=segment.alpha) +
    guides(color='none', fill='none', shape='none', linewidth='none', alpha='none') 

  p 
}

po.and.means = function(pop, dot.size=default.dot.size, mean.size=default.mean.size) { 
  potential.outcome.plot(pop, size=dot.size) + 
    po.means(pop, size=mean.size)
}
sampled.and.means = function(pop, dot.size=default.dot.size, mean.size=default.mean.size) { 
  sampled.potential.outcome.plot(pop, size=dot.size) + 
    po.means(pop, size=mean.size) +
    sampled.po.means(pop, size=mean.size)
}
randomized.and.means = function(pop, dot.size=default.dot.size, mean.size=default.mean.size) { 
  randomized.plot(pop, size=dot.size) + 
    po.means(pop, size=mean.size) +
    sampled.po.means(pop, size=mean.size) +
    randomized.means(pop, size=mean.size)
}

```


```{r}
annotated.scatter + scale
```


- Let's think back to the 2006 Michigan Primary we discussed in our first lecture. 
- But instead of mailing letters to increase turnout, we're trying to get donations. 
- We have a population of potential donors: a list of people who donated in 2004. 
- We're considering two ways of contacting them: [an email]{.groupa} or [a phone call]{.groupb}.
- We run a pilot study in which we ...
  - Sample without replacement to select `r n` potential donors from our list
  - Flip a coin for each one to choose between an email or call. 
  - Contact them as dictated by the coin and record their donation.
- The results, broken down by the donor's age, are shown above.

## What We Want From Our Pilot Study

```{r}
annotated.scatter + scale
```

- Calling costs more. We want to know whether it's worth it.
- We want to know how much higher our average donation would be higher if we called everyone.
- One simple thing we could do is look at the raw difference in mean donations between our two groups.

$$
\color{gray}
\begin{aligned}
\hat\Delta_{\text{raw}} =\ingroupb{\frac{1}{N_1} \sum_{i:W_i=1} Y_i} - \ingroupa{\frac{1}{N_0} \sum_{i:W_i=0} Y_i}  
\approx \ingroupb{`r round(mean(data$Y[data$W==1]), 2)`} - \ingroupa{`r round(mean(data$Y[data$W==0]), 2)`} \approx `r round(mean(data$Y[data$W==1]) - mean(data$Y[data$W==0]), 2)`
\end{aligned}
$$

- Do you think this works?

## What This Lecture Is About
```{r}

proportion.scale = 200
annotated.scatter + 
  geom_bar(aes(x=plotX(X,W), y=proportion.scale*after_stat(prop), color=factor(W), fill=factor(W)), 
    alpha=.1, linewidth=.1, position='identity') +
  scale_y_continuous(breaks=c(0,10,20), labels=scales::dollar_format(), 
    sec.axis=sec_axis(~./proportion.scale, name='Proportion')) +
  guides(color='none', fill='none')
```

- We're going to work out how to answer to 'what if' questions like this.
  - You can think of them as questions about *causality*.
  - The difference between what would happen if we call vs if we email is the *effect* of calling 
    - (instead of emailing).
- You'll learn ...
  - How to describe effects in formal mathematical language. 
  - How to use randomization to get unbiased estimates of those effects.
  - What this has to do with covariate shift.

# Analyzing Randomized Experiments

## The Potential Outcomes Formalism 

```{r}
y1 = c(6,0,4,7,8,2)
y0 = c(2,0,1,7,4,0)
x = c(55,55,55,75,75,75)
w = c(1,0,1,0,1,0)
yw = ifelse(w==1, y1, y0)
scale_x_tiny = list(scale_x_continuous(breaks=c(55,75)), coord_cartesian(xlim=c(50,80)))
tiny.pop = data.frame(X=x, Y1=y1, Y0=y0, jX=x+0*jitter.width*c(-1,0,1,-1,0,1), W=w, Yw=yw)
```


:::: {.columns}
::: {.column }
```{r}
#| fig-width: 7
#| fig-asp: .5

potential.outcome.plot(tiny.pop, segment.alpha=.2) + scale_x_tiny
```

:::
::: {.column}

$j$ | $x_j$ `| $y_j(1)$ | $y_j(0)$ | $\tau_j$ 
--- | --- | --- | --- | --- 
1 | `r x[1]` | `r y1[1]` | `r y0[1]` | `r y1[1] - y0[1]` 
2 | `r x[2]` | `r y1[2]` | `r y0[2]` | `r y1[2] - y0[2]` 
3 | `r x[3]` | `r y1[3]` | `r y0[3]` | `r y1[3] - y0[3]` 
4 | `r x[4]` | `r y1[4]` | `r y0[4]` | `r y1[4] - y0[4]` 
5 | `r x[5]` | `r y1[5]` | `r y0[5]` | `r y1[5] - y0[5]` 
6 | `r x[6]` | `r y1[6]` | `r y0[6]` | `r y1[6] - y0[6]` 

:::
::::

::: {.panel-tabset}
### The Framework 

- To reason about cause and effect formally, we use **potential outcomes**. 
  - $\ingroupa{y_j(0)}$ is the amount person $j$ would donate if they were emailed. 
  - $\ingroupb{y_j(1)}$ is the amount person $j$ would donate if they were called.
- *Terminology*. We call the actions we can take **treatments**.  
  - Each individual in our population has a potential outcome for each treatment. 
  - In this case, we have two treatments and therefore two potential outcomes---a pair---for each individual.
  - Each individual's potential outcomes are drawn as a connected pair of dots, one [red]{.groupa} and one [green]{.groupb}.

### The Target 
- How much higher our average donation would be if we [called]{.groupb} everyone vs. if we [emailed]{.groupa} everyone.
- *Question*. How do we write our target in terms of these potential outcomes?

::: {.fragment}
$$
\color{gray}
\begin{aligned}
\text{target} = \ingroupa{\frac1m \sum_{j=1}^m y_j(1)} - \ingroupb{\frac1m \sum_{j=1}^m y_j(0)} 
= \frac1m \sum_{j=1}^m \tau_j \quad \text{for} \quad \tau_j = y_j(1) - y_j(0)
\end{aligned}
$$

- We call the differences $\tau_j$ an *individual treatment effects*. 
  - $\tau_j$ is the effect of calling (vs. emailing) person $j$ in our population. 
  - We call the average of these individualized effects the *average treatment effect*.

:::

:::

## Potential Outcomes as Functions 

:::: {.columns}
::: {.column }
```{r}
#| fig-width: 7
#| fig-asp: .5

potential.outcome.plot(tiny.pop, segment.alpha=.2) + scale_x_tiny
```

:::
::: {.column}

$j$ | $x_j$ `| $y_j(1)$ | $y_j(0)$ | $\tau_j$ 
--- | --- | --- | --- | --- 
1 | `r x[1]` | `r y1[1]` | `r y0[1]` | `r y1[1] - y0[1]` 
2 | `r x[2]` | `r y1[2]` | `r y0[2]` | `r y1[2] - y0[2]` 
3 | `r x[3]` | `r y1[3]` | `r y0[3]` | `r y1[3] - y0[3]` 
4 | `r x[4]` | `r y1[4]` | `r y0[4]` | `r y1[4] - y0[4]` 
5 | `r x[5]` | `r y1[5]` | `r y0[5]` | `r y1[5] - y0[5]` 
6 | `r x[6]` | `r y1[6]` | `r y0[6]` | `r y1[6] - y0[6]` 

:::
::::

::: {.panel-tabset}
### The Idea 

- It's convenient to think of each individual's potential outcomes as *one function* instead of *two values*.
  - $y_1(\cdot)$ is the function defined like this.
$$
y_1(w) = \begin{cases} 
\underset{\veryfaint{y_1(1)}}{`r y1[1]`} & \text{if } w=1 \\ 
\underset{\veryfaint{y_1(0)}}{`r y0[1]`} & \text{if } w=0 
\end{cases}
$$
  - $y_j(\cdot)$ is the function defined like this. The only way the notation really lets us define it.  
$$
y_j(w) = \begin{cases}  
y_j(1) & \text{if } w=1 \\ 
y_j(0) & \text{if } w=0 
\end{cases}
$$

- That individual's **realized outcome** is ...
  - the value $y_j(w_j)$ their potential outcome function returns 
  - when we plug in the treatment $w_j$ that they actually receive.

### Why Think Like This? 

- It works for any number of treatments.
- It's useful for thinking about *random treatments*. 
  - If $W$ is a random variable taking on $k$ values, so is $y_j(W)$.
  - The alternative is to write $1_{=1}(W)y_j(1) + 1_{=0}(W)y_j(0)$ where we'd write $y_j(W)$. 
    - Or more generally $\sum_w 1_{=w}(W)y_j(w)$.
    - In function terms, that's using the *indicator trick* everywhere instead of just where we need it.
    - People do it, but it tends to make things more difficult than they need to be.  

### A Quick Exercise 
   
- Suppose $W$ is a coin flip: a random variable taking on the values $0$ and $1$ with equal probability.
- Write out a table describing the joint distribution of $W$ and $y_3(W)$.
- Step forward a slide to see the solution. It'll be on the right, so write yours on the left.

::: {.fragment}
:::: columns
::: {.column width="70%"}
\ 
:::
::: {.column width="30%"}
$\text{probability}$ | $W$ | $y_3(W)$
--- | --- | ---
1/2 | 0 | `r y0[3]`
1/2 | 1 | `r y1[3]`
:::
::::
:::

:::

## The Fundamental Problem of Causal Inference 

::::: columns 
::: column 

:::: r-stack
::: {.fragment .fade-out data-fragment-index="0"}

```{r}
#| fig-width: 7
#| fig-asp: .5

tiny.pop |> 
  potential.outcome.plot(shape=empty.circ, segment.alpha=.2)
```

:::
::: {.fragment .current-visible data-fragment-index="0"}

```{r}
#| fig-width: 7
#| fig-asp: .5

tiny.pop |> 
  mutate(sampled=TRUE) |>
  randomized.plot(segment.alpha=.2)
```

:::
::::

:::

::: column

:::: r-stack
::: {.fragment .fade-out data-fragment-index="0"}
$j$ | $x_j$ | $y_j(1)$ | $y_j(0)$ | $\tau_j$ | $w_j$ | $y_j(w_j)$ 
--- | --- | --- | --- | --- | --- | --- 
1 | `r x[1]` | `r y1[1]` | `r y0[1]` | `r y1[1] - y0[1]` | `r w[1]` | [`r yw[1]`]{.bgcolor} 
2 | `r x[2]` | `r y1[2]` | `r y0[2]` | `r y1[2] - y0[2]` | `r w[2]` | [`r yw[2]`]{.bgcolor} 
3 | `r x[3]` | `r y1[3]` | `r y0[3]` | `r y1[3] - y0[3]` | `r w[3]` | [`r yw[3]`]{.bgcolor} 
4 | `r x[4]` | `r y1[4]` | `r y0[4]` | `r y1[4] - y0[4]` | `r w[4]` | [`r yw[4]`]{.bgcolor} 
5 | `r x[5]` | `r y1[5]` | `r y0[5]` | `r y1[5] - y0[5]` | `r w[5]` | [`r yw[5]`]{.bgcolor} 
6 | `r x[6]` | `r y1[6]` | `r y0[6]` | `r y1[6] - y0[6]` | `r w[6]` | [`r yw[6]`]{.bgcolor} 
:::
::: {.fragment .current-visible data-fragment-index="0"}
$j$ | $x_j$ | $y_j(1)$ | $y_j(0)$ | $\tau_j$ | $w_j$ | $y_j(w_j)$ 
--- | --- | --- | --- | --- | --- | --- 
1 | `r x[1]` | `r y1[1]` | `r y0[1]` | `r y1[1] - y0[1]` | `r w[1]` | [`r yw[1]`]{.bgcolor} 
2 | `r x[2]` | `r y1[2]` | `r y0[2]` | `r y1[2] - y0[2]` | `r w[2]` | [`r yw[2]`]{.bgcolor} 
3 | `r x[3]` | `r y1[3]` | `r y0[3]` | `r y1[3] - y0[3]` | `r w[3]` | [`r yw[3]`]{.bgcolor} 
4 | `r x[4]` | `r y1[4]` | `r y0[4]` | `r y1[4] - y0[4]` | `r w[4]` | [`r yw[4]`]{.bgcolor} 
5 | `r x[5]` | `r y1[5]` | `r y0[5]` | `r y1[5] - y0[5]` | `r w[5]` | [`r yw[5]`]{.bgcolor} 
6 | `r x[6]` | `r y1[6]` | `r y0[6]` | `r y1[6] - y0[6]` | `r w[6]` | [`r yw[6]`]{.bgcolor} 
:::
::::

::: 
::::

::: {.panel-tabset}
### The Problem
- In concrete terms.
  - You can't both call and not call someone.
  - So you can see what happens when you call someone or what happens when you don't, but not both. 
- In abstract terms. 
  - Each individual can take only one treatment.^[This is inconsistent with everyday use of the term *treatment* because you can, e.g., take both ibuprofen and acetaminophen. In potential outcomes language, we'd say that ibuprofen alone, acetaminophen alone, and ibuprofen plus acetaminophen are three different treatments.]
  - So only one of each individual's potential outcomes is realized.
- That means that, even if we can choose everyone's treatment however we want:
  - We can't calculate anybody's *individual treatment effect*  $\tau_j = y_j(1) - y_j(0)$.
  - And we can't calculate the average of them, $\bar\tau = \frac1m \sum_{j=1}^m \tau_j$, either. 

### Check your Understanding 

- Look at the treatments $w_j$ in the table above. For each individual:
  - Fill in the realized outcome $y_j(w_j)$ for each row in the table.
  - Shade in the dot corresponding to the realized outcome in the plot.
  - Draw arrows to indicate the individual treatment effects $\tau_j = y_j(1) - y_j(0)$ in the plot.
    - make them point up for positive effects, down for negative effects.
- Step forward a slide to see the solution. 

:::

## The Fundamental Solution is Randomization

```{r}
randomization.mechanisms = list(
  shuffle=function(pop) { sample(1:nrow(pop)) <= nrow(pop)/2 },
  flip=function(pop) { rbinom(nrow(pop), 1, .5) })

plot.effect.estimate = function(pop, reps=10000, randomize=randomization.mechanisms$shuffle, ...) {
  estimates = 1:reps |> map_vec(function(.) { 
    W = randomize(pop)
    mean(pop$Y1[W==1]) - mean(pop$Y0[W==0])
  })
  
 ggplot() + 
  geom_bar(aes(x=estimates, y=after_stat(prop)), alpha=.3, linewidth=.1) +
  with_annotations(estimates, estimand=mean(pop$Y1 - pop$Y0), highlight.dots=0, 
                   interval.width=width(estimates), ...)
}
```

```{r}
#| layout-ncol: 2
#| fig-width: 7
#| fig-cap: 
#|   - The realized outcomes after randomizing treatment (once)
#|   - The sampling distribution when we randomize treatment (over and over). 
tiny.pop |> 
  mutate(sampled=TRUE) |>
  randomized.plot(segment.alpha=.2)
plot.effect.estimate(tiny.pop, height=.05)
```

::: {.panel-tabset}
### The Recipe 

- Step 1. *Randomize* treatment. 
  - Shuffle a deck of $m$ cards labeled 1...m. 
  - Treat individual $j$ if the card labeled $j$ is in the top half of the deck.^[This is sort of like a sampling-without-replacement version of flipping a coin for each individual. The marginal probability of treatment is the same ($1/2$), but treatments for different individuals aren't independent like they would be if we flipped a coin. We could flip coins instead, but we'd wind up treating everybody or nobody sometimes, especially when our population is small.]

$$
W_j =  \begin{cases} 0 & \text{with probability } 1/2 \\ 1   & \text{with probability } 1/2 \end{cases}  \qqtext{ and } \sum_{j=1}^m 1_{=w}(W_j) = m/2 \qfor w \in \{0,1\}
$$

- Step 2. *Compare* the mean realized outcome for groups receiving different treatments.
 
$$
\hat\tau = \frac{1}{m/2} \sum_{j:W_j=1} y_j(W_j) - \frac{1}{m/2} \sum_{j:W_j=0} y_j(W_j) 
$$
 

### Visualizing the Result 

- Above, we see the sampling distribution^[Some people prefer to say 'randomization distribution' in this context because we're not sampling in individuals from our population---we get a dot for each one.] of this estimator $\hat\tau$ with all the usual annotations.
  - The [target]{.target}, the average treatment effect $\bar\tau = \frac1m \sum_{j=1}^m \tau_j$, is drawn in green. 
  - The sampling distribution's mean is drawn in blue. 
- It looks like our estimator is unbiased. [|]{.blue .target}

:::

## Proving Unbiasedness

$$
\begin{aligned}
\hat\mu(w) 
&\explain{=}{This is our groups mean.} 
\frac{1}{m/2}\sum_{j:W_j=w}y_j(W_j) \\ 
&\explain{=}{We rewrite our subpopulation sums a a sum over the entire population using group indicators.}
\frac{1}{m/2}\sum_{j=1}^m y_j(W_j) \ 1_{=w}(W_j)  \\
&\explain{=}{We use the indicator trick. $f(W)1_{=w}(W) = f(w)1_{=w}(W)$ for any function $f$.}
\frac{1}{m/2}\sum_{j=1}^m y_j(w) 1_{=w}(W_j). \\ 
\E[\hat\mu(w)] 
&\explain{=}{distributing the expectation using linearity} 
\frac{1}{m/2}\sum_{j=1}^m y_j(w) \ \E[1_{=w}(W_j)] \\
&\explain{=}{expectations of indicators are probabilities and we're doing equal-probability randomization: $\E[1_{=w}(W)] = \P(W=w) = 1/2$}
\frac{1}{m/2}\sum_{j=1}^m y_j(w) \ \frac12 \\ 
&\explain{=}{multiplying: $1/(m/2) \times 1/2 = 1/m$}
\frac{1}{m} \sum_{j=1}^m y_j(w)  \\
\E[\hat\tau] 
&\explain{=}{distributing the expectation using linearity} 
\frac1m \sum_{j=1}^m y_j(1) - \frac1m \sum_{j=1}^m y_j(0) \\ 
&\explain{=}{grouping corresponding terms} 
\frac1m\sum_{j=1}^m \underset{\faint{\tau_j}}{\{ y_j(1) - y_j(0) \}} = \bar\tau.
\end{aligned}
$$

- If you want to see a similar proof that applies when we randomize via coin flips instead of shuffling, see @sec-unbiasedness-coin-flips.
- We'll do a proof that applies to both randomization mechanisms in the next section, but it'll look a little different.


# Combining Randomization and Sampling 

## The Idea

```{r}
#| label: two-groups-randomized-flip
n = 100
ages = c(55,75)

the.pop = pop[pop$X %in% ages, ] 
the.pop = the.pop[1:400,]
the.pop$jX = the.pop$X + jitter.width*runif(nrow(the.pop), -1,1)
the.pop$meanX = mean(the.pop$X)

sampled.1 = the.pop; sampled.1$sampled = (\() sample(1:nrow(the.pop)) <= n) |> with_seed(1)
sampled.2 = the.pop; sampled.2$sampled = (\() sample(1:nrow(the.pop)) <= n) |> with_seed(2)
sampled.3 = the.pop; sampled.3$sampled = (\() sample(1:nrow(the.pop)) <= n) |> with_seed(3)
randomized.1 = sampled.3; randomized.1$W = (\() rbinom(nrow(the.pop), 1, pi(the.pop$X))) |> with_seed(1)
randomized.2 = sampled.3; randomized.2$W = (\() rbinom(nrow(the.pop), 1, pi(the.pop$X))) |> with_seed(2)
randomized.3 = sampled.3; randomized.3$W = (\() rbinom(nrow(the.pop), 1, pi(the.pop$X))) |> with_seed(3)
scale_x = list(scale_x_continuous(breaks=c(55,75)), coord_cartesian(xlim=c(50,80)))
```
:::: {.columns}
::: {.column width="50%"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5
potential.outcome.plot(the.pop) + scale_x
```

:::
::: {.column width="50%"}

- Let's look at what happens when we randomize treatment to a *sample* drawn from our population.
- Let's start by looking at a subset of our population of potential donors: the ones aged 55 and 75.
- Everyone has two potential outcomes: the amount they donate if untreated (red) and the amount they donate if treated (green)

:::
::::

## Randomization and Sampling
:::: {.columns}
::: {.column width="50%"} 

:::: r-stack
::: {.fragment .fade-out fragment-index="0"}
```{r}
#| fig-width: 7
#| fig-asp: 1.5 
potential.outcome.plot(the.pop) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="0"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 
sampled.potential.outcome.plot(sampled.1) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="1"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 

sampled.potential.outcome.plot(sampled.2) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="2"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 

sampled.potential.outcome.plot(sampled.3) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="3"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5
randomized.plot(randomized.1) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="4"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 

randomized.plot(randomized.2) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="5"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 

randomized.plot(randomized.3) + scale_x
```

:::  
:::: 
::: 

::: {.column width="50%"}

::: {.fragment .semi-fade-out fragment-index="0"}
1. We start with two potential outcomes for each person in our population.
   - Those are the connected dots we see in the plot.
:::
::: {.fragment .fade-in fragment-index="0"}
::: {.fragment .semi-fade-out fragment-index="3"}
2. Then we sample from our population. 
   - The people who aren't sampled fade out in the plot.
   - This is **random**. So different things happen each time we do it.
:::
:::
::: {.fragment .fade-in fragment-index="3"}
3. Each person in our sample flips a coin to determine whether they're treated. 
   - &olcross; marks the potential outcomes that don't happen.
   - This is random, too. 
:::
:::
:::: 



## What Happens to the Means


:::: {.columns}
::: {.column width="50%"}

:::: r-stack
::: {.fragment .fade-out fragment-index="0"}
```{r}
#| fig-width: 7
#| fig-asp: 1.5 

po.and.means(the.pop) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="0"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 

sampled.and.means(sampled.1) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="1"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 

sampled.and.means(sampled.2) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="2"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 

sampled.and.means(sampled.3) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="3"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5

randomized.and.means(randomized.1) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="4"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 

randomized.and.means(randomized.2) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="5"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 

randomized.and.means(randomized.3) + scale_x
```

:::
::::

:::
::: {.column width="50%"}

::: {.fragment .semi-fade-out fragment-index="0"}
1. We start with two potential outcomes for each person in our population.
   - Those are the connected dots we see in the plot.
   - These have [a mean]{.counterfactual}. Often, that's what we're interested in. 
::::
::: {.fragment .fade-in fragment-index="0"}
::: {.fragment .semi-fade-out fragment-index="3"}
2. Then we sample from our population.
   - The people who aren't sampled fade out in the plot.
   - And we can look at the [the means]{.population} of the potential outcomes in the *sample*. They're random.
:::
:::
::: {.fragment .fade-in fragment-index="3"}
3. Each person in the sample flips a coin to determine whether they're treated. 
   - &olcross; marks the potential outcomes that don't happen.
   - We can look at [the means]{.sample} of people who actually flip '[heads]{.groupb}' and '[tails]{.groupa}', too. More randomness. 
:::

:::
::::


## What Happens to (Not Within-Group) Means

:::: {.columns}
::: {.column width="50%"}

:::: r-stack
::: {.fragment .fade-out fragment-index="0"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5

po.and.means(the.pop) + 
  aggregate.po.means(the.pop) + 
  scale_x
```

:::
::: {.fragment .current-visible fragment-index="0"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 

sampled.and.means(sampled.1) + 
  aggregate.po.means(sampled.1) + 
  sampled.aggregate.po.means(sampled.1) + 
  scale_x
```

:::
::: {.fragment .current-visible fragment-index="1"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 

sampled.and.means(sampled.2) + 
  aggregate.po.means(sampled.2) + 
  sampled.aggregate.po.means(sampled.2) +
  scale_x
```

:::
::: {.fragment .current-visible fragment-index="2"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 

sampled.and.means(sampled.3) + 
  aggregate.po.means(sampled.3) + 
  sampled.aggregate.po.means(sampled.3) +
  scale_x
```

:::
::: {.fragment .current-visible fragment-index="3"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 

randomized.and.means(randomized.1) + 
  aggregate.po.means(randomized.1) + 
  sampled.aggregate.po.means(randomized.1) +
  randomized.aggregate.means(randomized.1) +
  scale_x
```

:::
::: {.fragment .current-visible fragment-index="4"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 
randomized.and.means(randomized.2) + 
  aggregate.po.means(randomized.2) + 
  sampled.aggregate.po.means(randomized.2) +
  randomized.aggregate.means(randomized.2) +
  scale_x
```

:::
::: {.fragment .current-visible fragment-index="5"}
```{r}
#| fig-width: 7
#| fig-asp: 1.5 

randomized.and.means(randomized.3) + 
  aggregate.po.means(randomized.3) + 
  sampled.aggregate.po.means(randomized.3) +
  randomized.aggregate.means(randomized.3) +
  scale_x
```

:::
::::

:::
::: {.column width="50%"}

::: {.fragment .semi-fade-out fragment-index="0"}
1. We start with two potential outcomes for each person in our population.
   - Those are the connected dots we see in the plot.
   - These have [a mean]{.counterfactual} outright---ignoring $X$. Often, that's what we're interested in.
   
::::

::: {.fragment .fade-in fragment-index="0"}
::: {.fragment .semi-fade-out fragment-index="3"}
2. Then we sample from our population. 
    - This is random, too.
   - These have [a mean]{.population}, too. And it is, of course, random. It changes if our sample changes.
   
:::
:::
::: {.fragment .fade-in fragment-index="3"}
3. Each person in the sample flips a coin to determine whether they're treated.
   - &olcross; marks the potential outcomes that don't happen.
   - We can look at [the means]{.sample} of people who actually flip '[heads]{.groupb}' and '[tails]{.groupa}', too. More randomness. 
    
:::

::: {.fragment .fade-in fragment-index="4"}
How different are these means? 
:::
::: {.fragment .fade-in fragment-index="5"}
Not very. In fact, the means in the red and green subsamples are unbiased estimators of the means of the red and green potential outcomes.
$$
\E\qty[\frac{1}{N_w}\sum_{i:W_i=w} Y_i] = \frac{1}{m}\sum_{j=1}^m y_j(w) \qfor N_w = \sum_{i=1}^n 1_{=w}(W_i)
$$
:::

:::
::::

## Sampling and Randomization

:::: {.columns}

::: {.column}
```{r}
#| fig-width: 7
randomized.and.means(randomized.3)
```

:::

::: {.column}

$$
\begin{array}{cccc|cccc}
j & W_j & y_j(W_j) & x_j & y_j(1) & y_j(0) & \tau_j \\
\hline
1 &  &   & 55 & 6 & 5 & 1 \\
2 &  &   & 55 & 0 & 0 & 0 \\
3 &  &   & 55 & 4 & 4 & 0 \\
4 &  &   & 75 & 7 & 0 & 7 \\
5 &  &   & 75 & 4 & 0 & 4 \\
6 &  &   & 75 & 2 & 0 & 2 \\
\end{array}
$$

:::
::::

Let's review the process that gives us our observed treatment+covariate+outcome triples $(W_1, X_1, Y_1) \ldots (W_n,X_n,Y_n)$.

1. We draw covariate+potential-outcomes triples $\{X_i, Y_i(0), Y_i(1)\}$ uniformly-at-random from the population of all such triples $\{x_1, y_1(0), y_1(1)\}, \ldots, \{x_m, y_m(0), y_m(1)\}$.
$$
\{ X_i,Y_i(0), Y_i(1)\} = \{x_J, y_J(0), y_J(1)\} \qfor J=1 \ldots m \qqtext{ each with probability } 1/m
$$

2. We choose treatments $W_1 \ldots W_n$ by some random mechanism, independent of everything else. These determines the potential outcomes we observe.

$$
Y_i = Y_i(W_i) \qqtext{ for } W_1 \ldots W_n \qqtext{independent of} \{X_1,Y_1(0), Y_1(1)\} \ldots \{X_n, Y_n(0), Y_n(1)\}
$$

- What we visualized was a special case in which we ...
  - used sampling with replacement in Step 1 
  - used coin flips in Step 2
- What we looked at in the previous section was a special case in which we 
  - used sampling without replacement in Step 1.^[Why? If we draw a sample *of the same size as the population* without replacement, we get the whole population.]
  - used shuffling in Step 2.

## Causal Identification

```{r}
randomized.and.means(randomized.3) + 
  aggregate.po.means(randomized.3) + 
  sampled.aggregate.po.means(randomized.3) +
  randomized.aggregate.means(randomized.3) +
  scale_x
```


- With this sampling and randomization scheme ... 
  - we can rewrite our potential outcome means ...
  - as expected values involving the random variables that we observe.

$$
\frac{1}{m}\sum_{j=1}^m y_j(w) = \E[Y_i \mid W_i=w]
$$

::: {.panel-tabset}
### Proof 

$$
\begin{aligned}
\frac{1}{m}\sum_{j=1}^m y_j(w)
&\explain{=}{sampling uniformly-at-random} 
\E[Y_i(w)] \\
&\explain{=}{irrelevance of independent conditioning variables} 
\E[Y_i(w) \mid W_i=w] \\
&\explain{=}{if we've flipped $W_i=w$, $Y_i(W_i)=Y_i(w)$. It's a bit like the indicator trick.} 
\E[Y_i(W_i) \mid W_i=w] \\
&\explain{=}{Definition. $Y_i=Y_i(W_i)$.}
\E[Y_i \mid W_i=w] = \mu(w)
\end{aligned}
$$


### Terminology
- We call this rewriting process **identification**. 
  - We've 'identified' a summary of the potential outcomes if ...
  - ... we have an equivalent formula for it that doesn't involve potential outcomes.
  - We need randomization to have equivalences like this. 

:::

## Unbiasedness

```{r}
randomized.and.means(randomized.3) + 
  aggregate.po.means(randomized.3) + 
  sampled.aggregate.po.means(randomized.3) +
  randomized.aggregate.means(randomized.3) +
  scale_x
```

- This identification result reduces today's unbiasedness question to one we've addressed before: the unbiasedness of column means.

$$
\E[\hat\mu(w)] \explain{=}{unbiasedness of column means} \mu(w) \explain{=}{identification} \E[Y_i(w)]
$$

- When we addressed that in [Lecture 6](`r lecture(6)`), we assumed pairs $(W_i,Y_i)$ were sampled with replacement.
- We're being more general here, so I'll give you a direct proof of unbiasedness too. It's very similar to Lecture 6's.

$$
\begin{aligned}
\E\qty[\frac{1}{N_w}\sum_{i:W_i=w} Y_i ]
&\explain{=}{via the law of iterated expectations} 
\E\qty[ \E\qty[ \frac{1}{N_w}\sum_{i=1}^n Y_i \mid W_1 \ldots W_n] ] \\
&\explain{=}{linearity of conditional expectation} 
\E\qty[ \frac{1}{N_w}\sum_{i=1}^n \E[Y_i \mid W_1 \ldots W_n] \  1_{=w}(W_i) ] \\ 
&\explain{=}{indicator trick} 
\E\qty[ \frac{1}{N_w}\sum_{i=1}^n \E[Y_i(w) \mid W_1 \ldots W_n] \  1_{=w}(W_i) ] \\ 
&\explain{=}{irrelevance of independent conditioning variables. The assignments $W_1 \ldots W_n$ are independent of $Y_i(w)$. } 
\E\qty[ \frac{1}{N_w}\sum_{i=1}^n \E[Y_i(w)] \ 1_{=w}(W_i) ] \\
&\explain{=}{via linearity, i.e. pulling out the constant $E[Y_i(w)]$} 
\E[Y_i(w)] \E\qty[ \frac{1}{N_w}\sum_{i=1}^n 1_{=w}(W_i) ]  = \E[Y_i(w)]\frac{N_w}{N_w} = \E[Y_i(w)] \\
&\explain{=}{sampling uniformly-at-random} \frac{1}{m}\sum_{j=1}^m y_j(w) 
\end{aligned}
$$

# Conditionally Randomized Experiments 

## What is conditional randomization? 


```{r}
#| label: ggw-donations-data-conditional 


set.seed(1) 
m = 12000 
pi = function(x) { 1/(1+exp(-(x-60)/8 )) }
mu = function(w, x) { 10*((x-50)/30 + w*((x-50)/30)^2/2) }  

X = 50+round(30*runif(m))
W = 0 + (runif(m) <= pi(X))
epsilon0 = runif(m, min=-1, max=1)*mu(0,X) 
epsilon1 = runif(m, min=-1, max=1)*mu(1,X)
Y0 = mu(0,X) + epsilon0
Y1 = mu(1,X) + epsilon1

pop = data.frame(X=X, W=W, Y=Y0*(1-W) + Y1*W, Y0=Y0, Y1=Y1)
counterfactual.pop = data.frame(X=X, W=W, Y=Y0*W + Y1*(1-W))  

set.seed(1)
n = 1000
data = pop[sample(nrow(pop), n, replace=FALSE),]
```

```{r}
#| label: ggw-dontations-scatter-plots-conditional 
plotX = function(X,W,sep=.15) { X + sep*(2*W-1) }
pop.scatter = ggplot(pop) + 
  geom_point(aes(x=plotX(X,W), y=Y, color=factor(W)), alpha=.025, position=position_jitter(w=.1, h=0), size=.2) + 
  guides(color='none')
counterfactual.pop.scatter = ggplot(counterfactual.pop) +
  geom_point(aes(x=plotX(X,W), y=Y, color=factor(W)), alpha=.025, position=position_jitter(w=.1, h=0), size=.2) + 
  guides(color='none')
scatter = ggplot(data) + 
  geom_point(aes(x=plotX(X,W), y=Y, color=factor(W)), alpha=.3, size=.5, position=position_jitter(w=.1, h=0)) + 
  guides(color='none')

po.scatter = ggplot(pop) + 
  geom_point(aes(x=plotX(X,0), y=Y0), color=groupa, alpha=.025, size=.2, position=position_jitter(w=.1, h=0)) + 
  geom_point(aes(x=plotX(X,1), y=Y1), color=groupb, alpha=.025, size=.2, position=position_jitter(w=.1, h=0)) + 
  stat_summary(aes(x=plotX(X,0), y=Y0), geom="pointrange", alpha=.8, fun.data=mean_sd, size=.5, color=groupa) + 
  stat_summary(aes(x=plotX(X,1), y=Y1), geom="pointrange", alpha=.8, fun.data=mean_sd, size=.5, color=groupb) +
  guides(color='none')

big.coords = coord_cartesian(xlim=c(50,80), ylim=c(-1.5,2.5))
narrow.coords = coord_cartesian(xlim=c(60,70), ylim=c(-1.5,2.5))
tight.coords = coord_cartesian(xlim=c(50,80), ylim=c(0,1.5))

annotated.pop.scatter = pop.scatter +  
  stat_summary(aes(x=plotX(X,W), y=Y, color=factor(W)), geom="pointrange", alpha=.8,
                 fun=mean, 
                 fun.min = function(x) mean(x) - sd(x), 
                 fun.max = function(x) mean(x) + sd(x))

annotated.counterfactual.pop.scatter = counterfactual.pop.scatter +  
  stat_summary(aes(x=plotX(X,W), y=Y, color=factor(W)), geom="pointrange", alpha=.8,
                 fun=mean, 
                 fun.min = function(x) mean(x) - sd(x), 
                 fun.max = function(x) mean(x) + sd(x))


proportion.scale = 200
scale = list(scale_y_continuous(breaks = c(0, 10, 20), 
                           labels=scales::dollar_format(),
                           sec.axis=sec_axis(~./proportion.scale, name='Proportion')),
               scale_x_continuous(breaks=seq(50,80,by=10), labels=function(x) sprintf('19%d', x)))

annotated.scatter = scatter + 
  stat_summary(aes(x=plotX(X,W), y=Y, color=factor(W)), 
               geom="pointrange", alpha=.5, size=.35, fun.data=mean_sd) + 
  stat_summary(aes(x=plotX(X,W), y=Y, color=factor(W)), 
               geom="line", alpha=.8, linewidth=.5, fun.data=mean_sd) +
  guides(color='none', fill='none')

annotated.scatter.with.bars = annotated.scatter + 
  geom_bar(aes(x=X, y=20000*after_stat(prop), fill=factor(W)), 
           alpha=.2, linewidth=.1, position="identity") 
```


- So far, we have focused on the case that treatment is randomized without looking at anything else.
 - Formally: $W_1 \ldots W_n$ are independent of $\{X_1,Y_1(0), Y_1(1)\} \ldots \{X_n, Y_n(0), Y_n(1) \}$.
 - This is not the only way to randomize! 
- Suppose I suspect phone calls to older people to pay off more than calls to younger ones.
  - Maybe older people have more money to spend on campaigns.
  - Maybe they don't get as annoyed about getting phone calls as young people .
  - I might want to choose the probability a person gets treated (i.e. called) as a function of their age.
- In *conditionally randomized experiments*, that's what we do. 
  - We look at the covariates $X_1 \ldots X_n$ when we randomize. But only the covariates.
  - Formally: $W_1 \ldots W_n$ are *conditionally independent* of $\{Y_1(0), Y_1(1)\} \ldots \{Y_n(0), Y_n(1)\}$ given $X$.^[*Conditional independence* is a new term. We'll define it shortly.]
  - We'll focus on the case that each $W_i$ is a coin flip with heads probability depending on $X_i$.^[There are other ways, e.g. making a deck of cards for each age group, shuffling them, and treating the top $N_{w,x}$ cards on each deck.]

## The Idea
```{r}
#| label: two-groups-randomized
n = 100
ages = c(55,75)

the.pop = pop[pop$X %in% ages, ] 
the.pop = the.pop[1:400,]
the.pop$jX = the.pop$X + jitter.width*runif(nrow(the.pop), -1,1)
the.pop$meanX = mean(the.pop$X)

sampled.1 = the.pop; sampled.1$sampled = (\() sample(1:nrow(the.pop)) <= n) |> with_seed(1)
sampled.2 = the.pop; sampled.2$sampled = (\() sample(1:nrow(the.pop)) <= n) |> with_seed(2)
sampled.3 = the.pop; sampled.3$sampled = (\() sample(1:nrow(the.pop)) <= n) |> with_seed(3)
randomized.1 = sampled.3; randomized.1$W = (\() rbinom(nrow(the.pop), 1, pi(the.pop$X))) |> with_seed(1)
randomized.2 = sampled.3; randomized.2$W = (\() rbinom(nrow(the.pop), 1, pi(the.pop$X))) |> with_seed(2)
randomized.3 = sampled.3; randomized.3$W = (\() rbinom(nrow(the.pop), 1, pi(the.pop$X))) |> with_seed(3)
scale_x = list(scale_x_continuous(breaks=c(55,75)), coord_cartesian(xlim=c(50,80)))
```

:::: {.columns}
::: {.column width="50%"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5
#| 
potential.outcome.plot(the.pop) + scale_x
```

:::
::: {.column width="50%"}

- To make things simple, let's look at a subset of our population.
  - We have grouped everyone by age to give us two age groups---a binary covariate.
- As before, everyone has two potential outcomes--- treated and untreated. 

:::
::::

## Randomization and Sampling


:::: {.columns}
::: {.column width="50%"} 

:::: r-stack
::: {.fragment .fade-out fragment-index="0"}
```{r}
#| fig-width: 7
#| fig-asp: 1.5 

potential.outcome.plot(the.pop) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="0"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 
sampled.potential.outcome.plot(sampled.1) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="1"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 

sampled.potential.outcome.plot(sampled.2) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="2"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 

sampled.potential.outcome.plot(sampled.3) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="3"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5

randomized.plot(randomized.1) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="4"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 

randomized.plot(randomized.2) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="5"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 
randomized.plot(randomized.3) + scale_x
```

:::  
:::: 
::: 

::: {.column width="50%"}

::: {.fragment .semi-fade-out fragment-index="0"}
1. We start with two potential outcomes for each person in our population.
   - Those are the connected dots we see in the plot.
:::
::: {.fragment .fade-in fragment-index="0"}
::: {.fragment .semi-fade-out fragment-index="3"}
2. Then we sample from our population. 
    - This is random, too.
    
:::
:::
::: {.fragment .fade-in fragment-index="3"}
3.  Each one flips a **weighted coin** to determine whether they're treated. A coin that depends on their age. 
   - This is random. Different things happen each time. 
   - But the overall pattern is consistent. 
     - Most of the green dots are on the right. 
       - It's mostly 75-year-olds getting called.
       - The [heads]{.groupb} probability of their coin is 
       `r round(pi(75),2)`.
     - Most of the red dots are on the left. 
       - It's mostly 55-year-olds getting emailed.
       - The [heads]{.groupb} probability of their coin is 
       `r round(pi(55),2)`.
:::

:::
:::: 



## What Happens to Within-Group Means

:::: {.columns}
::: {.column width="50%"}

:::: r-stack
::: {.fragment .fade-out fragment-index="0"}
```{r}
#| fig-width: 7
#| fig-asp: 1.5

po.and.means(the.pop) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="0"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 

sampled.and.means(sampled.1) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="1"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 

sampled.and.means(sampled.2) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="2"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 

sampled.and.means(sampled.3) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="3"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 
randomized.and.means(randomized.1) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="4"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 
randomized.and.means(randomized.2) + scale_x
```

:::
::: {.fragment .current-visible fragment-index="5"}
```{r}
#| fig-width: 7
#| fig-asp: 1.5 
randomized.and.means(randomized.3) + scale_x
```

:::
::::

:::
::: {.column width="50%"}

::: {.fragment .semi-fade-out fragment-index="0"}
1. We start with two potential outcomes for each person in our population.
   - Those are the connected dots we see in the plot.
   - At each level of $X$, these potential outcomes have [a mean]{.counterfactual}.
::::
::: {.fragment .fade-in fragment-index="0"}
::: {.fragment .semi-fade-out fragment-index="3"}
2. Then we sample from our population. 
    - This is random, too.
   - These have [a mean]{.population}, too. And it is, of course, random. It changes if our sample changes.
:::
:::
::: {.fragment .fade-in fragment-index="3"}
3. Each person in the sample flips a **weighted** coin to determine whether they're treated.
   - &olcross; marks the potential outcomes that don't happen.
   - We can look at [the means]{.sample} of people who actually flip '[heads]{.groupb}' and '[tails]{.groupa}', too. More randomness. 
    - Our sample has [a mean]{.sample} within in group, too. A random one.
:::

::: {.fragment .fade-in fragment-index="4"}
How different are these means? 
:::
::: {.fragment .fade-in fragment-index="5"}
Not very different. Why?  
 
  - Because when we look at a single group, nothing has changed. 
  - What happens in the age=75 group when we flip an 80/20 coin there and a 20/80 coin in the age=55 group ...
  - ... will be the same thing that happens when we flip an 80/20 coin everywhere.
:::

:::
::::

## What Happens to (Not Within-Group) Means

:::: {.columns}
::: {.column width="50%"}

:::: r-stack
::: {.fragment .fade-out fragment-index="0"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5

po.and.means(the.pop) + 
  aggregate.po.means(the.pop) + 
  scale_x
```

:::
::: {.fragment .current-visible fragment-index="0"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 

sampled.and.means(sampled.1) + 
  aggregate.po.means(sampled.1) + 
  sampled.aggregate.po.means(sampled.1) + 
  scale_x
```

:::
::: {.fragment .current-visible fragment-index="1"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 

sampled.and.means(sampled.2) + 
  aggregate.po.means(sampled.2) + 
  sampled.aggregate.po.means(sampled.2) +
  scale_x
```

:::
::: {.fragment .current-visible fragment-index="2"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 

sampled.and.means(sampled.3) + 
  aggregate.po.means(sampled.3) + 
  sampled.aggregate.po.means(sampled.3) +
  scale_x
```

:::
::: {.fragment .current-visible fragment-index="3"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 

randomized.and.means(randomized.1) + 
  aggregate.po.means(randomized.1) + 
  sampled.aggregate.po.means(randomized.1) +
  randomized.aggregate.means(randomized.1) +
  scale_x
```

:::
::: {.fragment .current-visible fragment-index="4"}

```{r}
#| fig-width: 7
#| fig-asp: 1.5 
randomized.and.means(randomized.2) + 
  aggregate.po.means(randomized.2) + 
  sampled.aggregate.po.means(randomized.2) +
  randomized.aggregate.means(randomized.2) +
  scale_x
```

:::
::: {.fragment .current-visible fragment-index="5"}
```{r}
#| fig-width: 7
#| fig-asp: 1.5 

randomized.and.means(randomized.3) + 
  aggregate.po.means(randomized.3) + 
  sampled.aggregate.po.means(randomized.3) +
  randomized.aggregate.means(randomized.3) +
  scale_x
```

:::
::::

:::
::: {.column width="50%"}

::: {.fragment .semi-fade-out fragment-index="0"}
1. We start with two potential outcomes for each person in our population.
   - Those are the connected dots we see in the plot.
   - These have [a mean]{.counterfactual} outright---ignoring $X$. Often, that's what we're interested in.
   
::::

::: {.fragment .fade-in fragment-index="0"}
::: {.fragment .semi-fade-out fragment-index="3"}
2. Then we sample from our population. 
    - This is random, too.
   - These have [a mean]{.population}, too. And it is, of course, random. It changes if our sample changes.
   
:::
:::
::: {.fragment .fade-in fragment-index="3"}
3. Each person in the sample flips a **weighted** coin to determine whether they're treated.
   - &olcross; marks the potential outcomes that don't happen.
   - We can look at [the means]{.sample} of people who actually flip '[heads]{.groupb}' and '[tails]{.groupa}', too. More randomness. 
    
:::

::: {.fragment .fade-in fragment-index="4"}
How different are these means? 
:::
::: {.fragment .fade-in fragment-index="5"}
The counterfactual and post-randomization means are very different. Why?
:::

:::
::::

## Why? Covariate Shift.

```{r}
data = randomized.3
```

:::: r-stack
::: {.fragment .fade-out data-fragment-index="1"}
```{r}
#| layout-ncol: 2
#| fig-width: 7
ggplot(data[data$W==0,]) + 
  geom_point(aes(x=X, y=Y), alpha=.3, size=.5, position=position_jitter(w=.1, h=0), color=groupa) + 
  stat_summary(aes(x=X, y=Y), geom='pointrange', fun.data=mean_sd, color=groupa)
ggplot(data[data$W==0,]) + geom_bar(aes(x=X, y=after_stat(prop)), alpha=.3, linewidth=.1, fill=groupa, color=groupa)
```
:::
::: {.fragment .current-visible data-fragment-index="1"}
```{r}
#| layout-ncol: 2
#| fig-width: 7
ggplot(data[data$W==1,]) + 
  geom_point(aes(x=X, y=Y), 
  alpha=.3, size=.5, position=position_jitter(w=.1, h=0), color=groupb) + 
  stat_summary(aes(x=X, y=Y), geom='pointrange', fun.data=mean_sd, color=groupb)
ggplot(data[data$W==1,]) + geom_bar(aes(x=X, y=after_stat(prop)), alpha=.3, linewidth=.1, fill=groupb, color=groupb)
```
:::
::: {.fragment .current-visible data-fragment-index="2"}
```{r}
#| layout-ncol: 2
#| fig-width: 7

ggplot(data) + geom_point(aes(x=plotX2(W,X), y=Y, color=factor(W)), alpha=.3, size=.5, position=position_jitter(w=.1, h=0)) + 
  stat_summary(aes(x=plotX2(W,X), y=Y, color=factor(W)), geom='pointrange', fun.data=mean_sd) + 
  guides(color='none')
ggplot(data) + geom_bar(aes(x=X, y=after_stat(prop), color=factor(W), fill=factor(W)), 
  alpha=.3, linewidth=.1, position='identity') +
  guides(color='none', fill='none')
```
:::
::::

- When we switch from [emails]{.groupa} to [calls]{.groupb}, the distribution of ages shifts to the right.
- And the trend is that donations increase with age. 
- What does this mean for the raw difference in means?

$$
\color{gray}
\begin{aligned}
\hat\Delta_{\text{raw}} 
&= \ingroupb{\frac{1}{N_1} \sum_{i:W_i=1} Y_i} - \ingroupa{\frac{1}{N_0} \sum_{i:W_i=0} Y_i} \\
&= \sum_x \ingroupb{P_{x|1}} \ \ingroupb{\hat\mu(1,x)} - \sum_x \ingroupa{P_{x | 0}} \ {\ingroupa{\hat\mu(0,x)}} \\
&= \underset{\text{adjusted difference} \ \hat\Delta_1}{\sum_x \ingroupb{P_{x|1}} \{\ingroupb{\hat\mu(1,x)} - \ingroupa{\hat\mu(0,x)}\}} 
+ \qty{\underset{\text{covariate shift term}}{\sum_x \ingroupb{P_{x|1}} \ \ingroupa{\hat\mu(0,x)} - \sum_x \ingroupa{P_{x|0}} \  \ingroupa{\hat\mu(0,x)}}}
\end{aligned}
$$

## Covariate Shift in the Whole Dataset
```{r}
set.seed(1)
n = 1000
data = pop[sample(nrow(pop), n, replace=FALSE),]
```

:::: r-stack
::: {.fragment .fade-out data-fragment-index="1"}
```{r}
#| layout-ncol: 2
#| fig-width: 7
ggplot(data[data$W==0,]) + 
  geom_point(aes(x=X, y=Y), alpha=.3, size=.5, position=position_jitter(w=.1, h=0), color=groupa) + 
  stat_summary(aes(x=X, y=Y), geom='pointrange', fun.data=mean_sd, color=groupa)
ggplot(data[data$W==0,]) + geom_bar(aes(x=X, y=after_stat(prop)), alpha=.3, linewidth=.1, fill=groupa, color=groupa)
```
:::
::: {.fragment .current-visible data-fragment-index="1"}
```{r}
#| layout-ncol: 2
#| fig-width: 7
ggplot(data[data$W==1,]) + 
  geom_point(aes(x=X, y=Y), 
  alpha=.3, size=.5, position=position_jitter(w=.1, h=0), color=groupb) + 
  stat_summary(aes(x=X, y=Y), geom='pointrange', fun.data=mean_sd, color=groupb)
ggplot(data[data$W==1,]) + geom_bar(aes(x=X, y=after_stat(prop)), alpha=.3, linewidth=.1, fill=groupb, color=groupb)
```
:::
::: {.fragment .current-visible data-fragment-index="2"}
```{r}
#| layout-ncol: 2
#| fig-width: 7

ggplot(data) + geom_point(aes(x=plotX2(W,X), y=Y, color=factor(W)), alpha=.3, size=.5, position=position_jitter(w=.1, h=0)) + 
  stat_summary(aes(x=plotX2(W,X), y=Y, color=factor(W)), geom='pointrange', fun.data=mean_sd) + 
  guides(color='none')
ggplot(data) + geom_bar(aes(x=X, y=after_stat(prop), color=factor(W), fill=factor(W)), 
  alpha=.3, linewidth=.1, position='identity') +
  guides(color='none', fill='none')
```
:::
::::

$$
\color{gray}
\begin{aligned}
\hat\Delta_{\text{raw}} 
&= \ingroupb{\frac{1}{N_1} \sum_{i:W_i=1} Y_i} - \ingroupa{\frac{1}{N_0} \sum_{i:W_i=0} Y_i} \\
&= \sum_x \ingroupb{P_{x|1}} \ \ingroupb{\hat\mu(1,x)} - \sum_x \ingroupa{P_{x | 0}} \ {\ingroupa{\hat\mu(0,x)}} \\
&= \underset{\text{adjusted difference} \ \hat\Delta_1}{\sum_x \ingroupb{P_{x|1}} \{\ingroupb{\hat\mu(1,x)} - \ingroupa{\hat\mu(0,x)}\}} 
+ \qty{\underset{\text{covariate shift term}}{\sum_x \ingroupb{P_{x|1}} \ \ingroupa{\hat\mu(0,x)} - \sum_x \ingroupa{P_{x|0}} \  \ingroupa{\hat\mu(0,x)}}}
\end{aligned}
$$



## What should we do about this?

```{r}
#| layout-ncol: 2
#| fig-width: 7

ggplot(data) + geom_point(aes(x=plotX2(W,X), y=Y, color=factor(W)), alpha=.3, size=.5, position=position_jitter(w=.1, h=0)) + 
  stat_summary(aes(x=plotX2(W,X), y=Y, color=factor(W)), geom='pointrange', fun.data=mean_sd) + 
  guides(color='none')
ggplot(data) + geom_bar(aes(x=X, y=after_stat(prop), color=factor(W), fill=factor(W)), 
  alpha=.3, linewidth=.1, position='identity') +
  guides(color='none', fill='none')
```

- We know how to make comparisons that aren't influenced by covariate shift. Adjusted comparisons.

$$
\color{gray}
\begin{aligned}

\hat\Delta_1 
&= \frac{1}{N_1}\sum_{i:W_i=1} \qty{ \hat\mu(1, X_i) - \hat\mu(0,X_i) } 
= \sum_{x} P_{x \mid 1} \qty{ \hat\mu(1,x) - \hat\mu(0,x)} \\

\hat\Delta_0 
&= \frac{1}{N_0}\sum_{i:W_i=0} \qty{ \hat\mu(1, X_i) - \hat\mu(0,X_i) } 
= \sum_{x} P_{x \mid 0} \qty{ \hat\mu(1,x) - \hat\mu(0,x)} \\

\hat\Delta_{\text{all}}
&= \frac{1}{n}\sum_{i=1}^n \qty{ \hat\mu(1,X_i) - \hat\mu(0,X_i) } 
= \sum_{x} P_{x} \qty{ \hat\mu(1,x) - \hat\mu(0,x)}
\end{aligned}
$$

- What do these tell us about our treatment effects $\tau_j=y_j(1)-y_j(0)$? Let's find out.

## Identification in Conditionally Randomized Experiments

If treatment assignments are conditionally independent of the potential outcomes given covariates  
$$
\text{ i.e. if } \ W_i \qqtext{ is independent of } \{Y_i(0), Y_i(1)\} \qqtext{ conditional on } X_i
$$

then we can identify potential outcome means *within groups with the same covariate value*. \
It's a conditional version of the same formula.

$$
\begin{aligned}
\mu(w,x) &= \E[Y_i \mid W_i=w, X_i=x] = \E[Y_i(w) \mid X_i=x] = \frac{1}{m_x} \sum_{j:x_j=x} y_j(w) \\
\qfor &m_x = \frac{1}{n} \sum_{i=1}^n 1_{=x}(X_i) 
\end{aligned}
$$

::: {.panel-tabset}
### Conditional Independence and Irrelevance 

- Let's think about conditioning using multi-stage sampling.
  - Stage 1. We sample $X_i$
  - Stage 2. 
    - We sample $\{Y_i(0), Y_i(1)\}$ from the subpopulation with that level of $X_i$
    - We choose $W_i$ by flipping a coin with probability $\pi(X_i)$ of heads. 
    - What we observe is $W_i$, $X_i$, and $Y_i=Y_i(W_i)$.
 - Conditional independence is just independence in the probability distribution describing Stage 2. 
 - A consequence we'll use here is the *irrelevance of conditionally independent conditioning variables*.
 
$$
\E[Y \mid W, X] = \E[Y \mid X] \qqtext{ if } W \qqtext{ is independent of } \{Y(0), Y(1)\} \qqtext{ conditional on } X
$$

### Formula Derivation 

$$
\begin{aligned}
\E[Y_i \mid W_i=w, X_i=x] 
&\explain{=}{$Y_i=Y_i(W_i)$} 
\E[Y_i(W_i) \mid W_i=w, X_i=x] \\
&\explain{=}{$Y_i(W_i)=Y_i(w)$ when $W_i=w$} 
\E[Y_i(w) \mid W_i=w, X_i=x] \\
&\explain{=}{irrelevance of conditionally independent conditioning variables} 
\E[Y_i(w) \mid X_i=x] \\
&\explain{=}{$Y_i(w)$ is sampled uniformly at random from the potential outcomes $y_j(w)$ of the $m_x$ units with $X_i=x$} 
\frac{1}{m_x} \sum_{j:x_j=x} y_j(w) \qfor m_x = \frac{1}{n} \sum_{i=1}^n 1_{=x}(X_i) 
\end{aligned}
$$

### Consequence for Treatment Effects within Groups 

$$
\begin{aligned}
\mu(1,x) - \mu(0,x) 
&= \E[Y_i(1) \mid X_i=x] - \E[Y_i(0) \mid X_i=x] \\
&= \frac{1}{m_x} \sum_{j:x_j=x} y_j(1) - \sum_{j:x_j=x} y_j(0) \\
&= \frac{1}{m_x} \sum_{j:x_j=x} \qty{y_j(1) - y_j(0)} \\
&= \frac{1}{m_x} \sum_{i:x_i=x} \tau_j 
\end{aligned}
$$

- We call this the *Conditional Average Treatment Effect* (CATE). 
- We write it as $\tau(x)$ in mathematical notation.
- When we have conditional randomization, our adjusted comparisons are unbiased estimators of *averages of the CATE $\tau(x)$ over groups*. 
  - $\hat\Delta_1$ averages over the green dots, i.e., the treated individuals.
  - $\hat\Delta_0$ averages over the red dots, i.e., the untreated individuals.
  - $\hat\Delta_{\text{all}}$ averages over all the individuals.
- Let's prove that for $\hat\Delta_{\text{all}}$. We'll see the others for homework.
:::

## Unbiasedness of $\hat\Delta_{\text{all}}$

$$
\begin{aligned}
\E[\hat\Delta_\text{all}] 
&=\E\qty[\sum_x P_{x} \ \qty{\hat \mu(1,x) - \hat\mu(0,x) }] \\
&\explain{=}{law of iterated expectations}
\E\qty[\E\qty[ \sum_x P_{x} \ \qty{\hat \mu(1,x) - \hat\mu(0,x)}  \mid (W_1, X_1) \ldots (W_n, X_n)]] \\
&\explain{=}{linearity of conditional expectation} 
\E\qty[\sum_x P_{x} \ \qty{\E\qty[\hat\mu(1,x) \mid (W_1,X_1) \ldots (W_n,X_n)] -  \E\qty[\hat\mu(0,x) \mid (W_1,X_1) \ldots (W_n,X_n)]}] \\
&\explain{=}{unbiasedness of the sample mean} 
\E\qty[\sum_x P_{x} \ \qty{\mu(1,x)  -  \mu(0,x)}] \\
&\explain{=}{identification: $\tau(x)=\mu(1,x)-\mu(0,x)$} 
\E\qty[\sum_x P_{x} \ \tau(x)] \\
&\explain{=}{linearity of expectation} \sum_x \E[P_{x}] \ \tau(x) \\
&\explain{=}{unbiasedness of sample proportions and def of $\tau(x)$} 
\sum_x p_{x} \ \frac{1}{m_x}\sum_{j:x_j=x} \tau_j \qfor p_{x} = \E[P_{x}]=\frac{m_x}{m} \\
&\explain{=}{rewriting our sum of column sums as a single sum} 
\sum_x \frac{m_x}{m} \ \frac{1}{m_x}\sum_{j:x_j=x} \tau_j = \frac{1}{m}\sum_{j=1}^m \tau_j 
\end{aligned}
$$

- This is the average of the individual treatment effects $\tau_j$ 
over the whole population. 
- Or, for short, the *average treatment effect* or *ATE*.

# Appendix 

## Unbiasedness with Coin-Flip Randomization {#sec-unbiasedness-coin-flips}

- Proving unbiasedness little subtler when we randomize via coin flips vs. shuffling, as we have to ...
  - *condition on the number of times we flip heads*, $N_1$, to distribute the expectation
  - recognize that the *conditional probability* of any one flip being heads is the frequency of heads, $N_1/m$.

$$
\begin{aligned}
\E[\hat\mu(w)] 
&\explain{=}{This is our subsample mean.} 
\E\qty[\frac{1}{N_w}\sum_{j=1}^m y_j(W_j) \ 1_{=w}(W_j)] \\
&\explain{=}{via the law of iterated expectations} 
\E\qty[\E\qty[\frac{1}{N_w}\sum_{j=1}^m y_j(W_j) \ 1_{=w}(W_j) \mid N_w]] \\
&\explain{=}{the linearity of conditional expectation}  
\E\qty[\frac{1}{N_w}\sum_{j=1}^m y_j(W_j) \ \E[1_{=w}(W_j) \mid N_w]] \\ 
&\explain{=}{substituting the conditional probability} 
\E\qty[\frac{1}{N_w}\sum_{j=1}^m y_j(w) \ \frac{N_w}{m}] \\ 
&\explain{=}{multiplying} 
\E\qty[\frac{1}{m} \sum_{j=1}^m y_j(w)] \\
&\explain{=}{the expectation of a constant is that constant} 
\frac{1}{m} \sum_{j=1}^m y_j(w)
\end{aligned}
$$

- Why is $N_1/m$ the conditional probability of flipping heads?
  - The flips are identically distributed, so $\E[1_{=1}(W_j) \mid N_1]$ must be the same for each $j$.
  - And they sum to $N_1$. This lets us write an equation we can solve for the conditional expectation.

$$
\begin{aligned}
N_w
&\explain{=}{linearity of conditional expectation} 
\E[N_w \mid N_w] \\
&\explain{=}{plugging in the definition of $N_w$} 
\E\qty[\sum_{j=1}^m 1_{=w}(W_j) \mid N_w] \\
&\explain{=}{linearity of conditional expectation} 
\sum_{j=1}^m \E[1_{=w}(W_j) \mid N_w] \\ 
&\explain{=}{identical distribution} m \times \E[1_{=w}(W_j) \mid N_w].
\end{aligned}
$$






