---
title: "Inference for Complicated Summaries"
---

# Introduction 

## 
{{< include shared-code.qmd >}}
{{< include ca-income.qmd >}}

```{r scatter-plots}
pop.scatter = ggplot(pop.tab) + 
  geom_point(aes(x=education, y=income), color='#F7766D',size=dot.size/10, 
             alpha=.1, position=position_jitter(seed=0,w=.3)) +
  scale_x() + scale_y() + guides(color=FALSE)

sample.scatter = ggplot(sample.tab) + 
  geom_point(aes(x=education, y=income), color='#F7766D',size=dot.size, 
             alpha=.2, position=position_jitter(seed=0,w=.3)) +
  scale_x() + scale_y() + guides(color=FALSE)
```


```{r}
#| layout-ncol: 2
#| fig-width: 7
#| fig-asp: .4
pop.scatter
sample.scatter
```


:::: columns
::: column
### Population
$$
\small{
\begin{array}{ccccccc}
x_1 & y_1 & x_2 & y_2 &  \ldots & x_m & y_m \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\end{array}
}
$$
:::
::: column
### Sample
$$
\small{
\begin{array}{ccccccc}
X_1 & Y_1 & X_2 & Y_2 &  \ldots & X_n & Y_n \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\end{array}
}
$$
:::
::::

- Last class, we talked about summaries of the relationship between income and education in a population. 
- Today, we'll talk about estimating them by summarizing a sample drawn from it exactly the same way.
- This is a recipe we've been using since the start of the semester.
  - We've used sample means estimate population means.
$$
\frac1n\sum_{i=1}^n Y_i \qqtext{ estimates } \frac1m\sum_{j=1}^m y_j 
$$
  - We've used subsample means to estimate subpopulation means.
$$
\frac{1}{N_x}\sum_{i:X_i=x}Y_i \qqtext{ estimates } \frac{1}{m_x}\sum_{j:x_j=x}y_j \qfor N_x=\sum_{i:X_i=x} 1 \qand m_x = \sum_{j:x_j=x}1
$$

- What'll be new is that the summaries we'll be estimating will be more complex. 
  - For many of them, we already have the tools to understand estimation and inference.
  - For others, we'll need new tools that we'll cover today. 
  - Let's review a few examples.
  
## 
::: {.panel-tabset}
### Column Comparisons 
```{r}
summaries = sample.tab |> group_by(education) |>
  summarize(mean=mean(income), sd=sd(income), N=length(income))
coarsened.summaries = sample.tab[sample.tab$education %in% c(16,8:12),] |>
  group_by(education <= 12) |>
  summarize(mean=mean(income),
            sd=sd(income),
            mean.X = mean(education))

ggplot(sample.tab) + geom_point(aes(x=education, y=income, alpha=education %in% c(16,12), color=education > 12), 
                                  size=dot.size, position=position_jitter(seed=0,w=.3)) + 
                   geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries[summaries$education %in% c(16,12), ]) +
                   scale_x() + scale_y() + scale_alpha_manual(breaks=c(FALSE, TRUE), values=c(.01,.3)) + guides(color=FALSE, alpha=FALSE) 
```

$$
\hat\theta = \hat\mu(16) - \hat\mu(12) 
$$

- Everything from our [Lecture on Comparing Two Groups](https://qtm285-1.github.io/assets/lectures/Lecture6.html) applies. 
- We can 'forget' we have the columns we don't want.

### Comparison of Aggregates
```{r}
ggplot(sample.tab) + geom_point(aes(x=education, y=income, alpha=education %in% c(16,8:12), color=education > 12), 
                                  size=dot.size, position=position_jitter(seed=0,w=.3)) + 
                   geom_pointrange(aes(x=mean.X, y=mean, ymin=mean-sd, ymax=mean+sd), data=coarsened.summaries) +
                   scale_x() + scale_y() + scale_alpha_manual(breaks=c(FALSE, TRUE), values=c(.01,.3)) + guides(color=FALSE, alpha=FALSE) 
```
$$
\begin{aligned}
\hat\theta &= \hat\mu(16) - \hat\mu(\le 12) && \qfor \hat\mu(\le 12) = \frac{\sum_{i:X_i \le 12} \hat\mu(X_i)}{\sum_{i:X_i \le 12} 1}
\end{aligned}
$$

- Same thing. We can 'forget' that we have the *resolution* we don't want. 
- If our data had come to us dichotomized already, the tools we have would apply.
- The math doesn't care *who* dichotomized the data, so we can use the same tools.

###  Aggregated Comparisons
```{r}
earnings.jitter.hs = ggplot(sample.tab) + geom_point(aes(x=education, y=income, alpha=education <= 12), color='#F7766D',
                              size=dot.size, position=position_jitter(seed=0,w=.3)) +
                   scale_x() + scale_y() + scale_alpha_manual(breaks=c(FALSE, TRUE), values=c(.01,.3)) + guides(color=FALSE, alpha=FALSE)

                 
aggregate.comparison.plot = earnings.jitter.hs +
  geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries[summaries$education <= 12,]) +
  geom_line(aes(x=education, y=mean), data=summaries[summaries$education<=12,], color='black', linetype='dashed') + 
  geom_line(aes(x=education, y=mean), data=summaries[summaries$education %in% c(8,12),], color='blue', alpha=.5)

aggregate.comparison.plot 
```

$$
\begin{aligned}
\hat\theta &= \frac{\sum\limits_{x=9}^{12} \qty{ \hat\mu(x) - \hat\mu(x-1) }}{4}  && \qqtext{ the average over years} \\
\hat\theta &= \frac{\sum\limits_{i:X_i \in 9 \ldots 12} \qty{ \hat\mu(X_i) - \hat\mu(X_i-1) }}{\sum_\limits{i:X_i \in 9 \ldots 12} 1} && \qqtext{ the average over people}
\end{aligned}
$$

- These are different. There's no way to 'forget' into a situation we've already addressed.
- We'll need some new tools to deal with these. That's what we'll work on today.

:::

## Quiz {#quiz}

```{r}
#| fig-asp: .4
aggregate.comparison.plot + scale_x(c(7.5,12.5)) + scale_y(c(0, .7e5))
```
$$
\begin{aligned}
\hat\theta_A &= \frac{\sum\limits_{x=9}^{12} \qty{ \hat\mu(x) - \hat\mu(x-1) }}{4}  && \qqtext{ the average over years} \\
\hat\theta_B &= \frac{\sum\limits_{i:X_i \in 9 \ldots 12} \qty{ \hat\mu(X_i) - \hat\mu(X_i-1) }}{\sum_\limits{i:X_i \in 9 \ldots 12} 1} && \qqtext{ the average over people}
\end{aligned}
$$


1. One of these two estimators can be 'unrolled' into a simple comparison of two column means. 
$$
\inteal{\hat \theta = \frac{\hat \mu(12) - \hat \mu(8)}{4}} 
$$
Which is it?


2. What is the variance of this estimator $\inteal{\hat\theta}$? 
  - You may express your answer in terms of anything that appears in previous lectures. 
  - e.g., $\Var[\hat \mu(7)]$ can appear in your answer if you want it to. 


## Quiz Solution {#quiz-answers}

```{r}
aggregate.comparison.plot + scale_x(c(7.5,12.5)) + scale_y(c(0, .7e5))
```

::: {.panel-tabset}
### Part 1
It's the average over years that unrolls like this.
$$
\begin{aligned}
\frac{\sum_{x=9}^{12} \qty{ \hat\mu(x) - \hat\mu(x-1) }}{4} 
&= \frac{ \qty{\mu(12) - \mu(11)} + \qty{ \mu(11) - \mu(10) } + \qty{ \mu(10) - \mu(9) } + \qty{ \mu(9) - \mu(8) } }{4} \\
&= \frac{\mu(12) - \mu(8)}{4}
\end{aligned}
$$

### Part 2
It's $1/4^2 = 1/16$ times the variance of the difference of these two column means.

 - For any random variable $X$ and constant $a$ ...
$$
 \Var[aX] 
 \explain{=}{definition} \E[(aX-\E[aX])^2] 
 \explain{=}{linearity of expectation + arithmetic} a^2\E[(X-\E[X])^2] 
 \explain{=}{definition} a^2\Var[X].
$$

- In the special case $a=1/4$ and $X=\hat\mu(12) - \hat\mu(8)$ ...
$$
\begin{aligned}
\Var\qty[\frac{\hat\mu(12) - \hat\mu(8)}{4}] 
&= \frac{1}{4^2}\Var[\hat\mu(12) - \hat\mu(8)]  
&& \text{ is the answer I was looking for } \\
&= \frac{1}{16} \times \E\qty[ \frac{\sigma^2(12)}{N_12} + \frac{\sigma^2(8)}{N_8} ]         
&& \text{ using a formula from our lecture on comparing two groups.} 
\end{aligned}
$$

### More Along Those Lines
We know this estimator is unbiased. We can think about the difference in means for this, too. 
$$
\begin{aligned}
\E\qty[\frac{\hat\mu(12) - \hat\mu(8)}{4}] 
&\explain{=}{via linearity} 
\frac{\E[\hat\mu(12)] - \E[\hat\mu(8)]]}{4} 
\\ 
&\explain{=}{via unbiasedness of subsample means} 
\frac{\mu(12) - \mu(8)}{4}\\
&\explain{=}{'rolling it back up'} 
\frac{1}{4}\sum_{x=9}^{12}\qty{\mu(x) - \mu(x-1)}{4}.
\end{aligned}
$$ 

:::

## What This Tells Us About Inference 
```{r}
#| layout-ncol: 2
#| fig-width: 7
#| fig-cap:
#|  - Our estimator's actual sampling distribution 
#|  - Our estimate of that sampling distribution

Y = earnings$income
X = earnings$education
point.estimate = (mean(Y[X==12]) - mean(Y[X==8]))/4 
variance.estimate = (1/16) * (sd(Y[X==12])^2/sum(X==12) + sd(Y[X==8])^2/sum(X==8))
interval.estimate = point.estimate + c(-1,1)*1.96*sqrt(variance.estimate)

xx = seq(point.estimate-4*sqrt(variance.estimate), point.estimate+4*sqrt(variance.estimate), length.out=1000)
sampling.dist.estimate = dnorm(xx, mean=point.estimate, sd=sqrt(variance.estimate)) 
height = max(sampling.dist.estimate)
sampling.dist.draws = data.frame(x=rnorm(100,point.estimate,sqrt(variance.estimate)), 
                                 y=height*(1:100)/100, 
                                 w=1.96*sqrt(variance.estimate))

ggplot() + 
  annotate('rect', xmin=min(interval.estimate), xmax=max(interval.estimate), ymin=-Inf, ymax=Inf, fill='blue', alpha=.1) +
  geom_area(aes(x=xx, y=sampling.dist.estimate), color='blue', alpha=.1) +  
  geom_vline(aes(xintercept=point.estimate), color='green', linewidth=2, alpha=.5) + 
  geom_pointrange(aes(x=x, xmax=x+w/2, xmin=x-w/2, y=y), color='purple', alpha=.3, data=sampling.dist.draws) + 
  scale_x_continuous(breaks=c()) + coord_cartesian(xlim=range(xx))

ggplot() + 
  annotate('rect', xmin=min(interval.estimate), xmax=max(interval.estimate), ymin=-Inf, ymax=Inf, fill='blue', alpha=.1) +
  geom_area(aes(x=xx, y=sampling.dist.estimate), color='blue', alpha=.1) + 
  geom_pointrange(aes(x=point.estimate, y=height/3, xmin=min(interval.estimate), xmax=max(interval.estimate)), 
                      color='blue')  +
  scale_x_continuous(breaks=point.estimate + sqrt(variance.estimate)*c(-2,0,2)) + 
  coord_cartesian(xlim=range(xx))  
```

::: {.panel-tabset}
### What We Need for Calibration

To turn a point estimator into a confidence interval, we need to know two things about its sampling distribution.

- Where it's centered---relative to the estimation target.
  - In this case, it's exactly were we want it to be. 
  - Our estimator is unbiased, so it's centered at the target.
- How wide it is, i.e. how far out from center we need to go to cover 95% of draws. [`r circle`]{.purple}s
  - For this, we can use an estimate of the sampling distribution.
  - Using one based on *normal approximation*, that's $\pm 1.96 \ \text{estimated standard deviations}$

### Visualizing Calibration
- To illustrate what's going on, we can draw the picture on the left above. We ...
  - Place our estimated sampling distribution right at the target. It'll have roughly the same shape.
  - Think of our estimate as one of many draws [`r circle`]{.purple} from it and draw arms on each.
  - See that, for 95% of these draws, the arms touch the [target]{.target} at the distribution's center.
- We don't know where this center $\theta$ is. Otherwise we wouldn't bother estimating it.
  - That's why I haven't labeled the ticks on the $x$-axis.
  - But we don't need to know that to know our intervals are calibrated. 
  - Calibration is about where our draws are *relative to* our estimation target. Unbiasedness is enough.

### Anticipating Calibration

- We can *understand and anticipate* the width of these intervals using our formula for the estimator's variance.
- All we need are estimates of a few things in the formula: subpopulation standard deviations and subsample sizes.
- We can use this knowledge to design studies that give us a desired level of precision.
  - We can use a *pilot study* to estimate these things.
  - And use those estimates to choose an appropriate sample size for a second wave.
  - This is what we imagined doing when we talked about [the NSW Demonstration](https://qtm285-1.github.io/assets/lectures/Lecture6.html#the-national-supported-work-demonstration) a few weeks back.
  
:::: columns 
::: column 
$$
\begin{aligned}
\Var[\hat\theta] 
&= \frac{1}{16} \times \E\qty[ \frac{\sigma^2(12)}{N_{12}} + \frac{\sigma^2(8)} {N_8} ] \\
&\approx \frac{1}{16} \times \qty[\frac{\hat\sigma^2(12)}{N_{12}} + \frac{\hat\sigma^2(8)}{N_8} ] 
\end{aligned}
$$
:::
::: column
$x$ | $N_x$ |  $\hat \mu(x)$ | $\hat \sigma(x)$
--- | --- | --- | ---
8  | `r sum(X==8)` | `r thousand(mean(Y[X==8]), .1)` | `r thousand(sd(Y[X==8]), .1)`
12  | `r sum(X==12)` | `r thousand(mean(Y[X==12]), .1)` | `r thousand(sd(Y[X==12]), .1)`
:::
::::

:::

## Calibration for the Average over People
```{r}
#| cache: true

data = earnings

average.over.people = function(summaries) {
  muhat = function(x) { summaries$mean[summaries$education==x] }
  P = function(x) { summaries$N[summaries$education==x] / sum(summaries$N) }
  alphahat = function(x) { if (x==12) { P(12) } else if (x==8) { -P(9) } else { P(x) - P(x+1) } }  
  sapply(8:12, function(x) { alphahat(x) * muhat(x) }) %>% sum
}
average.over.years = function(summaries) {
  (summaries$mean[summaries$education==12] - summaries$mean[summaries$education==8])/4
}
run.boot = function(data, B=1000) { 
  purrr::map(1:B, function(i) {
  # use tryCatch to handle errors that happen when we get no  
  # units at a given education level in our bootstrap sample
  tryCatch({
  boot.sample = data[sample(1:nrow(data), replace=TRUE),]
  boot.summaries = boot.sample[boot.sample$education <= 12, ] |> 
    group_by(education) |> 
    summarize(mean=mean(income), N=length(income))
  data.frame(people=average.over.people(boot.summaries), 
             years=average.over.years(boot.summaries))
  }, error=function(e) NULL)
 }) |> list_rbind()
}

set.seed(1)
boots = run.boot(data, 1000)
summaries = data[data$education <= 12, ] %>% group_by(education) %>% 
  summarize(mean=mean(income), N=length(income), sd=sd(income))
```

```{r}
#| layout-ncol: 2
#| fig-width: 7
#| fig-subcap:
#|  - Bootstrap Sampling Distribution and Interval Estimate for the Average over Years
#|  - Bootstrap Sampling Distribution and Interval Estimate for the Average over People 

distribution.plot = function(boot.samples, point.estimate, w=width(boot.samples), interval.replications=0, point.replications=0) { 
  drange = c(0, dnorm(0, mean=0, sd=sd(boot.samples)))
  p = ggplot(data.frame(samples=boot.samples)) + 
    annotate('rect', xmin=mean(boot.samples) - w/2, xmax = mean(boot.samples) + w/2, ymin=-Inf, ymax=Inf, fill='blue', alpha=.1) +
    geom_histogram(aes(x=samples, y=after_stat(density)), alpha=.1, bins=50)  +
    geom_pointrange(aes(x=point.estimate, xmin=point.estimate - w/2, xmax=point.estimate + w/2, y=max(drange)/3), color='blue') + 
    geom_vline(xintercept=mean(boot.samples), color='blue')
  if(interval.replications > 0) { 
    K = interval.replications
    p = p + geom_pointrange(aes(y=y, x=x, xmin=xmin, xmax=xmax), 
          data=data.frame(y=seq(min(drange), max(drange), length.out=K), 
                  x=boot.samples[1:K], xmin=boot.samples[1:K] - w/2, xmax=boot.samples[1:K] + w/2), color='purple', alpha=.2)
  }
  if(point.replications > 0) { 
    K = point.replications
    p = p + geom_point(aes(y=y, x=x), 
          data=data.frame(y=seq(min(drange), max(drange), length.out=K), 
                  x=boot.samples[1:K], xmin=boot.samples[1:K] - w/2, xmax=boot.samples[1:K] + w/2), color='purple', alpha=.2)
  }
  p
}

distribution.plot(boots$years, average.over.years(summaries), point.replications=0) + 
  geom_line(aes(x=xx, y=sampling.dist.estimate), color='blue', alpha=.1) + 
  coord_cartesian(xlim=c(-1000, 15000), ylim=c(0, 3e-4))

distribution.plot(boots$people, average.over.people(summaries), point.replications=0) +
  coord_cartesian(xlim=c(-1000, 15000), ylim=c(0, 3e-4))
```

- We don't yet know these things about our other summary: the average over people.
- Our first step will be to show that it, too, is unbiased.
  - That's really enough for calibration. 
  - We can use the bootstrap sampling distribution to get the right width.
- Our second step will be to work out a formula for its variance.
  - This'll allow us to understand our estimator better, design properly sized-studies, etc.
  - And it'll be the key to understanding a phenomenon we can see in the plots above.
- The average over people is *harder to estimate* than the average over years. At least using this data.
  - We can see its bootstrap sampling distribution is at least twice as wide.
  - What's behind this? We'll find out. 

# Unbiasedness

## Warm-Up

```{r}
#| fig-asp: .4
aggregate.comparison.plot + scale_x(c(7.5,12.5)) + scale_y(c(0, .7e5))
```

- When analyzing estimators like this, it's helpful to rewrite them as a [linear combination of column means]{.teal}.
- That's what we were doing when we 'unrolled' our average over years to see that it was the [secant slope]{.blue}.
- And it's helpful, if possible, to work out formulas for its coefficients with as few cases as possible.
  - Why? Because when we do calculations, we have to analyze each case.
  - In this example, 3 cases is enough. It often is.
- Give it a shot. Fill in the cases below.

$$
\begin{aligned}
\frac{1}{4}\sum_{x\in 9 \ldots 12} \qty{ \hat \mu(x) - \hat \mu(x-1)} 
&= \frac{\hat\mu(12) -\hat\mu(8)}{4}
&= \inteal{\sum_{x}\hat\alpha(x)\hat\mu(x)} 
 \qfor \hat\alpha(x) = 
\begin{cases} 
  & \qqtext{if}   \\
  & \qqtext{if}   \\
  & \qqtext{if}  \\
\end{cases}
\end{aligned}
$$



## Step 1. Rewriting the Average over People in Standard Form

$$
\frac{\sum_{i:X_i \in 9 \ldots 12} \qty{ \hat \mu(X_i) - \hat \mu(X_i-1)}}{\sum_{i:X_i \in 9 \ldots 12} 1} = 
\sum_{x}\hat\alpha(x)\hat\mu(x) 
\qfor \hat\alpha(x)=
$$

::: {.panel-tabset}
### Exercise
- We'll do this in three steps. 
  - We'll rewrite it as a linear combination of *differences in subsample means*. 
    - This'll be a lot like what we did for aggregate means last class.
  - We'll 'unroll' that sum and group terms multiplying each subsample mean.
    - We did this for the average over years and got something very simple. 
    - All but two coefficients $\hat\alpha(x)$ were zero.
    - This one will be messier.
  - We'll try to recognize a pattern in the coefficients so we don't have to analyze every term by itself.

### Solution

$$
\begin{aligned}
\frac{\sum_{i:X_i \in 9 \ldots 12} \qty{ \hat\mu(X_i) - \hat\mu(X_i-1) }}{\sum_{i:X_i \in 9 \ldots 12} 1}  
&=  \sum_{x \in 9 \ldots 12} P_x \ \qty{ \hat\mu(x) - \hat\mu(x-1) } \quad \text{ for } \quad P_x = \frac{N_x}{\sum_{x \in 9 \ldots 12}N_x} \\
&= \sum_x \hat \alpha(x) \hat \mu(x) \qfor
\hat\alpha(x) = \begin{cases}
  P_{12} & \text{ if } x = 12 \\
  P_{x} - P_{x+1} & \text{ if } x \in \{9 \ldots 11\} \\
  -P_9 & \text{ if } x = 8 
\end{cases}
\end{aligned}
$$

:::

## Step 2. Estimator's Expectation in Standard Form


::: fragment
$$
\begin{aligned}
\E\qty[\sum_x \hat\alpha(x) \hat \mu(x)] 
&\explain{=}{linearity of expectations}
\sum_x \E\qty[\hat\alpha(x) \hat\mu(x)] \\
&\explain{=}{law of iterated expectations} 
\sum_x \E\qty{ \E\qty[\hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n] } \\
&\explain{=}{linearity of conditional expectations} 
\sum_x \E\qty{ \hat \alpha(x) \E[\hat \mu(x) \mid X_1 \ldots X_n] } \\
&\explain{=}{conditional unbiasedness of subsample means} 
\sum_x \E\qty{ \hat \alpha(x) \mu(x) }                    \\        
&\explain{=}{linearity of expectations} 
\sum_x \E\qty{ \hat \alpha(x) } \mu(x)                   
\end{aligned}
$$

- This tells us we have an unbiased estimator if 
 - the coefficients of our *estimation target* in standard form 
 - are the expected value of our estimator's coefficients.

:::

## Step 3. Dealing with Specifics
$$ 
\begin{aligned}
&\text{Does} \quad \E\qty{ \hat \alpha(x) } = \alpha(x) \qfor 
&&\hat\alpha(x) = \begin{cases}
  P_{12} & \text{ if } x = 12 \\
  P_{x} - P_{x+1} & \text{ if } x \in \{9 \ldots 11\} \\
  -P_9 & \text{ if } x = 8 
\end{cases} \\ 
\qqtext{and}
&&&\alpha(x) = \begin{cases}
  p_{12} & \text{ if } x = 12 \\
  p_{x} - p_{x+1} & \text{ if } x \in \{9 \ldots 11\} \\
  -p_9 & \text{ if } x = 8
\end{cases}
\end{aligned}
$$

- For this estimator, that question boils down to the unbiasedness of *sample proportions*. Almost. 
- These aren't proportions of our whole sample. They're proportions of the 'high school subsample'.
- But we we can show that $\E[P_x]=p_x$ for all $x$ via a sort of 'two-stage sampling argument'.
  - First flip a coin to decide if $X \in \{9 \ldots 12\}$ or not.
  - Then roll a die to decide which value you take within the group. 
  - $P_x$ is, conditional on the coin flip, a sample proportion. 
  - So, it's unbiased for the population proportion $p_x$.


## Beyond these two summaries 

```{r}
#| out-height: 200px
#| fig-asp: .2 

biasdata = data.frame(x=seq(-3,3,length.out=1000), fx=dnorm(seq(-3,3,length.out=1000)))
ggplot() + geom_area(aes(x=x,y=fx), alpha=.2, color='gray', data=biasdata) + 
           geom_vline(aes(xintercept=0), color='green', linewidth=2, alpha=.5) + 
           geom_point(aes(x=rnorm(100), y=(1:100)/(100*(sqrt(2*pi)))), color='purple', alpha=.3)
```

- We've shown that our two specific estimators are unbiased.
- More generally, we've shown that a linear combination of subsample means is unbiased ...
  - if our estimation target is a linear combination of subpopulation means ...
  - ... in which our estimator's coefficients are replaced by their expected values.

$$
\begin{aligned}
\E\qty[\sum_x \hat\alpha(x) \hat \mu(x) ] &= \sum_x \E[\hat\alpha(x)] \mu(x) \\
\end{aligned}
$$

- This can be used for all kinds of summaries. 
  - Almost every target we'll talk about this semester is a linear combination like this.
  - And the ones that aren't, like ratios, can be *approximated* by linear combinations. 

# Variance {#continued} 

**We'll continue from here.**

## Review: The Law of Total Variance

```{r}
#| layout-ncol: 2
#| fig-width: 7

Y = c(.75,1,1.25, 1,1.25,1.5)
X = 0 + (1:6 > 3)
Xdodge = X + rep(.02*c(-1,0,1),2)

mu = Vectorize(function(x) mean(Y[X==x]))
Y.wider = mu(X) + 3*(Y - mu(X))

ggplot() + geom_point(aes(x=Xdodge, y=Y.wider, color=factor(X)), alpha=.7, size=4) + 
  scale_y_continuous(breaks=seq(0,3,by=.5), limits=c(0,3)) + guides(color='none') 
ggplot() + geom_point(aes(x=Xdodge, y=Y+X, color=factor(X)), alpha=.7, size=4) + 
  scale_y_continuous(breaks=seq(0,3,by=.5), limits=c(0,3)) + guides(color='none') 
```

$$
\Var[Y] = \E\qty{\Var( Y \mid X ) } + \Var\qty{\E( Y \mid X ) }  
$$

- The Law of Total Variance breaks Variance into within-group and between-group terms.
- It's like the Law of Iterated Expectations, but for Variance.
- It's a useful way to decompose the variance of a random variable.
- Think about where most of the variance in $Y$ is coming from in the two populations shown above.


## A Variance Decomposition for our Estimator

$$
\Var\qty[\sum_x \hat\alpha(x) \hat\mu(x)]
= \textcolor{blue}{\sum_x \sigma^2(x) \times  \E\qty[ \frac{\hat\alpha(x)^2}{N_x} ]} + \textcolor{red}{\Var\qty[ \sum_x \mu(x) \hat \alpha(x)  ]} \qfor 
\sigma^2(x) = \Var[Y_i \mid X_i=x] 
$$

- We can use the *law of total variance* to decompose our estimator's variance into two parts.
  - [The first]{.fg style="color:blue"} reflects the randomness of our subsample means.
  - [The second]{.fg style="color:red"} reflects the randomness of our coefficients $\hat\alpha(x)$.
- Note that if our coefficients aren't random, i.e. $\hat\alpha(x)=\E[\hat\alpha(x)]$, the second term is zero.
- And often, even if they are random, the first term is much bigger.
- As a result, we can usually get away with thinking about the first part only.
  - Note that the second part is positive---it's the variance of something.
  - Ignoring it gives us a **lower bound** on the actual variance of our estimator.

## The Formula

We won't get deep into the derivation now. I'll guide you through it in the homework.


$$
\small{
\begin{aligned}
\Var\qty[ \sum_x \alpha(x) \hat \mu(x) ] 
&= \E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] \quad &&+ \quad \Var\qty[ \E\qty{\sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n} ] \\
&= \E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] \quad &&+ \quad \Var\qty[ \sum_x  \hat\alpha(x) \E\qty{ \hat \mu(x) \mid X_1 \ldots X_n} ] \\
&= \E\qty[ \Var\qty{ \sum_x \hat \alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] \quad &&+ \quad \Var\qty[ \sum_x \hat\alpha(x) \mu(x) ] \\
&= \textcolor{blue}{\sum_x \E\qty[ \hat\alpha(x)^2 \Var\qty{\hat\mu(x) \mid X_1 \ldots X_n}]} \quad &&+ \quad 
\textcolor{blue}{\sum_x \sum_{x'} \mu(x) \mu(x') \Cov\qty[\hat\alpha(x),\ \hat\alpha(x')]} \\
&=\textcolor{blue}{ \sum_x \sigma^2(x) \times \E\qty[ \frac{\hat\alpha(x)^2}{N_x} ]} \quad &&+ \quad 
\textcolor{blue}{\sum_x \sum_{x'} \mu(x) \mu(x') \Cov\qty[\hat\alpha(x),\ \hat\alpha(x')]} 
\end{aligned}
}
$$


## Variance Estimation

$$
\Var\qty[\sum_x \alpha(x) \hat \mu(x)] \ge \textcolor{blue}{\sum_x \sigma^2(x) \times \E\qty[ \frac{\hat\alpha(x)^2}{N_x} ] \quad \text{ where } \quad \sigma^2(x) = \Var[Y_i \mid X_i=x]}
$$

- LThe lower bound has two quantities we don't know, but can estimate easily.
  - The subpopulation variances $\sigma^2(x)$.
  - The expected value of the ratio of the coefficient squared and the sample size, $\hat\alpha(x)^2 / N_x$.
- The subsample variances are good esimates of the subpopulation variances
$$
\sigma^2(x) \approx \hat\sigma^2(x) := \frac{1}{N_x} \sum_{i:X_i=x} \qty{Y_i - \hat\mu(x)}^2
$$

- The squared coefficient / subsample size ratio is usually a good estimate of its expectation.

$$ \frac{\hat\alpha(x)^2}{N_x} \approx \E\qty[\frac{\hat\alpha(x)^2}{N_x}] $$

- Plugging them in gives us an estimate of our summary's variance lower bound.

$$
\Var[\hat\theta] \ge \sum_x \ \frac{\hat\sigma^2(x)\hat\alpha(x)^2}{N_x}
$$


## The Average Over Years
```{r}
#| out-height: 300px
#| fig-asp: .3
sample.scatter
```

```{r}
sigmahat = function(x) { sd(earnings$income[earnings$education==x]) }
alpha    = function(x) { case_match(x, 
  12 ~  1/4,
  8  ~ -1/4,
  .default = 0) 
}
N        = function(x) { sum(earnings$education==x) }
```

$$
\hat\sigma_{\hat\theta}^2 = \sum_x  \frac{\hat\alpha^2(x) \ \hat\sigma^2(x)}{N_x}
$$

where

$$
\begin{array}{l|lll} x & 8 & 9 & 10 & 11 & 12 & 13 & \ldots & 20 \\ \hline
\hat\alpha^2(x)    & `r approx(alpha(8)^2)` & `r approx(alpha(9)^2)` & `r approx(alpha(10)^2)` &
`r approx(alpha(11)^2)` & `r approx(alpha(12)^2)` & `r approx(alpha(13)^2)` &
\ldots & `r approx(alpha(20)^2)` \\
\hat\sigma^2(x) &    
`r approx(sigmahat(8)^2,1)` & `r approx(sigmahat(9)^2,1)` & `r approx(sigmahat(10)^2,1)` &
`r approx(sigmahat(11)^2,1)` & `r approx(sigmahat(12)^2,1)` & `r approx(sigmahat(13)^2,1)` &
\ldots & `r approx(sigmahat(20)^2,1)` \\
N_x  & 
`r N(8)` & `r N(9)` & `r N(10)` & 
`r N(11)` & `r N(12)` & `r N(13)` &
 \ldots & `r N(20)` \\
\frac{\alpha^2(x) \ \hat\sigma^2(x)}{N_x} &
`r approx(sigmahat(8)^2 * alpha(8)^2 / N(8))` & `r approx(sigmahat(9)^2 * alpha(9)^2 / N(9))` & `r approx(sigmahat(10)^2 * alpha(10)^2 / N(10))` &
`r approx(sigmahat(11)^2 * alpha(11)^2 / N(11))` & `r approx(sigmahat(12)^2 * alpha(12)^2 / N(12))` & `r approx(sigmahat(13)^2 * alpha(13)^2 / N(13))` &
\ldots & `r approx(sigmahat(20)^2 * alpha(20)^2 / N(20))` \\
\end{array}
$$

**Exercise**. Use a caricature of this table to approximate our estimator's variance.


## Comparing Our Two Summaries
```{r}
#| layout-ncol: 2
#| fig-width: 7
#| fig-asp: .3
#| fig-subcap:
#|  - Interval Estimate for the Average over Years
#|  - Interval Estimate for the Average over People 


x = seq(-10000,20000, by=10)
point.estimate.years = average.over.years(summaries)
w.years = 2*1.96*sd(boots$years) 
normal.distribution.plot.years = ggplot() + 
    geom_area(aes(x=x, y=dnorm(x, mean=mean(boots$years), sd=sd(boots$years))), color='purple', fill='purple', alpha=.08) +
    geom_vline(aes(xintercept=mean(boots$years)), color='purple', linetype='solid') +
    geom_vline(aes(xintercept=xi), data=data.frame(xi=qnorm(c(.025,.975), mean=mean(boots$years), sd=sd(boots$years))), color='purple', linetype='dotted') +
    geom_pointrange(aes(y=dnorm(2*sd(boots$years), mean=0, sd=sd(boots$years)), 
    x=point.estimate.years, xmin=point.estimate.years - w.years/2, xmax=point.estimate.years + w.years/2)) + 
    coord_cartesian(xlim=c(-5000,15000))

point.estimate.people = average.over.people(summaries)
w.people = 2*1.96*sd(boots$people) 
normal.distribution.plot.people = ggplot() + 
    geom_area(aes(x=x, y=dnorm(x, mean=mean(boots$people), sd=sd(boots$people))), color='purple', fill='purple', alpha=.08) +
    geom_vline(aes(xintercept=mean(boots$people)), color='purple', linetype='solid') +
    geom_vline(aes(xintercept=xi), data=data.frame(xi=qnorm(c(.025,.975), mean=mean(boots$people), sd=sd(boots$years))), color='purple', linetype='dotted') +
    geom_pointrange(aes(y=dnorm(2*sd(boots$people), mean=0, sd=sd(boots$people)), 
    x=point.estimate.people, xmin=point.estimate.people - w.people/2, xmax=point.estimate.people + w.people/2)) + 
    coord_cartesian(xlim=c(-5000,15000))

normal.distribution.plot.years
normal.distribution.plot.people
```

```{r}
earnings = sample.tab
N = function(x) { sum(earnings$education==x) }
P = function(x) { sum(earnings$education==x) / sum(earnings$education <= 12) }
alphahat = function(x) { if (x==12) { P(12) } else if (x==8) { -P(9) } else { P(x) - P(x+1) } }
sigmahat = function(x) { sd(earnings$income[earnings$education==x]) }
alpha    = function(x) { case_match(x, 
  12 ~  1/4,
  8  ~ -1/4,
  .default = 0) 
}
se.years = sd(boots$years)
se.people.EV = sqrt(sum(sapply(8:12, function(x) { alphahat(x)^2 * summaries$sd[summaries$education==x]^2 / sum(earnings$education==x) })))
se.people.EV.terms = sqrt(sapply(8:12, function(x) { alphahat(x)^2 * summaries$sd[summaries$education==x]^2 / sum(earnings$education==x) }))
```

$$
\Var\qty[\sum_x \hat\alpha(x) \hat\mu(x)] \ge \E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] = \sum_x \sigma^2(x) \times \E\qty[ \frac{\hat\alpha(x)^2}{N_x} ]
$$

- Let's think about what this means for our two examples.
$$
\small{
\begin{aligned}
\hat\theta_{\text{years}}  &= \frac14\sum_{x=9}^{12} \qty{ \hat\mu(x) - \hat\mu(x-1)} = \sum_x \alpha(x) \hat \mu(x)  \qfor \alpha(x) = \begin{cases} \hphantom{-}\frac14 & \text{ if } x = 12 \\ -\frac14 & \text{ if } x= 8 \\ 0 & \text{ otherwise} \end{cases} \\
\hat\theta_{\text{people}} &= \sum_{x=9}^{12} P_x \qty{ \hat\mu(x) - \hat\mu(x-1)} = \sum_x \hat \alpha(x) \hat \mu(x) \qfor
\hat\alpha(x) = \begin{cases}
  \hphantom{-}P_{12} & \text{ if } x = 12 \\
  P_{x} - P_{x+1} & \text{ if } x \in \{9 \ldots 11\} \\
  -P_9 & \text{ if } x = 8 
\end{cases}
\end{aligned}
}
$$

- We've seen that the variance of our estimate of the average-over-years summary has large variance.
- The problem was that it paired a relatively large weight, $\alpha(x)=-1/4$, with a very small subsample size $N_8=`r sum(earnings$education==8)`$.
- And we know the average-over-people summary has an even larger variance. Let's compare.

## A Comparison

$$ 
\begin{array}{l|c|ccccc|c}
& x & 8 & 9 & 10 & 11 & 12 & \sum \\
& N_x & `r N(8)` & `r N(9)` & `r N(10)` & `r N(11)` & `r N(12)` & `r sum(sapply(8:12, N))` \\
& \hat\sigma^2(x) & `r approx(sigmahat(8)^2,1)` & `r approx(sigmahat(9)^2,1)` & `r approx(sigmahat(10)^2,1)` & `r approx(sigmahat(11)^2,1)` & `r approx(sigmahat(12)^2,1)` & \\
&\hline \\ 
\text{for}\ \ \hat\theta_{\text{people}} 
&\hat\alpha^2(x) & `r approx(alphahat(8)^2)` & `r approx(alphahat(9)^2)` & `r approx(alphahat(10)^2)` & `r approx(alphahat(11)^2)` & `r approx(alphahat(12)^2)` & \\
&\frac{\hat\sigma^2(x)\alpha^2(x)}{N_x} & `r approx(sigmahat(8)^2*alphahat(8)^2/N(8),1)` & `r approx(sigmahat(9)^2*alphahat(9)^2/N(9),1)` & `r approx(sigmahat(10)^2*alphahat(10)^2/N(10),1)` & \textcolor{blue}{`r approx(sigmahat(11)^2*alphahat(11)^2/N(11),.1)`} & `r approx(sigmahat(12)^2*alphahat(12)^2/N(12),1)` 
& `r approx(sum(sapply(8:12, function(x) { sigmahat(x)^2 * alphahat(x)^2 / N(x) })),.1)` \\
\hline \\
\text{for}\ \ \hat\theta_{\text{years}} 
&\hat\alpha^2(x) & `r approx(alpha(8)^2)` & `r approx(alpha(9)^2)` & `r approx(alpha(10)^2)` & `r approx(alpha(11)^2)` & `r approx(alpha(12)^2)` &  \\
&\frac{\hat\sigma^2(x) \alpha^2(x)}{N_x} & \textcolor{blue}{`r approx(sigmahat(8)^2*alpha(8)^2/N(8),1)`} & `r approx(sigmahat(9)^2*alpha(9)^2/N(9),1)` & `r approx(sigmahat(10)^2*alpha(10)^2/N(10),1)` & `r approx(sigmahat(11)^2*alpha(11)^2/N(11),1)` & `r approx(sigmahat(12)^2*alpha(12)^2/N(12),1)` & `r approx(sum(sapply(8:12, function(x) { sigmahat(x)^2 * alpha(x)^2 / N(x) })),1)` \\
\\
\end{array}
$$

- For both summaries, *most* of the variance comes from [one term]{.fg style="color:blue"}.
- Let's try to compare their variances ignoring all but the biggest term in each. 
  - For the average over years, it's the $x=8$ subsample, with a sample size of $N_x=`r N(8)`$ and a coefficient of $-1/4$.
  - For the average over people, it's the $x=11$ subsample, with ...
    - a sample size $N_x=`r N(11)`$ that's roughly 4x bigger.
    - but a coefficient of $`r round(alphahat(11),3)`$ that's roughly `r round(alphahat(11)/alpha(8))`x bigger, too.
- Variance grows with the *square* of the coefficient divided by the sample size.
- So this approximation suggests variance would be roughly $3^2  / 4 \approx 2.25x$ bigger for the average over people.
  - That's not far off what we get when we compare bootstrap variance estimates. 
  - That's `r approx(sd(boots$people)^2/sd(boots$years)^2, .1)`x bigger.
- *Intuitive Conclusion*. 
  - Summaries that involve the means of small groups are hard to estimate.
  - But that's not all that matters. It matters *how much* they're used.

## What About The Other Part?

$$
\Var\qty[ \sum_x \alpha(x) \hat \mu(x) ] 
= \E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] \quad + \quad \Var\qty[ \E\qty{\sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n} ] 
$$

- What we've been thinking about so far is the first part of the variance.
  - That's zero when the coefficients are constants, but not when they're random.
  - Because this part was *larger* for the summary with random coefficients, that implied its variance would be larger.
  - If it had been *smaller*, that wouldn't have done much for us.  
- But to gain intuition about the term we left out, let's approximate it.  
  - We can use the bootstrap to estimate the variance of our summary---the whole thing.
  - We'll use the variance (draws from) the bootstrap sampling distribution. It's `r approx(sd(boots$people)^2,.01)`.
  - Our estimate for the first part only --- based on our table --- was `r approx(se.people.EV^2,.01)`.
  - That suggests that the second part should be roughly $`r approx(sd(boots$people)^2,.01)` - `r approx(se.people.EV^2,.01)` \approx `r approx(sd(boots$people)^2 - se.people.EV^2,100)`$.
  



## Summary

- When we're estimating is a linear combination of the *subpopulation means*,
the sample version is unbiased ... 
- *if* the sample coefficients are unbiased estimates of the population ones.

$$
\small{
\hat \theta = \sum_x \alpha(x) \hat\mu(x) \quad \text{ satisfies } \E[\hat\theta] = \theta
}
$$

- That's something you have to think about case-by-case. 
- But proportions---even proportions of subsamples---are, in fact, unbiased.

- Its variance is determined by the variance in each subpopulation, the coefficients, and the subsample sizes. 
- And we can estimate it. Or, at least, a lower bound.
$$
\small{
\Var[ \hat\theta ] \ge \sum_x \sigma^2(x) \alpha^2(x) \times \E\qty[ \frac{1}{N_x} ] \approx  \sum_x \hat\sigma^2(x)\alpha^2(x) \times \frac{1}{N_x}  =: \hat\sigma_{\hat\theta}^2
}
$$
- Fundamentally, we're going to get an imprecise estimate *if a small subsample has a large weight*. 
- By imprecise, we mean a large standard error and a wide confidence interval.

## Formula vs. Bootstrap

$$
\small{
\Var[ \hat\theta ] \ge \sum_x \sigma^2(x) \times \E\qty[\frac{\hat\alpha^2(x)}{N_x} ] \approx  \sum_x \hat\sigma^2(x) \times \frac{\alpha^2(x)}{N_x}  =: \hat\sigma_{\hat\theta}^2
}
$$

- We can use this formula construct a confidence interval based on normal approximation.

$$
\small{
\theta \in \qty[ \hat \theta - 1.96\hat\sigma_{\hat\theta}, \ \hat \theta + 1.96\hat\sigma_{\hat\theta}] \quad \text{ with probability } \approx .95 \qfor \hat\sigma_{\hat\theta}^2 = \sum_x \hat\sigma^2(x)\times \frac{\hat\alpha^2(x)}{N_x}  
}
$$

- Or we could use the bootstrap. Which do you prefer? And why?