---
title: "Lecture 11"
subtitle: "Least Squares Regression in Linear Models"
---


# A Problem With Subsample Averages

## Sometimes There Isn't Much to Average

{{< include shared-code.qmd >}}
{{< include ca-income.qmd >}}

```{r}
ggplot(earnings) + 
  geom_point(aes(x=education, y=income), color='#F7766D',size=dot.size, 
             alpha=.2, position=position_jitter(seed=0,w=.3)) +
  scale_x() + scale_y() + guides(color='none')
```

- Very few people in our California CPS sample have less than 11 years of education.
  - There's a good reason for this: CA law requires residents stay in school until they're 18.
- But that means our 8th-10th grade income averages are based on very few people.
- And that makes them very sensitive to who, exactly, is in the sample.

## Sensitivity

```{r}
#| cache: true

loo.means = \(rows) map(1:nrow(rows), \(i) mean(rows[-i,]$income)) |> unlist() 
loo = earnings |> group_by(education) |> 
                  group_modify(\(rows, group) 
                    data.frame(income = loo.means(rows)))

boot = map(1:100, \(bb) {
  boot.sam = earnings[sample(nrow(earnings), replace=TRUE),]
  boot.sam |> group_by(education) |> summarize(income=mean(income))
}) |> bind_rows()
```

```{r}
#| layout-ncol: 2
#| fig-width: 7
earnings$type = 'individual incomes'
loo$type = 'leave-one-out means'
boot$type = 'bootstrap means'
combined = rbind(earnings[, c('education','type','income')], loo, boot)
combined$type = factor(combined$type, levels=c('individual incomes', 'leave-one-out means', 'bootstrap means'))
education.levels = c(8,12,16,20)


ggplot(combined[combined$education %in% education.levels & 
                combined$type %in% c('individual incomes', 'leave-one-out means'), ]) + 
  geom_point(aes(x=factor(type), color=factor(type), y=income), size=2*dot.size, 
             alpha=.2, position=position_jitter(seed=0,w=.2)) + 
  facet_grid(cols=vars(education)) + guides(color='none') + scale_y(c(0,200000)) + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

ggplot(combined[combined$education %in% education.levels & 
                combined$type %in% c('individual incomes', 'bootstrap means'), ]) + 
  geom_point(aes(x=factor(type), color=factor(type), y=income), size=2*dot.size, 
             alpha=.2, position=position_jitter(seed=0,w=.2)) + 
  facet_grid(cols=vars(education)) + guides(color='none') + scale_y(c(0,200000)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```  

- If we leave even one person out of our 8th grade average, it changes a fair amount.
  - Each person's income is 1/`{r} sum(earnings$education==8)`th of that average.
  - We can see those changes in the 'leave-one-out means' plot on the left.
- That's not a problem with the 12th grade average.
  - Each person's income is 1/`{r} sum(earnings$education==12)`th of that average.
  - And the leave-one-out means barely change.
- We see a more 'statistical' version of this by looking at the bootstrap means plot on the right.
  - The spread of these bootstrap means is a good approximation to the spread \
    we get in the actual sample means when we sample from the population.
  - And it's *huge* for folks with 8th grade educations.


## Sometimes There's Nothing to Average
  
```{r}
#| out-height: 700px
#| fig-asp: .4
#| fig-width: 14
race.categories = c('White only', 'Black only', 'Asian only', 'Hawaiian/Pacific Islander only (HP)')
mean_sd = function(x) { list(y=mean(x), ymin=mean(x)-sd(x), ymax=mean(x)+sd(x)) }
ggplot(earnings[earnings$race %in% race.categories & earnings$education %in% education.levels, ]) + 
  geom_point(aes(x=race, y=income, color=race), size=dot.size, 
             alpha=.2, position=position_jitter(seed=0,w=.3)) +
  stat_summary(aes(x=race, y=income), geom="pointrange", fun.data=mean_sd, size=.5, alpha=.7) +

  facet_grid(cols=vars(education)) + guides(color='none', x=guide_axis(angle=45)) + scale_y(c(0,200000)) 
```

- When we break things out in additional dimensions, e.g. by race, we see a bigger problem.
- We can't even take an average of Black survey respondents with 8th grade educations.
- There's nobody in that column to average.

## Often.
```{r}
#| out-height: 700px
#| fig-asp: .4
#| fig-width: 14
counties = c('alameda', 'san francisco', 'los angeles', 'orange')
cearnings = earnings[earnings$county %in% counties, ]
cearnings$county = factor(cearnings$county, levels=counties) 
ggplot(cearnings[cearnings$education %in% education.levels & cearnings$county %in% counties, ]) +
  geom_point(aes(x=county, y=income, color=county), size=dot.size, 
             alpha=.2, position=position_jitter(seed=0,w=.05)) +
  stat_summary(aes(x=county, y=income), geom="pointrange", fun.data=mean_sd, size=.5, alpha=.7) +
  facet_grid(cols=vars(education)) +
  guides(color='none', x=guide_axis(angle=45)) + scale_y(c(0,200000)) 
```

- When we break things out by county, we have empty columns too.
- We've sampled nobody in San Francisco with an 8th grade education.
- This isn't an edge case. This is common.


## Extremely Common.
```{r}
#| out-height: 700px
#| fig-width: 14
ggplot(cearnings[cearnings$education %in% education.levels & 
                 cearnings$county %in% counties &
                 cearnings$race %in% race.categories, ]) + 
  geom_point(aes(x=race, y=income, color=race), size=2*dot.size, 
             alpha=.2, position=position_jitter(seed=0,w=.05)) +
  stat_summary(aes(x=race, y=income), geom="pointrange", fun.data=mean_sd, size=.5, alpha=.7) +
  facet_grid(cols=vars(county), rows=vars(education)) +
  guides(color='none', x=guide_axis(angle=45)) + scale_y(c(0,200000)) 
```

 - When we break things down in multiple dimensions, we have a lot of empty categories.
 - Even using 3 dimensions---race, county, and education---we have a lot of empties.
   - Those aren't necessarily empty categories in the population: we've only sampled 1/2500 people.
   - But if we need to sample people in those groups to make predictions about them, we're stuck.
     - We don't have anything we can say about these unsampled groups. 
     - So we can't estimate any population summaries involving them. 
 - So we need some way to make predictions for groups with no data.

## A Partial Solution

```{r}
cearnings$degree = ifelse(cearnings$education < 16, 'no 4-year degree', '4-year degree')
cearnings$degree = factor(cearnings$degree, levels=c('no 4-year degree', '4-year degree'))
cearnings$area = ifelse(cearnings$county %in% c('alameda', 'san francisco'), 'sf bay area', 'la area')
cearnings$area = factor(cearnings$area, levels=c('sf bay area', 'la area'))
ggplot(cearnings[cearnings$education %in% education.levels & 
                cearnings$county %in% counties &
                cearnings$race %in% race.categories, ]) + 
  geom_point(aes(x=race, y=income, color=race), size=2*dot.size, 
             alpha=.2, position=position_jitter(seed=0,w=.05)) +
  stat_summary(aes(x=race, y=income), geom="pointrange", fun.data=mean_sd, size=.5, alpha=.7) +
  facet_grid(cols=vars(area), rows=vars(degree)) +
  guides(color='none', x=guide_axis(angle=45)) + scale_y(c(0,200000)) 
```

- We've already seen a partial solution to this problem: coarsening.
- Here we've coarsened in two dimensions.
  - We've coarsened education into two categories: 4-year degree  (&geq; 16 years) vs. no 4-year degree (&lt; 16 years).
  - We've coarsened county into two areas: SF Bay Area (SF, Alameda) and LA Area (LA, Orange).
 - And it helps, but we have two problems.
   - We’ve renamed everything, so we’d have to redefine all our summaries. This is fixable.
   - We still have empty groups, although we have fewer. This is not.


## We Need A General Solution
```{r}
#| out-height: 800px
#| fig-width: 14
ggplot(earnings[earnings$education %in% education.levels & earnings$county != 'unknown', ]) + 
  geom_point(aes(x=race, y=income, color=race), size=2*dot.size, 
             alpha=.2, position=position_jitter(seed=0,w=.05)) +
  stat_summary(aes(x=race, y=income), geom="pointrange", fun.data=mean_sd, size=.5, alpha=.7) +
  facet_grid(cols=vars(county), rows=vars(education)) +
  guides(color='none', x=guide_axis(angle=45)) + scale_y(c(0,200000)) 
```

- This empty groups problem is a much bigger deal with this data
- We've been looking at some of the most populous counties and common racial identities.
- If we use all the counties in CA and all the identities the CPS includes, **most groups are empty**.
- And the range of coarsening options is overwhelming.

## Today's Plan 

```{r}
cearnings$degree = ifelse(cearnings$education < 16, 'no 4-year degree', '4-year degree')
cearnings$degree = factor(cearnings$degree, levels=c('no 4-year degree', '4-year degree'))
cearnings$area = ifelse(cearnings$county %in% c('alameda', 'san francisco'), 'sf bay area', 'la area')
cearnings$area = factor(cearnings$area, levels=c('sf bay area', 'la area'))
ggplot(cearnings[cearnings$education %in% education.levels & 
                cearnings$county %in% counties &
                cearnings$race %in% race.categories, ]) + 
  geom_point(aes(x=race, y=income, color=race), size=2*dot.size, 
             alpha=.2, position=position_jitter(seed=0,w=.05)) +
  stat_summary(aes(x=race, y=income), geom="pointrange", fun.data=mean_sd, size=.5, alpha=.7) +
  facet_grid(cols=vars(area), rows=vars(degree)) +
  guides(color='none', x=guide_axis(angle=45)) + scale_y(c(0,200000)) 
```


- Today, we're going to talk about how to deal with the problem of small or empty groups.
- To make this more manageable, we'll break things down into a two-step process.
  1. We'll decide on a **regression model**: a set of functions we could use to make predictions.
  2. We'll choose the best function from our model using the **least squares** criterion.
- We call the function we choose this way the **least squares predictor** within our model.

## Notation 
```{r}
#| layout-ncol: 2
#| fig-width: 7
#| fig-cap: 
#|  - The sample 
#|  - An illustration of the population 

pop.dot.size = .75
n = nrow(earnings)
m = nrow(pop.earnings)
pop.idx = c(1:4*n, (m-n+1):m) 

earnings.jitter = ggplot(earnings) + geom_point(aes(x=education, y=income, color=degree), 
                              size=dot.size, position=position_jitter(seed=0,w=.3), alpha=.2) +
                   scale_x() + scale_y() + guides(color=FALSE)
                    
earnings.jitter.pop = ggplot(pop.earnings[pop.idx, ]) + 
    geom_point(aes(x=education, y=income, color=degree), 
               size=pop.dot.size, position=position_jitter(seed=0,w=.3), alpha=.2) +
     scale_x() + scale_y() + guides(color='none')

pop.summaries = pop.earnings |> 
  group_by(education) |> 
  summarize(mean=mean(income), sd=sd(income), meanx=mean(education))

with.summaries = function(pop) { 
  pop.summaries = pop |> 
    group_by(education) |> 
  summarize(mean=mean(income), sd=sd(income), meanx=mean(education))
  geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), data=pop.summaries)
}

earnings.jitter + with.summaries(earnings)
earnings.jitter.pop + with.summaries(pop.earnings)
```


- So far, we've used $\mu(x)$ to refer to mean of a column in the population plot.

$$
\mu(x) = \frac{1}{m_x}\sum_{j:x_j=x} y_j \quad \text{ for } m_x = \sum_{j:x_j=x} 1
$$

- We've used $\hat\mu(x)$ for *one particular estimate* of $\mu(x)$: the mean of the corresponding column in the sample.

$$
\hat\mu(x) = \frac{1}{N_x}\sum_{i:X_i=x} Y_i \quad \text{ for } N_x = \sum_{i:X_i=x} 1
$$

- From today on, we'll be a bit more flexible with the meaning of $\hat\mu$.
  - We'll use $\hat\mu(x)$ to refer to *any* estimator of the subpopulation mean $\mu(x)$.
  - Which should be clear from context. 


# A Least Squares Interpretation of Means

## Reinterpreting the Sample Mean

```{r}
Y = earnings$income[earnings$education==8]
X = 1:length(Y) 

ggplot() + 
  geom_point(aes(x=X, y=Y), alpha=.5, size=3) +
  geom_hline(yintercept=mean(Y), color='red') +
  geom_segment(aes(x=X, xend=X, y=mean(Y), yend=Y), alpha=.2, color='red') +
  geom_hline(yintercept=median(Y), color='orange')  +
  geom_segment(aes(x=X+.1, xend=X+.1, y=median(Y), yend=Y), alpha=.2, color='orange') +
  geom_hline(yintercept=0, color='blue')  + 
  geom_segment(aes(x=X-.1, xend=X-.1, y=0, yend=Y), alpha=.2, color='blue') + 
  scale_y(c(0,60000)) 
```

- Let's stop thinking about the sample mean *procedurally*, i.e., how we calculate it
- Let's think of it as a *choice*, i.e., the best number for summarizing the data according to some criterion.
- Here are the outcomes $Y_i$ for sampled Californians with 8 years of education and three potential summaries of the location of those outcomes:
  1. The sample mode ([blue line]{.fg style="color:blue"})
  2. The sample median ([orange line]{.fg style="color:orange"})
  3. The sample mean ([red line]{.fg style="color:red"})
- All of these are reasonable summaries. And the best choice according to some reasonable criterion. 
- Criteria which, sensibly, look what's left over when we compare our observations to our location summary.

$$
\hat\varepsilon_i = Y_i - \hat\mu \qquad \text{ are the \textbf{residuals}}
$$


## Residuals

```{r}
signsq = function(x) { sign(x)*x^2 / 40000 }

ggplot() + 
  geom_point(aes(x=X, y=Y), alpha=.5, size=3) +
  geom_hline(yintercept=mean(Y), color='red') +
  geom_segment(aes(x=X, xend=X, y=mean(Y), yend=Y), alpha=.2, color='red') +
  geom_hline(yintercept=median(Y), color='orange')  +
  geom_segment(aes(x=X+.1, xend=X+.1, y=median(Y), yend=Y), alpha=.2, color='orange') +
  geom_hline(yintercept=0, color='blue')  + 
  geom_segment(aes(x=X-.1, xend=X-.1, y=0, yend=Y), alpha=.2, color='blue') + 
  scale_y(c(0,60000)) 

```

- Here we're looking at the residuals for each of our location summaries $\hat\mu$.
  1. $\hat\varepsilon_i = Y_i - \hat\mu$ when $\hat\mu$ is the mode ([blue points]{.fg style="color:blue"}). It has the largest number of *zero residuals*.
  2. $\hat\varepsilon_i = Y_i - \hat\mu$ when $\hat\mu$ is the sample median ([orange points]{.fg style="color:orange"}). It has the *smallest sum of absolute residuals*.
  3. $\hat\varepsilon_i = Y_i - \hat\mu$ when $\hat\mu$ is the sample mean ([red points]{.fg style="color:red"}). It has the *smallest sum of squared residuals*. 
- For better or for worse, the sum of squared residuals is the criterion we tend to use.

$$ 
\textcolor{red}{\hat \mu} = \argmin_{\text{numbers}\ m} \sum_{i=1}^n \qty{ Y_i - m }^2 \quad \text{ is the \textbf{least squares estimate}}
$$

### Terminology 

- The *argmin* of a function is the argument at which the function is minimized.
$$
\hat\mu = \argmin_{m} \ f(m) \quad \iff \quad f(\hat\mu) = \min_m \ f(m)
$$
- Here that function is the sum of squared residuals, $f(m) = \sum_{i=1}^n \qty{ Y_i - m }^2$.

## The Least Squares Location Summary

```{r}
#| layout-ncol: 2
#| fig-width: 7 
ggplot() + 
  geom_point(aes(x=X,y=Y-mean(Y)), size=3, color='red') +  
  geom_hline(yintercept=0, alpha=.1, linewidth=2) + scale_y(c(-40000, 60000)) +
  geom_segment(aes(x=X, xend=X, y=0, yend=signsq(Y-mean(Y))), alpha=.2)

ssr = Vectorize(\(m) sum((Y-m))^2)
x = seq(-40000, 80000, length.out=1000)
ggplot() + geom_line(aes(x=x, y=ssr(x)))  +
  geom_point(aes(x=mean(Y), y=ssr(mean(Y))), color='red', size=4) + 
  geom_point(aes(x=median(Y), y=ssr(median(Y))), color='orange', size=4) +
  geom_point(aes(x=0, y=ssr(0)), color='blue', size=4)  
```
$$ 
\hat \mu = \argmin_{\text{numbers}\ m} \sum_{i=1}^n \qty{ Y_i - m }^2 \quad \text{ satisfies the \textbf{zero-derivative condition} }
$$
$$ 
\begin{aligned}
0 
&= \frac{d}{dm} \sum_{i=1}^n \qty{ Y_i - m }^2  \ \ \mid_{m=\hat\mu} \\
&= \class{fragment}{\sum_{i=1}^n \frac{d}{dm} \qty{ Y_i - m }^2  \ \ \mid_{m=\hat\mu}} \\ 
&= \class{fragment}{\sum_{i=1}^n -2 \qty{ Y_i - m } \ \ \mid_{m=\hat\mu}} \\
&= \class{fragment}{\sum_{i=1}^n -2 \qty{ Y_i - \hat\mu }} 
\end{aligned}
$$

- This says the least squares residuals [$\hat\varepsilon = Y_i - \hat \mu$]{.fg style="color:red"} *sum to zero*. 
- With a bit of algebra, this tells us that the least squares estimator $\hat\mu$ is the sample mean.

$$
\begin{aligned}
&0 = \class{fragment}{\sum_{i=1}^n -2\qty{ Y_i - \hat\mu }} && \text{ when } \\ 
&\class{fragment}{2\sum_{i=1}^n Y_i = 2\sum_{i=1}^n \hat\mu = 2n\hat\mu} && \text{ and therefore when } \\
&\class{fragment}{\frac{1}{n}\sum_{i=1}^n Y_i = \hat\mu} 
\end{aligned}
$$

- It's not just the best choice of the 3 location summaries we've looked at. 
- It's the best choice --- in terms of the sum of squared errors criterion --- of all numbers outright.

## Two Locations: Regression with a Binary Covariate

```{r}
estimate_mode = function(x, na.rm = FALSE) {
  if(na.rm){
    x = x[!is.na(x)]
  }
  ux <- unique(x)
  return(ux[which.max(tabulate(match(x, ux)))])
}
Y = earnings$income[earnings$education %in% c(8,12)]
X = earnings$education[earnings$education %in% c(8,12)]

ggplot(data.frame(X=X,Y=Y)) + 
  geom_point(aes(x=X,y=Y), alpha=.2, size=1, position=position_jitter(w=.5)) + 
  stat_summary(aes(x=X, y=Y), geom="point", fun=mean, size=5, alpha=.7, color='red') +
  stat_summary(aes(x=X, y=Y), geom="point", fun=median, size=5, alpha=.7, color='orange') +
  stat_summary(aes(x=X, y=Y), geom="point", fun=estimate_mode, size=5, alpha=.7, color='blue') +
  scale_y(c(0,60000)) 
```

- Suppose we have two groups: people in our sample with 8 and 12 years of education. 
- We can find the best location for each using this criterion of minimizing squared residuals.
- We can think of the *column means* $\hat\mu(x)$ as the best **function of x** for predicting **y**.
  - That is, the best function of years of education for predicting income. 
  - Where *best*  means the one that minimizes the sum of squared residuals. 

$$ 
\hat \mu(x) = \argmin_{\text{functions}\ m(x)} \sum_{i=1}^n \qty{ Y_i - m(X_i) }^2 
$$


## Two Locations: Regression with a Binary Covariate
```{r}
#| fig-asp: .2
ggplot(data.frame(X=X,Y=Y)) + 
  geom_point(aes(x=X,y=Y), alpha=.2, size=1, position=position_jitter(w=.5)) + 
  stat_summary(aes(x=X, y=Y), geom="point", fun=mean, size=5, alpha=.7, color='red') +
  stat_summary(aes(x=X, y=Y), geom="point", fun=median, size=5, alpha=.7, color='orange') +
  stat_summary(aes(x=X, y=Y), geom="point", fun=estimate_mode, size=5, alpha=.7, color='blue') +
  scale_y(c(0,60000)) 
```

- Why? Recall that a function of $X$ is just a number for each value of $X$.
- So to see that the sub-sample means are the solution, we can break our sum into pieces where $X$ is the same.

$$
\hat \mu(x) = \argmin_{\text{functions}\ m(x)} \sum_{i:X_i=8} \{ Y_i - m(8) \}^2  + \sum_{i:X_i=12} \{ Y_i - m(12) \}^2
$$

- And observe that the first sum depends only on $m(8)$ and the second only on $m(12)$. 
  - So we get $\hat\mu(8)$ by minimizing squared residuals in the $X=8$ column.
  - And we get $\hat\mu(12)$ by minimizing squared residuals in the $X=12$ column.
- The minimizers --- which we can solve for exactly as before --- are the means in each column.

$$
\begin{aligned}
\hat \mu(8)   &= \argmin_{\text{numbers}\ m} \sum_{i:X_i=8} \{ Y_i - m \}^2  = \frac{\sum_{i:X_i=8} Y_i}{\sum_{i:X_i=8} 1 } \\
\hat \mu(12)  &= \argmin_{\text{numbers}\ m} \sum_{i:X_i=12} \{ Y_i - m \}^2 = \frac{\sum_{i:X_i=12} Y_i}{\sum_{i:X_i=12} 1 }
\end{aligned}
$$

## Many Locations: Regression with a Multiple or Many-Valued Covariates

```{r}
X = earnings$education
Y = earnings$income
ggplot(data.frame(X=X,Y=Y)) + 
  geom_point(aes(x=X,y=Y), alpha=.2, size=1, position=position_jitter(w=.3)) + 
  stat_summary(aes(x=X, y=Y), geom="point", fun=mean, size=5, alpha=.7, color='red') +
  stat_summary(aes(x=X, y=Y), geom="point", fun=median, size=5, alpha=.7, color='orange') +
  stat_summary(aes(x=X, y=Y), geom="point", fun=estimate_mode, size=5, alpha=.7, color='blue') +
  scale_y(c(0,150000)) + scale_x() 
```

- There's nothing special about two locations. We can do this for as many locations as we like.
- We break our sum into pieces where $X$ is the same.
$$
\hat \mu(x) = \argmin_{\text{functions}\ m(x)} \sum_x \sum_{i:X_i=x} \{ Y_i - m(x) \}^2 
$$

- And we get the best location at each $x$, in terms of squared residuals, by minimizing each piece.
- The solution is, as before, the mean of each subsample.

$$
\hat \mu(x) = \argmin_{\text{numbers}\ m} \sum_{i:X_i=x} \{ Y_i - m \}^2  = \frac{\sum_{i:X_i=x} Y_i}{\sum_{i:X_i=x} 1 }
$$

## Coarsening

```{r}
#| fig-asp: .2
X = earnings$education
Y = earnings$income
Xcoarse = ifelse(X < 16, mean(X[X<16]), mean(X[X>=16]))
df.small = data.frame(x=unique(X[X<16]), y=mean(Y[X<16]))
df.big = data.frame(x=unique(X[X>=16]), y=mean(Y[X>=16]))
ggplot(data.frame(X=X,Y=Y,Xcoarse=Xcoarse)) + 
  geom_point(aes(x=X,y=Y, color=factor(Xcoarse)), alpha=.2, size=1, position=position_jitter(w=.3)) + 
  stat_summary(aes(x=Xcoarse, y=Y), geom="point", fun=mean, size=5, alpha=.7, color='red') +
  stat_summary(aes(x=Xcoarse, y=Y), geom="point", fun=median, size=5, alpha=.7, color='orange') +
  stat_summary(aes(x=Xcoarse, y=Y), geom="point", fun=estimate_mode, size=5, alpha=.7, color='blue') +
  geom_point(aes(x=x, y=y), data=df.small, color='red') +
  geom_line(aes(x=x, y=y), data=df.small, color='red') +
  geom_point(aes(x=x, y=y), data=df.big, color='red') +
  geom_line(aes(x=x, y=y), data=df.big, color='red') + 
  coord_cartesian(ylim=c(0,150000)) + guides(color='none') 
```

- When we coarsen, we're in effect restricting the set of functions we're considering.
- We're restricting them to be functions of the coarsened groups $c(x)$, i.e. the color in the plot.
- Now we break our sum into pieces where the coarsened group $c(x)$ is the same.

$$
\begin{aligned}
\hat \mu(c(x)) 
&= \argmin_{\text{functions}\ m(c(x)) } \sum_{c(x)} \sum_{i:c(X_i)=c(x)} \{ Y_i - m(c(x)) \}^2 \\
&= \argmin_{\text{functions}\ m(c(x)) } \inred{\sum_{i:c(X_i)=\text{red}} \{ Y_i - m(\text{red}) \}^2} + \ingreen{\sum_{i:c(X_i)=\text{green}} \{ Y_i - m(\text{green}) \}^2}
\end{aligned}
$$

- And we get the best location at each coarsened $x$, in terms of squared error, by minimizing each piece.
- And the solution is the mean of each coarsened group.

$$
\hat \mu(x) = \argmin_{\text{numbers}\ m} \sum_{i:c(X_i)=c(x)} \{ Y_i - m \}^2  = \frac{\sum_{i:c(X_i)=c(x)} Y_i}{\sum_{i:c(X_i)=c(x)} 1 }
$$

- This says the prediction we make at each $x$ is the mean of the $y$ values in the coarsened group it belongs to.
  
## Least Squares Regression

```{r}
#| layout-ncol: 2
#| fig-width: 7
X = earnings$education
Y = earnings$income
dfXY = data.frame(X=X,Y=Y)
ggplot(dfXY) + 
  geom_point(aes(x=X,y=Y), alpha=.2, size=1, position=position_jitter(w=.3)) + 
  stat_summary(aes(x=X, y=Y), geom="point", fun=mean, size=5, alpha=.7, color='red') +
  stat_summary(aes(x=X, y=Y), geom="point", fun=median, size=5, alpha=.7, color='orange') +
  stat_summary(aes(x=X, y=Y), geom="point", fun=estimate_mode, size=5, alpha=.7, color='blue') +
  scale_y(c(0,150000)) + scale_x()

X = earnings$education
Y = earnings$income
Xcoarse = ifelse(X < 16, mean(X[X<16]), mean(X[X>=16]))
df.small = data.frame(x=unique(X[X<16]), y=mean(Y[X<16]))
df.big = data.frame(x=unique(X[X>=16]), y=mean(Y[X>=16]))
ggplot(data.frame(X=X,Y=Y,Xcoarse=Xcoarse)) + 
  geom_point(aes(x=X,y=Y, color=factor(Xcoarse)), alpha=.2, size=1, position=position_jitter(w=.3)) + 
  stat_summary(aes(x=Xcoarse, y=Y), geom="point", fun=mean, size=5, alpha=.7, color='red') +
  stat_summary(aes(x=Xcoarse, y=Y), geom="point", fun=median, size=5, alpha=.7, color='orange') +
  stat_summary(aes(x=Xcoarse, y=Y), geom="point", fun=estimate_mode, size=5, alpha=.7, color='blue') +
  geom_point(aes(x=x, y=y), data=df.small, color='red') +
  geom_line(aes(x=x, y=y), data=df.small, color='red') +
  geom_point(aes(x=x, y=y), data=df.big, color='red') +
  geom_line(aes(x=x, y=y), data=df.big, color='red') + 
  coord_cartesian(ylim=c(0,150000)) + guides(color='none') 
```

- We can think of both *uncoarsened* and *coarsened* subsample means as least squares estimators.
- Each is the best choice, within some set of functions, for predicting $Y$ from $X$.
- But the set of functions we're choosing from is different in each case.
  - In the uncoarsened case, we're choosing from *all* functions of $X$.
  - In the coarsened case, we're choosing from functions of the *coarsened* $X$.
  - We could consider other sets of functions as well.

$$
\hat\mu = \argmin_{m \in \model} \sum_{i=1}^n \{ Y_i - m(X_i) \}^2 \qquad \text{ where } \quad \mathcal{M} \ \ \text{ is our model.}
$$

# Least Squares Regression in Linear Models 

## Least Squares Regression
```{r}
muhat.monotone = as.stepfun(isoreg(X,Y))
muhat.lin = lm(Y ~ X)
muhat.piecewise = function(x) { ifelse(x < 16, mean(Y[X<16]), mean(Y[X>=16])) }
x = seq(0,20,length.out=1000)
lsq.plot = ggplot(data.frame(x=X, y=Y)) + geom_point(aes(x=X,y=Y), alpha=.1, size=1, position=position_jitter(w=.3)) + 
  stat_summary(aes(x=X+.1, y=Y), fun=mean, geom='point', color='red', alpha=.5) + 
  stat_summary(aes(x=X+.1, y=Y), fun=mean, geom='line', color='red', alpha=.3) + 
  geom_point(aes(X-.1, predict(muhat.lin)),  color='blue', alpha=.3) +
  geom_line(aes(X-.1, predict(muhat.lin)),  color='blue', alpha=.3) +
  geom_point(aes(x=X, y=muhat.monotone(X)), color='magenta', alpha=.3) + 
  geom_line(aes(x=X, y=muhat.monotone(X)), color='magenta', alpha=.3) + 
  geom_point(aes(x=x, y=muhat.piecewise(x)), color='cyan', alpha=.3) + 
  geom_line(aes(x=x, y=muhat.piecewise(x)), color='cyan', alpha=.3) + 
  scale_y(c(0,150000)) + scale_x()
lsq.plot
```

- Regression, generally, is choosing a function of covariates $X_i$ to predict outcomes $Y_i$. 
- We choose from a set of functions $\model$ that we call a **regression model**.

$$
Y_i \approx \hat\mu(X_i) \qfor \hat\mu \in \model 
$$

- In this class, we'll choose by **least squares**. We can do this for any model we like.

$$
\begin{aligned}
\hat\mu &= \argmin_{m \in \model} \sum_{i=1}^n \{ Y_i - m(X_i) \}^2 \qfor 
&& \textcolor{red}{\model} = \qty{ \text{all functions} \  m(x) } \\       
&&& \textcolor{blue}{\model} = \qty{ \text{all lines} \ m(x) = a + bx }  \\
&&& \textcolor{magenta}{\model} = \qty{ \text{all increasing functions} \  m(x) } \\ 
&&& \textcolor{cyan}{\model} = \qty{ \text{all functions of an indicator} \  m(x)=f(1_{\ge 16}(x))} \\    
\end{aligned}
$$

## Least Squares in Linear Models

```{r}
#| out-height: 400px
#| fig-asp: .4
lsq.plot
```

- We're going to focus on choosing from **linear models**.
  - These are sets of functions that are 'closed' under addition, subtraction, and scaling by constants.
  - This means when we add, subtract, and scale functions in the model, we get another function in the model.
$$
a(x), b(x) \in \model \implies a(x) + b(x), a(x) - b(x), c a(x) \in \model \quad \text{ for any constant $c$ }
$$

- **Q.** Which of these models are linear? 

$$
\begin{aligned}
\textcolor{red}{\model} &= \qty{ \text{all functions} \  m(x) }                               \\
\textcolor{blue}{\model} &= \qty{ \text{all lines} \ m(x)=a + bx }  \\
\textcolor{magenta}{\model} &= \qty{ \text{all increasing functions} \  m(x) }   \\
\textcolor{cyan}{\model} &= \qty{ \text{all functions of an indicator} \  m(x)=f(1_{\ge 16}(x))} \\    
\end{aligned}
$$

# Multivariate Regression Models
A few examples. 

## Additive Models 

```{r}
#| out-height: 400px
#| fig-asp: .4
 
grid = expand.grid(education=unique(earnings$education), race=race.categories) 
muhat = function(model) { function(ed,race) { predict(model, newdata=data.frame(education=ed, race=race)) } }

additive.fit   = lm(income ~ race+factor(education), data=earnings[earnings$race %in% race.categories,])
ggplot(earnings[earnings$race %in% race.categories, ]) + 
  geom_point(aes(x=education, y=income), size=1, 
             alpha=.1, position=position_jitter(seed=0,w=.3)) +
  geom_point(aes(x=education, y=muhat(additive.fit)(education,race)),  color='blue', data=grid) +
  geom_line(aes(x=education, y=muhat(additive.fit)(education,race)),  color='blue', data=grid, alpha=.5) +
  facet_grid(cols=vars(race)) + scale_y(c(0,150000))
```

- Here we let our predictions be arbitrary functions of education ($x$) as before.
- But we're requiring that race ($w$) only shift that function up and down---it can't change the function's shape.
- The result, in visual terms, are predictions for different values of $w$ are 'parallel curves'.

$$\small{
\begin{aligned}
\textcolor{blue}{\model} &= \qty{ m(w,x) = m_0(w) + m_1(x) \ \ \text{for univariate functions} \ \ m_0, m_1 }  &&\text{additive bivariate model} \\
\end{aligned}
}
$$

## The Parallel Lines Model 

```{r}
#| out-height: 400px
#| fig-asp: .4
 grid = expand.grid(education=unique(earnings$education), race=race.categories) 
muhat = function(model) { function(ed,race) { predict(model, newdata=data.frame(education=ed, race=race)) } }

additive.fit   = lm(income ~ race+factor(education), data=earnings[earnings$race %in% race.categories,])
additive.line.fit   = lm(income ~ race+education, data=earnings[earnings$race %in% race.categories,])
ggplot(earnings[earnings$race %in% race.categories, ]) + 
  geom_point(aes(x=education, y=income), size=1, 
             alpha=.1, position=position_jitter(seed=0,w=.3)) +
  geom_point(aes(x=education, y=muhat(additive.line.fit)(education,race)),  color='red', data=grid) +
  geom_line(aes(x=education, y=muhat(additive.line.fit)(education,race)),  color='red', data=grid, alpha=.5) +
  facet_grid(cols=vars(race)) + scale_y(c(0,150000))
```

- Here we impose the additional restriction that the function of education ($x$) is a line.
- This means that the predictions for different values of $w$ are parallel lines: same slope, different intercepts.

$$\small{
\begin{aligned}
\textcolor{red}{\model} &= \qty{ m(w,x) = a(w) + b x   }                 &&\text{parallel lines} \\
\end{aligned}
}
$$

## The Not-Necessarily-Parallel Lines Model 

```{r}
#| out-height: 400px
#| fig-asp: .4
 
additive.line.fit   = lm(income ~ race+education, data=earnings[earnings$race %in% race.categories,])
interacted.line.fit = lm(income ~ race*education, data=earnings[earnings$race %in% race.categories,])
ggplot(earnings[earnings$race %in% race.categories, ]) + 
  geom_point(aes(x=education, y=income), size=1, 
             alpha=.1, position=position_jitter(seed=0,w=.3)) +
  geom_point(aes(x=education, y=muhat(interacted.line.fit)(education,race)),  color='magenta', data=grid) +
  geom_line(aes(x=education, y=muhat(interacted.line.fit)(education,race)),   color='magenta', data=grid, alpha=.5) +
  facet_grid(cols=vars(race)) + scale_y(c(0,150000))
```

- Here we stick with the lines thing, but drop the additivity restriction.
- For each value of $w$, we can get a totally different line---different intercept and slope.

$$\small{
\begin{aligned}
\textcolor{magenta}{\model} &= \qty{ m(w,x) = a(w) + b(w) x  }                 &&\text{lines} \\
\end{aligned}
}
$$

## Where This Leaves Us

```{r}
#| out-height: 400px
#| fig-asp: .4
 
ggplot(earnings[earnings$race %in% race.categories, ]) + 
  geom_point(aes(x=education, y=income), size=1, 
             alpha=.1, position=position_jitter(seed=0,w=.3)) +
  geom_point(aes(x=education, y=muhat(interacted.line.fit)(education,race)),  color='magenta', data=grid) +
  geom_line(aes(x=education, y=muhat(interacted.line.fit)(education,race)),   color='magenta', data=grid, alpha=.5) +
  geom_point(aes(x=education, y=muhat(additive.line.fit)(education,race)),  color='red', data=grid) +
  geom_line(aes(x=education, y=muhat(additive.line.fit)(education,race)),  color='red', data=grid, alpha=.5) +
  geom_point(aes(x=education, y=muhat(additive.fit)(education,race)),  color='blue', data=grid) +
  geom_line(aes(x=education, y=muhat(additive.fit)(education,race)),  color='blue', data=grid, alpha=.5) +
  facet_grid(cols=vars(race)) + scale_y(c(0,150000))
```

- Each of these approaches allows us to make predictions everywhere---even in columns we haven't sampled.
- Fundamentally, it's because they're using information from one column to make predictions in another.
- **Q**. This solves the empty-groups problem, but it introduces a new one. What is it? 

::: fragment

- **A**. We're basing our predictions on assumptions about what $\mu$ looks like. 
- If we're wrong, our predictions will be wrong. It'll *bias*  our estimates of $\mu(x)$.

:::

## Where We're Going

```{r}
#| out-height: 400px
#| fig-asp: .4
 
ggplot(earnings[earnings$race %in% race.categories, ]) + 
  geom_point(aes(x=education, y=income), size=1, 
             alpha=.1, position=position_jitter(seed=0,w=.3)) +
  geom_point(aes(x=education, y=muhat(interacted.line.fit)(education,race)),  color='magenta', data=grid) +
  geom_line(aes(x=education, y=muhat(interacted.line.fit)(education,race)),   color='magenta', data=grid, alpha=.5) +
  geom_point(aes(x=education, y=muhat(additive.line.fit)(education,race)),  color='red', data=grid) +
  geom_line(aes(x=education, y=muhat(additive.line.fit)(education,race)),  color='red', data=grid, alpha=.5) +
  geom_point(aes(x=education, y=muhat(additive.fit)(education,race)),  color='blue', data=grid) +
  geom_line(aes(x=education, y=muhat(additive.fit)(education,race)),  color='blue', data=grid, alpha=.5) +
  facet_grid(cols=vars(race)) + scale_y(c(0,150000))
```

- There's no *right model* to use. Each has its own strengths and weaknesses.
- In the next couple weeks, we'll look into the issues surrounding model choice.
  - Implications of bad model choice.
  - A good model to use by default. 
  - How to choose a model automatically. 

# Practice 

## A Small Dataset

```{r}
#| eval: true
bone.data = data.frame(age        = c(1, 1, 1, 1, 1, 1, 1, 1, 0, 0, 0, 0),
                       supplement = c(1, 1, 0, 1, 1, 1, 0, 0, 0, 1, 0, 1),
                       bone_density = 3+c(-2, -3, -2, 0, -1, 0, -2, 0, 1, 1, -1, 1.5))
plotX = function(w,x) { x + 0.15*(w-0.5) } 

bone.plot = 
  ggplot(bone.data) + 
    geom_point(aes(x=plotX(supplement, age), y=bone_density, color=factor(supplement)),
               alpha=1, size=2, position=position_jitter(width=0.05, height=0, seed=2)) + 
    stat_summary(aes(x=plotX(supplement, age), y=bone_density, color=factor(supplement)),
                 geom='pointrange', fun.data=mean_sd, alpha=.5) + 
    geom_bar(aes(x=plotX(supplement, age), y=3*after_stat(prop), 
                 fill=factor(supplement), color=factor(supplement)), 
             position='identity', alpha=.1, linewidth=.2) + 
    scale_x_continuous(breaks=c(0,1), labels=c('Young', 'Old')) + 
    scale_y_continuous(breaks=c(0,1,2,3,4), sec.axis=sec_axis(~ . / 3, name='Proportion')) + 
    labs(x='', y='') + guides(color='none', fill='none')
     
bone.plot
```


- Researchers hypothesize that higher calcium intake causes greater bone density (stronger bones). 
- Bone density scores indicate how good someone’s bone density is.
  - 3 is typical, larger numbers are better, smaller numbers are worse 
  - a score of 0.5 or lower is used to diagnose osteoporosis. 
- Patients in our sample are classified as old (age 65+) or young (age 64-). 
  - We'll call this $X_i$: $X_i=0$ if patient $i$ is young, $X_i=1$ if patient $i$ is old.
- Each patient in our sample regularly takes a calcium supplement or not. 
  - We'll call this $W_i$: [$W_i=1$]{.groupb} if patient $i$ takes calcium supplement, [$W_i=0$]{.groupa} if they do not.

## The Horizontal Lines Model 

```{r}
prediction.function = function(model.fit) { function(w,x) { predict(model.fit, newdata=data.frame(supplement=w, age=x)) } }
plot.predictions = function(muhat) { 
  grid = expand.grid(supplement=c(0,1), age=c(0,1))
  bone.plot +
    geom_point(aes(x=age, y=muhat(supplement, age), color=factor(supplement)), data=grid) + 
    geom_line(aes(x=age,  y=muhat(supplement, age), color=factor(supplement)), data=grid) +
    guides(color='none', fill='none')
}
```

:::: r-stack 
::: {.fragment .fade-out data-fragment-index="0"}
```{r}
bone.plot 
```
:::
::: {.fragment .current-visible data-fragment-index="0"}
```{r}
horizontal.lines.model.fit = lm(bone_density ~ factor(supplement), data=bone.data)
plot.predictions(prediction.function(horizontal.lines.model.fit))
```
:::
::::

Sketch in the least squares predictor in the **horizontal lines model**.

$$
\mathcal{M} = \qty{ m(w,x) = a(w) \ \ \text{for univariate functions} \ \ a } 
$$

Flip a slide forward to check your answer.


## The Not-Necessarily-Parallel Lines Model 

:::: r-stack 
::: {.fragment .fade-out data-fragment-index="0"}
```{r}
bone.plot 
```
:::
::: {.fragment .fade-in data-fragment-index="0"}
```{r}
not.parallel.lines.model.fit = lm(bone_density ~ factor(supplement) + age + factor(supplement)*age, data=bone.data)
plot.predictions(prediction.function(not.parallel.lines.model.fit))
```
:::
::::

Sketch in the least squares predictor in the **not-necessarily-parallel lines model**.

$$
\mathcal{M} = \qty{ m(w,x) = a(w) + b(w) x \ \ \text{for univariate functions} \ \ a,b } 
$$

Flip a slide forward to check your answer.

## The Parallel Lines Model 

:::: r-stack 
::: {.fragment .fade-out data-fragment-index="0"}
```{r}
bone.plot 
```
:::
::: {.fragment .current-visible data-fragment-index="0"}
```{r}
parallel.lines.model.fit = lm(bone_density ~ factor(supplement) + age, data=bone.data)
plot.predictions(prediction.function(parallel.lines.model.fit))
```
:::
::::

Sketch in the least squares predictor in the **parallel lines model**.

$$
\mathcal{M} = \qty{ m(w,x) = a(w) + b x \ \ \text{for univariate functions} \ \ a \ \ \text{and constants} \ \ b } 
$$

Flip a slide forward to check your answer.



## The Additive Model 

:::: r-stack 
::: {.fragment .fade-out data-fragment-index="0"}
```{r}
bone.plot 
```
:::
::: {.fragment .fade-in data-fragment-index="0"}
```{r}
not.parallel.lines.model.fit = lm(bone_density ~ factor(supplement) + factor(age), data=bone.data)
plot.predictions(prediction.function(not.parallel.lines.model.fit))
```
:::
::::



Sketch in the least squares predictor in the **additive model**.
$$
\mathcal{M} = \qty{ m(w,x) = a(w) + b(x) \ \ \text{for univariate functions} \ \ a,b } 
$$

Flip a slide forward to check your answer. Is what you've drawn familiar? If so, why?

::: {.fragment .fade-in data-fragment-index="1"} 

 - It's the same as the parallel lines model. 
   - When $x$ takes on two values, all functions of $x$ can be written as lines.
   - Adding a function of $w$ to it, we get different---but parallel---lines for different values of $w$.

:::






