---
title: "Lecture 12"
subtitle: "The Behavior of Least Squares Predictors"
---

# Introduction 
This is a revised version of Lecture 12 as covered in class. It's a work in progress.
Parts may be useful as supplemental reading but it isn't yet entirely coherent or complete enough to be a replacement for the version from class.

## Today's Topic

{{< include ../_extensions/r-wasm/live/_knitr.qmd >}}
{{< include shared-code.qmd >}}


```{r}
#| label: fake-data
x = rep(c(20,30,40), each=5000)

pi.pop = function(x) { case_when(x < 25 ~ .25, x < 35 ~ .5, TRUE ~ .75) } 
mu.pop = function(w,x) case_when(
  w==1 & x <  25  ~ 10, 
  w==1 & x <= 35  ~ 10,
  w==1            ~ 20, 
  w==0 & x <  25  ~ 5, 
  w==0 & x <= 35  ~ 10,
  w==0            ~ 10)

set.seed(2)
w = 0 + (runif(length(x)) < pi.pop(x))
y = rpois(length(x), mu.pop(w,x))
fake.pop = data.frame(w=w, x=x, y=y)
```

```{r}
pop = fake.pop 
m = nrow(pop)
n = 200
set.seed(5)
J = sample(1:m, size=n, replace=TRUE)
sam = pop[J,]
```

```{r}
prop.scale = 40
default.shift = .5
plotX = function(w,x, shift=default.shift) { ifelse(w==0, x-shift, x+shift) }
lsq.plot = function(sam, formula=NULL, dot.size=2, shift=default.shift, residual.alpha=0) {
  sam$jitterX = runif(nrow(sam), -1, 1)*shift
  base.plot = ggplot(sam) + 
    geom_point(aes(x=plotX(w,x)+jitterX,y=y,color=factor(w)), 
      alpha=.2, size=dot.size) + 
    stat_summary(aes(x=plotX(w,x), y=y), 
      fun.data=mean_sd, geom='pointrange', alpha=.2) +
    geom_bar(aes(x=x, y=prop.scale*after_stat(prop)), 
      fill='purple', alpha=.1, linewidth=.05) +
    scale_y_continuous(sec.axis = sec_axis(~./prop.scale, name='proportion')) +
    scale_x_continuous(breaks=c(20,30,40)) +
    coord_cartesian(xlim=c(15,45), ylim=c(-1,40)) +
    guides(color='none') 
    
    if(is.null(formula)) { base.plot } else { 
        model = lm(formula, data=sam)
        gridx = seq(-10, 50, length.out=1000)
        muhat = function(w,x) predict(model, newdata=data.frame(w=w,x=x))   
        the.plot = base.plot + 
          geom_point(aes(x=plotX(w,x), y=muhat(w,x), color=factor(w)), 
            alpha=.7, shape='square') +
          geom_line(aes(x=plotX(w,x), y=muhat(w,x), color=factor(w)), 
            data = expand_grid(w=0:1, x=gridx), alpha=.7, linetype='dotted') 
        if(residual.alpha > 0) {
          the.plot = the.plot + 
            geom_segment(aes(x=plotX(w,x)+jitterX, xend=plotX(w,x), y=y, yend=muhat(w,x)), 
              alpha=residual.alpha, color='orange', linewidth=.1)
        }
        the.plot 
    }
}
```

```{r}
lsq.plot.1 = function(sam, formula, dot.size=1, color=groupa, residual.alpha=0) { 
  sam$jitterX = runif(nrow(sam), -1, 1)
  base.plot = ggplot(sam) + 
    geom_point(aes(x=x+jitterX,y=y, color=color), 
      alpha=.05, size=dot.size) + 
    stat_summary(aes(x=x+jitterX, y=y), 
      fun.data=mean_sd, geom='pointrange', alpha=.2) +
    geom_bar(aes(x=x, y=prop.scale*after_stat(prop)), 
      fill='purple', alpha=.1, linewidth=.05) +
    scale_y_continuous(sec.axis = sec_axis(~./prop.scale, name='proportion')) +
    guides(color='none') 
    
    if(is.null(formula)) { base.plot } else { 
        model = lm(formula, data=sam)
        muhat = function(x) predict(model, newdata=data.frame(x=x))   
        the.plot = base.plot + 
          geom_point(aes(x=x, y=muhat(x)), color=color, alpha=.7, shape=muhat.shape) +
          geom_line(aes(x=x, y=muhat(x)),  color=color, alpha=.7, shape=muhat.linetype) 
        if(residual.alpha > 0) {
          the.plot = the.plot + 
            geom_segment(aes(x=x+jitterX, xend=x+jitterX, y=y, yend=muhat(x)), 
              alpha=residual.alpha, color='orange', linewidth=.05)
        }
        the.plot 
    }
}
```


```{r}
#| cache: true
estimate = function(sam, formula) {
  X = sam$x
  W = sam$w
  model = lm(formula, data=sam)
  muhat = function(w,x) predict(model, newdata=data.frame(w=w,x=x)) 
  mean(muhat(1, X) - muhat(0, X))
}

samples = 1:4000 |> map(function(.) { 
  J = sample(1:nrow(pop), size=n, replace=TRUE)
  pop[J,]
})

target  = pop |> estimate(y~w*factor(x))
estimates = data.frame( 
  all.functions = samples |> map_vec(estimate, y~w*factor(x)),
  lines = samples |> map_vec(estimate, y~w*x))
```

```{r}
#| fig.width: 7
#| layout-ncol: 2
#| fig-cap: 
#|  - 'A sample drawn with replacement from a population.'
#|  - 'That population.'
sam |> lsq.plot(y~w*x)
pop |> lsq.plot()
```

We're going to talk about statistical inference when we use least squares in linear models. For example ...
$$
\color{gray}
\begin{aligned}
\theta &= \frac{1}{m}\sum_{j=1}^m \qty{ \mu(1,x_j) -  \mu(0,x_j) } \qfor \mu(w,x)=\frac{1}{m_{w,x}}\sum_{j:w_j=w,x_j=x} y_j \\
\hat\theta &= \frac{1}{n}\sum_{i=1}^n \qty{ \hat\mu(1,X_i) -  \hat\mu(0,X_i) } \qfor \hat\mu = \argmin_{m \in \model} \sum_{i=1}^n \qty{ Y_i - m(X_i) }^2
\end{aligned}
$$

- Does it look like this is a good estimator?
- If not, do you think it's too big or too small?

## Summary

```{r}
#| fig.width: 7
#| layout-ncol: 2
#| fig-cap: 
#|  - 'The sampling distribution of $\hat\theta$ and a perfectly calibrated interval estimate.'
#|  - 'The population least squares predictor $\tilde\mu$ fit to the population.'
#| 
theta.hat = estimate(sam, y~w*x)
theta.tilde = estimate(pop, y~w*x)
theta.sampling.dist.plot =ggplot(estimates) + 
  geom_histogram(aes(x=all.functions, y=after_stat(density)), bins=50, alpha=.05, fill=pink, color='gray', linewidth=.05) + 
  geom_histogram(aes(x=lines, y=after_stat(density)), bins=50, alpha=.2, fill=teal, color='gray', linewidth=.05) + 
  with_annotations(c(theta.hat,estimates$lines), estimand=target, highlight.dots=1, interval.width=width(estimates$lines)) + 
  geom_vline(xintercept=theta.tilde, color=pink, alpha=.3, linewidth=2.5)

plot.pop = pop[sample(1:nrow(pop), size=10000),]
pop.lsq.plot = plot.pop |> lsq.plot(y~w*x)

theta.sampling.dist.plot
pop.lsq.plot
```

::: {.panel-tabset}
### What happened?
- In this case, $\hat\theta$ is too small. And that's not just a chance occurrence. 
  - When we look at its sampling distribution, we see that it's *systematically* too small.
  - It's *biased*. In roughly 95% of samples, $\hat\theta$ will be smaller than $\theta$.
- It boils down to a bad choice of regression model. 
  - To make sense of this, think about what happens when everything else is perfect.
  - i.e. if we fit this model to the whole population, then average over the population.
  - The result is *nonrandom*. And it's right in the middle of our estimator's sampling distribution.
$$
\inpink{\tilde\theta} = \frac{1}{m}\sum_{j=1}^m \qty{ \tilde\mu(1,x_j) -  \tilde\mu(0,x_j) } \qfor \tilde\mu = \argmin_{m \in \model} \frac{1}{m}\sum_{j=1}^m \qty{y_j - m(w_j,x_j)}^2 
$$

### What happens generally 

We can think of our estimator's error as a sum of two things.

$$
\hat\theta - \theta = 
\underset{\text{sampling error}}{\qty{\hat\theta - \inpink{\tilde\theta}}} + 
\underset{\text{modeling error}}{\qty{\inpink{\tilde\theta} - \theta}}
$$

1. *Sampling error* is the 'variance part'. It's random and has mean zero. 
  - It's the error we'd have had if what we'd really wanted to estimate was $\tilde\theta$.
  - We can get a good estimate of this term's distribution using the bootstrap. 
2. *Modeling error* is the 'bias part'. It's non-random---it's the bias of our estimator. 
  - It's the error we'd have had if we'd been given the whole population as our sample.
  - This is zero *if* your model contains the subpopulation mean function $\mu(w,x)$.
    - e.g. when you use the 'all functions' model.

  - That we have a nice, clean breakdown of our error like this isn't obvious.
    - You can, of course, add and subtract whatever you want. We chose $\inpink{\tilde\theta}$.
    - But it's not obvious that the sampling error term $\hat\theta - \inpink{\tilde\theta}$ has mean zero.

### Terminology
  
  - We say a model is *correctly specified* when it contains $\mu$ 
  - We say it is *misspecified* when it doesn't contain $\mu$.

:::
  
## Plan 

$$
\hat\theta - \theta = 
\underset{\text{sampling error}}{\qty{\hat\theta - \inpink{\tilde\theta}}} + 
\underset{\text{modeling error}}{\qty{\inpink{\tilde\theta} - \theta}}
$$

- Today, we'll do two things. 
 1. We'll prove that the sampling error term has mean zero. 
    - In fact, we'll prove that it's an average of mean-zero random variables.
    - That'll take a little calculus. It's not a hard proof, but it's a little clever.
 2. Then, we'll try to get a handle on what modeling error looks like for different models. 
  - This'll be a visualization exercise to help you develop some intuition.
  - This intuition comes in very handy because a lot of people don't have it.^[Blame us teachers. In a lot of classes, you do the math assuming you're using a correctly specified model. Then you learn some techniques for choosing a model that isn't obviously misspecified. The message people get is to put a little effort into choosing a model, then act as if it were correctly specified. That doesn't work.]
  - With even pretty basic intuition, you can catch a lot of mistakes and get them fixed.

# Orthogonality of Residuals
*The* tool for understanding least squares.

## Review: Least Squares and the Sample Mean
```{r}
#| layout-ncol: 2
#| fig-width: 7 
#| fig-cap:  
#|  - 'Some data with horizontal lines at the mean (red), median (orange), and mode (blue).' 
#|  - 'The sum of squared deviations from $m$ (on the x-axis) with the mean (red), median (orange), and mode (blue) indicated by dots.' 
Y = sam$y
X = 1:length(Y)
signsq = function(x) { sign(x)*x^2 / 40000 }
themode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}
ggplot() + 
  geom_hline(yintercept=mean(Y), alpha=.4, linewidth=1, color='red') + 
  geom_hline(yintercept=median(Y), alpha=.4, linewidth=1, color='orange') + 
  geom_hline(yintercept=themode(Y), alpha=.4, linewidth=1, color='blue') + 
  geom_point(aes(x=X,y=Y), size=2, alpha=.2) +  
  geom_segment(aes(x=X, xend=X, y=mean(Y), yend=mean(Y)+signsq(Y-mean(Y))), alpha=.2, color='red', linewidth=.5) + 
  geom_segment(aes(x=X+.1, xend=X+.1, y=median(Y), yend=median(Y)+signsq(Y-median(Y))), alpha=.2, color='orange', linewidth=.5) +
  geom_segment(aes(x=X-.1, xend=X-.1, y=themode(Y), yend=themode(Y)+signsq(Y-themode(Y))), alpha=.2, color='blue', linewidth=.5)

ssr = Vectorize(\(m) sum((Y-m))^2)
x = seq(mean(Y)-2*sd(Y), mean(Y)+2*sd(Y), length.out=1000)
ggplot() + geom_line(aes(x=x, y=ssr(x)))  +
  geom_point(aes(x=mean(Y), y=ssr(mean(Y))), color='red', size=4) + 
  geom_point(aes(x=median(Y), y=ssr(median(Y))), color='orange', size=4) +
  geom_point(aes(x=themode(Y), y=ssr(themode(Y))), color='blue', size=4)  
```

$$
\hat \mu = \argmin_{\text{numbers}\ m} SSR(m) \qfor SSR(m) =\sum_{i=1}^n \qty{ Y_i - m }^2 \quad \text{ satisfies the \textbf{zero-derivative condition} }
$$
 
$$
\begin{aligned}
0 
&= \frac{d}{dm} \sum_{i=1}^n \qty{ Y_i - m }^2  \ \ \mid_{m=\hat\mu}  
&&= \sum_{i=1}^n \frac{d}{dm} \qty{ Y_i - m }^2  \ \ \mid_{m=\hat\mu}  \\ 
&= \sum_{i=1}^n -2 \qty{ Y_i - m } \ \ \mid_{m=\hat\mu} 
&&= \sum_{i=1}^n -2 \qty{ Y_i - \hat\mu } 
\end{aligned}
$$

- This says the least squares residuals [$\hat\varepsilon = Y_i - \hat \mu$]{.fg style="color:red"} *sum to zero*. 
- With a bit of algebra, this tells us that the least squares estimator $\hat\mu$ is the sample mean.

$$
\begin{aligned}
&0 = \sum_{i=1}^n -2\qty{ Y_i - \hat\mu } && \text{ when } \\ 
&2\sum_{i=1}^n Y_i = 2\sum_{i=1}^n \hat\mu = 2n\hat\mu && \text{ and therefore when } \\
&\frac{1}{n}\sum_{i=1}^n Y_i = \hat\mu
\end{aligned}
$$

## Least Squares and Subsample Means

```{r}
#| layout-ncol: 2
#| fig-width: 7 
#| fig-cap:  
#|  - 'Some data with horizontal lines at the means of the two groups.' 
#|  - 'A contour plot of the sum of squared residuals.'

Y = sam$y
X = 1:length(Y)
W = sam$w

signsq = function(x) { sign(x)*x^2 / 40000 }
themode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}
subsample.means.plot = ggplot() + 
  geom_hline(aes(yintercept=mean(Y[W==0])), alpha=.4, linewidth=1, color=groupa) +
  geom_hline(aes(yintercept=mean(Y[W==1])), alpha=.4, linewidth=1, color=groupb) +
  geom_hline(aes(yintercept=mean(Y[W==0])), alpha=.4, linetype='dotted', color='red') +
  geom_hline(aes(yintercept=mean(Y[W==1])), alpha=.4, linetype='dotted', color='red') +
  geom_hline(aes(yintercept=median(Y[W==0])), alpha=.4, linewidth=1, color=groupa) +
  geom_hline(aes(yintercept=median(Y[W==1])), alpha=.4, linewidth=1, color=groupb) +
  geom_hline(aes(yintercept=median(Y[W==0])), alpha=.4, linetype='dashed', color='orange') +
  geom_hline(aes(yintercept=median(Y[W==1])), alpha=.4, linetype='dashed', color='orange') +
  geom_hline(aes(yintercept=themode(Y[W==0])), alpha=.4, linewidth=1, color=groupa) +
  geom_hline(aes(yintercept=themode(Y[W==1])), alpha=.4, linewidth=1, color=groupb) +
  geom_hline(aes(yintercept=themode(Y[W==0])), alpha=.4, linetype='dotted', color='blue') +
  geom_hline(aes(yintercept=themode(Y[W==1])), alpha=.4, linetype='dotted', color='blue') +
  geom_point(aes(x=X,y=Y,color=factor(W)), size=2, alpha=.2) + guides(color='none')

xr = seq(mean(Y)-1.5*sd(Y), mean(Y)+1.5*sd(Y), length.out=100)
ssr = Vectorize(\(m0,m1) sum((Y-m0*(W==0)-m1*(W==1))^2))
grid = expand.grid(m0=xr, m1=xr)
grid$y = ssr(grid$m0, grid$m1)

two.groups.ssr.plot=ggplot() + geom_contour_filled(aes(x=m0, y=m1, z=y), 
          data=grid, breaks=6000 + c(0,100*2^(0:8))) + 
           geom_point(aes(x=mean(Y[W==0]), y=mean(Y[W==1])), color='red', size=4) +
           geom_point(aes(x=median(Y[W==0]), y=median(Y[W==1])), color='orange', size=4) +
           geom_point(aes(x=themode(Y[W==0]), y=themode(Y[W==1])), color='blue', size=4) + 
           coord_cartesian(xlim=range(xr), ylim=range(xr)) 

subsample.means.plot 
two.groups.ssr.plot + class.theme + labs(x='m(0)', y='m(1)', fill='SSR(m)')
```

$$
\hat \mu = \argmin_{\text{functions}\ m(w)} SSR(m) \qfor SSR(m) =\sum_{i=1}^n \qty{ Y_i - m(W_i) }^2
$$

- How do we write a zero-derivative condition for this problem?
  - i.e., how do we talk about the derivative with respect to a function?
  - Like many things in multivariable calculus, we think one 'direction' at a time.
- Let's start with something simpler. How do we visualize this problem? 
- We can think of each function $m$ as a location in a two-dimensional space. 
  - One dimension is the value of $m(0)$, the other is the value of $m(1)$. 
- And we can use a *contour plot* to visualize the sum of squared residuals at each location. 
  - If you hike or run, you probably see these plots all the time. They're used to show *elevation*.
  - So you can think of the function we're minimizing, $SSR$, as the elevation of some landscape. 
- The point we're looking for, $\hat\mu$, is the lowest point in that landscape.
  - I've drawn in dots showing the height at the within-group [means]{.fg style="color:red"}, [medians]{.fg style="color:orange"}, and [modes]{.fg style="color:blue"}. 
  - The means are the lowest point. Let's prove it.

## Least Squares and Subsample Means

```{r}
slope = (median(Y[W==1])-mean(Y[W==1]))/(median(Y[W==0])-mean(Y[W==0]))
intercept = mean(Y[W==1]) - slope*mean(Y[W==0])
median.path = geom_abline(aes(slope=slope, intercept=intercept), color='orange', linewidth=.5)
contour.labs = list(class.theme, labs(x='m(0)', y='m(1)', fill='SSR(m)'))
two.groups.ssr.plot + median.path + contour.labs
```

::: {.panel-tabset}
### Paths
$$
\hat \mu = \argmin_{\text{functions}\ m(w)} SSR(m) \qfor SSR(m) =\sum_{i=1}^n \qty{ Y_i - m(W_i) }^2
$$

- To write a zero-derivative condition, let's keep our landscape metaphor going. 
  - Let's think about what happens if you start at the lowest point --- the means --- and hike uphill.
  - Each point on the *path*^[This isn't just my own metaphorical language. These are really called *paths* in statistical theory.] you take is a location in our landscape: a function $m$ in our model.
- We could, for example, hike straight toward the groupwise medians. I've drawn that path in orange.
  - At 'time' $t=0$---when we start---we're at [the means.]{.fg style="color:red"}
  - At 'time' $t=1$---when we finish---we're at the [medians.]{.fg style="color:orange"}^[There's no need to restrict $t$ to $[0,1]$. For $t>1$, we keep going past the medians. For $t<0$, we're walking in the opposite direction. The whole path is shown as an orange line on the contour plot.]
$$
m_t = \hat \mu + t (m - \hat \mu) \qqtext{ is 'where we are' at time } t.
$$

- When we take this path, what's our elevation --- the height of the $SSR$ landscape ---at time $t$?

### The Lowest Point

:::: columns 
::: column 
$$
\begin{aligned}
SSR(m_t) 
&= \sum_{i=1}^n \qty{ Y_i - m_t(W_i) }^2 \\
&= \sum_{i=1}^n \qty{ Y_i - \hat\mu(W_i) - t (m(W_i) - \hat\mu(W_i)) }^2
\end{aligned}
$$

- This is a *univariate function*---a function of $t$. 
- It's minimized at $t=0$ because we're starting at \
the lowest point: the least squares predictor $\hat\mu$. 
- That tells us that its derivative is zero at $t=0$. 

$$
\begin{aligned}
0 = \frac{d}{dt}\mid_{t=0} SSR(m_t) 
\end{aligned}
$$

:::
::: column 
```{r} 
#| fig.width: 7
x = seq(-1, 1,length.out=1000)
f = function(x) ssr(mean(Y[W==0])*(1-x) + median(Y[W==0])*x, 
                    mean(Y[W==1])*(1-x) + median(Y[W==1])*x)
ggplot() + geom_line(aes(x=x, y=f(x)), color='orange')  +
  geom_point(aes(x=0,y=f(0)), color='red',   size=4) + 
  geom_point(aes(x=1,y=f(1)), color='orange', size=4) + class.theme +
  labs(x='t', y='SSR(m_t)')
```
:::
::::

::: 

## Our Zero-Derivative Condition

```{r}
two.groups.ssr.plot + 
  geom_hline(aes(yintercept=mean(Y[W==1])), alpha=.4, linewidth=1, color=groupa) +
  geom_vline(aes(xintercept=mean(Y[W==0])), alpha=.4, linewidth=1, color=groupb) + 
  geom_abline(aes(slope=1, intercept=mean(Y[W==1]) - mean(Y[W==0])), color='orange', linewidth=.5) +
  class.theme + labs(x='m(0)', y='m(1)', fill='SSR(m)')
```

- To make the math a bit easier, we'll define our paths slightly differently. 
- We'll think of $m$ as a *direction* instead of a destination. 

$$
m_t = \hat\mu + t m \qqtext{ is 'where we are' at time } t \color{lightgray} \qqtext{ instead of } m_t = \hat\mu + t (m - \hat\mu).
$$

- No matter what direction $m \in \model$ we look at, our whole path $\hat\mu + tm$ is in the model.
  - Why? Because our model is linear, $\hat\mu + tm \in \model$ whenever $\hat\mu \in \model$ and $m \in \model$.
  - And because $\hat\mu$ is the lowest point in our whole landscape (i.e. our model $\model$), it has to be the lowest point in our path.

- Now let's see what our zero-derivative for $SSR(m_t)$ tells us. 

$$
\begin{aligned}
0 = \frac{d}{dt}SSR(m_t) 
  &= \frac{d}{dt}\mid_{t=0} \sum_{i=1}^n \qty{ Y_i - \hat\mu(W_i) - tm(W_i) }^2   \\
  &= \sum_{i=1}^n 2 \times \qty{ Y_i - \hat\mu(W_i) } \times \frac{d}{dt}\mid_{t=0} \ \qty{ Y_i - \hat\mu(W_i) - tm(W_i) } \\
  &= \sum_{i=1}^n -2 \times \qty{ Y_i - \hat\mu(W_i) } \times m(W_i) \\
\end{aligned}
$$ 

- Below, I've drawn this function for three different directions $m \in \model$.

$$
\underset{\text{left}}{\textcolor{red}{m(w)}} = 1_{=0}(w) = \begin{cases} 1 & \text{ if } w=0 \\ 0 & \text{ if } w=1 \end{cases} \qquad 
\underset{\text{middle}}{\textcolor{magenta}{m(w)}} = 1_{=1}(w) = \begin{cases} 0 & \text{ if } w=0 \\ 1 & \text{ if } w=1 \end{cases} \qquad 
\underset{\text{right}}{\textcolor{orange}{m(w)}} = 1
$$

```{r}
#| fig.width: 3.5
#| layout-ncol: 3
x = seq(-2*sd(Y), 2*sd(Y), length.out=1000)
x0 = mean(Y[W==1])
f = function(x) ssr(mean(Y[W==0]), x0+x)
ggplot() + geom_line(aes(x=x, y=f(x)), color='red')  +
  geom_point(aes(x=0,y=f(0)), color='red', size=4)  

x0 = mean(Y[W==0])
g = function(x) ssr(x0+x, mean(Y[W==1]))
ggplot() + geom_line(aes(x=x, y=g(x)), color='magenta') +
  geom_point(aes(x=0,    y=g(0)),   color='red', size=4)  

h = function(x) ssr(mean(Y[W==0])+x, mean(Y[W==1])+x)
ggplot() + geom_line(aes(x=x, y=h(x)), color='orange') + 
  geom_point(aes(x=0, y=h(0)), color='red', size=4) 
```

- By working with these, we can show some familiar properties of the residuals. 
  - We know that $\hat\mu$ is the minimizer of $SSR(\hat\mu + tm)$ for any $m \in \model$ ...
  - so the derivative of $SSR(\hat\mu + tm)$ with respect to $t$ must be zero at $t=0$.

$$
\begin{aligned}
0 &= \frac{d}{dt}\mid_{t=0} \ \sum_{i=1}^n \qty{ Y_i - \hat\mu(W_i) + tm(W_i) }^2   \\
  &= \sum_{i=1}^n 2 \times \qty{ Y_i - \hat\mu(W_i) } \times \frac{d}{dt}\mid_{t=0} \ \qty{ Y_i - \hat\mu(W_i) - tm(W_i) } \\
  &= \sum_{i=1}^n -2 \times \qty{ Y_i - \hat\mu(W_i) } \times m(W_i) \\
\end{aligned}
$$ 

- For the case $m(w) = 1_{=1}(w)$, this says the residuals sum to zero within group $w=1$.
- For the case $m(w) = 1_{=0}(w)$, this says the residuals sum to zero within group $w=0$.
- For the case $m(w) = 1$, this says the residuals sum to zero outright.

## Orthogonality of Residuals in General

- There's nothing specific to this model in the math we just did. It works for any linear model.
- No matter which linear model we use, we can always think one direction at a time. 

$$
\begin{aligned}
\hat\mu &\qqtext{minimizes} \sum_{i=1}^n \qty{ Y_i - m(W_i,X_i) }^2 &&\qqtext{ over } m \in \model \implies \\
\hat\mu &\qqtext{minimizes} \sum_{i=1}^n \qty{ Y_i - m(W_i,X_i) }^2 &&\qqtext{ over } m \in \{ m_t = \hat\mu + t m \ : \ t \in \mathbb{R} \} \qqtext{ for any } m \in \model
\end{aligned}
$$

- Writing mean squared error as a function of $t$ in this model ... 
  - We can observe that it's minimized at $t=0$, since that's when $\hat\mu + tm=\hat\mu$
  - And we can set the derivative to zero there to get the orthogonality condition we want.
  
$$
\begin{aligned}
0 &= \frac{d}{dt}\mid_{t=0} \sum_{i=1}^n \qty{ Y_i - \hat\mu(W_i,X_i) - t m(W_i,X_i) }^2 \\
  &= \sum_{i=1}^n 2 \times \qty{ Y_i - \hat\mu(W_i,X_i) } \frac{d}{dt}\mid_{t=0} \qty{ Y_i - \hat\mu(W_i,X_i) - t m(W_i,X_i) }^2 \\
  &= \sum_{i=1}^n 2 \times \qty{ Y_i - \hat\mu(W_i,X_i) } \times -m(W_i,X_i) \\
\end{aligned}
$$

# The Expected Value of the Least Squares Predictor $\hat\mu(w,x)$

## The Population Least Squares Predictor

```{r}
#| fig.width: 7
#| layout-ncol: 2
#| fig-cap: 
#|  - 'A sample drawn with replacement from a population and the least squares predictor.'
#|  - 'That population and the population least squares predictor.'
sam |> lsq.plot(y~w*x)
plot.pop |> lsq.plot(y~w+x)
```


::: {.panel-tabset}
### Population Least Squares 

- To make it easier to think about what happens when we do least squares, let's do a thought experiment. 
  - Let's take all the randomness out of the picture. 
  - Let's think about what would happen if we chose the function in our model $\model$ that minimizes the sum of squared error *over the whole population*.
- We call this the *population least squares predictor*.

$$
\begin{aligned}
\tilde\mu 
&= \argmin_{m \in \model} \sum_{j=1}^m \qty{ y_j - m(w_j,x_j) }^2 \\
&\explain{=}{When we sample uniformly-at-random from the population, this expectation is $1/m$ times the sum above.}
\argmin_{m \in \model} \E\qty[ \qty{ Y_i - m(W_i,X_i) }^2 ]   
\end{aligned}
$$

- It's reasonable to expect it's a good approximation to the least squares predictor we get from a large sample.
  - After all, the law of large numbers tells us that sample averages converge to a population averages. 
  - Including the sample average of squared errors.^[We are talking about minimizing the sum and average of squared errors interchangeably. Why does that difference not matter? *Hint*. Why is the value of $x$ that minimizes $f(x)=x^2$ the same as the value of $x$ that minimizes $10x^2$ and $x^2/10$?]
$$
\frac{1}{n}\sum_{i=1}^n \qty{ Y_i - \hat\mu(W_i,X_i) }^2 \to  \E\qty[ \qty{ Y_i - \hat\mu(W_i,X_i) }^2 ] \qqtext{ as } n \to \infty.
$$


### 'Unbiasedness' of the Least Squares Predictor

- It turns out that we can say even more than this. It's also the *expectation* of the least squares predictor.
  - In other words, if the population least squares prediction $\tilde\mu(w,x)$ were our estimation target...
  - ... then the least squares predictor we get from a sample would be an *unbiased* estimator for it.

$$
\E[\hat\mu(w,x)] = \tilde\mu(w,x)
$$


- To prove it, we'll think about the *population least squares residuals* $\tilde\varepsilon_j = y_j - \tilde\mu(w_j,x_j)$.
- These satisfy an orthogonality property analogous to the one we saw for sample residuals.
$$
\begin{aligned}
0 &= \veryfaint{\frac{1}{n}}\sum_{i=1}^n \overset{\textcolor{lightgray}{\text{sample residuals}}}{\lcb{ Y_i - \hat\mu(W_i,X_i) }} \ m(W_i,X_i)  
&& \qqtext{ for all } m \in \model \\
0 &= \veryfaint{\frac{1}{m}}\sum_{j=1}^m \overset{\textcolor{lightgray}{\text{population residuals}}}{\lcb{ y_j - \tilde\mu(w_j,x_j) }} \ m(w_j,x_j) 
&& \qqtext{ for all } m \in \model \\
&= \E\qty[ \qty{ Y_i - \tilde\mu(W_i,X_i) } \ m(W_i,X_i) ] 
\end{aligned}
$$

- To prove 'unbiasedness', we'll combine these two orthogonality properties in a clever way.

:::

## The Claim and Proof Sketch 

**Claim**. 
$$
\begin{aligned}
\tilde\mu(w,x) &= \mu_E(w,x) 
&& \qfor \mu_E(w,x) = \E[\hat\mu(W_i,X_i) \mid W_i=w, X_i=x] \\
&&& \qand \tilde\mu = \argmin_{m \in \model} \sum_{j=1}^m \qty{ y_j - m(w_j,x_j) }^2
\end{aligned}
$$

**Proof Sketch**.
1. We'll show that orthogonality of residuals implies a kind of orthogonality of the difference $\hat\mu - \tilde\mu$.
$$
0=\E\left[ \lcb{ \hat\mu(W_i,X_i) - \tilde\mu(W_i,X_i) } \ m(W_i,X_i) \right] \qqtext{ for all } m \in \model
$$

2. We'll plug in a clever choice of $m \in \model$: $m=\mu_E - \tilde \mu$ and show, using the law of iterated expectations, that this tells us the mean-squared-difference between $\mu_E$ and $\tilde\mu$ is zero.^[That implies $\mu_E(W_i,X_i) - \tilde\mu(W_i, X_i)=0$. That's how squares work. If something isn't zero, its mean-square is positive.]
$$
\begin{aligned}
0 &= \E\left[ \lcb{ \hat\mu(W_i,X_i) - \tilde\mu(W_i,X_i) } \ \qty{ \mu_E(W_i,X_i) - \tilde\mu(W_i, X_i) } \right] \\
  &= \E\left[ \lcb{ \mu_E(W_i,X_i) - \tilde\mu(W_i,X_i) }^2 \right]
\end{aligned}
$$


## Proof: Part 1

- From the orthogonality of the sample residuals, all $m \in\model$ satisfy ...

$$ 
\begin{aligned}
0 &= \frac{1}{n}\sum_{i=1}^n \lcb{ Y_i -\hat\mu(W_i,X_i) } \ m(W_i,X_i) \qqtext{ and therefore } \\
0 &= \E \left[ \frac{1}{n}\sum_{i=1}^n \lcb{ Y_i -\hat\mu(W_i,X_i) } \ m(W_i,X_i) \right] \\ 
  &=  \E\left[ \lcb{ Y_i - \hat\mu(W_i,X_i) } \ m(W_i,X_i) \right]. 
\end{aligned}
$$

- From the orthogonality of the population residuals, all $m \in\model$ satisfy ...

$$
\begin{aligned}
0 &= \E\left[ \lcb{ Y_i - \tilde \mu(W_i,X_i)} \ m(W_i,X_i) \right] \\
\end{aligned}
$$

- Subtracting the first from the second, we find that all $m \in \model$ satisfy ...

$$
\begin{aligned}
0 &= \E\left[ \tcb{ \lcb{ Y_i- \tilde\mu(W_i,X_i) } - \lcb{ Y_i - \hat\mu(W_i,X_i) }} \ m(W_i,X_i) \right] \\
  &= \E\left[ \tcb{ \hat\mu(W_i,X_i) - \tilde\mu(W_i,X_i) } \ m(W_i,X_i) \right]
\end{aligned}
$$

## Proof: Part 2

- Because $\mu_E$ and $\tilde\mu$ are in the model^[See @sec-check-in-model], so is $m=\mu_E - \tilde\mu$. And therefore ...
$$
\begin{aligned}
0 &= \E\left[ \lcb{ \hat\mu(W_i,X_i) - \tilde\mu(W_i,X_i) } \ m(W_i,X_i) \right] \qqtext{ for all } m \in \model \implies \\
0 &= \E\left[ \lcb{ \hat\mu(W_i,X_i) - \tilde\mu(W_i,X_i) } \ \lcb{ \mu_E(W_i,X_i) - \tilde\mu(W_i,X_i) } \right] \qfor \mu_E(w,x) = \E[\hat\mu(W_i,X_i) \mid W_i=w, X_i=x]
\end{aligned}
$$

- We'll use the law of iterated expectations to simplify this expectation.

$$
\begin{aligned}
0 
&\explain{=}{Step 1 + Substitution} 
\E\left[ \lcb{ \hat\mu(W_i,X_i) - \tilde\mu(W_i,X_i) } \lcb{ \mu_E(W_i,X_i) - \tilde\mu(W_i,X_i) } \right] \\
&\explain{=}{Law of Iterated Expectations} 
  \E\left[ \color{blue}\E\left[ \color{black}  \left\{ \hat\mu(W_i,X_i) - \tilde\mu(W_i,X_i) \right\} \left\{ \mu_E(W_i,X_i) - \tilde\mu(W_i,X_i) \right\} \color{blue} \mid W_i, X_i\right] \color{black} \right] \\
&\explain{=}{Linearity of Conditional Expectations} 
  \E\left[  \left\{ \color{blue}\E\left[ \color{black} \hat\mu(W_i,X_i) \color{blue} \mid W_i, X_i\right] \color{black} - \tilde\mu(W_i,X_i) \right\} \left\{ \mu_E(W_i,X_i) - \tilde\mu(W_i,X_i) \right\}  \right] \\
&\explain{=}{Recognizing $\mu_E$'s definition} 
\E\qty[ \qty{ \color{blue} \mu_E(W_i,X_i) \color{black} - \tilde\mu(W_i,X_i) }\qty{ \mu_E(W_i,X_i)  - \tilde\mu(W_i,X_i) } ] \\
&\explain{=}{Recognizing a square} 
\E\left[ \left\{ \mu_E(W_i,X_i) - \tilde\mu(W_i,X_i) \right\}^2 \right] 
\end{aligned}
$$

- That concludes the proof. 
  - If $\mu_E(W_i,X_i) - \tilde\mu(W_i,X_i)$ weren't always zero ...
  - then the expected value of its square would be positive, not zero.

# Implications for Complex Targets

## Breaking Down the Problem

```{r}
#| fig.width: 7
#| layout-ncol: 2
#| fig-cap: 
#|  - 'A sample drawn with replacement from a population.'
#|  - 'That population.'
sam |> lsq.plot(y~w*x)
pop.lsq.plot
```


- The summaries we've been estimating are complex, but made up of simple pieces.
- In particular, they're all linear combinations of subpopulation means $\mu(w,x)$. 

$$
\theta = \frac{1}{m}\sum_{j=1}^m \qty{ \mu(1,x_j) -  \mu(0,x_j) } \qfor \mu(w,x)=\frac{1}{m_{w,x}}\sum_{j:w_j=w,x_j=x} y_j
$$

- Let's start by looking at the error we get when we estimate a single one of these. 
- We'll break it down into two pieces. 
  - One piece is the error of a *population least squares predictor* $\inpink{\tilde\mu}$.
  - The other is the difference between that and our actual predictor $\hat\mu$. 

$$
\hat\mu(w,x) - \mu(w,x) = \qty{\hat\mu(w,x) - \inpink{\tilde\mu(w,x)}} + \qty{\inpink{\tilde\mu(w,x)} - \mu(w,x)}
$$

## Estimating Subpopulation Means 

```{r}
#| cache: true
grid = expand.grid(w=0:1, x=c(20,30,40))
mutilde=function(w,x) { predict(lm(y~w*x, data=pop), newdata=data.frame(w=w,x=x)) }
mu=function(w,x) { predict(lm(y~w*factor(x), data=pop), newdata=data.frame(w=w,x=x)) }
mu.hat.grid = samples |> map(function(sam) { 
  newgrid=grid 
  newgrid$muhat=predict(lm(y~w*x, data=sam), newdata=grid)
  newgrid 
}) |> bind_rows()
```
```{r}
#| layout-ncol: 3
#| layout-nrow: 2
#| fig-width: 3.5

plothist = function(w,x, color=color) { 
  ggplot() + 
    geom_histogram(aes(x=muhat, y=after_stat(density)), 
      alpha=.2, bins=50, fill=color, data=mu.hat.grid[mu.hat.grid$w==w & mu.hat.grid$x==x,]) +
    geom_vline(aes(xintercept=mutilde(w,x)), alpha=.2, linewidth=2, color=pink) +
    geom_vline(aes(xintercept=mu(w,x)), alpha=.2, linewidth=2, color='green')
}

plothist(0,20, color=groupa)
plothist(0,30, color=groupa)
plothist(0,40, color=groupa)
plothist(1,20, color=groupb)
plothist(1,30, color=groupb)
plothist(1,40, color=groupb)
```

$$
\hat\mu(w,x) - \mu(w,x) = \qty{\hat\mu(w,x) - \inpink{\tilde\mu(w,x)}} + \qty{\inpink{\tilde\mu(w,x)} - \mu(w,x)}
$$

- What we see looks a lot like what we saw when we were estimating $\hat\mu$ using the 'all functions' model.
- But there's one significant difference: bias. 
  - The sampling distribution of $\hat\mu(w,x)$ isn't centered on the population mean $\mu(w,x)$.
  - It's centered on the population least squares predictor $\inpink{\tilde\mu(w,x)}$.
$$
\E[\hat\mu(w,x)] = \tilde\mu(w,x) \qfor \tilde\mu = \argmin_{m \in \model} \sum_{j=1}^m \qty{y_j - m(w_j,x_j)}^2
$$

- That's what we're going to prove now. 

**Q.** If the distribution of $\hat\mu$ is centered on the population least squares predictor $\tilde\mu$, \
why is $\hat\mu(w,x)$ unbiased when we use the 'all functions model'?

## Implications for Complex Targets

```{r}
#| fig.width: 7
#| layout-ncol: 2
#| fig-cap: 
#|  - 'A sample drawn with replacement from a population.'
#|  - 'That population.'
sam |> lsq.plot(y~w*x)
pop.lsq.plot
```


- Let's consider a more complex estimation target, e.g., the adjusted difference in means $\hat\Delta_{\text{all}}$.
  - This, like most of our targets, is a linear combination of column means.
  - To keep things simple, let's suppose that the weights are non-random.
  - e.g., if we were able to get the distribution of $w,x$ in the population from a census or voter file.
- This is how we've been estimating it. 
$$
\begin{aligned}
\hat\theta 
&= \frac{1}{m}\sum_{j=1}^m \qty{ \hat\mu(1,x_j) -  \hat\mu(0,x_j) } 
&&\qfor \hat\mu = \argmin_{m \in \model} \sum_{i=1}^n \qty{ Y_i - m(X_i) }^2 \\
&= \sum_{w,x} \alpha(w,x) \hat\mu(w,x) 
&&\qfor \alpha(w,x) = \begin{cases} p_{x} & \text{ for } w=1 \\
                                    -p_{x} & \text{ for } w=0. \end{cases}
\end{aligned}
$$

- We can decompose our estimator's error by comparing to $\tilde\mu$ term-by-term. 
$$
\begin{aligned}
\hat\theta-\theta 
&= \sum_{w,x} \alpha(w,x) \qty{ \hat\mu(w,x) - \mu(w,x) } \\ 
&= \underset{\faint{\text{sampling error}}}{\sum_{w,x} \alpha(w,x) \qty{ \hat\mu(w,x) - \tilde\mu(w,x) }} + \underset{\faint{\text{modeling error}}}{\sum_{w,x} \alpha(w,x) \qty{ \tilde\mu(w,x) - \mu(w,x) }} \\ 
\end{aligned}
$$

- This breaks our error down into two pieces. 
  - The *sampling error* has mean zero. Why?
  - The *modeling error* is deterministic. It's our estimator's bias. 

# Conclusion 

## What Do We Do With This?
```{r}
#| layout-ncol: 2
#| fig-width: 7
theta.sampling.dist.plot
pop.lsq.plot
```
$$
\begin{aligned}
\hat\theta-\theta 
&= \underset{\faint{\text{sampling error}}}{\sum_{w,x} \alpha(w,x) \qty{ \hat\mu(w,x) - \tilde\mu(w,x) }} + \underset{\faint{\text{modeling error}}}{\sum_{w,x} \alpha(w,x) \qty{ \tilde\mu(w,x) - \mu(w,x) }} \\ 
\end{aligned}
$$

- We can estimate the distribution of the sampling error as usual.
  - Meaning using the bootstrap.
  - There are variance formulas, but they're a little complicated.
- If there were no modeling error, this'd mean we could calibrate interval estimates for 95% coverage. With modeling error, we can't do that. 
- This leaves us with some bad options. 
  1. We act as if there's no modeling error. 
  2. We act as if what we really wanted to estimate was $\tilde\theta$.
- Because these choices are so popular, it's important to understand the implications of doing them.
- But we do have some good options.
  1. Use the all-functions model, so there really isn't any modeling error.
  2. Use *inverse probability weighting*, a special way estimation-target-specific way of estimating $\mu$ that can eliminate or at least significant reduce modeling error. That's what we'll discuss next class.
  
## Putting It All Together: An Example

```{r}
#| layout-ncol: 3
#| layout-nrow: 2
#| fig-width: 3.5

plothist(0,20, color=groupa)
plothist(0,30, color=groupa)
plothist(0,40, color=groupa)
plothist(1,20, color=groupb)
plothist(1,30, color=groupb)
plothist(1,40, color=groupb)
```

- Above, we see the sampling distribution of $\hat\mu(w,x)$ for $w=0,1$ and $x=20,30,40$.
- Roughly what's the bias of $\hat\theta$? 
  - Could you have guessed that from the plot of $\hat\mu$ on top of the sample? 
  - Or did you need all the population information displayed here?

$$
\hat\theta = \frac{1}{m}\sum_{j=1}^m \qty{ \hat\mu(1,x_j) -  \hat\mu(0,x_j) } 
$$

```{r}
#| fig.width: 7
#| layout-ncol: 2
#| fig-cap: 
#|  - 'A sample drawn with replacement from a population.'
#|  - 'That population.'
sam |> lsq.plot(y~w*x)
pop.lsq.plot
```

# Appendix  

## Checking that $m=\mu_E - \tilde\mu$ is in the model {#sec-check-in-model}

- In our proof, we derive this orthogonality condition. 
$$
0 = \E\left[ \tcb{ \hat\mu(W_i,X_i) - \tilde\mu(W_i,X_i) } \ m(W_i,X_i) \right] \qqtext{ for all } m \in \model
$$
- And we plug $m=\mu_E - \tilde\mu$ to get this equation.
$$
0 = \E\left[ \tcb{ \hat\mu(W_i,X_i) - \tilde\mu(W_i,X_i) } \ \tcb{ \mu_E(W_i,X_i) - \tilde\mu(W_i,X_i) } \right]
$$
- But this only follows if we $m=\mu_E - \tilde\mu$ is in the model $\model$.
  - It is, but to really prove our claim, we have to check.
  - This ultimately boils down to linearity of the model and linearity of expectation.
- $\mu_E$ is a (probability-weighted) linear combination of functions in the model, so it's in the model.
  
$$
\begin{aligned}
\mu_E &= \sum_{\substack{(w_1,x_1,y_1) \ldots (w_n,x_n,y_n) \\ w_i=w, x_i=x }} 
\hat\mu \times P\qty{ (W_1,X_1,Y_1) \ldots (W_n,X_n,Y_n) = (w_1,x_1,y_1) \ldots (w_n,x_n,y_n) \mid W_i=w, X_i=x} \\
&\qfor \hat\mu = \argmin_{m \in \model} \sum_{i=1}^n \qty{ y_i - m(w_i,x_i) }^2
\end{aligned}
$$
