---
title: "Lecture 3"
subtitle: "Calibrating Interval Estimates"
format:
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---

```{r setup}
library(ggplot2)
library(scales)
library(knitr)
library(magrittr)
library(tidyverse)

percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }
```

```{r ggplot-themes}
class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
            panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
            axis.ticks.x = element_blank(),
            axis.ticks.y = element_blank(),
            axis.text.x  = element_text(colour = "#aaaaaa"),
            axis.text.y  = element_text(colour = "#aaaaaa"),
            axis.title.x  = element_text(colour = "#aaaaaa"),
            axis.title.y  = element_text(colour = "#aaaaaa", angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
                axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
             panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(simplified.theme)
```

# Review

## Polling
::: {.hidden}

```{r population}
our.seed = 0
set.seed(our.seed)
N=7233584
p=4999960/N
y= c(rep(1,p*N), rep(0, (1-p)*N))
y=sample(y) # shuffle the order 
population.proportion = p
m=N
```

```{r sample}
set.seed(our.seed)

n = 625
J = sample(N, n, replace=TRUE)
Y = y[J]
sample.proportion = mean(Y)
```

:::

- We want to estimate the proportion of *all registered voters* --- our population --- who will vote.
- To do this, we use the proportion of *polled voters* --- our sample --- who said they would.
- When the election occurs, we get to see who turns out to vote.
  - `r approx(p*N)` people, or roughly `r percent(mean(Y))` of registered voters, actually vote.
 
\ \

:::: columns
::: column
### Before the Election 

$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}  \\
Y_i  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$

:::
::: column
### After the Election

$$
\begin{array}{r|rrrrrr|r}
j & 1 & 2 & 3 & 4 & \dots &   `r approx(N)` & \bar{y}_{`r approx(N)`}  \\
y_{j} & `r y[1]` & `r y[2]` & `r y[3]` & `r y[4]` & \dots & `r y[N]`   & `r approx(mean(y))`  \\
\end{array}
$$
:::
::::


## The Sampling Distribution of our Estimate

```{r}
#| out-height: 400px
#| fig-asp: .3 

width = function(samples, alpha=.05) {
  covg.dif = function(w) { 
    actual = mean( mean(samples) - w/2 <= samples & samples <= mean(samples) + w/2)
    nominal = 1-alpha 
    actual - nominal  
  }
  uniroot(covg.dif, interval=c(0, 2*max(abs(samples-mean(samples)))))$root
}

set.seed(our.seed)
Js = array(dim=c(10000, n))
Ys = array(dim=c(10000, n))
for(rr in 1:10000) {
  Js[rr,] = sample(N, n, replace=TRUE)
  Ys[rr,] = y[Js[rr,]]
}
meanY.samples = rowMeans(Ys)
height = .04
scale_x = scale_x_continuous(breaks=seq(.6,.8, by=.025), limits=c(.6, .8)) 
scale_y = scale_y_continuous(breaks=seq(0,height,by=.01), limits=c(0, height))
annotate.histogram = function(samples) { 
    w66 = width(samples, alpha=1/3)
    w95 = width(samples, alpha=.05)
    ggplot() + geom_bar(aes(x=samples, y=after_stat(prop)), position='identity', alpha=.3) +
           geom_vline(aes(xintercept=mean(y)), color='green', size=3, alpha=.7) +
           geom_vline(aes(xintercept=mean(samples)), color="blue") +
           geom_vline(aes(xintercept=mean(samples)+c(-1,1)*w66/2), color="blue", linetype='dashed') +
           geom_vline(aes(xintercept=mean(samples)+c(-1,1)*w95/2), color="blue", linetype='dotted') +
           scale_x + scale_y
}

with.points = function(histplot, samples, K=100) { 
    K=100
    set.seed(our.seed)
    heights = sample(height*(1:K)/K)
    histplot + 
        geom_point(aes(y=heights[5:K],  x=samples[5:K]), alpha=.1, color='purple') + 
        geom_point(aes(y=heights[1],    x=samples[1]), alpha=1, color='black') +  
        geom_point(aes(y=heights[2],    x=samples[2]), alpha=1, color='magenta') +
        geom_point(aes(y=heights[3],    x=samples[3]), alpha=1, color='cyan') +
        geom_point(aes(y=heights[4],    x=samples[4]), alpha=1, color='gray') 
}

with.intervals = function(histplot, samples, w=width(samples, alpha=.05), K=100) {
    K=100
    set.seed(our.seed)
    heights = sample(height*(1:K)/K)
    histplot + 
        geom_pointrange(aes(y=heights[5:K], x=samples[5:K],
                            xmin = samples[5:K]-w/2, xmax = samples[5:K]+w/2), alpha=.1, color='purple') +
        geom_pointrange(aes(y=heights[1], x=samples[1], 
                            xmin = samples[1]-w/2, xmax = samples[1]+w/2), alpha=1, color='black') +  
        geom_pointrange(aes(y=heights[2], x=samples[2],
                            xmin = samples[2]-w/2, xmax = samples[2]+w/2), alpha=1, color='magenta') +  
        geom_pointrange(aes(y=heights[3], x=samples[3],   
                            xmin = samples[3]-w/2, xmax = samples[3]+w/2), alpha=1, color='cyan') + 
        geom_pointrange(aes(y=heights[4], x=samples[4],                                   
                            xmin = samples[4]-w/2, xmax = samples[4]+w/2), alpha=1, color='gray') 
}

annotated.histogram = annotate.histogram(meanY.samples)
with.points(annotated.histogram, meanY.samples)
```

```{r}
J1 = Js[1,]
J2 = Js[2,]
J3 = Js[3,]
JM = Js[4,]
```

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{pollster} & J_1 & Y_1 & J_2 & Y_2   & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\hline
\color{black}   1   
    & \color{black}   `r J=J1; Y=y[J]; J[1]`    & \color{black}  `r Y[1]` 
    & \color{black}   `r J[2]`   & \color{black}   `r Y[2]`           & \color{black}   \dots 
    & \color{black}   `r J[n]`   & \color{black}   `r Y[n]`           & \color{black}   `r approx(mean(Y))` \\
\color{magenta} 2   
    & \color{magenta} `r J=J2; Y=y[J]; J[1]`    & \color{magenta}  `r Y[1]` 
    & \color{magenta} `r J[2]`   & \color{magenta}  `r Y[2]`          & \color{magenta} \dots 
    & \color{magenta} `r J[n]`   & \color{magenta} `r Y[n]`           & \color{magenta} `r approx(mean(Y))` \\
\color{cyan}    3   
    & \color{cyan}    `r J=J3; Y=y[J]; J[1]`    & \color{cyan}     `r Y[1]` 
    & \color{cyan}    `r J[2]`   & \color{cyan}    `r Y[2]`           & \color{cyan}    \dots 
    & \color{cyan}    `r J[n]`   & \color{cyan}    `r Y[n]`           & \color{cyan}    `r approx(mean(Y))` \\
{\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots} &  {\vdots} \\ 
\color{gray}    1M   
    & \color{gray}    `r J=JM; Y=y[J]; J[1]`    & \color{gray}     `r Y[1]` 
    & \color{gray}    `r J[2]`   & \color{gray}    `r Y[2]`           & \color{gray}    \dots 
    & \color{gray}    `r J[n]`   & \color{gray}    `r Y[n]`           & \color{gray}    `r approx(mean(Y))` \\
 {\vdots} & {\vdots} & {\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots}  & {\vdots} \\ 
\end{array}
$$


## Interval Estimation

```{r}
#| out-height: 400px
#| fig-asp: .3 

w = .02
with.intervals(annotated.histogram, meanY.samples, w=w)
```

- To give a sense of what this tells us about the population proportion, we can report an interval estimate.
    - i.e., we can draw 'arms' on our point estimate $\bar Y_{`r n`}$. 
    - Here, we've used arms of width .02. i.e., we've drawn in the interval $\bar Y_{`r n`} \pm .01$.
- But this is only useful if we have a sense of how likely this interval is to be 'right.'
- That is, how often, when we report and interval, it contains [the population proportion]{.fg style="color:limegreen"}.
- We call this a *coverage probability*. In this case, it's bad. It's about **`r percent(mean(abs(meanY.samples - mean(y)) <= .01))`**.
- If you run polls and report intervals like this, you'll make misleading claims about turnout **`r percent(1-mean(abs(meanY.samples - mean(y)) <= .01))`** of the time. 
- That's not a good look. To fix this, we need to **calibrate** our interval estimates.

## A Useful Counting Trick

```{r}
#| out-height: 400px
#| fig-asp: .3 

adhoc.width.histogram = annotated.histogram + 
    geom_rect(aes(xmin=mean(y)-w/2, xmax=mean(y)+w/2, ymin=0, ymax=Inf), alpha=.15, fill='green')
with.intervals(adhoc.width.histogram, meanY.samples, w=.02)
```

- We don't really need to look at every interval---and its arms---to calculate coverage.
- Instead, we can draw arms of the same width around the population proportion.
- An interval estimate covers the [population proportion]{.fg style="color:limegreen"} if and only if ...
- the corresponding point estimate is between the population proportion's arms.

## A Useful Counting Trick

```{r}
#| out-height: 400px
#| fig-asp: .3 

with.points(adhoc.width.histogram, meanY.samples)
```

- So all we need to do is count how many of our point estimates are between the population proportion's arms.
- Or, equivalently, calculate the **mass of our histogram** between the population proportion's arms.
- That is, we calculate the area of the verticals bars in the [shaded region]{.fg style="color:limegreen"}.
- You can see it's roughly half the area of the histogram itself. Hence, roughly 50% coverage.

## Calibrating Interval Estimates

```{r}
#| out-height: 400px
#| fig-asp: .3 

w=width(meanY.samples, alpha=.05)
calibrated.width.histogram = annotated.histogram + 
    geom_rect(aes(xmin=mean(y)-w/2, xmax=mean(y)+w/2, ymin=0, ymax=Inf), alpha=.15, fill='green')
with.points(calibrated.width.histogram, meanY.samples)
```

- To increase the coverage of our intervals estimates, we need to make the arms wider.
- To **calibrate** this width to give us 95% coverage---which is the convention---we can use our counting trick.
- We expand our [shaded region]{.fg style="color:limegreen"} until it covers 95% of the histogram's mass, i.e., 95% of our point estimates.
- If we use intervals of this width around our point estimates, we'll get 95% coverage.

## Calibrating Interval Estimates
```{r}
#| out-height: 400px
#| fig-asp: .3 

with.intervals(calibrated.width.histogram, meanY.samples, w=w)
```

- To increase the coverage of our intervals estimates, we need to make the arms wider.
- To **calibrate** this width to give us 95% coverage---which is the convention---we can use our counting trick.
- We expand our [shaded region]{.fg style="color:limegreen"} until it covers 95% of the histogram's mass, i.e., 95% of our point estimates.
- If we use intervals of this width around our point estimates, we'll get 95% coverage.

## The Problem

```{r}
#| out-height: 400px
#| fig-asp: .3 

with.intervals(calibrated.width.histogram, meanY.samples, w=w)
```

- Unless we've seen the population, i.e. unless it's after the election, we can't run simulated polls.
- That means we can't calibrate our interval using the sampling distribution of our point estimate. 
- That's something we don't know. The good news is that it's something we can estimate.
- So we'll actually do calibration using **an estimate of the sampling distribution**.

# Real-Life Calibration

## Calibration using a Sampling Distribution Estimate

```{r}
#| out-height: 400px
#| fig-asp: .3 

set.seed(our.seed)
estimated.dist.samples = rbinom(10000, n, mean(Y)) / n 
w.est=width(estimated.dist.samples, alpha=.05)
annotated.histogram.estimated = annotate.histogram(estimated.dist.samples)
calibrated.width.histogram.estimated = annotated.histogram.estimated + 
    geom_rect(aes(xmin=mean(Y)-w.est/2, xmax=mean(Y)+w.est/2, ymin=0, ymax=Inf), alpha=.15, fill='red')
with.intervals(calibrated.width.histogram.estimated, estimated.dist.samples)
```

- Here we've estimated the sampling distribution --- we'll get to how --- and used it to find an interval width.
- In particular, we've used a width which ...
    - if we draw arms around the mean of our estimated sampling distribution
    - will cover 95% of draws from our estimated sampling distribution.
- The one apparent sticking point is that our estimated sampling distribution is a bit off-center.
- Its mean is a bit lower than [the population proportion]{.fg style="color:limegreen"}. It's actually our sample proportion $\bar Y$.
- It turns out that this doesn't matter much. Why?

## Calibration Comparison

```{r}
#| out-height: 400px
#| fig-asp: .3 
calibration.comparison.histogram = 
    calibrated.width.histogram + 
    geom_rect(aes(xmin=mean(y)-w.est/2, xmax=mean(y)+w.est/2, ymin=0, ymax=Inf), alpha=.15, fill='red')

K=100
set.seed(our.seed)
heights = sample(height*(1:K)/K)

calibration.comparison.histogram +  
    geom_pointrange(aes(y=heights[1]+3, x=meanY.samples[1], 
                        xmin = meanY.samples[1]-w.est/2, xmax = meanY.samples[1]+w.est/2), alpha=1, color='red')  +
    geom_pointrange(aes(y=heights[1], x=meanY.samples[1], 
                        xmin = meanY.samples[1]-w/2, xmax = meanY.samples[1]+w/2), alpha=1, color='black')   
```

- It doesn't matter because we're not putting arms on draws from the estimated sampling distribution.
- We're putting arms on our point estimate, which is a draws from the **actual sampling distribution**.
- All that matters is the width---not the center---of the estimated sampling distribution.
- And that's very close to the width we get from the actual sampling distribution.
- Above, I've drawn in shaded regions corresponding to two versions of the the [population proportion's]{.fg style="color:limegreen"} arms.
    - The [green region]{.fg style="color:limegreen"} is the one we get from the actual sampling distribution. We can't use this.
    - The [red region]{.fg style="color:red"} is the one we get from the estimated sampling distribution. We can --- and do --- use this.
- And I've drawn in interval estimates based on our poll using both widths. They're almost the same.


## Our Samping Distribution is Binomial

```{r}
#| out-height: 300px
#| fig-asp: .25 
p = mean(y)
xx= seq(0,n)

with.points(calibrated.width.histogram, meanY.samples) +
            geom_line(aes(x=xx/n, y=dbinom(xx, n, mean(y))), color='black')
```

- To estimate the sampling distribution, it helps to know something about the actual sampling distribution.
    - It's the distribution of the proportion heads in `r n` flips of a coin with probability $p\approx `r approx(p)`$ of heads.
    - where $p \approx `r approx(p)`$ is the proportion of registered voters who will vote.
- There's a name for that: it's the binomial distribution. 
    - To be more precise, the binomial distribution is for the number of heads.
    - And we scale that by $1/`r n`$ to get a distribution for the proportion.

$$
`r n` \times \bar{Y}_{`r n`} = \sum_{i=1}^{`r n`} Y_i \sim \text{Binomial}(`r n`, p) \qfor p = \bar y_{`r approx(m)`} = \frac{1}{`r approx(m)`}\sum_{j=1}^{`r approx(m)`} y_j 
$$

- There's a formula for this probability, so we don't have to draw samples and histogram if we don't want to.
- In `R`, it's `dbinom(x, n, p)`, where `x` is the number of heads, `n` is the number of flips, and `p` is the probability of heads.

## Estimating the Sampling Distribution

```{r}
#| out-height: 300px
#| layout-ncol: 2 
#| fig-asp: .5 
#| fig-cap:
#|   - "Actual Sampling Distribution"
#|   - "Estimated Sampling Distribution"

p = mean(y)

with.points(calibrated.width.histogram, meanY.samples) +
            geom_line(aes(x=xx/n, y=dbinom(xx, n, mean(y))), color='black')

with.points(calibrated.width.histogram.estimated, estimated.dist.samples) +
            geom_line(aes(x=xx/n, y=dbinom(xx, n, mean(Y))), color='black')
```

$$
`r n` \times \bar{Y}_{`r n`} = \sum_{i=1}^{`r n`} Y_i \sim \text{Binomial}(`r n`, p) \qfor p = \bar y_{`r approx(m)`} = \frac{1}{`r approx(m)`}\sum_{j=1}^{`r approx(m)`} y_j 
$$


- What we don't know is $p=\bar y$, the proportion of registered voters who turn out.
- But that's what we're using our sample proportion $\bar Y$ to estimate. 
- And we can just plug that in to get an estimate of the sampling distribution.
- To draw samples from this estimated sampling distribution, we use `rbinom`.

```{r}
#| echo: true
#| eval: false
samples = rbinom(10000, 625, mean(Y)) / 625
``` 

- That is exactly how I got the estimated sampling distribution histogram I was using for calibration. 
    - If we prefer, we can calibrate using the actual estimated binomial distribution, i.e., using `dbinom`.
    - We'll explore this a bit at some point in the future, but it doesn't make much difference.

## The Estimate Works. Why Does it Work?



```{r}
#| out-height: 300px
#| fig-asp: .6 
#| layout-ncol: 2
K = 3
ps = quantile(meanY.samples, c(.025, .5, .975))

density.stability.plot = ggplot() + 
           geom_line(aes(x=x/n, y=dbinom(x,n,p)), data=expand.grid(p=mean(y), x=0:n), alpha=.5) +
           geom_vline(aes(xintercept=qbinom(c(.025,.975), n,mean(y))/n), linetype='dotted') +
           geom_line(aes(x=x/n, y=dbinom(x,n,p), color=factor(p)), data=expand.grid(p=ps, x=0:n), alpha=.5) +
           geom_pointrange(aes(x=p, xmin=qbinom(.025,n,p)/n, xmax=qbinom(.975,n,p)/n, y=.6*height*k/K, color=factor(p)), data=data.frame(p=ps, k=1:K)) +
           xlim(.6,.8) + guides(color=FALSE) + geom_vline(xintercept=mean(y), color='green', size=2, alpha=.5)
centered.density.stability.plot = ggplot() + 
           geom_line(aes(x=x/n-p, y=dbinom(x,n,p)), data=expand.grid(p=mean(y), x=0:n), alpha=.5) +
           geom_vline(aes(xintercept=qbinom(c(.025,.975), n,mean(y))/n-mean(y)), linetype='dotted') +
           geom_line(aes(x=x/n-p, y=dbinom(x,n,p), color=factor(p)), data=expand.grid(p=ps, x=0:n), alpha=.5) +
           geom_pointrange(aes(x=0, xmin=qbinom(.025,n,p)/n-p, xmax=qbinom(.975,n,p)/n-p, y=.6*height*k/K, color=factor(p)), data=data.frame(p=ps, k=1:K)) +
           xlim(-.1,.1) + guides(color=FALSE)

density.stability.plot
centered.density.stability.plot
```

- Why are we getting a good estimate of the sampling distribution?
- The Binomial distribution is *continuous* as a function of $p$---when $p$ changes little, the distribution changes little.
- This means that, if we're not far off in the proportion, the estimated and actual sampling distributions are similar.
- The relevant difference (after centering) is even smaller because *the way* the binomial changes is mostly location.
- Here I'm showing three estimates of the sampling distribution based on three polls.
    - with centering (right) and without (left). 
- In particular, polls giving estimates at the 2.5th, 50th, and 97.5th percentiles of the actual sampling distribution. 

## The Estimate Works. Why Does it Work?
```{r}
#| out-height: 300px
#| fig-asp: .6 
#| layout-ncol: 2

density.stability.plot
centered.density.stability.plot
```

- You can think of this as a sort of 'confidence interval' for our estimate of the sampling distribution.
    - 95% of the time, you'll get an estimate somewhere between the [red]{.fg style="color:red"} and [blue]{.fg style="color:blue"} ones.
    - And, as a result, a width for your interval estimate somewhere between the [red]{.fg style="color:red"} and [blue]{.fg style="color:blue"} widths.
- As you can see, the widths we get barely change. What changes is the location.
- But because we're using calibrated 95% intervals, we still cover the population proportion. Or, at least, almost do. 

## The Bootstrap Interpretation

- There's another way to think about this.
- The estimated sampling distribution is $\text{Binomial}(n,p)/n$ for $n=`r n`$ and $p=\bar Y\approx `r mean(Y)`$.
- That is, it's the distribution of the proportion of heads in ...
    - `r n` flips of a coin 
    - that has probability $\bar Y\approx `r mean(Y)`$ of being heads.
- What's the distribution of the proportion of yeses if we poll a population of `r n` people ...
- ... among whom the proportion of yeses is exactly $\bar Y\approx `r mean(Y)`$?

::: fragment

- It's the same distribution. It's $\text{Binomial}(n,p)/n$.
- That means we can get a draw from our estimated sampling distribution by
- running a 'poll' of the people in our sample, i.e.
    - rolling a 625-sided die 625 times
    - calling up the corresponding person in our sample
    - and counting up the yeses we hear
:::


## Bootstrapping

```{r}
set.seed(our.seed)
```

```{r}
#| echo: true
#| out-height: 400px 
#| fig-asp: .3 

meanY.boot = array(dim=10000)
for(rr in 1:10000) {
    I = sample(1:n, n, replace=TRUE)
    meanY.boot[rr] = sum(Y[I])/n
}

ggplot() + geom_bar(aes(x=meanY.boot, y=after_stat(prop)), alpha=.3, position='identity') + 
           geom_line(aes(x=(0:n)/n, y=dbinom(0:n, n, mean(Y))), color='black') + 
           xlim(.6,.8)
```


