---
title: "QTM 220 Lecture 11"
subtitle: "Statistical Inference with Non-Binary Covariates"
format: 
  clean-revealjs:
    controls: true
    controls-layout: bottom-right
    embed-resources: true
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    html-math-method: mathjax
    include-in-header: 
      text: |
        <link rel="stylesheet" href="https://cdn.jsdelivr.net/gh/mdbassit/Coloris@latest/dist/coloris.min.css"/>
        <script src="https://cdn.jsdelivr.net/gh/mdbassit/Coloris@latest/dist/coloris.min.js"></script>
        <script src="slide-canvas.js"></script>
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---

## Last Week

{{< include shared-code.qmd >}}
{{< include ca-income.qmd >}}

```{r scatter-plots}
pop.scatter = ggplot(pop.tab) + 
  geom_point(aes(x=education, y=income), color='#F7766D',size=dot.size/10, 
             alpha=.1, position=position_jitter(seed=0,w=.3)) +
  scale_x() + scale_y() + guides(color=FALSE)

sample.scatter = ggplot(sample.tab) + 
  geom_point(aes(x=education, y=income), color='#F7766D',size=dot.size, 
             alpha=.2, position=position_jitter(seed=0,w=.3)) +
  scale_x() + scale_y() + guides(color=FALSE)
```


```{r}
#| out-height: 400px
sample.scatter
```

- We talked summaries of the relationship between income and education *in a sample*.
- We talked about a number of types of summaries. 
- What you'd use depends on what you're interested in.


## Three Types

```{r}
#| layout-ncol: 3
#| out-height: 400px
#| fig-asp: 1
#| fig-subcap:
#|  - A Fine Group Comparison $$\hat\mu(16) - \hat\mu(12)$$
#|  - An Aggregate Group Comparison $$\hat\mu(16) - \hat\mu(\le 12)$$
#|  - An Aggregated Comparison $$\frac{1}{4}\sum_{x= 9}^{12} \qty{ \hat\mu(x) - \hat\mu(x-1) } $$

summaries = sample.tab |> group_by(education) |>
  summarize(mean=mean(income), sd=sd(income), N=length(income))
coarsened.summaries = sample.tab[sample.tab$education %in% c(16,8:12),] |>
  group_by(education <= 12) |>
  summarize(mean=mean(income),
            sd=sd(income),
            mean.X = mean(education))

# Plot the fine-group comparison
ggplot(sample.tab) + geom_point(aes(x=education, y=income, alpha=education %in% c(16,12), color=education > 12), 
                                  size=dot.size, position=position_jitter(seed=0,w=.3)) + 
                   geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries[summaries$education %in% c(16,12), ]) +
                   scale_x() + scale_y() + scale_alpha_manual(breaks=c(FALSE, TRUE), values=c(.01,.3)) + guides(color=FALSE, alpha=FALSE) 

# Plot the aggregate comparison
ggplot(sample.tab) + geom_point(aes(x=education, y=income, alpha=education %in% c(16,8:12), color=education > 12), 
                                  size=dot.size, position=position_jitter(seed=0,w=.3)) + 
                   geom_pointrange(aes(x=mean.X, y=mean, ymin=mean-sd, ymax=mean+sd), data=coarsened.summaries) +
                   scale_x() + scale_y() + scale_alpha_manual(breaks=c(FALSE, TRUE), values=c(.01,.3)) + guides(color=FALSE, alpha=FALSE) 

# Plot the comparison of aggregates
earnings.jitter.hs = ggplot(sample.tab) + geom_point(aes(x=education, y=income, alpha=education <= 12), color='#F7766D',
                              size=dot.size, position=position_jitter(seed=0,w=.3)) +
                   scale_x() + scale_y() + scale_alpha_manual(breaks=c(FALSE, TRUE), values=c(.01,.3)) + guides(color=FALSE, alpha=FALSE)

aggregate.comparison.plot = earnings.jitter.hs +
  geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries[summaries$education <= 12,]) +
  geom_line(aes(x=education, y=mean), data=summaries[summaries$education<=12,], color='black', linetype='dashed') 

aggregate.comparison.plot
```


## Today

```{r}
#| out-height: 250px
#| layout-ncol: 2
#| cache: true

sample.scatter
pop.scatter
```


:::: columns
::: column
### Sample
$$
\small{
\begin{array}{ccccccc}
X_1 & Y_1 & X_2 & Y_2 &  \ldots & X_n & Y_n \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\end{array}
}
$$
:::
::: column
### Population
$$
\small{
\begin{array}{ccccccc}
x_1 & y_1 & x_2 & y_2 &  \ldots & x_m & y_m \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\end{array}
}
$$
:::
::::

- We'll talk about *statistical inference* based on these sample summaries.
- That is, we'll talk about what they tell us about the population our sample was drawn from.
  - A population like we illustrate on the right. Small dots, small letters, etc.
  - A sample like we illustrate on the left. Big dots, big letters, etc.
- To keep us focused, we'll talk mostly about two different *aggregated comparisons*.
  - Both are summaries of the value of a year of high school. 
  - They're different averages of the year-on-year increments in mean income.

## Our Comparisons

```{r}
#| out-height: 400px
#| fig-asp: .3
aggregate.comparison.plot + xlim(c(7.5,12.5))
```

1. The average, over the four years of high school, of the increments.
$$
\frac{1}{4}\sum_{x=9}^{12} \qty{ \hat\mu(x) - \hat\mu(x-1) } = \frac{\hat\mu(12) - \hat\mu(8)}{4}
$$

2. The average, over people with &ge;1 year of high school, of the increments *affecting them* in their last year. 
$$
\frac{\sum_{i:X_i \in 9 \ldots 12} \qty{ \hat\mu(X_i) - \hat\mu(X_i-1) }}{\sum_{i:X_i \in 9 \ldots 12} 1} 
=  \sum_{x \in 9 \ldots 12} P_x \ \qty{ \hat\mu(x) - \hat\mu(x-1) } \quad \text{ for } \quad P_x = \frac{N_x}{\sum_{x \in 9 \ldots 12}N_x}.
$$


## Why This is Enough

- These aren't the only summaries we might be interested in. We've talked about many others.
- But every one was a linear combination --- coefficient weighted sum --- of the subsample means.
- Some, like the average over years, involved coefficients that didn't depend on the covariates.

$$
\small{
\frac14\sum_{x=9}^{12} \qty{ \hat\mu(x) - \hat\mu(x-1)} = \sum_x \alpha(x) \hat \mu(x)  \qfor \alpha(x) 
= \begin{cases} 
\hphantom{-}\frac14 & \text{ if } x = 12 \\ 
          -\frac14  & \text{ if } x = 8 \\ 
          0 & \text{ otherwise} \end{cases}
}
$$

- Others, like the average over people with &ge;1 year of high school, involved coefficients that did.

$$
\small{
\sum_{x=9}^{12} P_x \qty{ \hat\mu(x) - \hat\mu(x-1)} = \sum_x \hat \alpha(x) \hat \mu(x) \qfor
\hat\alpha(x) = \begin{cases}
  P_{12} & \text{ if } x = 12 \\
  P_{x} - P_{x+1} & \text{ if } x \in \{9 \ldots 11\} \\
  -P_9 & \text{ if } x = 8 
\end{cases}
}
$$

- We'll tackle these two general cases, using our two examples as illustrations.


# What We're Doing

## Our Data

```{r}
#| out-height: 700px
#| fig-ncol: 2
sample.plot
pop.plot
```



- As before, we'll work with data from the 2022 Current Population Survey.
- We'll look at California residents age 25-35 with at least an 8th-grade education.
  - This sample includes `r nrow(earnings)` people. 
  - We'll think of them as having been sampled with replacement from a population.
  - That is, from the population of all California residents meeting these criteria.
  - That's an approximation---the CPS is a complex survey---but it's not too bad.

::: aside
As discussed earlier in the course, these locations are fictitious. The maps are just an illustration of the relationship between sample and population.
:::

## What We're Estimating {.smaller}

```{r}
#| out-height: 250px
#| layout-ncol: 2
#| cache: true

sample.scatter
pop.scatter
```


:::: columns
::: column
### Sample
$$
\small{
\begin{array}{ccccccc}
X_1 & Y_1 & X_2 & Y_2 &  \ldots & X_n & Y_n \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\end{array}
}
$$
:::
::: column
### Population
$$
\small{
\begin{array}{ccccccc}
x_1 & y_1 & x_2 & y_2 &  \ldots & x_m & y_m \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\end{array}
}
$$
:::
::::

- Qualitatively, the sample and the population are the same kind of thing. 
- They're both collections of people with incomes and education levels. 
- That means we can summarize them in the same ways.  
- It's natural to think of sample summaries as *estimates* of the corresponding population summaries.

$$
\small{
\begin{aligned}
&\hat \mu(x) = \frac{1}{N_x}\ \sum_{i:X_i=x} Y_i \qfor N_x = \sum_{i:X_i=x} 1 
&& \text{ estimates } \quad \mu(x) = \frac{1}{m_x}\ \sum_{j:x_j=x} y_j \qfor m_x = \sum_{j:x_j=x} 1 \\
&\frac{1}{4}\sum_{x=9}^{12} \qty{ \hat \mu(x) - \hat\mu(x-1) }                       
&& \text{ estimates } \quad \frac{1}{4}\sum_{x=9}^{12} \qty{ \mu(x) - \mu(x-1) }   \\
&\sum_{x=9}^{12} P_x \qty{ \hat \mu(x) - \hat\mu(x-1) } \qfor P_x = \frac{N_x}{\sum_{x=9}^{12} N_x}                       
&& \text{ estimates } \quad \sum_{x=9}^{12} p_x \qty{ \mu(x) - \mu(x-1) }   \qfor p_x = \frac{m_x}{\sum_{x=9}^{12} m_x} \\
\end{aligned}
}
$$

## Unbiasedness
```{r}
#| out-height: 200px
#| fig-asp: .2 

biasdata = data.frame(x=seq(-3,3,length.out=1000), fx=dnorm(seq(-3,3,length.out=1000)))
ggplot() + geom_area(aes(x=x,y=fx), alpha=.2, color='gray', data=biasdata) + 
           geom_vline(aes(xintercept=0), color='green', linewidth=2, alpha=.5)
```


- These estimates are *unbiased*.
  - That is, their expected value is the population summary they're meant to estimate.
  - So, loosely speaking, their sampling distribution is centered at the right **location**.
- We've seen this before in the binary-covariate case.
  - Subsample means are unbiased for subpopulation means---even conditional on $X_1 \ldots X_n$.
  - This extends, via linearity of expectations, to combinations like the differences of subsample means.
- All this still generalizes, via linearity, to the non-binary covariate case.
  - It's true for combinations with non-random coeffificents, like the average over years.
  - It's true for combinations with random, but *unbiased* coefficients, like the average over people.

$$
\small{
\begin{aligned}
&\E\qty[\hat \mu(x)] = \E\qty[\frac{1}{N_x}\ \sum_{i:X_i=x} Y_i] \qfor N_x = \sum_{i:X_i=x} 1 
&& \quad \textbf{is equal to}  \quad \mu(x) = \frac{1}{m_x}\ \sum_{j:x_j=x} y_j \qfor m_x = \sum_{j:x_j=x} 1 \\
&\E\qty[\frac{1}{4}\sum_{x=9}^{12} \qty{ \hat \mu(x) - \hat\mu(x-1) }]                       
&& \quad \textbf{is equal to} \quad \frac{1}{4}\sum_{x=9}^{12} \qty{ \mu(x) - \mu(x-1) }   \\
&\E\qty[\sum_{x=9}^{12} P_x \qty{ \hat \mu(x) - \hat\mu(x-1) }] \qfor P_x = \frac{N_x}{\sum_{x=9}^{12} N_x}                       
&& \quad \textbf{is equal to} \quad \sum_{x=9}^{12} p_x \qty{ \mu(x) - \mu(x-1) }   \qfor p_x = \frac{m_x}{\sum_{x=9}^{12} m_x} \\
\end{aligned}
}
$$

## The Average Over Years

1. To show unbiasedness, we expand our estimator into a linear combination of subsample means.
$$
\frac{1}{4}\sum_{x=9}^{12} \qty{ \hat\mu(x) - \hat\mu(x-1) } 
= \frac{\hat\mu(12) - \hat\mu(8)}{4} \\
= \sum_x \alpha(x) \hat \mu(x)  \qfor \alpha(x) = \begin{cases} 
\hphantom{-}\frac14 & \text{ if } x = 12 \\ 
          -\frac14  & \text{ if } x = 8 \\ 
          0 & \text{ otherwise} \end{cases}
$$

2. Then we calculate its expectation in this general form.

$$
\begin{aligned}
\E\qty[\sum_x \alpha(x) \hat \mu(x)] 
&= \sum_x \alpha(x) \E\qty[\hat\mu(x)] && \text{ via linearity } \\
&= \sum_x \alpha(x) \mu(x)         && \text{ because $\hat\mu(x)$ is unbiased } \\
\end{aligned}
$$

3. Then we reverse our expansion to get something more familiar-looking.

$$
\begin{aligned}
\sum_x \alpha(x) \mu(x)
&= \frac{\mu(12) - \mu(8)}{4} && \text{plugging in specific coefficients} \\
&= \frac{1}{4}\sum_{x=9}^{12} \qty{ \mu(x) - \mu(x-1) } && \text{'untelescoping'}
\end{aligned}
$$

- In the middle, we have the general case.
    - A linear combination of subsample means with non-random coefficients is an unbiased estimator ...
    - of the same linear combination of subpopulation means.
- At the top and bottom, we translate between this general form and a more familiar-looking one.


## The Average Over Years, Faster
- We could, in fact, have done the whole thing without translation.

$$
\begin{aligned}
\E\qty[\frac{1}{4}\sum_{x=9}^{12} \qty{ \hat\mu(x) - \hat\mu(x-1) }] 
&= \frac{1}{4}\sum_{x=9}^{12} \qty{ \E[\hat\mu(x)] - \E[\hat\mu(x-1)] } && \text{via linearity }\\
&= \frac{1}{4}\sum_{x=9}^{12} \qty{ \mu(x) - \mu(x-1) } && \text{via unbiasedness of subsample means }\\
\end{aligned}
$$

- But we'll need the general form later, for variance calculation, so we took this opportunity to introduce it.


## The Average Over People

1. Again, we expand our estimator into a linear combination of subsample means.
$$
\begin{aligned}
\frac{\sum_{i:X_i \in 9 \ldots 12} \qty{ \hat\mu(X_i) - \hat\mu(X_i-1) }}{\sum_{i:X_i \in 9 \ldots 12} 1}  
&=  \sum_{x \in 9 \ldots 12} P_x \ \qty{ \hat\mu(x) - \hat\mu(x-1) } \quad \text{ for } \quad P_x = \frac{N_x}{\sum_{x \in 9 \ldots 12}N_x} \\
&= \sum_x \hat \alpha(x) \hat \mu(x) \qfor
\hat\alpha(x) = \begin{cases}
  P_{12} & \text{ if } x = 12 \\
  P_{x} - P_{x+1} & \text{ if } x \in \{9 \ldots 11\} \\
  -P_9 & \text{ if } x = 8 
\end{cases}
\end{aligned}
$$

2. We calculate is expectation in this general form. There's a new wrinkle: our coefficients are random.
$$
\begin{aligned}
\E\qty[\sum_x \hat\alpha(x) \hat \mu(x)]    
&= \sum_x \E\qty[\hat\alpha(x) \hat\mu(x)] && \text{ linearity of expectations } \\
&= \sum_x \E\qty{ \E\qty[\hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n] } && \text{ law of iterated expectations } \\
&= \sum_x \E\qty{ \hat \alpha(x) \E[\hat \mu(x) \mid X_1 \ldots X_n] } && \text{ linearity of cond. expectations }\\
&= \sum_x \E\qty{ \hat \alpha(x) \mu(x) }                              && \text{ conditional unbiasedness of sample means} \\
&= \sum_x \E\qty{ \hat \alpha(x) } \mu(x)                              && \text{ linearity of expectations }.
\end{aligned}
$$

3. If our coefficients are 'unbiased' ...
  - i.e. if their expectations are the coefficients $\alpha(x)$ in the population summary
  - then we can reverse the expansion to get the population summary in familiar form.

$$
\begin{aligned}
\sum_x \E \qty{\hat\alpha(x)}\mu(x) 
      &= \sum_x \alpha(x) \mu(x)   
&& \qqtext{if}  \E\qty{ \hat \alpha(x) } = \alpha(x) \\
      &= \sum_{x=9}^{12} p_x \qty{ \mu(x) - \mu(x-1) } \qfor p_x = \frac{m_x}{\sum_{x=9}^{12} m_x} 
&& \qfor \alpha(x) = \begin{cases}
  p_{12} & \text{ if } x = 12 \\
  p_{x} - p_{x+1} & \text{ if } x \in \{9 \ldots 11\} \\
  -p_9 & \text{ if } x = 8
\end{cases} 
\end{aligned}
$$

## Does This Work Out?

$$ 
\begin{aligned}
&\text{Does} \quad \E\qty{ \hat \alpha(x) } = \alpha(x) \qfor 
\hat\alpha(x) = \begin{cases}
  P_{12} & \text{ if } x = 12 \\
  P_{x} - P_{x+1} & \text{ if } x \in \{9 \ldots 11\} \\
  -P_9 & \text{ if } x = 8 
\end{cases} \qqtext{and} 
\alpha(x) = \begin{cases}
  p_{12} & \text{ if } x = 12 \\
  p_{x} - p_{x+1} & \text{ if } x \in \{9 \ldots 11\} \\
  -p_9 & \text{ if } x = 8
\end{cases}
\end{aligned} \qqtext{?}
$$

- Yes, it does. We can show that $\E[P_x]=p_x$ for all $x$ via a sort of 'two-stage sampling argument'.
  - First flip a coin to decide if $X \in \{9 \ldots 12\}$ or not.
  - Then roll a die to decide which value you take within the group. 
  - $P_x$ is, conditional on the coin flip, a sample proportion. 
  - So, it's unbiased for the population proportion $p_x$.
- This last part isn't something I'd expect you to be able to do on an exam.
- But it is important to get a satisfying conclusion in this example.

## Summarizing Unbiasedness 

```{r}
#| out-height: 200px
#| fig-asp: .2 

biasdata = data.frame(x=seq(-3,3,length.out=1000), fx=dnorm(seq(-3,3,length.out=1000)))
ggplot() + geom_area(aes(x=x,y=fx), alpha=.2, color='gray', data=biasdata) + 
           geom_vline(aes(xintercept=0), color='green', linewidth=2, alpha=.5)
```

- We've shown that our two specific estimators are unbiased.
- More generally, we've shown that a linear combination of subsample means is unbiased ...
  - for the same linear combination of subpopulation means, if the coefficients are non-random.
  - for an analogous linear combination of subpopulation means, using the **expected coefficients**, if they are random.

$$
\begin{aligned}
\E\qty[\sum_x \alpha(x) \hat \mu(x) ] &= \sum_x \alpha(x) \mu(x) \\
\E\qty[\sum_x \hat\alpha(x) \hat \mu(x) ] &= \sum_x \E[\hat\alpha(x)] \mu(x) \\
\end{aligned}
$$


## Interval Estimation

```{r}
#| cache: true

data = sample.tab

average.over.people = function(summaries) {
  muhat = function(x) { summaries$mean[summaries$education==x] }
  P = function(x) { summaries$N[summaries$education==x] / sum(summaries$N) }
  alphahat = function(x) { if (x==12) { P(12) } else if (x==8) { -P(9) } else { P(x) - P(x+1) } }  
  sapply(8:12, function(x) { alphahat(x) * muhat(x) }) %>% sum
}
average.over.years = function(summaries) {
  (summaries$mean[summaries$education==12] - summaries$mean[summaries$education==8])/4
}
run.boot = function(data, B=1000) { purrr::map(1:B, function(i) {
  # use tryCatch to handle errors that happen when we get no  
  # units at a given education level in our bootstrap sample
  tryCatch({
  boot.sample = data[sample(1:nrow(data), replace=TRUE),]
  boot.summaries = boot.sample[boot.sample$education <= 12, ] |> 
    group_by(education) |> 
    summarize(mean=mean(income), N=length(income))
  data.frame(people=average.over.people(boot.summaries), 
             years=average.over.years(boot.summaries))
  }, error=function(e) NULL)
 }) |> list_rbind()
}

set.seed(1)
boots = run.boot(data, 1000)
summaries = data[data$education <= 12, ] %>% group_by(education) %>% 
  summarize(mean=mean(income), N=length(income), sd=sd(income))
```

```{r}
#| layout-ncol: 2
#| out-height: 300px
#| fig-asp: .3 
#| fig-subcap:
#|  - Interval Estimate for the Average over Years
#|  - Interval Estimate for the Average over People 

distribution.plot = function(boot.samples, point.estimate, w=width(boot.samples), interval.replications=0, point.replications=0) { 
  drange = c(0, dnorm(0, mean=0, sd=sd(boot.samples)))
  p = ggplot(data.frame(samples=boot.samples)) + 
    geom_histogram(aes(x=samples, y=after_stat(density)), alpha=.3, bins=50)  +
    geom_pointrange(aes(y=dnorm(2*sd(samples), mean=0, sd=sd(samples)), 
      x=point.estimate, xmin=point.estimate - w/2, xmax=point.estimate + w/2)) + 
      geom_vline(xintercept=mean(boot.samples), color='blue') + 
      geom_vline(xintercept=quantile(boot.samples, c(.025,.975)), color='blue', linetype='dotted') 
  if(interval.replications > 0) { 
    K = interval.replications
    p = p + geom_pointrange(aes(y=y, x=x, xmin=xmin, xmax=xmax), 
          data=data.frame(y=seq(min(drange), max(drange), length.out=K), 
                  x=boot.samples[1:K], xmin=boot.samples[1:K] - w/2, xmax=boot.samples[1:K] + w/2), color='purple', alpha=.2)
  }
  if(point.replications > 0) { 
    K = point.replications
    p = p + geom_point(aes(y=y, x=x), 
          data=data.frame(y=seq(min(drange), max(drange), length.out=K), 
                  x=boot.samples[1:K], xmin=boot.samples[1:K] - w/2, xmax=boot.samples[1:K] + w/2), color='purple', alpha=.2)
  }
  p
}

distribution.plot(boots$years, average.over.years(summaries), point.replications=100) + xlim(-10000,15000)
distribution.plot(boots$people, average.over.people(summaries), point.replications=100) + xlim(-10000,15000)
```

- It's nice to know that our estimators are unbiased. That's a start. But we want to know more.
- We want to know what an individual sample actually tells us about a population summary.
- To do this, we'll use a calibrated interval estimate, i.e. a **confidence interval**.
- This looks exactly like it did in the binary-covariate case we discussed earlier in the semester.
  1. We estimate the sampling distribution of our estimator.
  2. We draw arms on our point estimate wide enough to cover 95% of this distribution---95/100 [purple dots]{.fg style="color:purple"}.
- And, as before, we can use the bootstrap sampling distribution as our sampling distribution estimate.
- But then, looking at our two summaries of the average value of a year of high school, we see something.
- Our estimate of the average over years is much more precise. Its arms are much shorter. Let's understand why.

## Understanding Sources of Error

```{r}
#| layout-ncol: 2
#| out-height: 300px
#| fig-asp: .3 
#| fig-subcap:
#|  - Interval Estimate for the Average over Years
#|  - Interval Estimate for the Average over People 

x = seq(-20000,20000, length.out=1000)
distribution.plot(boots$years, average.over.years(summaries), point.replications=100) + 
  geom_area(aes(x=x, y=fx), fill='green', alpha=.2, color='black',
                data=data.frame(x=x, fx=dnorm(x, mean=average.over.years(summaries), sd=sd(boots$years)))) + 
  xlim(-10000,15000)

distribution.plot(boots$people, average.over.people(summaries), point.replications=100) + 
  geom_area(aes(x=x, y=fx), fill='green', alpha=.2, color='black',
                data=data.frame(x=x, fx=dnorm(x, mean=average.over.people(summaries), sd=sd(boots$people)))) + 
  xlim(-10000,15000)
```


- To understand why our intervals look like they do, we think about an interval based on normal approximation.
- The interval's width is determined by the estimator's **standard error**, i.e the square root of its variance.

$$
\text{population summary} \in \text{point estimate} \pm 1.96 \times \text{standard error}
$$

- To make sense of this, what we need is a variance formula. Here it is.

$$
\Var\qty[\sum_x \hat\alpha(x) \hat\mu(x)]
= \textcolor{blue}{\sum_x \sigma^2(x) \times  \E\qty[ \frac{\hat\alpha(x)^2}{N_x} ]} + \textcolor{red}{\Var\qty[ \sum_x \mu(x) \hat \alpha(x)  ]} \qfor 
\sigma^2(x) = \Var[Y_i \mid X_i=x] 
$$

- This breaks the variance of our estimator down into two parts.
  - [The first]{.fg style="color:blue"} reflects the randomness of our subsample means.
  - [The second]{.fg style="color:red"} reflects the randomness of our coefficients $\hat\alpha(x)$.
- Note that if our coefficients aren't random, i.e. $\hat\alpha(x)=\alpha(x)$, the second term is zero.
- And often, even if they are random, the first term is much bigger.
- As a result, we can usually get away with thinking about the first part only.
  - Note that the second part is positive---it's the variance of something.
  - Ignoring it gives us a **lower bound** on the actual variance of our estimator.

## The Formula

- This two-part structure arises from the law of total variance.
  - We won't get deep into the calculations here. 
  - That's 'appendix stuff' as far as this class is concerned.

$$
\small{
\begin{aligned}
\Var\qty[ \sum_x \alpha(x) \hat \mu(x) ] 
&= \E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] \quad &&+ \quad \Var\qty[ \E\qty{\sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n} ] \\
&= \E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] \quad &&+ \quad \Var\qty[ \sum_x  \hat\alpha(x) \E\qty{ \hat \mu(x) \mid X_1 \ldots X_n} ] \\
&= \E\qty[ \Var\qty{ \sum_x \hat \alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] \quad &&+ \quad \Var\qty[ \sum_x \hat\alpha(x) \mu(x) ] \\
&= \textcolor{blue}{\sum_x \E\qty[ \hat\alpha(x)^2 \Var\qty{\hat\mu(x) \mid X_1 \ldots X_n}]} \quad &&+ \quad 
\textcolor{blue}{\sum_x \sum_{x'} \mu(x) \mu(x') \Cov\qty[\hat\alpha(x),\ \hat\alpha(x')]} \\
&=\textcolor{blue}{ \sum_x \sigma^2(x) \times \E\qty[ \frac{\hat\alpha(x)^2}{N_x} ]} \quad &&+ \quad 
\textcolor{blue}{\sum_x \sum_{x'} \mu(x) \mu(x') \Cov\qty[\hat\alpha(x),\ \hat\alpha(x')]} 
\end{aligned}
}
$$


## Variance Estimation

$$
\Var\qty[\sum_x \alpha(x) \hat \mu(x)] \ge \textcolor{blue}{\sum_x \sigma^2(x) \times \E\qty[ \frac{\hat\alpha(x)^2}{N_x} ] \quad \text{ where } \quad \sigma^2(x) = \Var[Y_i \mid X_i=x]}
$$

- The lower bound has two quantities we don't know.
  - The subpopulation variances $\sigma^2(x)$.
  - The expected value of the ratio of the coefficient squared and the sample size, $\hat\alpha(x)^2 / N_x$.
- But we can estimate both easily.
- The subsample variances are good esimates of the subpopulation variances
$$
\sigma^2(x) \approx \hat\sigma^2(x) := \frac{1}{N_x} \sum_{i:X_i=x} \qty{Y_i - \hat\mu(x)}^2
$$

- The squared coefficient / subsample size ratio is usually a good estimate of its expectation.

$$ \frac{\hat\alpha(x)^2}{N_x} \approx \E\qty[\frac{\hat\alpha(x)^2}{N_x}] $$

- Plugging them in gives us an estimate of our summary's variance.

$$
\widehat{\text{standard error}}^2 = \sum_x \ \frac{\hat\sigma^2(x)\hat\alpha(x)^2}{N_x}
$$


## The Average Over Years
```{r}
#| out-height: 300px
#| fig-asp: .3
sample.scatter
```

```{r}
sigmahat = function(x) { sd(earnings$income[earnings$education==x]) }
alpha    = function(x) { case_match(x, 
  12 ~  1/4,
  8  ~ -1/4,
  .default = 0) 
}
N        = function(x) { sum(earnings$education==x) }
```

$$
\hat\sigma_{\hat\theta}^2 = \sum_x  \frac{\alpha^2(x) \ \hat\sigma^2(x)}{N_x}
$$

where

$$
\begin{array}{l|lll} x & 8 & 9 & 10 & 11 & 12 & 13 & \ldots & 20 \\ \hline
\alpha^2(x)    & `r approx(alpha(8)^2)` & `r approx(alpha(9)^2)` & `r approx(alpha(10)^2)` &
`r approx(alpha(11)^2)` & `r approx(alpha(12)^2)` & `r approx(alpha(13)^2)` &
\ldots & `r approx(alpha(20)^2)` \\
\hat\sigma^2(x) &    
`r approx(sigmahat(8)^2,1)` & `r approx(sigmahat(9)^2,1)` & `r approx(sigmahat(10)^2,1)` &
`r approx(sigmahat(11)^2,1)` & `r approx(sigmahat(12)^2,1)` & `r approx(sigmahat(13)^2,1)` &
\ldots & `r approx(sigmahat(20)^2,1)` \\
N_x  & 
`r N(8)` & `r N(9)` & `r N(10)` & 
`r N(11)` & `r N(12)` & `r N(13)` &
 \ldots & `r N(20)` \\
\frac{\alpha^2(x) \ \hat\sigma^2(x)}{N_x} &
`r approx(sigmahat(8)^2 * alpha(8)^2 / N(8))` & `r approx(sigmahat(9)^2 * alpha(9)^2 / N(9))` & `r approx(sigmahat(10)^2 * alpha(10)^2 / N(10))` &
`r approx(sigmahat(11)^2 * alpha(11)^2 / N(11))` & `r approx(sigmahat(12)^2 * alpha(12)^2 / N(12))` & `r approx(sigmahat(13)^2 * alpha(13)^2 / N(13))` &
\ldots & `r approx(sigmahat(20)^2 * alpha(20)^2 / N(20))` \\
\end{array}
$$


## Comparing Our Two Summaries
```{r}
#| layout-ncol: 2
#| out-height: 300px
#| fig-asp: .3
#| fig-subcap:
#|  - Interval Estimate for the Average over Years
#|  - Interval Estimate for the Average over People 


x = seq(-5000,10000, by=10)
point.estimate = (summaries$mean[summaries$education==12] - summaries$mean[summaries$education==8]) / 4
w = 2*1.96*sd(boots$years) 
normal.distribution.plot.years = ggplot() + 
    geom_area(aes(x=x, y=dnorm(x, mean=mean(boots$years), sd=sd(boots$years))), color='purple', fill='purple', alpha=.08) +
    geom_vline(aes(xintercept=mean(boots$years)), color='purple', linetype='solid') +
    geom_vline(aes(xintercept=xi), data=data.frame(xi=qnorm(c(.025,.975), mean=mean(boots$years), sd=sd(boots$years))), color='purple', linetype='dotted') +
    geom_pointrange(aes(y=dnorm(2*sd(boots$years), mean=0, sd=sd(boots$years)), x=point.estimate, xmin=point.estimate - w/2, xmax=point.estimate + w/2)) + 
    xlim(-5000,10000)

point.estimate = average.over.people(summaries)
w = 2*1.96*sd(boots$people) 
normal.distribution.plot.people = ggplot() + 
    geom_area(aes(x=x, y=dnorm(x, mean=mean(boots$people), sd=sd(boots$people))), color='purple', fill='purple', alpha=.08) +
    geom_vline(aes(xintercept=mean(boots$people)), color='purple', linetype='solid') +
    geom_vline(aes(xintercept=xi), data=data.frame(xi=qnorm(c(.025,.975), mean=mean(boots$people), sd=sd(boots$years))), color='purple', linetype='dotted') +
    geom_pointrange(aes(y=dnorm(2*sd(boots$people), mean=0, sd=sd(boots$people)), x=point.estimate, xmin=point.estimate - w/2, xmax=point.estimate + w/2)) + 
    xlim(-2000,8000)

normal.distribution.plot.years
normal.distribution.plot.people
```

```{r}
earnings = sample.tab
N = function(x) { sum(earnings$education==x) }
P = function(x) { sum(earnings$education==x) / sum(earnings$education <= 12) }
alphahat = function(x) { if (x==12) { P(12) } else if (x==8) { -P(9) } else { P(x) - P(x+1) } }
sigmahat = function(x) { sd(earnings$income[earnings$education==x]) }
alpha    = function(x) { case_match(x, 
  12 ~  1/4,
  8  ~ -1/4,
  .default = 0) 
}
se.years = sd(boots$years)
se.people.EV = sqrt(sum(sapply(8:12, function(x) { alphahat(x)^2 * summaries$sd[summaries$education==x]^2 / sum(earnings$education==x) })))
se.people.EV.terms = sqrt(sapply(8:12, function(x) { alphahat(x)^2 * summaries$sd[summaries$education==x]^2 / sum(earnings$education==x) }))
```

$$
\Var\qty[\sum_x \hat\alpha(x) \hat\mu(x)] \ge \E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] = \sum_x \sigma^2(x) \times \E\qty[ \frac{\hat\alpha(x)^2}{N_x} ]
$$

- Let's think about what this means for our two examples.
$$
\small{
\begin{aligned}
\hat\theta_{\text{years}}  &= \frac14\sum_{x=9}^{12} \qty{ \hat\mu(x) - \hat\mu(x-1)} = \sum_x \alpha(x) \hat \mu(x)  \qfor \alpha(x) = \begin{cases} \hphantom{-}\frac14 & \text{ if } x = 12 \\ -\frac14 & \text{ if } x= 8 \\ 0 & \text{ otherwise} \end{cases} \\
\hat\theta_{\text{people}} &= \sum_{x=9}^{12} P_x \qty{ \hat\mu(x) - \hat\mu(x-1)} = \sum_x \hat \alpha(x) \hat \mu(x) \qfor
\hat\alpha(x) = \begin{cases}
  \hphantom{-}P_{12} & \text{ if } x = 12 \\
  P_{x} - P_{x+1} & \text{ if } x \in \{9 \ldots 11\} \\
  -P_9 & \text{ if } x = 8 
\end{cases}
\end{aligned}
}
$$

- We've seen that our estimate of the average-over-years summary of the value of a year of high school has large variance.
- The problem was that it paired a relatively large weight--$\alpha(x)=-1/4$---with a very small sample size $N_8=`r sum(earnings$education==8)`$.
- And we know the average-over-people summary has an even larger variance. Let's compare.

## A Comparison

$$ 
\begin{array}{l|c|ccccc|c}
& x & 8 & 9 & 10 & 11 & 12 & \sum \\
& N_x & `r N(8)` & `r N(9)` & `r N(10)` & `r N(11)` & `r N(12)` & `r sum(sapply(8:12, N))` \\
& \hat\sigma^2(x) & `r approx(sigmahat(8)^2,1)` & `r approx(sigmahat(9)^2,1)` & `r approx(sigmahat(10)^2,1)` & `r approx(sigmahat(11)^2,1)` & `r approx(sigmahat(12)^2,1)` & \\
&\hline \\ 
\text{for}\ \ \hat\theta_{\text{people}} 
&\hat\alpha^2(x) & `r approx(alphahat(8)^2)` & `r approx(alphahat(9)^2)` & `r approx(alphahat(10)^2)` & `r approx(alphahat(11)^2)` & `r approx(alphahat(12)^2)` & \\
&\frac{\hat\sigma^2(x)\alpha^2(x)}{N_x} & `r approx(sigmahat(8)^2*alphahat(8)^2/N(8),1)` & `r approx(sigmahat(9)^2*alphahat(9)^2/N(9),1)` & `r approx(sigmahat(10)^2*alphahat(10)^2/N(10),1)` & \textcolor{blue}{`r approx(sigmahat(11)^2*alphahat(11)^2/N(11),.1)`} & `r approx(sigmahat(12)^2*alphahat(12)^2/N(12),1)` 
& `r approx(sum(sapply(8:12, function(x) { sigmahat(x)^2 * alphahat(x)^2 / N(x) })),.1)` \\
\hline \\
\text{for}\ \ \hat\theta_{\text{years}} 
&\alpha^2(x) & `r approx(alpha(8)^2)` & `r approx(alpha(9)^2)` & `r approx(alpha(10)^2)` & `r approx(alpha(11)^2)` & `r approx(alpha(12)^2)` &  \\
&\frac{\hat\sigma^2(x) \alpha^2(x)}{N_x} & \textcolor{blue}{`r approx(sigmahat(8)^2*alpha(8)^2/N(8),1)`} & `r approx(sigmahat(9)^2*alpha(9)^2/N(9),1)` & `r approx(sigmahat(10)^2*alpha(10)^2/N(10),1)` & `r approx(sigmahat(11)^2*alpha(11)^2/N(11),1)` & `r approx(sigmahat(12)^2*alpha(12)^2/N(12),1)` & `r approx(sum(sapply(8:12, function(x) { sigmahat(x)^2 * alpha(x)^2 / N(x) })),1)` \\
\\
\end{array}
$$

- For both summaries, *most* of the variance comes from [one term]{.fg style="color:blue"}---from the variance of one subsample mean.
- For the average over years, it's the $x=8$ subsample, with a sample size of $N_x=`r N(8)`$ and a coefficient of $-1/4$.
- For the average over people, it's the $x=11$ subsample, with ...
  - a sample size $N_x=`r N(11)`$ that's roughly 4x bigger.
  - but a coefficient of $`r round(alphahat(11),3)`$ that's roughly `r round(alphahat(11)/alpha(8))`x bigger, too.
- Variance grows with the *square* of the coefficient divided by the sample size ... 
- So this suggests variance would be roughly $3^2  / 4 \approx 2.25x$ bigger.
- There's actually a bit more to it---$\sigma^2(11) \approx 2 \times \sigma^2(8)$ --- so it's really `r approx(sd(boots$people)^2/sd(boots$years)^2, 1)`x bigger.

## What About The Other Part?

$$
\Var\qty[ \sum_x \alpha(x) \hat \mu(x) ] 
= \E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] \quad + \quad \Var\qty[ \E\qty{\sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n} ] 
$$

- What we've been thinking about so far is the first part of the variance.
  - That's zero when the coefficients are constants, but not when they're random.
  - Because this part was *larger* for the summary with random coefficients, that implied its variance would be larger.
  - But are we missing something? Are we underestimating how much larger it'll be by ignoring the second part?
- Basically no. 
  - We can estimate the variance of our summary---the whole thing---by the variance of bootstrap sampling distribution. It's `r approx(sd(boots$people)^2,.1)`.
  - Our estimate for the first part only --- based on our table --- was `r approx(se.people.EV^2,.1)`.
  - That suggests that the second part should be roughly $`r approx(sd(boots$people)^2,.1)` - `r approx(se.people.EV^2,.1)` \approx `r approx(sd(boots$people)^2 - se.people.EV^2,100)`$.
  - That's impossible---the part is the variance of something, so it can't be negative---but it suggests it's not very big.
  - Let's see why.
  



## Summary

- If what we're estimating is a linear combination of the *subpopulation means*, with non-random coefficients
$$ 
\small{
\theta = \sum_x \alpha(x) \mu(x)
}
$$
- Then the same linear combination of *subsample means* is unbiased.
$$
\small{
\hat \theta = \sum_x \alpha(x) \hat\mu(x) \quad \text{ satisfies } \E[\hat\theta] = \theta
}
$$
- Its variance is determined by the variance in the subpopulation, the coefficients, and the subsample sizes. And we can estimate it.
$$
\small{
\Var[ \hat\theta ] = \sum_x \sigma^2(x) \alpha^2(x) \times \E\qty[ \frac{1}{N_x} ] \approx  \sum_x \hat\sigma^2(x)\alpha^2(x) \times \frac{1}{N_x}  =: \hat\sigma_{\hat\theta}^2
}
$$
- We can use this to construct a confidence interval based on normal approximation.
$$
\small{
\theta \in \qty[ \hat \theta - 1.96\hat\sigma_{\hat\theta}, \ \hat \theta + 1.96\hat\sigma_{\hat\theta}] \quad \text{ with probability } \approx .95 \qfor \hat\sigma_{\hat\theta}^2 = \sum_x \hat\sigma^2(x)\alpha^2(x) \times \frac{1}{N_x}  
}
$$

- Fundamentally, we're going to get an imprecise estimate *if a small subsample has a large weight*. 
- By imprecise, we mean a large standard error and a wide confidence interval.


## Application {.smaller}

```{r}
#| out-height: 150px
#| fig-asp: .2

summaries = earnings %>% group_by(education) %>% 
          summarize(mean=mean(income), sd=sd(income))
estimate = (summaries$mean[summaries$education==12] - summaries$mean[summaries$education==8])/4
se = sqrt( summaries$sd[summaries$education==12]^2 / (16 * sum(earnings$education==12)) + summaries$sd[summaries$education==8]^2 / (16 * sum(earnings$education==8)) )

blueline = data.frame(x=c(8,12), y=c(summaries$mean[summaries$education==8], summaries$mean[summaries$education==12]))
aggregate.comparison.plot + geom_line(aes(x=x, y=y), data=blueline, color='blue', alpha=.4) + xlim(7.7,12.5)
```

- Let's use this to estimate the value of a year of high school in our population.
- The one we've addressed is the average, over the four years of high school, of the increments.
$$
\small{
\theta = \sum_{x=9}^{12} \frac14 \qty{ \mu(x) - \mu(x-1) } = \frac14 \qty{ \mu(12) - \mu(8) }
}
$$

- That's the slope of the secant---blue line---in an analogous plot for the population.
- We're estimating it by the slope of the secant in the plot of our sample above.
$$
\small{
\hat\theta = \sum_{x=9}^{12} \frac14 \qty{ \hat\mu(x) - \hat\mu(x-1) } = \frac14 \qty{ \hat\mu(12) - \hat\mu(8) } = `r approx(estimate)`
}
$$

- This is our estimate of its standard error, i.e., the square root of our estimate of its variance.

$$
\small{
\hat\sigma_{\hat\theta}^2 
= \frac{\hat\sigma^2(12) \times \qty(\frac{1}{4})^2}{ N_{12} } + \frac{\hat\sigma^2(8) \times \qty(\frac{1}{4})^2}{ N_{8} } 
= \frac{`r approx(summaries$sd[summaries$education==12])`^2}{16 \times `r sum(earnings$education==12)`} 
+ \frac{`r approx(summaries$sd[summaries$education==8])`^2}{16 \times `r sum(earnings$education==8)`} = `r approx(se^2)` \approx `r approx(se)`^2 
}
$$

## This is the resulting interval.
```{r}
#| layout-ncol: 2
#| out-height: 300px
#| fig-asp: .3 
#| fig-subcap:
#|  - Interval Estimate for the Average over Years
#|  - Interval Estimate for the Average over People 

summaries = earnings[earnings$education <= 12, ] %>% group_by(education) %>% 
  summarize(mean=mean(income), N=length(income), sd=sd(income))
years.estimate = (summaries$mean[summaries$education==12] - summaries$mean[summaries$education==8])/4

distribution.plot(boots$years, years.estimate) + xlim(-5000,10000)
distribution.plot(boots$people, average.over.people(summaries)) + xlim(-5000,10000)
```

$$
\small{
\qty[ \hat \theta - 1.96\hat\sigma_{\hat\theta}, \ \hat \theta + 1.96\hat\sigma_{\hat\theta}] = \qty[`r approx(estimate - 1.96*se)` , `r approx(estimate + 1.96*se)`]
}
$$

- This is for the average over years---the plot on the left.
- At the moment, we don't know why it's narrower than the interval for the average over people.
- That's because we haven't yet work out how to get a normal approximation for its distribution.
- It's a linear combinations with coefficients that *depend on the covariate*. We'll do that next time.

$$
\small{
\text{ average over people} = \sum_{x=9}^{12} P_x \qty{ \hat\mu(x) - \hat\mu(x-1)} = \sum_x \hat \alpha(x) \hat \mu(x) \qfor
\hat\alpha(x) = \begin{cases}
  P_{12} & \text{ if } x = 12 \\
  P_{x} - P_{x+1} & \text{ if } x \in \{9 \ldots 11\} \\
  -P_9 & \text{ if } x = 8 
\end{cases}
}
$$


# Appendix


## Unbiasedness of Subsample Means

- We'll start with something we've essentially shown before --- that $\hat\mu(x)$ is an unbiased estimator of $\mu(x)$. 
  - $\E[\hat\mu(x)] = \mu(x)$
- We'll show something stronger than that. We'll show that it's *covariate-conditional mean* is $\mu(x)$.
  - $\E[\hat\mu(x) \mid X_1 \ldots X_n] = \mu(x)$
- Unbiasedness follows from this by the law of iterated expectations.
  - $\E[\hat\mu(x)] = \E[\E[\hat\mu(x) \mid X_1 \ldots X_n]] = \E[\mu(x)] = \mu(x)$


::: fragment
$$
\small{
\begin{aligned}
&\E\qty[\hat\mu(x) \mid X_1 \ldots X_n] \\
&= \E\qty[\frac{\sum_{i:X_i=x} Y_i}{\sum_{i:X_i=x} 1} \mid X_1 \ldots X_n]  && \\
&= \E\qty[\frac{\sum_{i=1}^n 1_x(X_i) Y_i}{\sum_{i=1}^n 1_x(X_i)} \mid X_1 \ldots X_n] && \qfor 1_x(X) = \begin{cases} 1 & \text{ if } x = X  \\ 0 & \text{ otherwise} \end{cases} \\
&= \frac{\sum_{i=1}^n 1_x(X_i) \E[ Y_i \mid X_1 \ldots X_n]}{\sum_{i=1}^n 1_x(X_i)} && \text{ by linearity of conditional expectations } \\
&= \frac{\sum_{i=1}^n 1_x(X_i) \E[ Y_i \mid X_i]}{\sum_{i=1}^n 1_x(X_i)}          && \text{  by irrelevance of independent conditioning variables } \\
&= \frac{\sum_{i=1}^n 1_x(X_i) \mu(X_i)}{\sum_{i=1}^n 1_x(X_i)}                     && \text{  is a matter of notation } \\
&= \frac{\sum_{i=1}^n 1_x(X_i) \mu(x)}{\sum_{i=1}^n 1_x(X_i)}                       && \text{ by the \emph{indicator trick} } 1_x(X_i)\mu(X_i) = 1_x(X_i)\mu(x) \\
&= \mu(x) \frac{\sum_{i=1}^n 1_x(X_i)}{\sum_{i=1}^n 1_x(X_i)} = \mu(x) 
\end{aligned}
}
$$


## Variance Calculation Step 1: Total Variance


- We'll use the law of total variance to break the variance of our estimator into two parts.
- One of these parts --- $\Var\E[\ldots]$ part --- has always been zero before.
  - It isn't now. This part is how the *randomness of the coefficients* contributes.
  - It was zero last time because the coefficients were constants.
- What we'll get for the remaining $\E[\Var[\ldots]]$ part is almost exactly what we did before.
  - We *sum* the conditional variances of the individual terms, then take the expectation of the result.
  - All that changes is that, the coefficients being random, we can't pull them out of the expectation

$$
\small{
\begin{aligned}
\Var\qty[ \sum_x \alpha(x) \hat \mu(x) ] 
&= \E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] \quad &&+ \quad \Var\qty[ \E\qty{\sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n} ] \\
&= \E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] \quad &&+ \quad \Var\qty[ \sum_x  \hat\alpha(x) \E\qty{ \hat \mu(x) \mid X_1 \ldots X_n} ] \\
&= \E\qty[ \Var\qty{ \sum_x \hat \alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] \quad &&+ \quad \Var\qty[ \sum_x \hat\alpha(x) \mu(x) ] \\
&= \textcolor{blue}{\sum_x \E\qty[ \hat\alpha(x)^2 \Var\qty{\hat\mu(x) \mid X_1 \ldots X_n}]} \quad &&+ \quad 
\textcolor{blue}{\sum_x \sum_{x'} \mu(x) \mu(x') \Cov\qty[\hat\alpha(x),\ \hat\alpha(x')]} \\
&=\textcolor{blue}{ \sum_x \sigma^2(x) \times \E\qty[ \frac{\hat\alpha(x)^2}{N_x} ]} \quad &&+ \quad 
\textcolor{blue}{\sum_x \sum_{x'} \mu(x) \mu(x') \Cov\qty[\hat\alpha(x),\ \hat\alpha(x')]} 
\end{aligned}
}
$$

## Variance Calculation Step 2a: $\E[\Var[\ldots]]$ 

$$\small{
\begin{aligned}
\E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] 
&=\E\qty[ \E\qty{ \qty(\sum_x \hat\alpha(x) \hat\mu(x) - \E\qty[\sum_x \alpha(x) \hat\mu(x) \mid X_1 \ldots X_n])^2 \mid X_1 \ldots X_n }] \\
&=\E\qty[ \E\qty{ \qty(\sum_x \hat\alpha(x) \hat\mu(x) - \alpha(x) \E\qty[\hat\mu(x) \mid X_1 \ldots X_n])^2 \mid X_1 \ldots X_n }] \\
&=\E\qty[ \E\qty{ \qty(\sum_x \hat\alpha(x) \qty{\hat\mu(x) - \E\qty[\hat\mu(x) \mid X_1 \ldots X_n]})^2 \mid X_1 \ldots X_n }] \\
&=\E\qty[ \E\qty{ \qty(\sum_x \hat\alpha(x) Z_x)^2  \mid X_1 \ldots X_n }] \quad \text{ for } \quad  Z_x=\hat\mu(x) - \E\qty[\hat\mu(x) \mid X_1 \ldots X_n] 
\end{aligned}
}
$$

- Here $Z_x$ is a *conditionally centered* version of our subsample mean $\hat\mu(x)$.
  - The *conditional variance* of $\hat\mu(x)$ is the conditional expectation of its square $Z_x^2$
  - And the conditional expectation of *products* $Z_x Z_{x'}$ for different subsamples is zero.

$$
\begin{aligned} 
\E[Z_x^2 \mid X_1 \ldots X_n] &= \Var[\hat\mu(x) \mid X_1 \ldots X_n] \\
\E[Z_x Z_{x'} \mid X_1 \ldots X_n] &= 0 \quad \text{ for } \quad x \neq x'
\end{aligned}
$$

## Variance Calculation Step 2b: $\E[\Var[\ldots]]$ and FOIL
- To make progress from here, we have to square out our sum. We get one term for each combination of $x$s.

$$
\small{
\qty(\sum_x \hat\alpha_x Z_x)^2 = \sum_x\sum_{x'} \hat\alpha_x Z_x \ \hat\alpha_{x'} Z_{x'} 
}
$$

- Making this substitution, we can boil things down to something we've calculated before.

$$\small{
\begin{aligned}
\E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] 
&=\E\qty[ \E\qty{ \qty(\sum_x \hat\alpha(x) Z_x)^2  \mid X_1 \ldots X_n }]  \\
&=\E\qty[ \E\qty{ \sum_x\sum_{x'} \hat\alpha(x) \hat\alpha(x') Z_x Z_{x'} \mid X_1 \ldots X_n }] \\ 
&=\E\qty[ \sum_x\sum_{x'} \hat\alpha(x) \hat \alpha(x') \E\qty{ Z_x Z_{x'} \mid X_1 \ldots X_n}] \\ 
&=\E\qty[ \sum_x\sum_{x'} \hat\alpha(x) \hat \alpha(x') \begin{cases} \Var\qty{\hat\mu(x) \mid X_1 \ldots X_n} & \text{ if } x=x' \\ 0 & \text{ otherwise} \end{cases} ] \\
&=\E\qty[ \sum_x \hat\alpha(x)^2 \Var\qty{\hat\mu(x) \mid X_1 \ldots X_n}] \\
\end{aligned}
}
$$

## Conclusion for *This Part* of the Variance

```{r}
#| layout-ncol: 2
#| out-height: 300px
#| fig-asp: .3
sample.scatter
pop.scatter
```

- In the end, this part of the variance is essentially what it was last time.
- It's a linear combination of the conditional variances of the subsample means.
- We worked out how to [calculate those]{.fg style="color:blue"} before the midterm.

$$ 
\begin{aligned}
\E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] 
&=  \E\qty[\sum_x  \hat\alpha(x)^2 \textcolor{blue}{\Var\qty{\hat\mu(x) \mid X_1 \ldots X_n}}] \\
&=  \E\qty[\sum_x  \hat\alpha(x)^2  \textcolor{blue}{\frac{\sigma^2(x)}{N_x}}] \quad \text{ where } \quad \sigma^2(x) = \Var[Y_i \mid X_i=x]
\\
&= \sum_x \sigma^2(x) \times  \E\qty[ \frac{\hat\alpha(x)^2}{N_x} ]
\end{aligned}
$$


## Estimating This Part

```{r}
#| layout-ncol: 2
#| out-height: 300px
#| fig-asp: .3
sample.scatter 
pop.scatter
```

$$
\E\qty[ \Var\qty{ \sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] =
\sum_x \sigma^2(x) \times \E\qty[ \frac{\hat\alpha(x)^2}{N_x} ] \quad \text{ where } \quad \sigma^2(x) = \Var[Y_i \mid X_i=x]
$$

:::: columns
::: column

- This has two quantities we don't know.
  - The subpopulation variances $\sigma^2(x)$.
  - The expected coefficient-squared over subsample size $\hat\alpha^2(x) / N_x$.
- But there are natural estimators for them.
  - Subsample variances are estimates of the subpopulation variances.
  - A coefficient-squared over subsample size estimates its expectation.
- Plugging them in gives us an estimate of *this part* of our summary's variance.
- This is more useful than it sounds because ...
  - It's a lower bound on the whole variance---the other part is positive.
  - It's often the big part---the other part is often more-or-less irrelevant.

::: 
::: column

$$
\begin{aligned}
\sigma^2(x) &\approx \hat\sigma^2(x) := \frac{1}{N_x} \sum_{i:X_i=x} \qty{Y_i - \hat\mu(x)}^2 \\
\E\qty[\frac{\hat\alpha(x)^2}{N_x}] &\approx \frac{\hat\alpha(x)^2}{N_x} \\
\hat\sigma_{\hat\theta}^2 &= \sum_x \frac{ \hat\alpha^2(x) \ \hat\sigma^2(x) }{ N_x } \quad + \quad ?
\end{aligned}
$$
:::
::::

## Variance Calculation Step 3: The $\Var[\E[\ldots]]$ Part.

- Now we'll calculate the new part --- the $\Var[\E[\ldots]]$ part. 
- This **additional variance** is what we pay for using *estimates* of the population coefficients $\alpha(x)$.
- We're going to use a formula for the variance of a linear combination of random variables that you may have seen before.
$$
\Var\qty[\sum_x a_x Z_x] = \sum_x \sum_{x'} a_x a_x' \Cov(Z_x,\ Z_y) \qfor \Cov(Z_x,\ Z_{x'}) = \E\qty[ \qty{Z_x - \E[Z_x]} \qty{Z_{x'} - \E[Z_{x'}]} ]
$$

- You may have seen a special case instead, where the $Z_x$ are *independent*---i.e. where $\Cov(Z_x,\ Z_{x'})=0$ for $x \neq x'$.
- This gives us the simple statement *the variance of a sum is the sum of the variances*---but that's true only when the $Z_x$ are independent.
- And ours, generally, won't be. $\hat\alpha(12)=P_{12}$ and $\hat\alpha(11)=P_{11}-P_{12}$ are *not* independent.
- This general version is easy enough to prove, but we'll skip that. Here's what it tells us.

$$
\Var\qty[ \E{\sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n} ] 
= \Var\qty[ \sum_x \mu(x) \hat \alpha(x)  ] \\
= \sum_x \sum_{x'} \mu(x) \mu(x') \Cov\qty[\hat\alpha(x),\ \hat\alpha(x')] 
$$

## Variance Calculation Step 4?

$$
\Var\qty[ \E{\sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n} ] 
= \Var\qty[ \sum_x \mu(x) \hat \alpha(x)  ] \\
= \sum_x \sum_{x'} \mu(x) \mu(x') \Cov\qty[\hat\alpha(x),\ \hat\alpha(x')] 
$$

- These covariances are easy enough to calculate, but doing all of them is a bit tedious.
- We'll just do enough to get a handle on the size of this additional variance.
- And we'll use a simple *upper bound* on the covariances to do it.
- The covariance of two random variables is no larger than *the product of their standard deviations*.
$$
\Cov\qty[\hat\alpha(x), \ \hat\alpha(x')] \le \sqrt{\Var[\hat\alpha(x)]} \times \sqrt{\Var[\hat\alpha(x')]}
$$

- Now let's calculate one of these standard deviations. We'll stick with a simple one: $\hat\alpha(12)=P_{12}$.
- This is roughly the square root of $p_x (1-p_x) / \sum_x N_x$ for $x=12$
  -  i.e., the standard deviation of an average of Bernoulli random variables with expectation $p_{12}$.
- What's important is that *the denominator here is the sum of the sample sizes*. 
  - It's not small like the denominators in the other part.
  - As a result, this additional part of the variance is *much smaller* than the other one.

## Variance Calculation Step 4: Never Worth It?

$$
\Var\qty[ \E{\sum_x \hat\alpha(x) \hat \mu(x) \mid X_1 \ldots X_n} ] 
= \Var\qty[ \sum_x \mu(x) \hat \alpha(x)  ] \\
= \sum_x \sum_{x'} \mu(x) \mu(x') \Cov\qty[\hat\alpha(x),\ \hat\alpha(x')] 
$$

- I've just argued that the second part of the variance is small.
- That had to do with the weights $\hat\alpha(x)$ not being all that variable.
- This is the case for this example, but it's not always the case.
- We'll see one or two later in the semester 
