---
title: "Point and Interval Estimates"
---

# Sampling Review

{{< include shared-code.qmd >}}

```{r}
#| label: define-population
m = 6
pop = data.frame(j=1:6, 
                 y=c(0,1,1,1,0,1))
                 
scales = list(scale_x_continuous(breaks = 1:m),
              scale_y_continuous(breaks = (0:3/3), labels=sprintf("%d/3", 0:3))) 

pop.plot = ggplot(pop, aes(x = j, y = y)) + 
           geom_point(shape='circle', alpha=.1, size=5) + 
           scales               
```


:::: columns
::: column

```{r}
#| fig-width: 7
y= pop$y
pop.plot + geom_point(aes(x=j, y=y), size=5, data=pop[c(3,5,6),], color=polla)
```

:::
::: column

$$
\color{gray}
\begin{array}{r|rrrrrr|r}
j & 1 & 2 & 3 & 4 & 5 & 6 & 
\bar{y}  \\
y_{j} & `r y[1]` & `r y[2]` & \inpolla{`r y[3]`} & `r y[4]` & \inpolla{`r y[5]`} & \inpolla{`r y[6]`} 
& \textcolor{black}{`r sum(y)`/`r length(y)`}  \\
\end{array}
$$

:::
::::
On Friday, we talked about how to use sampling to summarize a population without surveying everyone in it. We estimated the proportion of people in a population of six who prefer chocolate to sour candy. In the fake population shown above, `r sum(pop$y)`/`r m` people prefer chocolate. Those who prefer chocolate are in the top row, at y=1. Those who prefer sour candy are below at y=0.

We tried a few different sampling schemes: sampling with replacement, sampling without replacement, coin-flip randomization, and convenience sampling. Ultimately we chose coin-flip randomization. We flipped a coin for each person in our population to decide whether they would be sampled. Of the [3 people who flipped heads]{.polla}, 2 prefer chocolate to sour candy. So our estimate of the proportion of the population who preferred chocolate was 2/3.


:::: columns
::: column

```{r}
#| fig-width: 7
#| layout-nrow: 3
y= pop$y
sam1 = pop[c(3,5,6),]
sam2 = pop[c(4,6),]
sam3 = pop[c(1,3,4,6),]
pop.plot + geom_point(aes(x=j, y=y), size=5, data=sam1, color=polla) + 
           geom_point(aes(x=j, y=y-.05),     size=5, data=sam2, color=pollb) + 
           geom_point(aes(x=j, y=y-.1), size=5,  data=sam3, color=pollc)
```

:::
::: column

$$
\color{gray}
\begin{array}{r|rrrrrr|r}
j & 1 & 2 & 3 & 4 & 5 & 6 & 
\bar{Y}  \\
y_{j} & `r y[1]` & `r y[2]` & \inpolla{`r y[3]`} & `r y[4]` & \inpolla{`r y[5]`} & \inpolla{`r y[6]`} 
& \inpolla{`r sum(sam1$y)`/`r length(sam1$y)`}  \\
y_{j} & `r y[1]` & `r y[2]` & `r y[3]` & \inpollb{`r y[4]`} & `r y[5]` & \inpollb{`r y[6]`} 
& \inpollb{`r sum(sam2$y)`/`r length(sam2$y)`}  \\
y_{j} & \inpollc{`r y[1]`} & `r y[2]` & \inpollc{`r y[3]`} & \inpollc{`r y[4]`} & `r y[5]` & \inpollc{`r y[6]`} 
& \inpollc{`r sum(sam3$y)`/`r length(sam3$y)`}  \\
\end{array}
$$

:::
::::

Sampling was a huge time-saver.  But is our estimate trustworthy? How close was our estimate to the actual proportion in the population of 6? We looked into it ahead of time by running a **simulation** of our survey on a fake population. We did exactly what we did in our actual survey, but we did it over and over. This let us look at what our estimator does, knowing what we want it to do: be $\textcolor{gray}{ \approx 4/6}$. Above, we've shown the results of three of these simulations. We call the distribution of estimates we get like this our estimator's **sampling distribution**. These three estimates are three *draws* from the sampling distribution. That's three equally-likely outcomes of our survey. And we do ok in these three. But that's just three. To get a better sense of what's likely, we should look at a lot more.


```{r}
#| label: sampling-mechanisms
estimator = function(sam) { mean(sam$y) }
with.replacement = function(pop, sampling.rate=1/2) { 
  N = round(m * sampling.rate) 
  J = sample(1:m, N, replace=TRUE)
  sam = pop[J,]
  estimator(sam)
}
without.replacement = function(pop, sampling.rate=1/2) {
  N = round(m * sampling.rate) 
  J = sample(1:m, N, replace=FALSE)
  sam = pop[J,]
  estimator(sam)
}
coinflip.randomization = function(pop, sampling.rate=1/2) {
  flips = rbinom(m, 1, sampling.rate)
  sam = pop[flips==1,]
  estimator(sam)
}
convenience = function(pop, sampling.rate=1/2) {
  N = round(m * sampling.rate) 
  J = 1:N 
  sam = pop[J,]
  estimator(sam)
}
```

```{r}
#| label: candy-sampling-distributions

set.seed(1) 
sampling.rate = 1/2
all.samples = rep(list(pop), 1000) |> map(function(pop) {
  data.frame(with.replacement=with.replacement(pop, sampling.rate),
             without.replacement=without.replacement(pop, sampling.rate),
             coinflip.randomization=coinflip.randomization(pop, sampling.rate),
             convenience=convenience(pop, sampling.rate))
}) |> bind_rows() |> 
      pivot_longer(cols=everything(), 
      names_to='sampling', values_to='estimate') 
all.samples$y = runif(nrow(all.samples))
```

::: {.panel-tabset group="candy"}

### Using Dots
```{r}
coinflip.samples = all.samples |> filter(sampling == 'coinflip.randomization')
scale_x = scale_x_continuous(breaks = (0:3)/3, labels=sprintf("%d/3", 0:3), limits=c(-.2, 1.2))
scale_y = scale_y_continuous(breaks = seq(0, 1, by=.25), limits=c(0,1))

sampling.dist.plot = ggplot(coinflip.samples) +
  geom_vline(xintercept=mean(y), color='green', linewidth=2, alpha=.5) + 
  geom_point(aes(x=estimate, y=y), color='purple', alpha=.2, size=.5) 
sampling.dist.plot + scale_x + scale_y
```

Above, we plot 1000 draws from the sampling distribution as 1000 [`r circle`]{.purple}s. To highlight what we want, we've drawn the population proportion as a green line. Eyeballing it, we can see that estimates are usually pretty close, but not always. To be more precise, we could count the dots in each column. But it's easier to ask our computer to do it for us.

### Using Bars 
```{r}
sampling.dist.plot + 
  geom_bar(aes(x=estimate, y=after_stat(prop)), alpha=.2) + scale_x + scale_y
```

We can draw in the proportion of dots in each column as a bar graph. If you really want to count dots in columns, this is the way to do it. But they can be a bit counterintuitive visually because of the uneven spacing of the columns. Most of the dots are in the middle, but there are a lot of columns there. A lot of slightly different estimates. As a result, no bar there is particularly high. The highest bar is out at $x=1$. How do we visualize the distribution of dots without this problem?

### Using Histograms
```{r}
histogram_overlay = geom_histogram(aes(x=estimate, y=after_stat(density)/3), breaks=(-1:6)/6 + 1/12, closed='right', alpha=.2) 
sampling.dist.plot + histogram_overlay + scale_x + scale_y
```

We can group together nearby columns into *equal-width bins* and count the dots in each bin. This gives you a sense of the *density* of estimates near each value of $x$. We call this kind of plot a *histogram*. If you want to know the fraction of dots in some interval, this makes it easy, as long as it's the interval between two bin edges.

:::

## Sampling Distributions in Real Studies

```{r}
#| layout-ncol: 2
#| fig-width: 7 
#| fig-cap:
#|  - Your Estimator in a Simulated Study
#|  - Your Estimator in a Real Study 

sampling.dist.plot + histogram_overlay + scale_x + scale_y
ggplot(coinflip.samples[1,]) + geom_point(aes(x=estimate, y=1/2)) + scale_x + scale_y
```
It's easy to think about sampling distributions when we're running a **simulation** of our study. We can run our simulation as many times as we want and plot the sampling distribution. And we can use our knowledge of the population to see if we're happy with it.

In a real study, we don't have any of this. All we get is one estimate. But that doesn't stop us from filling in the rest of the picture by thinking of it as one draw from our estimator's sampling distribution, working out where that sampling distribution is in relation to the estimation target, and using the data we have to estimate the sampling distribution.


## What Do We Do with a Sampling Distribution?
```{r}
sampling.dist.plot + 
  geom_point(aes(x=estimate, y=1/2), data=coinflip.samples[1,], color=midnight) + 
  histogram_overlay + scale_x + scale_y
```

```{r}
estimates = all.samples |> filter(sampling == 'coinflip.randomization') |> pull(estimate)
```

A sampling distribution is an odd summary of the proportion of people who prefer chocolate. It's not what you asked for. You asked for a single number describing a population. What you got was 1000 numbers describing 1000 samples. 

But it does tell you something if you look at it right. How often is the estimate exactly equal to the population proportion? `r percent(mean(estimates == mean(y), na.rm=TRUE))` of the time. How often is it off by $\\inteal{1/3}$ or less? `r percent(mean(abs(estimates - mean(y)) <= 1/3+.01, na.rm=TRUE))` of the time.

To make this sound like a statement about the population, we report two things.

1. An **interval** you can expect the population proportion to be in. For example, we think it's in the interval $\\inmidnight{2/3} \\pm \\inteal{1/3}$ because $\\inmidnight{2/3}$ is our sample proportion. 
2. The degree of **confidence** you have that it's actually in it. For example, it's in *an interval calculated exactly like this* in `r percent(mean(abs(estimates - mean(y)) <= 1/3+.01, na.rm=TRUE))` of surveys *run exactly like this*.

## Sampling Mechanisms

```{r}
#| label: fig-compare-sampling
four.sampling.dist.plot = ggplot(all.samples) +
  geom_vline(xintercept=mean(y), color='green', linewidth=2, alpha=.5) + 
  geom_point(aes(x=estimate, y=y), color='purple', alpha=.2, size=.5) +
  geom_histogram(aes(x=estimate, y=after_stat(density)/3), breaks=(-1:3)/3 + 1/6, alpha=.2)  
four.sampling.dist.plot + scale_x + scale_y + facet_grid(cols=vars(sampling)) + guides(fill=FALSE) 
```

Let's take a moment to get a sense of how different sampling schemes affect our estimate. To do this, we'll look at 1000 draws from the sampling distribution we get using each. To help it a bit easier, we'll display a histogram of each too (@fig-compare-sampling).

3/4 of the mechanisms we've plotted give us fairly similar sampling distributions. They all have a peak around 2/3, which is the population proportion. That's where most of our estimates wind up being---where we want them.

The one that's different is convenience sampling. It was most convenient for me to sample the first three people in our population. That doesn't change, so neither does our estimate. In this case, it's a good one. That happens sometimes. But you can't count on it. Without a random mechanism, it's hard to *know* your estimate will be good. Using multiple, meaningfully different fake populations can help you catch stuff that works by luck in one. Today, we'll focus on *sampling with replacement*. 

```{r}
#| label: fig-with-replacement-dist
ggplot(all.samples |> filter(sampling=='with.replacement')) +
  geom_vline(xintercept=mean(y), color='green', linewidth=2, alpha=.5) + 
  geom_point(aes(x=estimate, y=y), color='purple', alpha=.2, size=.5) +
  geom_histogram(aes(x=estimate, y=after_stat(density)/3), breaks=(-1:3)/3 + 1/6, alpha=.2)  +
  scale_x + scale_y + guides(fill=FALSE)
```

That's just to keep things simple and concrete. Most of what we'll say will apply to any sampling scheme. We'll really be talking about the relationship between three things.

1. Our estimator
2. Its sampling distribution
3. Our estimation target

Later, we'll look into this more generally. We'll talk about how we want these three things to be related and when we know they're related that way.

# Polling

{{< include ga-turnout.qmd >}}

```{r}
height = pmf.height = max(dbinom(0:n, n, mean(y))) 

scale_x = scale_x_continuous(breaks=seq(.6,.8, by=.025)) 
scale_y = scale_y_continuous(breaks=seq(0, pmf.height,by=.01))
coord = coord_cartesian(xlim=c(.6,.8), ylim=c(0, pmf.height))

bar.plot = ggplot() + geom_bar(aes(x=meanY.samples, y=after_stat(prop)), alpha=.3) +
                      geom_vline(aes(xintercept=mean(y)), color='green', size=2, alpha=.7) + 
                      scale_x + scale_y
```

Suppose that, week before the 2020 presidential election, you did some polling. You use a list of the $m \approx `r approx(m)`$ people registered to vote in Georgia. And you make $n=`r n`$ phone calls. Each call, you select a voter uniformly at random from the list, e.g. by rolling a `r approx(m)` -sided die. And then you ask the potential voter whether they plan to vote.

Suppose also that all these registered voters will pick up the phone when called, respond honestly to your questions, and not change their minds about voting. That is, suppose they tell us whether they do ultimately vote on election day.

You put your polling results in a table. It has one column for each call. In that columns, you record the call number $i$---a number from $1 \ldots `r n`$, and the response $Y_i$ of the person you called---$1$ for 'yes and $0$ for no'.

$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}  \\
Y_i  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$ 

You summarized the results with an extra column: the mean of the responses. Remember that the mean of a binary variable is a frequency. Or a proportion.^[This is one of those instances where language is complicated. We say the 'frequency a person said yes' or the 'proportion of people who said yes'. If we're leaving the people out of the sentence, we say either: 'the sample frequency' or 'the sample proportion'. Using only one or the other really limits how you phrase things.] You found that `r percent(mean(Y))` of the people polled said they would vote.

We want to estimate the proportion of *all registered voters* --- our population --- who will vote. To do this, we use the proportion of *polled voters* --- our sample --- who said they would. When the election occurs, we get to see who turns out to vote. `r approx(sum(y))` people, or roughly `r percent(mean(y))` of registered voters, actually vote. 

Outcomes for the population of registered voters. To enumerate our population, we give each registered voter a number $j \in 1 \ldots `r approx(m)`$ and write $y_j$ for the turnout of the person with ID $j$.

:::: columns
::: column
### Before the Election 

$$\
\begin{array}{r|rrrr|r}\ni    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}  \\
Y_i  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}\
$$\

:::
::: column
### After the Election

$$
\begin{array}{r|rrrrrr|r}
j & 1 & 2 & 3 & 4 & \dots &   `r approx(m)` & \bar{y}_{`r approx(m)`}  \\
y_{j} & `r y[1]` & `r y[2]` & `r y[3]` & `r y[4]` & \dots & `r y[m]`   & `r approx(mean(y))`  \\
\end{array}
$$
:::
::::

Our sample proportion $`r sum(Y)` / `r n` \approx `r approx(mean(Y))`$ is close to the population proportion $`r approx(sum(y))` / `r approx(m)` \approx `r approx(mean(y))`$

That's pretty accurate. Our reputation as a turnout pollster is intact for now. But unless we're looking to retire, one success isn't enough. We're going to poll again. If our methods aren't reliable, we've got work to do fixing them. Even if they are, if we overstate our accuracy, we're going to have to answer for it. For example, we could say we'll be off by at most 2%, since that's what happened this time. But if it's 4% next time, that's not going to look good.

### Important Questions

1. Was it luck that we got as close as we did?  
2. Could we have predicted how close we'd get *before* the election happened?


To find answers, we'll think about what'd happen if our friends had run identical polls. Each friend will choose have a different random sample $Y_1 \ldots Y_{`r n`}$ and estimate the population proportion using the proportion in their sample. We'll see how accurate these estimates tend to be.
  

This 'friends' stuff is just an informal way of talking about the *sampling distribution* of our estimator. The sampling distribution is the *probability distribution* of our estimator, i.e., distribution of the turnout frequency in a sample of size $n=`r n`$ drawn with replacement from the population $y_1 \ldots y_{`r approx(m)`}$. Each friend's estimate is, like ours, a *random variable* with this probability distribution. I've illustrated each of our samples and their estimates below.^[The locations of the dots are fake. This is just to illustrate the sampling scheme. We'll talk more about this in a later lecture.]

::: {.panel-tabset}

### Our Poll

```{r}
#| label: fig-our-poll-map
J = J1
sam = pop[J,]
sam$label = J
Y = sam$y

state.plot + geom_point(aes(x = long, y = lat), color=polla, alpha=.7, size=3, data=sam)
```


$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{variable}  & J_1 & Y_1 & J_2 & Y_2 & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\text{value}    & `r J[1]` & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`} 
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`} 
                  & `r approx(mean(Y))` \\
\end{array}
$$

### Our First Friend's Poll

```{r}
#| label: fig-friend1-poll-map
J = J2
sam = pop[J,]
sam$label = J
Y = sam$y

state.plot + geom_point(aes(x = long, y = lat), color=pollb, alpha=.7, size=3, data=sam)
```

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{variable}  & J_1 & Y_1 & J_2 & Y_2 & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\text{value}     & `r J[1]` & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`} 
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`} 
                  & `r approx(mean(Y))` \\
\end{array}
$$

### Our Second Friend's Poll

```{r}
#| label: fig-friend2-poll-map
J = J3
sam = pop[J,]
sam$label = J
Y = sam$y

state.plot + geom_point(aes(x = long, y = lat), color=pollc, alpha=.7, size=3, data=sam)
```

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{variable}  & J_1 & Y_1 & J_2 & Y_2 & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\text{value}     & `r J[1]` & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`} 
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`} 
                  & `r approx(mean(Y))` \\
\end{array}
$$

:::

Here's how we run our polls. For each call $i$, we randomly select a voter with an id we'll call $J_i$. And we record as that call's outcome the turnout of that voter: $Y_i=y_{J_i}$. On each call, each registered voter has a $1/`r approx(m)`$ chance of being called. This is called sampling with replacement because we could call the same person twice. In our poll, this is unlikely because we're making a small number of calls relative to population size.


## The Sampling Distribution of our Estimate

```{r}
#| echo: false
set.seed(default.seed)
```

We run 10,000 simulated polls and store who we call (J) and what they say (Y).
```{r}
#| echo: true
Js = array(dim=c(10000, n))
Ys = array(dim=c(10000, n))
for(rr in 1:10000) {
  Js[rr,] = sample(m, n, replace=TRUE)
  Ys[rr,] = y[Js[rr,]]
}
```

We calculate the sample proportion for each poll.
```{r}
#| echo: true
meanY.samples = rowMeans(Ys)
```

And histogram the result. That's the human-readable version. If you're a computer or a person who just really like tables, you can just put all the numbers---including our sample proportions---in a table. Before we look at the real thing, let's sketch the histogram we expect to see. 

If you take it seriously, sketching is a great way to develop your intution. 
As you sketch your histogram, think about what you're doing. Are you drawing your histogram in the *location* you think it should be in? What about the *shape*? Is its *spread* what you really think it should be?

::: {.panel-tabset}
### A Sketch

```{r} 
ggplot() + geom_bar(aes(x=meanY.samples, y=after_stat(prop)), fill=NA, color=NA)
```
 


### A Histogram
```{r}
ggplot() + geom_bar(aes(x=meanY.samples, y=after_stat(prop)), alpha=.2)  
```
### A Table

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{pollster} & J_1 & Y_1 & J_2 & Y_2   & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\hline
\polla   1   
    & \polla   `r J=J1; Y=y[J]; J[1]`    & \polla  `r Y[1]` 
    & \polla   `r J[2]`   & \polla   `r Y[2]`           & \polla   \dots 
    & \polla   `r J[n]`   & \polla   `r Y[n]`           & \polla   `r approx(mean(Y))` \\
\pollb 2   
    & \pollb `r J=J2; Y=y[J]; J[1]`    & \pollb  `r Y[1]` 
    & \pollb `r J[2]`   & \pollb  `r Y[2]`          & \pollb \dots 
    & \pollb `r J[n]`   & \pollb `r Y[n]`           & \pollb `r approx(mean(Y))` \\
\pollc    3   
    & \pollc    `r J=J3; Y=y[J]; J[1]`    & \pollc     `r Y[1]` 
    & \pollc    `r J[2]`   & \pollc    `r Y[2]`           & \pollc    \dots 
    & \pollc    `r J[n]`   & \pollc    `r Y[n]`           & \pollc    `r approx(mean(Y))` \\
{\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots} &  {\vdots} \\\\ 
\polld    1M   
    & \polld    `r J=JM; Y=y[J]; J[1]`    & \polld     `r Y[1]` 
    & \polld    `r J[2]`   & \polld    `r Y[2]`           & \polld    \dots 
    & \polld    `r J[n]`   & \polld    `r Y[n]`           & \polld    `r approx(mean(Y))` \\\\
 {\vdots} & {\vdots} & {\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots}  & {\vdots} \\\\ 
\end{array}
$$
:::

There are a few features of our sampling distribution that we may want to highlight. In @fig-annotated-hist below, the *mean of the sampling distribution* is the solid blue line. The middle 2/3 of the sampling distribution lies between the dashed blue lines. The middle 95% of the sampling distribution lies between the dotted blue lines. 

Our estimation target --- the turnout frequency in the population --- is drawn as a wide green line. It's in exactly the same place as the solid blue line.^[That's why it looks like the solid blue line is 'highlighted' in green.] What does this tell us about our estimator?

```{r}
#| label: fig-annotated-hist
#| out-height: 500px
samples = meanY.samples
estimand = mean(y)
Y=y[J1] 
samples[1] = mean(Y)
samples[2] = mean(y[J2])
samples[3] = mean(y[J3])
samples[4] = mean(y[JM])

dot.height = .7*height
dot.size = 2
scale_x = scale_x_continuous(breaks=seq(.6,.8, by=.025), limits=c(.6, .8))

turnout.bar = ggplot() + 
  geom_bar(aes(x=samples, y=after_stat(prop)), alpha=.2) + scale_x 
turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=1, dot.size=dot.size, height=dot.height)

p.as.close    = mean(abs(samples - mean(y)) <= abs(mean(Y)-mean(y))) 
w66 = width(samples, alpha=1/3)
w95 = width(samples, alpha=.05)
```

**Observation**. Our point estimate (the `r circle` in @fig-annotated-hist) is close to our estimation target. It's within `r percent(abs(mean(Y)-mean(y)))`. Did we get lucky? Not really. In 68% of polls, the estimator is within `r percent(w66/2)` of the target. In 95% of polls, the estimator is within `r percent(w95/2)` of the target.
  

Could we have predicted how close we'd get before the election happened?
Yes, in a sense. We will use an interval estimate---a *range of values* the estimation target is likely to be in. The *width* of this interval speaks to the 'how close' question. The *coverage probability* --- the probability our estimate is actually that close --- qualifies this answer.


## Interval Estimation

```{r}
w = .02 
```

Our *point estimate* of the turnout frequency in our population is the turnout frequency in our sample: $\overline{Y}_{`r n`}$. So let's try an interval of width `r w` centered on it: $\overline{Y}_{`r n`} \pm `r w/2`$. This is just a width we chose arbitarily. Maybe it's wishful thinking. Being off by at most 1\% sounds good. What we want is for our interval to **cover** our estimation target, i.e. for the population frequency to be in our interval. This one doesn't. Is that just bad luck? Or is it typical of $\pm `r w/2`$ intervals? Let's see what happens when our friends try intervals like this.

::: {.panel-tabset}

### Our Poll

```{r}
#| label: fig-interval-est-our-poll
#| out-height: 500px

turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=1, interval.width=w, 
                   dot.size=dot.size, height=dot.height)
```

### Friends' Polls

```{r}
#| label: fig-interval-est-friends-polls
#| out-height: 500px
turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=3, interval.width=w, 
                   dot.size=dot.size, height=dot.height)
```


### 100 Polls

```{r}
#| label: fig-interval-est-100-polls
#| out-height: 500px
imaginary.polls =  turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=100, interval.width=w, 
                   dot.size=dot.size, height=dot.height)
imaginary.polls
```

:::

Flip to the second tab above (@fig-interval-est-friends-polls). It shows the interval estimates our first two friends. That is, the interval estimates based on the [pink]{.pollb} and [teal]{.pollc} rows in our sampling distribution table. One of these intervals covers the estimation target. The [teal]{.pollc} one. So between ours and our two friends, we're covering 1/3 of the time. Not great. But that's just three polls. 


To get a better sense of how often this happens, we can do it for a hundred. That's what the third tab is for (@fig-interval-est-100-polls). $`r sum(meanY.samples[1:100]-w/2 <= mean(y) & mean(y) <= meanY.samples[1:100]+w/2)`/100$ of these intervals cover the estimation target. This gives us a sense of the *probability* that our interval covers the target. If we want to be more precise, we could do the same for millions of different polls. Let's not. Instead, let's find a more direct way to calculate the **coverage probability**.

## Calculating the Coverage Probability 
```{r}
#| label: fig-coverage-prob-setup
#| out-height: 500px
turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=100, interval.width=w, 
                   dot.size=dot.size, height=dot.height, 
                   highlighted.dot.alpha=.15, unhighlighted.dot.alpha=.15)
```

 **Activity**. Explain how to calculate this *coverage probability* using the *sampling distribution* of your point estimate $\bar Y_{`r n`}$. 

::: {.panel-tabset}
### No Guidance at All
### Step 1. Counting

Suppose you want to use a diagram like this to count how many of our 100 intervals cover the estimation target. But you don't want to look at the horizontal segments for each poll. They're small and hard too see. You just want to use the dots representing each poll's point estimate $\bar Y_{`r n`}$. **Sketch** something on top of our diagram to help you count.

### + Hint 

What should you sketch? Think of an interval estimate as a point --- a 'body' --- with 'arms' of a certain length. It's no so different from you. Suppose you had identical twin. And you're not sure whether you're standing close enough to touch them. But you don't want to put your arms out. You're tired from all that polling. Could your twin check for you?


### Step 2. From Counting to Probability

Now you've worked out how to count how many of 100 intervals cover the estimation target. What would you do if you had a million? Or a billion? That'd be good enough. If x% of a billion intervals cover, you're pretty safe saying that x% is the *coverage probability*. You can't look at a billion dots one by one, but you can look *a histogram* of a billion dots. **Explain** how to use that histogram to calculate the coverage probability. Use your sketch from Step 1.

### A Solution 

```{r}
#| label: fig-coverage-calc-viz
#| out-height: 500px
coverage = mean(mean(y)- w/2 <= meanY.samples & meanY.samples <= mean(y) + w/2)
coverage.hint.plot = turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=100, interval.width=w, 
                   dot.size=dot.size, height=dot.height, 
                   unhighlighted.dot.alpha=.15, highlighted.dot.alpha=.15,
                   shade.width=w, shade.color='green', shade.alpha=.2)
coverage.hint.plot
```

Shade in an interval of width .02 centered on the estimation target. This gives it 'arms' the same length as our interval estimates have. And its arms touch a point estimate *if and only if* the point estimates' arms touch it. That means we can count the intervals that cover by counting the point estimates between the dotted lines. 

What, in terms of the sampling distribution of the point estimate, is the coverage probability? It's the probability that a random draw from the sampling distribution lies in the [green]{.green} shaded area.
$$
\text{coverage probability} = P\qty(\overline{Y}_{`r n`} \in \overline{y}_{`r approx(m)`} \pm .01) 
$$

And it's about `r percent(coverage)`. We can get that by counting dots.

```{r}
#| echo: true
mean(mean(y) - .01 <= meanY.samples & meanY.samples <= mean(y) + .01)
```

Or by finding the area of the histogram that's shaded [green]{.green}.

:::

## Calibrating Interval Estimates

::: {.panel-tabset}
### $\pm `r w/2`$ Intervals 
```{r}
#| label: fig-adhoc-interval
#| out-height: 500px
coverage.hint.plot 
```
### 95% Intervals
```{r}
#| out-height: 500px

w = width(samples, alpha=.05)
dotted.width = w
coverage.good.plot = turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=100, interval.width=w, 
                   unhighlighted.dot.alpha=.15, highlighted.dot.alpha=.15,
                   shade.width=w, shade.color='green', shade.alpha=.2)
coverage.good.plot
```
:::

What we just did was choose a width and calculate a coverage probability. The coverage probability we found --- `r percent(coverage)` --- probably wasn't what we wanted. Think of how you'd advertise your polling services. 'I'm right about half the time. Actually, a bit less than that'. 95% sounds a lot better. For that, we'll have to use a wider interval.

Instead of choosing a width and calculating the coverage probability, let's go backward. We'll choose a coverage probability --- 95% is conventional. And we'll calculate the width we need to get it. An interval estimate *calibrated* like this---to have a given coverage---is called a *confidence interval*. Let's think about how to do that. Again, we'll use the sampling distribution of our point estimate. Let's take a look at our annotated histogram of point estimates again.

**Question**. Suppose you and your friends want to draw 95% confidence intervals around your point estimates. How wide do you have to make them to actually get 95% coverage?

::: {.panel-tabset}
### Review---Our Annotations
The *mean of the sampling distribution* is the solid blue line---and is the same as the estimation target. The middle 2/3 of the sampling distribution lies between the dashed blue lines. The middle 95\% of the sampling distribution lies between the dotted blue lines. 

### Answer
The width of a 95% interval should be the width between the dotted blue lines. That's the width of the 'arms' containing 95% of the estimates drawn from the sampling distribution. You can check that these intervals have the coverage we want.


```{r}
#| echo: true
#|  
coverage = mean(mean(y)- dotted.width/2 <= meanY.samples & meanY.samples <= mean(y) + dotted.width/2)
coverage
```

:::

### A Problem 

```{r}
#| label: fig-comm-mistake
#| out-height: 500px
imaginary.polls.calibrated =  turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=100, 
                   interval.width=w, highlight.dots=1,
                   dot.size=dot.size, height=dot.height)
imaginary.polls.calibrated
```

We can't calibrate intervals like this in real life. When we run our a poll, we get a single point estimate $\bar Y_{`r n`}$ based on our sample. We don't know the sampling distribution of this point estimate until election day. But what we actually do is *almost* the same. We *do* the same thing. But we use an *estimate* of the sampling distribution in place of the thing itself. That's what we'll talk about next class.

## Communication

Talking about calibrated interval estimates, a.k.a. confidence intervals, has some advantages. It focuses on what we actually want to know: where the estimation target is. It reminds us that we're not (usually) going to be able to know it exactly. It gives us a sense of how close we can expect that we've gotten.

But there's something about them that is a bit infuriating when you're not used to them. Fundamentally, you're talking about what would happen in surveys you aren't doing. You can imagine someone saying 'I don't have time for imaginary surveys. What did this one tell you?' And being pretty unhappy when the answer is 'If that's how you want to think about it, almost nothing.'

This isn't a problem with intervals. This is something that's fundamentally uncomfortable about sampling. It might feel like a miracle that you can say anything about `r approx(length(y))` people after 625 phone calls. But once someone thinks you can, it can be hard for them to accept that you also kind of can't. We really are just saying 'I think it's between `r percent(mean(Y)-w/2)` and `r percent(mean(Y)+w/2)`, but I'm wrong sometimes.' We are being clear about what 'sometimes' means. But that feels like it's about you more than it is about what they want to know.

### Don't Make this Mistake!

Talking about surveys you aren't doing is pretty awkward. So awkward that you'll want to try to avoid it. It's tempting to say something nonsensical, e.g. "the turnout frequency in the election is in my confidence interval, `r round(mean(Y)-w/2,2)`-`r round(mean(Y)+w/2,2)`, 95% of the time." This a really weird thing to say. It's like saying that 95% of the time 2 is between 1 and 3. It either is or it isn't. The only way it could make sense is if the turnout frequency were random. Because `r round(mean(Y)-w/2,2)` and `r round(mean(Y)+w/2,2)` aren't. People do that. Don't. And please don't encourage it by saying stuff like this.

This doesn't come up that often because you can just say 'confidence interval' most of the time. So you really only have to deal with all of this awkwardness in two situations. 

1. When you say 'confidence interval' and someone asks what that means.
2. When someone else says 'confidence interval' but clearly doesn't know what it means.

It takes a bit of practice to get good at this. And you'll get some.
