---
title: "Lecture 3"
subtitle: "Calibrating Interval Estimates with Binary Observations"
---


# Sampling and Calibration Review
A Polling Nightmare

## Before the Election

{{< include shared-code.qmd >}}
{{< include ga-turnout.qmd >}}

```{r}
#| out-height: 400px
#| fig-asp: .3 

set.seed(default.seed)
Js = array(dim=c(10000, n))
Ys = array(dim=c(10000, n))
for(rr in 1:10000) {
  Js[rr,] = sample(m, n, replace=TRUE)
  Ys[rr,] = y[Js[rr,]]
}
meanY.samples = rowMeans(Ys)
pmf.height = max(dbinom(0:n, n, mean(y))) 

scale_x = scale_x_continuous(breaks=seq(.6,.8, by=.025)) 
scale_y = scale_y_continuous(breaks=seq(0, pmf.height,by=.01))
coord = coord_cartesian(xlim=c(.6,.8), ylim=c(0, pmf.height))

bar.plot = ggplot() + geom_bar(aes(x=meanY.samples, y=after_stat(prop)), alpha=.3) +
                      geom_vline(aes(xintercept=mean(y)), color='green', linewidth=2, alpha=.7) + 
                      scale_x + scale_y
                      
```

```{r}
#| label: choosing-an-appropriately-mediocre-estimate
JM = Js[504,]
Y = y[JM]
```

```{r}
binom.width = function(p,n,...) { width_cdf(\(x)pbinom(n*x, n, p), p, limits=c(0,1), ...) }

your.widths = c(.02, binom.width(mean(Y), n, alpha=.1), binom.width(mean(Y), n, alpha=.05))
opacities = c(1,.25,.25)
interval.data = data.frame(x=mean(Y), w=your.widths,
                          y=pmf.height*c(.57,.6, .63), alpha=opacities)
your.interval = 
    geom_pointrange(aes(x=x, xmin=x-w/2, xmax=x+w/2, y=y, alpha=I(alpha)), 
                    data=interval.data[1,], color=polld)
your.intervals = 
    geom_pointrange(aes(x=x, xmin=x-w/2, xmax=x+w/2, y=y, alpha=I(alpha)), 
                    data=interval.data, color=polld)

ggplot() + 
    your.interval + 
    scale_x + scale_y + coord
```



:::: columns
::: column
### Your Sample 

$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}  \\
Y_i  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$

:::
::: column
### The Population

$$
\color{lightgray}
\begin{array}{r|rrrrrr|r}
j & 1 & 2 & 3 & 4 & \dots &   `r approx(m)` & \bar{y}_{`r approx(m)`}  \\
y_{j} & `r y[1]` & `r y[2]` & `r y[3]` & `r y[4]` & \dots & `r y[m]`   & `r approx(mean(y))`  \\
\end{array}
$$
:::
::::

- You want to estimate the proportion of *all registered voters* --- the population --- who will vote.
- To do this, you use the proportion of *polled voters* --- your sample --- who said they would.
    - You're probably not going to match the population proportion exactly ... 
    - ... so you report an *interval estimate*---a range of values we claim the population proportion is in.
    - You're know you're not going to be right 100% of the time you make with claims like this ...
    - ... so you state a *nominal coverage probability*---how often you're right about claims like this.^[You say that's how often, anyway. Hence 'nominal'.]
    
 - Without thinking too hard about it, you ...
    - ... report, as your interval, [your sample proportion (`r percent(mean(Y))`) &plusmn; `r percent(your.widths[1]/2)`]{.polld}. Because it sounds good.
    - ... and say your coverage probability is 95%. Because that's what everybody else says.

## After the Election

```{r}
ggplot() + 
    geom_vline(xintercept=mean(y), color='green', size=2, alpha=.7) + 
    your.intervals + 
    scale_x + scale_y + coord
```



:::: columns
::: column
### Your Sample

$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}  \\
Y_i  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$

:::
::: column

### The Population

$$
\begin{array}{r|rrrrrr|r}
j & 1 & 2 & 3 & 4 & \dots &   `r approx(m)` & \bar{y}_{`r approx(m)`}  \\
y_{j} & `r y[1]` & `r y[2]` & `r y[3]` & `r y[4]` & \dots & `r y[m]`   & `r approx(mean(y))`  \\
\end{array}
$$
:::
::::

- When the election occurs, we get to see who turns out to vote.
  - `r approx(sum(y))` people, or roughly `r percent(mean(y))` of registered voters, actually vote.
  - You --- and future employers --- can see how well you did. And how well everybody else did.
- [Your interval]{.polld} missed the target. It doesn't contain [the turnout proportion]{.target}.
  - Your point estimate is only off by `r percent(abs(mean(y) - mean(Y)))`. But you overstated your precision. 
  - Now you're kicking yourself. You'd briefly considered saying &plusmn; `r percent(your.widths[2]/2)` or &plusmn; `r percent(your.widths[3]/2)`. 
  - That would've done it. But that didn't sound as good, so you went for &plusmn; `r percent(your.widths[1]/2)`.
- You hope you're not the only one who missed. So you check out the competition.

## The Competition 

::: {.panel-tabset}
### Your Competitors' Polls

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{poll} & J_1 & Y_1 & J_2 & Y_2   & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\hline
\polla   1   
    & \polla   `r J=J1; Y=y[J]; J[1]`    & \polla  `r Y[1]` 
    & \polla   `r J[2]`   & \polla   `r Y[2]`           & \polla   \dots 
    & \polla   `r J[n]`   & \polla   `r Y[n]`           & \polla   `r approx(mean(Y))` \\
\pollb 2   
    & \pollb `r J=J2; Y=y[J]; J[1]`    & \pollb  `r Y[1]` 
    & \pollb `r J[2]`   & \pollb  `r Y[2]`          & \pollb \dots 
    & \pollb `r J[n]`   & \pollb `r Y[n]`           & \pollb `r approx(mean(Y))` \\
\pollc    3   
    & \pollc    `r J=J3; Y=y[J]; J[1]`    & \pollc     `r Y[1]` 
    & \pollc    `r J[2]`   & \pollc    `r Y[2]`           & \pollc    \dots 
    & \pollc    `r J[n]`   & \pollc    `r Y[n]`           & \pollc    `r approx(mean(Y))` \\
{\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots} &  {\vdots} \\ 
\polld    1M   
    & \polld    `r J=JM; Y=y[J]; J[1]`    & \polld     `r Y[1]` 
    & \polld    `r J[2]`   & \polld    `r Y[2]`           & \polld    \dots 
    & \polld    `r J[n]`   & \polld    `r Y[n]`           & \polld    `r approx(mean(Y))` \\
 {\vdots} & {\vdots} & {\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots}  & {\vdots} \\ 
\end{array}
$$

### Visualizing the Competition
```{r}
all.competitors.widths = meanY.samples |> map_vec(\(p)binom.width(p, n, alpha=.05))
competitors.widths = all.competitors.widths[1:100]

ggplot() + 
    with_annotations(meanY.samples, estimand=mean(y), 
                     height=pmf.height, highlight.dots=3, interval.width=competitors.widths, vline.alpha=NA) +
    your.interval + 
    scale_x + scale_y + coord 
```

:::

- It turns out that everyone got their point estimate exactly like you did.
    - Each of them rolled their `r approx(m)`-sided die `r n` times.
    - And after making their `r n` calls, they reported their sample proportion.
- You're not the only one who missed, but you're part of a pretty small club.
    - 5% of the other polls got it wrong.
- But you are the only one who claimed they'd get within 1%.
    - Everyone else said claimed they'd get within roughly `r percent(your.widths[3]/2)`. 
    - Most were right. About 95%. And a lot of them had worse point estimates than you.
- What you have is a *calibration* problem. 


## Calibration using the Sampling Distribution

```{r}
#| out-height: 400px
#| fig-asp: .3 

bar.plot = ggplot() + geom_bar(aes(x=meanY.samples, y=after_stat(prop)), alpha=.3) 
calibration.plot = bar.plot + 
    with_annotations(meanY.samples, estimand=mean(y), interval.width=competitors.widths,
                     height=pmf.height, highlight.dots=3) +
    your.interval + 
    scale_x + scale_y + coord

calibration.plot
```


- You'd have seen that your interval was miscalibrated if you knew what you do now.
    1. Your competitors' point estimates, which they got exactly like you got yours.
    2. Who actually turned out, so you could simulate even more polls if you wanted to.
- You'd have drawn &plusmn; `r percent(your.widths[1]/2)` intervals around all these point estimates.^[In technical terms, these estimates are draws from your estimator's *sampling distribution*.]
    - And noticed that many of these intervals cover [the target]{.target}. About `r percent(mean(abs(meanY.samples - mean(y)) <= your.widths[1]/2))` do.
    - Or, if leaving the target out of it, that only `r percent(mean(abs(meanY.samples - mean(meanY.samples)) <= your.widths[1]/2))` of them cover [the mean]{.blue} of all these estimates.
    - Same thing. The mean and target are right on top of one another. Like this: [|]{.blue .target}.
- And you'd have known how to choose the right interval width. One that makes 95% of these intervals cover. Right? 

## Quiz

```{r}
quiz.widths = c(binom.width(mean(y), 2*n, alpha=.5), 
                binom.width(mean(y), 2*n, alpha=.05), 
                binom.width(mean(y), 2*n, alpha=.003))
quiz.data = data.frame(x=mean(Y), w=quiz.widths,
                       y=pmf.height*c(.5,.4,.3), 
                       color=c(polla, pollb, pollc), 
                       alpha=c(1,1,1))
quiz.samples = rbinom(10000, 2*n, mean(y))/(2*n)
pmf.height = max(dbinom(0:(2*n), 2*n, mean(y)))

quiz.plot = ggplot() + geom_bar(aes(x=quiz.samples, y=after_stat(prop)), alpha=.3) +   
    with_annotations(quiz.samples, estimand=mean(y), highlight.dots=0, height=pmf.height) +
    geom_pointrange(aes(x=x, xmin=x-w/2, xmax=x+w/2, y=y, alpha=I(alpha), color=I(color)), 
                    data=quiz.data)
quiz.plot
```

- I've drawn the sampling distribution of an estimator, 100 draws from it as [`r circle`]{.purple}s, and 3 interval estimates.
- One of these interval estimates is calibrated to have exactly 95% coverage. Which is it?
  - A. The one around the [`r circle`]{.polla} on top. 
  - B. The one around the [`r circle`]{.pollb} in the middle. 
  - C. The one around the [`r circle`]{.pollc} on the bottom.
- *Submit* your answer [here](`r pollurl`).
- *Hint*. You can reach your twin with your arms if your twin can reach you with theirs. 
- *Hint*. Maybe you look like this [`r circle`]{.purple} and your twin like this [|]{.blue .target}.


## Calibration in Reality

```{r}
#| fig-width: 7
#| 
#| layout-ncol: 2
#| fig-cap: 
#| - "What You Know Now"
#| - "What You Knew Then" 
w95 = binom.width(mean(y), n, alpha=.05)
calibration.plot.shaded = bar.plot + 
    with_annotations(meanY.samples, estimand=mean(y), shade.width=w95, shade.color='green', 
                     interval.width=competitors.widths,
                     height=pmf.height, highlight.dots=3) +
    your.interval +
    scale_x + scale_y + coord

calibration.plot.shaded
ggplot() + your.interval + scale_x + scale_y + coord
```

- But you didn't know any of this. You just knew [your own point estimate]{.polld}.
- Without all that post-election information, you didn't know how to calibrate an interval estimate. 
- But everyone else did. Their widths were almost the same as the width you'd choose now. 

:::: columns 
::: column
```{r}
#| fig-width: 7
#| fig-cap: 
#| - Your competitors' intervals, recentered for easy comparison of width to the sampling distribution's.
ggplot() + 
    geom_rect(aes(xmin=mean(y)-w95/2, xmax=mean(y) + w95/2, ymin=-Inf, ymax=Inf), alpha=.15, fill='green') +
    geom_vline(xintercept=mean(y), color='green', size=2, alpha=.7) +
    geom_linerange(aes(y=pmf.height*seq(0,1,length.out=100), 
                        xmin=mean(y)-competitors.widths/2, xmax=mean(y)+competitors.widths/2), 
                        color='purple', alpha=.6) + scale_x + scale_y + coord 
```

:::
::: column

- It's as if they knew how to approximate their estimators' sampling distribution. They did.
- Today, we're going to work out how.
- We'll need to know a bit about sums of \
  independent binary random variables.

:::
::::

# Probability Review 
Sums of Independent Binary Random Variables

## From Sampling to Coin Flips

:::: columns
::: {.column width="30%"}
### The Population

$j$ | $y_j$
--|---
$1$ | `r y[1]`
$2$ | `r y[2]`
$3$ | `r y[3]`
$4$ | `r y[4]`
&vellip; | &vellip;
$m$ | `r y[m]`

:::
::: {.column width="30%"}
### The Joint

$p$ | $J_1$ | $Y_1$
--|---|---
$1/m$ | $1$ |  $`r y[1]`$
$1/m$ | $2$ |  $`r y[2]`$
$1/m$ | $3$ |  $`r y[3]`$
$1/m$ | $4$ |  $`r y[4]`$
&vellip; | &vellip; | &vellip;
$1/m$ | $m$ |  $`r y[m]`$

:::
::: {.column width="40%"}
### The Marginal

$p$ | $Y_1$
--|---
$\underset{\color{gray}\approx `r approx(mean(y==0))`}{\sum\limits_{j:y_j=0} \frac{1}{m}}$ | $0$
$\underset{\color{gray}\approx `r approx(mean(y==1))`}{\sum\limits_{j:y_j=1}  \frac{1}{m}}$ | $1$

:::
::::

- To find the distribution of one response $Y_1$, we start with the probability distribution of our dice rolls.
  - The person we call, $J_1$, is equally likely to be anyone in the population of $m \approx `r approx(m)`$ people.
  - It takes on each value $1 \ldots m$ with probability $1/m$. 
- Once we know who we're calling, we know their response, $Y_1$. 
    - If we call person $j$, their response is $y_j$. So we can add that column to our table 'for free'. 
    - The probability that we roll $j$ and hear the response $y_j$ is still $1/m$.
- And we can *marginalize* to find the distribution of $Y_1$.
    - We sum the probabilities in rows where $Y_1=0$ and $Y_1=1$. And the result is intuitive.
    - The *probability* of hearing a response $y \in \{0,1\}$ is the *frequency* of that response in the population. 
    - We'll call these frequencies $\theta_0$ and $\theta_1$. This is a little redundant because $\theta_0 = 1 - \theta_1$.
    - Our estimation target is just the 'Yes' frequency $\theta_1$. Some people call it the *success rate*. 

## From Sampling to Coin Flip Sums: Two Responses

:::: columns
::: {.column width="30%"}
### The Joint

$p$ | $J_1$ | $J_2$ | $Y_1$ | $Y_2$ | $Y_1 + Y_2$
--|---|---|---|---
$\frac{1}{m^2}$ |  $1$ | $1$ | $y_1$ | $y_1$ | $y_1+y_1$
$\frac{1}{m^2}$ |  $1$ | $2$ | $y_1$ | $y_2$ | $y_1+y_2$
&vellip; | &vellip; | &vellip; | &vellip; | &vellip;
$\frac{1}{m^2}$ |  $1$ | $m$ | $y_1$ | $y_m$ | $y_1+y_m$ 
$\frac{1}{m^2}$ |  $2$ | $1$ | $y_2$ | $y_1$ | $y_2+y_1$
$\frac{1}{m^2}$ |  $2$ | $2$ | $y_2$ | $y_2$ | $y_2+y_2$
&vellip; | &vellip; | &vellip; | &vellip; | &vellip;
$\frac{1}{m^2}$ |  $m$ | $m$ | $y_m$ | $y_m$ | $y_m+y_m$
:::
::: {.column width="42%"}
### Partially Marginalized

$p$ | $Y_1$ | $Y_2$ | $Y_1 + Y_2$
--|---|---|---
$\inpollb{\sum\limits_{j_1,j_2 : y_{j_1},y_{j_2} = 0,0} \frac{1}{m^2}}$ | $0$ | $0$ | $\inpollb{0}$
$\inpollc{\sum\limits_{j_1,j_2 : y_{j_1},y_{j_2} = 0,1} \frac{1}{m^2}}$ | $0$ | $1$ | $\inpollc{1}$
$\inpollc{\sum\limits_{j_1,j_2 : y_{j_1},y_{j_2} = 1,0} \frac{1}{m^2}}$ | $1$ | $0$ | $\inpollc{1}$
$\inpolld{\sum\limits_{j_1,j_2 : y_{j_1},y_{j_2} = 1,1} \frac{1}{m^2}}$ | $1$ | $1$ | $\inpolld{1}$

:::
::: {.column width="28%"}
### Fully Marginalized
$p$ | $Y_1 + Y_2$
--|---
$\inpollb{\sum\limits_{j_1,j_2: y_{j_1}+y_{j_2} = 0} \frac{1}{m^2}}$ | $\inpollb{0}$
$\inpollc{\sum\limits_{j_1,j_2: y_{j_1}+y_{j_2} = 1} \frac{1}{m^2}}$ | $\inpollc{1}$
$\inpolld{\sum\limits_{j_1,j_2: y_{j_1}+y_{j_2} = 2} \frac{1}{m^2}}$ | $\inpolld{2}$
:::
::::

::: {.panel-tabset}
### Two-Step Approach
- To find the distribution of a sum of two responses, $Y_1+Y_2$, we do the same thing.
- We start with the joint distribution of two dice rolls.
    - Our rolls are equally likely to be any pair of numbers in $1\ldots m$.
    - And there are $m^2$ pairs, so the probability is $1/m^2$ for each pair.
- Then we add columns that are functions of the rolls. 
    - This is 'for free'. The rolls are the rolls, so the probabilities don't change.
    - If we know who we're calling, we know their responses. So we can add columns for $Y_1$,$Y_2$, and $Y_1+Y_2$.
- Then we *marginalize* to find the distribution of the sum. We'll do this in two steps.
    
    1. We *partially marginalize* over $J_1,J_2$, to find the joint distribution of $Y_1, Y_2$. 
    2. We *fully marginalize* the result to find the distribution of $Y_1+Y_2$.

### Step 1. Partial Marginalization

- To sum over pairs of responses, we sum over one response then the other.
    - This exactly like what we did for one response. Except twice.
    - And what we get is pretty similar too.
- The probability of observing the pair of responses is determined by their frequencies in the population.
    - It's the *product* of the frequences of those responses.
    - For 'Yes,Yes' it's $\theta_1^2$. For 'No,No' it's $\theta_0^2$. 
    - And for 'Yes,No' and 'No,Yes' it's $\theta_1\theta_0$.
- This is, for what it's worth, the *product* of the marginal probabilities of those responses. 
    - When this happens for any pair of random variables, we say they are *independent*. More on that soon.
- And the probability of any response pair $a,b$ depends only on its sum $a+b$.
      
$$ 
\begin{aligned}
\sum\limits_{j_1,j_2: y_{j_1},y_{j_2} = a,b} \frac{1}{m^2} \\
&= \sum\limits_{j_1 : y_{j_1}=1} \qty{ \sum\limits_{j_2 : y_j_2=b} \frac{\1}{m^2} } \\ 
&= \sum\limits_{j_1 : y_{j_1}=1} \times \qty{ m_b \times \frac{1}{m^2} } \qfor m_y = \sum\limits_{j:y_j=y} 1
&= m_a \times m_b \times \frac{1}{m} = \underset{\color{gray}\theta_a}{\frac{m_a}{m}}\underset{\color{gray}\theta_b}{\frac{m_b}{m}} = \theta_0^{s} \theta_1^{1-s} \qfor s = a+b
\end{aligned}
$$

### Step 2. Full Marginalization 

- To calculate the marginal probability that $Y_1+Y_2=s$, we sum the probabilities of all pairs with that sum.
    - This is pretty easy because there aren't too many pairs. 
    - And I've color coded the pairs to make it even easier.

:::

## Exercise: Two Calls to A Small Population 
```{r}
yex = c(0,0,1)
```

:::: columns 
::: {.column width="40%"}

$j$ | $y_j$
--|---
$1$ | $`r yex[1]`$
$2$ | $`r yex[2]`$
$3$ | $`r yex[3]`$

:::
::: {.column width="60%"}
```{r}
#| fig-height: 2.7
#| fig-width:  7
ggplot(data.frame(j=1:length(yex), y=yex[1:3])) + 
    geom_point(aes(x=j, y=y), size=5, alpha=.3) + 
    scale_y_continuous(breaks=c(0,1)) + 
    scale_x_continuous(breaks=1:length(yex))  
```

:::
::::

::: {.panel-tabset}
### Instructions

1. Make a table for the joint distribution of three rolls of a 3-sided die.  Add columns for the responses.
2. Partially marginalize to get the joint distribution of the responses. Add a column for their sum.
3. Marginalize to get the distribution of the sum.


::: {.footnotestyle} 
Aren't you happy that, even though you can't make a 3-sided die, I didn't ask you to use 4-sided one? 
:::

### Blank Space


### Solution

:::: columns

::: {.column width="40%"}

$p$ | $J_1$ | $J_2$ | $Y_1$ | $Y_2$ 
---|---|---|---|---
$\pollb 1/9$ |  $1$ | $1$ | $`r yex[1]`$ | $`r yex[1]`$
$\pollb 1/9$ |  $1$ | $2$ | $`r yex[1]`$ | $`r yex[2]`$
$\pollc 1/9$ |  $1$ | $3$ | $`r yex[1]`$ | $`r yex[3]`$
$\pollb 1/9$ |  $2$ | $1$ | $`r yex[2]`$ | $`r yex[1]`$
$\pollb 1/9$ |  $2$ | $2$ | $`r yex[2]`$ | $`r yex[2]`$
$\pollc 1/9$ |  $2$ | $3$ | $`r yex[2]`$ | $`r yex[3]`$
$\pollc 1/9$ |  $3$ | $1$ | $`r yex[3]`$ | $`r yex[1]`$
$\pollc 1/9$ |  $3$ | $2$ | $`r yex[3]`$ | $`r yex[2]`$
$\polld 1/9$ |  $3$ | $3$ | $`r yex[3]`$ | $`r yex[3]`$
:::

::: {.column width="30%"}

$p$ | $Y_1$ | $Y_2$ | $Y_1 + Y_2$
---|---|---|---
$\pollb \frac{4}{9}$ | $0$ | $0$ | $0$
$\pollc \frac{2}{9}$ | $0$ | $1$ | $1$
$\pollc \frac{2}{9}$ | $1$ | $0$ | $1$
$\polld \frac{1}{9}$ | $1$ | $1$ | $2$

:::
::: {.column width="30%"}

$p$ | $Y_1 + Y_2$
---|---
$\pollb \frac{4}{9}$ | $0$
$\pollc \frac{4}{9}$ | $1$
$\polld \frac{1}{9}$ | $2$

:::
::::
:::

## From Sampling to Coin Flip Sums: $n$ Responses

::: {.panel-tabset}
### Two-Step Approach
- To find the distribution of a sum $Y_1 + \ldots + Y_n$, we do the same thing.
- We start by writing out the joint distribution of $n$ dice rolls.
    - We're equally likely to roll any sequence of $n$ numbers in $1\ldots m$.
    - And there are $m^n$ sequences, so the probability is $1/m^n$ for each sequence.
    - Then we add columns for the corresponding responses 'for free'. And the sums.
- Then we marginalize in two steps.
    
    1. Sum over roll sequences leading to the same response sequence $a_1 \ldots a_n$.
    2. Sum over response sequences leading to the same sum $s=a_1+\ldots+a_n$.
    
### Step 1. Partial Marginalization

- To sum over sequences of responses, we sum over one response then the other.
    - This exactly like what we did for one response. Except over and over.
    - And what we get is pretty similar to the two-response case.
- The probability of observing the sequence is determined by the frequency of Yeses and Nos in the population.
    - It's the *product* of the frequences of those responses.
    - For 'Yes,Yes,...,Yes' it's $\theta_1^n$. For 'No,No,...,No' it's $\theta_0^n$. 
    - And if we have $s$ Yeses and therefore $n-s$ Nos, it's $\theta_1^s\theta_0^{n-s}$.
- This is, again, the *product* of the marginal probabilities of those $n$ responses.

$$
\begin{aligned}
\sum\limits_{\substack{j_1 \ldots j_n : \\ y_{j_1} \ldots y_{j_n} = a_1 \ldots a_n}} \frac{1}{m^n} 
&= \sum\limits_{\substack{j_1 : \\ y_{j_1}=a_1}} \qty[ \ldots \qty { \sum\limits_{\substack{j_{n-1} : \\ y_{j_{n-1}}=a_{n-1}}}  \qty( \sum\limits_{\substack{j_n : \\y_{j_n}=a_n}} \frac{1}{m^n} )}] \\
&= \sum\limits_{\substack{j_1 : \\ y_{j_1}=a_1}} \qty[ \ldots \qty{\sum\limits_{\substack{j_{n-1} : \\ y_{j_{n-1}}=a_{n-1}}}  \qty( m_{a_n} \times \frac{1}{m^n}) } ]\\
&= \sum\limits_{\substack{j_1 : \\ y_{j_1}=a_1}} \qty[ \ldots \qty{ m_{a_{n-1}} m_{a_n} \times \frac{1}{m^n} }] \\
&= m_{a_1} \ldots m_{a_n} \times \frac{1}{m^n}  = \theta_{a_i} \ldots \theta_{a_n} \\
&= \prod_{i:a_i=1} \theta_1 \prod_{i:a_i=0} \theta_0 = \theta_1^{s} \theta_0^{n-s} \qfor s = a_1 + \ldots + a_n
\end{aligned}
$$

### Step 2. Full Marginalization

- The full marginalization step is a bit trickier. But we can take advantage of our partial marginalization.
- To calculate the probability that the sum takes on the value $s$, we order our sum of probabilities thoughtfully.
    - We sum the over response sequences $a_1 \ldots a_n$ with the same number of Yeses and Nos.
    - And within that, sum over calls where we'd hear those responses.
- The inner sum we've done. That was our partial marginalization.
- And the outer sum boils down to counting the number of sequences with $s$ Yeses.
    - The probability of all these sequences is the same: $\theta_1^s\theta_0^{n-s}$. So summing is counting and multiplying.
    - You may have done the counting part in high school in a unit on 'permutations and combinations'.
    - In any case, that count has a name. It's spoken '$n$ choose $s$' and written $\binom{n}{s}$.

$$
\begin{aligned}
\sum_\limits_{\substack{j_1 \ldots j_n : \\ y_{j_1} + \ldots + y_{j_n} = s}} \frac{1}{m^n} 
&= \sum\limits_{\substack{a_1 \dots a_n : \\ a_1 + \ldots + a_n = s}} \sum_{y_1=a_1 \ldots y_n=a_n} \frac{1}{m^n} \\
& =\sum\limits_{a_1 \dots a_n : a_1 + \ldots + a_n = s} \theta_1^{s}\theta_0^{n-s} \\
&= \binom{s}{n} \theta_1^{s} \theta_0^{n-s} \qqtext{ where } \binom{s}{n} = \sum\limits_{a_1 \dots a_n : a_1 + \ldots + a_n = s} 1
\end{aligned} 
$$

### Using **R**

$$
\begin{aligned}
P(\sum_{i=1}^n Y_i = s) &=  \binom{s}{n} \theta_a^{m_a}\theta_b^{m_b} 
&\qqtext{ where } \binom{s}{n} \theta_1^{s}\theta_0^{n-s} &= \qqtext{ is the number of binary sequences $a_1 \ldots a_n$ summing to $s$.}
\end{aligned}
$$

- STUFF
- You probably did all this rearrangement-counting in high school at some point.
    - But it's not something we tend to do ourselves all that much.
    - **R** will do it for us.
- We can get the rearrangement count using the `choose` function: $\binom{s}{n}$ is `choose(s,n)`.
- We can get the whole probability using the `dbinom` function: $\binom{s}{n}\theta_1^s\theta_0^{n-s}$ is  
`dbinom(s, n, theta_1)`.
- And we can draw samples from this distribution using the `rbinom` function: `rbinom(1000, n, theta_1)` gives us a vector of 1000 samples.
  
:::

## The Binomial Distribution
```{r}

```


$$
P(\sum_{i=1}^n Y_i = s) &=  \binom{s}{n} \theta_a^{m_a}\theta_b^{m_b} 
$$

- These functions have 'binom' in their name because we call this distribution the *Binomial distribution*. 
    - The Binomial distribution on $n$ *trials* with *success probability* $\theta$ is our name for ...
    - the distribution of the sum of $n$ independent binary random variables with probability $\theta$ of being 1.
    - e.g., the distribution of the number of heads in $n$ coin flips is Binomial with $n$ trials and success probability $1/2$.
- And it's the sampling distribution of the *sum of responses*, $Y_1 + \ldots + Y_n$, when we sample with replacement ...
- ... from a population of binary responses $y_1 \ldots y_m$ in which $\theta$ is the frequency of ones.
- It's convenient that ...
    1. It depends only on one thing we don't know --- $\theta$ 
    2. And that's exactly the thing we're trying to estimate.

- This'll make it pretty easy to use for calibration. That's why it's what we're talking about first.

## Other Targets and Sampling Schemes


## Do I Need To Do All This?

- No.
- If you can work out what properties of the pop are relevant, you use simulation to estimate the sampling distn.
- And there are other methods for estimating the sampling distn that don't require you to know

# Calibration with the Binomial Distribution 


# Appendix

## The Binomial Distribution for 3 Flips {#three-flips}

::::
::: {.column width="60%"}
$p$ | $Y_1$ | $Y_2$ | $Y_3$ | $Y_1 + Y_2 + Y_3$
---|---|---|---|---
$\inpollb{\theta_0^3}$ | 0 | 0 | 0 | [0]{.pollb}
$\inpollc{\theta_1 \theta_0^2}$          | 0 | 0 | 1 | [1]{.pollc}
$\inpollc{\theta_1 \theta_0^2}$          | 0 | 1 | 0 | [1]{.pollc}
$\inpollc{\theta_1 \theta_0^2}$          | 1 | 0 | 0 | [1]{.pollc}
$\inpolld{\theta_1^2 \theta_0}$          | 0 | 1 | 1 | [2]{.polld}
$\inpolld{\theta_1^2 \theta_0}$          | 1 | 1 | 0 | [2]{.polld}
$\inpolld{\theta_1^2 \theta_0}$          | 1 | 0 | 1 | [2]{.polld}
$\inpolle{\theta_1^3 }$                  | 1 | 1 | 1 | [3]{.polle}

:::
::: {.column width="40%"}

$p$  | $Y_1 + Y_2 + Y_3$
---|---
$\inpollb{\theta_0^3}$              | [0]{.pollb}
$\inpollc{3\theta_1 \theta_0^2}$    | [1]{.pollc}
$\inpolld{3\theta_1^2 \theta_0}$    | [2]{.polld}
$\inpolle{\theta_1^3 }$             | [3]{.polle} 

:::
::::