---
title: "Lecture 13"
subtitle: "Causal Inference in Conditionally Randomized Experiments"
---


## An Example

{{< include shared-code.qmd >}}
{{< include ca-income.qmd >}}


```{r}
#| out-height: 400px
#| fig-asp: .4 
 
set.seed(1) 
m = 1e4 
pi = function(x) { 1/(1+exp(-(x-60)/8 )) }
mu = function(w, x) { 1 - (x-50)/30 + w*((x-50)/30)^2/2 }  

X = 50+round(30*runif(m))
W = 0 + (runif(m) <= pi(X))
epsilon0 = rnorm(m)
epsilon1 = rnorm(m)
Y0 = mu(0,X) + .25*epsilon0
Y1 = mu(1,X) + .25*epsilon1

pop = data.frame(X=X, W=W, Y=Y0*(1-W) + Y1*W, Y0=Y0, Y1=Y1)
counterfactual.pop = data.frame(X=X, W=W, Y=Y0*W + Y1*(1-W))  

n = 5000
data = pop[sample(nrow(pop), n, replace=TRUE),]

plotX = function(X,W) { X + .15*(2*W-1) }
pop.scatter = ggplot(pop) + 
  geom_point(aes(x=plotX(X,W), y=Y, color=factor(W)), alpha=.025, position=position_jitter(w=.15, h=0), size=.2) + 
  guides(color=FALSE)
counterfactual.pop.scatter = ggplot(counterfactual.pop) +
  geom_point(aes(x=plotX(X,W), y=Y, color=factor(W)), alpha=.025, position=position_jitter(w=.15, h=0), size=.2) + 
  guides(color=FALSE)
scatter = ggplot(data) + 
  geom_point(aes(x=plotX(X,W), y=Y, color=factor(W)), alpha=.1, position=position_jitter(w=.15, h=0)) + 
  guides(color=FALSE)

big.coords = coord_cartesian(xlim=c(50,80), ylim=c(-1.5,2.5))
narrow.coords = coord_cartesian(xlim=c(60,70), ylim=c(-1.5,2.5))
tight.coords = coord_cartesian(xlim=c(50,80), ylim=c(0,1.5))

annotated.pop.scatter = pop.scatter +  
  stat_summary(aes(x=plotX(X,W), y=Y, color=factor(W)), geom="pointrange", alpha=.8,
                 fun=mean, 
                 fun.min = function(x) mean(x) - sd(x), 
                 fun.max = function(x) mean(x) + sd(x))

annotated.counterfactual.pop.scatter = counterfactual.pop.scatter +  
  stat_summary(aes(x=plotX(X,W), y=Y, color=factor(W)), geom="pointrange", alpha=.8,
                 fun=mean, 
                 fun.min = function(x) mean(x) - sd(x), 
                 fun.max = function(x) mean(x) + sd(x))
annotated.pop.scatter + big.coords
```


- Researchers hypothesize that MHT increases bone density. 
- We'll think about how to look into this.
- Above, we've got a population of `r approx(m,1)` people eligible for some form of MHT.
  - Their age at the time the study begins is on the $x$-axis.
  - Bone density, measured 2 years later, is on the $y$-axis.
  - Color indicates whether they [did]{.twocolor-green} or [didn't]{.twocolor-red} do MHT.
- What we want to know is how well MHT works, on average, for the people who take it.
- That means comparing green folks---people who take MHT---to versions of themselves who didn't.

::: aside
To make things easier to look at, I'm drawing people who did and didn't receive treatment side-by-side. 
This shift isn't meaningful. Age is a round number: 60, 61, ...
:::


## Counterfactuals

```{r}
#| out-height: 400px
#| layout-ncol: 2 
#| fig-subcap:
#|   - "Real Population: Green People Did MHT, Red People Didn't"
#|   - "Counterfactual Population: Green People Didn't Do MHT, Red People Did" 
annotated.pop.scatter + big.coords
annotated.counterfactual.pop.scatter + big.coords
```

- We can't observe these *counterfactual* versions of people.
- But we can think about what the data would look like if we did.
  - On the left, we've got our real population.
  - On the right, we've got a hypothetical one where everyone who did MHT didn't and vice versa.
  - Note we're keeping the colors the same---counterfactual green people are still green.
- We can think of the average difference as the *average effect of MHT for those who took it*.

$$
\begin{aligned}
\frac{\sum_{j:w_j=1} y_j(1) - y_j(0)}{\sum_{j:w_j=1} 1} = 
\frac{\sum_{j:w_j=1} y_j}{\sum_{j:w_j=1} 1} - \textcolor{blue}{\frac{\sum_{j:w_j=1} y_j(0)}{\sum_{j:w_j=1} 1}}
\end{aligned}
$$

- We call this the *average treatment effect on the treated* (ATT). It's a difference of two means. 
- One is observable (if we have the population). [The other isn't]{.fg style="color:blue"}.


## Counterfactuals

```{r}
#| out-height: 400px
#| layout-ncol: 2 
#| fig-subcap:
#|   - "Real Population: Green People Did MHT, Red People Didn't"
#|   - "Counterfactual Population: Green People Didn't Do MHT, Red People Did" 

annotated.pop.scatter + big.coords
annotated.counterfactual.pop.scatter + big.coords
```

- We can't observe these *counterfactual* versions of people.
- But we can think about what the data would look like if we did.
  - On the left, we've got our real population.
  - On the right, we've got a hypothetical one where everyone who did MHT didn't and vice versa.
  - Note we're keeping the colors the same---counterfactual green people are still green.
- We can think of the average difference as the *average effect of MHT for those who took it*.

$$
\begin{aligned}
\frac{\sum_{j:w_j=1} y_j(1) - y_j(0)}{\sum_{j:w_j=1} 1} = 
\frac{\sum_{j:w_j=1} y_j}{\sum_{j:w_j=1} 1} - \textcolor{blue}{\frac{\sum_{j:w_j=1} y_j(0)}{\sum_{j:w_j=1} 1}}
\end{aligned}
$$

- We call this the *average treatment effect on the treated* (ATT). It's a difference of two means. 
- One is observable (if we have the population). [The other isn't]{.fg style="color:blue"}.

## Summarizing these Populations

```{r}
#| out-height: 400px
#| layout-ncol: 2 
#| fig-asp: .4 
#| fig-subcap:
#|   - "Real Population: Green People Did MHT, Red People Didn't"
#|   - "Counterfactual Population: Green People Didn't Do MHT, Red People Did" 
#| 
mean.shape=21

summarized.pop = pop.scatter + 
  stat_summary(aes(x=plotX(X,W), y=Y), geom="point", color='black', shape=mean.shape, fun=mean) + 
  stat_summary(aes(x=plotX(X,W), y=Y, color=factor(W)), fun=mean, geom="line") +
  geom_bar(aes(x=X, y=10*after_stat(prop), fill=factor(W), color=factor(W)), 
    position='identity', alpha=.1, linewidth=.2, data=pop) +
    guides(fill='none', color='none') 

summarized.counterfactual.pop = counterfactual.pop.scatter + 
  stat_summary(aes(x=plotX(X,W), y=Y), geom="point", color='black', shape=mean.shape, fun=mean) + 
  stat_summary(aes(x=plotX(X,W), y=Y, color=factor(W)), fun=mean, geom="line") +
  geom_bar(aes(x=X, y=10*after_stat(prop), fill=factor(W), color=factor(W)), 
    position='identity', alpha=.1, linewidth=.2, data=pop) + 
    guides(fill='none', color='none') 

summarized.pop + tight.coords
summarized.counterfactual.pop + tight.coords
```

- The scatter plot is hard to read. To help, I've overlaid some summaries.
  - Mean bone density for each age and treatment group.
  - Histograms of age for each treatment group.
- We can think of what we're doing in terms of these summaries. Here's the observable mean.

$$
\begin{aligned}
\text{observable mean} 
= \frac{\sum_{j:w_j=1} y_j}{\sum_{j:w_j=1} 1} 
= \sum_x \frac{\sum_{j:w_j=1, x_j=x} 1}{\sum_{j:w_j=1} 1} \frac{\sum_{j:w_j=1, x_j=x} y_j}{\sum_{j:w_j=1, x_j=x} 1} 
= \sum_x \ingreen{p(1,x)}\ingreen{\mu(1,x)}
\end{aligned}
$$

- It's the average of the green curve over the distribution of the green points---i.e., the green histogram.
- The counterfactual mean is the same thing, but in the counterfactual plot.


## Working With Summaries

```{r}
#| out-height: 400px
#| layout-ncol: 2 
#| fig-asp: .4 
#| fig-subcap:
#|   - "Real Population: Green People Did MHT, Red People Didn't"
#|   - "Counterfactual Population: Green People Didn't Do MHT, Red People Did" 
summarized.pop + 
  stat_summary(aes(x=plotX(X,W), y=Y, color=factor(W)), summary=mean, geom="line", linetype='dashed', data=counterfactual.pop) +
  tight.coords
summarized.counterfactual.pop + tight.coords
```


- This seems like it'd be a problem, since we don't actually see the counterfactual plot. 
- But, comparing our plots, we see something helpful. 
  - The green curve on the right looks like the red curve on the left.
  - To make this easier to see, I've drawn it in as a dashed line.
- We can use *what actually happened to the red people* to estimate *what would have happened to the green people*.

$$
\begin{aligned}
\text{observable mean} &= \sum_x \ingreen{p(1,x)}\ingreen{\mu(1,x)} \\
\text{counterfactual mean} &= \sum_x \ingreen{p(1,x)}\inred{\mu(0,x)}
\end{aligned}
$$

- In real data, there's no counterfactual plot, so we can't do this comparison visually.
- But *we can prove* that these curves look the same *in conditionally randomized experiments*.

## Sampling

```{r}
#| out-height: 400px
#| layout-ncol: 2 
#| fig-asp: .4 
summarized.scatter = scatter + 
  stat_summary(aes(x=plotX(X,W), y=Y), geom="point", color='black', shape=mean.shape, fun=mean) + 
  stat_summary(aes(x=plotX(X,W), y=Y, color=factor(W)), geom="line") +
  geom_bar(aes(x=X, y=10*after_stat(prop), fill=factor(W), color=factor(W)), 
    position='identity', alpha=.1, linewidth=.2)  +
    guides(fill='none', color='none')
  
summarized.scatter + tight.coords
summarized.pop + tight.coords
```

- The other part of our problem is the **statistical element**.
  - In reality, we don't have the whole population. 
  - We just have a sample of, say, `r approx(n,1)` people. 
- So we have to ask whether what we see in the sample is true of the population.

## Plan
  
- We're going to talk about how to answer these questions.
- You've seen the main ideas before in simpler contexts.
- All we've got to do is generalize them and put them together.

# Conditionally Randomized Experiments

## The Idea

```{r}
#| out-height: 500px
#| fig-asp: .4
#| 
ages = c(60,70)
twolevel.pop = pop[pop$X %in% ages, ]
twolevel.pop = twolevel.pop[1:200, ]

jitter.width = 1.5
plotX2 = function(W,X) { X + 1.5*(2*W-1) }
twolevel.pop$jX = twolevel.pop$X + jitter.width*runif(nrow(twolevel.pop), -1,1)
twolevel.pop$fW = factor(twolevel.pop$W)
w0 = twolevel.pop$fW[twolevel.pop$W==0][1]
w1 = twolevel.pop$fW[twolevel.pop$W==1][1]

twolevel.coord = coord_cartesian(ylim=c(-.5,1.5))
twolevel.pop.scatter = ggplot(twolevel.pop) + 
  geom_point(aes(x=plotX2(W,jX), y=Y, color=factor(W)), alpha=.5, size=3) + 
  guides(color='none') + scale_x_continuous(breaks=ages) 
twolevel.pop.scatter + twolevel.coord
```

- To make things simple, let's start by looking at a subset of our population.
  - We'll look at participants who were either `{r} min(ages)` or `{r} max(ages)` when the study began.
  - We'll ignore the others. This gives us two age groups---a binary covariate.
- Suppose that each person flips a coin to determine whether they do MHT.
  - The coins are weighted differently are different age groups.
  - That's why different proportions of people do MHT in the two groups.
- We call this a **conditionally randomized experiment**.
  - a.k.a. a **blocked** or **stratified** randomized experiment.

## Randomization and Sampling

:::: r-stack

::: {.fragment .fade-out fragment-index="0"}
```{r}
#| out-height: 500px
#| fig-asp: .4 

potential.outcome.plot = ggplot(twolevel.pop) + 
  geom_point(aes(x=plotX2(0,jX), y=Y0, color=w0), alpha=.5, size=3) + 
  geom_point(aes(x=plotX2(1,jX), y=Y1, color=w1), alpha=.5, size=3) + 
  geom_segment(aes(x=plotX2(0,jX), xend=plotX2(1,jX), y=Y0, yend=Y1), alpha=.05, linewidth=.5) +
  guides(color='none') + scale_x_continuous(breaks=ages) + twolevel.coord

population.plot = function(seed=1) {
  twolevel.pop$W =(\() rbinom(nrow(twolevel.pop), 1, pi(twolevel.pop$X))) |> with_seed(seed)
  twolevel.pop$fW = factor(twolevel.pop$W)
  twolevel.pop$Y = twolevel.pop$Y0*(1-twolevel.pop$W) + twolevel.pop$Y1*twolevel.pop$W
  ggplot(twolevel.pop) + 
    geom_point(aes(x=plotX2(0,jX), y=Y0, color=w0), alpha=.5, size=3, shape=1) + 
    geom_point(aes(x=plotX2(1,jX), y=Y1, color=w1), alpha=.5, size=3, shape=1) + 
    geom_segment(aes(x=plotX2(0,jX), xend=plotX2(1,jX), y=Y0, yend=Y1), alpha=.05, linewidth=.5) +
    geom_point(aes(x=plotX2(W,jX), y=Y, color=fW), alpha=.5, size=3) +  
    guides(color='none') + scale_x_continuous(breaks=ages) + twolevel.coord
}

sample.plot = function(randomization.seed=1, sample.seed=1) {  
  twolevel.pop$W =(\() rbinom(nrow(twolevel.pop), 1, pi(twolevel.pop$X))) |> with_seed(randomization.seed)
  twolevel.pop$fW = factor(twolevel.pop$W)
  twolevel.pop$Y = twolevel.pop$Y0*(1-twolevel.pop$W) + twolevel.pop$Y1*twolevel.pop$W

  sample.ind = (\() sample(nrow(twolevel.pop), 50, replace=TRUE)) |> with_seed(sample.seed)
  twolevel.pop$sampled = sapply(1:nrow(twolevel.pop), function(i) sum(sample.ind==i))

  ggplot(twolevel.pop) + 
    geom_point(aes(x=plotX2(0,jX), y=Y0, color=w0), alpha=.5, size=3, shape=1, data=twolevel.pop[twolevel.pop$sampled==0,]) + 
    geom_point(aes(x=plotX2(1,jX), y=Y1, color=w1), alpha=.5, size=3, shape=1, data=twolevel.pop[twolevel.pop$sampled==0,]) + 
    geom_segment(aes(x=plotX2(0,jX), xend=plotX2(1,jX), y=Y0, yend=Y1), alpha=.05, linewidth=.5) +
    geom_point(aes(x=plotX2(W,jX), y=Y, color=fW), alpha=.5, size=3, shape=13, data=twolevel.pop[twolevel.pop$sampled==0,]) +  
    geom_point(aes(x=plotX2(W,jX), y=Y, color=fW, size=sampled), alpha=.8) +  
    guides(color='none', size='none') + scale_x_continuous(breaks=ages) + scale_size_area() + twolevel.coord
  
}

potential.outcome.plot
```

:::
::: {.fragment .current-visible fragment-index="0"}

```{r}
#| out-height: 500px
#| fig-asp: .4 

population.plot(1)
```

:::
::: {.fragment .current-visible fragment-index="1"}

```{r}
#| out-height: 500px
#| fig-asp: .4 

population.plot(2)
```

:::
::: {.fragment .current-visible fragment-index="2"}

```{r}
#| out-height: 500px
#| fig-asp: .4 

population.plot(3)
```

:::
::: {.fragment .current-visible fragment-index="3"}

```{r}
#| out-height: 500px
#| fig-asp: .4 

sample.plot(3, 1)
```

:::
::: {.fragment .current-visible fragment-index="4"}
```{r}
#| out-height: 500px
#| fig-asp: .4 

sample.plot(3, 2)
```

:::
::: {.fragment .current-visible fragment-index="5"}
```{r}
#| out-height: 500px
#| fig-asp: .4 

sample.plot(3, 3)
```

:::
::::

::: {.fragment .semi-fade-out fragment-index="0"}
1. We start with two potential outcomes for each person in our population.
   - Those are the connected dots we see in the plot.
:::
::: {.fragment .fade-in fragment-index="0"}
::: {.fragment .semi-fade-out fragment-index="3"}
2. Each one flips a coin --- one with weight depending on their age --- to determine whether they're treated.
   - The winning outcomes --- the ones that actually happen --- are the ones that don't fade out.
   - This is, of course, random. So different this happen each time we do it.
:::
:::
::: {.fragment .fade-in fragment-index="3"}
3. Then we sample --- with replacement --- from our population. 
    - &olcross; marks a person who isn't sampled.
    - small dots mark people sampled once.
    - bigger dots mark people sampled 2,3,... times.
    - This is random, too.
:::

## What Happens to Means

:::: r-stack

::: {.fragment .fade-out fragment-index="0"}
```{r}
#| out-height: 500px
#| fig-asp: .4 

potential.outcome.plot + 
  stat_summary(aes(x=plotX2(0,X), y=Y0, color=w0), geom="pointrange", fun.data=mean_sd)
```

:::
::: {.fragment .current-visible fragment-index="0"}

```{r}
#| out-height: 500px
#| fig-asp: .4 

population.plot(1) 
```

:::
::: {.fragment .current-visible fragment-index="1"}

```{r}
#| out-height: 500px
#| fig-asp: .4 

population.plot(2)
```

:::
::: {.fragment .current-visible fragment-index="2"}

```{r}
#| out-height: 500px
#| fig-asp: .4 

population.plot(3)
```

:::
::: {.fragment .current-visible fragment-index="3"}

```{r}
#| out-height: 500px
#| fig-asp: .4 

sample.plot(3, 1)
```

:::
::: {.fragment .current-visible fragment-index="4"}
```{r}
#| out-height: 500px
#| fig-asp: .4 

sample.plot(3, 2)
```

:::
::: {.fragment .current-visible fragment-index="5"}
```{r}
#| out-height: 500px
#| fig-asp: .4 

sample.plot(3, 3)
```

:::
::::

::: {.fragment .semi-fade-out fragment-index="0"}
1. We start with two potential outcomes for each person in our population.
   - Those are the connected dots we see in the plot.
:::
::: {.fragment .fade-in fragment-index="0"}
::: {.fragment .semi-fade-out fragment-index="3"}
2. Each one flips a coin --- one with weight depending on their age --- to determine whether they're treated.
   - The winning outcomes --- the ones that actually happen --- are the ones that don't fade out.
   - This is, of course, random. So different this happen each time we do it.
:::
:::
::: {.fragment .fade-in fragment-index="3"}
3. Then we sample --- with replacement --- from our population. 
    - &olcross; marks a person who isn't sampled.
    - small dots mark people sampled once.
    - bigger dots mark people sampled 2,3,... times.
    - This is random, too.
:::





## Implications (TODO: IMPROVE/MAKE VISUAL)

$$
\begin{array}{cccc|ccc}
i & X_i & W_i & Y_i & Y_i(1) & Y_i(0) & Y_i(1)-Y_i(0) \\
\hline
1 & \textcolor{blue}{60} &1 & 70 & \textcolor{blue}{70} & \alert{70} & \alert{0} \\
2 & \textcolor{magenta}{70} &1 & 80 & \textcolor{magenta}{80} & \alert{70} & \alert{10} \\
3 & \textcolor{blue}{60} & 0 & 60 & \alert{60} & \textcolor{blue}{60} & \alert{0}\\
4 & \textcolor{magenta}{70} &0& 70 & \alert{80} & \textcolor{magenta}{70} & \alert{10} \\
5 & \textcolor{blue}{60} & 0&70 & \alert{70} & \textcolor{blue}{70} & \alert{0} \\
6 & \textcolor{blue}{60}& 0 & 80 & \alert{80} & \textcolor{blue}{80} & \alert{0} \\
\end{array}
$$

- For each age group, we have a randomized experiment. 
- Among those with $\textcolor{blue}{X=60}$ and $\textcolor{magenta}{X=70}$ ...
  - Everything we've said before about randomized experiments is true.
  - We can think of the people who do and don't do MHT as being drawn from the same population.
  - That means that --- if we ignore $W_i$ and $Y_i$ --- they have the same distribution.
  - We have to ignore this stuff to get idenitical distribution because it's *downstream of randomization*.
- This allows us to get at the *distribution of our missing data*. 
- [Remember red numbers are not observed]{.fg style="color:red"}.

## Filling in Missing Data

$$
\begin{array}{cccc|ccc}
i & X_i & W_i & Y_i & Y_i(1) & Y_i(0) & Y_i(1)-Y_i(0) \\
\hline
1 & \textcolor{blue}{60} &1 & 70 & \textcolor{blue}{70} & \alert{70} & \alert{0} \\
2 & \textcolor{magenta}{70} &1 & 80 & \textcolor{magenta}{80} & \alert{70} & \alert{10} \\
3 & \textcolor{blue}{60} & 0 & 60 & \alert{60} & \textcolor{blue}{60} & \alert{0}\\
4 & \textcolor{magenta}{70} &0& 70 & \alert{80} & \textcolor{magenta}{70} & \alert{10} \\
5 & \textcolor{blue}{60} & 0&70 & \alert{70} & \textcolor{blue}{70} & \alert{0} \\
6 & \textcolor{blue}{60}& 0 & 80 & \alert{80} & \textcolor{blue}{80} & \alert{0} \\
\end{array}
$$

What is the average $Y_i(1)$ among the participants with $X=60$?

::: fragment

$$
\frac{\textcolor{blue}{70} +\alert{60} +\alert{70} + \alert{80}}{\textcolor{blue}{1} + \alert{3}}  = 70
$$

:::


What is the average $Y_i(1)$ among the *treated* ($D=1$) participants with $X=60$?

::: fragment
$$
\frac{\textcolor{blue}{70} }{\textcolor{blue}{1}}  = 70
$$
:::

What is the average $Y_i(0)$ among the participants with $X=60$?

::: fragment
$$
\frac{\textcolor{red}{70} +\textcolor{blue}{60} +\textcolor{blue}{70}+\textcolor{blue}{80}}{4}  = 70
$$
:::


What is the average $Y_i(0)$ among the *untreated* ($D=0$) participants with $X=60$?

::: fragment
$$
\frac{\textcolor{blue}{60} +\textcolor{blue}{70}+\textcolor{blue}{80}}{3}  = 70
$$

:::

::: aside
This is a stylized example. Randomization implies equivalance *in distribution*, but not in any given sample.
:::


## Conditional ATE for $\textcolor{blue}{X=60}$

$$
\begin{array}{cccc|ccc}
i & X_i & W_i & Y_i & Y_i(1) & Y_i(0) & Y_i(1)-Y_i(0) \\
\hline
1 & \textcolor{blue}{60} &1 & 70 & \textcolor{blue}{70} & \alert{70} & \alert{0} \\
2 & \textcolor{magenta}{70} &1 & 80 & \textcolor{magenta}{80} & \alert{70} & \alert{10} \\
3 & \textcolor{blue}{60} & 0 & 60 & \alert{60} & \textcolor{blue}{60} & \alert{0}\\
4 & \textcolor{magenta}{70} &0& 70 & \alert{80} & \textcolor{magenta}{70} & \alert{10} \\
5 & \textcolor{blue}{60} & 0&70 & \alert{70} & \textcolor{blue}{70} & \alert{0} \\
6 & \textcolor{blue}{60}& 0 & 80 & \alert{80} & \textcolor{blue}{80} & \alert{0} \\
\end{array}
$$

- Because we have a randomized experiment within the  $\textcolor{blue}{X=60}$ group ...
- The difference in *subpopulation means* is equal to the average effect of treatment within that group.
- We call this the **conditional average treatment effect** ($CATE(60)$).
$$
CATE(60) = \E[Y_i(1) - Y_i(0) \mid X_i=0] = \E[Y_i \mid W_i=1, X_i=60] - \E[Y_i \mid W_i=0, X_i=60]
$$
- The difference in subsample means $\textcolor{blue}{X=60}$ is an unbiased estimator of 
this quantity.

$$
\begin{aligned}
\widehat{CATE}(60) &=\frac{\textcolor{blue}{70} }{1}  - \frac{\textcolor{blue}{60} +\textcolor{blue}{70}+\textcolor{blue}{80}}{3} \\
&= \frac{\textcolor{blue}{70} +\alert{60} +\alert{70} + \alert{80}}{4} - \frac{\textcolor{red}{70} +\textcolor{blue}{60} +\textcolor{blue}{70}+\textcolor{blue}{80}}{4} \\
&= \frac{\alert{0} +\alert{0} +\alert{0} + \alert{0}}{4}
\end{aligned}
$$



## Conditional ATE for $\textcolor{magenta}{X=70}$

$$
\begin{array}{cccc|ccc}
i & X_i & W_i & Y_i & Y_i(1) & Y_i(0) & Y_i(1)-Y_i(0) \\
\hline
1 & \textcolor{blue}{60} &1 & 70 & \textcolor{blue}{70} & \alert{70} & \alert{0} \\
2 & \textcolor{magenta}{70} &1 & 80 & \textcolor{magenta}{80} & \alert{70} & \alert{10} \\
3 & \textcolor{blue}{60} & 0 & 60 & \alert{60} & \textcolor{blue}{60} & \alert{0}\\
4 & \textcolor{magenta}{70} &0& 70 & \alert{80} & \textcolor{magenta}{70} & \alert{10} \\
5 & \textcolor{blue}{60} & 0&70 & \alert{70} & \textcolor{blue}{70} & \alert{0} \\
6 & \textcolor{blue}{60}& 0 & 80 & \alert{80} & \textcolor{blue}{80} & \alert{0} \\
\end{array}
$$

- Because we have a randomized experiment within the  $\textcolor{magenta}{X=70}$ group ...
- The difference in *subpopulation means* is equal to the average effect of treatment within that group.
- We call this the **conditional average treatment effect** ($CATE(70)$).
$$
CATE(70) = \E[Y_i(1) - Y_i(0) \mid X_i=0] = \E[Y_i \mid W_i=1, X_i=70] - \E[Y_i \mid W_i=0, X_i=70]
$$
- The difference in subsample means $\textcolor{magenta}{X=70}$ is an unbiased estimator of 
this quantity.

$$
\begin{aligned}
\widehat{CATE}(70) &= \frac{\textcolor{magenta}{80} }{1}  - \frac{\textcolor{magenta}{70} }{1} \\
&=\frac{\textcolor{magenta}{80} +\alert{80} }{2} - \frac{\alert{70}  + \textcolor{magenta}{70}}{2} \\
&= \frac{\alert{10} +\alert{10}}{2}
\end{aligned}
$$

## The ATE

$$
\begin{array}{cccc|ccc}
i & X_i & W_i & Y_i & Y_i(1) & Y_i(0) & Y_i(1)-Y_i(0) \\
\hline
1 & \textcolor{blue}{60} &1 & 70 & \textcolor{blue}{70} & \alert{70} & \alert{0} \\
2 & \textcolor{magenta}{70} &1 & 80 & \textcolor{magenta}{80} & \alert{70} & \alert{10} \\
3 & \textcolor{blue}{60} & 0 & 60 & \alert{60} & \textcolor{blue}{60} & \alert{0}\\
4 & \textcolor{magenta}{70} &0& 70 & \alert{80} & \textcolor{magenta}{70} & \alert{10} \\
5 & \textcolor{blue}{60} & 0&70 & \alert{70} & \textcolor{blue}{70} & \alert{0} \\
6 & \textcolor{blue}{60}& 0 & 80 & \alert{80} & \textcolor{blue}{80} & \alert{0} \\
\end{array}
$$

- To get an *aggregate summary* of the effect of MHT, we can average the two within-subpopulation effects.
$$
\begin{aligned}
\text{ATE} 
&= \E\qty[ \E\qty{ Y_i(1) - Y_i(0) \mid X_i } ] \\
&= \E\qty[ \E\qty{ Y_i \mid W_i=1, X_i } - \E\qty{ Y_i \mid W_i=0, X_i } ] \\
&= \E\qty[ CATE(X_i) ] 
\end{aligned}
$$

- We can estimate this with a weighted average of our two estimates.
$$
\begin{aligned}
\widehat{ATE} &= \textcolor{blue}{\frac{4}{6}} \widehat{CATE}(60) + \textcolor{magenta}{\frac{2}{6}} \widehat{CATE}(70) \\
&=\textcolor{blue}{\frac{4}{6}} \cdot \left\{\frac{70}{1} -  \frac{60+70+80}{3}\right\} + \textcolor{magenta}{\frac{2}{6}} \cdot \left\{\frac{80}{1} -  \frac{70}{1}\right\} \\
&= \frac{20}{6}
\end{aligned}
$$

## The ATT and ATC

$$
\begin{array}{cccc|ccc}
i & X_i & W_i & Y_i & Y_i(1) & Y_i(0) & Y_i(1)-Y_i(0) \\
\hline
1 & \textcolor{blue}{60} &1 & 70 & \textcolor{blue}{70} & \alert{70} & \alert{0} \\
2 & \textcolor{magenta}{70} &1 & 80 & \textcolor{magenta}{80} & \alert{70} & \alert{10} \\
3 & \textcolor{blue}{60} & 0 & 60 & \alert{60} & \textcolor{blue}{60} & \alert{0}\\
4 & \textcolor{magenta}{70} &0& 70 & \alert{80} & \textcolor{magenta}{70} & \alert{10} \\
5 & \textcolor{blue}{60} & 0&70 & \alert{70} & \textcolor{blue}{70} & \alert{0} \\
6 & \textcolor{blue}{60}& 0 & 80 & \alert{80} & \textcolor{blue}{80} & \alert{0} \\
\end{array}
$$

- We can, if we prefer, average the CATE over different distributions of units.
- This changes our weights. And therefore gives us different summaries of the effect.

:::: columns
::: column

$$
\begin{aligned}
\text{ATT} 
&= \E\qty[ \E\qty{ Y_i(1) - Y_i(0) \mid X_i } \mid W_i = 1] \\
&= \E\qty[ \E\qty{ Y_i \mid W_i=1, X_i } - \E\qty{ Y_i \mid W_i=0, X_i } \mid W_i=1] \\
&= \E\qty[ CATE(X_i) \mid W_i=1] \\
& \\
\widehat{ATT} &= \textcolor{blue}{\frac{1}{2}} \widehat{CATE}(60) + \textcolor{magenta}{\frac{1}{2}} \widehat{CATE}(70) \\
&=\textcolor{blue}{\frac{1}{2}} \cdot \left\{\frac{70}{1} -  \frac{60+70+80}{1}\right\} + \textcolor{magenta}{\frac{1}{2}} \cdot \left\{\frac{80}{1} -  \frac{70}{1}\right\} \\
&= \frac{10}{2}
\end{aligned}
$$

:::
::: column

$$
\begin{aligned}
\text{ATC} 
&= \E\qty[ \E\qty{ Y_i(1) - Y_i(0) \mid X_i } \mid W_i = 0] \\
&= \E\qty[ \E\qty{ Y_i \mid W_i=0, X_i } - \E\qty{ Y_i \mid W_i=0, X_i } \mid W_i=0] \\
&= \E\qty[ CATE(X_i) \mid W_i=0] \\
& \\
\widehat{ATC} &= \textcolor{blue}{\frac{3}{4}} \widehat{CATE}(60) + \textcolor{magenta}{\frac{1}{4}} \widehat{CATE}(70) \\
&=\textcolor{blue}{\frac{3}{4}} \cdot \left\{\frac{70}{1} -  \frac{60+70+80}{1}\right\} + \textcolor{magenta}{\frac{1}{4}} \cdot \left\{\frac{80}{1} -  \frac{70}{1}\right\} \\
&= \frac{10}{4}
\end{aligned}
$$

:::
::::

## Working With All the Data
```{r}
#| out-height: 400px
#| layout-ncol: 2 
#| fig-asp: .4 
summarized.pop
summarized.counterfactual.pop
```

- This formalizes what we were talking about earlier.
- It tells us we can use subpopulation means *that we can estimate* in place of *counterfactual means* we (otherwise) cannot.

$$
\begin{aligned}
\E[Y(1) \mid W_i=1] &=  \E[ \E[Y_i \mid W_i=1, X_i]] = \sum_x \ingreen{p(1,x)\mu(1,x)}  \approx \sum_x \ingreen{\hat p(1,x) \hat\mu(1,x)} \\
\E[Y(0) \mid W_i=1] &=  \E[ \E[Y_i \mid W_i=0, X_i]] = \sum_x \ingreen{p(1,x)}\inred{\mu(0,x)} \approx \sum_x \ingreen{\hat p(1,x)} \inred{\hat\mu(0,x)} \\
\text{ATT} &= \sum_{x} \ingreen{p(1,x)}\qty{ \ingreen{\mu(1,x)} - \inred{\mu(0,x)} } \approx \sum_{x} \ingreen{\hat p(1,x)}\qty{ \ingreen{\hat\mu(1,x)} - \inred{\hat\mu(0,x)} } && \text{ and similarly } \\
\text{ATC} &= \sum_{x} \inred{p(0,x)}\qty{ \ingreen{\mu(1,x)} - \inred{\mu(0,x)} } \approx \sum_{x} \ingreen{\hat p(1,x)}\qty{ \ingreen{\hat\mu(1,x)} - \inred{\hat\mu(0,x)} }.
\end{aligned}
$$

- So the adjusted gap we were talking about last time *is* the ATC.
- Given randomization. And we can calculate an estimate pretty easily. 

```{r}
x = ages
Y = data$Y
X = data$X
W = data$W
```

```{r}
#| echo: true
mu.hat = function(w,x) { mean(Y[X==x & W==w]) }         |> Vectorize()
p.hat  = function(w,x) { sum(X==x & W==w) / sum(W==w) } |> Vectorize()
ATC.hat = sum(p.hat(0,x)*(mu.hat(1,x) - mu.hat(0,x)))
ATT.hat = sum(p.hat(1,x)*(mu.hat(1,x) - mu.hat(0,x)))
```

- The ATT and ATC are `{r} approx(ATT.hat)` and `{r} approx(ATC.hat)` respectively.







