---
title: "Lecture 4"
subtitle: "Calibrating Interval Estimates using the Bootstrap"
---

# Review

## Point and Interval Estimation

{{< include shared-code.qmd >}}
{{< include ga-turnout.qmd >}}

```{r}
#| fig-asp: .3 

set.seed(default.seed)
Js = array(dim=c(10000, n))
Ys = array(dim=c(10000, n))
for(rr in 1:10000) {
  Js[rr,] = sample(m, n, replace=TRUE)
  Ys[rr,] = y[Js[rr,]]
}
meanY.samples = rowMeans(Ys)
pmf.height = max(dbinom(0:n, n, mean(y))) 

scale_x = scale_x_continuous(breaks=seq(.6,.8, by=.025)) 
scale_y = scale_y_continuous(breaks=seq(0, pmf.height,by=.01))

bar.plot = ggplot() + geom_bar(aes(x=meanY.samples, y=after_stat(prop)), alpha=.3) +
                      geom_vline(aes(xintercept=mean(y)), color='green', size=2, alpha=.8) + 
                      scale_x + scale_y
                      
bar.plot + with_annotations(meanY.samples, interval.width=width(meanY.samples), height=pmf.height) + 
           coord_cartesian(xlim=c(.6,.8), ylim=c(0, pmf.height))
```

- In last week, we talked about estimating population frequencies.
- Our **point estimates** were frequencies in a random sample drawn from the population. 
- To characterize our uncertaintly, we used **interval estimates**.  
  - In particular, *calibrated* interval estimates. Confidence intervals.  
  - To get these intervals, we add 'arms' of equal length to our point estimate.
  - We choose the arm length so that, in 95% of surveys done like ours, the estimation target is within them.

## Calibration of Interval Estimates
```{r}
#| fig-asp: .3

bar.plot + with_annotations(meanY.samples, interval.width=width(meanY.samples), 
                                          shade.width=width(meanY.samples), height=pmf.height) +
           coord_cartesian(xlim=c(.6,.8), ylim=c(0, pmf.height))
```

- We can do that using the **sampling distribution** of our point estimate.
- We'd draw arms out from the sampling distribution's until they span 95% of point estimates.
- We can see that this is the right length by a shift of perspective.
  - A point estimate can touch the mean with its arms iff^[if and only if] the mean can touch it with equally long arms.
  - So 95% of intervals cover if and only if 95% of point estimates are within the mean's arms.
- In practice, we can't do exactly that. We don't know the sampling distribution.
- But we can use an estimate of the sampling distribution in its place. 
- If it's a good one, we'll get approximately 95% coverage.

## A Parametric Estimate of the Sampling Distribution 
```{r}
#| label: sampling-dist-comparison-binomial
#| fig-width: 7
#| fig-asp: .7
#| layout-ncol: 2 
#| fig-cap:
#| - Our point estimate and its sampling distribution
#| - Our estimate of its sampling distribution 

estimated.dist.samples = rbinom(10000, n, mean(Y)) / n 
bar.plot.estimated = ggplot() + 
  geom_bar(aes(x=estimated.dist.samples, y=after_stat(prop)), alpha=.3) +
  geom_vline(aes(xintercept=mean(y)), color='green', size=2, alpha=.7) + 
  scale_x + scale_y 

xx = seq(0,n)
p.parametric.truth = bar.plot + 
  with_annotations(meanY.samples, interval.width=width(meanY.samples), highlight.dots=1, height=pmf.height) +
  geom_line(aes(x=xx/n, y=dbinom(xx, n, mean(y))), color='black', alpha=.6) 
p.parametric.estimate = bar.plot.estimated  +
  with_annotations(c(mean(Y), estimated.dist.samples), highlight.dots=1, height=pmf.height,
                   shade.width=width(estimated.dist.samples), shade.color='red') +
  geom_line(aes(x=xx/n, y=dbinom(xx, n, mean(Y))), color='black') 

coord  = coord_flip(xlim=c(.6,.8), ylim=c(0, pmf.height))
p.parametric.truth + coord 
p.parametric.estimate + coord
```

 - Our estimate was based on knowledge of the **parametric form** of the sampling distribution.
 - When we sample with replacement, it's **binomial**.
    - It's the distribution of the frequency of heads in `r n` flips of a coin with probability $\theta$
    - where $\theta$ is the 'frequency of heads' in the population.
    - e.g. $\theta\approx `r approx(mean(y))`$ is the proportion of **registered voters** who will vote in our turnout example.

 $$
 P_\theta(\hat\theta = t) = \binom{n}{nt} \theta^{nt} (1-\theta)^{n(1-t)} \qqtext{ is the probability of the sample frequency being $t$}
 $$


- By *parametric form*, I mean a formula in terms of some parameters. 
  - That tells us what parameters we need to estimate to estimate the sampling distribution.
  - In this case, the only (unknown) parameter is the $\theta$, the frequency of heads in the population.
  - We plugged in the sample frequency, $\hat\theta$, to get our estimate of the sampling distribution.

$$
\hat P_\theta(\hat\theta = t) = \binom{n}{nt} \hat\theta^{nt} (1-\hat\theta)^{n(1-t)} \qqtext{ is our estimate of the same thing}
$$


## Sampling without Replacement

```{r}
#| label: sampling-dist-comparison-hypergeometric
#| fig-width: 7
#| fig-asp: .7
#| layout-ncol: 2 
#| fig-cap:
#| - Our point estimate and its sampling distribution
#| - Our estimate of its sampling distribution 

mh = 200
theta = 1/2
nh = 100
dist.samples = rhyper(10000, mh*theta, mh*(1-theta), nh)/ nh 
theta.hat = dist.samples[1]
estimated.dist.samples = rhyper(10000, round(mh*theta.hat), round(mh*(1-theta.hat)), nh) / nh

pmf.height.hyper = max(dhyper(0:n, mh*theta, mh*(1-theta), nh))
scale_x = scale_x_continuous(breaks=seq(.6,.8, by=.025)) 
scale_y = scale_y_continuous(breaks=seq(0, pmf.height.hyper,by=.01))
bar.plot.hyper = ggplot() + geom_bar(aes(x=dist.samples, y=after_stat(prop)), alpha=.3) +
                      geom_vline(aes(xintercept=theta), color='green', size=1.5, alpha=.7) + 
                      scale_x + scale_y 

bar.plot.estimated = ggplot() + 
  geom_bar(aes(x=estimated.dist.samples, y=after_stat(prop)), alpha=.3) +
  geom_vline(aes(xintercept=theta), color='green', size=2, alpha=.7) + 
  scale_x + scale_y 

xx = seq(0,nh)
p.parametric.truth = bar.plot.hyper + 
  with_annotations(dist.samples, interval.width=width(dist.samples), highlight.dots=1, height=pmf.height.hyper) +
  geom_line(aes(x=xx/nh, y=dhyper(xx, mh*theta, mh*(1-theta), nh)), color='black', alpha=.6) 
p.parametric.estimate = bar.plot.estimated  +
  with_annotations(c(theta.hat, estimated.dist.samples), highlight.dots=1, height=pmf.height.hyper,
                   shade.width=width(estimated.dist.samples), shade.color='red') +
  geom_line(aes(x=xx/nh, y=dhyper(xx, mh*theta.hat, mh*(1-theta.hat), nh)), color='black') 

coord  = coord_flip(xlim=c(.35,.65), ylim=c(0, pmf.height.hyper))
p.parametric.truth + coord 
p.parametric.estimate + coord
```

- We can do the same thing when we sample *without replacement*.
- We just have to use a different parametric form: the **hypergeometric distribution**.

$$
 P_\theta(\hat\theta = t) =  \binom{n}{nt} \frac{\{m(1-\theta)\}!}{\{m(1-\theta)-n(1-t)\}!} \times \frac{(m\theta)!}{(m\theta-nt)!} \times \frac{(m-n)!}{m!} 
$$
 

- Again, the only unknown parameter is $\theta$. And we can plug in our point estimate to estimate this distribution.

$$
 \hat P_\theta(\hat\theta = t) =  \binom{n}{nt} \frac{\{m(1-\hat\theta)\}!}{\{m(1-\hat\theta)-n(1-t)\}!} \times \frac{(m\hat\theta)!}{(m\hat\theta-nt)!} \times \frac{(m-n)!}{m!} 
$$

# An Exercise

## A Survey
```{webr-r}
n=6
Y = c(0,0,0,1,1,1)

ggplot() + geom_point(aes(x=1:6, y=Y))
```

::: {.panel-tabset}
### The Idea
- We're going to run a survey on Candy Preferences.
  - I'll draw a sample of size $n=6$, with replacement, from the people in the room.
  - They'll pick candy and write their selection into a sample table on the board.
  - In particular, we'll write out whether they chose Chocolate ($Y=1$) or sour candy ($Y=0$).
- Then we're going to estimate the sampling distribution in a new way.
  - It'll be a lot like calculating the actual sampling distribution.
  - We'll draw a sample of size $n$ with replacement, use it to calculate our estimator, and repeat.
- What's different is the population we're drawing our sample from.
- We don't have responses from the whole population, so we'll draw it from the closest thing we've got: the sample.
  - We call this *bootstrapping*. 
  - The estimate of the sampling distribution we get is called the *bootstrap sampling distribution*.

### Instructions

- I've drawn the sample we got from our survey on the board.
- Here's how you draw a sample from the bootstrap sampling distribution of our point estimator.
  1. Roll your die 6 times to draw a sample of size $n=6$ *from the sample*.
  2. Calculate the sample frequency. That's one draw from the bootstrap sampling distribution. Write it down.
- If we all do this 5 times, we'll have a bunch of draws.
- We'll tally them up on the board and visualize the result as a histogram.
- A histogram of draws from the *bootstrap sampling distribution*.

:::

## Simulating a Few More Draws

```{webr-r}
draws = 1:10000 |> map_vec(function(.) { 
  Jstar = sample(1:n, n, replace=TRUE)
  Ystar = Y[Jstar]
  mean(Ystar)
})
computer.tally = as.data.frame(table(draws)) |>
  rename(theta.hat=draws, count=Freq)
```

## Discussion

```{webr-r}
hand.tally = data.frame(theta.hat = c(0,1,2,3,4,5,6), 
                        count     = c(0,0,0,1,0,0,0))

tally=computer.tally
ggplot() + geom_col(aes(x=(0:n)/n, y=dbinom(0:n, n, mean(Y))), color='blue', fill='blue', linewidth=.1, alpha=.2) + 
           geom_col(aes(x=(0:n)/n, y=tally$count/sum(tally$count)), fill='red', color='red', linewidth=.1, alpha=.2)
```

::: {.panel-tabset}
### Success!
- If all has gone to plan, [our histogram]{.fg style="color:red"} looks a lot like our [binomial estimate]{.fg style="color:blue"} of the sampling distribition.
  - If we substitute in a computer tally of 10,000 draws from the bootstrap sampling distribution, we nail it.
  - It looks like the bootstrap sampling distribution *is* the binomial estimate. 
- **Q**. It is. How do you know?

### A Big Hint.
  - We know the distribution of the frequency of 1s in a sample of size $n$ drawn with replacement ...
    - ... from a population $y_1\ldots y_m$ of binary responses with frequency $\theta$.
  - It's $\text{Binomial}(n, \theta)$. To estimate it, we plug in $\hat\theta$, the frequency of 1s in our sample. 
  - So our Binomial estimate is the distribution of the frequency of 1s in a sample of size $n$ drawn with replacement ...
    - ... from 'a population' $Y_1 \ldots Y_n$ with frequency $\hat\theta$.
  - What's a sample from the bootstrap sampling distribution?

### Answer.

- Each bootstrap sample is the frequency of 1s in a sample of size $n$ drawn with replacement ...
  - ... from  'a population' $Y_1 \ldots Y_n$ in which the frequency of 1s is $\hat\theta$,  the frequency of 1s in our sample. 
  - Because that 'population' is our sample.

:::

# The Bootstrap

## The Bootstrap Interpretation in Our Turnout Poll
```{r}
#| fig-asp: .8 
#| fig-width: 4.5
#| layout-ncol: 3

boot.ind = sample(1:n, n, replace=TRUE)

displaypop = pop[1:50000,]
pop$sam.replicated   = as.numeric(table(c(1:m, Js[1,])) - 1)
pop$boot.replicated  = as.numeric(table(c(1:m, Js[1, boot.ind])) - 1)

pop.map  = state.plot + geom_point(aes(x = long, y = lat), 
                                   alpha=.05, data=displaypop, size=.025, color=polla) + 
                        scale_size_area() + guides(size='none')
sam.map  = state.plot + geom_point(aes(x = long, y = lat, size=sam.replicated),   
                                   alpha=.2, data=pop[pop$sam.replicated > 0, ], color=polla) + 
                        scale_size_area(max_size=3) + guides(size='none')
boot.map = state.plot + geom_point(aes(x = long, y = lat, size=boot.replicated),   
                                   alpha=.2, data=pop[pop$boot.replicated > 0, ], color=polla)   +
                        scale_size_area(max_size=6) + guides(size='none')

pop.map
sam.map
boot.map
```


:::: r-stack
::: {.fragment .fade-out data-fragment-index="0"}

- The sampling distribution estimate we've used was $\text{Binomial}(n,\hat\theta)$ for $n=`r n`$ and $\hat\theta \approx `r approx(mean(Y))`$.
  - It's the distribution of the proportion heads in `r n` flips of a coin with probability $\hat\theta \approx `r approx(mean(Y))`$ of heads.
  - where $\hat\theta \approx `r approx(mean(Y))`$ is the proportion of **voters we've polled** who will vote. 
- That is, it's the sampling distribution of a 'poll' of the people in our sample, i.e.
    - roll a 625-sided die 625 times
    - call up the corresponding person in our sample
    - and counting up the yeses we hear
- Note that this is random because we're drawing with replacement.
  - Each time we run this poll, we call each person in our sample 0,1,2,... times
  - And the number of times we call them is random.

:::
::: {.fragment .current-visible data-fragment-index="0"}

- If we plot our voters on a map, you can see the idea in visual terms.
    - On the left, we have the population. 
    - In the middle, we have our sample. It's drawn, with replacement, from the population.
    - On the right, we have something else. A new sample drawn, with replacement, *from the sample*.
- Each 'call' that a person receives increases the size of their dot: $\text{circle area} \propto \text{number of calls}$.
- In the sample, even though it's drawn with replacement, all dots are the same size.
  - Because we draw from such a large population, nobody gets called twice.
- In the bootstrap sample, dots vary in size.
  - Because we draw $n$ people from a sample of size $n$, it's almost impossible not to call somebody twice. 

:::
::::

## Bootstrapping

- Before the election, we don't observe the population. But we do observe the sample. 
  - And we can *sample from the sample* act as if it were the population.
  - We'll take repeated random samples of size `r n`, with replacement, from our sample of size `r n`.
- We call these *bootstrap samples* and estimates based on them *bootstrap estimates*.
  - The distribution of these estimates is called the *bootstrap sampling distribution*.
  - If the sample is like the population, this should be like our estimator's actual sampling distribution.

```{r bootstrap.sample}
J.star = sample(n, size=n, replace=TRUE)
Y.star = Y[J.star]
```

:::: columns
::: column
The Sample
$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}  \\
Y_i  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$

The Bootstrap Sample

$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}^*  \\
Y_i^*  &  `r Y.star[1]` & `r Y.star[2]` & \dots & `r Y.star[n]` & `r approx(mean(Y.star))` \\
\end{array}
$$
:::

::: column
The Population

$$
\begin{array}{r|rrrr|r}
j & 1 & 2 & \dots &   `r approx(m)` & \bar{y}_{`r approx(m)`}  \\
y_{j} & `r y[1]` & `r y[2]` & \dots & `r y[m]`   & `r approx(mean(y))`  \\
\end{array}
$$

The 'Bootstrap Population' --- The Sample
$$
\begin{array}{r|rrrr|r}
j        &   1 & 2 & \dots & `r n` & \bar{y}^*_{`r n`}  \\
y_j^*  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$

:::
::::

::: aside
Notation. We use stars to distinguish the bootstrap sample from our original sample---we write $Y_i^*$ and $\bar Y^*$.
:::

## The Bootstrap is *Nonparametric* 

```{r}
#| echo: true

bootstrap.samples = array(dim=10000)
for(rr in 1:10000) {
    Y.star = Y[sample(1:n, n, replace=TRUE)]
    bootstrap.samples[rr] = mean(Y.star) 
}
```

```{r}
#| out-height: 400px 
#| fig-asp: .3 
ggplot() + geom_bar(aes(x=bootstrap.samples, y=after_stat(prop)), alpha=.3, position='identity') + 
           geom_line(aes(x=(0:n)/n, y=dbinom(0:n, n, mean(Y))), color='black') + 
           xlim(.6,.8)
```

- We do not need to know the parametric form of our sampling distribution to use it.
- All we do is re-run our poll *acting as if* our sample were the population.
  - We can do this no matter what we're estimating. 
  - So let's try it out on some stuff where we don't know the parametric form.
  
  
## Comparing Black and Non-Black Turnout

$$
\small{
\begin{array}{r|rr|rr|r|rr|rrrrr}
\text{call} &  1  & & 2  & & \dots & `r n`  & & & & & & \\
\text{question}  & X_1 & Y_1 &  X_2 & Y_2 & \dots &  X_{`r n`} & Y_{`r n`} & \overline{X}_{`r n`} & \overline{Y}_{`r n`} &\frac{\sum_{i:X_i=0} Y_i}{\sum_{i:X_i=0} 1} & \frac{\sum_{i:X_i=1} Y_i}{\sum_{i:X_i=1} 1} & \text{difference} \\
\text{outcome}    &  \underset{\textcolor{gray}{x_{`r J[1]`}}}{`r X[1]`} & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & \underset{\textcolor{gray}{x_{`r J[2]`}}}{`r X[2]`} & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`}
                  & \dots 
                  & \underset{\textcolor{gray}{x_{`r J[n]`}}}{`r X[n]`} & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`}
                  & `r approx(mean(X))` & `r approx(mean(Y))` & `r approx(mean(Y[X==0]))` & `r approx(mean(Y[X==1]))` 
                  & `r approx(mean(Y[X==1])-mean(Y[X==0]))` \\
\end{array}
}
$$


:::: r-stack
::: {.fragment .fade-out data-fragment-index="0"}

- It's useful to predict turnout among Black voters because they tend to vote differently than non-Black voters.
- Let's suppose that we're interested in the difference in turnout between Black and non-Black voters.
- This is a bit reductive, but it's the beginning of the semester, so we're keeping things simple. 
- We know from the voter file that `r percent(mean(x==1))` of registered voters are Black.
- Let's suppose we asked the people we polled if they were Black, recording their answer in a *covariate* $X_i$.
- And we found that...
  - `r sum(X==1)` of the people we called (`r percent(mean(X==1))`) were Black with a turnout rate of `r approx(mean(Y[X==1]))`
  - `r sum(X==0)` of the people we called (`r percent(mean(X==0))`) were non-Black with a turnout rate of `r approx(mean(Y[X==0]))`  
- So we'd estimate the *difference* to be $`r approx(mean(Y[X==1]))`-`r approx(mean(Y[X==0]))` \approx  `r approx(mean(Y[X==1])- mean(Y[X==0]))`$. 
  
:::
::: {.fragment .current-visible data-fragment-index="0"}

- After the election, we found that ...
  - Turnout among Black registered voters was `r approx(mean(y[x==1]))` 
  - Turnout among non-Black registered voters was `r approx(mean(y[x==0]))`.
- The actual difference was $`r approx(mean(y[x==1]))`-`r approx(mean(y[x==0]))` \approx  `r approx(mean(y[x==1]) - mean(y[x==0]))`$.
- Depending on what you're doing, that point estimate of `r approx(mean(Y[X==1])- mean(Y[X==0]))` may or may not have been accurate enough.
- It would've been nice to have a confidence interval to tell us what kind of precision we could expect.
- For that, we'll need to estimate the sampling distribution of this difference.

:::
::::

## A Table of Imaginary Polls

$$
\begin{array}{r|rr|rr|r|rr|rrrr}
\text{call} &  1 & & 2 & &  \dots & `r n` & & & & & & \\
\text{poll} &  X_1 & Y_1 &  X_2 & Y_2   & \dots &  X_{`r n`} & Y_{`r n`}  & \overline{X} & \overline{Y}  &\frac{\sum_{i:X_i=0} Y_i}{\sum_{i:X_i=0} 1} & \frac{\sum_{i:X_i=1} Y_i}{\sum_{i:X_i=1} 1} & \text{difference} \\
\hline
\polla   1   
    & \polla   `r J=J1; Y=y[J]; X=x[J]; X[1]`      & \polla `r Y[1]` 
    & \polla   `r X[2]`                            & \polla `r Y[2]`      
    & \polla   \dots 
    & \polla   `r X[n]`                            & \polla `r Y[n]`       
    & \polla   `r approx(mean(X))`                 & \polla `r approx(mean(Y))`   
    & \polla   `r approx(mean(Y[X==0]))`           & \polla `r approx(mean(Y[X==1]))` 
    & \polla   `r approx(mean(Y[X==1]) - mean(Y[X==0]))` \\

\pollb   1   
    & \pollb   `r J=J2; Y=y[J]; X=x[J]; X[1]`      & \pollb `r Y[1]` 
    & \pollb   `r X[2]`                            & \pollb `r Y[2]`      
    & \pollb   \dots 
    & \pollb   `r X[n]`                            & \pollb `r Y[n]`       
    & \pollb   `r approx(mean(X))`                 & \pollb `r approx(mean(Y))`   
    & \pollb   `r approx(mean(Y[X==0]))`           & \pollb `r approx(mean(Y[X==1]))` 
    & \pollb   `r approx(mean(Y[X==1]) - mean(Y[X==0]))` \\
    
\pollc   1   
    & \pollc   `r J=J3; Y=y[J]; X=x[J]; X[1]`      & \pollc `r Y[1]` 
    & \pollc   `r X[2]`                            & \pollc `r Y[2]`      
    & \pollc   \dots 
    & \pollc   `r X[n]`                            & \pollc `r Y[n]`       
    & \pollc   `r approx(mean(X))`                 & \pollc `r approx(mean(Y))`   
    & \pollc   `r approx(mean(Y[X==0]))`           & \pollc `r approx(mean(Y[X==1]))` 
    & \pollc   `r approx(mean(Y[X==1]) - mean(Y[X==0]))` \\
    
\vdots & \vdots & \vdots &  \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\

\polld   1   
    & \polld   `r J=JM; Y=y[J]; X=x[J]; X[1]`      & \polld `r Y[1]` 
    & \polld   `r X[2]`                            & \polld `r Y[2]`      
    & \polld   \dots 
    & \polld   `r X[n]`                            & \polld `r Y[n]`       
    & \polld   `r approx(mean(X))`                 & \polld `r approx(mean(Y))`   
    & \polld   `r approx(mean(Y[X==0]))`           & \polld `r approx(mean(Y[X==1]))` 
    & \polld   `r approx(mean(Y[X==1]) - mean(Y[X==0]))` \\
    
\end{array}
$$


## The Sampling Distribution 

```{r}
#| echo: true
#| eval: false
difference.samples = array(dim=10000)
for(rr in 1:10000) {
    I = sample(1:m, n, replace=TRUE)
    X = x[I]
    Y = y[I]
    difference.samples[rr] = mean(Y[X==1]) - mean(Y[X==0])
}
```

```{r}
set.seed(default.seed)
difference.samples = array(dim=10000)
difference.sd.samples = array(dim=10000)
for(rr in 1:10000) {
    I = sample(1:m, n, replace=TRUE)
    X = x[I]
    Y = y[I]
    difference.samples[rr] = mean(Y[X==1]) - mean(Y[X==0])
    p1 = mean(Y[X==1])
    p0 = mean(Y[X==0])
    difference.sd.samples[rr] = sqrt(p1*(1-p1)/sum(X==1) + p0*(1-p0)/sum(X==0))
}
```

```{r}
#| out-height: 400px 
#| fig-asp: .3 
difference.histogram = ggplot() + geom_histogram(aes(x=difference.samples, y=after_stat(density)), bins=100, alpha=.3)
difference.histogram + 
  with_annotations(difference.samples, shade.width=width(difference.samples), interval.width=width(difference.samples)) 
```

- Here's what the **sampling distribution** of this difference in turnout looks like.
- As before, if we can estimate it we can use that estimate to get a 95\% confidence interval.
- But, unlike before, we don't really know the parametric form of its sampling distribution.
- Or---at least---it'd be a pain to work it out. So we'll use the bootstrap to estimate it.

## Making a Table of Bootstrap Polls

```{r}
sam$label = 1:n
set.seed(default.seed)
boot.ind1 = sample(1:n, n, replace=TRUE)
boot.ind2 = sample(1:n, n, replace=TRUE)
boot.ind3 = sample(1:n, n, replace=TRUE)
bootstrap.sam = 
  rbind(sam[boot.ind1,], 
        sam[boot.ind2,], 
        sam[boot.ind3,])
bootstrap.sam$color=rep(c(pollb,pollc,polld), each=n)
bootstrap.sam$i=rep(1:n, 3)
```

:::: {.r-stack}
::: {.fragment .fade-out data-fragment-index="0"}
```{r}
#| fig-width: 7
#| fig-asp: .5
#| layout-ncol: 2
#| fig-cap:
#|  - Our Sample
#|  - Three Bootstrap Samples---The First 'Call'
state.plot + geom_text(aes(x = long, y = lat, label=label), color=polla, alpha=.7, size=3, data=sam) 
state.plot + geom_text(aes(x = long, y = lat, label=label, color=I(color)), alpha=1, size=3, data=bootstrap.sam[bootstrap.sam$i == 1, ])
```
:::
::: {.fragment .current-visible data-fragment-index="0"}
```{r}
#| fig-width: 7
#| fig-asp: .5
#| layout-ncol: 2
#| fig-cap:
#|  - Our Sample
#|  - Three Bootstrap Samples---The First and Second 'Call'
state.plot + geom_text(aes(x = long, y = lat, label=label), color=polla, alpha=.7, size=3, data=sam) 
state.plot + geom_text(aes(x = long, y = lat, label=label, color=I(color)), alpha=1, size=3, data=bootstrap.sam[bootstrap.sam$i %in% c(1,2), ])
```
:::
::: {.fragment .current-visible data-fragment-index="1"}
```{r}
#| fig-width: 7
#| fig-asp: .5
#| layout-ncol: 2
#| fig-cap:
#|  - Our Sample
#|  - Three Bootstrap Samples---The First, Second, and Last 'Call'
state.plot + geom_text(aes(x = long, y = lat, label=label), color=polla, alpha=.7, size=3, data=sam) 
state.plot + geom_text(aes(x = long, y = lat, label=label, color=I(color)), alpha=1, size=3, data=bootstrap.sam[bootstrap.sam$i %in% c(1,2,n), ])
```
:::
::::

$$
\small{
\begin{array}{r|rr|rr|r|rr|rrrr}
\text{call} &  1 & & 2 & & \dots & `r n` & & & & & & \\
\text{poll} &  X_1 & Y_1 &  X_2 & Y_2   & \dots &  X_{`r n`} & Y_{`r n`}  & \overline{X} & \overline{Y}  &\frac{\sum_{i:X_i=0} Y_i}{\sum_{i:X_i=0} 1} & \frac{\sum_{i:X_i=1} Y_i}{\sum_{i:X_i=1} 1} & \text{difference} \\
\hline
\polla   1   
    & \polla   `r J=J1; Y=y[J]; X=x[J]; X[1]`      & \polla `r Y[1]` 
    & \polla                         `r X[2]`      & \polla `r Y[2]`      
    & \polla   \dots 
    & \polla                         `r X[n]`      & \polla `r Y[n]`       
    & \polla   `r approx(mean(X))`   & \polla `r approx(mean(Y))`   
    & \polla  `r approx(mean(Y[X==0]))` & `r approx(mean(Y[X==1]))` 
    & \polla `r approx(mean(Y[X==1]) - mean(Y[X==0]))` 
\end{array}
}
$$

$$
\small{
\begin{array}{r|rr|rr|r|rr|rrrr}
\text{`call'} &  1 & &  2 & &  \dots & `r n` & & & & & & \\
\text{`poll'} &  X_1^* & Y_1^* &  X_2^* & Y_2^*   & \dots &  X^*_{`r n`} & Y^*_{`r n`}  & \overline{X}^* & \overline{Y}^*  &\frac{\sum_{i:X_i^*=0} Y_i^*}{\sum_{i:X_i^*=0} 1} & \frac{\sum_{i:X_i^*=1} Y_i^*}{\sum_{i:X_i^*=1} 1} & \text{difference} \\
\hline
\pollb   `r I=boot.ind1; 1`
    & \pollb   X_{`r I[1]`}  & \pollb Y_{`r I[1]`} 
    & &
    & \pollb   \dots
    & &
    & & & & & & \\
    & \pollb   `r X[I[1]]`   & \pollb `r Y[I[1]]`
    & &
    & \pollb   \dots
    & &
    & & & & & & \\
\pollc   `r I=boot.ind2; 2`
    & \pollc   X_{`r I[1]`}  & \pollc Y_{`r I[1]`} 
    & &
    & \pollc   \dots
    & &
    & & & & & & \\
    & \pollc   `r X[I[1]]`   & \pollc `r Y[I[1]]`
    & &
    & \pollc   \dots
    & &
    & & & & & & \\
\polld   `r I=boot.ind3; 3`
    & \polld   X_{`r I[1]`}  & \polld Y_{`r I[1]`} 
    & &
    & \polld   \dots
    & &
    & & & & & & \\
    & \polld   `r X[I[1]]`   & \polld `r Y[I[1]]`
    & &
    & \polld   \dots
    & &
    & & & & & & \\
\end{array}
}
$$


## A Completed Table of Bootstrap Polls

```{r}
#| fig-width: 7
#| fig-asp: .5
#| layout-ncol: 2
#| fig-cap:
#|  - Our Sample
#|  - A Few Bootstrap Samples

state.plot + geom_text(aes(x = long, y = lat, label=label), alpha=.7, size=3, data=sam) 
state.plot + geom_text(aes(x = long, y = lat, label=label, color=I(color)), alpha=.5, size=3, data=bootstrap.sam)
```


$$
\small{
\begin{array}{r|rr|rr|r|rr|rrrr}
\text{call} &  1 & & 2 & & \dots & `r n` & & & & & & \\
\text{poll} &  X_1 & Y_1 &  X_2 & Y_2   & \dots &  X_{`r n`} & Y_{`r n`}  & \overline{X} & \overline{Y}  &\frac{\sum_{i:X_i=0} Y_i}{\sum_{i:X_i=0} 1} & \frac{\sum_{i:X_i=1} Y_i}{\sum_{i:X_i=1} 1} & \text{difference} \\
\hline
\polla   1   
    & \polla   `r J=J1; Y=y[J]; X=x[J]; X[1]`      & \polla `r Y[1]` 
    & \polla                         `r X[2]`      & \polla `r Y[2]`      
    & \polla   \dots 
    & \polla                         `r X[n]`      & \polla `r Y[n]`       
    & \polla   `r approx(mean(X))`   & \polla `r approx(mean(Y))`   
    & \polla  `r approx(mean(Y[X==0]))` & `r approx(mean(Y[X==1]))` 
    & \polla `r approx(mean(Y[X==1]) - mean(Y[X==0]))` 
\end{array}
}
$$

$$
\small{
\begin{array}{r|rr|rr|r|rr|rrrr}
\text{`call'} &  1 & &  2 & &  \dots & `r n` & & & & & & \\
\text{`poll'} &  X_1^* & Y_1^* &  X_2^* & Y_2^*   & \dots &  X^*_{`r n`} & Y^*_{`r n`}  & \overline{X}^* & \overline{Y}^*  &\frac{\sum_{i:X_i^*=0} Y_i^*}{\sum_{i:X_i^*=0} 1} & \frac{\sum_{i:X_i^*=1} Y_i^*}{\sum_{i:X_i^*=1} 1} & \text{difference} \\
\hline
\pollb   `r I=boot.ind1; 2`
    & \pollb   X_{`r I[1]`}  & \pollb Y_{`r I[1]`} 
    & \pollb   X_{`r I[2]`}  & \pollb Y_{`r I[2]`}      
    & \pollb   \dots
    & \pollb   X_{`r I[n]`}  & \pollb Y_{`r I[n]`} 
    & & & & & & \\
    & \pollb   `r X[I[1]]`   & \pollb `r Y[I[1]]`
    & \pollb   `r X[I[2]]`      & \pollb `r Y[I[2]]`      
    & \pollb   \dots
    & \pollb   `r X[I[n]]`     & \pollb `r Y[I[n]]` 
    & \pollb   `r approx(mean(X[I]))`   & \pollb `r approx(mean(Y[I]))`   
    &  \pollb  `r approx(mean(Y[I][X[I]==0]))` & \pollb `r approx(mean(Y[I][X[I]==1]))` 
    & \pollb `r approx(mean(Y[I][X[I]==1]) - mean(Y[I][X[I]==0]))` \\
\pollc   `r I=boot.ind2; 2`
    & \pollc   X_{`r I[1]`}  & \pollc Y_{`r I[1]`} 
    & \pollc   X_{`r I[2]`}  & \pollc Y_{`r I[2]`}      
    & \pollc   \dots
    & \pollc   X_{`r I[n]`}  & \pollc Y_{`r I[n]`} 
    & & & & & & \\
    & \pollc   `r X[I[1]]`   & \pollc `r Y[I[1]]`
    & \pollc   `r X[I[2]]`      & \pollc `r Y[I[2]]`      
    & \pollc   \dots
    & \pollc   `r X[I[n]]`     & \pollc `r Y[I[n]]` 
    & \pollc   `r approx(mean(X[I]))`   & \pollc `r approx(mean(Y[I]))`   
    & \pollc  `r approx(mean(Y[I][X[I]==0]))` & \pollc `r approx(mean(Y[I][X[I]==1]))` 
    & \pollc `r approx(mean(Y[I][X[I]==1]) - mean(Y[I][X[I]==0]))` \\
\polld   `r I=boot.ind3; 1`M
    & \polld   X_{`r I[1]`}  & \polld Y_{`r I[1]`} 
    & \polld   X_{`r I[2]`}  & \polld Y_{`r I[2]`}      
    & \polld   \dots
    & \polld   X_{`r I[n]`}  & \polld Y_{`r I[n]`} 
    & & & & & & \\
    & \polld   `r X[I[1]]`   & \polld `r Y[I[1]]`
    & \polld   `r X[I[2]]`      & \polld `r Y[I[2]]`      
    & \polld   \dots
    & \polld   `r X[I[n]]`     & \polld `r Y[I[n]]` 
    & \polld   `r approx(mean(X[I]))`   & \polld `r approx(mean(Y[I]))`   
    & \polld  `r approx(mean(Y[I][X[I]==0]))` & \polld `r approx(mean(Y[I][X[I]==1]))` 
    & \polld `r approx(mean(Y[I][X[I]==1]) - mean(Y[I][X[I]==0]))` \\
\end{array}
}
$$


## The Difference's Bootstrap Sampling Distribution 

```{r}
#| echo: true

difference.bootstrap.samples = array(dim=10000)
for(rr in 1:10000) {
    I = sample(1:n, n, replace=TRUE)
    Xstar = X[I]
    Ystar = Y[I]
    difference.bootstrap.samples[rr] = mean(Ystar[Xstar==1]) - mean(Ystar[Xstar==0])
}
```

```{r}
#| fig-width: 7
#| fig-asp: .5 
#| layout-ncol: 2 

scale_x_dif = scale_x_continuous(limits=c(-.1,.2))
w = width(difference.samples, alpha=.05)
wboot=width(difference.bootstrap.samples, alpha=.05)

bootstrap.difference.histogram = ggplot() + 
  geom_histogram(aes(x=difference.bootstrap.samples, y=after_stat(density)), alpha=.3, bins=100) 

difference.histogram + 
  with_annotations(difference.samples, interval.width=wboot, highlight.dots=1) + 
  coord_flip() + scale_x_dif

bootstrap.difference.histogram + 
  with_annotations(c(difference.samples[1], difference.bootstrap.samples), highlight.dots=1, shade.width=wboot, shade.color='red') + 
  coord_flip() + scale_x_dif
```

- It looks like it works in this case. 
- But we're no longer able to argue that it should work the way we did before.
  - For that, we took advantage of our knowledge of our estimator's parametric form.
  - And we don't have that now.
- We'll get there. But we'll need a few new tools we'll develop in the coming weeks.
  - Normal approximation---a parametric form for an *approximation* to our estimator's sampling distribution.
  - Techniques for variance calculation. This'll help us understand the parameters that go into it.