---
title: "Lecture 3"
subtitle: "Calibrating Interval Estimates with Binary Observations"
---


# Review
## Polling

{{< include shared-code.qmd >}}
{{< include ga-turnout.qmd >}}


:::

- We want to estimate the proportion of *all registered voters* --- our population --- who will vote.
- To do this, we use the proportion of *polled voters* --- our sample --- who said they would.
- When the election occurs, we get to see who turns out to vote.
  - `r approx(sum(y))` people, or roughly `r percent(mean(y))` of registered voters, actually vote.
 
\ \

:::: columns
::: column
### Before the Election 

$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}  \\
Y_i  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$

:::
::: column
### After the Election

$$
\begin{array}{r|rrrrrr|r}
j & 1 & 2 & 3 & 4 & \dots &   `r approx(m)` & \bar{y}_{`r approx(m)`}  \\
y_{j} & `r y[1]` & `r y[2]` & `r y[3]` & `r y[4]` & \dots & `r y[m]`   & `r approx(mean(y))`  \\
\end{array}
$$
:::
::::


## The Sampling Distribution of our Estimate

```{r}
#| out-height: 400px
#| fig-asp: .3 

set.seed(default.seed)
Js = array(dim=c(10000, n))
Ys = array(dim=c(10000, n))
for(rr in 1:10000) {
  Js[rr,] = sample(m, n, replace=TRUE)
  Ys[rr,] = y[Js[rr,]]
}
meanY.samples = rowMeans(Ys)
pmf.height = max(dbinom(0:n, n, mean(y))) 

scale_x = scale_x_continuous(breaks=seq(.6,.8, by=.025)) 
scale_y = scale_y_continuous(breaks=seq(0, pmf.height,by=.01))
coord = coord_cartesian(xlim=c(.6,.8), ylim=c(0, pmf.height))

bar.plot = ggplot() + geom_bar(aes(x=meanY.samples, y=after_stat(prop)), alpha=.3) +
                      geom_vline(aes(xintercept=mean(y)), color='green', size=2, alpha=.7) + 
                      scale_x + scale_y
                      
bar.plot + with_annotations(meanY.samples, height=pmf.height, highlight.dots=4) + coord
```

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{poll} & J_1 & Y_1 & J_2 & Y_2   & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\hline
\polla   1   
    & \polla   `r J=J1; Y=y[J]; J[1]`    & \polla  `r Y[1]` 
    & \polla   `r J[2]`   & \polla   `r Y[2]`           & \polla   \dots 
    & \polla   `r J[n]`   & \polla   `r Y[n]`           & \polla   `r approx(mean(Y))` \\
\pollb 2   
    & \pollb `r J=J2; Y=y[J]; J[1]`    & \pollb  `r Y[1]` 
    & \pollb `r J[2]`   & \pollb  `r Y[2]`          & \pollb \dots 
    & \pollb `r J[n]`   & \pollb `r Y[n]`           & \pollb `r approx(mean(Y))` \\
\pollc    3   
    & \pollc    `r J=J3; Y=y[J]; J[1]`    & \pollc     `r Y[1]` 
    & \pollc    `r J[2]`   & \pollc    `r Y[2]`           & \pollc    \dots 
    & \pollc    `r J[n]`   & \pollc    `r Y[n]`           & \pollc    `r approx(mean(Y))` \\
{\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots} &  {\vdots} \\ 
\polld    1M   
    & \polld    `r J=JM; Y=y[J]; J[1]`    & \polld     `r Y[1]` 
    & \polld    `r J[2]`   & \polld    `r Y[2]`           & \polld    \dots 
    & \polld    `r J[n]`   & \polld    `r Y[n]`           & \polld    `r approx(mean(Y))` \\
 {\vdots} & {\vdots} & {\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots}  & {\vdots} \\ 
\end{array}
$$


## Interval Estimation

```{r}
#| out-height: 400px
#| fig-asp: .3 

w = .02
bar.plot + with_annotations(meanY.samples, interval.width=w, height=pmf.height, highlight.dots=4) + coord 
```

- To give a sense of what this tells us about the population proportion, we can report an interval estimate.
    - i.e., we can draw 'arms' on our point estimate $\bar Y_{`r n`}$. 
    - Here, we've used arms of width .02. i.e., we've drawn in the interval $\bar Y_{`r n`} \pm `{r} w/2`$.
- But this is only useful if we have a sense of how likely this interval is to be 'right.'
- That is, how often, when we report and interval, it contains [the population proportion]{.fg style="color:limegreen"}.
- We call this a *coverage probability*. In this case, it's bad. It's about **`r percent(mean(abs(meanY.samples - mean(y)) <= w/2))`**.
- If you run polls and report intervals like this, you'll make misleading claims about turnout **`r percent(1-mean(abs(meanY.samples - mean(y)) <= w/2))`** of the time. 
- That's not a good look. To fix this, we need to **calibrate** our interval estimates.

## A Useful Counting Trick

```{r}
#| out-height: 400px
#| fig-asp: .3 

bar.plot + with_annotations(meanY.samples, interval.width=w, shade.width=w, height=pmf.height, highlight.dots=4) + coord 
```

- We don't really need to look at every interval---and its arms---to calculate coverage.
- Instead, we can draw arms of the same width around the population proportion.
- An interval estimate covers the [population proportion]{.fg style="color:limegreen"} if and only if ...
- the corresponding point estimate is between the population proportion's arms.

## A Useful Counting Trick

```{r}
#| out-height: 400px
#| fig-asp: .3 

bar.plot + with_annotations(meanY.samples, interval.width=w, shade.width=w, height=pmf.height, highlight.dots=4) + coord 
```

- So all we need to do is count how many of our point estimates are between the population proportion's arms.
- Or, equivalently, calculate the **mass of our histogram** between the population proportion's arms.
- That is, we calculate the area of the verticals bars in the [shaded region]{.fg style="color:limegreen"}.
- You can see it's roughly half the area of the histogram itself. Hence, roughly 50% coverage.

## Calibrating Interval Estimates

```{r}
#| out-height: 400px
#| fig-asp: .3 

oldw = w
w=width(meanY.samples, alpha=.05)
bar.plot + with_annotations(meanY.samples, interval.width=oldw, shade.width=w, height=pmf.height, highlight.dots=4) + coord 
```

- To increase the coverage of our intervals estimates, we need to make the arms wider.
- To **calibrate** this width to give us 95% coverage---which is the convention---we can use our counting trick.
- We expand our [shaded region]{.fg style="color:limegreen"} until it covers 95% of the histogram's mass, i.e., 95% of our point estimates.
- If we use intervals of this width around our point estimates, we'll get 95% coverage.

## Calibrating Interval Estimates
```{r}
#| out-height: 400px
#| fig-asp: .3 

bar.plot + with_annotations(meanY.samples, interval.width=w, shade.width=w, height=pmf.height, highlight.dots=4) + coord 
```

- To increase the coverage of our intervals estimates, we need to make the arms wider.
- To **calibrate** this width to give us 95% coverage---which is the convention---we can use our counting trick.
- We expand our [shaded region]{.fg style="color:limegreen"} until it covers 95% of the histogram's mass, i.e., 95% of our point estimates.
- If we use intervals of this width around our point estimates, we'll get 95% coverage.

## The Problem

```{r}
#| out-height: 400px
#| fig-asp: .3 

bar.plot + with_annotations(meanY.samples, interval.width=w, shade.width=w, height=pmf.height, highlight.dots=4) + coord 
```

- Unless we've seen the population, i.e. unless it's after the election, we can't run simulated polls.
- That means we can't calibrate our interval using the sampling distribution of our point estimate. 
- That's something we don't know. The good news is that it's something we can estimate.
- So we'll actually do calibration using **an estimate of the sampling distribution**.

# Real-Life Calibration

## Calibration using a Sampling Distribution Estimate

```{r}
#| out-height: 400px
#| fig-asp: .3 

set.seed(default.seed)
estimated.dist.samples = rbinom(10000, n, mean(Y)) / n 
w.est=width(estimated.dist.samples, alpha=.05)

bar.plot.estimated = ggplot() + 
    geom_bar(aes(x=estimated.dist.samples, y=after_stat(prop)), alpha=.3) +
    geom_vline(aes(xintercept=mean(y)), color='green', size=2, alpha=.7) + 
    scale_x + scale_y
bar.plot.estimated + with_annotations(estimated.dist.samples, interval.width=w.est, highlight.dots = 0,  
                                      shade.width=w.est, shade.color='red', height=pmf.height) + coord
```

- Here we've estimated the sampling distribution --- we'll get to how --- and used it to find an interval width.
- In particular, we've used a width which ...
    - if we draw arms around the mean of our estimated sampling distribution
    - will cover 95% of draws from our estimated sampling distribution.
- The one apparent sticking point is that our estimated sampling distribution is a bit off-center.
- Its mean is a bit lower than [the population proportion]{.fg style="color:limegreen"}. It's actually our sample proportion $\bar Y$.
- It turns out that this doesn't matter much. Why?

## Calibration Comparison

```{r}
#| out-height: 400px
#| fig-asp: .3 


K = 100
heights = annotation_dot_heights(K, 3, pmf.height)
calibration.comparison.histogram = bar.plot +
    geom_rect(aes(xmin=mean(y)-w/2, xmax=mean(y)+w/2, ymin=0, ymax=1.25*heights[1]), alpha=.15, fill='green') +
    geom_rect(aes(xmin=mean(y)-w.est/2, xmax=mean(y)+w.est/2, ymin=1.25*heights[1], ymax=Inf), alpha=.15, fill='red')


calibration.comparison.histogram +  
    geom_pointrange(aes(y=heights[1]*1.5, x=meanY.samples[1], 
                        xmin = meanY.samples[1]-w.est/2, xmax = meanY.samples[1]+w.est/2), alpha=1, color='black')  +
    geom_pointrange(aes(y=heights[1], x=meanY.samples[1], 
                        xmin = meanY.samples[1]-w/2, xmax = meanY.samples[1]+w/2), alpha=.4, color='black')   
```

- It doesn't matter because we're not putting arms on draws from the estimated sampling distribution.
- We're putting arms on our point estimate, which is a draw from the **actual sampling distribution**.
- To do this, we're using the width---but not the center---of the estimated sampling distribution.
- And it works because that's very close to the width we'd get from the actual sampling distribution.
- Above, I've drawn in shaded regions corresponding to two versions of the the [population proportion's]{.fg style="color:limegreen"} arms.
    - The [green region]{.fg style="color:limegreen"} is the one we get from the actual sampling distribution. We can't use this.
    - The [red region]{.fg style="color:red"} is the one we get from the estimated sampling distribution. We can --- and do --- use this.
- And I've drawn in interval estimates based on our poll using both widths. They're almost the same.


## Our Samping Distribution is Binomial

```{r}
#| out-height: 300px
#| fig-asp: .25 
p = mean(y)
x= seq(0,n)

bar.plot + with_annotations(meanY.samples, shade.width=w, highlight.dots=0) + 
            geom_line(aes(x=x/n, y=dbinom(x, n, mean(y))), color='black') + coord
```

- To estimate the sampling distribution, it helps to know something about the actual sampling distribution.
    - It's the distribution of the proportion heads in `r n` flips of a coin with probability $p\approx `r approx(p)`$ of heads.
    - where $p \approx `r approx(p)`$ is the proportion of registered voters who will vote.
- There's a name for that: it's the binomial distribution. 
    - To be more precise, the binomial distribution is for the number of heads.
    - And we scale that by $1/`r n`$ to get a distribution for the proportion.

$$
`r n` \times \bar{Y}_{`r n`} = \sum_{i=1}^{`r n`} Y_i \sim \text{Binomial}(`r n`, p) \qfor p = \bar y_{`r approx(m)`} = \frac{1}{`r approx(m)`}\sum_{j=1}^{`r approx(m)`} y_j 
$$

- There's a formula for this probability, so we don't have to draw samples and histogram if we don't want to.
- In `R`, it's `dbinom(x, n, p)`, where `x` is the number of heads, `n` is the number of flips, and `p` is the probability of heads.

## Estimating the Sampling Distribution

```{r}
#| out-height: 300px
#| layout-ncol: 2 
#| fig-asp: .4 
#| fig-cap:
#|   - "Actual Sampling Distribution"
#|   - "Estimated Sampling Distribution"


bar.plot + with_annotations(meanY.samples, shade.width=w, highlight.dots=0) + 
            geom_line(aes(x=x/n, y=dbinom(x, n, mean(y))), color='gray') + coord
bar.plot.estimated + with_annotations(estimated.dist.samples, shade.width=w.est, shade.color='red', highlight.dots=0) + 
            geom_line(aes(x=x/n, y=dbinom(x, n, mean(Y))), color='black') + coord
```

$$
`r n` \times \bar{Y}_{`r n`} = \sum_{i=1}^{`r n`} Y_i \sim \text{Binomial}(`r n`, p) \qfor p = \bar y_{`r approx(m)`} = \frac{1}{`r approx(m)`}\sum_{j=1}^{`r approx(m)`} y_j 
$$


- What we don't know is $p=\bar y$, the proportion of registered voters who turn out.
- But that's what we're using our sample proportion $\bar Y$ to estimate. 
- And we can just plug that in to get an estimate of the sampling distribution.
- To draw samples from this estimated sampling distribution, we use `rbinom`.

```{r}
#| echo: true
#| eval: false
samples = rbinom(10000, 625, mean(Y)) / 625
``` 

- That is exactly how I got the estimated sampling distribution histogram I was using for calibration. 
    - If we prefer, we can calibrate using the actual estimated binomial distribution, i.e., using `dbinom`.
    - We'll explore this a bit at some point in the future, but it doesn't make much difference.

## Each of Our Friends Gets Their Own Estimate

```{r}
#| cache: true 
#| out-height: 300px
#| layout-ncol: 2 
#| fig-asp: .4 
#| fig-cap:
#|   - "Actual Sampling Distribution"
#|   - "Estimated Sampling Distribution (Poll 1)"
#|   - "Estimated Sampling Distribution (Poll 2)"
#|   - "Estimated Sampling Distribution (Poll 3)"

plot.interval = function(Y, height, color) {
    set.seed(default.seed)
    samples = rbinom(10000, n, mean(Y)) / n 
    w=width(samples, alpha=.05)
    geom_pointrange(aes(x=mean(Y), y=height, xmin=mean(Y)-w/2, xmax=mean(Y)+w/2), color=color)
}

dh = .2*heights[1]
bar.plot + with_annotations(meanY.samples, shade.width=w, highlight.dots=3, highlighted.dot.alpha=.4) +
               geom_line(aes(x=x/n, y=dbinom(x, n, mean(y))), color='gray') +
               plot.interval(y[J1], dh+heights[1], color=polla) +
               plot.interval(y[J2], dh+heights[2], color=pollb) +
               plot.interval(y[J3], dh+heights[3], color=pollc) + coord

calibration.diagram = function(Y, color, height) {
    set.seed(default.seed)
    samples = rbinom(10000, n, mean(Y)) / n 
    w=width(samples, alpha=.05)
    histplot = ggplot() + geom_bar(aes(x=samples, y=after_stat(prop)), alpha=.3) +
                          geom_vline(aes(xintercept=mean(y)), color='green', size=3, alpha=.7) + 
                          with_annotations(samples, shade.width=w, shade.color='red', height=pmf.height, highlight.dots=0) +
    geom_line(aes(x=x/n, y=dbinom(x, n, mean(Y))), color=color) +
    geom_pointrange(aes(x=mean(Y), xmin=mean(Y)-w/2, xmax=mean(Y)+w/2, y=height), color=color)
}

calibration.diagram(Y,     polla, heights[1]) + coord
calibration.diagram(y[J2], pollb, heights[2]) + coord
calibration.diagram(y[J3], pollc, heights[3]) + coord
```

- This means each of our friends estimates a different width for their interval.
- But they're all close to the width we get using the actual sampling distribution. You can see it above.
    - I've plotted the intervals they'd use --- based on their sampling distribution estimates --- in bold colors.
    - And the intervals they wish they could --- based on the actual sampling distribution --- more faintly. 

## Evidence That It Works 

```{r}
#| out-height: 400px
K = 100
ps = meanY.samples[1:K]

ggplot() + geom_line(aes(x=x/n-p, y=n*dbinom(x,n,p)), data=expand.grid(p=mean(y), x=0:n)) +
           geom_vline(aes(xintercept=qbinom(c(.025,.975), n,mean(y))/n-mean(y)), linetype='dotted') +
           geom_line(aes(x=x/n-p, y=n*dbinom(x,n,p), color=factor(p)), data=expand.grid(p=ps, x=0:n), alpha=.2) +
           geom_pointrange(aes(x=0, xmin=qbinom(.025,n,p)/n-p, xmax=qbinom(.975,n,p)/n-p, y=20*k/K, color=factor(p)), data=data.frame(p=ps, k=1:K)) +
           xlim(-.1,.1) + guides(color=FALSE)
```

- We see the same thing when we look at 100 polls.
- Our friends' sampling distribution estimates are all close to the actual sampling distribution.
- The resulting interval-widths are near-indistinguishable from the width we got using the sampling distribution.
  - The estimated widths are shown as points with arms.
  - That sampling distribution's actual width is the space between the dotted lines.
- You can see estimating the width like this works reliably.

## Why Does it Work?

```{r}
#| out-height: 300px
#| fig-asp: .6 
#| layout-ncol: 2
height = pmf.height 
K = 3
ps = quantile(meanY.samples, c(.025, .5, .975))

density.stability.plot = ggplot() + 
           geom_line(aes(x=x/n, y=dbinom(x,n,p)), data=expand.grid(p=mean(y), x=0:n), alpha=.5) +
           geom_vline(aes(xintercept=qbinom(c(.025,.975), n,mean(y))/n), linetype='dotted') +
           geom_line(aes(x=x/n, y=dbinom(x,n,p), color=factor(p)), data=expand.grid(p=ps, x=0:n), alpha=.5) +
           geom_pointrange(aes(x=p, xmin=qbinom(.025,n,p)/n, xmax=qbinom(.975,n,p)/n, y=.6*height*k/K, color=factor(p)), data=data.frame(p=ps, k=1:K)) +
           xlim(.6,.8) + guides(color=FALSE) + geom_vline(xintercept=mean(y), color='green', size=1, alpha=.5)
centered.density.stability.plot = ggplot() + 
           geom_line(aes(x=x/n-p, y=dbinom(x,n,p)), data=expand.grid(p=mean(y), x=0:n), alpha=.5) +
           geom_vline(aes(xintercept=qbinom(c(.025,.975), n,mean(y))/n-mean(y)), linetype='dotted') +
           geom_line(aes(x=x/n-p, y=dbinom(x,n,p), color=factor(p)), data=expand.grid(p=ps, x=0:n), alpha=.5) +
           geom_pointrange(aes(x=0, xmin=qbinom(.025,n,p)/n-p, xmax=qbinom(.975,n,p)/n-p, y=.6*height*k/K, color=factor(p)), data=data.frame(p=ps, k=1:K)) +
           xlim(-.1,.1) + guides(color=FALSE)

density.stability.plot
centered.density.stability.plot
```

- Why are we getting a good estimate of the sampling distribution?
- The Binomial distribution is *continuous* as a function of $p$---when $p$ changes little, the distribution changes little.
- This means that, if we're not far off in the proportion, the estimated and actual sampling distributions are similar.
- The relevant difference (after centering) is even smaller because *the way* the binomial changes is mostly location.
- Here I'm showing three estimates of the sampling distribution based on three polls.
    - with centering (right) and without (left). 
- In particular, polls giving estimates at the 2.5th, 50th, and 97.5th percentiles of the actual sampling distribution. 

## The Estimate Works. Why Does it Work?
```{r}
#| out-height: 300px
#| fig-asp: .6 
#| layout-ncol: 2

density.stability.plot
centered.density.stability.plot
```

- You can think of this as a sort of 'confidence interval' for our estimate of the sampling distribution.
    - 95% of the time, you'll get an estimate somewhere between the [red]{.fg style="color:red"} and [blue]{.fg style="color:blue"} ones.
    - And, as a result, a width for your interval estimate somewhere between the [red]{.fg style="color:red"} and [blue]{.fg style="color:blue"} widths.
- As you can see, the widths we get barely change. What changes is the location.
- But because we're using calibrated 95% intervals, we still cover the population proportion 95% of the time.
    - i.e., we cover it when our sample proportion is between the [red]{.fg style="color:red"} and [blue]{.fg style="color:blue"} ones.
- Or, accounting for [slight underestimates]{.fg style="color:blue"} of the sampling distribution's width, almost cover it. 

## I'm Repeating the Last Slide To Demonstrate How to 'Unlink' Zoom on Multiple Plots When the Default is to Link Them
::: {.free-zoom}
```{r}
#| out-height: 300px
#| fig-asp: .6 
#| layout-ncol: 2

density.stability.plot
centered.density.stability.plot
```
:::

- You can think of this as a sort of 'confidence interval' for our estimate of the sampling distribution.
    - 95% of the time, you'll get an estimate somewhere between the [red]{.fg style="color:red"} and [blue]{.fg style="color:blue"} ones.
    - And, as a result, a width for your interval estimate somewhere between the [red]{.fg style="color:red"} and [blue]{.fg style="color:blue"} widths.
- As you can see, the widths we get barely change. What changes is the location.
- But because we're using calibrated 95% intervals, we still cover the population proportion 95% of the time.
    - i.e., we cover it when our sample proportion is between the [red]{.fg style="color:red"} and [blue]{.fg style="color:blue"} ones.
- Or, accounting for [slight underestimates]{.fg style="color:blue"} of the sampling distribution's width, almost cover it.
  

## I'm Repeating it Again To Demonstrate How to 'Link' Zoom on Multiple Plots When the Default is not to Link Them 
::: {.linked-zoom}
```{r}
#| out-height: 300px
#| fig-asp: .6 
#| layout-ncol: 2

density.stability.plot
centered.density.stability.plot
```
:::

- You can think of this as a sort of 'confidence interval' for our estimate of the sampling distribution.
    - 95% of the time, you'll get an estimate somewhere between the [red]{.fg style="color:red"} and [blue]{.fg style="color:blue"} ones.
    - And, as a result, a width for your interval estimate somewhere between the [red]{.fg style="color:red"} and [blue]{.fg style="color:blue"} widths.
- As you can see, the widths we get barely change. What changes is the location.
- But because we're using calibrated 95% intervals, we still cover the population proportion 95% of the time.
    - i.e., we cover it when our sample proportion is between the [red]{.fg style="color:red"} and [blue]{.fg style="color:blue"} ones.
- Or, accounting for [slight underestimates]{.fg style="color:blue"} of the sampling distribution's width, almost cover it.


## The Bootstrap Interpretation

```{r}
boot.ind = sample(1:n, n, replace=TRUE)

displaypop = pop[1:50000,]
pop$sam.replicated   = as.numeric(table(c(1:m, Js[1,])) - 1)
pop$boot.replicated  = as.numeric(table(c(1:m, Js[1, boot.ind])) - 1)

pop.map  = state.plot + geom_point(aes(x = long, y = lat), 
                                   alpha=.05, data=displaypop, size=.025, color=polla) + 
                        scale_size_area() + guides(size='none')
sam.map  = state.plot + geom_point(aes(x = long, y = lat, size=sam.replicated),   
                                   alpha=.2, data=pop[pop$sam.replicated > 0, ], color=polla) + 
                        scale_size_area(max_size=3) + guides(size='none')
boot.map = state.plot + geom_point(aes(x = long, y = lat, size=boot.replicated),   
                                   alpha=.2, data=pop[pop$boot.replicated > 0, ], color=polla)   +
                        scale_size_area(max_size=6) + guides(size='none')
```

```{r}
#| out-height: 300px
#| fig-asp: 1 
#| layout-ncol: 3
pop.map
sam.map
boot.map
```


- There's another way to think about this.
- The estimated sampling distribution is $\text{Binomial}(n,p)/n$ for $n=`r n`$ and $p=\bar Y=`{r} sum(Y)`/`{r} n` \approx `r approx(mean(Y))`$.
- That is, it's the distribution of the proportion of heads in ...
    - `r n` flips of a coin 
    - that has probability $\bar Y=`{r} sum(Y)`/`{r} n`$ of being heads.
- What's the distribution of the proportion of yeses if we poll a population of `r n` people ...
- ... among whom the proportion of yeses is exactly $\bar Y = `{r} sum(Y)` / `{r} n`$?


## The Bootstrap Interpretation

```{r}
#| out-height: 300px
#| fig-asp: 1 
#| layout-ncol: 3

pop.map
sam.map
boot.map
```

- It's the same distribution. It's $\text{Binomial}(n,p)/n$.
- That means we can get a draw from our estimated sampling distribution by
- running a 'poll' of the people in our sample, i.e.
    - rolling a 625-sided die 625 times
    - 'calling' the corresponding person in our sample
    - and counting up the yeses we hear
- We call the sample we get from this new 'poll' a **bootstrap sample**.

## The Bootstrap Interpretation

```{r}
#| out-height: 300px
#| fig-asp: 1 
#| layout-ncol: 3

pop.map
sam.map
boot.map
```

- If we plot our voters on a map, you can see the idea in visual terms.
    - On the left, we have the population. 
    - In the middle, we have our sample. It's drawn, with replacement, from the population.
    - On the right, we have something else. A new sample drawn, with replacement, *from the sample*.
- Each 'call' that a person receives increases the size of their dot: $\text{circle area} \propto \text{number of calls}$.
- In the sample, even though it's drawn with replacement, all dots are the same size.
  - Because we draw from such a large population, nobody gets called twice.
- In the bootstrap sample, dots vary in size.
  - Because we draw $n$ people from a sample of size $n$, it's almost impossible not to call somebody twice..

## Bootstrapping Our Turnout Estimate

```{r}
set.seed(default.seed)
```

```{r}
#| echo: true
#| out-height: 400px 
#| fig-asp: .3 

meanY.boot = array(dim=10000)
for(rr in 1:10000) {
    I = sample(1:n, n, replace=TRUE)
    meanY.boot[rr] = sum(Y[I])/n
}

ggplot() + geom_bar(aes(x=meanY.boot, y=after_stat(prop)), alpha=.3) + 
           geom_line(aes(x=(0:n)/n, y=dbinom(0:n, n, mean(Y))), color='black') + 
           xlim(.6,.8)
```


