---
title: "Point and Interval Estimates"
format: html
---


# Review

## Friday's Lab

{{< include shared-code.qmd >}}

```{r}
#| label: define-population
m = 6
pop = data.frame(j=1:6, 
                 y=c(0,1,1,1,0,1))
                 
scales = list(scale_x_continuous(breaks = 1:m),
              scale_y_continuous(breaks = (0:3/3), labels=sprintf("%d/3", 0:3))) 

pop.plot = ggplot(pop, aes(x = j, y = y)) + 
           geom_point(shape='circle', alpha=.1, size=5) + 
           scales               
```

:::: columns
::: column

```{r}
#| fig-width: 7
y= pop$y
pop.plot + geom_point(aes(x=j, y=y), size=5, data=pop[c(3,5,6),], color=polla)
```

:::
::: column

$$
\color{gray}
\begin{array}{r|rrrrrr|r}
j & 1 & 2 & 3 & 4 & 5 & 6 & 
\bar{y}  \\
y_{j} & `r y[1]` & `r y[2]` & \inpolla{`r y[3]`} & `r y[4]` & \inpolla{`r y[5]`} & \inpolla{`r y[6]`} 
& \textcolor{black}{`r sum(y)`/`r length(y)`}  \\
\end{array}
$$

:::
::::

On Friday, we explored how sampling can be used to summarize a population without surveying everyone in it. In our example, we estimated the proportion of people in a population of six who prefer chocolate to sour candy. Looking at the fake population shown above, we can see that `r sum(pop$y)`/`r m` people prefer chocolate. Those who prefer chocolate are represented in the top row at y=1, while those who prefer sour candy are shown below at y=0.

We experimented with several different sampling schemes, including sampling with replacement, sampling without replacement, coin-flip randomization, and convenience sampling. Ultimately, we settled on coin-flip randomization as our method. This involved flipping a coin for each person in our population to determine whether they would be included in our sample. Among the three people who flipped heads, two preferred chocolate to sour candy. This led us to estimate that 2/3 of the population preferred chocolate.

## Fake Data Simulation 

:::: columns
::: column

```{r}
#| fig-width: 7
#| layout-nrow: 3
y= pop$y
sam1 = pop[c(3,5,6),]
sam2 = pop[c(4,6),]
sam3 = pop[c(1,3,4,6),]
pop.plot + geom_point(aes(x=j, y=y), size=5, data=sam1, color=polla) + 
           geom_point(aes(x=j, y=y-.05),     size=5, data=sam2, color=pollb) + 
           geom_point(aes(x=j, y=y-.1), size=5,  data=sam3, color=pollc)
```

:::
::: column

$$
\color{gray}
\begin{array}{r|rrrrrr|r}
j & 1 & 2 & 3 & 4 & 5 & 6 & 
\bar{Y}  \\
y_{j} & `r y[1]` & `r y[2]` & \inpolla{`r y[3]`} & `r y[4]` & \inpolla{`r y[5]`} & \inpolla{`r y[6]`} 
& \inpolla{`r sum(sam1$y)`/`r length(sam1$y)`}  \\
y_{j} & `r y[1]` & `r y[2]` & `r y[3]` & \inpollb{`r y[4]`} & `r y[5]` & \inpollb{`r y[6]`} 
& \inpollb{`r sum(sam2$y)`/`r length(sam2$y)`}  \\
y_{j} & \inpollc{`r y[1]`} & `r y[2]` & \inpollc{`r y[3]`} & \inpollc{`r y[4]`} & `r y[5]` & \inpollc{`r y[6]`} 
& \inpollc{`r sum(sam3$y)`/`r length(sam3$y)`}  \\
\end{array}
$$

:::
::::

While this approach saved us significant time, we needed to consider whether our estimate was trustworthy. How close was our estimate to the actual proportion in the population of six? To investigate this, we ran a simulation of our survey on a fake population. We repeated exactly what we did in our actual survey multiple times, allowing us to examine what our estimator does while knowing what we want it to achieve: being approximately 4/6. Above, we've shown the results of three of these simulations.

We refer to the distribution of estimates we obtain in this way as our estimator's sampling distribution. These three estimates represent three draws from the sampling distribution - three equally likely outcomes of our survey. While we performed well in these three instances, three samples aren't enough to get a comprehensive sense of what's likely. To better understand the probabilities involved, we need to examine many more samples.

## Visualizing the Sampling Distribution

```{r}
#| label: sampling-mechanisms
estimator = function(sam) { mean(sam$y) }
with.replacement = function(pop, sampling.rate=1/2) { 
  N = round(m * sampling.rate) 
  J = sample(1:m, N, replace=TRUE)
  sam = pop[J,]
  estimator(sam)
}
without.replacement = function(pop, sampling.rate=1/2) {
  N = round(m * sampling.rate) 
  J = sample(1:m, N, replace=FALSE)
  sam = pop[J,]
  estimator(sam)
}
coinflip.randomization = function(pop, sampling.rate=1/2) {
  flips = rbinom(m, 1, sampling.rate)
  sam = pop[flips==1,]
  estimator(sam)
}
convenience = function(pop, sampling.rate=1/2) {
  N = round(m * sampling.rate) 
  J = 1:N 
  sam = pop[J,]
  estimator(sam)
}
```

```{r}
#| label: candy-sampling-distributions

set.seed(1) 
sampling.rate = 1/2
all.samples = rep(list(pop), 1000) |> map(function(pop) {
  data.frame(with.replacement=with.replacement(pop, sampling.rate),
             without.replacement=without.replacement(pop, sampling.rate),
             coinflip.randomization=coinflip.randomization(pop, sampling.rate),
             convenience=convenience(pop, sampling.rate))
}) |> bind_rows() |> 
      pivot_longer(cols=everything(), 
      names_to='sampling', values_to='estimate') 
all.samples$y = runif(nrow(all.samples))
```

::: {.panel-tabset group="candy"}

### Using Dots
```{r}
coinflip.samples = all.samples |> filter(sampling == 'coinflip.randomization')
scale_x = scale_x_continuous(breaks = (0:3)/3, labels=sprintf("%d/3", 0:3), limits=c(-.2, 1.2))
scale_y = scale_y_continuous(breaks = seq(0, 1, by=.25), limits=c(0,1))

sampling.dist.plot = ggplot(coinflip.samples) +
  geom_vline(xintercept=mean(y), color='green', linewidth=2, alpha=.5) + 
  geom_point(aes(x=estimate, y=y), color='purple', alpha=.2, size=.5) 
sampling.dist.plot + scale_x + scale_y
```

In the visualization above, we plot 1000 draws from the sampling distribution as 1000 purple circles. To highlight our target, we've drawn the population proportion as a green line. By examining the plot, we can see that estimates are typically quite close to the target, though not always. While we could manually count the dots in each column to be more precise, it's more efficient to let our computer handle this task.

### Using Bars 
```{r}
sampling.dist.plot + 
  geom_bar(aes(x=estimate, y=after_stat(prop)), alpha=.2) + scale_x + scale_y
```

We can represent the proportion of dots in each column as a bar graph. This approach is useful if you want to count dots in columns, but it can be somewhat counterintuitive visually due to the uneven spacing of the columns. Most of the dots cluster in the middle, but because there are many columns in that region representing slightly different estimates, no single bar appears particularly high. In fact, the highest bar appears at x=1. This raises the question: how can we better visualize the distribution of dots without encountering this problem?

### Using Histograms
```{r}
histogram_overlay = geom_histogram(aes(x=estimate, y=after_stat(density)/3), breaks=(-1:6)/6 + 1/12, closed='right', alpha=.2) 
sampling.dist.plot + histogram_overlay + scale_x + scale_y
```

The solution lies in grouping nearby columns into equal-width bins and counting the dots in each bin. This approach gives us a better sense of the density of estimates near each value of x, creating what we call a histogram. This type of visualization makes it easy to determine the fraction of dots in any interval between two bin edges - it's simply the fraction of the histogram's area that falls between them.

:::

## Sampling Distributions in Real Studies

```{r}
#| layout-ncol: 2
#| fig-width: 7 
#| fig-cap:
#|  - Your Estimator in a Simulated Study
#|  - Your Estimator in a Real Study 

sampling.dist.plot + histogram_overlay + scale_x + scale_y
ggplot(coinflip.samples[1,]) + geom_point(aes(x=estimate, y=1/2)) + scale_x + scale_y
```

Understanding sampling distributions is relatively straightforward when we're running a simulation of our study. In this context, we can run our simulation as many times as we want, allowing us to plot the sampling distribution and use our knowledge of the population to evaluate its performance. However, in a real study, we don't have these luxuries - we only get a single estimate.

Despite this limitation, we can still develop a complete picture by following three steps. First, we think of our single estimate as one draw from our estimator's sampling distribution. Second, we work out where that sampling distribution is in relation to the estimation target. Finally, we use the data we have to estimate the sampling distribution itself.

## What Do We Do with a Sampling Distribution?
```{r}
sampling.dist.plot + 
  geom_point(aes(x=estimate, y=1/2), data=coinflip.samples[1,], color=midnight) + 
  histogram_overlay + scale_x + scale_y
```

```{r}
estimates = all.samples |> filter(sampling == 'coinflip.randomization') |> pull(estimate)
```

A sampling distribution might seem like an unusual way to summarize the proportion of people who prefer chocolate. After all, you asked for a single number describing the population, but what you got was 1000 numbers describing 1000 different samples. However, when examined properly, this distribution can tell us quite a bit about our estimation process.

For instance, we can determine how often our estimate exactly matches the population proportion - in this case, it happens `r percent(mean(estimates == mean(y), na.rm=TRUE))` of the time. We can also see how often our estimate is off by 1/3 or less - this occurs `r percent(mean(abs(estimates - mean(y)) <= 1/3+.01, na.rm=TRUE))` of the time.

To translate this information into a statement about the population, we report two key pieces of information. First, we provide an interval where we expect the population proportion to lie. For example, we might say we think it's in the interval 2/3 ± 1/3 because 2/3 is our sample proportion. Second, we specify our confidence level - the probability that the true value falls within this interval. In this case, we can say that in `r percent(mean(abs(estimates - mean(y)) <= 1/3+.01, na.rm=TRUE))` of surveys run exactly like this, the true value would fall within our calculated interval.

# Other Sampling Schemes

## One Sampling Distribution per Sampling Scheme
```{r}
four.sampling.dist.plot = ggplot(all.samples) +
  geom_vline(xintercept=mean(y), color='green', linewidth=2, alpha=.5) + 
  geom_point(aes(x=estimate, y=y), color='purple', alpha=.2, size=.5) +
  geom_histogram(aes(x=estimate, y=after_stat(density)/3, fill=sampling), breaks=(-1:3)/3 + 1/6, alpha=.2)  
four.sampling.dist.plot + scale_x + scale_y + facet_grid(cols=vars(sampling)) + guides(fill=FALSE) 
```

Let's take a moment to examine how different sampling schemes affect our estimates. To do this, we'll look at 1000 draws from the sampling distribution we get using each scheme, and we'll display a histogram for each to make the comparison easier.

Three out of the four mechanisms we've plotted produce fairly similar sampling distributions. They all show a peak around 2/3, which is the population proportion - exactly where we want most of our estimates to be. The exception is convenience sampling, which behaves quite differently. In this case, it was most convenient for me to sample the first three people in our population. Since this selection doesn't change, neither does our estimate. While this particular instance yielded a good estimate, that's not something we can count on. Without a random mechanism, it's difficult to be confident that our estimate will be accurate. This is why using multiple, meaningfully different fake populations can be helpful - it helps us identify cases where a method works well by chance rather than by design.

## We'll Focus on One
```{r}
ggplot(all.samples |> filter(sampling=='with.replacement')) +
  geom_vline(xintercept=mean(y), color='green', linewidth=2, alpha=.5) + 
  geom_point(aes(x=estimate, y=y), color='purple', alpha=.2, size=.5) +
  geom_histogram(aes(x=estimate, y=after_stat(density)/3), breaks=(-1:3)/3 + 1/6, alpha=.2)  +
  scale_x + scale_y + guides(fill=FALSE)
```

For the purposes of today's discussion, we'll focus on sampling with replacement. This choice is made primarily to keep things simple and concrete, though most of what we'll discuss applies to any sampling scheme. Our focus will be on understanding the relationship between three key elements: our estimator, its sampling distribution, and our estimation target.

As we progress through the semester, we'll explore these relationships more generally. We'll discuss how we want these three elements to be related and when we can be confident that they are related in the way we desire.


# Polling

{{< include ga-turnout.qmd >}}

```{r}
height = pmf.height = max(dbinom(0:n, n, mean(y))) 

scale_x = scale_x_continuous(breaks=seq(.6,.8, by=.025)) 
scale_y = scale_y_continuous(breaks=seq(0, pmf.height,by=.01))
coord = coord_cartesian(xlim=c(.6,.8), ylim=c(0, pmf.height))

bar.plot = ggplot() + geom_bar(aes(x=meanY.samples, y=after_stat(prop)), alpha=.3) +
                      geom_vline(aes(xintercept=mean(y)), color='green', size=2, alpha=.7) + 
                      scale_x + scale_y
```

## Context

Let's consider a scenario where, a week before the 2020 presidential election, you conducted some polling. You used a list of approximately `r approx(m)` people registered to vote in Georgia and made `r n` phone calls. For each call, you selected a voter uniformly at random from the list, perhaps by rolling a `r approx(m)`-sided die, and then asked the potential voter whether they planned to vote.

For the sake of this example, let's assume that all registered voters will pick up the phone when called, respond honestly to your questions, and not change their minds about voting. In other words, we'll assume they tell us whether they ultimately vote on election day.

## Polling Results

You organized your polling results in a table with one column for each call. In each column, you recorded two pieces of information: the call number i (a number from 1 to `r n`) and the response Y_i of the person you called (1 for 'yes' and 0 for 'no').

$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}  \\
Y_i  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$ 

You summarized these results with an additional column showing the mean of the responses. Remember that the mean of a binary variable represents a frequency or proportion. In this case, you found that `r percent(mean(Y))` of the people polled said they would vote.

## Sample Versus Population

Our goal is to estimate the proportion of all registered voters - our population - who will vote. To do this, we use the proportion of polled voters - our sample - who said they would. When the election occurs, we get to see the actual turnout: `r approx(sum(y))` people, or roughly `r percent(mean(y))` of registered voters, actually vote.

:::: columns
::: column
### Before the Election 

$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}  \\
Y_i  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$

Outcomes for our sample of polled voters. \
To enumerate our sample, we ...

- give each call a number $i \in 1 \ldots `r n`$.
- write $Y_i$ for the turnout of the $i$th person we called.
:::
::: column
### After the Election

$$
\begin{array}{r|rrrrrr|r}
j & 1 & 2 & 3 & 4 & \dots &   `r approx(m)` & \bar{y}_{`r approx(m)`}  \\
y_{j} & `r y[1]` & `r y[2]` & `r y[3]` & `r y[4]` & \dots & `r y[m]`   & `r approx(mean(y))`  \\
\end{array}
$$
Outcomes for the population of registered voters. \
To enumerate our population, we ... 

- give each registered voter a number $j \in 1 \ldots `r approx(m)`$.
- write $y_j$ for the turnout of the person with ID $j$.
:::
::::

## Success!

Our sample proportion of `r sum(Y)`/`r n` ≈ `r approx(mean(Y))` is close to the population proportion of `r approx(sum(y))`/`r approx(m)` ≈ `r approx(mean(y))`. This level of accuracy helps maintain our reputation as a turnout pollster - at least for now.

However, one successful poll isn't enough unless we're planning to retire. We'll need to conduct more polls, and if our methods aren't reliable, we'll need to work on improving them. Even if they are reliable, overstating our accuracy could lead to problems. For example, we might be tempted to claim we'll be off by at most 2% based on this result, but if we're off by 4% next time, that could damage our credibility.

::: {.panel-tabset} 
### Important Questions

1. Was it luck that we got as close as we did?  
2. Could we have predicted how close we'd get *before* the election happened?

### A Plan

To find answers to these questions, we'll consider what would happen if our friends ran identical polls. Each friend would choose a different random sample Y_1 ... Y_{`r n`} and estimate the population proportion using the proportion in their sample. This will help us understand how accurate these estimates tend to be.

### On Terminology

This "friends" approach is simply an informal way of discussing the sampling distribution of our estimator. The sampling distribution represents the probability distribution of our estimator - specifically, the distribution of the turnout frequency in a sample of size n=`r n` drawn with replacement from the population y_1 ... y_{`r approx(m)`}. Each friend's estimate is, like ours, a random variable with this probability distribution.

:::

## Review: Connecting Sample and Population

When we conduct our polling, we follow a systematic process for each call. For each call i, we randomly select a voter with an ID we designate as J_i, and we record that voter's turnout as our call's outcome: Y_i = y_{J_i}. The selection process gives each registered voter a 1/`r approx(m)` chance of being called. We refer to this as sampling with replacement because it's possible to call the same person twice, though this is unlikely in our poll given that we're making a relatively small number of calls compared to the population size.

:::: {.r-stack}
::: {.fragment .fade-out data-fragment-index="0"}

```{r}
#| out-height: 300px 
sam$label = J
state.plot + geom_text(aes(x = long, y = lat, label=label), color=polla, alpha=.7, size=3, data=sam[c(1,2,n),]) 
```

:::
::: {.fragment .current-visible data-fragment-index="0"}
```{r}
#| out-height: 300px
state.plot + geom_text(aes(x = long, y = lat, label=label), color=polla, alpha=.7, size=3, data=sam)
```
:::
::: {.fragment .current-visible data-fragment-index="1"}
```{r}
#| out-height: 300px
state.plot + geom_point(aes(x = long, y = lat), color=polla, alpha=.7, size=3, data=sam)
```
:::
::::

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{variable}  & J_1 & Y_1 & J_2 & Y_2 & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\text{value}    & `r J[1]` & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`} 
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`} 
                  & `r approx(mean(Y))` \\
\end{array}
$$ 

:::: aside
Note: These locations are artificial and are used solely to illustrate the sampling scheme.
::::

## Our Poll

Let's examine our initial polling results. The visualization below shows the locations of the voters we sampled in our poll, represented by points on the map. Each point corresponds to a specific voter we contacted, and their responses are recorded in the accompanying table.

```{r}
J = J1
sam = pop[J,]
sam$label = J
Y = sam$y

state.plot + geom_point(aes(x = long, y = lat), color=polla, alpha=.7, size=3, data=sam)
```

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{variable}  & J_1 & Y_1 & J_2 & Y_2 & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\text{value}    & `r J[1]` & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`} 
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`} 
                  & `r approx(mean(Y))` \\
\end{array}
$$

## Our First Friend's Poll

To understand the variability in polling results, let's look at our first friend's poll. They conducted their survey using the same methodology but naturally contacted a different set of voters. The pink points on the map show the locations of the voters they sampled.

```{r}
J = J2
sam = pop[J,]
sam$label = J
Y = sam$y

state.plot + geom_point(aes(x = long, y = lat), color=pollb, alpha=.7, size=3, data=sam)
```

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{variable}  & J_1 & Y_1 & J_2 & Y_2 & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\text{value}     & `r J[1]` & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`} 
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`} 
                  & `r approx(mean(Y))` \\
\end{array}
$$

## Our Second Friend's Poll

Our second friend also conducted their own poll, providing us with yet another sample to analyze. The teal points on the map represent the voters they contacted, offering another perspective on how random sampling can lead to different but equally valid results.

```{r}
J = J3
sam = pop[J,]
sam$label = J
Y = sam$y

state.plot + geom_point(aes(x = long, y = lat), color=pollc, alpha=.7, size=3, data=sam)
```

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{variable}  & J_1 & Y_1 & J_2 & Y_2 & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\text{value}     & `r J[1]` & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`} 
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`} 
                  & `r approx(mean(Y))` \\
\end{array}
$$

## The Sampling Distribution of our Estimate

To better understand the variability in our polling results, let's examine the sampling distribution of our estimates. The table below shows the results from multiple polls, including ours and our friends', as well as many other simulated polls. Each row represents a different pollster's results, showing which voters they contacted and the responses they received.

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{pollster} & J_1 & Y_1 & J_2 & Y_2   & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\hline
\polla   1   
    & \polla   `r J=J1; Y=y[J]; J[1]`    & \polla  `r Y[1]` 
    & \polla   `r J[2]`   & \polla   `r Y[2]`           & \polla   \dots 
    & \polla   `r J[n]`   & \polla   `r Y[n]`           & \polla   `r approx(mean(Y))` \\
\pollb 2   
    & \pollb `r J=J2; Y=y[J]; J[1]`    & \pollb  `r Y[1]` 
    & \pollb `r J[2]`   & \pollb  `r Y[2]`          & \pollb \dots 
    & \pollb `r J[n]`   & \pollb `r Y[n]`           & \pollb `r approx(mean(Y))` \\
\pollc    3   
    & \pollc    `r J=J3; Y=y[J]; J[1]`    & \pollc     `r Y[1]` 
    & \pollc    `r J[2]`   & \pollc    `r Y[2]`           & \pollc    \dots 
    & \pollc    `r J[n]`   & \pollc    `r Y[n]`           & \pollc    `r approx(mean(Y))` \\
{\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots} &  {\vdots} \\ 
\polld    1M   
    & \polld    `r J=JM; Y=y[J]; J[1]`    & \polld     `r Y[1]` 
    & \polld    `r J[2]`   & \polld    `r Y[2]`           & \polld    \dots 
    & \polld    `r J[n]`   & \polld    `r Y[n]`           & \polld    `r approx(mean(Y))` \\
 {\vdots} & {\vdots} & {\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots}  & {\vdots} \\ 
\end{array}
$$

## The Sampling Distribution in **R**

To analyze this more systematically, we'll use R to simulate a large number of polls and examine their results. First, we'll set a random seed to ensure reproducibility of our results:

```{r}
#| echo: false
set.seed(default.seed)
```

We begin by running 10,000 simulated polls, storing both who we call (J) and their responses (Y):

```{r}
#| echo: true
Js = array(dim=c(10000, n))
Ys = array(dim=c(10000, n))
for(rr in 1:10000) {
  Js[rr,] = sample(m, n, replace=TRUE)
  Ys[rr,] = y[Js[rr,]]
}
```

Next, we calculate the sample proportion for each poll:

```{r}
#| echo: true
meanY.samples = rowMeans(Ys)
```

Finally, we create a histogram of these results to visualize the distribution of our estimates:

```{r}
#| echo: true
#| eval: false
ggplot() + geom_bar(aes(x=meanY.samples, y=after_stat(prop)), alpha=.2)  
```

::: {.panel-tabset }
### A Sketch

Before we look at the actual distribution, let's consider what we expect to see. Take a moment to think about where the histogram should be centered and what shape it might take. Consider:

```{r} 
ggplot() + geom_bar(aes(x=meanY.samples, y=after_stat(prop)), fill=NA, color=NA)
```
 
- Is the histogram centered where you would expect it to be?
- Does the shape make sense given what we know about sampling?
- Is the spread of the distribution what you would anticipate?

### The Real Thing

Now let's look at the actual distribution of our simulated polls:

```{r}
ggplot() + geom_bar(aes(x=meanY.samples, y=after_stat(prop)), alpha=.2)  
```

:::


## Annotating the Sampling Distribution Histogram 

```{r}
#| out-height: 500px
samples = meanY.samples
estimand = mean(y)
Y=y[J1] 
samples[1] = mean(Y)
samples[2] = mean(y[J2])
samples[3] = mean(y[J3])
samples[4] = mean(y[JM])

dot.height = .7*height
dot.size = 2
scale_x = scale_x_continuous(breaks=seq(.6,.8, by=.025), limits=c(.6, .8))

turnout.bar = ggplot() + 
  geom_bar(aes(x=samples, y=after_stat(prop)), alpha=.2) + scale_x 

turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=0)
```

In examining our sampling distribution histogram, several key features warrant attention. The solid blue line represents the mean of the sampling distribution, while the dashed blue lines demarcate the middle two-thirds of the distribution. The dotted blue lines encompass the middle 95% of the sampling distribution. Perhaps most notably, our estimation target - the actual turnout frequency in the population - is shown as a wide green line. Its perfect alignment with the solid blue line (making the blue line appear highlighted in green) provides important insights about the properties of our estimator.

##  Revisiting our Questions

```{r}
#| out-height: 500px
p.as.close    = mean(abs(samples - mean(y)) <= abs(mean(Y)-mean(y))) 
w66 = width(samples, alpha=1/3)
w95 = width(samples, alpha=.05)

turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=1, dot.size=dot.size, height=dot.height)
```

Looking at our results, we can observe that our estimate (represented by the black dot) comes remarkably close to our estimation target, with a difference of only `r percent(abs(mean(Y)-mean(y)))`. This naturally raises two important questions: Did we simply get lucky? And could we have predicted this level of accuracy before conducting the poll?

The evidence suggests that luck played only a minor role in our success. Our analysis shows that in 68% of polls, the estimator falls within `r percent(w66/2)` of the target, and in 95% of polls, it lands within `r percent(w95/2)` of the target. These findings indicate a consistent level of accuracy across multiple samples.

As for predicting our accuracy beforehand, the answer is yes, albeit with some qualifications. We can use interval estimation - specifying a range of values where we expect the estimation target to fall. The width of this interval addresses the question of accuracy, while the coverage probability (the likelihood that our interval contains the true value) provides important context for interpreting our results.

## Interval Estimation

```{r}  
#| out-height: 500px

w=.02
turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=1, interval.width=w, 
                   dot.size=dot.size, height=dot.height)
```

Our point estimate for the turnout frequency in our population is simply the turnout frequency in our sample: $\overline{Y}_{`r n`}$. To account for uncertainty, we can construct an interval of width `r w` centered on this estimate: $\overline{Y}_{`r n`} \pm `r w/2`$. While this width is somewhat arbitrary (chosen perhaps with the optimistic hope that being off by at most 1% would be acceptable), what matters is whether this interval successfully covers our estimation target - that is, whether the population frequency falls within our interval.

In this particular case, our interval doesn't contain the true value. This raises an important question: Is this just an unfortunate coincidence, or is it typical of intervals with width ±`r w/2`? To answer this, we should examine what happens when our friends apply the same approach to their polls.

## Interval Estimation: Our Friends' Polls

```{r}
#| out-height: 500px
turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=3, interval.width=w, 
                   dot.size=dot.size, height=dot.height)
```

Examining the interval estimates from our first two friends' polls provides additional insight. These estimates correspond to the pink and teal rows in our sampling distribution table. Among these three attempts, only the teal interval successfully covers the estimation target, giving us a coverage rate of just 1/3. While this might seem concerning, we should remember that three polls represent a very small sample. To get a more reliable sense of the coverage probability, we need to examine many more polls.

## Interval Estimation: 100 Polls

```{r}
#| out-height: 500px
imaginary.polls =  turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=100, interval.width=w, 
                   dot.size=dot.size, height=dot.height)
imaginary.polls
```

When we expand our analysis to include 100 polls, we find that `r sum(meanY.samples[1:100]-w/2 <= mean(y) & mean(y) <= meanY.samples[1:100]+w/2)`/100 of the intervals successfully cover the estimation target. This gives us a better approximation of the probability that our interval contains the true value. While we could achieve even greater precision by analyzing millions of polls, there's a more direct approach to calculating this coverage probability.

##  Coverage Probability and Sampling Distributions
```{r}
#| out-height: 500px
turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=100, interval.width=w, 
                   dot.size=dot.size, height=dot.height, 
                   highlighted.dot.alpha=.15, unhighlighted.dot.alpha=.15)
```

To understand how to calculate coverage probability using the sampling distribution of our point estimate $\bar Y_{`r n`}$, let's break this down into steps:

First, imagine you want to count how many of our 100 intervals cover the estimation target, but you'd prefer not to examine each horizontal segment individually (they're small and difficult to see). Instead, you can focus on the dots representing each poll's point estimate $\bar Y_{`r n`}$.

Think of an interval estimate as a point with "arms" of a certain length - similar to a person's reach. Just as you might check if you can touch someone by having them check if they can reach you, we can use this analogy to count successful coverage. This approach becomes particularly valuable when dealing with large numbers of polls.

## Calculating Coverage
```{r}
#| out-height: 500px
coverage = mean(mean(y)- w/2 <= meanY.samples & meanY.samples <= mean(y) + w/2)
coverage.hint.plot = turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=100, interval.width=w, 
                   dot.size=dot.size, height=dot.height, 
                   unhighlighted.dot.alpha=.15, highlighted.dot.alpha=.15,
                   shade.width=w, shade.color='green', shade.alpha=.2)
coverage.hint.plot
```

To simplify our coverage calculations, we've shaded an interval of width .02 centered on the estimation target in green. This shaded region has arms the same length as our interval estimates, and a point estimate falls within this region if and only if its corresponding interval estimate contains the target. This allows us to count successful coverage simply by counting the point estimates that fall within the shaded area.

## The Coverage Probability
```{r}
#| out-height: 500px
coverage.hint.plot
``` 

The coverage probability can be defined as the probability that a random draw from the sampling distribution falls within the green shaded area. Mathematically, we express this as:

$$
\text{coverage probability} = P\qty(\overline{Y}_{`r n`} \in \overline{y}_{`r approx(m)`} \pm .01) 
$$

Our calculations show this probability is approximately `r percent(coverage)`. We can verify this either by counting dots:

```{r}
#| echo: true
mean(mean(y) - .01 <= meanY.samples & meanY.samples <= mean(y) + .01)
```

Or by measuring the area of the histogram that falls within the green shaded region.

## Ad-hoc Interval Estimates
```{r}
#| out-height: 500px
coverage.hint.plot 
```

Our approach thus far has been to choose a width and then calculate a coverage probability. However, the coverage probability we found - `r percent(coverage)` - likely falls short of what we'd want. Consider how you would market your polling services: "I'm right about half the time - actually, slightly less than that" doesn't inspire much confidence. A 95% accuracy rate would be far more impressive, but achieving this would require wider intervals.

## Calibrated Interval Estimates
```{r}
#| out-height: 500px
coverage.hint.plot 
```

Instead of starting with a width and calculating coverage probability, we can work backward: choose a desired coverage probability (conventionally 95%) and determine what width we need to achieve it. An interval estimate calibrated in this way - to have a specific coverage probability - is called a confidence interval.

To calculate this width, we'll use the sampling distribution of our point estimate. Let's examine our annotated histogram again to understand how this works.

## Using the Sampling Distribution for Calibration

```{r}
#| out-height: 500px
coverage.hint.plot 
```

When constructing 95% confidence intervals around point estimates, we need to determine the appropriate width to achieve this coverage probability. To help us understand this, let's review our histogram's annotations:

- The solid blue line marks the mean of the sampling distribution (which aligns with our estimation target)
- The dashed blue lines encompass the middle two-thirds of the sampling distribution
- The dotted blue lines contain the middle 95% of the sampling distribution

## Using the Sampling Distribution for Calibration

```{r}
#| out-height: 500px

w = width(samples, alpha=.05)
dotted.width = w
coverage.good.plot = turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=100, interval.width=w, 
                   dot.size=dot.size, height=dot.height, 
                   unhighlighted.dot.alpha=.15, highlighted.dot.alpha=.15,
                   shade.width=w, shade.color='green', shade.alpha=.2)
coverage.good.plot
```

For a 95% confidence interval, we should use the width between the dotted blue lines - the span containing 95% of the estimates drawn from the sampling distribution. We can verify that these intervals achieve our desired coverage:

```{r}
#| echo: true
#|  
coverage = mean(mean(y)- dotted.width/2 <= meanY.samples & meanY.samples <= mean(y) + dotted.width/2)
coverage
```

## A Problem 
```{r}
#| out-height: 500px

imaginary.polls.calibrated =  turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=100, interval.width=w, 
                   dot.size=dot.size, height=dot.height)
imaginary.polls.calibrated
```

There's a significant challenge in applying this approach in real-world situations: when we conduct an actual poll, we only get one point estimate $\bar Y_{`r n`}$ based on our sample. We don't have access to the sampling distribution until after election day. However, our solution is quite elegant - we perform essentially the same process but use an estimate of the sampling distribution instead of the actual distribution. We'll explore this approach in our next class.

# Communication

## Why Confidence Intervals?

```{r}
#| out-height: 500px

imaginary.polls.calibrated
```

The use of calibrated interval estimates, or confidence intervals, offers several distinct advantages in statistical analysis. First, it focuses our attention on what we truly want to know: the location of the estimation target. Second, it acknowledges the inherent uncertainty in our estimates, reminding us that precise values are rarely attainable. Third, it provides a meaningful measure of how close we can expect our estimates to be.

However, confidence intervals can be conceptually challenging for those new to statistics. The fundamental issue lies in discussing hypothetical surveys that weren't actually conducted. Someone might reasonably ask, "I don't have time for imaginary surveys - what did this one tell you?" They might be frustrated to hear that, viewed in isolation, a single survey tells us very little with certainty.

This discomfort isn't a flaw in interval estimation - it reflects a fundamental challenge in sampling. While it might seem remarkable that we can draw meaningful conclusions about `r approx(length(y))` people from just 625 phone calls, we must also accept the limitations of this approach. When we say "we think it's between `r percent(mean(Y)-w/2)` and `r percent(mean(Y)+w/2)`, but we're wrong sometimes," we're being transparent about our uncertainty. We're precise about what "sometimes" means, even though this might feel more like a statement about our method than about what people want to know.

## Don't Make this Mistake

```{r}
#| out-height: 500px
 
imaginary.polls.calibrated 
```

The awkwardness of discussing hypothetical surveys often tempts people to make problematic statements. For instance, you might hear someone say "the turnout frequency in the election is in my confidence interval, `r round(mean(Y)-w/2,2)`-`r round(mean(Y)+w/2,2)`, 95% of the time." This statement is nonsensical - it's like saying "95% of the time, 2 is between 1 and 3." A value either is or isn't between two numbers; there's no probability involved unless the value itself is random. The bounds `r round(mean(Y)-w/2,2)` and `r round(mean(Y)+w/2,2)` are fixed, not random.

Fortunately, this issue rarely arises in practice because we can usually just say "confidence interval" without elaborating. The concept only needs careful explanation in two situations:

1. When someone asks what "confidence interval" means after you use the term
2. When you encounter someone using the term "confidence interval" without understanding its meaning

You'll have opportunities to practice explaining these concepts in homework assignments, quizzes, and possibly exams.
