---
title: "Lecture 2"
subtitle: "Interval Estimates for a Proportion"
format:
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---

```{r setup}
library(usmap)
library(sf)

library(ggplot2)

library(scales)
library(knitr)

percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }
```

```{r ggplot-themes}
class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
        panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x  = element_text(colour = "#aaaaaa"),
        axis.text.y  = element_text(colour = "#aaaaaa"),
        axis.title.x  = element_text(colour = "#aaaaaa"),
        axis.title.y  = element_text(colour = "#aaaaaa", angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
              axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
           panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(simplified.theme)
```

# Review and Warm-Up

## How Many Odd Numbers Are There Between 1 and 6?

```{r}
m = 6
number = 1:6
y = number %% 2

set.seed(1)
n = 5
J = sample(number, 5, replace=TRUE)
Y = y[J]

scale_dice=scale_x_continuous(breaks=(0:5)/5, labels=sprintf('%d/5', 0:5), limits=c(0,1))
```

- You could count them all, but that's tedious. 

$$
\begin{array}{r|rrrrrr|r}
j & 1 & 2 & 3 & 4 &  \dots &   `r m` & \bar{y}  \\
y_{j} & `r y[1]` & `r y[2]` & `r y[3]` & `r y[4]` & \dots & `r y[m]`   & \textcolor{blue}{`r approx(mean(y))`}  \\
\end{array}
$$

- Instead, you do a little experiment.
  - You roll a die 5 times.
  - You record the proportion of odds you get.

$$
\begin{array}{r|rrrrr|r}
i & 1 & 2 & 3 & 4 & 5 & \bar{Y}  \\
J_i & `r J[1]` & `r J[2]` & `r J[3]` & `r J[4]` & `r J[5]` & \\
Y_i & `r Y[1]` & `r Y[2]` & `r Y[3]` & `r Y[4]` & `r Y[5]` & \textcolor{red}{`r approx(mean(Y))`} \\
\end{array}
$$

- Much faster. And if you do bother to count them all... 
- You'll find [our estimate]{.fg text="style:red"} is pretty close to [the actual proportion]{.fg text="style:blue"}
- But is this approach trustworthy? After all, the answer we got *is* random.
- To talk about this, we'll think about the *sampling distribution* of our estimate.

## The Sampling Distribution ...

```{r}
#| out-height: 500px
Js = t(replicate(10000, sample(number, n, replace=TRUE)))
Ys = array(dim=dim(Js))
for (i in 1:10000) {
  Ys[i,] = y[Js[i,]]
}
dice.data = data.frame(p=rowMeans(Ys), y=runif(nrow(Ys)))
ggplot(dice.data) + geom_bar(aes(x=p, y=after_stat(prop))) + 
           geom_point(aes(x=mean(Ys[1,]), y=.1), color='red') +
           geom_point(aes(x=mean(Ys[2,]), y=.2), color='green') + scale_dice 
```

:::: {.columns}
::: {.column}
- is the distribution of estimates we'd get if we **replicated** our experiment \ over and over.
- Think of lots of different people rolling the dice and reporting what they got.
- We'll think about this by imagining that our friends did the same experiment. 
  - We'll give each friend a row in a table.
  - And we'll give their estimates a color so we can tell them apart.
:::
::: {.column}
$$
\begin{array}{r|rrrrr|r}
i & 1 & 2 & 3 & 4 & 5 & \bar{Y}  \\
\hline
J_i & `r J=Js[1,]; J[1]` & `r J[2]` & `r J[3]` & `r J[4]` & `r J[5]` & \\
Y_i & `r Y=y[J];   Y[1]` & `r Y[2]` & `r Y[3]` & `r Y[4]` & `r Y[5]` & \textcolor{red}{`r approx(mean(Y))`} \\
\hline
J_i & `r J=Js[2,]; J[1]` & `r J[2]` & `r J[3]` & `r J[4]` & `r J[5]` & \\
Y_i & `r Y=y[J];   Y[1]` & `r Y[2]` & `r Y[3]` & `r Y[4]` & `r Y[5]` & \textcolor{green}{`r approx(mean(Y))`} \\
\hline
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\end{array}
$$
:::
::::


## What Do We Do With This
```{r}
#| out-height: 500px
#| layout-ncol: 2
#| 
Js = t(replicate(10000, sample(number, n, replace=TRUE)))
Ys = array(dim=dim(Js))
for (i in 1:10000) {
  Ys[i,] = y[Js[i,]]
}
dice.data = data.frame(p=rowMeans(Ys), y=runif(nrow(Ys)))

ggplot(dice.data) + geom_bar(aes(x=p, y=after_stat(prop))) +
           geom_point(aes(x=mean(Ys[1,]), y=.1), color='red') +
           geom_point(aes(x=mean(Ys[2,]), y=.2), color='green') + scale_dice

## change sampling distn

Js = t(replicate(1000, sample(number, n, replace=TRUE)))
Ys = array(dim=dim(Js))
for (i in 1:1000) {
  Ys[i,] = Js[i,]==1
}
dice.data2 = data.frame(p=rowMeans(Ys), y=runif(nrow(Ys)))

ggplot(dice.data2) + geom_bar(aes(x=p, y=after_stat(prop))) +
           geom_point(aes(x=mean(Ys[1,]), y=.1), color='red') +
           geom_point(aes(x=mean(Ys[2,]), y=.2), color='green') + scale_dice
```


- This is a little counterintuitive as a summary of the proportion of odds between 1 and 6.
  - But it does tell us something. 
  - It's pretty different from what we'd get if there were only one odd number.
- One way we can make this sound more like the proportion we want is to say
  - what we've estimated
  - how close we tend to be
- For example, 
  - in `r percent(mean(dice.data$p==1/2))` of replications, we'll be exactly right
  - in `r percent(mean(abs(dice.data$p-1/2) <= 1/6))` of replications, we'll be off by no more than 1/6
- This gives us ...
  - an **interval** you can expect the proportion is in 
  - a statement about how much **confidence** you should have about that.

# Polling


```{r population}
our.seed = 0
set.seed(our.seed)
N=7233584
p=4999960/N
y= c(rep(1,p*N), rep(0, (1-p)*N))
y=sample(y) # shuffle the order 
population.proportion = p
```

```{r sample}
set.seed(our.seed)

n = 625
J = sample(N, n, replace=TRUE)
Y = y[J]
sample.proportion = mean(Y)
```


```{r histograms}
width = function(samples, alpha=.05) {
  covg.dif = function(w) { 
    actual = mean( mean(samples) - w/2 <= samples & samples <= mean(samples) + w/2)
    nominal = 1-alpha 
    actual - nominal  
  }
  uniroot(covg.dif, interval=c(0, 2*max(abs(samples-mean(samples)))))$root
}

height = .04
scale_x = scale_x_continuous(breaks=seq(.6,.8, by=.025), limits=c(.6, .8)) 
scale_y = scale_y_continuous(breaks=seq(0,height,by=.01), limits=c(0, height))
annotate.histogram = function(samples) { 
    w66 = width(samples, alpha=1/3)
    w95 = width(samples, alpha=.05)
    ggplot() + geom_bar(aes(x=samples, y=after_stat(prop)), alpha=.3) +
           geom_vline(aes(xintercept=mean(y)), color='green', size=3, alpha=.7) +
           geom_vline(aes(xintercept=mean(samples)), color="blue") +
           geom_vline(aes(xintercept=mean(samples)+c(-1,1)*w66/2), color="blue", linetype='dashed') +
           geom_vline(aes(xintercept=mean(samples)+c(-1,1)*w95/2), color="blue", linetype='dotted') +
           scale_x + scale_y
}

K=100
heights = sample(height*(1:K)/K)
default.highlight.colors = c('black', 'magenta', 'cyan', 'gray')
with.points = function(histplot, samples, K=100, highlight=4) {
    set.seed(our.seed)
    highlight.colors = if(highlight) default.highlight.colors[1:highlight] else c()
    heights = sample(height*(1:K)/K)
    data = data.frame(y=heights, x=samples[1:K], 
                      color=c(highlight.colors, rep('purple', K-length(highlight.colors))),
                      alpha=c(rep(1, length(highlight.colors)), rep(.1, K-length(highlight.colors))))
    histplot + geom_point(aes(y=y, x=x, color=I(color), alpha=I(alpha)), data=data)
}

with.intervals = function(histplot, samples, w=width(samples, alpha=.05), K=100, 
                          highlight=4, highlight.alpha=1) { 
    set.seed(our.seed)
    highlight.colors = if(highlight) default.highlight.colors[1:highlight] else c()
    heights = sample(height*(1:K)/K)
    data = data.frame(y=heights, x=samples[1:K], w=w,
                      color=c(highlight.colors, rep('purple', K-length(highlight.colors))),
                      alpha=c(rep(highlight.alpha, length(highlight.colors)), rep(.1, K-length(highlight.colors))))
    histplot + 
        geom_pointrange(aes(y=y, x=x, xmin=x-w/2, xmax=x+w/2, color=I(color), alpha=I(alpha)), data=data)
}
```

```{r mapping}
# Create fake locations for the people in our population
# by taking random convex combinations of state vertices
set.seed(0)
ga.shape = us_map(include = c("GA"))

# To reduce computation, we'll have 100K locations + jitter
M = 100000 
V = nrow(ga.shape)
weights = abs(rt(M*V,1))
dim(weights) = c(M,V)
weights = (1/rowSums(weights)) * weights 
base.x = weights %*% ga.shape$x
base.y = weights %*% ga.shape$y
base.ind = 1 + ( (0:(N-1)) %% M )
locations = data.frame(x=base.x[base.ind] + rnorm(N, sd=.01),
                       y=base.y[base.ind] + rnorm(N, sd=.01))

ga = ggplot() + geom_polygon(data=ga.shape, aes(x = x, y = y), fill="white", color="black") + coord_fixed() + blank.theme
```

## Context

- Suppose that, week before the 2020 presidential election, you did some polling.
  - You use a list of the $N \approx `r approx(N)`$ people registered to vote in Georgia.
  - And you make $n=`r n`$ phone calls.
  - Each call, you select a voter uniformly at random from the list, e.g. by rolling a `r approx(N)` -sided die.
  - And then you ask the potential voter whether they plan to vote.
- Suppose also that all these registered voters will
  - Pick up the phone will called
  - Respond honestly to your questions
  - Not change their minds about voting

## Polling Results

- You put your polling results in a table.
- It has one column for each call.
- In that columns, you record ...
  - the call number $i$---a number from $1 \ldots `r n`$.
  - the response $Y_i$ of the person you called---$1$ for 'yes and $0$ for no'.

$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}  \\
Y_i  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$ 

- You summarize the results with an extra column: the mean of the responses.
  - Remember that the mean of a binary variable is a proportion. 
  - You found that `r percent(mean(Y))` of the people polled said they would vote.

## Sample Versus Population

- We want to estimate the proportion of *all registered voters* --- our population --- who will vote.
- To do this, we use the proportion of *polled voters* --- our sample --- who said they would.
- When the election occurs, we get to see who turns out to vote.
  - `r approx(p*N)` people, or roughly `r percent(mean(Y))` of registered voters, actually vote.
 
\ \

:::: columns
::: column
### Before the Election 

$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}  \\
Y_i  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$

Outcomes for our sample of polled voters. \
To enumerate our sample, we ...

- give each call a number $i \in 1 \ldots `r n`$.
- write $Y_i$ for the turnout of the $i$th person we called.
:::
::: column
### After the Election

$$
\begin{array}{r|rrrrrr|r}
j & 1 & 2 & 3 & 4 & \dots &   `r approx(N)` & \bar{y}_{`r approx(N)`}  \\
y_{j} & `r y[1]` & `r y[2]` & `r y[3]` & `r y[4]` & \dots & `r y[N]`   & `r approx(mean(y))`  \\
\end{array}
$$
Outcomes for the population of registered voters. \
To enumerate our population, we ... 

- give each registered voter a # $j \in 1 \ldots `r approx(N)`$.
- write $y_j$ for the turnout of the person with ID $j$.
:::
::::

## Two Questions

Our sample proportion $`r sum(Y)` / `r n` \approx `r approx(mean(Y))`$ 
is close to the population proportion $`r approx(sum(y))` / `r approx(N)` \approx `r approx(mean(y))`$

1. Did we get lucky?
2. Could we have predicted how close we'd get before the election happened?


::: fragment

- Answers: No and Yes.
- To get these answers, we'll think about what'd happen if our friends had run identical polls.
  - Each friend will choose have a different random sample.
  - And estimate the population proportion by a different sample proportion.
- We'll see how close these estimates tend to be.

:::

## Review: Connecting Sample and Population

- For each call $i$, we randomly select a voter with an id we'll call $J_i$.
- And we record as that call's outcome the turnout of that voter: $Y_i=y_{J_i}$.
- On each call, each registered voter has a $1/`r approx(N)`$ chance of being called.
  - This is called sampling with replacement because we could call the same person twice.
  - In our poll, this is unlikely because we're making a small number of calls relative to population size.

:::: {.r-stack}
::: {.fragment .fade-out data-fragment-index="0"}

```{r}
#| out-height: 300px
sample.locations = locations[J, ]
sample.locations$label = J
ga + geom_text(aes(x = x, y = y, label=label), alpha=.5, data=sample.locations[c(1,2,n),]) 
```

:::
::: {.fragment .current-visible data-fragment-index="0"}

```{r}
#| out-height: 300px
ga + geom_text(aes(x = x, y = y, label=label), alpha=.2, data=sample.locations) 
```
:::
::: {.fragment .current-visible data-fragment-index="1"}
```{r}
#| out-height: 300px
ga + geom_point(aes(x = x, y = y), alpha=.2, data=sample.locations) 
```
:::
::::

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{select/record}  & J_1 & Y_1 & J_2 & Y_2 & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\text{outcome}    & `r J[1]` & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`} 
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`} 
                  & `r approx(mean(Y))` \\
\end{array}
$$ 

:::: aside
Obviously these locations are fake. Most people in Georgia live near Atlanta, not where these dots are. This is just to illustrate the sampling scheme. 
:::

## Our Poll

```{r}
# Now we'll repeat our sampling so we have polls for a few friends
set.seed(our.seed)
J1 = sample(N, n, replace=TRUE)
J2 = sample(N, n, replace=TRUE)
J3 = sample(N, n, replace=TRUE)
JM = sample(N, n, replace=TRUE)
```

```{r}
J = J1
Y = y[J1]

sample.locations = locations[J, ]
sample.locations$label = J
ga + geom_text(aes(x = x, y = y, label=label), alpha=.2, data=sample.locations) 
```


$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{select/record}  & J_1 & Y_1 & J_2 & Y_2 & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\text{outcome}    & `r J[1]` & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`} 
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`} 
                  & `r approx(mean(Y))` \\
\end{array}
$$

## Our First Friend's Poll

```{r}
J = J2
Y = y[J2]

sample.locations = locations[J, ]
sample.locations$label = J
ga + geom_text(aes(x = x, y = y, label=label), color='magenta', alpha=.2, data=sample.locations)
```

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{select/record}  & J_1 & Y_1 & J_2 & Y_2 & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\text{outcome}    & `r J[1]` & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`} 
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`} 
                  & `r approx(mean(Y))` \\
\end{array}
$$

## Our Second Friend's Poll

```{r}
J = J3
Y = y[J3]

sample.locations = locations[J,]
sample.locations$label = J
ga + geom_text(aes(x = x, y = y, label=label), color='cyan', alpha=.2, data=sample.locations)
```

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{select/record}  & J_1 & Y_1 & J_2 & Y_2 & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\text{outcome}    & `r J[1]` & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`} 
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`} 
                  & `r approx(mean(Y))` \\
\end{array}
$$

## The Sampling Distribution of our Estimate

\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{pollster} & J_1 & Y_1 & J_2 & Y_2   & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\hline
\color{black}   1   
    & \color{black}   `r J=J1; Y=y[J]; J[1]`    & \color{black}  `r Y[1]` 
    & \color{black}   `r J[2]`   & \color{black}   `r Y[2]`           & \color{black}   \dots 
    & \color{black}   `r J[n]`   & \color{black}   `r Y[n]`           & \color{black}   `r approx(mean(Y))` \\
\color{magenta} 2   
    & \color{magenta} `r J=J2; Y=y[J]; J[1]`    & \color{magenta}  `r Y[1]` 
    & \color{magenta} `r J[2]`   & \color{magenta}  `r Y[2]`          & \color{magenta} \dots 
    & \color{magenta} `r J[n]`   & \color{magenta} `r Y[n]`           & \color{magenta} `r approx(mean(Y))` \\
\color{cyan}    3   
    & \color{cyan}    `r J=J3; Y=y[J]; J[1]`    & \color{cyan}     `r Y[1]` 
    & \color{cyan}    `r J[2]`   & \color{cyan}    `r Y[2]`           & \color{cyan}    \dots 
    & \color{cyan}    `r J[n]`   & \color{cyan}    `r Y[n]`           & \color{cyan}    `r approx(mean(Y))` \\
{\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots} &  {\vdots} \\ 
\color{gray}    1M   
    & \color{gray}    `r J=JM; Y=y[J]; J[1]`    & \color{gray}     `r Y[1]` 
    & \color{gray}    `r J[2]`   & \color{gray}    `r Y[2]`           & \color{gray}    \dots 
    & \color{gray}    `r J[n]`   & \color{gray}    `r Y[n]`           & \color{gray}    `r approx(mean(Y))` \\
 {\vdots} & {\vdots} & {\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots}  & {\vdots} \\ 
\end{array}


## The Sampling Distribution in **R**

```{r}
#| echo: false
set.seed(our.seed)
```

We run 10,000 simulated polls and store who we call (J) and what they say (Y).
```{r}
#| echo: true
Js = array(dim=c(10000, n))
Ys = array(dim=c(10000, n))
for(rr in 1:10000) {
  Js[rr,] = sample(N, n, replace=TRUE)
  Ys[rr,] = y[Js[rr,]]
}
```

We calculate the sample proportion for each poll.
```{r}
#| echo: true
meanY.samples = rowMeans(Ys)
```

Histogram the result
```{r}
#| echo: true
#| out-height: 400px
ggplot() + geom_bar(aes(x=meanY.samples, y=after_stat(prop)), alpha=.3)  
```

## Annotating the Sampling Distribution Histogram 

```{r}
#| out-height: 500px

annotated.histogram = annotate.histogram(meanY.samples)
annotated.histogram
```

- There are a few features we may want to highlight.
  - The *mean of the sampling distribution* is the solid blue line.
  - The middle 2/3 of the sampling distribution lies between the dashed blue lines.
  - The middle 95\% of the sampling distribution lies between the dotted blue lines. 
- I've also drawn the *population proportion* as a wide green line.
  - We see that it appears as a kind of highlight on the solid blue line. 
  - It's in exactly the same place. What does this mean? 


##  Revisiting our First Question

```{r}
#| out-height: 500px
Y=y[J1] 
p.as.close    = mean(abs(meanY.samples - mean(y)) <= abs(mean(Y)-mean(y))) 
w66 = width(meanY.samples, alpha=1/3)
w95 = width(meanY.samples, alpha=.05)

annotated.histogram + geom_point(aes(x=mean(Y), y=heights[1]))
```


**Observation**. Our sample proportion --- the black dot --- is close to our population proportion.


**Q**. Did we get lucky? 

::: fragment
Not that lucky.

  - In `r percent(p.as.close)` of polls, we come as close as our estimate.
  - In 68% of polls, the sample proportion is within `r percent(w66/2)` of the population proportion.
  - In 95% of polls, the sample proportions are within  `r percent(w95/2)` of the population proportion.
  
:::

##  Revisiting our Second Question

```{r}
#| out-height: 500px
annotated.histogram + geom_point(aes(x=mean(Y), y=heights[1]))
```

**Q**. Could we have predicted how close we'd get before the election happened?

::: fragment
Yes, in sense.

  - We will use an interval estimate---a *range of values* the population proportion is likely to be in.
  - The *width* of this interval speaks to the 'how close' question. 
  - The *coverage probability* --- the probability we're right --- qualifies this answer.
:::

## Interval Estimation

```{r}  
#| out-height: 500px
w = .02
p.blank = ggplot() + geom_vline(aes(xintercept=population.proportion), color='green', size=1) +
  scale_x + scale_y
p.nofriends = p.blank + geom_pointrange(aes(y=heights[1], x=mean(Y), xmin=mean(Y)-w/2, xmax=mean(Y)+w/2))
p.nofriends
```

- Our *point estimate* of the population proportion is the sample proportion $\overline{Y}_{`r n`}$.
- So let's try an interval of w .02 centered on it: $\ \ \overline{Y}_{`r n`} \pm .01$.
- Ours covers --- includes --- the population proportion.
- Is that just chance? Let's check out our friends' intervals.

## Interval Estimation: Our Friends' Polls

```{r}
#| out-height: 500px
Y2 = y[J2]
Y3 = y[J3]
p.with2friends = p.nofriends + 
                  geom_pointrange(aes(y=heights[2], x=mean(Y2), xmin=mean(Y2)-w/2, xmax=mean(Y2)+w/2), color='magenta') +
                  geom_pointrange(aes(y=heights[3], x=mean(Y3), xmin=mean(Y3)-w/2, xmax=mean(Y3)+w/2), color='cyan') 
p.with2friends
```

- Lets's look at the interval estimates our first two friends.
  - i.e. estimates based on the magenta and cyan rows in our sampling distribution table.
-  One of these interval covers the population proportion.
- To get a better sense of how often this happens, let's do this for a hundred more polls.

## Interval Estimation: 100 Polls

```{r}
#| out-height: 500px
p.with2friends + geom_pointrange(aes(y=heights, x=meanY.samples[1:K], xmin=meanY.samples[1:K]-w/2, xmax=meanY.samples[1:K]+w/2), 
                                 alpha=.1, color='purple') 
```

- $`r sum(meanY.samples[1:K]-w/2 <= p & p <= meanY.samples[1:K]+w/2)`/100$ cover the population proportion.
- This gives us a sense of the *probability* that our interval covers the population proportion.
- If we want to be more precise, we could do the same for millions.
- Let's not. Instead, let's work out a trick for counting intervals that cover.


##  Coverage Probability and Sampling Distributions
```{r}
#| out-height: 500px
 
blank.histogram = ggplot() + geom_bar(aes(x=meanY.samples, y=after_stat(prop)), alpha=.2) + 
  scale_x + scale_y 
p.coverage.ex = with.intervals(blank.histogram, meanY.samples, w=.02, highlight=0) 
p.coverage.ex
```

 **Activity**. Calculate this *coverage probability* using the *sampling distribution* of our point estimate $\bar Y_{`r n`}$.\
 \ \
 
 Here's a hint.

  - Suppose we want to use a diagram like this to count the number of intervals that cover.
  - But we don't want to draw in the horizontal segments for each poll.
  - We just want to draw dots representing each poll's point estimate $\bar Y_{`r n`}$.
  - Can we draw in a few lines that'll help us count?


## Using Our Hint
```{r}
#| out-height: 500px
p.coverage.hint = p.coverage.ex + 
  geom_vline(aes(xintercept=population.proportion+c(-.01,.01)), color="red", linetype='dotted') +
  geom_rect(aes(xmin=population.proportion-w/2, xmax=population.proportion+w/2, 
                ymin=0, ymax=Inf,-w/2), alpha=.1, fill='red')
p.coverage.hint
```

- Let's draw an interval of width .02 centered on the population proportion---as a pair of dotted lines.
  - This gives it 'arms' the same length as our interval estimates have.
  - And its arms touch a point estimate *if and only if* the point estimates' arms touch it.
- That means we can count the intervals that cover by counting the point estimates between the dotted lines.
- What, in terms of the sampling distribution of the point estimate, is the coverage probability?

## The Coverage Probability
```{r}
#| out-height: 500px
p.coverage.hint
```

It's the probability that a random draw from the sampling distribution lies between the dotted lines.
$$
\text{coverage probability} = P\qty(\overline{Y}_{`r n`} \in \overline{y}_{`r approx(N)`} \pm .01) 
$$
And it's about 40\%.
```{r}
#| echo: true
mean(mean(y) - w/2 <= meanY.samples & meanY.samples <= mean(y) + w/2)
```

## Ad-hoc Interval Estimates
```{r}
#| out-height: 500px
p.coverage.hint
```

What we just did was choose a width and calculate a coverage probability.

  - The coverage probability we found --- 40% --- probably wasn't what we wanted.
  - Think of how you'd advertise your polling services.
  - 'I'm right about half the time. Actually, a bit less than that'.
  - 95% sounds a lot better. For that, we'll have to use a wider interval.

## Calibrated Interval Estimates
```{r}
#| out-height: 500px
with.intervals(annotated.histogram, meanY.samples, highlight=0) +
  geom_vline(aes(xintercept=population.proportion+c(-w/2,w/2)), color="red", linetype='dotted') +
  geom_rect(aes(xmin=population.proportion-w/2, xmax=population.proportion+w/2, ymin=0, ymax=Inf), alpha=.1, fill='red')
```

What we want to do is choose a coverage probability and calculate the right width.

  - An interval estimate calibrated like this---to have a given coverage---is called a *confidence interval*.
  - Let's think about how to do that. Again, we'll use the sampling distribution of our point estimate.
  - Let's take a look at our annotated histogram of point estimates again.


## Using the Sampling Distribution for Calibration

```{r}
#| out-height: 500px
with.intervals(annotated.histogram, meanY.samples, highlight=0) 
```

Remember what we've drawn in.

  - The *mean of the sampling distribution* is the solid blue line---and is the same as the population proportion.
  - The middle 2/3 of the sampling distribution lies between the dashed blue lines.
  - The middle 95\% of the sampling distribution lies between the dotted blue lines. 

## Using the Sampling Distribution for Calibration

```{r}
#| out-height: 500px
with.intervals(annotated.histogram, meanY.samples, highlight=0) 
```

**Activity**. 

  - Suppose you and your friends want to draw in intervals around your point estimates.
  - How wide should they be if you want 95\% of them to cover the population proportion?
  - Think about 'arms'. 

## Using the Sampling Distribution for Calibration

```{r}
#| out-height: 500px

w = width(meanY.samples, .05)
with.intervals(annotated.histogram, meanY.samples, highlight=0) + 
  geom_rect(aes(xmin=population.proportion-w/2, xmax=population.proportion+w/2, ymin=0, ymax=Inf), alpha=.1, fill='green')

dotted.width = w
```

**Answer**.

- The width of a 95% interval should be the width between the dotted blue lines.
- That's the width of the 'arms' containing 95% of the estimates drawn from the sampling distribution.
- You can check that these intervals have the coverage we want.

```{r}
#| echo: true
#|  
coverage = mean(mean(y)- dotted.width/2 <= meanY.samples & meanY.samples <= mean(y) + dotted.width/2)
coverage
```


## Summary

```{r}
#| out-height: 500px

with.intervals(annotated.histogram, meanY.samples, highlight=1) +  
  geom_rect(aes(xmin=population.proportion-w/2, xmax=population.proportion+w/2, ymin=0, ymax=Inf), alpha=.1, fill='green') 
```

- Our sample proportion `r approx(mean(Y))` is close to the population proportion `r approx(mean(y))`. Did we get lucky? 
  - Not that lucky. In `r percent(p.as.close)` of polls, we come as close as our estimate.
- Could we have predicted how close we would get before seeing the `r approx(mean(y))`? 
  - Yes! We can use a calibrated interval estimate---a Confidence Interval.
- Except that so far, we've cheated. We've used the sampling distribution of our point estimate.

## A Problem 
```{r}
#| out-height: 500px

with.intervals(annotated.histogram, meanY.samples, highlight=1) +
 geom_rect(aes(xmin=population.proportion-w/2, xmax=population.proportion+w/2, ymin=0, ymax=Inf), alpha=.1, fill='green')
```

- We can't calibrate intervals like this in real life.
  - When we run our a poll, we get a single point estimate $\bar Y_{`r n`}$ based on our sample.
  - We don't know the sampling distribution of this point estimate until election day.
- But what we actually do is *almost* the same. 
  - We *do* the same thing.
  - But we use an *estimate* of the sampling distribution in place of the thing itself.
- That's what we'll talk about next time.
