---
title: "Lecture 7"
subtitle: "Working With Non-Binary Outcomes"
format: 
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---


```{r load-libraries}
library(purrr)
library(tibble)
library(dplyr)
library(magrittr)
library(stringr)
library(ggplot2)
library(htmltools)
library(kableExtra)
library(scales)
library(cpsR)
library(mapdata)

options(knitr.kable.NA = '')

percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }

library(cpsR)
census.api.key = '8084f81e4f71e4d32638dac15f188b424f72d063'
# for cps variable interpretation see https://www2.census.gov/programs-surveys/cps/datasets/2022/march/asec2022_ddl_pub_full.pdf
```

```{r ggplot-themes}
class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
        panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x  = element_text(colour = "#aaaaaa"),
        axis.text.y  = element_text(colour = "#aaaaaa"),
        axis.title.x  = element_text(colour = "#aaaaaa"),
        axis.title.y  = element_text(colour = "#aaaaaa", angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
              axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
           panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(simplified.theme)
```

```{r table-styling}
ellipsis <- function(df, nhead=5L, ntail=2) {
    rownames(df) = as.character(1:nrow(df))
    stopifnot("data.frame" %in% class(df))
    els <- rep(NA, ncol(df)) %>% 
        matrix(nrow=1, dimnames=list(NULL, names(df))) %>% 
        data.frame(stringsAsFactors=FALSE) %>% 
        tibble::as_tibble()
    rownames(els) = "⋮"
    dplyr::bind_rows(head(df, nhead), els, tail(df, ntail))
}
```


```{r get-california-data}
#| cache: true
cps.sampling.ratio = 1/2500
plotting.cps.sampling.ratio=1/25
california.state.code = 6
cps.data = cpsR::get_asec(year=2022, vars=c('pearnval', 'gestfips', 'h_numper', 'htotval', 'a_age', 'a_hga','gereg', 'a_maritl', 'a_sex','peafever','a_wkstat','gtco'), key=census.api.key)
```

```{r process-california-data}
all.earnings = data.frame(age=cps.data$a_age,
                          state.code = cps.data$gestfips,
                          county.code = cps.data$gtco,
                          income=cps.data$pearnval, 
                          education=case_match(cps.data$a_hga,
                                          0  ~ 0,   # child
                                          31 ~ 0,   # < grade 1
                                          32 ~ 4,   #  grade 1-4
                                          33 ~ 6,   #  grade 5-6
                                          34 ~ 8,   #  grade 7-8 
                                          35 ~ 9,   #  grade 9
                                          36 ~ 10,  #  grade 10
                                          37 ~ 11,  #  grade 11
                                          38 ~ 11,  #  grade 12 no diploma
                                          39 ~ 12,  #  high school grad
                                          40 ~ 13,  #  some college
                                          41 ~ 14,  #  associate degree(vocational)
                                          42 ~ 14,  #  associate degree (academic)
                                          43 ~ 16,  #  bachelors degree
                                          44 ~ 18,  #  masters degree
                                          45 ~ 20,  #  professional school degree
                                          46 ~ 20)) #  doctorate

in.interval = function(X, interval) { interval[1] <= X & X <= interval[2] }
all.earnings$employed<- ifelse(cps.data$a_wkstat >5, 0,ifelse(cps.data$a_wkstat<2,NA,1))
all.earnings$degree = factor(all.earnings$education >= 16,
                    levels=c(FALSE, TRUE), 
                    labels=c('no 4-year degree','4-year degree')) 

earnings = all.earnings[all.earnings$age %>% in.interval(c(25,35)) & 
                        all.earnings$state.code == california.state.code &
                        all.earnings$education > 8, ]

rownames(earnings)=NULL
earnings$person = 1:nrow(earnings)
n = nrow(earnings)
```



```{r maps}
#| cache: true

## All of this is just to make maps showing the sample and population.
## And this code will be used later, so some of what's written here isn't 
## used in the lecture. I wouldn't read it unless you're interested in drawing maps in R.

# county info from https://en.wikipedia.org/wiki/List_of_counties_in_California
# converted to csv using https://wikitable2csv.ggor.de
to.number = function(s) { as.numeric(str_trim(gsub(",", "", s))) }
ca_county_info = read.csv('california_county_info.csv') %>%
  mutate(subregion = tolower(sub(' County', '', County))) %>%
  mutate(pop = to.number(Population..2022.)) %>%
  mutate(area = to.number(sub('sq.*', '', Area))) 

earnings.county = left_join(earnings, ca_county_info[,c('FIPS.code', 'subregion')], by=c('county.code' = 'FIPS.code'))
earnings$county = earnings.county$subregion
earnings$county[is.na(earnings$county)]='unknown'
  
# To give our survey recipients a bit more reality, we'll give them fake locations on a map
# And we'll make up a population of California residents that they're sampled from
# maps are based on code from https://eriqande.github.io/rep-res-eeb-2017/map-making-in-R.html

counties = map_data("county")
ca_county = counties %>%
  filter(region == "california")
ca_county_centers = ca_county %>%
  group_by(subregion) %>%
  summarize(county.lat = mean(lat), county.long=mean(long))

ca_county_more = left_join(ca_county_centers, ca_county_info, by="subregion")
ca_all = left_join(ca_county, ca_county_more, by="subregion")


ca.plot = ggplot(data=ca_all, mapping=aes(x=long, y=lat, group=group)) + 
  coord_quickmap() +
  geom_polygon(fill='gray', color = "white", alpha=.4) +
      scale_fill_gradient(trans="log10") + 
      theme_void()

ca.pop.plot = ggplot(data=ca_all, mapping=aes(x=long, y=lat, group=group)) + 
  coord_quickmap() +
  geom_polygon(aes(fill = pop/area), color = "white") +
      scale_fill_gradient(trans="log10") + 
      theme_void()

## we'll make up a population as follows
# 1. choose the number of people to be the number of people in our sample scaled up to account for the cps sampling ratio
# 2. for each, sample a county with probability proportional to its finite.population
# 3. for each, sample a latitude and longitude by taking a randomly weighted combination of the    
#    county's vertices. As long as the county is convex, these points will be inside the county.
# 4. we'll give them a degree too, chosen independently of all else from the empirical distribution on the sample
npop = nrow(earnings) / plotting.cps.sampling.ratio

# takes a list of subregions, possibly with duplicates
# for each element, chooses random lat/long coordinates within the sample.subregion
# by taking a random convex combination of the subregion's vertices
sample.within.subregion = function(subregions) {
  output = data.frame(subregion=subregions, lat=NA, long=NA)
  for(subregion in unique(subregions)) { 
      vertices = ca_county[ca_county$subregion==subregion, c('lat','long')]
      U = sum(subregions==subregion)
      V = nrow(vertices)
      weights = abs(rt(U*V, 1))
      dim(weights) = c(U,V)
      weights = (1/rowSums(weights)) * weights
      output[subregions==subregion, c('lat', 'long')] = 
                 cbind(weights %*% vertices$lat, weights %*% vertices$long)
    }
    output
}

finite.population.subregion = sample(ca_county_more$subregion, size=npop, replace=TRUE, prob=ca_county_more$pop)
finite.population = sample.within.subregion(finite.population.subregion)
finite.population$degree = sample(earnings$degree, nrow(finite.population), replace=TRUE)
finite.population$group=1 # for plotting

display.population = sample_n(finite.population, min(100000, nrow(finite.population)))

# and we'll force the degree-distribution in the sample to match our survey responses
our.sample = sample_n(finite.population, nrow(earnings))
our.sample$degree=earnings$degree

pop.plot = ca.plot + geom_point(aes(x=long, y=lat), data=display.population, size=.02, alpha=.1) 
sample.plot = ca.plot + geom_point(aes(x=long, y=lat), data=our.sample, size=.2, alpha=.5)
pop.plot.inf = ca.pop.plot + geom_point(aes(x=long, y=lat), data=display.population, size=.01, alpha=.1)  
sample.plot.inf = ca.pop.plot + geom_point(aes(x=long, y=lat), data=our.sample, size=.2, alpha=.5)
```



## So Far

::: {.hidden}
$$
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\V}{V}
\DeclareMathOperator{\Var}{\V}
\newcommand{\inred}[1]{\textcolor{red}{#1}}
\newcommand{\inblue}[1]{\textcolor{blue}{#1}}
$$ 

:::

```{r}
our.seed = 0
width = function(samples, alpha=.05) {
  covg.dif = function(w) { 
    actual = mean( mean(samples) - w/2 <= samples & samples <= mean(samples) + w/2)
    nominal = 1-alpha 
    actual - nominal  
  }
  uniroot(covg.dif, interval=c(0, 2*max(abs(samples-mean(samples)))))$root
}

default.height = .04
scale_x = scale_x_continuous(breaks=seq(.6,.8, by=.025), limits=c(.6, .8)) 
scale_y = scale_y_continuous(breaks=seq(0,default.height,by=.01), limits=c(0, default.height))

annotate.histogram = function(histplot, samples, shade=FALSE, 
                              shade.width=width(samples, alpha=.05), shade.color='green', shade.alpha=.15) { 
    w66 = width(samples, alpha=1/3)
    w95 = width(samples, alpha=.05)
    p = histplot + 
           geom_vline(aes(xintercept=mean(samples)), color="blue") +
           geom_vline(aes(xintercept=mean(samples)+c(-1,1)*w66/2), color="blue", linetype='dashed') +
           geom_vline(aes(xintercept=mean(samples)+c(-1,1)*w95/2), color="blue", linetype='dotted') 
    if(shade) { 
      p + geom_rect(aes(xmin=mean(samples)-shade.width/2, 
                        xmax=mean(samples)+shade.width/2, 
                        ymin=0, ymax=Inf), alpha=shade.alpha, fill=shade.color)
    } else { p }
}

annotated.bar = function(samples, ...) { 
    p = ggplot() + geom_bar(aes(x=samples, y=after_stat(prop)), alpha=.3) +
        geom_vline(aes(xintercept=mean(y)), color='green', size=3, alpha=.7)
    annotate.histogram(p, samples, ...) + scale_x + scale_y
}

default.highlight.colors = c('black', 'magenta', 'cyan', 'gray')
with.points = function(histplot, samples, K=100, highlight=4, height=default.height) {
    set.seed(our.seed)
    highlight.colors = if(highlight) default.highlight.colors[1:highlight] else c()
    heights = sample(height*(1:K)/K)
    data = data.frame(y=heights, x=samples[1:K], 
                      color=c(highlight.colors, rep('purple', K-length(highlight.colors))),
                      alpha=c(rep(1, length(highlight.colors)), rep(.1, K-length(highlight.colors))))
    histplot + geom_point(aes(y=y, x=x, color=I(color), alpha=I(alpha)), data=data)
}

with.intervals = function(histplot, samples, w=width(samples, alpha=.05), K=100, 
                          highlight=4, highlight.alpha=1, height=default.height) { 
    set.seed(our.seed)
    highlight.colors = if(highlight) default.highlight.colors[1:highlight] else c()
    heights = sample(height*(1:K)/K)
    data = data.frame(y=heights, x=samples[1:K], w=w,
                      color=c(highlight.colors, rep('purple', K-length(highlight.colors))),
                      alpha=c(rep(highlight.alpha, length(highlight.colors)), rep(.1, K-length(highlight.colors))))
    histplot + 
        geom_pointrange(aes(y=y, x=x, xmin=x-w/2, xmax=x+w/2, color=I(color), alpha=I(alpha)), data=data)
}
```

```{r}
#| cache: true
social.pressure = read.csv('social-pressure-data.csv') 
keep = social.pressure$treatment %in% c(' Control', ' Neighbors')
D = as.numeric(social.pressure[keep,'treatment'] == ' Neighbors')
Y = as.numeric(social.pressure[keep,'voted'] == 'Yes')
n = length(Y)

difference = mean(Y[D==1]) - mean(Y[D==0])
difference.bootstrap.samples = array(dim=10000)
for(rr in 1:10000) {
    I = sample(1:n, n, replace=TRUE)
    Dstar = D[I]
    Ystar = Y[I]
    difference.bootstrap.samples[rr] = mean(Ystar[Dstar==1]) - mean(Ystar[Dstar==0])
}
```

```{r}
#| out-height: 250px
#| fig-asp: .25 
 
xx = seq(.05, .1, length=1000)
scale_x_dif = scale_x_continuous(limits=c(.07,.095), labels=label_percent())
wboot=width(difference.bootstrap.samples, alpha=.05)

difference.bootstrap.samples = difference.bootstrap.samples - mean(difference.bootstrap.samples) + difference
pboot=ggplot() + geom_histogram(aes(x=difference.bootstrap.samples, y=after_stat(density)), alpha=.3, bins=100) 
annotated.bootstrap.difference.histogram = annotate.histogram(pboot, difference.bootstrap.samples,
  shade=FALSE)

with.points(annotated.bootstrap.difference.histogram, difference.bootstrap.samples, height=150, highlight=0) + 
  geom_pointrange(aes(x=x,y=y, xmin=x-w/2, xmax=x+wboot/2), 
                  data=data.frame(x=difference, y=50, w=wboot), color='black')  +
  geom_line(aes(x=xx, y=dnorm(xx, mean(difference.bootstrap.samples), sd(difference.bootstrap.samples))), color='cyan') +
  scale_x_dif
``` 


- We've worked with binary outcomes. In particular, the outcome of voting.
- We've estimated turnout: the proportion of people in our **population** of registered voters who do actually vote.
- We've talked about how to estimate this using a **random sample** of these voters, e.g., from a telephone survey.
- To summarize how close our estimate is to the actual turnout rate, we've used confidence intervals.
  - We calibrate them so that, 95% of the time, they contain what we're trying to estimate.
  - For calibration, we've used the **bootstrap sampling distribution** as an estimate of the sampling distribution.
- And we've talked about using **normal approximation** to understand what it'll all look like.
  - e.g. to estimate the precision we'll get from a given sample size.
- We've done all this for *turnout differences* between groups of voters, too. 
  - e.g. Black and non-Black voters.
  - e.g. Voters who received a letter pressuring them to vote and voters who didn't.

::: aside
Shown above: an estimate of the difference in turnout between people who received a letter pressuring them to vote and people who didn't.
:::

## Today

```{r visualization-defaults}
scale_y = function(limits=c(0, 4e5)) { 
  scale_y_continuous(labels=label_dollar(), limits=limits)
}

dot.size = 2
```

:::: columns

::: column
```{r map-sample}
#| fig-width: 5
#| out-height: 300px
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=our.sample, size=.1, alpha=.5) + guides(color=FALSE)
```
:::

::: column

```{r table}
tab = earnings[sample(nrow(earnings)), ]
tab$income = sprintf('$%dk', round(as.numeric(tab$income) / 1000))
ellipsis.tab = tab[, c('income', 'degree', 'age', 'county')] %>% ellipsis(nhead=4,ntail=2)
ellipsis.tab %>%
  kable %>% kable_styling() %>% 
  row_spec(which(ellipsis.tab$education >= 16),  background=twocolor.green) %>%
  row_spec(which(ellipsis.tab$education <  16),  background=twocolor.red) 
```
:::

::::

- We'll do the same thing, but for **non-binary outcomes**.
- Using data from the Current Population Survey, we'll estimate ...
  - the mean income in our population.
  - the mean income among people in our population [with 4-year degrees]{.twocolor-green} 
  - the difference in mean income between people [with 4-year degrees]{.twocolor-green} and people [without them]{.twocolor-red}.
- Our population will be the set of California residents 
    - between the ages of 25 and 35
    - with at least an 8th-grade education


::: aside
The locations shown on the map are made up. They're not the actual locations of the people in the sample. \
The survey includes some location information for some people, but as you can see in the table, not for everyone.
:::


## Sampling

```{r pop-sample-maps}
#| out-height: 500px
#| fig-asp: .8 
#| layout-ncol: 2 
#| 
ca.plot + 
 geom_point(aes(x=long, y=lat, color=degree), data=display.population, size=.02, alpha=.1) + 
 guides(color=FALSE)
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=our.sample, size=.1, alpha=.5) 
```

- We'll act as if our data were sampled, with replacement, from this population.
  - On the left, I've illustrated the population.
  - On the right, I've illustrated our sample.
- Our working assumption: each dot on the right was chosen, from those on the left, by rolling a big die.
- That's not quite right, but we'll stick with it throughout the semester.

## Sample and Population Means

```{r}
#| out-height: 300px
#| layout-ncol: 2
#| 
pop.earnings = earnings[sample(nrow(earnings), nrow(display.population), replace=TRUE), ]
pop.earnings$income = pop.earnings$income + 3000*rnorm(nrow(pop.earnings))

ggplot(pop.earnings) + 
  geom_point(aes(x=0, y=income),  color='blue',
            size=.2, position=position_jitter(seed=0,w=.3), alpha=.05) +
  stat_summary(aes(x=0, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x))  +
  scale_y() 

ggplot(earnings) + 
  geom_point(aes(x=0, y=income), color='blue', 
            size=2, position=position_jitter(seed=0,w=.3), alpha=.2) +
  stat_summary(aes(x=0, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x))  +
  scale_y() 
```

- One thing we often want to estimate is the mean of our population.

$$
\mu = \frac{1}{m}\sum_{j=1}^m y_j 
$$

- We'll think of the mean in our sample as an estimate of it.

$$ 
\hat \mu = \frac{1}{n}\sum_{i=1}^n Y_i 
$$


## Subsample and Subpopulation means

```{r}
#| out-height: 300px
#| layout-ncol: 2 

earnings.jitter.pop = ggplot(pop.earnings) + 
    geom_point(aes(x=degree, y=income, color=degree), 
               size=.2, position=position_jitter(seed=0,w=.3), alpha=.1) +
     scale_y() + guides(color='none')

earnings.jitter = ggplot(earnings) + 
    geom_point(aes(x=degree, y=income, color=degree), 
               size=dot.size, position=position_jitter(seed=0,w=.3), alpha=.1) +
     scale_y() + guides(color='none')

earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

- We'll often be interested in the mean income in subpopulations, too.
  - We'll think about the subsample with $X_i=x$ for some value $x$. 
  - e.g. $X_i=1$ for people [with a 4-year degree]{.twocolor-green} and $X_i=0$ for people [without]{.twocolor-red}.
$$
\mu(x) = \frac{1}{m_x}\sum_{j:x_j = x } y_j \quad \text{ where } \quad m_x = \sum_{j:x_j=x} 1.
$$

- We'll use the mean in the corresponding subsample to estimate it.

$$ 
\hat \mu(x) = \frac{1}{N_x}\sum_{i:X_i=x} Y_i \quad \text{ where } \quad N_x = \sum_{i:X_i=1} 1.
$$


## Observations as Random Variables

```{r}
#| out-height: 300px
#| layout-ncol: 2 

summaries = earnings %>% group_by(degree) %>% summarize(mean=mean(income), sd=sd(income))

earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

- To talk about the relationship between the sample and population, we'll use **random variables**.
- The notation $y_j$ refers to a number.
  - It's the income of the $j$th person in our population. That's just some person. 
- The notation $Y_i$ refers to something else.
  - It's the income of the $i$th person we call in our survey. That's not a person.
  - That's the result of a random process---the roll of a die.
  - And that person is person $1 \ldots m$ each with probability $1/m$.
- To summarize this result, we talk about the **probability distribution** of $Y_i$.

## Probability Distributions: The Binary Case
```{r}
set.seed(0)
tab = earnings[sample(nrow(earnings)), ]
tab$income.exceeds.100k = tab$income > 100000
ellipsis.tab = tab[, c('income.exceeds.100k', 'degree', 'age', 'county')] %>% ellipsis(nhead=4,ntail=2)
ellipsis.tab %>%
  kable %>% kable_styling() %>% 
  row_spec(which(ellipsis.tab$education >= 16),  background=twocolor.green) %>%
  row_spec(which(ellipsis.tab$education <  16),  background=twocolor.red) 

```

- When our outcomes are binary, it's easy to describe this distribution.
  - All we need to know is the probability that $Y_i$ is 1. Why?
  - To do that, we **sum the probabilities** of the rolls that result in it being one.

$$
P(Y_i = 1) = \sum_{j: y_j = 1} P(\text{roll}_i = j) = \sum_{j: y_j = 1} \frac{1}{m} \times y_j = \mu.
$$

- This **collapses** the information about what happens in our random process---the probability we roll each number in 1...m---into the information that's relevant to $Y_i$.
- That collapsing is what gives us our 'weighted coin flip'.


## Probability Distributions: The General Case
```{r}
set.seed(0)
tab = earnings[sample(nrow(earnings)), ]
tab$income = sprintf('$%dk', round(as.numeric(tab$income) / 1000))
ellipsis.tab = tab[, c('income', 'degree', 'age', 'county')] %>% ellipsis(nhead=4,ntail=2)
ellipsis.tab %>%
  kable %>% kable_styling() %>% 
  row_spec(which(ellipsis.tab$education >= 16),  background=twocolor.green) %>%
  row_spec(which(ellipsis.tab$education <  16),  background=twocolor.red) 
```

- When our outcomes are are nonbinary, it's a bit more complicated.
  - We need to know the probability that $Y_i$ takes on each possible value.
  - But we calculate it the same way. We sum the probabilities of the rolls that result in it taking on those values.

$$
P(Y_i = y) = \sum_{j: y_j = y} P(\text{roll}_i = j) 
$$

- We're still collapsing out irrelevant information, but what we're left with is more complicated.
- It's a **weighted die roll**, with one face for each possible value of $Y_i$.
- And sometimes we're effectively back where we started: if each person's income is different, then there's nothing to collapse out: $P(Y_i = y) = 1/m$ for all $y$.

## Expectations

```{r}
set.seed(0)
tab = earnings[sample(nrow(earnings)), ]
tab$income = sprintf('$%dk', round(as.numeric(tab$income) / 1000))
ellipsis.tab = tab[, c('income', 'degree', 'age', 'county')] %>% ellipsis(nhead=4,ntail=2)
ellipsis.tab %>%
  kable %>% kable_styling() %>% 
  row_spec(which(ellipsis.tab$education >= 16),  background=twocolor.green) %>%
  row_spec(which(ellipsis.tab$education <  16),  background=twocolor.red) 
```

- That said, we often don't need to know the whole distribution.
- Often the only thing we care about is the expected value of $Y_i$. Or some related quantity.
  - The expected value of $Y_i$ is the **probability-weighted average** of the values it can take on.
  - What's nice about this is that we can think of this in 'uncollapsed' terms.
- When we sample as usual, this is just the population mean.
- The collapsed form is, in a sense, just summing in a different order.

$$
\begin{aligned}
\E[Y_i] = \sum_y P(Y_i = y) \times y
&= \sum_y \qty(\sum_{j: y_j = y} \frac{1}{m}) \times y \\
&= \frac{1}{m}\sum_y\sum_{j:y_j = y} y \\
&= \frac{1}{m}\sum_{j=1}^m y_j
\end{aligned}
$$

- What's nice about the binary case is, in fact, that it's an expectation.
- When $Y_i$ is binary, the probability it's 1 is its expected value $Y_i$.

$$
P(Y_i=1) = \sum_{j: y_j = 1} P(\text{roll}_i = j) = \frac{1}{m}\sum_{j:1}^m \begin{cases} 1 & \text{ if } y_j = 1 \\ 0 & \text{ otherwise} \\ \end{cases} = \frac{1}{m}\sum_{j:1}^m y_j = \E[Y_i].
$$


## Independence 

- We say two random variables are **independent** if knowing the value of one doesn't tell us anything about the value of the other.
- When this is the case, their joint probability distribution is just the product of their individual ones.

$$
P(Y_1=y_1 \quad Y_2=y_2) = P(Y_1=y_1)P(Y_2=y_2).
$$

- That's what happens when the randomness in $Y_1$ and $Y_2$ comes from two different rolls of the die.
- And since that's how we're doing our sampling --- that's sampling with replacement---it'll be true in our sample.
- Because we get each of our observations the same way, they also have the same probability distribution. 
- We say they're **independent and indentically distributed.**

## Conditioning

```{r}
#| out-height: 300px
#| layout-ncol: 2 

earnings.jitter.pop = ggplot(pop.earnings) + 
    geom_point(aes(x=degree, y=income, color=degree), 
               size=.2, position=position_jitter(seed=0,w=.3), alpha=.1) +
     scale_y() + guides(color='none')

earnings.jitter = ggplot(earnings) + 
    geom_point(aes(x=degree, y=income, color=degree), 
               size=dot.size, position=position_jitter(seed=0,w=.3), alpha=.1) +
     scale_y() + guides(color='none')

earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

- **Conditioning** is, in effect, a way of thinking about sampling as a two-stage process.
  - First, we choose the color of our dot, i.e., the value of $X_i$, according to it frequency in the population.
  - Then, we choose a specific one of those dots, i.e. $J_i$, from those with that color---with equal probability.
- Because this is just a way of thinking, each person still gets chosen with probability $1/m$.
$$
P(J_i=j) = 
\begin{cases}
   \frac{m_{green}}{m} \ \ \ \ \ \times \frac{1}{m_{green}} \ \ \ \ \ =\ \frac{1}{m} & \text{if the $j$th dot is green  ($x_j=1$) } \\
    \frac{m-m_{green}}{m} \times \frac{1}{m-m_{green}} \ = \ \frac{1}{m} & \text{otherwise}
\end{cases}
$$
  


## Conditioning

```{r}
#| out-height: 300px
#| layout-ncol: 2 

earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```


- The **Conditional Probability** of $Y_i$ is the probability resulting from the second stage.
- It's a function of the result of the first. 
  - $P(Y_i=y \mid X_i=1) is the probability distribution of $Y_i$ when we're rolling the 'green die'.
  - $P(Y_i=y \mid X_i=0) is the probability distribution of $Y_i$ when we're rolling the 'red die'.
- And the **Conditional Expectation** of $Y_i$ is the 'second stage expected value' in the same sense.
  - $\E[Y_i \mid X_i=1]$ is the expected value of $Y_i$ when we're rolling the 'green die'. 
  - That is, the mean value $\mu(1)$ of $y_j$ in the subpopulation drawn as little green dots.
  - $\E[Y_i \mid X_i=0]$ is the expected value of $Y_i$ when we're rolling the 'red die'.
  - That is, the mean value $\mu(0)$ of $y_j$ in the subpopulation drawn as little red dots.

## The Dry Part Begins

- Up to this point in the class, we've mostly focused on concepts and intuition.
- We've done some calculations, but there have been parts that haven't been totally rigorous.
- Today, and from here on, we won't be able to do that.
- Let's take a few minutes to review some of the math we'll need.

# Review: Expectations and Conditional Expectations

## Conditioning

```{r}
#| out-height: 300px
#| layout-ncol: 2 

earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

### The law of iterated expectations
$$
E \{ E( Y \mid X ) \} \quad \text{ for any random variables $X, Y$} 
$$

To calculate the mean of $Y$, we can average within subpopulations, then across subpopulations.

### Irrelevance of independent conditioning variables
$$ E( Y \mid X, X' ) = E( Y \mid X ) \quad \text{ when $X'$ is independent of $X,Y$ } $$

If $X'$ is unrelated to $X$ and $Y$, holding it constant doesn't affect the relationship between them.

## Linearity of Expectations

```{r}
#| out-height: 300px
#| layout-ncol: 2 

earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

$$ 
\begin{aligned}
E ( a Y + b Z ) 
&= E (aY) + E (bZ) \\
&= aE(Y) + bE(Z)  \\
& \text{ for random variables $Y, Z$ and numbers $a,b$ }
\end{aligned}
$$

  - There are two things going on here.
    - To average a sum of two things, we can take two averages and sum.
    - To average a constant times a random variable, we can multiply the random variable's average by the constant.
  - In other words, we can *distribute* expectations and can pull constants out of them.
  
## Proof

- In essence, it comes down to the fact that all we're doing is summing.
	- Expectations are probability-weighted sums.
	- And we're looking at the expectation of a sum.
- And we can change the order we sum in without changing what we get.

$$
\small{
\begin{aligned}
\E\qty( a Y + b Z ) 
&= \sum_{y}\sum_z (a y + b z) \ P(Y=y, Z=z) && \text{ by definition of expectation} \\
&= \sum_{y}\sum_z a y \ P(Y=y, Z=z) + \sum_{z}\sum_y b z \ P(Y=y, Z=z) && \text{changing the order in which we sum} \\
&= \sum_{y} a y \ \sum_z P(Y=y,Z=z) + \sum_{z} b z \ \sum_y P(Y=y,Z=z) && \text{pulling constants out of the inner sums} \\
&= \sum_{y} a y \ P(Y=y) + \sum_{z} b z \ \sum_y P(Z=z)  && \text{summing to get marginal probabilities from our joint } \\
&= a\sum_{y} y \ P(Y=y) + b\sum_{z} z \ \sum_y P(Z=z)  && \text{ pulling constants out of the remaining sum } \\
&= a\E Y + b \E Z  && \text{by definition}
\end{aligned}
}
$$

## Linearity of **Conditional** Expectations

```{r}
#| out-height: 300px
#| layout-ncol: 2 

earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

$$ 
\begin{aligned}
E\{ a(X) Y + b(X) Z \mid X \} 
&= E\{a(X)Y \mid X\} + E\{ b(X)Z \mid X\} \\
&= a(X)E(Y \mid X) + b(X)E(Z \mid X)
\end{aligned}
$$

- This is like linearity of expectations, but with a twist.
   - When we condition on $X$, we're working with subpopulations in which $X$ is constant.
   - This means we can act as if functions of $X$ **are** constants.
   - So we can pull them out of expectations that are conditional on $X$.
   
## It's important to distinguish between two things

```{r}
#| out-height: 250px
#| fig-asp: .25
Y = c(.75,1,1.25, 1,1.25,1.5)
X = 0 + (1:6 > 3)
Xdodge = X + rep(.02*c(-1,0,1),2)
mu = function(x) { mean(Y[X==x]) }

tinypop.plot = ggplot() + geom_point(aes(x=Xdodge, y=Y, color=factor(X)), alpha=.7, size=2) + 
  scale_x_continuous(breaks=c(0,1), labels=c('0','1')) + scale_y_continuous(breaks=seq(.25,1.75,by=.25)) +
  guides(color=FALSE) 
tinypop.plot
```

  1.  The **conditional expectation function** $\mu(x)=E[Y \mid X=x]$.
      - $\mu$ is a function; evaluated at $x$, it's a number. It's not random.
      - It's the mean of the subpopulation of people with $X=x$.
   
  2.  The conditional expectation $\mu(X)=E[Y \mid X]$.
      - $\mu(X)$ is the mean of a **random subpopulation** of people.
      - It's the conditional expectation function **evaluated at** the random variable $X$.
      - This is the sort of thing that shows up when we use the law of iterated expectations.

## Check Your Understanding {.smaller}

```{r}
#| out-height: 250px
#| fig-asp: .25
tinypop.plot
```

- Suppose we sample a point $(X, Y)$  uniformly at random from the population of 6 points above.
  - What is the conditional expectation function $\mu(x)$ at [$x=0$]{.twocolor-red} and [$x=1$]{.twocolor-green}?
  - What is the conditional expectation $\mu(X)$?
  
::: fragment
  - [$\mu(0)=`r mu(0)`$]{.twocolor-red} and [$\mu(1)=`r mu(1)`$]{.twocolor-green}
:::
::: fragment
  - $\mu(X)$ is a random variable taking on these two values.
  $$ \mu(X) = \begin{cases}
      `r mu(0)` & \text{ when } X=0 \\
      `r mu(1)` & \text{ when } X=1
      \end{cases}
  $$
:::

## The Indicator Trick
```{r}
#| out-height: 250px
#| fig-asp: .25
tinypop.plot
```

- Suppose we sample a point $(X, Y)$  uniformly at random from the population of 6 points above.
  - What is $X\mu(X)$?

::: fragment

- $X\mu(X)$ is a random variable taking on these two values.
  $$ X\mu(X) = \begin{cases}
      0 \times \mu(0) = 0 \times `r mu(0)` & \text{ when } X=0 \\
      1 \times \mu(1) = 1 \times `r mu(1)` & \text{ when } X=1
      \end{cases}
  $$

- We can write it equivalently as $X\mu(1)$ because either ...
  - $X=0$, so $X\mu(X) = X\mu(1)=0$ 
  - $X=1$, so $X\mu(X)=X\mu(1)=1\times \mu(1)$
- This comes up a lot working with subsample means.
- We'll swap $X\mu(X)$ for $X\mu(1)$ often, referring to the **indicator trick**.

:::


## A More Realistic Example

```{r}
#| out-height: 250px
#| fig-asp: .25
#| 
mean.has.degree = round(summaries$mean[summaries$degree=='4-year degree'] / 10000) * 10
mean.no.degree  = round(summaries$mean[summaries$degree=='no 4-year degree'] / 10000) * 10
degree.proportion = round(mean(earnings$degree=='4-year degree'),1)
earnings.jitter + geom_pointrange(aes(x=degree, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries) + scale_y()
```

 - Let's think about a random person drawn from this population
    - $Y$ is their income 
    - $X$ is an indicator for having a four-year degree. 
  - Suppose the subpopulation means are `r approx(mean.has.degree, 10)` for people with degrees and `r approx(mean.no.degree, 10)` for people without.
  - And that `r approx(degree.proportion*10,1)`/10 of our population has degrees.
  
- Questions 
  - What is $E(Y \mid X)$? 
  - What is $E(Y)$?

:::: r-stack
::: {.fragment .fade-out data-fragment-index="1"}

- $E(Y \mid X)$ is a random variable that is **either** `r mean.has.degree`K or `r mean.no.degree`K
  - It's `r mean.has.degree`K with probability `r degree.proportion*10`/10,       when $X=1$. 
  - It's `r mean.no.degree`K  with probability `r (1-degree.proportion)*10`/10,   when $X=0$. 

::: 

::: {.fragment .current-visible data-fragment-index="1"}

- $\E(Y)$ is the expectation of this random variable---a number.
- We can calculate it using the law of iterated expectations.

$$
\begin{aligned} 
E\{ E( Y \mid X ) \} 
&= E(Y|X=1)P(X=1) + E(Y|X=0)P(X=0) \\
&= `r mean.has.degree`K \cdot `r degree.proportion*10`/10 +
   `r mean.no.degree`K  \cdot `r (1-degree.proportion)*10`/10
=  `r mean.has.degree*degree.proportion`K +
   `r mean.no.degree*(1-degree.proportion)`K 
=  `r mean.has.degree*degree.proportion + mean.no.degree*(1-degree.proportion)`K
\end{aligned}
$$

:::
::::

## Review Exercise 1
```{r}
#| out-height: 250px
#| fig-asp: .25
tinypop.plot
```

- Suppose we sample a point $(X, Y)$ in the plot above uniformly at random.
  - Ignore the jitter; just think of the points as being at $X=0$ and $X=1$.
  - What is $\E\{ \mu(X) \}$?

::: fragment
  - There are two ways of thinking about calculating $\E\{\mu(X)\}$.
      1. We just calculate the expectation of it, thinking of it as an arbitrary random variable.
      2. We use the law of iterated expectations to show it's the unconditional mean of $Y$.

$$
\small{
\begin{aligned}
\E\qty{ \mu(X) } &= \frac{1}{2}\mu(0) + \frac{1}{\mu(1)}= \frac{1}{2} \cdot `r mu(0)` + \frac{1}{2} \cdot `r mu(1)` = 1.05 && \text{ the first way } \\
\E\qty{ \mu(X) } &= \E\qty{\E\qty(Y \mid X)} = \E\qty{Y} = \frac{1}{6} \cdot `r Y[1]` + \frac{1}{6} \cdot `r Y[2]` + \ldots && \text{ the second way }
\end{aligned}
}
$$

:::

## Review Exercise 2
```{r}
#| out-height: 250px
#| fig-asp: .25
tinypop.plot
```

- Suppose we sample a point $(X, Y)$ in the plot above uniformly at random.
  - Ignore the jitter; just think of the points as being at $X=0$ and $X=1$.
  - What is $\E\{ X\mu(X) \}/\E(X)$?

::: fragment

- It's $\mu(1)$. $X\mu(X)=X\mu(1)$, so ...
$$
\begin{aligned}
\E\{ X\mu(X) \} 
&= \E\{ X\mu(1) \} \\
&= \mu(1) \ \E\{ X \} \\
\end{aligned}
$$

:::

## Review Exercise 3

```{r}
#| out-height: 250px
#| fig-asp: .25
tinypop.plot
```

- Suppose we sample a point $(X, Y)$ in the plot above uniformly at random.
  - Ignore the jitter; just think of the points as being at $X=0$ and $X=1$.
  - What is $\E\{ (1-X)\mu(X) \}/\E(1-X)$?

::: fragment

- It's $\mu(0)$. It's analogous to the last one. $(1-X)\mu(X)=(1-X)\mu(0)$, so ...
$$
\begin{aligned}
\E\{ (1-X)\mu(X) \} 
&= \E\{ (1-X)\mu(0) \} \\
&= \mu(0) \ \E\{ 1-X \} \\
\end{aligned}
$$

:::


# Unbiasedness of Sample and Subsample Means

## Bias

```{r}
#| layout-nrow: 2
#| 
xx = seq(-4,4,by=.01)
heights = seq(0,.4,length.out=100)
normals = rnorm(10000)

ggplot() + geom_histogram(aes(x=normals, y=after_stat(density)), alpha=.3, bins=100) +
           geom_vline(xintercept=0, color='green', linewidth=3, alpha=.4) + 
           geom_vline(xintercept=0, color='blue') + 
           geom_vline(xintercept=c(-1,1), color='blue', linetype='dashed') +
           geom_vline(xintercept=c(-2,2), color='blue', linetype='dotted') +
           geom_point(aes(y=heights, x=normals[1:length(heights)]), color='purple', alpha=.3)

ggplot() + geom_histogram(aes(x=normals+2, y=after_stat(density)), alpha=.3, bins=100) +
           geom_vline(xintercept=0, color='green', linewidth=3, alpha=.4) + 
           geom_vline(xintercept=2+0, color='blue') + 
           geom_vline(xintercept=2+c(-1,1), color='blue', linetype='dashed') +
           geom_vline(xintercept=2+c(-2,2), color='blue', linetype='dotted') +
           geom_point(aes(y=heights, x=normals[1:length(heights)]+2), color='purple', alpha=.3) +
           xlim(-4,4)
```

- The **bias** of an estimator is the difference between its expected value and the value of the thing it's estimating.
- We haven't talked about this yet because there hasn't been any. We've been using **unbiased estimators**.
- You can see, looking at the sampling distribution plot on the right, that bias might cause some problems with coverage.
- If we calibrate interval estimates to cover **the estimator's mean** 95% of the time---and that's what we've really been doing---how often will they cover the [thing we're estimating]{.fg style="color:green"}?
- This means biased estimators can cause big problems. We'll talk about this more later in the semester.
- But for now, let's just show the estimators that we've been using are, in fact, unbiased.

## The Sample Mean

This one's easy. What is the expected value of the sample mean?

$$
\E\qty[\frac1n\sum_{i=1}^n Y_i] \ = \  ?
$$

## The Mean: Answered

The expected value of the sample mean is the population mean.

$$
\begin{aligned}
\E\qty[\frac1n\sum_{i=1}^n Y_i] 
&= \frac1n\sum_{i=1}^n \E[Y_i] && \text{ via linearity (Rule 1)} \\
&= \frac1n\sum_{i=1}^n \mu     && \text{ via uniform sampling } \\
&= \frac1n \times n \times \mu = \mu.
\end{aligned}
$$

## A Subsample Mean

**Claim.** The subsample mean is unbiased for the subpopulation mean.
$$ \E[\hat\mu(1)] = \mu(1) $$

**Sketch.**

  - Use the Law of Iterated Expectations, conditioning on $X_1 \ldots X_n$.
  - Then the linearity of conditional expectations to push the conditional expectation into the sum.
  - Then irrelevance of independent conditioning variables to write things in terms of the *random variable* $\mu(X_i)$  
  - Then the **indicator trick**. What's $X_i \mu(X_i)$? How is it related to $\mu(1)$?

::: fragment

$$
\small{
\begin{aligned}
\E[\hat\mu(1)] 
&=\E\qty[ E\qty{\frac{\sum_{i=1}^{n}X_i Y_{i}}{\sum_{i=1}^{n}X_{i}} \mid X_1 \ldots X_n}] \\
&=E\qty[\frac{\sum_{i=1}^{n} X_i \E\qty{ X_{i} \mid X_1 \ldots X_n}}{\sum_{i=1}^{n}X_{i}}] \\
&=E\qty[\frac{\sum_{i=1}^{n}X_i\E\qty{ Y_{i} \mid X_i}}{\sum_{i=1}^{n}X_{i}}] \\
&=E\qty[\frac{\sum_{i=1}^{n}X_i \mu(X_i)}{\sum_{i=1}^{n}X_{i}}] \\
&=E\qty[\frac{\sum_{i=1}^{n}X_i\mu(1)}{\sum_{i=1}^{n}X_{i}}] \\
&=E\qty[\mu(1) \times \frac{\sum_{i=1}^{n}X_i}{\sum_{i=1}^{n}X_{i}}] = \E[\mu(1)]=\mu(1) 
\end{aligned}
}
$$

:::

## Looking Back

$$
\small{
\begin{aligned}
\E[\hat\mu(1)] 
&=\E\qty[ E\qty{\frac{\sum_{i=1}^{n}X_i Y_{i}}{\sum_{i=1}^{n}X_{i}} \mid X_1 \ldots X_n}] \\
&=E\qty[\frac{\sum_{i=1}^{n} X_i \E\qty{ X_{i} \mid X_1 \ldots X_n}}{\sum_{i=1}^{n}X_{i}}] \\
&=E\qty[\frac{\sum_{i=1}^{n}X_i\E\qty{ Y_{i} \mid X_i}}{\sum_{i=1}^{n}X_{i}}] \\
&=E\qty[\frac{\sum_{i=1}^{n}X_i \mu(X_i)}{\sum_{i=1}^{n}X_{i}}] \\
&=E\qty[\frac{\sum_{i=1}^{n}X_i\mu(1)}{\sum_{i=1}^{n}X_{i}}] \\
&=E\qty[\mu(1) \times \frac{\sum_{i=1}^{n}X_i}{\sum_{i=1}^{n}X_{i}}] = \E[\mu(1)]=\mu(1) 
\end{aligned}
}
$$

- Now let's look back at our proof. 
- What's the *conditional expectation* of our estimator, $\E[\hat\mu(1) \mid X_1 \ldots X_n]$?
- This'll be useful next time.

::: fragment

 - It's $\mu(1)$, the subpopulation mean, just like the unconditional expecation.
 - In our proof, we used the law of iterated expectations to write $\E[\hat\mu(1)]$ as the expectation of $\E[\hat\mu(1) \mid X_1 \ldots X_n]$.
  - And we showed that the latter is *the constant* $\mu(1)$.
  - So the former is the expectation of that constant---the same constant.

:::


# Review: Variance and Conditional Variance

# The Variance of Sample and Subsample Means

## Variance as Excess {.smaller}

$$
\small{
\begin{aligned}
\Var(Y) &= \E \qty[ \qty{ Y - \E(Y) }^2 ] && \text{ the \emph{spread} idea: the mean square of a \emph{centered version} of $Y$ } \\
&= \E\qty(Y^2) - \qty{\E\qty(Y)}^2         && \text{ the \emph{excess} idea: the average amount $Y^2$ \emph{exceeds} the square of its mean}.
\end{aligned}
}
$$

```{r}
#| out-height: 250px
#| fig-asp: .25
tinypop.square.plot = ggplot() + geom_point(aes(x=Xdodge, y=Y, color=factor(X)), alpha=.5, size=3*dot.size) +
                                 geom_point(aes(x=Xdodge, y=Y^2, color=factor(X)), shape=0, size=3*dot.size) + 
                                 geom_segment(aes(x=Xdodge, xend=Xdodge, y=Y, yend=Y^2), alpha=.3) + guides(color=FALSE)
tinypop.square.plot + xlim(min(Xdodge[X==0])-sd(Xdodge[X==0]), max(Xdodge[X==0] + sd(Xdodge[X==0])))
```

Let's think, in terms of the plot, about why there is excess in the square.

- Suppose $Y$ takes on three values --- the dots ---with equal probability.
    - Its expectation is the average height of the black dots.
    - And its expectation, squared, is the square of this.
- Its square $Y^2$ takes on the the square of those values---the squares --- with equal probability.
    - Its expectation is the average height of the squares.
- Why is the second one bigger?

::: fragment

- Because squaring *increases big draws of Y* more than it *decreases small ones*.
- Visually, the line from $Y$ to $Y^2$ is longer when $Y$ is bigger than when it's smaller.

:::


## Equivalence Proof

::: fragment

$$
\small{
\begin{aligned}
\E \qty[ \qty{ Y - \E(Y) }^2 ] 
&= \E \qty[ Y^2 - 2 Y \E(Y) + \qty{\E(Y)}^2 ] && \text{ FOIL } \\
&= \E[ Y^2] - \E\qty[2 Y \E(Y)] + \E\qty[\qty{\E(Y)}^2] && \text{Distributing expectations (linearity)} \\
&= \E[ Y^2] - 2\E(Y)\E\qty[Y] + \qty{\E(Y)}^2\E[1] && \text{Pulling constants out of expectations (linearity)} \\
&= \E[ Y^2] - 2\qty{\E(Y)}^2 + \qty{\E(Y)}^2] && \text{Recognizing a product of something with itself as a square} \\
&= \E[ Y^2] - \cancel{2}\qty{\E(Y)}^2 + \cancel{\qty{\E(Y)}^2}] && \text{Subtracting} 
\end{aligned}
}
$$

:::

## Conditional Variance {.smaller}

- The **conditional variance** describes the spread of one random variable *within groups*. 
- Like the conditional expectation of the random variable, there are two related concepts.

1. The **conditional variance function**.
$$
\sigma^2(x) := \Var(Y \mid X=x) = \E\qty[ \textcolor{blue}{\qty{ Y-\E(Y \mid X) }^2} \mid X=x]
$$
    - This is the variance of $Y$, **within the subpopulation** of people with $X=x$.
    - It is the *conditional expectation function* of the random variable $\textcolor{blue}{\{Y-\E(Y \mid X)\}^2}$.
    - We call it 'the conditional variance of $Y$ given $X=x$'

2. The conditional variance $\sigma^2(X) = \Var(Y \mid X)$
$$
\sigma^2(X) := \Var(Y \mid X)   = \E\qty[ \textcolor{blue}{\qty{ Y-\E(Y \mid X) }^2} \mid X]
$$

    - This is the variance of $Y$ within **a random subpopulation** of people.
        - It's the conditional variance function **evaluated at** the random variable $X$.
        - And  the conditional expectation of the random variable $\textcolor{blue}{\{Y-\E(Y \mid X)\}^2}$.
    - We call it 'the conditional variance of $Y$ given $X$.
    - This is the sort of thing that shows up when you use the law of iterated expectations.



## Conditional Variance as Excess {.smaller}

$$
\small{
\begin{aligned}
\Var(Y \mid X) &= \E \qty[ \qty{ Y - \E(Y\mid X)}^2 \mid X ] && \text{ the \emph{spread} idea: the mean square of a conditionally centered version of $Y$ } \\
&= \E\qty(Y^2 \mid X) - \qty{\E\qty(Y \mid X)}^2              && \text{ the \emph{excess} idea: the average amount $Y^2$ exceeds the square of its cond. mean}
\end{aligned}
}
$$


```{r}
#| out-height: 500px
#| fig-asp: .5
tinypop.square.plot 
```

- We can think of the Conditional Variance as excess, too.
- It's the same as the unconditional case, but *within subpopulations*.


## The Law of Total Variance

$$
\Var[Y] = \E\qty{\Var( Y \mid X ) } + \Var\qty{\E( Y \mid X ) }  
$$

- The Law of Total Variance relates Conditional and Unconditional Variance
- It's like the Law of Total Expectation, but for Variance
- It's a useful way to decompose the variance a random variable.
- Think about where most of the variance in $Y$ is coming from in the two populations shown below.

```{r}
#| out-height: 500px
#| fig-ncol: 2

ggplot() + geom_point(aes(x=Xdodge, y=Y, color=factor(X)), alpha=.7, size=4) + 
  scale_y_continuous(breaks=seq(0,3,by=.5), limits=c(0,3)) + guides(color=FALSE) 
ggplot() + geom_point(aes(x=Xdodge, y=Y+X, color=factor(X)), alpha=.7, size=4) + 
  scale_y_continuous(breaks=seq(0,3,by=.5), limits=c(0,3)) + guides(color=FALSE) 

```


## An Activity 

```{r}
#| out-height: 250px
#| fig-asp: .25

mean.has.degree = round(summaries$mean[summaries$degree=='4-year degree'] / 10000) * 10
mean.no.degree  = round(summaries$mean[summaries$degree=='no 4-year degree'] / 10000) * 10
degree.proportion = round(mean(earnings$degree=='4-year degree'),1)
earnings.jitter + geom_pointrange(aes(x=degree, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries) + scale_y()
```

- Working with your neighbor, think about why we call it this.
- Try to put the three terms into words.
    - $\Var[Y]$ is the ...
    - $\E\qty{ \Var(Y \mid X) }$ is the ...
    - $\Var\qty{\E( Y \mid X)}$ is the ...
- Do this in two ways.
    1. Using 'Random Variables Language', 
        - e.g. 'the expected value of the conditional variance of Y given X'
    2. Using 'Population Language'
        - e.g. 'the variance of income (Y) within the random subpopulation with degree status X'


## Proof

**Claim.**
$$
\Var[Y] = \textcolor{blue}{\E\qty[\Var(Y \mid X)]} + \textcolor{green}{\Var\qty[\E(Y \mid X)]} 
$$

1. Rewrite the variances, conditional and unconditional, in *excess* form.

::: fragment
$$
\Var[Y]  = \textcolor{blue}{\E\qty[ \E(Y^2 \mid X) - \qty{\E\qty(Y \mid X)}^2]} 
                       + \textcolor{green}{\E\qty{ \E(Y \mid X)^2}  - \qty(\E\qty{\E\qty(Y \mid X)})^2} 
$$

:::

2. Distribute the outer expectations in the first term.

::: fragment
$$
\Var[Y] = \textcolor{blue}{\E\qty[ \E(Y^2 \mid X)] - \E\qty[\qty{\E\qty(Y \mid X)}^2]} 
                       + \textcolor{green}{\E \qty[ \qty{\E(Y \mid X)^2} ] - \qty(\E\qty{\E\qty(Y \mid X)})^2}
$$

:::

3. Cancel matching terms

::: fragment
$$
\Var[Y]  = \textcolor{blue}{\E\qty[ \E(Y^2 \mid X)] \cancel{- \E\qty[\qty{\E\qty(Y \mid X)}^2]}}  
                       + \textcolor{green}{\cancel{\E \qty[ \qty{\E(Y \mid X)^2} ]} -  \qty(\E\qty{\E\qty(Y \mid X)})^2}
$$

:::

4. Use the law of iterated expectations to simplify the remaining terms.

::: fragment
$$
\Var[Y]  = \textcolor{blue}{E[Y^2]  \cancel{- \E\qty[\qty{\E\qty(Y \mid X)}^2]}}
                       + \textcolor{green}{\cancel{\E \qty[ \qty{\E(Y \mid X)^2} ]} - (\E\qty{Y})^2}
$$

:::

## The Variance of the Subsample Mean

**Claim.** The variance of our point estimate is the **expected value** of the subpopulation variance divided by the number of people in the subsample.
$$ \Var[\hat\mu(1)] = \E\qty[ \frac{\sigma^2(1)}{N_1} ] \quad \text{ for } \quad N_1 = \sum_{i=1}^n X_i $$

- This is like what we saw in the unconditional case in Lecture 3, with a twist.
- The number of people in the subsample is random, so instead of $1/n$, we have $\E[ 1/N_1 ]$.
- Let's prove it.
  - It's a bit long. 
  - To keep organized, I'll remind you where we left off at the top of each slide.
  - Except the first one, where we're starting from the beginning.

## Proof, Part 1

- We'll start by using the Law of Total Variance, conditioning on $X_1 \ldots X_n$.
- We 'looked back' just now for a reason---we saw that one of these terms is zero.

$$
\small{
\begin{aligned}
\Var\qty[ \hat \mu(1) ] &= \E \qty[ \Var\qty{ \hat \mu(1) \mid X_1 \ldots X_n }] + \Var\qty[ \E{\hat \mu(1) \mid X_1 \ldots X_n} ] \\
&= \E \qty[ \Var\qty{ \hat \mu(1) \mid X_1 \ldots X_n }] + \Var[\mu(1)] \\
&= \E \qty[ \Var\qty{ \hat \mu(1) \mid X_1 \ldots X_n }] + 0
\end{aligned}
}
$$



## Proof, Part 2 {.smaller}

$$
\Var\qty[ \hat \mu(1) ] 
= \E\qty[ \Var\qty{ \hat \mu(1) \mid X_1 \ldots X_n }] 
$$

- Now we'll substitute in definitions and simplify.
- We'll end having to work with the conditional mean of a *squared sum*.

$$
\small{
\begin{aligned}
\Var\qty[ \hat \mu(1) ] 
&= \E\qty[ \Var\qty{ \hat \mu(1) \mid X_1 \ldots X_n }] \\
&= \E\qty[ \E\qty{  \qty(\hat \mu(1)-\mu(1))^2 \mid X_1 \ldots X_n }] \\
&= \E\qty[ \E\qty{  \qty(\frac{\sum_{i=1}^{n}X_i Y_{i}}{\sum_{i=1}^{n}X_{i}} - \mu(1))^2 \mid X_1 \ldots X_n }] \\
&= \E\qty[ \E\qty{  \qty(\frac{\sum_{i=1}^{n}X_i Y_{i}}{\sum_{i=1}^{n}X_{i}} - \frac{\mu(1)\sum_{i=1}^n X_i}{\sum_{i=1}^n X_i })^2 \mid X_1 \ldots X_n }] \\
&= \E\qty[ \E\qty{  \qty(\frac{\sum_{i=1}^{n}X_i \qty{Y_{i}-\mu(1)}}{\sum_{i=1}^{n}X_{i}})^2 \mid X_1 \ldots X_n }] \\
&= \E\qty[ \frac{\E\qty{ \qty(\sum_{i=1}^{n}X_i \qty{Y_{i}-\mu(1)})^2 \mid X_1 \ldots X_n}}{\qty(\sum_{i=1}^{n}X_{i})^2}]
\end{aligned}
}
$$

## Proof, Part 3 {.smaller}

$$
\Var\qty[ \hat \mu(1) ]  = \E\qty[ \frac{\E\qty{ \qty(\sum_{i=1}^{n}X_i \qty{Y_{i}-\mu(1)})^2 \mid X_1 \ldots X_n}}{\qty(\sum_{i=1}^{n}X_{i})^2}]
$$

- To work with this squared sum, we do our FOIL-like expansion. This is ...

$$ \small{ 
\begin{aligned}
\qty(\sum_{i}Z_i)^2 = \qty(\sum_{i}Z_i) \qty(\sum_{j}Z_i) = \sum_i\sum_j Z_i Z_j \quad && \text{ for } \quad Z_i = X_i \qty{Y_{i}-\mu(1)} \\
\text{or equivalently, using the \textbf{indicator trick}} && \text{ for } \quad Z_i = X_i\qty{Y_i - \mu(X_i)}
\end{aligned}
}
$$

- This lets us push the conditional mean *all the way into the sum*, so it meets the individual terms.
- And when we've done that, we can use the irrelevance of independent conditioning variables to simplify.

$$
\small{
\begin{aligned}
\Var\qty[ \hat \mu(1) ] 
&= \E\qty[ \frac{\E\qty{ \qty(\sum_{i=1}^{n}Z_i )^2 \mid X_1 \ldots X_n}}{\qty(\sum_{i=1}^{n}X_{i})^2}] \\
&= \E\qty[ \frac{\E\qty{ \sum_{i=1}^{n}\sum_{j=1}^n Z_i Z_j \mid X_1 \ldots X_n}}{\qty(\sum_{i=1}^{n}X_{i})^2}] \\
&= \E\qty[ \frac{\sum_{i=1}^{n}\sum_{j=1}^n \E\qty{Z_i Z_j \mid X_1 \ldots X_n}}{\qty(\sum_{i=1}^{n}X_{i})^2}] \\
&= \E\qty[ \frac{\sum_{i=1}^{n}\sum_{j=1}^n \E\qty{Z_i Z_j \mid X_i, X_j}}{\qty(\sum_{i=1}^{n}X_{i})^2}] \\
\end{aligned}
}
$$

## Proof, Part 4 {.smaller}

$$
\begin{aligned} 
\Var\qty[ \hat \mu(1) ] 
&= \E\qty[ \frac{\sum_{i=1}^{n}\sum_{j=1}^n \E\qty{Z_i Z_j \mid X_i, X_j}}{\qty(\sum_{i=1}^{n}X_{i})^2}] \quad \text{ for } \quad Z_i=Y_i-\mu(X_i)
\end{aligned}
$$


- Let's take a look at one of the terms in the numerator, $\E\qty{Z_i Z_j \mid X_i, X_j}$.
  - We're going to show that the cross-terms --- the ones for $i \neq j$ --- are zero.
  - This will let us drop them from the sum.
  - We'll use the fact that $Z_i$ and $Z_j$ are independent for $i \neq j$.
- Let's rewrite a cross-term using our usual stuff
  - the law of iterated expectations, conditioning on $Z_i$
  - linearity of expectations;
  - irrelevance of independent conditioning variables.

$$\small{
\begin{aligned}
\E\qty{Z_i Z_j \mid X_i, X_j} 
&= \E\qty{ \E(Z_i Z_j \mid Z_i,X_i,X_j) } \\
&= \E\qty{ Z_i \E(Z_j \mid X_i,X_j) } \\
&= \E\qty{ Z_i \textcolor{blue}{\E(Z_j \mid X_j)} } \quad \text{ where } \\
\textcolor{blue}{\E(Z_j \mid X_j)} 
&= \E\qty[X_j \qty{Y_j - \mu(X_j)}] \\ 
&= \E\qty{ \E\qty[X_j \qty{Y_j - \mu(X_j)} \mid X_j] } \\
&= \E\qty[X_j \qty{ \E\qty[Y_j \mid X_j] - \mu(X_j)}] \\
&= \E\qty[X_j \cross{\qty{ \mu(X_j) - \mu(X_j)}}] = 0. \\
\end{aligned}
}$$


## Proof, Part 5 {.smaller}

$$
\begin{aligned} 
\Var\qty[ \hat \mu(1) ] 
&= \E\qty[ \frac{\sum_{i=1}^{n}\sum_{j=1}^n \E\qty{Z_i Z_j \mid X_i, X_j}}{\qty(\sum_{i=1}^{n}X_{i})^2}] \quad \text{ for } \quad Z_i=Y_i-\mu(X_i)  \\
&= \E\qty[ \frac{\sum_{i=1}^{n} \E\qty{Z_i^2 \mid X_i}}{\qty(\sum_{i=1}^{n}X_{i})^2}] \quad \text{because cross-terms are zero}.
\end{aligned}
$$

- All we've got to do now is recognize what we have. 
- Let's look at a term in the numerator.
  - We'll use iterated expecations and linearity.
  - Then the indicator trick.

$$
\small{
\begin{aligned}
\E\qty[Z_i^2 \mid X_i] 
&= \E\qty[ X_i\qty{Y_i - \mu(X_i)}^2 \mid X_i] 
= X_i\E\qty[\qty{Y_i - \mu(X_i)}^2 \mid X_i] 
= X_i\Var\qty[Y_i \mid X_i] = X_i \sigma^2(X_i) 
= X_i\sigma^2(1) 
\end{aligned}
}
$$

- We finish up by substituting this back into what we started with.

$$
\small{ 
\begin{aligned}
\Var\qty[ \hat \mu(1) ] 
&= \E\qty[ \frac{\sum_{i=1}^{n} \E\qty{Z_i^2 \mid X_i}}{\qty(\sum_{i=1}^{n}X_{i})^2}]  
= \E\qty[ \frac{\sum_{i=1}^{n} X_i \sigma^2(1)}{\qty(\sum_{i=1}^{n}X_{i})^2}] 
= \E\qty[ \frac{\cancel{\sum_{i=1}^{n} X_i }\sigma^2(1)}{\qty(\sum_{i=1}^{n}X_{i})^{\cancel{2}}}] 
= \E\qty[ \frac{\sigma^2(1)}{\sum_{i=1}^{n}X_{i}}] = \sigma^2(1) \times \E\qty[ \frac{1}{N_1}]. 
\end{aligned}
}
$$


## Our Estimator's Sampling Distribution 
```{r}
#| out-height: 300px
#| fig-asp: .3 


muhat.1 = mean(Y[X==1])
se = sqrt(var(Y[X==1])/sum(X))
breaks = seq(-4*se, 4*se, 2*se) 
signs  = case_when(
  breaks < 0 ~ '-', 
  breaks == 0 ~ '',
  breaks > 0 ~ '+')
offsets = case_when(
  breaks != 0 ~ approx(breaks),
  breaks == 0 ~ '')
labels.conrete = sprintf('%1.2f', breaks+muhat.1)
labels.abstract = sprintf('μ %s %s',     signs, offsets)

x=seq(-4,4,by=.01)
sampling.dist.plot = ggplot() + geom_area(aes(x=x,y=dnorm(x)), alpha=.1, color='black', fill='green') + 
       geom_vline(aes(xintercept=qnorm(c(.16,.84))), linetype='dashed', color='blue') +
       geom_vline(aes(xintercept=qnorm(c(.025,.975))), linetype='dotted', color='blue') +
       scale_x_continuous(breaks=seq(-4,4,by=2), labels=labels.abstract) 

sampling.dist.plot
```

- It's approximately this normal, with
  - mean $\mu(1)$, the subpopulation mean for people with 4-year degrees.
  - standard deviation $\sqrt{\E \frac{\sigma^2(1)}{N_1}}$
- 95% of draws from this distribution lie between the dotted blue lines.
  - Which are $\mu(1) \pm 1.96 \sqrt{\E \frac{\sigma^2(1)}{N_1}}$.
- With a little arithmetic, this tells us we can get a calibrated interval by ...
  - Starting at our point estimate
  - Drawing *arms* on either side of length $1.96 \sqrt{\E \frac{\sigma^2(1)}{N_1}}$

## Forming the CI {.smaller}

```{r}
#| out-height: 250px
#| fig-asp: .25

sampling.dist.plot
```

$$ \tiny{ 
  \text{an approximately calibrated 95\% interval:} \quad  \qty[ \hat\mu(1) - 1.96 \sqrt{\E \frac{\sigma^2(1)}{N_1}} \quad  \ldots \quad \hat\mu(1) + 1.96 \sqrt{\E \frac{\sigma^2(1)}{N_1}} ] 
} $$

- We can *almost* do that.
  - We know our point estimate.
  - And although we *don't know* the standard error we need to draw the arms, it's easy enough to estimate.
- We'll substiute the *subsample variance* $\hat\sigma(1)$ for the *subpopulation variance* $\sigma(1)$.
$$ \tiny{
\hat\sigma^2(1) = \frac{1}{N_1}\sum_{i:X_i=1} \qty{ Y_i - \hat\mu(1) }^2
} $$

- And the inverse of the *subsample size*, $1 / N_1$, for its *expected value*.

$$ \tiny{
\text{another approximately calibrated 95\% interval:} \quad  \qty[ \hat\mu(1) - 1.96 \sqrt{\frac{\hat\sigma^2(1)}{N_1}} \quad  \ldots \quad \hat\mu(1) + 1.96 \sqrt{\frac{\hat\sigma^2(1)}{N_1}} ] = \qty[`r approx(muhat.1 - 1.96*se)` \quad \ldots \quad `r approx(muhat.1 + 1.96*se)`] 
} $$

- This is the interval we'll use.  Why am I not drawing it in?

## I can't

```{r}
#| out-height: 250px 
#| fig-ncol: 2
sampling.dist.plot
ggplot() + geom_pointrange(aes(x=muhat.1, y=.25, xmin=muhat.1-1.96*se, xmax=muhat.1+1.96*se)) 
```

- I can draw the sampling distribution---in abstract terms, centered on the subpopulation man $\mu(1)$.
- I can draw the interval---in concrete terms, centered on our point estimate $\hat\mu(1)$.
- But I don't know the precise relationship between these two points.
- That means I can't 'align the axes' and overlay one on the other.

\ \ 

- When I did this for turnout, I was using *post-election knowledge* of the population.
- I'm never going to have that knowledge about income in California.
- So we'll never know if our interval covers the subpopulation mean.

## Interpretation
```{r}
#| out-height: 250px
#| fig-asp: .25 

set.seed(0)
x=seq(-4,4,by=.01)
muhat = rnorm(20)

sampling.dist.plot + 
       geom_pointrange(aes(x=muhat, y=(1:20)/40, xmin=muhat-1.96, xmax=muhat+1.96, color=abs(muhat) <= 1.96), alpha=.4) +
       scale_color_discrete(type=c('#F50041', '#6615A3')) + guides(color=FALSE)
```

- We do think---based on our mathematical reasoning---that our interval is calibrated.
- Recall what calibration means.
  - A calibrated interval contains what it's supposed to
  - in the case, the subpopulation mean $\mu(1)$
  - roughly 95% of the time --- 19/20 times 
- So if we want to picture our interval in reference to $\mu$, we can in a sense.
  - We can think, in rough terms, of it being *one of these*.
  - We don't know which and they're all equally likely.
  - Most of the time, it'll be one of the [good ones]{.fg style="color:#6615A3"}
  - But sometimes it'll be the [bad one]{.fg style="color:#F50041"}.
