---
title: "QTM 220 Lecture 15"
subtitle: "Bias and Least Squares Regression"
---

# Review

## Least Squares Regression

{{< include shared-code.qmd >}}
{{< include ca-income.qmd >}}

```{r}
#| label: precision
options(scipen = 1, digits = 3)
```

```{r}
#| out-height: 400px
#| fig-asp: .4

race.categories = c('White only', 'Black only', 'Asian only', 'Hawaiian/Pacific Islander only (HP)')
X = earnings$education
Y = earnings$income
muhat.monotone = as.stepfun(isoreg(X,Y))
muhat.lin = lm(Y ~ X)
lsq.plot = ggplot(data.frame(x=X, y=Y)) + geom_point(aes(x=X,y=Y), alpha=.1, size=1, position=position_jitter(w=.3)) + 
  stat_summary(aes(x=X+.1, y=Y), fun=mean, geom='point', color='red', alpha=.5) + 
  stat_summary(aes(x=X+.1, y=Y), fun=mean, geom='line', color='red', alpha=.3) + 
  geom_point(aes(X-.1, predict(muhat.lin)),  color='blue', alpha=.3) +
  geom_line(aes(X-.1, predict(muhat.lin)),  color='blue', alpha=.3) +
  geom_point(aes(x=X, y=muhat.monotone(X)), color='magenta', alpha=.3) + 
  geom_line(aes(x=X, y=muhat.monotone(X)), color='magenta', alpha=.3) + 
  scale_y(c(0,150000)) + scale_x()
lsq.plot
```

- Regression, generally, is choosing a function of covariates $X_i$ to predict outcomes $Y_i$. 
- We choose from a set of functions $\model$ that we call a **regression model**.

$$
Y_i \approx \hat\mu(X_i) \qfor \hat\mu \in \model 
$$

- Least Squares Regression is a special case in which we choose by minimizing the sum of squared residuals.

$$
\hat \mu = \argmin_{m \in \model} \sum_{i=1}^n \qty{ Y_i - m(X_i) }^2
$$

- Given a model, we can find the least squares estimator by doing this minimization.
- And we can do this for any model we like, so long as we can solve the minimization problem. 
$$
\begin{aligned}
\textcolor{red}{\model} &= \qty{ \text{all functions} \  m(x) }                               \\
\textcolor{blue}{\model} &= \qty{ \text{all lines} \ m(x)=ax + b }  \\
\textcolor{magenta}{\model} &= \qty{ \text{all increasing functions} \  m(x) }   \\
\end{aligned}
$$

## Linear Models

```{r}
#| out-height: 400px
#| fig-asp: .4
lsq.plot
```

- In this class, we're going to focus on **linear models**.
  - These are sets of functions that are 'closed' under addition, subtraction, and scaling by constants.
  - This means when we add, subtract, and scale functions in the model, we get another function in the model.
$$
a(x), b(x) \in \model \implies a(x) + b(x), a(x) - b(x), c a(x) \in \model \quad \text{ for any constant $c$ }
$$

- **Q.** Which of these models are linear? 

$$
\begin{aligned}
\textcolor{red}{\model} &= \qty{ \text{all functions} \  m(x) }                               \\
\textcolor{blue}{\model} &= \qty{ \text{all lines} \ m(x)=ax + b }  \\
\textcolor{magenta}{\model} &= \qty{ \text{all increasing functions} \  m(x) }   \\
\end{aligned}
$$


## The Additive Bivariate Model 

```{r}
#| out-height: 400px
#| fig-asp: .4
 
grid = expand.grid(education=unique(earnings$education), race=race.categories) 
muhat = function(model) { function(ed,race) { predict(model, newdata=data.frame(education=ed, race=race)) } }

additive.fit   = lm(income ~ race+factor(education), data=earnings[earnings$race %in% race.categories,])
ggplot(earnings[earnings$race %in% race.categories, ]) + 
  geom_point(aes(x=education, y=income), size=1, 
             alpha=.1, position=position_jitter(seed=0,w=.3)) +
  geom_point(aes(x=education, y=muhat(additive.fit)(education,race)),  color='blue', data=grid) +
  geom_line(aes(x=education, y=muhat(additive.fit)(education,race)),  color='blue', data=grid, alpha=.5) +
  facet_grid(cols=vars(race)) + scale_y(c(0,150000))
```

- Here we let our predictions be arbitrary functions of education ($x$) as before.
- But we're requiring that race ($w$) only shift that function up and down---it can't change the function's shape.
- The result, in visual terms, are predictions for different values of $w$ are 'parallel curves'.

$$\small{
\begin{aligned}
\textcolor{blue}{\model} &= \qty{ m(w,x) = m_0(w) + m_1(x) \ \ \text{for univariate functions} \ \ m_0, m_1 }  &&\text{additive bivariate model} \\
\end{aligned}
}
$$


## The Parallel Lines Model 

```{r}
#| out-height: 400px
#| fig-asp: .4
 
additive.line.fit   = lm(income ~ race+education, data=earnings[earnings$race %in% race.categories,])
ggplot(earnings[earnings$race %in% race.categories, ]) + 
  geom_point(aes(x=education, y=income), size=1, 
             alpha=.1, position=position_jitter(seed=0,w=.3)) +
  geom_point(aes(x=education, y=muhat(additive.line.fit)(education,race)),  color='red', data=grid) +
  geom_line(aes(x=education, y=muhat(additive.line.fit)(education,race)),  color='red', data=grid, alpha=.5) +
  facet_grid(cols=vars(race)) + scale_y(c(0,150000))
```

- Here we impose the additional restriction that the function of education ($x$) is a line.
- This means that the predictions for different values of $w$ are parallel lines.

$$\small{
\begin{aligned}
\textcolor{red}{\model} &= \qty{ m(w,x) = a(w) + b x   }                 &&\text{parallel lines} \\
\end{aligned}
}
$$

## The Not-Necessarily-Parallel Lines Model 

```{r}
#| out-height: 400px
#| fig-asp: .4
 
additive.line.fit   = lm(income ~ race+education, data=earnings[earnings$race %in% race.categories,])
interacted.line.fit = lm(income ~ race*education, data=earnings[earnings$race %in% race.categories,])
ggplot(earnings[earnings$race %in% race.categories, ]) + 
  geom_point(aes(x=education, y=income), size=1, 
             alpha=.1, position=position_jitter(seed=0,w=.3)) +
  geom_point(aes(x=education, y=muhat(interacted.line.fit)(education,race)),  color='magenta', data=grid) +
  geom_line(aes(x=education, y=muhat(interacted.line.fit)(education,race)),   color='magenta', data=grid, alpha=.5) +
  facet_grid(cols=vars(race)) + scale_y(c(0,150000))
```

- Here we stick with the lines thing, but drop the additivity restriction.
- For each value of $w$, we can get a totally different line.

$$\small{
\begin{aligned}
\textcolor{magenta}{\model} &= \qty{ m(w,x) = a(w) + b(w) x  }                 &&\text{lines} \\
\end{aligned}
}
$$


## Where This Leaves Us

```{r}
#| out-height: 400px
#| fig-asp: .4
 
ggplot(earnings[earnings$race %in% race.categories, ]) + 
  geom_point(aes(x=education, y=income), size=1, 
             alpha=.1, position=position_jitter(seed=0,w=.3)) +
  geom_point(aes(x=education, y=muhat(interacted.line.fit)(education,race)),  color='magenta', data=grid) +
  geom_line(aes(x=education, y=muhat(interacted.line.fit)(education,race)),   color='magenta', data=grid, alpha=.5) +
  geom_point(aes(x=education, y=muhat(additive.line.fit)(education,race)),  color='red', data=grid) +
  geom_line(aes(x=education, y=muhat(additive.line.fit)(education,race)),  color='red', data=grid, alpha=.5) +
  geom_point(aes(x=education, y=muhat(additive.fit)(education,race)),  color='blue', data=grid) +
  geom_line(aes(x=education, y=muhat(additive.fit)(education,race)),  color='blue', data=grid, alpha=.5) +
  facet_grid(cols=vars(race)) + scale_y(c(0,150000))
```

- Each of these approaches allows us to make predictions everywhere---even in columns we haven't sampled.
- Fundamentally, it's because they're using information from one column to make predictions in another.
- This solves the empty-groups problem, but it introduces a new one. What is it?


# Bias

## What It Is 

```{r}
#| out-height: 500px
mu = 1
biasdata = data.frame(x=seq(-4,4,length.out=1000), fx=dnorm(seq(-4,4,length.out=1000), mean=mu))
ggplot() + geom_area(aes(x=x,y=fx), alpha=.2, color='gray', data=biasdata) + 
           geom_vline(aes(xintercept=mu), color='black', linewidth=2, alpha=.5) + 
           geom_vline(aes(xintercept=0),  color='green', linewidth=2, alpha=.5)
```

- Bias is the difference between an estimator's expected value and what we mean to estimate.
- Intuitively, it's the **systematic part** of our estimator's error.
- Visually, it's the difference between ...
  - The location of our estimator's sampling distribution. The black line.
  - The location of what we're trying to estimate. The green line.


**Q.** What is bias here?

::: fragment

It's the difference in the location of the black and green lines: $\text{bias} = 1 - \textcolor{green}{0} = 1.$

:::

## Bias and Consistency 

```{r}
#| out-height: 500px
set.seed(1234)
error = 2*rnorm(100000)
bias = list(a=function(n) 0, b=function(n) sqrt(10)/sqrt(n), c=function(n) 1)

plotdata = expand_grid(estimator=names(bias), n=c(10,40,160)) %>%
  pmap(function(estimator, n) {
    thebias = bias[[estimator]](n)
    data.frame(estimator=estimator, n=n, muhat=thebias + error / sqrt(n))
    }) %>% bind_rows() 
summaries = plotdata %>% group_by(n, estimator) %>% summarize(mean=mean(muhat))

breaks = seq(-4, 4, by=2)/2 
signs  = case_when(
  breaks < 0 ~ '-', 
  breaks == 0 ~ '',
  breaks > 0 ~ '+')
offsets = case_when(
  breaks != 0 ~ sprintf('%d', abs(breaks)),
  breaks == 0 ~ '')
labels.abstract = sprintf('μ %s %s', signs, offsets)

ggplot(plotdata) + geom_histogram(aes(x=muhat, y=after_stat(density)), bins=100, alpha=.3) + 
                   geom_vline(aes(xintercept=mean), data=summaries, linewidth=1, alpha=.6) + 
                   geom_vline(aes(xintercept=0), color='green', linewidth=1, alpha=.6) + 
                   scale_x_continuous(breaks=breaks, labels=labels.abstract, limits=range(breaks)) + 
                   facet_grid(n ~ estimator, labeller=labeller(n=label_both, estimator=label_both)) + 
                   theme(axis.text.x = element_text(angle = -90, vjust = 0.5, hjust=1)) + 
                   xlab('') + ylab('')
```

- In Practice Midterm 1, we looked at the sampling distribution of three estimators at three sample sizes.
  - *Estimator a*  was unbiased.
  - *Estimator b* was biased, but its bias decreased with sample size.
  - *Estimator c* was biased and its bias didn't change with sample size.
- Constant bias means our estimator isn't going to much closer to the truth as we get more data. 
- We call estimators like this---estimators that don't get arbitrarily close to the truth in large samples---**inconsistent**.

## A Biased Least Squares Estimator

```{r}
#| layout-ncol: 2
#| out-height: 400px
#| fig-asp: .8

lm.plot = function(data, formula, dot.size=2) { 
  model = lm(formula, data=data)
  muhat = function(ed) predict(model, newdata=data.frame(education=ed))
  ggplot(data, aes(x=education, y=income)) + 
    geom_point(alpha=.1, position=position_jitter(w=.3), size=dot.size) + 
    stat_summary(fun.data=mean_se, geom='pointrange', color='red', alpha=.5) +
    geom_point(aes(x=education, y=muhat(education)),  color='blue', alpha=.3) +
    geom_line(aes(x=education, y=muhat(education)),  color='blue', alpha=.3) 
}
lm.plot(earnings, 'income~education')
lm.plot(earnings, 'income~education') + scale_x(c(7.5,10.5)) + scale_y(c(-10000,40000))
```

$$
\begin{aligned}
\textcolor{red}{\model} &= \qty{ \text{all functions} \  m(x) }                               \\
\textcolor{blue}{\model} &= \qty{ \text{all lines} \ m(x)=ax + b }  \\
\end{aligned}
$$

- We can see that, when we use the [all lines]{.fg style="color:blue"} model, our least squares estimate $\hat\mu(8)$ is biased.
- Why? Because the best line, trying to get the slope right in the middle where most of the data is, gets it wrong at the edges.
- Typically, we can't **know** an estimator is biased, but we can **suspect** it is.
  - e.g., because it's nowhere near the [subsample means]{.fg style="color:red"}.
- That's because bias is a property of the estimator's sampling distribution. And we can't observe that directly.
- But in this case something gives it away: [$\hat\mu(8)$]{.fg style="color:blue"} is a negative estimate of a positive quantity.

## Bias and Fake Data

```{r}
#| layout-ncol: 2
#| out-height: 400px
#| fig-asp: .8

n = nrow(earnings)
fake.pop = pop.earnings
fake.sample = fake.pop[sample(1:nrow(fake.pop), size=n, replace=TRUE), ]

model = lm(income~education, data=fake.sample)
muhat = function(ed) predict(model, newdata=data.frame(education=ed))

lm.plot(fake.sample, 'income~education', dot.size=1)
lm.plot(fake.pop, 'income~education',  dot.size=.1)
```


- To really see bias, we have to work with fake data. 
- We'll work with a fake version of the earnings data we've been using.
  - Earlier in the semester, we've used a fake version of our income data for illustration.
  - Now we're actually going to sample from it.

```{r}
#| echo: true
#| eval: false
fake.sample = fake.pop[sample(1:nrow(fake.pop), size=n, replace=TRUE), ]
```

- And we can actually **calculate** our estimator's error.

```{r}
#| echo: true
error = mean(fake.pop$income[fake.pop$education==8]) - muhat(8)
```

- And therefore its bias, the **mean** of our estimator's error **over many samples**.

```{r}
#| cache: true
set.seed(1)
muhat8.sampling.distribution = replicate(1000, {
  fake.sample = fake.pop[sample(1:nrow(fake.pop), size=n, replace=TRUE), ]
  model = lm(income~education, data=fake.sample)
  predict(model, newdata=data.frame(education=8))
}) 
```

## Bias and Fake Data

```{r}
#| layout-ncol: 2
#| out-height: 400px
#| fig-asp: .8
lm.plot(fake.sample, 'income~education', dot.size=1)
ggplot() + geom_histogram(aes(x=muhat8.sampling.distribution, y=after_stat(density)), alpha=.3, bins=100) +
           geom_vline(aes(xintercept=mean(muhat8.sampling.distribution)), color='black', alpha=.5) + 
           geom_vline(aes(xintercept=mean(fake.pop$income[fake.pop$education==8])), color='green', alpha=.5) 
```

- We can see, looking at our estimator's sampling distribution, that it has downward (i.e. negative) bias.

```{r}
#| echo: true
#| eval: false
muhat8.sampling.distribution = replicate(1000, {
  fake.sample = fake.pop[sample(1:nrow(fake.pop), size=n, replace=TRUE), ]
  model = lm(income~education, data=fake.sample)
  predict(model, newdata=data.frame(education=8))
})
```

**Q.** Does this go away with more data?

# It Does Not
```{r}
#| cache: true 
muhat8.sampling.distribution = function(sample.size) { 
  replicate(1000, {
    fake.sample = fake.pop[sample(1:nrow(fake.pop), size=sample.size, replace=TRUE), ]
    model = lm(income~education, data=fake.sample)
    predict(model, newdata=data.frame(education=8))
  })
}
dist.quartern   = muhat8.sampling.distribution(round(n/4))
dist.n          = muhat8.sampling.distribution(n)
dist.fourn      = muhat8.sampling.distribution(4*n) 
```

```{r}
#| layout-ncol: 3
#| out-height: 400px
#| fig-asp: .8 
plot.sampling.dist = function(muhat8.sampling.distribution) {
  ggplot() + geom_histogram(aes(x=muhat8.sampling.distribution, y=after_stat(density)), alpha=.3, bins=100) +
             geom_vline(aes(xintercept=mean(muhat8.sampling.distribution)), color='black', alpha=.5, linewidth=2) + 
             geom_vline(aes(xintercept=mean(fake.pop$income[fake.pop$education==8])), color='green', alpha=.5, linewidth=2) + 
             xlim(-20000,20000)
}
plot.sampling.dist(dist.quartern) + coord_flip()
plot.sampling.dist(dist.n) + coord_flip()
plot.sampling.dist(dist.fourn) + coord_flip()
```


```{r}
#| echo: true
#| eval: false
muhat8.sampling.distribution = function(sample.size) { 
  replicate(1000, {
    fake.sample = fake.pop[sample(1:nrow(fake.pop), size=sample.size, replace=TRUE), ]
    model = lm(income~education, data=fake.sample)
    predict(model, newdata=data.frame(education=8))
  })
}

dist.quartern   = muhat8.sampling.distribution(round(n/4))
dist.n          = muhat8.sampling.distribution(n)
dist.fourn      = muhat8.sampling.distribution(4*n) 
```

## What This Is About

```{r}

model.quartern = lm(income~education, data=fake.pop[sample(1:nrow(fake.pop), size=round(n/4), replace=TRUE), ])
model.n        = lm(income~education, data=fake.pop[sample(1:nrow(fake.pop), size=n, replace=TRUE), ])
model.fourn    = lm(income~education, data=fake.pop[sample(1:nrow(fake.pop), size=4*n, replace=TRUE), ])
preds = function(model,x) { data.frame(x=x, y=predict(model, newdata=data.frame(education=x))) }

x = unique(fake.pop$education)
lm.plot(fake.pop, 'income~education',  dot.size=.025) +
  geom_point(aes(x=x, y=y), data=preds(model.quartern, x), color='cyan', size=1, alpha=.5) +
  geom_line(aes(x=x, y=y), data=preds(model.quartern, x), color='cyan', alpha=.7) +
  geom_point(aes(x=x, y=y), data=preds(model.n, x), color='magenta', size=1, alpha=.5) +
  geom_line(aes(x=x, y=y), data=preds(model.n, x), color='magenta', alpha=.5) +
  geom_point(aes(x=x, y=y), data=preds(model.fourn, x), color='orange', size=1, alpha=.4) +
  geom_line(aes(x=x, y=y), data=preds(model.fourn, x), color='orange', alpha=.6) + 
  scale_y(c(0,150000)) + scale_x()
```

- Bias doesn't go away with more data. What we're seeing here is ...
  - In [blue]{.fg style="color:blue"}, our model's fit to the population.
  - In [cyan]{.fg style="color:cyan"}, its fit to a sample of size $n/4$.
  - In [magenta]{.fg style="color:magenta"}, its fit to a sample of size $n$.
  - In [orange]{.fg style="color:orange"}, its fit to a sample of size $4n$.
- As sample size increases, we get closer to the [blue line]{.fg style="color:blue"}.
- But the blue line *isn't* close to the subpopulation (i.e. column) means $\textcolor{red}{\hat\mu(x)}$.
  
## Least Squares Bias

```{r}
#| layout-ncol: 3
#| out-height: 400px
#| fig-asp: .8 

pop.model = lm(income~education, data=fake.sample)
pop.pred = predict(pop.model, newdata=data.frame(education=8))

plot.sampling.dist(dist.quartern) + geom_vline(aes(xintercept=pop.pred), color='blue', alpha=.5, linewidth=2) + coord_flip()
plot.sampling.dist(dist.n) + geom_vline(aes(xintercept=pop.pred), color='blue', alpha=.5, linewidth=2) + coord_flip()
plot.sampling.dist(dist.fourn) + geom_vline(aes(xintercept=pop.pred), color='blue', alpha=.5, linewidth=2) + coord_flip()
```

- This isn't great, but at least we know what's going on.
- In fact, we can *prove* that the **bias** of the least squares estimator $\hat\mu(8)$ ...
  - is the **error** we get when we fit the model to the whole population
  - no matter what sample size we're using
- As sample size increases ...
  - our estimator's sampling distribution gets narrower and narrower
  - but---if we have bias---it get narrower around the wrong value.


## It's Not Just About Lines

```{r}
#| layout-ncol: 2
#| out-height: 400px
#| fig-asp: .8
#| fig-cap:
#| - Additive bivariate model fit to a sample from the fake population.
#| - Additive bivariate model fit to the fake population itself.
lm.plot2 = function(data, formula, dot.size=2) { 
  data = data[data$race %in% c('Black only', 'White only'),]
  model = lm(formula, data=data)
  muhat = function(r, ed) predict(model, newdata=data.frame(education=ed, race=r))
  ggplot(data, aes(x=education, y=income)) + 
    geom_point(alpha=.1, position=position_jitter(w=.3), size=dot.size) + 
    stat_summary(fun.data=mean_se, geom='pointrange', color='red', alpha=.5) +
    geom_point(aes(x=education, y=muhat(race, education)),  color='blue', alpha=.3) +
    geom_line(aes(x=education, y=muhat(race, education)),  color='blue', alpha=.3) + facet_grid(cols=vars(race))
}

lm.plot2(fake.sample, 'income~factor(education)+race', dot.size=.5) + scale_y(c(0,150000)) 
lm.plot2(fake.pop,    'income~factor(education)+race', dot.size=.01) + scale_y(c(0,150000))
```

- When we use the [additive bivariate model]{.fg style="color:blue"} to predict income from education and race ...
  - some [subpopulation means]{.fg style="color:red"} are still far from the [model's predictions]{.fg style="color:blue"}
  - i.e. some predictions $\inblue{\hat\mu(x)}$ are still biased estimators of $\inred{\mu(x)}$
    
  
## But Lines Aren't Great

```{r}
#| layout-ncol: 2
#| out-height: 400px
#| fig-asp: .8
#| fig-cap:
#| - Additive bivariate model fit to a sample from the fake population.
#| - Additive bivariate model fit to the fake population itself.
lm.plot2 = function(data, formula, dot.size=2) { 
  data = data[data$race %in% c('Black only', 'White only'),]
  model = lm(formula, data=data)
  muhat = function(r, ed) predict(model, newdata=data.frame(education=ed, race=r))
  ggplot(data, aes(x=education, y=income)) + 
    geom_point(alpha=.1, position=position_jitter(w=.3), size=dot.size) + 
    stat_summary(fun.data=mean_se, geom='pointrange', color='red', alpha=.5) +
    geom_point(aes(x=education, y=muhat(race, education)),  color='blue', alpha=.3) +
    geom_line(aes(x=education, y=muhat(race, education)),  color='blue', alpha=.3) + facet_grid(cols=vars(race))
}

lm.plot2(fake.sample, 'income~education+race', dot.size=.5) + scale_y(c(0,150000)) 
lm.plot2(fake.pop,    'income~education+race', dot.size=.01) + scale_y(c(0,150000))
```

- That said, when we use the [additive lines model]{.fg style="color:blue"} to predict income from education and race, we tend to do worse.
  - The **misspecification**, or 'wrongness', of a model is a matter of degree.
  - And this one is more wrong.

# Bias and Summaries

## Two Versions of Our MHT Experiment

```{r}
#| layout-ncol: 2
#| out-height: 400px
#| fig-asp: .8 
 
set.seed(1) 
m = 1e4 
pi = function(x) { 1/(1+exp(-(x-60)/8 )) }
pi2 = function(x) { rep(1/2, length(x)) }
mu = function(w, x) { 1 - (x-50)/30 + w*((x-50)/30)^2/2 }  

X = 50+round(30*runif(m))
W = 0 + (runif(m) <= pi(X))
W2 = 0 + (runif(m) <= pi2(X))
epsilon0 = rnorm(m)
epsilon1 = rnorm(m)
Y0 = mu(0,X) + .25*epsilon0
Y1 = mu(1,X) + .25*epsilon1

pop = data.frame(X=X, W=W, Y=Y0*(1-W) + Y1*W, Y0=Y0, Y1=Y1)
pop2 = data.frame(X=X, W=W2, Y=Y0*(1-W2) + Y1*W2, Y0=Y0, Y1=Y1)

n = 5000
data = pop[sample(nrow(pop), n, replace=TRUE),]
data2 = pop2[sample(nrow(pop), n, replace=TRUE),]

plotX = function(X,W) { X + .15*(2*W-1) }
pop.scatter = ggplot(pop) + 
  geom_point(aes(x=plotX(X,W), y=Y, color=factor(W)), alpha=.05, position=position_jitter(w=.15, h=0), size=.5) + 
  guides(color=FALSE)
pop.scatter2 = ggplot(pop2) + 
  geom_point(aes(x=plotX(X,W), y=Y, color=factor(W)), alpha=.05, position=position_jitter(w=.15, h=0), size=.5) + 
  guides(color=FALSE)
scatter = ggplot(data) + 
  geom_point(aes(x=plotX(X,W), y=Y, color=factor(W)), alpha=.1, position=position_jitter(w=.15, h=0)) + 
  guides(color=FALSE)
scatter2 = ggplot(data2) + 
  geom_point(aes(x=plotX(X,W), y=Y, color=factor(W)), alpha=.1, position=position_jitter(w=.15, h=0), size=.2) + 
  guides(color=FALSE)

big.coords = coord_cartesian(xlim=c(50,80), ylim=c(-1.5,2.5))
narrow.coords = coord_cartesian(xlim=c(60,70), ylim=c(-1.5,2.5))
tight.coords = coord_cartesian(xlim=c(50,80), ylim=c(0,1.5))

annotated.pop.scatter = pop.scatter +  
  stat_summary(aes(x=plotX(X,W), y=Y, color=factor(W)), geom="pointrange", alpha=.3,
                 fun=mean, 
                 fun.min = function(x) mean(x) - sd(x), 
                 fun.max = function(x) mean(x) + sd(x)) +
  geom_bar(aes(x=plotX(X,W), y=10*after_stat(prop), color=factor(W)), position='identity', alpha=.1) 

annotated.pop.scatter2 = pop.scatter2 +  
  stat_summary(aes(x=plotX(X,W), y=Y, color=factor(W)), geom="pointrange", alpha=.3,
                 fun=mean, 
                 fun.min = function(x) mean(x) - sd(x), 
                 fun.max = function(x) mean(x) + sd(x)) +
  geom_bar(aes(x=plotX(X,W), y=10*after_stat(prop), color=factor(W)), position='identity', alpha=.1) 

annotated.pop.scatter  + tight.coords
annotated.pop.scatter2 + tight.coords
```


- Researchers hypothesize that MHT increases bone density. We want to know is how well it works, on average, for the people who take it.
- Above, we've got a population of `r approx(m,1)` people eligible for some form of MHT randomized two ways.
  - Their age at the time the study begins is on the $x$-axis.
  - Bone density, measured 2 years later, is on the $y$-axis.
  - Color indicates whether they [did]{.twocolor-green} or [didn't]{.twocolor-red} do MHT.
  - On the left, randomization probabilities depend on age. On the right, they don't.
- How does our approach randomization affect the bias of our estimator when we use a bad model?

## Version 1: Age-Dependent Randomization
```{r}
#| out-height: 400px
#| layout-ncol: 2
#| fig-asp: .8


general.model  = lm(Y ~ factor(X) * W,  data=pop)
additive.model  = lm(Y ~ factor(X) + W,  data=pop)
additive.lines  = lm(Y ~ X + W, data=pop)

model.plots = function(model, linetype) { 
  grid = expand.grid(X=unique(pop$X), W=c(0,1))
  grid$muhat = predict(model, grid)
  list(geom_point(aes(x=X, y=muhat, color=factor(W)), alpha=.5, data=grid),  
       geom_line(aes(x=X,  y=muhat, color=factor(W)), alpha=.7, linetype=linetype, data=grid))
} 
muhat.general  = function(W, X) { predict(general.model,  newdata=data.frame(X=X, W=W)) }
muhat.additive = function(W, X) { predict(additive.model, newdata=data.frame(X=X, W=W)) }
muhat.lines    = function(W, X) { predict(additive.lines, newdata=data.frame(X=X, W=W)) }
X1 = pop$X[pop$W==1]

annotated.pop.scatter + model.plots(additive.model, 'dashed')  + tight.coords
annotated.pop.scatter + model.plots(additive.lines, 'dashed')  + tight.coords
```

$$
\begin{aligned}
\ingreen{\frac{1}{m_1}\sum_{j:w_j=1}} \qty{ \ingreen{\hat\mu(1,x_j)} - \inred{\hat\mu(0,x_j)} }
&= `r mean(muhat.general(1,X))` - `r mean(muhat.general(0,X))` = `r mean(muhat.general(1,X) - muhat.general(0,X))`  && \text{all functions} \\
&= `r mean(muhat.additive(1,X))` - `r mean(muhat.additive(0,X))` = `r mean(muhat.additive(1,X) - muhat.additive(0,X))`  && \text{additive} \\
&= `r mean(muhat.lines(1,X))` - `r mean(muhat.lines(0,X))` = `r mean(muhat.lines(1,X) - muhat.lines(0,X))`    && \text{lines} \\
\end{aligned}
$$



## Version 2: Age-Independent Randomization

```{r}
#| out-height: 400px
#| layout-ncol: 2
#| fig-asp: .8


general.model  = lm(Y ~ factor(X) * W,  data=pop2)
additive.model  = lm(Y ~ factor(X) + W,  data=pop2)
additive.lines  = lm(Y ~ X + W, data=pop2)

model.plots = function(model, linetype) { 
  grid = expand.grid(X=unique(pop$X), W=c(0,1))
  grid$muhat = predict(model, grid)
  list(geom_point(aes(x=X, y=muhat, color=factor(W)), alpha=.5, data=grid),  
       geom_line(aes(x=X,  y=muhat, color=factor(W)), alpha=.7, linetype=linetype, data=grid))
} 
muhat.general  = function(W, X) { predict(general.model,  newdata=data.frame(X=X, W=W)) }
muhat.additive = function(W, X) { predict(additive.model, newdata=data.frame(X=X, W=W)) }
muhat.lines    = function(W, X) { predict(additive.lines, newdata=data.frame(X=X, W=W)) }
X1 = pop2$X[pop2$W==1]

annotated.pop.scatter2 + model.plots(additive.model, 'dashed')  + tight.coords
annotated.pop.scatter2 + model.plots(additive.lines, 'dashed')  + tight.coords
```

$$
\begin{aligned}
\ingreen{\frac{1}{m_1}\sum_{j:w_j=1}} \qty{ \ingreen{\hat\mu(1,x_j)} - \inred{\hat\mu(0,x_j)} }
&= `r mean(muhat.general(1,X))` - `r mean(muhat.general(0,X))` = `r mean(muhat.general(1,X) - muhat.general(0,X))`  && \text{all functions} \\
&= `r mean(muhat.additive(1,X))` - `r mean(muhat.additive(0,X))` = `r mean(muhat.additive(1,X) - muhat.additive(0,X))`  && \text{additive} \\
&= `r mean(muhat.lines(1,X))` - `r mean(muhat.lines(0,X))` = `r mean(muhat.lines(1,X) - muhat.lines(0,X))`    && \text{lines} \\
\end{aligned}
$$