---
title: "Lecture 1"
subtitle: "Descriptive Regression with a Binary Covariate"
format:
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---

```{r libraries}
library(ggplot2)
library(dplyr)
library(magrittr)
library(knitr)
library(scales)
options(knitr.kable.NA = '')

library(cpsR)
library(usmap)
library(sf)

library(ggplot2)

library(scales)
library(knitr)

percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }
census.api.key = '8084f81e4f71e4d32638dac15f188b424f72d063'
# for cps variable interpretation see https://www2.census.gov/programs-surveys/cps/datasets/2022/march/asec2022_ddl_pub_full.pdf
```

```{r ggplot-themes}
class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
		    panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
		    axis.ticks.x = element_blank(),
		    axis.ticks.y = element_blank(),
		    axis.text.x  = element_text(colour = "#aaaaaa"),
		    axis.text.y  = element_text(colour = "#aaaaaa"),
		    axis.title.x  = element_text(colour = "#aaaaaa"),
		    axis.title.y  = element_text(colour = "#aaaaaa", angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
   	         axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
	         panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(simplified.theme)
```


```{r table-styling}
ellipsis <- function(df, nhead=5L, ntail=2) {
    rownames(df) = as.character(1:nrow(df))
    stopifnot("data.frame" %in% class(df))
    els <- rep(NA, ncol(df)) %>% 
        matrix(nrow=1, dimnames=list(NULL, names(df))) %>% 
        data.frame(stringsAsFactors=FALSE) %>% 
        tibble::as_tibble()
    rownames(els) = "⋮"
    dplyr::bind_rows(head(df, nhead), els, tail(df, ntail))
}
```


```{r get-california-data}
#| cache: true

california.state.code = 6
cps.data = cpsR::get_asec(year=2022, vars=c('pearnval', 'gestfips', 'h_numper', 'htotval', 'a_age', 'a_hga','gereg', 'a_maritl', 'a_sex','peafever','a_wkstat'), key=census.api.key)
```

```{r process-california-data}
all.earnings = data.frame(age=cps.data$a_age,
                          state.code = cps.data$gestfips,
                          income=cps.data$htotval, 
                          education=case_match(cps.data$a_hga,
                                          0  ~ 0,   # child
                                          31 ~ 0,   # < grade 1
                                          32 ~ 4,   #  grade 1-4
                                          33 ~ 6,   #  grade 5-6
                                          34 ~ 8,   #  grade 7-8 
                                          35 ~ 9,   #  grade 9
                                          36 ~ 10,  #  grade 10
                                          37 ~ 11,  #  grade 11
                                          38 ~ 11,  #  grade 12 no diploma
                                          39 ~ 12,  #  high school grad
                                          40 ~ 13,  #  some college
                                          41 ~ 14,  #  associate degree(vocational)
                                          42 ~ 14,  #  associate degree (academic)
                                          43 ~ 16,  #  bachelors degree
                                          44 ~ 18,  #  masters degree
                                          45 ~ 20,  #  professional school degree
                                          46 ~ 20)) #  doctorate

in.interval = function(X, interval) { interval[1] <= X & X <= interval[2] }
all.earnings$employed<- ifelse(cps.data$a_wkstat >5, 0,ifelse(cps.data$a_wkstat<2,NA,1))
all.earnings$fouryrdegree = all.earnings$education >=16
all.earnings$degree = factor(all.earnings$fouryrdegree,
                    levels=c(FALSE, TRUE), 
                    labels=c('no 4-year degree','4-year degree')) 

earnings = all.earnings[all.earnings$age %>% in.interval(c(25,35)) & 
                        all.earnings$state.code == california.state.code &
                        all.earnings$education > 8, ]

rownames(earnings)=NULL
earnings$person = 1:nrow(earnings)
```

## Today

- We'll talk majorly about two examples you'll be seeing a lot in this class ...
  - 2022 CPS wage data 
  - Sampling examples which would include sum of two dice and poling
- Our goal for today is to understand and learn how to summarize the data
  - location
  - spread
  - distribution 
- We'll also talk about (in relation to sampling) ...
  - population means
  - subpopulation means 
  - differences in subpopulation means

## Context: 2022 CPS Wage Data
- Today, to stay grounded, I want to focus on some real survey data.
- The Current Population Survey is a survey run by Census Bureau and Bureau of Labor Statistics
    - Each year, about 1/2500 US residents is surveyed.
    - It's approximately correct to think of these residents as being chosen randomly 
    from the set of all residents with equal probability.
    - This approximation gets better if you focus on a single state.
- We'll look at surveyed California residents 
    - between the ages of 25 and 35
    - with at least an 8th-grade education
- This sample includes `r nrow(earnings)` people. Here's a partial view of their responses.

 \ \

```{r}
cols = c('income', 'age', 'education')
earnings[,cols] %>% ellipsis(nhead=4, ntail=2) %>% kable() 
```

## Here is a Complete View of Their Incomes

```{r}
lims = c(-.5e5, 5e5)
scatter = ggplot(earnings) + 
            geom_point(aes(x=person, y=income), alpha=.05) +
            scale_y_continuous(limits=lims, labels=label_dollar())
scatter
```

- Each person gets a dot.
- The y-coordinate of the dot shows their income.
- The x-coordinate of the dot is arbitrary.
    - It's their row in the table.
- The plot, in effect, shows our people 
    - lined up in some arbitrary order
    - standing as tall as their income.
- The problem with this is that it's hard to make sense of by looking at it.
- It often helps to look at *data summaries*, a.k.a. *descriptive statistics*.

# Context: Sum of Two Dice and Odd Rolls

::: {.hidden}

```{r dice_simulation}
set.seed(123)  # Setting seed for reproducibility

# Simulating the sum of two dice rolls
roll_two_dice <- function() {
  die1 <- sample(1:6, 1, replace = TRUE)
  die2 <- sample(1:6, 1, replace = TRUE)
  return(die1 + die2)
}

# Generating a list of simulated dice sums for 10 runs
num_rolls <- 10
dice_sums <- replicate(num_rolls, roll_two_dice())

# Counting the number of odd rolls
num_odd_rolls <- sum(dice_sums %% 2 == 1)

# Displaying the results
cat("Results for Sum of Two Dice Rolls:\n")
cat("List of Dice Sums:", dice_sums, "\n")
cat("Number of Odd Rolls:", num_odd_rolls, "\n")
```    

:::
- Let's consider a sum of two dice
  - You use a list of simulated dice sums obtained by rolling two dice r num_rolls times.
  - You also count the number of odd rolls in these 10 runs.

## Context : Poling data

::: {.hidden}

```{r population}
our.seed = 0
set.seed(our.seed)
N=7233584
p=4999960/N
y= c(rep(1,p*N), rep(0, (1-p)*N))
y=sample(y) # shuffle the order 
population.proportion = p
```

```{r sample}
set.seed(our.seed)

n = 625
J = sample(N, n, replace=TRUE)
Y = y[J]
sample.proportion = mean(Y)
```

:::
- Suppose that, week before the 2020 presidential election, you did some polling.
  - You use a list of the $N \approx `r approx(N)`$ people registered to vote in Georgia.
  - And you make $n=`r n`$ phone calls.
  - Each call, you select a voter uniformly at random from the list, e.g. by rolling a `r approx(N)` -sided die.
  - And then you ask the potential voter whether they plan to vote.
- Suppose also that all these registered voters will
  - Pick up the phone will called
  - Respond honestly to your questions
  - Not change their minds about voting

## Polling Results

- You put your polling results in a table.
- It has one column for each call.
- In that columns, you record ...
  - the call number $i$---a number from $1 \ldots `r n`$.
  - the response $Y_i$ of the person you called---$1$ for 'yes and $0$ for no'.

$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}  \\
Y_i  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$ 

- You summarize the results with an extra column: the mean of the responses.
  - Remember that the mean of a binary variable is a proportion. 
  - You found that `r percent(mean(Y))` of the people polled said they would vote.

## Sample Versus Population

- We want to estimate the proportion of *all registered voters* --- our population --- who will vote.
- To do this, we use the proportion of *polled voters* --- our sample --- who said they would.
- When the election occurs, we get to see who turns out to vote.
  - `r approx(p*N)` people, or roughly `r percent(mean(Y))` of registered voters, actually vote.
 
\ \

:::: columns
::: column
### Before the Election 

$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}  \\
Y_i  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$

Outcomes for our sample of polled voters. \
To enumerate our sample, we ...

- give each call a number $i \in 1 \ldots `r n`$.
- write $Y_i$ for the turnout of the $i$th person we called.
:::
::: column
### After the Election

$$
\begin{array}{r|rrrrrr|r}
j & 1 & 2 & 3 & 4 & \dots &   `r approx(N)` & \bar{y}_{`r approx(N)`}  \\
y_{j} & `r y[1]` & `r y[2]` & `r y[3]` & `r y[4]` & \dots & `r y[N]`   & `r approx(mean(y))`  \\
\end{array}
$$
Outcomes for the population of registered voters. \
To enumerate our population, we ... 

- give each registered voter a # $j \in 1 \ldots `r approx(N)`$.
- write $y_j$ for the turnout of the person with ID $j$.
:::
::::

## Two Questions

Our sample proportion $`r sum(Y)` / `r n` \approx `r approx(mean(Y))`$ 
is close to the population proportion $`r approx(sum(y))` / `r approx(N)` \approx `r approx(mean(y))`$

1. Did we get lucky?
2. Could we have predicted how close we'd get before the election happened?


::: fragment

- Answers: No and Yes.
- To get these answers, we'll think about what'd happen if our friends had run identical polls.
  - Each friend will choose have a different random sample.
  - And estimate the population proportion by a different sample proportion.
- We'll see how close these estimates tend to be.

:::

## Connecting Sample and Population

- For each call $i$, we randomly select a voter with an id we'll call $J_i$.
- And we record as that call's outcome the turnout of that voter: $Y_i=y_{J_i}$.
- On each call, each registered voter has a $1/`r approx(N)`$ chance of being called.
  - This is called sampling with replacement because we could call the same person twice.
  - In our poll, this is unlikely because we're making a small number of calls relative to population size.
  



```{r mapping}
# Create fake locations for the people in our population
# by taking random convex combinations of state vertices
set.seed(0)
ga.shape = us_map(include = c("GA"))

# To keep all this from being overwhelming, we'll have 10000 fake locations
# The jth person is at the (j %% M + 1)st location
M = 10000 
V = nrow(ga.shape)
weights = abs(rt(M*V,1))
dim(weights) = c(M,V)
weights = (1/rowSums(weights)) * weights
locations = data.frame(x=weights %*% ga.shape$x, y=weights %*% ga.shape$y)

ga = ggplot() + geom_polygon(data=ga.shape, aes(x = x, y = y), fill="white", color="black") + coord_fixed() + blank.theme
```
`

:::: {.r-stack}
::: {.fragment .fade-out data-fragment-index="0"}

```{r}
#| out-height: 300px
sample.locations = locations[J %% M + 1, ]
sample.locations$label = J
ga + geom_text(aes(x = x, y = y, label=label), alpha=.5, data=sample.locations[c(1,2,n),]) 
```

:::
::: {.fragment .current-visible data-fragment-index="0"}

```{r}
#| out-height: 300px
ga + geom_text(aes(x = x, y = y, label=label), alpha=.2, data=sample.locations) 
```
:::
::: {.fragment .current-visible data-fragment-index="1"}
```{r}
#| out-height: 300px
ga + geom_point(aes(x = x, y = y), alpha=.2, data=sample.locations) 
```
:::
::::

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{select/record}  & J_1 & Y_1 & J_2 & Y_2 & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\text{outcome}    & `r J[1]` & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`} 
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`} 
                  & `r approx(mean(Y))` \\
\end{array}
$$ 

:::: aside
Obviously these locations are fake. Most people in Georgia live near Atlanta, not where these dots are. This is just to illustrate the sampling scheme. 
:::

## Our Poll

```{r}
# Now we'll repeat our sampling so we have polls for a few friends
set.seed(our.seed)
J1 = sample(N, n, replace=TRUE)
J2 = sample(N, n, replace=TRUE)
J3 = sample(N, n, replace=TRUE)
JM = sample(N, n, replace=TRUE)
```

```{r}
J = J1
Y = y[J1]

sample.locations = locations[J %% M +1, ]
sample.locations$label = J
ga + geom_text(aes(x = x, y = y, label=label), alpha=.2, data=sample.locations) 
```


$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{select/record}  & J_1 & Y_1 & J_2 & Y_2 & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\text{outcome}    & `r J[1]` & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`} 
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`} 
                  & `r approx(mean(Y))` \\
\end{array}
$$

## Our First Friend's Poll

```{r}
J = J2
Y = y[J2]

sample.locations = locations[J %% M + 1, ]
sample.locations$label = J
ga + geom_text(aes(x = x, y = y, label=label), color='magenta', alpha=.2, data=sample.locations)
```

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{select/record}  & J_1 & Y_1 & J_2 & Y_2 & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\text{outcome}    & `r J[1]` & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`} 
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`} 
                  & `r approx(mean(Y))` \\
\end{array}
$$

## Our Second Friend's Poll

```{r}
J = J3
Y = y[J3]

sample.locations = locations[J %% M + 1, ]
sample.locations$label = J
ga + geom_text(aes(x = x, y = y, label=label), color='cyan', alpha=.2, data=sample.locations)
```

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{select/record}  & J_1 & Y_1 & J_2 & Y_2 & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\text{outcome}    & `r J[1]` & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`} 
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`} 
                  & `r approx(mean(Y))` \\
\end{array}
$$

## The Sampling Distribution of our Estimate

\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{pollster} & J_1 & Y_1 & J_2 & Y_2   & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\hline
\color{black}   1   
    & \color{black}   `r J=J1; Y=y[J]; J[1]`    & \color{black}  `r Y[1]` 
    & \color{black}   `r J[2]`   & \color{black}   `r Y[2]`           & \color{black}   \dots 
    & \color{black}   `r J[n]`   & \color{black}   `r Y[n]`           & \color{black}   `r approx(mean(Y))` \\
\color{magenta} 2   
    & \color{magenta} `r J=J2; Y=y[J]; J[1]`    & \color{magenta}  `r Y[1]` 
    & \color{magenta} `r J[2]`   & \color{magenta}  `r Y[2]`          & \color{magenta} \dots 
    & \color{magenta} `r J[n]`   & \color{magenta} `r Y[n]`           & \color{magenta} `r approx(mean(Y))` \\
\color{cyan}    3   
    & \color{cyan}    `r J=J3; Y=y[J]; J[1]`    & \color{cyan}     `r Y[1]` 
    & \color{cyan}    `r J[2]`   & \color{cyan}    `r Y[2]`           & \color{cyan}    \dots 
    & \color{cyan}    `r J[n]`   & \color{cyan}    `r Y[n]`           & \color{cyan}    `r approx(mean(Y))` \\
{\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots} &  {\vdots} \\ 
\color{gray}    1M   
    & \color{gray}    `r J=JM; Y=y[J]; J[1]`    & \color{gray}     `r Y[1]` 
    & \color{gray}    `r J[2]`   & \color{gray}    `r Y[2]`           & \color{gray}    \dots 
    & \color{gray}    `r J[n]`   & \color{gray}    `r Y[n]`           & \color{gray}    `r approx(mean(Y))` \\
 {\vdots} & {\vdots} & {\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots}  & {\vdots} \\ 
\hline
{f_{\cdot}(\cdot)} & \hphantom{Ber(`r approx(p)`)} & & \hphantom{Ber(`r approx(p)`)} &  & & \hphantom{Ber(`r approx(p)`)} & \hphantom{\frac{Bin(`r n`,`r approx(p)`)}{`r n`}} \\
\end{array}

- What is the distribution of $Y_1$?
- What is the distribution of $\sum_{i}^{`r n`}Y_i$ \ ?

## The Sampling Distribution of our Estimate

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{pollster} & J_1 & Y_1 & J_2 & Y_2   & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\hline
\color{black}   1   
    & \color{black}   `r J=J1; Y=y[J]; J[1]`    & \color{black}    `r Y[1]` 
    & \color{black}   `r J[2]`   & \color{black}   `r Y[2]`           & \color{black}   \dots 
    & \color{black}   `r J[n]`   & \color{black}   `r Y[n]`           & \color{black}   `r approx(mean(Y))` \\
\color{magenta} 2   
    & \color{magenta} `r J=J2; Y=y[J]; J[1]`    & \color{magenta}  `r Y[1]` 
    & \color{magenta} `r J[2]`   & \color{magenta}  `r Y[2]`          & \color{magenta} \dots 
    & \color{magenta} `r J[n]`   & \color{magenta} `r Y[n]`           & \color{magenta} `r approx(mean(Y))` \\
\color{cyan}    3   
    & \color{cyan}    `r J=J3; Y=y[J]; J[1]`    & \color{cyan}     `r Y[1]` 
    & \color{cyan}    `r J[2]`   & \color{cyan}    `r Y[2]`           & \color{cyan}    \dots 
    & \color{cyan}    `r J[n]`   & \color{cyan}    `r Y[n]`           & \color{cyan}    `r approx(mean(Y))` \\
{\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots} &  {\vdots} \\ 
\color{gray}    1M   
    & \color{gray}    `r J=JM; Y=y[J]; J[1]`    & \color{gray}     `r Y[1]` 
    & \color{gray}    `r J[2]`   & \color{gray}    `r Y[2]`           & \color{gray}    \dots 
    & \color{gray}    `r J[n]`   & \color{gray}    `r Y[n]`           & \color{gray}    `r approx(mean(Y))` \\
 {\vdots} & {\vdots} & {\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots}  & {\vdots} \\ 
\hline
{f_{\cdot}(\cdot)} & Ber(`r approx(p)`) & & Ber(`r approx(p)`) &  & & Ber(`r approx(p)`) & \frac{Bin(`r n`,`r approx(p)`)}{`r n`} \\
\end{array}
$$

- $Y_1$ has a Bernoulli distribution 
- $\sum_{i}^{`r n`}Y_i$, a sum of *independent and identically distributed* Bernoullis, has a Binomial distribution.

## Bernoulli and Binomial Distributions

- Our individual responses $Y_1...Y_{`r n`}$ are Bernoulli Random Variables 
$$
\begin{aligned}
f_{Y_1}(y) &= (`r approx(p)`)^{y} (`r approx(1-p)`)^{1-y} \qfor  y=0,1\\
\vdots \quad  &=   \quad\quad\quad\quad\quad\quad \vdots  \\
f_{Y_{`r n`}}(y) &= (`r approx(p)`)^{y} (`r approx(1-p)`)^{1-y} \qfor  y=0,1
\end{aligned}
$$

- Our sum of responses, $S_{`r n`} = \sum_{i=1}^{`r n`} Y_i$, is a Binomial Random Variable
$$
\begin{aligned}
f_{S_{`r n`}}(s) = {`r n` \choose s} 
(`r approx(1-p)`)^{s} 
(`r approx(1-p)`)^{`r n`-s} 
\qfor  s=0 \ldots `r n`
\end{aligned}
$$

- Our sample proportion, $\overline{Y}_{`r n`}$, has the distribution of this Binomial Random Variable divided by $`r n`$.

## What are descriptive statistics?

-   Descriptive statistics are summaries of the data we have.
-   The data we have might be
    -   a random sample (e.g. an exit poll) from a population (e.g. voters) we're interested in.
    -   the whole population (e.g., administrative vote counts). Sometimes we call this a census.
-   Having the whole population is becoming more common in the era of Big Data.
    -   In particular, companies like facebook, google, and amazon are mostly interested in their users.
    -   And, for the most part, they don't have to poll them. They're always recording.
-   That said, if you are interested in the causal inference, that's impossible.
    -   We tend to phrase causal questions in terms of multiple versions of the same person.
    -   Versions who take different drugs, are exposed to different policies etc.
    -   And you can never see all of them. There's only one real version of each person.
-   More on that in a few weeks.


## What makes a good summary?

-  It should be brief
    -   Maybe a number or two.
    -   Or an-easy-to-understand plot.
-  It should tell you what you want to know.
    -   This one's more context specific.
-  Let's look at a couple common types of summary.
    - Location
    - Spread
    - Distribution

## Location

```{r}
#| out-height: 500px
x.med  = 1300
x.mean = 1400
size.mean = 3
size.med = 3

summaries = earnings %>% 
    summarize(mean=mean(income), 
              median=median(income), 
              sd=sd(income), 
              mad=mad(income, constant=1))
scatter + geom_point(aes(x=x.mean, y=mean), size=size.mean, color='red', data=summaries) + 
          geom_hline(aes(yintercept=mean), color='red', data=summaries) +
          geom_point(aes(x=x.med, y=median), size=size.med, color='orange', data=summaries) +
          geom_hline(aes(yintercept=median), color='orange', data=summaries)
```

- The **location** of the data is where it is.
- It's about approximating the data by a *constant*.
$$ Y_i \approx \mu \qfor i=1,\dots,n $$

- Depending on your criteria, you may get different answers.
    - e.g. the [Median]{.fg style="--col: orange"}
    - e.g. the [Mean]{.fg style="--col: red"}
- More on this later today.

## Spread

```{r}
#| out-height: 500px
scatter + geom_pointrange(aes(x=x.mean, y=mean, ymin=mean-sd, ymax=mean+sd), color='red', data=summaries)  +
          geom_hline(aes(yintercept=mean), color='red', data=summaries) +
          geom_pointrange(aes(x=x.med, y=median, ymin=median-mad, ymax=median+mad), color='orange', data=summaries) +
          geom_hline(aes(yintercept=median), color='orange', data=summaries)
```

- The **spread** of the data is how far it is from its location.
- Different measures of spread tend to be used with different measures of location.
    - the [Median Absolute Deviation]{.fg style="--col: orange"} is used with the Median.
    - the [Standard Deviation]{.fg style="--col: red"} is used with the Mean.


## Spread and Residuals
```{r}
#| layout-ncol: 2
#| layout-nrow: 2

scatter + geom_hline(aes(yintercept=median(income)), color='orange') +
          geom_segment(aes(x=person, xend=person, y=income, yend=median(income)), color='gray', alpha=.1)
scatter + geom_hline(aes(yintercept=mean(income)), color='red') + 
          geom_segment(aes(x=person, xend=person, y=income, yend=mean(income)), color='gray', alpha=.1)

res.lims = c(-2.5e5, 2.5e5)
ggplot(earnings) + geom_point(aes(x=person, y=income-median(income)), color='orange', alpha=.3) +
    scale_y_continuous(labels=label_dollar(), limits=res.lims) 
ggplot(earnings) + geom_point(aes(x=person, y=income-mean(income), size=abs(income-mean(income))), color='red', alpha=.3) +
    scale_y_continuous(labels=label_dollar(), limits=res.lims) + scale_size_area() + guides(size=FALSE) 
```

- Spread summarizes the size of the *residuals* $\hat\varepsilon_i = Y_i - \hat\mu$ left over after constant approximation.
- The median absolute deviation is the median size of the residuals.
- The standard deviation is the root mean squared size of the residuals.
  - variance
  - think: size-weighted avg of size of the residuals.
  - that's what we're showing in the plot: area of the dots is proportional to the size of the residuals
  - so two dots at 100k count like ...?

## Distribution

```{r}
#| layout-ncol: 2
h=ggplot(earnings) + geom_histogram(aes(x=income, y=after_stat(density)), 
                                  colour="black", fill="blue") + coord_flip() + xlim(0,5e5)
h 
scatter + ylim(0,5e5)
```

- This is called a histogram
- It tells us where ...
    - This is California. There are people off the the right, but who cares.
    - Mark Zuckerberg, Marc Benioff, etc. All the Marks.
- You can more or less read location and spread off histogram
    - It's instructive to try...

$$

# Subsample means and jitter plots

## Subsample Means

-   In this class, we focus on a single 'outcome' variable. 
    - We'll write $Y$ for this.
-   Typically, we'll look at the mean of this outcome $Y$ within subsamples.
-   We call this *Regression Analysis*
-   We define those subsamples in terms of our other variables called covariates. 
    - We'll write $X$ for this.
    - Often, we'll work with several covariates, so we'll have to think of $X$ as a vector.
    - But to keep things simple early in the semester, we'll use only one, so $X$ is a scalar.
-   Today we'll consider *binary* covariate $X$ that indicates whether a person has a 4-year degree.
    - $X_i=0$ if person $i$ doesn't have a 4-year degree.
    - $X_i=1$ if they do.
    
## Our Data as A Table

:::: columns
::: column
### In English
$$
\begin{array}{r|r r}
\text{person} & \text{household income} & \text{4-year degree}   \\
\hline
1 & \$250000  & \text{yes} \\
2 & \$32086   & \text{yes} \\
\vdots & \vdots & \vdots   \\
11 & \$300640  & \text{no} \\
\vdots & \vdots  & \vdots\\
2258 & \$24000     & \text{yes} \\
\hline
\end{array}
$$
:::

::: column
### In Mathematical Notation
$$
\begin{array}{r|r r}
i & Y_i & X_i \\
\hline
1 & 250000  & 1  \\
2 & 32086   & 1  \\
\vdots & \vdots & \vdots \\
11 & 300640  & 0 \\
\vdots & \vdots  & \vdots\\
2258 & 24000 & 1  \\
\hline
\end{array}
$$

:::
::::

## A Scatter Plot of Our Data [Handled more or less above, but fix/explain jitter vs person-index stuff]
```{r}
attach(earnings)
summaries = earnings %>% group_by(degree) %>%
                         summarize(mean=mean(income), sd=sd(income))

ggplot(earnings) + geom_point(aes(x=degree, y=income, color=degree), shape=1, alpha=.1) +
                   geom_point(aes(x=degree, y=mean), data=summaries) +
                   scale_y_continuous(limits=c(-.5e5,5e5)) + guides(color=FALSE)
```

- This is a way of displaying everyone in our sample. Each person is a dot.
    - Their horizontal position---their $x$-coordinate---is determined by the covariate $X_i$.
    - Their vertical position---their $y$ coordinate---is determined by the outcome $Y_i$.
- We've overlaid a black dot to show the subsample means.
- It's a bit hard to see what's going on because everyone is plotted on top of one another.
- That's because $X_i$ is binary---it takes on only two values. But it's an easy fix.


## An Idea for plots
-  don't put everyone on top of each other
-  line People up in Arbitary Order to create spacing
-  This lets us see variation in our outcome within groups
-  Can be useful to add summary; subsample standard deviation 

## A Jitter Plot of Our Data

```{r}
attach(earnings)

ggplot(earnings) +  geom_point(aes(x=degree, y=income, color=degree), 
                                position=position_jitter(width=.3), alpha=.1) + 
                   geom_point(aes(x=degree, y=mean), data=summaries) +
                   scale_y_continuous(limits=c(-.5e5,5e5)) + guides(color=FALSE)
```

- This plot tweaks the x-coordinate to give the dots more space. 
  - We have the same dots, but with a bit of random noise added to each one's x-coordinate.
  - This makes it easier to see how many dots there are at a given income level (i.e. y-coordinate).
  - We keep the noise small so we can still see the real value of $X_i$
  - Each vertical tube is a set of people with the same degree status.


## Subsample Standard Deviations

```{r}
ggplot(earnings) +  geom_point(aes(x=degree, y=income, color=degree), 
                                position=position_jitter(width=.3, height=0), alpha=.1) + 
                   geom_pointrange(aes(x=degree, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries) +
                   scale_y_continuous(limits=c(-.5e5,5e5)) + guides(color=FALSE)
```

- To summarize spread, we can draw in the subsample standard deviations.
- I've drawn in bars extending one standard deviation above and below the mean.


## Special Case: Binary Outcomes

- Let's revisit our data thinking of *employment* as the outcome.
- That's binary. You're either employed or not.
- So we have a binary covariate *and* a binary outcome.

$$
$$

:::: columns

::: column
### In English
$$
\begin{array}{r|rr}
\text{Household} & \text{Degree} & \text{Employed}  \\
\hline
1 & 0 & 1 \\
2 & 1  & 1 \\
\vdots & \vdots & \vdots \\
2245 & 0 & 1\\
\hline
\end{array}
$$

:::

::: column
### In Mathematical Notation
$$
\begin{array}{r|rr}
i & X_i & Y_i  \\
\hline
1 & 0 & 1 \\
2 & 1  & 1 \\
\vdots & \vdots & \vdots \\
2245 & 0 & 1\\
\hline
\end{array}
$$
:::

::::


## Binary Outcome Regression

```{r}
laborforce<- subset(earnings,!(is.na(employed)))

summaries = laborforce %>% group_by(fouryrdegree) %>%
                         summarize(mean=mean(employed), sd=sd(employed))

ggplot(laborforce) +  geom_point(aes(x=fouryrdegree, y=employed, color=degree), 
                                position=position_jitter(width=.3, height=.01), alpha=.1) + 
                      geom_pointrange(aes(x=fouryrdegree, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries) +
                    guides(color=FALSE)

```
To give our dots a little space, we've added little vertical jitter here as well.

**Activity**. In pairs, discuss ...

- The substantive interpretation of the subsample means and their difference.
- Why is the SD smaller for those with four-year degrees than those without? 
    - Hint: look at the variance formula for binary variables.
  
