---
title: "Lecture 18"
subtitle: "Inverse Probability Weighting"
---

## None of these models fits the data well.

{{< include shared-code.qmd >}}

```{r}
set.seed(1) 
m = 1e6 
pi = function(x) { ifelse(x <= 65, .2, .8) }
mu = function(w, x) { 2 - (x-50)/30 + w*((x-50)/30)^2/2 }  

X = 50+round(30*runif(m))
W = 0 + (runif(m) <= pi(X))
epsilon0 = rnorm(m)
epsilon1 = rnorm(m)
Y0 = mu(0,X) + .15*epsilon0
Y1 = mu(1,X) + .15*epsilon1

pop = data.frame(X=X, W=W, Y=Y0*(1-W) + Y1*W, Y0=Y0, Y1=Y1)

n = 10000
data = pop[sample(nrow(pop), n, replace=TRUE),]

prediction.function = function(model) { 
  function(w, x) { 
    predict(model, newdata=data.frame(W=w, X=x))
  }
}

fits = list(
B =  lm(Y ~ X, data=data),
D =  lm(Y ~ W, data=data), 
E =  lm(Y ~ W + I(X <= 65), data=data),
A =  lm(Y ~ W + X, data=data),
C =  lm(Y ~ W * I(X <= 65), data=data),
F =  lm(Y ~ W * X, data=data))
muhats = fits |> map(prediction.function)

grid = expand_grid(W=c(0,1), X=seq(50,80,by=.01))
predictions = muhats |> map(\(muhat) { 
  grid$muhat = muhat(grid$W, grid$X)
  grid
}) |> bind_rows(.id='model') 


plotX = \(w,x) x + .15*(2*w-1)
```

```{r} 
#| out-height: 600

mht.plot = scatter = ggplot(data) + 
  geom_point(aes(x=plotX(W,X), y=Y, color=factor(W)), alpha=.025, size=.05, position=position_jitter(w=.15, h=0)) + 
  stat_summary(aes(x=plotX(W,X), y=Y, color=factor(W)), geom="pointrange", alpha=.3, size=.5,
                 fun=mean, 
                 fun.min = function(x) mean(x) - sd(x), 
                 fun.max = function(x) mean(x) + sd(x)) +
  geom_bar(aes(x=plotX(W,X), y=15*after_stat(prop), fill=factor(W), color=factor(W)), 
          alpha=.5, linewidth=.2) + 
          xlab('') + ylab('') + 
          guides(color='none', fill='none') +
          scale_x_continuous(breaks=seq(50,80,by=5))

mht.plot + 
  geom_line(data=predictions, aes(x=X, y=muhat, color=factor(W)), linewidth=.5) + 
  facet_wrap(~model, ncol=3)
```


$$
\begin{aligned}
\model_1 &= \qty{ m(w,x) = a    + bx \ : \ \text{ any numbers }\  a,b }   && \text{ one line } \\
\model_2 &= \qty{ m(w,x) = a(w)        : \ \text{ any function }\  a(w) }   && \text{ horizontal lines } \\
\model_3 &= \qty{ m(w,x) = a(w) + bx \ : \ \text{ any function }\ a(w) \ \text{ and number }\ b } && \text{ parallel lines } \\
\model_4 &= \qty{ m(w,x) = a(w) + b(w)x \ : \ \text{ any functions }\ a(w),\ b(w) } && \text{ lines } \\
\model_5 &= \qty{ m(w,x) = a(w) + b \operatorname*{1_{\le 65}}(x)  \ : \ \text{ any function } a(w) \ \text{ and number }\ b } && \text{ additive piecewise-constant } \\
\model_6 &= \qty{ m(w,x) = a(w) + b(w) \operatorname*{1_{\le 65}}(x)  \ : \ \text{ any functions } a(w),\ b(w) } && \text{ piecewise-constant } 
\end{aligned}
$$

  
## But one does give us an unbiased estimate of the ATT.

```{r}
#| cache: true
set.seed(1) 

estimates = function(rep) { 
m = 1e6 
pi = function(x) { ifelse(x <= 65, .2, .8) }
mu = function(w, x) { 2 - (x-50)/30 + w*((x-50)/30)^2/2 }  

X = 50+round(30*runif(m))
W = 0 + (runif(m) <= pi(X))
epsilon0 = rnorm(m)
epsilon1 = rnorm(m)
Y0 = mu(0,X) + .15*epsilon0
Y1 = mu(1,X) + .15*epsilon1

n = 10000
pop = data.frame(X=X, W=W, Y=Y0*(1-W) + Y1*W, Y0=Y0, Y1=Y1)
data = pop[sample(nrow(pop), n, replace=TRUE),]

fits = list(
B =  lm(Y ~ X, data=pop),
D =  lm(Y ~ W, data=pop), 
E =  lm(Y ~ W + I(X <= 65), data=pop),
A =  lm(Y ~ W + X, data=pop),
C =  lm(Y ~ W * I(X <= 65), data=pop),
F =  lm(Y ~ W * X, data=pop))
muhats = fits |> map(prediction.function)

ATT.estimates = muhats |> map(\(muhat) { 
  data.frame(estimate=mean(muhat(1, pop$X[pop$W==1]) - muhat(0, pop$X[pop$W==1])))
}) |> bind_rows(.id='model')
ATT.estimates$true = mean(mu(1,pop$X[pop$W==1]) - mu(0,pop$X[pop$W==1]))
ATT.estimates
}

ests = 1:100 |> map(estimates) |> bind_rows(.id='rep')
```

```{r}
#| out-height: 600px
theATT = mean(ests$true)
ggplot(ests) +
        geom_histogram(aes(x=estimate, y=after_stat(density)), bins=30, alpha=.3) + 
        geom_vline(xintercept=theATT, color='green') + 
        facet_wrap(~model, ncol=3, scales='free')
```

## Why?

```{r} 
#| out-height: 500px
mht.plot 
```

- Let's think about what we're estimating: the ATT.
- This is a weighted average of column means.

$$
\begin{aligned}
\text{ATT} 
&= \frac{1}{m_1}\sum_{j:w_j=1} \qty{ \mu(1,x_j) - \mu(0,x_j)} \qfor \mu(w,x) = \frac{\sum_{j: w_j,x_j=w,x } y_j}{\sum_{j: w_j,x_j=w} 1} \qand m_w = \sum_{j: w_j=w} 1 \\
&= \sum_{x} \ingreen{p(1,x)} \qty{ \ingreen{\mu(1,x)} - \inred{\mu(0,x)}} \qfor p(w,x) = \frac{\sum_{j: w_j=w, x_j=x} 1}{\sum_{j: w_j=w} 1 } \\
&= \sum_{w,x} \alpha(w,x) \mu(w,x) \qfor \alpha(w,x) = \begin{cases}+p(1,x) & \text{ if } w=1 \\ -p(1,x) & \text{ if } w=0 \end{cases}
\end{aligned}
$$

- What you may notice is that, because these proportions $p(w,x)$ are piecewise-constant, they're in the model.

## 
### A Claim

- You'll get an unbiased estimate of the summary $\sum_{w,x}\alpha(w,x) \mu(w,x)$
- if the regression model $\model$ contains at least one of ...
    - the subpopulation mean function $\mu(w,x)$ 
    - the weights $\alpha(w,x)$

### Proof Ideas 

1. Remember that our sampling distribution's mean is our summary with the *population least squares fit* $\tilde\mu$ plugged in for $\mu$.
$$
\E\qty[\widehat{\text{ATT}}] = \widetilde{\text{ATT}} := \ingreen{p(1,x)} \qty{ \ingreen{\tilde \mu(1,x)} - \inred{\tilde \mu(0,x)}}
$$

2. Think about what we know about *population least squares residuals*.

$$
\begin{aligned}
\frac{1}{n}\sum_{i=1}^n \qty{Y_i - \hat\mu(W_i,X_i)} m(W_i,X_i) &= 0 \qqtext{ for every } m \in \model && \text{\textcolor{gray}{(sample least squares residuals)}} \\
\E\qty[ \qty{Y_i - \tilde\mu(W_i,X_i)} m(W_i,X_i)] &= 0 \qqtext{ for every } m \in \model  && \text{\textcolor{gray}{(population least squares residuals)}} \\
\end{aligned} 
$$


## Weighted Least Squares



- This sort of feels like an 'a stopped clock is right twice a day' thing.
- We chose a 'bad model' for $\mu$ but it worked because it was a good model for $\alpha$.
- But we can make this more systematic by using **[weight]{style="color:blue"}ed least squares**.

$$
\begin{aligned}
\hat \mu = \argmin_{m \in \model} \frac{1}{n}\sum_{i=1}^n \inblue{\gamma(W_i,X_i)} \qty{Y_i - m(W_i,X_i)}^2 \quad \text{ satisfies } \quad &\frac{1}{n}\sum_{i=1}^n \inblue{\gamma(W_i,X_i)} \qty{Y_i - \hat\mu(W_i,X_i)} m(W_i,X_i)  = 0 && \qqtext{ for every } m \in \model  \\
\tilde\mu = \argmin_{m \in \model} \E\qty[ \inblue{\gamma(W_i,X_i)} \qty{Y_i - m(W_i,X_i)}^2 ] \quad \text{ satisfies } \quad &\E\qty[ \inblue{\gamma(W_i,X_i)} \qty{Y_i - \tilde\mu(W_i,X_i)} m(W_i,X_i)]  = 0 && \qqtext{ for every } m \in \model 
\end{aligned}
$$

- Our weights should reflect our summary.
- e.g. if we take $\gamma(w,x) = \abs{\alpha(w,x)}$, we're good if our model contains $m(w,x) = \text{sign}(\alpha(w,x))$.
- For treatment effects, that means --- if we choose the right weights --- we just need to have indicators for treatment.

## Inverse Probability Weighting

```{r} 
#| out-height: 500px
mht.plot 
```

- This just sort of worked out for us 'by chance' with our block-randomized experiment.
- But let's think about how to get the right weights in general. What are we doing?
- We're averaging the the red and green curves over the distribution of the green points.

$$
$$

:::: columns
::: {.column width="33%"}

$$
\begin{aligned}
\text{ATT}
&= \sum_{x} \ingreen{p(1,x)} \qty{ \ingreen{\mu(1,x)} - \inred{\mu(0,x) }} \\
&= \underset{\text{\ingreen{matched} \textcolor{gray}{term}}}{\sum_{x} \ingreen{p(1,x) \tilde\mu(1,x)}} - \underset{\text{\ingreen{misma}\inred{tched} \textcolor{gray}{term}}}{\sum_{x} \ingreen{p(1,x)} \inred{\tilde\mu(0,x)}} \\
\end{aligned}
$$

:::
::: {.column width="33%"}

$$
\begin{aligned}
\widetilde{\text{ATT}}
&= \sum_{x} \ingreen{p(1,x)} \qty{ \ingreen{\tilde\mu(1,x)} - \inred{\tilde\mu(0,x) }} \\
&= \underset{\text{\ingreen{matched} \textcolor{gray}{term}}}{\sum_{x} \ingreen{p(1,x) \tilde\mu(1,x)}} - \underset{\text{\ingreen{misma}\inred{tched} \textcolor{gray}{term}}}{\sum_{x} \ingreen{p(1,x)} \inred{\tilde\mu(0,x)}} \\
\end{aligned}
$$

:::
::: {.column width="33%"}

$$
\begin{aligned}
\text{bias}
&= \widetilde{\text{ATT}} - \text{ATT} \\
&= \underset{\text{\ingreen{matched} \textcolor{gray}{term}}}{\faint{\sum_x \quad ? \quad}} - \underset{\text{\ingreen{misma}\inred{tched} \textcolor{gray}{term}}}{\faint{\sum_x \quad ? \quad }}
\end{aligned}
$$

:::
::::

## Does the Matched Term Cause Any Trouble?

```{r} 
#| out-height: 250px
#| fig-asp: .25
mht.plot 
```

$$
\text{bias} = \sum_{x} \ingreen{p(1,x) \qty{\tilde\mu(1,x) - \mu(1,x)}}  \faint{ - \sum_{x} p(1,x) \qty{\tilde\mu(0,x)-\mu(0,x)}} 
$$

- If our model has treatment-group indicators, no.

$$
\begin{aligned}
0 &= \E\qty[ \qty{Y_i - \tilde\mu(W_i,X_i)} 1(W_i=1) ] \\
&=\E\qty[ \qty{\mu(W_i,X_i) - \tilde\mu(W_i,X_i)} 1(W_i=1) ] \\
&=\E\qty[ \qty{\mu(1,X_i) - \tilde\mu(1,X_i)} 1(W_i=1) ] \\
&=\E\qty[ \qty{\mu(1,X_i) - \tilde\mu(1,X_i)} \mid W_i=1 ] P(W_i=1) \\
&=\sum_{x} \ingreen{p(1,x) \qty{\mu(1,x) - \hat\mu(1,x)}} \times P(W_i=1)
\end{aligned}
$$

- Intuitively, this is because we're 'fitting the data in the right place'
- We're averaging over the same distribution of points we've fit our curve to.
- So, even if our model is 'bad' most places, its average is still right.


## What About the Mismatched Term? 

```{r} 
#| out-height: 250px
#| fig-asp: .25
mht.plot 
```

$$
\text{bias} = \faint{\sum_{x} \tilde p(1,x) \qty{ \tilde\mu(1,x) - \mu(1,x)} -} \sum_{x} \ingreen{\tilde p(1,x)} \inred{\qty{\tilde\mu(0,x) - \mu(0,x)}} 
$$

- Even when our model has treatment-group indicators, it's still trouble. Let's try what we just did.

$$
\begin{aligned}
0 &= \E\qty[ \qty{Y_i - \tilde\mu(W_i,X_i)} 1(W_i=0) ] \\
&=\E\qty[ \qty{\mu(W_i,X_i) - \tilde\mu(W_i,X_i)} 1(W_i=0) ] \\
&=\E\qty[ \qty{\mu(0,X_i) - \tilde\mu(0,X_i)} 1(W_i=1) ] \\
&=\E\qty[ \qty{\mu(0,X_i) - \tilde\mu(0,X_i)} \mid W_i=0 ] P(W_i=0) \\
&= \sum_{x} \inred{p(0,x) \qty{\mu(0,x) - \tilde\mu(0,x)}} \times P(W_i=0)
\end{aligned}
$$

- This is not the right average. It's the one that occurs in the 'unadjusted gap', not the ATT.
- We need to adjust for the fact that we're averaging over a different distribution of points.


## The Second: A Weighted Version
```{r} 
#| out-height: 250px
#| fig-asp: .25
mht.plot 
```

$$
\text{bias} = \faint{\sum_{x} \tilde p(1,x) \qty{ \tilde\mu(1,x) - \mu(1,x)} -} \sum_{x} \ingreen{\tilde p(1,x)} \inred{\qty{\tilde\mu(0,x) - \mu(0,x)}} 
$$

- Recall what we know about weighted least squares and its residuals.
$$
\tilde \mu = \argmin_{m \in \model} \E\qty[ \gamma(W_i,X_i) \qty{Y_i - m(W_i,X_i)}^2 ] \quad \text{ satisfies } \quad \E\qty[ \gamma(W_i,X_i) \qty{Y_i - \tilde\mu(W_i,X_i)} m(W_i,X_i)]  = 0 \qqtext{ for every } m \in \model
$$

- Let's try 'upweighting' the red points on the right side by the ratio of the number of green points to red ones.

$$
\begin{aligned}
0 &= \E\qty[ \gamma(X_i) \qty{Y_i - \tilde\mu(W_i,X_i)} 1(W_i=0) ] \qfor \gamma(x)= \frac{\ingreen{p(1,x)}}{\inred{p(0,x)}} \\
&=\E\qty[ \gamma(X_i)\qty{\mu(W_i,X_i) - \tilde\mu(W_i,X_i)} 1(W_i=0) ] \\
&=\E\qty[ \gamma(X_i)\qty{\mu(0,X_i) - \tilde\mu(0,X_i)} 1(W_i=1) ] \\
&=\E\qty[ \gamma(X_i)\qty{\mu(,X_i) - \tilde\mu(0,X_i)} \mid W_i=0 ] \faint{P(W_i=0)} \\
&= \sum_{x} \inred{p(0,x)} \gamma(0,x) \inred{\qty{\mu(0,x) - \tilde\mu(0,x)}} \faint{\times P(W_i=0)} \\
&= \sum_{x} \inred{p(0,x)} \frac{\ingreen{p(1,x)}}{\inred{p(0,x)}}  \inred{\qty{\mu(0,x) - \tilde\mu(0,x)}} \faint{\times P(W_i=0)}\\
&= \sum_{x} \ingreen{p(1,x)} \inred{\qty{\mu(0,x) - \tilde\mu(0,x)}} \faint{\times P(W_i=0)}
\end{aligned}
$$

- This works. Intuitively, using these weights means that we're 'fitting the data in the right place' again.

## Summary

```{r} 
#| out-height: 500px
mht.plot.ipw = ggplot(data) + 
  geom_point(aes(x=plotX(W,X), y=Y, color=factor(W)), alpha=.025, position=position_jitter(w=.15, h=0), size=.05) + 
  stat_summary(aes(x=plotX(W,X), y=Y, color=factor(W), size=W + (1-W)*pi(X)/(1-pi(X))), geom="pointrange", alpha=.15, 
                 fun=mean, 
                 fun.min = function(x) mean(x) - sd(x), 
                 fun.max = function(x) mean(x) + sd(x)) +
  geom_bar(aes(x=plotX(W,X), y=15*after_stat(prop), fill=factor(W), color=factor(W)), 
          alpha=.5, linewidth=.2) + 
          xlab('') + ylab('') + 
          guides(color='none', fill='none', size='none') +
          scale_x_continuous(breaks=seq(50,80,by=5)) + scale_size_area(max_size=1)

mht.plot.ipw
```


$$
\begin{aligned}
\widehat{\text{ATT}} 
= \sum_{x} \ingreen{\hat p(1,x)} \qty{ \ingreen{\hat\mu(1,x)} - \inred{\hat\mu(0,x)}} \qfor \hat \mu = \argmin_{m \in \model} \sum_{i=1}^n \gamma(W_i,X_i) \qty{Y_i - m(W_i,X_i)}^2 \qfor \frac{\ingreen{p(1,x)}}{p(w,x)}
\end{aligned}
$$

- This is the 'inverse probability weighted least squares' estimate of the ATT.
- We give equal weight to the green points because that's the distribution we're averaging over.
- But we weight the red points by the ratio of the number of green points to red ones so that ...
  - if we 'count them by weight' ...
  - then they look like they have that distribution too.
  


## Summary

```{r} 
#| out-height: 500px
data$weights = ifelse(data$W, 1, pi(data$X)/(1-pi(data$X)))

ipwfits = list(
B =  lm(Y ~ X, data=data, weights=weights),
D =  lm(Y ~ W, data=data, weights=weights),
E =  lm(Y ~ W + I(X <= 65), data=data, weights=weights),
A =  lm(Y ~ W + X, data=data, weights=weights),
C =  lm(Y ~ W * I(X <= 65), data=data, weights=weights),
F =  lm(Y ~ W * X, data=data, weights=weights))

muhats.ipw = ipwfits |> map(prediction.function)

ATT.estimates.ipw = muhats.ipw |> map(\(muhat) { 
  data.frame(estimate=mean(muhat(1, pop$X[pop$W==1]) - muhat(0, pop$X[pop$W==1])))
}) |> bind_rows(.id='model')


fits = list(
B =  lm(Y ~ X, data=pop),
D =  lm(Y ~ W, data=pop), 
E =  lm(Y ~ W + I(X <= 65), data=pop),
A =  lm(Y ~ W + X, data=pop),
C =  lm(Y ~ W * I(X <= 65), data=pop),
F =  lm(Y ~ W * X, data=pop))
muhats = fits |> map(prediction.function)

ATT.estimates = muhats |> map(\(muhat) { 
  data.frame(estimate=mean(muhat(1, pop$X[pop$W==1]) - muhat(0, pop$X[pop$W==1])))
}) |> bind_rows(.id='model')
ATT.estimates$true = mean(mu(1,pop$X[pop$W==1]) - mu(0,pop$X[pop$W==1]))

att.comparison = left_join(ATT.estimates, ATT.estimates.ipw, by='model') |> 
    rename(unweighted=estimate.x, weighted=estimate.y)

grid = expand_grid(W=c(0,1), X=seq(50,80,by=.01))
predictions.ipw = muhats.ipw |> map(\(muhat) { 
  grid$muhat = muhat(grid$W, grid$X)
  grid
}) |> bind_rows(.id='model') 
predictions = muhats |> map(\(muhat) { 
  grid$muhat = muhat(grid$W, grid$X)
  grid
}) |> bind_rows(.id='model') 


mht.plot.ipw + 
  geom_line(data=predictions.ipw, aes(x=X, y=muhat, color=factor(W)), linewidth=.5) + 
  geom_line(data=predictions, aes(x=X, y=muhat, color=factor(W)), linewidth=.5, linetype='dashed') + 
  facet_wrap(~model, ncol=3)
```

$$
\begin{aligned}
\widehat{\text{ATT}} 
= \sum_{x} \ingreen{\hat p(1,x)} \qty{ \ingreen{\hat\mu(1,x)} - \inred{\hat\mu(0,x)}} \qfor \hat \mu = \argmin_{m \in \model} \sum_{i=1}^n \gamma(W_i,X_i) \qty{Y_i - m(W_i,X_i)}^2 \qfor \frac{\ingreen{p(1,x)}}{p(w,x)}
\end{aligned}
$$



:::: columns
::: {.column width="30%"}
```{r}
o = order(att.comparison$model)
att.comparison[o , c('model', 'true', 'unweighted', 'weighted')] |> kable(digits=2)
```

:::
::: {.column width="70%"}
- This is the 'inverse probability weighted least squares' estimate of the ATT.
- We give equal weight to the green points because that's the distribution we're averaging over.
- But we weight the red points by the ratio of the number of green points to red ones so they look like they have that distribution too.

:::
::::

## This Works For Any Distribution of Assignments

```{r}

set.seed(1) 
m = 1e6 
pi = function(x) { .1 + .8*(x-50)/30 } 
mu = function(w, x) { 2 - (x-50)/30 + w*((x-50)/30)^2/2 }  

X = 50+round(30*runif(m))
W = 0 + (runif(m) <= pi(X))
epsilon0 = rnorm(m)
epsilon1 = rnorm(m)
Y0 = mu(0,X) + .15*epsilon0
Y1 = mu(1,X) + .15*epsilon1

pop = data.frame(X=X, W=W, Y=Y0*(1-W) + Y1*W, Y0=Y0, Y1=Y1)

n = 10000
data = pop[sample(nrow(pop), n, replace=TRUE),]

prediction.function = function(model) { 
  function(w, x) { 
    predict(model, newdata=data.frame(W=w, X=x))
  }
}


mht.plot.ipw = ggplot(data) + 
  geom_point(aes(x=plotX(W,X), y=Y, color=factor(W)), alpha=.025, position=position_jitter(w=.15, h=0), size=.05) + 
  stat_summary(aes(x=plotX(W,X), y=Y, color=factor(W), size=W + (1-W)*pi(X)/(1-pi(X))), geom="pointrange", alpha=.15, 
                 fun=mean, 
                 fun.min = function(x) mean(x) - sd(x), 
                 fun.max = function(x) mean(x) + sd(x)) +
  geom_bar(aes(x=plotX(W,X), y=15*after_stat(prop), fill=factor(W), color=factor(W)), 
          alpha=.5, linewidth=.2) + 
          xlab('') + ylab('') + 
          guides(color='none', fill='none', size='none') +
          scale_x_continuous(breaks=seq(50,80,by=5)) + scale_size_area(max_size=1)

data$weights = ifelse(data$W, 1, pi(data$X)/(1-pi(data$X)))
ipwfits = list(
B =  lm(Y ~ X, data=data, weights=weights),
D =  lm(Y ~ W, data=data, weights=weights),
E =  lm(Y ~ W + I(X <= 65), data=data, weights=weights),
A =  lm(Y ~ W + X, data=data, weights=weights),
C =  lm(Y ~ W * I(X <= 65), data=data, weights=weights),
F =  lm(Y ~ W * X, data=data, weights=weights))

ATT.estimates.ipw = muhats.ipw |> map(\(muhat) { 
  data.frame(estimate=mean(muhat(1, pop$X[pop$W==1]) - muhat(0, pop$X[pop$W==1])))
}) |> bind_rows(.id='model')


fits = list(
B =  lm(Y ~ X, data=pop),
D =  lm(Y ~ W, data=pop), 
E =  lm(Y ~ W + I(X <= 65), data=pop),
A =  lm(Y ~ W + X, data=pop),
C =  lm(Y ~ W * I(X <= 65), data=pop),
F =  lm(Y ~ W * X, data=pop))
muhats = fits |> map(prediction.function)

ATT.estimates = muhats |> map(\(muhat) { 
  data.frame(estimate=mean(muhat(1, pop$X[pop$W==1]) - muhat(0, pop$X[pop$W==1])))
}) |> bind_rows(.id='model')
ATT.estimates$true = mean(mu(1,pop$X[pop$W==1]) - mu(0,pop$X[pop$W==1]))

att.comparison = left_join(ATT.estimates, ATT.estimates.ipw, by='model') |> 
    rename(unweighted=estimate.x, weighted=estimate.y)

grid = expand_grid(W=c(0,1), X=seq(50,80,by=.01))
predictions.ipw = muhats.ipw |> map(\(muhat) { 
  grid$muhat = muhat(grid$W, grid$X)
  grid
}) |> bind_rows(.id='model') 
predictions = muhats |> map(\(muhat) { 
  grid$muhat = muhat(grid$W, grid$X)
  grid
}) |> bind_rows(.id='model') 


mht.plot.ipw + 
  geom_line(data=predictions.ipw, aes(x=X, y=muhat, color=factor(W)), linewidth=.5) + 
  geom_line(data=predictions, aes(x=X, y=muhat, color=factor(W)), linewidth=.5, linetype='dashed') + 
  facet_wrap(~model, ncol=3)
```


```{r}
att.comparison[o, c('model', 'true', 'unweighted', 'weighted')] |> kable(digits=2)
```

