---
title: "Lecture 2"
subtitle: "Population Inference for a Proportion"
format:
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---

```{r setup}
library(usmap)
library(sf)

library(ggplot2)

library(scales)
library(knitr)

percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }
```

```{r ggplot-themes}
class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
		    panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
		    axis.ticks.x = element_blank(),
		    axis.ticks.y = element_blank(),
		    axis.text.x  = element_text(colour = "#aaaaaa"),
		    axis.text.y  = element_text(colour = "#aaaaaa"),
		    axis.title.x  = element_text(colour = "#aaaaaa"),
		    axis.title.y  = element_text(colour = "#aaaaaa", angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
   	         axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
	         panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(simplified.theme)
```

## Review


- Last lecture, we talked about how to summarize a sample. 
- We talked about ...
  - sample means
  - subsample means 
  - differences in subsample means
- We talked about a few ways of thinking about means
  - as the balance point on a seesaw
  - as the point that minimizes the sum of squared residuals 
- Throughout, we were thinking about the sample on its own terms.

## Today

- Today we'll consider a context in which ...
  - we have a random sample from a large population 
  - we have measured a binary outcome
- Our goal will be to summarize the population.
- We'll talk about ...
  - population means
  - subpopulation means 
  - differences in subpopulation means
- As you might expect, this is harder than summarizing the sample.
  - Most of the data we're summarizing is data we don't have.
  - That is, it's about people in the population who aren't in the sample.
    
## Context

::: {.hidden}

```{r population}
our.seed = 0
set.seed(our.seed)
N=7233584
p=4999960/N
y= c(rep(1,p*N), rep(0, (1-p)*N))
y=sample(y) # shuffle the order 
population.proportion = p
```

```{r sample}
set.seed(our.seed)

n = 625
J = sample(N, n, replace=TRUE)
Y = y[J]
sample.proportion = mean(Y)
```

:::

- Suppose that, week before the 2020 presidential election, you did some polling.
  - You use a list of the $N \approx `r approx(N)`$ people registered to vote in Georgia.
  - And you make $n=`r n`$ phone calls.
  - Each call, you select a voter uniformly at random from the list, e.g. by rolling a `r approx(N)` -sided die.
  - And then you ask the potential voter whether they plan to vote.
- Suppose also that all these registered voters will
  - Pick up the phone will called
  - Respond honestly to your questions
  - Not change their minds about voting

## Polling Results

- You put your polling results in a table.
- It has one column for each call.
- In that columns, you record ...
  - the call number $i$---a number from $1 \ldots `r n`$.
  - the response $Y_i$ of the person you called---$1$ for 'yes and $0$ for no'.

$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}  \\
Y_i  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$ 

- You summarize the results with an extra column: the mean of the responses.
  - Remember that the mean of a binary variable is a proportion. 
  - You found that `r percent(mean(Y))` of the people polled said they would vote.

## Sample Versus Population

- We want to estimate the proportion of *all registered voters* --- our population --- who will vote.
- To do this, we use the proportion of *polled voters* --- our sample --- who said they would.
- When the election occurs, we get to see who turns out to vote.
  - `r approx(p*N)` people, or roughly `r percent(mean(Y))` of registered voters, actually vote.
 
\ \

:::: columns
::: column
### Before the Election 

$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}  \\
Y_i  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$

Outcomes for our sample of polled voters. \
To enumerate our sample, we ...

- give each call a number $i \in 1 \ldots `r n`$.
- write $Y_i$ for the turnout of the $i$th person we called.
:::
::: column
### After the Election

$$
\begin{array}{r|rrrrrr|r}
j & 1 & 2 & 3 & 4 & \dots &   `r approx(N)` & \bar{y}_{`r approx(N)`}  \\
y_{j} & `r y[1]` & `r y[2]` & `r y[3]` & `r y[4]` & \dots & `r y[N]`   & `r approx(mean(y))`  \\
\end{array}
$$
Outcomes for the population of registered voters. \
To enumerate our population, we ... 

- give each registered voter a # $j \in 1 \ldots `r approx(N)`$.
- write $y_j$ for the turnout of the person with ID $j$.
:::
::::

## Two Questions

Our sample proportion $`r sum(Y)` / `r n` \approx `r approx(mean(Y))`$ 
is close to the population proportion $`r approx(sum(y))` / `r approx(N)` \approx `r approx(mean(y))`$

1. Did we get lucky?
2. Could we have predicted how close we'd get before the election happened?


::: fragment

- Answers: No and Yes.
- To get these answers, we'll think about what'd happen if our friends had run identical polls.
  - Each friend will choose have a different random sample.
  - And estimate the population proportion by a different sample proportion.
- We'll see how close these estimates tend to be.

:::

## Review: Connecting Sample and Population

- For each call $i$, we randomly select a voter with an id we'll call $J_i$.
- And we record as that call's outcome the turnout of that voter: $Y_i=y_{J_i}$.
- On each call, each registered voter has a $1/`r approx(N)`$ chance of being called.
  - This is called sampling with replacement because we could call the same person twice.
  - In our poll, this is unlikely because we're making a small number of calls relative to population size.
  



```{r mapping}
# Create fake locations for the people in our population
# by taking random convex combinations of state vertices
set.seed(0)
ga.shape = us_map(include = c("GA"))

# To keep all this from being overwhelming, we'll have 10000 fake locations
# The jth person is at the (j %% M + 1)st location
M = 10000 
V = nrow(ga.shape)
weights = abs(rt(M*V,1))
dim(weights) = c(M,V)
weights = (1/rowSums(weights)) * weights
locations = data.frame(x=weights %*% ga.shape$x, y=weights %*% ga.shape$y)

ga = ggplot() + geom_polygon(data=ga.shape, aes(x = x, y = y), fill="white", color="black") + coord_fixed() + blank.theme
```
`

:::: {.r-stack}
::: {.fragment .fade-out data-fragment-index="0"}

```{r}
#| out-height: 300px
sample.locations = locations[J %% M + 1, ]
sample.locations$label = J
ga + geom_text(aes(x = x, y = y, label=label), alpha=.5, data=sample.locations[c(1,2,n),]) 
```

:::
::: {.fragment .current-visible data-fragment-index="0"}

```{r}
#| out-height: 300px
ga + geom_text(aes(x = x, y = y, label=label), alpha=.2, data=sample.locations) 
```
:::
::: {.fragment .current-visible data-fragment-index="1"}
```{r}
#| out-height: 300px
ga + geom_point(aes(x = x, y = y), alpha=.2, data=sample.locations) 
```
:::
::::

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{select/record}  & J_1 & Y_1 & J_2 & Y_2 & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\text{outcome}    & `r J[1]` & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`} 
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`} 
                  & `r approx(mean(Y))` \\
\end{array}
$$ 

:::: aside
Obviously these locations are fake. Most people in Georgia live near Atlanta, not where these dots are. This is just to illustrate the sampling scheme. 
:::

## Our Poll

```{r}
# Now we'll repeat our sampling so we have polls for a few friends
set.seed(our.seed)
J1 = sample(N, n, replace=TRUE)
J2 = sample(N, n, replace=TRUE)
J3 = sample(N, n, replace=TRUE)
JM = sample(N, n, replace=TRUE)
```

```{r}
J = J1
Y = y[J1]

sample.locations = locations[J %% M +1, ]
sample.locations$label = J
ga + geom_text(aes(x = x, y = y, label=label), alpha=.2, data=sample.locations) 
```


$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{select/record}  & J_1 & Y_1 & J_2 & Y_2 & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\text{outcome}    & `r J[1]` & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`} 
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`} 
                  & `r approx(mean(Y))` \\
\end{array}
$$

## Our First Friend's Poll

```{r}
J = J2
Y = y[J2]

sample.locations = locations[J %% M + 1, ]
sample.locations$label = J
ga + geom_text(aes(x = x, y = y, label=label), color='magenta', alpha=.2, data=sample.locations)
```

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{select/record}  & J_1 & Y_1 & J_2 & Y_2 & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\text{outcome}    & `r J[1]` & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`} 
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`} 
                  & `r approx(mean(Y))` \\
\end{array}
$$

## Our Second Friend's Poll

```{r}
J = J3
Y = y[J3]

sample.locations = locations[J %% M + 1, ]
sample.locations$label = J
ga + geom_text(aes(x = x, y = y, label=label), color='cyan', alpha=.2, data=sample.locations)
```

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{select/record}  & J_1 & Y_1 & J_2 & Y_2 & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\text{outcome}    & `r J[1]` & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`} 
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`} 
                  & `r approx(mean(Y))` \\
\end{array}
$$

## The Sampling Distribution of our Estimate

\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{pollster} & J_1 & Y_1 & J_2 & Y_2   & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\hline
\color{black}   1   
    & \color{black}   `r J=J1; Y=y[J]; J[1]`    & \color{black}  `r Y[1]` 
    & \color{black}   `r J[2]`   & \color{black}   `r Y[2]`           & \color{black}   \dots 
    & \color{black}   `r J[n]`   & \color{black}   `r Y[n]`           & \color{black}   `r approx(mean(Y))` \\
\color{magenta} 2   
    & \color{magenta} `r J=J2; Y=y[J]; J[1]`    & \color{magenta}  `r Y[1]` 
    & \color{magenta} `r J[2]`   & \color{magenta}  `r Y[2]`          & \color{magenta} \dots 
    & \color{magenta} `r J[n]`   & \color{magenta} `r Y[n]`           & \color{magenta} `r approx(mean(Y))` \\
\color{cyan}    3   
    & \color{cyan}    `r J=J3; Y=y[J]; J[1]`    & \color{cyan}     `r Y[1]` 
    & \color{cyan}    `r J[2]`   & \color{cyan}    `r Y[2]`           & \color{cyan}    \dots 
    & \color{cyan}    `r J[n]`   & \color{cyan}    `r Y[n]`           & \color{cyan}    `r approx(mean(Y))` \\
{\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots} &  {\vdots} \\ 
\color{gray}    1M   
    & \color{gray}    `r J=JM; Y=y[J]; J[1]`    & \color{gray}     `r Y[1]` 
    & \color{gray}    `r J[2]`   & \color{gray}    `r Y[2]`           & \color{gray}    \dots 
    & \color{gray}    `r J[n]`   & \color{gray}    `r Y[n]`           & \color{gray}    `r approx(mean(Y))` \\
 {\vdots} & {\vdots} & {\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots}  & {\vdots} \\ 
\hline
{f_{\cdot}(\cdot)} & \hphantom{Ber(`r approx(p)`)} & & \hphantom{Ber(`r approx(p)`)} &  & & \hphantom{Ber(`r approx(p)`)} & \hphantom{\frac{Bin(`r n`,`r approx(p)`)}{`r n`}} \\
\end{array}

- What is the distribution of $Y_1$?
- What is the distribution of $\sum_{i}^{`r n`}Y_i$ \ ?

## The Sampling Distribution of our Estimate

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{pollster} & J_1 & Y_1 & J_2 & Y_2   & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\hline
\color{black}   1   
    & \color{black}   `r J=J1; Y=y[J]; J[1]`    & \color{black}    `r Y[1]` 
    & \color{black}   `r J[2]`   & \color{black}   `r Y[2]`           & \color{black}   \dots 
    & \color{black}   `r J[n]`   & \color{black}   `r Y[n]`           & \color{black}   `r approx(mean(Y))` \\
\color{magenta} 2   
    & \color{magenta} `r J=J2; Y=y[J]; J[1]`    & \color{magenta}  `r Y[1]` 
    & \color{magenta} `r J[2]`   & \color{magenta}  `r Y[2]`          & \color{magenta} \dots 
    & \color{magenta} `r J[n]`   & \color{magenta} `r Y[n]`           & \color{magenta} `r approx(mean(Y))` \\
\color{cyan}    3   
    & \color{cyan}    `r J=J3; Y=y[J]; J[1]`    & \color{cyan}     `r Y[1]` 
    & \color{cyan}    `r J[2]`   & \color{cyan}    `r Y[2]`           & \color{cyan}    \dots 
    & \color{cyan}    `r J[n]`   & \color{cyan}    `r Y[n]`           & \color{cyan}    `r approx(mean(Y))` \\
{\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots} &  {\vdots} \\ 
\color{gray}    1M   
    & \color{gray}    `r J=JM; Y=y[J]; J[1]`    & \color{gray}     `r Y[1]` 
    & \color{gray}    `r J[2]`   & \color{gray}    `r Y[2]`           & \color{gray}    \dots 
    & \color{gray}    `r J[n]`   & \color{gray}    `r Y[n]`           & \color{gray}    `r approx(mean(Y))` \\
 {\vdots} & {\vdots} & {\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots}  & {\vdots} \\ 
\hline
{f_{\cdot}(\cdot)} & Ber(`r approx(p)`) & & Ber(`r approx(p)`) &  & & Ber(`r approx(p)`) & \frac{Bin(`r n`,`r approx(p)`)}{`r n`} \\
\end{array}
$$

- $Y_1$ has a Bernoulli distribution 
- $\sum_{i}^{`r n`}Y_i$, a sum of *independent and identically distributed* Bernoullis, has a Binomial distribution.

## Review of Bernoulli and Binomial Distributions

- Our individual responses $Y_1...Y_{`r n`}$ are Bernoulli Random Variables 
$$
\begin{aligned}
f_{Y_1}(y) &= (`r approx(p)`)^{y} (`r approx(1-p)`)^{1-y} \qfor  y=0,1\\
\vdots \quad  &=   \quad\quad\quad\quad\quad\quad \vdots  \\
f_{Y_{`r n`}}(y) &= (`r approx(p)`)^{y} (`r approx(1-p)`)^{1-y} \qfor  y=0,1
\end{aligned}
$$

- Our sum of responses, $S_{`r n`} = \sum_{i=1}^{`r n`} Y_i$, is a Binomial Random Variable
$$
\begin{aligned}
f_{S_{`r n`}}(s) = {`r n` \choose s} 
(`r approx(1-p)`)^{s} 
(`r approx(1-p)`)^{`r n`-s} 
\qfor  s=0 \ldots `r n`
\end{aligned}
$$

- Our sample proportion, $\overline{Y}_{`r n`}$, has the distribution of this Binomial Random Variable divided by $`r n`$.

## The Sampling Distribution in **R**

Resample the number of voters 1,000,000 times and store these 1,000,000 numbers in a vector.

```{r}
#| echo: true
set.seed(our.seed)
sumY.samples <- rbinom(1000000, size=n, prob=mean(y))
```

Divide all of these numbers by the sample size

```{r}
#| echo: true
meanY.samples <- sumY.samples/n
```

Histogram the result
```{r}
#| echo: true
ggplot() + geom_histogram(aes(x=meanY.samples, y=after_stat(density)), binwidth=.001) 
```

## Annotating the Sampling Distribution Histogram 

```{r}
#| echo: true
#| out-height: 400px

ggplot() + geom_histogram(aes(x=meanY.samples, y=after_stat(density)), binwidth=.001, alpha=.2) +
           geom_vline(aes(xintercept=mean(meanY.samples)), color="blue") +
           geom_vline(aes(xintercept=quantile(meanY.samples, probs=c(.16, .84))), color="blue", linetype='dashed') + 
           geom_vline(aes(xintercept=quantile(meanY.samples, probs=c(.025, .975))), color="blue", linetype='dotted') +
           geom_vline(aes(xintercept=mean(y)), color='green', size=3, alpha=.7) +
           scale_x_continuous(breaks=seq(.6,.8, by=.025), limits=c(.6, .8)) +
           scale_y_continuous(breaks=seq(0,40,  by=10), limits=c(0, 40))
```

- There are a few features we may want to highlight.
  - The *mean of the sampling distribution* is the solid blue line.
  - The middle 2/3 of the sampling distribution lies between the dashed blue lines.
  - The middle 95\% of the sampling distribution lies between the dotted blue lines. 
- I've also drawn the *population proportion* as a wide green line.
  - We see that it appears as a kind of highlight on the solid blue line. 
  - It's in exactly the same place. What does this mean? 


##  Revisiting our First Question

```{r}
#| out-height: 500px
Y=y[J1] 

dist.68   = max(abs(quantile(meanY.samples, probs=c(.16,   .84))-mean(y)))
dist.95   = max(abs(quantile(meanY.samples, probs=c(.025, .975))-mean(y)))
p.as.close    = mean(abs(meanY.samples - mean(y)) < abs(mean(Y)-mean(y))) 

annotated.histogram = ggplot() + geom_histogram(aes(x=meanY.samples, y=after_stat(density)), binwidth=.001, alpha=.2) +
           geom_vline(aes(xintercept=mean(meanY.samples)), color="blue") +
           geom_vline(aes(xintercept=quantile(meanY.samples, probs=c(.16, .84))), color="blue", linetype='dashed') + 
           geom_vline(aes(xintercept=quantile(meanY.samples, probs=c(.025, .975))), color="blue", linetype='dotted') +
           geom_vline(aes(xintercept=mean(y)), color='green', size=3, alpha=.7) +
           scale_x_continuous(breaks=seq(.6,.8, by=.025), limits=c(.6, .8)) + 
           scale_y_continuous(breaks=seq(0,40,  by=10), limits=c(0, 40))
annotated.histogram + geom_point(aes(x=mean(Y), y=0))
```


**Observation**. Our sample proportion --- the black dot --- is close to our population proportion.


**Q**. Did we get lucky? 

::: fragment
No, not really.

  - In 68% of polls, the sample proportion is within `r round(dist.68,3)` of the population proportion.
  - In 95% of polls, the sample proportions are within  `r round(dist.95,3)` of the population proportion.
  - In `r percent(p.as.close)` of polls, we come as close as our estimate.
  
:::

##  Revisiting our Second Question

```{r}
#| out-height: 500px
annotated.histogram + geom_point(aes(x=mean(Y), y=0))
```

**Q**. Could we have predicted how close we'd get before the election happened?

::: fragment
Yes, in sense.

  - We will use an interval estimate---a *range of values* the population proportion is likely to be in.
  - The *width* of this interval speaks to the 'how close' question. 
  - The *coverage probability* --- the probability we're right --- qualifies this answer.
:::

## Interval Estimation

```{r}  
#| out-height: 500px
width = .02
p.blank = ggplot() + geom_vline(aes(xintercept=population.proportion), color='green', size=1) +
                     scale_x_continuous(breaks=seq(.6,.8, by=.025), limits=c(.6,.8)) +
                     scale_y_continuous(breaks=seq(0,40,  by=10), limits=c(0, 40))
p.nofriends = p.blank + geom_pointrange(aes(y=0, x=mean(Y), xmin=mean(Y)-width/2, xmax=mean(Y)+width/2))
p.nofriends
```

- Our *point estimate* of the population proportion is the sample proportion $\overline{Y}_{`r n`}$.
- So let's try an interval of width .02 centered on it: $\ \ \overline{Y}_{`r n`} \pm .01$.
- Ours covers --- includes --- the population proportion.
- Is that just chance? Let's check out our friends' intervals.

## Interval Estimation: Our Friends' Polls

```{r}
#| out-height: 500px
Y2 = y[J2]
Y3 = y[J3]
p.with2friends = p.nofriends + 
                  geom_pointrange(aes(y=15, x=mean(Y2), xmin=mean(Y2)-width/2, xmax=mean(Y2)+width/2), color='magenta') +
                  geom_pointrange(aes(y=30, x=mean(Y3), xmin=mean(Y3)-width/2, xmax=mean(Y3)+width/2), color='cyan') 
p.with2friends
```

- Lets's look at the interval estimates our first two friends.
  - i.e. estimates based on the magenta and cyan rows in our sampling distribution table.
-  One of these interval covers the population proportion.
- To get a better sense of how often this happens, let's do this for a hundred more polls.

## Interval Estimation: 100 Polls

```{r}
#| out-height: 500px
K=100
p.with2friends + geom_pointrange(aes(y=30*(1:K)/K, x=meanY.samples[1:K], xmin=meanY.samples[1:K]-width/2, xmax=meanY.samples[1:K]+width/2), 
                                 alpha=.1, color='purple') 
```

- $`r sum(meanY.samples[1:K]-width/2 <= p & p <= meanY.samples[1:K]+width/2)`/100$ cover the population proportion.
- This gives us a sense of the *probability* that our interval covers the population proportion.
- If we want to be more precise, we could do the same for millions.
- Let's not. Instead, let's work out a trick for counting intervals that cover.


##  Coverage Probability and Sampling Distributions
```{r}
#| out-height: 500px
p.blank + geom_pointrange(aes(y=30*(1:K)/K, x=meanY.samples[1:K], xmin=meanY.samples[1:K]-width/2, xmax=meanY.samples[1:K]+width/2), 
                                 alpha=.1, color='purple') + 
                 geom_histogram(aes(x=meanY.samples, y=after_stat(density)), binwidth=.001, alpha=.2) 
```

 **Activity**. Calculate this *coverage probability* using the *sampling distribution* of our point estimate $\bar Y_{`r n`}$.\
 \ \
 
 Here's a hint.

  - Suppose we want to use a diagram like this to count the number of intervals that cover.
  - But we don't want to draw in the horizontal segments for each poll.
  - We just want to draw a dot representing $\bar Y_{`r n`}$.
  - Can we draw in a few lines that'll help us count?


## Using Our Hint
```{r}
#| out-height: 500px

p.blank + geom_pointrange(aes(y=30*(1:K)/K, x=meanY.samples[1:K], xmin=meanY.samples[1:K]-width/2, xmax=meanY.samples[1:K]+width/2), 
                                 alpha=.1, color='purple') + 
                 geom_histogram(aes(x=meanY.samples, y=after_stat(density)), binwidth=.001, alpha=.2) + 
                 geom_vline(aes(xintercept=population.proportion+c(-.01,.01)), color="red", linetype='dotted', size=1) 
```

- Let's draw an interval of width .02 centered on the population proportion---as a pair of dotted lines. 
  - This gives it 'arms' the same length as our interval estimates have.
  - And its arms touch a point estimate *if and only if* the point estimates' arms touch it.
- That means we can count the intervals that cover by counting the point estimates between the dotted lines.
- What, in terms of the sampling distribution of the point estimate, is the coverage probability?

## The Coverage Probability
```{r}
#| out-height: 500px
K=100
p.blank + geom_point(aes(y=30*(1:K)/K, x=meanY.samples[1:K]), alpha=.2, color='purple') + 
                 geom_histogram(aes(x=meanY.samples, y=after_stat(density)), binwidth=.001, alpha=.2) + 
                 geom_vline(aes(xintercept=population.proportion+c(-.01,.01)), color="red", linetype='dotted', size=1) 
```

It's the probability that a random draw from the sampling distribution lies between the dotted lines.
$$
\text{coverage probability} = P\qty(\overline{Y}_{`r n`} \in \overline{y}_{`r approx(N)`} \pm .01) 
$$
And it's about 40\%.
```{r}
#| echo: true
mean(mean(y) - .01 <= meanY.samples & meanY.samples <= mean(y) + .01)
```

## Ad-hoc Interval Estimates
```{r}
#| out-height: 500px
K=100
p.blank + geom_point(aes(y=30*(1:K)/K, x=meanY.samples[1:K]), alpha=.2, color='purple') + 
                 geom_histogram(aes(x=meanY.samples, y=after_stat(density)), binwidth=.001, alpha=.2) + 
                 geom_vline(aes(xintercept=population.proportion+c(-.01,.01)), color="red", linetype='dotted', size=1) 
```

What we just did was choose a width and calculate a coverage probability.

  - The coverage probability we found --- 40% --- probably wasn't what we wanted.
  - Think of how you'd advertise your polling services.
  - 'I'm right about half the time. Actually, a bit less than that'.
  - 95% sounds a lot better. For that, we'll have to use a wider interval.

## Calibrated Interval Estimates
```{r}
#| out-height: 500px
K=100
p.blank + geom_point(aes(y=30*(1:K)/K, x=meanY.samples[1:K]), alpha=.2, color='purple') + 
                 geom_histogram(aes(x=meanY.samples, y=after_stat(density)), binwidth=.001, alpha=.2) + 
                 geom_vline(aes(xintercept=population.proportion+c(-.01,.01)), color="red", linetype='dotted', size=1) 
```

What we want to do is choose a coverage probability and calculate the right width.

  - An interval estimate calibrated like this---to have a given coverage---is called a *confidence interval*.
  - Let's think about how to do that. Again, we'll use the sampling distribution of our point estimate.
  - Let's take a look at our annotated histogram of point estimates again.


## Using the Sampling Distribution for Calibration

```{r}
#| out-height: 500px
annotated.histogram + geom_point(aes(y=30*(1:K)/K, x=meanY.samples[1:K]), alpha=.2, color='purple')       
```

Remember what we've drawn in.

  - The *mean of the sampling distribution* is the solid blue line---and is the same as the population proportion.
  - The middle 2/3 of the sampling distribution lies between the dashed blue lines.
  - The middle 95\% of the sampling distribution lies between the dotted blue lines. 

## Using the Sampling Distribution for Calibration

```{r}
#| out-height: 500px
annotated.histogram + geom_point(aes(y=30*(1:K)/K, x=meanY.samples[1:K]), alpha=.2, color='purple')       
```
**Activity**. 

  - Suppose you and your friends want to draw in intervals around your point estimates.
  - How wide should they be if you want 95\% of them to cover the population proportion?
  - Think about 'arms'. You may assume the sampling distribution is symmetric.

## Using the Sampling Distribution for Calibration

```{r}
#| out-height: 500px

width = quantile(meanY.samples, probs=c(.975)) - quantile(meanY.samples, probs=c(.025))
coverage = mean(population.proportion - width/2 <= meanY.samples & meanY.samples <= population.proportion + width/2)
annotated.histogram + geom_pointrange(aes(y=30*(1:K)/K, x=meanY.samples[1:K], xmin=meanY.samples[1:K]-width/2, xmax=meanY.samples[1:K]+width/2), alpha=.2, color='purple') 
```

**Answer**.

- The width of a 95% interval should be the width between the blue lines.
- That's the width of the 'arms' containing 95% of the estimates drawn from the sampling distribution.
- These intervals have a coverage probability of `r percent(coverage)`. 

```{r}
#| echo: true
width = quantile(meanY.samples, probs=c(.975)) - quantile(meanY.samples, probs=c(.025))
coverage = mean(mean(y)- width/2 <= meanY.samples & meanY.samples <= mean(y) + width/2)
```

## A Problem 
```{r}
#| out-height: 500px

width = quantile(meanY.samples, probs=c(.975)) - quantile(meanY.samples, probs=c(.025))
coverage = mean(population.proportion - width/2 <= meanY.samples & meanY.samples <= population.proportion + width/2)
annotated.histogram + geom_pointrange(aes(y=30*(1:K)/K, x=meanY.samples[1:K], xmin=meanY.samples[1:K]-width/2, xmax=meanY.samples[1:K]+width/2), alpha=.2, color='purple') 
```

- We can't calibrate intervals like this in real life.
  - When we run our a poll, we get a single point estimate $\bar Y_{`r n`}$ based on our sample.
  - We don't know the sampling distribution of this point estimate until election day.
- But what we actually do is *almost* the same. 
  - We *do* the same thing.
  - But we use an *estimate* of the sampling distribution in place of the thing itself.
- Today we'll talk about an estimate based on *the bootstrap*.

## The Bootstrap Idea

- Before the election, we don't observe the population or population proportion.
  - But we do observe our sample and sample proportion. 
  - And we can *sample from the sample* act as if it were the population.
- We'll take repeated random samples of size `r n`, with replacement, from our sample of size `r n`.
  - We call these *bootstrap samples*. 
  - We use stars to distinguish them from our original sample---we write $Y_i^*$.

```{r bootstrap.sample}
J.star = sample(n, size=n, replace=TRUE)
Y.star = Y[J.star]
```

:::: columns
::: column
The Sample
$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}  \\
Y_i  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$

The Bootstrap Sample

$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}^*  \\
Y_i^*  &  `r Y.star[1]` & `r Y.star[2]` & \dots & `r Y.star[n]` & `r approx(mean(Y.star))` \\
\end{array}
$$
:::

::: column
The Population

$$
\begin{array}{r|rrrr|r}
j & 1 & 2 & \dots &   `r approx(N)` & \bar{y}_{`r approx(N)`}  \\
y_{j} & `r y[1]` & `r y[2]` & \dots & `r y[N]`   & `r approx(mean(y))`  \\
\end{array}
$$

The 'Bootstrap Population' --- The Sample
$$
\begin{array}{r|rrrr|r}
j        &   1 & 2 & \dots & `r n` & \bar{y}^*_{`r n`}  \\
y_j^*  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$

:::
::::

## The Bootstrap Sampling Distribution

- In this case, it's easy to work out the bootstrap sampling distribution.
- Think about what happens we draw our sample from the population.
- The distribution of our sample proportion is determined by two things.
    - The number of draws we were taking, `r n`.
    - The proportion of voters in the population, `r approx(mean(y))`.
- It's Binomial(`r n`, `r approx(mean(y))`)/`r n`.

```{r}
#| echo: true
#| execute: false
meanY.samples = rbinom(1000000, size=n, prob=mean(y)) / n
```  

- What happens when we draw our bootstrap sample from the sample is analogous.
    - The number of draws we were taking, `r n`.
    - The proportion of voters in the sample, `r approx(mean(Y))`
- It's Binomial(`r n`, `r approx(mean(Y))`)/`r n`.

```{r}
#| echo: true
meanY.bootstrap.samples = rbinom(1000000, size=n, prob=mean(Y)) / n
```


## The Bootstrap Sampling Distribution 

```{r}
#| echo: true
#| out-height: 400px

ggplot() + geom_histogram(aes(x=meanY.bootstrap.samples, y=after_stat(density)), binwidth=.001, alpha=.2) +
           geom_vline(aes(xintercept=mean(meanY.bootstrap.samples)), color="blue") +
           geom_vline(aes(xintercept=quantile(meanY.bootstrap.samples, probs=c(.16, .84))), color="blue", linetype='dashed') + 
           geom_vline(aes(xintercept=quantile(meanY.bootstrap.samples, probs=c(.025, .975))), color="blue", linetype='dotted') +
           geom_vline(aes(xintercept=sample.proportion), color='green', size=3, alpha=.7) +
           scale_x_continuous(breaks=seq(.6,.8, by=.025), limits=c(.6, .8)) +
           scale_y_continuous(breaks=seq(0,40,  by=10), limits=c(0, 40))
```

- There are a few features we may want to highlight.
  - The *mean of the bootstrap sampling distribution* is the solid blue line.
  - The middle 2/3 of the bootstrap sampling distribution lies between the dashed blue lines.
  - The middle 95\% of the bootstrap sampling distribution lies between the dotted blue lines. 
- I've also drawn the *sample proportion* as a wide green line.
  - We see that it appears as a kind of highlight on the solid blue line. 
  - It's in exactly the same place. What does this mean? 


## A Comparison

```{r}
#| out-height: 300px
#| layout-nrow: 2
#| fig-asp: .3
#| fig-subcap: 
#|  - "The Bootstrap Sampling Distribution"
#|  - "The Sampling Distribution"
annotated.bootstrap.histogram = ggplot() + geom_histogram(aes(x=meanY.bootstrap.samples, y=after_stat(density)), binwidth=.001, alpha=.2) +
           geom_vline(aes(xintercept=mean(meanY.bootstrap.samples)), color="blue") +
           geom_vline(aes(xintercept=quantile(meanY.bootstrap.samples, probs=c(.16, .84))), color="blue", linetype='dashed') + 
           geom_vline(aes(xintercept=quantile(meanY.bootstrap.samples, probs=c(.025, .975))), color="blue", linetype='dotted') +
           geom_vline(aes(xintercept=mean(Y)), color='green', size=3, alpha=.7) +
           scale_x_continuous(breaks=seq(.6,.8, by=.025), limits=c(.6, .8)) +
           scale_y_continuous(breaks=seq(0,40,  by=10), limits=c(0, 40))
         
annotated.bootstrap.histogram
annotated.histogram
```

- We can see these two distributions are pretty similar. 
- And we're only using the width, so we don't care about differences in the mean.
- To do a visual comparison without worrying about the mean, we can center them.

## A Comparison --- Centered

```{r}
#| out-height: 300px
#| layout-nrow: 2
#| fig-asp: .3
#| fig-subcap: 
#|  - "The Bootstrap Sampling Distribution"
#|  - "The Sampling Distribution"

cmeanY.bootstrap.samples = meanY.bootstrap.samples - mean(meanY.bootstrap.samples) + mean(meanY.samples)
annotated.bootstrap.histogram = ggplot() + geom_histogram(aes(x=cmeanY.bootstrap.samples, y=after_stat(density)), binwidth=.001, alpha=.2) +
           geom_vline(aes(xintercept=mean(cmeanY.bootstrap.samples)), color="blue") +
           geom_vline(aes(xintercept=quantile(cmeanY.bootstrap.samples, probs=c(.16, .84))), color="blue", linetype='dashed') + 
           geom_vline(aes(xintercept=quantile(cmeanY.bootstrap.samples, probs=c(.025, .975))), color="blue", linetype='dotted') +
           geom_vline(aes(xintercept=mean(y)), color='green', size=3, alpha=.7) +
           scale_x_continuous(breaks=seq(.6,.8, by=.025), limits=c(.6, .8)) +
           scale_y_continuous(breaks=seq(0,40,  by=10), limits=c(0, 40))
         
annotated.bootstrap.histogram
annotated.histogram
```

- After centering, they look even more similar. 
- But this is just *one* bootstrap based on one sample.
- Maybe we got lucky? Lets look at some empirical evidence.
- Let's suppose 1000 of our friends did the same thing.

## Empirical Evidence for the Bootstrap

```{r}
K = 100
ps = meanY.samples[1:K]

ggplot() + geom_line(aes(x=x/n-p, y=n*dbinom(x,n,p)), data=expand.grid(p=mean(y), x=0:n)) +
           geom_vline(aes(xintercept=qbinom(c(.025,.975), n,mean(y))/n-mean(y)), linetype='dotted') +
           geom_line(aes(x=x/n-p, y=n*dbinom(x,n,p), color=factor(p)), data=expand.grid(p=ps, x=0:n), alpha=.2) +
           geom_pointrange(aes(x=0, xmin=qbinom(.025,n,p)/n-p, xmax=qbinom(.975,n,p)/n-p, y=20*k/K, color=factor(p)), data=data.frame(p=ps, k=1:K)) +
           xlim(-.1,.1) + guides(color=FALSE)
```

- Our friends' bootstrap sampling distributions are almost indistinguishable from the sampling distribution.
- The resulting interval-widths are near-indistinguishable from the width we got using the sampling distribution.
  - The former are shown as points with arms.
  - That latter is the space between the dotted lines.
- Not luck!  This works reliably.

## It Works. Why Does it Work?

- Why is the bootstrap sampling distribution a good estimate of the sampling distribution?
- Here's an argument specific to the binary-outcome case we've been talking about today.
- We know the sampling distribution and bootstrap sampling distribution.
- They're the distributions of means of `r n` independent Bernoullis---$\frac{Binomial(`r n`,p)}{`r n`}$---with different vote probabilities $p$.
  - In the first case, $p$ is the proportion of voters in the population.
  - In the second case, $p$ is the proportion of voters in the sample.
  - And the difference between these proportions is small in almost all polls.
- The Binomial distribution is *continuous* as a function of $p$---when $p$ changes little, the distribution changes little.
- This means that, in almost all polls, bootstrap and actual sampling distributions are very similar.
- The relevant difference (after centering) is even smaller because *the way* the binomial changes is mostly location.

```{r}
#| out-height: 200px
#| layout-ncol: 2
K = 3
ps = quantile(meanY.samples, c(.025, .5, .975))

ggplot() + geom_line(aes(x=x/n, y=n*dbinom(x,n,p)), data=expand.grid(p=mean(y), x=0:n), alpha=.5) +
           geom_vline(aes(xintercept=qbinom(c(.025,.975), n,mean(y))/n), linetype='dotted') +
           geom_line(aes(x=x/n, y=n*dbinom(x,n,p), color=factor(p)), data=expand.grid(p=ps, x=0:n), alpha=.5) +
           geom_pointrange(aes(x=p, xmin=qbinom(.025,n,p)/n, xmax=qbinom(.975,n,p)/n, y=20*k/K, color=factor(p)), data=data.frame(p=ps, k=1:K)) +
           xlim(.6,.8) + guides(color=FALSE)

ggplot() + geom_line(aes(x=x/n-p, y=n*dbinom(x,n,p)), data=expand.grid(p=mean(y), x=0:n), alpha=.5) +
           geom_vline(aes(xintercept=qbinom(c(.025,.975), n,mean(y))/n-mean(y)), linetype='dotted') +
           geom_line(aes(x=x/n-p, y=n*dbinom(x,n,p), color=factor(p)), data=expand.grid(p=ps, x=0:n), alpha=.5) +
           geom_pointrange(aes(x=0, xmin=qbinom(.025,n,p)/n-p, xmax=qbinom(.975,n,p)/n-p, y=20*k/K, color=factor(p)), data=data.frame(p=ps, k=1:K)) +
           xlim(-.1,.1) + guides(color=FALSE)
```

::: aside
Bootstrap Sampling Distributions with Interval Showing Middle 95% for p at the 2.5th, 50th, and 95th percentiles of the sampling distribution. \
With centering (right) and without (left). 
:::

## Summary

- Our sample proportion `r approx(mean(Y))` is close to the population proportion `r approx(mean(y))`. Did we get lucky? 
  - No! In a million runs, almost all are within .05.
- Could we have predicted how close we would get before seeing the `r approx(mean(y))`? 
  - Yes! We can use a calibrated interval estimate---a Confidence Interval.
