---
title: "Lecture 3"
subtitle: "Calibrating Interval Estimates with Binary Observations"
---


# Sampling and Calibration Review
A Polling Nightmare

## Before the Election

{{< include shared-code.qmd >}}
{{< include ga-turnout.qmd >}}

```{r}
#| out-height: 400px
#| fig-asp: .3 

set.seed(default.seed)
Js = array(dim=c(10000, n))
Ys = array(dim=c(10000, n))
for(rr in 1:10000) {
  Js[rr,] = sample(m, n, replace=TRUE)
  Ys[rr,] = y[Js[rr,]]
}
meanY.samples = rowMeans(Ys)
pmf.height = max(dbinom(0:n, n, mean(y))) 

scale_x = scale_x_continuous(breaks=seq(.6,.8, by=.025)) 
scale_y = scale_y_continuous(breaks=seq(0, pmf.height,by=.01))
coord = coord_cartesian(xlim=c(.6,.8), ylim=c(0, pmf.height))

bar.plot = ggplot() + geom_bar(aes(x=meanY.samples, y=after_stat(prop)), alpha=.3) +
                      geom_vline(aes(xintercept=mean(y)), color='green', linewidth=2, alpha=.7) + 
                      scale_x + scale_y
                      
```

```{r}
#| label: choosing-an-appropriately-mediocre-estimate
JM = Js[504,]
Y = y[JM]
```

```{r}
binom.width = function(p,n,...) { width_cdf(\(x)pbinom(n*x, n, p), p, limits=c(0,1), ...) }

your.widths = c(.02, binom.width(mean(Y), n, alpha=.1), binom.width(mean(Y), n, alpha=.05))
opacities = c(1,.25,.25)
interval.data = data.frame(x=mean(Y), w=your.widths,
                          y=pmf.height*c(.57,.6, .63), alpha=opacities)
your.interval = 
    geom_pointrange(aes(x=x, xmin=x-w/2, xmax=x+w/2, y=y, alpha=I(alpha)), 
                    data=interval.data[1,], color=polld)
your.intervals = 
    geom_pointrange(aes(x=x, xmin=x-w/2, xmax=x+w/2, y=y, alpha=I(alpha)), 
                    data=interval.data, color=polld)

ggplot() + 
    your.interval + 
    scale_x + scale_y + coord
```



:::: columns
::: column
### Your Sample 

$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}  \\
Y_i  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$

:::
::: column
### The Population

$$
\color{lightgray}
\begin{array}{r|rrrrrr|r}
j & 1 & 2 & 3 & 4 & \dots &   `r approx(m)` & \bar{y}_{`r approx(m)`}  \\
y_{j} & `r y[1]` & `r y[2]` & `r y[3]` & `r y[4]` & \dots & `r y[m]`   & `r approx(mean(y))`  \\
\end{array}
$$
:::
::::

- You want to estimate the proportion of *all registered voters* --- the population --- who will vote.
- To do this, you use the proportion of *polled voters* --- your sample --- who said they would.
    - You're probably not going to match the population proportion exactly ... 
    - ... so you report an *interval estimate*---a range of values we claim the population proportion is in.
    - You're know you're not going to be right 100% of the time you make with claims like this ...
    - ... so you state a *nominal coverage probability*---how often you're right about claims like this.^[You say that's how often, anyway. Hence 'nominal'.]
    
 - Without thinking too hard about it, you ...
    - ... report, as your interval, [your sample proportion (`r percent(mean(Y))`) &plusmn; `r percent(your.widths[1]/2)`]{.polld}. Because it sounds good.
    - ... and say your coverage probability is 95%. Because that's what everybody else says.

## After the Election

```{r}
ggplot() + 
    geom_vline(xintercept=mean(y), color='green', size=2, alpha=.7) + 
    your.intervals + 
    scale_x + scale_y + coord
```



:::: columns
::: column
### Your Sample

$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}  \\
Y_i  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$

:::
::: column

### The Population

$$
\begin{array}{r|rrrrrr|r}
j & 1 & 2 & 3 & 4 & \dots &   `r approx(m)` & \bar{y}_{`r approx(m)`}  \\
y_{j} & `r y[1]` & `r y[2]` & `r y[3]` & `r y[4]` & \dots & `r y[m]`   & `r approx(mean(y))`  \\
\end{array}
$$
:::
::::

- When the election occurs, we get to see who turns out to vote.
  - `r approx(sum(y))` people, or roughly `r percent(mean(y))` of registered voters, actually vote.
  - You --- and future employers --- can see how well you did. And how well everybody else did.
- [Your interval]{.polld} missed the target. It doesn't contain [the turnout proportion]{.target}.
  - Your point estimate is only off by `r percent(abs(mean(y) - mean(Y)))`. But you overstated your precision. 
  - Now you're kicking yourself. You'd briefly considered saying &plusmn; `r percent(your.widths[2]/2)` or &plusmn; `r percent(your.widths[3]/2)`. 
  - That would've done it. But that didn't sound as good, so you went for &plusmn; `r percent(your.widths[1]/2)`.
- You hope you're not the only one who missed. So you check out the competition.

## The Competition 

::: {.panel-tabset}
### Your Competitors' Polls

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{poll} & J_1 & Y_1 & J_2 & Y_2   & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\hline
\polla   1   
    & \polla   `r J=J1; Y=y[J]; J[1]`    & \polla  `r Y[1]` 
    & \polla   `r J[2]`   & \polla   `r Y[2]`           & \polla   \dots 
    & \polla   `r J[n]`   & \polla   `r Y[n]`           & \polla   `r approx(mean(Y))` \\
\pollb 2   
    & \pollb `r J=J2; Y=y[J]; J[1]`    & \pollb  `r Y[1]` 
    & \pollb `r J[2]`   & \pollb  `r Y[2]`          & \pollb \dots 
    & \pollb `r J[n]`   & \pollb `r Y[n]`           & \pollb `r approx(mean(Y))` \\
\pollc    3   
    & \pollc    `r J=J3; Y=y[J]; J[1]`    & \pollc     `r Y[1]` 
    & \pollc    `r J[2]`   & \pollc    `r Y[2]`           & \pollc    \dots 
    & \pollc    `r J[n]`   & \pollc    `r Y[n]`           & \pollc    `r approx(mean(Y))` \\
{\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots} &  {\vdots} \\ 
\polld    1M   
    & \polld    `r J=JM; Y=y[J]; J[1]`    & \polld     `r Y[1]` 
    & \polld    `r J[2]`   & \polld    `r Y[2]`           & \polld    \dots 
    & \polld    `r J[n]`   & \polld    `r Y[n]`           & \polld    `r approx(mean(Y))` \\
 {\vdots} & {\vdots} & {\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots}  & {\vdots} \\ 
\end{array}
$$

### Visualizing the Competition
```{r}
all.competitors.widths = meanY.samples |> map_vec(\(p)binom.width(p, n, alpha=.05))
competitors.widths = all.competitors.widths[1:100]

ggplot() + 
    with_annotations(meanY.samples, estimand=mean(y), 
                     height=pmf.height, highlight.dots=3, interval.width=competitors.widths, vline.alpha=NA) +
    your.interval + 
    scale_x + scale_y + coord 
```

:::

- It turns out that everyone got their point estimate exactly like you did.
    - Each of them rolled their `r approx(m)`-sided die `r n` times.
    - And after making their `r n` calls, they reported their sample proportion.
- You're not the only one who missed, but you're part of a pretty small club.
    - 5% of the other polls got it wrong.
- But you are the only one who claimed they'd get within 1%.
    - Everyone else said claimed they'd get within roughly `r percent(your.widths[3]/2)`. 
    - Most were right. About 95%. And a lot of them had worse point estimates than you.
- What you have is a *calibration* problem. 


## Calibration using the Sampling Distribution

```{r}
#| out-height: 400px
#| fig-asp: .3 

bar.plot = ggplot() + geom_bar(aes(x=meanY.samples, y=after_stat(prop)), alpha=.3) 
calibration.plot = bar.plot + 
    with_annotations(meanY.samples, estimand=mean(y), interval.width=competitors.widths,
                     height=pmf.height, highlight.dots=3) +
    your.interval + 
    scale_x + scale_y + coord

calibration.plot
```


- You'd have seen that your interval was miscalibrated if you knew what you do now.
    1. Your competitors' point estimates, which they got exactly like you got yours.
    2. Who actually turned out, so you could simulate even more polls if you wanted to.
- You'd have drawn &plusmn; `r percent(your.widths[1]/2)` intervals around all these point estimates.^[In technical terms, these estimates are draws from your estimator's *sampling distribution*.]
    - And noticed that many of these intervals cover [the target]{.target}. About `r percent(mean(abs(meanY.samples - mean(y)) <= your.widths[1]/2))` do.
    - Or, if leaving the target out of it, that only `r percent(mean(abs(meanY.samples - mean(meanY.samples)) <= your.widths[1]/2))` of them cover [the mean]{.blue} of all these estimates.
    - Same thing. The mean and target are right on top of one another. Like this: [|]{.blue .target}.
- And you'd have known how to choose the right interval width. One that makes 95% of these intervals cover. Right? 

## Quiz

```{r}
quiz.widths = c(binom.width(mean(y), 2*n, alpha=.5), 
                binom.width(mean(y), 2*n, alpha=.05), 
                binom.width(mean(y), 2*n, alpha=.003))
quiz.data = data.frame(x=mean(Y), w=quiz.widths,
                       y=pmf.height*c(.5,.4,.3), 
                       color=c(polla, pollb, pollc), 
                       alpha=c(1,1,1))
quiz.samples = rbinom(10000, 2*n, mean(y))/(2*n)
pmf.height = max(dbinom(0:(2*n), 2*n, mean(y)))

quiz.plot = ggplot() + geom_bar(aes(x=quiz.samples, y=after_stat(prop)), alpha=.3) +   
    with_annotations(quiz.samples, estimand=mean(y), highlight.dots=0, height=pmf.height) +
    geom_pointrange(aes(x=x, xmin=x-w/2, xmax=x+w/2, y=y, alpha=I(alpha), color=I(color)), 
                    data=quiz.data)
quiz.plot
```

- I've drawn the sampling distribution of an estimator, 100 draws from it as [`r circle`]{.purple}s, and 3 interval estimates.
- One of these interval estimates is calibrated to have exactly 95% coverage. Which is it?
  - A. The one around the [`r circle`]{.polla} on top. 
  - B. The one around the [`r circle`]{.pollb} in the middle. 
  - C. The one around the [`r circle`]{.pollc} on the bottom.
- *Submit* your answer [here](`r pollurl`).
- *Hint*. You can reach your twin with your arms if your twin can reach you with theirs. 
- *Hint*. Maybe you look like this [`r circle`]{.purple} and your twin like this [|]{.blue .target}.


## Calibration in Reality

```{r}
#| fig-width: 7
#| 
#| layout-ncol: 2
#| fig-cap: 
#| - "What You Know Now"
#| - "What You Knew Then" 
w95 = binom.width(mean(y), n, alpha=.05)
calibration.plot.shaded = bar.plot + 
    with_annotations(meanY.samples, estimand=mean(y), shade.width=w95, shade.color='green', 
                     interval.width=competitors.widths,
                     height=pmf.height, highlight.dots=3) +
    your.interval +
    scale_x + scale_y + coord

calibration.plot.shaded
ggplot() + your.interval + scale_x + scale_y + coord
```

- But you didn't know any of this. You just knew [your own point estimate]{.polld}.
- Without all that post-election information, you didn't know how to calibrate an interval estimate. 
- But everyone else did. Their widths were almost the same as the width you'd choose now. 

:::: columns 
::: column
```{r}
#| fig-width: 7
#| fig-cap: 
#| - Your competitors' intervals, recentered for easy comparison of width to the sampling distribution's.
ggplot() + 
    geom_rect(aes(xmin=mean(y)-w95/2, xmax=mean(y) + w95/2, ymin=-Inf, ymax=Inf), alpha=.1, fill='green') +
    geom_vline(xintercept=mean(y), color='green', size=2, alpha=.15) +
    geom_linerange(aes(y=pmf.height*seq(0,1,length.out=100), 
                        xmin=mean(y)-competitors.widths/2, xmax=mean(y)+competitors.widths/2), 
                        color='purple', alpha=.6) + scale_x + scale_y + coord 
```

:::
::: column

- It's as if they knew how to approximate their estimators' sampling distribution. They did.
- Today, we're going to work out how.
- We'll need to know a bit about sums of \
  independent binary random variables.

:::
::::

# Probability Review 
Sums of Independent Binary Random Variables

::: {.footnotestyle}
This'll be a slog. I'll explain why we've bothered when we get to the end.
:::

## From Sampling to Coin Flips

:::: columns
::: {.column width="30%"}
### The Population

$j$ | $y_j$
--|---
$1$ | `r y[1]`
$2$ | `r y[2]`
$3$ | `r y[3]`
$4$ | `r y[4]`
&vellip; | &vellip;
$m$ | `r y[m]`

:::
::: {.column width="30%"}
### The Joint

$p$ | $J_1$ | $Y_1$
--|---|---
$1/m$ | $1$ |  $`r y[1]`$
$1/m$ | $2$ |  $`r y[2]`$
$1/m$ | $3$ |  $`r y[3]`$
$1/m$ | $4$ |  $`r y[4]`$
&vellip; | &vellip; | &vellip;
$1/m$ | $m$ |  $`r y[m]`$

:::
::: {.column width="40%"}
### The Marginal

$p$ | $Y_1$
--|---
$\underset{\color{gray}\approx `r approx(mean(y==0))`}{\sum\limits_{j:y_j=0} \frac{1}{m}}$ | $0$
$\underset{\color{gray}\approx `r approx(mean(y==1))`}{\sum\limits_{j:y_j=1}  \frac{1}{m}}$ | $1$

:::
::::

- To find the distribution of one response $Y_1$, we start with the probability distribution of our dice rolls.
  - The person we call, $J_1$, is equally likely to be anyone in the population of $m \approx `r approx(m)`$ people.
  - It takes on each value $1 \ldots m$ with probability $1/m$. 
- Once we know who we're calling, we know their response, $Y_1$. 
    - If we call person $j$, their response is $y_j$. So we can add that column to our table 'for free'. 
    - The probability that we roll $j$ and hear the response $y_j$ is still $1/m$.
- And we can *marginalize* to find the distribution of $Y_1$.
    - We sum the probabilities in rows where $Y_1=0$ and $Y_1=1$. And the result is intuitive.
    - The *probability* of hearing a response $y \in \{0,1\}$ is the *frequency* of that response in the population. 
    - We'll call these frequencies $\theta_0$ and $\theta_1$. This is a little redundant because $\theta_0 = 1 - \theta_1$.
    - Our estimation target is just the 'Yes' frequency $\theta_1$. Some people call it the *success rate*. 

## From Sampling to Coin Flip Sums: Two Responses {#sec-sampling-to-sums-two}

:::: columns
::: {.column width="30%"}
### The Joint

$p$ | $J_1$ | $J_2$ | $Y_1$ | $Y_2$ | $Y_1 + Y_2$
--|---|---|---|---
$\frac{1}{m^2}$ |  $1$ | $1$ | $y_1$ | $y_1$ | $y_1+y_1$
$\frac{1}{m^2}$ |  $1$ | $2$ | $y_1$ | $y_2$ | $y_1+y_2$
&vellip; | &vellip; | &vellip; | &vellip; | &vellip;
$\frac{1}{m^2}$ |  $1$ | $m$ | $y_1$ | $y_m$ | $y_1+y_m$ 
$\frac{1}{m^2}$ |  $2$ | $1$ | $y_2$ | $y_1$ | $y_2+y_1$
$\frac{1}{m^2}$ |  $2$ | $2$ | $y_2$ | $y_2$ | $y_2+y_2$
&vellip; | &vellip; | &vellip; | &vellip; | &vellip;
$\frac{1}{m^2}$ |  $m$ | $m$ | $y_m$ | $y_m$ | $y_m+y_m$
:::
::: {.column width="42%"}
### Partially Marginalized

$p$ | $Y_1$ | $Y_2$ | $Y_1 + Y_2$
--|---|---|---
$\inpollb{\sum\limits_{j_1,j_2 : y_{j_1},y_{j_2} = 0,0} \frac{1}{m^2}}$ | $0$ | $0$ | $\inpollb{0}$
$\inpollc{\sum\limits_{j_1,j_2 : y_{j_1},y_{j_2} = 0,1} \frac{1}{m^2}}$ | $0$ | $1$ | $\inpollc{1}$
$\inpollc{\sum\limits_{j_1,j_2 : y_{j_1},y_{j_2} = 1,0} \frac{1}{m^2}}$ | $1$ | $0$ | $\inpollc{1}$
$\inpolld{\sum\limits_{j_1,j_2 : y_{j_1},y_{j_2} = 1,1} \frac{1}{m^2}}$ | $1$ | $1$ | $\inpolld{1}$

:::
::: {.column width="28%"}
### Fully Marginalized
$p$ | $Y_1 + Y_2$
--|---
$\inpollb{\sum\limits_{j_1,j_2: y_{j_1}+y_{j_2} = 0} \frac{1}{m^2}}$ | $\inpollb{0}$
$\inpollc{\sum\limits_{j_1,j_2: y_{j_1}+y_{j_2} = 1} \frac{1}{m^2}}$ | $\inpollc{1}$
$\inpolld{\sum\limits_{j_1,j_2: y_{j_1}+y_{j_2} = 2} \frac{1}{m^2}}$ | $\inpolld{2}$
:::
::::

::: {.panel-tabset}
### Two-Step Approach
- To find the distribution of a sum of two responses, $Y_1+Y_2$, we do the same thing.
- We start with the joint distribution of two dice rolls.
    - Our rolls are equally likely to be any pair of numbers in $1\ldots m$.
    - And there are $m^2$ pairs, so the probability is $1/m^2$ for each pair.
- Then we add columns that are functions of the rolls. 
    - This is 'for free'. The rolls are the rolls, so the probabilities don't change.
    - If we know who we're calling, we know their responses. So we can add columns for $Y_1$,$Y_2$, and $Y_1+Y_2$.
- Then we *marginalize* to find the distribution of the sum. We'll do this in two steps.
    
    1. We *partially marginalize* over $J_1,J_2$, to find the joint distribution of $Y_1, Y_2$. 
    2. We *fully marginalize* the result to find the distribution of $Y_1+Y_2$.

### Step 1. Partial Marginalization

- To sum over pairs of responses, we sum over one response then the other.
    - This exactly like what we did for one response. Except twice.
    - And what we get is pretty similar too.
- The probability of observing the pair of responses is determined by their frequencies in the population.
    - It's the *product* of the frequences of those responses.
    - For 'Yes,Yes' it's $\theta_1^2$. For 'No,No' it's $\theta_0^2$. 
    - And for 'Yes,No' and 'No,Yes' it's $\theta_1\theta_0$.
- This is, for what it's worth, the *product* of the marginal probabilities of those responses. 
    - When this happens for any pair of random variables, we say they are *independent*. More on that soon.
- And the probability of any response pair $a,b$ depends only on its sum $a+b$.
      
$$ 
\begin{aligned}
\sum\limits_{j_1,j_2: y_{j_1},y_{j_2} = a,b} \frac{1}{m^2} \\
&= \sum\limits_{j_1 : y_{j_1}=1} \qty{ \sum\limits_{j_2 : y_j_2=b} \frac{\1}{m^2} } \\ 
&= \sum\limits_{j_1 : y_{j_1}=1} \times \qty{ m_b \times \frac{1}{m^2} } \qfor m_y = \sum\limits_{j:y_j=y} 1
&= m_a \times m_b \times \frac{1}{m} = \underset{\color{gray}\theta_a}{\frac{m_a}{m}}\underset{\color{gray}\theta_b}{\frac{m_b}{m}} = \theta_0^{s} \theta_1^{1-s} \qfor s = a+b
\end{aligned}
$$

### Step 2. Full Marginalization 

- To calculate the marginal probability that $Y_1+Y_2=s$, we sum the probabilities of all pairs with that sum.
    - This is pretty easy because there aren't too many pairs. 
    - And I've color coded the pairs to make it even easier.

:::

## Exercise: Two Calls to A Small Population {#sec-two-calls-to-small}
```{r}
yex = c(0,0,1)
```

:::: columns 
::: {.column width="40%"}

$j$ | $y_j$
--|---
$1$ | $`r yex[1]`$
$2$ | $`r yex[2]`$
$3$ | $`r yex[3]`$

:::
::: {.column width="60%"}
```{r}
#| fig-height: 2.7
#| fig-width:  7
ggplot(data.frame(j=1:length(yex), y=yex[1:3])) + 
    geom_point(aes(x=j, y=y), size=5, alpha=.3) + 
    scale_y_continuous(breaks=c(0,1)) + 
    scale_x_continuous(breaks=1:length(yex))  
```

:::
::::

::: {.panel-tabset}
### Instructions

1. Make a table for the joint distribution of three rolls of a 3-sided die.  Add columns for the responses.
2. Partially marginalize to get the joint distribution of the responses. Add a column for their sum.
3. Marginalize to get the distribution of the sum.


::: {.footnotestyle} 
Aren't you happy that, even though you can't make a 3-sided die, I didn't ask you to use 4-sided one? 
:::

### Blank Space


### Solution

:::: columns

::: {.column width="40%"}

$p$ | $J_1$ | $J_2$ | $Y_1$ | $Y_2$ 
---|---|---|---|---
$\pollb 1/9$ |  $1$ | $1$ | $`r yex[1]`$ | $`r yex[1]`$
$\pollb 1/9$ |  $1$ | $2$ | $`r yex[1]`$ | $`r yex[2]`$
$\pollc 1/9$ |  $1$ | $3$ | $`r yex[1]`$ | $`r yex[3]`$
$\pollb 1/9$ |  $2$ | $1$ | $`r yex[2]`$ | $`r yex[1]`$
$\pollb 1/9$ |  $2$ | $2$ | $`r yex[2]`$ | $`r yex[2]`$
$\pollc 1/9$ |  $2$ | $3$ | $`r yex[2]`$ | $`r yex[3]`$
$\pollc 1/9$ |  $3$ | $1$ | $`r yex[3]`$ | $`r yex[1]`$
$\pollc 1/9$ |  $3$ | $2$ | $`r yex[3]`$ | $`r yex[2]`$
$\polld 1/9$ |  $3$ | $3$ | $`r yex[3]`$ | $`r yex[3]`$
:::

::: {.column width="30%"}

$p$ | $Y_1$ | $Y_2$ | $Y_1 + Y_2$
---|---|---|---
$\pollb \frac{4}{9}$ | $0$ | $0$ | $0$
$\pollc \frac{2}{9}$ | $0$ | $1$ | $1$
$\pollc \frac{2}{9}$ | $1$ | $0$ | $1$
$\polld \frac{1}{9}$ | $1$ | $1$ | $2$

:::
::: {.column width="30%"}

$p$ | $Y_1 + Y_2$
---|---
$\pollb \frac{4}{9}$ | $0$
$\pollc \frac{4}{9}$ | $1$
$\polld \frac{1}{9}$ | $2$

:::
::::
:::

## From Sampling to Coin Flip Sums: $n$ Responses


:::: columns
::: {.column width="45%"}
:::: {.panel-tabset}
### The Joint

$p$ | $J_1$ | &hellip; | $J_n$ | $Y_1$ | &hellip; | $Y_n$ | $Y_1 + \ldots + Y_n$
--|---|---|---|--|-|-
$\frac{1}{m^n}$ |  $1$ | &hellip; | $1$ | $y_1$ | &hellip; | $y_1$ | $y_1+\ldots+y_1$
$\frac{1}{m^n}$ |  $1$ | &hellip; | $2$ | $y_1$ | &hellip; |$y_2$ |  $y_1+\ldots+y_2$
&vellip; | &vellip; | &dtdot; | &vellip; | &vellip; | &dtdot; | &vellip; | &vellip;
$\frac{1}{m^n}$ |  $1$ | &hellip; | $m$ | $y_1$ | &hellip; | $y_m$ | $y_1+\ldots+y_m$ 
$\frac{1}{m^n}$ |  $2$ | &hellip; |$1$ | $y_2$ | &hellip; |$y_1$ | $y_2+\ldots +y_1$
$\frac{1}{m^n}$ |  $2$ | &hellip; |$2$ | $y_2$ | &hellip; |$y_2$ | $y_2+\ldots +y_2$
&vellip; | &vellip; | &dtdot; |&vellip; | &vellip; | &dtdot; | &vellip; | &vellip;
$\frac{1}{m^n}$ |  $m$ | &hellip; |$m$ | $y_m$ | &hellip; |$y_m$ | $y_m+\ldots+y_m$

### Fully Marginalized
$p$ | $Y_1 + \ldots + Y_n$
--|---
$\pollb \sum\limits_{\substack{a_1 \ldots a_n \\ a_1 + \ldots + a_n = 0}} \sum\limits_{\substack{j_1 \ldots j_n \\ y_{j_1} \ldots y_{j_n} = a_1 \ldots a_n}} \frac{1}{m^n}$ | $\pollb 0$
$\pollc \sum\limits_{\substack{a_1 \ldots a_n \\ a_1 + \ldots + a_n = 0}} \sum\limits_{\substack{j_1 \ldots j_n \\ y_{j_1} \ldots y_{j_n} = a_1 \ldots a_n}} \frac{1}{m^n}$ | $\pollc 1$
&vellip; | &vellip; 
$\polld \sum\limits_{\substack{a_1 \ldots a_n \\ a_1 + \ldots + a_n = n-1}} \sum\limits_{\substack{j_1 \ldots j_n \\ y_{j_1} \ldots y_{j_n} = a_1 \ldots a_n}} \frac{1}{m^n}$ | $\polld n-1$
$\polle \sum\limits_{\substack{a_1 \ldots a_n \\ a_1 + \ldots + a_n = n}} \sum\limits_{\substack{j_1 \ldots j_n \\ y_{j_1} \ldots y_{j_n} = a_1 \ldots a_n}} \frac{1}{m^n}$ | $\polle n$

::::
:::

::: {.column width="55%"}
### Partially Marginalized

$p$ | $Y_1$ |  $Y_2$ |  &hellip; | $Y_n$ | $Y_1 + \ldots + Y_n$
--|---|---|---|---|-
$\pollb \sum\limits_{\substack{j_1 \ldots j_n \\ y_{j_1}, y_{j_2} \ldots y_{j_n} = 0, 0 \ldots 0}} \frac{1}{m^n}$ | 0 | 0 | &hellip; | $0$ |  $\pollb 0$ 
$\pollc \sum\limits_{\substack{j_1 \ldots j_n \\ y_{j_1}, y_{j_2} \ldots y_{j_n} = 0, 0 \ldots 1}} \frac{1}{m^n}$ | 0 | 0 | &hellip; | $1$ |  $\pollc 1$
&vellip; | &vellip; | &dtdot; | &vellip; | &vellip; | &dtdot; | &vellip; | &vellip;
$\pollc \sum\limits_{\substack{j_1 \ldots j_n \\ y_{j_1}, y_{j_2} \ldots y_{j_n} = 0, 1 \ldots 0}} \frac{1}{m^n}$ | 0 | 1 | &hellip; | $0$ |  $\pollc 1$
&vellip; | &vellip; | &dtdot; | &vellip; | &vellip; | &dtdot; | &vellip; | &vellip;
$\polld \sum\limits_{\substack{j_1 \ldots j_n \\ y_{j_1}, y_{j_2} \ldots y_{j_n} = 0, 1 \ldots 1}} \frac{1}{m^n}$ | 0 | 1 | &hellip; | $1$ |  $\polld n-1$
$\polle \sum\limits_{\substack{j_1 \ldots j_n \\ y_{j_1}, y_{j_2} \ldots y_{j_n} = 1, 1 \ldots 1}} \frac{1}{m^n}$ | 1 | 1 | &hellip; | $1$ |  $\polle n$

:::
::::

::: {.panel-tabset}
### Two-Step Approach
- To find the distribution of a sum $Y_1 + \ldots + Y_n$, we do the same thing.
- We start by writing out the joint distribution of $n$ dice rolls.
    - We're equally likely to roll any sequence of $n$ numbers in $1\ldots m$.
    - And there are $m^n$ sequences, so the probability is $1/m^n$ for each sequence.
    - Then we add columns for the corresponding responses 'for free'. And the sums.
- Then we marginalize in two steps.
    
    1. Sum over roll sequences leading to the same response sequence $a_1 \ldots a_n$.
    2. Sum over response sequences leading to the same sum $s=a_1+\ldots+a_n$.
    
### Step 1. Partial Marginalization

- To sum over sequences of responses, we sum over one response then the other.
    - This exactly like what we did for one response. Except over and over.
    - And what we get is pretty similar to the two-response case.
- The probability of observing the sequence is determined by the frequency of Yeses and Nos in the population.
    - It's the *product* of the frequences of those responses.
    - For 'Yes,Yes,...,Yes' it's $\theta_1^n$. For 'No,No,...,No' it's $\theta_0^n$. 
    - And if we have $s$ Yeses and therefore $n-s$ Nos, it's $\theta_1^s\theta_0^{n-s}$.
- This is, again, the *product* of the marginal probabilities of those $n$ responses.

$$
\begin{aligned}
\sum\limits_{\substack{j_1 \ldots j_n  \\ y_{j_1} \ldots y_{j_n} = a_1 \ldots a_n}} \frac{1}{m^n} 
&= \sum\limits_{\substack{j_1 \\ y_{j_1}=a_1}}  \ \ldots \  \sum\limits_{\substack{j_{n-1} \\ y_{j_{n-1}}=a_{n-1}}} \ \sum\limits_{\substack{j_n \\y_{j_n}=a_n}} \frac{1}{m^n}  \\
&= \sum\limits_{\substack{j_1 \\ y_{j_1}=a_1}} \  \ldots \ \sum\limits_{\substack{j_{n-1} \\ y_{j_{n-1}}=a_{n-1}}}  m_{a_n} \times \frac{1}{m^n} \\
&= \sum\limits_{\substack{j_1 \\ y_{j_1}=a_1}} m_{a_2} \ldots m_{a_{n-1}} m_{a_n} \times \frac{1}{m^n}  \\
&= m_{a_1} \ldots m_{a_n} \times \frac{1}{m^n}  = \theta_{a_i} \ldots \theta_{a_n} \\
&= \prod_{i:a_i=1} \theta_1 \prod_{i:a_i=0} \theta_0 = \theta_1^{s} \theta_0^{n-s} \qfor s = a_1 + \ldots + a_n
\end{aligned}
$$

### Step 2. Full Marginalization

- The full marginalization step is a bit trickier. But we can take advantage of our partial marginalization.
- To calculate the probability that the sum takes on the value $s$, we order our sum of probabilities thoughtfully.
    - We sum the over response sequences $a_1 \ldots a_n$ with the same number of Yeses and Nos.
    - And within that, sum over calls where we'd hear those responses.
- The inner sum we've done. That was our partial marginalization.
- And the outer sum boils down to counting the number of sequences with $s$ Yeses.
    - The probability of all these sequences is the same: $\theta_1^s\theta_0^{n-s}$. So summing is counting and multiplying.
    - You may have done the counting part in high school in a unit on 'permutations and combinations'.
    - In any case, that count has a name. It's spoken '$n$ choose $s$' and written $\binom{n}{s}$.

$$
\begin{aligned}
\sum\limits_{\substack{j_1 \ldots j_n \\ y_{j_1} + \ldots + y_{j_n} = s}} \frac{1}{m^n} 
&= \sum\limits_{\substack{a_1 \dots a_n \\ a_1 + \ldots + a_n = s}} \ \ \sum_{\substack{j_1 \ldots j_n \\ y_{j_1} \ldots y_{j_n} = a_1 \ldots a_n}} \frac{1}{m^n} \\
& =\sum\limits_{\substack{a_1 \dots a_n \\ a_1 + \ldots + a_n = s}} \theta_1^{s}\theta_0^{n-s} \\
&= \binom{s}{n} \theta_1^{s} \theta_0^{n-s} \qqtext{ where } \binom{s}{n} = \sum\limits_{\substack{a_1 \dots a_n \\ a_1 + \ldots + a_n = s}} 1
\end{aligned} 
$$

### Using **R**

$$
P\qty(\sum_{i=1}^n Y_i = s) =  \binom{s}{n} \theta_1^{s}\theta_0^{n-s} 
\ \ \text{ where } \ \ \binom{s}{n} \text{ is the number of binary sequences $a_1 \ldots a_n$ summing to $s$.}
$$

- We don't actually have to do all this sequence-summing-to-$s$ counting ourselves.
- The `choose` function in **R** will do it for us: $\binom{s}{n}$ is `choose(n,s)`.
- The `dbinom` function will give us the whole probability: $\binom{s}{n}\theta_1^s\theta_0^{n-s}$ is `dbinom(s, n, theta_1)`.
- The `rbinom` function will draw samples from this distribution: `rbinom(10000, n, theta_1)` gives us 10,000. 

```{r}
#| echo: true
theta_1 = .7
n = 625
p = dbinom(0:n, n, theta_1)
S = rbinom(10000, n, theta_1)

ggplot() + geom_area(aes(x=0:n,   y=p), color=pollc, alpha=.2) +
           geom_histogram(aes(x=S), bins=60, alpha=.3) +
           geom_point(aes(x=S[1:100],   y=max(p)*seq(0,1,length.out=100)), 
                      color='purple', alpha=.2)
```

:::

## The Binomial Distribution
:::: columns
::: {.column width="50%"}
```{r}
#| fig-width: 7
 
n = 625
thetas = c(.68, .70, .72)
p1 = dbinom(0:n, n, thetas[1])
p2 = dbinom(0:n, n, thetas[2])
p3 = dbinom(0:n, n, thetas[3])

binomial.dist.plot = 
	ggplot() + geom_area(aes(x=0:n, y=p1), fill=pollb,    color=pollb,    alpha=.3) + 
           geom_area(aes(x=0:n, y=p2), fill=pollc,    color=pollc,    alpha=.3) + 
           geom_area(aes(x=0:n, y=p3), fill=polld,    color=polld,    alpha=.3) +
           scale_x_continuous(breaks=seq(0,1,by=.1)*n, labels=sprintf('%1.1fn', seq(0,1,by=.1)), limits=c(.5,.9)*n)
binomial.dist.plot
```

:::
::: {.column width="50%"}
$$
\begin{aligned}
&\overset{\color{gray}=P\qty(\frac{1}{n}\sum_{i=1}^n Y_i = \frac{s}{n})}{P\qty(\sum_{i=1}^n Y_i = s)} =  \binom{s}{n} \theta_1^{s}\theta_0^{n-s} \\
&\qfor n =`r n` \\ 
&\qand \theta_1 \in \{\inpollb{`r thetas[1]`}, \inpollc{`r thetas[2]`}, \inpolld{`r thetas[3]`} \}
\end{aligned}
$$

:::
::::

- These functions have 'binom' in their name because we call this distribution the *Binomial distribution*. 
    - The Binomial distribution on $n$ *trials* with *success probability* $\theta$ is our name for ...
    - the distribution of the sum of $n$ independent binary random variables with probability $\theta$ of being $1$.
    - E.g. the number of heads in $n$ coin flips is Binomial with $n$ trials and success probability $1/2$.
- We've shown that's the sampling distribution of the *sum of responses*, $Y_1 + \ldots + Y_n$, when we sample ...
    - ... with replacement from a population of binary responses $y_1 \ldots y_m$ in which $\theta$ is the frequency of ones.
- We're interested in the *mean of responses*, so we divide by $n$: $\color{gray} \sum_{i=1}^n Y_i = s \ \text{ when } \ \frac{1}{n}\sum_{i=1}^n Y_i = s/n$.
- It's easy to estimate --- and talk about estimating --- this sampling distribution.
    - It depends only on one thing we don't know --- the population frequency $\theta$.
    - And that's exactly the thing we're trying to estimate anyway.
- That's why we've started here---the case of sampling with replacement from a population of binary responses.

## Payoff 

```{r}
set.seed(1)
samples = rbinom(10000, n, mean(Y))/n
your.widths = c(width(samples, alpha=.05), 
                binom.width(mean(Y), n, alpha=.05),
                binom.width(mean(y), n, alpha=.05))               
opacities = c(1,1,1)
interval.data = data.frame(x=mean(Y), w=your.widths,
                           y=pmf.height*c(.57,.63,.6), alpha=opacities)
your.calibrated.interval = geom_pointrange(aes(x=x, xmin=x-w/2, xmax=x+w/2, y=y, alpha=I(alpha)), 
                                            data=interval.data[1,], color=polld)

your.calibrated.interval.plot = ggplot() + 
    with_annotations(meanY.samples, estimand=mean(y), 
                     height=pmf.height, highlight.dots=3, interval.width=competitors.widths, vline.alpha=NA) +
    your.calibrated.interval + 
    scale_x + scale_y + coord 

your.calibration.picture = ggplot() + geom_bar(aes(x=samples, y=after_stat(prop)), alpha=.3) + 
    with_annotations(samples, height=pmf.height, highlight.dots=0, 
                     shade.width = your.widths[1], shade.color='red', shade.alpha=.15) +
    your.calibrated.interval + 
    scale_x + scale_y + coord
```

```{r}
#| layout-ncol: 2
#| fig-width: 7
#| fig-cap:
#| - "You vs. Your Competitors Now"
#| - "How You're Calibrating"
your.calibrated.interval.plot
your.calibration.picture
```

- To estimate this sampling distribution, you plug your point estimate $\hat\theta$ into the Binomial formula.
$$
\hat P\qty(\sum_{i=1}^n Y_i = s) = \binom{s}{n} \hat\theta^{s} (1-\hat\theta)^{n-s} \qqtext{ estimates }
P\qty(\sum_{i=1}^n Y_i = s) = \binom{s}{n} \theta^{s} (1-\theta)^{n-s}
$$

- To calibrate [your interval estimate]{.polld}, you ...
  
    - use `rbinom` to draw 10,000 samples from this estimate of the sampling distribution. 
    - Even remembering to divide by $n$.
    - And use the function `width` from the [Week 1 Homework](https://qtm285-1.github.io/assets/homework/week1-homework.html#exr-interval-estimate) to find an interval that covers 95% of them.

```{r}
#| echo: true
#| eval: false
theta.hat = mean(Y)
samples = rbinom(10000, n, theta.hat) / n
interval.width = width(samples, alpha=.05)
```


- You nail it. Your interval covers the estimation target $\theta$ just like 95% of your competitors' do.

:::: columns 
::: column
```{r}
#| fig-width: 7
#| fig-cap: 
#| - You and your competitors' intervals, recentered for easy comparison of width to your estimators' sampling distribution's middle-95% width.
ggplot() + 
    geom_rect(aes(xmin=mean(y)-w95/2, xmax=mean(y) + w95/2, ymin=-Inf, ymax=Inf), alpha=.15, fill='green') +
    geom_vline(xintercept=mean(y), color='green', size=2, alpha=.7) +
    geom_linerange(aes(y=pmf.height*seq(0,1,length.out=100), 
                        xmin=mean(y)-competitors.widths/2, xmax=mean(y)+competitors.widths/2), 
                        color='purple', alpha=.15) + 
    geom_linerange(aes(y=pmf.height*.59, 
                        xmin=mean(y)-your.widths[1]/2, xmax=mean(y)+your.widths[1]/2), color=polld) +
                        scale_x + scale_y + coord 
```

:::
::: column

- It's not just that you've widened your interval enough.
- You've widened it almost exactly the right amount. 
- Just like your competitors. Almost as if you all knew ...
- how to estimate your estimator's sampling distribution.

:::
::::

# Did We Have To Do All This?

## This Was A Lot of Work

:::: columns

::: {.column width="40%"}

$p$ | $J_1$ | $J_2$ | $Y_1$ | $Y_2$ 
---|---|---|---|---
$\pollb 1/9$ |  $1$ | $1$ | $`r yex[1]`$ | $`r yex[1]`$
$\pollb 1/9$ |  $1$ | $2$ | $`r yex[1]`$ | $`r yex[2]`$
$\pollc 1/9$ |  $1$ | $3$ | $`r yex[1]`$ | $`r yex[3]`$
$\pollb 1/9$ |  $2$ | $1$ | $`r yex[2]`$ | $`r yex[1]`$
$\pollb 1/9$ |  $2$ | $2$ | $`r yex[2]`$ | $`r yex[2]`$
$\pollc 1/9$ |  $2$ | $3$ | $`r yex[2]`$ | $`r yex[3]`$
$\pollc 1/9$ |  $3$ | $1$ | $`r yex[3]`$ | $`r yex[1]`$
$\pollc 1/9$ |  $3$ | $2$ | $`r yex[3]`$ | $`r yex[2]`$
$\polld 1/9$ |  $3$ | $3$ | $`r yex[3]`$ | $`r yex[3]`$
:::

::: {.column width="30%"}

$p$ | $Y_1$ | $Y_2$ | $Y_1 + Y_2$
---|---|---|---
$\pollb \frac{4}{9}$ | $0$ | $0$ | $0$
$\pollc \frac{2}{9}$ | $0$ | $1$ | $1$
$\pollc \frac{2}{9}$ | $1$ | $0$ | $1$
$\polld \frac{1}{9}$ | $1$ | $1$ | $2$

:::
::: {.column width="30%"}

$p$ | $Y_1 + Y_2$
---|---
$\pollb \frac{4}{9}$ | $0$
$\pollc \frac{4}{9}$ | $1$
$\polld \frac{1}{9}$ | $2$

:::
::::

- You might be wondering if we really had to do all this.
- In some other classes on these topics, e.g. QTM220, you'd just be told a couple things we worked out.

1. When you sample with replacement from a binary population $y_1 \ldots y_m$ ...
    - your observations $Y_1 \ldots Y_n$ are independent and have the same distribution.^[We say *independent and identically distributed* or *iid* for short.]
    - each one is like the result of a coin flip that comes up heads with probability $\theta=\frac{1}{m}\sum_{i=j}^m y_j$,\
      the frequency of ones in the population.^[We call these *Bernoulli($\theta$)* random variables.]
2. The sum of $n$ independent flips like this is the *Binomial($n,\theta$)* distribution.

- But this doesn't apply to you unless you're ...
  - looking at the sample mean
  - after sampling with replacement 
  - from a binary population 

## What It Bought Us

:::: columns
::: {.column width="50%"}
```{r}
#| fig-width: 7
 
binomial.dist.plot
```

:::
::: {.column width="50%"}
$$
\begin{aligned}
&\overset{\color{gray}=P\qty(\frac{1}{n}\sum_{i=1}^n Y_i = \frac{s}{n})}{P\qty(\sum_{i=1}^n Y_i = s)} =  \binom{s}{n} \theta_1^{s}\theta_0^{n-s} \\
&\qfor n =`r n` \\ 
&\qand \theta_1 \in \{\inpollb{`r thetas[1]`}, \inpollc{`r thetas[2]`}, \inpolld{`r thetas[3]`} \}
\end{aligned}
$$

:::
::::

- What we did 'always works'.
- To find the sampling distribution of any estimator after any kind of random sampling, you can ...
  1. start a the joint distribution describing your sampling mechanism, e.g. our sequence of $n$ calls $J_1 \ldots J_n$
  2. add a column for your estimator. That's 'for free' because if you know your sample, you know your estimate.
  3. marginalize over all the other columns.
- If you do this symbolically, with placeholders $y_1 \ldots y_m$ instead of actual numbers
  - e.g. as in @sec-sampling-to-sums-two vs @sec-two-calls-to-small
  - then you'll find your estimator's sampling distribution as a function of your population.
- What you'd hope for is this.
  1. You can actually do it. You can organize the sums so you get a useful description of a distribution.
  2. It depends on your population in a relatively simple way, e.g. through the 'Yes frequency' $\theta$.
- When it works out, we get a *formula* for our sampling distribution in terms of *parameters* summarizing our population. 
  - We don't know what the sampling distribution looks like exactly because we don't know those parameters. 
  - But our formula tells us what it can look like and how that that depends on the population. 
- We call this formula the distribution's *parametric form*. Knowing this is a start. 
  - We can estimate the parameters and plug them in, as you saw a moment ago. 	
  - We call estimators that rely on this kind of knowledge *parametric*. 

## Why 'Always' and not Always?

```{r}
binomial.dist.plot
```

- The bad news is that it doesn't work out like this all that often. 
- Often doing all this marginalization symbolically is too hard or doesn't leave you with an interpretable formula.
- The good news is that there are a variety of ways to work around this.
- If it 'works out halfway', meaning you know the parameter(s) it depends on but not the formula ... 
  - .. you can calculate the sampling distribution as accurately as you want by simulation. 
  - You'll get a chance to try that in this week's homework. 
- Failing that, you can work with a simpler *approximation* to your estimator's sampling distribution.
  - Ideally, you'd have some mathematical result telling you that approximation is accurate. 
  - And proving that yourself would involve some thinking about this marginalization process.
- That's what people usually do. Including us later this semester.
- This usually works and often, though not always, we have mathematical proof.

### Why This Still Matters 

- There are, however, important instances in which the usual approximations can fail pretty badly.
  1. When we're running an *adaptive randomized trial*. These let us experiment with new treatments with minimal risk of a negatively impacting patients who'd have done better on the old one.
  2. When we're estimating *effect variability*. These help understand whether a treatment that doesn't work on average might still be good for some people.
  3. When we're studying people who are connected in some way, e.g. in a social network or in a classroom. 
- This is hard stuff. We're not going to get there in this class. 
- But knowing how inference works from the ground up will hopefully help you in the future.
    - That might mean working out a new estimator's sampling distribution from scratch.
    - Or maybe  knowing what to look for or ask about when the usual stuff isn't working.
- If you want to know more about any of the three estimation problems above, let me know.
    - I can give you reading or maybe even a project idea.


# Calibration using the Binomial Distribution 

## Terminology: Sums vs. Means

:::: columns
::: {.column width="50%"}
```{r}
#| fig-width: 7
 
n = 625
thetas = c(.68, .70, .72)
p1 = dbinom(0:n, n, thetas[1])
p2 = dbinom(0:n, n, thetas[2])
p3 = dbinom(0:n, n, thetas[3])

binomial.dist.plot = 
	ggplot() + geom_area(aes(x=0:n, y=p1), fill=pollb,    color=pollb,    alpha=.3) + 
           geom_area(aes(x=0:n, y=p2), fill=pollc,    color=pollc,    alpha=.3) + 
           geom_area(aes(x=0:n, y=p3), fill=polld,    color=polld,    alpha=.3) +
           scale_x_continuous(breaks=seq(0,1,by=.1)*n, labels=sprintf('%1.1fn', seq(0,1,by=.1)), limits=c(.5,.9)*n)
binomial.dist.plot
```

:::
::: {.column width="50%"}
$$
\begin{aligned}
&\overset{\color{gray}=P\qty(\frac{1}{n}\sum_{i=1}^n Y_i = \frac{s}{n})}{P\qty(\sum_{i=1}^n Y_i = s)} =  \binom{s}{n} \theta^{s}(1-\theta)^{n-s} \\
&\qfor n =`r n` \\ 
&\qand \theta \in \{\inpollb{`r thetas[1]`}, \inpollc{`r thetas[2]`}, \inpolld{`r thetas[3]`} \}
\end{aligned}
$$

:::
::::

- Blah
- This visualization is kind of a hedge. 


## Looking Back on Your Success

```{r}
#| layout-ncol: 2
#| fig-width: 7
#| fig-cap:
#| - "What You Know Now---Plus What You Used For Calibration"
#| - "Your Calibration Picture---Plus the Estimation Target"
your.calibrated.interval.plot + geom_bar(aes(x=samples, y=after_stat(prop)), fill='red', alpha=.3)  
your.calibration.picture + geom_vline(xintercept=mean(y), color='green', size=2, alpha=.7)
```

- Remember your sucessfully calibrated interval estimate from a moment ago?
    - You did that by plugging your point estimate $\hat\theta$ into the Binomial formula.
    - And you got a nicely calibrated interval estimate. That was great.
- But let's take a closer look. Let's compare your estimate of the sampling distribution to the actual thing.
    - We can do that because we have a bunch of draws from the real thing---all the others polls. [`r circle`]{.purple}s.
    - And if we want more draws, since it's after the election, we can simulate as many polls as we want.
- It doesn't look great. It's off center.
    - Its mean is a bit lower than [the population proportion]{.target} $\theta$.
    - It's actually our sample proportion $\hat\theta$. That makes sense. 
- The mean of the Binomial distribution with success rate $\theta$ is $\theta$ and we're using $\hat\theta$ in its place.
- It turns out that this doesn't matter much. It worked just fine for calibration. Why?

## Calibration Comparison

```{r}
#| out-height: 400px
#| fig-asp: .3 

w.est = binom.width(mean(Y), n, alpha=.05)
w = binom.width(mean(y), n, alpha=.05)
K = 100
heights = annotation_dot_heights(K, 3, pmf.height)
calibration.comparison.histogram = bar.plot +
    geom_rect(aes(xmin=mean(y)-w/2, xmax=mean(y)+w/2, ymin=0, ymax=1.25*heights[1]), alpha=.15, fill='green') +
    geom_rect(aes(xmin=mean(y)-w.est/2, xmax=mean(y)+w.est/2, ymin=1.25*heights[1], ymax=Inf), alpha=.15, fill='red') + 
    geom_vline(xintercept=mean(y), color='green', size=.5)


calibration.comparison.histogram +  
    geom_pointrange(aes(y=heights[1]*1.5, x=mean(Y), 
                        xmin = mean(Y)-w.est/2, xmax = mean(Y)+w.est/2), alpha=1, color='black')  +
    geom_pointrange(aes(y=heights[1], x=mean(Y), 
                        xmin = mean(Y)-w/2, xmax = mean(Y)+w/2), alpha=.4, color='black') +
                        scale_x + scale_y + coord  
```

- It doesn't matter because we're not putting arms on draws from the estimated sampling distribution.
- We're putting arms on our point estimate^[We're coloring it black instead of [green]{.polld} here. It's hard to see [green on a green background]{.polld .target}], which is a draw from the **actual sampling distribution**.
- To do this, we're using the width---but not the center---of the estimated sampling distribution.
- And it works because that's very close to the width we'd get from the actual sampling distribution.
- Above, I've drawn in shaded regions corresponding to two versions of the the population proportion's arms.
    - The [green region]{.target} is the one we get from the actual sampling distribution. We can't use this.
    - The [red region]{.shadedred} is the one we get from the estimated sampling distribution. We do use this.
- And I've drawn in a version of our interval estimate calibrated each way. They're almost the same.

## Our Competitors' Calibration

```{r}
#| fig-width: 7
#| layout-ncol: 2 
#| fig-asp: .4 
#| fig-cap:
#|   - "Actual Sampling Distribution"
#|   - "Estimated Sampling Distribution (Poll 1)"
#|   - "Estimated Sampling Distribution (Poll 2)"
#|   - "Estimated Sampling Distribution (Poll 3)"

plot.interval = function(Y, height, color) {
    n = length(Y)
    w = binom.width(mean(Y), n, alpha=.05)
    geom_pointrange(aes(x=mean(Y), y=height, xmin=mean(Y)-w/2, xmax=mean(Y)+w/2), color=color)
}

calibration.diagram = function(Y, color, height) {
    set.seed(default.seed)
    n = length(Y)
    samples = rbinom(10000, n, mean(Y)) / n 
    w = binom.width(mean(Y), n, alpha=.05)
    histplot = ggplot() + geom_bar(aes(x=samples, y=after_stat(prop)), alpha=.3) +
                          geom_vline(aes(xintercept=mean(y)), color='green', size=3, alpha=.7) + 
                          with_annotations(samples, shade.width=w, shade.color='red', height=pmf.height, highlight.dots=0) +
    geom_line(aes(x=x/n, y=dbinom(x, n, mean(Y))), color=color) +
    geom_pointrange(aes(x=mean(Y), xmin=mean(Y)-w/2, xmax=mean(Y)+w/2, y=height), color=color)
}

calibration.diagram(y[J1], polla, heights[1]) + coord
calibration.diagram(y[J2], pollb, heights[2]) + coord
calibration.diagram(y[J3], pollc, heights[3]) + coord
```

```{r}
dh = .2*heights[1]
our.calibration.plot = ggplot() + 
    geom_line(aes(x=x/n, y=dbinom(x, n, mean(y))), color='gray') 

our.calibration.plot
```
bar.plot + geom_line(aes(x=x/n, y=dbinom(x, n, mean(y))), color='gray') +
               plot.interval(y[J1], dh+heights[1], color=polla) +
               plot.interval(y[J2], dh+heights[2], color=pollb) +
               plot.interval(y[J3], dh+heights[3], color=pollc) + coord
```

This means each of our friends estimates a different width for their interval.
- But they're all close to the width we get using the actual sampling distribution. You can see it above.
    - I've plotted the intervals they'd use --- based on their sampling distribution estimates --- in bold colors.
    - And the intervals they wish they could --- based on the actual sampling distribution --- more faintly. 


## Evidence That It Works 

```{r}
#| out-height: 400px
K = 100
ps = meanY.samples[1:K]

ggplot() + geom_line(aes(x=x/n-p, y=n*dbinom(x,n,p)), data=expand.grid(p=mean(y), x=0:n)) +
           geom_vline(aes(xintercept=qbinom(c(.025,.975), n,mean(y))/n-mean(y)), linetype='dotted') +
           geom_line(aes(x=x/n-p, y=n*dbinom(x,n,p), color=factor(p)), data=expand.grid(p=ps, x=0:n), alpha=.2) +
           geom_pointrange(aes(x=0, xmin=qbinom(.025,n,p)/n-p, xmax=qbinom(.975,n,p)/n-p, y=20*k/K, color=factor(p)), data=data.frame(p=ps, k=1:K)) +
           xlim(-.1,.1) + guides(color=FALSE)
```

- We see the same thing when we look at 100 polls.
- Our friends' sampling distribution estimates are all close to the actual sampling distribution.
- The resulting interval-widths are near-indistinguishable from the width we got using the sampling distribution.
  - The estimated widths are shown as points with arms.
  - That sampling distribution's actual width is the space between the dotted lines.
- You can see estimating the width like this works reliably.

## Why Does it Work?

```{r}
#| out-height: 300px
#| fig-asp: .6 
#| layout-ncol: 2
height = pmf.height 
K = 3
ps = quantile(meanY.samples, c(.025, .5, .975))

density.stability.plot = ggplot() + 
           geom_line(aes(x=x/n, y=dbinom(x,n,p)), data=expand.grid(p=mean(y), x=0:n), alpha=.5) +
           geom_vline(aes(xintercept=qbinom(c(.025,.975), n,mean(y))/n), linetype='dotted') +
           geom_line(aes(x=x/n, y=dbinom(x,n,p), color=factor(p)), data=expand.grid(p=ps, x=0:n), alpha=.5) +
           geom_pointrange(aes(x=p, xmin=qbinom(.025,n,p)/n, xmax=qbinom(.975,n,p)/n, y=.6*height*k/K, color=factor(p)), data=data.frame(p=ps, k=1:K)) +
           xlim(.6,.8) + guides(color=FALSE) + geom_vline(xintercept=mean(y), color='green', size=1, alpha=.5)
centered.density.stability.plot = ggplot() + 
           geom_line(aes(x=x/n-p, y=dbinom(x,n,p)), data=expand.grid(p=mean(y), x=0:n), alpha=.5) +
           geom_vline(aes(xintercept=qbinom(c(.025,.975), n,mean(y))/n-mean(y)), linetype='dotted') +
           geom_line(aes(x=x/n-p, y=dbinom(x,n,p), color=factor(p)), data=expand.grid(p=ps, x=0:n), alpha=.5) +
           geom_pointrange(aes(x=0, xmin=qbinom(.025,n,p)/n-p, xmax=qbinom(.975,n,p)/n-p, y=.6*height*k/K, color=factor(p)), data=data.frame(p=ps, k=1:K)) +
           xlim(-.1,.1) + guides(color=FALSE)

density.stability.plot
centered.density.stability.plot
```

- Why are we getting a good estimate of the sampling distribution?
- The Binomial distribution is *continuous* as a function of $p$---when $p$ changes little, the distribution changes little.
- This means that, if we're not far off in the proportion, the estimated and actual sampling distributions are similar.
- The relevant difference (after centering) is even smaller because *the way* the binomial changes is mostly location.
- Here I'm showing three estimates of the sampling distribution based on three polls.
    - with centering (right) and without (left). 
- In particular, polls giving estimates at the 2.5th, 50th, and 97.5th percentiles of the actual sampling distribution. 

## The Estimate Works. Why Does it Work?
```{r}
#| out-height: 300px
#| fig-asp: .6 
#| layout-ncol: 2

density.stability.plot
centered.density.stability.plot
```

- You can think of this as a sort of 'confidence interval' for our estimate of the sampling distribution.
    - 95% of the time, you'll get an estimate somewhere between the [red]{.fg style="color:red"} and [blue]{.fg style="color:blue"} ones.
    - And, as a result, a width for your interval estimate somewhere between the [red]{.fg style="color:red"} and [blue]{.fg style="color:blue"} widths.
- As you can see, the widths we get barely change. What changes is the location.
- But because we're using calibrated 95% intervals, we still cover the population proportion 95% of the time.
    - i.e., we cover it when our sample proportion is between the [red]{.fg style="color:red"} and [blue]{.fg style="color:blue"} ones.
- Or, accounting for [slight underestimates]{.fg style="color:blue"} of the sampling distribution's width, almost cover it. 


# Appendix

## The Binomial Distribution for 3 Flips {#three-flips}

::::
::: {.column width="60%"}
$p$ | $Y_1$ | $Y_2$ | $Y_3$ | $Y_1 + Y_2 + Y_3$
---|---|---|---|---
$\inpollb{\theta_0^3}$ | 0 | 0 | 0 | [0]{.pollb}
$\inpollc{\theta_1 \theta_0^2}$          | 0 | 0 | 1 | [1]{.pollc}
$\inpollc{\theta_1 \theta_0^2}$          | 0 | 1 | 0 | [1]{.pollc}
$\inpollc{\theta_1 \theta_0^2}$          | 1 | 0 | 0 | [1]{.pollc}
$\inpolld{\theta_1^2 \theta_0}$          | 0 | 1 | 1 | [2]{.polld}
$\inpolld{\theta_1^2 \theta_0}$          | 1 | 1 | 0 | [2]{.polld}
$\inpolld{\theta_1^2 \theta_0}$          | 1 | 0 | 1 | [2]{.polld}
$\inpolle{\theta_1^3 }$                  | 1 | 1 | 1 | [3]{.polle}

:::
::: {.column width="40%"}

$p$  | $Y_1 + Y_2 + Y_3$
---|---
$\inpollb{\theta_0^3}$              | [0]{.pollb}
$\inpollc{3\theta_1 \theta_0^2}$    | [1]{.pollc}
$\inpolld{3\theta_1^2 \theta_0}$    | [2]{.polld}
$\inpolle{\theta_1^3 }$             | [3]{.polle} 

:::
::::
