---
title: "Lecture 12"
subtitle: "Multivariate Analysis and Regression Adjustment"
---

# Multivariate Analysis


{{< include shared-code.qmd >}}
{{< include ca-income.qmd >}}

```{r}
pop = pop.earnings 
sam = earnings 

xco = function(w,x) {
  w=as.numeric(w)
  x + .15*(2*w-1)
}

sam.dot = list(size=1,    alpha=.15)
pop.dot = list(size=.25,    alpha=.05)
scatter.by.group = function(data, dot=sam.dot, width=sd, 
                            highlight=NULL, 
                            highlight.color='yellow', highlight.alpha=.25,
                            group.means.alpha=NULL)
                            { 
  p = ggplot(data)
  if(!is.null(highlight)) {
    p = p + annotate('rect', 
                      xmin=min(highlight)-.5,
                      xmax=min(highlight)+.5,
                      ymin=-Inf, ymax=Inf, 
                      fill=highlight.color, alpha=highlight.alpha)
  } 
  if(!is.null(group.means.alpha)) {
    summaries = data |> group_by(group) |> summarize(income=mean(income))
    p = p + geom_hline(aes(yintercept=income, color=group), alpha=group.means.alpha, data=summaries) 
  }
  p + geom_point(aes(x=xco(group, education),  y=income, color=group), 
               size=dot$size, position=position_jitter(seed=0,w=.15), alpha=dot$alpha) +
      stat_summary(aes(x=xco(group, education), y=income, color=group), geom="pointrange", alpha=.7,
                 fun=mean, 
                 fun.min = function(x) mean(x) - width(x), 
                 fun.max = function(x) mean(x) + width(x))  +
    scale_x() + scale_y() + scale_alpha(range=c(0,1)) + guides(color='none')
}

scatter.by.sex.only = function(data, dot=sam.dot, width=sd) {
    ggplot(data) + 
      geom_point(aes(x=sex,  y=income, color=group), 
                 size=dot$size, position=position_jitter(seed=0,w=.15), alpha=dot$alpha) +
    stat_summary(aes(x=sex, y=income, color=group), geom="pointrange", alpha=.7,
                 fun=mean, 
                 fun.min = function(x) mean(x) - width(x), 
                 fun.max = function(x) mean(x) + width(x)) + 
    scale_y() + guides(color='none')
}

money = function(x) { sprintf('%2.1fk', round(x/1000, digits=1)) }  
```

## Multivariate Analysis

```{r}
sam$group = sam$sex == 'female'
scatter.by.group(sam)
```

```{r}
muhat.male = mean(sam$income[sam$sex=='male']) 
muhat.female = mean(sam$income[sam$sex=='female'])
```

- Today, we're going to talk about how to analyze data when we have multiple covariates.
- An important application is analyzing disparities to think about discrimination.
- Above, we see the kind of data we'll be working with.
- Each dot represents a California resident, age 25-35, who responded to the CPS.
  - $Y_i$, their income, is on the $y$-axis.
  - $X_i$, their education level, is on the $x$-axis.
  - $W_i$, their sex, is indicated by color. And a slight shift along the $x$-axis.
    - $\ingroupa{W_i=0}$ indicates male, $\ingroupb{W_i=1}$ indicates female.
- As usual, we've drawn in each group's income mean and standard deviation too.
- We'll write $\ingroupa{\hat\mu(0,x)}$ and $\ingroupb{\hat\mu(1,x)}$ for the mean income of male and female respondents with $x$ years of education. 


## A Few Summaries 

::: free-zoom

```{r}
#| fig-width: 7
#| layout-ncol: 2 
#| fig-cap: 
#|  - income vs. sex (aggregated)
#|  - income vs. education and sex (disaggregated)
scatter.by.sex.only(sam)
scatter.by.group(sam, highlight=12)
```

:::

```{r}
Y = sam$income
X = sam$education
W = sam$sex == 'female'
N1 = sum(W==1)
N0 = sum(W==0)
muhat = Vectorize(function(w,x) { mean(Y[W==w & X==x]) })
```


:::: columns
::: {.column width="48%"}
### Female Respondents
... earn \$`r money(muhat.male-muhat.female)` less, on average, than male respondents.
$$
\color{gray}
\begin{aligned}
\ingroupb{\hat\mu(1)} \ - \  \ingroupa{\hat\mu(0)}  
&\approx \ingroupb{`r money(muhat.female)`} \ - \ \ingroupa{`r money(muhat.male)`} \\
&\approx `r money(muhat.female - muhat.male)` 
\end{aligned}
$$
... earn `r 100*round(muhat.female/muhat.male, 2)` cents for every dollar a male one does.
$$
\color{gray}
\begin{aligned}
\ingroupb{\hat\mu(1)} \ / \ \ingroupa{\hat\mu(0)} 
&\approx \ingroupb{`r money(muhat.female)`} \ / \  \ingroupa{`r money(muhat.male)`} \\
&\approx {`r round(muhat.female/muhat.male, 2)`} \\
\end{aligned}
$$
:::
::: {.column width="52%"}
### ... with High School Educations 
... earn \$`r money(muhat(0, 12)-muhat(1, 12))` 
less than similarly-educated male ones.
$$
\color{gray}
\begin{aligned}
\ingroupb{\hat\mu(1, 12)} \ - \  \ingroupa{\hat\mu(0, 12)}  
&\approx \ingroupb{`r money(muhat(1, 12))`} \ - \ \ingroupa{`r money(muhat(0, 12))`} \\
&\approx `r money(muhat(1, 12)-muhat(0, 12))`
\end{aligned}
$$

... earn `r 100*round(muhat(1, 12)/muhat(0, 12), 2)` cents for every dollar a male one does.
$$
\color{gray}
\begin{aligned}
\ingroupb{\hat\mu(1, 12)} \ / \ \ingroupa{\hat\mu(0, 12)} 
&\approx \ingroupb{`r money(muhat(1, 12))`} \ / \  \ingroupa{`r money(muhat(0, 12))`} \\
&\approx {`r round(muhat(1, 12)/muhat(0, 12), 2)`} \\
\end{aligned}
$$
:::
::::

- These comparisons are relatively simple in that we're just reading two numbers off a plot.
- On the left, we compared aggregates---we used our plot of the aggregated data.
  - This is nice because the result is simple---we get one number.
- On the right, we compared fine-grained groups in the disaggregated data.
  - This is nice because it's more of a 'like-with-like' comparison.

##  What If We Want Simplicity *and* Comparability?

```{r}
scatter.by.group(sam)
```

**Exercise.** Describe how, using information in this plot, you would summarize the typical difference in income ...

  - between [female respondents]{.groupb} and **similarly educated** [male respondents]{.groupa}
  - using only one number.
  
  Use mathematical notation, e.g., $\ingroupa{\hat\mu(0, x)}$ and $\ingroupb{\hat\mu(1, x)}$ for within-group means.
  
  *Tip.*
  If you're having trouble with the notation, start by describing how you'd calculate it using the plot. 
  

## Solution 1. An Average over Levels of Education

```{r}
scatter.by.group(sam)
```

```{r}
years = sort(unique(sam$education))
years.str = paste(years, collapse=', ') 
```

$$
\color{gray}
\begin{aligned}
\text{summary} 
&= \frac{1}{`r length(years)`}\sum_{x \in \text{years}}  
\qty{ \ingroupb{\hat\mu(1,x)} - \ingroupa{\hat \mu(0, x)} } \qfor \text{years} =\{ `r years.str ` \}  \quad (`r length(years)` \text{ levels}) \\
&= \frac{\{\ingroupb{\hat\mu(1, 8)} - \ingroupa{\hat\mu(0, 8)}\} + \{\ingroupb{\hat\mu(1, 9)} - \ingroupa{\hat\mu(0, 9)}\} + \ \ldots \ + \{\ingroupb{\hat\mu(1, 20)} - \ingroupa{\hat\mu(0, 20)}\}}{`r length(years)`} \\
&\approx \frac{  \{ \ingroupb{`r money(muhat(1, 8))`}-\ingroupa{`r money(muhat(0, 8))`}\} \ + \ \{\ingroupb{`r money(muhat(1, 9))`}-\ingroupa{`r money(muhat(0, 9))`} \} + \ \ldots \ + \{ \ingroupb{`r money(muhat(1, 20))`}-\ingroupa{`r money(muhat(0, 20))`} \} }{`r length(years)`} \\
&\approx \frac{  \{`r money(muhat(1, 8) - muhat(0, 8))`\} + \{`r money(muhat(1, 9) - muhat(0, 9))`\} \ + \ldots \ + \{`r money(muhat(1, 20) - muhat(0, 20))`\}}{`r length(years)`} \\
&\approx `r money(mean(muhat(1, years) - muhat(0, years)))` 
\end{aligned}
$$

- This is, coincidentally, the same as the difference for respondents with 12 years of education.^[After rounding to the nearest \$100.]
- But is it a good summary of how much less female respondents are making? Probably not. 
  - Very few---only `r sum(X[W==1]==8)`---have 8 years of education.
  - Many---`r sum(X[W==1]==12)`---have 12 years of education.
  - And we're giving the same weight to the mean difference at each level of education.
- What we get isn't necessarily typical *for female respondents*.
- We've counted the ones with 8 years of education 
$`r sum(X[W==1]==12)`/`r sum(X[W==1]==8)` \approx `r round(sum(X[W==1]==12)/sum(X[W==1]==8))`$ times more than the ones with 12. 

## Solution 2. An Average over Respondents
```{r}
scatter.by.group(sam)
```

$$
\color{gray}
\begin{aligned}
\text{summary} 
&= \ingroupb{\frac{1}{N_1}\sum_{i: W_i=1}} \qty{ \ingroupb{\hat \mu(1, X_i) - \ingroupa{\hat \mu(0,X_i)} } }  \qfor N_w = \sum_{i: W_i=w} 1 \\
&= \frac{ \{ \ingroupb{\hat \mu(1, `r X[W==1][1]`)} - \ingroupa{\hat \mu(0, `r X[W==1][1]`)} \} + 
\{\ingroupb{\hat \mu(1, `r X[W==1][2]`)} - \ingroupa{\hat \mu(0, `r X[W==1][2]`)} \} + 
\ \ldots \ +
\{\ingroupb{\hat \mu(1, `r X[W==1][N1]`)} - \ingroupa{\hat \mu(0, `r X[W==1][N1]`)} \} }{`r N1`} \\
&\approx \frac{ \{\ingroupb{`r money(muhat(0, X[W==1][1]))`} - \ingroupa{`r money(muhat(1, X[W==1][1]))`}\} +
\{ \ingroupb{`r money(muhat(1, X[W==1][2]))`} - \ingroupa{`r money(muhat(0, X[W==1][2]))`} \} +
\ \ldots \ +
\{ \ingroupb{`r money(muhat(1, X[W==1][N1]))`} - \ingroupa{`r money(muhat(0, X[W==1][N1]))`}\}  }{`r N1`} \\
&\approx \frac{ \{`r money(muhat(1, X[W==1][1]) - muhat(0, X[W==1][1]))`\} +
\{`r money(muhat(1, X[W==1][2]) - muhat(0, X[W==1][2]))`\} +
\ \ldots \ +
\{`r money(muhat(1, X[W==1][N1]) - muhat(0, X[W==1][N1]))`\} }{`r N1`} \\
& \approx `r money(mean(muhat(1,X[W==1]) - muhat(0,X[W==1])))`
\end{aligned}
$$

- This is a comparison that focuses on what's typical *for our female respondents*.
- Each female respondent gets equal weight in 'deciding' what is typical. 
  - That's why $X_i=13$ occurs twice in the three (of `r N1`) terms we actually write out above.
  - It's a fairly common level of education.
    
## Reinterpreting Solution 2

```{r}
#| out-width: 100%
scatter.by.group(sam)
```

- We can, in fact, think of this as the average difference between the income of
  - each female respondent's income, $\ingroupb{Y_i}$ 
  - the mean income among males at their level of education, $\ingroupa{\hat \mu(0, X_i)}$.

$$
\color{gray}
\begin{aligned}
\text{summary} 
&= \ingroupb{\frac{1}{N_1}\sum_{i: W_i=1}} \qty{ \ingroupb{\hat \mu(1,X_i)} - \ingroupa{\hat \mu(0, X_i)} }  
= \ingroupb{\frac{1}{N_1}\sum_{i: W_i=1}} \qty{ \ingroupb{Y_i} - \ingroupa{\hat \mu(0, X_i)} }
\end{aligned}
$$

- These two expressions will always be equal, no matter what the actual values of $W_i,X_i,Y_i$ are.
- **Q.** Why are the equal?
- *Hint*. We've seen something like this this earlier in the semester.
  - Think about what happens when we change the second into the first. 
  - In our average over [female respondents]{.groupb}, ...
  - we replace each respondents' income $\ingroupb{Y_i}$ with their column's mean income $\ingroupb{\hat \mu(1, X_i)}$.

::: fragment

- **A.** We can think of the first one as a version of the second where we sum in a certain order. \

:::: columns

::: {.column width="50%"}
```{r}
#| fig-width: 7
scatter.by.group(sam)
```

[The same plot as above, for convenience.]{.fg style="color:gray"}
:::
::: {.column width="50%"}
$$
\color{gray}
\begin{aligned}
\sum_{i: W_i=w} Y_i 
&\explain{=}{We sum first over dots in a column, then over columns.}
\sum_{x} \sum_{\substack{i: W_i=w, \ X_i=x}} Y_i    \\
&\explain{=}{The column sum is the column's mean times its number of dots.}
\sum_{x} \hat \mu(w, x) \times N_{w,x} \qfor N_{w,x} = \sum_{\substack{i: W_i=w, \ X_i=x}} 1  \\
&\explain{=}{Or equivalently, its mean summed over the dots.}
\sum_{x} \sum_{\substack{i: W_i=w, \  X_i=x}} \hat \mu(w,x) \\ 
&\mathexplain{=}{\hat\mu(w,x)=\hat\mu(w,X_i) \text{ within a column}.} 
\sum_{x} \sum_{\substack{i: W_i=w, \ X_i=x}} \hat \mu(w, X_i) \\
&\explain{=}{We 'unorder' that sum to get back to the form we started with.} 
\sum_{i: W_i=w} \hat \mu(w, X_i)       
\end{aligned}
$$

:::
::::
:::

## Variations

- If we prefer, we could summarize this disparity by focusing on what's typical for *male respondents*.
  - Or for respondents in general.
  - Or for respondents who, e.g., have not attended college.
    
### Exercise

  - Here are a few summaries of the income disparity between similarly-educated male and female respondents.
  - They are all averages of the same [within-column]{.groupb} [differences]{.groupa}.
  - But they're averages over different groups of dots (i.e. respondents).
  - For each of these summaries, describe the group we're averaging over. Do it in two ways.
    1. As you'd describe the people in them, e.g. 'all female respondents'.
    2. As you'd describe the dots in the plot, e.g. 'all green dots' 

:::: columns  
::: column
```{r}
#| fig-width: 7
scatter.by.group(sam)
```

:::
::: column
$$
\color{gray}
\begin{aligned}
\Delta_1 
&\explain{=}{All male respondents. All red dots} 
\ingroupa{\frac{1}{N_0}\sum_{i: W_i=0}} \qty{ \ingroupb{\hat\mu(1,X_i)} - \ingroupa{\hat \mu(0, X_i)} } \\
\Delta_2 
&\explain{=}{All female respondents. All green dots} 
\ingroupb{\frac{1}{N_1}\sum_{i: W_i=1}} \qty{ \ingroupb{\hat\mu(1,X_i)} - \ingroupa{\hat \mu(0, X_i)} } \\
\Delta_3 
&\explain{=}{All respondents who didn't attend college. All dots in the leftmost 4 (pairs of) columns.} 
\frac{1}{N_{x \le 12}}\sum_{i: X_i \le 12} \qty{ \ingroupb{\hat\mu(1,X_i)} - \ingroupa{\hat \mu(0, X_i)} } \qfor N_{x \le 12} = \sum_{i: X_i \le 12} 1  \\
\Delta_4 
&\explain{=}{All respondents outright. All dots.} 
\frac{1}{n} \sum_{i=1}^n \qty{ \ingroupb{\hat\mu(1,X_i)} - \ingroupa{\hat \mu(0, X_i)} } 
\end{aligned}
$$

:::
::::

### Terminology

- It's common to use the same words to describe each of these summaries.
  - 'The average difference in income between male and female respondents, **adjusted for education**'.
  - Or sometimes '..., *controlling for education*'.^[This can be confusing because it has a different --- but related --- meaning when when we're talking about designing an experiment.]
- This results in a lot of confusion. 
- There is more precise terminology out there. 
  - But it tends to be used when we're talking about differences between groups *receiving different treatments*.
  - We'll get into that more precise terminology next time. 
  - And talk briefly about how it gets used in contexts like this, where we're not talking about treatments.
  - Today, when we talk about the adjusted difference, that'll mean the average over [female respondents]{.groupb}.

# Covariate Shift

## Comparisons with and without Adjustment

```{r}
#| fig-width: 7
#| layout-ncol: 2
#| fig-cap:
#|  - income vs. sex (aggregated)
#|  - income vs. education and sex (disaggregated)

scatter.by.sex.only(sam)
scatter.by.group(sam, group.means.alpha=.2)
```

:::: columns
::: {.column width="48%"}
### The Raw Difference
Female respondents earn \$`r money(muhat.male-muhat.female)` less, \
on average, than male respondents.
$$
\color{gray}
\ingroupb{\frac{1}{N_1}\sum_{i:W_i=1} Y_i} \ - \ \ingroupa{\frac{1}{N_0}\sum_{i:W_i=0} Y_i}  
\approx `r money(muhat.female - muhat.male)` 
$$

:::
::: {.column width="52%"}
### The Adjusted Difference
 \
And they earn `r money(mean(Y[W==1] - muhat(0,X[W==1])))` less than similarly-educated ones.
$$
\color{gray}
\ingroupb{\frac{1}{N_1}\sum_{i: W_i=1}}  \qty{ \ingroupb{Y_i} \ - \ \ingroupa{\hat \mu(0, X_i)} }
\approx `r money(mean(Y[W==1] - muhat(0,X[W==1])))` 
$$

:::
::::

- When people talk about comparisons like this, they're often pretty imprecise with their language.
- Here are a few things you might want to think about when hear about a comparison between two groups.
  1. What's being reported.
    - A raw difference or an adjusted one?
    - What variable(s) are being adjusted for? 
    - What group is being averaged over?
  2. How the things *they could report* might differ.
  3. Why they're reporting what they're reporting.
- 1 & 3 are very important skills, but you really need to practice them in context.
- Today we're going to focus on 2.
  - i.e., developing skills to help us understand and communicate about different comparisons ... 
  - that seem like they're answers to the same question but may not be.

::: aside
The averages shown in the plot on the left ---  the average incomes of all [male]{.groupa} and [female]{.groupb}
respondents --- are included as faint horizontal lines in the plot on the right.
:::


## What's the Difference?

```{r}
scatter.by.group(sam, group.means.alpha=.2)
```

:::: columns

::: {.column width="52%"}
### The Raw Difference

$$
\color{gray}
\ingroupb{\frac{1}{N_1}\sum_{i:W_i=1} Y_i} \ - \ \ingroupa{\frac{1}{N_0}\sum_{i:W_i=0} Y_i}  
\approx `r money(muhat.female - muhat.male)` 
$$

is what we get when we ...
  
::: {.incremental}
  1. Look at the height of each [red dot]{.groupa}.
  2. Subtract the average height of *all of the [green dots]{.groupb}*.
  3. Average these differences.
:::
:::

::: {.column width="48%"}
### The Adjusted Difference
$$
\color{gray}
\ingroupb{\frac{1}{N_1}\sum_{i: W_i=1}}  \qty{ \ingroupb{Y_i} \ - \ \ingroupa{\hat \mu(0, X_i)} }
\approx `r money(mean(Y[W==1] - muhat(0,X[W==1])))` 
$$

is what we get when we ...

  1. Look at the height of each [red dot]{.groupa}.
  2. Subtract the average height of the \
  [green dots]{.groupb} in the same column.
  3. Average these differences.
:::
::::


 - **Q**. Why is the adjusted difference is bigger in magnitude (i.e. smaller, being negative)?
 - *Hint*. $\color{gray} \text{same} - \text{bigger} = \text{smaller}$
   
::: fragment
 - **A.** For most [female respondents]{.groupa}, the average [male]{.groupb} income *at their education level* \
 is bigger than the average for *all male respondents*. Let's see if we can work out why.
:::

## Covariate Shift

:::: r-stack
::: {.fragment .fade-out data-fragment-index="0"} 
```{r}
scatter.by.group(sam) 
```
:::
::: {.fragment .current-visible data-fragment-index="0"}
```{r}
scaled.interleaved.bars = list(geom_bar(aes(x=xco(group, education), y=300000*after_stat(prop), fill=group, color=group), 
                        position='identity', alpha=.05, linewidth=.4), 
                        guides(color='none', fill='none'))
interleaved.bars = list(geom_bar(aes(x=xco(group, education), y=after_stat(prop), fill=group, color=group), 
                        position='identity', alpha=.05, linewidth=.4), 
                        guides(color='none', fill='none'))
bars= list(geom_bar(aes(x=education, y=after_stat(prop), fill=group, color=group), 
                     position='identity', alpha=.05, linewidth=.4), 
            guides(color='none', fill='none'))

scatter.by.group(sam) + scaled.interleaved.bars
```
:::
::::

- This happens because the [male respondents]{.groupa} tend to be less educated than the [female ones]{.groupb}.
  - We call this phenomenon **covariate shift**. 
  - The distribution of $X$ *shifts* when we go from looking at one group (i.e. sex) to another.
- We can see this shift --- with a lot of effort --- by counting the dots in each column.
- But it's easier to see if we draw that information into the plot. &nbsp; [Do it!]{.button .navigate-next} 
- We draw in bars showing the proportion of people, within a group, that have each level of education. 
    - The bars' absolute height is meaningless---they're scaled to fit into the plot.
    - But the relative height tells us how education is distributed within each group.
    - The [red]{.groupa} and [green]{.groupb} bars show how education is distributed among [male]{.groupa} and [female]{.groupb} respondents.
    - *Notation.* We write $\ingroupa{\hat p_{0}(x)}$ and $\ingroupb{\hat p_{1}(x)}$ for these proportions at education level $x$.
- And we see that the [green bars]{.groupb} tend to be higher than the [red bars]{.groupa} on the right side of the plot, esp. for $x > 12$.
  - *Interpretation*. Higher levels of education are more common for female respondents.
  - This can be easier to see in different visualizations of the covariate distributions.
  
### Another Visualization of Covariate Shift

```{r}
facet.by.group = list(facet_grid(row=vars(group)), theme(strip.background=element_blank(), strip.text=element_blank()))
ggplot(sam) + bars + facet.by.group
``` 

- This is exactly the same information we had overlaid on the previous plot. They're exactly the same bars.
- What's changed is *layout*. Our groups are shown one above the other rather than interleaved as before.
- This makes it easier to see that the [red one]{.groupa} has more mass *on the left* side than the [green one]{.groupb}.


## Covariate Shift *reduces* this Disparity
::: free-zoom
```{r}
#| layout-ncol: 2
#| fig-width: 7
scatter.by.group(sam) + scaled.interleaved.bars
ggplot(sam) + bars + facet.by.group
```
:::

- Our plot shows that, at virtually all levels of education, male respondents out-earn female ones.
- And the typical magnitude of this disparity is something like \$15k.
  - Sometimes it's about \$20k (e.g. at 11, 14, and 18 years of education)
  - Sometimes it's about \$10k (e.g. at 16 years)
- That's summarized reasonably by the adjusted difference, \$`r money(abs(mean(Y[W==1] - muhat(0,X[W==1]))))`.
  The raw difference, \$`r money(abs(mean(Y[W==1]) - mean(Y[W==0])))`, is much smaller. 
- So we're seeing that covariate shift *reduces* the disparity --- it makes it smaller than what we see within levels.
- That's two factors working together.
  1. *A Downward Shift*. Male respondents tend to have less education than female ones.
  2. *An Upward Trend*.   Income tends be higher at higher levels of education.
- The relationship between the raw and adjusted difference depends on both. 

## Covariate Shift *increases* this other Disparity
```{r}
adj.mf.ca = mean(Y[W==1] - muhat(0,X[W==1]))
raw.mf.ca = mean(Y[W==1]) - mean(Y[W==0])
```

```{r}
#| cache: true
georgia.state.code = 13
ga.sam = income(georgia.state.code)$sample
```

```{r}
#| out-width: 100%
this.sam = ga.sam |> mutate(group = race == 'Black only') |> filter(race %in% c('Black only', 'White only'))

W = this.sam$group
X = this.sam$education
Y = this.sam$income
N1 = sum(W==1)
N0 = sum(W==0)
muhat = Vectorize(function(w,x) { mean(Y[W==w & X==x]) })
```

::: free-zoom
```{r}
#| layout-ncol: 2
#| fig-width: 7
scatter.by.group(this.sam) + scaled.interleaved.bars
ggplot(this.sam) + bars + facet.by.group
```
:::


- Here we're making the same comparison, but for [Black]{.groupb} and [white]{.groupa} respondents in Georgia.
- The within-group differences are similar (using the same dot colors) to the ones we've just looked at.
  - The trend is still upward --- more education means more income.
  - And [Black]{.groupb} respondents make less at most levels of education, esp. the common ones.
- But the covariate shift is in the opposite direction.
  - [Black]{.groupb} respondents tend to have fewer years of education than [white ones]{.groupa}. 
- As a result, the raw difference is *larger* than the adjusted one. Covariate shift *increases* the disparity.
  - The adjusted difference is about \$`r money(abs(mean(Y[W==1] - muhat(0,X[W==1]))))`. The raw one is about \$`r money(abs(mean(Y[W==1]) - mean(Y[W==0])))`.
- This is a bigger shift --- and in the opposite direction --- compared to what we saw in the previous example.
  - That was \$`r money(abs(adj.mf.ca))` adjusted &rarr; \$`r money(abs(raw.mf.ca))` raw.  

## Sometimes there isn't much Covariate Shift

```{r}
this.sam = sam 
this.sam$group = (\() runif(nrow(this.sam)) <= .5) |> with_seed(10)

W = this.sam$group
X = this.sam$education
Y = this.sam$income
N1 = sum(W==1)
N0 = sum(W==0)
muhat = Vectorize(function(w,x) { mean(Y[W==w & X==x]) })
```

::: free-zoom
```{r}
#| layout-ncol: 2
#| fig-width: 7
scatter.by.group(this.sam) + scaled.interleaved.bars
ggplot(this.sam) + bars + facet.by.group
```
:::

- We're back in California comparing respondents in two *randomly assigned* groups.
  - I flipped a coin for each to decide whether they're [red]{.groupa} or [green]{.groupb}.
- Because of randomization, there isn't really any covariate shift.
  - The distribution of education is the same in both groups. The bars are the same height.
  - As a result, the raw disparity (\$`r money(abs(mean(Y[W==1] - muhat(0,X[W==1]))))`) is 
  about the same as the adjusted one (\$`r money(abs(mean(Y[W==1]) - mean(Y[W==0])))`).
- There's also little difference in mean income --- at any level --- between the groups. That's why both are about \$0.
- But if we did something as a result of the coin flip, e.g. gave one group a treatment, that might *induce* a difference.
- Next class, we'll see how randomization can help us estimate this induced difference---the treatment's effect.


## Sometimes Covariate Shift is Complicated
```{r}
this.sam = sam |> filter(race %in% c('Black only', 'White only')) |> 
                  mutate(group = race == 'Black only') 

W = this.sam$group
X = this.sam$education
Y = this.sam$income
N1 = sum(W==1)
N0 = sum(W==0)
muhat = Vectorize(function(w,x) { mean(Y[W==w & X==x]) })
```

::: free-zoom
```{r}
#| layout-ncol: 2
#| fig-width: 7
scatter.by.group(this.sam) + scaled.interleaved.bars
ggplot(this.sam) + bars + facet.by.group
```
:::

- Now we're comparing [Black]{.groupb} and [white]{.groupa} respondents in California.
- Here we see a more complicated pattern of covariate shift---a shift toward the center.
  - [Black]{.groupb} respondents tend to have *less extreme* levels of education than [white ones]{.groupa}.
  - Visually, the [green bars]{.groupb} are higher in the middle than the [red ones]{.groupa}.
- In a sense, the  rightward shift on the left side compensates for a leftward shift on the left side.
- As a result, we don't see a big difference in the raw and adjusted differences.
  - The adjusted difference is about \$`r money(abs(mean(Y[W==1] - muhat(0,X[W==1]), na.rm=TRUE)))` 
  - The raw difference is about \$`r money(abs(mean(Y[W==1]) - mean(Y[W==0])))`.

## What Would You Report?

```{r}
this.sam = ga.sam |> mutate(group = race == 'Black only') |> filter(race %in% c('Black only', 'White only'))

W = this.sam$group
X = this.sam$education
Y = this.sam$income
N1 = sum(W==1)
N0 = sum(W==0)
muhat = Vectorize(function(w,x) { mean(Y[W==w & X==x]) })
```

::: {.panel-tabset}
### California
```{r}
scatter.by.group(sam) + scaled.interleaved.bars
```
### Georgia
```{r}
scatter.by.group(this.sam) + scaled.interleaved.bars
```
:::

### Exercise

*Context*. You're writing an article about income disparities in the US. 

1. Flip a coin. \

  - **Tails**. It's on the disparity between [female]{.groupb} and [male]{.groupa} residents of California.  
  - **Heads**. It's on the disparity between [Black]{.groupb} and [white]{.groupa} residents of Georgia.
    
2. You're going to include one summary in your headline---either the raw or the adjusted difference. Decide which.

3. Write down, in a sentence or two, why you chose what you did.

## Thoughts on the Exercise

::: {.panel-tabset}
### California
```{r}
scatter.by.group(sam) + scaled.interleaved.bars
```
### Georgia
```{r}
scatter.by.group(this.sam) + scaled.interleaved.bars
```
:::


- There's no right answer---either in general or in any particular case.
- And  reasons may be context-specific, e.g.  
- Theory beyond ...; stats is about measurement, whereas your decision might be about ...
- fsdf


## Plan for the rest of class

1. How extreme the impact of covariate shift can be.
2. Statistical Inference.
  - We've focused on an income disparity among *CPS respondents* living in California. 
    - That's a sample.
    - We're probably not particularly interested in that.
  - What we really want to know about is the disparity among *all residents* of California.
    - That's the population we've sampled from.
    - And we can use our sample to learn about it.
  - This'll be quick because it's *exactly* the same as what we've been doing.
3. Practice with Visualization and Explanation
4. What to Adjust For.
  - More than One Thing?
  - Matching 
  - Audit Studies 


# Testing the Limits
## A Famously Paradoxical Example
```{r}
Y = c(rep(c(0,1), c(4704, 3738)), 
      rep(c(0,1), c(2827, 1494)))
W = c(rep(0, 4704+3738), rep(1, 2827+1494))

berkeley = data.frame(admit=Y, sex=factor(c('male', 'female'), levels=c('male','female'))[W+1])

dot = pop.dot
ggplot(berkeley) + 
      geom_point(aes(x=sex,  y=admit, color=sex), 
                 size=dot$size, position=position_jitter(seed=0,w=.15,h=.075), alpha=dot$alpha) +
    stat_summary(aes(x=sex, y=admit, color=sex), geom="pointrange", alpha=.7,
                 fun=mean, 
                 fun.min = function(x) mean(x) - width(x), 
                 fun.max = function(x) mean(x) + width(x)) + guides(color='none') + 
                 coord_cartesian(ylim=c(-.1,1.1)) + 
                 scale_y_discrete(breaks=c(0,1), labels=c('Rejected', 'Admitted'))
```


- Let's shift our attention from income today to graduate school admissions in the 1970s.
- Berkeley graduate programs were admitting a much higher proportion of [male applicants]{.groupa} than [female ones]{.groupb}.
  - $\ingroupa{`r sum(Y[W==0])`/`r sum(W==0)`\approx `r round(100*mean(Y[W==0]))`}$% of male applicants were admitted
  vs.  $\ingroupb{`r sum(Y[W==1])`/`r sum(W==1)`\approx `r round(100*mean(Y[W==1]))`}$% of female ones. 
  - People were concerned that this difference was evidence of discrimination against female applicants.
$$
\color{gray}
\text{raw difference} = \ingroupb{\frac{1}{N_1}\sum_{i: W_i=1} Y_i} - \ingroupa{\frac{1}{N_0}\sum_{i: W_i=0} Y_i} \approx `r round(mean(Y[W==1]) - mean(Y[W==0]), 2)`
$$

- But, on closer examination, most departments were admitting a larger proportion of their female applicants than of their male ones.
- What's going on?  Let's see if we can figure it out by working with some stylized data.


## 

```{r}
x = c(1,1,1,0,0,0,  0,1,1)
y = c(1,1,1,0,0,0,  1,0,1)
w = c(0,0,1,0,1,1,  1,0,0)   
jitter.y = .07*c(0,0,0,0,0,0,    0,0,-1)
jitter.x = .035*c(-1,0,1,-1,0,1,  0,0,0)  
data = data.frame(x=x, jitter.x=jitter.x, y=y, jitter.y=jitter.y, w=w)

exercise.grid = function(data) { 
  ggplot(data) + geom_point(aes(x=x+jitter.x, y=y+jitter.y), shape=1, size=8) +
    scale_y_continuous(breaks=c(0, 1/3, 2/3, 1), labels=c('0', '1/3', '2/3', '1')) + scale_x_continuous(breaks=c(0,1))
}

solution.grid = function(data) { 
  exercise.grid(data) +
    geom_point(aes(x=x+jitter.x, y=y+jitter.y, color=factor(w)), size=8, alpha=.5, data=data) + 
    stat_summary(aes(x=x, y=y, color=factor(w)), geom="point", size=12, shape=0, data=data) +
    geom_hline(aes(yintercept=mean.y, color=factor(w)), 
      data=data |> group_by(w) |> summarize(mean.y=mean(y))) + 
    guides(color='none')
}
```

:::: free-zoom 

```{r}
#| fig-asp: .5
#| fig-width: 10

exercise.grid(data[1:6,])
```

- Above is a stylized version of the Berkeley data in which we have two departments.
  - At the department on the left ($x=0$),  getting in is hard.
  - At the department on the right ($x=1$), getting in is easy.
- We'll draw dots for each applicant. 
  - The $y$-coordinate indicates whether they were admitted ($y=1$) or not ($y=0$).

::: {.panel-tabset}

### Warm-up Exercise
- Color three dots red and three green so that ...
  1. Within each column, the mean of the green dots is *equal to* the mean of the red dots.
  2. The mean of all the green dots is *lower* than the mean of the red ones. 
- In red and green ...
  1. Draw squares to show the within-column means of that color's dots.
  2. Draw a horizontal line to show the mean of all the dots of that color.
    
### Follow-up Exercise
- Add three more dots --- two red and one green --- so that ...
  1. Within each column, the mean of the green dots is *now higher* the mean of the red ones. 
  2. The mean of all the red dots is *still lower* than the mean of the red ones. 
- Move the squares and lines so these new dots are included.

### Warm-up Solution

```{r}
#| fig-asp: .5
#| fig-width: 10
solution.grid(data[1:6,])
```

### Follow-up Solution

```{r}
#| fig-asp: .5
#| fig-width: 10
solution.grid(data)
```


:::
::::

## Simpson's Paradox

```{r}
#| layout-ncol: 2
#| fig-width: 7

solution.grid(data)
ggplot(data) + geom_bar(aes(x=x, y=after_stat(prop), fill=factor(w), color=factor(w)), 
                     position='identity', alpha=.05, linewidth=.4) +
              guides(color='none', fill='none') +
              facet_grid(rows=vars(factor(w))) +
              theme(strip.background=element_blank(), strip.text=element_blank())
```


- What we're seeing here is covariate shift so extreme that it flips the disparity.




# Stuff

## Yup

```{r}
filtered = \(rows) rows[rows$income < 5e5 & rows$race %in% c('White only', 'Black only'),]
  
pop = pop.earnings |> filtered()
sam = earnings |> filtered()
```

```{r}
xco = function(w,x) {
  w=as.numeric(w)
  x + .15*(2*w-1)
}

sam.dot = list(size=1,    alpha=.2)
pop.dot = list(size=.25,    alpha=.05)

scatter.by.race = function(rows, 
                           dot=sam.dot,
                           width=sd) { 
  ggplot(rows[rows$race %in% c('Black only', 'White only'), ]) + 
    geom_point(aes(x=xco(race=='White only',education), y=income, color=race), 
               size=dot$size, position=position_jitter(seed=0,w=.15), alpha=dot$alpha) + 
    stat_summary(aes(x=xco(race=='White only',education), y=income, color=race), geom="pointrange", alpha=.7,
                 fun=mean, 
                 fun.min = function(x) mean(x) - width(x), 
                 fun.max = function(x) mean(x) + width(x))  +
    scale_x() + scale_y() + guides(color='none')
}
```

```{r}
#| out-height: 350px
#| fig-asp: .4

W = as.numeric(sam$race == 'White only')
X = sam$education
Y = sam$income 
muhat = Vectorize(function(w,x) { mean(Y[W==w & X==x]) })

a = money
scatter.by.race(sam)
```

- Let's take a look at at income ($y$) broken down in terms of two variables.
  1. Education ($x$), in years of schooling, is indicated by column in the plot.
  2. Race ($w$) is indicated by color in the plot.  [$w=0$ for Black people]{.twocolor-red}  and [$w=1$ for white people]{.twocolor-green} in our sample.
- To keep things simple, we'll look at the subset of CPS recipients who identify as either [Black only]{.twocolor-red} or [white only]{.twocolor-green}.
- At most education levels, the [white people]{.twocolor-green} in our sample make more money than the [Black people]{.twocolor-red}.
- We know that's a thing. If we want, we can report a summary or two.

::: {#def-conditional-proportion}
$$
\hat p_{w}(x) = \frac{N_{w,x}}{N_w} \qqtext{ where }  \frac{N_{w,x}}{N_w} = \frac{\sum_{i: W_i=w, X_i=x} 1}{\sum_{i: W_i=w} 1}.
$$ 
:::

$
\begin{array}{lll}
\text{raw income gap} \quad & \quad \text{adjusted income gap} & \\
\displaystyle \hat \Delta = \ingreen{\frac{1}{N_1}\sum_{i:W_i=1} Y_i} - \inred{\frac{1}{N_0}\sum_{i:W_i=0} Y_i} 
\quad& \displaystyle \hat \tau = \inred{\frac{1}{N_0}\sum_{i:W_i=0}} \ingreen{\hat \mu(1, X_i)} - \inred{\frac{1}{N_0}\sum_{i:W_i=0}} \inred{Y_i}
\quad& \text{symbolically} \\
        `{r} money(mean(Y[W==1]) - mean(Y[W==0]))`  
        = \ingreen{`{r} a(mean(Y[W==1]))`} - \inred{`{r} a(mean(Y[W==0]))`}   
\quad&  `{r} a(mean(muhat(1, X[W==0]) - muhat(0, X[W==0])))` 
       = \ingreen{`{r} a(mean(muhat(1, X[W==0])))`} - \inred{`{r} a(mean(muhat(0, X[W==0])))`}   
\quad& \text{numerically}
\end{array} 
$$


## Statistical Inference

```{r}
W = as.numeric(sam$race == 'White only')
X = sam$education
Y = sam$income 
muhat = Vectorize(function(w,x) { mean(Y[W==w & X==x]) })
the.adjusted.gap = mean(muhat(1, X[W==0]) - mean(muhat(0, X[W==0])))
the.raw.gap = mean(Y[W==1]) - mean(Y[W==0])

adjusted.gap = \(people) people |> group_by(education, race) |> 
                    summarize(muhat=mean(income), N=n(), .groups='drop') |> 
                  group_by(race) |>
                    mutate(P = N/sum(N)) |>
                   group_by(education) |>
                    group_map(\(s,e) s$P[s$race=='Black only'] * 
                    (s$muhat[s$race=='White only'] - s$muhat[s$race=='Black only'])) |>
                  unlist() |> sum()
raw.gap = \(people) mean(people$income[people$race=='White only']) - 
                    mean(people$income[people$race=='Black only'])
                    
stopifnot(abs(the.adjusted.gap - adjusted.gap(sam)) < 1e-6)
stopifnot(abs(the.raw.gap - raw.gap(sam)) < 1e-6)

set.seed(0)
B = 200
bootstrap.gaps = \(sam) 1:B |> map(\(B) {
  bootsam = sam[sample(1:nrow(sam), replace=TRUE), ]
  list(raw=raw.gap(bootsam), adjusted=adjusted.gap(bootsam))
}) |> bind_rows()
```

```{r}
#| cache: true
ca.bootstrap.gaps = bootstrap.gaps(sam)
```

```{r}
#| out-height: 350px
#| fig-asp: .5
#| layout-ncol: 2 
#| fig-cap: 
#|  - "Sample"
#|  - "Imagined Population" 
scatter.by.race(sam)
scatter.by.race(pop, dot=pop.dot)
```

- As usual, we can interpret these sample summaries as estimates of corresponding population summaries. 
- And we can construct 95% confidence intervals as usual---by bootstrapping. Nothing new here.
- What I want to talk about today is what these summaries mean and how they differ.
  - Incidentally, in California the raw and adjusted gaps are very similar.
  - But they are conceptually different and can be very different elsewhere. Let's look at Georgia.

$$
\begin{array}{lll}
\text{raw income gap} \quad & \quad \text{adjusted income gap} & \\
\displaystyle \hat \Delta = \ingreen{\frac{1}{N_1}\sum_{i:W_i=1} Y_i} - \inred{\frac{1}{N_0}\sum_{i:W_i=0} Y_i} 
\quad& \displaystyle \hat \tau = \inred{\frac{1}{N_0}\sum_{i:W_i=0}} \ingreen{\hat \mu(1, X_i)} - \inred{\frac{1}{N_0}\sum_{i:W_i=0}} \inred{Y_i}
\quad& \text{in sample} \\
\displaystyle \Delta = \ingreen{\frac{1}{m_1}\sum_{j:w_j=1} y_j} - \inred{\frac{1}{m_0}\sum_{j:w_j=0} y_j} 
\quad&\displaystyle \tau = \inred{\frac{1}{m_0}\sum_{j:w_j=0}} \ingreen{\mu(1, x_j)} - \inred{\frac{1}{m_0}\sum_{j:w_j=0}} \inred{y_j}
\quad& \text{in population} \\
        `{r} a(mean(Y[W==1]) - mean(Y[W==0]))` \pm `r a(2*sd(ca.bootstrap.gaps$raw))`
\quad&  `{r} a(mean(muhat(1, X[W==0]) - muhat(0, X[W==0])))` \pm `r a(2*sd(ca.bootstrap.gaps$adjusted))` 
\quad& \text{in 95\% of surveys}
\end{array} 
$$


## What Happens in Georgia 
```{r}
#| cache: true
georgia.state.code = 13
ga.sam = income(georgia.state.code)$sample
ga.bootstrap.gaps = bootstrap.gaps(ga.sam)
```

```{r}
#| out-height: 350px
#| fig-asp: .4
scatter.by.race(ga.sam)
```

- In Georgia, they're different. What are these two summaries?
  1. **Raw**. How much more the average white person makes than the average Black person.
  2. **Adjusted**. How much more, *for the average Black person*, a similarly-educated white person tends to make.
- The adjusted gap explains *part of* the raw gap. But not all of it.
- What's left over is, conceptually, about differences in the distribution of education.

$$
\begin{array}{lll}
\text{raw income gap} \quad & \quad \text{adjusted income gap} & \\
\displaystyle \hat \Delta = \ingreen{\frac{1}{N_1}\sum_{i:W_i=1} Y_i} - \inred{\frac{1}{N_0}\sum_{i:W_i=0} Y_i} 
\quad& \displaystyle \hat \tau = \inred{\frac{1}{N_0}\sum_{i:W_i=0}} \ingreen{\hat \mu(1, X_i)} - \inred{\frac{1}{N_0}\sum_{i:W_i=0}} \inred{Y_i}
\quad& \text{in sample} \\
\displaystyle \Delta = \ingreen{\frac{1}{m_1}\sum_{j:w_j=1} y_j} - \inred{\frac{1}{m_0}\sum_{j:w_j=0} y_j} 
\quad&\displaystyle \tau = \inred{\frac{1}{m_0}\sum_{j:w_j=0}} \ingreen{\mu(1, x_j)} - \inred{\frac{1}{m_0}\sum_{j:w_j=0}} \inred{y_j}
\quad& \text{in population} \\
        `{r} a(raw.gap(ga.sam))` \pm `r a(2*sd(ga.bootstrap.gaps$raw))`
\quad&  `{r} a(adjusted.gap(ga.sam))` \pm `r a(2*sd(ga.bootstrap.gaps$adjusted))` 
\quad& \text{in 95\% of surveys}
\end{array} 
$$


## Revisiting A Computational Trick

```{r}
#| out-height: 350px
#| fig-asp: .4
#| layout-ncol: 2 
#| fig-cap:
#|  - "Georgia"
#|  - "California"

scatter.with.hist = scatter.by.race(ga.sam) +
  geom_bar(aes(x=xco(race=='White only', education), y=400000*after_stat(prop), fill=race, color=race), 
                 position='identity', alpha=.05, linewidth=.4) + guides(fill='none') + scale_y(c(0,150000))
scatter.with.hist
scatter.by.race(sam) +
  geom_bar(aes(x=xco(race=='White only', education), y=400000*after_stat(prop), fill=race, color=race), 
                 position='identity', alpha=.05, linewidth=.4) + guides(fill='none') + scale_y(c(0,150000))
```

- To make sense of our two summaries, it's helpful to revisit a counting trick we used before.
  - When we want to average over the people in a sample, we can use a histogram to help us count.
  - This reduces a sum over hundreds or thousands of people to a sum over a few bins.

$$
\begin{aligned}
\hat \Delta &= \sum_{x} \qty{ \ingreen{P_{1,x} \hat\mu(1,x)} \quad - \quad \inred{P_{0,x} \ \hat\mu(0,x)} }            && \qqtext{where} \ P_{w,x} = \frac{N_{w,x}}{N_w} \\
\hat \tau   &= \sum_{x} \qty{ \inred{P_{0,x}} \ \ingreen{\hat\mu(1,x)} \quad - \quad \inred{P_{0,x} \ \hat\mu(0,x)} }    && \qqtext{for} \ \ \ N_w = \sum_{i:W_i=w} 1 \\ 
\hat \Delta - \hat \tau &= \sum_{x} \qty{\ingreen{P_{1,x}} - \inred{P_{0,x}}} \ingreen{\hat\mu(1,x)}                         && \qqtext{and} \ \ \ N_{w,x} = \sum_{i:W_i=w, X_i=x} 1
\end{aligned}
$$

- Looking at what happens in each bin can help us understand what's going on.
- Why were our raw and adjusted gaps similar in California but different in Georgia?
  

## Derivation: The Raw Gap
```{r}
#| out-height: 350px
#| fig-asp: .4
scatter.with.hist
```

$$
\begin{aligned}
\hat \Delta &= \frac{1}{N_1}\sum_{i:W_i=1} Y_i \quad -  \quad \frac{1}{N_0}\sum_{i:W_i=0} Y_i  && \qfor  N_{w} = \sum_{i:W_i=w} 1\\
            &= \sum_{x} \frac{1}{N_1}\sum_{i:W_i=1, X_i=x} Y_i   \quad - \quad \sum_{x}\frac{1}{N_0}\sum_{i:W_i=0, X_i=x} Y_i && \qqtext{summing within then over bins} \\
            &= \sum_{x} \frac{1}{N_1} \ N_{1,x} \ \hat\mu(1,x)  \quad - \quad \sum_{x} \frac{1}{N_0} \ N_{0,x} \ \hat\mu(0,x)     && \qfor N_{w,x} = \sum_{i:W_i=w, X_i=x} 1  \\
            &= \sum_{x} \ingreen{P_{1,x} \ \hat\mu(1,x)} \quad - \quad \sum_x \inred{P_{0,x} \ \hat\mu(0,x)}  && \qfor P_{w,x} = \frac{N_{w,x}}{N_w} \\
\end{aligned}
$$

## Derivation: The Adjusted Gap

```{r}
#| out-height: 350px
#| fig-asp: .4
scatter.with.hist
```


## Derivation: The Adjusted Gap
```{r}
#| out-height: 350px
#| fig-asp: .4
scatter.with.hist
```

$$
\begin{aligned}
\hat \tau &= \inred{\frac{1}{N_0}\sum_{i:W_i=0}} \ingreen{\hat \mu(1, X_i)} - \inred{\frac{1}{N_0}\sum_{i:W_i=0} Y_i} \\
          &= \sum_{x} \frac{1}{N_0}\sum_{i:W_i=1, X_i=x} \hat\mu(1,x)   - \sum_{x}\frac{1}{N_0}\sum_{i:W_i=0, X_i=x} Y_i && \qqtext{summing within then over bins} \\
          &= \sum_{x} \frac{1}{N_0} \ N_{0,x} \ \hat\mu(1,x)  - \sum_{x} \frac{1}{N_0} \ N_{0,x} \ \hat\mu(0,x)             && \qfor N_{w,x} = \sum_{i:W_i=w, X_i=x} 1  \\
          &= \sum_{x} \ingreen{P_{1,x} \ \hat\mu(1,x)} \quad - \quad \sum_x \inred{P_{0,x} \ \hat\mu(0,x)}  && \qfor P_{w,x} = \frac{N_{w,x}}{N_w} \\ \\
\end{aligned}
$$

## What To Adjust For?

```{r}
#| out-height: 350px
#| fig-asp: .4
#| layout-ncol: 2 
#| fig-cap:
#|  - "Georgia"
#|  - "California"

scatter.with.hist = scatter.by.race(ga.sam) +
  geom_bar(aes(x=xco(race=='White only', education), y=400000*after_stat(prop), fill=race, color=race), 
                 position='identity', alpha=.05, linewidth=.4) + guides(fill='none') + scale_y(c(0,150000))
scatter.with.hist
scatter.by.race(sam) +
  geom_bar(aes(x=xco(race=='White only', education), y=400000*after_stat(prop), fill=race, color=race), 
                 position='identity', alpha=.05, linewidth=.4) + guides(fill='none') + scale_y(c(0,150000))
```

- If we're trying to think about discrimination, it's not entirely clear what we should adjust for.
  - By looking at the adjusted gap, we're trying to make the people we compare 'similarly qualified'
  - But discrimination also *causes* differences in education.
  - So we're *adjusting away* some of the effect we might want to measure.
- We can try to formalize our comparisons in terms of potential outcomes. 
- We'll imagine that for each person in our population ... 
  -  there's one income they'd have if they were 'treated' by being raced white.
  -  and another income they'd have if they were 'treated' by being raced Black.

$$
\begin{aligned}
y_j(1) & \text{ income person $j$ would have if they were 'treated' by being raced white} \\
y_j(0) & \text{ income person $j$ would have if they were 'treated' by being raced Black}
\end{aligned}
$$


## What Do We Do With These Potential Outcomes?

$$
\begin{aligned}
y_j(1) & \text{ income person $j$ would have if they were 'treated' by being raced white} \\
y_j(0) & \text{ income person $j$ would have if they were 'treated' by being raced Black}
\end{aligned}
$$

- The individual effect of this treatment is the difference between these two.
- We can summarize this by the average of these differences for the Black people in our population.

$$
\begin{aligned}
\text{individual effect}_j 
&= y_j(1) - y_j(0) \\
\text{average effect} 
&= \frac{1}{m_0}\sum_{j:w_j=0} \qty{ y_j(1) - y_j(0) }  = \frac{1}{m_0}\sum_{j:w_j=0} y_j(1) - \frac{1}{m_0}\sum_{j:w_j=0} y_j(0) 
\end{aligned}
$$

- And $y_j(1)$ being unobservable for the people in this average, we try to find some proxy for it.
  - The adjusted-gap approach is to use the average income of white people with the same education level.
  - But what's the justification for this? Or something like this? There really isn't one. 
  - We've just disguised a fundamental problem in mathematical notation that works in other contexts. 
- You could argue that the problem is this 'treatment' isn't something you can assign, e.g., by randomization.
- But that's not really the problem. The problem is that the potential outcomes don't make sense.
- What does it mean to be person $j$,  except raced differently? It's meaningless.

## A Non-Solution --- Audit Studies

- People try to work around this by reframing the question in terms of *perception* of race. 
  - Because that's manipulable. You can, for example, choose the name on a job application.
  - You can randomize job applications to be 'Black' or 'white' by flipping a coin...
  - and then, on the basis of that, selecting the name of a Black or white census respondent at random.
- You can then mail these out and see who gets called back for an interview.

$$
\begin{aligned}
y_j(1) & \text{ whether application $j$ would get an interview when using a `random white name'} \\
y_j(0) & \text{ whether application $j$ would get an interview when using a `random Black name'}
\end{aligned}
$$

- Fundamentally, this is a version of our adjusted gap, but with more levels of $X$---one per application.
- The difference in interview rates is the called the 'effect of race' on *the decision-maker's actions*.
- But really, it's just the effect of a specific race signifier on the decision-maker's actions. 
  - If the idea of person $j$ except raced differently means nothing in the actual world...
  - ... then why do we imagine that it means something in the decision-maker's head?
- That doesn't mean that studies like this are useless. 
  - But we shouldn't act as if they get around the fundamental problem.
  - i.e., as if it works to think about discrimination using statistical tools alone.

## A Reference

::: {style="z-index:1000"}
{{< video  https://www.youtube.com/watch?v=l_LeJfn_qW0 width="854px" height="480px">}}
:::

- This critique of potential outcomes for thinking about discrimination is paraphrased
 from \
 a paper by [Issa Kohler-Hausmann](https://law.yale.edu/issa-kohler-hausmann). Roughly paraphased. I encourage you to read it yourself.
- The title refers to a 1984 SNL sketch in which Eddie Murphy conducts a revealing audit study. 

[Eddie Murphy and the Dangers of Counterfactual Causal Thinking About Detecting Racial Discrimination](https://openyls.law.yale.edu/bitstream/handle/20.500.13051/18009/Kohler-Hausmann,%20Eddie%20Murphy%20and%20the%20Dangers%20of%20Counterfactual%20Causal%20Thinking%20about%20Detecting%20Racial%20Discrimination.pdf?sequence=1) \
[Kohler-Hausmann](https://law.yale.edu/issa-kohler-hausmann), 2018
