---
title: "Lecture 14"
subtitle: "Inverse Probability Weighting"
---

## Summary  

{{< include shared-code.qmd >}}
{{< include ca-income.qmd >}}

```{r}
ca = earnings 
sam = data.frame(
  w = 0 + (ca$sex == 'female'),
  x = ca$education,
  y = ca$income)
```

```{r}
summaries = sam |> 
    group_by(w,x) |> 
    summarize(muhat = mean(y), sigmahat = sd(y), Nwx = n(), .groups='drop')
  
summary.lookup = function(column.name, summaries, default=NA) { #<1>
  function(w,x) { 
    out = left_join(data.frame(w=w,x=x),                              #<2>
                    summaries[, c('w','x',column.name)],              #<2>
                    by=c('w','x')) |>                                 #<2>
      pull(column.name)                                               #<2>
    out[is.na(out)] = default                                         #<3>
    out
  }
}

muhat    = summary.lookup('muhat',    summaries)
sigmahat = summary.lookup('sigmahat', summaries)
Nwx      = summary.lookup('Nwx',      summaries)

W = sam$w
X = sam$x
Y = sam$y
x = unique(X)
n = length(Y)
```

```{r}
proportion.scale = 350000
plotX = function(w,x, shift=default.shift) { ifelse(w==0, x-shift, x+shift) }
interleaved.bars = list(
  geom_bar(aes(x=plotX(w,x), y=proportion.scale*after_stat(prop), 
               fill=factor(w), width=w), 
           position='identity', alpha=.1, linewidth=.05, color='black'))
purple.bars = geom_bar(aes(x=x, y=proportion.scale*after_stat(prop)), 
                       alpha=.1, linewidth=.05, fill='purple', color='black', width=.2)
scales = list(
  scale_y(sec.axis=sec_axis(~./proportion.scale, name='proportion')), 
  scale_x(),
  coord_cartesian(ylim=c(0,1.5e5)),
  guides(color='none', fill='none', size='none') 
)
```

```{r}
m0x.mild     = rep(c(10,12,14,16,18), rev(c(80, 90, 100, 110, 120)))
m0x.moderate = rep(c(10,12,14,16,18), rev(c(50, 75, 100, 125, 150)))
m0x.extreme  = rep(c(10,12,14,16,18), rev(c(2,49, 100, 149, 198)))
m0x.flat = rep(c(10,12,14,16,18), c(100,100,100, 100,100))
m1x.mild     = rep(c(10,12,14,16,18), c(80, 90, 100, 110, 120))
m1x.moderate = rep(c(10,12,14,16,18), c(50, 75, 100, 125, 150))
m1x.extreme  = rep(c(10,12,14,16,18), c(2,49, 100, 149, 198))

exercise.pop = function(m0x, m1x, mu=NULL) {
  if(is.null(mu)) { 
    fit = lm(y ~ w*factor(x), data=sam)
    mu = function(w,x) { predict(fit, newdata=data.frame(w=w, x=x)) }
  }
  eps0 = rnorm(length(m0x))*sigmahat(0,m0x)/3
  eps1 = rnorm(length(m1x))*sigmahat(1,m1x)/3
  for(uu in unique(m1x)) { 
    eps0[m0x==uu] = eps0[m0x==uu] - mean(eps0[m0x==uu])
    eps1[m1x==uu] = eps1[m1x==uu] - mean(eps1[m1x==uu])
  }
  rbind(
    data.frame(w=0, x=m0x, y=mu(0,m0x)+eps0),
    data.frame(w=1, x=m1x, y=mu(1,m1x)+eps1))
}

exercise.pop.mild = exercise.pop(m0x.mild, m1x.mild)
exercise.pop.moderate = exercise.pop(m0x.moderate, m1x.moderate)
exercise.pop.extreme = exercise.pop(m0x.extreme, m1x.extreme)

set.seed(0)
J.sam = sample(1:nrow(exercise.pop.moderate), 200, replace=TRUE)
exercise.sam = exercise.pop.moderate[J.sam,]
n = nrow(exercise.sam)
```

```{r}
Delta0.target = function(pop=NULL) {
  function(mu, sam) { 
    if(is.null(pop)) { pop = sam }
    mean(mu(1,pop$x[pop$w==0]) - mu(0,pop$x[pop$w==0]))
  }
}
with.constant.weights = function(sam) { 
  sam$weights = 1  
  sam
}
with.ipw = function(sam, from=sam) { 
  summaries = from |> group_by(w,x) |> summarize(Nwx=n(), .groups='drop')
  Nwx = summary.lookup('Nwx', summaries)
  sam$weights = Nwx(0,sam$x) / Nwx(sam$w,sam$x)
  sam 
} 

plot.sampling.distribution = function(pop, formula, n = nrow(sam), 
target=Delta0.target(), pop.version=FALSE, replications=1000,
with.weights=with.constant.weights, seed=1) { 
    estimates = with_seed(function() { 
      1:replications |> map_vec(function(rep) { 
        J = sample(1:nrow(pop), size=n, replace=TRUE)
        sam = pop[J,] |> with.weights() 

        fitted.model = lm(formula, data=sam, weights=weights) 
        muhat = function(w,x) { predict(fitted.model, newdata=data.frame(w=w, x=x)) } 
        target(muhat, sam)
    })}, seed=seed)
    
    mu.model = lm(y~w*factor(x), data=pop)
    mu = function(w,x) { predict(mu.model, newdata=data.frame(w=w, x=x)) }
    the.target = target(mu, pop)
    

    mutilde.model = lm(formula, data=pop |> with.weights(), weights=weights)
    mutilde = function(w,x) { predict(mutilde.model, newdata=data.frame(w=w, x=x)) }
    the.pop.version = target(mutilde, pop)
    
    p = ggplot() + 
     geom_vline(xintercept = the.target, color='green', linewidth=2, alpha=.5) 
    if(pop.version) { p = p + geom_vline(xintercept = the.pop.version, color=pink, linewidth=2, alpha=.5) }
    p +
     geom_vline(xintercept = mean(estimates), color='blue', linewidth=.75, alpha=.9) +
     geom_histogram(aes(x=estimates, y=after_stat(density)), bins=50, alpha=.2)
}
```

```{r}
default.shift = .15
biggest.max.weight = exercise.pop.extreme |> with.ipw() |> pull(weights) |> max()
lsq.plot = function(sam, formula=NULL, shift=default.shift, line.alpha=.7, size.scale=.5) {
  if(is.null(sam$weights)) { 
    sam$weights = rep(1, nrow(sam))
  }
  max.weight = max(sam$weights)
  max.size = size.scale * max.weight 
  
  base.plot = ggplot(sam) + 
    geom_point(aes(x=plotX(w,x), y=y,color=factor(w), size=I(weights)), 
      position=position_jitter(seed=0, width=shift),
      alpha=.3) + 
    stat_summary(aes(x=plotX(w,x), y=y, fill=factor(w)), 
      fun.data=mean_sd, geom='pointrange', 
      shape=21, alpha=.5, color='black') +
    interleaved.bars + purple.bars + scales
    
    if(is.null(formula)) { base.plot } else { 
        model = lm(formula, data=sam, weights=weights)
        gridx = seq(-10, 50, length.out=1000)
        muhat = function(w,x) predict(model, newdata=data.frame(w=w,x=x))   
        
        the.plot = base.plot 
        if(line.alpha > 0) { 
            the.plot = base.plot + 
            geom_line(aes(x=plotX(w,x), y=muhat(w,x), color=factor(w)), 
              data = expand_grid(w=0:1, x=gridx), alpha=line.alpha, linetype='dashed') 
        }
        the.plot + 
          geom_point(aes(x=plotX(w,x), y=muhat(w,x), fill=factor(w)),  
            color='black', alpha=.5, shape=23)
    }
}
```


```{r}
J = sample(1:nrow(exercise.pop.moderate), 20)
exercise.pop.moderate[J,] |> with.ipw(from=exercise.pop.moderate) |> lsq.plot(y ~ 1+w) 
```

- Today, we're going to talk about *inverse probability weighting*.
- This is a simple technique for getting unbiased estimates ... 
  - of targets like adjusted differences in means... 
  - even when your model is badly misspecified. 
- You can use it even when it's impossible to avoid bias using the 'all functions' model. 
  - e.g., when you have empty columns. This often happens when you have a small sample.
  - Or when you have many columns, i.e. many covariates to adjust for or one with many levels.
- The essential idea is to fit the data *where it's used in your comparison*, not where most of the data is.

## A Simplification

::: {.panel-tabset}
### Sample Least Squares Predictor 
```{r}
exercise.sam |> lsq.plot(y ~ 1+w)
```

### Population Least Squares Predictor 
```{r}
exercise.pop.moderate |> lsq.plot(y ~ 1+w)
```

### The Sampling Distribution of $\hat\Delta_0$
::: {#fig-sampling-Delta0-lsq}
```{r}
exercise.pop.moderate |> plot.sampling.distribution(y ~ 1+w, 
  pop.version=TRUE, target=Delta0.target(exercise.pop.moderate))
```
The sampling distribution of the adjusted difference in means $\hat\Delta_0$
when we use $\hat\mu$ as an estimator of $\mu$. Its center $\E\hat\Delta_0$ is shown in blue. The target $\Delta_0$ is shown in green. And the population version of the estimator, $\tilde\Delta_0$, is shown in pink. You can see that the sampling distribution of $\hat\Delta_0$ is centered on $\tilde\Delta_0$.
:::
:::

- To keep things simple, we'll start by thinking about the *population least squares predictor*. 

$$
\begin{aligned}
\tilde\mu 
&\explain{=}{it minimizes the average of the squared errors we make on the population}
  \argmin_{m \in \model} \frac{1}{m}\sum_{j=1}^m \qty{ y_j - m(w_j,x_j) }^2  \\
&\explain{=}{this average can be rewritten as the expected squared error for $(W_i,X_i,Y_i)$ drawn uniformly-at-random from the population}
  \argmin_{m \in \model} \E\qty[ \qty{ Y_i - m(W_i,X_i) }^2 ] \\
\end{aligned}
$$

- We saw in [Lecture 12](`r lecture(12)`) that $\tilde\mu(w,x)$ is the *expected value* of the (sample) least squares predictor \
when $(W_i,X_i,Y_i)$ are drawn uniformly-at-random from the population.

$$
\tilde\mu(w,x) = \E[ \hat\mu(w,x) ] \qfor \hat\mu = \argmin_{m \in \model} \frac{1}{n}\sum_{i=1}^n \qty{ Y_i - m(W_i,X_i) }^2.
$$

- So if we're estimating a target like $\Delta_0$, the error we get using $\tilde\mu$ is the *bias* we get using $\hat\mu$.
$$
\E[\hat\Delta_0] - \Delta_0 = \tilde\Delta_0 - \Delta_0 
$$


## Least Squares and 'Noise': A Warm-up Exercise 
```{r}
exercise.pop.moderate |> lsq.plot(y ~ 1+w) 
```

- When we're doing this, it'll be helpful to know something about the population least squares predictor $\tilde\mu(w,x)$.
  - It depends *only* on the column means $\mu(w,x)$.
  - The rest---how our observations vary around them---doesn't matter. 

$$
\begin{aligned}
\tilde\mu 
&= \argmin_{m \in \model} \sum_{j=1}^n \left\{ y_j - m(w_j,x_j) \right\}^2 \\
&= \argmin_{m \in \model} \sum_{j=1}^n \left\{ \mu(w_j,x_j) - m(w_j,x_j) \right\}^2. 
\end{aligned}
$$

::: {.panel-tabset}
### Notation and Terminology
- That's why, in Machine Learning tradition, we call $\mu(w,x)$ the *signal* and everything else the *noise*.
$$
y_j = \underset{\text{signal}}{\mu(w_j, x_j)} + \underset{\text{noise}}{\varepsilon_j} \qfor \mu(w,x)=\frac{1}{m_{w,x}}\sum_{j:w_j=w, x_j=x} y_j
$$
- When we talk about a random variable $(W_i,X_i,Y_i)$ drawn uniformly-at-random from the population, we can phrase this in terms of *conditioning*.

$$
Y_i = \mu(W_i,X_i) + \varepsilon_i \qqtext{ where } \E[\varepsilon_i \mid W_i, X_i] = 0. 
$$

- $\mu$ here is the same function.
- $\varepsilon_i$ is drawn uniformly-at-random from the set of deviations from the column mean in the population.
$$
\varepsilon_i \text{ is drawn uniformly-at-random from } \{ \varepsilon_j \ : \ w_j = W_i, x_j = X_i \}.
$$

###

### Proof 

- Let's look at what $\tilde\mu$ minimizes in random variable terms. 
$$
\begin{aligned}
\text{MSE}(m)
&= \E\qty[ \qty{ Y_i - m(W_i,X_i) }^2 ] \\ 
&= \E\qty[ \qty{ \varepsilon_i + \mu(W_i,X_i) - m(W_i,X_i) }^2 ] \\ 
&= \E\qty[ \qty{ \varepsilon_i + \qty{\mu(W_i,X_i) - m(W_i,X_i)}}^2 ] \\ 
&= \E\qty[ \varepsilon_i^2 + 2 \varepsilon_i \qty{\mu(W_i,X_i) - m(W_i,X_i)} + \qty{\mu(W_i,X_i) - m(W_i,X_i)}^2 ] \\
&= \E[\varepsilon_i^2] + 0 + \E[\qty{\mu(W_i,X_i) + m(W_i,X_i)}^2 ]
\end{aligned}
$$
- The difference between the version with $Y$ and the version with $\mu$ is a constant. 
- It doesn't depend on $m$. So it doesn't influence which $m$ minimizes the (population) mean squared prediction error $\text{MSE}(m)$.

:::

# The Idea
Intuition and Empirical Evidence. 

## Least Squares and Covariate Shift 
::: {.panel-tabset}

### Mild Covariate Shift 
::: {#fig-mildshift-lsq}
```{r}
formula = y ~ 1+w
exercise.pop(m0x.mild, m1x.mild) |> lsq.plot(formula)
```
A population with mild covariate shift and the population least squares predictor $\tilde\mu$ in the *horizontal lines model* ([eq @eq-horizontal-lines-model]).
:::

### Moderate Covariate Shift 
::: {#fig-modshift-lsq}
```{r}
exercise.pop(m0x.moderate, m1x.moderate) |> lsq.plot(formula)
```
A population with moderate covariate shift and the population least squares predictor $\tilde\mu$ in the *horizontal lines model* ([eq @eq-horizontal-lines-model])
:::

### Extreme Covariate Shift 
::: {#fig-extshift-lsq}
```{r}
exercise.pop(m0x.extreme, m1x.extreme) |> lsq.plot(formula)
```
A population with extreme covariate shift and the population least squares predictor $\tilde\mu$ in the *horizontal lines model* ([eq @eq-horizontal-lines-model])
:::
::: 


$$
\color{gray}
\begin{aligned}
\Delta_{\text{raw}} 
&= 
\ingroupb{\frac{1}{m_1}\sum_{j:w_j=1} y_j} - \ingroupa{\frac{1}{m_0}\sum_{j:w_j=0} y_j} 
&&\explain{=}{translated into histogram form} \quad 
\sum_{x} \ingroupb{p_{x \mid 1} \ \mu(1,x)} - \sum_{x} \ingroupa{p_{x \mid 0} \ \mu(0,x)}  \\
\tilde\Delta_{0} 
&= 
\ingroupa{\frac{1}{m_0}\sum_{j:w_j=0}} \lcb{ \ingroupb{\tilde \mu(1,x)} - \ingroupa{\tilde\mu(0,x) }} 
&&\explain{=}{translated into histogram form} \quad
\sum_x \ingroupa{p_{x \mid 0}} \ \ingroupb{\tilde \mu(1,x)} - 
\sum_x \ingroupa{p_{x \mid 0}} \ \ingroupa{\tilde\mu(0,x) }.
\end{aligned}
$$

::: {.panel-tabset}
### Question 1
How does the *raw difference in means* $\Delta_{\text{raw}}$ compare to the adjusted difference $\tilde\Delta_{0}$ \
when our adjustment uses the least squares predictor in the *all functions model*?


See @sec-q1-answer for the answer.


### Question 2 
How does the *raw difference in means* $\Delta_{\text{raw}}$ compare to the adjusted difference $\tilde\Delta_{0}$ \
when our adjustment uses the least squares predictor in the *horizontal lines* model?

See @sec-q2-answer for the answer.

### Question 3 
Why, in intuitive terms, is the constant [$\tilde\mu(1,x)$]{.groupa} *not* the right prediction to use in this comparison? *Hint*. Think about the extreme covariate shift case. Which green dots does fit well? Which does it not fit well? Why?

See @sec-q3-answer for the answer.

### The Essential Problem 

What we want $\tilde\mu(1,x)$ to be good at (when we're using it in $\tilde\Delta_{0}$) and what we're asking it to be good at (when we're doing least squares) are different things. When there's a lot of covariate shift, they're very different things. 

:::


## A Very Simple Improvement 

::: {.panel-tabset}
### Mild Covariate Shift 

::: {#fig-mildshift-ipw}
```{r}
formula = y ~ 1+w
exercise.pop.mild |> with.ipw() |> lsq.plot(formula)
```
A population with mild covariate shift and the *inverse probability weighted* population least squares predictor $\tilde\mu$ in the *horizontal lines model* ([eq @eq-horizontal-lines-model]) after inverse probability weighting.
:::

### Moderate Covariate Shift 
::: {#fig-modshift-ipw}
```{r}
exercise.pop.moderate |> with.ipw() |> lsq.plot(formula)
```
A population with moderate covariate shift and the *inverse probability weighted* population least squares predictor $\tilde\mu$ in the *horizontal lines model* ([eq @eq-horizontal-lines-model]) after inverse probability weighting.
:::

### Extreme Covariate Shift 
::: {#fig-extshift-ipw}
```{r}
exercise.pop(m0x.extreme, m1x.extreme) |> with.ipw() |> lsq.plot(formula)
```
A population with extreme covariate shift and the *weighted* population least squares predictor $\tilde\mu$ in the *horizontal lines model* ([eq @eq-horizontal-lines-model]).
:::
:::

```{r}
summaries = exercise.pop(m0x.moderate, m1x.moderate) |> 
  group_by(w,x) |> summarize(Nwx=n(), .groups='drop')
Nwx = summary.lookup('Nwx', summaries)
```

::: {.panel-tabset}
### Pretending our way out of covariate shift 
- Misspecification is a problem *because of* covariate shift. We're fitting the wrong green dots.
- And we can't really make the shift go away. In the moderate case above...
  - At $x=10$, we have [`r Nwx(0,10)`]{.groupa} red dots and [`r Nwx(1,10)`]{.groupb} green dots.
  - At $x=12$, we have [`r Nwx(0,12)`]{.groupa} red dots and [`r Nwx(1,12)`]{.groupb} green dots.
  - At $x=14$, we have [`r Nwx(0,14)`]{.groupa} red dots and [`r Nwx(1,14)`]{.groupb} green dots.
- But we can pretend that we see each green dot we do have multiple times. 
We pretend that ...
  - we see each one in the $x=10$ column $\ingroupa{`r Nwx(0,10)`} / \ingroupb{`r Nwx(1,10)`} = `r round(Nwx(0,10)/Nwx(1,10), 2)`$ times.
  - we see each one in the $x=12$ column $\ingroupa{`r Nwx(0,12)`} / \ingroupb{`r Nwx(1,12)`} \approx `r round(Nwx(0,12)/Nwx(1,12), 2)`$ times.
  - we see each one in the $x=14$ column $\ingroupa{`r Nwx(0,14)`} / \ingroupb{`r Nwx(1,14)`} = `r round(Nwx(0,14)/Nwx(1,14), 2)`$ time.  
- Or possibly a fewer times. Since that's fewer times than 1, fewer means a fraction. 
  - We pretend that we see each one in the $x=16$ column $\ingroupa{`r Nwx(0,16)`} / \ingroupb{`r Nwx(1,16)`} \approx `r round(Nwx(0,16)/Nwx(1,16),2)`$ times. 
  - We pretend that we see each one in the $x=18$ column $\ingroupa{`r Nwx(0,18)`} / \ingroupb{`r Nwx(1,18)`} \approx `r round(Nwx(0,18)/Nwx(1,18),2)`$ times.
  - Those are the ratios of red dots to green dots in those columns.

### Visualization 
- In the plots above, we scale each dot by the number of times we pretend we see it.
  - In particular, we scale each dot's *area*. 
  - The total area of the red and green dots in each column is the same. 
- When we have extreme shift, that means we get some really big green dots.
  - That's because we're pretending that each observation is *a lot* of observations. 

### What it means to pretend 

- Let's think about what the population least squares predictor $\tilde\mu(w,x)$ looks like in histogram form. 
$$
\begin{aligned}
\tilde\mu = \argmin_{m \in \model} \ \text{MSE}(m) \qfor
\text{MSE}(m) 
&= \frac{1}{m}\sum_{wx}\sum_{j:w_j=w, x_j=x} \lcb{ \mu(w_j,x_j) - m(w_j,x_j) }^2 \\
&= \frac{1}{m}\sum_{wx}\sum_{j:w_j=w, x_j=x}  \lcb{ \mu(w, x) - m(w,x) }^2 \\
&= \frac{1}{m}\sum_{wx} m_{wx} \lcb{ \mu(w, x) - m(w,x) }^2 \\
\end{aligned}
$$

- To *pretend* that we have the same number of green dots as red dots in each column ...
  - we just replace the number of dots $m_{wx}$ that are actually there ... 
  - with the number of dots we want to *pretend* there are: $m_{0x}$. 
- Equivalently, we *weight* each term by a factor of $m_{0x}/m_{wx}$. We'll call this weight $\gamma(w,x)$.
$$
\begin{aligned}
\tilde\mu^{\text{IPW}} = \argmin_{m \in \model} \ \text{WMSE}(m) \qfor
\text{WMSE}(m) 
&= \frac{1}{m}\sum_{wx} \ingroupa{m_{0x}} \lcb{ \mu(w, x) - m(w,x) }^2 \\
&= \frac{1}{m}\sum_{wx} \gamma(w,x) m_{wx}\lcb{ \mu(w, x) - m(w,x) }^2  \\
\qfor &\gamma(w,x) = \frac{m_{0x}}{m_{wx}} = \begin{cases} 1 & \text{if } w=0 \\ \frac{\ingroupa{m_{0x}}}{\ingroupb{m_{1x}}} & \text{if } w=1 \end{cases}
\end{aligned}
$$

- And if we want to think about this as a sum over the population ...
- we can just give each person in our sum the weight $\gamma(w_j,x_j)$ for their column.

$$
\begin{aligned}
\tilde\mu^{\text{IPW}} = \argmin_{m \in \model} \ \text{WMSE}(m) \qfor
\text{WMSE}(m) 
&= \frac{1}{m}\sum_{j=1}^m \gamma(w_j,x_j) \lcb{ \mu(w_j, x_j) - m(w_j,x_j) }^2  
\end{aligned}
$$

:::

## A Very Simple Improvement: Flat vs. Mild {#flat-vs-mild}
::: {.panel-tabset}
### Mild Covariate Shift 
```{r}
formula = y ~ 1+w
exercise.pop(m0x.flat, m1x.mild)|> with.ipw() |> lsq.plot(formula)
```
### Moderate Covariate Shift 
```{r}
exercise.pop(m0x.flat, m1x.moderate) |> with.ipw() |> lsq.plot(formula)
```
### Extreme Covariate Shift 
```{r}
exercise.pop(m0x.flat, m1x.extreme) |> with.ipw() |> lsq.plot(formula)
```
### 
:::

- Here's a version of the same comparison when we ...
  - make the shift in [$p_{x \mid 1}$]{.groupb} more and more extreme 
  - but keep [$p_{x \mid 0}$]{.groupa} flat.
- Notice that the least squares predictor $\tilde\mu$ doesn't change at all.
  - That's because it's doing what we asked it to do: minimize error over the red distribution. 
  - That's not changing, so neither are our least squares predictors.

## Using a Sample

::: {.panel-tabset}
### Mild Covariate Shift 
```{r}
#| layout-ncol: 2
#| fig-width: 7
#| fig-cap:
#|  - A sample and the weighted least squares predictor in the horizontal lines model. 
#|  - The population it's drawn from and corresponding weighted population least squares predictor. 
formula = y ~ 1+w
J = J.sam
exercise.pop(m0x.flat, m1x.mild)[J,] |> with.ipw(from=exercise.pop(m0x.flat, m1x.mild)) |> lsq.plot(formula)
exercise.pop(m0x.flat, m1x.mild) |> with.ipw() |> lsq.plot(formula)
```

### Moderate Covariate Shift 
```{r}
#| layout-ncol: 2
#| fig-width: 7
#| fig-cap:
#|  - A sample and the weighted least squares predictor in the horizontal lines model. 
#|  - The population it's drawn from and corresponding weighted population least squares predictor. 
exercise.pop(m0x.flat, m1x.moderate)[J,] |> with.ipw(from=exercise.pop(m0x.flat, m1x.moderate)) |> lsq.plot(formula)
exercise.pop(m0x.flat, m1x.moderate) |> with.ipw() |> lsq.plot(formula)
```

### Extreme Covariate Shift 
```{r}
#| layout-ncol: 2
#| fig-width: 7
#| fig-cap:
#|  - A sample and the weighted least squares predictor in the horizontal lines model. 
#|  - The population it's drawn from and corresponding weighted population least squares predictor. 
exercise.pop(m0x.flat, m1x.extreme)[J,] |> with.ipw(from=exercise.pop(m0x.flat, m1x.extreme)) |> lsq.plot(formula)
exercise.pop(m0x.flat, m1x.extreme) |> with.ipw() |> lsq.plot(formula)
```
:::


- In reality, we don't have the population. We have (hopefully) a sample drawn uniformly-at-random from it.
- To make this work like a sample drawn from the population we've been pretending to have ....
- ... we weight the people in our sample exactly the same way when we estimate $\hat\mu$.
$$
\begin{aligned}
\hat\Delta_0 &= \frac{1}{m_0} \sum_{j:w_j=0} \qty{ \hat\mu(1,x_j) - \hat\mu(0,x_j) } \\
& \qfor \hat\mu = \argmin_{m \in \model} \frac{1}{n}\sum_{i=1}^n \gamma(W_i,X_i) \lcb{ Y_i - m(W_i,X_i) }^2 \\
& \qand \gamma(w,x) = \frac{\ingroupa{m_{0x}}}{m_{wx}} = \begin{cases} 1 & \text{if } w=0 \\ \frac{\ingroupa{m_{0x}}}{\ingroupb{m_{1x}}} & \text{if } w=1 \end{cases}
\end{aligned}
$$

- Note that the weights $\gamma(w,x)$ we are based on knowledge of the *population's covariate distribution*. 
- Sometimes we have that information. 
  - E.g. when we survey registered voters, some information about every registered voter is public record. 
- Sometimes we don't, so we have to use an estimate $\hat\gamma$ of these weights. 
  - E.g., we can use the corresponding ratios of red and green dots in the sample.

$$
\hat\gamma(w,x) = \frac{\ingroupa{N_{0x}}}{N_{wx}} = \begin{cases} 1 & \text{if } w=0 \\ \frac{\ingroupa{N_{0x}}}{\ingroupb{N_{1x}}} & \text{if } w=1 \end{cases} \qfor N_{wx} = \sum_{i:W_i=w, X_i=x} 1
$$

## This Works *Perfectly* 

::: {.panel-tabset}
### Mild Covariate Shift 
```{r}
#| layout-ncol: 2
#| fig-width: 7
#| fig-cap:
#|  - A sample and the weighted least squares predictor in the horizontal lines model. 
#|  - The sampling distribution of the corresponding adjusted difference in means estimator. 
formula = y ~ 1+w
J = J.sam

xrange = c(-25000, 0)
xbreaks = seq(xrange[1], xrange[2], by=5000)
scalex.hist = scale_x_continuous(breaks=xbreaks, limits=range(xbreaks), labels=sprintf("$%dk", xbreaks/1000))

exercise.pop(m0x.flat, m1x.mild)[J,] |> with.ipw(from=exercise.pop(m0x.flat, m1x.mild)) |> lsq.plot(formula) 
exercise.pop(m0x.flat, m1x.mild) |> plot.sampling.distribution(formula, 
  target=Delta0.target(exercise.pop(m0x.flat, m1x.mild)), 
  with.weights=\(sam)with.ipw(sam, from=exercise.pop(m0x.flat, m1x.mild))) + scalex.hist
```

### Moderate Covariate Shift 
```{r}
#| layout-ncol: 2
#| fig-width: 7
#| fig-cap: 
#|  - A sample and the weighted least squares predictor in the horizontal lines model. 
#|  - The sampling distribution of the corresponding adjusted difference in means estimator. 
exercise.pop(m0x.flat, m1x.moderate)[J,] |> with.ipw(from=exercise.pop(m0x.flat, m1x.moderate)) |> lsq.plot(formula)
exercise.pop(m0x.flat, m1x.moderate) |> plot.sampling.distribution(formula, 
  target=Delta0.target(exercise.pop(m0x.flat, m1x.moderate)), 
  with.weights=\(sam)with.ipw(sam, from=exercise.pop(m0x.flat, m1x.moderate))) + scalex.hist
```

### Extreme Covariate Shift 
```{r}
#| layout-ncol: 2
#| fig-width: 7
#| fig-cap: 
#|  - A sample and the weighted least squares predictor in the horizontal lines model. 
#|  - The sampling distribution of the corresponding adjusted difference in means estimator. 
exercise.pop(m0x.flat, m1x.extreme)[J,] |> with.ipw(from=exercise.pop(m0x.flat, m1x.extreme)) |> lsq.plot(formula)
exercise.pop(m0x.flat, m1x.extreme) |> plot.sampling.distribution(formula, 
  target=Delta0.target(exercise.pop(m0x.flat, m1x.extreme)), 
  with.weights=\(sam)with.ipw(sam, from=exercise.pop(m0x.flat, m1x.extreme))) + scalex.hist
```
:::


- When we weight based on the *population covariate distribution*, we get an unbiased estimator.^[Here, for the sake of clarity, we're talking about $\hat\Delta_0$ defined in @sec-deltahat0-pop-version.]

$$
\E[\hat\Delta_0] = \Delta_0 \qfor \Delta_0 = \frac{1}{m_0} \sum_{j:w_j=0} \qty{ \mu(1,x_j) - \mu(0,x_j) }
$$


## Or Almost *Perfectly*. 

::: {.panel-tabset}
### Mild Covariate Shift 
```{r}
#| layout-ncol: 2
#| fig-width: 7
#| fig-cap:
#|  - The sampling distribution of our adjusted difference in means estimator (using sample proportions). 
#|  - The sampling distribution of our adjusted difference in means estimator (using population proportions). 
formula = y ~ 1+w
J = J.sam
exercise.pop(m0x.flat, m1x.mild) |> plot.sampling.distribution(formula, 
  target=Delta0.target(), 
  with.weights=with.ipw) + scalex.hist
exercise.pop(m0x.flat, m1x.mild) |> plot.sampling.distribution(formula, 
  target=Delta0.target(exercise.pop(m0x.flat, m1x.mild)), 
  with.weights=\(sam)with.ipw(sam, from=exercise.pop(m0x.flat, m1x.mild))) + scalex.hist
```

### Moderate Covariate Shift 
```{r}
#| layout-ncol: 2
#| fig-width: 7
#| fig-cap: 
#|  - The sampling distribution of our adjusted difference in means estimator (using sample proportions). 
#|  - The sampling distribution of our adjusted difference in means estimator (using population proportions). 
exercise.pop(m0x.flat, m1x.moderate) |> plot.sampling.distribution(formula, 
  target=Delta0.target(), 
  with.weights=with.ipw) + scalex.hist
exercise.pop(m0x.flat, m1x.moderate) |> plot.sampling.distribution(formula, 
  target=Delta0.target(exercise.pop(m0x.flat, m1x.moderate)), 
  with.weights=\(sam)with.ipw(sam, from=exercise.pop(m0x.flat, m1x.moderate))) + scalex.hist
```

### Extreme Covariate Shift 
```{r}
#| layout-ncol: 2
#| fig-width: 7
#| fig-cap: 
#|  - The sampling distribution of our adjusted difference in means estimator (using sample proportions). 
#|  - The sampling distribution of our adjusted difference in means estimator (using population proportions). 
exercise.pop(m0x.flat, m1x.extreme) |> plot.sampling.distribution(formula, 
  target=Delta0.target(), 
  with.weights=with.ipw) + scalex.hist
exercise.pop(m0x.flat, m1x.extreme) |> plot.sampling.distribution(formula, 
  target=Delta0.target(exercise.pop(m0x.flat, m1x.extreme)), 
  with.weights=\(sam)with.ipw(sam, from=exercise.pop(m0x.flat, m1x.extreme))) + scalex.hist
```
:::

- We actually do better when we weight using the *sample covariate distribution*.^[Here, for the sake of clarity, we're talking about $\hat\Delta_0$ defined in @sec-deltahat0-sample-version. 
  - It has lower variance than the population-based version. 
  - Especially when there's a lot of covariate shift.
  - This is pretty interesting, but we won't have time to get into it.

# Generalizing the Idea
An Exercise.

## $\Delta_0$ Review

## $\Delta_1$

## $\Delta_{\text{all}}$ 

# Proving Unbiasedness 
DO IT! 

::: {.panel-tabset}
### Setup
- Suppose we have this inverse probability weighted population least squares predictor $\tilde\mu$.
$$
\begin{aligned}
\tilde\mu= \argmin_{m \in \model} \ WMSR(m) \qfor
WMSR(m) 
&= \sum_{wx} \frac{\ingroupa{m_{0x}}}{m_{wx}} m_{wx} \lcb{ \mu(w, x) - m(w,x) }^2 \\
\end{aligned}
$$
- Show that when we plug it in to our formula for $\tilde\Delta_0$, we get our estimation target $\Delta_0$.
$$
\begin{aligned}
\tilde\Delta_0 &= \frac{1}{m_0} \sum_{j:w_j=0} \qty{ \tilde\mu(1,x_j) - \tilde\mu(0,x_j) }  && \text{ the population version of our estimator} \\
\Delta_0 &= \frac{1}{m_0} \sum_{j:w_j=0} \qty{ \mu(1,x_j) - \mu(0,x_j) } && \text{ our target}
\end{aligned}
$$

- *Tips*. 
  - Write things in histogram form. 
  - What orthogonality condition to get for weighted least squares?

### Space 

### Questions 

:::

## It Does Work for Other Models 

```{r}
pop = exercise.pop(m0x.flat, m1x.moderate)
J = sample(1:nrow(pop), 200)
sam = pop[J,]
```

::: {.panel-tabset}
### Horizontal Lines Model 
```{r}
#| layout-ncol: 2
#| fig-width: 7
formula = y ~ 1+w
sam |> with.ipw() |> lsq.plot(formula) 
pop |> plot.sampling.distribution(formula, with.weights=with.ipw)
```

### Lines Model
```{r}
#| layout-ncol: 2
#| fig-width: 7
formula = y ~ w*x
sam |> with.ipw() |> lsq.plot(formula) 
pop |> plot.sampling.distribution(formula, with.weights=with.ipw)
```

### Parallel Lines Model
```{r}
#| layout-ncol: 2
#| fig-width: 7
formula = y ~ w*x
sam |> with.ipw() |> lsq.plot(formula) 
pop |> plot.sampling.distribution(formula, with.weights=with.ipw)
```

### Additive Model
```{r}
#| layout-ncol: 2
#| fig-width: 7
formula = y ~ w+factor(x) 
sam |> with.ipw() |> lsq.plot(formula, line.alpha=0) 
pop |> plot.sampling.distribution(formula, with.weights=with.ipw)
```
:::

## Why This is Called Inverse Probability Weighting 

STUFF

## Conclusion

2x2 COMPARISON PLOTS

Comments on Science 

# Appendix: Answers 

## Q1 Answer {#sec-q1-answer}
The raw difference is bigger. In comparison with the adjusted one, the average of $\ingroupb{\mu(1,x)}$ is taken over a covariate distribution that's shifted to the right, where it's bigger.

## Q2 Answer {#sec-q2-answer}
They're the same. When we use the horizontal lines model, $\tilde\mu(1,x)$ and $\tilde\mu(0,x)$ are constant and equal to the within-group means. So in the first term in $\tilde\Delta_{0}$, we're taking the average of a constant---and that constant is the first term of $\Delta_\text{raw}$. And the same happens for the second term.

## Q3 Answer {#sec-q3-answer}
It fits the green dots on the right well because that's where most of the green dots are. But we're averaging it over the distribution of the red dots, which are on the left.

# Appendix: Two Versions of $\hat\Delta$

## The version where we use the population covariate distribution {#sec-deltahat0-pop-version}

$$
\begin{aligned}
\hat\Delta_0 &= \frac{1}{m_0} \sum_{j:w_j=0} \qty{ \hat\mu(1,x_j) - \hat\mu(0,x_j) } \\
& \qfor \hat\mu = \argmin_{m \in \model} \frac{1}{n}\sum_{i=1}^n \gamma(W_i,X_i) \lcb{ Y_i - m(W_i,X_i) }^2 \\
& \qand \gamma(w,x) = \frac{\ingroupa{m_{0x}}}{m_{wx}} = \begin{cases} 1 & \text{if } w=0 \\ \frac{\ingroupa{m_{0x}}}{\ingroupb{m_{1x}}} & \text{if } w=1 \end{cases}
\end{aligned}
$$ 

## The version where we use the sample covariate distribution {#sec-deltahat0-sample-version}

$$
\begin{aligned}
\hat\Delta_0 &= \frac{1}{N_0} \sum_{i:W_i=0} \qty{ \hat\mu(1,X_i) - \hat\mu(0,X_i) } \\
& \qfor \hat\mu = \argmin_{m \in \model} \frac{1}{n}\sum_{i=1}^n \gamma(W_i,X_i) \lcb{ Y_i - m(W_i,X_i) }^2 \\
& \qand \gamma(w,x) = \frac{\ingroupa{N_{0x}}}{N_{wx}} = \begin{cases} 1 & \text{if } w=0 \\ \frac{\ingroupa{N_{0x}}}{\ingroupb{N_{1x}}} & \text{if } w=1 \end{cases}
\end{aligned}
$$ 

# Appendix: Models

## The Horizontal Lines Model

```{r}
sam = exercise.pop(m0x.moderate, m1x.moderate)
```

```{r}
#| fig-cap: A sample and the least squares fit in the model. 
sam |> lsq.plot(y ~ w*x) 
```

$$
\model=\{m(w,x)=a(w) \qqtext{ for functions } a\}
$$ {#eq-horizontal-lines-model}

```{r}
#| echo: true
formula = y ~ w
```

## The Parallel Lines Model

```{r}
#| fig-cap: A sample and the least squares fit in the model. 
sam |> lsq.plot(y ~ w+x) 
```
$$
\model=\{m(w,x)=a(w)+bx \qqtext{ for functions } a,b\}
$$ {#eq-parallel-lines-model}

```{r}
#| echo: true
formula = y ~ w+x
```

## The (Not-Necessarily-Parallel) Lines Model

```{r}
#| fig-cap: A sample and the least squares fit in the model. 
sam |> lsq.plot(y ~ w*x) 
```

$$
\model=\{m(w,x)=a(w)+b(w)x \qqtext{ for functions } a,b\}
$$ {#eq-lines-model}

```{r}
#| echo: true
formula = y ~ w*x
```



## The Additive Model 

```{r}
#| fig-cap: A sample and the least squares fit in the model. 
sam |> lsq.plot(y ~ w + factor(x), line.alpha=0) 
```

$$
\model=\{m(w,x)=a(w)+b(x) \qqtext{ for functions } a,b\}
$$ {#eq-additive-model}

```{r}
#| echo: true
formula = y ~ w + factor(x)
```