---
title: "QTM 220 Lecture 10"
subtitle: "Estimation with Non-Binary Covariates"
---

## Population and Sample

{{< include shared-code.qmd >}}
{{< include ca-income.qmd >}}


```{r}
style.tab = function(tab, cols=c('income', 'education', 'sex'), nhead=3, ntail=1) {
  for(name in colnames(tab)) { 
    if(str_starts(name, 'income')) { 
      tab[[name]] = sprintf('$%dk', round(as.numeric(tab[[name]]) / 1000))
    }
  }
  ellipsis.tab = if(nhead + ntail >= nrow(tab)) { 
    tab[, cols]
  } else {
    ellipsis.tab = tab[, cols] %>% ellipsis(nhead=nhead,ntail=ntail)
  }
  if('sex' %in% colnames(ellipsis.tab)) { 
  ellipsis.tab %>%
    kable %>% kable_styling() %>% 
    row_spec(which(ellipsis.tab$sex == 'male'),  background=twocolor.red) %>%
    row_spec(which(ellipsis.tab$sex != 'male'),  background=twocolor.green) 
  } else {
    ellipsis.tab %>%
    kable %>% kable_styling()
  }
}
```



```{r}
#| out-height: 300px
#| fig-width: 5 
#| layout-ncol: 2 
#| 
ca.plot + 
 geom_point(aes(x=long, y=lat, color=sex), data=display.population, size=.02, alpha=.1) + 
 guides(color='none')
ca.plot + geom_point(aes(x=long, y=lat, color=sex), data=our.sample, size=.1, alpha=.5) +
  guides(color='none')
```

:::: columns
::: column
```{r}
style.tab(pop.tab, nhead=2)
```
:::
::: column
```{r}
style.tab(sample.tab, nhead=2)
```
:::
::::

- Today we're going to talk about *estimation* when we have non-binary covariates.
- We've got a sample (right) which we'll think of as being drawn with replacement from a population (left).
- And we'll use the sample to estimate various summaries of the population.

::: aside
As usual, the population plot and table are made up. All we really have is the sample. The population stuff is included for the sake of illustration.
:::


## Summaries
```{r}
#| out-height: 200px 
#| layout-ncol: 2
#| 

scale_y = function() { coord_cartesian(ylim=c(0, 200000)) } 
pop.scatter = ggplot(pop.tab) +
  geom_point(aes(x=education, y=income),  color='blue',
            size=.2, position=position_jitter(seed=0,w=.3), alpha=.03) +
  stat_summary(aes(x=education, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x))  + 
  scale_y() 

sample.scatter = ggplot(sample.tab) +
  geom_point(aes(x=education, y=income), color='blue', 
            size=2, position=position_jitter(seed=0,w=.3), alpha=.13) +
  stat_summary(aes(x=education, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x))  +
  scale_y() 

pop.scatter
sample.scatter
```

:::: columns
::: column
```{r}
styled.pop.tab = pop.tab |> 
  group_by(education) |> 
  summarize(number=length(income), income.mean=mean(income), income.sd=sd(income)) |> 
  style.tab(cols=c('education', 'income.mean', 'number'), nhead=2)
styled.pop.tab
```
$$
\mu(x)  = \frac{1}{m_x}\sum_{j:x_j=x} y_j \qfor m_x = \sum_{j:x_j=x} 1
$$

:::
::: column
```{r}
styled.sample.tab = sample.tab |> 
  group_by(education) |> 
  summarize(number=length(income), income.mean=mean(income), income.sd=sd(income)) |> 
  style.tab(cols=c('education', 'income.mean', 'number'), nhead=2)
styled.sample.tab
```
$$
\hat\mu(x) = \frac{1}{N_{x}}\sum_{i:X_i=x} Y_i \qfor N_{x} = \sum_{i:X_i=x} 1
$$
:::
::::

- We summarize the relationship between education and income by the mean income for each level of education.
  - When we do this for the population, we get the stuff on the left. 
  - When we do this for the sample, we get the stuff on the right.
- **Exercise**. What is the mean income for California residents in our population with 10 years of education? 
- Locate it in the plot, write it in mathematical notation, then find the value in the table.

## Summaries Based on Multiple Covariates
```{r}
#| out-height: 200px
#| layout-ncol: 2
#| 

xco = function(education, sex) {
  education + .15*(2*(sex=='male')-1)
}

pop.scatter.bivariate = ggplot(pop.tab) + 
  geom_point(aes(x=xco(education,sex), y=income, color=sex), 
            size=.2, position=position_jitter(seed=0,w=.15), alpha=.05) +
  stat_summary(aes(x=xco(education,sex), y=income, color=sex), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x))  + 
  scale_y() 

sample.scatter.bivariate = ggplot(sample.tab) +
  geom_point(aes(x=xco(education,sex), y=income, color=sex), 
            size=2, position=position_jitter(seed=0,w=.1), alpha=.15) +
  stat_summary(aes(x=xco(education,sex), y=income, color=sex), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x))  +
  scale_y() 

pop.scatter.bivariate + guides(color='none')
sample.scatter.bivariate + guides(color='none')
```

:::: columns
::: column
```{r}
styled.pop.tab.bysex = pop.tab |> 
  group_by(education, sex) |> 
  summarize(number=length(income), income.mean=mean(income), income.sd=sd(income)) |> 
  style.tab(cols=c('education', 'sex', 'income.mean',  'number'), nhead=2)
styled.pop.tab.bysex
```
$$
\mu(w,x)  = \frac{1}{m_{w,x}}\sum_{\substack{j\\w_j=w\\x_j=x}} y_j \qfor m_{w,x} = \sum_{\substack{j \\ w_j=w \\ x_j=x}} 1
$$
:::
::: column
```{r}
styled.sample.tab.bysex = sample.tab |> 
  group_by(education, sex) |> 
  summarize(number=length(income), income.mean=mean(income), income.sd=sd(income)) |> 
  style.tab(cols=c('education', 'sex', 'income.mean', 'number'), nhead=2)
styled.sample.tab.bysex
```
$$
\hat\mu(w,x) = \frac{1}{N_{w,x}}\sum_{\substack{i \\ W_i=w \\ X_i=x}} Y_i \qfor N_{w,x} = \sum_{\substack{i \\ W_i=w \\ X_i=x}} 1
$$
:::
::::

- We can work with finer groups defined using multiple covariates. Here, we use sex ($w$) and education ($x$).
- **Exercise**. What is the mean income for female California residents in our population with 9 years of education? 
- Locate it in the plot, write it in mathematical notation, then find the value in the table.

## Unbiasedness

:::: columns
::: column
```{r}
styled.pop.tab.bysex = pop.tab |> 
  group_by(education, sex) |> 
  summarize(number=length(income), income.mean=mean(income), income.sd=sd(income)) |> 
  style.tab(cols=c('sex','education', 'income.mean',  'number'), nhead=2)
styled.pop.tab.bysex
```
$$
\mu(x)  = \frac{1}{m_{w,x}}\sum_{\substack{j\\w_j=w\\x_j=x}} y_j \qfor m_{w,x} = \sum_{\substack{j \\ w_j=w \\ x_j=x}} 1
$$
:::
::: column
```{r}
styled.sample.tab.bysex = sample.tab |> 
  group_by(education, sex) |> 
  summarize(number=length(income), income.mean=mean(income), income.sd=sd(income)) |> 
  style.tab(cols=c( 'sex','education','income.mean', 'number'), nhead=2)
styled.sample.tab.bysex
```
$$
\hat\mu(w,x) = \frac{1}{N_{w,x}}\sum_{\substack{i \\ W_i=w \\ X_i=x}} Y_i \qfor N_{w,x} = \sum_{\substack{i \\ W_i=w \\ X_i=x}} 1
$$
:::
::::

- Subsample means are, in general, unbiased estimators of the corresponding subpopulation mean.

$$
\E[ \hat \mu(w,x) ] = \mu(w,x)
$$

- The proof is exactly the same as it was for a single binary covariate, so we won't repeat it here.


## Unbiasedness

:::: columns
::: column
```{r}
styled.pop.tab.bysex
```
$$
\mu(x)  = \frac{1}{m_{w,x}}\sum_{\substack{j\\w_j=w\\x_j=x}} y_j \qfor m_{w,x} = \sum_{\substack{j \\ w_j=w \\ x_j=x}} 1
$$
:::
::: column
```{r}
styled.sample.tab.bysex
```
$$
\hat\mu(w,x) = \frac{1}{N_{w,x}}\sum_{\substack{i \\ W_i=w \\ X_i=x}} Y_i \qfor N_{w,x} = \sum_{\substack{i \\ W_i=w \\ X_i=x}} 1
$$
:::
::::

- **Linearity of expectations** justifies using **linear combinations** of subsample means to estimate corresponding combinations of subpopulation means.
- For example, the value of a 4-year degree vs. a high school diploma for male California residents.
$$
\E[\hat \mu(\text{m},16)-\hat\mu(\text{m}, 12)] = \mu(\text{m}, 16) - \mu(\text{m}, 12) 
$$

- What population summary does this estimate? Describe it in mathematical notation then in words.
$$
\qty{\hat\mu(\text{m},16)-\hat\mu(\text{m}, 12)} - \qty{\hat\mu(\text{f},16)-\hat\mu(\text{f}, 12)}
$$

# Activity

## Exercise 1

- Today we're going to do a few exercises to practice translating between plain english, plots, mathematical notation, and actual numerical values. 
- They'll be versions of of the party game `telephone'. The first one is in teams of 4. 

1. Person 1 is given a description, in plain english, of some population summary. They translate it into a description in terms of 
a plot, write it down, and pass it to person 2.
2. Person 2 translates this into a mathematical formula, writes it down, and passes it to person 3.
3. Person 3 translates this back into a description in terms of a plot, writes it down, and passes it to person 4.
4. Person 4 translates this back into a description in plain english and writes that down.

- We'll then compare the original description to the final one to see how well the information was transmitted.
- If they don't match, we'll review the chain of descriptions together to see how well the information was transmitted.


## Exercise 2

- We'll do a second version of the game with a different end.
- As before, we'll start with a description of a population summary in plain english.
- But this time, we'll end with an **estimate**.

1. Person 1 is given a description, in plain english, of some population summary. They translate it into a description in mathematical notation, write it down, and pass it to person 2.
2. Person 2 writes down a sample summary that *estimates* the population one, in mathematical notation, and passes it to person 3.
3. Person 3 calculates the estimate, writes it down, and passes it to person 4. 
4. Person 4, who was given the same description as Person 1, collects all intermediate descriptions and checks the team's work.

- As before, we'll review these together to see how well the information was transmitted.

## Possible Estimation Targets

- The mean income for California residents with 12-year of education.
- The difference in mean incomes between male and female California residents with graduate degrees.
- The average incremental value of 10th, 11th, and 12th grade.
- The average value, for female California residents with a 10th-12th grade educations, of their last year of education.

## Bivariate Sample Summaries

:::: columns
::: column
```{r}
#| out-height: 500px
#| fig-asp: 1 
sample.scatter.bivariate + guides(color='none')
```
:::
::: column
```{r}
sample.tab |> group_by(education, sex) |> summarize(number = length(income), income.mean=mean(income)) |> 
  style.tab(cols=c('sex', 'education', 'income.mean', 'number'), nhead=Inf)
```
:::
::::


## Bivariate Population Summaries

:::: columns
::: column
```{r}
#| out-height: 500px
#| fig-asp: 1 
pop.scatter.bivariate + guides(color='none')
```
:::
::: column
```{r}
pop.tab |> group_by(education, sex) |> summarize(income.mean=mean(income), number = length(income)) |> 
  style.tab(cols=c('sex', 'education', 'income.mean', 'number'), nhead=Inf)
```
:::
::::
