---
title: "Lecture 12"
subtitle: "Multivariate Analysis and Regression Adjustment"
---

# Introduction

## Multivariate Analysis

{{< include shared-code.qmd >}}
{{< include ca-income.qmd >}}

```{r}
pop = pop.earnings 
sam = earnings 

xco = function(w,x) {
  w=as.numeric(w)
  x + .15*(2*w-1)
}

sam.dot = list(size=1,    alpha=.15)
pop.dot = list(size=.25,    alpha=.05)
scatter.by.group = function(data, dot=sam.dot, width=sd) { 
  ggplot(data) + 
    geom_point(aes(x=xco(group, education),  y=income, color=group), 
               size=dot$size, position=position_jitter(seed=0,w=.15), alpha=dot$alpha) +
    stat_summary(aes(x=xco(group, education), y=income, color=group), geom="pointrange", alpha=.7,
                 fun=mean, 
                 fun.min = function(x) mean(x) - width(x), 
                 fun.max = function(x) mean(x) + width(x))  +
    scale_x() + scale_y() + guides(color='none')
}

money = function(x) { sprintf('%2.1fk', round(x/1000, digits=1)) }  
```

```{r}
muhat.male = mean(sam$income[sam$sex=='male']) 
muhat.female = mean(sam$income[sam$sex=='female'])
sam$group = sam$sex == 'male'
scatter.by.group(sam)
```


- Today, we're going to talk about how to analyze data when we have multiple covariates.
- An important application is analyzing disparities to think about discrimination.
- Above, we see the kind of data we'll be working with.
- Each dot represents a California resident, age 25-35, who responded to the CPS.
  - $Y_i$, their income, is on the $y$-axis.
  - $X_i$, their education level, is on the $x$-axis.
  - $W_i$, their sex, is indicated by color. And a slight shift along the $x$-axis.
    - $\ingroupa{W_i=0}$ indicates female, $\ingroupb{W_i=1}$ indicates male.
- As usual, we've drawn in each group's income mean and standard deviation too.
- We'll write $\ingroupa{\hat\mu(0,x)}$ and $\ingroupb{\hat\mu(1,x)}$ for the mean income of female and male respondents with $x$ years of education. 


## A Few Summaries 

::: free-zoom

```{r}
#| out-width: 50%
#| layout-ncol: 2 
#| fig-cap: 
#|  - income vs. sex (aggregated)
#|  - income vs. education and sex (disaggregated)
 
Y = sam$income
X = sam$education
W = sam$sex == 'male'
muhat = Vectorize(function(w,x) { mean(Y[W==w & X==x]) })

dot = sam.dot
scatter.by.sex.only = ggplot(sam) + 
    geom_point(aes(x=sex,  y=income, color=group), 
               size=dot$size, position=position_jitter(seed=0,w=.15), alpha=dot$alpha) +
    stat_summary(aes(x=sex, y=income, color=group), geom="pointrange", alpha=.7,
                 fun=mean, 
                 fun.min = function(x) mean(x) - width(x), 
                 fun.max = function(x) mean(x) + width(x)) + 
    scale_y() + guides(color='none')

scatter.by.sex.only
scatter.by.group(sam)
```

:::

:::: columns
::: {.column width="48%"}
### Female Respondents
...earn \$`r money(muhat.male-muhat.female)` less, on average, than male respondents.
$$
\color{gray}
\begin{aligned}
\ingroupa{\hat\mu(0)} \ - \  \ingroupb{\hat\mu(1)}  
&\approx \ingroupa{`r money(muhat.female)`} \ - \ \ingroupb{`r money(muhat.male)`} \\
&\approx `r money(muhat.female - muhat.male)` 
\end{aligned}
$$
... earn `r 100*round(muhat.female/muhat.male, 2)` cents for every dollar a male one does.
$$
\color{gray}
\begin{aligned}
\ingroupa{\hat\mu(0)} \ / \ \ingroupb{\hat\mu(1)} 
&\approx \ingroupa{`r money(muhat.female)`} \ / \  \ingroupb{`r money(muhat.male)`} \\
&\approx {`r round(muhat.female/muhat.male, 2)`} \\
\end{aligned}
$$
:::
::: {.column width="52%"}
### ... with High School Educations 
... earn \$`r money(muhat(1, 12)-muhat(0, 12))` 
less than similarly-educated male ones.
$$
\color{gray}
\begin{aligned}
\ingroupa{\hat\mu(0, 12)} \ - \  \ingroupb{\hat\mu(1, 12)}  
&\approx \ingroupa{`r money(muhat(0, 12))`} \ - \ \ingroupb{`r money(muhat(1, 12))`} \\
&\approx `r money(muhat(0, 12)-muhat(1, 12))`
\end{aligned}
$$

... earn `r 100*round(muhat(0, 12)/muhat(1, 12), 2)` cents for every dollar a male one does.
$$
\color{gray}
\begin{aligned}
\ingroupa{\hat\mu(0, 12)} \ / \ \ingroupb{\hat\mu(1, 12)} 
&\approx \ingroupa{`r money(muhat(0, 12))`} \ / \  \ingroupb{`r money(muhat(1, 12))`} \\
&\approx {`r round(muhat(0, 12)/muhat(1, 12), 2)`} \\
\end{aligned}
$$
:::
::::

- These comparisons are relatively simple in that we're just reading two numbers off a plot.
- On the left, we compared aggregates---we used our plot of the aggregated data.
  - This is nice because the result is simple---we get one number.
- On the right, we compared fine-grained groups in the disaggregated data.
  - This is nice because it's more of a 'like-with-like' comparison.
- We can combine these advantages by **aggregating** like-with-like **comparisons**.
  - We call this *regression adjustment*.


## An Adjusted Comparison

```{r}
#| out-width: 100%
 
sam$group = sam$sex == 'male'
scatter.by.group(sam)
```

- Female respondents earn \$`r money(-mean(Y[W==0] - muhat(1,X[W==0])))` less, on average, than **similarly educated** male respondents.
$$
\color{gray}
\begin{aligned}
\text{adjusted difference} 
&= \ingroupa{\frac{1}{N_0}\sum_{i: W_i=0}}  \qty{ \ingroupa{Y_i} - \ingroupb{\hat \mu(1, X_i)} }
\approx `r money(mean(Y[W==0] - muhat(1,X[W==0])))` && \qfor 
N_w = \sum_{i: W_i=w} 1. \\
\end{aligned}
$$

- This is the average---over female respondents---of the difference between their income ...
- ... and the average income of male respondents with the same level of education.
- This is bigger than the difference we looked at a moment ago, of the average male and female incomes.
$$
\color{gray}
\begin{aligned}
\text{raw difference} 
&= \ingroupa{\frac{1}{N_0}\sum_{i: W_i=0} Y_i} - \ingroupb{\frac{1}{N_1}\sum_{i: W_i=1} Y_i}  
\approx `r money(mean(Y[W==0]) - mean(Y[W==1]))`.
\end{aligned}
$$

- What's going on here?  You may have a guess. But let's work through it in steps.

## Step 1. Using *dot language* to find what's changed

```{r}
#| out-width: 100%
 
scatter.by.group(sam)
```

:::: columns
::: {.column width="48%"}
### The Adjusted Difference
$$
\color{gray}
\ingroupa{\frac{1}{N_0}\sum_{i: W_i=0}}  \qty{ \ingroupa{Y_i} - \ingroupb {\hat \mu(1, X_i)} } 
$$
is what we get when we ...

  1. Look at the height of each [red dot]{.groupa}.
  2. Subtract the average height of the \
  [green dots]{.groupb} in the same column.
  3. Average these differences.
:::

::: {.column width="52%"}
### The Raw Difference
$$
\color{gray}
\ingroupa{\frac{1}{N_0}\sum_{i: W_i=0}} \qty{ \ingroupa{Y_i} - \ingroupb{\frac{1}{N_1}\sum_{i: W_i=1} Y_i } } \qfor N_w = \sum_{i: W_i=w} 1.
$$
is what we get when we ...
  
::: {.incremental}
  1. Look at the height of each [red dot]{.groupa}.
  2. Subtract the average height of *all of the [green dots]{.groupb}*.
  3. Average these differences.
:::
:::
::::

**Q.** Why is the adjusted difference is bigger in magnitude (i.e. smaller, being negative)?

::: fragment
**A.** For the average [female respondent]{.groupa}, [males]{.groupb} *at their education level* make more than the average *for all male respondents*. 
$$
\color{gray}
\text{same} - \text{bigger} = \text{smaller}
$$
:::
Why is that the case?

## Step 2. Making sense of the change

:::: r-stack
::: {.fragment .fade-out data-fragment-index="0"} 
```{r}
#| out-width: 100%
scatter.by.group(sam) 
```
:::
::: {.fragment .current-visible data-fragment-index="0"}
```{r}
#| out-width: 100%
scatter.by.group(sam) + 
  geom_bar(aes(x=xco(group, education), y=300000*after_stat(prop), fill=group, color=group), 
                 position='identity', alpha=.05, linewidth=.4) + guides(fill='none')
```
:::
::::

- It's the case because the [female respondents]{.groupa} tend to be better-educated than the [male ones]{.groupb}.
  - We call this phenomenon **covariate shift**. 
  - The distribution of $X$ *shifts* when we go from looking at one group (i.e. sex) to another.
- We can see this shift --- with a lot of effort --- by counting the dots in each column.
- But it's easier to see if we draw that information into the plot. &nbsp; [Do it!]{.button .navigate-next} 
- We draw in bars showing the proportion of people, within a group, that have each level of education. 
    - The bars' absolute height is meaningless---they're scaled to fit into the plot.
    - But the relative height tells us how education is distributed within each group.
    - The [red]{.groupa} and [green]{.groupb} bars show how education is distributed among [female]{.groupa} and [male respondents]{.groupb}.
    - *Notation.* We write $\ingroupa{\hat p_{0}(x)}$ and $\ingroupb{\hat p_{1}(x)}$ for these proportions at $x=x$.
- And we see that the [red bars]{.groupa} tend to be higher than the [green bars]{.groupb} on the right side of the plot, esp. for $x > 12$.
  - This may be a bit easier to see when we draw our bar plots side-by-side instead of interleaved @fig-covariate-shift-in-isolation.
  - *Interpretation*. Higher levels of education are more common for female respondents.


::: {#def-conditional-proportion}
$$
\hat p_{w}(x) = \frac{N_{w,x}}{N_w} \qqtext{ where }  \frac{N_{w,x}}{N_w} = \frac{\sum_{i: W_i=w, X_i=x} 1}{\sum_{i: W_i=w} 1}.
$$ 
:::

```{r}
#| out-width: 100%
#| label: fig-covariate-shift-in-isolation
#| fig-cap: The distribution of education among male and female respondents.
ggplot(sam) + geom_bar(aes(x=education, y=after_stat(prop), fill=group, color=sex), 
                 position='identity', alpha=.05, linewidth=.4) + 
                 guides(fill='none', color='none') + facet_wrap(~sex) 
``` 


## Covariate Shift and the Cumulative Distribution

```{r}
#| out-width: 100%
#| fig-asp: 1 
#| label: fig-covariate-shift-with-cdf
ggplot(sam) + geom_bar(aes(x=education, y=after_stat(prop), fill=sex, color=sex), 
                 position='identity', alpha=.05, linewidth=.2) + 
                 guides(fill='none', color='none') + stat_ecdf(aes(x=education, color=sex), geom='step', linetype='dashed')
``` 

- When we're looking for covariate shift, it can be helpful to look at the cumulative distribution.
  - Our bar plots show shows the proportion of people with *exactly* $x$ years of education.
  - The cumulative distribution is the proportion of people with *no more than* $x$ years of education.
  - That's the height of all bars up to and including the one at $x$, stacked. These are drawn in using dashed lines.
- What we see is that the [green one]{.groupb} is consistently higher than the [red one]{.groupa}. What does this mean? 

### A Thought Experiment [that needs refinement]
  
- Suppose we have equal numbers of male and female respondents. Close enough.
  - We line each group up, starting with the people with the least education and ending with those with the most. 
  - And put them side-by-side, effectively pairing up each male respondent with a female one and vice versa.
  - What do we know about the pairs?
- If the male respondents' cumulative distribution lies above the females' *at some $x$*, that means ...
  - there are too many male respondents with $\le x$ years of education to stand a female one next to each.
  - and conversely, too many female respondents with $> x$ years of education to stand a male one next to each.
- If this happens for every $x$, then *every female respondent* is standing next to a male respondent that is not more educated.
  - To convince ourselves, we move left to right ....
 [ Proof ?]
  - You could see why they'd be pissed if they grew up thinking that education is *the key* to making money. 
  - They're looking at someone who 'should' be making less.
  - But that's not what's happening.


## Homework Idea: Visualization Strengths and Weaknesses
```{r}
#| out-width: 100%
#| layout-ncol: 2 
#| layout-nrow: 3
#| label: fig-covariate-shift-with-cdf-hp
hp.ed.dist.overlaid = ggplot(sam[sam$race %in% 'Hawaiian/Pacific Islander only (HP)', ]) +
  geom_bar(aes(x=education, y=after_stat(prop), fill=sex, color=sex), 
                 position='identity', alpha=.05, linewidth=.2) + 
                 guides(fill='none', color='none') + 
                 stat_ecdf(aes(x=education, color=sex), geom='step', linetype='dashed')
hp.ed.dist = ggplot(sam[sam$race %in% 'Hawaiian/Pacific Islander only (HP)', ]) +
  geom_bar(aes(x=xco(sex, education), y=after_stat(prop), fill=sex, color=sex), 
                 position='identity', alpha=.05, linewidth=.2) + 
                 guides(fill='none', color='none') + 
                 stat_ecdf(aes(x=education, color=sex), geom='step', linetype='dashed')
hp.ed.dist 
hp.ed.dist.overlaid 
hp.ed.dist + facet_grid(rows=vars(sex))  
hp.ed.dist.overlaid + facet_grid(rows=vars(sex))
hp.ed.dist + facet_grid(cols=vars(sex))  
hp.ed.dist.overlaid + facet_grid(cols=vars(sex))
``` 

- This is not a real lecture slide. I'm just writing down an idea for a homework exercise.
- The idea is to compare different approaches to visualizing covariate distributions.
  - Interleaved bar plots  (with/without overlaid CDFS).
  - Side-by-side bar plots (with/without overlaid CDFs).
  - Faceted so we can compare heights easily vs. so we can compare center of mass easily.
  - etc.
- Some of these choices work together well. Some don't. Some are better for some purposes than others.
- On this slide I've subsetted the data to Hawaiian/Pacific Islanders because the covariate shift is more prominent.
- Might make sense to do this for a different group/the whole sample too and prompt them to think about why some visualizations are better for some samples, e.g. the CDF or interleaved histograms make it easier to do level-by-level comparisons which is helpful if the distributions are too similar to see a meaningful shift 
from a side-by-side comparison.


## Plan for Today

- Work out how to see what's going on by eye and relate it to our calculations.
- Famous Berkeley Example, Simpson's paradox.
- Finally, we'll talk a little about how to make sense of this stuff in the context of discrimination.
- A Warning---Statistics alone won't do it. Statistics is about measurement. 
- All it does for is is tell us how to use a sample to estimate population summaries.
- It doesn't tell us what these summaries mean.


- To get a feel for this stuff, let's work with a very small, stylized dataset.

# Covariate Shift in Stylized Data


```{r}
x = c(1,1,1,0,0,0)
y = c(0,3,3,0,0,3)
w = c(0,1,1,0,0,1)
jitter.y = .1*c(0,1,-1,1,-1,0)

exercise.grid = ggplot() + geom_point(aes(x=x, y=y+jitter.y), shape=1, size=5) + blank.theme
```

```{r}
#| out-height: 500px
#| fig-asp: .4

exercise.grid
```

- Draw in the means for the two columns.
- Then color 3 points red and 3 green so that the red 

the points so that
- Draw in the means for the two columns. 
- Color the points so that

blah blah blah.


# Main Bit

## The Black/White Income Gap in California


```{r}
filtered = \(rows) rows[rows$income < 5e5 & rows$race %in% c('White only', 'Black only'),]
  
pop = pop.earnings |> filtered()
sam = earnings |> filtered()
```

```{r}
xco = function(w,x) {
  w=as.numeric(w)
  x + .15*(2*w-1)
}

sam.dot = list(size=1,    alpha=.2)
pop.dot = list(size=.25,    alpha=.05)

scatter.by.race = function(rows, 
                           dot=sam.dot,
                           width=sd) { 
  ggplot(rows[rows$race %in% c('Black only', 'White only'), ]) + 
    geom_point(aes(x=xco(race=='White only',education), y=income, color=race), 
               size=dot$size, position=position_jitter(seed=0,w=.15), alpha=dot$alpha) + 
    stat_summary(aes(x=xco(race=='White only',education), y=income, color=race), geom="pointrange", alpha=.7,
                 fun=mean, 
                 fun.min = function(x) mean(x) - width(x), 
                 fun.max = function(x) mean(x) + width(x))  +
    scale_x() + scale_y() + guides(color='none')
}
```

```{r}
#| out-height: 350px
#| fig-asp: .4

W = as.numeric(sam$race == 'White only')
X = sam$education
Y = sam$income 
muhat = Vectorize(function(w,x) { mean(Y[W==w & X==x]) })

a = money
scatter.by.race(sam)
```

- Let's take a look at at income ($y$) broken down in terms of two variables.
  1. Education ($x$), in years of schooling, is indicated by column in the plot.
  2. Race ($w$) is indicated by color in the plot.  [$w=0$ for Black people]{.twocolor-red}  and [$w=1$ for white people]{.twocolor-green} in our sample.
- To keep things simple, we'll look at the subset of CPS recipients who identify as either [Black only]{.twocolor-red} or [white only]{.twocolor-green}.
- At most education levels, the [white people]{.twocolor-green} in our sample make more money than the [Black people]{.twocolor-red}.
- We know that's a thing. If we want, we can report a summary or two.


$$
\begin{array}{lll}
\text{raw income gap} \quad & \quad \text{adjusted income gap} & \\
\displaystyle \hat \Delta = \ingreen{\frac{1}{N_1}\sum_{i:W_i=1} Y_i} - \inred{\frac{1}{N_0}\sum_{i:W_i=0} Y_i} 
\quad& \displaystyle \hat \tau = \inred{\frac{1}{N_0}\sum_{i:W_i=0}} \ingreen{\hat \mu(1, X_i)} - \inred{\frac{1}{N_0}\sum_{i:W_i=0}} \inred{Y_i}
\quad& \text{symbolically} \\
        `{r} money(mean(Y[W==1]) - mean(Y[W==0]))`  
        = \ingreen{`{r} a(mean(Y[W==1]))`} - \inred{`{r} a(mean(Y[W==0]))`}   
\quad&  `{r} a(mean(muhat(1, X[W==0]) - muhat(0, X[W==0])))` 
       = \ingreen{`{r} a(mean(muhat(1, X[W==0])))`} - \inred{`{r} a(mean(muhat(0, X[W==0])))`}   
\quad& \text{numerically}
\end{array} 
$$


## Statistical Inference

```{r}
W = as.numeric(sam$race == 'White only')
X = sam$education
Y = sam$income 
muhat = Vectorize(function(w,x) { mean(Y[W==w & X==x]) })
the.adjusted.gap = mean(muhat(1, X[W==0]) - mean(muhat(0, X[W==0])))
the.raw.gap = mean(Y[W==1]) - mean(Y[W==0])

adjusted.gap = \(people) people |> group_by(education, race) |> 
                    summarize(muhat=mean(income), N=n(), .groups='drop') |> 
                  group_by(race) |>
                    mutate(P = N/sum(N)) |>
                   group_by(education) |>
                    group_map(\(s,e) s$P[s$race=='Black only'] * 
                    (s$muhat[s$race=='White only'] - s$muhat[s$race=='Black only'])) |>
                  unlist() |> sum()
raw.gap = \(people) mean(people$income[people$race=='White only']) - 
                    mean(people$income[people$race=='Black only'])
                    
stopifnot(abs(the.adjusted.gap - adjusted.gap(sam)) < 1e-6)
stopifnot(abs(the.raw.gap - raw.gap(sam)) < 1e-6)

set.seed(0)
B = 200
bootstrap.gaps = \(sam) 1:B |> map(\(B) {
  bootsam = sam[sample(1:nrow(sam), replace=TRUE), ]
  list(raw=raw.gap(bootsam), adjusted=adjusted.gap(bootsam))
}) |> bind_rows()
```

```{r}
#| cache: true
ca.bootstrap.gaps = bootstrap.gaps(sam)
```

```{r}
#| out-height: 350px
#| fig-asp: .5
#| layout-ncol: 2 
#| fig-cap: 
#|  - "Sample"
#|  - "Imagined Population" 
scatter.by.race(sam)
scatter.by.race(pop, dot=pop.dot)
```

- As usual, we can interpret these sample summaries as estimates of corresponding population summaries. 
- And we can construct 95% confidence intervals as usual---by bootstrapping. Nothing new here.
- What I want to talk about today is what these summaries mean and how they differ.
  - Incidentally, in California the raw and adjusted gaps are very similar.
  - But they are conceptually different and can be very different elsewhere. Let's look at Georgia.

$$
\begin{array}{lll}
\text{raw income gap} \quad & \quad \text{adjusted income gap} & \\
\displaystyle \hat \Delta = \ingreen{\frac{1}{N_1}\sum_{i:W_i=1} Y_i} - \inred{\frac{1}{N_0}\sum_{i:W_i=0} Y_i} 
\quad& \displaystyle \hat \tau = \inred{\frac{1}{N_0}\sum_{i:W_i=0}} \ingreen{\hat \mu(1, X_i)} - \inred{\frac{1}{N_0}\sum_{i:W_i=0}} \inred{Y_i}
\quad& \text{in sample} \\
\displaystyle \Delta = \ingreen{\frac{1}{m_1}\sum_{j:w_j=1} y_j} - \inred{\frac{1}{m_0}\sum_{j:w_j=0} y_j} 
\quad&\displaystyle \tau = \inred{\frac{1}{m_0}\sum_{j:w_j=0}} \ingreen{\mu(1, x_j)} - \inred{\frac{1}{m_0}\sum_{j:w_j=0}} \inred{y_j}
\quad& \text{in population} \\
        `{r} a(mean(Y[W==1]) - mean(Y[W==0]))` \pm `r a(2*sd(ca.bootstrap.gaps$raw))`
\quad&  `{r} a(mean(muhat(1, X[W==0]) - muhat(0, X[W==0])))` \pm `r a(2*sd(ca.bootstrap.gaps$adjusted))` 
\quad& \text{in 95\% of surveys}
\end{array} 
$$


## What Happens in Georgia 
```{r}
#| cache: true
georgia.state.code = 13
ga.sam = income(georgia.state.code)$sample
ga.bootstrap.gaps = bootstrap.gaps(ga.sam)
```

```{r}
#| out-height: 350px
#| fig-asp: .4
scatter.by.race(ga.sam)
```

- In Georgia, they're different. What are these two summaries?
  1. **Raw**. How much more the average white person makes than the average Black person.
  2. **Adjusted**. How much more, *for the average Black person*, a similarly-educated white person tends to make.
- The adjusted gap explains *part of* the raw gap. But not all of it.
- What's left over is, conceptually, about differences in the distribution of education.

$$
\begin{array}{lll}
\text{raw income gap} \quad & \quad \text{adjusted income gap} & \\
\displaystyle \hat \Delta = \ingreen{\frac{1}{N_1}\sum_{i:W_i=1} Y_i} - \inred{\frac{1}{N_0}\sum_{i:W_i=0} Y_i} 
\quad& \displaystyle \hat \tau = \inred{\frac{1}{N_0}\sum_{i:W_i=0}} \ingreen{\hat \mu(1, X_i)} - \inred{\frac{1}{N_0}\sum_{i:W_i=0}} \inred{Y_i}
\quad& \text{in sample} \\
\displaystyle \Delta = \ingreen{\frac{1}{m_1}\sum_{j:w_j=1} y_j} - \inred{\frac{1}{m_0}\sum_{j:w_j=0} y_j} 
\quad&\displaystyle \tau = \inred{\frac{1}{m_0}\sum_{j:w_j=0}} \ingreen{\mu(1, x_j)} - \inred{\frac{1}{m_0}\sum_{j:w_j=0}} \inred{y_j}
\quad& \text{in population} \\
        `{r} a(raw.gap(ga.sam))` \pm `r a(2*sd(ga.bootstrap.gaps$raw))`
\quad&  `{r} a(adjusted.gap(ga.sam))` \pm `r a(2*sd(ga.bootstrap.gaps$adjusted))` 
\quad& \text{in 95\% of surveys}
\end{array} 
$$


## Revisiting A Computational Trick

```{r}
#| out-height: 350px
#| fig-asp: .4
#| layout-ncol: 2 
#| fig-cap:
#|  - "Georgia"
#|  - "California"

scatter.with.hist = scatter.by.race(ga.sam) +
  geom_bar(aes(x=xco(race=='White only', education), y=400000*after_stat(prop), fill=race, color=race), 
                 position='identity', alpha=.05, linewidth=.4) + guides(fill='none') + scale_y(c(0,150000))
scatter.with.hist
scatter.by.race(sam) +
  geom_bar(aes(x=xco(race=='White only', education), y=400000*after_stat(prop), fill=race, color=race), 
                 position='identity', alpha=.05, linewidth=.4) + guides(fill='none') + scale_y(c(0,150000))
```

- To make sense of our two summaries, it's helpful to revisit a counting trick we used before.
  - When we want to average over the people in a sample, we can use a histogram to help us count.
  - This reduces a sum over hundreds or thousands of people to a sum over a few bins.

$$
\begin{aligned}
\hat \Delta &= \sum_{x} \qty{ \ingreen{P_{1,x} \hat\mu(1,x)} \quad - \quad \inred{P_{0,x} \ \hat\mu(0,x)} }            && \qqtext{where} \ P_{w,x} = \frac{N_{w,x}}{N_w} \\
\hat \tau   &= \sum_{x} \qty{ \inred{P_{0,x}} \ \ingreen{\hat\mu(1,x)} \quad - \quad \inred{P_{0,x} \ \hat\mu(0,x)} }    && \qqtext{for} \ \ \ N_w = \sum_{i:W_i=w} 1 \\ 
\hat \Delta - \hat \tau &= \sum_{x} \qty{\ingreen{P_{1,x}} - \inred{P_{0,x}}} \ingreen{\hat\mu(1,x)}                         && \qqtext{and} \ \ \ N_{w,x} = \sum_{i:W_i=w, X_i=x} 1
\end{aligned}
$$

- Looking at what happens in each bin can help us understand what's going on.
- Why were our raw and adjusted gaps similar in California but different in Georgia?
  

## Derivation: The Raw Gap
```{r}
#| out-height: 350px
#| fig-asp: .4
scatter.with.hist
```

$$
\begin{aligned}
\hat \Delta &= \frac{1}{N_1}\sum_{i:W_i=1} Y_i \quad -  \quad \frac{1}{N_0}\sum_{i:W_i=0} Y_i  && \qfor  N_{w} = \sum_{i:W_i=w} 1\\
            &= \sum_{x} \frac{1}{N_1}\sum_{i:W_i=1, X_i=x} Y_i   \quad - \quad \sum_{x}\frac{1}{N_0}\sum_{i:W_i=0, X_i=x} Y_i && \qqtext{summing within then over bins} \\
            &= \sum_{x} \frac{1}{N_1} \ N_{1,x} \ \hat\mu(1,x)  \quad - \quad \sum_{x} \frac{1}{N_0} \ N_{0,x} \ \hat\mu(0,x)     && \qfor N_{w,x} = \sum_{i:W_i=w, X_i=x} 1  \\
            &= \sum_{x} \ingreen{P_{1,x} \ \hat\mu(1,x)} \quad - \quad \sum_x \inred{P_{0,x} \ \hat\mu(0,x)}  && \qfor P_{w,x} = \frac{N_{w,x}}{N_w} \\
\end{aligned}
$$

## Derivation: The Adjusted Gap

```{r}
#| out-height: 350px
#| fig-asp: .4
scatter.with.hist
```


## Derivation: The Adjusted Gap
```{r}
#| out-height: 350px
#| fig-asp: .4
scatter.with.hist
```

$$
\begin{aligned}
\hat \tau &= \inred{\frac{1}{N_0}\sum_{i:W_i=0}} \ingreen{\hat \mu(1, X_i)} - \inred{\frac{1}{N_0}\sum_{i:W_i=0} Y_i} \\
          &= \sum_{x} \frac{1}{N_0}\sum_{i:W_i=1, X_i=x} \hat\mu(1,x)   - \sum_{x}\frac{1}{N_0}\sum_{i:W_i=0, X_i=x} Y_i && \qqtext{summing within then over bins} \\
          &= \sum_{x} \frac{1}{N_0} \ N_{0,x} \ \hat\mu(1,x)  - \sum_{x} \frac{1}{N_0} \ N_{0,x} \ \hat\mu(0,x)             && \qfor N_{w,x} = \sum_{i:W_i=w, X_i=x} 1  \\
          &= \sum_{x} \ingreen{P_{1,x} \ \hat\mu(1,x)} \quad - \quad \sum_x \inred{P_{0,x} \ \hat\mu(0,x)}  && \qfor P_{w,x} = \frac{N_{w,x}}{N_w} \\ \\
\end{aligned}
$$

## What To Adjust For?

```{r}
#| out-height: 350px
#| fig-asp: .4
#| layout-ncol: 2 
#| fig-cap:
#|  - "Georgia"
#|  - "California"

scatter.with.hist = scatter.by.race(ga.sam) +
  geom_bar(aes(x=xco(race=='White only', education), y=400000*after_stat(prop), fill=race, color=race), 
                 position='identity', alpha=.05, linewidth=.4) + guides(fill='none') + scale_y(c(0,150000))
scatter.with.hist
scatter.by.race(sam) +
  geom_bar(aes(x=xco(race=='White only', education), y=400000*after_stat(prop), fill=race, color=race), 
                 position='identity', alpha=.05, linewidth=.4) + guides(fill='none') + scale_y(c(0,150000))
```

- If we're trying to think about discrimination, it's not entirely clear what we should adjust for.
  - By looking at the adjusted gap, we're trying to make the people we compare 'similarly qualified'
  - But discrimination also *causes* differences in education.
  - So we're *adjusting away* some of the effect we might want to measure.
- We can try to formalize our comparisons in terms of potential outcomes. 
- We'll imagine that for each person in our population ... 
  -  there's one income they'd have if they were 'treated' by being raced white.
  -  and another income they'd have if they were 'treated' by being raced Black.

$$
\begin{aligned}
y_j(1) & \text{ income person $j$ would have if they were 'treated' by being raced white} \\
y_j(0) & \text{ income person $j$ would have if they were 'treated' by being raced Black}
\end{aligned}
$$


## What Do We Do With These Potential Outcomes?

$$
\begin{aligned}
y_j(1) & \text{ income person $j$ would have if they were 'treated' by being raced white} \\
y_j(0) & \text{ income person $j$ would have if they were 'treated' by being raced Black}
\end{aligned}
$$

- The individual effect of this treatment is the difference between these two.
- We can summarize this by the average of these differences for the Black people in our population.

$$
\begin{aligned}
\text{individual effect}_j 
&= y_j(1) - y_j(0) \\
\text{average effect} 
&= \frac{1}{m_0}\sum_{j:w_j=0} \qty{ y_j(1) - y_j(0) }  = \frac{1}{m_0}\sum_{j:w_j=0} y_j(1) - \frac{1}{m_0}\sum_{j:w_j=0} y_j(0) 
\end{aligned}
$$

- And $y_j(1)$ being unobservable for the people in this average, we try to find some proxy for it.
  - The adjusted-gap approach is to use the average income of white people with the same education level.
  - But what's the justification for this? Or something like this? There really isn't one. 
  - We've just disguised a fundamental problem in mathematical notation that works in other contexts. 
- You could argue that the problem is this 'treatment' isn't something you can assign, e.g., by randomization.
- But that's not really the problem. The problem is that the potential outcomes don't make sense.
- What does it mean to be person $j$,  except raced differently? It's meaningless.

## A Non-Solution --- Audit Studies

- People try to work around this by reframing the question in terms of *perception* of race. 
  - Because that's manipulable. You can, for example, choose the name on a job application.
  - You can randomize job applications to be 'Black' or 'white' by flipping a coin...
  - and then, on the basis of that, selecting the name of a Black or white census respondent at random.
- You can then mail these out and see who gets called back for an interview.

$$
\begin{aligned}
y_j(1) & \text{ whether application $j$ would get an interview when using a `random white name'} \\
y_j(0) & \text{ whether application $j$ would get an interview when using a `random Black name'}
\end{aligned}
$$

- Fundamentally, this is a version of our adjusted gap, but with more levels of $X$---one per application.
- The difference in interview rates is the called the 'effect of race' on *the decision-maker's actions*.
- But really, it's just the effect of a specific race signifier on the decision-maker's actions. 
  - If the idea of person $j$ except raced differently means nothing in the actual world...
  - ... then why do we imagine that it means something in the decision-maker's head?
- That doesn't mean that studies like this are useless. 
  - But we shouldn't act as if they get around the fundamental problem.
  - i.e., as if it works to think about discrimination using statistical tools alone.

## A Reference

::: {style="z-index:1000"}
{{< video  https://www.youtube.com/watch?v=l_LeJfn_qW0 width="854px" height="480px">}}
:::

- This critique of potential outcomes for thinking about discrimination is paraphrased
 from \
 a paper by [Issa Kohler-Hausmann](https://law.yale.edu/issa-kohler-hausmann). Roughly paraphased. I encourage you to read it yourself.
- The title refers to a 1984 SNL sketch in which Eddie Murphy conducts a revealing audit study. 

[Eddie Murphy and the Dangers of Counterfactual Causal Thinking About Detecting Racial Discrimination](https://openyls.law.yale.edu/bitstream/handle/20.500.13051/18009/Kohler-Hausmann,%20Eddie%20Murphy%20and%20the%20Dangers%20of%20Counterfactual%20Causal%20Thinking%20about%20Detecting%20Racial%20Discrimination.pdf?sequence=1) \
[Kohler-Hausmann](https://law.yale.edu/issa-kohler-hausmann), 2018
