---
title: "Sampling, Estimators, and Sampling Distributions"
---


# Review of Terms 

## Our E-Words

{{< include shared-code.qmd >}}
::: {.r-fit-text}
- *Estimand*: what we want to know about some population of interest 

- *Estimate*: our "guess" at the value of our estimand, formed through our procedure 

- *Estimator*: the procedure by which we generate our estimate 



- Why do we need an estimate? Why can't we directly measure the estimand? 
:::

## Population and Sampling 

- Estimands are a *population* level quantity 
  - For example, the proportion of all registered voters in GA who turn out to vote 
  - Or the average weight of all Chihuahuas in the world 
  - Or the difference in average income for all Americans with and without 4-year college degrees 
  
. . . 
  
- How would we measure these estimands? 

. . . 

- Let's take a *sample* and estimate the quantity instead! 

## Sampling Procedures 

- A *sample* is the subset of the population 

- There are lots of different ways to sample. In this course, we will use random sampling *with replacement* 
  - Sampling with replacement: each sampled unit is returned to the population (a.k.a replaced) before the next unit is drawn 
  
 . . . 
 
- Why do we sample with replacement? 
  - Imagine we are sampling Chihuahuas to weigh. About 1/4 of the population of Chihuahuas are obese (extremely overweight), about 1/4 are juveniles or underweight, and the rest are around average. 
  - We sample one Chihuahua at random from the population. What is the probability it is obese? 

 . . .
 
  - To see the importance of sampling with replacement, let's imagine we are sampling *without replacement*. Our first sampled Chihuahua was obese. What is the probability the next Chihuahua is obese? Is it 1/4? 

 . . . 
 
  - We sample with replacement so that the probability of selecting any particular unit is the same, regardless of whether it is the first draw or the 100th. This means the sampled units are *independent*
  
## Getting an Estimate 

- We took a random sample (with replacement) from our population... now what? 
- We want an estimate of our estimand (population quantity of interest)

- We need an *estimator* --a procedure we apply to our sample to generate our estimate 
- Examples of estimators? 

. . . 

- Any procedure that takes information from our sample and gives us a number (our estimate) is an estimator. But not all estimators might be *good* 
  - We will define what makes a good estimator more rigorously later in the semester 
  
. . . 

-For our Chihuahua example, our estimand is the average weight of all Chihuahuas so let's use the sample mean as our estimator 

$$ \frac{1}{n} \sum_i^n Y_i$$
where $n$ is the number of Chihuahuas in our sample and $Y$ is each sampled Chihuahua's weight 

## Aside: Sample Versus Population Notation 
-How can you tell whether we are talking about the population or sample? 
  - For population quantities of interest (estimands), we often use greek letters 
    - For example, $\mu$ is the population mean and $\sigma^2$ is its variance 
  - For estimates of those quantities, we use hats 
    - For example, $\hat{\mu}$ is the estimate of our population mean 

. . . 

- When talking about estimators applied to samples, we specify that it is the sample version 
  - For example, we calculate the sample mean and sample variance 
  - As distinguished from the mean and variance (of the population)

## Chihuahua Samples 

- On the next slide will be a table of Chihuahua weights
  - Let's pretend this is the whole population of interest 
  
- With a partner, you will sample randomly with replacement from this population. Your sample size is 5.
  - How to sample randomly: use a random number generator (1-20), a 20-sided die,  or a number spinner (1-20). That way each of the 20 Chihuahuas has an equal probability of being sampled 
  - Whatever number comes up for you, select that numbered Chihuahua for your sample 

- When you have your sample of 5 Chihuahuas, calculate the sample mean weight and submit it to the poll 
- Repeat! Keep taking new samples, calculating new sample means, and submitting them until we call stop! 

## Chihuahua Weights 

Table of 20 Chihuahua Weights (fake) for sampling 

(if possible add QR code for poll everywhere or open it separately)

Show poll everywhere histogram of sample means 

## Sampling Distributions 

- Was any individual estimate equal to the estimand? 

- We ended up with a *distribution* of estimates 
  - This is the *sampling distribution* of our estimator (the sample mean)
  
- What do we notice about the sampling distribution we just created? 

## Why Sampling Distributions Matter 

- No estimator is perfect: whenever we work with samples, our estimates will be too large or too small--maybe by a little bit, maybe by a lot 

- The sampling distribution tells us about the relationship between our estimates and our estimand. 
  - Is the sampling distribution centered around the estimand? 
  - Is it wide (disperse) or narrow? 
    - The wider the sampling distribution, the more likely we are to have an estimate far from the estimand 
    - We quantify the spread of the sampling distribution with the **standard error**, the standard deviation of the sampling distribution  
    
 . . .
 
- But we have the population of Chihuahuas... why did we make a sampling distribution? 

. . .

- When we have a single sample and a single estimate, we want to quantify our uncertainty about that estimate. Is it likely to be close to the estimand? How could we know, since we don't know the estimand? 
  - We will estimate a sampling distribution to quantify the uncertainty around our single estimate
  - We create a true sampling distribution from the population to compare our estimated versions to when we are learning 
    - Otherwise, how would we know if it the estimation procedures work? We need to compare them to the population to check ourselves, then we can use our procedures on future samples where we don't have the population 

# Polling Example 

{{< include ga-turnout.qmd >}}

```{r}
height = pmf.height = max(dbinom(0:n, n, mean(y))) 

scale_x = scale_x_continuous(breaks=seq(.6,.8, by=.025)) 
scale_y = scale_y_continuous(breaks=seq(0, pmf.height,by=.01))
coord = coord_cartesian(xlim=c(.6,.8), ylim=c(0, pmf.height))

bar.plot = ggplot() + geom_bar(aes(x=meanY.samples, y=after_stat(prop)), alpha=.3) +
                      geom_vline(aes(xintercept=mean(y)), color='green', size=2, alpha=.7) + 
                      scale_x + scale_y
```

## Context

- Suppose that, week before the 2024 presidential election, you did some polling.
  - You use a list of the $m \approx `r approx(m)`$ people registered to vote in Georgia.
  - And you make $n=`r n`$ phone calls.
  - Each call, you select a voter uniformly at random from the list, e.g. by rolling a `r approx(m)` -sided die.
  - And then you ask the potential voter whether they plan to vote.
- Suppose also that all these registered voters will
  - Pick up the phone will called
  - Respond honestly to your questions
  - Not change their minds about voting

## Polling Results

- You put your polling results in a table.
- It has one column for each call.
- In that columns, you record ...
  - the call number $i$---a number from $1 \ldots `r n`$.
  - the response $Y_i$ of the person you called---$1$ for 'yes and $0$ for no'.

$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}  \\
Y_i  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$ 

- You summarize the results with an extra column: the mean of the responses.
  - Remember that the mean of a binary variable is a proportion. 
  - You found that `r percent(mean(Y))` of the people polled said they would vote.

## Sample Versus Population

- What was our estimand? 
- What estimator are we using? 
- What is the estimate? 

. . . 

- When the election occurs, we get to see who turns out to vote.
  - `r approx(sum(y))` people, or roughly `r percent(mean(y))` of registered voters, actually vote.
 
\ \

:::: columns
::: column
### Before the Election 

$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}  \\
Y_i  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$

Outcomes for our sample of polled voters. \
To enumerate our sample, we ...

- give each call a number $i \in 1 \ldots `r n`$.
- write $Y_i$ for the turnout of the $i$th person we called.
:::
::: column
### After the Election

$$
\begin{array}{r|rrrrrr|r}
j & 1 & 2 & 3 & 4 & \dots &   `r approx(m)` & \bar{y}_{`r approx(m)`}  \\
y_{j} & `r y[1]` & `r y[2]` & `r y[3]` & `r y[4]` & \dots & `r y[m]`   & `r approx(mean(y))`  \\
\end{array}
$$
Outcomes for the population of registered voters. \
To enumerate our population, we ... 

- give each registered voter a number $j \in 1 \ldots `r approx(m)`$.
- write $y_j$ for the turnout of the person with ID $j$.
:::
::::

## Two Questions

Our sample proportion $`r sum(Y)` / `r n` \approx `r approx(mean(Y))`$ 
is close to the population proportion $`r approx(sum(y))` / `r approx(m)` \approx `r approx(mean(y))`$

1. Did we get lucky?
2. Could we have predicted how close we'd get *before* the election happened? In other words, could we quantify the uncertainty of our sample estimate before the estimand was revealed after the election? 


::: fragment

- To get these answers, we'll think about our sampling distribution.
  - What if our friends had run identical polls? 
  - Each friend will choose a different random sample.
  - And estimate the population mean (turnout proportion) by a different sample mean (turnout proportion in our sample)

:::

## Our Sample

- For each call $i$, we randomly select a voter with a Voter ID $J_i$.
- And we record as that call's outcome the turnout of that voter: $Y_i=y_{J_i}$.
- On each call, each registered voter has a $1/`r approx(m)`$ chance of being called.
  - Is this sampling with replacement?  

:::: {.r-stack}
::: {.fragment .fade-out data-fragment-index="0"}

```{r}
#| out-height: 300px 
sam$label = J
state.plot + geom_text(aes(x = long, y = lat, label=label), color=polla, alpha=.7, size=3, data=sam[c(1,2,n),]) 
```

:::
::: {.fragment .current-visible data-fragment-index="0"}

```{r}
#| out-height: 300px
state.plot + geom_text(aes(x = long, y = lat, label=label), color=polla, alpha=.7, size=3, data=sam)
```
:::
::: {.fragment .current-visible data-fragment-index="1"}
```{r}
#| out-height: 300px
state.plot + geom_point(aes(x = long, y = lat), color=polla, alpha=.7, size=3, data=sam)
```
:::
::::

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
  & J_1 & Y_1 & J_2 & Y_2 & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\text{value}    & `r J[1]` & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`} 
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`} 
                  & `r approx(mean(Y))` \\
\end{array}
$$ 


## Our First Friend's Poll

```{r}
J = J2
sam = pop[J,]
sam$label = J
Y = sam$y

state.plot + geom_point(aes(x = long, y = lat), color=pollb, alpha=.7, size=3, data=sam)
```

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
 & J_1 & Y_1 & J_2 & Y_2 & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\text{value}    & `r J[1]` & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`} 
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`} 
                  & `r approx(mean(Y))` \\
\end{array}
$$

## Our Second Friend's Poll

```{r}
J = J3
sam = pop[J,]
sam$label = J
Y = sam$y

state.plot + geom_point(aes(x = long, y = lat), color=pollc, alpha=.7, size=3, data=sam)
```

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
 & J_1 & Y_1 & J_2 & Y_2 & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\text{value}    & `r J[1]` & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`} 
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`} 
                  & `r approx(mean(Y))` \\
\end{array}
$$

## The Sampling Distribution of our Estimate

\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{pollster} & J_1 & Y_1 & J_2 & Y_2   & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\hline
\polla   1   
    & \polla   `r J=J1; Y=y[J]; J[1]`    & \polla  `r Y[1]` 
    & \polla   `r J[2]`   & \polla   `r Y[2]`           & \polla   \dots 
    & \polla   `r J[n]`   & \polla   `r Y[n]`           & \polla   `r approx(mean(Y))` \\
\pollb 2   
    & \pollb `r J=J2; Y=y[J]; J[1]`    & \pollb  `r Y[1]` 
    & \pollb `r J[2]`   & \pollb  `r Y[2]`          & \pollb \dots 
    & \pollb `r J[n]`   & \pollb `r Y[n]`           & \pollb `r approx(mean(Y))` \\
\pollc    3   
    & \pollc    `r J=J3; Y=y[J]; J[1]`    & \pollc     `r Y[1]` 
    & \pollc    `r J[2]`   & \pollc    `r Y[2]`           & \pollc    \dots 
    & \pollc    `r J[n]`   & \pollc    `r Y[n]`           & \pollc    `r approx(mean(Y))` \\
{\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots} &  {\vdots} \\ 
\polld    1M   
    & \polld    `r J=JM; Y=y[J]; J[1]`    & \polld     `r Y[1]` 
    & \polld    `r J[2]`   & \polld    `r Y[2]`           & \polld    \dots 
    & \polld    `r J[n]`   & \polld    `r Y[n]`           & \polld    `r approx(mean(Y))` \\
 {\vdots} & {\vdots} & {\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots}  & {\vdots} \\ 
\end{array}



## The Sampling Distribution in **R**

```{r}
#| echo: false
set.seed(default.seed)
```

We run 10,000 simulated polls and store who we call (J) and what they say (Y).
```{r}
#| echo: true
Js = array(dim=c(10000, n))
Ys = array(dim=c(10000, n))
for(rr in 1:10000) {
  Js[rr,] = sample(m, n, replace=TRUE)
  Ys[rr,] = y[Js[rr,]]
}
```

We calculate the sample proportion for each poll.
```{r}
#| echo: true
meanY.samples = rowMeans(Ys)
```

Before we histogram the result: What is the x-axis? What is the y-axis? What shape do you expect? 

. . . 
 
```{r}
#| echo: true
#| out-height: 400px
ggplot() + geom_bar(aes(x=meanY.samples, y=after_stat(prop)), alpha=.2)  
```
 

## Annotating the Sampling Distribution Histogram 

```{r}
#| out-height: 500px
samples = meanY.samples
estimand = mean(y)
Y=y[J1] 
samples[1] = mean(Y)
samples[2] = mean(y[J2])
samples[3] = mean(y[J3])
samples[4] = mean(y[JM])

dot.height = .7*height
dot.size = 2
scale_x = scale_x_continuous(breaks=seq(.6,.8, by=.025), limits=c(.6, .8))

turnout.bar = ggplot() + 
  geom_bar(aes(x=samples, y=after_stat(prop)), alpha=.2) + scale_x 

turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=0)
```

- There are a few features we may want to highlight.
  - The *mean of the sampling distribution* is the solid blue line.
  - The middle 2/3 of the sampling distribution lies between the dashed blue lines.
  - The middle 95\% of the sampling distribution lies between the dotted blue lines. 
- The *population proportion* is drawn as a wide green line.
  - It's in exactly the same place as the solid blue line. What does this mean? 


##  Revisiting our Questions

```{r}
#| out-height: 500px
p.as.close    = mean(abs(samples - mean(y)) <= abs(mean(Y)-mean(y))) 
w66 = width(samples, alpha=1/3)
w95 = width(samples, alpha=.05)

turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=1, dot.size=dot.size, height=dot.height)
```


**Observation**. Our sample mean --- the black dot --- is close to our population mean.


**Q**. Did we get lucky? 

::: fragment
Not really.

  - In 68% of polls, the sample means are within `r percent(w66/2)` of the population mean. 
  - In 95% of polls, the sample means are within  `r percent(w95/2)` of the population mean.
  
:::

##  Revisiting our Questions

```{r}
#| out-height: 500px
turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=1, dot.size=dot.size, height=dot.height)
```

**Q**. Could we have predicted how close we'd get before the election happened?

::: fragment
Yes, in sense.

- We could use an estimate of the standard error to describe the spread of our sampling distribution-- how different our estimates could be from the estimand 
  - We will use an interval estimate---a *range of values* the population proportion is likely to be in---based on our sample.
  - The *width* of this interval speaks to the 'how close' question. 
  - The *coverage probability* --- the probability our estimate is actually that close --- qualifies this answer.
:::

## Interval Estimation

```{r}  
#| out-height: 500px

w=.02
turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=1, interval.width=w, 
                   dot.size=dot.size, height=dot.height)
```

- Our *point estimate* of the population mean is the sample mean $\overline{Y}_{`r n`}$.
- We know our estimate isn't "correct" -- even before we saw the true mean
- So let's create an interval of width `r w` centered on it: $\ \ \overline{Y}_{`r n`} \pm `r w/2`$. 
  - We just arbitrarily picked this width 
- Ours doesn't cover --- does not include --- the population mean. 
- Is that just chance? Let's check out our friends' intervals.

## Interval Estimation: Our Friends' Polls

```{r}
#| out-height: 500px
turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=3, interval.width=w, 
                   dot.size=dot.size, height=dot.height)
```

- Lets's look at the interval estimates our first two friends.
- That is, the estimates based on the [pink]{.pollb} and [teal]{.pollc} rows in our sampling distribution table.
- One of these interval covers the population proportion.
- But that was just two friends... let's do this for a hundred more polls.

## Interval Estimation: 100 Polls

```{r}
#| out-height: 500px
turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=100, interval.width=w, 
                   dot.size=dot.size, height=dot.height)
```

- $`r sum(meanY.samples[1:100]-w/2 <= mean(y) & mean(y) <= meanY.samples[1:100]+w/2)`/100$ cover the population mean.
- This gives us a sense of the *probability* that our interval covers the population proportion.
- If we want to be more precise, we could do the same for millions of different polls.
- Let's not. Instead, let's find a more direct way to calculate coverage probability. 


##  Coverage Probability and Sampling Distributions
```{r}
#| out-height: 500px
turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=100, interval.width=w, 
                   dot.size=dot.size, height=dot.height, 
                   highlighted.dot.alpha=.15, unhighlighted.dot.alpha=.15)
```

Calculate the *coverage probability* of $\ \ \overline{Y}_{`r n`} \pm `r w/2`$ using the *sampling distribution* of our point estimate $\bar Y_{`r n`}$.

. . .
 
Hint 1: 

- For each poll's point estimate, we want to know if it's horizontal segments ("arms") that represent the interval touch the population mean. 
- But we don't want to have to draw in the horizontal segments for each poll.
- Is there another way to tell if the intervals will touch the population mean? 

. . . 

Hint 2: 

- Instead of making an interval around each of the polls, make an interval around the population mean.
- How wide should it be? 

## Calculating Coverage
```{r}
#| out-height: 500px
coverage.hint.plot = turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=100, interval.width=w, 
                   dot.size=dot.size, height=dot.height, 
                   unhighlighted.dot.alpha=.15, highlighted.dot.alpha=.15,
                   shade.width=w, shade.color='green', shade.alpha=.2)
coverage.hint.plot
```

- Let's shade in an interval of width .02 centered on the population proportion.
  - This gives it 'arms' the same length as our interval estimates have.
  - And its arms touch a point estimate *if and only if* the point estimates' arms touch it.
- That means we can count the intervals that cover by counting the point estimates between the dotted lines.
- What, in terms of the sampling distribution of the point estimate, is the coverage probability?

## The Coverage Probability
```{r}
#| out-height: 500px
coverage.hint.plot
``` 

It's the probability that a random draw from the sampling distribution lies in the green area.
$$
\text{coverage probability} = P\qty(\overline{Y}_{`r n`} \in \overline{y}_{`r approx(m)`} \pm .01) 
$$
And it's about 40\%.
```{r}
#| echo: true
mean(mean(y) - w/2 <= meanY.samples & meanY.samples <= mean(y) + w/2)
```

## Ad-hoc Interval Estimates
```{r}
#| out-height: 500px
coverage.hint.plot 
```

What we just did was choose a width and calculate a coverage probability.

  - The coverage probability we found --- 40% --- probably wasn't what we wanted.
  - 95% sounds a lot better. 

## Calibrated Interval Estimates
```{r}
#| out-height: 500px
coverage.hint.plot 
```

So instead of choosing a width and calculating the probability of coverage, we are going to choose our coverage probability (95%) and calculate the width that will give us that coverage

  - An interval estimate calibrated like this---to have a given coverage---is called a *confidence interval*.

## Using the Sampling Distribution for Calibration

```{r}
#| out-height: 500px
coverage.hint.plot 
```

**Activity**. 

  - You want to draw intervals around some poll's point estimates.
  - How wide should they be if you want 95\% of them to cover the population mean?
  
  . . . 
  
Hint: remember how we annotated the histogram before. What do the vertical lines mean? 

## Using the Sampling Distribution for Calibration

```{r}
#| out-height: 500px

w = width(samples, alpha=.05)
dotted.width = w
coverage.good.plot = turnout.bar + 
  with_annotations(samples, estimand=estimand, dots=100, interval.width=w, 
                   dot.size=dot.size, height=dot.height, 
                   unhighlighted.dot.alpha=.15, highlighted.dot.alpha=.15,
                   shade.width=w, shade.color='green', shade.alpha=.2)
coverage.good.plot
```

**Answer**.

- The width of a 95% interval should be the width between the dotted blue lines.
- That's the width of the 'arms' containing 95% of the estimates drawn from the sampling distribution.
  - We can check that these intervals have the coverage we want.

```{r}
#| echo: true
#|  
coverage = mean(mean(y)- dotted.width/2 <= meanY.samples & meanY.samples <= mean(y) + dotted.width/2)
coverage
```


## Confidence Intervals 

- Calibrated Intervals (a.k.a. confidence intervals) are helpful as they
  - Remind us that our estimate is never "correct" -- there are a range of values that could be the estimand 
  - Shows us how precise and stable our estimate is 
    - if the arms are really long (the interval is wide), that is implying our sampling distribution is wide--- our estimates would vary a lot depending on which sample we happened to draw 
    
- Calibrated Intervals (a.k.a. confidence intervals) are frustrating as they
  - can be difficult to interpret 
  - are calibrated based on the sampling distribution/population *which we usually don't have* 
    - and therefore must be *approximated* or *estimated* 

## A Problem 
```{r}
#| out-height: 500px

coverage.good.plot
```

- We can't calibrate intervals like this in real life.
  - When we run our a poll, we get a single point estimate $\bar Y_{`r n`}$ based on our sample.
  - We don't know the sampling distribution of this point estimate until election day.
- But what we actually do is *almost* the same. 
  - We *do* the same thing.
  - But we use an *estimate* of the sampling distribution in place of the thing itself.
- That's what we'll talk about next class 
