---
title: "Lecture 3"
subtitle: "Calibrating Interval Estimates"
---


# Review
## Polling

{{< include shared-code.qmd >}}
{{< include ga-turnout.qmd >}}



- We want to estimate the proportion of *all registered voters* --- our population --- who will vote.
- To do this, we use the proportion of *polled voters* --- our sample --- who said they would.
- When the election occurs, we get to see who turns out to vote.
  - `r approx(sum(y))` people, or roughly `r percent(mean(y))` of registered voters, actually vote.

- While we only collected one sample of 625 voters, we thought about all the samples of 625 voters that could have been chosen and the resulting sample means from each of those to create a *sampling distribution*  


## The Sampling Distribution of our Estimate

```{r}
#| out-height: 400px
#| fig-asp: .3 

set.seed(default.seed)
Js = array(dim=c(10000, n))
Ys = array(dim=c(10000, n))
for(rr in 1:10000) {
  Js[rr,] = sample(m, n, replace=TRUE)
  Ys[rr,] = y[Js[rr,]]
}
meanY.samples = rowMeans(Ys)
pmf.height = max(dbinom(0:n, n, mean(y))) 

scale_x = scale_x_continuous(breaks=seq(.6,.8, by=.025)) 
scale_y = scale_y_continuous(breaks=seq(0, pmf.height,by=.01))
coord = coord_cartesian(xlim=c(.6,.8), ylim=c(0, pmf.height))

bar.plot = ggplot() + geom_bar(aes(x=meanY.samples, y=after_stat(prop)), alpha=.3) +
                      geom_vline(aes(xintercept=mean(y)), color='green', size=2, alpha=.7) + 
                      scale_x + scale_y
                      
bar.plot + with_annotations(meanY.samples, height=pmf.height, highlight.dots=4) + coord
```

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{poll} & J_1 & Y_1 & J_2 & Y_2   & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\hline
\polla   1   
    & \polla   `r J=J1; Y=y[J]; J[1]`    & \polla  `r Y[1]` 
    & \polla   `r J[2]`   & \polla   `r Y[2]`           & \polla   \dots 
    & \polla   `r J[n]`   & \polla   `r Y[n]`           & \polla   `r approx(mean(Y))` \\
\pollb 2   
    & \pollb `r J=J2; Y=y[J]; J[1]`    & \pollb  `r Y[1]` 
    & \pollb `r J[2]`   & \pollb  `r Y[2]`          & \pollb \dots 
    & \pollb `r J[n]`   & \pollb `r Y[n]`           & \pollb `r approx(mean(Y))` \\
\pollc    3   
    & \pollc    `r J=J3; Y=y[J]; J[1]`    & \pollc     `r Y[1]` 
    & \pollc    `r J[2]`   & \pollc    `r Y[2]`           & \pollc    \dots 
    & \pollc    `r J[n]`   & \pollc    `r Y[n]`           & \pollc    `r approx(mean(Y))` \\
{\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots} &  {\vdots} \\ 
\polld    1M   
    & \polld    `r J=JM; Y=y[J]; J[1]`    & \polld     `r Y[1]` 
    & \polld    `r J[2]`   & \polld    `r Y[2]`           & \polld    \dots 
    & \polld    `r J[n]`   & \polld    `r Y[n]`           & \polld    `r approx(mean(Y))` \\
 {\vdots} & {\vdots} & {\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots}  & {\vdots} \\ 
\end{array}
$$


## Interval Estimation

```{r}
#| out-height: 400px
#| fig-asp: .3 

w = .02
bar.plot + with_annotations(meanY.samples, interval.width=w, height=pmf.height, highlight.dots=4) + coord 
```

- To give a sense of what this tells us about the population mean, we can report an interval estimate.
    - We started with an arbitrary width of .02. i.e., we've drawn in the interval $\bar Y_{`r n`} \pm `{r} w/2`$.
- But we really care about an interval's *coverage probability*. 
- We need to **calibrate** our interval estimates to give us a coverage probability of 95%


## Calibrating Interval Estimates

```{r}
#| out-height: 400px
#| fig-asp: .3 

oldw = w
w=width(meanY.samples, alpha=.05)
bar.plot + with_annotations(meanY.samples, interval.width=oldw, shade.width=w, height=pmf.height, highlight.dots=4) + coord 
```

- Our arbitrary width of 0.02 didn't give us the coverage we wanted
- To **calibrate** this width to give us 95% coverage---which is the convention--- we expand our [shaded region]{.fg style="color:limegreen"} until it covers 95% of the histogram's mass, i.e., 95% of our point estimates.
- If we use intervals of this width around our point estimates, we'll get 95% coverage.

## Calibrating Interval Estimates
```{r}
#| out-height: 400px
#| fig-asp: .3 

bar.plot + with_annotations(meanY.samples, interval.width=w, shade.width=w, height=pmf.height, highlight.dots=4) + coord 
```

- Our arbitrary width of 0.02 didn't give us the coverage we wanted
- To **calibrate** this width to give us 95% coverage---which is the convention--- we expand our [shaded region]{.fg style="color:limegreen"} until it covers 95% of the histogram's mass, i.e., 95% of our point estimates.
- If we use intervals of this width around our point estimates, we'll get 95% coverage.

## The Problem

```{r}
#| out-height: 400px
#| fig-asp: .3 

bar.plot + with_annotations(meanY.samples, interval.width=w, shade.width=w, height=pmf.height, highlight.dots=4) + coord 
```

- Unless we've seen the population, i.e. unless it's after the election, we can't run these simulated polls to create the sampling distribution.
- That means we can't calibrate our interval using the sampling distribution of our point estimate. 
- So we'll actually do calibration using **an estimate of the sampling distribution**.

# Real-Life Calibration

## Why we can calibrate using an Estimated Sampling Distribution
```{r}
#| out-height: 400px
#| fig-asp: .3 

set.seed(default.seed)
estimated.dist.samples = rbinom(10000, n, mean(Y)) / n 
w.est=width(estimated.dist.samples, alpha=.05)

bar.plot.estimated = ggplot() + 
    geom_bar(aes(x=estimated.dist.samples, y=after_stat(prop)), alpha=.3) +
    geom_vline(aes(xintercept=mean(y)), color='green', size=2, alpha=.7) + 
    scale_x + scale_y
bar.plot.estimated + with_annotations(estimated.dist.samples, interval.width=w.est, highlight.dots = 0,  
                                      shade.width=w.est, shade.color='red', height=pmf.height) + coord
```

- Here we've estimated the sampling distribution --- how is coming next --- and used it to find an interval width.
- We found the width which if we draw arms around the mean of our estimated sampling distribution will cover 95% of draws from our estimated sampling distribution.
- Anything look weird here? 

. . . 

- Our estimated sampling distribution is a bit off--- the estimated sampling distribution is not centered on the estimand (population mean)
- Because it's centered on our sample mean $\bar Y$.
- It turns out that this doesn't matter much. Why not?
  - What does calibration tell us? 

## Calibration Comparison

```{r}
#| out-height: 400px
#| fig-asp: .3 


K = 100
heights = annotation_dot_heights(K, 3, pmf.height)
calibration.comparison.histogram = bar.plot +
    geom_rect(aes(xmin=mean(y)-w/2, xmax=mean(y)+w/2, ymin=0, ymax=1.25*heights[1]), alpha=.15, fill='green') +
    geom_rect(aes(xmin=mean(y)-w.est/2, xmax=mean(y)+w.est/2, ymin=1.25*heights[1], ymax=Inf), alpha=.15, fill='red')


calibration.comparison.histogram +  
    geom_pointrange(aes(y=heights[1]*1.5, x=meanY.samples[1], 
                        xmin = meanY.samples[1]-w.est/2, xmax = meanY.samples[1]+w.est/2), alpha=1, color='black')  +
    geom_pointrange(aes(y=heights[1], x=meanY.samples[1], 
                        xmin = meanY.samples[1]-w/2, xmax = meanY.samples[1]+w/2), alpha=.4, color='black')   
```

- Calibration is about choosing the width of our interval, not where it is located! 
  - We're not putting arms on draws from the estimated sampling distribution.
  - We're putting arms on our point estimate, which is a draw from the **actual sampling distribution**.
- It works because that's very close to the width we'd get from the actual sampling distribution.
- The shaded regions correspond to two versions of the the [population proportion's]{.fg style="color:limegreen"} arms.
    - The [green region]{.fg style="color:limegreen"} is the one we get from the actual sampling distribution. We can't use this because in reality we won't have the actual sampling distribution.
    - The [red region]{.fg style="color:red"} is the one we get from the estimated sampling distribution. We can --- and do --- use this.
- The interval estimates centered on our poll (sample mean) using both widths shows they are almost the same.

## Estimating Sampling Distributions 
- The sampling distribution tells us 
  - the standard error 
  - the calibrated interval width for 95% coverage (or whatever coverage we want)

- An estimated sampling distribution tells us 
  - the **estimated** standard error 
  - the **approximately** calibrated interval width for whatever coverage we want 
  
. . . 

- We will discuss two ways to estimate the sampling distribution 
  - Bootstrapping 
  - Normal Approximation 

## Estimating the Sampling Distribution with the Bootstrap 

- Bootstrapping is a *non-parametric* procedure for estimating the sampling distribution
  - Non-parametric means we aren't relying on any assumptions about the form or parameters of the underlying distribution
  - It is very flexible-- can be used with lots of different estimators and sampling designs 
  - But needs a sufficiently large sample size and can be computationally intensive 
    - Has become a lot more popular as estimators get more complicated and computers get better! 
    
. . . 

The procedure: 

- Sample with replacement from *your sample* using the same sample size as your sample. This is called a **bootstrap sample**.
- Calculate your estimate for your bootstrap sample. This is your **bootstrap estimate**
- Repeat many times (10,000 times usually)
- The resulting distribution of bootstrap estimates is your **bootstrap estimated sampling distribution** 

## Bootstrap Estimated Sampling Distribution 

```{r}
#| echo: true

bootstrap.samples = array(dim=10000)
for(rr in 1:10000) {
    Y.star = Y[sample(1:n, n, replace=TRUE)]
    bootstrap.samples[rr] = mean(Y.star) 
}
```

```{r}
#| out-height: 400px 
#| fig-asp: .3 
ggplot() + geom_bar(aes(x=bootstrap.samples, y=after_stat(prop)), alpha=.3, position='identity') 
```


- We are *acting as if* our sample were the population to estimate the sampling distribution 
- Note that this is random because we're drawing with replacement
  - Each time we re-sample from our sample, we call each person in our sample 0,1,2,... times
    - The number of times we call them is random
    
## Practice Bootstrap 

- Back to our silly Chihuahuas! We have a random sample of 10 Chihuahuas (a random sample of our population of 20 from last class). With a partner, draw a bootstrap sample and calculate the sample mean (your bootstrap estimate). 
  - What size should your bootstrap sample be? 
- Keep taking new bootstrap samples and submitting your bootstrap estimates to the poll until we call stop. 
  - What should the distribution look like? Where should it be centered? 
  

## Sample Chihuahua Weights 

This is a small sample so you can do it by hand, but we need a larger sample size really conduct bootstrapping 

You will practice in bootstrapping in R in lab and in homework. 

## Bootstrapped Poll

- Let's go back to our Georgia voter poll. Our sample size of 625 is sufficient to apply our bootstrap procedure to estimate the sampling distribution. 
- Here is our bootstrapping code 

```{r}
#| echo: true

meanY.boot = array(dim=10000)
for(rr in 1:10000) {
    I = sample(1:n, n, replace=TRUE)
    meanY.boot[rr] = sum(Y[I])/n
}
```

- What is $n$? 
- What is I? 
- What is meanY.boot? 

## Identifying our Bootstrap Estimated Sampling Distribution 

```{r}
#| out-height: 400px
#| fig-asp: .7
#| layout-ncol: 2 
ggplot() + geom_bar(aes(x=bootstrap.samples, y=after_stat(prop)), alpha=.3, position='identity') 

ggplot() + geom_bar(aes(x=dfsam$y, y=after_stat(prop)), alpha=.3, position='identity') 
```

- One of these plots is the distribution of $Y$-- the turnout of voters in our sample. One of these plots is the bootstrap estimated sampling distribution of $\bar{Y}$. 
- Which is which? How do you know? 

## Using the Estimated Sampling Distribution

- What was the point of estimating the sampling distribution? What do we DO with it?

. . . 

- We want to estimate our standard error ... and 
- We want to calibrate the width of our interval around our point estimate so that it is the same width as the middle 95% of our sampling distribution. 

. . . 

- Review: what is standard error? (poll)

## Estimated Standard Error 

- We want the dispersion of the sampling distribution... 
- We have the estimated sampling distribution... 
- Let's find the dispersion of the estimated sampling distribution! 

. . . 

```{r}
#| echo: true

meanY.boot = array(dim=10000)
for(rr in 1:10000) {
    I = sample(1:n, n, replace=TRUE)
    meanY.boot[rr] = sum(Y[I])/n
}

sd(meanY.boot)
```

## Approximately Calibrated Interval 
- We want the width of the middle 95% of the sampling distribution 
- Let's use the middle 95% of the estimated sampling distribution to approximate it 

```{r}
#| out-height: 400px
#| fig-asp: .3 

set.seed(default.seed)
estimated.dist.samples = rbinom(10000, n, mean(Y)) / n 
w.est=width(estimated.dist.samples, alpha=.05)

calibration.histogram = ggplot() + geom_bar(aes(x=bootstrap.samples, y=after_stat(prop)), alpha=.3, position='identity') +
    geom_rect(aes(xmin=mean(Y)-w.est/2, xmax=mean(Y)+w.est/2, ymin=0, ymax=Inf), alpha=.15, fill='red') + geom_vline(aes(xintercept=mean(Y)), color='blue', size=1, alpha=.5)

calibration.histogram +
    geom_pointrange(aes(y=heights[1], x=mean(Y), 
                        xmin = mean(Y)-w.est/2, xmax = mean(Y)+w.est/2), alpha=.4, color='black')   
```

- How do we get this exact width? 

. . . 

```{r}
#| echo: true

meanY.boot = array(dim=10000)
for(rr in 1:10000) {
    I = sample(1:n, n, replace=TRUE)
    meanY.boot[rr] = sum(Y[I])/n
}

quantile(meanY.boot, probs=c(0.025,0.975))
```

## How did we do? 

```{r}
#| out-height: 300px
#| layout-ncol: 2 
#| fig-cap:
#|   - "Actual Sampling Distribution"
#|   - "Bootstrap Estimated Sampling Distribution "
bar.plot + with_annotations(meanY.samples, height=pmf.height, highlight.dots=4) + coord +
geom_rect(aes(xmin=mean(y)-w/2, xmax=mean(y)+w/2, ymin=0, ymax=Inf), alpha=.15, fill='green') + geom_pointrange(aes(y=heights[1], x=mean(y), 
                        xmin = mean(y)-w/2, xmax = mean(y)+w/2), alpha=.4, color='black') 

calibration.histogram +
    geom_pointrange(aes(y=heights[1], x=mean(Y), 
                        xmin = mean(Y)-w.est/2, xmax = mean(Y)+w.est/2), alpha=.4, color='black')   

```

```{r}
#| echo: true

#Middle 95% of Sampling Distribution 
quantile(meanY.samples, probs=c(0.025,0.975))

#Middle 95% of Bootstrap Estimated Sampling Distribution 
quantile(meanY.boot, probs=c(0.025,0.975))

```


## Normal Approximation 

- Besides the bootstrap, we can also use a Normal approximation to estimate the sampling distribution 
  - Normal as in our old friend, the Normal Distribution 
  - To fully describe a Normal Distribution, we need two parameters. What are they? 
  
. . . 

- $\mu$, the mean/location 
- $\sigma^2$, the variance/spread 

. . . 

- Most things involving averages are approximately normal. This is a consequence of the **Central Limit Theorem**. 
  - There are a lot of versions of the Central Limit Theorem with slightly different proofs. We will not spend time on that here. 
  
## Normal Approximation for the Sampling Distribution 

- We want to estimate the sampling distribution of our sample mean 
- We know from the Central Limit Theorem it should be approximately normal, $N(\mu,\sigma^2)$, but we don't know what to plug in for $\mu$ and $\sigma^2$ 

- Where should we center our estimated sampling distribution? 

. . .  

- The sample mean seems like a good estimate for $\mu$. We will prove that it is next time. 

- What is the spread of our sampling distribution? What should we plug in for $\sigma^2$? 

. . . 

- We don't want the variance of $Y$, we want the variance of $\bar{Y}$, our estimator
  - Sometimes the variance of our estimator is straightforward to derive, sometimes it is not 
  - We will derive the variance of $\bar{Y}$ next week 
  - Spoiler alert: its $\frac{\sigma^2}{n}$ where $\sigma^2$ is the variance of $y$ and $n$ is the sample size 
  
## Estimated Standard Error (Normal Approximation)
- We want the dispersion of the sampling distribution... 
- We know the variance of the sampling distribution of our sample mean is $\frac{\sigma^2}{n}$ 
  - So its standard deviation is $\frac{\sigma}{\sqrt{n}}$
- But we still don't know $\sigma$, the standard deviation of $y$ 

. . . 

- We can use $s$, the standard deviation of our sample, as an estimate of $\sigma$, the standard deviation of the population 
- Is $s$ a good approximation for $\sigma$? 
  - If the sample size is large (and random), probably 

. . . 

Our estimated standard error: $\frac{s}{\sqrt{n}}$

```{r}
#| echo: true
sd(Y)/sqrt(length(Y))
```

  
## The Middle 95% of the Normal 

- We got our estimated standard error, but what about calibrating our 95% Confidence Interval? 
- Our sampling distribution is approximately normal by CLT. How do we get the middle 95% of a normal curve? [add image with 68% 95% etc]

. . .  

- 95% of the area under the normal is within 1.96 standard deviations from the center
- So the middle 95% will be $\mu \pm 1.96* \frac{\sigma}{\sqrt{n}}$ 
- We don't have $\mu$ or $\sigma$ but we can use our estimates! 

. . . 

- Normal approximation 95% Confidence Interval: $\bar{Y} \pm 1.96* \frac{s}{\sqrt{n}}$

```{r}
#| echo: true
mean(Y) - 1.96*sd(Y)/sqrt(length(Y))
mean(Y) + 1.96*sd(Y)/sqrt(length(Y))
```

## How did we do? 

```{r}
#| echo: true

#Middle 95% of Sampling Distribution 
quantile(meanY.samples, probs=c(0.025,0.975))

#Middle 95% of Estimated Sampling Distribution using Normal Approximation 
mean(Y) - 1.96*sd(Y)/sqrt(length(Y))
mean(Y) + 1.96*sd(Y)/sqrt(length(Y))

```


## Using a Sample 

- We used our sample to make these estimates/approximations and it seemed to work okay

. . . 

- But what if we had a different sample? Did we just get lucky with a "good" sample? 



## Each of Our Friends Gets Their Own Estimate

```{r}
#| out-height: 300px
#| layout-ncol: 2 
#| fig-asp: .4 
#| fig-cap:
#|   - "Actual Sampling Distribution"
#|   - "Estimated Sampling Distribution (Poll 1)"
#|   - "Estimated Sampling Distribution (Poll 2)"
#|   - "Estimated Sampling Distribution (Poll 3)"

plot.interval = function(Y, height, color) {
    set.seed(default.seed)
    samples = rbinom(10000, n, mean(Y)) / n 
    w=width(samples, alpha=.05)
    geom_pointrange(aes(x=mean(Y), y=height, xmin=mean(Y)-w/2, xmax=mean(Y)+w/2), color=color)
}

dh = .2*heights[1]
bar.plot + with_annotations(meanY.samples, shade.width=w, highlight.dots=3, highlighted.dot.alpha=.4) +
               geom_line(aes(x=x/n, y=dbinom(x, n, mean(y))), color='gray') +
               plot.interval(y[J1], dh+heights[1], color=polla) +
               plot.interval(y[J2], dh+heights[2], color=pollb) +
               plot.interval(y[J3], dh+heights[3], color=pollc) + coord

calibration.diagram = function(Y, color, height) {
    set.seed(default.seed)
    samples = rbinom(10000, n, mean(Y)) / n 
    w=width(samples, alpha=.05)
    histplot = ggplot() + geom_bar(aes(x=samples, y=after_stat(prop)), alpha=.3) +
                          geom_vline(aes(xintercept=mean(y)), color='green', size=3, alpha=.7) + 
                          with_annotations(samples, shade.width=w, shade.color='red', height=pmf.height, highlight.dots=0) +
    geom_line(aes(x=x/n, y=dbinom(x, n, mean(Y))), color=color) +
    geom_pointrange(aes(x=mean(Y), xmin=mean(Y)-w/2, xmax=mean(Y)+w/2, y=height), color=color)
}

calibration.diagram(Y,     polla, heights[1]) + coord
calibration.diagram(y[J2], pollb, heights[2]) + coord
calibration.diagram(y[J3], pollc, heights[3]) + coord
```

- This means each of our friends estimates a different width for their interval because it is based on their individual sample.
- But they're all close to the 95% width we want using the actual sampling distribution. 
- Were our friends lucky too? 


## Evidence That It Works 

```{r}
#| out-height: 400px
K = 100
ps = meanY.samples[1:K]

ggplot() + geom_line(aes(x=x/n-p, y=n*dbinom(x,n,p)), data=expand.grid(p=mean(y), x=0:n)) +
           geom_vline(aes(xintercept=qbinom(c(.025,.975), n,mean(y))/n-mean(y)), linetype='dotted') +
           geom_line(aes(x=x/n-p, y=n*dbinom(x,n,p), color=factor(p)), data=expand.grid(p=ps, x=0:n), alpha=.2) +
           geom_pointrange(aes(x=0, xmin=qbinom(.025,n,p)/n-p, xmax=qbinom(.975,n,p)/n-p, y=20*k/K, color=factor(p)), data=data.frame(p=ps, k=1:K)) +
           xlim(-.1,.1) + guides(color=FALSE)
```

- We see the same thing when we look at 100 polls.
- All 100 estimated sampling distribution are close to the actual sampling distribution.
- The resulting interval-widths are near-indistinguishable from the width we got using the sampling distribution.
  - The estimated widths are shown as points with arms.
  - That sampling distribution's actual width is the space between the dotted lines.
- You can see estimating the width works reliably!

## Interpretating Confidence Intervals in Practice 

- So we have our point estimate (e.g. sample mean) and use one of our procedures to approximately calibrate a 95% Confidence Interval 
- Can we say for sure whether our interval includes the estimand (e.g. population mean)?

. . . 

- No! What if we got a sample from the tails of the sampling distribution? 

```{r}
#| out-height: 400px
bar.plot + with_annotations(meanY.samples, height=pmf.height, highlight.dots=4)  
```

- But remember this is unlikely. How unlikely? 

## Interpreting Confidence Intervals in Practice 

- If we had the perfectly calibrated 95% Confidence Interval width from the sampling distribution, we could say there is a 95% chance our sample came from the middle 95% and therefore there is a 95% chance our interval includes the estimand 

- **But we don't** 

- We don't have the sampling distribution for perfect calibration, we had to *approximate* the width of our interval for 95% calibration 

- So we are left with the clunky interpretation "if we repeated our estimator (sample mean of our poll) infinite times, our intervals constructed this way would include the population mean 95% of the time" 
  - Because our approximation of the width works in the limit (infinite repetitions), but we can't be sure it is right for an individual estimate 

