---
title: "Lecture 3"
subtitle: "Calibrating Interval Estimates"
---

```{r setup}
library(usmap)
library(sf)
library(ggplot2)
library(scales)
library(knitr)
library(magrittr)
library(tidyverse)

percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }
```

```{r ggplot-themes}
class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
            panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
            axis.ticks.x = element_blank(),
            axis.ticks.y = element_blank(),
            axis.text.x  = element_text(colour = "#aaaaaa"),
            axis.text.y  = element_text(colour = "#aaaaaa"),
            axis.title.x  = element_text(colour = "#aaaaaa"),
            axis.title.y  = element_text(colour = "#aaaaaa", angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
                axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
             panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(simplified.theme)
```

# Review

## Polling
::: {.hidden}

```{r population}
our.seed = 0
set.seed(our.seed)
N=7233584
p=4999960/N
y= c(rep(1,p*N), rep(0, (1-p)*N))
y=sample(y) # shuffle the order 
population.proportion = p
m=N
```

```{r sample}
set.seed(our.seed)

n = 625
J = sample(N, n, replace=TRUE)
Y = y[J]
sample.proportion = mean(Y)
```

:::

- We want to estimate the proportion of *all registered voters* --- our population --- who will vote.
- To do this, we use the proportion of *polled voters* --- our sample --- who said they would.
- When the election occurs, we get to see who turns out to vote.
  - `r approx(p*N)` people, or roughly `r percent(mean(Y))` of registered voters, actually vote.
 
\ \

:::: columns
::: column
### Before the Election 

$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}  \\
Y_i  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$

:::
::: column
### After the Election

$$
\begin{array}{r|rrrrrr|r}
j & 1 & 2 & 3 & 4 & \dots &   `r approx(N)` & \bar{y}_{`r approx(N)`}  \\
y_{j} & `r y[1]` & `r y[2]` & `r y[3]` & `r y[4]` & \dots & `r y[N]`   & `r approx(mean(y))`  \\
\end{array}
$$
:::
::::


## The Sampling Distribution of our Estimate

```{r}
#| out-height: 400px
#| fig-asp: .3 

width = function(samples, alpha=.05) {
  covg.dif = function(w) { 
    actual = mean( mean(samples) - w/2 <= samples & samples <= mean(samples) + w/2)
    nominal = 1-alpha 
    actual - nominal  
  }
  uniroot(covg.dif, interval=c(0, 2*max(abs(samples-mean(samples)))))$root
}

set.seed(our.seed)
Js = array(dim=c(10000, n))
Ys = array(dim=c(10000, n))
for(rr in 1:10000) {
  Js[rr,] = sample(N, n, replace=TRUE)
  Ys[rr,] = y[Js[rr,]]
}
meanY.samples = rowMeans(Ys)
height = .04
scale_x = scale_x_continuous(breaks=seq(.6,.8, by=.025), limits=c(.6, .8)) 
scale_y = scale_y_continuous(breaks=seq(0,height,by=.01), limits=c(0, height))

annotate.histogram = function(samples) { 
    w66 = width(samples, alpha=1/3)
    w95 = width(samples, alpha=.05)
    ggplot() + geom_bar(aes(x=samples, y=after_stat(prop)), alpha=.3) +
           geom_vline(aes(xintercept=mean(y)), color='green', size=3, alpha=.7) +
           geom_vline(aes(xintercept=mean(samples)), color="blue") +
           geom_vline(aes(xintercept=mean(samples)+c(-1,1)*w66/2), color="blue", linetype='dashed') +
           geom_vline(aes(xintercept=mean(samples)+c(-1,1)*w95/2), color="blue", linetype='dotted') +
           scale_x + scale_y
}

default.highlight.colors = c('black', 'magenta', 'cyan', 'gray')
with.points = function(histplot, samples, K=100, highlight=4) {
    set.seed(our.seed)
    highlight.colors = if(highlight) default.highlight.colors[1:highlight] else c()
    heights = sample(height*(1:K)/K)
    data = data.frame(y=heights, x=samples[1:K], 
                      color=c(highlight.colors, rep('purple', K-length(highlight.colors))),
                      alpha=c(rep(1, length(highlight.colors)), rep(.1, K-length(highlight.colors))))
    histplot + geom_point(aes(y=y, x=x, color=I(color), alpha=I(alpha)), data=data)
}

with.intervals = function(histplot, samples, w=width(samples, alpha=.05), K=100, 
                          highlight=4, highlight.alpha=1) { 
    set.seed(our.seed)
    highlight.colors = if(highlight) default.highlight.colors[1:highlight] else c()
    heights = sample(height*(1:K)/K)
    data = data.frame(y=heights, x=samples[1:K], w=w,
                      color=c(highlight.colors, rep('purple', K-length(highlight.colors))),
                      alpha=c(rep(highlight.alpha, length(highlight.colors)), rep(.1, K-length(highlight.colors))))
    histplot + 
        geom_pointrange(aes(y=y, x=x, xmin=x-w/2, xmax=x+w/2, color=I(color), alpha=I(alpha)), data=data)
}

annotated.histogram = annotate.histogram(meanY.samples)
with.points(annotated.histogram, meanY.samples)
```

```{r}
J1 = Js[1,]
J2 = Js[2,]
J3 = Js[3,]
JM = Js[4,]
```

$$
\begin{array}{r|rr|rr|r|rr|r}
\text{call} &  1 & & 2 & & \dots & `r n` & &  \\
\text{pollster} & J_1 & Y_1 & J_2 & Y_2   & \dots & J_{`r n`} & Y_{`r n`}  & \overline{Y}_{`r n`}  \\
\hline
\color{black}   1   
    & \color{black}   `r J=J1; Y=y[J]; J[1]`    & \color{black}  `r Y[1]` 
    & \color{black}   `r J[2]`   & \color{black}   `r Y[2]`           & \color{black}   \dots 
    & \color{black}   `r J[n]`   & \color{black}   `r Y[n]`           & \color{black}   `r approx(mean(Y))` \\
\color{magenta} 2   
    & \color{magenta} `r J=J2; Y=y[J]; J[1]`    & \color{magenta}  `r Y[1]` 
    & \color{magenta} `r J[2]`   & \color{magenta}  `r Y[2]`          & \color{magenta} \dots 
    & \color{magenta} `r J[n]`   & \color{magenta} `r Y[n]`           & \color{magenta} `r approx(mean(Y))` \\
\color{cyan}    3   
    & \color{cyan}    `r J=J3; Y=y[J]; J[1]`    & \color{cyan}     `r Y[1]` 
    & \color{cyan}    `r J[2]`   & \color{cyan}    `r Y[2]`           & \color{cyan}    \dots 
    & \color{cyan}    `r J[n]`   & \color{cyan}    `r Y[n]`           & \color{cyan}    `r approx(mean(Y))` \\
{\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots} &  {\vdots} \\ 
\color{gray}    1M   
    & \color{gray}    `r J=JM; Y=y[J]; J[1]`    & \color{gray}     `r Y[1]` 
    & \color{gray}    `r J[2]`   & \color{gray}    `r Y[2]`           & \color{gray}    \dots 
    & \color{gray}    `r J[n]`   & \color{gray}    `r Y[n]`           & \color{gray}    `r approx(mean(Y))` \\
 {\vdots} & {\vdots} & {\vdots} & {\vdots} & {\vdots}  & {\vdots} & {\vdots} & {\vdots}  & {\vdots} \\ 
\end{array}
$$


## Interval Estimation

```{r}
#| out-height: 400px
#| fig-asp: .3 

w = .02
with.intervals(annotated.histogram, meanY.samples, w=w)
```

- To give a sense of what this tells us about the population proportion, we can report an interval estimate.
    - i.e., we can draw 'arms' on our point estimate $\bar Y_{`r n`}$. 
    - Here, we've used arms of width .02. i.e., we've drawn in the interval $\bar Y_{`r n`} \pm .01$.
- But this is only useful if we have a sense of how likely this interval is to be 'right.'
- That is, how often, when we report and interval, it contains [the population proportion]{.fg style="color:limegreen"}.
- We call this a *coverage probability*. In this case, it's bad. It's about **`r percent(mean(abs(meanY.samples - mean(y)) <= .01))`**.
- If you run polls and report intervals like this, you'll make misleading claims about turnout **`r percent(1-mean(abs(meanY.samples - mean(y)) <= .01))`** of the time. 
- That's not a good look. To fix this, we need to **calibrate** our interval estimates.

## A Useful Counting Trick

```{r}
#| out-height: 400px
#| fig-asp: .3 

adhoc.width.histogram = annotated.histogram + 
    geom_rect(aes(xmin=mean(y)-w/2, xmax=mean(y)+w/2, ymin=0, ymax=Inf), alpha=.15, fill='green')
with.intervals(adhoc.width.histogram, meanY.samples, w=.02)
```

- We don't really need to look at every interval---and its arms---to calculate coverage.
- Instead, we can draw arms of the same width around the population proportion.
- An interval estimate covers the [population proportion]{.fg style="color:limegreen"} if and only if ...
- the corresponding point estimate is between the population proportion's arms.

## A Useful Counting Trick

```{r}
#| out-height: 400px
#| fig-asp: .3 

with.points(adhoc.width.histogram, meanY.samples)
```

- So all we need to do is count how many of our point estimates are between the population proportion's arms.
- Or, equivalently, calculate the **mass of our histogram** between the population proportion's arms.
- That is, we calculate the area of the verticals bars in the [shaded region]{.fg style="color:limegreen"}.
- You can see it's roughly half the area of the histogram itself. Hence, roughly 50% coverage.

## Calibrating Interval Estimates

```{r}
#| out-height: 400px
#| fig-asp: .3 

w=width(meanY.samples, alpha=.05)
calibrated.width.histogram = annotated.histogram + 
    geom_rect(aes(xmin=mean(y)-w/2, xmax=mean(y)+w/2, ymin=0, ymax=Inf), alpha=.15, fill='green')
with.points(calibrated.width.histogram, meanY.samples)
```

- To increase the coverage of our intervals estimates, we need to make the arms wider.
- To **calibrate** this width to give us 95% coverage---which is the convention---we can use our counting trick.
- We expand our [shaded region]{.fg style="color:limegreen"} until it covers 95% of the histogram's mass, i.e., 95% of our point estimates.
- If we use intervals of this width around our point estimates, we'll get 95% coverage.

## Calibrating Interval Estimates
```{r}
#| out-height: 400px
#| fig-asp: .3 

with.intervals(calibrated.width.histogram, meanY.samples, w=w)
```

- To increase the coverage of our intervals estimates, we need to make the arms wider.
- To **calibrate** this width to give us 95% coverage---which is the convention---we can use our counting trick.
- We expand our [shaded region]{.fg style="color:limegreen"} until it covers 95% of the histogram's mass, i.e., 95% of our point estimates.
- If we use intervals of this width around our point estimates, we'll get 95% coverage.

## The Problem

```{r}
#| out-height: 400px
#| fig-asp: .3 

with.intervals(calibrated.width.histogram, meanY.samples, w=w)
```

- Unless we've seen the population, i.e. unless it's after the election, we can't run simulated polls.
- That means we can't calibrate our interval using the sampling distribution of our point estimate. 
- That's something we don't know. The good news is that it's something we can estimate.
- So we'll actually do calibration using **an estimate of the sampling distribution**.

# Real-Life Calibration

## Calibration using a Sampling Distribution Estimate

```{r}
#| out-height: 400px
#| fig-asp: .3 

set.seed(our.seed)
estimated.dist.samples = rbinom(10000, n, mean(Y)) / n 
w.est=width(estimated.dist.samples, alpha=.05)
annotated.histogram.estimated = annotate.histogram(estimated.dist.samples)
calibrated.width.histogram.estimated = annotated.histogram.estimated + 
    geom_rect(aes(xmin=mean(Y)-w.est/2, xmax=mean(Y)+w.est/2, ymin=0, ymax=Inf), alpha=.15, fill='red')
with.intervals(calibrated.width.histogram.estimated, estimated.dist.samples, highlight=FALSE)
```

- Here we've estimated the sampling distribution --- we'll get to how --- and used it to find an interval width.
- In particular, we've used a width which ...
    - if we draw arms around the mean of our estimated sampling distribution
    - will cover 95% of draws from our estimated sampling distribution.
- The one apparent sticking point is that our estimated sampling distribution is a bit off-center.
- Its mean is a bit lower than [the population proportion]{.fg style="color:limegreen"}. It's actually our sample proportion $\bar Y$.
- It turns out that this doesn't matter much. Why?

## Calibration Comparison

```{r}
#| out-height: 400px
#| fig-asp: .3 
calibration.comparison.histogram = annotated.histogram +
    geom_rect(aes(xmin=mean(y)-w/2, xmax=mean(y)+w/2, ymin=0, ymax=1.25*heights[1]), alpha=.15, fill='green') +
    geom_rect(aes(xmin=mean(y)-w.est/2, xmax=mean(y)+w.est/2, ymin=1.25*heights[1], ymax=Inf), alpha=.15, fill='red')

K=100
set.seed(our.seed)
heights = sample(height*(1:K)/K)

calibration.comparison.histogram +  
    geom_pointrange(aes(y=heights[1]*1.5, x=meanY.samples[1], 
                        xmin = meanY.samples[1]-w.est/2, xmax = meanY.samples[1]+w.est/2), alpha=1, color='black')  +
    geom_pointrange(aes(y=heights[1], x=meanY.samples[1], 
                        xmin = meanY.samples[1]-w/2, xmax = meanY.samples[1]+w/2), alpha=.4, color='black')   
```

- It doesn't matter because we're not putting arms on draws from the estimated sampling distribution.
- We're putting arms on our point estimate, which is a draw from the **actual sampling distribution**.
- To do this, we're using the width---but not the center---of the estimated sampling distribution.
- And it works because that's very close to the width we'd get from the actual sampling distribution.
- Above, I've drawn in shaded regions corresponding to two versions of the the [population proportion's]{.fg style="color:limegreen"} arms.
    - The [green region]{.fg style="color:limegreen"} is the one we get from the actual sampling distribution. We can't use this.
    - The [red region]{.fg style="color:red"} is the one we get from the estimated sampling distribution. We can --- and do --- use this.
- And I've drawn in interval estimates based on our poll using both widths. They're almost the same.


## Our Samping Distribution is Binomial

```{r}
#| out-height: 300px
#| fig-asp: .25 
p = mean(y)
x= seq(0,n)

with.points(calibrated.width.histogram, meanY.samples, highlight=FALSE) +
            geom_line(aes(x=x/n, y=dbinom(x, n, mean(y))), color='black')
```

- To estimate the sampling distribution, it helps to know something about the actual sampling distribution.
    - It's the distribution of the proportion heads in `r n` flips of a coin with probability $p\approx `r approx(p)`$ of heads.
    - where $p \approx `r approx(p)`$ is the proportion of registered voters who will vote.
- There's a name for that: it's the binomial distribution. 
    - To be more precise, the binomial distribution is for the number of heads.
    - And we scale that by $1/`r n`$ to get a distribution for the proportion.

$$
`r n` \times \bar{Y}_{`r n`} = \sum_{i=1}^{`r n`} Y_i \sim \text{Binomial}(`r n`, p) \qfor p = \bar y_{`r approx(m)`} = \frac{1}{`r approx(m)`}\sum_{j=1}^{`r approx(m)`} y_j 
$$

- There's a formula for this probability, so we don't have to draw samples and histogram if we don't want to.
- In `R`, it's `dbinom(x, n, p)`, where `x` is the number of heads, `n` is the number of flips, and `p` is the probability of heads.

## Estimating the Sampling Distribution

```{r}
#| out-height: 300px
#| layout-ncol: 2 
#| fig-asp: .4 
#| fig-cap:
#|   - "Actual Sampling Distribution"
#|   - "Estimated Sampling Distribution"

p = mean(y)

with.points(calibrated.width.histogram, meanY.samples, highlight=1) +
            geom_line(aes(x=x/n, y=dbinom(x, n, mean(y))), color='gray')

with.points(calibrated.width.histogram.estimated, estimated.dist.samples, highlight=0) +
            geom_line(aes(x=x/n, y=dbinom(x, n, mean(Y))), color='black') +
            geom_point(aes(x=mean(Y), y=heights[1]))
```

$$
`r n` \times \bar{Y}_{`r n`} = \sum_{i=1}^{`r n`} Y_i \sim \text{Binomial}(`r n`, p) \qfor p = \bar y_{`r approx(m)`} = \frac{1}{`r approx(m)`}\sum_{j=1}^{`r approx(m)`} y_j 
$$


- What we don't know is $p=\bar y$, the proportion of registered voters who turn out.
- But that's what we're using our sample proportion $\bar Y$ to estimate. 
- And we can just plug that in to get an estimate of the sampling distribution.
- To draw samples from this estimated sampling distribution, we use `rbinom`.

```{r}
#| echo: true
#| eval: false
samples = rbinom(10000, 625, mean(Y)) / 625
``` 

- That is exactly how I got the estimated sampling distribution histogram I was using for calibration. 
    - If we prefer, we can calibrate using the actual estimated binomial distribution, i.e., using `dbinom`.
    - We'll explore this a bit at some point in the future, but it doesn't make much difference.

## Each of Our Friends Gets Their Own Estimate

```{r}
#| out-height: 300px
#| layout-ncol: 2 
#| fig-asp: .4 
#| fig-cap:
#|   - "Actual Sampling Distribution"
#|   - "Estimated Sampling Distribution (Poll 1)"
#|   - "Estimated Sampling Distribution (Poll 2)"
#|   - "Estimated Sampling Distribution (Poll 3)"

plot.interval = function(Y, height, color) {
    set.seed(our.seed)
    samples = rbinom(10000, n, mean(Y)) / n 
    w=width(samples, alpha=.05)
    geom_pointrange(aes(x=mean(Y), y=height, xmin=mean(Y)-w/2, xmax=mean(Y)+w/2), color=color)
}

dh = .2*heights[1]
with.intervals(calibrated.width.histogram, meanY.samples, highlight=3, highlight.alpha=.4) +
               geom_line(aes(x=x/n, y=dbinom(x, n, mean(y))), color='gray') +
               plot.interval(y[J1], dh+heights[1], 'black') +
               plot.interval(y[J2], dh+heights[2], 'magenta') +
               plot.interval(y[J3], dh+heights[3], 'cyan') 

calibration.diagram = function(Y, color, height) {
    set.seed(our.seed)
    samples = rbinom(10000, n, mean(Y)) / n 
    w=width(samples, alpha=.05)
    histplot = annotate.histogram(samples)
    histplot + geom_rect(aes(xmin=mean(Y)-w/2, 
                             xmax=mean(Y)+w/2, ymin=0, ymax=Inf), 
                             alpha=.15, fill='red') + 
    geom_line(aes(x=x/n, y=dbinom(x, n, mean(Y))), color=color) +
    geom_point(aes(x=mean(Y), y=height), color=color)
}

calibration.diagram(Y,     'black',   heights[1])
calibration.diagram(y[J2], 'magenta', heights[2])
calibration.diagram(y[J3], 'cyan',    heights[3])
```

- This means each of our friends estimates a different width for their interval.
- But they're all close to the width we get using the actual sampling distribution. You can see it above.
    - I've plotted the intervals they'd use --- based on their sampling distribution estimates --- in bold colors.
    - And the intervals they wish they could --- based on the actual sampling distribution --- more faintly. 

## Evidence That It Works 

```{r}
#| out-height: 400px
K = 100
ps = meanY.samples[1:K]

ggplot() + geom_line(aes(x=x/n-p, y=n*dbinom(x,n,p)), data=expand.grid(p=mean(y), x=0:n)) +
           geom_vline(aes(xintercept=qbinom(c(.025,.975), n,mean(y))/n-mean(y)), linetype='dotted') +
           geom_line(aes(x=x/n-p, y=n*dbinom(x,n,p), color=factor(p)), data=expand.grid(p=ps, x=0:n), alpha=.2) +
           geom_pointrange(aes(x=0, xmin=qbinom(.025,n,p)/n-p, xmax=qbinom(.975,n,p)/n-p, y=20*k/K, color=factor(p)), data=data.frame(p=ps, k=1:K)) +
           xlim(-.1,.1) + guides(color=FALSE)
```

- We see the same thing when we look at 100 polls.
- Our friends' sampling distribution estimates are all close to the actual sampling distribution.
- The resulting interval-widths are near-indistinguishable from the width we got using the sampling distribution.
  - The estimated widths are shown as points with arms.
  - That sampling distribution's actual width is the space between the dotted lines.
- You can see estimating the width like this works reliably.

## Why Does it Work?

```{r}
#| out-height: 300px
#| fig-asp: .6 
#| layout-ncol: 2
K = 3
ps = quantile(meanY.samples, c(.025, .5, .975))

density.stability.plot = ggplot() + 
           geom_line(aes(x=x/n, y=dbinom(x,n,p)), data=expand.grid(p=mean(y), x=0:n), alpha=.5) +
           geom_vline(aes(xintercept=qbinom(c(.025,.975), n,mean(y))/n), linetype='dotted') +
           geom_line(aes(x=x/n, y=dbinom(x,n,p), color=factor(p)), data=expand.grid(p=ps, x=0:n), alpha=.5) +
           geom_pointrange(aes(x=p, xmin=qbinom(.025,n,p)/n, xmax=qbinom(.975,n,p)/n, y=.6*height*k/K, color=factor(p)), data=data.frame(p=ps, k=1:K)) +
           xlim(.6,.8) + guides(color=FALSE) + geom_vline(xintercept=mean(y), color='green', size=1, alpha=.5)
centered.density.stability.plot = ggplot() + 
           geom_line(aes(x=x/n-p, y=dbinom(x,n,p)), data=expand.grid(p=mean(y), x=0:n), alpha=.5) +
           geom_vline(aes(xintercept=qbinom(c(.025,.975), n,mean(y))/n-mean(y)), linetype='dotted') +
           geom_line(aes(x=x/n-p, y=dbinom(x,n,p), color=factor(p)), data=expand.grid(p=ps, x=0:n), alpha=.5) +
           geom_pointrange(aes(x=0, xmin=qbinom(.025,n,p)/n-p, xmax=qbinom(.975,n,p)/n-p, y=.6*height*k/K, color=factor(p)), data=data.frame(p=ps, k=1:K)) +
           xlim(-.1,.1) + guides(color=FALSE)

density.stability.plot
centered.density.stability.plot
```

- Why are we getting a good estimate of the sampling distribution?
- The Binomial distribution is *continuous* as a function of $p$---when $p$ changes little, the distribution changes little.
- This means that, if we're not far off in the proportion, the estimated and actual sampling distributions are similar.
- The relevant difference (after centering) is even smaller because *the way* the binomial changes is mostly location.
- Here I'm showing three estimates of the sampling distribution based on three polls.
    - with centering (right) and without (left). 
- In particular, polls giving estimates at the 2.5th, 50th, and 97.5th percentiles of the actual sampling distribution. 

## The Estimate Works. Why Does it Work?
```{r}
#| out-height: 300px
#| fig-asp: .6 
#| layout-ncol: 2

density.stability.plot
centered.density.stability.plot
```

- You can think of this as a sort of 'confidence interval' for our estimate of the sampling distribution.
    - 95% of the time, you'll get an estimate somewhere between the [red]{.fg style="color:red"} and [blue]{.fg style="color:blue"} ones.
    - And, as a result, a width for your interval estimate somewhere between the [red]{.fg style="color:red"} and [blue]{.fg style="color:blue"} widths.
- As you can see, the widths we get barely change. What changes is the location.
- But because we're using calibrated 95% intervals, we still cover the population proportion 95% of the time.
    - i.e., we cover it when our sample proportion is between the [red]{.fg style="color:red"} and [blue]{.fg style="color:blue"} ones.
- Or, accounting for [slight underestimates]{.fg style="color:blue"} of the sampling distribution's width, almost cover it. 

## The Bootstrap Interpretation
```{r mapping}
#| cache: true

# Create fake locations for the people in our population
# by taking random convex combinations of state vertices
set.seed(0)
ga.shape = us_map(include = c("GA"))

# To reduce computation, we'll have 10K locations + jitter
M = 10000 
V = nrow(ga.shape)
weights = abs(rt(M*V,1))
dim(weights) = c(M,V)
weights = (1/rowSums(weights)) * weights 
base.x = weights %*% ga.shape$x
base.y = weights %*% ga.shape$y
base.ind = 1 + ( (0:(N-1)) %% M )
locations = data.frame(x=base.x[base.ind] + rnorm(N, sd=.01),
                       y=base.y[base.ind] + rnorm(N, sd=.01))

ga = ggplot() + geom_polygon(data=ga.shape, aes(x = x, y = y), fill="white", color="black") + coord_fixed() + blank.theme
```

```{r}
#| out-height: 300px
#| fig-asp: 1 
#| layout-ncol: 3
sample.locations = locations[J, ]
boot.ind = sample(1:n, n, replace=TRUE)
resampled.locations = sample.locations[boot.ind, ]
ga + geom_point(aes(x = x, y = y), alpha=.05, data=locations[1:M,], size=.05)
ga + geom_point(aes(x = x, y = y), alpha=.2, data=sample.locations) 
ga + geom_point(aes(x = x, y = y), alpha=.2, data=resampled.locations)
```

- There's another way to think about this.
- The estimated sampling distribution is $\text{Binomial}(n,p)/n$ for $n=`r n`$ and $p=\bar Y\approx `r mean(Y)`$.
- That is, it's the distribution of the proportion of heads in ...
    - `r n` flips of a coin 
    - that has probability $\bar Y\approx `r mean(Y)`$ of being heads.
- What's the distribution of the proportion of yeses if we poll a population of `r n` people ...
- ... among whom the proportion of yeses is exactly $\bar Y\approx `r mean(Y)`$?


## The Bootstrap Interpretation

```{r}
#| out-height: 300px
#| fig-asp: 1 
#| layout-ncol: 3
ga + geom_point(aes(x = x, y = y), alpha=.05, data=locations[1:M,], size=.05)
ga + geom_point(aes(x = x, y = y), alpha=.2, data=sample.locations) 
ga + geom_point(aes(x = x, y = y), alpha=.2, data=resampled.locations)
```

- It's the same distribution. It's $\text{Binomial}(n,p)/n$.
- That means we can get a draw from our estimated sampling distribution by
- running a 'poll' of the people in our sample, i.e.
    - rolling a 625-sided die 625 times
    - 'calling' the corresponding person in our sample
    - and counting up the yeses we hear
- We call the sample we get from this new 'poll' a **bootstrap sample**.

## The Bootstrap Interpretation

```{r}
#| out-height: 300px
#| fig-asp: 1 
#| layout-ncol: 3
ga + geom_point(aes(x = x, y = y), alpha=.05, data=locations[1:M,], size=.05)
ga + geom_point(aes(x = x, y = y), alpha=.2, data=sample.locations, size=1) 
ga + geom_point(aes(x = x, y = y), alpha=.2, data=resampled.locations, size=1)
```

- If we plot our voters on a map, you can see the idea in visual terms.
    - On the left, we have the population. 
    - In the middle, we have our sample.
    - On the right, we have locations drawn, with replacement, from the sample.
- Each 'call' is one transparent dot, so we see darker dots ...
    - or actually multiple copies of the same dot overlaid
    - where we've drawn someone multiple times.
- 2x darker means we've called them twice, 3x darker means we've called them three times, etc.

## Bootstrapping Our Turnout Estimate

```{r}
set.seed(our.seed)
```

```{r}
#| echo: true
#| out-height: 400px 
#| fig-asp: .3 

meanY.boot = array(dim=10000)
for(rr in 1:10000) {
    I = sample(1:n, n, replace=TRUE)
    meanY.boot[rr] = sum(Y[I])/n
}

ggplot() + geom_bar(aes(x=meanY.boot, y=after_stat(prop)), alpha=.3) + 
           geom_line(aes(x=(0:n)/n, y=dbinom(0:n, n, mean(Y))), color='black') + 
           xlim(.6,.8)
```


