---
title: "Lecture 6"
subtitle: "Working With Non-Binary Outcomes: Location"
---

## So Far

{{< include shared-code.qmd >}}
{{< include ca-income.qmd >}}

```{r}
#| cache: true
social.pressure = read.csv('data/social-pressure-data.csv') 
keep = social.pressure$treatment %in% c(' Control', ' Neighbors')
D = as.numeric(social.pressure[keep,'treatment'] == ' Neighbors')
Y = as.numeric(social.pressure[keep,'voted'] == 'Yes')
n = length(Y)

difference = mean(Y[D==1]) - mean(Y[D==0])
difference.bootstrap.samples = replicate(1000, {
    I = sample(1:n, n, replace=TRUE)
    Dstar = D[I]
    Ystar = Y[I]
    mean(Ystar[Dstar==1]) - mean(Ystar[Dstar==0])
})
```

```{r}
#| out-height: 250px
#| fig-asp: .25 
 
xx = seq(.05, .1, length=1000)
scale_x_dif = scale_x_continuous(limits=c(.07,.095), labels=label_percent())
wboot=width(difference.bootstrap.samples, alpha=.05)

difference.bootstrap.samples = difference.bootstrap.samples - mean(difference.bootstrap.samples) + difference
pboot=ggplot() + geom_histogram(aes(x=difference.bootstrap.samples, y=after_stat(density)), alpha=.2, bins=100) 
pboot + with_annotations(difference.bootstrap.samples, highlight.dots=0) + 
        geom_pointrange(aes(x=x,y=y, xmin=x-w/2, xmax=x+wboot/2), 
                  data=data.frame(x=difference, y=50, w=wboot), color='black')  +
        geom_line(aes(x=xx, y=dnorm(xx, mean(difference.bootstrap.samples), sd(difference.bootstrap.samples))), color='cyan') +
        scale_x_dif
``` 


- We've worked with binary outcomes. In particular, the outcome of voting.
- We've estimated turnout: the proportion of people in our **population** of registered voters who do actually vote.
- We've talked about how to estimate this using a **random sample** of these voters, e.g., from a telephone survey.
- To summarize how close our estimate is to the actual turnout rate, we've used confidence intervals.
  - We calibrate them so that, 95% of the time, they contain what we're trying to estimate.
  - For calibration, we've used the **bootstrap sampling distribution** as an estimate of the sampling distribution.
- And we've talked about using **normal approximation** to understand what it'll all look like.
  - e.g. to estimate the precision we'll get from a given sample size.
- We've done all this for *turnout differences* between groups of voters, too. 
  - e.g. Black and non-Black voters.
  - e.g. Voters who received a letter pressuring them to vote and voters who didn't.

::: aside
Shown above: an estimate of the difference in turnout between people who received a letter pressuring them to vote and people who didn't.
:::

## Today


:::: columns

::: column
```{r map-sample}
#| fig-width: 5
#| out-height: 300px
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=earnings, size=.1, alpha=.5) + guides(color='none')
```
:::

::: column

```{r table}
style.tab = function(tab, cols=c('income', 'education', 'county'), nhead=3, ntail=1) {
  tab$income50k = 0+(tab$income > 50000)
  tab$income = sprintf('$%dk', round(as.numeric(tab$income) / 1000))
  tab$county = ifelse(tab$county=='los angeles', 'LA',  tab$county)
  ellipsis.tab = tab[, cols] %>% ellipsis(nhead=nhead,ntail=ntail)
  ellipsis.tab %>%
    kable %>% kable_styling() %>% 
    row_spec(which(ellipsis.tab$education >= 16),  background=twocolor.green) %>%
    row_spec(which(ellipsis.tab$education <  16),  background=twocolor.red) 
}

style.tab(sample.tab, nhead=4)
```
:::

::::

- We'll do the same thing, but for **non-binary outcomes**.
- Using data from the Current Population Survey, we'll estimate ...
  - the mean income in our population.
  - the mean income among people in our population [with 4-year degrees]{.twocolor-green} 
  - the difference in mean income between people [with 4-year degrees]{.twocolor-green} and people [without them]{.twocolor-red}.
- Our population will be the set of California residents 
    - between the ages of 25 and 35
    - with at least an 8th-grade education


::: aside
The locations shown on the map are made up. They're not the actual locations of the people in the sample. \
The survey includes some location information for some people, but as you can see in the table, not for everyone.
:::


## Imagining a Population

```{r}
#| out-height: 300px
#| fig-width: 5 
#| layout-ncol: 2 
#| 
ca.plot + 
 geom_point(aes(x=long, y=lat, color=degree), data=pop.earnings, size=.02, alpha=.1) + 
 guides(color='none')
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=earnings, size=.1, alpha=.5) +
  guides(color='none')
```

:::: columns
::: column
```{r}
style.tab(pop.tab)
```
:::
::: column
```{r}
style.tab(sample.tab)
```
:::
::::

- We don't have much information about the people in our population. Except the ones in our sample. 
- But, for the sake of visualization, I've made some up. 
  - We'll put visualizations of this fake population on the left.
  - Visualizations of the (real) sample will be on the right.



## Sampling

```{r pop-sample-maps}
#| out-height: 500px
#| fig-asp: .8 
#| layout-ncol: 2 
#| 
ca.plot + 
 geom_point(aes(x=long, y=lat, color=degree), data=display.population, size=.02, alpha=.1) + 
 guides(color='none')
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=our.sample, size=.1, alpha=.5) 
```

- We'll act as if our data were sampled, with replacement, from this population.
  - On the left, I've illustrated the population.
  - On the right, I've illustrated our sample.
- Our working assumption: each dot on the right was chosen, from those on the left, by rolling a big die.
- That's not quite right, but we'll stick with it throughout the semester.

## Sample and Population Means


```{r}
#| out-height: 300px
#| layout-ncol: 2
#| 

ggplot(pop.earnings) + 
  geom_point(aes(x=0, y=income),  color='blue',
            size=.2, position=position_jitter(seed=0,w=.3), alpha=.05) +
  stat_summary(aes(x=0, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x))  + 
  scale_y() 

ggplot(earnings) + 
  geom_point(aes(x=0, y=income), color='blue', 
            size=2, position=position_jitter(seed=0,w=.3), alpha=.2) +
  stat_summary(aes(x=0, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x))  +
  scale_y() 
```

- One thing we often want to estimate is the mean of our population.

$$
\mu = \frac{1}{m}\sum_{j=1}^m y_j 
$$

- We'll think of the mean in our sample as an estimate of it.

$$ 
\hat \mu = \frac{1}{n}\sum_{i=1}^n Y_i 
$$


## Subsample and Subpopulation Means

```{r}
#| out-height: 300px
#| layout-ncol: 2 

earnings.jitter.pop = ggplot(pop.earnings) + 
    geom_point(aes(x=degree, y=income, color=degree), 
               size=.2, position=position_jitter(seed=0,w=.3), alpha=.1) +
     scale_y() + guides(color='none')

earnings.jitter = ggplot(earnings) + 
    geom_point(aes(x=degree, y=income, color=degree), 
               size=dot.size, position=position_jitter(seed=0,w=.3), alpha=.1) +
     scale_y() + guides(color='none')

earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

- We'll often be interested in the mean income in subpopulations, too.
  - We'll think about the subsample with $X_i=x$ for some value $x$. 
  - e.g. $X_i=1$ for people [with a 4-year degree]{.twocolor-green} and $X_i=0$ for people [without]{.twocolor-red}.
$$
\mu(x) = \frac{1}{m_x}\sum_{j:x_j = x } y_j \quad \text{ where } \quad m_x = \sum_{j:x_j=x} 1.
$$

- We'll use the mean in the corresponding subsample to estimate it.

$$ 
\hat \mu(x) = \frac{1}{N_x}\sum_{i:X_i=x} Y_i \quad \text{ where } \quad N_x = \sum_{i:X_i=x} 1.
$$


## Are These Good Estimators?

```{r}
xx = seq(-6,6,by=.01)
K = 100
heights = seq(0,.4,length.out=K)
normals = rnorm(10000)

unbiased.plot = ggplot() + geom_histogram(aes(x=normals, y=after_stat(density)), alpha=.3, bins=100) +
           geom_vline(xintercept=0, color='green', linewidth=3, alpha=.4) + 
           geom_vline(xintercept=0, color='blue') + 
           geom_vline(xintercept=c(-1,1), color='blue', linetype='dashed') +
           geom_vline(xintercept=c(-2,2), color='blue', linetype='dotted') +
           geom_pointrange(aes(y=heights, x=normals[1:K], xmin=normals[1:K]-1.96, xmax=normals[1:K]+1.96), 
                           color='purple', alpha=.15) + xlim(min(xx),max(xx))
           
biased.plot = ggplot() + geom_histogram(aes(x=normals+2, y=after_stat(density)), alpha=.3, bins=100) +
           geom_vline(xintercept=0, color='green', linewidth=3, alpha=.4) + 
           geom_vline(xintercept=2+0, color='blue') + 
           geom_vline(xintercept=2+c(-1,1), color='blue', linetype='dashed') +
           geom_vline(xintercept=2+c(-2,2), color='blue', linetype='dotted') +
           geom_pointrange(aes(y=heights, x=2+normals[1:K], xmin=2+normals[1:K]-1.96, xmax=2+normals[1:K]+1.96), 
                           color='purple', alpha=.15) + xlim(min(xx),max(xx))


```

```{r}
#| layout-ncol: 2
unbiased.plot
biased.plot
```

- Today, we'll focus on the **location** of our estimator. 
  - That is, on where its sampling distribution is centered.
  - We'll think about the **precision** of our estimator next time.
- We want to know if it's centered at our [estimation target]{.fg style="text:green"}.
  - If things look like they do on the left, we're in good shape.
  - But if they look like they do on the right, we're not.
  - You can see that a good number of our samples on the right are far off-target. 
  - Further away, e.g., than the width of the distribution.
- To think about this, we'll use **random variables** to describe
  the relationship between the sample and population.

# Random Variables

## Observations as Random Variables

```{r}
#| out-height: 300px
#| layout-ncol: 2 

summaries = earnings %>% group_by(degree) %>% summarize(mean=mean(income), sd=sd(income))

earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

- The notation $y_j$ refers to a number.
  - It's the income of the $j$th person in our population. That's just some person. 
- The notation $Y_i$ refers to something else.
  - It's the income of the $i$th person we call in our survey. That's not a person.
  - That's the result of a random process---the roll of a die.
- To summarize this result, we talk about the **probability distribution** of $Y_i$.

### Notation Conventions

- Random variables are written in **uppercase**: $X$, $Y$, $Z$, etc. Constants are written in **lowercase**: $x$, $y$, $z$, etc.
- We'll use the same letter for a random variable and the value it takes on. $x$ is a possible value of $X$, $y$ of $Y$, etc.
- Estimators are also random variables, but instead of uppercase we use a **hat**: $\hat \mu$, $\hat \sigma$, etc.
- We'll use the same letter for the estimator and what it's meant to estimate. $\hat \mu$ is an estimate of $\mu$.



## Probability Distributions: The Binary Case

```{r}
cols = c('income50k', 'income', 'education', 'county')
sample.tab$roll = sample.tab$J
```

:::: columns
::: column
```{r}
style.tab(pop.tab, cols=cols)
```
:::
::: column
```{r}
style.tab(sample.tab, cols=c('roll', cols))
```
:::
::::

- When our outcomes are binary, it's easy to describe this distribution.
  - All we need to know is the probability that $Y_i$ is 1. Why?
  - To do that, we **sum the probabilities** of the rolls that result in it being one.

$$
P(Y_i = 1) = \sum_{j: y_j = 1} P(\text{roll}_i = j) = \sum_{j: y_j = 1} \frac{1}{m} \times y_j = \mu.
$$

- This **collapses** out information about our random process that's irrelevant to $Y_i$.
- That is, it collapses the probability we roll each number in 1...m into our 'weighted coin flip'.


## Probability Distributions: The General Case

:::: columns
::: column
```{r}
style.tab(pop.tab, cols=cols)
```
:::
::: column
```{r}
style.tab(sample.tab, cols=c('roll', cols))
```
:::
::::


- When our outcomes are are nonbinary, it's a bit more complicated.
  - We need to know the probability that $Y_i$ takes on each possible value.But we calculate it the same way. 
  - We sum the probabilities of the rolls that result in it taking on those values.

$$
P(Y_i = y) = \sum_{j: y_j = y} P(\text{roll}_i = j) 
$$

- We're still collapsing out irrelevant information, but what we're left with is more complicated.
- It's a **weighted die roll**, with one face for each possible value of $Y_i$. 
- And If each person's income is different, then there's nothing to collapse out. Then, ...
$$ 
P(Y_i = y) = \begin{cases} \frac{1}{m} & \qqtext{for} y \in y_1 \ldots y_m \\ 
                            0 & \qqtext{otherwise}
             \end{cases}
$$

## Expectations

:::: columns
::: column
```{r}
style.tab(pop.tab, cols=cols)
```
:::
::: column
```{r}
style.tab(sample.tab, cols=c('roll', cols))
```
:::
::::


- That said, we often don't need to know the whole distribution.
- Often the only thing we care about is the expected value of $Y_i$. Or some related quantity.
  - The expected value of $Y_i$ is the **probability-weighted average** of the values it can take on.
  - What's nice about this is that we can think of this in 'uncollapsed' terms.
- When we sample as usual, this is just the population mean.
- The collapsed form is, in a sense, just summing in a different order.

$$
\begin{aligned}
\E[Y_i] = \sum_y P(Y_i = y) \times y
= \sum_y \qty(\sum_{j: y_j = y} \frac{1}{m}) \times y 
= \frac{1}{m}\sum_y\sum_{j:y_j = y} y 
= \frac{1}{m}\sum_{j=1}^m y_j
\end{aligned}
$$

## Expectations

:::: columns
::: column
```{r}
style.tab(pop.tab, cols=cols)
```
:::
::: column
```{r}
style.tab(sample.tab, cols=c('roll', cols))
```
:::
::::


- What's nice about the binary case is, in fact, that it's an expectation.
- When $Y_i$ is binary, the probability it's 1 is its expected value $Y_i$.

$$
\begin{aligned}
P(Y_i=1) = \sum_{j: y_j = 1} P(\text{roll}_i = j) 
         &= \sum_{j:y_j=1} \frac{1}{m} \\
         &= \sum_{j=1}^m \frac{1}{m} \times \begin{cases} 1 & \ \ \text{ if } y_j = 1 \\ 0 & \ \ \text{ otherwise} \\ \end{cases} \\
         &= \sum_{j=1}^m \frac{1}{m} \times y_j = \E[Y_i].
\end{aligned}
$$

- In fact, we like expectations so much that we often use them to work with probabilities.

$$
P(Z \in A) = \E[1_A(Z)] \qfor 1_A(z) = 
\begin{cases} 
1 & \qqtext{ for } z \in A \\ 0 & \qqtext{ otherwise} 
\end{cases}
$$


## Independence 


```{r}
#| out-height: 300px
#| layout-ncol: 2 

summaries = earnings %>% group_by(degree) %>% summarize(mean=mean(income), sd=sd(income))

earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

- We say random variables are **independent** if knowing the value of one doesn't tell us anything about the value of the other.
- When this is the case, their joint probability distribution is just the product of their individual ones.

$$
P(Y_1=y_1 \qand Y_2=y_2) = P(Y_1=y_1)P(Y_2=y_2).
$$

- That's what happens when the randomness in $Y_1$ and $Y_2$ comes from two different rolls of the die.
- And since that's how we're doing our sampling --- that's sampling with replacement---it'll be true in our sample.
- Because we get each of our observations the same way, they also have the same probability distribution. 
- We say they're **independent and indentically distributed.**

## Conditioning

```{r}
#| out-height: 300px
#| layout-ncol: 2 

earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

- **Conditioning** is, in effect, a way of thinking about sampling as a two-stage process.
  - First, we choose the color of our dot, i.e., the value of $X_i$, according to it frequency in the population.
  - Then, we choose a specific one of those dots, i.e. $J_i$, from those with that color---with equal probability.
- Because this is just a way of thinking, each person still gets chosen with probability $1/m$.
$$
P(J_i=j) = 
\begin{cases}
   \frac{m_{green}}{m} \ \ \ \ \ \times \frac{1}{m_{green}} \ \ \ \ \ =\ \frac{1}{m} & \text{if the $j$th dot is green  ($x_j=1$) } \\
    \frac{m-m_{green}}{m} \times \frac{1}{m-m_{green}} \ = \ \frac{1}{m} & \text{otherwise}
\end{cases}
$$
  


## Conditioning

```{r}
#| out-height: 300px
#| layout-ncol: 2 

earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```


- The **Conditional Probability** of $Y_i$ is the probability resulting from the second stage.
- It's a function of the result of the first. 
  - $P(Y_i=y \mid X_i=1)$ is the probability distribution of $Y_i$ when we're rolling the 'green die'.
  - $P(Y_i=y \mid X_i=0)$ is the probability distribution of $Y_i$ when we're rolling the 'red die'.
- And the **Conditional Expectation** of $Y_i$ is the 'second stage expected value' in the same sense.
  - $\E[Y_i \mid X_i=1]$ is the expected value of $Y_i$ when we're rolling the 'green die'. 
  - That is, the mean value $\mu(1)$ of $y_j$ in the subpopulation drawn as little green dots.
  - $\E[Y_i \mid X_i=0]$ is the expected value of $Y_i$ when we're rolling the 'red die'.
  - That is, the mean value $\mu(0)$ of $y_j$ in the subpopulation drawn as little red dots.


# Working with Expectations 

## Conditioning

```{r}
#| out-height: 300px
#| layout-ncol: 2 

earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

### The law of iterated expectations
$$
E \{ E( Y \mid X ) \} \quad \text{ for any random variables $X, Y$} 
$$

To calculate the mean of $Y$, we can average within subpopulations, then across subpopulations.

### Irrelevance of independent conditioning variables
$$ E( Y \mid X, X' ) = E( Y \mid X ) \quad \text{ when $X'$ is independent of $X,Y$ } $$

If $X'$ is unrelated to $X$ and $Y$, holding it constant doesn't affect the relationship between them.

## Linearity of Expectations

```{r}
#| out-height: 300px
#| layout-ncol: 2 

earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

$$ 
\begin{aligned}
E ( a Y + b Z ) 
&= E (aY) + E (bZ) \\
&= aE(Y) + bE(Z)  \\
& \text{ for random variables $Y, Z$ and numbers $a,b$ }
\end{aligned}
$$

  - There are two things going on here.
    - To average a sum of two things, we can take two averages and sum.
    - To average a constant times a random variable, we can multiply the random variable's average by the constant.
  - In other words, we can *distribute* expectations and can pull constants out of them.
  
## Proof

- In essence, it comes down to the fact that all we're doing is summing.
	- Expectations are probability-weighted sums.
	- And we're looking at the expectation of a sum.
- And we can change the order we sum in without changing what we get.

$$
\small{
\begin{aligned}
\E\qty( a Y + b Z ) 
&= \sum_{y}\sum_z (a y + b z) \ P(Y=y, Z=z) && \text{ by definition of expectation} \\
&= \sum_{y}\sum_z a y \ P(Y=y, Z=z) + \sum_{z}\sum_y b z \ P(Y=y, Z=z) && \text{changing the order in which we sum} \\
&= \sum_{y} a y \ \sum_z P(Y=y,Z=z) + \sum_{z} b z \ \sum_y P(Y=y,Z=z) && \text{pulling constants out of the inner sums} \\
&= \sum_{y} a y \ P(Y=y) + \sum_{z} b z \ \sum_y P(Z=z)  && \text{summing to get marginal probabilities from our joint } \\
&= a\sum_{y} y \ P(Y=y) + b\sum_{z} z \ \sum_y P(Z=z)  && \text{ pulling constants out of the remaining sum } \\
&= a\E Y + b \E Z  && \text{by definition}
\end{aligned}
}
$$

## Linearity of **Conditional** Expectations

```{r}
#| out-height: 300px
#| layout-ncol: 2 

earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

$$ 
\begin{aligned}
E\{ a(X) Y + b(X) Z \mid X \} 
&= E\{a(X)Y \mid X\} + E\{ b(X)Z \mid X\} \\
&= a(X)E(Y \mid X) + b(X)E(Z \mid X)
\end{aligned}
$$

- This is like linearity of expectations, but with a twist.
   - When we condition on $X$, we're working with subpopulations in which $X$ is constant.
   - This means we can act as if functions of $X$ **are** constants.
   - So we can pull them out of expectations that are conditional on $X$.
   
## It's important to distinguish between two things

```{r}
#| out-height: 250px
#| fig-asp: .25
Y = c(.75,1,1.25, 1,1.25,1.5)
X = 0 + (1:6 > 3)
Xdodge = X + rep(.02*c(-1,0,1),2)
mu = function(x) { mean(Y[X==x]) }

tinypop.plot = ggplot() + geom_point(aes(x=Xdodge, y=Y, color=factor(X)), alpha=.7, size=2) + 
  scale_x_continuous(breaks=c(0,1), labels=c('0','1')) + scale_y_continuous(breaks=seq(.25,1.75,by=.25)) +
  guides(color='none') 
tinypop.plot
```

  1.  The **conditional expectation function** $\mu(x)=E[Y \mid X=x]$.
      - $\mu$ is a function; evaluated at $x$, it's a number. It's not random.
      - It's the mean of the subpopulation of people with $X=x$.
   
  2.  The conditional expectation $\mu(X)=E[Y \mid X]$.
      - $\mu(X)$ is the mean of a **random subpopulation** of people.
      - It's the conditional expectation function **evaluated at** the random variable $X$.
      - This is the sort of thing that shows up when we use the law of iterated expectations.

## Check Your Understanding {.smaller}

```{r}
#| out-height: 250px
#| fig-asp: .25
tinypop.plot
```

- Suppose we sample a point $(X, Y)$  uniformly at random from the population of 6 points above.
  - What is the conditional expectation function $\mu(x)$ at [$x=0$]{.twocolor-red} and [$x=1$]{.twocolor-green}?
  - What is the conditional expectation $\mu(X)$?
  
::: fragment
  - [$\mu(0)=`r mu(0)`$]{.twocolor-red} and [$\mu(1)=`r mu(1)`$]{.twocolor-green}
:::
::: fragment
  - $\mu(X)$ is a random variable taking on these two values.
  $$ \mu(X) = \begin{cases}
      `r mu(0)` & \text{ when } X=0 \\
      `r mu(1)` & \text{ when } X=1
      \end{cases}
  $$
:::

## The Indicator Trick
```{r}
#| out-height: 250px
#| fig-asp: .25
tinypop.plot
```

- Suppose we sample a point $(X, Y)$  uniformly at random from the population of 6 points above.
  - What is $X\mu(X)$?

::: fragment

- $X\mu(X)$ is a random variable taking on these two values.
  $$ X\mu(X) = \begin{cases}
      0 \times \mu(0) = 0 \times `r mu(0)` & \text{ when } X=0 \\
      1 \times \mu(1) = 1 \times `r mu(1)` & \text{ when } X=1
      \end{cases}
  $$

- We can write it equivalently as $X\mu(1)$ because either ...
  - $X=0$, so $X\mu(X) = X\mu(1)=0$ 
  - $X=1$, so $X\mu(X)=X\mu(1)=1\times \mu(1)$
- This comes up a lot working with subsample means.
- We'll swap $X\mu(X)$ for $X\mu(1)$ often, referring to the **indicator trick**.

:::


## A More Realistic Example

```{r}
#| out-height: 250px
#| fig-asp: .25
#| 
pop.summaries = pop.earnings %>% group_by(degree) %>% summarize(mean=mean(income), sd=sd(income))
mean.has.degree = round(pop.summaries$mean[pop.summaries$degree=='4-year degree'] / 10000) * 10
mean.no.degree  = round(pop.summaries$mean[pop.summaries$degree=='no 4-year degree'] / 10000) * 10
degree.proportion = round(mean(pop.earnings$degree=='4-year degree'),1)
earnings.jitter.pop + geom_pointrange(aes(x=degree, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries) 
```

 - Let's think about a random person drawn from this population
    - $Y$ is their income 
    - $X$ is an indicator for having a four-year degree. 
  - Suppose the subpopulation means are `r approx(mean.has.degree, 10)`k for people with degrees and `r approx(mean.no.degree, 10)`k for people without.
  - And that `r approx(degree.proportion*10,1)`/10 of our population has degrees.
  
- **Q.**  What is $E(Y \mid X)$? And what is $E(Y)$?

:::: columns
::: column
::: fragment

- $E(Y \mid X)$ is a random variable that is **either** `r mean.has.degree`k or `r mean.no.degree`k
  - It's `r mean.has.degree`k with probability `r degree.proportion*10`/10,       when $X=1$. 
  - It's `r mean.no.degree`k  with probability `r (1-degree.proportion)*10`/10,   when $X=0$. 

::: 
:::
::: column
::: fragment

- $\E(Y)$ is its expectation. A number.
- We'll calculate it using iterated expectations.

$$
\begin{aligned} 
E\{ E( Y \mid X ) \} 
&= E(Y|X=1)P(X=1) + E(Y|X=0)P(X=0) \\
&= `r mean.has.degree`k \cdot `r degree.proportion*10`/10 +
   `r mean.no.degree`k  \cdot `r (1-degree.proportion)*10`/10
=  `r mean.has.degree*degree.proportion`k +
   `r mean.no.degree*(1-degree.proportion)`k 
=  `r mean.has.degree*degree.proportion + mean.no.degree*(1-degree.proportion)`k
\end{aligned}
$$

:::
:::
::::

## Review Exercise 1
```{r}
#| out-height: 250px
#| fig-asp: .25
tinypop.plot
```

- Suppose we sample a point $(X, Y)$ in the plot above uniformly at random.
  - Ignore the jitter; just think of the points as being at $X=0$ and $X=1$.
  - What is $\E\{ \mu(X) \}$?

::: fragment
  - There are two ways of thinking about calculating $\E\{\mu(X)\}$.
      1. We just calculate the expectation of it, thinking of it as an arbitrary random variable.
      2. We use the law of iterated expectations to show it's the unconditional mean of $Y$.

$$
\small{
\begin{aligned}
\E\qty{ \mu(X) } &= \frac{1}{2}\mu(0) + \frac{1}{2}\mu(1)= \frac{1}{2} \cdot `r mu(0)` + \frac{1}{2} \cdot `r mu(1)` = `r (mu(0)+mu(1))/2` && \text{ the first way } \\
\E\qty{ \mu(X) } &= \E\qty{\E\qty(Y \mid X)} = \E\qty{Y} = \frac{1}{6} \cdot `r Y[1]` + \frac{1}{6} \cdot `r Y[2]` + \ldots && \text{ the second way }
\end{aligned}
}
$$

:::

## Review Exercise 2
```{r}
#| out-height: 250px
#| fig-asp: .25
tinypop.plot
```

- Suppose we sample a point $(X, Y)$ in the plot above uniformly at random.
  - Ignore the jitter; just think of the points as being at $X=0$ and $X=1$.
  - What is $\E\{ X\mu(X) \}/\E(X)$?

::: fragment

- It's $\mu(1)$. $X\mu(X)=X\mu(1)$, so ...
$$
\begin{aligned}
\E\{ X\mu(X) \} 
&= \E\{ X\mu(1) \} \\
&= \mu(1) \ \E\{ X \} \\
\end{aligned}
$$

:::

## Review Exercise 3

```{r}
#| out-height: 250px
#| fig-asp: .25
tinypop.plot
```

- Suppose we sample a point $(X, Y)$ in the plot above uniformly at random.
  - Ignore the jitter; just think of the points as being at $X=0$ and $X=1$.
  - What is $\E\{ (1-X)\mu(X) \}/\E(1-X)$?

::: fragment

- It's $\mu(0)$. It's analogous to the last one. $(1-X)\mu(X)=(1-X)\mu(0)$, so ...
$$
\begin{aligned}
\E\{ (1-X)\mu(X) \} 
&= \E\{ (1-X)\mu(0) \} \\
&= \mu(0) \ \E\{ 1-X \} \\
\end{aligned}
$$

:::


# Unbiasedness of Means

## Bias

```{r}
#| layout-ncol: 2
unbiased.plot
biased.plot
```

- The **bias** of an estimator is the difference between its expected value and the value of the thing it's estimating.
- We haven't talked about this yet because there hasn't been any. We've been using **unbiased estimators**.
- You can see, looking at the sampling distribution on the right, that bias might cause some problems with coverage.
- If we calibrate interval estimates to cover **the estimator's mean** 95% of the time
  - and that's what we've really been doing
  how often will they cover the [thing we're estimating]{.fg style="color:green"}?
- This means biased estimators can cause big problems. We'll talk about this more later in the semester.
- But for now, let's just show the estimators that we've been using are, in fact, unbiased.

## The Sample Mean

**Claim.** The sample mean is an unbiased estimator of the population mean.
$$ \E[\hat\mu] = \mu $$

::: fragment

$$
\begin{aligned}
\E\qty[\frac1n\sum_{i=1}^n Y_i] 
&= \frac1n\sum_{i=1}^n \E[Y_i] && \text{ via linearity } \\
&= \frac1n\sum_{i=1}^n \mu     && \text{ via equal-probability sampling } \\
&= \frac1n \times n \times \mu = \mu.
\end{aligned}
$$

:::

## A Subsample Mean

**Claim.** The subsample mean is unbiased for the subpopulation mean.
$$ \E[\hat\mu(1)] = \mu(1) $$

**Sketch.**

  - Use the Law of Iterated Expectations, conditioning on $X_1 \ldots X_n$.
  - Then the linearity of conditional expectations to push the conditional expectation into the sum.
  - Then irrelevance of independent conditioning variables to write things in terms of the *random variable* $\mu(X_i)$  
  - Then the **indicator trick**. What's $X_i \mu(X_i)$? How is it related to $\mu(1)$?

::: fragment

$$
\small{
\begin{aligned}
\E[\hat\mu(1)] 
&=\E\qty[ E\qty{\frac{\sum_{i=1}^{n}X_i Y_{i}}{\sum_{i=1}^{n}X_{i}} \mid X_1 \ldots X_n}] \\
&=E\qty[\frac{\sum_{i=1}^{n} X_i \E\qty{ X_{i} \mid X_1 \ldots X_n}}{\sum_{i=1}^{n}X_{i}}] \\
&=E\qty[\frac{\sum_{i=1}^{n}X_i\E\qty{ Y_{i} \mid X_i}}{\sum_{i=1}^{n}X_{i}}] \\
&=E\qty[\frac{\sum_{i=1}^{n}X_i \mu(X_i)}{\sum_{i=1}^{n}X_{i}}] \\
&=E\qty[\frac{\sum_{i=1}^{n}X_i\mu(1)}{\sum_{i=1}^{n}X_{i}}] \\
&=E\qty[\mu(1) \times \frac{\sum_{i=1}^{n}X_i}{\sum_{i=1}^{n}X_{i}}] = \E[\mu(1)]=\mu(1) 
\end{aligned}
}
$$

:::

## Differences in Subsample Means

**Claim.** The difference in subsample means is unbiased for the difference in subpopulation means.

$$ \E[\hat\mu(1) - \hat\mu(0)] = \mu(1) - \mu(0) $$

::: fragment
This follows from the linearity of expectations and unbiasedness of the subsample means.

$$
\E[\hat\mu(1) - \hat\mu(0)] = \E[\hat\mu(1)] - \E[\hat\mu(0)] =  \mu(1) - \mu(0) 
$$

:::

## Summary

```{r}
#| layout-ncol: 2
#| 
xx = seq(-4,4,by=.01)
heights = seq(0,.4,length.out=100)
normals = rnorm(10000)

ggplot() + geom_histogram(aes(x=normals, y=after_stat(density)), alpha=.3, bins=100) +
           geom_vline(xintercept=0, color='green', linewidth=3, alpha=.4) + 
           geom_vline(xintercept=0, color='blue') + 
           geom_vline(xintercept=c(-1,1), color='blue', linetype='dashed') +
           geom_vline(xintercept=c(-2,2), color='blue', linetype='dotted') +
           geom_point(aes(y=heights, x=normals[1:length(heights)]), color='purple', alpha=.3)

ggplot() + geom_histogram(aes(x=normals+2, y=after_stat(density)), alpha=.3, bins=100) +
           geom_vline(xintercept=0, color='green', linewidth=3, alpha=.4) + 
           geom_vline(xintercept=2+0, color='blue') + 
           geom_vline(xintercept=2+c(-1,1), color='blue', linetype='dashed') +
           geom_vline(xintercept=2+c(-2,2), color='blue', linetype='dotted') +
           geom_point(aes(y=heights, x=normals[1:length(heights)]+2), color='purple', alpha=.3) +
           xlim(-4,4)
```

- So far, we've been using unbiased estimators. 
- Their sampling distributions, in relation to the estimation target, look like the one on the left.
- So now that we've worked out their location --- and are pleased --- it's time to talk about spread.
- That's what we'll do next time.




# This is Where The Lecture on Unbiasedness Used To End


## In the Next Section, We Start What Used To Be the Next Lecture (On Spread)

### TODO

- Move the probability review stuff out into a separate lecture.
- Merge the stuff on bias and spread here into one coherent lecture.


# Review. This is the first section of the old Spread Lecture
## Where We Are

:::: columns

::: column
```{r}
#| fig-width: 5
#| out-height: 300px
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=earnings, size=.1, alpha=.5) + guides(color='none')
```
:::

::: column

```{r}
style.tab = function(tab, cols=c('income', 'education', 'county'), nhead=3, ntail=1) {
  tab$income50k = 0+(tab$income > 50000)
  tab$income = sprintf('$%dk', round(as.numeric(tab$income) / 1000))
  tab$county = ifelse(tab$county=='los angeles', 'LA',  tab$county)
  ellipsis.tab = tab[, cols] |> ellipsis(nhead=nhead,ntail=ntail)
  ellipsis.tab |>
    kable() |> kable_styling() |>
    row_spec(which(ellipsis.tab$education >= 16),  background=twocolor.green) |>
    row_spec(which(ellipsis.tab$education <  16),  background=twocolor.red) 
}

sample.tab |> style.tab(nhead=4)
```
:::

::::

- We've talked about **estimating the mean of a population**. 
- And means of subpopulations, too. And differences in those.
- We've been looking the incomes of California residents [with]{.twocolor-green} and [without]{.twocolor-red} 4-year degrees.
  - Our population is the set of all California residents meeting some age and education criteria.
  - And we'll think of earnings as being drawn with replacement from that population.
- To think carefully about the relationship between earnings and the population, we'll use **random variables**.

## Imagining a Population

```{r}
#| out-height: 300px
#| fig-width: 5 
#| layout-ncol: 2 
#| 
ca.plot + 
 geom_point(aes(x=long, y=lat, color=degree), data=display.population, size=.02, alpha=.1) + 
 guides(color='none')
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=earnings, size=.1, alpha=.5) +
  guides(color='none')
```


:::: columns
::: column
```{r}
pop.tab |> style.tab()
```
:::
::: column
```{r}
sample.tab |> style.tab()
```
:::
::::


- We don't have much information about the people in our population. Except the ones in earnings. 
- But, for the sake of visualization, I've made some up. 
  - We'll put visualizations of this fake population on the left.
  - Visualizations of the (real) sample will be on the right.

## Unbiasedness

```{r}
xx = seq(-6,6,by=.01)
K = 100
heights = seq(0,.4,length.out=K)
normals = rnorm(10000)

unbiased.plot = ggplot() + 
  geom_histogram(aes(x=normals, y=after_stat(density)), alpha=.2, bins=100) + with_annotations(normals, estimand=0, dots=100, highlight.dots=0, interval.width=1.96) + xlim(min(xx),max(xx))

biased.plot = ggplot() + 
  geom_histogram(aes(x=normals+2, y=after_stat(density)), alpha=.2, bins=100) + with_annotations(normals+2, estimand=0, dots=100, highlight.dots=0, interval.width=1.96) + xlim(min(xx),max(xx))
```

```{r}
#| layout-ncol: 2
unbiased.plot
biased.plot
```

- We've shown that using sample means as estimators of population means gives us **unbiased estimators**.
  - That is, estimators whose expected values are the thing we want to estimate.
  - Intuitively, that means their sampling distributions are where we want them to be.
  - Like the plot on the left. Not the one on the right. 
- This is essentially necessary to get correctly calibrated interval estimates.
- It's more or less sufficient too. 
  - Calibration via bootstrap works for every unbiased estimator we'll talk about this semester. 
  - Things are more complicated for fancy machine learning stuff, but that's beyond the scope.

## We Can Trust This Interval Estimate

```{r}
#| cache: true
X = as.numeric(earnings$education >= 16)
Y = earnings$income 
n = length(Y)

difference = mean(Y[X==1]) - mean(Y[X==0])
samples = array(dim=10000)
for(rr in 1:10000) {
    I = sample(1:n, n, replace=TRUE)
    Xstar = X[I]
    Ystar = Y[I]
    samples[rr] = mean(Ystar[Xstar==1]) - mean(Ystar[Xstar==0])
}
```

```{r}
#| out-height: 250px
#| fig-asp: .25 
 
interval.plot = function(samples, point.estimate, interval) { 
  xx = seq(min(samples), max(samples), length.out=1000)
  sample.data=data.frame(samples=samples)
  interval.data=data.frame(x = point.estimate, 
                           xmin = min(interval),
                           xmax = max(interval), 
                           y=.4*default_height(samples))
  ggplot() +
    geom_histogram(aes(x=samples, y=after_stat(density)), 
                   alpha=.2, bins=100, data=sample.data) +
    with_annotations(samples, dots=100, highlight.dots=0) +
    geom_pointrange(aes(x=x,y=y, xmin=xmin, xmax=xmax), 
                   data=interval.data, color='black')
}

wboot=width(samples, alpha=.05)
cali.interval = difference + c(-1,1)*wboot/2
cali.interval.plot = interval.plot(samples, difference, cali.interval)
cali.interval.plot
``` 

$$
\mu(1)-\mu(0) \in `r approx(difference)` \pm `r approx(wboot/2)` \qqtext{is a 95\% confidence interval}
$$ 

- This is a 95% **confidence interval** for the difference in mean income between California residents [with]{.twocolor-green} and [without]{.twocolor-red} 4-year degrees.
- In 95% of surveys, it'll contain the **different in population means**. That's the point of all of this.

## The NSW version

```{r}
#| cache: true
nsw = read.csv('https://davidahirshberg.bitbucket.io/teaching/regression/fall2021/data/nsw.csv')
nsw.experimental = nsw[nsw$dataset=='nsw', ]
X = as.numeric(0+nsw.experimental$treated)
Y=as.numeric(nsw.experimental$income78)
n = length(Y)

difference = mean(Y[X==1]) - mean(Y[X==0])
samples = array(dim=10000)
for(rr in 1:10000) {
    I = sample(1:n, n, replace=TRUE)
    Xstar = X[I]
    Ystar = Y[I]
    samples[rr] = mean(Ystar[Xstar==1]) - mean(Ystar[Xstar==0])
}
```

```{r}
#| out-height: 250px
#| fig-asp: .25 
wboot=width(samples, alpha=.05)
nsw.interval = difference + c(-1,1)*wboot/2
nsw.interval.plot = interval.plot(samples, difference, nsw.interval)
nsw.interval.plot
``` 


$$
\mu(1)-\mu(0) \in `r approx(difference)` \pm `r approx(wboot/2)` \qqtext{is a 95\% confidence interval}
$$ 

- This is a 95% **confidence interval** for the difference in mean income between California residents [with]{.twocolor-green} and [without]{.twocolor-red} 4-year degrees.
- In 95% of surveys, it'll contain the **different in population means**. That's the point of all of this.

## The NSW version

```{r}
#| out-height: 250px
#| fig-asp: .25 
inflation.factor = 3.5
nsw.interval.plot
``` 

$$
\mu(1)-\mu(0) \in `r approx(difference)` \pm `r approx(wboot/2)` \qqtext{is a 95\% confidence interval}
$$ 

- This is a 95% **confidence interval** for the difference in mean income between treated and untreated people in the National Supported Work experiment.
- That's in 1978 dollars. Accounting for inflation, it's about $`r round(inflation.factor,1)`$ times that in 2022 dollars.

$$
\mu(1)-\mu(0) \in `r approx(inflation.factor*difference)` \pm `r approx(inflation.factor*wboot/2)` \qqtext{is a 95\% confidence interval (in 2022 dollars)}
$$ 

## What's The Difference?

```{r}
#| layout-ncol: 2
#| fig-cap:
#| - "California 4-year degree vs no 4-year degree"
#| - "NSW treated vs untreated (adjusted for inflation)"

cali.interval.plot
nsw.interval.plot + scale_x_continuous(labels=function(x){x*inflation.factor})
```

- Do these intervals tell the same story about the income differences between the groups involved?

::: fragment

- In my opinion, no. 
- The California interval tells us there's a big difference between these groups in the population.
  - It's `r approx(cali.interval[1])` to `r approx(cali.interval[2])`. That's a big difference no matter which end of the interval the population difference is on.
- The NSW interval tells us we don't know.
  - It's `r approx(inflation.factor*nsw.interval[1])` to `r approx(inflation.factor*nsw.interval[2])` in 2022 dollars. 
  - That's a smallish difference at one end of the interval and a big one at the other.
- What we want is to be able to anticipate this and design our studies to avoid it. 
- The key to this is understanding the **spread** of the distribution of our estimator. That's what we'll talk about today.

:::

# Probability Stuff

## Review: Observations as Random Variables

```{r}
#| out-height: 300px
#| layout-ncol: 2 

earnings.jitter.pop = ggplot(pop.earnings) + 
    geom_point(aes(x=degree, y=income, color=degree), 
               size=.2, position=position_jitter(seed=0,w=.3), alpha=.1) +
     scale_y() + guides(color='none')

earnings.jitter = ggplot(earnings) + 
    geom_point(aes(x=degree, y=income, color=degree), 
               size=dot.size, position=position_jitter(seed=0,w=.3), alpha=.1) +
     scale_y() + guides(color='none')
```


```{r}
#| out-height: 300px
#| layout-ncol: 2 
summaries = earnings %>% group_by(degree) %>% summarize(mean=mean(income), sd=sd(income))

earnings.jitter.pop.withspread = earnings.jitter.pop +
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x), size=1, alpha=.5) 
earnings.jitter.withspread =  earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x), size=1, alpha=.5) 

earnings.jitter.pop.withspread
earnings.jitter.withspread
```

- The notation $y_j$ refers to a number.
  - It's the income of the $j$th person in our population. That's just some person. 
- The notation $Y_i$ refers to something else.
  - It's the income of the $i$th person we call in our survey. That's not a person.
  - That's the result of a random process---the roll of a die.
- To summarize this result, we talk about the **probability distribution** of $Y_i$.

### Notation Conventions

- Random variables are written in **uppercase**: $X$, $Y$, $Z$, etc. Constants are written in **lowercase**: $x$, $y$, $z$, etc.
- We'll use the same letter for a random variable and the value it takes on. $x$ is a possible value of $X$, $y$ of $Y$, etc.
- Estimators are also random variables, but instead of uppercase we use a **hat**: $\hat \mu$, $\hat \sigma$, etc.
- We'll use the same letter for the estimator and what it's meant to estimate. $\hat \mu$ is an estimate of $\mu$.


## Review: Expectations

```{r}
cols = c('income50k', 'income', 'education', 'county')
sample.tab$roll = sample.tab$J
```

:::: columns
::: column
```{r}
style.tab(pop.tab)
```
:::
::: column
```{r}
style.tab(sample.tab, cols=c('roll', cols))
```
:::
::::

- Often the thing we care about is the expected value of $Y_i$. Or some related quantity.
  - The expected value of $Y_i$ is the **probability-weighted average** of the values it can take on.
  - What's nice about this is that we can think of this in 'uncollapsed' terms.
- When we sample as usual, this is just the population mean.
- The collapsed form is, in a sense, just summing in a different order.

$$
\begin{aligned}
\E[Y_i] = \sum_y P(Y_i = y) \times y
= \sum_y \qty(\sum_{j: y_j = y} \frac{1}{m}) \times y 
= \frac{1}{m}\sum_y\sum_{j:y_j = y} y 
= \frac{1}{m}\sum_{j=1}^m y_j
\end{aligned}
$$

## Rules for Expectations

### The law of iterated expectations
$$
E \{ E( Y \mid X ) \} \quad \text{ for any random variables $X, Y$} 
$$

To calculate the mean of $Y$, we can average within subpopulations, then across subpopulations.

### Irrelevance of independent conditioning variables
$$ E( Y \mid X, X' ) = E( Y \mid X ) \quad \text{ when $X'$ is independent of $X,Y$ } $$

If $X'$ is unrelated to $X$ and $Y$, holding it constant doesn't affect the relationship between them.

### Linearity of Expectations
$$ 
E ( a Y + b Z ) = aE(Y) + bE(Z)  \quad \text{ for random variables $Y, Z$ and numbers $a,b$ }
$$

We can distribute expectations through sums and multiplication by constants.

### Linearity of Conditional Expectations

$$
E\{ a(X) Y + b(X) Z \mid X \} = a(X)E(Y \mid X) + b(X)E(Z \mid X)
$$

When we condition on $X$, we can distribute as if functions of $X$ were constants.

## An Important Consequence.

The expectation of a product of **independent** random variables is the product of their expectations.
$$
\E[YZ] = \E[Y]\E[Z] \qqtext{when $Y$ and $Z$ are independent}
$$

This comes up a lot in variance calculations. Let's prove it. It'll be good practice.

::: fragment

$$
\begin{aligned}
\E[YZ] &= \E\qty[ \E\qty{YZ \mid Z}] && \text{iterated expectations} \\
&= \E\qty[ Z\E\qty{Y \mid Z} ] && \text{linearity of conditional expectations} \\
&= \E\qty[ Z \E\qty{Y} ]       && \text{irrelevance of independent conditioning variables} \\
&= \E\qty{Y}\E[Z]              && \text{linearity of expectations} 
\end{aligned}
$$

:::

## Spread

```{r}
#| out-height: 300px
#| layout-ncol: 2 
#| 
ggplot(earnings) + 
  geom_point(aes(x=0, y=income), color='blue', 
            size=2, position=position_jitter(seed=0,w=.3), alpha=.2) +
  stat_summary(aes(x=0, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) +
          scale_y() 

earnings.jitter.withspread = 
  earnings.jitter + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) +
          scale_y()

earnings.jitter.withspread
```

- To summarize the **spread** of a distribution, we often talk about variance.
- The variance is an expectation: it's the **expected squared difference** between a random variable and its expectation.
$$
\Var(Y) = \E \qty[ \qty{ Y - \E(Y) }^2 ]
$$

- To put that back on the same scale as $Y$---instead of the scale of $Y^2$---we take it's square root. That's the standard deviation.

$$
\text{standard deviation}(Y) = \sqrt{\Var(Y)}
$$

- We can talk about spread outright (left) --- that's (unconditional) variance.
- Or spread within groups (right) --- that's conditional variance.



## Conditional Variance {.smaller}

```{r}
Y = c(.75,1,1.25, 1,1.25,1.5)
X = 0 + (1:6 > 3)
Xdodge = X + rep(.02*c(-1,0,1),2)
mu = function(x) { mean(Y[X==x]) }

tinypop.plot = ggplot(data.frame(X=X,Y=Y,Xdodge=Xdodge)) + 
  geom_point(aes(x=Xdodge, y=Y, color=factor(X)), alpha=.5, size=3*dot.size) + 
  guides(color='none')
tinypop.square.plot = tinypop.plot + 
  geom_point(aes(x=Xdodge, y=Y^2, color=factor(X)), shape=0, size=3*dot.size) + 
  geom_segment(aes(x=Xdodge, xend=Xdodge, y=Y, yend=Y^2), alpha=.3) 
tinypop.plot.withspread = tinypop.plot + 
  stat_summary(aes(x=X, y=Y), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x), size=.5, alpha=.5)

tinylims.0 = xlim(min(Xdodge[X==0])-sd(Xdodge[X==0]), max(Xdodge[X==0] + sd(Xdodge[X==0])))
tinylims = xlim(min(Xdodge)-sd(Xdodge), max(Xdodge + sd(Xdodge)))
```

```{r}
#| out-height: 250px
#| fig-asp: .5
#| layout-ncol: 2 
tinypop.plot.withspread
earnings.jitter.pop.withspread
```

- The **conditional variance** describes the spread of one random variable *within groups*. 
- Like the conditional expectation of the random variable, there are two related concepts.

1. The **conditional variance function**. This is a function of $x$.
$$
\sigma^2(x) := \Var(Y \mid X=x) = \E\qty[ \textcolor{blue}{\qty{ Y-\E(Y \mid X) }^2} \mid X=x]
$$

2. The conditional variance $\sigma^2(X) = \Var(Y \mid X)$. This is a random variable.
$$
\sigma^2(X) := \Var(Y \mid X)   = \E\qty[ \textcolor{blue}{\qty{ Y-\E(Y \mid X) }^2} \mid X]
$$


## The Conditional Variance Function

```{r}
#| out-height: 250px
#| fig-asp: .4
#| layout-ncol: 2 

tinypop.plot 
earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

$$
\sigma^2(x) := \Var(Y \mid X=x) = \E\qty[ \textcolor{blue}{\qty{ Y-\E(Y \mid X) }^2} \mid X=x]
$$

- This is the variance of $Y$ within the subpopulation of people with $X=x$.
  - $\sigma(1)$ is a number describing the spread of the green dots in our population.
  - $\sigma(0)$ is a number describing the spread of the red dots in our population.
- It's the conditional expectation function of the random variable $\textcolor{blue}{\{Y-\E(Y \mid X)\}^2}$.
- We call it 'the conditional variance function of $Y$ given $X=x$.

## The Conditional Variance 
```{r}
#| out-height: 250px
#| fig-asp: .4
#| layout-ncol: 2 

tinypop.plot 
earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

$$
\sigma^2(X) := \Var(Y \mid X)   = \E\qty[ \textcolor{blue}{\qty{ Y-\E(Y \mid X) }^2} \mid X]
$$

- This is the variance of $Y$ within **a random subpopulation** of people.
    - It's the conditional variance function **evaluated at** the random variable $X$.
    - And  the conditional expectation of the random variable $\textcolor{blue}{\{Y-\E(Y \mid X)\}^2}$.
- We call it 'the conditional variance of $Y$ given $X$'.
- This is the sort of thing that shows up when you use the law of iterated expectations.


## The Law of Total Variance

$$
\Var[Y] = \E\qty{\Var( Y \mid X ) } + \Var\qty{\E( Y \mid X ) }  
$$

- The Law of Total Variance relates Conditional and Unconditional Variance
- It's like the Law of Total Expectation, but for Variance
- It's a useful way to decompose the variance a random variable.
- Think about where most of the variance in $Y$ is coming from in the two populations shown below.

```{r}
#| out-height: 500px
#| fig-ncol: 2

ggplot() + geom_point(aes(x=Xdodge, y=Y, color=factor(X)), alpha=.7, size=4) + 
  scale_y_continuous(breaks=seq(0,3,by=.5), limits=c(0,3)) + guides(color='none') 
ggplot() + geom_point(aes(x=Xdodge, y=Y+X, color=factor(X)), alpha=.7, size=4) + 
  scale_y_continuous(breaks=seq(0,3,by=.5), limits=c(0,3)) + guides(color='none') 

```


## An Activity 

```{r}
#| out-height: 250px
#| fig-asp: .25

mean.has.degree = round(summaries$mean[summaries$degree=='4-year degree'] / 10000) * 10
mean.no.degree  = round(summaries$mean[summaries$degree=='no 4-year degree'] / 10000) * 10
degree.proportion = round(mean(earnings$degree=='4-year degree'),1)
earnings.jitter.pop + 
  stat_summary(aes(x=degree, y=income), geom="pointrange", fun=mean, 
          fun.min = function(x) mean(x) - sd(x), fun.max = function(x) mean(x) + sd(x)) 
```

- Working with your neighbor, think about why we call it this.
- Try to put the three terms into words.
    - $\Var[Y]$ is the ...
    - $\E\qty{ \Var(Y \mid X) }$ is the ...
    - $\Var\qty{\E( Y \mid X)}$ is the ...
- Do this in two ways.
    1. Using 'Random Variables Language', 
        - e.g. 'the expected value of the conditional variance of Y given X'
    2. Using 'Population Language'
        - e.g. 'the variance of income (Y) within the random subpopulation with degree status X'


# Proving the Law of Total Variance

## Starting Point: Interpreting Variance as Excess {.smaller}

$$
\small{
\begin{aligned}
\Var(Y) &= \E \qty[ \qty{ Y - \E(Y) }^2 ] && \text{ the \emph{spread} idea: the mean square of a \emph{centered version} of $Y$ } \\
&= \E\qty(Y^2) - \qty{\E\qty(Y)}^2         && \text{ the \emph{excess} idea: the average amount $Y^2$ \emph{exceeds} the square of its mean}.
\end{aligned}
}
$$

```{r}
#| out-height: 250px
#| fig-asp: .25

tinypop.square.plot + tinylims 
```

Let's think, in terms of the plot, about why there is excess in the square.

- Suppose $Y$ takes on three values --- the dots ---with equal probability.
    - Its expectation is the average height of the black dots.
    - And its expectation, squared, is the square of this.
- Its square $Y^2$ takes on the the square of those values---the squares --- with equal probability.
    - Its expectation is the average height of the squares.
- Why is the second one bigger?
  - Because squaring *increases big draws of Y* more than it *decreases small ones*.
  - Visually, the line from $Y$ to $Y^2$ is longer when $Y$ is bigger than when it's smaller.
- We'll mostly think about variance in terms of spread, but the 'excess' formula  comes up in calculations.

## Equivalence Proof

**Claim.** The spread and excess formulas for variance are equivalent.
$$  
\E \qty[ \qty{ Y - \E(Y) }^2 ] =\E[ Y^2] - \qty{\E(Y)}^2 
$$

$$
\small{
\begin{aligned}
\E \qty[ \qty{ Y - \E(Y) }^2 ] 
&= \E \qty[ Y^2 - 2 Y \E(Y) + \qty{\E(Y)}^2 ] && \text{ FOIL } \\
&= \E[ Y^2] - \E\qty[2 Y \E(Y)] + \E\qty[\qty{\E(Y)}^2] && \text{Distributing expectations (linearity)} \\
&= \E[ Y^2] - 2\E(Y)\E\qty[Y] + \qty{\E(Y)}^2\E[1] && \text{Pulling constants out of expectations (linearity)} \\
&= \E[ Y^2] - 2\qty{\E(Y)}^2 + \qty{\E(Y)}^2] && \text{Recognizing a product of something with itself as a square} \\
&= \E[ Y^2] - \cancel{2}\qty{\E(Y)}^2 + \cancel{\qty{\E(Y)}^2}] && \text{Subtracting} 
\end{aligned}
}
$$


## Conditional Variance as Excess {.smaller}

$$
\small{
\begin{aligned}
\Var(Y \mid X) &= \E \qty[ \qty{ Y - \E(Y\mid X)}^2 \mid X ] && \text{ the \emph{spread} idea: the mean square of a conditionally centered version of $Y$ } \\
&= \E\qty(Y^2 \mid X) - \qty{\E\qty(Y \mid X)}^2              && \text{ the \emph{excess} idea: the average amount $Y^2$ exceeds the square of its cond. mean}
\end{aligned}
}
$$


```{r}
#| out-height: 500px
#| fig-asp: .5
tinypop.square.plot + tinylims
```

- We can think of the Conditional Variance as excess, too.
- It's the same as the unconditional case, but *within subpopulations*.


## Proof

**Claim.**
$$
\Var[Y] = \textcolor{blue}{\E\qty[\Var(Y \mid X)]} + \textcolor{green}{\Var\qty[\E(Y \mid X)]} 
$$

1. Rewrite the variances, conditional and unconditional, in *excess* form.

::: fragment
$$
\Var[Y]  = \textcolor{blue}{\E\qty[ \E(Y^2 \mid X) - \qty{\E\qty(Y \mid X)}^2]} 
                       + \textcolor{green}{\E\qty{ \E(Y \mid X)^2}  - \qty(\E\qty{\E\qty(Y \mid X)})^2} 
$$

:::

2. Distribute the outer expectations in the first term.

::: fragment
$$
\Var[Y] = \textcolor{blue}{\E\qty[ \E(Y^2 \mid X)] - \E\qty[\qty{\E\qty(Y \mid X)}^2]} 
                       + \textcolor{green}{\E \qty[ \qty{\E(Y \mid X)^2} ] - \qty(\E\qty{\E\qty(Y \mid X)})^2}
$$

:::

3. Cancel matching terms

::: fragment
$$
\Var[Y]  = \textcolor{blue}{\E\qty[ \E(Y^2 \mid X)] \cancel{- \E\qty[\qty{\E\qty(Y \mid X)}^2]}}  
                       + \textcolor{green}{\cancel{\E \qty[ \qty{\E(Y \mid X)^2} ]} -  \qty(\E\qty{\E\qty(Y \mid X)})^2}
$$

:::

4. Use the law of iterated expectations to simplify the remaining terms.

::: fragment
$$
\Var[Y]  = \textcolor{blue}{E[Y^2]  \cancel{- \E\qty[\qty{\E\qty(Y \mid X)}^2]}}
                       + \textcolor{green}{\cancel{\E \qty[ \qty{\E(Y \mid X)^2} ]} - (\E\qty{Y})^2}
$$

:::

# Variance Calculations

## The Variance of The Mean

$$
\begin{aligned}
\Var[\bar Y] 
&= \E\qty[ \qty{ \frac{1}{n}\sum_i Y_i - \E \qty( \frac{1}{n}\sum_i Y_i ) }^2 ] && \text{definition}\\
&= \E\qty[ \qty{ \frac{1}{n}\sum_i (Y_i - \E Y_i) }^2 ]  && \text{Linearity} \\
&= \E\qty[ \qty{ \frac{1}{n}\sum_i Z_i }^2 ] && \text{for} \ \ Z_i = Y_i - \E Y_i \\
&= \E\qty[ \frac{1}{n^2}\sum_i \sum_j Z_i Z_j ]   && \text{via FOIL}
\end{aligned}
$$

### FOIL?

- Remember FOIL from high school algebra? 
- When we square a sum, we get *all products* of terms in the sum.
  
$$ 
(a+b)^2 = \inblue{(a+b)}\inred{(a+b)} 
= \inblue{a}\inred{a} + \inblue{a}\inred{b} + \inblue{b}\inred{a} + \inblue{b}\inred{b}
= a^2 + 2ab + b^2
$$

- Same deal for bigger sums.

$$ 
\qty{ \sum_i Z_i }^2 = \inblue{\sum_i Z_i}\inred{\sum_j Z_j} = \sum_i \sum_j \inblue{Z_i}\inred{Z_j}
$$

## Completing the Calculation

$$
\begin{aligned}
\Var[\bar Y] &= \E\qty[ \frac{1}{n^2}\sum_i \sum_j Z_i Z_j ]  \quad \text{ for } \quad Z_i = Y_i - \E Y_i \quad && \text{via FOIL}
\end{aligned}
$$

**Activity**. Complete the calculation to get a simple expression. 

:::: columns
::: column
 **Hints**.

  1. Use linearity to push the expectation inside the sum.
  2. Think about what $\E Z_i Z_j$ is when $i=j$ and $i\neq j$.
  3. Recall that when $i \neq j$ they're independent.
  4. What is $\E Z_i^2$?  
:::
::: column
**Rules**.
$$
\small{
\begin{aligned}
\text{1.} \quad & \E\qty[ \sum_i a_i Y_i] =  \sum_i a_i \E Y_i \quad \text{ for RVs $Y_i$ and constants $a_i$.} \\
\text{2.} \quad & \E\qty[ Z_i Z_j] = \E[Z_i] \E[ Z_j]  \quad \text{if $Z_i$ and $Z_j$ are independent.} \\
\end{aligned}
}
$$

:::
::::

## The Variance of The Mean: Answered
$$
\begin{aligned}
\Var[\bar Y] 
&= \E\qty[ \frac{1}{n^2}\sum_i \sum_j Z_i Z_j ]  && \text{via FOIL} \\
&= \frac{1}{n^2} \sum_i \sum_j \E Z_i Z_j    && \text{via Linearity } \\
&= \frac{1}{n^2} \sum_i\sum_j \begin{cases} \E Z_i^2 & \text{if } i=j \\ \E Z_i \E Z_j & \text{if } i\neq j \end{cases} && \text{using the independence of $Z_i$ and $Z_j$} \\
&= \frac{1}{n^2} \sum_i\sum_j \begin{cases}  \Var(Y_i) & \text{if } i=j \\ 0 \times 0 & \text{if } i\neq j \end{cases} && \text{since $Z_i=Y_i-\E Y_i$} \\
&= \frac{1}{n^2} \times n \times \Var(Y_i) = \frac{\Var(Y_i)}{n} = \frac{\sigma^2}{n}.   && \text{counting the copies of $\Var(Y_i)$ in the sum} \\
\end{aligned}
$$

## The Variance of the Subsample Mean

**Claim.** The variance of our point estimate is the **expected value** of the subpopulation variance divided by the number of people in the subsample.
$$ \Var[\hat\mu(1)] = \E\qty[ \frac{\sigma^2(1)}{N_1} ] \quad \text{ for } \quad N_1 = \sum_{i=1}^n X_i $$

- This is like what we saw in the unconditional case, with a twist.
- The number of people in the subsample is random, so instead of $1/n$, we have $\E[ 1/N_1 ]$.
- Let's prove it.
  - It's not hard, but it's a bit long. The key is organization. 
  - We'll break things down into manageable pieces and focus on one piece at a time. 


## Step 1. Total Variance 

```{r}
#| out-height: 250px
#| fig-asp: .25
tinypop.plot + tinylims
```

- We'll start by using the Law of Total Variance, conditioning on $X_1 \ldots X_n$.
- And then we'll have to calculate two terms.
  1. The expected value of the conditional variance
  2. The variance of the conditional expectation 
$$
\begin{aligned}
\Var\qty[ \hat \mu(1) ] &= \E \qty[ \Var\qty{ \hat \mu(1) \mid X_1 \ldots X_n }] + \Var\qty[ \E\qty{\hat \mu(1) \mid X_1 \ldots X_n} ] \\
\end{aligned}
$$

- Plugging in the definitions of variance and conditional variance, these terms are 

$$
\begin{aligned}
\Var\qty{ \hat \mu(1) \mid X_1 \ldots X_n}  
&= \E\qty( \qty[ \hat \mu(1) - \textcolor{blue}{\E\qty{\hat \mu(1) \mid X_1 \ldots X_n}} ]^2 \mid X_1 \ldots X_n) \\
\Var\qty[\E\qty{\hat \mu(1) \mid X_1 \ldots X_n}] 
&= \E\qty( \qty{\textcolor{blue}{\E\qty{\hat \mu(1) \mid X_1 \ldots X_n}} - \E\qty[\E\qty{\hat \mu(1) \mid X_1 \ldots X_n}]}^2 ) \\
\end{aligned}
$$

- We see that the [conditional expectation of our estimator]{.fg style="color:blue"} comes up in both terms. 
- I've claimed it's just $\mu(1)$. Let's take a look at that.  

## Step 2. The Estimator's Conditional Mean {.smaller}

- To calculate the conditional mean of our estimator, we'll revisit an argument we've seen before.
- We'll look at our argument that our argument that the subsample mean $\hat\mu(1)$ is unbiased. 
  - We'll see that it's **conditionally unbiased**.
  - That is, its expected value is $\mu(1)$, the thing we want to estimate 
  - *conditional on* the values of the covariates $X_1 \ldots X_n$.

$$
\begin{aligned}
\E\qty[\hat\mu(1) \mid X_1 \ldots X_n] 
&=\E\qty[\frac{\sum_{i=1}^{n}X_i Y_{i}}{\sum_{i=1}^{n}X_{i}} \mid X_1 \ldots X_n] \\
&=\frac{\sum_{i=1}^{n} X_i \E\qty{ Y_{i} \mid X_1 \ldots X_n}}{\sum_{i=1}^{n}X_{i}} \\
&=\frac{\sum_{i=1}^{n}X_i\E\qty{ Y_{i} \mid X_i}}{\sum_{i=1}^{n}X_{i}}  \\
&=\frac{\sum_{i=1}^{n}X_i \mu(X_i)}{\sum_{i=1}^{n}X_{i}}  \\
&=\frac{\sum_{i=1}^{n}X_i\mu(1)}{\sum_{i=1}^{n}X_{i}} \\
&=\mu(1) \times \frac{\sum_{i=1}^{n}X_i}{\sum_{i=1}^{n}X_{i}} = \mu(1) 
\end{aligned}
$$

## Step 3. Simplifying Our Total Variance Decomposition

$$
\begin{aligned}
\Var\qty[ \hat \mu(1) ] &= \E \qty[ \Var\qty{ \hat \mu(1) \mid X_1 \ldots X_n }] + \Var\qty[ \E\qty{\hat \mu(1) \mid X_1 \ldots X_n} ] 
\end{aligned}
$$

where 

$$
\begin{aligned}
\Var\qty{ \hat \mu(1) \mid X_1 \ldots X_n}  
&= \E\qty( \qty[ \hat \mu(1) - \textcolor{blue}{\E\qty{\hat \mu(1) \mid X_1 \ldots X_n}} ]^2 \mid X_1 \ldots X_n) \\
&= \E\qty( \qty[ \hat \mu(1) - \textcolor{blue}{\mu(1)} ]^2 \mid X_1 \ldots X_n) \\
\Var\qty[\E\qty{\hat \mu(1) \mid X_1 \ldots X_n}] 
&= \E\qty( \qty{\textcolor{blue}{\E\qty{\hat \mu(1) \mid X_1 \ldots X_n}} - \E\qty[\E\qty{\hat \mu(1) \mid X_1 \ldots X_n}]}^2 ) \\
&= \E\qty( \qty{\textcolor{blue}{\mu(1)} - \E\qty[\mu(1)] }^2 )  = 0\\
\end{aligned}
$$

so

$$
\Var\qty[ \hat \mu(1) ] = \E \qty[ \E\qty{ \qty[\hat \mu(1) - \textcolor{blue}{\mu(1)} ]^2 \mid X_1 \ldots X_n }]
$$

## Step 4. Rewriting as a Sum of Mean-Zero Random Variables {.smaller}

$$
\Var\qty[ \hat \mu(1) ] 
= \E \qty[ \E\qty{ \qty[\hat \mu(1) - \textcolor{blue}{\mu(1)} ]^2 \mid X_1 \ldots X_n }]
$$

- Now we'll substitute in definitions and simplify.
- We'll end having to work with the conditional mean of a *squared sum*.

$$
\small{
\begin{aligned}
\Var\qty[ \hat \mu(1) ] 
&= \E\qty[ \E\qty{  \qty(\hat \mu(1)-\mu(1))^2 \mid X_1 \ldots X_n }] \\
&= \E\qty[ \E\qty{  \qty(\frac{\sum_{i=1}^{n}X_i Y_{i}}{\sum_{i=1}^{n}X_{i}} - \mu(1))^2 \mid X_1 \ldots X_n }] \\
&= \E\qty[ \E\qty{  \qty(\frac{\sum_{i=1}^{n}X_i Y_{i}}{\sum_{i=1}^{n}X_{i}} - \frac{\mu(1)\sum_{i=1}^n X_i}{\sum_{i=1}^n X_i })^2 \mid X_1 \ldots X_n }] \\
&= \E\qty[ \E\qty{  \qty(\frac{\sum_{i=1}^{n}X_i \qty{Y_{i}-\mu(1)}}{\sum_{i=1}^{n}X_{i}})^2 \mid X_1 \ldots X_n }] \\
&= \E\qty[ \frac{\E\qty{ \qty(\sum_{i=1}^{n}X_i \qty{Y_{i}-\mu(1)})^2 \mid X_1 \ldots X_n}}{\qty(\sum_{i=1}^{n}X_{i})^2}]
\end{aligned}
}
$$

## Step 5 {.smaller}

$$
\Var\qty[ \hat \mu(1) ]  = \E\qty[ \frac{\E\qty{ \qty(\sum_{i=1}^{n}X_i \qty{Y_{i}-\mu(1)})^2 \mid X_1 \ldots X_n}}{\qty(\sum_{i=1}^{n}X_{i})^2}]
$$

- To work with this squared sum, we do our FOIL-like expansion. This is ...

$$ \small{ 
\begin{aligned}
\qty(\sum_{i}Z_i)^2 = \qty(\sum_{i}Z_i) \qty(\sum_{j}Z_i) = \sum_i\sum_j Z_i Z_j \quad && \text{ for } \quad Z_i = X_i \qty{Y_{i}-\mu(1)} \\
\text{or equivalently, using the \textbf{indicator trick}} && \text{ for } \quad Z_i = X_i\qty{Y_i - \mu(X_i)}
\end{aligned}
}
$$

- This lets us push the conditional mean *all the way into the sum*, so it meets the individual terms.
- And when we've done that, we can use the irrelevance of independent conditioning variables to simplify.

$$
\small{
\begin{aligned}
\Var\qty[ \hat \mu(1) ] 
&= \E\qty[ \frac{\E\qty{ \qty(\sum_{i=1}^{n}Z_i )^2 \mid X_1 \ldots X_n}}{\qty(\sum_{i=1}^{n}X_{i})^2}] \\
&= \E\qty[ \frac{\E\qty{ \sum_{i=1}^{n}\sum_{j=1}^n Z_i Z_j \mid X_1 \ldots X_n}}{\qty(\sum_{i=1}^{n}X_{i})^2}] \\
&= \E\qty[ \frac{\sum_{i=1}^{n}\sum_{j=1}^n \E\qty{Z_i Z_j \mid X_1 \ldots X_n}}{\qty(\sum_{i=1}^{n}X_{i})^2}] \\
&= \E\qty[ \frac{\sum_{i=1}^{n}\sum_{j=1}^n \E\qty{Z_i Z_j \mid X_i, X_j}}{\qty(\sum_{i=1}^{n}X_{i})^2}] \\
\end{aligned}
}
$$

## Step 6 {.smaller}

$$
\begin{aligned} 
\Var\qty[ \hat \mu(1) ] 
&= \E\qty[ \frac{\sum_{i=1}^{n}\sum_{j=1}^n \E\qty{Z_i Z_j \mid X_i, X_j}}{\qty(\sum_{i=1}^{n}X_{i})^2}] \quad \text{ for } \quad Z_i=Y_i-\mu(X_i)
\end{aligned}
$$


- Let's take a look at one of the terms in the numerator, $\E\qty{Z_i Z_j \mid X_i, X_j}$.
  - We're going to show that the cross-terms --- the ones for $i \neq j$ --- are zero.
  - This will let us drop them from the sum.
  - We'll use the fact that $Z_i$ and $Z_j$ are independent for $i \neq j$.
- Let's rewrite a cross-term using our usual stuff
  - the law of iterated expectations, conditioning on $Z_i$
  - linearity of expectations;
  - irrelevance of independent conditioning variables.

$$\small{
\begin{aligned}
\E\qty{Z_i Z_j \mid X_i, X_j} 
&= \E\qty{ \E(Z_i Z_j \mid Z_i,X_i,X_j) } \\
&= \E\qty{ Z_i \E(Z_j \mid X_i,X_j) } \\
&= \E\qty{ Z_i \textcolor{blue}{\E(Z_j \mid X_j)} } \quad \text{ where } \\
\textcolor{blue}{\E(Z_j \mid X_j)} 
&= \E\qty[X_j \qty{Y_j - \mu(X_j)}] \\ 
&= \E\qty{ \E\qty[X_j \qty{Y_j - \mu(X_j)} \mid X_j] } \\
&= \E\qty[X_j \qty{ \E\qty[Y_j \mid X_j] - \mu(X_j)}] \\
&= \E\qty[X_j \cross{\qty{ \mu(X_j) - \mu(X_j)}}] = 0. \\
\end{aligned}
}
$$

## Step 7 {.smaller}

$$
\begin{aligned} 
\Var\qty[ \hat \mu(1) ] 
&= \E\qty[ \frac{\sum_{i=1}^{n}\sum_{j=1}^n \E\qty{Z_i Z_j \mid X_i, X_j}}{\qty(\sum_{i=1}^{n}X_{i})^2}] \quad \text{ for } \quad Z_i=Y_i-\mu(X_i)  \\
&= \E\qty[ \frac{\sum_{i=1}^{n} \E\qty{Z_i^2 \mid X_i}}{\qty(\sum_{i=1}^{n}X_{i})^2}] \quad \text{because cross-terms are zero}.
\end{aligned}
$$

- All we've got to do now is recognize what we have. 
- Let's look at a term in the numerator.
  - We'll use iterated expecations and linearity.
  - Then the indicator trick.

$$
\small{
\begin{aligned}
\E\qty[Z_i^2 \mid X_i] 
&= \E\qty[ X_i\qty{Y_i - \mu(X_i)}^2 \mid X_i] 
= X_i\E\qty[\qty{Y_i - \mu(X_i)}^2 \mid X_i] 
= X_i\Var\qty[Y_i \mid X_i] = X_i \sigma^2(X_i) 
= X_i\sigma^2(1) 
\end{aligned}
}
$$

- We finish up by substituting this back into what we started with.

$$
\small{ 
\begin{aligned}
\Var\qty[ \hat \mu(1) ] 
&= \E\qty[ \frac{\sum_{i=1}^{n} \E\qty{Z_i^2 \mid X_i}}{\qty(\sum_{i=1}^{n}X_{i})^2}]  
= \E\qty[ \frac{\sum_{i=1}^{n} X_i \sigma^2(1)}{\qty(\sum_{i=1}^{n}X_{i})^2}] 
= \E\qty[ \frac{\cancel{\sum_{i=1}^{n} X_i }\sigma^2(1)}{\qty(\sum_{i=1}^{n}X_{i})^{\cancel{2}}}] 
= \E\qty[ \frac{\sigma^2(1)}{\sum_{i=1}^{n}X_{i}}] = \sigma^2(1) \times \E\qty[ \frac{1}{N_1}]. 
\end{aligned}
}
$$

## The Variance of **Differences in Means**

**Claim.** The variance of the difference in subsample means is the **sum** of the variances of the subsample means themselves.
$$ 
\Var[\hat\mu(1) - \hat\mu(0)] = \E\qty[ \frac{\sigma^2(1)}{N_1} + 
\frac{\sigma^2(0)}{N_0}] \quad \text{ for } \quad N_x = \sum_{i=1}^n 1_{x}(X_i) = \begin{cases} 1 &\text{if} X_i=x \\ 0 &\text{otherwise}\end{cases} 
$$

 **Prove it.**  This is about understanding the logic of the proof we just did. Start with the law of total variance.

$$
\Var[\hat\mu(1) - \hat\mu(0)] = \E\qty[ \Var[\hat\mu(1) - \hat\mu(0) \mid X_1 \ldots X_n] ] + \Var[\E[\hat\mu(1) - \hat\mu(0) \mid X_1 \ldots X_n] ]
$$

Here are some things to think about.

 - What is the conditional expectation of the estimator given $X_1 \ldots X_n$, i.e. $\E[\hat\mu(1) - \hat\mu(0) \mid X_1 \ldots X_n]$?  And what is its variance? 
 - $\Var[\hat\mu(1) - \hat\mu(0) \mid X_1 \ldots X_n]$ is the conditional mean $\E[Z^2 \mid X_i]$ of some random variable $Z$. 
  - What is that random variable?
  - Can we write it in a way so that, when we square it, the cross terms are zero?  
  - Hint: $X_i(1-X_i)=0$.
 
