---
title: "Lecture 4"
subtitle: "Estimating Sampling Distributions using the Bootstrap"
format:
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---

```{r setup}
library(usmap)
library(sf)

library(ggplot2)

library(scales)
library(knitr)

percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }
```

```{r ggplot-themes}
class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
        panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x  = element_text(colour = "#aaaaaa"),
        axis.text.y  = element_text(colour = "#aaaaaa"),
        axis.title.x  = element_text(colour = "#aaaaaa"),
        axis.title.y  = element_text(colour = "#aaaaaa", angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
              axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
           panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(simplified.theme)
```

```{r population}
our.seed = 0
set.seed(our.seed)
N=7233584
p=4999960/N
y= c(rep(1,p*N), rep(0, (1-p)*N))
y=sample(y) # shuffle the order 
population.proportion = p
m=N
```

```{r sample}
set.seed(our.seed)

n = 625
J = sample(N, n, replace=TRUE)
Y = y[J]
sample.proportion = mean(Y)
```

```{r mapping}
#| cache: true

# Create fake locations for the people in our population
# by taking random convex combinations of state vertices
set.seed(0)
ga.shape = us_map(include = c("GA"))

# To reduce computation, we'll have 10K locations + jitter
M = 10000 
V = nrow(ga.shape)
weights = abs(rt(M*V,1))
dim(weights) = c(M,V)
weights = (1/rowSums(weights)) * weights 
base.x = weights %*% ga.shape$x
base.y = weights %*% ga.shape$y
base.ind = 1 + ( (0:(N-1)) %% M )
locations = data.frame(x=base.x[base.ind] + rnorm(N, sd=.01),
                       y=base.y[base.ind] + rnorm(N, sd=.01))

ga = ggplot() + geom_polygon(data=ga.shape, aes(x = x, y = y), fill="white", color="black") + coord_fixed() + blank.theme
```


## Point and Interval Estimation


```{r}
#| out-height: 400px
#| fig-asp: .3 

width = function(samples, alpha=.05) {
  covg.dif = function(w) { 
    actual = mean( mean(samples) - w/2 <= samples & samples <= mean(samples) + w/2)
    nominal = 1-alpha 
    actual - nominal  
  }
  uniroot(covg.dif, interval=c(0, 2*max(abs(samples-mean(samples)))))$root
}

set.seed(our.seed)
Js = array(dim=c(10000, n))
Ys = array(dim=c(10000, n))
for(rr in 1:10000) {
  Js[rr,] = sample(N, n, replace=TRUE)
  Ys[rr,] = y[Js[rr,]]
}
meanY.samples = rowMeans(Ys)
default.height = .04
scale_x = scale_x_continuous(breaks=seq(.6,.8, by=.025), limits=c(.6, .8)) 
scale_y = scale_y_continuous(breaks=seq(0,default.height,by=.01), limits=c(0, default.height))

annotate.histogram = function(histplot, samples, shade=FALSE, 
                              shade.width=width(samples, alpha=.05), shade.color='green', shade.alpha=.15) { 
    w66 = width(samples, alpha=1/3)
    w95 = width(samples, alpha=.05)
    p = histplot + 
           geom_vline(aes(xintercept=mean(samples)), color="blue") +
           geom_vline(aes(xintercept=mean(samples)+c(-1,1)*w66/2), color="blue", linetype='dashed') +
           geom_vline(aes(xintercept=mean(samples)+c(-1,1)*w95/2), color="blue", linetype='dotted') 
    if(shade) { 
      p + geom_rect(aes(xmin=mean(samples)-shade.width/2, 
                        xmax=mean(samples)+shade.width/2, 
                        ymin=0, ymax=Inf), alpha=shade.alpha, fill=shade.color)
    } else { p }
}

annotated.bar = function(samples, ...) { 
    p = ggplot() + geom_bar(aes(x=samples, y=after_stat(prop)), alpha=.3) +
        geom_vline(aes(xintercept=mean(y)), color='green', size=3, alpha=.7)
    annotate.histogram(p, samples, ...) + scale_x + scale_y
}

default.highlight.colors = c('black', 'magenta', 'cyan', 'gray')
with.points = function(histplot, samples, K=100, highlight=4, height=default.height) {
    set.seed(our.seed)
    highlight.colors = if(highlight) default.highlight.colors[1:highlight] else c()
    heights = sample(height*(1:K)/K)
    data = data.frame(y=heights, x=samples[1:K], 
                      color=c(highlight.colors, rep('purple', K-length(highlight.colors))),
                      alpha=c(rep(1, length(highlight.colors)), rep(.1, K-length(highlight.colors))))
    histplot + geom_point(aes(y=y, x=x, color=I(color), alpha=I(alpha)), data=data)
}

with.intervals = function(histplot, samples, w=width(samples, alpha=.05), K=100, 
                          highlight=4, highlight.alpha=1, height=default.height) { 
    set.seed(our.seed)
    highlight.colors = if(highlight) default.highlight.colors[1:highlight] else c()
    heights = sample(height*(1:K)/K)
    data = data.frame(y=heights, x=samples[1:K], w=w,
                      color=c(highlight.colors, rep('purple', K-length(highlight.colors))),
                      alpha=c(rep(highlight.alpha, length(highlight.colors)), rep(.1, K-length(highlight.colors))))
    histplot + 
        geom_pointrange(aes(y=y, x=x, xmin=x-w/2, xmax=x+w/2, color=I(color), alpha=I(alpha)), data=data)
}

annotated.histogram = annotated.bar(meanY.samples)
with.intervals(annotated.histogram, meanY.samples)
```

- Last week, we talked about estimating population proportions.
- Our **point estimates** were proportions of a random sample, drawn with replacement, from the population. 
- To we'd be, we used **interval estimates**. In particular, *calibrated* interval estimates. Confidence intervals.  
  - To get these intervals, we add 'arms' of equal length to our point estimate.
  - We choose the arm length so that, 95% of the time, the population proportion is within them.

## Calibration of Interval Estimates
```{r}
#| out-height: 400px
#| fig-asp: .3
p = mean(y)
xx= seq(0,n)

w=width(meanY.samples, alpha=.05)
calibrated.width.histogram = annotated.histogram + 
    geom_rect(aes(xmin=mean(y)-w/2, xmax=mean(y)+w/2, ymin=0, ymax=Inf), alpha=.15, fill='green')
with.intervals(calibrated.width.histogram, meanY.samples)
```

- We can do that using the **sampling distribution** of our point estimate.
- We'd draw arms out from the sampling distribution's until they span 95% of point estimates.
- We can see that this is the right length by a shift of perspective.
  - A point estimate can touch the mean with its arms iff^[if and only if] the mean can touch it with equally long arms.
  - So 95% of intervals cover if and only if 95% of point estimates are within the mean's arms.
- In practice, we can't do exactly that. We don't know the sampling distribution.
- But we can use an estimate of the sampling distribution in its place. 
- If it's a good one, we'll get approximately 95% coverage.

## A Parametric Estimate of the Sampling Distribution
```{r}
#| out-height: 400px
#| fig-asp: .7
#| layout-ncol: 2 
#| fig-cap:
#| - Our point estimate and its sampling distribution
#| - Our estimate of its sampling distribution 

estimated.dist.samples = rbinom(10000, n, mean(Y)) / n 
w.est=width(estimated.dist.samples, alpha=.05)
calibrated.width.histogram.estimated = annotated.bar(estimated.dist.samples, shade=TRUE, shade.color='red')
calibrated.width.histogram.noshade   = annotated.bar(meanY.samples, shade=FALSE)

set.seed(our.seed)
K = 100
heights = sample(default.height*(1:K)/K)

p.parametric.truth = 
  with.intervals(calibrated.width.histogram.noshade, meanY.samples, w=w.est, highlight=1) +
            geom_line(aes(x=xx/n, y=dbinom(xx, n, mean(y))), color='black', alpha=.6) + coord_flip()
p.parametric.estimate = 
  with.intervals(calibrated.width.histogram.estimated, estimated.dist.samples, highlight=0) +
            geom_line(aes(x=xx/n, y=dbinom(xx, n, mean(Y))), color='black') +
            geom_point(aes(x=mean(Y), y=heights[1])) + coord_flip()
p.parametric.truth
p.parametric.estimate
```

 - It helps to know the **parametric form** of the sampling distribution: it's **binomial**.
    - It's the distribution of the proportion heads in `r n` flips of a coin with probability $p \approx `r approx(p)`$ of heads.
    - where $p \approx `r approx(p)`$ is the proportion of **registered voters** who will vote.
- By *parametric form*, I mean a formula in terms of some parameters. 
  - That tells us what parameters we need to estimate to estimate the sampling distribution.
  - In this case, the only (unknown) parameter is the probability of heads, $p$.

::: aside

- To be more precise, the binomial distribution is for the number of heads.
- And we scale that by $1/`r n`$ to get a distribution for the proportion.

:::

## A Parametric Estimate of the Sampling Distribution
```{r}
#| out-height: 400px
#| fig-asp: .7
#| layout-ncol: 2 
#| fig-cap:
#| - Our point estimate and its sampling distribution
#| - Our estimate of its sampling distribution 

p.parametric.truth
p.parametric.estimate
```

- In `R`, `dbinom(x, n, p)`, calculates this formula where ...
  - `x` is the number of heads
  - `n` is the number of flips
  - `p` is the probability of heads, i.e., the proportion $\bar y$ of registered voters who will vote.
- To estimate the sampling distribution, we plug in our point estimate $\bar Y$ for $p$.
- There's another way to think about this that generalizes to more complicated situations.


## The Bootstrap Interpretation
```{r}
#| out-height: 300px
#| fig-asp: .8 
#| layout-ncol: 3

sample.locations = locations[J, ]
boot.ind = sample(1:n, n, replace=TRUE)
resampled.locations = sample.locations[boot.ind, ]

ga + geom_point(aes(x = x, y = y), alpha=.05, data=locations[1:M,], size=.05)
ga + geom_point(aes(x = x, y = y), alpha=.2, data=sample.locations, size=1) 
ga + geom_point(aes(x = x, y = y), alpha=.2, data=resampled.locations, size=1)
```

- The estimated sampling distribution is $\text{Binomial}(n,p)/n$ for $n=`r n`$ and $p=\bar Y\approx `r approx(mean(Y))`$.
  - It's the distribution of the proportion heads in `r n` flips of a coin with probability $p \approx `r approx(mean(Y))`$ of heads.
  - where $p \approx `r approx(mean(Y))`$ is the proportion of **voters we've polled** who will vote. 
- That is, it's the sampling distribution of a 'poll' of the people in our sample, i.e.
    - roll a 625-sided die 625 times
    - call up the corresponding person in our sample
    - and counting up the yeses we hear
- Note that this is random because we're drawing with replacement.
  - Each time we run this poll, we call each person in our sample 0,1,2,... times
  - And the number of times we call them is random.



## The Bootstrap Interpretation

```{r}
#| out-height: 300px
#| fig-asp: .8
#| layout-ncol: 3
ga + geom_point(aes(x = x, y = y), alpha=.05, data=locations[1:M,], size=.05)
ga + geom_point(aes(x = x, y = y), alpha=.2, data=sample.locations, size=1) 
ga + geom_point(aes(x = x, y = y), alpha=.2, data=resampled.locations, size=1)
```

- If we plot our voters on a map, you can see the idea in visual terms.
    - On the left, we have the population. 
    - In the middle, we have our sample.
    - On the right, we have locations drawn, with replacement, from the sample.
- Each 'call' is one transparent dot, so we see darker dots ...
    - or actually multiple copies of the same dot overlaid
    - where we've drawn someone multiple times.
- 2x darker means we've called them twice, 3x darker means we've called them three times, etc.

## Bootstrapping

- Before the election, we don't observe the population. But we do observe the sample. 
  - And we can *sample from the sample* act as if it were the population.
  - We'll take repeated random samples of size `r n`, with replacement, from our sample of size `r n`.
- We call these *bootstrap samples* and estimates based on them *bootstrap estimates*.
  - The distribution of these estimates is called the *bootstrap sampling distribution*.
  - If the sample is like the population, this should be like our estimator's actual sampling distribution.

```{r bootstrap.sample}
J.star = sample(n, size=n, replace=TRUE)
Y.star = Y[J.star]
```

:::: columns
::: column
The Sample
$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}  \\
Y_i  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$

The Bootstrap Sample

$$
\begin{array}{r|rrrr|r}
i    &   1 & 2 & \dots & `r n` & \bar{Y}_{`r n`}^*  \\
Y_i^*  &  `r Y.star[1]` & `r Y.star[2]` & \dots & `r Y.star[n]` & `r approx(mean(Y.star))` \\
\end{array}
$$
:::

::: column
The Population

$$
\begin{array}{r|rrrr|r}
j & 1 & 2 & \dots &   `r approx(N)` & \bar{y}_{`r approx(N)`}  \\
y_{j} & `r y[1]` & `r y[2]` & \dots & `r y[N]`   & `r approx(mean(y))`  \\
\end{array}
$$

The 'Bootstrap Population' --- The Sample
$$
\begin{array}{r|rrrr|r}
j        &   1 & 2 & \dots & `r n` & \bar{y}^*_{`r n`}  \\
y_j^*  &  `r Y[1]` & `r Y[2]` & \dots & `r Y[n]` & `r approx(mean(Y))` \\
\end{array}
$$

:::
::::

::: aside
Notation. We use stars to distinguish the bootstrap sample from our original sample---we write $Y_i^*$ and $\bar Y^*$.
:::

## The Bootstrap is *Nonparametric* 

```{r}
#| echo: true

bootstrap.samples = array(dim=10000)
for(rr in 1:10000) {
    Y.boot = Y[sample(1:n, n, replace=TRUE)]
    bootstrap.samples[rr] = sum(Y.boot)/n 
}
```

```{r}
#| out-height: 400px 
#| fig-asp: .3 
ggplot() + geom_bar(aes(x=bootstrap.samples, y=after_stat(prop)), alpha=.3, position='identity') + 
           geom_line(aes(x=(0:n)/n, y=dbinom(0:n, n, mean(Y))), color='black') + 
           xlim(.6,.8)
```

- We do not need to know the parametric form of our sampling distribution to use it.
- All we do is re-run our poll *acting as if* our sample were the population.
  - We can do this no matter what we're estimating. 
  - So let's try it out on some other stuff.
  
## Comparing Black and Non-Black Turnout

```{r population-and-sample-black-vs-nonblack}
#| echo: false 
our.seed = 0
set.seed(4)
N = 7233584
n = 625

b0 = floor(.7*N)
b1 = ceiling(.3*N)
y0 = rbinom(n=b0,size=1,prob=.68)
y1 = rbinom(n=b1,size=1,prob=.74)
dfpop = data.frame(y=c(y0,y1), x=c(rep(0,b0),rep(1,b1)))
dfpop = dfpop[sample(1:N),] # shuffle the rows
x = dfpop$x
y = dfpop$y


# Generate Sample Data for us (J1) and three friends (J2,J3,JM)
set.seed(our.seed)
J1 = sample(1:N,size=n,replace=T)
J2 = sample(1:N,size=n,replace=T)
J3 = sample(1:N,size=n,replace=T)
JM = sample(1:N,size=n,replace=T)

J = J1
dfsam = dfpop[J,]
Y = dfsam$y
X = dfsam$x
```

$$
\small{
\begin{array}{r|rr|rr|r|rr|rrrr}
\text{call} &  1  & & 2  & & \dots & `r n`  & & & & & & \\
\text{question}  & X_1 & Y_1 &  X_2 & Y_2 & \dots &  X_{`r n`} & Y_{`r n`} & \overline{X}_{`r n`} & \overline{Y}_{`r n`} &\frac{\sum_{i:X_i=1} Y_i}{\sum_{i:X_i=1} 1} & \frac{\sum_{i:X_i=0} Y_i}{\sum_{i:X_i=0} 1} \\
\text{outcome}    &  \underset{\textcolor{gray}{x_{`r J[1]`}}}{`r X[1]`} & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & \underset{\textcolor{gray}{x_{`r J[2]`}}}{`r X[2]`} & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`}
                  & \dots 
                  & \underset{\textcolor{gray}{x_{`r J[n]`}}}{`r X[n]`} & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`}
                  & `r approx(mean(X))` & `r approx(mean(Y))` & `r approx(mean(Y[X==1]))` & `r approx(mean(Y[X==0]))` \\
\end{array}
}
$$


- It's useful to predict turnout among Black voters because they tend to vote differently than non-Black voters.
- Let's suppose that we're interested in the difference in turnout between Black and non-Black voters.
- This is a bit reductive, but it's the beginning of the semester, so we're keeping things simple. 
- We know from the voter file that `r percent(mean(x==1))` of registered voters are Black.
- Let's suppose we asked the people we polled if they were Black, recording their answer in a *covariate* $X_i$.

## Comparing Black and Non-Black Turnout

$$
\small{
\begin{array}{r|rr|rr|r|rr|rrrr}
\text{call} &  1  & & 2  & & \dots & `r n`  & & & & & & \\
\text{question}  & X_1 & Y_1 &  X_2 & Y_2 & \dots &  X_{`r n`} & Y_{`r n`} & \overline{X}_{`r n`} & \overline{Y}_{`r n`} &\frac{\sum_{i:X_i=1} Y_i}{\sum_{i:X_i=1} 1} & \frac{\sum_{i:X_i=0} Y_i}{\sum_{i:X_i=0} 1} \\
\text{outcome}    &  \underset{\textcolor{gray}{x_{`r J[1]`}}}{`r X[1]`} & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & \underset{\textcolor{gray}{x_{`r J[2]`}}}{`r X[2]`} & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`}
                  & \dots 
                  & \underset{\textcolor{gray}{x_{`r J[n]`}}}{`r X[n]`} & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`}
                  & `r approx(mean(X))` & `r approx(mean(Y))` & `r approx(mean(Y[X==1]))` & `r approx(mean(Y[X==0]))` \\
\end{array}
}
$$

- In our poll, we found that...
  - `r sum(X==1)` of the people we called (`r percent(mean(X==1))`) were Black with a turnout rate of `r approx(mean(Y[X==1]))`
  - `r sum(X==0)` of the people we called (`r percent(mean(X==0))`) were non-Black with a turnout rate of `r approx(mean(Y[X==0]))`   
- After the election, we found that ...
  - Turnout among Black registered voters was `r approx(mean(y[x==1]))` 
  - Turnout among non-Black registered voters was `r approx(mean(y[x==0]))`.
- We'd estimate the *difference* to be $`r approx(mean(Y[X==1]))`-`r approx(mean(Y[X==0]))` \approx  `r approx(mean(Y[X==1])- mean(Y[X==0]))`$. 
- The actual difference was $`r approx(mean(y[x==1]))`-`r approx(mean(y[x==0]))` \approx  `r approx(mean(y[x==1]) - mean(y[x==0]))`$.
- Depending on what you're doing, that point estimate might be good enough or might not be.
- It'd be nice to have an interval estimate.

## A Table of Imaginary Polls

$$
\small{
\begin{array}{r|rr|rr|r|rr|rrrr}
\text{call} &  1 & & 2 & &  \dots & `r n` & & & & & & \\
\text{pollster} &  X_1 & Y_1 &  X_2 & Y_2   & \dots &  X_{`r n`} & Y_{`r n`}  & \overline{X} & \overline{Y}  &\frac{\sum_{i:X_i=1} Y_i}{\sum_{i:X_i=1} 1} & \frac{\sum_{i:X_i=0} Y_i}{\sum_{i:X_i=0} 1} & \text{difference} \\
\hline
\color{black}   1   
    & \color{black}   `r J=J1; Y=y[J]; X=x[J]; X[1]`      & \color{black} `r Y[1]` 
    &                            \color{black}  `r X[2]`      & \color{black} `r Y[2]`      
                                                        & \color{black}   \dots 
    &                            \color{black}   `r X[n]`     & \color{black} `r Y[n]`       
    & \color{black}   `r approx(mean(X))`   & \color{black} `r approx(mean(Y))`   &  \color{black}  `r approx(mean(Y[X==1]))` & `r approx(mean(Y[X==0]))` & \color{black} `r approx(mean(Y[X==1]) - mean(Y[X==0]))` \\
\color{magenta}   2   
    & \color{magenta}   `r J=J2; Y=y[J]; X=x[J]; X[1]`      & \color{magenta} `r Y[1]` 
    & \color{magenta}   `r X[2]`      & \color{magenta} `r Y[2]`      
                                                        & \color{magenta}   \dots 
    & \color{magenta}   `r X[n]`     & \color{magenta} `r Y[n]`       
    & \color{magenta}   `r approx(mean(X))`   & \color{magenta} `r approx(mean(Y))`   &  \color{magenta}  `r approx(mean(Y[X==1]))` & \color{magenta} `r approx(mean(Y[X==0]))` &  \color{magenta} `r approx(mean(Y[X==1]) - mean(Y[X==0]))`  \\
\color{cyan}    3   
    & \color{cyan}   `r J=J3; Y=y[J]; X=x[J]; X[1]`      & \color{cyan} `r Y[1]`
    & \color{cyan}   `r X[2]`      & \color{cyan} `r Y[2]`      
                                                        & \color{cyan}   \dots
    & \color{cyan}   `r X[n]`     & \color{cyan} `r Y[n]`
    & \color{cyan}   `r approx(mean(X))`   & \color{cyan} `r approx(mean(Y))`   &  \color{cyan}  `r approx(mean(Y[X==1]))` & \color{cyan} `r approx(mean(Y[X==0]))` & \color{cyan} `r approx(mean(Y[X==1]) - mean(Y[X==0]))` \\
\vdots & \vdots & \vdots &  \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{gray}    1M   
    & \color{gray}   `r J=JM; Y=y[J]; X=x[J]; X[1]`      & \color{gray} `r Y[1]`
    & \color{gray}   `r X[2]`      & \color{gray} `r Y[2]`      
                                                        & \color{gray}   \dots
    & \color{gray}   `r X[n]`     & \color{gray} `r Y[n]`
    & \color{gray}   `r approx(mean(X))`   & \color{gray} `r approx(mean(Y))`   &  \color{gray}  `r approx(mean(Y[X==1]))` & \color{gray} `r approx(mean(Y[X==0]))` & \color{gray} `r approx(mean(Y[X==1]) - mean(Y[X==0]))` \\
\end{array}
}
$$


## The Sampling Distribution 

```{r}
#| echo: true

set.seed(our.seed)
difference.samples = array(dim=10000)
difference.sd.samples = array(dim=10000)
for(rr in 1:10000) {
    I = sample(1:m, n, replace=TRUE)
    X = x[I]
    Y = y[I]
    difference.samples[rr] = mean(Y[X==1]) - mean(Y[X==0])
    p1 = mean(Y[X==1])
    p0 = mean(Y[X==0])
    difference.sd.samples[rr] = sqrt(p1*(1-p1)/sum(X==1) + p0*(1-p0)/sum(X==0))
}
```

```{r}
#| out-height: 400px 
#| fig-asp: .3 
w=width(difference.samples, alpha=.05)
difference.histogram=ggplot() + geom_histogram(aes(x=difference.samples, y=after_stat(density)), bins=100, alpha=.3)
annotated.difference.histogram = annotate.histogram(difference.histogram, difference.samples, 
  shade=TRUE, shade.color='green') 
with.intervals(annotated.difference.histogram, difference.samples, height=10)
```

- Here's what the **sampling distribution** of this difference in turnout looks like.
- As before, if we can estimate it we can use that estimate to get a 95\% confidence interval.
- But, unlike before, we don't really know the parametric form of its sampling distribution.
- Or---at least---it'd be a pain to work it out. So we'll use the bootstrap to estimate it.

## A Table of Bootstrap Polls

```{r}
#| out-height: 300px
#| fig-asp: .5
#| layout-ncol: 2
#| fig-cap:
#|  - Our Sample
#|  - Three Bootstrap Samples---The First Person 'Polled'
sample.locations$label = 1:n
set.seed(our.seed)
boot.ind1 = sample(1:n, n, replace=TRUE)
boot.ind2 = sample(1:n, n, replace=TRUE)
boot.ind3 = sample(1:n, n, replace=TRUE)
bootstrap.locations = 
  rbind(sample.locations[boot.ind,], 
        sample.locations[boot.ind2,], 
        sample.locations[boot.ind3,])
bootstrap.locations$color=rep(c('magenta','cyan', 'gray'), each=n)
bootstrap.locations$i=rep(1:n, 3)

ga + geom_text(aes(x = x, y = y, label=label), alpha=.5, size=3, data=sample.locations) 
ga + geom_text(aes(x = x, y = y, label=label, color=I(color)), alpha=.5, size=3, data=bootstrap.locations[bootstrap.locations$i == 1, ])
```


$$
\small{
\begin{array}{r|rr|rr|r|rr|rrrr}
\text{call} &  1 & & 2 & & \dots & `r n` & & & & & & \\
\text{pollster} &  X_1 & Y_1 &  X_2 & Y_2   & \dots &  X_{`r n`} & Y_{`r n`}  & \overline{X} & \overline{Y}  &\frac{\sum_{i:X_i=1} Y_i}{\sum_{i:X_i=1} 1} & \frac{\sum_{i:X_i=0} Y_i}{\sum_{i:X_i=0} 1} & \text{difference} \\
\hline
\color{black}   1   
    & \color{black}   `r J=J1; Y=y[J]; X=x[J]; X[1]`      & \color{black} `r Y[1]` 
    &                            \color{black}  `r X[2]`      & \color{black} `r Y[2]`      
                                                        & \color{black}   \dots 
    &                            \color{black}   `r X[n]`     & \color{black} `r Y[n]`       
    & \color{black}   `r approx(mean(X))`   & \color{black} `r approx(mean(Y))`   &  \color{black}  `r approx(mean(Y[X==1]))` & `r approx(mean(Y[X==0]))` & \color{black} `r approx(mean(Y[X==1]) - mean(Y[X==0]))` \\
\end{array}
}
$$

$$
\small{
\begin{array}{r|rr|rr|r|rr|rrrr}
\text{`call'} &  1 & &  2 & &  \dots & `r n` & & & & & & \\
\text{`pollster'} &  X_1^* & Y_1^* &  X_2^* & Y_2^*   & \dots &  X^*_{`r n`} & Y^*_{`r n`}  & \overline{X}^* & \overline{Y}^*  &\frac{\sum_{i:X_i^*=1} Y_i}{\sum_{i:X_i^*=1} 1} & \frac{\sum_{i:X_i^*=0} Y_i}{\sum_{i:X_i^*=0} 1} & \text{difference} \\
\hline
\color{magenta}   `r I=boot.ind1; 1`
    & \color{magenta}   X_{`r I[1]`}  & \color{magenta} Y_{`r I[1]`} 
    & &
    & \color{magenta}   \dots
    & &
    & & & & & & \\
    & \color{magenta}   `r X[I[1]]`   & \color{magenta} `r Y[I[1]]`
    & &
    & \color{magenta}   \dots
    & &
    & & & & & & \\
\color{cyan}   `r I=boot.ind2; 2`
    & \color{cyan}   X_{`r I[1]`}  & \color{cyan} Y_{`r I[1]`} 
    & &
    & \color{cyan}   \dots
    & &
    & & & & & & \\
    & \color{cyan}   `r X[I[1]]`   & \color{cyan} `r Y[I[1]]`
    & &
    & \color{cyan}   \dots
    & &
    & & & & & & \\
\color{gray}   `r I=boot.ind3; 3`
    & \color{gray}   X_{`r I[1]`}  & \color{gray} Y_{`r I[1]`} 
    & &
    & \color{gray}   \dots
    & &
    & & & & & & \\
    & \color{gray}   `r X[I[1]]`   & \color{gray} `r Y[I[1]]`
    & &
    & \color{gray}   \dots
    & &
    & & & & & & \\
\end{array}
}
$$


## A Table of Bootstrap Polls

```{r}
#| out-height: 300px
#| fig-asp: .5
#| layout-ncol: 2
#| fig-cap:
#|  - Our Sample
#|  - Three Bootstrap Samples

ga + geom_text(aes(x = x, y = y, label=label), alpha=.5, size=3, data=sample.locations) 
ga + geom_text(aes(x = x, y = y, label=label, color=I(color)), alpha=.5, size=3, data=bootstrap.locations)
```


$$
\small{
\begin{array}{r|rr|rr|r|rr|rrrr}
\text{call} &  1 & & 2 & & \dots & `r n` & & & & & & \\
\text{pollster} &  X_1 & Y_1 &  X_2 & Y_2   & \dots &  X_{`r n`} & Y_{`r n`}  & \overline{X} & \overline{Y}  &\frac{\sum_{i:X_i=1} Y_i}{\sum_{i:X_i=1} 1} & \frac{\sum_{i:X_i=0} Y_i}{\sum_{i:X_i=0} 1} & \text{difference} \\
\hline
\color{black}   1   
    & \color{black}   `r J=J1; Y=y[J]; X=x[J]; X[1]`      & \color{black} `r Y[1]` 
    &                            \color{black}  `r X[2]`      & \color{black} `r Y[2]`      
                                                        & \color{black}   \dots 
    &                            \color{black}   `r X[n]`     & \color{black} `r Y[n]`       
    & \color{black}   `r approx(mean(X))`   & \color{black} `r approx(mean(Y))`   &  \color{black}  `r approx(mean(Y[X==1]))` & `r approx(mean(Y[X==0]))` & \color{black} `r approx(mean(Y[X==1]) - mean(Y[X==0]))` \\
\end{array}
}
$$

$$
\small{
\begin{array}{r|rr|rr|r|rr|rrrr}
\text{`call'} &  1 & &  2 & &  \dots & `r n` & & & & & & \\
\text{`pollster'} &  X_1^* & Y_1^* &  X_2^* & Y_2^*   & \dots &  X^*_{`r n`} & Y^*_{`r n`}  & \overline{X}^* & \overline{Y}^*  &\frac{\sum_{i:X_i^*=1} Y_i}{\sum_{i:X_i^*=1} 1} & \frac{\sum_{i:X_i^*=0} Y_i}{\sum_{i:X_i^*=0} 1} & \text{difference} \\
\hline
\color{magenta}   `r I=boot.ind1; 2`
    & \color{magenta}   X_{`r I[1]`}  & \color{magenta} Y_{`r I[1]`} 
    & \color{magenta}   X_{`r I[2]`}  & \color{magenta} Y_{`r I[2]`}      
    & \color{magenta}   \dots
    & \color{magenta}   X_{`r I[n]`}  & \color{magenta} Y_{`r I[n]`} 
    & & & & & & \\
    & \color{magenta}   `r X[I[1]]`   & \color{magenta} `r Y[I[1]]`
    & \color{magenta}   `r X[I[2]]`      & \color{magenta} `r Y[I[2]]`      
    & \color{magenta}   \dots
    & \color{magenta}   `r X[I[n]]`     & \color{magenta} `r Y[I[n]]` 
    & \color{magenta}   `r approx(mean(X[I]))`   & \color{magenta} `r approx(mean(Y[I]))`   
    &  \color{magenta}  `r approx(mean(Y[I][X[I]==1]))` & \color{magenta} `r approx(mean(Y[I][X[I]==0]))` 
    & \color{magenta} `r approx(mean(Y[I][X[I]==1]) - mean(Y[I][X[I]==0]))` \\
\color{cyan}   `r I=boot.ind2; 3`
    & \color{cyan}   X_{`r I[1]`}  & \color{cyan} Y_{`r I[1]`} 
    & \color{cyan}   X_{`r I[2]`}  & \color{cyan} Y_{`r I[2]`}      
    & \color{cyan}   \dots
    & \color{cyan}   X_{`r I[n]`}  & \color{cyan} Y_{`r I[n]`} 
    & & & & & & \\
    & \color{cyan}   `r X[I[1]]`   & \color{cyan} `r Y[I[1]]`
    & \color{cyan}   `r X[I[2]]`      & \color{cyan} `r Y[I[2]]`      
    & \color{cyan}   \dots
    & \color{cyan}   `r X[I[n]]`     & \color{cyan} `r Y[I[n]]` 
    & \color{cyan}   `r approx(mean(X[I]))`   & \color{cyan} `r approx(mean(Y[I]))`   
    &  \color{cyan}  `r approx(mean(Y[I][X[I]==1]))` & \color{cyan} `r approx(mean(Y[I][X[I]==0]))` 
    & \color{cyan} `r approx(mean(Y[I][X[I]==1]) - mean(Y[I][X[I]==0]))` \\
\color{gray}   `r I=boot.ind3; 1`
    & \color{gray}   X_{`r I[1]`}  & \color{gray} Y_{`r I[1]`} 
    & \color{gray}   X_{`r I[2]`}  & \color{gray} Y_{`r I[2]`}      
    & \color{gray}   \dots
    & \color{gray}   X_{`r I[n]`}  & \color{gray} Y_{`r I[n]`} 
    & & & & & & \\
    & \color{gray}   `r X[I[1]]`   & \color{gray} `r Y[I[1]]`
    & \color{gray}   `r X[I[2]]`      & \color{gray} `r Y[I[2]]`      
    & \color{gray}   \dots
    & \color{gray}   `r X[I[n]]`     & \color{gray} `r Y[I[n]]` 
    & \color{gray}   `r approx(mean(X[I]))`   & \color{gray} `r approx(mean(Y[I]))`   
    &  \color{gray}  `r approx(mean(Y[I][X[I]==1]))` & \color{gray} `r approx(mean(Y[I][X[I]==0]))` 
    & \color{gray} `r approx(mean(Y[I][X[I]==1]) - mean(Y[I][X[I]==0]))` \\
\end{array}
}
$$


## The Difference's Bootstrap Sampling Distribution 

```{r}
#| echo: true

difference.bootstrap.samples = array(dim=10000)
for(rr in 1:10000) {
    I = sample(1:n, n, replace=TRUE)
    Xstar = X[I]
    Ystar = Y[I]
    difference.bootstrap.samples[rr] = mean(Ystar[Xstar==1]) - mean(Ystar[Xstar==0])
}
```

```{r}
#| out-height: 300px 
#| fig-asp: .5
#| layout-ncol: 2 

scale_x_dif = scale_x_continuous(limits=c(-.1,.2))
wboot=width(difference.bootstrap.samples, alpha=.05)
annotated.difference.histogram.noshade = annotate.histogram(difference.histogram, difference.samples, shade=FALSE) 

pboot=ggplot() + geom_histogram(aes(x=difference.bootstrap.samples, y=after_stat(density)), alpha=.3, bins=100) 
annotated.bootstrap.difference.histogram = annotate.histogram(pboot, difference.bootstrap.samples,
  shade=TRUE, shade.color='red')

with.points(annotated.difference.histogram.noshade, difference.samples, height=10, highlight=0) + 
  geom_pointrange(aes(x=x,y=y, xmin=x-w/2, xmax=x+w/2), 
                  data=data.frame(x=difference.samples[1], y=5, w=wboot), color='black') + 
                  coord_flip() + scale_x_dif

with.points(annotated.bootstrap.difference.histogram, 
            c(difference.samples[1], difference.bootstrap.samples), 
            height=10, highlight=4) + coord_flip() + scale_x_dif
```

- Not bad. Our bootstrap sampling distribution looks a lot like our actual sampling distribution.
- And, in particular, the intervals widths we calculate are very close: [`r round(w,3)`]{.fg style='color:limegreen'} vs. [`r round(wboot,3)`]{.fg style='color:red'}.
- And that's really all that matters. **Q**. Why? 

::: fragment

- Because that's what we use to calibrate our interval estimate.
  - It's our point estimate --- the **black dot** 
  - With arms wide enough to span 95% of the bootstrap sampling distribution --- the [**red-shaded interval**]{.fg style="color:red"}.

:::

# Why Does The Bootstrap Work?

## Our Binomial-Specific Argument 
```{r}
#| out-height: 300px
#| fig-asp: .6 
#| layout-ncol: 2
#| 
K = 3
ps = quantile(meanY.samples, c(.025, .5, .975))
p = population.proportion

density.stability.plot = ggplot() + 
           geom_line(aes(x=x/n, y=dbinom(x,n,p)), data=expand.grid(p=p, x=0:n), alpha=.5) +
           geom_vline(aes(xintercept=qbinom(c(.025,.975), n,mean(y))/n), linetype='dotted') +
           geom_line(aes(x=x/n, y=dbinom(x,n,p), color=factor(p)), data=expand.grid(p=ps, x=0:n), alpha=.5) +
           geom_pointrange(aes(x=p, xmin=qbinom(.025,n,p)/n, xmax=qbinom(.975,n,p)/n, y=.6*default.height*k/K, color=factor(p)), data=data.frame(p=ps, k=1:K)) +
           xlim(.6,.8) + guides(color=FALSE) + geom_vline(xintercept=mean(y), color='green', size=2, alpha=.5)
centered.density.stability.plot = ggplot() + 
           geom_line(aes(x=x/n-p, y=dbinom(x,n,p)), data=expand.grid(p=mean(y), x=0:n), alpha=.5) +
           geom_vline(aes(xintercept=qbinom(c(.025,.975), n,mean(y))/n-mean(y)), linetype='dotted') +
           geom_line(aes(x=x/n-p, y=dbinom(x,n,p), color=factor(p)), data=expand.grid(p=ps, x=0:n), alpha=.5) +
           geom_pointrange(aes(x=0, xmin=qbinom(.025,n,p)/n-p, xmax=qbinom(.975,n,p)/n-p, y=.6*default.height*k/K, color=factor(p)), data=data.frame(p=ps, k=1:K)) +
           xlim(-.1,.1) + guides(color=FALSE)

density.stability.plot
centered.density.stability.plot
```

- Last class, we argued that we could estimate the sampling distribution of a proportion well enough.
- Our argument used the parametric form of the sampling distribution: the binomial distribution.
  - The width of the binomial probability distribution doesn't change quickly as the probability of heads changes. 
  - So if we have an ok estimate of the probability of heads,      i.e. the sample proportion $\bar Y$ ...
  - then we can use that instead of the actual probability of heads, i.e. the population proportion $\bar y$.

$$
\bar Y \approx \bar y \implies \text{Binomial}(n,\bar Y) \approx \text{Binomial}(n,\bar y)
$$

- This told us the bootstrap worked because the bootstrap was doing exactly this.
- That is, the bootstrap sampling distribution *is* Binomial with heads probability $\bar Y$.
- But the bootstrap still works. Why? Ultimately, it's a subtle question with a lot of different answers.
- I'm going to give you a simple one. One that's very similar to the one we used last class.
- Instead of *equivalance* to a parametric estimate, we'll argue using *approximation* by one.

## The Same Idea With Normal Approximation
```{r}
#| out-height: 300px
#| fig-asp: .6 
#| layout-ncol: 2

 
K = 3
ps = quantile(meanY.samples, c(.025, .5, .975))
dbinomapprox = function(x,n,p) { 
  dnorm(x+.5, n, sqrt(p*(1-p))) - dnorm(x-.5, n, sqrt(n*p*(1-p)))
}
qbinomapprox = function(x,n,p) {
  qnorm(x,n*p, sqrt(n*p*(1-p)))
}

density.stability.plot.approx = ggplot() + 
           geom_line(aes(x=x/n, y=dbinom(x,n,p)), data=expand.grid(p=p, x=0:n), alpha=.5) +
           geom_vline(aes(xintercept=qbinom(c(.025,.975), n,mean(y))/n), linetype='dotted') +
           geom_line(aes(x=x/n, y=dbinomapprox(x,n,p), color=factor(p)), data=expand.grid(p=ps, x=0:n), alpha=.5) +
           geom_pointrange(aes(x=p, xmin=qbinomapprox(.025,n,p)/n, xmax=qbinom(.975,n,p)/n, y=.6*default.height*k/K, color=factor(p)), data=data.frame(p=ps, k=1:K)) +
           xlim(.6,.8) + guides(color=FALSE) + geom_vline(xintercept=mean(y), color='green', size=2, alpha=.5)
centered.density.stability.plot.approx = ggplot() + 
           geom_line(aes(x=x/n-p, y=dbinom(x,n,p)), data=expand.grid(p=mean(y), x=0:n), alpha=.5) +
           geom_vline(aes(xintercept=qbinom(c(.025,.975), n,mean(y))/n-mean(y)), linetype='dotted') +
           geom_line(aes(x=x/n-p, y=dbinomapprox(x,n,p), color=factor(p)), data=expand.grid(p=ps, x=0:n), alpha=.5) +
           geom_pointrange(aes(x=0, xmin=qbinomapprox(.025,n,p)/n-p, xmax=qbinom(.975,n,p)/n-p, y=.6*default.height*k/K, color=factor(p)), data=data.frame(p=ps, k=1:K)) +
           xlim(-.1,.1) + guides(color=FALSE)

density.stability.plot
centered.density.stability.plot
```

- Here I've drawn the same diagram as before, but with the binomial distribution replaced by a **normal approximation** to it.
$$
f_{\mu,\sigma}(x) = \frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{(x-\mu)^2}{2\sigma^2}} \qfor \mu=p \qand \sigma=\sqrt{p(1-p)/n}
$$

- This is a good approximation. I can't tell the difference between this diagram and the last one.
- And we can repeat our argument using the normal distribution's parametric form.
  - The width of the normal distribution doesn't change too fast as its standard deviation $\sigma$ does.
  - So if we have a good estimate of that standard deviation, e.g. $\sqrt{p(1-p)/n}$ where $p$ is the sample proportion $\bar Y$ ...
  - then we can use that instead of the actual standard deviation, which is the same where $p$ is the population proportion $\bar y$.
- This argument generalizes well. It works, e.g., for the difference in proportions.
- It works whenever the actual and boostrap sampling distributions ...
  - are both approximately normal
  - and have similar standard deviations.

## Approximate Normality for the Difference in Turnout

```{r}
#| out-height: 300px 
#| fig-asp: .6
#| layout-ncol: 2 
#| fig-cap:
#| - A histogram of draws from the actual sampling distribution and its normal approximation.
#| - A histogram of draws from the bootstrap sampling distribution and its normal approximation. 
 
xx = seq(-.2,.3, length=1000)
annotated.difference.histogram + 
  geom_line(aes(x=xx, y=dnorm(xx, mean(difference.samples), sd(difference.samples))), color='black') +
  coord_flip() + scale_x_dif

annotated.bootstrap.difference.histogram +
  geom_line(aes(x=xx, y=dnorm(xx, mean(difference.bootstrap.samples), sd(difference.bootstrap.samples))), color='black') + 
  coord_flip() + scale_x_dif
```

- The actual sampling distribution of the difference in turnout is approximately normal. 
- So is the bootstrap sampling distribution. And they do have similar standard deviations.
- Most things involving averages are approximately normal. This is a consequence of the **Central Limit Theorem**.
- We won't prove that, so you'll have to take my word for it.
  - There's a very nice proof that doesn't really require any background you don't have.
  - It uses an argument called *Stein's method* that's very interesting and very widely applicable.
  - But it'd take us a couple of lectures and we don't have time for that.

## The Same Idea for the Difference in Turnout

```{r}
#| out-height: 300px
#| fig-asp: .3

 
difs   = quantile(difference.samples, c(.025, .5, .975))
sigmas = quantile(difference.sd.samples, c(.025, .5, .975))
xgrid = seq(-.2,.3, length=1000)
height = dnorm(0,0,min(sigmas))

density.stability.plot.dif = ggplot() + 
           geom_line(aes(x=x, y=dnorm(x,d,s), color=factor(d), linetype=factor(s)), 
                     data=expand.grid(d=difs, s=sigmas, x=xgrid), alpha=.5) +
           geom_pointrange(aes(x=d, xmin=qnorm(.025,d,s), xmax=qnorm(.975,d,s), 
                            y=.6*height*k/9, color=factor(d), linetype=factor(s)), 
                            data=cbind(expand.grid(s=sigmas, d=difs), data.frame(k=1:9))) +
           xlim(min(xgrid), max(xgrid)) + guides(color=FALSE, linetype=FALSE) + 
           geom_vline(xintercept=mean(y[x==1])-mean(y[x==0]), color='green', size=2, alpha=.5)

density.stability.plot.dif
```

- We can use almost the same picture to argue that the bootstrap works for the difference in turnout.
- But there is a difference. 
  - The spread of a proportion's sampling distribution is determined by that proportion.
  - But the spread of the difference isn't determined by the difference. 
  - It's determined by the two proportions we're comparing.
- This means we can't summarize the range of interval estimates we tend to get as we did before.
- i.e., by drawing a single interval around the [2.5th]{.fg style="color:red"}, [50th]{.fg style="color:green"}, and [97.5th]{.fg style="color:blue"} percentiles of our sampling distribution.
- Instead, we need to draw intervals at a range of widths. 
- Above, I've drawn intervals three intervals around each point estimate.
- One each at the 2.5th, 50th, and 97.5th percentiles of the sampling distribution of interval estimate widths.

## Summary

```{r}
#| out-height: 300px
#| fig-asp: .3


density.stability.plot.dif
```

- Most of the time, the bootstrap sampling distribution is a good approximation of the actual sampling distribution.
- This means we have a general procedure we can use to calibrating interval estimates: we use the bootstrap.
- The downside is that it's a bit opaque. 
  - We can use it to approximate our point estimate's sampling distribution *after* we've run our experiment.
  - But that doesn't tell us what it will look like *before* we've run it.
- That matters when we're designing experiments. It's important to know what we will and won't be able to learn.
  - e.g., it's important to make our experiments big enough that we get the precision we need.
  - This is called *power analysis* and it's a big part of experimental design.
- Next class, we'll talk about how to use normal approximation to do that.