---
title: "QTM 220: Lecture 12"
subtitle: "Causal Inference with a Multivalued Treatment"
format: 
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---


```{r load-libraries}
library(purrr)
library(tibble)
library(dplyr)
library(magrittr)
library(stringr)
library(ggplot2)
library(htmltools)
library(kableExtra)
library(scales)

percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }
```

```{r ggplot-themes}
class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
        panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x  = element_text(colour = "#aaaaaa"),
        axis.text.y  = element_text(colour = "#aaaaaa"),
        axis.title.x  = element_text(colour = "#aaaaaa"),
        axis.title.y  = element_text(colour = "#aaaaaa", angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
              axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
           panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(simplified.theme)
```

```{r table-styling}
ellipsis <- function(df, nhead=5L, ntail=2) {
    rownames(df) = as.character(1:nrow(df))
    stopifnot("data.frame" %in% class(df))
    els <- rep(NA, ncol(df)) %>% 
        matrix(nrow=1, dimnames=list(NULL, names(df))) %>% 
        data.frame(stringsAsFactors=FALSE) %>% 
        tibble::as_tibble()
    rownames(els) = "⋮"
    dplyr::bind_rows(head(df, nhead), els, tail(df, ntail))
}
```


## Today

::: {.hidden}
$$
\newcommand{\alert}[1]{\textcolor{red}{#1}}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\V}{V}
\DeclareMathOperator{\Var}{\V}
\newcommand{\inred}[1]{\textcolor{red}{#1}}
\newcommand{\inblue}[1]{\textcolor{blue}{#1}}
$$ 
:::

- We're going to talk about causal inference with a multivalued treatment.
- Our context will be a randomized experiment with three treatments.
- A large employer is experimenting with a new benefit for their employees: subsized gym memberships.
- They're randomly selecting employees to participate in a trial program.
- Then randomizing these participants into three plans.
  - Plan 1: No gym membership. It costs them \$0/month.
  - Plan 2: A gym membership that covers entry but not classes. It costs \$50/month.
  - Plan 3: A gym membership that covers entry but and classes. It costs \$100/month.
- And surveying them to see how often they go to the gym, e.g. in days/month.
- They'll be able to use this information to decide what plan to offer everyone next year.

## How We Think

- We'll think about this using potential outcomes.
- We imagine, for each employee, that there's a number of times they'd go to the gym given each plan.
  - $y_j(0)$: The number of days/month they'd go to the gym if they had no membership.
  - $y_j(50)$: The number of days/month they'd go to the gym if they had a \$50 membership.
  - $y_j(100)$: The number of days/month they'd go to the gym if they had a \$100 membership.
- And we'll think about the effect of each plan on each employee.
  - $\tau_j(50)  = y_j(50)  - y_j(0)$: The effect of a \$50 gym membership vs none.
  - $\tau_j(100) = y_j(100) - y_j(50)$: The effect of a \$100 gym membership vs \$50.

$$
\small{
\begin{array}{c|ccc|cc}
j & y_j(0) & y_j(50) & y_j(100) & \tau_j(50) & \tau_j(100) \\
1 & 5  & 10 & 10 & 5 & 0 \\
\color{blue}2 & 10 & 10 & 10 & 0 & 0 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{green}327 & 0 & 0  &  20 & 0 & 20 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{purple}358 & 15 & 18 & 20 & 3 & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
m & 10 & 10 & 5  & 0 & -5 \\
\end{array}
}
$$

## What We Want To Know

- What we want to know is the population-level average treatment effect.
  - i.e. the effect of offering these plans to every employee.

$$
\begin{aligned}
\tau(50)  &= \frac{1}{m} \sum_{j=1}^m \tau_j(50)  \\
\tau(100) &= \frac{1}{m} \sum_{j=1}^m \tau_j(100)
\end{aligned}
$$

- What's hard about this?
  - *Sampling*. We only observe information about a fraction of these people.
  - *Missing Data*. For that fraction, we only observe one of the potential outcomes.

Gray out missing potential outcomes, too. No colors.
$$
\small{
\begin{array}{c|ccc|cc}
j & y_j(0) & y_j(50) & y_j(100) & \tau_j(50) & \tau_j(100) \\
\color{gray}1 & 5  & 10 & 10 & 5 & 0 \\
\color{blue}2 & 10 & 10 & 10 & 0 & 0 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{green}327 & 0 & 0  &  20 & 0 & 20 \\
\color{gray}\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{gray}358 & 15 & 18 & 20 & 3 & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{gray}m & \color{gray}10 & \color{gray}10 & \color{gray}5  & \color{gray}0 & \color{gray}-5 \\
\end{array}
}
$$

## Sampling

$$
\small{
\begin{array}{c|ccc|cc}
j & y_j(0) & y_j(50) & y_j(100) & \tau_j(50) & \tau_j(100) \\
1 & 5  & 10 & 10 & 5 & 0 \\
\color{blue}2 & 10 & 10 & 10 & 0 & 0 \\
\vdots & \vdots  & \vdots & \vdots & \vdots & \vdots \\
\color{purple}327 & 4 & 8 & 16 & 3 & 2 \\
\vdots & \vdots  & \vdots & \vdots & \vdots & \vdots \\
\color{green}358 & 0 & 0  &  20 & 0 & 20 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
m & 10 & 10 & 5  & 0 & -5 \\
\end{array}
}
$$
:::
::: column

$$
\small{
\begin{array}{c|cc|ccc|cc}
i & D_i & Y_i   & Y_i(0) & Y_i(50) & Y_i(100) & \tau_i(50) & \tau_i(100) \\
\color{blue}1   &  & & 10 & 10 & 10 & 0 & 0 \\
\color{green}2  &  & & 0  & 0  &  20 & 0 & 20 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{purple}n & &  & 15 & 18 & 20 & 3 & 2 \\
\end{array}
}
$$

:::
::::

- We don't want to trouble all of our employees with a survey, so we sample a subset of them.
- Each employee in our sample is drawn, with replacement, from our population of employees---all $m$ of them.
- That means each row of our *sample table* is a random row from our *population table*.
  - I've colored them so we can track them from population into the sample.

## Missing Data

- We assign a treatment to each sampled employee.
  - That tells us which potential outcome we observe.
  - It's the one corresponding to the treatment that actually happens: $Y_i=Y_i(D_i)$.

$$
\small{
\begin{array}{c|cc|ccc|cc}
i & D_i & Y_i   & Y_i(0) & Y_i(50) & Y_i(100) & \tau_i(50) & \tau_i(100) \\
\color{blue}1   &  & & 10 & 10 & 10 & 0 & 0 \\
\color{green}2  &  & & 0  & 0  &  20 & 0 & 20 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{purple}n & &  & 15 & 18 & 20 & 3 & 2 \\
\end{array}
}
$$

## This is What We're Left With

- How do we recover we need?

[ Two column table ]

- If we had all the potential outcomes, we could approximate the population-level average treatment effect by a sample average.
$$
\frac{1}{m} \sum_j y_j(50) - y_j(0) \approx \frac{1}{n} \sum_i Y_i(50) - Y_i(0)
$$

- This works because of the law of large numbers. Sample averages approximate population averages.
- But we don't have all the potential outcomes. We only have one of them for each person.
- One thing we can do is average the outcomes we do have. Does this work?

$$
\begin{aligned}
\frac{1}{n} \sum_i Y_i(50) &\overset{?}{\approx} \frac{1}{N_{50}} \sum_{i:D_i=50}Y_i(50) = \frac{1}{N_{50}} \sum_{i:D_i=50}Y_i \\
\frac{1}{n} \sum_i Y_i(0)  &\overset{?}{\approx} \frac{1}{N_{0}} \sum_{i:D_i=0}Y_i(0)    = \frac{1}{N_{0}} \sum_{i:D_i=0}Y_i
\end{aligned}
$$

- This works because of randomization. The coin-flip that determines treatment *has nothing to do with the potential outcomes*. Or anything else.
- So we are, in effect, just sampling again when we choose the rows with $D_i=50$ and $D_i=0$. This time, from the sample.

[ Table split into two tables ]

## Formal Identification

- All this is getting at the idea that we can write the population-level average treatment effect 
  in terms of means of the observed outcome in subpopulations.
$$
\begin{aligned}
\frac{1}{m}\sum_j y_j(50) - y_j(0) 
&= \E[Y_i(50) - Y_i(0)] \\
&= \E[Y_i(50) \mid D_i=50] -  \E[Y_i(0) \mid D_i=0] \\
&= \E[Y_i \mid D_i=50] -  \E[Y_i \mid D_i=0] \\
&= \mu(50) - \mu(0).
\end{aligned}
$$

- Now we're done with the causal stuff. 
- We've worked out how to answer our question in terms of things we can estimate.
- And the particular form --- a linear combination of subpopulation means --- is something we've done before.



# Old Shit

## Potential Outcomes and Treatment Effects 

- Potential Outcomes
  - There are three levels of treatment. We'll write them in terms of what the employer pays.d
- Individual-Level Treatment Effect
  - There are two natural individual-level treatment effects to think about.
  - The effect of a \$50 gym membership vs none:  $\tau_i(50)  = Y_i(50) - Y_i(0)$
  - The effect of a \$100 gym membership vs \$50:  $\tau_i(100) = Y_i(100) - Y_i(50)$

$$\begin{array}{c|ccc|cc}
i & Y_i(0) & Y_i(50) & Y_i(100) & \tau_i(50) & \tau_i(100) \\
\end{array}
$$


## What We See 

- Observed Outcomes
  - We only observe one of these potential outcomes: $Y_i = Y_i(D_i)$
  - That's not enough to tell use about the value of any one of these effects.
- Randomization
  - The employer assigns participants to these three plans randomly. 
  - They do it by having the employees roll two dice.
    - If they roll an 11 or 12, they get the \$100 plan.
    - If they roll 7,8,9, or 10 they get the \$50 plan.
    - If they roll 1-6, they get the \$0 plan---i.e. nothing.

## Population-Level Treatment Effects

- What they care about is the population-level treatment effect.
  - i.e. the effect of offering these plans to every employee.
- That'll have an effect on the rate they're able to negotiate with their insurance provider.
- We can think of our employees as having potential outcomes associated with each level of treatment.
- And the effects we're interested in are population-level average treatment effects.

:::: columns

::: column
$$
\small{
\begin{array}{c|ccc|cc}
j & y_j(0) & y_j(50) & y_j(100) & \tau_j(50) & \tau_j(100) \\
1 & 5  & 10 & 10 & 5 & 0 \\
\color{blue}2 & 10 & 10 & 10 & 0 & 0 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{green}327 & 0 & 0  &  20 & 0 & 20 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{purple}358 & 15 & 18 & 20 & 3 & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
m & 10 & 10 & 5  & 0 & -5 \\
\end{array}
}
$$

:::
::: column

$$
\small{
\begin{aligned}
\tau(50)  &= \frac{1}{m} \sum_{j=1}^m \tau_j(50)  \\
\tau(100) &= \frac{1}{m} \sum_{j=1}^m \tau_j(100) 
\end{aligned}
}
$$

:::
::::


## Sampling

- We don't want to trouble all of our employees with a survey, so we sample a subset of them.
- In the table, I've colored a few employees so we can track them from population into the sample.
- To have some concrete behaviors in mind, we'll keep an eye on them.
  - In blue, we have an employee doesn't care about subsidies. 
    - They're going to the gym 10 days/month no matter what.
  - In purple, we have one who is a bit price-sensitive.
    - They'll go once a week with no subsidy, twice a week if it's free, and 4x/week with free classes.
  - In green, we have one who loves classes but hates spending money even more. 
    - They'll go 20 days/month with free classes, but won't go at all otherwise.
  
:::: columns
::: column
$$
\small{
\begin{array}{c|ccc|cc}
j & y_j(0) & y_j(50) & y_j(100) & \tau_j(50) & \tau_j(100) \\
1 & 5  & 10 & 10 & 5 & 0 \\
\color{blue}2 & 10 & 10 & 10 & 0 & 0 \\
\vdots & \vdots  & \vdots & \vdots & \vdots & \vdots \\
\color{purple}327 & 4 & 8 & 16 & 3 & 2 \\
\vdots & \vdots  & \vdots & \vdots & \vdots & \vdots \\
\color{green}358 & 0 & 0  &  20 & 0 & 20 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
m & 10 & 10 & 5  & 0 & -5 \\
\end{array}
}
$$
:::
::: column

$$
\small{
\begin{array}{c|cc|ccc|cc}
i & D_i & Y_i   & Y_i(0) & Y_i(50) & Y_i(100) & \tau_i(50) & \tau_i(100) \\
\color{blue}1   &  & & 10 & 10 & 10 & 0 & 0 \\
\color{green}2  &  & & 0  & 0  &  20 & 0 & 20 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{purple}n & &  & 15 & 18 & 20 & 3 & 2 \\
\end{array}
}
$$

:::
::::

- Each row in this table is chosen, independently with probability $1/m$, from the population table.
- That tells us that, in particular, expectations involving the quantities in these rows coincide with population averages.
- And if we have a large enough sample, we can approximate these expectations with sample averages.

$$
\begin{aligned}
\E Y_i(50) = \frac{1}{m}\sum_{j=1}^m y_j(50) \\
\E \tau_i(50) = \frac{1}{m}\sum_{j=1}^m y_j(50) - y_j(0) = \tau(50) \\
\end{aligned}
$$

## Randomization 

- We roll the dice for each of our sampled employees.
- In this case, the people we're tracking get what it takes for them to go to the gym.
  - The blue one rolls a 3.    They get nothing. But they're maintaining their schedule no matter what.
  - The purple one rolls an 8. They get the $\50 plan, so they're at the gym twice a week.
  - The green one rolls a 12.  They get the $\100 plan, so they're at the gym 20 days/month.
  
$$
\begin{array}{c|cc|ccc|cc}
i & D_i & Y_i   & Y_i(0) & Y_i(50) & Y_i(100) & \tau_i(50) & \tau_i(100) \\
\color{blue}1   & 0    & 10 & 10 & 10 & 10 & 0 & 0 \\
\color{green}2  & 100  & 20 & 0  & 0  &  20 & 0 & 20 \\
\vdots & \vdots &  \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{purple}n & 50   & 18 & 15 & 18 & 20 & 3 & 2 \\
\end{array}
$$

## Alternate Randomization 

- That could happen. But so could this.
- In this case, the subsidies that go to our three people aren't that encouraging.
  - The blue one rolls a 9.    They get the \$100 plan, but that makes no difference to their behavior.
  - The purple one rolls an 2. They get the \$0 plan,   so they're only in the gym once a week.
  - The green one rolls a 7.   They get the \$50 plan, but it's worth nothing to them. No free classes, no gym!
  
$$
\begin{array}{c|cc|ccc|cc}
i & D_i & Y_i   & Y_i(0) & Y_i(50) & Y_i(100) & \tau_i(50) & \tau_i(100) \\
\color{blue}1   &100  &  & 10 & 10 & 10 & 0 & 0 \\
\color{green}2  & 50   & 0 & 0  & 0  &  20 & 0 & 20 \\
\vdots & \vdots &  \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{purple}n & 0   & 15 & 15 & 18 & 20 & 3 & 2 \\
\end{array}
$$

## An In-Class Randomization

- Let's do a real one. Let's roll some dice and fill in the table.
- What's the story here?
  - The blue one rolls a  ...
  - The purple one rolls a ...
  - The green one rolls a ...

$$
\begin{array}{c|cc|ccc|cc}
i & D_i & Y_i   & Y_i(0) & Y_i(50) & Y_i(100) & \tau_i(50) & \tau_i(100) \\
\color{blue}1   &  &  & 10 & 10 & 10 & 0 & 0 \\
\color{green}2  &  &  & 0  & 0  &  20 & 0 & 20 \\
\vdots & \vdots &  \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{purple}n &     &  & 15 & 18 & 20 & 3 & 2 \\
\end{array}
$$

## Implications of Sampling and Randomization --- Informal

- We've looked at what happened to 3 employees in a few different randomizations. 
- This was just a thought experiment. Only one randomization actually happens.
- But if our sample size is large enough, we'll have multiple employees of each 'type'.
  - Some who don't respond to subsidies, like our blue one.
  - Some who respond only if they get the biggest subsidy, like our purple one.
  - Some who respond a bit at each level, like our green one.
  - And other types, too.
- And, in effect, all of the different rolls we've looked at will happen at once.
- They'll happen to different people, of course, but people who behave more or less the same. 
- And this randomization ensures that the *mix of types* is the same in in each treatment group.
    - i.e., we'll have the same proportion of blue, purple, and green types among the people with $D_i=0,50,100$.
- This means that the distribution of *each potential outcome* is the same within each group and in the whole sample.
- And that *includes the realized outcome*, which is one of them.
- And because we're looking at averages...

$$
\begin{array}{c|cc|ccc|cc}
i & D_i & Y_i   & Y_i(0) & Y_i(50) & Y_i(100) & \tau_i(50) & \tau_i(100) \\
\color{blue}1   & 0    & 10 & 10 & 10 & 10 & 0 & 0 \\
\color{green}2  & 100  & 20 & 0  & 0  &  20 & 0 & 20 \\
\vdots & \vdots &  \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{purple}27 & 50   & 18 & 15 & 18 & 20 & 3 & 2 \\
\color{blue}36   &100  &  & 10 & 10 & 10 & 0 & 0 \\
\vdots & \vdots &  \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{purple}56 & 0   & 15 & 15 & 18 & 20 & 3 & 2 \\
\color{blue}57   &  &  & 10 & 10 & 10 & 0 & 0 \\
\color{green}58  & 50   & 0 & 0  & 0  &  20 & 0 & 20 \\
\color{green}59  &  &  & 0  & 0  &  20 & 0 & 20 \\
\vdots & \vdots &  \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{purple}n &     &  & 15 & 18 & 20 & 3 & 2 \\
\end{array}
$$

## Implications of Sampling and Randomization --- Formal

- Randomization ensures that the potential outcomes have the same distribution ...
  - within each treatment group, i.e., for people with $D_i=0$,$D_i=50$,$D_i=100$.
  - in the whole sample.
- And in particular, the *realized outcome* 
- And because we're looking at averages...
$$
E[Y_i(d)] = E[Y_i(d)|D_i=d] = \underset{\mu(d)}{E[Y_i \mid D_i=d]} \approx \underset{\hat\mu(d)}{\frac{1}{N_d}\sum_{i:D_i=d} Y_i}
$$