---
title: "QTM 220: Lecture 12"
subtitle: "Causal Inference with a Multivalued Treatment"
format: 
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---


```{r load-libraries}
library(purrr)
library(tibble)
library(dplyr)
library(magrittr)
library(stringr)
library(ggplot2)
library(htmltools)
library(kableExtra)
library(scales)

percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }
```

```{r ggplot-themes}
fontsize = 16
textcolor = "#888888"

class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
        panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x  = element_text(colour = textcolor, size=fontsize),
        axis.text.y  = element_text(colour = textcolor, size=fontsize),
        axis.title.x  = element_text(colour = textcolor),
        axis.title.y  = element_text(colour = textcolor, angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
              axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
           panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(simplified.theme)
```


```{r table-styling}
ellipsis <- function(df, nhead=5L, ntail=2) {
    rownames(df) = as.character(1:nrow(df))
    stopifnot("data.frame" %in% class(df))
    els <- rep(NA, ncol(df)) %>% 
        matrix(nrow=1, dimnames=list(NULL, names(df))) %>% 
        data.frame(stringsAsFactors=FALSE) %>% 
        tibble::as_tibble()
    rownames(els) = "⋮"
    dplyr::bind_rows(head(df, nhead), els, tail(df, ntail))
}
```


## Today

::: {.hidden}
$$
\newcommand{\alert}[1]{\textcolor{red}{#1}}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\V}{V}
\DeclareMathOperator{\Var}{\V}
\newcommand{\inred}[1]{\textcolor{red}{#1}}
\newcommand{\inblue}[1]{\textcolor{blue}{#1}}
$$ 
:::

- We're going to talk about causal inference with a multivalued treatment.
- Our context will be a randomized experiment with three treatments.
- A large employer is experimenting with a new benefit for their employees: subsized gym memberships.
- They're randomly selecting employees to participate in a trial program.
- Then randomizing these participants into three plans.
  - Plan 1: No gym membership. It costs them \$0/month.
  - Plan 2: A gym membership that covers entry but not classes. It costs \$50/month.
  - Plan 3: A gym membership that covers entry but and classes. It costs \$100/month.
- And surveying them to see how often they go to the gym, e.g. in days/month.
- They'll be able to use this information to decide what plan to offer everyone next year.

## How We Think

- We'll think about this using potential outcomes.
- We imagine, for each employee, that there's a number of times they'd go to the gym given each plan.
  - $y_j(0)$: The number of days/month they'd go to the gym if they had no membership.
  - $y_j(50)$: The number of days/month they'd go to the gym if they had a \$50 membership.
  - $y_j(100)$: The number of days/month they'd go to the gym if they had a \$100 membership.
- And we'll think about the effect of each plan on each employee.
  - $\tau_j(50)  = y_j(50)  - y_j(0)$: The effect of a \$50 gym membership vs none.
  - $\tau_j(100) = y_j(100) - y_j(50)$: The effect of a \$100 gym membership vs \$50.

$$
\small{
\begin{array}{c|ccc|cc}
j & y_j(0) & y_j(50) & y_j(100) & \tau_j(50) & \tau_j(100) \\
1 & 5  & 10 & 10 & 5 & 0 \\
2 & 10 & 10 & 10 & 0 & 0 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
327 & 0 & 0  &  20 & 0 & 20 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
358 & 15 & 18 & 20 & 3 & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
m & 10 & 10 & 5  & 0 & -5 \\
\end{array}
}
$$

## What We Want To Know

- What we want to know is the population-level average treatment effect.
  - i.e. the average increase in monthly gym visits if we offered these plans to every employee.

$$
\small{
\begin{aligned}
\tau(50)  &= \frac{1}{m} \sum_{j=1}^m \tau_j(50)  && \quad \text{this is the first increment---the effect of \$50 vs. \$0} \\
\tau(100) &= \frac{1}{m} \sum_{j=1}^m \tau_j(100) && \quad \text{this is the second increment---the effect of \$100 vs. \$50} \\
\end{aligned}
}
$$

- What's hard about this?
  - *Sampling*. We only observe information about a fraction of these people.
  - *Missing Data*. For that fraction, we only observe one of the potential outcomes.


$$
\small{
\begin{array}{c|ccc|cc}
j & y_j(0) & y_j(50) & y_j(100) & \tau_j(50) & \tau_j(100) \\
\color{gray}1 & \color{gray}5  & \color{gray}10 & \color{gray}10 & \color{gray}5 & \color{gray}0 \\
2 & 10 & \color{gray}10 & \color{gray}10 & \color{gray}0 & \color{gray}0 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
327 & \color{gray}0 & \color{gray}0  &  20 & \color{gray}0 & \color{gray}20 \\
\color{gray}\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
358 & \color{gray}15 & 18 & \color{gray}20 & \color{gray}3 & \color{gray}2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{gray}m & \color{gray}10 & \color{gray}10 & \color{gray}5  & \color{gray}0 & \color{gray}-5 \\
\end{array}
}
$$

## Sampling

:::: columns
::: column
### Our Population 
$$
\small{
\begin{array}{c|ccc|cc}
j & y_j(0) & y_j(50) & y_j(100) & \tau_j(50) & \tau_j(100) \\
\color{gray} 1 & \color{gray}5  & \color{gray}10 & \color{gray}10 & \color{gray}5 & \color{gray}0 \\ 
\color{blue}2 & \color{blue}10 & \color{blue}10 & \color{blue}10 & \color{blue}0 & \color{blue}0 \\
\vdots & \vdots  & \vdots & \vdots & \vdots & \vdots \\
\color{purple}327 & \color{purple}15 & \color{purple}18 & \color{purple}20 & \color{purple}3 & \color{purple}2 \\
\vdots & \vdots  & \vdots & \vdots & \vdots & \vdots \\
\color{green}358 & \color{green}0 & \color{green}0  &  \color{green}20 & \color{green}0 & \color{green}20 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{gray}m & \color{gray}10 & \color{gray}10 & \color{gray}5  & \color{gray}0 & \color{gray}-5 \\
\end{array}
}
$$

:::
::: column
### Our Sample
$$
\small{
\begin{array}{c|ccc|cc}
i &  Y_i(0) & Y_i(50) & Y_i(100) & \tau_i(50) & \tau_i(100) \\
\color{blue}1 & \color{blue}10 & \color{blue}10 & \color{blue}10 & \color{blue}0 & \color{blue}0 \\
\color{green}2 & \color{green}0 & \color{green}0  &  \color{green}20 & \color{green}0 & \color{green}20 \\
\vdots & \vdots  & \vdots & \vdots & \vdots \\
\color{purple}n & \color{purple}15 & \color{purple}18 & \color{purple}20 & \color{purple}3 & \color{purple}2 \\
\end{array}
}
$$

:::
::::

- We don't want to trouble all of our employees with a survey, so we sample a subset of them.
- Each employee in our sample is drawn, with replacement, from our population of employees---all $m$ of them.
- That means each row of our *sample table* is a random row from our *population table*.
  - I've colored them so we can track them from population into the sample.

## Missing Data

- We assign a treatment to each sampled employee.
  - That tells us which potential outcome we observe.
  - It's the one corresponding to the treatment that actually happens: $Y_i=Y_i(D_i)$.

$$
\small{
\begin{array}{c|cc|ccc|cc}
i & D_i & Y_i   & Y_i(0) & Y_i(50) & Y_i(100) & \tau_i(50) & \tau_i(100) \\
\color{blue}1   &  \color{blue}0 & \color{blue}10 & \color{blue}10 & \color{gray}10 &  \color{gray}10 &  \color{gray}0 &  \color{gray}0 \\
\color{green}2  &  \color{green}100& \color{green} 20 &  \color{gray}0  &  \color{gray}0  &  \color{green}20 &  \color{gray}0 &  \color{gray}20 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{purple}n &  \color{purple}50& \color{purple}18  & \color{gray}15 &  \color{purple}18 &  \color{gray}20 &  \color{gray}3 &  \color{gray}2 \\
\end{array}
}
$$

- We'll do this *randomly*, e.g. by rolling a die. In effect, this is just *another sampling step*.
- Think of the group that receives each treatment as being drawn from our population by a two-stage process.
  - First, roll an $m$-sided die to choose a row from our population table *to consider*.
  - Then, roll a 3-sided die to decide whether we're included in the subsample with $D_i=d$.


:::: columns
::: {.column width="30%"}
$$
\small{
\begin{array}{c|cc}
i & D_i & Y_i    \\
\color{blue}1   &  \color{blue}0 & \color{blue}10 \\
\vdots & 0 & \vdots  \\
\end{array}
}
$$
:::
::: {.column width="30%"}
$$
\small{
\begin{array}{c|cc}
i & D_i & Y_i    \\
\vdots & 50 & \vdots \\
\color{purple}n &  \color{purple}50& \color{purple}18 \\
\end{array}
}
$$
:::
::: {.column width="30%"}
$$
\small{
\begin{array}{c|cc}
i & D_i & Y_i    \\
\vdots & 100 & \vdots \\
\color{green}2  &  \color{green}100& \color{green} 20  \\
\end{array}
}
$$
:::
::::



## Getting By With What We Have


- If we had all the potential outcomes... 
- We could approximate a population-level average treatment effect by a sample average.

$$
\small{
\frac{1}{m} \sum_{j=1}^m y_j(50) - y_j(0) \approx \frac{1}{n} \sum_{i=1}^n Y_i(50) - Y_i(0) 
}
$$

- This works because of the law of large numbers. Sample averages approximate population averages.
- By the same logic, we can do the same using *each treatment group*. 
- When we randomize, these are random samples from the population, too.

$$
\small{
\frac{1}{m} \sum_{j=1}^m y_j(50) - y_j(0) \approx \frac{1}{N_d} \sum_{i:D_i=d}^n Y_i(50) - Y_i(0) = \frac{1}{N_d} \sum_{i:D_i=d} Y_i(50) - \frac{1}{N_d}\sum_{i:D_i=d} Y_i(0)
}
$$ 

- This still isn't something we can use. We don't have all the potential outcomes. But it just takes one tweak. 
- We use *different subsamples* to approximate the averages of diferent potential outcomes. 
- In particular, we use the subsample in which that potential outcome is observed.

$$
\small{
\frac{1}{m} \sum_{j=1}^m y_j(50) - y_j(0) \approx \frac{1}{N_d} \sum_{i:D_i=50} Y_i(50) - \frac{1}{N_0}\sum_{i:D_i=0} Y_i(0)  = \frac{1}{N_d} \sum_{i:D_i=50} Y_i - \frac{1}{N_0}\sum_{i:D_i=0} Y_i
}
$$


## Formal Identification

- Here, in formal language, is the crux of this argument. As a result of sampling and randomization, ...
- Subsample averages of realized outcomes are *unbiased and consistent* estimators of population averages of potential outcomes.
- That's because *randomized assignment* ensures that the treatment $D_i$ is *independent* of the potential outcomes.

$$
\small{
\begin{aligned}
\frac{1}{m}\sum_j y_j(d) 
  &= \E[Y_i(d)]            && \text{sampling} \\
  &= \E[Y_i(d) \mid D_i=d] && \text{irrelevance of independent conditioning variables} \\
  &= \E[Y_i \mid D_i=d]=\mu(d)    && \text{$Y_i(d)=Y_i$ \ when \ $D_i=d$ } \\
&\approx \hat\mu(d) = \frac{1}{N_d}\sum_{i:D_i=d} Y_i && \text{law of large numbers}
\end{aligned}
}
$$

- If we can express what we want to estimate as a linear combination of potential outcome means...
- This means we can *rewrite it* as a linear combination of *subpopulation means*.

$$
\small{
\begin{aligned}
\tau(50) &= \frac{1}{m}\sum_j y_j(50) - y_j(0) = \mu(50) - \mu(0) \\
\tau(100) &= \frac{1}{m}\sum_j y_j(100) - y_j(50) = \mu(100) - \mu(50) \\
&\frac{1}{m}\sum_j \alpha(100)y_j(100) + \alpha(50)y_j(50) + \alpha(0)y_j(0) = \alpha(100)\mu(100) + \alpha(50)\mu(50) + \alpha(0)\mu(0) \\
\end{aligned}
}
$$

## Where This Leaves Us

$$
\small{
\begin{aligned}
\tau(50) &= \frac{1}{m}\sum_j y_j(50) - y_j(0) = \mu(50) - \mu(0) \\
\tau(100) &= \frac{1}{m}\sum_j y_j(100) - y_j(50) = \mu(100) - \mu(50) \\
&\frac{1}{m}\sum_j \alpha(100)y_j(100) + \alpha(50)y_j(50) + \alpha(0)y_j(0) = \alpha(100)\mu(100) + \alpha(50)\mu(50) + \alpha(0)\mu(0) \\
\end{aligned}
}
$$

- Once we've done this, we can put away our causal thinking.
- We know, in terms of the distribution of our observations ($D_i,Y_i$), what we want to estimate.
- And there's nothing especially causal about that. We can just use the techniques we've already learned.


## Today's Focus
```{r}
#| out-height: 300px
#| fig-asp: .3 
set.seed(0)
n=250
Yalways = rbinom(n, 30, .25)
Y0   = ifelse(rbinom(n,1,.5),   Yalways, rbinom(n, 30, .05))
Y50  = ifelse(rbinom(n,1,.5),   Yalways, rbinom(n, 30, .1))
Y100 = Y50
R = sample(c(1:6), size=n, replace=TRUE) + sample(c(1:6), size=n, replace=TRUE)
D = case_match(R,
  1:6~0,
  7:10~50,
  11:12~100)
Y = case_match(D,
  0~Y0,
  50~Y50,
  100~Y100)

real.data = tibble(D=D, Y=Y)

summaries = tibble(D=D, Y=Y) %>% group_by(D) %>% summarize(mean=mean(Y), sd=sd(Y), n=length(Y))

scatter = ggplot() + geom_point(aes(x=D, y=Y), color="red", data=tibble(D=D,Y=Y),
                                     position=position_jitter(w=10,h=0), alpha=.3) +
           geom_pointrange(aes(x=D, y=mean, ymin=mean-2*sd, ymax=mean+2*sd), data=summaries) +
           scale_x_continuous(breaks=c(0,50,100)) + labs(x="D", y="Y") + ylim(-3,15)
scatter
``` 

- Let's talk about estimating $\tau(100)$, the effect of a \$100 gym membership vs. a \$50 membership.
  - We really don't want to spend \$100/month if we don't have to.
  - So we're hoping it's not much better than spending \$50/month.
- The statistical challenge is we've leaned into that preference in our experiment too.
- We've designed our experiment so not many people are in the \$100 group.
- In particular, for each employee in our sample, we're rolling two dice
  - If the sum is 11 or 12, they get the \$100 plan.
  - If the sum if 7,8,9, or 10, they get the \$50 plan.
  - If the sum is 1-6, they get the \$0 plan.

## We Know How To Do This
```{r}
#| out-height: 300px
#| fig-asp: .3 
point.estimate = summaries$mean[summaries$D==100] - summaries$mean[summaries$D==50]
scatter
``` 

$$
\tau(100) = \mu(100) - \mu(50)
$$

- We can use a simple difference in subsample means as a point estimate.

$$
\hat\tau(100) = \hat\mu(100) - \hat\mu(50) = `r approx(summaries$mean[summaries$D==100])` - `r approx(summaries$mean[summaries$D==50])` \approx `r approx(point.estimate)`
$$

- And we can use an estimate of its sampling distribution to calibrate an interval estimate.

```{r}
data = tibble(D=D, Y=Y)
boot.distn = function(data, B=1000) { purrr::map(1:B, function(i) {
  data.boot = data[sample(1:nrow(data), replace=TRUE),]
  summaries.boot = data.boot %>% group_by(D) %>% summarize(mean=mean(Y), sd=sd(Y), n=length(Y))
  data.frame(dif=summaries.boot$mean[summaries.boot$D==100] - summaries.boot$mean[summaries.boot$D==50])
}) %>% bind_rows()
}
boots = boot.distn(data)
```

```{r plotting-code}
width = function(samples, center=mean(samples), alpha=.05) {
  covg.dif = function(w) { 
    actual = mean( center - w/2 <= samples & samples <= center + w/2)
    nominal = 1-alpha 
    actual - nominal  
  }
  uniroot(covg.dif, interval=c(0, 2*max(abs(samples-center))))$root
}

distribution.plot = function(boot.samples, point.estimate, w=width(boot.samples), interval.replications=0, point.replications=0) { 
  drange = c(0, dnorm(0, mean=0, sd=sd(boot.samples)))
  p = ggplot(data.frame(samples=boot.samples)) + 
    geom_histogram(aes(x=samples, y=after_stat(density)), alpha=.3, bins=50)  +
    geom_pointrange(aes(y=dnorm(2*sd(samples), mean=0, sd=sd(samples)), 
      x=point.estimate, xmin=point.estimate - w/2, xmax=point.estimate + w/2)) + 
      geom_vline(xintercept=mean(boot.samples), color='blue') + 
      geom_vline(xintercept=quantile(boot.samples, c(.025,.975)), color='blue', linetype='dotted') 
  if(interval.replications > 0) { 
    K = interval.replications
    p = p + geom_pointrange(aes(y=y, x=x, xmin=xmin, xmax=xmax), 
          data=data.frame(y=seq(min(drange), max(drange), length.out=K), 
                  x=boot.samples[1:K], xmin=boot.samples[1:K] - w/2, xmax=boot.samples[1:K] + w/2), color='purple', alpha=.2)
  }
  if(point.replications > 0) { 
    K = point.replications
    p = p + geom_point(aes(y=y, x=x), 
          data=data.frame(y=seq(min(drange), max(drange), length.out=K), 
                  x=boot.samples[1:K], xmin=boot.samples[1:K] - w/2, xmax=boot.samples[1:K] + w/2), color='purple', alpha=.2)
  }
  p
}
```

```{r}
#| out-height: 250px
#| fig.asp: .25
#| layout-ncol: 2

se = sqrt( var(Y[D==100])/sum(D==100) + var(Y[D==50])/sum(D==50) )
x = seq(point.estimate-4*se, point.estimate+4*se, length.out=1000)
normal.distribution.plot = ggplot() + 
    geom_area(aes(x=x, y=dnorm(x, mean=point.estimate, sd=se)), color='purple', fill='purple', alpha=.08) +
    geom_vline(aes(xintercept=point.estimate), color='purple', linetype='solid') +
    geom_vline(aes(xintercept=point.estimate + c(-1,1)*1.96*se), color='purple', linetype='dotted') +
    geom_pointrange(aes(y=dnorm(2*se, mean=0, sd=se), x=point.estimate, xmin=point.estimate - 1.96*se, xmax=point.estimate + 1.96*se))

distribution.plot(boots$dif, point.estimate) + xlim(min(x), max(x))
normal.distribution.plot + xlim(min(x), max(x))
``` 

## But We're Not Happy With Our Precision
```{r}
#| out-height: 250px
#| fig.asp: .25
normal.distribution.plot + xlim(min(x), max(x))
``` 

- It's worth it to spend the extra money if it gets us at least 1 day/month for the average employee.
  - And we can't rule that out.
  - But we can't 'rule it in' either. 
- We just don't have compelling evidence in one direction or the other.
- So let's look at why that is. What did we do wrong when designing this experiment? 


## Variance
```{r}
#| out-height: 250px
#| fig.asp: .25
normal.distribution.plot + xlim(min(x), max(x))
``` 

- Let's calculate the variance of our point estimate to see where the problem is.
$$
\Var\qty[\sum_d \alpha(d) \hat\mu(d)]  = \sum_d \sigma^2(d) \alpha^2(d) \times \E\qty[ \frac{1}{N_d} ]
$$

- We can use this table.

$$
\begin{array}{c|ccc}
d & 0 & 50 & 100 \\
\hline
\hat\sigma^2(d) & `r approx(var(Y[D==0]))` & `r approx(var(Y[D==50]))` & `r approx(var(Y[D==100]))` \\
\alpha^2(d) &  &  &  \\
N_d & `r approx(sum(D==0))` & `r approx(sum(D==50))` & `r approx(sum(D==100))` \\
\frac{\hat\sigma^2(d) \alpha^2(d)}{N_d} &  &  &  \\
\end{array}
$$

- What do you see?

## Variance
```{r}
#| out-height: 250px
#| fig.asp: .25
normal.distribution.plot + xlim(min(x), max(x))
``` 


$$
\begin{array}{c|ccc}
d & 0 & 50 & 100 \\
\hline
\hat\sigma^2(d) & `r approx(var(Y[D==0]))` & `r approx(var(Y[D==50]))` & `r approx(var(Y[D==100]))` \\
\alpha^2(d) & 0 & 1 & 1 \\
N_d & `r approx(sum(D==0))` & `r approx(sum(D==50))` & `r approx(sum(D==100))` \\
\frac{\hat\sigma^2(d) \alpha^2(d)}{N_d} & 0 & `r approx(var(Y[D==50])/sum(D==50))` & `r approx(var(Y[D==100])/sum(D==100))` \\
\end{array}
$$

- What I see is that we don't have enough people randomized to the $D=100$ treatment.
  - The variance we get from just that one term is $1/4$.
  - Meaning the contribution to our standard error is $\sqrt{1/4}=1/2$
  - And since we multiply standard error by 1.96 to get an 'arm' for our interval ...
  - This ensures our arms are at least $1.96 \times 1/2 \approx 1$ day wide.

## If We Could Do It Again
```{r}
#| out-height: 300px
#| fig-asp: .3 

n = 250
D = sample(c(50,100), size=n, replace=TRUE) 
Y = case_match(D,
  0~Y0,
  50~Y50,
  100~Y100)

summaries.new = tibble(D=D, Y=Y) %>% group_by(D) %>% summarize(mean=mean(Y), sd=sd(Y), n=length(Y))
point.estimate = summaries.new$mean[summaries.new$D==100] - summaries.new$mean[summaries.new$D==50]

scatter.new = ggplot() + geom_point(aes(x=D, y=Y), color="red", data=tibble(D=D,Y=Y),
                                     position=position_jitter(w=10,h=0), alpha=.3) +
           geom_pointrange(aes(x=D, y=mean, ymin=mean-2*sd, ymax=mean+2*sd), data=summaries) +
           scale_x_continuous(breaks=c(0,50,100)) + labs(x="D", y="Y") + ylim(-3,15)
scatter.new
```

- We'd assign more people to the $D=100$ treament to get a more precise estimate.
- Here's what we'd get if we assigned people to each of our 2 *relevant* treatments with probability $1/2$.
- This does what we need it to do: it gives us a confidence interval that to the left of $1$.

```{r}
data = tibble(D=D, Y=Y)
boot.distn = function(data, B=1000) { purrr::map(1:B, function(i) {
  data.boot = data[sample(1:nrow(data), replace=TRUE),]
  summaries.boot = data.boot %>% group_by(D) %>% summarize(mean=mean(Y), sd=sd(Y), n=length(Y))
  data.frame(dif=summaries.boot$mean[summaries.boot$D==100] - summaries.boot$mean[summaries.boot$D==50])
}) %>% bind_rows()
}
boots = boot.distn(data)
```

```{r}
#| out-height: 250px
#| fig.asp: .25
#| layout-ncol: 2

se = sqrt( var(Y[D==100])/sum(D==100) + var(Y[D==50])/sum(D==50) )
x = seq(point.estimate-4*se, point.estimate+4*se, length.out=1000)
normal.distribution.plot = ggplot() + 
    geom_area(aes(x=x, y=dnorm(x, mean=point.estimate, sd=se)), color='purple', fill='purple', alpha=.08) +
    geom_vline(aes(xintercept=point.estimate), color='purple', linetype='solid') +
    geom_vline(aes(xintercept=point.estimate + c(-1,1)*1.96*se), color='purple', linetype='dotted') +
    geom_pointrange(aes(y=dnorm(2*se, mean=0, sd=se), x=point.estimate, xmin=point.estimate - 1.96*se, xmax=point.estimate + 1.96*se))

distribution.plot(boots$dif, point.estimate) + xlim(min(x), max(x))
normal.distribution.plot + xlim(min(x), max(x))
``` 

## But We Can't

```{r}
#| out-height: 300px
#| fig-asp: .3 


model=lm(Y ~ D, data=real.data)
mu.hat = predict(model, newdata=data.frame(D=c(0,50,100)))
slope.point.estimate = mu.hat[3]-mu.hat[2]

scatter + geom_line(aes(x=c(0,50,100), y=mu.hat), color='blue', linetype='dashed')
```

- This is our data. We have to do what we can. So we get clever.
- Instead of estimating each subpopulation mean separately, we *fit a line* to the data.
- We take $\hat\mu(d)$ to be the height of the line at $d$ and $\hat\tau(d)=\hat\mu(100)-\hat\mu(50)$ as before.
- And when we bootstrap it ... 
  - We get a much narrower interval than the one we had before.
  - But it's not in the same place at all.

```{r}
boot.distn = function(data, B=1000) { purrr::map(1:B, function(i) {
  data.boot = data[sample(1:nrow(data), replace=TRUE),]
  summaries.boot = data.boot %>% group_by(D) %>% summarize(mean=mean(Y), sd=sd(Y), n=length(Y))
  model = lm(Y ~ D, data=data.boot)
  data.frame(dif=summaries.boot$mean[summaries.boot$D==100] - summaries.boot$mean[summaries.boot$D==50],
             slope=model$coefficients[2]*50)
}) %>% bind_rows()
}
boots = boot.distn(real.data)
```

```{r}
#| out-height: 250px
#| fig.asp: .25
#| layout-ncol: 2 
#| fig-subcap:
#|   - "Difference in Sample Means Sampling Dist"
#|   - "Slope-Based Sampling Dist" 

real.point.estimate = mean(real.data$Y[real.data$D==100]) - mean(real.data$Y[real.data$D==50])
x = seq(slope.point.estimate-7*se, slope.point.estimate+7*se, length.out=1000)
distribution.plot(boots$dif,   real.point.estimate) + xlim(min(x), max(x))
distribution.plot(boots$slope, slope.point.estimate) + xlim(min(x), max(x))
```






