---
title: "QTM 220: Lecture 12"
subtitle: "Causal Inference with a Multivalued Treatment"
format: 
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---


```{r load-libraries}
library(purrr)
library(tibble)
library(dplyr)
library(magrittr)
library(stringr)
library(ggplot2)
library(htmltools)
library(kableExtra)
library(scales)

percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }
```

```{r ggplot-themes}
class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
        panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x  = element_text(colour = "#aaaaaa"),
        axis.text.y  = element_text(colour = "#aaaaaa"),
        axis.title.x  = element_text(colour = "#aaaaaa"),
        axis.title.y  = element_text(colour = "#aaaaaa", angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
              axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
           panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(simplified.theme)
```

```{r table-styling}
ellipsis <- function(df, nhead=5L, ntail=2) {
    rownames(df) = as.character(1:nrow(df))
    stopifnot("data.frame" %in% class(df))
    els <- rep(NA, ncol(df)) %>% 
        matrix(nrow=1, dimnames=list(NULL, names(df))) %>% 
        data.frame(stringsAsFactors=FALSE) %>% 
        tibble::as_tibble()
    rownames(els) = "⋮"
    dplyr::bind_rows(head(df, nhead), els, tail(df, ntail))
}
```


## Today

::: {.hidden}
$$
\newcommand{\alert}[1]{\textcolor{red}{#1}}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\V}{V}
\DeclareMathOperator{\Var}{\V}
\newcommand{\inred}[1]{\textcolor{red}{#1}}
\newcommand{\inblue}[1]{\textcolor{blue}{#1}}
$$ 
:::

- We're going to talk about causal inference with a multivalued treatment.
- Our context will be a randomized experiment with three treatments.
- A large employer is experimenting with a new benefit for their employees: subsized gym memberships.
- They're randomizing employees they hire this year to three plans.
  - Plan 1: No gym membership. It costs them $\0/month.
  - Plan 2: A gym membership that covers entry but not classes. It costs \$50/month.
  - Plan 3: A gym membership that covers entry but and classes. It costs \$100/month.
- What they're interested in is how often employees go to the gym when provided these plans. 
- This idea is that they'll be able to use this information to decide what plan to offer everyone next year.

## Potential Outcomes and Treatment Effects 

- Potential Outcomes
  - There are three levels of treatment. We'll write them in terms of what the employer pays: $D_i=0,50,100$.
  - As a result, each employee has three potential outcomes: $Y_i(0)$, $Y_i(50)$, and $Y_i(100)$.
- Individual-Level Treatment Effect
  - There are two natural individual-level treatment effects to think about.
  - The effect of a \$50 gym membership vs none:  $\tau_i(50)  = Y_i(50) - Y_i(0)$
  - The effect of a \$100 gym membership vs \$50:  $\tau_i(100) = Y_i(100) - Y_i(50)$
- Observed Outcomes
  - We only observe one of these potential outcomes: $Y_i = Y_i(D_i)$
  - That's not enough to tell use about the value of any one of these effects.
- Randomization
  - The employer assigns its new employees to these three plans randomly, each with probability $1/3$.
  - As a result of this randomization, the distribution of potential outcomes is the same in each group.

## Causal Inference on the Population Level

- What we're really interested in is not the effects of these memberships on the people in our sample.
- We're thinking long-term---we're interested in the effects of these memberships on the people we *might* hire in the future.
- That's the population of interest. We're interested in a population-level causal effect.

:::: columns
::: column
### Sample

:::
::: column
### Population


:::
::::



- What we've just described is a sa

## A Table

$$
\begin{array}{ccc|ccc}
i & D_i & Y_i & Y_i(1) & Y_i(0) & \tau_i \\
\hline
1 & 1 & 70 & 70 & \alert{?} & \alert{?} \\
2 & 1 & 80 & 80 & \alert{?} & \alert{?} \\
3 & 0 & 60 & \alert{?} & 60 & \alert{?} \\
4 & 0 & 70 & \alert{?} & 70 & \alert{?} \\
5 & 0 & 70 & \alert{?}& 70 & \alert{?} \\
6 & 0 & 80 & \alert{?} & 80 & \alert{?} \\
\end{array}
$$


- Suppose we're administering ADHD patients at a clinic. 
    - Two are being treated with stimulants ($D=1$) 
    - Four are not ($D=0$). 
- We're interested in the treatment's effect on resting heart rate ($Y$).
- And we can't, of course, observe the counterfactual outcomes.
  - $Y_i(1)$ when $D_i=0$
  - $Y_i(0)$ when $D_i=1$
- This is what we see.

## Going Forward: From Potential Outcomes to Observed Outcomes

$$
\begin{array}{ccc|ccc}
i & D_i & Y_i & Y_i(1) & Y_i(0) & \tau_i \\
\hline
1 & ? & ?\  & \alert{70} & \alert{70} & \alert{0} \\
2 & ? & ?\  & \alert{80} & \alert{70} & \alert{10} \\
3 & ? & ?\  & \alert{70} & \alert{60} & \alert{10} \\
4 & ? & ?\  & \alert{80} & \alert{70} & \alert{10} \\
5 & ? & ?\  & \alert{70} & \alert{70} & \alert{0} \\
6 & ? & ?\  & \alert{80} & \alert{80} & \alert{0} \\
\end{array}
$$

- Let's think of our study as it unfolds over time.
- At the beginning, nobody's been assigned a treatment.
- But---as an idea---the potential outcomes already exist.
- And, as a result, so do the individual-level treatment effects.
- And all this stuff---the stuff to the right of the vertical bar---has nothing to do with treatment.

## Going Forward: From Potential Outcomes to Observed Outcomes

$$
\begin{array}{ccc|ccc}
i & D_i & Y_i & Y_i(1) & Y_i(0) & \tau_i \\
\hline
1 1 ? & ?\  & \alert{70} & \alert{70} & \alert{0} \\
2 1 ? & ?\  & \alert{80} & \alert{70} & \alert{10} \\
3 0 ? & ?\  & \alert{70} & \alert{60} & \alert{10} \\
4 0 ? & ?\  & \alert{80} & \alert{70} & \alert{10} \\
5 0 ? & ?\  & \alert{70} & \alert{70} & \alert{0} \\
6 0 ? & ?\  & \alert{80} & \alert{80} & \alert{0} \\
\end{array}
$$

- Next, we assign treatment.
- Outcomes "haven't happened yet", so they're not in our table, but we know what they will be.
- Let's fill them in. The assigned treatment $D_i$ tells us *which* potential outcome we observe.

## Going Forward: From Potential Outcomes to Observed Outcomes

$$
\begin{array}{ccc|ccc}
i & D_i & Y_i & Y_i(1) & Y_i(0) & \tau_i \\
\hline
1 & 1 & 70 & 70         & \alert{70} & \alert{0} \\
2 & 1 & 80 & 80         & \alert{70} & \alert{10} \\
3 & 0 & 60 & \alert{70} & 60 & \alert{10} \\
4 & 0 & 70 & \alert{80} & 70 & \alert{10} \\
5 & 0 & 70 & \alert{70} & 70 & \alert{0} \\
6 & 0 & 80 & \alert{80} & 80 & \alert{0} \\
\end{array}
$$

- Finally, we observe the outcomes.
- This reveals what *some* of the potential outcomes are. But not all of them.
- And it revals *none* of the individual-level treatment effects.

## Going Backward: From Observed Outcomes to Potential Outcomes

$$
\begin{array}{ccc|ccc}
i & D_i & Y_i & Y_i(1) & Y_i(0) & \tau_i \\
\hline
1 & 1 & 70 & 70 & \alert{70} & \alert{0} \\
2 & 1 & 80 & 80 & \alert{70} & \alert{10} \\
3 & 0 & 60 & \alert{70} & 60 & \alert{10} \\
4 & 0 & 70 & \alert{80} & 70 & \alert{10} \\
5 & 0 & 70 & \alert{70} & 70 & \alert{0} \\
6 & 0 & 80 & \alert{80} & 80 & \alert{0} \\
\end{array}
$$

- Now let's try to go backwards. In general, this is impossible.
- But let's suppose that the treatment was **randomized**. Then we can get part-way there.
- It means we can think of the treated and subsamples as being drawn from our population by a two-stage process.
  - First, we draw a random sample of people.
  - Then, for each, we flip a coin to decide whether it's in the subsample.
- This is **the same process** for both the treated and control groups.
- As a result, people in those two subsamples have the same distribution ...
  - if we ignore the stuff downstream of treatment assignment.
  - i.e. if we only look at what's to the right of the vertical bar.

## Going Backward: From Observed Outcomes to Potential Outcomes

$$
\begin{array}{ccc|ccc}
i & D_i & Y_i & Y_i(1) & Y_i(0) & \tau_i \\
\hline
1 & 1 & 70 & 70 & \alert{70} & \alert{0} \\
2 & 1 & 80 & 80 & \alert{70} & \alert{10} \\
3 & 0 & 60 & \alert{70} & 60 & \alert{10} \\
4 & 0 & 70 & \alert{80} & 70 & \alert{10} \\
5 & 0 & 70 & \alert{70} & 70 & \alert{0} \\
6 & 0 & 80 & \alert{80} & 80 & \alert{0} \\
\end{array}
$$

- In particular, it tells us that conditioning on treatment **does not affect** the potential outcome means.

$$ 
\begin{aligned}
\E[Y_i(1)|D_i=1] &= \E[\alert{Y_i(1)|D_i=0}] = \E[Y_i(1)] \\
\E[\alert{Y_i(0)|D_i=1}] &= \E[Y_i(0)|D_i=0] = \E[Y_i(0)] \\
\end{aligned}
$$

- And, because one of these potential outcomes is observed within each group, this has a very useful implication.
- The expectations of our potential outcomes coincide with *conditional* expectations **of observed outcomes**.
$$ 
\begin{aligned}
\E[Y_i\cancel{(1)}|D_i=1] &= \E[\alert{Y_i(1)|D_i=0}] = \E[Y_i(1)] \\
\E[\alert{Y_i(0)|D_i=1}] &= \E[Y_i\cancel{(0)}|D_i=0] = \E[Y_i(0)] \\
\end{aligned}
$$


## Implications for Causal Inference

:::: columns
::: column

- Randomization tells us that an real-world thing and a conceptual thing match
  - Real-world: within-group expectations 
  - Conceptual: counterfactual expectations
- Let's see what this tells us about our table.
:::
::: column

$$
\begin{aligned}
\E[Y_i(1)] &=\E[Y_i|D_i=1] \\
\E[Y_i(0)] &=\E[Y_i|D_i=0]
\end{aligned}
$$
:::
::::

\ \

### Let's see what this tells us about our table.

:::: columns
::: column
$$
\begin{array}{ccc|ccc}
i & D_i & Y_i & Y_i(1) & Y_i(0) & \tau_i \\
\hline
1 & 1 & 70 & 70 & \alert{70} & \alert{0} \\
2 & 1 & 80 & 80 & \alert{70} & \alert{10} \\
3 & 0 & 60 & \alert{70} & 60 & \alert{10} \\
4 & 0 & 70 & \alert{80} & 70 & \alert{10} \\
5 & 0 & 70 & \alert{70} & 70 & \alert{0} \\
6 & 0 & 80 & \alert{80} & 80 & \alert{0} \\
\end{array}
$$
:::
::: {.column .incremental} 

What is the average of ...

- $Y_i(1)$?
  - $\frac{70 + 80 +\alert{70} + \alert{80} +\alert{70} + \alert{80}}{6}  = 75$
- $Y_i(1)$ among people with $D=1$?
  - $\frac{70 + 80 }{2}  = 75$
- $Y_i$ among people with $D=1$?
  - $\frac{70 + 80 }{2}  = 75$
:::
::::



## Implications for Causal Inference

:::: columns
::: column

- Randomization tells us that an real-world thing and a conceptual thing match
  - Real-world: within-group expectations 
  - Conceptual: counterfactual expectations
:::
::: column

$$
\begin{aligned}
\E[Y_i(1)] &=\E[Y_i|D_i=1] \\
\E[Y_i(0)] &=\E[Y_i|D_i=0]
\end{aligned}
$$
:::
::::

\ \

### Let's see what this tells us about our table.

:::: columns
::: column
$$
\begin{array}{ccc|ccc}
i & D_i & Y_i & Y_i(1) & Y_i(0) & \tau_i \\
\hline
1 & 1 & 70 & 70 & \alert{70} & \alert{0} \\
2 & 1 & 80 & 80 & \alert{70} & \alert{10} \\
3 & 0 & 60 & \alert{70} & 60 & \alert{10} \\
4 & 0 & 70 & \alert{80} & 70 & \alert{10} \\
5 & 0 & 70 & \alert{70} & 70 & \alert{0} \\
6 & 0 & 80 & \alert{80} & 80 & \alert{0} \\
\end{array}
$$
:::
::: {.column .incremental} 

What is the average of ...

-  $Y_i(0)$?
  - $\frac{\alert{70} +\alert{70} +60+ 70 +70 + 80}{6}  = 70$
- $Y_i(0)$ among people with $D=0$?
   - $\frac{60+ 70 +70 + 80}{4}  = 70$
- $Y_i$ among people with $D=0$?
  - $\frac{60+ 70 +70 + 80}{4}  = 70$
:::
::::

## This is All Too Perfect

- This fake data is a little contrived.
- We're seeing stuff balance out perfectly in sample. It won't. 
- But we know that the difference between sample and population means is small in large samples.
  - In particular, it's proportional to $1/\sqrt{n}$
- So what we're seeing with our contrived data is close to what we'd really see.

## The ATE 

- It's the Difference in Average Potential Outcomes

$$
\begin{array}{ccc|ccc}
i & D_i & Y_i & Y_i(1) & Y_i(0) & \tau_i \\
\hline
1 & 1 & 70 & 70 & \alert{70} & \alert{0} \\
2 & 1 & 80 & 80 & \alert{70} & \alert{10} \\
3 & 0 & 60 & \alert{70} & 60 & \alert{10} \\
4 & 0 & 70 & \alert{80} & 70 & \alert{10} \\
5 & 0 & 70 & \alert{70} & 70 & \alert{0} \\
6 & 0 & 80 & \alert{80} & 80 & \alert{0} \\
\end{array}
$$

- What is the average $Y_i(1)$?
- 
$$
\frac{70 + 80 +\alert{70} + \alert{80} +\alert{70} + \alert{80}}{6}  = 75
$$

- What is the average $Y_i(0)$?
$$
\frac{\alert{70} +\alert{70} +60+ 70 +70 + 80}{6}  = 70
$$

- What is the average $\tau_i = Y_i(1) - Y_i(0)$?
$$
75-70 = \frac{\alert{0} + \alert{10} + \alert{10}+  \alert{10} +\alert{0} +\alert{0}}{6}  = 5
$$



## Implications for Linear Regression

```{r}
#| out-height: 250px
#| fig-asp: .25

eval$X = 0 + (eval$degree=='treated')
earnings.jitter = ggplot(eval) + geom_point(aes(x=X, y=income, color=degree), 
                              size=dot.size, position=position_jitter(seed=0,w=.3), alpha=.1) +
                   scale_y(limits=c(-.5e4,4e4)) + guides(color=FALSE)
earnings.summaries = eval %>% group_by(X) %>% summarize(mean=mean(income), sd=sd(income), meanX = mean(X))

earnings.jitter + geom_pointrange(aes(x=X, y=mean, ymin=mean-sd, ymax=mean+sd), data=earnings.summaries) + 
                  geom_line(aes(x=meanX, y=mean), linetype='dashed', data=earnings.summaries) 
```
$$
\hat a, \hat b = \operatorname*{argmin}_{a,b} \sum_{i=1}^n \qty{ aX_i + b - Y_i }^2
$$

- The slope $\hat b$ is just the difference in subsample means. As discussed in Lecture 1.
- So it's our difference-in-means estimate of the ATE. Nothing going on here.
- I just wanted to make the connection because people sometimes phrase it this way.



## Descriptive vs Causal and Sample vs Pop {.smaller}

:::: columns
::: column
$$
\tiny{
\begin{array}{ccc|ccc}
i & D_i & Y_i & Y_i(1) & Y_i(0) & \tau_i \\
\hline
\text{Sample} & \text{Descriptive} & \text{Statistics} & \text{Sample} & \text{Causal} & \text{Inference} \\
\hline
1 & 1 & 70 & 70 & ? & ? \\
2 & 1 & 80 & 80 & ? & ? \\
3 & 0 & 60 & ? & 60 & ? \\
4 & 0 & 70 & ? & 70 & ? \\
5 & 0 & 70 & ? & 70 & ? \\
6 & 0 & 80 & ? & 80 & ? \\
\hline
\text{Population} & \text{Descriptive} & \text{Inference} & \text{Population} & \text{Causal} & \text{Inference} \\
\hline
7 & ? & ? & ? & ? & ? \\
8 & ? & ? & ? & ? & ? \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\end{array}
}
$$
:::
::: column

- This table lays out 4 things we can work at.
 - Top Left: Describing the Sample
 - Bottom Left: Describing the Population
 - Top Right: Causal Inference about the Sample
 - Top Left: Causal Inference about the Population
- All we ever have is the sample.
- So we need to work out what it tells us about the rest.
- Obviously that involves assumptions.
  - Downward: Random Sampling
  - Rightward: Randomization of Treatment
- In class, we'll talk about everything but the top-right.
:::
::::

## What does the ATE tell us about Individual Effects?

Let's play CausalDoku! 

- Randomization implies that the *probability distribution* of both potential outcomes is the same in both groups
- Let's take this to an extreme.
  - We'll pretend it implies the *observed distribution* (i.e. the set of values) is the same
  - All we get to to fiddle with is *who* we give each value to
- Let's make this a puzzle.
  - The empty $Y_i(1)$ cells for $i=4,5,6$ must be filled with a 4 a 3 and a 2 (in some order)...
  - And the empty $Y_i(0)$ cells for $i = 1,2,3$ must be filled with a 3 and a 2 and a 1 (in some order).

$$
\begin{array}{ccc||c|c|c}
% Treatment & Outcome  & Outcome if Treated  & Outcome if Control  & Causal Effect  \\
i & D_i & Y_i & Y_i(1) & Y_i(0) & \tau_i=Y_i(1)-Y_i(0) \\
\hline
\hline
1&1 & 4 & 4 & &  \\
\hline
2&1 & 3 & 3 & & \\
\hline
3&1 & 2 & 2 & & \\
\hline
\hline
4&0 & 3 &   & 3 & \\
\hline
5&0 & 2 &   & 2 & \\
\hline
6& 0 & 1 &   & 1 & \\
\hline
\hline
\end{array}
$$




## CausalDoku 

Find two solutions:

- with constant treatment effects
- with both negative and positive effects


$$
\begin{array}{ccc||c|c|c}
% Treatment & Outcome  & Outcome if Treated  & Outcome if Control  & Causal Effect  \\
i & D_i & Y_i & Y_i(1) & Y_i(0) & \tau_i=Y_i(1)-Y_i(0) \\
\hline
\hline
1&1 & 4 & 4 & &  \\
\hline
2&1 & 3 & 3 & & \\
\hline
3&1 & 2 & 2 & & \\
\hline
\hline
4&0 & 3 &   & 3 & \\
\hline
5&0 & 2 &   & 2 & \\
\hline
6& 0 & 1 &   & 1 & \\
\hline
\hline
\end{array}
$$


- Note that both solutions imply the same ATE! 
 - We focus on ATEs because linearity of ATEs implies *we can* estimate them.
 - But on an individual level, these two scenarios are very different.
- With modern stuff, we can get part of the way but not all the way there...


