---
title: "QTM 220: Lecture 13"
subtitle: "Misspecification and Statistical Inference"
format: 
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---


```{r load-libraries}
library(purrr)
library(tibble)
library(dplyr)
library(magrittr)
library(stringr)
library(ggplot2)
library(htmltools)
library(kableExtra)
library(scales)

percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }
```

```{r ggplot-themes}
fontsize = 16
textcolor = "#888888"

class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
        panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x  = element_text(colour = textcolor, size=fontsize),
        axis.text.y  = element_text(colour = textcolor, size=fontsize),
        axis.title.x  = element_text(colour = textcolor),
        axis.title.y  = element_text(colour = textcolor, angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
              axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
           panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(simplified.theme)
```


```{r table-styling}
ellipsis <- function(df, nhead=5L, ntail=2) {
    rownames(df) = as.character(1:nrow(df))
    stopifnot("data.frame" %in% class(df))
    els <- rep(NA, ncol(df)) %>% 
        matrix(nrow=1, dimnames=list(NULL, names(df))) %>% 
        data.frame(stringsAsFactors=FALSE) %>% 
        tibble::as_tibble()
    rownames(els) = "⋮"
    dplyr::bind_rows(head(df, nhead), els, tail(df, ntail))
}
```

```{r fake-data}
set.seed(3)

# Sampling 250 employees. Our population is an even mix of ...
#   'blue guys', who go 1/4 days no matter what
#   'green guys', who go more if it's free but --- unlike the ones in our table earlier --- don't care about classes

n=250
is.blue = rbinom(n,1, 1/2)
Yblue   = rbinom(n, 30, 1/4) 
Ygreen0 = rbinom(n, 30, 1/10)
Ygreen50 = rbinom(n, 30, 1/5)

Y0   = ifelse(is.blue, Yblue, Ygreen0)
Y50  = ifelse(is.blue, Yblue, Ygreen50)
Y100 = Y50   ### <--- this tells us we've got no effect.

# Randomizing by rolling two dice. R is their sum
R = sample(c(1:6), size=n, replace=TRUE) + sample(c(1:6), size=n, replace=TRUE)
D = case_match(R, 1:6~0, 7:10~50, 11:12~100)
Y = case_match(D, 0~Y0, 50~Y50, 100~Y100)

# Randomizing by rolling two dice. R is their sum
R = sample(c(1:6), size=n, replace=TRUE) + sample(c(1:6), size=n, replace=TRUE)
D = case_match(R,
  1:6~0,
  7:10~50,
  11:12~100)
Y = case_match(D,
  0~Y0,
  50~Y50,
  100~Y100)

real.data = tibble(D=D, Y=Y)
summaries = tibble(D=D, Y=Y) %>% group_by(D) %>% summarize(mean=mean(Y), sd=sd(Y), n=length(Y))

scatter = ggplot() + geom_point(aes(x=D, y=Y), color="red", data=tibble(D=D,Y=Y),
                                     position=position_jitter(w=10,h=0, seed=0), alpha=.3) +
           geom_pointrange(aes(x=D, y=mean, ymin=mean-2*sd/sqrt(n), ymax=mean+2*sd/sqrt(n)), data=summaries, alpha=.8) +
           scale_x_continuous(breaks=c(0,50,100)) + labs(x="D", y="Y") + ylim(-3,15)

model=lm(Y ~ D, data=tibble(D=D, Y=Y))
predictions = tibble(D=c(0,50,100), muhat=predict(model, newdata=data.frame(D=c(0,50,100))))

dif.point.estimate = mean(Y[D==100]) - mean(Y[D==50])
slope.point.estimate = predictions$muhat[3]-predictions$muhat[2]
``` 

```{r}
boot.distn = function(data, B=1000) { 
  rows = purrr::map(1:B, function(i) {
    data.boot = data[sample(1:nrow(data), replace=TRUE),]
    summaries.boot = data.boot %>% group_by(D) %>% summarize(mean=mean(Y), sd=sd(Y), n=length(Y))
    model = lm(Y ~ D, data=data.boot)
    muhat = predict(model, newdata=data.frame(D=c(0,50,100)))

    tauhat.dif = mean(data.boot$Y[data.boot$D==100]) - mean(data.boot$Y[data.boot$D==50]) 
    tauhat.slope = muhat[3]-muhat[2]
    data.frame(dif=tauhat.dif, slope=tauhat.slope)
  })
  rows %>% bind_rows() 
}
```

```{r plotting-code}
width = function(samples, center=mean(samples), alpha=.05) {
  covg.dif = function(w) { 
    actual = mean( center - w/2 <= samples & samples <= center + w/2)
    nominal = 1-alpha 
    actual - nominal  
  }
  uniroot(covg.dif, interval=c(0, 2*max(abs(samples-center))))$root
}

distribution.plot = function(boot.samples, point.estimate, w=width(boot.samples), interval.replications=0, point.replications=0) { 
  drange = c(0, dnorm(0, mean=0, sd=sd(boot.samples)))
  p = ggplot(data.frame(samples=boot.samples)) + 
    geom_histogram(aes(x=samples, y=after_stat(density)), alpha=.3, bins=50)  +
    geom_pointrange(aes(y=dnorm(2*sd(samples), mean=0, sd=sd(samples)), 
      x=point.estimate, xmin=point.estimate - w/2, xmax=point.estimate + w/2)) + 
      geom_vline(xintercept=mean(boot.samples), color='blue') + 
      geom_vline(xintercept=quantile(boot.samples, c(.025,.975)), color='blue', linetype='dotted') 
  if(interval.replications > 0) { 
    K = interval.replications
    p = p + geom_pointrange(aes(y=y, x=x, xmin=xmin, xmax=xmax), 
          data=data.frame(y=seq(min(drange), max(drange), length.out=K), 
                  x=boot.samples[1:K], xmin=boot.samples[1:K] - w/2, xmax=boot.samples[1:K] + w/2), color='purple', alpha=.2)
  }
  if(point.replications > 0) { 
    K = point.replications
    p = p + geom_point(aes(y=y, x=x), 
          data=data.frame(y=seq(min(drange), max(drange), length.out=K), 
                  x=boot.samples[1:K], xmin=boot.samples[1:K] - w/2, xmax=boot.samples[1:K] + w/2), color='purple', alpha=.2)
  }
  p
}
```


# Review

## Estimating An Average Treatment Effect

::: {.hidden}
$$
\newcommand{\alert}[1]{\textcolor{red}{#1}}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\V}{V}
\DeclareMathOperator{\Var}{\V}
\DeclareMathOperator*{\argmin}{argmin}
\newcommand{\inred}[1]{\textcolor{red}{#1}}
\newcommand{\inblue}[1]{\textcolor{blue}{#1}}
$$ 
:::

```{r}
#| out-height: 400px
#| fig-asp: .4
scatter 
```


- We reduced the problem to estimating a difference in subpopulation means.
- And we tried estimating that using a difference in subsample means. 
- I've drawn these, $\pm$ two standard errors, in black.

$$
\tau(100) = \frac{1}{m_{100}}\sum_{j:d_j=100} y_j - \frac{1}{m_{50}}\sum_{j:d_j=50} y_j 
          \approx \hat\tau(100) = \frac{1}{N_{100}}\sum_{i:D_i=100} Y_i - \frac{1}{N_{50}}\sum_{i:D_i=50} Y_i
$$          


## Dissapointing Results

```{r}
#| out-height: 400px
#| fig-asp: .4
scatter 
```

- We weren't happy with the resulting precision. 
    - The problem was that we had too few observations with the expensive treatment ($D_i=100$).
    - This made the subsample mean a high-variance estimate of the corresponding subpopulation mean.

$$
\Var[\hat\tau(100)] = \Var[\hat\mu(50)] + \Var[\hat\mu(100)] \ge \Var[\hat\mu(100)] = \frac{\Var[Y \mid D_i=100]}{N_{100}} \approx `r approx(var(Y[D==100])/sum(D==100))` \approx `r approx(sqrt(var(Y[D==100])/sum(D==100)))`^2
$$

## Uninformative Interval Estimates 

```{r}
#| out-height: 400px
#| fig-asp: .4
scatter 
```

- Our confidence intervals, whether calibrated via bootstrap or normal approximation, are wide.
  - One thing people find notable is that they contain both positive and negative numbers.
  - This means we can't say, with confidence, whether our treatment effect is positive or negative. 

```{r}
#| out-height: 200px
#| fig.asp: .2
#| layout-ncol: 2 
#| fig-subcap:
#|   - "Bootstrap Sampling Distribution for the Difference-in-Means"
#|   - "Normal Approximation to the Sampling Distribution for the Difference-in-Means"

point.estimate = dif.point.estimate
se = sqrt( var(Y[D==100])/sum(D==100) + var(Y[D==50])/sum(D==50) )
x = seq(point.estimate-4*se, point.estimate+4*se, length.out=1000)
normal.distribution.plot = ggplot() + 
    geom_area(aes(x=x, y=dnorm(x, mean=point.estimate, sd=se)), color='purple', fill='purple', alpha=.08) +
    geom_vline(aes(xintercept=point.estimate), color='purple', linetype='solid') +
    geom_vline(aes(xintercept=point.estimate + c(-1,1)*1.96*se), color='purple', linetype='dotted') +
    geom_pointrange(aes(y=dnorm(2*se, mean=0, sd=se), x=point.estimate, xmin=point.estimate - 1.96*se, xmax=point.estimate + 1.96*se))

boots = boot.distn(real.data)
limits = c(min(x), max(x))
breaks = floor(min(limits)):ceiling(max(limits))
scale_x = scale_x_continuous(breaks=breaks, limits=limits)
distribution.plot(boots$dif, point.estimate) + scale_x
normal.distribution.plot + scale_x
```


## An Attempt At Being Clever

```{r}
#| out-height: 400px
#| fig-asp: .4
scatter.with.lsq = scatter +  
  geom_line(aes(x=D, y=muhat), color='blue', linetype='dashed', data=predictions) + 
  geom_point(aes(x=D,y=muhat), color='blue', data=predictions)
scatter.with.lsq
```

- To work around this, we tried estimating our subpopulation means using the [least-squares line]{.fg style="color:blue"}. 
$$
\hat\tau(100) = \hat\mu(100) - \hat\mu(50) \qqtext{where} \hat\mu(x) = \hat a x + \hat b \qfor \hat a,\hat b = \argmin_{a,b} \sum_{i=1}^n \qty(aX_i + b - Y_i)^2
$$ 

## An Attempt At Being Clever
```{r}
#| out-height: 400px
#| fig-asp: .4
scatter.with.lsq = scatter +  
  geom_line(aes(x=D, y=muhat), color='blue', linetype='dashed', data=predictions) + 
  geom_point(aes(x=D,y=muhat), color='blue', data=predictions)
scatter.with.lsq
```

:::: columns
::: {.column width="40%"}
```{r}
#| out-height: 200px
#| fig.asp: .2
#| layout-nrow: 2 
#| fig-subcap:
#|   - "Difference in Means Bootstrap Sampling Distribution"
#|   - "Fitted-Line-Based Bootstrap Sampling Distribution" 

distribution.plot(boots$dif,   dif.point.estimate) + scale_x
distribution.plot(boots$slope, slope.point.estimate)  + scale_x
```

:::
::: {.column width="60%"}

- This *did* reduce variance. 
    - There are lots of observations for cheaper treatments.
    - These helped us get a more precise estimate of the line's slope.
- But the line gave us pretty different estimates of the subpopulation means.
    - We were higher at $D_i=100$ and low at $D_i=50$.
    - So the difference, high - low, was bigger.
:::
::::


## Inconsistent Interval Estimates

```{r}
#| out-height: 200px
#| fig.asp: .2
#| layout-nrow: 2 
#| fig-subcap:
#|   - "Difference in Means Bootstrap Sampling Distribution"
#|   - "Fitted-Line-Based Bootstrap Sampling Distribution" 

distribution.plot(boots$dif,   dif.point.estimate) + scale_x
distribution.plot(boots$slope, slope.point.estimate)  + scale_x
```

- These estimates are so different that our confidence intervals lead to meaningfully different claims.
- In particular, they say different things about whether out treatment effect is positive.
- The difference in subsample means approach would say we don't know. 
    - A bootstrap-calibrated confidence interval based on that estimate covers zero.
- The fitted line approach would say we do know that it's positive. 
    - A bootstrap-calibrated confidence interval based on that approach is to the right of zero.


## That's Wrong

- This is fake data, so we can see that by looking at the code generating it.
- In the data we're looking at ...
  - The actual effect is exactly zero.
  - The potential outcomes $Y_i(50)$ and $Y_i(100)$ are *exactly the same*.

```{r}
#| eval: false
#| echo: true

# Sampling 250 employees. Our population is an even mix of ...
#   'blue guys', who go 1/4 days no matter what
#   'green guys', who go more if it's free but --- unlike the ones in our table earlier --- don't care about classes

n=250
is.blue = rbinom(n,1, 1/2)
Yblue   = rbinom(n, 30, 1/4) 
Ygreen0 = rbinom(n, 30, 1/10)
Ygreen50 = rbinom(n, 30, 1/5)

Y0   = ifelse(is.blue, Yblue, Ygreen0)
Y50  = ifelse(is.blue, Yblue, Ygreen50)
Y100 = Y50   ### <--- this tells us we've got no effect.

# Randomizing by rolling two dice. R is their sum
R = sample(c(1:6), size=n, replace=TRUE) + sample(c(1:6), size=n, replace=TRUE)
D = case_match(R, 1:6~0, 7:10~50, 11:12~100)
Y = case_match(D, 0~Y0, 50~Y50, 100~Y100)
``` 

## The Line-Based Estimator is **Biased**

```{r}

line.based.estimates = purrr::map(1:1000, function(b) {
  n=250
  is.blue = rbinom(n,1, 1/2)
  Yblue   = rbinom(n, 30, 1/4) 
  Ygreen0 = rbinom(n, 30, 1/10)
  Ygreen50 = rbinom(n, 30, 1/5)

  Y0   = ifelse(is.blue, Yblue, Ygreen0)
  Y50  = ifelse(is.blue, Yblue, Ygreen50)
  Y100 = Y50   ### <--- this tells us we've got no effect.

  # Randomizing by rolling two dice. R is their sum
  R = sample(c(1:6), size=n, replace=TRUE) + sample(c(1:6), size=n, replace=TRUE)
  D = case_match(R, 1:6~0, 7:10~50, 11:12~100)
  Y  = case_match(D, 0~Y0, 50~Y50, 100~Y100)

  model=lm(Y ~ D, data=tibble(D=D, Y=Y))
  mu.hat = predict(model, newdata=data.frame(D=c(0,50,100)))
  data.frame(estimate =  mu.hat[3]-mu.hat[2], se = sqrt(vcov(model)[2,2])*50)
}) %>% bind_rows()

```

```{r}
#| out-height: 400px
#| fig-asp: .4
#| 
line.based.estimates$i = 1:nrow(line.based.estimates)
slope.se = sqrt(vcov(model)[2,2])*50
line.sampling.dist = ggplot() + 
  geom_histogram(aes(x=estimate, y=after_stat(density)), bins=50, alpha=.2, data=line.based.estimates) + 
  geom_vline(aes(xintercept=mean(estimate)), color='blue', data=line.based.estimates) +
  geom_pointrange(aes(x=slope.point.estimate, y=dnorm(2*slope.se, mean=0, sd=slope.se), 
                      xmin=slope.point.estimate - 1.96*slope.se, 
                      xmax=slope.point.estimate + 1.96*slope.se)) + 
                      geom_vline(aes(xintercept=0), color='green') + scale_x 
line.sampling.dist
``` 

- We can do a little simulation to see that.
- We'll generate 1000 fake datasets like the one we've been looking at.
  - And calculate our estimator for each one. 
  - This gives us draws from our estimator's *actual* sampling distribution. 
  - I've used them to draw a histogram.
- The mean of these point estimates is $`r approx(mean(line.based.estimates$estimate))`$. 
    - Since our actual effect is zero, that's our bias. The blue line.
- Our actual point estimate is off in the same direction---but a bit more than usual.
  - It's shown in black with a corresponding 95% interval.
  - An interval that fails to cover the actual effect of zero. The green line.

## The Resulting Coverage Is **Bad**

```{r}
#| out-height: 400px
#| fig-asp: .4
#| 
covers =  abs(line.based.estimates$estimate) <= 1.96*line.based.estimates$se 
line.sampling.dist + geom_pointrange(aes(x=estimate, y=(8*i/100)*dnorm(2*slope.se, mean=0, sd=slope.se), 
                      xmin=estimate - 1.96*se, 
                      xmax=estimate + 1.96*se), data=line.based.estimates[1:100,], color='purple', alpha=.3) 
```

- We can construct 95% confidence intervals the same way in each fake dataset.
  - I've drawn 100 of these intervals in purple. 
  - You can see that they sometimes cover zero, but not as often as you'd like.
- `r sum(covers)` / `r nrow(line.based.estimates)` &asymp; `r percent(mean(covers))` of them cover zero. That's not good. 
- If that's typical of our analysis, people really shouldn't trust us. 
- So why have statisticians gotten away with fitting lines for so long?
- In essence, it's by pretending we were trying to do something else.

## What We Pretend To Do

```{r}
#| out-height: 400px
#| fig-asp: .4

dicepmf = (1/36) * c((1:7)-1, 13-(8:12))
px = c(sum(dicepmf[1:6]), sum(dicepmf[7:10]), sum(dicepmf[11:12]))
mu = (1/2)*(30/4) + (1/2)*c(30/10, 30/5, 30/5)
model.pop = lm(Y~D, data=tibble(D=c(0,50,100), Y=mu), weights=px)
mutilde = predict(model.pop, newdata=tibble(D=c(0,50,100)))
pop.summary = 50*coef(model.pop)[2]

covers =  abs(line.based.estimates$estimate-pop.summary) <= 1.96*line.based.estimates$se 
line.sampling.dist + geom_pointrange(aes(x=estimate, y=(8*i/100)*dnorm(2*slope.se, mean=0, sd=slope.se), 
                      xmin=estimate - 1.96*se, 
                      xmax=estimate + 1.96*se), data=line.based.estimates[1:100,], color='purple', alpha=.4) + 
                      geom_vline(aes(xintercept=pop.summary), color='red') +
                      scale_x 
```

- Our estimator --- 50 times the slope of the least squares line --- *is* a good estimator of something.
- It's a good estimate of the analogous population summary --- 50 times the *population least squares slope*. The red line.
- `r sum(covers)` / `r nrow(line.based.estimates)` &asymp; `r percent(mean(covers))` of our intervals cover it. That makes it sound pretty good. 

:::: columns
::: column

$$
\small{
\begin{aligned}
\hat\tau(100) &= \hat\mu(100) - \hat\mu(50) = \qty(100 \hat a + \hat b) - \qty(50 \hat a + \hat b) = 50\hat a \\
\qqtext{where} & \hat\mu(x) = \hat a x + \hat b \qfor \hat a,\hat b = \argmin_{a,b} \sum_{i=1}^n \qty(aX_i + b - Y_i)^2
\end{aligned}
}
$$ 

:::
::: column

$$
\small{
\begin{aligned}
\tilde\tau(100) &= \tilde\mu(100) - \tilde\mu(50)) = \qty(100 \tilde a + \tilde b) - \qty(50 \tilde a + \tilde b) = 50\tilde a \\ 
\qqtext{where} & \tilde\mu(x) = \tilde a x + \tilde b \qfor \tilde a,\tilde b = \argmin_{a,b} \E\qty[\qty(aX_i + b - Y)^2]
\end{aligned}
}
$$

:::
::::


## **Problem**: That's Not What We Wanted To Estimate

```{r}
#| out-height: 400px
#| fig-asp: .4

pop.predictions = tibble(D=c(0,50,100), mu=mutilde) 
pop.summaries = tibble(D=c(0,50,100), mu=mu)
pop.scatter = scatter.with.lsq +
          geom_line(aes(x=D, y=mutilde), color='red', linetype='dashed',  pop.predictions) + 
          geom_point(aes(x=D, y=mutilde), color='red',  pop.predictions) + 
          geom_point(aes(x=D, y=mu), color='green', size=3, data=pop.summaries) +
          geom_line(aes(x=D, y=mu), color='green', data=pop.summaries) + coord_cartesian(ylim=c(2,10))
pop.scatter
```

- The effects we want to estimate are, as a result of randomization, differences in subpopulation means.
$$
\tau(50) = \mu(50) - \mu(0) \qand \tau(100) = \mu(100) - \mu(50) 
$$

- The population [least squares line]{.fg style="color:red"} *doesn't go through these means*. It can't. 
- They don't lie along a line. They lie along a [hockey-stick shaped curve]{.fg style="color:LimeGreen"}. 
- This means no line---including the population least squares line---can tell us what we want to know.
- This problem has a history of getting buried in technically correct but opaque language. 


## Language Games

```{r}
#| out-height: 400px
#| fig-asp: .4

pop.scatter
```


- The widespread use of causal language is a recent development.
- In the past, people would do a little linguistic dance to avoid talking about causality.
- They'd talk about what they were really estimating as if it were what you wanted to know.
- This not only buried issues of causality, it buried issues of bias due to *misspecification*.
  - e.g., due to trying to use a line to estimate a hockey-stick shaped curve.
- They'd say 'The regression coefficient for $Y$ on $D$ is $`r approx(coef(model)[2])`$.'
  - You'd be left to intepret this as saying the treatment effect was roughly $50 \times `r approx(coef(model)[2])`$.
  - That's wrong, but it's your fault for interpreting it that way. **They set you up**.

# Today

## We'll Think About Misspecification

```{r}
#| out-height: 400px
#| fig-asp: .4

line.sampling.dist + geom_pointrange(aes(x=estimate, y=(8*i/100)*dnorm(2*slope.se, mean=0, sd=slope.se), 
                      xmin=estimate - 1.96*se, 
                      xmax=estimate + 1.96*se), data=line.based.estimates[1:100,], color='purple', alpha=.4) + 
                      geom_vline(aes(xintercept=pop.summary), color='red') +
                      scale_x 
```


- In particular, we'll think about its implications for statistical inference.
- This boils down to what happens when our estimator's sampling distribution
is not centered on what we want to estimate.
- We can see that, if our sampling distribution is narrow enough, the coverage of interval estimates is *awful*.
- Today, we'll think about how sample size impacts that. Something interesting happens. 
- From a point-estimation perspective, more data is always good. Our estimators do get more accurate.
- But from an inferential perspective, it can be very bad. When we use misspecified moedls, our coverage claims will get increasingly inaccurate.

## The Impact of Misspecification as Sample Size Grows

```{r}

sampling.dist = function(n) { 
  purrr::map(1:1000, function(b) {
  is.blue = rbinom(n,1, 1/2)
  Yblue   = rbinom(n, 30, 1/4) 
  Ygreen0 = rbinom(n, 30, 1/10)
  Ygreen50 = rbinom(n, 30, 1/5)

  Y0   = ifelse(is.blue, Yblue, Ygreen0)
  Y50  = ifelse(is.blue, Yblue, Ygreen50)
  Y100 = Y50   ### <--- this tells us we've got no effect.

  # Randomizing by rolling two dice. R is their sum
  R = sample(c(1:6), size=n, replace=TRUE) + sample(c(1:6), size=n, replace=TRUE)
  D = case_match(R, 1:6~0, 7:10~50, 11:12~100)
  Y  = case_match(D, 0~Y0, 50~Y50, 100~Y100)

  model = lm(Y~D)
  mu.hat = predict(model, newdata=data.frame(D=c(0,50,100)))
  rbind(data.frame(estimate =  mu.hat[3]-mu.hat[2], se = sqrt(vcov(model)[2,2])*50, type='line'),
        data.frame(estimate = mean(Y[D==100])-mean(Y[D==50]), se = sqrt(var(Y[D==100])/sum(D==100) + var(Y[D==50])/sum(D==50)), type='dif'))
}) %>% bind_rows()
}

bootstrap.sampling.dist.plot = function(estimates, type='histogram', integration.range=NULL) { 
  drange = c(0, dnorm(0, mean=0, sd=sd(estimates$estimate, na.rm=TRUE)))
  if(type=='histogram') { 
    base = ggplot() + geom_histogram(aes(x=estimate, y=after_stat(density)), bins=50, alpha=.2, data=estimates) 
  } else { 
    x = mean(estimates$estimate,na.rm=T) + seq(-1,1,by=.01)*5*sd(estimates$estimate,na.rm=T)
    normal.approx = data.frame(x=x, fx=dnorm(x, mean=mean(estimates$estimate,na.rm=T), sd=sd(estimates$estimate,na.rm=T)))
    base = ggplot() + geom_area(aes(x=x, y=fx), data=normal.approx, color='purple', fill='purple', alpha=.08)
    if(!is.null(integration.range)) { 
      in.range = integration.range[1] <= x & x <= integration.range[2]
      base = base +  geom_area(aes(x=x, y=fx), data=normal.approx[in.range,], color=NA, fill='yellow', alpha=.4)
    }
    estimates$se = sd(estimates$estimate,na.rm=T)
  }
  base +
    geom_vline(aes(xintercept=mean(estimate)), color='blue', data=estimates) +
    geom_pointrange(aes(x=estimate, y=dnorm(2*se, mean=0, sd=se), 
                        xmin=estimate - 1.96*se, 
                        xmax=estimate + 1.96*se), 
                        data=estimates[1,]) +
    geom_pointrange(aes(y=seq(min(drange), max(drange), length.out=99), x=estimate, 
                        xmin=estimate - 1.96*se, 
                        xmax=estimate + 1.96*se), 
                        data=estimates[2:100,], alpha=.1, color='purple') +        
    geom_vline(aes(xintercept=0), color='green') + geom_vline(aes(xintercept=pop.summary), color='red') 
}
```


```{r}
#| layout-nrow: 3
#| layout-ncol: 2 
#| out-height: 250px
#| fig-asp: .25
set.seed(10)

limits = c(-5,5)
breaks = floor(min(limits)):ceiling(max(limits))
scale_x = scale_x_continuous(breaks=breaks, limits=limits) 
estimates = sampling.dist(50)
bootstrap.sampling.dist.plot(estimates[estimates$type=='dif',]) + scale_x
bootstrap.sampling.dist.plot(estimates[estimates$type=='line',]) + scale_x
estimates50 = estimates
estimates50$n = 50

estimates = sampling.dist(100)
bootstrap.sampling.dist.plot(estimates[estimates$type=='dif',]) + scale_x
bootstrap.sampling.dist.plot(estimates[estimates$type=='line',]) + scale_x
estimates100 = estimates
estimates100$n = 100

estimates = sampling.dist(200)
bootstrap.sampling.dist.plot(estimates[estimates$type=='dif',]) + scale_x
bootstrap.sampling.dist.plot(estimates[estimates$type=='line',]) + scale_x
estimates200 = estimates
estimates200$n = 200
```

- Here we're seeing the sampling distributions of our two estimators at three sample sizes.
  - Left/Right: Difference in Subsample Means / Difference in Values of Fitted Line
  - Top/Middle/Bottom: Sample size 50,100,200
- The green line is the actual effect; the red one is the effect estimate we get using the population least squares line.
- The red one is the effect estimate we get using the population least squares line: 50 times the slope of the red line.
- I've drawn 100 confidence intervals based on normal approximation for each estimator.
- We can see that the line-based estimator's coverage gets increasingly bad as sample size increases. 

## Coverage

```{r}
#| out-height: 500px
#| 
estimates = rbind(estimates50, estimates100, estimates200)
covg = estimates %>% group_by(n,type) %>% summarize(coverage = mean(abs(estimate) <= 1.96*se, na.rm=T))

ggplot(covg) + geom_line(aes(x=n, y=coverage, color=type)) + geom_point(aes(x=n, y=coverage, color=type))  + 
    scale_y_continuous(limits=c(0,1), breaks=seq(0,1,by=.05)) + labs(y='Coverage', x='Sample Size') +
    geom_hline(aes(yintercept=.95), color='black', linetype='dashed') 
```

- Here are the coverage rates for each estimator at each sample size.
- The difference in sample means works more or less as we'd like.
  - Coverage is always pretty good and is roughly 95% --- what we claim --- in larger samples.
- The line-based estimator does the opposite.
  - Coverage is bad in small samples and worse in larger ones.
- Now, we're going to do some exercises. We're going to work out how to calculate coverage by hand.
  - The point is to get some perspective on what it means to have 'just a little bit of bias'.
  - This'll come in handy when people around you say a bit of misspecification isn't a big deal.

# Exercises

## Set Up

```{r}
#| layout-ncol: 2
#| output-height: 400px
limits = c(-2,3)
breaks = floor(min(limits)):ceiling(max(limits))
scale_x = scale_x_continuous(breaks=breaks, limits=limits) 
bootstrap.sampling.dist.plot(estimates[estimates$type=='line' & estimates$n == 50,]) + scale_x
bootstrap.sampling.dist.plot(estimates[estimates$type=='line' & estimates$n == 50,], type='normal') + scale_x
```

- Above, I've drawn the sampling distribution for the estimator $\hat\tau$ and a normal approximation to it.
- We're going to calculate that estimator's coverage rate. Approximately. We'll make some simplifications.
- We'll act as if the normal approximation to the sampling distribution were perfect.
- We'll assume our intervals are perfectly calibrated.
  - i.e., for a draw $\hat\tau$ from the sampling distribution ...
  - the interval is $\hat\tau \pm 1.96\sigma$ where $\sigma$ is our estimator's actual standard error.


## Exercise 1

```{r}
#| output-height: 300px
#| fig-asp: .3
se = sd(estimates[estimates$type=='line' & estimates$n == 50,]$estimate)
bootstrap.sampling.dist.plot(estimates[estimates$type=='line' & estimates$n == 50,], type='normal') + scale_x +
  geom_vline(aes(xintercept=x), data=tibble(x=1.96*se*c(-1,1)), color='green', linetype='dashed', size=1)
```

- Let's think about things from the treatment effect's perspective. The solid green line. 
- Let's give it arms the same length as our intervals have. They go out to the dashed green lines.
- An interval's arms touch the effect if and only if the effect touches the interval's center, i.e., the corresponding point estimate.
- So we're calculating the proportion of point estimates covered by *the treatment effect's arms*.
- **Exercise**. Write this in terms of the sampling distribution's [probability density f(x)]{.fg color="purple"}
  - This is a calculation you can illustrate with a sketch on top of the plot above. You may want to start there.

## Exercise 1

```{r}
#| output-height: 300px
#| fig-asp: .3
se = sd(estimates[estimates$type=='line' & estimates$n == 50,]$estimate)
bootstrap.sampling.dist.plot(estimates[estimates$type=='line' & estimates$n == 50,], 
                             type='normal', integration.range=1.96*se*c(-1,1)) +
  geom_vline(aes(xintercept=x), data=tibble(x=1.96*se*c(-1,1)), color='green', linetype='dashed', size=1) + scale_x
```

- To calculate the proportion of point estimates covered by *the treatment effect's arms* 
- We integrate the sampling distribution's probability density from one to the other.

$$
\text{coverage} = \int_{-1.96\sigma}^{1.96\sigma} f(x) dx \qqtext{ where } f(x)=\frac{1}{\sqrt{2\pi} \sigma} e^{-\frac{(x-\tilde\tau)^2}{2\sigma^2}}
$$

- Here I've written in the actual formula of the density of the normal I drew.
  - The one with mean $\tilde\tau$ and standard deviation $\sigma$.
  - I'm not expecting you to have that memorized, but it does come in handy from time to time.
- This doesn't yet give us all that much *generalizable* intution. 
- But we'll derive some by, to some degree, simplifying that integral.

## Exercise 2

```{r}
#| output-height: 300px
#| fig-asp: .3
se = sd(estimates[estimates$type=='line' & estimates$n == 50,]$estimate)
bootstrap.sampling.dist.plot(estimates[estimates$type=='line' & estimates$n == 50,], 
                             type='normal', integration.range=1.96*se*c(-1,1)) +
  geom_vline(aes(xintercept=x), data=tibble(x=1.96*se*c(-1,1)), color='green', linetype='dashed', size=1) + scale_x
```

- Now we'll rewrite our coverage as an integral of the [standard normal density]{.fg style="color:blue"}
- To do it, we're going back to basic calculus. We're making a *substitution* $u=(x-\tilde\tau)/\sigma$.
- This means we have to change the limits.
  - As $x$ goes from $-1.96\sigma$ to $1.96\sigma$ ...
  - $u$ goes from $-1.96-\tilde\tau/\sigma$ to $1.96-\tilde\tau/\sigma$.
- And make a substitution into the integrand.
  - For $(x-\tilde\tau)^2/\sigma^2$, we subtitute $u^2$.
- And multiply by $dx/du$ so we get to integrate over $u$.
  - $x= u\sigma + \tilde\tau$ so $dx=\sigma du$

:::: r-stack
::: {.fragment .fade-out data-fragment-index="0"}
$$
\text{coverage} = \int_{-1.96\sigma}^{1.96\sigma} e^{-\frac{(x-\tilde\tau)^2}{2\sigma^2}} dx 
\phantom{ =\int_{-1.96-\tilde\tau/\sigma}^{1.96-\tilde\tau/\sigma} \frac{1}{\sqrt{2\pi} \sigma} e^{-\frac{u^2}{2}} \sigma du = \int_{-1.96-\tilde\tau/\sigma}^{1.96+\tilde\tau/\sigma} \textcolor{blue}{\frac{1}{\sqrt{2\pi}} e^{-\frac{u^2}{2}}} du} 
$$

:::
::: {.fragment .current-visible data-fragment-index="0"}
$$
\text{coverage} 
= \int_{-1.96\sigma}^{1.96\sigma} e^{-\frac{(x-\tilde\tau)^2}{2\sigma^2}} dx \\
= \int_{-1.96-\tilde\tau/\sigma}^{1.96+\tilde\tau/\sigma} \frac{1}{\sqrt{2\pi} \sigma} e^{-\frac{u^2}{2}} \sigma du \\
= \int_{-1.96-\tilde\tau/\sigma}^{1.96+\tilde\tau/\sigma} \textcolor{blue}{\frac{1}{\sqrt{2\pi}} e^{-\frac{u^2}{2}}} du 
$$

:::
::::

## Implications

```{r}
#| layout-nrow: 2
#| output-height: 300px
#| fig-asp: .3
lines50 = estimates[estimates$type=='line' & estimates$n == 50,]
se = sd(lines50$estimate, na.rm=T)
bootstrap.sampling.dist.plot(lines50, type='normal', integration.range=1.96*se*c(-1,1)) +
  geom_vline(aes(xintercept=x), data=tibble(x=1.96*se*c(-1,1)), color='green', linetype='dashed', size=1) + scale_x

tau.tilde = pop.summary
x = mean(lines50$estimate, na.rm=T)/se + seq(-1,1,by=.01)*5
xin = x[ -1.96 - tau.tilde/se <= x & x <= 1.96 - tau.tilde/se ]
breaks = sort(c(seq(-3,2,by=1), c(-1.96-tau.tilde/se, 1.96-tau.tilde/se)))
names(breaks) = NULL
# TODO: label nicely
scale_x_standardized = scale_x_continuous(breaks=breaks, limits=(limits-tau.tilde)/se) 
ggplot() + geom_area(aes(x=x, y=dnorm(x)), color='purple', fill='purple', alpha=.2) +
  geom_area(aes(x=xin, y=dnorm(xin)), color=NA, fill='yellow', alpha=.4) + scale_x_standardized
```


