---
title: "Exam 2 (November 9/13)"
subtitle: "QTM 220"
execute:
  echo: false
  warning: false
format: 
  pdf:
    include-in-header: 
      text: |
        \usepackage{amsmath}
        \usepackage{setspace}
        \onehalfspacing
        \usepackage{amssymb}
        \usepackage{physics}
        \usepackage{color}
        \usepackage{fullpage}
        \usepackage{graphicx}
        \usepackage{cancel}
        \usepackage{physics}
        \usepackage{fullpage}
        \DeclareMathOperator{\E}{E}
        \DeclareMathOperator{\Var}{V}
        \DeclareMathOperator{\bias}{bias}
        \newenvironment{answer}{\ \\ \color{red}}{\color{black}}
        % replace 'red' with 'white' above  to hide solutions (in printout)
---

```{r}
#| echo: false
#| warning: false
library(ggplot2)
library(tidyverse)
library(patchwork)
```

\subsection*{Problem 1}
Suppose we have a sample of $Y_1 \ldots Y_n$ and $X_1 \ldots X_n$  stored in an \textbf{R} dataframe `df' with $n$ rows. $X$ can take the values of 1,2,3,4 or 5. We run this code.

```{r}
#| echo: false
set.seed(123)
n = 1000
Y= rep(NA,n)
X = c(rep(1,350), rep(2,300),rep(3,200),rep(4,100),rep(5,50))
for(i in 1:n){
  Y[i] = X[i] + rnorm(1,0,2) 
}
df<- data.frame(x=X,y=Y)

```

```{r}
#| echo: true
B=10000
theta<- rep(NA,B)
for( b in 1:B ){
bootind = sample(1:n,size=n,replace=T)
dfboot = df[bootind,]
theta[b] = (1/4)*(mean(dfboot$y[dfboot$x==5]) - mean(dfboot$y[dfboot$x==1]))
}
```

```{r}
#| out-height: 3.5in
#| fig-ncol: 2
#| fig-subcap:
#|  -  Histogram of the vector Y's elements
#|  -  Jittered Scatterplot of X and Y
ggplot() + geom_histogram(aes(x=Y,     y=after_stat(density)), alpha=.3) + 
           geom_vline(aes(xintercept=mean(Y)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(Y)+c(-1,1)*sd(Y)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(Y), xmin=mean(Y)-1.96*sd(Y), xmax=mean(Y)+1.96*sd(Y), y=.025)) + theme_bw() + xlab('') + ylab('')

ggplot(df, aes(x=x, y=y)) +geom_jitter(shape=1,alpha=.3) + geom_point(x=1,y=mean(df$y[df$x==1]),size=3) + geom_point(x=2,y=mean(df$y[df$x==2]),size=3) + geom_point(x=3,y=mean(df$y[df$x==3]),size=3)+ geom_point(x=4,y=mean(df$y[df$x==4]),size=3) + geom_point(x=5,y=mean(df$y[df$x==5]),size=3) + geom_segment(aes(x=1,y=mean(y[x==1])-sd(y[x==1]), xend=1, yend=mean(y[x==1])+sd(y[x==1]))) + geom_segment(aes(x=2,y=mean(y[x==2])-sd(y[x==2]), xend=2, yend=mean(y[x==2])+sd(y[x==2]))) + geom_segment(aes(x=3,y=mean(y[x==3])-sd(y[x==3]), xend=3, yend=mean(y[x==3])+sd(y[x==3]))) + geom_segment(aes(x=4,y=mean(y[x==4])-sd(y[x==4]), xend=4, yend=mean(y[x==4])+sd(y[x==4]))) + geom_segment(aes(x=5,y=mean(y[x==5])-sd(y[x==5]), xend=5, yend=mean(y[x==5])+sd(y[x==5]))) 
  

```

```{r}
#| out-height: 3.5in
#| fig-ncol: 2
#| fig-subcap:
#|  -  Histogram of the theta's elements
ggplot() + geom_histogram(aes(x=theta, y=after_stat(density)), alpha=.3) + 
           geom_vline(aes(xintercept=mean(theta)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(theta)+c(-1,1)*sd(theta)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(theta), xmin=mean(theta)-1.96*sd(theta), xmax=mean(theta)+1.96*sd(theta), y=1)) + theme_bw() + xlab('') + ylab('')

```


\subsection*{Part a}
What does this code calculate? State it in words and mark it with an 'a', and an arrow or bracket
if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
mean(Y)
```

\begin{answer}
The sample mean (unconditioned). It's indicated by the solid vertical line on the sample histogram.
\end{answer}

\subsection*{Part b}
What does this code calculate? State it in words and mark it with a 'b', and an arrow or bracket if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
sd(Y)
```

\begin{answer}
The sample standard deviation. It's the distance between the solid vertical and either dashed vertical line on the sample histogram.
\end{answer}


\subsection*{Part c}
What does this code calculate? State it in words and mark it with a 'c', and an arrow or bracket
if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
mean(Y[X==1])
```

\begin{answer}
The conditional sample mean of Y among X=1. It's indicated by the black dot on the scatterplot where X=1.
\end{answer}

\subsection*{Part d}
What does this code calculate? State it in words and mark it with a 'd', and an arrow or bracket if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
sd(theta)
```

\begin{answer}
The bootstrap sampling distribution's standard deviation for the average increment estimator. 
It's the distance between the solid vertical and either dashed vertical line on the sampling distribution histogram.
\end{answer}

\subsection*{Part e}
What does this code calculate? State it in words and mark it with a 'e', and an arrow or bracket if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
(1/4)(mean(Y[X=5]) - mean(Y[X=1])) 
- 1.96* sqrt((1/4)^2*var(Y[X=5])/length(Y[X=5]) + (-1/4)^2*var(Y[X=1])/length(Y[X=1]))

(1/4)(mean(Y[X=5]) - mean(Y[X=1]))
+ 1.96* sqrt((1/4)^2*var(Y[X=5])/length(Y[X=5]) + (-1/4)^2*var(Y[X=1])/length(Y[X=1]))
```

\begin{answer}
A 95\% confidence interval --- based on normal approximation --- for the average incremental difference in Y from X=1 to X=5 in the population from which our data was drawn. This is indicated by the horizontal segment (with a dot in the middle) on the sampling distribution histogram (theta).
\end{answer}

\subsection*{Part f}
How could you calculate roughly the same thing as in part e using 'sd(theta)' replacing \begin{verbatim} sqrt((1/4)^2*var(Y[X=5])/length(Y[X=5]) + (-1/4)^2*var(Y[X=1])/length(Y[X=1])) \end{verbatim}

\begin{answer}

We can use the standard deviation of the bootstrap sampling distribution as our estimate of standard error.

\begin{verbatim}
(1/4)(mean(Y[X=5]) - mean(Y[X=1]))  - 1.96 * sd(theta)
(1/4)(mean(Y[X=5]) - mean(Y[X=1]))  + 1.96 * sd(theta)
\end{verbatim}

\end{answer}


\subsection*{Problem 2}

In the following parts, choose the figure with the appropriate regression line(s).


```{r, fig.width=10,fig.height=4}
#| echo: false
#| warning: false
n <- 500
zz <- c(rep(0,n/2),rep(1,n/2))
set.seed(220220)
xx <-  1 + 2*zz + rnorm(n,mean=0,sd = 1)
yy <- 2*zz*xx - xx + rnorm(n)


modAct1 <- lm(yy ~ xx)
modAct2 <- lm(yy ~ xx + zz)


df <- data.frame(xx,yy,zz)

p <- ggplot(data = df, aes(x = xx, y = yy)) +
  geom_point(aes(shape = factor(zz)),size = 1,alpha=.5) +
  scale_shape_manual(values = c(1,19), label = c("z=0","z=1")) +
  xlab("x") +
  ylab("y") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        legend.box.just = "left",
        legend.justification = c("left","top"),
        legend.position = c(0,1),
        legend.margin = margin(6,6,6,6),
        legend.title = element_blank(),
        legend.box.background = element_rect(color="black",size=1))

coef1 <- modAct1$coefficients
coef2 <- modAct2$coefficients
mod3 <- lm(yy ~ xx+zz+xx*zz, data = df)
coef3 <- coef(mod3)
# simple regression line y ~ x --------


```


```{r}
#| echo: false
#ggplot(data = df, aes(x = xx, y = yy)) +
#  geom_point(size = 1,alpha=.5) +
#  xlab("x") +
#  ylab("y") +
#  theme_bw() + geom_abline(intercept = coef1[1], slope = coef1[2]) + ggtitle("A3")
```

\subsubsection*{Part A}

Choose the figure with the two dashed lines associated with the additive multiple regression ($y\sim x+z$).

```{r,fig.width=10,fig.height=4}
#| echo: false
# multiple additive regression y ~ x + z
addA <- p + geom_abline(intercept = coef2[1], slope = coef2[2], linetype = 'dashed') +
  geom_abline(intercept = coef2[1] + coef2[3], slope = coef2[2], linetype = 'dashed') +
  ggtitle("A3")

addB <- p + geom_abline(intercept = 0.5, slope = coef3[2]*1.5,
                linetype = 'dashed')+ geom_abline(intercept = 0.5, slope = -1,
                linetype = 'dashed') + ggtitle("A1")

addC <- p + geom_curve(aes(x = -2,xend=6,y=-6,yend=-1),
               curvature = -0.8,
               angle = 70,
               linetype = 'dashed') +
  geom_curve(aes(x=-2,y=0,xend=4,yend=-7),
             curvature = -0.5,
             linetype = 'dashed') +
  ggtitle("A2")

#pdf(file="Q1partBplot.pdf",width=10,height=4)
 addB + addC + addA
#dev.off()
```

\color{red} A3 \color{black}

\subsubsection*{Part B}

Choose the figure with the two dotted lines associated with the interactive multiple regression ($y\sim x*z$).

```{r,fig.width=10,fig.height=4}
#| echo: false
# interactive model

intC2 <- p + geom_abline(intercept = coef3[1], slope = coef3[2],
                linetype = 'dotted') +
  geom_abline(intercept = coef3[1] + coef3[3],
              slope = coef3[2] + coef3[4],
              linetype = 'dotted') +
  ggtitle("B2")


intC1 <- p + geom_abline(intercept = coef3[1], slope = coef3[2]*1.5,
                linetype = 'dotted') +
  geom_abline(intercept = 5,
              slope = coef3[2]*1.5,
              linetype = 'dotted') +
  ggtitle("B1")

intC3 <- p + geom_curve(aes(x = -2,y = -2.5,xend=3,yend=5),
               curvature = 0.3,
               #angle = 30,
               linetype = 'dotted') +
  geom_curve(aes(x=0,y=5,xend=6,yend=-5),
             curvature = 0.3,
             angle=95,
             linetype = 'dotted') +
  ggtitle("B3")

#pdf(file="Q1partCplot.pdf",width=10,height=4)
intC1 + intC2 + intC3
#dev.off()
```

\color{red} B2 \color{black}

\subsection*{Problem 3}

\subsubsection*{Part A} 

```{r}
#| echo: false

p3n <- 200
z <- c(rep(0,p3n/2),rep(1,p3n/2))
set.seed(220220)
x <-   z + rnorm(p3n,mean=1,sd=1) + z*rnorm(p3n,mean=3,sd = 1)
y <- (-0.5)*x + (4)*z + rnorm(p3n)


dfp3 <- data.frame(y,x,z)

q1 <- ggplot(data = dfp3, aes(x = x, y = y)) +
  geom_point(aes(shape = factor(z)),size = 1,alpha=.5) +
  scale_shape_manual(values = c(1,19), label = c("z=0","z=1")) +
  xlab("x") +
  ylab("y") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        legend.box.just = "left",
        legend.justification = c("left","top"),
        legend.position = c(0,1),
        legend.margin = margin(6,6,6,6),
        legend.title = element_blank(),
        legend.box.background = element_rect(color="black",size=1))

q2<- ggplot(data = dfp3, aes(x = x, y = z)) +
  geom_jitter(height=.25) +  xlab("x") +
  ylab("z") +
  theme_bw()



modone<- lm(y~x+z,data=dfp3)
modtwo<- lm(z ~ x, data=dfp3)

```

We have collected $y$, $x$, and a binary variable $z$. The scatter plots below show our data. 

```{r}
#| out-height: 3.5in
#| fig-ncol: 2
#| fig-subcap:
#|  -  Scatterplot 1
#|  -  Scatterplot 2 (jittered)
q1
q2

```

We then run the regressions $lm(y \sim x+ z)$ and $lm(z \sim x)$, which give us the following output: 

```{r}
summary(modone)

summary(modtwo)

```
\bigskip 

What is the coefficient on $x$ from the simple regression $lm(y\sim x)$? Explain your steps and define any notation you use. 


\bigskip 

\color{red} 
Let the simple regression of $y$ on $x$ be $y = \beta_0^S + \beta_1^S x$ 

Let the multiple regression of $y$ on $x$ and $z$ be $y= \beta_0 + \beta_1 x + \beta_2 z$ 

Let the regression of $z$ on $x$ be $z = \gamma_0 + \gamma_1 x$ where $\gamma_1 = \frac{\sum_{i=1}^n (x_i-\bar{x})z_i}{\sum_{i=1}^n (x_i - \bar{x})^2}$

From our OVB formula, we know $\beta_1^S = \beta_1 + \beta_2 * \gamma_1$ 

The regression outputs indicate that $\beta_1 = -0.5517$, $\beta_2 = 4.21$, and $\gamma_1 = 0.1845$

$\beta_1^S = -0.5517 + (4.21)*(0.1845) = 0.225$ 

\color{black}



\subsubsection*{Part B}

Suppose we want $\beta_1$ from $y_i = \beta_0 + \beta_1 x_i + \beta_2 z_i +  \beta_3 w_i + e_i$ but we run the regression $lm (y\sim x)$ without $z$ or $w$. We have de-meaned our variables $x_i$, $z_i$, and $w_i$ such that $\bar{x}=\sum_{i=1}^n x_i =0$, $\bar{z}=\sum_{i=1}^n z_i =0$, and $\bar{w}= \sum_{i=1}^n w_i =0$. Under what condition(s) will the coefficient on $x$ from $lm (y\sim x)$, which is $\frac{\sum_{i=1}^n x_i y_i}{\sum_{i=1}^n x_i^2 }$, be equal to $\beta_1$? Derive what your coefficient on $x$ will be in terms of $\beta$s and justify your steps.

\bigskip

\color{red}

The coefficient on $x$ from $lm (y\sim x)$ is $$\frac{\sum_{i=1}^n x_i y_i}{\sum_{i=1}^n x_i^2 }$$

Plug in $y_i$ defined by the multivariate regression: $y_i = \beta_0 + \beta_1 x_i + \beta_2 z_i +  \beta_3 w_i + e_i$

$$\frac{\sum_{i=1}^n x_i (\beta_0 + \beta_1 x_i + \beta_2 z_i +  \beta_3 w_i + e_i)}{\sum_{i=1}^n x_i^2 }$$

Distribute

$$\frac{\sum_{i=1}^n x_i \beta_0 }{\sum_{i=1}^n x_i^2 } + \frac{\sum_{i=1}^n x_i \beta_1 x_i}{\sum_{i=1}^n x_i^2 } + \frac{\sum_{i=1}^n x_i \beta_2 z_i }{\sum_{i=1}^n x_i^2 } + \frac{\sum_{i=1}^n x_i \beta_3 w_i }{\sum_{i=1}^n x_i^2 }+  \frac{\sum_{i=1}^n x_i  e_i}{\sum_{i=1}^n x_i^2 }$$ Let's work through each term.

$$\frac{\sum_{i=1}^n x_i \beta_0 }{\sum_{i=1}^n x_i^2 } = \frac{ \beta_0 \sum_{i=1}^n x_i }{\sum_{i=1}^n x_i^2 } = \frac{ \beta_0 *0 }{\sum_{i=1}^n x_i^2 } = 0$$ 

as $x$ is demeaned, $\sum x_i =0$ 


\bigskip

$$\frac{\sum_{i=1}^n x_i \beta_1 x_i}{\sum_{i=1}^n x_i^2 } = \frac{  \beta_1 \sum_{i=1}^n x_i x_i}{\sum_{i=1}^n x_i^2 } = \beta_1$$

\bigskip

$$\frac{\sum_{i=1}^n x_i \beta_2 z_i }{\sum_{i=1}^n x_i^2 } = \frac{\beta_2 \sum_{i=1}^n x_i z_i }{\sum_{i=1}^n x_i^2 }$$

\bigskip

$$\frac{\sum_{i=1}^n x_i \beta_3 w_i }{\sum_{i=1}^n x_i^2 } = \frac{\beta_3 \sum_{i=1}^n x_i w_i }{\sum_{i=1}^n x_i^2 }$$

\bigskip

$$\frac{\sum_{i=1}^n x_i  e_i}{\sum_{i=1}^n x_i^2 } = \frac{0}{\sum_{i=1}^n x_i^2 } =0$$ we know that $\sum_{i=1}^n x_i e_i =0$ from the first order conditions on our least squares estimator (we chose the intercept and coefficient(s) that minimize the sum of squared residuals)

\bigskip

Thus we have

$$\frac{\sum_{i=1}^n x_i y_i}{\sum_{i=1}^n x_i^2 } = \beta_1 + \frac{\beta_2 \sum_{i=1}^n x_i z_i }{\sum_{i=1}^n x_i^2 } + \frac{\beta_3 \sum_{i=1}^n x_i w_i }{\sum_{i=1}^n x_i^2 }$$ The coefficient on $x$ from the simple regression of $y$ on $x$ will be equal to $\beta_1$ if: (1) $\beta_2$ or the coefficient on $x$ from the simple regression from $z$ on $x$ is 0 and (2) $\beta_3$ or the coefficient on $x$ from the simple regression of $w$ on $x$ if 0. If either of these terms is non-zero, our simple regression will not give us the same result as our regression that includes both $x$, $z$, and $w$. If $x$ is randomized (like a random treatment assignment), it should not be correlated with $z$ or $w$, so our short regression coefficient should give us $\beta_1$. 
\color{black}

\bigskip 

\subsection*{Problem 4}

Consider the following \texttt{R} code:

```{=tex}
\begin{verbatim}
1. lm(y ~ residuals(lm(z~x)))
2. lm(y ~ residuals(lm(x~z)))
3. lm(residuals(lm(y~z)) ~ x)
4. lm(residuals(lm(y~x)) ~ z)
5. lm(residuals(lm(y~x)) ~ residuals(lm(z~x)))
6. lm(residuals(lm(y~x)) ~ residuals(lm(x~z)))
7. lm(residuals(lm(y~z)) ~ residuals(lm(x~z)))
8. lm(residuals(lm(y~z)) ~ residuals(lm(z~x)))
 \end{verbatim}
```
```{=tex}
\begin{itemize}
\item[a)]  Choose one of the above that will produce the coefficient on $x$ from the multiple regression of $y \sim x + z$\\
\color{red} 2 or 7 \color{black}

\bigskip

\item[b)]  Choose one of the above that will produce the coefficient on $z$ from the multiple regression of $y \sim x + z$\\
\color{red} 5 or 1 \color{black}

\bigskip 

\item[c)] Choose one of the above that will produce the same residuals as the residuals from the multiple regression of $y \sim x + z$\\
 \color{red}  7 or 5 \color{black}
\end{itemize}


