---
title: "Midterm 2 Practice Problems"
format: pdf
editor: visual
---

```{r}
#| echo: false
#| warning: false
library(ggplot2)
library(tidyverse)
library(patchwork)
```

\section*{Problem 1}

In the following three parts, choose the figure with the appropriate regression line(s).
\subsubsection*{Part A}
Choose the figure with the solid line that represents the least squares fit of the simple univariate regression model, i.e. $\hat\mu(x) = \hat\beta_0 + \hat\beta_1 x$, to the whole sample.

```{r, fig.width=10,fig.height=4}
#| echo: false
#| warning: false
n <- 1000
zz <- c(rep(0,n/2),rep(1,n/2))
set.seed(123456)
xx <-  3 - 2*zz + rnorm(n,mean=0,sd = 1)
yy <- 2*zz*xx - xx + rnorm(n)

#pdf("midtermQ1.pdf")
#plot(xx,yy,xlab="x",ylab="y")
#points(xx[zz==0],yy[zz==0],pch=1)
#points(xx[zz==1],yy[zz==1],pch=19)
#legend("topleft",legend=c("z=0","z=1"),pch=c(1,19))
#dev.off()


#pdf("midtermQ1sol.pdf")
#plot(xx,yy,xlab="x",ylab="y")
#points(xx[zz==0],yy[zz==0],pch=1)
#points(xx[zz==1],yy[zz==1],pch=19)
#legend("topleft",legend=c("z=0","z=1"),pch=c(1,19))
modAct1 <- lm(yy ~ xx)
modAct2 <- lm(yy ~ xx + zz)
#abline(modAct1,lwd=2)
#abline(a=modAct2$coef[1],b=modAct2$coef[2],lty=2,lwd=2)
#abline(a=modAct2$coef[1]+modAct2$coef[3],b=modAct2$coef[2],lty=2,lwd=2)
#abline(lm(yy[zz==0]~xx[zz==0]),lty=3,lwd=2)
#abline(lm(yy[zz==1]~xx[zz==1]),lty=3,lwd=2)
#dev.off()


df <- data.frame(xx,yy,zz)

p <- ggplot(data = df, aes(x = xx, y = yy)) +
  geom_point(aes(shape = factor(zz)),size = 1,alpha=.2) +
  scale_shape_manual(values = c(1,19), label = c("z=0","z=1")) +
  xlab("x") +
  ylab("y") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        legend.box.just = "left",
        legend.justification = c("left","top"),
        legend.position = c(0,1),
        legend.margin = margin(6,6,6,6),
        legend.title = element_blank(),
        legend.box.background = element_rect(color="black",size=1))

coef1 <- modAct1$coefficients
coef2 <- modAct2$coefficients
# simple regression line y ~ x --------
simpleA <- p + geom_abline(intercept = coef1[1], slope = coef1[2]) + ggtitle("A3")

simpleB <- p + geom_hline(yintercept = mean(df$yy)) + ggtitle("A1")

mod3 <- lm(yy ~ xx:zz, data = df)
simpleC <- p + geom_abline(intercept = coef(mod3)[1], slope = coef(mod3)[2]) +
  ggtitle("A2")


#pdf(file="Q1partAplot.pdf",width=10,height=4)
simpleB + simpleC + simpleA
#dev.off() 

```

\subsubsection*{Part B}
Choose the figure with the two dashed lines associated with the least squares fit of the additive bivariate regression model, i.e., $\hat\mu(x,z)=\hat\beta_0 + \hat\beta_1 x+ \hat\beta_2 z$.

```{r,fig.width=10,fig.height=4}
#| echo: false
# multiple additive regression y ~ x + z
addA <- p + geom_abline(intercept = coef2[1], slope = coef2[2], linetype = 'dashed') +
  geom_abline(intercept = coef2[1] + coef2[3], slope = coef2[2], linetype = 'dashed') +
  ggtitle("B1")

addB <- p + geom_vline(xintercept = 1, linetype = 'dashed') +
  geom_vline(xintercept = 3.25, linetype = 'dashed') +
  ggtitle("B2")

addC <- p + geom_curve(aes(x = -2,xend=6,y=-6,yend=-1),
               curvature = -0.8,
               angle = 70,
               linetype = 'dashed') +
  geom_curve(aes(x=-2,y=0,xend=4,yend=-7),
             curvature = -0.5,
             linetype = 'dashed') +
  ggtitle("B3")

#pdf(file="Q1partBplot.pdf",width=10,height=4)
addA + addB + addC
#dev.off()
```

\subsubsection*{Part C}
Choose the figure with the two dotted lines associated with the least squares fit of the fully-interacted regression model, i.e., $\hat\mu(x,z) = \hat\beta_0 + \hat\beta_1 x + \hat\beta_2 z + \hat\beta_3 xz$.

```{r,fig.width=10,fig.height=4}
#| echo: false
# interactive model
mod3 <- lm(yy ~ xx*zz, data = df)
coef3 <- coef(mod3)
intC2 <- p + geom_abline(intercept = coef3[1], slope = coef3[2],
                linetype = 'dotted') +
  geom_abline(intercept = coef3[1] + coef3[3],
              slope = coef3[2] + coef3[4],
              linetype = 'dotted') +
  ggtitle("C2")


intC1 <- p + geom_abline(intercept = coef3[1], slope = coef3[2]*1.5,
                linetype = 'dotted') +
  geom_abline(intercept = coef3[1] + coef3[3]*4,
              slope = coef3[2] + coef3[4]*4,
              linetype = 'dotted') +
  ggtitle("C1")

intC3 <- p + geom_curve(aes(x = -2,y = -2.5,xend=3,yend=5),
               curvature = 0.3,
               #angle = 30,
               linetype = 'dotted') +
  geom_curve(aes(x=0,y=5,xend=6,yend=-5),
             curvature = 0.3,
             angle=95,
             linetype = 'dotted') +
  ggtitle("C3")

#pdf(file="Q1partCplot.pdf",width=10,height=4)
intC1 + intC2 + intC3
#dev.off()
```



\section*{Problem 2}

Suppose we have a sample $Y_1 \ldots Y_n$ stored on an \textbf{R} vector $Y$ of length $n$. We run this code.

```{r}
#| echo: false
set.seed(123)
n = 1000
Y = rnorm(n)
```

```{r}
#| echo: true
B=10000
Y.bar = rep(NA,B)
for(b in 1:B) {
  Y.bar[b]= mean(sample(Y,size=n,replace=TRUE))
}
```

```{r}
#| echo: false
#| warning: false
#| out-height: 6in
#| fig-ncol: 2
#| fig-subcap:
#|  -  Histogram of the vector Y's elements
#|  -  Histogram of the Y.bar's elements
ggplot() + geom_histogram(aes(x=Y,     y=after_stat(density)), alpha=.3) + 
           geom_vline(aes(xintercept=mean(Y)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(Y)+c(-1,1)*sd(Y)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(Y), xmin=mean(Y)-1.96*sd(Y), xmax=mean(Y)+1.96*sd(Y), y=.15)) + theme_bw() + xlab('') + ylab('')
ggplot() + geom_histogram(aes(x=Y.bar, y=after_stat(density)), alpha=.3) + 
           geom_vline(aes(xintercept=mean(Y.bar)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(Y.bar)+c(-1,1)*sd(Y.bar)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(Y.bar), xmin=mean(Y.bar)-1.96*sd(Y)/sqrt(n), xmax=mean(Y.bar)+1.96*sd(Y)/sqrt(n), y=5)) + theme_bw() + xlab('') + ylab('')
```


 \vspace{.5in}

\subsection*{Part a}

What does this code calculate? State it in words and mark it with an 'a', and an arrow or bracket if necessary for clarity, on the appropriate plot above.

```{r}
#| echo: true
#| eval: false
mean(Y) - 1.96* sd(Y)/sqrt(n)
mean(Y) + 1.96* sd(Y)/sqrt(n)
```

\vspace{.5in}

\subsection*{Part b}

How could you calculate roughly the same thing as in part a using 'sd(Y.bar)' and not 'sd(Y)'?

\vspace{.5in}

\section*{Problem 3}
Suppose $Y_1 \ldots Y_n$ are sampled with replacement from a population with mean $\mu=1$ and variance $\sigma^2=1$.
In Midterm 1, we discussed the estimator $\tilde{Y}_k = \frac{1}{n-k} \sum_{i=1}^n Y_i$ for $k=2$. Here, we'll
talk about that case and the case $k=\sqrt{n}$. For both estimators $\tilde{Y}_2$ and $\tilde{Y}_{\sqrt{n}}$, 
for each of the three sample sizes $n=10$, $n=10^2=100$, and $n=10^4=10,000$, do the following.
\begin{itemize}
\item Calculate the estimator's mean $\hat\mu$ and standard error $\hat\sigma_n$. 
\item Sketch the estimator's sampling distribution. Label your axes clearly and include $\mu=1$ in your sketch.
\item On top of your sketches, illustrate how you've calculate the coverage probability of the 
      interval estimate $\hat\mu \pm 1.96 \hat\sigma_n$.
\item Give a rough approximation of this coverage probability. You don't have to do this well---just do it by eye. Is it about 95\%? Is it above 50\%? 
      That's the kind of precision I'm looking for.
\end{itemize}

You should have 6 sketches with corresponding means, standard errors, and coverage probabilities. That's more than we'd ask for on the real exam. If you get the idea and want to save time by skipping the $n=10^2$ case, that's fine.

\section*{Problem 4}

We take a random sample of $n$ individuals and record their years of education and income. $Y_i$ is an individual $i$'s income and $X_i$ is that individual's number of years of education. Let
$$\hat{\mu}(16) = \frac{1}{\sum_{i=1}^n 1_{16}(X_i)} \sum_{i=1}^n 1_{16}(X_i)Y_i
\quad \text{where} \quad 1_{16}(x) = \begin{cases} 1 & \text{if } x = 16 \\ 0 & \text{otherwise} \end{cases}$$
Show that $\hat{\mu}(16)$ is an unbiased estimator of $\mu(16)$, the subpopulation mean income for those with 16 years of education. Justify your steps. 



\section*{Problem 5}

In the below visualization, the x-axis is an individual's years of education, the y-axis is income, the red line is the least squares line, and the black dots indicate the mean income within each year of education. You have three options to summarize the change in income across an additional year of education: the average over years, the average weighted by the proportion of individuals with the years of education $P_x$, and the slope of the least squares line. Rank these options smallest to largest regarding the magnitude of the estimate you expect and explain your reasoning. 

\begin{figure}[H]
\includegraphics{test.png}
\caption*{Income and Years of Education}
\end{figure}

\section*{Problem 6}
Suppose we want the slope parameter $\hat\gamma_1$ from the least squares fit of the additive model,
$\hat \mu_2(x,z) = \hat\gamma_0 + \hat\gamma_1 x + \hat\gamma_2 z$. But we instead have the slope parameter $\hat\beta_1$ from the least squares fit of the univariate regression model, $\hat\mu_1(x)=\hat \beta_0 + \hat\beta_1 x$. And suppose we've centered our covariates so $\bar X_i$ and $\bar Z_i$ are zero. Will 
the slope paramter we want ($\hat\gamma_1$) be equal to the one we have ($\hat\beta_1$)? Derive a formula for $\hat\gamma_1$ in terms of $\hat\beta_1$ and explain the steps in your derivation. 

\paragraph{Hint} Start with the formula for the slope $\hat\gamma_1$ in a univariate regression and use the expansion $Y_i = \hat\mu_2(X_i,Z_i) + \hat\varepsilon_i$ where $\hat\varepsilon_i = Y_i-\hat\mu_2(X_i)$ are the residuals from the fit of the the additive model. What do we know about these residuals and why do we know it?
