---
title: "Exam 2 (November 13)"
subtitle: "QTM 220"
execute:
  echo: false
  warning: false
format: 
  pdf:
    include-in-header: 
      text: |
        \usepackage{amsmath}
        \usepackage{setspace}
        \onehalfspacing
        \usepackage{amssymb}
        \usepackage{physics}
        \usepackage{color}
        \usepackage{fullpage}
        \usepackage{graphicx}
        \usepackage{cancel}
        \usepackage{physics}
        \usepackage{fullpage}
        \DeclareMathOperator{\E}{E}
        \DeclareMathOperator{\Var}{V}
        \DeclareMathOperator{\bias}{bias}
        \newenvironment{answer}{\ \\ \color{red}}{\color{black}}
        % replace 'red' with 'white' above  to hide solutions (in printout)
---

```{r}
#| echo: false
#| warning: false
library(ggplot2)
library(tidyverse)
library(patchwork)
```

\subsection*{Problem 1}
Suppose we have a sample of $(X_1,Y_1) \ldots (X_n,Y_n)$ drawn with replacement from a population. It's stored in an \textbf{R} dataframe `df' with $n$ rows. $X$ can take the values of 1,2,3,4 or 5. We run this code.

```{r}
#| echo: false
set.seed(123)
n = 1000
Y= rep(NA,n)
X = c(rep(1,350), rep(2,300),rep(3,200),rep(4,100),rep(5,50))
for(i in 1:n){
  Y[i] = X[i] + rnorm(1,0,2) 
}
df<- data.frame(x=X,y=Y)

```

```{r}
#| echo: true
B=10000
theta<- rep(NA,B)
for( b in 1:B ){
bootind = sample(1:n,size=n,replace=T)
dfboot = df[bootind,]
theta[b] = (1/4)*(mean(dfboot$y[dfboot$x==5]) - mean(dfboot$y[dfboot$x==1]))
}
```

```{r}
#| out-height: 3.5in
#| fig-ncol: 2
#| fig-subcap:
#|  -  Histogram of the vector Y's elements
#|  -  Jittered Scatterplot of X and Y
ggplot() + geom_histogram(aes(x=Y,     y=after_stat(density)), alpha=.3) + 
           geom_vline(aes(xintercept=mean(Y)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(Y)+c(-1,1)*sd(Y)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(Y), xmin=mean(Y)-1.96*sd(Y), xmax=mean(Y)+1.96*sd(Y), y=.025)) + theme_bw() + xlab('') + ylab('')

ggplot(df, aes(x=x, y=y)) +geom_jitter(shape=1,alpha=.3) + geom_point(x=1,y=mean(df$y[df$x==1]),size=3) + geom_point(x=2,y=mean(df$y[df$x==2]),size=3) + geom_point(x=3,y=mean(df$y[df$x==3]),size=3)+ geom_point(x=4,y=mean(df$y[df$x==4]),size=3) + geom_point(x=5,y=mean(df$y[df$x==5]),size=3) + geom_segment(aes(x=1,y=mean(y[x==1])-sd(y[x==1]), xend=1, yend=mean(y[x==1])+sd(y[x==1]))) + geom_segment(aes(x=2,y=mean(y[x==2])-sd(y[x==2]), xend=2, yend=mean(y[x==2])+sd(y[x==2]))) + geom_segment(aes(x=3,y=mean(y[x==3])-sd(y[x==3]), xend=3, yend=mean(y[x==3])+sd(y[x==3]))) + geom_segment(aes(x=4,y=mean(y[x==4])-sd(y[x==4]), xend=4, yend=mean(y[x==4])+sd(y[x==4]))) + geom_segment(aes(x=5,y=mean(y[x==5])-sd(y[x==5]), xend=5, yend=mean(y[x==5])+sd(y[x==5]))) + theme_bw() 

```

```{r}
#| out-height: 3.5in
#| fig-ncol: 2
#| fig-subcap:
#|  -  Histogram of the theta's elements
ggplot() + geom_histogram(aes(x=theta, y=after_stat(density)), alpha=.3) + 
           geom_vline(aes(xintercept=mean(theta)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(theta)+c(-1,1)*sd(theta)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(theta), xmin=mean(theta)-1.96*sd(theta), xmax=mean(theta)+1.96*sd(theta), y=1)) + theme_bw() + xlab('') + ylab('')

```


\subsection*{Part a}
What does this code calculate? State it in words and mark it with an 'a', and an arrow or bracket
if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
mean(Y)
```

\vspace{1 in}


\subsection*{Part b}
What does this code calculate? State it in words and mark it with a 'b', and an arrow or bracket if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
sd(Y)
```

\vspace{1 in}


\subsection*{Part c}
What does this code calculate? State it in words and mark it with a 'c', and an arrow or bracket
if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
mean(Y[X==1])
```

\vspace{1 in}

\vspace{1 in}

\subsection*{Part d}
What does this code calculate? State it in words and mark it with a 'd', and an arrow or bracket if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
sd(theta)
```

\newpage

\subsection*{Part e}
What does this code calculate? State it in words and mark it with a 'e', and an arrow or bracket if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
(1/4)(mean(Y[X=5]) - mean(Y[X=1])) 
- 1.96* sqrt((1/4)^2*var(Y[X=5])/length(Y[X=5]) + (-1/4)^2*var(Y[X=1])/length(Y[X=1]))

(1/4)(mean(Y[X=5]) - mean(Y[X=1]))
+ 1.96* sqrt((1/4)^2*var(Y[X=5])/length(Y[X=5]) + (-1/4)^2*var(Y[X=1])/length(Y[X=1]))
```

\vspace{2 in}

\subsection*{Part f}
How could you calculate roughly the same thing as in part e using 'sd(theta)' replacing \begin{verbatim} sqrt((1/4)^2*var(Y[X=5])/length(Y[X=5]) + (-1/4)^2*var(Y[X=1])/length(Y[X=1])) \end{verbatim}

\vspace{1 in}


\subsection*{Problem 2}

In the following parts, choose the figure with the appropriate regression line(s). Clearly mark one choice, no explanation needed.


```{r, fig.width=10,fig.height=4}
#| echo: false
#| warning: false
n <- 500
zz <- c(rep(0,n/2),rep(1,n/2))
set.seed(220220)
xx <-  1 + 2*zz + rnorm(n,mean=0,sd = 1)
yy <- 2*zz*xx - xx + rnorm(n)


modAct1 <- lm(yy ~ xx)
modAct2 <- lm(yy ~ xx + zz)


df <- data.frame(xx,yy,zz)

p <- ggplot(data = df, aes(x = xx, y = yy)) +
  geom_point(aes(shape = factor(zz)),size = 1,alpha=.5) +
  scale_shape_manual(values = c(1,19), label = c("z=0","z=1")) +
  xlab("x") +
  ylab("y") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        legend.box.just = "left",
        legend.justification = c("left","top"),
        legend.position = c(0,1),
        legend.margin = margin(6,6,6,6),
        legend.title = element_blank(),
        legend.box.background = element_rect(color="black",size=1))

coef1 <- modAct1$coefficients
coef2 <- modAct2$coefficients
mod3 <- lm(yy ~ xx+zz+xx*zz, data = df)
coef3 <- coef(mod3)
# simple regression line y ~ x --------


```


```{r}
#| echo: false
#ggplot(data = df, aes(x = xx, y = yy)) +
#  geom_point(size = 1,alpha=.5) +
#  xlab("x") +
#  ylab("y") +
#  theme_bw() + geom_abline(intercept = coef1[1], slope = coef1[2]) + ggtitle("A3")
```

\subsubsection*{Part A}

Choose the figure with the two dashed lines associated with the additive multiple regression $\hat\mu(x,z)=\hat\beta_0 + \hat\beta_1 x + \hat\beta_2 z$.

```{r,fig.width=10,fig.height=4}
#| echo: false
# multiple additive regression y ~ x + z
addA <- p + geom_abline(intercept = coef2[1], slope = coef2[2], linetype = 'dashed') +
  geom_abline(intercept = coef2[1] + coef2[3], slope = coef2[2], linetype = 'dashed') +
  ggtitle("A3")

addB <- p + geom_abline(intercept = 0.5, slope = coef3[2]*1.5,
                linetype = 'dashed')+ geom_abline(intercept = 0.5, slope = -1,
                linetype = 'dashed') + ggtitle("A1")

addC <- p + geom_curve(aes(x = -2,xend=6,y=-6,yend=-1),
               curvature = -0.8,
               angle = 70,
               linetype = 'dashed') +
  geom_curve(aes(x=-2,y=0,xend=4,yend=-7),
             curvature = -0.5,
             linetype = 'dashed') +
  ggtitle("A2")

#pdf(file="Q1partBplot.pdf",width=10,height=4)
 addB + addC + addA
#dev.off()
```


\subsubsection*{Part B}

Choose the figure with the two dotted lines associated with the interactive multiple regression $\hat\mu(x,z)=\hat\beta_0 + \hat\beta_1 x + \hat\beta_2 z + \hat\beta_3 xz$.

```{r,fig.width=10,fig.height=4}
#| echo: false
# interactive model

intC2 <- p + geom_abline(intercept = coef3[1], slope = coef3[2],
                linetype = 'dotted') +
  geom_abline(intercept = coef3[1] + coef3[3],
              slope = coef3[2] + coef3[4],
              linetype = 'dotted') +
  ggtitle("B2")


intC1 <- p + geom_abline(intercept = coef3[1], slope = coef3[2]*1.5,
                linetype = 'dotted') +
  geom_abline(intercept = 5,
              slope = coef3[2]*1.5,
              linetype = 'dotted') +
  ggtitle("B1")

intC3 <- p + geom_curve(aes(x = -2,y = -2.5,xend=3,yend=5),
               curvature = 0.3,
               #angle = 30,
               linetype = 'dotted') +
  geom_curve(aes(x=0,y=5,xend=6,yend=-5),
             curvature = 0.3,
             angle=95,
             linetype = 'dotted') +
  ggtitle("B3")

#pdf(file="Q1partCplot.pdf",width=10,height=4)
intC1 + intC2 + intC3
#dev.off()
```


\subsection*{Problem 3}

\subsubsection*{Part A} 

```{r}
#| echo: false

p3n <- 200
z <- c(rep(0,p3n/2),rep(1,p3n/2))
set.seed(220220)
x <-   z + rnorm(p3n,mean=1,sd=1) + z*rnorm(p3n,mean=3,sd = 1)
y <- (-0.5)*x + (4)*z + rnorm(p3n)


dfp3 <- data.frame(y,x,z)

q1 <- ggplot(data = dfp3, aes(x = x, y = y)) +
  geom_point(aes(shape = factor(z)),size = 1,alpha=.5) +
  scale_shape_manual(values = c(1,19), label = c("z=0","z=1")) +
  xlab("x") +
  ylab("y") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        legend.box.just = "left",
        legend.justification = c("left","top"),
        legend.position = c(0,1),
        legend.margin = margin(6,6,6,6),
        legend.title = element_blank(),
        legend.box.background = element_rect(color="black",size=1))

q2<- ggplot(data = dfp3, aes(x = x, y = z)) +
  geom_jitter(height=.25) +  xlab("x") +
  ylab("z") +
  theme_bw()



modone<- lm(y~x+z,data=dfp3)
modtwo<- lm(z ~ x, data=dfp3)

```

We have collected $y$, $x$, and a binary variable $z$. The scatter plots below show our data. 

```{r}
#| out-height: 3.5in
#| fig-ncol: 2
#| fig-subcap:
#|  -  Scatterplot 1
#|  -  Scatterplot 2 (jittered)
q1
q2

```

\subsubsection*{Part A}
Do you expect the coefficient on $x$ from the simple regression of $y$ on $x$, $\hat\mu(x)=\hat\beta_0 + \hat\beta_1 x$, to be: (choose one, no explanation needed)

\begin{itemize}
\item Positive

\item Negative

\item Zero 
\end{itemize}
\newpage
\subsubsection*{Part B}
We then run the regression $\hat\mu(x,z) = \hat\beta_0 + \hat\beta_1 x + \hat\beta_2 z$ which gives us the following output: 
```{r}
summary(modone)
```
And we run the regression $lm(z \sim x)$ which gives us the following output: 
```{r}
summary(modtwo)
```


How would you use these outputs to calculate the coefficient on $x$ from the simple regression $\hat\mu(x)=\hat\beta_0 + \hat\beta_1 x$? Explain your steps and define any notation you use. Plug in the estimates from your output (you can leave it as an equation). 

\vspace{2 in}

\newpage
\subsubsection*{Part C (Bonus)}
Suppose we want $\hat\beta_1$ from the additive regression of $y$ on three variables, $\hat\mu(x,z,w) = \hat\beta_0 + \hat\beta_1 x + \hat\beta_2 z +  \hat\beta_3 w$, but we run the univariate regression $lm (y\sim x)$ without $z$ or $w$. Under what condition(s) will the coefficient on $x$ from the univariate regression, which is $\frac{\sum_{i=1}^n (X_i-\bar X) Y_i}{\sum_{i=1}^n (X_i-\bar X)^2 }$, be equal to $\hat\beta_1$ from the additive model? Explain your answer. You'll probably want to derive a formula for the univariate regression coefficient in terms of $\hat\beta_0 \ldots \hat\beta_3$. 


\newpage

\subsection*{Problem 4}

Let's reconsider what we were doing in Problem 1. In that problem, we were using a difference in means to estimate some population summary. Suppose that, instead of that difference in means, we used the slope of a least squares line fit to the data in Figure 2.
This gives us a \emph{biased} estimator of the same population summary. I've plotted the sampling distribution of that estimator below, with its mean indicated by the thin solid line; dashed lines indicating the mean plus or minus two standard errors; and the population summary we're estimating indicated by a thick black line.


```{r}
bias = -1
se = 1
mu=0
breaks = seq(-5,5,1)
x=seq(min(breaks), max(breaks), length.out=1000)
ggplot() + 
    geom_area(aes(x=x, y=dnorm(x, mean=mu+bias, sd=se)), alpha=.1) +
    geom_vline(aes(xintercept=mu+bias)) +
    geom_vline(aes(xintercept=mu+bias+c(-2,2)*se), linetype='dotted') + 
    geom_vline(aes(xintercept=mu), color='black', size=1) +
    scale_x_continuous(breaks=breaks, limits=range(breaks)) +
    theme_bw() + theme(axis.text.x = element_blank(), axis.text.y=element_blank()) +
    xlab('') + ylab('')
```

\subsubsection*{Part A}
Letting $\hat\theta$ be our point estimator, i.e. a draw from the sampling distribution above, explain how to calculate the coverage of the confidence interval $\hat\theta \pm 2\  \text{se}$ in terms of the sampling distribution's density $f(x)$.  Is the coverage closer to 0\%, 50\%, or 95\%?  You do not need to calculate the coverage any more precisely than that---just explain how to do it.

\subsubsection*{Part B}
Suppose the data we have is from a pilot study. We're considering re-running the study with 4 times more data. 
If we use intervals $\hat\theta \pm 2\  \text{se}$ based on the same point estimator, will our coverage be closer to 0\%, 50\%, or 95\%?  What if we used 16 times more data?
You may assume the point estimator's bias is the same at all sample sizes and its standard error $\text{se}$ is directly proportional to $1/\sqrt{n}$, i.e., that
$\text{se}$ is $1/2$ as big as it is above when we have 4 times more data and $1/4$ as big as it is above when we have 16 times more data.
\emph{Hint.} Consider drawing diagrams like the one above at these two sample sizes.