---
title: "Midterm 2"
subtitle: "QTM 220"
---

```{r}
#| echo: false
library(tidyverse) 
library(scales)

theme_set(theme_minimal())
options(scipen = 1, digits = 2)
```

::: hidden
$$
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\Var}{V}
\DeclareMathOperator{\hVar}{\widehat{V}}
\DeclareMathOperator{\bias}{bias}
\DeclareMathOperator*{argmin}{argmin}
\newcommand{\model}{\mathcal{M}}
$$
:::

```{r}
#| echo: false
#| warning: false
library(ggplot2)
library(tidyverse)
library(patchwork)
```

# Problem 1

Suppose we have a sample of $(X_1,Y_1) \ldots (X_n,Y_n)$ drawn with replacement from a population. It's stored in vectors $X$ and $Y$ of length $n$. $X_i$ can take the values of 1,2,3,4 or 5. 
```{r}
#| echo: false
set.seed(123)
n = 1000
X = c(rep(1,350), rep(2,300),rep(3,200),rep(4,100),rep(5,50))
Y = X + rnorm(n,0,2)
df<- data.frame(x=X,y=Y)
```

Here's what the data looks like.

```{r}
#| out-height: 3.5in
#| fig-ncol: 2
#| fig-subcap:
#|  -  Histogram of the vector Y's elements
#|  -  Jittered Scatterplot of X and Y
ggplot() + geom_histogram(aes(x=Y,     y=after_stat(density)), alpha=.2, bins=50) + 
           geom_vline(aes(xintercept=mean(Y)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(Y)+c(-1,1)*sd(Y)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(Y), xmin=mean(Y)-1.96*sd(Y), xmax=mean(Y)+1.96*sd(Y), y=.025)) + theme_bw() + xlab('') + ylab('')

ggplot() + geom_point(aes(x=X, y=Y), alpha=.1, position=position_jitter(w=.2, h=0)) + xlab('') + ylab('') +
           stat_summary(aes(x=X, y=Y), geom='pointrange', 
              fun=mean, 
              fun.min=function(x) mean(x)-sd(x),
              fun.max=function(x) mean(x)+sd(x))  
```


We run this code.

```{r}
#| echo: true
B=1000

thetastar =  rep(NA,B)
for( b in 1:B ){
  Jstar = sample(1:n,size=n,replace=TRUE)
  Xstar = X[Jstar]
  Ystar = Y[Jstar]
  thetastar[b] = (1/4)*(mean(Ystar[Xstar==5]) - mean(Ystar[Xstar==1]))
}
```

```{r}
#| out-height: 3.5in
#| fig-cap: Histogram of the thetastar's elements
ggplot() + geom_histogram(aes(x=thetastar, y=after_stat(density)), alpha=.2) + 
           geom_vline(aes(xintercept=mean(thetastar)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(thetastar)+c(-1,1)*sd(thetastar)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(thetastar), xmin=mean(thetastar)-1.96*sd(thetastar), xmax=mean(thetastar)+1.96*sd(thetastar), y=1)) + xlab('') + ylab('')

```


## Part A
What does this code calculate? State it in words and mark it with an 'a', and an arrow or bracket
if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
mean(Y)
```


## Part B
What does this code calculate? State it in words and mark it with a 'b', and an arrow or bracket if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
sd(Y)
```

## Part C
What does this code calculate? State it in words and mark it with a 'c', and an arrow or bracket
if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
mean(Y[X==1])
```

## Part D   
What does this code calculate? State it in words and mark it with a 'd', and an arrow or bracket if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
sd(thetastar)
```

## Part E
What does this code calculate? State it in words and mark it with a 'e', and an arrow or bracket if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
(1/4)(mean(Y[X=5]) - mean(Y[X=1])) 
- 1.96* sqrt((1/4)^2*var(Y[X=5])/length(Y[X=5]) + (-1/4)^2*var(Y[X=1])/length(Y[X=1]))

(1/4)(mean(Y[X=5]) - mean(Y[X=1]))
+ 1.96* sqrt((1/4)^2*var(Y[X=5])/length(Y[X=5]) + (-1/4)^2*var(Y[X=1])/length(Y[X=1]))
```

## Part F
How could you calculate roughly the same thing as in Part E if we knew `sd(thetastar)` 
but not `Var(Y[X=5])` and `Var(Y[X=1])`?


# Problem 2

Let's reconsider what we were doing in Problem 1. In that problem, we were using a difference in means to estimate some population summary $\theta$ (theta). 
Suppose that, instead of that difference in means, we used the slope of a line fit by least squares to the data. That is, we use this point estimate.

$$ 
\frac{\hat \mu(5) - \hat \mu(1)}{4} = \frac{\qty{ \alpha + 5 \hat \beta} - \qty{\alpha + 1\hat\beta}}{4} = \hat\beta \qqtext { where } \hat\mu(x) = \hat \alpha + \hat \beta x 
$$

is the least squares fit of the least squares fit of the 'lines' model. This line.

```{r}
lines.fit = lm(Y~X, data=df)
muhat = function(x) { predict(lines.fit, newdata=data.frame(X=x)) }
x = seq(.5,5.5,by=.01)
ggplot() + geom_point(aes(x=X, y=Y), alpha=.15, position=position_jitter(w=.2, h=0)) + xlab('') + ylab('') +
           stat_summary(aes(x=X, y=Y), geom='pointrange', 
              fun=mean, 
              fun.min=function(x) mean(x)-sd(x),
              fun.max=function(x) mean(x)+sd(x))  +
           geom_line(aes(x=x, y=muhat(x))) 
```

This gives us a *biased* estimator of the same population summary. In **one of the two plots below**, I've shown the sampling distribution of that estimator, with its mean indicated by the thin solid line; dashed lines indicating the mean plus or minus two standard errors; and the population summary we're estimating indicated by a thicker black line. 

## Part A

Which plot is it? Mark it with an 'A'. 

```{r}
#| out-height: 3.5in
#| fig-ncol: 2

point.estimate = (muhat(5) - muhat(1))/4
se = sd(thetastar)
bias = point.estimate - (mean(Y[X==5]) - mean(Y[X==1]))/4

dist.plot = function(bias) { 
  theta = bias
  breaks = theta + seq(-5,5, by=1)
  x = theta  + seq(-5,5, by=.01)
    labels.abstract = case_when(
      breaks == theta  ~ 'theta',
      breaks == theta+1  ~ 'theta + se',
      breaks == theta-1 ~ 'theta - se',
      breaks > theta+1  ~ sprintf('theta + %s se', abs(breaks-theta)),
      breaks < theta-1  ~ sprintf('theta - %s se', abs(breaks-theta)))
ggplot() + 
    geom_area(aes(x=x, y=dnorm(x, mean=theta+bias, sd=1)), alpha=.1) +
    geom_vline(aes(xintercept=theta+bias)) +
    geom_vline(aes(xintercept=theta+bias+c(-2,2)), linetype='dotted') + 
    geom_vline(aes(xintercept=theta+bias+c(-1,1)), linetype='dashed') + 
    geom_vline(aes(xintercept=theta), color='black', size=1.5) +
    scale_x_continuous(breaks=breaks, limits=range(breaks), labels=labels.abstract) +
    theme(axis.text.x=element_text(angle=60)) +
    xlab('') + ylab('')
}
dist.plot(bias/se)
dist.plot(-bias/se)
```

## Part B 

Letting $\hat\theta$ be our point estimator, i.e. a draw from the correct sampling distribution above, explain how to calculate the coverage of the confidence interval $\hat\theta \pm 2\  \text{se}$ in terms of the sampling distribution's density $f(x)$.  Is the coverage closer to 0\%, 50\%, or 95\%?  You do not need to calculate the coverage any more precisely than that. Just explain how to do it.

::: {.callout-tip}
If you're uncertain of your answer to Part A, don't worry. Because of the symmetry of the normal distribution, coverage is the same in both possibilities plotted above.
:::

## Part C
Suppose the data we have is from a pilot study. We're considering re-running the study with 4 times more data. 
If we use intervals $\hat\theta \pm 2\  \text{se}$ based on the same point estimator, will our coverage be closer to 0\%, 50\%, or 95\%?  What if we used 16 times more data? You may assume the point estimator's bias is the same at all sample sizes and its standard error $\text{se}$ is directly proportional to $1/\sqrt{n}$, i.e., that $\text{se}$ is $1/2$ as big as it is above when we have 4 times more data and $1/4$ as big as it is above when we have 16 times more data. 

::: {.callout-tip} 
Consider drawing diagrams like the one above at these two sample sizes. 
:::

# Problem 3

Below, I've shown data from a randomized experiment on the effect of aspirin on headaches. Here $X$ is the self-reported severity of the headache
at the time of the initial complain on a scale of 1-5, $Y$ an analogous severity score an hour an hour after treatment with aspirin or a placebo, and $W$ a binary indicator of whether the patient was assigned to treatment. We've let them be a bit more precise after treatment, so $Y$ will not necessarily be an integer as $X$ was.

In this experiment, patients in our population were assigned to aspirin ($w=1$, triangles) or placebo ($w=0$, circles) with probability depending only on whether their pre-treatment headache severity ($x$) **was 3 or higher**. Then, to get the patients who actually participated in our experiment, we sampled with replacement from this population.

Here's a plot of the data. As usual, the bars show the proportion of patients with a given level of treatment with the corresponding value of $x$, i.e.
$$ 
\text{bar height} = N_{w,x} / N_w \qfor N_{w,x} = \sum_{i:W_i=w \text{ and } X_i=x} 1 \qand  N_w = \sum_{i: W_i=w} 1.
$$

```{r}
#| out-width: 100% 
 
set.seed(1) 
m = 1e6 
pi = function(x) { ifelse(x >= 3, .8, .2) }
mu = function(w, x) { 1 + (x-1)^1.7/2.7 - 3.5*w*(x-1)/10 }  

X = sample(1:5, size=m, replace=TRUE)
W = 0 + (runif(m) <= pi(X))
epsilon0 = rnorm(m)
epsilon1 = rnorm(m)
Y0 = mu(0,X) + .15*epsilon0
Y1 = mu(1,X) + .15*epsilon1

pop = data.frame(X=X, W=W, Y=Y0*(1-W) + Y1*W, Y0=Y0, Y1=Y1)

n = 1000
data = pop[sample(nrow(pop), n, replace=TRUE),]

plotX = function(W,X) { X + .1*(2*W-1) }
aspirin.plot = scatter = ggplot(data) + 
  geom_point(aes(x=plotX(W,X), y=Y, shape=factor(W)), alpha=.05, size=1, position=position_jitter(w=.075, h=0)) + 
  stat_summary(aes(x=plotX(W,X), y=Y, shape=factor(W)), geom="pointrange", alpha=.3, size=1,
                 fun=mean, 
                 fun.min = function(x) mean(x) - sd(x), 
                 fun.max = function(x) mean(x) + sd(x)) +
  geom_bar(aes(x=plotX(W,X), y=4*after_stat(prop)), 
          alpha=.5, linewidth=.2) + 
          xlab('') + ylab('') + 
          guides(shape='none', color='none', fill='none', linetype='none') +
          scale_x_continuous(breaks=seq(50,80,by=5)) + scale_y_continuous(breaks=seq(1,5,by=1))
aspirin.plot

estimate.ATT = function(formula, data) {
  model.fit = lm(formula, data=data)
  muhat = function(w,x) { predict(model.fit, newdata=data.frame(W=w, X=x)) }
  mean(muhat(1, data$X[data$W==1])) - mean(muhat(0, data$X[data$W==1]))
}

model.formulas = list(
horizontal = Y ~ W,
parallel = Y ~ W + X,
all = Y ~ W * factor(X),
piecewise = Y ~ W * I(X >= 3))

muhat.parallel = function(w,x) { predict(lm(Y~W+X, data=data), newdata=data.frame(W=w, X=x)) }
point.estimates = model.formulas |> map(\(formula) estimate.ATT(formula, data)) |> bind_rows(.id='model')
```

```{r}
width = function(samples, center=mean(samples), alpha=.05) {
  covg.dif = function(w) { 
    actual = mean( center - w/2 <= samples & samples <= center + w/2)
    nominal = 1-alpha 
    actual - nominal  
  }
  uniroot(covg.dif, interval=c(0, 2*max(abs(samples-center))))$root
}
bootstrap.widths = model.formulas |> map(\(formula) { 
  bootstrap.samples = replicate(100, { estimate.ATT(formula, data[sample(nrow(data), n, replace=TRUE), ]) })
  width(bootstrap.samples)
}) |> bind_rows(.id='model')
```

## Part A

Your cousin, who likes to be very formal and statistical, calculates a 95% confidence interval for the difference in mean outcomes for the subpopulations assigned to aspirin and placebo. For
$$
\hat \Delta = \frac{1}{N_1}\sum_{i:W_i=1} Y_i - \frac{1}{N_0}\sum_{i:W_i=0} Y_i,
$$
they report this interval estimate:
$$
\hat \Delta \pm 1.96 \ \text{se} = `{r} point.estimates$horizontal` \pm `{r} bootstrap.widths$horizontal/2`.
$$

On the basis of this interval estimate, they throw out their bottle of aspirin and go to the pharmacy to buy empty capsules to make placebos. Explain, in terms they can understand, why this is a mistake.

## Part B

Your uncle, who owns a very fancy adjustable-width ruler, decides to estimate the average treatment effect on the treated subpopulation (ATT) using the **parallel-lines regression model**. Here's their least squares prediction function, which they plot as a solid line for the placebo treatment ($w=0$)
and a dashed line for the aspirin treatment ($w=1$).

$$
\begin{aligned}
\hat\mu(w,x) &= \argmin_{m \in \model} \sum_{i=1}^n \qty{ Y_i - m(W_i,X_i) }^2 \qfor \\
\model &= \qty{ m(w,x) = a(w) + bx \text{ for functions } \ a(w) \ \text{ and numbers } \ b } 
\end{aligned}
$$

```{r}
#| out-width: 100%
grid = expand_grid(W=c(0,1), X=seq(min(data$X), max(data$X), by=.01))

aspirin.plot + 
  geom_line(aes(x=X, y=muhat.parallel(W,X), linetype=factor(W)), data=grid)
```

They report a point estimate of `{r} point.estimates$parallel`, which they calculated like this.

$$
\widehat{ATT} = \frac{1}{N_1}\sum_{i:W_i=1} \qty{ \hat\mu(1,X_i) - \hat\mu(0,X_i) } \qfor  N_w = \sum_{i: W_i=w} 1. 
$$

They bootstrap it to calculate a 95% confidence interval of $`{r} point.estimates$parallel` \pm `{r} bootstrap.widths$parallel/2`$.

They fish your cousin's aspirin bottle out of the trash and say that, while they shouldn't necessarily expect a `full point' of improvement on the 1-5 scale, they still think the aspirin is worth taking. Explain, in terms they can understand, why they're underestimating the effect.

## Part C (Extra Credit)

This was a bad question. Everyone gets free points.