---
title: "Midterm 2"
subtitle: "QTM 220"
---

```{r}
#| echo: false
library(tidyverse) 
library(scales)

theme_set(theme_minimal())
options(scipen = 1, digits = 2)
```

::: hidden
$$
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\Var}{V}
\DeclareMathOperator{\hVar}{\widehat{V}}
\DeclareMathOperator{\bias}{bias}
\newcommand{\model}{\mathcal{M}}
$$
:::

```{r}
#| echo: false
#| warning: false
library(ggplot2)
library(tidyverse)
library(patchwork)
```

# Problem 1

Suppose we have a sample of $(X_1,Y_1) \ldots (X_n,Y_n)$ drawn with replacement from a population. It's stored in an \textbf{R} dataframe `df' with $n$ rows. $X$ can take the values of 1,2,3,4 or 5. 
```{r}
#| echo: false
set.seed(123)
n = 1000
X = c(rep(1,350), rep(2,300),rep(3,200),rep(4,100),rep(5,50))
Y = X + rnorm(n,0,2)
df<- data.frame(x=X,y=Y)
```

Here's what the data looks like.

```{r}
#| out-height: 3.5in
#| fig-ncol: 2
#| fig-subcap:
#|  -  Histogram of the vector Y's elements
#|  -  Jittered Scatterplot of X and Y
ggplot() + geom_histogram(aes(x=Y,     y=after_stat(density)), alpha=.2, bins=50) + 
           geom_vline(aes(xintercept=mean(Y)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(Y)+c(-1,1)*sd(Y)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(Y), xmin=mean(Y)-1.96*sd(Y), xmax=mean(Y)+1.96*sd(Y), y=.025)) + theme_bw() + xlab('') + ylab('')

ggplot() + geom_point(aes(x=X, y=Y), alpha=.1, position=position_jitter(w=.2, h=0)) + xlab('') + ylab('') +
           stat_summary(aes(x=X, y=Y), geom='pointrange', 
              fun=mean, 
              fun.min=function(x) mean(x)-sd(x),
              fun.max=function(x) mean(x)+sd(x))  
```


We run this code.

```{r}
#| echo: true
B=1000

thetastar =  rep(NA,B)
for( b in 1:B ){
  Jstar = sample(1:n,size=n,replace=T)
  Xstar = X[Jstar]
  Ystar = Y[Jstar]
  thetastar[b] = (1/4)*(mean(Ystar[Xstar==5]) - mean(Ystar[Xstar==1]))
}
```

```{r}
#| out-height: 3.5in
#| fig-cap: Histogram of the thetastar's elements
ggplot() + geom_histogram(aes(x=thetastar, y=after_stat(density)), alpha=.2) + 
           geom_vline(aes(xintercept=mean(thetastar)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(thetastar)+c(-1,1)*sd(thetastar)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(thetastar), xmin=mean(thetastar)-1.96*sd(thetastar), xmax=mean(thetastar)+1.96*sd(thetastar), y=1)) + xlab('') + ylab('')

```


## Part A
What does this code calculate? State it in words and mark it with an 'a', and an arrow or bracket
if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
mean(Y)
```


## Part B
What does this code calculate? State it in words and mark it with a 'b', and an arrow or bracket if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
sd(Y)
```

## Part C
What does this code calculate? State it in words and mark it with a 'c', and an arrow or bracket
if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
mean(Y[X==1])
```

## Part D   
What does this code calculate? State it in words and mark it with a 'd', and an arrow or bracket if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
sd(thetastar)
```

## Part E
What does this code calculate? State it in words and mark it with a 'e', and an arrow or bracket if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
(1/4)(mean(Y[X=5]) - mean(Y[X=1])) 
- 1.96* sqrt((1/4)^2*var(Y[X=5])/length(Y[X=5]) + (-1/4)^2*var(Y[X=1])/length(Y[X=1]))

(1/4)(mean(Y[X=5]) - mean(Y[X=1]))
+ 1.96* sqrt((1/4)^2*var(Y[X=5])/length(Y[X=5]) + (-1/4)^2*var(Y[X=1])/length(Y[X=1]))
```

## Part F
How could you calculate roughly the same thing as in Part E if we knew `sd(thetastar)` 
but not `Var(Y[X=5])` and `Var(Y[X=1])`?


# Problem 2

Let's reconsider what we were doing in Problem 1. In that problem, we were using a difference in means to estimate some population summary $\theta$ (theta). 
Suppose that, instead of that difference in means, we used the slope of a line fit by least squares to the data. That is, we use this point estimate.

$$ 
\qty{\hat \mu(5) - \hat \mu(1)}/4 = \hat \beta_1/4 \qqtext { where } \hat\mu(x) = \hat \alpha + \hat \beta 
$$

is the least squares fit of the least squares fit of the 'lines' model. This line.

```{r}
lines.fit = lm(Y~X, data=df)
muhat = function(x) { predict(lines.fit, newdata=data.frame(X=x)) }
x = seq(.5,5.5,by=.01)
ggplot() + geom_point(aes(x=X, y=Y), alpha=.15, position=position_jitter(w=.2, h=0)) + xlab('') + ylab('') +
           stat_summary(aes(x=X, y=Y), geom='pointrange', 
              fun=mean, 
              fun.min=function(x) mean(x)-sd(x),
              fun.max=function(x) mean(x)+sd(x))  +
           geom_line(aes(x=x, y=muhat(x))) 
```

This gives us a *biased* estimator of the same population summary. In **one of the two plots below**, I've shown the sampling distribution of that estimator, with its mean indicated by the thin solid line; dashed lines indicating the mean plus or minus two standard errors; and the population summary we're estimating indicated by a thicker black line. 

## Part A

Which plot is it? Mark it with an 'A'. 

```{r}
#| out-height: 3.5in
#| fig-ncol: 2

point.estimate = (muhat(5) - muhat(1))/4
se = sd(thetastar)
bias = point.estimate - (mean(Y[X==5]) - mean(Y[X==1]))/4

dist.plot = function(bias) { 
  theta = bias
  breaks = theta + seq(-5,5, by=1)
  x = theta  + seq(-5,5, by=.01)
    labels.abstract = case_when(
      breaks == theta  ~ 'theta',
      breaks == theta+1  ~ 'theta + se',
      breaks == theta-1 ~ 'theta - se',
      breaks > theta+1  ~ sprintf('theta + %s se', abs(breaks-theta)),
      breaks < theta-1  ~ sprintf('theta - %s se', abs(breaks-theta)))
ggplot() + 
    geom_area(aes(x=x, y=dnorm(x, mean=theta+bias, sd=1)), alpha=.1) +
    geom_vline(aes(xintercept=theta+bias)) +
    geom_vline(aes(xintercept=theta+bias+c(-2,2)), linetype='dotted') + 
    geom_vline(aes(xintercept=theta+bias+c(-1,1)), linetype='dashed') + 
    geom_vline(aes(xintercept=theta), color='black', size=1.5) +
    scale_x_continuous(breaks=breaks, limits=range(breaks), labels=labels.abstract) +
    theme(axis.text.x=element_text(angle=60)) +
    xlab('') + ylab('')
}
dist.plot(bias/se)
dist.plot(-bias/se)
```

## Part B 

Letting $\hat\theta$ be our point estimator, i.e. a draw from the correct sampling distribution above, explain how to calculate the coverage of the confidence interval $\hat\theta \pm 2\  \text{se}$ in terms of the sampling distribution's density $f(x)$.  Is the coverage closer to 0\%, 50\%, or 95\%?  You do not need to calculate the coverage any more precisely than that---just explain how to do it.

::: {.callout-tip}
If you're uncertain of your answer to Part A, don't worry. Because of the symmetry of the normal distribution, coverage is the same in both possibilities plotted above.
:::

## Part C
Suppose the data we have is from a pilot study. We're considering re-running the study with 4 times more data. 
If we use intervals $\hat\theta \pm 2\  \text{se}$ based on the same point estimator, will our coverage be closer to 0\%, 50\%, or 95\%?  What if we used 16 times more data? You may assume the point estimator's bias is the same at all sample sizes and its standard error $\text{se}$ is directly proportional to $1/\sqrt{n}$, i.e., that $\text{se}$ is $1/2$ as big as it is above when we have 4 times more data and $1/4$ as big as it is above when we have 16 times more data. 

::: {.callout-tip} 
Consider drawing diagrams like the one above at these two sample sizes. 
:::