---
title: "Solutions: Midterm 2 Practice Problems"
format: pdf
editor: visual
---
```{r}
#| echo: false
#| warning: false
library(ggplot2)
library(tidyverse)
library(patchwork)
```


\section*{Problem 1}

In the following three parts, choose the figure with the appropriate regression line(s).
\subsubsection*{Part A}
Choose the figure with the solid line that represents the simple regression line ($y\sim x$).

```{r, fig.width=10,fig.height=4}
#| echo: false
#| warning: false
n <- 1000
zz <- c(rep(0,n/2),rep(1,n/2))
set.seed(123456)
xx <-  3 - 2*zz + rnorm(n,mean=0,sd = 1)
yy <- 2*zz*xx - xx + rnorm(n)


modAct1 <- lm(yy ~ xx)
modAct2 <- lm(yy ~ xx + zz)


df <- data.frame(xx,yy,zz)

p <- ggplot(data = df, aes(x = xx, y = yy)) +
  geom_point(aes(shape = factor(zz)),size = 1,alpha=.5) +
  scale_shape_manual(values = c(1,19), label = c("z=0","z=1")) +
  xlab("x") +
  ylab("y") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        legend.box.just = "left",
        legend.justification = c("left","top"),
        legend.position = c(0,1),
        legend.margin = margin(6,6,6,6),
        legend.title = element_blank(),
        legend.box.background = element_rect(color="black",size=1))

coef1 <- modAct1$coefficients
coef2 <- modAct2$coefficients
# simple regression line y ~ x --------
simpleA <- p + geom_abline(intercept = coef1[1], slope = coef1[2]) + ggtitle("A3")

simpleB <- p + geom_hline(yintercept = mean(df$yy)) + ggtitle("A1")

mod3 <- lm(yy ~ xx:zz, data = df)
simpleC <- p + geom_abline(intercept = coef(mod3)[1], slope = coef(mod3)[2]) +
  ggtitle("A2")


#pdf(file="Q1partAplot.pdf",width=10,height=4)
simpleB + simpleC + simpleA
#dev.off() 

```

\color{red} A3 \color{black}


\subsubsection*{Part B}
Choose the figure with the two dashed lines associated with the additive multiple regression ($y\sim x+z$).

```{r,fig.width=10,fig.height=4}
#| echo: false
# multiple additive regression y ~ x + z
addA <- p + geom_abline(intercept = coef2[1], slope = coef2[2], linetype = 'dashed') +
  geom_abline(intercept = coef2[1] + coef2[3], slope = coef2[2], linetype = 'dashed') +
  ggtitle("B1")

addB <- p + geom_vline(xintercept = 1, linetype = 'dashed') +
  geom_vline(xintercept = 3.25, linetype = 'dashed') +
  ggtitle("B2")

addC <- p + geom_curve(aes(x = -2,xend=6,y=-6,yend=-1),
               curvature = -0.8,
               angle = 70,
               linetype = 'dashed') +
  geom_curve(aes(x=-2,y=0,xend=4,yend=-7),
             curvature = -0.5,
             linetype = 'dashed') +
  ggtitle("B3")

#pdf(file="Q1partBplot.pdf",width=10,height=4)
addA + addB + addC
#dev.off()
```

\color{red} B1 \color{black}



\subsubsection*{Part C}
Choose the figure with the two dotted lines associated with the interactive multiple regression ($y\sim x*z$).

```{r,fig.width=10,fig.height=4}
#| echo: false
# interactive model
mod3 <- lm(yy ~ xx*zz, data = df)
coef3 <- coef(mod3)
intC2 <- p + geom_abline(intercept = coef3[1], slope = coef3[2],
                linetype = 'dotted') +
  geom_abline(intercept = coef3[1] + coef3[3],
              slope = coef3[2] + coef3[4],
              linetype = 'dotted') +
  ggtitle("C2")


intC1 <- p + geom_abline(intercept = coef3[1], slope = coef3[2]*1.5,
                linetype = 'dotted') +
  geom_abline(intercept = coef3[1] + coef3[3]*4,
              slope = coef3[2] + coef3[4]*4,
              linetype = 'dotted') +
  ggtitle("C1")

intC3 <- p + geom_curve(aes(x = -2,y = -2.5,xend=3,yend=5),
               curvature = 0.3,
               #angle = 30,
               linetype = 'dotted') +
  geom_curve(aes(x=0,y=5,xend=6,yend=-5),
             curvature = 0.3,
             angle=95,
             linetype = 'dotted') +
  ggtitle("C3")

#pdf(file="Q1partCplot.pdf",width=10,height=4)
intC1 + intC2 + intC3
#dev.off()
```


\color{red} C2 \color{black}



\section*{Problem 2}
Suppose we have a sample $Y_1 \ldots Y_n$ stored on an \textbf{R} vector $Y$ of length $n$.
We run this code.

```{r}
#| echo: false
set.seed(123)
n = 1000
Y = rnorm(n)
```

```{r}
#| echo: true
B=10000
Y.bar = rep(NA,B)
for(b in 1:B) {
  Y.bar[b]= mean(sample(Y,size=n,replace=TRUE))
}
```

```{r}
#| echo: false
#| warning: false
#| out-height: 6in
#| fig-ncol: 2
#| fig-subcap:
#|  -  Histogram of the vector Y's elements
#|  -  Histogram of the Y.bar's elements
ggplot() + geom_histogram(aes(x=Y,     y=after_stat(density)), alpha=.3) + 
           geom_vline(aes(xintercept=mean(Y)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(Y)+c(-1,1)*sd(Y)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(Y), xmin=mean(Y)-1.96*sd(Y), xmax=mean(Y)+1.96*sd(Y), y=.15)) + theme_bw() + xlab('') + ylab('')
ggplot() + geom_histogram(aes(x=Y.bar, y=after_stat(density)), alpha=.3) + 
           geom_vline(aes(xintercept=mean(Y.bar)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(Y.bar)+c(-1,1)*sd(Y.bar)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(Y.bar), xmin=mean(Y.bar)-1.96*sd(Y)/sqrt(n), xmax=mean(Y.bar)+1.96*sd(Y)/sqrt(n), y=5)) + theme_bw() + xlab('') + ylab('')
```




\subsection*{Part a}
What does this code calculate? State it in words and mark it with an 'a', and an arrow or bracket if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
mean(Y) - 1.96* sd(Y)/sqrt(n)
mean(Y) + 1.96* sd(Y)/sqrt(n)
```


A 95\% confidence interval --- based on normal approximation --- for the mean of the population our sample $Y_1 \ldots Y_n$ is drawn from. This is indicated by the horizontal segment (with a dot in the middle) on the right histogram.
\vspace{.5in}


\subsection*{Part b}
How could you calculate roughly the same thing as in part a using 'sd(Y.bar)' and not 'sd(Y)'?


We can use the standard deviation of the bootstrap sampling distribution, rather than 'sd(Y)/sqrt(n)', as our estimate of standard error.

\begin{verbatim}
mean(Y) - 1.96 * sd(Y.bar)
mean(Y) + 1.96 * sd(Y.bar)
\end{verbatim}


\section*{Problem 3}
Consider the following \texttt{R} code:
\begin{verbatim}
1. lm(residuals(lm(y~z)) ~ residuals(lm(z~x)))
2. lm(y ~ residuals(lm(z~x)))
3. lm(residuals(lm(y~z)) ~ residuals(lm(x~z)))
4. lm(residuals(lm(y~z)) ~ x)
5. lm(residuals(lm(y~x)) ~ residuals(lm(x~z)))
6. lm(residuals(lm(y~x)) ~ z)
7. lm(residuals(lm(y~x)) ~ residuals(lm(z~x)))
8. lm(y ~ residuals(lm(x~z)))
 \end{verbatim}

\begin{itemize}
\item[a)]  Which one or more of the above will produce the coefficient on $x$ from the multiple regression of $y \sim x + z$?\\
\color{red} 3 and 8  \color{black}
\item[b)] Which one or more of the above will produce the same residuals as the residuals from the multiple regression of $y \sim x + z$?\\
 \color{red}  3 and 7 \color{black}
\end{itemize}


\section*{Problem 4}

$Y_i$ is a individual $i$'s income and $X_i$ is that individual's number of years of education. We take a random sample of individuals and record their years of education and income. Define $\hat{\mu}(16)$ as the subsample mean income among those with 16 years of education. Define the indicator function for $x=16$ as $$ 1_{16}(X) = \left\{  \begin{array} {ll}
1 & if \ x = 16 \\
0 & otherwise
\end{array}
\right. $$

Show that $\hat{\mu}(16)$ is an unbiased estimator of $\mu(16)$, the subpopulation mean income for those with 16 years of education. Justify your steps. 


\bigskip 

We want to show that $E[\hat{\mu}(16)] = \mu(16)$

Subsample mean income estimator: $\hat{\mu}(16) = \frac{\sum_{i=1}^n 1_{16} (X_i)Y_i }{\sum_{i=1}^n 1_{16} (X_i)}$ where $$ 1_{16}(X) = \left\{  \begin{array} {ll}
1 & if \ x = 16 \\
0 & otherwise
\end{array}
\right. $$ By law of iterated expectations, $E[\hat{\mu}(16)] = E[E[\frac{\sum_{i=1}^n 1_{16} (X_i)Y_i }{\sum_{i=1}^n 1_{16} (X_i)} | X_1....X_n]]$

$$ E[\frac{1}{\sum_{i=1}^n 1_{16} (X_i)}  \ E[\sum_{i=1}^n 1_{16} (X_i)Y_i  | X_1....X_n]]$$ because when we condition the indicator sum is a constant

$$ E[\frac{1}{\sum_{i=1}^n 1_{16} (X_i)}  \ \sum_{i=1}^n E[ 1_{16} (X_i)Y_i  | X_1....X_n]]$$ expectation of sum = sum of expectations

$$ E[\frac{1}{\sum_{i=1}^n 1_{16} (X_i)}  \ \sum_{i=1}^n 1_{16} (X_i) \ E[ Y_i  | X_1....X_n]]$$ $$ E[\frac{1}{\sum_{i=1}^n 1_{16} (X_i)}  \ \sum_{i=1}^n 1_{16} (X_i) \ E[ Y_i  | X_i]]$$ by irrelevance of independent conditioning variables

We know $E[Y_i|X_i] = \mu(X_i)$ by definition

$$ E[\frac{1}{\sum_{i=1}^n 1_{16} (X_i)}  \ \sum_{i=1}^n 1_{16} (X_i) \ \mu(X_i)]$$

for all $\mu(X_i)$ that are not $\mu(16)$ (i.e. $\mu(12)$ and $\mu(18)$), the indicator function will multiply them by 0 so we can remove them

$$ E[\frac{1}{\sum_{i=1}^n 1_{16} (X_i)}  \ \sum_{i=1}^n 1_{16} (X_i) \ \mu(16)]$$ $\mu(16)$ is a constant

$$ E[\frac{1}{\sum_{i=1}^n 1_{16} (X_i)}  \ \mu(16) \ \sum_{i=1}^n 1_{16} (X_i)]$$

The sums cancel

$$E[\mu(16)] = \mu(16) $$ as expectation of a constant is that constant

Thus the subsample mean of income among those with 16 years of education is an unbiased estimator of the subpopulation mean income among those with 16 years of education.



\section*{Problem 5}

In the below visualization, the x-axis is an individual's years of education, the y-axis is income, the red line is the least squares line, and the black dots indicate the mean income within each year of education. You have three options to summarize the change in income across an additional year of education: the average over years, the average weighted by the proportion of individuals with the years of education $P_x$, and the slope of the least squares line. Rank these options smallest to largest regarding the magnitude of the estimate you expect and explain your reasoning. 

\begin{figure}[H]
\includegraphics{test.png}
\caption*{Income and Years of Education}
\end{figure}

\section*{Problem 6}
Suppose we want $\gamma_1$ from $y_i = \gamma_0 + \gamma_1 x_i + \gamma_2 w_i + e_i$ but we run the regression $lm (y\sim x)$ without $w$. We have de-meaned our variables $x_i$ and $w_i$ such that $\bar{x}=0$ and $\bar{w}=0$. Will the coefficient on $x$ from $lm (y\sim x)$ be equal to $\gamma_1$? Derive what your coefficient on $x$ will be in terms of $\gamma$s and justify your steps. 


