---
title: "Lecture 1 CPS"
subtitle: "Descriptive Regression with a Binary Covariate"
format: 
  beamer:
     header-includes: |
       \setbeamertemplate{navigation symbols}{}
       \setbeamertemplate{footline}[page number]
editor: visual
---

```{r}
library(foreign)
library(ggplot2)
library(tinytex)

#students change working directory? 
#setwd("/Users/acuttne/Documents/GitHub/qtm220/lectures/lec1")

# for cps variable interpretation see https://www2.census.gov/programs-surveys/cps/datasets/2022/march/asec2022_ddl_pub_full.pdf
census.api.key = '8084f81e4f71e4d32638dac15f188b424f72d063'
library(cpsR)
library(dplyr)
library(magrittr)
#library(kableExtra)
library(scales)

#cps.data = cpsR::get_asec(year=2022, vars=c('h_numper', 'htotval', 'a_age', 'a_hga'), key=census.api.key)
cps.data = cpsR::get_asec(year=2022, vars=c('h_numper', 'htotval', 'a_age', 'a_hga','gereg', 'a_maritl', 'a_sex','peafever'), key=census.api.key)

all.earnings = data.frame(age=cps.data$a_age,
                          income=cps.data$htotval, 
                          education=case_match(cps.data$a_hga,
                                           0  ~ 0,   # child
                                           31 ~ 0,   # < grade 1
                                           32 ~ 4,   #  grade 1-4
                                           33 ~ 6,   #  grade 5-6
                                           34 ~ 8,   #  grade 7-8 
                                           35 ~ 9,   #  grade 9
                                           36 ~ 10,  #  grade 10
                                           37 ~ 11,  #  grade 11
                                           38 ~ 11,  #  grade 12 no diploma
                                           39 ~ 12,  #  high school grad
                                           40 ~ 13,  #  some college
                                           41 ~ 14,  #  associate degree (vocational)
                                           42 ~ 14,  #  associate degree (academic)
                                           43 ~ 16,  #  bachelors degree
                                           44 ~ 18,  #  masters degree
                                           45 ~ 20,  #  professional school degree
                                           46 ~ 20)) #  doctorate

#in.interval = function(X, interval) { interval[1] <= X & X <= interval[2] }
#earnings = all.earnings[all.earnings$age %>% in.interval(c(25,35)) & 
                        #all.earnings$education >= 11, ]

earnings <- all.earnings

earnings$raw.income = earnings$income
earnings$degree = factor(earnings$education >= 16, 
                     levels=c(FALSE, TRUE), 
                     labels=c('no 4-year degree','4-year degree')) 
rownames(earnings)=NULL


```

## Outline

```{=tex}
\begin{itemize}
\item Univariate Descriptive Statistics 
\item Descriptive Regression with a Binary Covariate
\end{itemize}
```
## What are descriptive statistics?

```{=tex}
\begin{itemize}
\item Descriptive statistics are summaries of the data we have. \pause
\item The data we have might be a sample (e.g., an exit poll) or it might be the whole population (e.g., administrative vote counts), sometimes known as a census. \pause
\item Having the whole population is becoming more common in the era of Big Data.\pause
\item Note: if you are interested in the causal effects of a treatment (a drug, a policy, etc.), you can never really have the whole population (we will discuss this in a few weeks).
\end{itemize}
```
## What makes a good summary?

What are the two things that make a good summary?

\vspace{.5cm}

```{=tex}
\visible<2>{
From the google definition:
\begin{enumerate}
\item brief
\item includes main points
\end{enumerate}}
```
## What makes a good descriptive statistic?

```{=tex}
\begin{enumerate}
\item brief: at most a few numbers \pause
\item includes main points: \pause
\begin{itemize}
\item sufficient to describe main characteristics of the data
\begin{itemize}
\item there is a technical definition of sufficiency that I can discuss in office hours for those interested
\end{itemize} \pause
\item minimizes the information lost by reducing the data
\begin{itemize}
\item I'm using ``information'' in a loose sense here, will formalize a bit later
\item reducing the data means taking the data, which are often represented by $n*v$ numbers (with $n$ the number of observations and $v$ the number of variables) and reducing to a descriptive statistic that is represented by a few numbers.
\end{itemize}
\end{itemize}
\end{enumerate} \pause
```
Enough forest. Trees (examples) are coming!

## 

\vspace{5cm}

\includegraphics[scale=.4]{figs/forest_tree_cartoon}

It is easy in a statistics class to focus on the specifics. I want you to remember why we are doing what we are doing.

## Outline

Univariate Descriptive Statistics

## Univariate Example: CPS Wage Data

```{=tex}
\begin{itemize}
\item Data on 2022 household income from the CPS 
\end{itemize}
```
```{=tex}
\begin{table}[h]
\begin{center}
\begin{tabular}{r|r}
Household & income   \\
\hline
1 & 44220  \\
2 & 82752 \\
\vdots & \vdots \\
152732 & 32300 \\
\hline
\end{tabular}
\end{center}
\end{table}
```
## Describing Univariate Data

```{r}
attach(earnings)
hist(income,freq=FALSE,col="blue",ylim=c(-0.0000001,.0000075), xlab="Income")
lines(density(income),col="red",lwd=2)
rug(income,lwd=1.5)
```

## Description/Summarization

```{=tex}
\begin{itemize}
\item Even for univariate data, we may want to focus on one or two characteristics of the population instead of describing it in its entirety. \pause
\item We typically focus on measures of center and spread.\pause
\item We will discuss the rationale for this over the coming weeks.
\end{itemize}
```
## Univariate Activity

```{r}
hist(income,freq=FALSE,col="blue",ylim=c(-0.0000001,.0000075),xlab="Income")
lines(density(income),col="red",lwd=2)
```

Find the rough location of the mean and median.

## Univariate Activity Solution

```{r}
hist(income,freq=FALSE,col="blue",ylim=c(-0.0000001,.0000075), xlab="Income")
lines(density(income),col="red",lwd=2)
points(x=mean(income),y=-.00000015,pch=2,cex=3,lwd=2,col="red")
points(x=median(income),y=-.00000015,pch="|",cex=3,lwd=2,col="red")
```


## Mean/Median Discussion

In pairs, discuss what the mean and median represent in the context of the household income variable. What does it mean for the mean to be larger than the median in this context?

## Mean as a balance point

\vspace{5cm}

\includegraphics[scale=.5]{figs/seesaw}

## Choosing a Descriptive Statistic

How might we choose between these descriptive statistics?

```{=tex}
\begin{enumerate}
\item brief: at most a few numbers \pause
\item includes main points: \pause
\begin{itemize}
\item sufficient to describe main characteristics of the data
\item minimizes the information lost by reducing the data
\end{itemize}
\end{enumerate}
```
## A Least Squares Summary

Suppose we want to pick a single number ($m$) for the middle of the data that summarizes all the values for $y$, by minimizing the sum of squared residuals (i.e., least squares). $$
SSR(\tilde{m}) = \sum_{i=1}^N (y_i - \tilde{m})^2
$$ \pause

Activity: Using calculus, derive the least squares estimator

```{=tex}
\begin{enumerate}
  \item Calculate the derivative of $SSR$ with respect to $\tilde{m}$
  \item Set the derivative equal to 0
  \item Solve for $m$
\end{enumerate}
```

## The Objective Function for SSR Income

```{r}
y <- income
mu.tilde <- seq(-2000000,2000000, by=500)
S <- function(mu.tilde,y){
	sum((y-mu.tilde)^2)
	}
S.mu.tilde <- rep(0,length(mu.tilde))
for(i in 1:length(mu.tilde)){
	S.mu.tilde[i] <- S(mu.tilde[i],y)
	}

plot(mu.tilde,S.mu.tilde,type="l",lwd=2,col="red",xlab= expression(tilde(m)),ylab=expression(paste("S(",tilde(m),")")),main="Sum of Squared Residuals")
```
## 

```{=tex}
\begin{enumerate}
\item \begin{align*}
SSR(\tilde{m}) 
  &= \sum_{i=1}^N (y_i -
  \tilde{m})^2\\
  &= \sum_{i=1}^N (y_i^2 - 2 y_i \tilde{m} + \tilde{m}^2)
\end{align*} 
 $$
\frac{\partial SSR(\tilde{m})}{\partial
  \tilde{m}} = \sum_{i=1}^N (-2 y_i + 2 \tilde{m})
$$
\end{enumerate}
```
## The Slope of the Tangent Line for SSR Income

```{r}
S1 <- function(mu.tilde,y){
	sum(-2*y + 2*mu.tilde)
}

plot(mu.tilde,S.mu.tilde,type="l",lwd=2,col="red",xlab= expression(tilde(m)),ylab=expression(paste("S(",tilde(m),")")),main="Sum of Squared Residuals")
abline(a=S(mu.tilde=1000000,y=y) - 1000000*S1(mu.tilde=1000000,y=y),b=S1(mu.tilde=1000000,y=y),lwd=2,col="blue")
text(x=1000000,y=0,labels="|",col="blue",cex=2)
```

## The Slope of the Tangent Line for SSR Income

```{r}
plot(mu.tilde,S.mu.tilde,type="l",lwd=2,col="red",xlab= expression(tilde(m)),ylab=expression(paste("S(",tilde(m),")")),main="Sum of Squared Residuals")
abline(a=S(mu.tilde=500000,y=y) - 500000*S1(mu.tilde=500000,y=y),b=S1(mu.tilde=500000,y=y),lwd=2,col="blue")
text(x=500000,y=0,labels="|",col="blue",cex=2)
```

## The Slope of the Tangent Line for SSR Income

```{r}
plot(mu.tilde,S.mu.tilde,type="l",lwd=2,col="red",xlab= expression(tilde(m)),ylab=expression(paste("S(",tilde(m),")")),main="Sum of Squared Residuals")
abline(a=S(mu.tilde=mean(income),y=y) - mean(income)*S1(mu.tilde=mean(income),y=y),b=S1(mu.tilde=mean(income),y=y),lwd=2,col="blue")
text(x=mean(income),y=0,labels="|",col="blue",cex=2)
```

## 

```{=tex}
\begin{enumerate}
\item \begin{align*}
SSR(\tilde{m}) 
  &= \sum_{i=1}^N (y_i -
  \tilde{m})^2\\
  &= \sum_{i=1}^N (y_i^2 - 2 y_i \tilde{m} + \tilde{m}^2)
\end{align*} 
 $$
\frac{\partial S(m)}{\partial
  \tilde{m}} = \sum_{i=1}^N (-2 y_i + 2\tilde{m})
$$

\item $$
0 = \sum_{i=1}^N (-2 y_i + 2 m)
$$
\item $$
m = \frac{1}{N}\sum_{i=1}^N  y_i 
$$
\end{enumerate}
\vspace{.1cm}
```
Therefore, the mean is a least squares summary.


## An Least Absolute Value Summary

Suppose we want to pick a single number ($m$) for the middle of the data that summarizes all the values for $y$, by minimizing the sum of absolute deviations. $$
SAV(\tilde{m}) = \sum_{i=1}^N |y_i - \tilde{m}|
$$ \pause

```{=tex}
\begin{itemize}
  \item Why is this harder to solve?
  \item Any guesses?
\end{itemize}
```
## Why we focus on means

```{=tex}
\begin{itemize}
\item Means often provide a ``good'' summary of the data (and it is relatively easy to tell when they provide bad summaries).  \pause
\item The properties of means are relatively easy to describe.  \pause
\item Randomized experiments only identify mean causal effects (as do regression and matching when the appropriate assumptions hold).
\end{itemize}
```
## Variance and Standard Deviation

```{=tex}
\begin{itemize}
\item Mean is the least squares predictor in terms of SSR.
\item How small is ``least'' (i.e., how much spread around the mean)?
\item Variance and Standard Deviation are useful transformations of SSR for the mean.
\end{itemize} \pause
```
Variance \begin{align*}
S^2 &= \frac{\sum_{i=1}^N (y_i-\bar{y})^2}{N-1} \\ 
&=  \frac{SSR}{N-1}
\end{align*} \pause

Standard Deviation \begin{align*}
S &= \sqrt{S^2}
\end{align*}

## Mean and Standard Deviation

```{r}
hist(income,freq=FALSE,col="blue",ylim=c(-0.0000001,.0000075), xlab="Income")
lines(density(income),col="red",lwd=2)
points(x=mean(income),y=0,pch=2,cex=3,lwd=2,col="red")
points(x=mean(income)-sd(income),y=0,pch="[",cex=2,lwd=2,col="red")
points(x=mean(income)+sd(income),y=0,pch="]",cex=2,lwd=2,col="red")
```

## Variance as turning force

\vspace{5cm}

\includegraphics[scale=.6]{figs/turning_effect_01}

## Why Variance and Standard Deviation?

```{=tex}
\begin{itemize}
\item Natural measures of accuracy for means.  \pause
\item Means and standard deviations are sufficient to describe the distribution of normally distributed data. \pause
\item For non-normal data, means and standard deviations provide information via Chebyshev's inequality and other concentration inequalities. 
\end{itemize}
```
## Normally-Shaped Histograms

\vspace{6cm}

\includegraphics[scale=.55]{figs/empiricalRule.png}

## Chebyshev for Non-Normal Histograms

```{r}
hist(income,freq=FALSE,col="blue",ylim=c(-0.0000001,.0000075),xlab="Income",main="At most 25% outside of 2 SDs")
lines(density(income),col="red",lwd=2)
points(x=mean(income),y=0,pch=2,cex=3,lwd=2,col="red")
points(x=mean(income)-2*sd(income),y=0,pch="[",cex=2,lwd=2,col="red")
points(x=mean(income)+2*sd(income),y=0,pch="]",cex=2,lwd=2,col="red")
```
## Special Case: Binary Variables

```{r}
earnings$fouryrdegree<- ifelse(earnings$degree=="4-year degree", 1,0)
```


```{=tex}
\begin{table}[h]
\begin{center}
\begin{tabular}{r|r}
Household & Education (4-year degree)   \\
\hline
1 & 0  \\
\vdots & \vdots\\
25 & 1 \\
26 & 0 \\
\vdots & \vdots \\
152732 & 0 \\
\hline
Proportion & 38983/152732\\
\end{tabular}
\end{center}
\end{table}
```
## Mean and Variance for 4-Year Degree Variable 

```{=tex}
\begin{align*}
\bar{y} &= \frac{1}{N}\sum_{i=1}^{N} y_i \\
&= \frac{0 + \dots+ 1+ 0 + \dots + 1 + 0 + \dots+ 0}{152732} \\
&= \frac{38983}{152732}\\
&\\
S^2 &= \frac{\sum_{i=1}^{N} (y_i-\bar{y})^2}{N-1} \\ 
&= \frac{(0-.255)^2 + \dots+ (1-.255)^2}{152732-1}  \\ 
&=  \left(\frac{152732}{152732-1}\right)[\bar{y} \cdot(1-\bar{y})]
\end{align*}
```
## Outline

Descriptive Regression with a Binary Covariate

