---
title: "QTM 220 Lecture 7"
subtitle: "Descriptive Statistics with Non-Binary Covariates"
format: 
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---

```{r load-libraries}
library(purrr)
library(tibble)
library(dplyr)
library(magrittr)
library(stringr)
library(ggplot2)
library(htmltools)
library(kableExtra)
library(scales)
library(cpsR)
library(mapdata)

options(knitr.kable.NA = '')

compact_number = label_number(accuracy=.1, scale_cut=cut_short_scale())
ellipsis <- function(df, nhead=5L, ntail=2) {
    rownames(df) = as.character(1:nrow(df))
    stopifnot("data.frame" %in% class(df))
    els <- rep(NA, ncol(df)) %>% 
        matrix(nrow=1, dimnames=list(NULL, names(df))) %>% 
        data.frame(stringsAsFactors=FALSE) %>% 
        tibble::as_tibble()
    rownames(els) = "⋮"
    dplyr::bind_rows(head(df, nhead), els, tail(df, ntail))
}

percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }
```

```{r ggplot-themes}
fontsize = 12
textcolor = "#888888"

class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
        panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x  = element_text(colour = textcolor, size=fontsize),
        axis.text.y  = element_text(colour = textcolor, size=fontsize),
        axis.title.x  = element_text(colour = textcolor),
        axis.title.y  = element_text(colour = textcolor, angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
              axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
           panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(simplified.theme)
```




```{r get-data}
#| cache: true
#| 
# for cps variable interpretation see https://www2.census.gov/programs-surveys/cps/datasets/2022/march/asec2022_ddl_pub_full.pdf
census.api.key = '8084f81e4f71e4d32638dac15f188b424f72d063'
cps.sampling.ratio = 1/25 # Not true. 1/2500 households are surveyed. But for the sake of plotting, it's ok.

california.state.code = 6
year = 2020
cps.data = results=get_asec(year=year, vars=c('pearnval', 'a_age', 'a_hga', 'gestfips', 'gtco', 'gtcbsasz'), key=census.api.key)
```

```{r transform-data}
#| cache: false
#| 
all.earnings = data.frame(state.code = cps.data$gestfips,
                          county.code = cps.data$gtco,
                          age=cps.data$a_age,
                          income=cps.data$pearnval,
                          education.code=cps.data$a_hga,
                          education=case_match(cps.data$a_hga,
                                           0  ~ 0,   # child
                                           31 ~ 0,   # < grade 1
                                           32 ~ 4,   #  grade 1-4
                                           33 ~ 6,   #  grade 5-6
                                           34 ~ 8,   #  grade 7-8 
                                           35 ~ 9,   #  grade 9
                                           36 ~ 10,  #  grade 10
                                           37 ~ 11,  #  grade 11
                                           38 ~ 11,  #  grade 12 no diploma
                                           39 ~ 12,  #  high school grad
                                           40 ~ 13,  #  some college
                                           41 ~ 14,  #  associate degree (vocational)
                                           42 ~ 14,  #  associate degree (academic)
                                           43 ~ 16,  #  bachelors degree
                                           44 ~ 18,  #  masters degree
                                           45 ~ 20,  #  professional school degree
                                           46 ~ 20)) #  doctorate

all.earnings$degree = factor(all.earnings$education >= 16, 
                     levels=c(FALSE, TRUE), 
                     labels=c('no 4-year degree','4-year degree')) 

coarsen = function(x, group, seed=0) {
  current.seed = .Random.seed
  set.seed(seed)
  x.coarsened = rep(NA, length(x))
  for(gg in unique(group)) {
    x.coarsened[group==gg] = runif(sum(group==gg), min(x[group==gg]), max(x[group==gg]))
  }
  set.seed(current.seed)
  x.coarsened
} 

in.interval = function(X, interval) { interval[1] <= X & X <= interval[2] }
earnings = filter(all.earnings, education >= 8 & 
                                state.code == california.state.code & 
                                age %>% in.interval(c(25,35)))
Y = earnings$income
X = earnings$education
```


```{r get-county-info}
# county info from https://en.wikipedia.org/wiki/List_of_counties_in_California
# converted to csv using https://wikitable2csv.ggor.de

to.number = function(s) { as.numeric(str_trim(gsub(",", "", s))) }
ca_county_info = read.csv('california_county_info.csv') %>%
  mutate(subregion = tolower(sub(' County', '', County))) %>%
  mutate(pop = to.number(Population..2022.)) %>%
  mutate(area = to.number(sub('sq.*', '', Area))) 

earnings.county = left_join(earnings, ca_county_info[,c('FIPS.code', 'subregion')], by=c('county.code' = 'FIPS.code'))
earnings$county = earnings.county$subregion
earnings$county[is.na(earnings$county)]='unknown'
```


```{r make-maps}
#| cache: true
#| 

# To give our survey recipients a bit more reality, we'll give them fake locations on a map
# And we'll make up a population of California residents that they're sampled from
# maps are based on code from https://eriqande.github.io/rep-res-eeb-2017/map-making-in-R.html

counties = map_data("county")
ca_county = counties %>%
  filter(region == "california")
ca_county_centers = ca_county %>%
  group_by(subregion) %>%
  summarize(county.lat = mean(lat), county.long=mean(long))

ca_county_more = left_join(ca_county_centers, ca_county_info, by="subregion")
ca_all = left_join(ca_county, ca_county_more, by="subregion")


ca.plot = ggplot(data=ca_all, mapping=aes(x=long, y=lat, group=group)) + 
  coord_quickmap() +
  geom_polygon(fill='gray', color = "white", alpha=.4) +
      scale_fill_gradient(trans="log10") + 
      theme_void()

ca.pop.plot = ggplot(data=ca_all, mapping=aes(x=long, y=lat, group=group)) + 
  coord_quickmap() +
  geom_polygon(aes(fill = pop/area), color = "white") +
      scale_fill_gradient(trans="log10") + 
      theme_void()

## we'll make up a population as follows
# 1. choose the number of people to be the number of people in our sample scaled up to account for the cps sampling ratio
# 2. for each, sample a county with probability proportional to its finite.population
# 3. for each, sample a latitude and longitude by taking a randomly weighted combination of the    
#    county's vertices. As long as the county is convex, these points will be inside the county.
# 4. we'll give them a degree too, chosen independently of all else from the empirical distribution on the sample
n = nrow(earnings) / cps.sampling.ratio

# takes a list of subregions, possibly with duplicates
# for each element, chooses random lat/long coordinates within the sample.subregion
# by taking a random convex combination of the subregion's vertices
sample.within.subregion = function(subregions) {
  output = data.frame(subregion=subregions, lat=NA, long=NA)
  for(subregion in unique(subregions)) { 
      vertices = ca_county[ca_county$subregion==subregion, c('lat','long')]
      U = sum(subregions==subregion)
      V = nrow(vertices)
      weights = abs(rt(U*V, 1))
      dim(weights) = c(U,V)
      weights = (1/rowSums(weights)) * weights
      output[subregions==subregion, c('lat', 'long')] = 
                 cbind(weights %*% vertices$lat, weights %*% vertices$long)
    }
    output
}

finite.population.subregion = sample(ca_county_more$subregion, size=n, replace=TRUE, prob=ca_county_more$pop)
finite.population = sample.within.subregion(finite.population.subregion)
finite.population$degree = sample(earnings$degree, nrow(finite.population), replace=TRUE)
finite.population$group=1 # for plotting

display.population = sample_n(finite.population, min(100000, nrow(finite.population)))
```


```{r make-sample}
# and we'll force the degree-distribution in the sample to match our survey responses
our.sample = sample_n(finite.population, nrow(earnings))
our.sample$degree=earnings$degree

pop.plot = ca.plot + geom_point(aes(x=long, y=lat), data=display.population, size=.02, alpha=.1) 
sample.plot = ca.plot + geom_point(aes(x=long, y=lat), data=our.sample, size=.2, alpha=.5)
pop.plot.inf = ca.pop.plot + geom_point(aes(x=long, y=lat), data=display.population, size=.01, alpha=.1)  
sample.plot.inf = ca.pop.plot + geom_point(aes(x=long, y=lat), data=our.sample, size=.2, alpha=.5)
```

```{r visualization-defaults}
scale_y = function(limits=c(-.3e5, 3e5)) { 
  scale_y_continuous(labels=label_dollar(), limits=limits)
}

years = 8:20
scale_x = function(limits=c(min(years)-.5, max(years)+.5)) { 
  scale_x_continuous(breaks=years, labels=sprintf('%dy', years), limits=limits)
}

dot.size = 2
```


```{r define-colors}
twocolor.red   = 'rgb(248,118,109)'
twocolor.green = 'rgb(0,191,196)'
```

## Today

::: {.hidden}
$$
% color defs aren't making it into MathJax
%\definecolor{twocolor-red}{RGB}{248, 118, 109}   % red   used by ggplot in plots w/ 2 colors 
%\definecolor{twocolor-green}{RGB}{0, 191, 196}   % green used by ggplot in plots w/ 2 colors
% so create commands for using those colors instead
\newcommand{\inred}[1]{\textcolor[RGB]{248,118,109}{#1}}
\newcommand{\ingreen}[1]{\textcolor[RGB]{0,191,196}{#1}}
\newcommand{\inblue}[1]{\textcolor{blue}{#1}}

\newcommand{\ind}[1]{1_{\{#1\}}}
\newcommand{\with}{\ingreen{\text{degree}}}
\newcommand{\without}{\inred{\text{no degree}}}

\DeclareMathOperator{\E}{E}degredegre
\DeclareMathOperator{\Var}{Var}
$$ 

:::

```{r scatter-plots}

earnings.jitter = ggplot(earnings) + geom_point(aes(x=education, y=income, color=degree), 
                              size=dot.size, position=position_jitter(seed=0,w=.3), alpha=.1) +
                   scale_x() + scale_y() + guides(color=FALSE)
                    
earnings.scatter = ggplot(earnings) + geom_point(aes(x=education, y=income, color=degree), 
                              size=dot.size, alpha=.1) +
               scale_x() + scale_y() + guides(color=FALSE) 
```

```{r}
#| out-width: 1290px
#| fig-asp: .32
earnings.jitter
```

- We're going to look at the relationship between income and education.
- We've done that before. When we did, we thought of education in dichotomous terms.
  - We had some people with 4-year degrees --- the [green dots]{.twocolor-green}.
  - We had some people without them --- the [red dots]{.twocolor-red}.
- Today, we'll be a bit more granular. We'll think of education in terms of *years of schooling*.
    - 8+  years  = finished middle school
    - 12+ years = high school diploma
    - 16+ years = 4-year college degree

::: aside
Ignore the gaps at 15 and 17 years for now. We'll come back to them.
:::

## Our Sample

- As before, we'll work with data from the 2022 Current Population Survey.
- We'll look at California residents age 25-35 with at least an 8th-grade education.
- This sample includes `r nrow(earnings)` people. 
  - To visualize it, I've marked each one on a map. 
  - To display some of the information we have, I've made a table.
- I've used color to emphasize the dichotomous view of education we've been using.
  
:::: columns

::: column
```{r map-sample}
#| fig-width: 5
#| out-height: 500px
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=our.sample, size=.1, alpha=.5) + guides(color=FALSE)
```
:::

::: column

```{r table-no-color}
tab = earnings[sample(nrow(earnings)), ]
tab$income = sprintf('$%dk', round(as.numeric(tab$income) / 1000))
ellipsis.tab = tab[, c('income', 'education', 'age', 'county')] %>% ellipsis(nhead=8,ntail=2)
ellipsis.tab %>%
  kable %>% kable_styling() %>% 
  row_spec(which(ellipsis.tab$education >= 16), background=twocolor.green) %>%
  row_spec(which(ellipsis.tab$education <  16),  background=twocolor.red) 
```
:::

::::

::: aside
The locations shown on the map are made up. They're not the actual locations of the people in the sample. \
The survey includes some location information for some people, but as you can see in the table, not for everyone.
:::

## Visualization

```{r}
#| out-height: 500px
earnings.jitter
```

- Our primary visualization will be a *scatter plot*. It shows a dot for each person in our sample. 
- The person's income is on the y-axis and their education on the x-axis.
- We add a bit of fuzz to the displayed x-coordinate to space the dots out.
  - But not so much that it's not clear what the real value of $X$.
  - This is called 'jittering'. Sometimes a plot like this is called a 'jitter plot'.
- As before, we can use color to highlight the categories of people [with]{.twocolor-green} and [without]{.twocolor-red} 4-year degrees.

## The Same Thing Without Fuzz

```{r}
#| out-height: 500px
earnings.scatter
```

## Summarization
```{r}
#| out-height: 500px

binary.summaries = earnings %>% 
          group_by(degree) %>% 
          summarize(mean=mean(income), 
                    sd=sd(income), 
                    meanx=mean(education), 
                    meanx.dichotomous=mean(range(education)))

earnings.jitter + geom_pointrange(aes(x=meanx, y=mean, ymin=mean-sd, ymax=mean+sd), data=binary.summaries)
```

- Looking at the dots alone, we can get rough answers to some simple questions.
  - Which group tends to make more money---people [with]{.twocolor-green} or [without]{.twocolor-red} 4-year degrees?
- But it's hard to be precise just eyeballing things. It helps to look at numerical summaries.
- I've overlaid --- in black --- the mean income $\pm$ one standard deviation for each group.
  - That's shown by the dot's position and its 'arms' on the y-axis.
  - Its position on the x-axis shows the *mean years of schooling* within the group.

## The Dichotomous Version

```{r}
#| out-height: 500px

ggplot(earnings) + geom_point(aes(x=coarsen(education, degree), y=income, color=degree), 
                              size=dot.size, position=position_jitter(seed=0,w=.3), alpha=.1) +
                   scale_x() + scale_y() + guides(color=FALSE) + 
                   geom_pointrange(aes(x=meanx.dichotomous, y=mean, ymin=mean-sd, ymax=mean+sd), data=binary.summaries)
```

- It's worth visualizing the information lost by dichotomizing education.
- To do this, we can look at a similar plot without that information. This one.
- In it, I've replaced the x-coordinates with random values that tell us nothing beyond group membership.
- People [without]{.twocolor-red} 4-year degrees get random values uniformly distributed between 8 and 14. 
    - That's the range of *years of schooling* in that group.
- People [with]{.twocolor-green} 4-year degrees get random values between 16 and 20. Same deal.

## Fine Comparisons

```{r}
#| out-height: 500px

earnings$foreground = TRUE
jitter.nocolor = function(earnings) { 
  ggplot(earnings) + geom_point(aes(x=education, y=income, alpha=foreground), 
                                size=dot.size, position=position_jitter(seed=0,w=.3), color='#F7766D') +
                                scale_x() + scale_y() + scale_alpha_manual(breaks=c(FALSE, TRUE), values=c(.01,.2)) + 
                                guides(color=FALSE, alpha=FALSE)
} 

summaries = earnings %>% 
          group_by(education) %>% 
          summarize(mean=mean(income),
                    sd=sd(income))
                   
jitter.nocolor(earnings) + geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries)
```

- Let's think about what we can do with the additional information we have now.
- One simple option is to work with the finest grouping we can.
- We can, e.g., look at the income mean and standard deviation for people with 8,9,10, etc. years of schooling.
- Or compare any pair of those groups. It's natural to look at groups that are in some sense adjacent.
    - 18 vs. 16: the value of a masters' degree (only) vs. a 4-year one (only).
    - 16 vs. 14: the value of a 4-year degree (only) vs. a 2-year degree (only).
    - 14 vs. 12: the value of a 2-year degree (only) vs. a high school degree (only).

## Fine Comparisons

```{r}
#| out-height: 500px
jitter.nocolor(earnings) + geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries)
```

- This is essentially the same as what we did when we dichotomized education.
  - We just 'forget' the other groups. We don't look at them.
  - That's great if we know exactly what we're interested in---and it's something like that.
- But it's not so great if we're a bit less focused.
  - That'd mean we might have to think about/report a lot of different numbers.
  - It can get to be too much. We're not summarizing enough.

## Coarsening

```{r}
#| out-height: 500px

groups = factor(1:4, 
               levels=1:4, 
               labels=c('no college degree', '2-year degree', '4-year degree', 'graduate degree'))

group = function(x) { 
  case_match(x,
    8:13  ~ groups[1],
    14    ~ groups[2],
    16    ~ groups[3],
    18:20 ~ groups[4])
}


coarsened.summaries = data.frame(X=X, Y=Y, G=group(X)) %>% 
  group_by(G) %>% 
  summarize(mean=mean(Y), sd=sd(Y), mean.X = mean(coarsen(X, G)))


ggplot() + geom_point(aes(x=coarsen(X,group(X)), y=Y), 
                      color='#F7766D', position=position_jitter(w=.3, seed=0), alpha=.3) +
           geom_pointrange(aes(x=mean.X, y=mean, ymin=mean-sd, ymax=mean+sd), data=coarsened.summaries) + scale_x() + scale_y()
```

:::: r-stack
::: {.fragment .fade-out data-fragment-index="0"}
- One remedy is to go back to working with coarser groupings. 
- We can make up own by *aggregating*---combining---the fine groups we have.
- Here, for example, is a breakdown in 4 groups.
    - &lt; 14 years: no college degree.
    - =14 years: 2-year degree. 
    - =16 years: 4-year degree.
    - &gt; 16 years: graduate degree.
:::
::: {.fragment .current-visible data-fragment-index="0"}
- This is more information than a two-group comparison but perhaps little enough to be manageable.
  - We can report the set of group-specific summaries.
  - Or visualize them in a plot like this one.
:::
::::

## Coarsening is Always Happening
```{r}
#| eval: false
#| echo: true 
                          education=case_match(cps.data$a_hga,
                                           0  ~ 0,   # child
                                           31 ~ 0,   # < grade 1
                                           32 ~ 4,   #  grade 1-4
                                           33 ~ 6,   #  grade 5-6
                                           34 ~ 8,   #  grade 7-8 
                                           35 ~ 9,   #  grade 9
                                           36 ~ 10,  #  grade 10
                                           37 ~ 11,  #  grade 11
                                           38 ~ 11,  #  grade 12 no diploma
                                           39 ~ 12,  #  high school grad
                                           40 ~ 13,  #  some college
                                           41 ~ 14,  #  associate degree (vocational)
                                           42 ~ 14,  #  associate degree (academic)
                                           43 ~ 16,  #  bachelors degree
                                           44 ~ 18,  #  masters degree
                                           45 ~ 20,  #  professional school degree
                                           46 ~ 20)  #  doctorate
```

:::: r-stack
::: {.fragment .fade-out data-fragment-index="0"}

- Often, the data we have is already coarsened to some degree.
  - Our data does not, for example, distinguish between people with 7 and 8 years of schooling.
  - Similarly, it doesn't tell us years for people with graduate degrees. We've made that up.
- And to fit it into our analysis, we often have to coarsen it further.
  - To fit education into our 'years of schooling' framework, we've grouped together people with ...
    - vocational and academic associate degrees. (12+2=14 years)
    - professional and doctoral degrees. (16 + 4ish ≈ 20 years)
  - Our choices about some of these made-up numbers create the gaps we see at 15,17,and 19 years.

:::
::: {.fragment .current-visible data-fragment-index="0"}

- Keeping all this in mind can be overwhelming.
- It's often a good idea to pick and abstraction --- like years of schooling --- and go with it.
- But it's important to come back to this at some point to make sure your conclusions mean what you think they do.
- Suppose you see a bigger income jump between '8' and 9 years of education than between 9 and 10.
  - That might be because the group with '8' years actually includes people with 7 and 8 years.
  - So many of the people with '8' years have 2 fewer years of schooling than the people with 9.

:::
::::

## Dichotomization

```{r}
#| out-height: 250px
#| fig-asp: .25
jitter.nocolor(earnings) + geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries)
```

- Often, when we want to report a numeric summary, we do resort to dichotomization.
- But *having* the more granular information allows us to choose the groups we want.
- There are many meaningful choices.
    - &gt;16: people with graduate degrees. 
    - &le;12: people who haven't been to college.

- **Activity.**  Put these comparisons into words and --- for mean income --- mathematical notation and **R** code.
    -  &gt;16 vs. 16 
    -  14 vs. &le;12


## A Coarsened Comparison: &gt;16 vs. 16

```{r}
#| out-height: 250px
#| fig-asp: .25

dsummaries = earnings[earnings$education >= 16, ] %>% 
          group_by(education > 16 ) %>% 
          summarize(mean=mean(income),
                    sd=sd(income), mean.X = mean(education)) 
dichotomous.gt16vs16 = ggplot(earnings) + geom_point(aes(x=education, y=income, color = education > 16, alpha=education >= 16),
                                size=dot.size, position=position_jitter(seed=0,w=.3)) +
                                geom_pointrange(aes(x=mean.X, y=mean, ymin=mean-sd, ymax=mean+sd), data=dsummaries) +
                                scale_x() + scale_y() + scale_alpha_manual(breaks=c(FALSE, TRUE), values=c(.01,.2)) + 
                                guides(color=FALSE, alpha=FALSE)
dichotomous.gt16vs16
```

### In words

::: fragment

People with graduate degrees vs. people with 4-year degrees (only). 
:::

## A Coarsened Comparison: &gt;16 vs. 16

```{r}
#| out-height: 250px
#| fig-asp: .25
dichotomous.gt16vs16
```

### In math

::: fragment

- Option 1. A compact, readable version
$$
\frac{\sum_{i:X_i > 16} Y_i}{\sum_{i:X_i > 16} 1} - \frac{\sum_{i:X_i = 16} Y_i}{\sum_{i:X_i = 16} 1} 
$$

- Option 2. A verbose version that's easier to use in probability calculations
$$
\frac{\sum_{i=1}^n \ind{18,20}(X_i) Y_i}{\sum_{i=1}^n \ind{18, 20}(X_i)} - \frac{\sum_{i=1}^n\ind{16}(X_i)Y_i}{\sum_{i=1}^n \ind{16}(X_i)} 
\ \ \text{ where } \ \ 1_S(x) = \begin{cases} 1 & \text{ if } x \in S \\ 0 & \text{ otherwise} \end{cases}.
\\
$$

:::

## A Coarsened Comparison: &gt;16 vs. 16

```{r}
#| out-height: 250px
#| fig-asp: .25
dichotomous.gt16vs16
```

### In code

::: fragment

- Option 1. A compact, readable version
```{r}
#| echo: true
mean(Y[X > 16]) - mean(Y[X==16])
```

- Option 2. A verbose version that's easier to combine with more abstract code
```{r}
#| echo: true
groups = list(a = c(18,20), b=16)
mean(Y[X %in% groups$a]) - mean(Y[X %in% groups$b]) 
```

:::

## A Coarsened Comparison: 14 vs &le;12

```{r}
#| out-height: 250px
#| fig-asp: .25
dsummaries = earnings[earnings$education %in% c(8:12, 14), ] %>% 
          group_by(education == 14) %>% 
          summarize(mean=mean(income),
                    sd=sd(income), mean.X = mean(education)) 
dichotomous.14vsle12 = ggplot(earnings) + geom_point(aes(x=education, y=income, color = education == 14, alpha=education %in% c(8:12,14)),
                                size=dot.size, position=position_jitter(seed=0,w=.3)) +
                                geom_pointrange(aes(x=mean.X, y=mean, ymin=mean-sd, ymax=mean+sd), data=dsummaries) +
                                scale_x() + scale_y() + scale_alpha_manual(breaks=c(FALSE, TRUE), values=c(.01,.2)) + 
                                guides(color=FALSE, alpha=FALSE)

dichotomous.14vsle12
```

### In words

::: fragment
People with 2-year degrees (only) vs. people who haven't completed a year of college.
:::

## A Coarsened Comparison: 14 vs &le;12

```{r}
#| out-height: 250px
#| fig-asp: .25
dichotomous.14vsle12
```

### In math

::: fragment

- Option 1. A compact, readable version
$$
\frac{\sum_{i:X_i = 14} Y_i}{\sum_{i:X_i = 14} 1} - \frac{\sum_{i:X_i \le 12} Y_i}{\sum_{i:X_i \le 12} 1} \\
$$

- Option 2. A verbose version that's easier to use in probability calculations 

$$
\frac{\sum_{i=1}^n \ind{14}(X_i) Y_i}{\sum_{i=1}^n \ind{14}(X_i)} - \frac{\sum_{i=1}^n\ind{8 \ldots 12}(X_i)Y_i}{\sum_{i=1}^n \ind{8 \ldots 12}(X_i)} 
\ \ \text{ where } \ \ 1_S(x) = \begin{cases} 1 & \text{ if } x \in S \\ 0 & \text{ otherwise} \end{cases}.
$$

:::

## A Coarsened Comparison: 14 vs &le;12

```{r}
#| out-height: 250px
#| fig-asp: .25
dichotomous.14vsle12
```

### In code

::: fragment

- Option 1. A compact, readable version
```{r}
#| echo: true
mean(Y[X == 14]) - mean(Y[X<=12])
```

- Option 2. A verbose version that's easier to combine with more abstract code
```{r}
#| echo: true
groups = list(a = c(14), b=8:12)
mean(Y[X %in% groups$a]) - mean(Y[X %in% groups$b]) 
```

:::

## The Relationship Between Coarse and Fine Means

```{r}
#| out-height: 250px
#| fig-asp: .25
dichotomous.14vsle12 + geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), 
                                       color='blue', alpha=.5, data=summaries[summaries$education %in% c(8:12),])  
```


- We can express the means in our *aggregate groups* in terms of the means in our *fine groups*.
- Let's define some notation for describing our fine groups to make this easier.
  - $N_x$ will be the number of people with $X=x$.
  - $\hat \mu(x)$ will be the mean income of people with $X=x$.

$$
\hat \mu(x) = \frac{1}{N_x}\sum_{i:X_i=x} Y_i \quad \text{ where } \quad N_x = \sum_{i:X_i=x} 1
$$

- **Q**. What, on the plot, corresponds to $\hat\mu(14)$? What about $N_{12}$?

::: fragment

- **A**. The black dot in the $x=14$ column. And the number of red dots in the $x=12$ column.
:::

## The Relationship Between Coarse and Fine Means

```{r}
#| out-height: 250px
#| fig-asp: .25
dichotomous.14vsle12 + geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), 
                                       color='blue', alpha=.5, data=summaries[summaries$education %in% c(8:12),])  
```

- **Q**. What, in terms $\mu(x)$, is the mean outcome among people in our sample who did not attend college?
- **Strategy**. Where we see a sum over the people in the aggregate group,  we rewrite it as 
  - a *sum over the fine groups* 
  - of sums over the people in them.
  
::: fragment
- **Answer**. It's a weighted average of the fine-group means $\mu(x)$.
$$
\begin{aligned}
\frac{\sum_{i:X_i \in 8 \ldots 12} Y_i}{\sum_{i:X_i \in 8 \ldots 12} 1} 
&= \frac{\sum_{x \in 8 \ldots 12}\sum_{i:X_i=x} Y_i}{\sum_{x \in 8 \ldots 12}\sum_{i:X_i=x} 1} 
= \frac{\sum_{x \in 8 \ldots 12} N_x \times N_x^{-1}\sum_{i:X_i=x} Y_i}{\sum_{x \in 8 \ldots 12}N_x} \\
&= \sum_{x \in 8 \ldots 12} P_x \times \hat \mu(x)  \quad \text{ for } \quad P_x = \frac{N_x}{\sum_{x \in 8 \ldots 12}N_x}.
\end{aligned}
$$

- The weight $P_x$ is the proportion of the people in the aggregate group with $X_i=x$.

:::

## Aggregating Comparisons

```{r}
#| out-height: 250px
#| fig-asp: .25 
earnings$foreground = earnings$education <= 12
aggregate.comparison.plot = jitter.nocolor(earnings) + 
  geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries[summaries$education <= 12,]) +
  geom_line(aes(x=education, y=mean), data=summaries[summaries$education %in% c(8,12),], color='blue', alpha=.4) +
  geom_line(aes(x=education, y=mean), data=summaries[summaries$education<=12,], color='black', linetype='dashed') +
  scale_y(c(0, .7e5))
aggregate.comparison.plot
```

- Another option---rather than comparing aggregates---is aggregating comparisons.
- We can summarize the 'value' of each year of high school by the associated increment in mean income.
  - Visually, that's the slope of the dashed black line leading up to it.
- To reduce these to a single number, we can take the average of the increments.
  - Visually, that's the slope of the *secant*---the blue one. In a sense.
  - If that's not obvious to you, don't worry. We'll come back to it in a minute.
- Why in a sense? Because there are two things we might mean by 'the average of the increments'.
  
## Different Ways of Aggregating

```{r}
education.to.summary.row = rep(NA, max(summaries$education))
education.to.summary.row[summaries$education] = 1:nrow(summaries)
muhat = function(x) { 
  summaries$mean[education.to.summary.row[x]]
}
increments = muhat(9:12)-muhat(8:11)
average.increment = mean(increments)

hs.subsample = earnings$education[earnings$education %in% 9:12]
sample.average.increment = mean(muhat(hs.subsample) - muhat(hs.subsample-1))
proportions = table(hs.subsample) / length(hs.subsample)
```


```{r}
#| out-height: 250px
#| fig-asp: .25  
aggregate.comparison.plot + scale_x(c(8,12)) + scale_y(c(0, .7e5))
```


1. The average, over the four years of high school, of the increments. Let's write it out. 

::: fragment

$$
\frac{1}{4}\sum_{x=9}^{12} \qty{ \hat\mu(x) - \hat\mu(x-1) }
$$

- That's the slope of the *secant* from $\hat\mu(8)$ to $\hat \mu(12)$---the blue line.
  - To see this, we can write out our four terms and see what cancels. 
  - It helps to count backward from $x=12$.

:::

::: fragment
$$ 
\frac{ \qty{\hat \mu(12) - \hat\mu(11)} + \qty{ \hat\mu(11) - \hat\mu(10) } + \qty{ \hat \mu(10) - \hat\mu(9) } + \qty{ \hat \mu(9) - \hat\mu(8) } }{4}
= \frac{\hat\mu(12) - \hat\mu(8)}{4}
$$

- As we go from $x=8$ to $x=12$, the 'rise' is $\hat\mu(12)-\hat\mu(8)$ and the 'run' is $12-8=4$; slope= rise / run.

:::

## Different Ways of Aggregating

```{r}
#| out-height: 250px
#| fig-asp: .25  
aggregate.comparison.plot + scale_x(c(8,12)) + scale_y(c(0, .7e5))
```

2. The average, over people with &ge;1 year of high school, of the increments *affecting them* in their last year. 
    - That's something different. For the majority of these people, their last year was their fourth.
    - So most of the terms we're averaging are the increment from the third to fourth year. 

::: fragment

$$
\frac{\sum_{i:X_i \in 9 \ldots 12} \qty{ \hat\mu(X_i) - \hat\mu(X_i-1) }}{\sum_{i:X_i \in 9 \ldots 12} 1} 
=  \sum_{x \in 9 \ldots 12} P_x \ \qty{ \hat\mu(x) - \hat\mu(x-1) } \quad \text{ for } \quad P_x = \frac{N_x}{\sum_{x \in 9 \ldots 12}N_x}.
$$

We can try the 'cancellation trick' we used before, but it doesn't work very well.

:::

::: fragment

$$\small{ 
\begin{aligned}
& P_{12} \qty{ \hat\mu(12) - \hat \mu(11) } + P_{11} \qty{ \hat\mu(11) - \hat \mu(10) } + P_{10} \qty{ \hat\mu(10) - \hat \mu(9) } + P_{9}  \qty{ \hat\mu(9) - \hat \mu(8) } \\
&= P_{12}\hat\mu(12)  + (P_{11}-P_{12}) \hat\mu(11) + (P_{10}-P_{11}) \hat\mu(10) + (P_9 - P_{10}) \hat \mu(9)  - P_9\hat \mu(8)
\end{aligned}}
$$

- We do get a linear combination of the fine-group means, but it's not an interpretable one.
- It will, however, be useful in calculations next time when we're talking about inference.

:::


## Different Results

```{r}
#| out-height: 250px
#| fig-asp: .25  
aggregate.comparison.plot + scale_x(c(8,12)) + scale_y(c(0, .7e5))
```

$$
\small{
\begin{array}{r|rrrr}
\text{Grade Increment}  & 8 \to 9 & 9 \to 10 & 10 \to 11 & 11 \to 12 \\
\text{Mean Income Increment} & `r approx(increments[1])` & `r approx(increments[2])` & `r approx(increments[3])` & `r approx(increments[4])` \\
\text{Proportion of People} & `r approx(proportions[1])` & `r approx(proportions[2])` & `r approx(proportions[3])` & `r approx(proportions[4])` \\
\end{array}
}
$$

**Question.** Which is bigger---the average over years or the average over people?

::: fragment

**Answer.** The average over people is bigger. `r approx(average.increment)` vs `r approx(sample.average.increment)`.

:::    

## The Least Squares Slope

```{r}
#| out-height: 250px
#| fig-asp: .25  
earnings$foreground = earnings$education <= 12
model = lm(income ~ education, data=earnings[earnings$education <= 12,])

lm.predictions = data.frame(education=seq(8,12,by=.01))
lm.predictions$muhat = predict(model, newdata=lm.predictions)

aggregate.comparison.plot + scale_x(c(8,12)) + scale_y(c(0, .7e5)) +
    geom_line(aes(x=education, y=muhat), data=lm.predictions, color='red', alpha=.4)
```

- Another thing people use as a summary is the *slope of the least squares line*---the red line.
- **Q.** Is that the same as either of the two summaries we've been talking about? You may want to zoom in.

::: fragment

- **A.** No. We can see that visually. The slope of the least squares line is smaller than the secant slope.
- And the secant slope --- the average over years --- is smaller than the average over people.
  - least squares slope &lt; secant slope &lt; average over people
:::

## Calculating The Least Squares Slope

- The least squares line is the line that minimizes the sum of squared errors.

$$
\small{
\hat ax + \hat b = \operatorname*{argmin}_{a,b} \sum_{i} \qty(aX_i + b - Y_i)^2 \quad \text{ implicitly summing over } \quad i:X_i \in 8 \ldots 12
}
$$

- We can solve for $\hat a$ and $\hat b$ by taking derivatives and setting them to zero.
- And we'll find that the slope is the subsample covariance of $X$ and $Y$ divided by the subsample variance of $X$.

$$ 
\small{
\hat b = \frac{\frac{1}{N_{HS}}\sum_i (X_i - \bar X) Y_i}{\frac{1}{N_{HS}}\sum_i (X_i - \bar X)^2} \quad \text{ where } \quad \bar X = \frac{1}{N_{HS}}\sum_i X_i \quad \text{ and } \quad N_{HS} = \sum_i 1.
}
$$

- We can rewrite this as a weighted average of the subsample means $\hat\mu(x)$.

::: fragment

$$
\small{
\begin{aligned}
\hat b 
&= \frac{1}{N_{HS}}\sum_{i} w(X_i)Y_i =
\frac{1}{N_{HS}}\sum_{x\in 8\ldots 12} \sum_{i:X_i=x} w(x) Y_i 
\quad 
&& \text{ for } \quad w(x) = \frac{x - \bar X}{\frac{1}{N_{HS}}\sum_{i} (X_i-\bar X)^2} \\
&= \frac{1}{N_{HS}}\sum_{x\in 8 \ldots 12} w(x) N_x \ \times \  \frac{1}{N_x}\sum_{i:X_i=x} Y_i 
= \sum_{x \in 8 \ldots 12} \tilde w (x) \ \hat\mu(x)  \quad 
&& \text{ for } \quad \tilde w(x)
= \frac{N_x}{N_{HS}} \times w(x) 
= \frac{N_x (x-\bar X)}{\sum_{i} (X_i-\bar X)^2} 
\end{aligned}
}
$$
:::


## Our Weights

```{r}
#| out-height: 500px
P = function(x) { mean(hs.subsample == x) }
tw = function(x) { sum(hs.subsample==x)*(x - mean(hs.subsample)) / sum((hs.subsample - mean(hs.subsample))^2) }

years.weights  = c(1/4, 0, 0, 0, -1/4) 
people.weights = c(P(12), P(11)-P(12), P(10)-P(11), P(9)-P(10), -P(9))
lsq.weights    = c(tw(12), tw(11), tw(10), tw(9), tw(8))
x= rev(seq(8,12))

weights = data.frame(x=rep(x,3), 
                     w=c(years.weights, people.weights, lsq.weights), 
                     type=rep(c('over years', 'over people', 'least squares'), each=length(x)))


ggplot(weights) + geom_point(aes(x=x, y=w, color=type)) + 
                  geom_line(aes(x=x,  y=w, color=type), linetype='dotted') + 
                  scale_x(c(8,12)) + theme_update(legend.position = c(0.1, 0.1)) + scale_color_discrete(name="")
```

- We can see the weights we're using in the three summaries we've been talking about.
- If you look at them hard enough, there's a lot to see. But we won't go into it here.
- What's obvious is that they focus, to different degrees, on different comparisons.
  - The average over years puts all its emphasis on a long-term comparison: $\hat\mu(12)$ vs. $\hat\mu(8)$.
  - The average over people puts most of its emphasis on a short-term one: $\hat\mu(12)$ vs. $\hat\mu(11)$.
  - The least squares slope spreads emphasis over comparisons between $\hat\mu(12)$ and $\hat\mu(9) \ldots \hat \mu(11)$.

## The Impact of Variation in Slope

```{r}
#| out-height: 500px

earnings2 = earnings
earnings2$income[earnings2$education==10] = earnings2$income[earnings2$education==10]*3 
summaries2 = earnings2 %>% group_by(education) %>% summarize(mean=mean(income), sd=sd(income), mean.X = mean(education))

aggregate.comparison.plot.2 = jitter.nocolor(earnings2) + 
  geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries2[summaries2$education <= 12,]) +
  geom_line(aes(x=education, y=mean), data=summaries2[summaries2$education %in% c(8,12),], color='blue', alpha=.4) +
  geom_line(aes(x=education, y=mean), data=summaries2[summaries2$education<=12,], color='black', linetype='dashed') +
  scale_y(c(0, .7e5))

model = lm(income ~ education, data=earnings2[earnings2$education <= 12,])
lm.predictions = data.frame(education=seq(8,12,by=.01))
lm.predictions$muhat = predict(model, newdata=lm.predictions)

aggregate.comparison.plot.2 + scale_x(c(8,12)) + scale_y(c(0, .7e5)) +
    geom_line(aes(x=education, y=muhat), data=lm.predictions, color='red', alpha=.4)
```

- While our three summaries estimates are based on very different comparisons, we got similar answers.
- That was, fundamentally, because every comparison tells us the same thing. 
  - If you draw the lines from $\hat\mu(8) \ldots \hat \mu(11)$ to $\hat\mu(12)$, you'd get roughly the same slope.
- If that weren't the case, it'd be a very different story. It'd look more like this.
- In short, sometimes the choice of summary --- among qualitatively similar ones --- matters a lot. 
- Sometimes it doesn't. But that's not something you should bet much on.
- We'll talk more about choosing the right summary next week.

## Summary
```{r}
#| out-height: 500px
jitter.nocolor(earnings) 
```

- Working with non-binary covariates, e.g., years of schooling, opens up a lot of possibilities.
  - We can make and report comparisons on the finest level available, e.g. grade 12 vs grade 11.
  - We can work with coarser groups, e.g. grade 12 vs grade < 12, to simplify what we have to report.
  - We can make a number of fine comparisons and aggregate them into a simple summary.
- For the most part, ultimately we'll be looking at *linear combinations* of the subsample means.
- This will be convenient next time when we're seeing what the sample summaries tell us about our population.
- It means that, in abstract terms, we'll only need to understand the statistical behavior of one thing.