$$
\begin{aligned} 
\bar Y_{\with} &= \frac{1}{7} \sum_{i=1}^n 1(X_i=\with) Y_i \\
\bar Y_{\without} &= \frac{1}{3} \sum_{i=1}^n 1(X_i=\without) Y_i 
\end{aligned}
$$

## Iterated Means 
If we like, we can calculate means in two steps.
1. We can calculate means *within groups*.
2. We take a *group-size-weighted average* of them *across groups*.
This goes for both sample and population means.

:::: {.r-stack}



::: {.fragment .fade-out data-fragment-index="0"}
Suppose that in a sample of ten people and 7 have 4-year degrees. 
1. We calculate the mean income of our subsamples people with and without degrees.
$$
\begin{aligned} 
\bar Y_{\with} &= \frac{1}{7} \sum_{i=1}^n 1(X_i=\with) Y_i \\
\bar Y_{\without} &= \frac{1}{3} \sum_{i=1}^n 1(X_i=\without) Y_i 
\end{aligned}
$$
2. We take a weighted mean of these two averages.
$$
\begin{aligned} 
\bar Y &= \frac{7}{10} \bar Y_{\with} + \frac{3}{10} \bar Y_{\without} \\
       &= \frac{7}{10}\frac{1}{7}\sum_{i=1}^n 1_\with(X_i) Y_i + \frac{3}{10}\frac{1}{3}\sum_{i=1}^n 1_\without(X_i) Y_i \\
       &= \frac{1}{10} \sum_{i=1}^n 1_\with(X_i) Y_i + \frac{1}{10} 1_\without(X_i) Y_i \\
       &= \frac{1}{10} \sum_{i=1}^n \qty{ 1_\with(X_i) + 1_\without(X_i)} Y_i \\
       &= \frac{1}{10} \sum_{i=1}^n Y_i \\
\end{aligned}
$$
This gives us the mean of our whole sample.
:::

::: {.fragment .current-visible data-fragment-index="0"}
Suppose that in an infinite population, the probability of a person having a 4-year degree is 7/10.
1. We calculate the mean income of people with and without degrees.
$$
\begin{aligned} 
\mu_\with = \E[ Y \mid X_i=\with]    &= \\frac{1}{7} \sum_{i=1}^n 1(X_i=\with) Y_i \\
\mu_\without = \E[ Y \mid X_i=\without] &= \frac{1}{3} \sum_{i=1}^n 1(X_i=\without) Y_i 
\end{aligned}
$$



## The Indicator Trick

 - When we want to calculate the expectation of means in terms of *conditional means* use the law of iterated expectations, 
 the conditional expectation we get is a random variable. Like $\mu(X)=\E[Y \mid X]$.
   shows up. The random variable.
 - How do we make the *conditional expectation* for a fixed subpopulation,
    $\mu(x)=\E[Y \mid X=x]$, appear?
 - There's a trick to it. We multiply the conditional expectation by an indicator.
    - Let $1_x(X_i)$ be a random variable that's one when $X_i=x$ and zero otherwise.
    - If $\mu$ appears multiplied by this, it doesn't matter whether it's evaluated
      at the random variable $X_i$ or the constant $x$.

 $$
\mu(X_i)1_x(X_i) = \mu(x)1_x(X_i)
 $$
 


## The Indicator Trick

- When we're doing calculations, we often want to write *random values* of a function in terms of fixed values.
- There's a trick to it. To illustrate it, let's start with a dice analogy.
- Which do you prefer?
  1. Every time I roll a $5$, I pay you what I've rolled --- that many dollars. Otherwise, I pay you nothing.
  2. Every time I roll a $5$, I pay you five dollars. Otherwise, I pay you nothing.
- 
::: fragment
- They're the same. But they correspond to different mathematical formulas.
  1. The first is  $X_i 1_{x}(X_i)$.
  2. The second is 
  
   = x 1_{x}(X_i)$
2. $f(x)

It's the same.

:::
  

::: {.notes}

  - $\ind{x}(X_i)$ is an *indicator* that is 1 if $X_i=x$ and 0 otherwise.
  - More generally, $1_A(X_i)$ is in indicator that $X_i$ is in the set $A$.
  - In this case, we're talking about the set containing the single element $x$ 

:::

## An infinite-population approximation

:::: {.r-stack}

::: {.fragment .fade-out data-fragment-index="0"}

We can think of the people in our finite population as a large number of independent draws

  - from an infinite population of people 
  - meeting those critera
  - that *could exist*. 

Our sample is a small number of these people, chosen randomly.

```{r map-sample-and-infinite-pop}
#| layout-ncol: 2
#| layout-valign: bottom
sample.plot
pop.plot.inf
```
:::

::: {.fragment .current-visible data-fragment-index="0"}

This *implies* that our sample is a smaller number of draws from the same infinite population.

  - The infinite population isn't real; it's an abstraction we use to make some things simple.
  - We tend to estimate summaries of this (abstract) infinite population.
  - Then interpret them in relation to our (real) finite population.

Interpretation aside, we don't need the finite population at all.

```{r map-sample-over-infinite-pop}
#| layout-ncol: 2
#| layout-valign: bottom
sample.plot.inf + guides(fill=FALSE)
ca.pop.plot
```
:::

::::


::: {.notes}
Let's think about where the people in our infinite population tend to live.

- I've mapped the *probability density* of their locations.
- A person drawn from this infinite population, 
  i.e. one in our finite population,
  tends to live in areas of high density (SF,LA).
- A person in the sample *is* a person in the finite population;
  they'll live in the same places with the same probabilities.

:::

## Maybe something about finite vs infinite population summaries
  - maybe to clarify why they look different
    - one would look like a bigger sample average
    - the other one is going to be an expectation
  - perhaps say something about what makes the infinite population model convenient
    - identical distribution lets us think of every observation as (distributionally) the same
    - that makes it easier to summarize the population
    - e.g. there is a function $\mu(x)=\E[Y_i \mid X_i=x]$ that doesn't depend on $i$.
  - probably less that that, though
- A quick review of the binary case including
  - *descriptive statistics*: summaries of the data.
  - *inferential statistics*: summaries of a related population.


## Outline


- Descriptive statistics in many-valued case.
- Inferential statistics in the many-valued case.
    - A more calculation-focused review of the binary case.
    - Generalizations for the many-valued case 
      - We'll do this as a set of in-class exercises.
      - I'll introduce techniques that'll make this straightforward.
      

## 

:::: columns

::: column
### The sample
We'll look at the survey recipients who

  - were California residents 
  - between 25 and 35
  - who lived alone

This sample includes `r nrow(earnings)` people.
:::

::: column
### The finite population
It's roughly correct to think of them as  

- being chosen randomly
- with equal probability
- from the set of all people like this

This includes about `r compact_number(nrow(earnings) / cps.sampling.ratio)` people.
:::
::::



```{r map-subsamples}
#| layout-ncol: 2
#| layout-valign: bottom
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=our.sample, alpha=1)  + guides(color=FALSE)
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=display.population, alpha=.01) + guides(color=FALSE)
```

ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=display.population, alpha=.01) + guides(color=FALSE)


## Subsamples and subpopulations

:::: {.r-stack}

::: {.fragment .fade-out data-fragment-index="0"}
  - Our survey doesn't tell us much about location. We've made that up.
  - But it does tell us some things about education.
    - It tells us whether the respondent [has]{.twocolor-green} 
      or [doesn't have]{.twocolor-red} a four-year degree.
    - That's what we've been using so far.
  - This lets us break our sample into **subsamples** of people 
    with [with]{.twocolor-green} and [without]{.twocolor-red} degrees.
  - And think of our population in terms of corresponding **subpopulations**
    they're drawn from.
:::

::: {.fragment .current-visible data-fragment-index="0"}
  - The **subsample** of [green dots]{.twocolor-green} on the left are survey recipients [with 4-year degrees]{.twocolor-green}.
  - They're selected randomly from the **subpopulation** drawn as [green dots]{.twocolor-green} on the right.
    - i.e. the population of California residents, 25-35 and living alone, [with 4 year-degrees]{.twocolor-green}. 
  - Same deal with the [red dots]{.twocolor-red} and recipients [without 4-year degrees]{.twocolor-red}.
:::

::::

```{r map-subsamples}
#| layout-ncol: 2
#| layout-valign: bottom
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=our.sample, alpha=1)  + guides(color=FALSE)
```


::: {.fragment .fade-out data-fragment-index="0"}
  - But it does tell us some things about education.
    - It tells us whether the respondent [has]{.twocolor-green} 
      or [doesn't have]{.twocolor-red} a four-year degree.
    - That's what we've been using so far.
  - This lets us break our sample into **subsamples** of people 
    with [with]{.twocolor-green} and [without]{.twocolor-red} degrees.
  - And think of our population in terms of corresponding **subpopulations**
    they're drawn from.
:::

::: {.fragment .current-visible data-fragment-index="0"}
  - The **subsample** of [green dots]{.twocolor-green} on the left are survey recipients [with 4-year degrees]{.twocolor-green}.
  - They're selected randomly from the **subpopulation** drawn as [green dots]{.twocolor-green} on the right.
    - i.e. the population of California residents, 25-35 and living alone, [with 4 year-degrees]{.twocolor-green}. 
  - Same deal with the [red dots]{.twocolor-red} and recipients [without 4-year degrees]{.twocolor-red}.
:::

::::

```{r map-subsamples}
#| layout-ncol: 2
#| layout-valign: bottom
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=our.sample, alpha=1)  + guides(color=FALSE)
```

:: {.notes}

- Neither of these is a complete success.
  - The table is huge, so you can't look at it all at once.
      - This is just the top of it.
  - The scatter plot is dense, so you can't see individual points.
- But there are tricks to making them work.

:::


## Making Scatter 

:::: {.r-stack }

::: {.fragment .fade-out data-fragment-index="0"}
```{r binary-scatter-no-scaling}
#| out-height: 300px
dot.size = 5
earnings.scatter = ggplot(earnings) + 
  geom_point(aes(x=degree, y=income, color=degree), size=dot.size, alpha=.2) + 
  scale_y_continuous(labels=label_dollar()) + 
  guides(color=FALSE) + xlab('') + class.theme
earnings.scatter
```

- The major downside of scatter plots is that they can be dense.
  - I've plotted one dot for each person. 
  - They overlap to the point we can't see individual dots.
- Part of problem is wasted space. Everyone is clustered at the bottom.
- We can zoom in there, cutting off a few high-earners, to get a better look.

:::

::: {.fragment .current-visible data-fragment-index="0"}
```{r binary-scatter-scaled}
#| out-height: 300px
earnings.scatter + scale_y_continuous(labels=label_dollar(), limits=c(0,3e5))  
```

- Now I've cut off people who make more than \$300k
  - This is better, but the dots are still overlapping too much.
- That's still a problem of wasted space, but now the problem is mostly horizontal.
  - There are only two values of $X_i$, so we've got two vertical lines and a lot of blank space.
  - We can spread them out a little horizontally to get a better look. 
:::

::: {.fragment .current-visible data-fragment-index="1"}
```{r binary-jittered}
#| out-height: 300px
earnings.scatter.jittered =  ggplot(earnings) + 
  geom_point(aes(x=degree, y=income, color=degree), 
             position=position_jitter(w=.2), size=dot.size, alpha=.2) + 
  guides(color=FALSE) + xlab('') + class.theme
earnings.scatter.jittered + scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) 
```

- To use the horizontal space, we use a technique called *jittering*. 
  - We just give the points some arbitrary horizontal spread.
  - There's no meaning to it. It's just to use the space better.
- Now it's easier to see where there are more dots and where there are less.
- That can give you a rough sense of how much most people make.
:::

::::

## Answering Questions using Scatter Plots

```{r binary-scatter-jitter-2}
#| out-height: 300px
earnings.scatter.jittered + scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) 
```


- What does this plot tell us about who makes more?  
  - Is it people [with]{.twocolor-green} or [without]{.twocolor-red} four year degrees?
  - And roughly how much more do they make?

::: notes
- People [with 4-year degrees]{.twocolor-green} in the plot tend to make more than most [without]{.twocolor-red}. 
  - Now we're looking at all but a couple people.
  - So that's true about our sample as a whole.
- And it looks like they make roughly \$50k more.
  - Shifting up the dense cluster of [red dots]{.twocolor-red} by \$50k will
    roughly align it with the dense cluster of [green ones]{.twocolor-green}.
:::


## Summary Statistics

```{r binary-mean-sd}
summaries = earnings %>% 
          group_by(degree) %>% 
          summarize(mean=mean(income), sd=sd(income))


difference.in.means = as.numeric(summaries[summaries$degree=='4-year degree', 'mean'] - 
                                 summaries[summaries$degree=='no 4-year degree', 'mean'])
```

- Looking at the raw data helps us get a sense of what's going on.
- It's a good way to get rough answers to vague questions.
- But when we want to describe a sample or subsample precisely, we use *summary statistics*.
    - e.g., the mean and standard deviation
    - e.g., a histogram
- One way to pose the 'how much more' question precisely to ask for the difference in subsample means
- It's \$`r compact_number(difference.in.means)`. That's close to our visual estimate of \$50k.

## Mean and Standard Deviation

```{r binary-mean-sd-plot}
#| out-height: 250px
#| fig-asp: .3 
earnings.scatter.jittered + 
  geom_pointrange(aes(x=degree, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries) +
  ylim(-1e5, 4e5)
```

:::: {.r-stack}

::: {.fragment .fade-out data-fragment-index="0"}
### Means
- The subsample means are location summaries. 
  - They tell us where our subsamples are centered.
$$
\hat\mu(x) = \frac{1}{N_x} \sum_{i:X_i=x} Y_i \qfor N_x = \sum_{i:X_i=x} 1
$$
  - The black dots in our plot are these means.
  - We've got one for $x=\with$ and another for $x=\without$.

:::

::: {.fragment .current-visible data-fragment-index="0"}
### Standard Deviations
- The subsample standard deviations are spread summaries. 
  - They tell us how spread out our subsamples are.
$$
\hat\sigma(x) = \sqrt{ \frac{1}{N_x} \sum_{i: X_i=x}  \qty{ Y_i - \hat\mu(x) }^2 }  \qfor N_x = \sum_{i : X_i=x} 1
$$
  - It's the square root of the *mean squared distance* between the incomes in our subsample and their mean
  - The black bars in our plot show one standard deviation above and below the mean
  - We expect most of the incomes in our subsample to be within these bars.
:::

::::


## Issues

:::: columns

::: column
```{r binary-mean-sd-col}
earnings.scatter.jittered + 
  geom_pointrange(aes(x=degree, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries) +
  ylim(-1e5, 4e5)
```
:::

::: column
```{r binary-mean-sd-col-tall}
earnings.scatter.jittered + 
  geom_pointrange(aes(x=degree, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries) 
```
:::

::::

- Means and standard deviations are strongly affected by extreme observations
  - That's why the mean for [people with degrees]{.twocolor-green} is a bit
    bigger than the typical value we estimated by eye.
  - It's also why their standard deviation is so big.
- There are many alternatives, but we'll mostly focus on one this semester: histograms.

## Histograms

```{r}
#| out-height: 300px
#| fig-asp: .5 
earnings.histogram = ggplot(earnings) + 
  geom_histogram(aes(x=income, y=after_stat(count/sum(count))), alpha=.5, binwidth=25000) + 
  coord_flip() + facet_wrap(~degree, strip.position='bottom') + aes(fill=degree) +
  guides(fill=FALSE) + scale_y_continuous(labels=label_percent()) + ylab('') + class.theme
```

:::: {.r-stack}

::: {.fragment .fade-out data-fragment-index="0"}
```{r}
#| out-height: 300px
#| fig-asp: .5 
earnings.histogram + scale_x_continuous(labels=label_dollar(), limits=c(0,3e5)) 
```

- In these histograms, we show the fraction of people
  that make roughly a given amount of money.
- To be precise, we show the *subsample means*
  of **indicators** for different income ranges.
- The ranges we're using are $[0 \ldots 25k)$, $[25k \ldots 50k)$, $[50k \ldots 75k)$, etc.

:::

::: {.fragment .current-visible data-fragment-index="0"}

```{r}
#| out-height: 300px
#| fig-asp: .5 
earnings.histogram + scale_x_continuous(labels=label_dollar())
```

- What we get in one income range doesn't affect the others.
- That means our summaries of incomes in most ranges aren't
  affected by how much our high-income folks make.
- We just weren't showing the high-income folks in our last plot.

:::

::::

$$
\begin{aligned}
\text{frac}_{[a \ldots b)} &= \frac{1}{N_x}\sum_{i:X_i=x} 1_{[a \ldots b)}(Y_i) \qfor  1_{[a,b)}(y) = \begin{cases} 1 & \text{ if } a \leq y < b \\
                             0 & \text{ otherwise } \end{cases}
\end{aligned}
$$


## Population Summaries
 
 - Often, what we want isn't a summary of our sample.  We want to summarize our population instead.
    - e.g., we don't want the mean income of the [people with 4-year degrees]{.twocolor-green} on the left map---the ones we've actually surveyed. 
    - We want to know the now the mean income of [people with 4-year degrees]{.twocolor-green} on the right map---who we mostly haven't.
$$
$$

```{r map-subsamples-again}
#| layout-ncol: 2
#| layout-valign: bottom
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=our.sample, alpha=1)  + guides(color=FALSE)
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=display.population, alpha=.01) + guides(color=FALSE)
```

 - We can't do this perfectly because we're trying to summarize data we don't have.
 - But we can **estimate** these population summaries using our sample.

## Maybe something about finite vs infinite population summaries
  - maybe to clarify why they look different
    - one would look like a bigger sample average
    - the other one is going to be an expectation
  - perhaps say something about what makes the infinite population model convenient
    - identical distribution lets us think of every observation as (distributionally) the same
    - that makes it easier to summarize the population
    - e.g. there is a function $\mu(x)=\E[Y_i \mid X_i=x]$ that doesn't depend on $i$.
  - probably less that that, though


## Point and Interval Estimation 

- We've covered two types of estimate in this class.
  - A **point estimate** is a single number that we think is close to the population summary.
  - An **interval estimate** is a range of numbers that we think contains the population summary.
- We'll focus on point estimates today and interval estimates next class.


## Subsample and Subpopulation Means 

- We can think of *subsample means* as point estimates of the corresponding *subpopulation means*.
- When we do this, we're thinking of our sample as a random draw from our population.
- This means that summary statistics, like the subsample mean $\hat\mu(x)$, are random variables. 
  - Why? It means the observations in our sample are random variables.
  - And anything we calculate in terms of random variables is itself a random variable.

:::: columns

::: column
### Subsample Mean
$$
\hat\mu(x) = \frac{1}{N_x} \sum_{i:X_i=x} Y_i \qfor N_x = \sum_{i:X_i=x} 1
$$
:::

::: column
### (Infinite) Subpopulation Mean
$$
\mu(x) = \E[ Y_i \mid X_i = x]
$$

:::

::::

## Sampling Distributions

- A summary statistic, like all random variables, has a distribution.
- We call this its **sampling distribution**.
- That's the distribution the summary would have if we were to redraw the sample, over and over, from the population.
  - i.e., if we were to send out the Current Population Survey many times
  - each time randomly selecting the people who receive it survey the same way.
- In real life, we usually can't do that. 
  - That means we see only one draw from the sampling distribution.
  - We can, however, try to *estimate the sampling distribution* using only our sample.
- When we're simulating data, we can redraw our sample as many times as we like.
  - That means we can see what the sampling distribution looks like.
  - This allows us to check, to some degree, whether our estimates work.


  
## Sampling Distributions: A Picture
### TODO 
Sequence of histograms of some summary statistic
  - Maybe show K samples as scatter plots w/ mean plotted
  - And a histogram the K subsample means
  - e.g. for K=9, 9^2 [we can do 3x3 and 9x9 displays of the sample]


### Expectations

- Because a summary statistics is a random variable, it has a mean, standard deviation, etc.
- These are the mean and standard deviation of a draw from the summary's sampling distribution.

### TODO: illustrate this as an overlay on one of the sampling distribution pictures


## Subsample Means as Point Estimators
  - They are unbiased, i.e., their expectation is what we're trying to estimate.
  $$
  \E \hat\mu(x) = \mu(x)
  $$
  - They are convergent, i.e., they get closer to that expectation as the sample gets bigger.
    - In particular, their standard deviations are proportional to the inverse square root of the sample size.
  $$
  \begin{aligned}
  \sqrt{\Var \hat\mu(x)} = \E\qty{ \frac{ \sigma^2(x) }{\sqrt{N_x}} } \qfor \sigma^2(x) &= \Var(Y_i \mid X_i=x) %\\ &= \E \qty[ \qty{ Y_i - \mu(x) }^2 \mid X_i=x] 
  \end{aligned}
 $$ 
 - They are approximately normal, i.e., their sampling distribution is close to normal for large samples.
   - This is a consequence of the central limit theorem.
   - It's a good thing because it lets us base **interval estimates** on these normal approximations.

---
# Maybe use this stuff in the next class
# 
### Interval Estimates
#
#  - Estimate of where the pop summary is that accounts for uncertainty
#  - Based on the sampling distribution of our estimator
#  - Intention: in 95/100 replications, the interval estimate contains the pop summary
#  - Coverage: the fraction of replications where the interval estimate actually does contain
#  - We'll focus on point estimation today and interval estimation next class
---

## Comparing Means
- If we want to compare means for two subpopulations, we can do the same thing.
- The difference between the corresponding subsample means is an unbiased estimate.
$$ \E\qty{ \hat \mu(\with) - \hat \mu(\without) } = \mu(\with) - \mu(\without)  $$
- It's convergent and approximately normal, too.
- We'll review all this later today.

## Limitations of what we've done
  - We've done a fairly complete story of the relationship between income and education
    - We've talked about descriptive statistics, i.e., sample summaries
    - We've talked about inferential statistics, i.e., population summaries
  - But we've based all of this on very coarse idea of education
    - We've just asked whether people have a 4-year degree or not
  - This means there are really only a couple questions to answer
    - How much people with and without them usually make?
    - How different that is between the two groups.
    - How much income varies, overall or within these two groups.
  - We can say much more if we think of education as being many-valued.
    - e.g. 0 years, 1 year, 2 years, ... 20 years of formal education.
    - There are many more questions we can ask when we do that.
  - That's what we'll talk about now.

# Beyond Binary Comparisons: Descriptive Analysis




:::: columns

::: column

```{r short-binary-table-2}
style.table( earnings[random.rows, c('income', 'degree')] )
```

:::

::: column

- What does this table tell us about who makes more?  
    - Is it people [with]{.twocolor-green} or [without]{.twocolor-red} four year degrees?
    - Roughly how much more do they tend to make?

:::

::::

::: notes
- Of the randomly-chosen people we're looking at, 
  most [with 4-year degrees]{.twocolor-green} make more than most [without]{.twocolor-red}
- But the incomes we're seeing are all over the place, so
  - it's not clear that this holds up in the sample 
  - it's hard to quantify the difference
:::



# Junk after this point; maybe can be cut into something

# Question 1

##
```{r}
#earnings.scatter + scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) 
```

**Q**. Which group tends to make more money? 

::: {.fragment}
People with 4 year degrees tend to make more.
:::

# Question 2

##
```{r}
#earnings.histogram
```

**Q**. Roughly how much more?

::: {.fragment}
 Typically \$50k more, comparing the two peaks in the histogram.
:::

# Question 3

##

```{r}
#earnings.scatter + scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) 
```

How would we calculate the difference in **mean income** between people in our sample with and without 4-year degrees?

### Outline
::: {.fragment}
1. We calculate the mean income among the *subsample* [with]{.twocolor-green} 4-year degrees.
2. We do the same for the subsample [without]{.twocolor-red} degrees. 
3. We subtract.
:::

## Step 1

```{r}
earnings.scatter + scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) 
```

We calculate the mean income among the *subsample* [with]{.twocolor-green} 4-year degrees.

:::: {.r-stack}

::: {.fragment .current-visible data-frame-index="1"}

We can write it out in mathematical notation.

$$
\hat\mu(\text{with}) = \frac{\sum_{i : X_i=\text{with}} Y_i }{ \sum_{i : X_i=\text{with}} 1 }
$$
:::

::: {.fragment .current-visible data-frame-index="2"}
```{r}
Y = earnings$income
X = earnings$degree
```

Writing it in *R* actually gives us an answer.

```{r}
#| echo: true
mu.hat.with = mean(Y[X == '4-year degree'])
```

What we get is roughly \$`{r} as.integer(round(mu.hat.with, digits=-3))`.
:::

:::: 

## Step 2 
```{r}
earnings.scatter + scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) 
```

We do the same for the subsample [without]{.twocolor-red} degrees. 

:::: {.r-stack}

::: {.fragment .current-visible data-frame-index="1"}

We can write it out in mathematical notation.

$$
\hat\mu(\text{with}) = \frac{\sum_{i : X_i=\without} Y_i }{ \sum_{i : X_i=\without} 1 }
$$
:::

::: {.fragment .current-visible data-frame-index="2"}

Writing it in *R* actually gives us an answer.

```{r}
#| echo: true
mu.hat.without = mean(Y[X != '4-year degree'])
```

What we get is roughly \$`{r} as.integer(round(mu.hat.without, digits=-3))`.
:::

:::: 


## Step 3
```{r}
earnings.scatter + scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) 
```

We subtract.

:::: {.r-stack}

::: {.fragment .current-visible data-frame-index="1"}
Here's the math.
$$
\hat \Delta = \hat\mu(\text{with}) - \hat\mu(\text{without})  \vphantom{ \frac{\sum_{i : X_i=\text{without}} Y_i} { \sum_{i : X_i=\text{without}} 1 } }
$$
:::

::: {.fragment .current-visible data-frame-index="2"}
Here's the code.
```{r}
#| echo: true
difference = mu.hat.with - mu.hat.without
```

What we get is roughly \$`{r} as.integer(round(difference, digits=-3))`.
:::

::::

# Question 4

## 
```{r}
earnings.histogram
```
What *Summary of the Population* does this estimate?

::: fragment

- The difference in mean income among *subpopulations* with and without 4-year degrees. \
- That's a difference in conditional expectations.
$$
\Delta = \mu(\with)-\mu(\text{without}) \quad \text{ where } \quad \mu(x)=\E[Y_i \mid X_i=x]
$$
:::

# Question 5

##

$$\text{What is our estimator's bias,} \ \E \hat\Delta - \Delta ? $$

::: {.fragment }
#### It's zero. Why?
:::

$$
$$

:::: {.r-stack}

::: {.fragment .current-visible data-frame-index="2"}
#### Part of the story: subsample means are **unbiased** estimators of subpopulation means.

  - In particular, $\hat\mu(x)$ is an unbiased estiate of $\mu(x)$.
  $$
  \E\hat\mu(x)=\mu(x)
  $$
  - We proved this in Homework ? using the law of iterated expectations.
  
$$
TODO: Proof
$$

:::

::: {.fragment .current-visible data-frame-index="3"}
#### The rest: Expectation is **linear**. 

  - The expectation of a difference is the difference in expectations.
  - Subtracting unbiased estimates of our subpopulation means gives us \
    an unbiased estimate of the **difference in those subpopulation means**.
$$ 
\begin{aligned}
\E \hat\Delta 
&= \E \left\{ \hat\mu(\text{with}) - \hat\mu(\text{without}) \right\} && \text{definition} \\
&= \E \hat\mu(\text{with}) - \E \hat \mu(\text{without}) && \text{linearity} \\
&= \mu(\text{with}) - \mu(\text{without})  && \text{unbiasedness} \\
&= \Delta && \text{definition} 
\end{aligned}
$$
:::

::::

# Question 6

##

How can we approximate the *sampling distribution* of this estimator?

$$
$$

TODO: Picture of sampling dist vs approximation

### Outline 

::: fragment
1. We find normal approximations to the sampling distributions of each of our subsample means.
2. We characterize their difference as another normal distribution. 
3. We calculate its mean.
4. We calculate its variance.
:::

## Step 1

We observe that our subsample means are *approximately independent and normal* 
$$
\hat\mu(x) \approx Z_x \quad \text{ for } \quad Z_x \sim N\left( \E \hat\mu(x), \ \Var \hat\mu(x) \right)
$$
and
$$
Z_{x} \quad \text{ is independent of } \quad Z_{x'}  \quad  \text{ when } \quad x \neq x'. 
$$

They're approximately independent because our subsamples are *disjoint*, i.e., nobody is in both.


## Step 2

We'll approximate the difference in our subsample means by 
the corresponding difference of *independent normals*.
$$
\hat\mu(\text{with}) - \hat\mu(\text{without}) \approx Z_{\text{with}} - Z_{\text{without}}
$$
Then use the rule for subtracting independent normals: subtract the means; add the variances.
$$
\begin{aligned}
Z_{\text{with}} - Z_{\text{without}} &\sim  N\left( \mu_{\Delta}, \sigma^2_\Delta \right) \\
\text{ where } \quad    \mu_{\Delta} &= \E \hat\mu(\text{with})  - \E \hat\mu(\text{without}) \\
\text{ and  } \ \ \quad \sigma^2_\Delta &= \Var \hat\mu(\text{with}) + \Var \hat\mu(\text{without}) 
\end{aligned}
$$ 

## Step 3
We subtract the means, i.e., we calculate this difference. 
$$
\mu_{\Delta} = \E \hat\mu(\text{with})  - \E \hat\mu(\text{without}). 
$$
We know, from our last question, that this is the population summary we want.
$$\mu_{\Delta} = \hat\mu(\text{with})  - \hat\mu(\text{without}) = \Delta$$

## Step 4
We add the variances, i.e., we calculate this sum.
$$ 
\sigma^2_\Delta = \Var \hat\mu(\text{with}) + \Var \hat\mu(\text{without})
$$
The **variance** of each subsample mean $\hat\mu(x)$ is determined by 

1. The conditional variance of the terms in it
2. The size of the subsample.

$$ 
\sigma^2_\Delta = \frac{\Var(Y_i \mid X_i = \text{with})}{n_\text{with}} + \frac{\Var(Y_i \mid X_i = \text{without})}{n_{\text{without}}} 
$$

TODO: Do we define $n_\text{with}$ and $n_\text{with}$ properly as expectations of inverses of proportions?

## Summary.

We can approximate the sampling distribution of our estimator
by that of a *normal random variable* with the mean we want.
$$ 
\begin{aligned}
\hat\mu(\text{with}) - \hat\mu(\text{without}) &\approx Z_{\Delta} \quad \text{ for } \quad Z_{\Delta} \sim N(\Delta, \sigma^2_\Delta) \\
\text{ where } \quad \sigma^2_\Delta &= \frac{\Var(Y_i \mid X_i = \text{with})}{n_\text{with}} + \frac{\Var(Y_i \mid X_i = \text{without})}{n_{\text{without}}}
\end{aligned}
$$


# Question 7

##
How can we use this to calculate a 95\% confidence interval for this summary $\Delta$?

$$ $$

### Outline
::: fragment

1. We transform our estimate into a *pivot*: a random variable with an known sampling distribution.
2. We choose an interval that contains the pivot with 95\% probability.
3. We invert our transformation to get an interval for our estimate.

TODO: Fix up to deal with approximate pivots.
:::


## Step 1
t-stat calculation + picture of std normal reference distn

## Step 2
formula + picture of interval in reference distribution

## Step 3
formula + picture of transformed reference distributon

## Summary
something


## Most of the Remainder
We'll answer the questions

## Hopefully this fits at the end
- We'll explain why we might need an unsaturated model and how that results in bias.
- We'll imagine a reality in which we know whether there are 1 or 2 year masters degrees
  and think of the bias associated with grouping them for Question 3.
