---
title: "Working with Non-Binary Covariates"
subtitle: "QTM 220 Lecture 5"
format: 
  clean-revealjs:
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    show-notes: separate-page
    pdf-separate-fragments: true
    fig-format: svg
    fig.width: 10
    fig.asp: 2
    fig-align: left
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
execute: 
  cache: true
---

```{r setup}
library(purrr)
library(tibble)
library(dplyr)
library(magrittr)
library(stringr)
library(ggplot2)
library(htmltools)
library(kableExtra)
library(scales)
library(cpsR)
library(mapdata)

source('../../shared-code/plotting.R')
theme_set(simplified.theme)
set.seed(1)

compact_number = label_number(accuracy=.1, scale_cut=cut_short_scale())
```

```{r data-setup}
# for cps variable interpretation see https://www2.census.gov/programs-surveys/cps/datasets/2022/march/asec2022_ddl_pub_full.pdf
census.api.key = '8084f81e4f71e4d32638dac15f188b424f72d063'

california.state.code = 6
years = 2022
cps.data = map_dfr(years, function(year) {
  get_asec(year=year, vars=c('h_numper', 'htotval', 'a_age', 'a_hga', 'gestfips', 'gtcbsasz'), key=census.api.key)
})
```

```{r}
cps.sampling.ratio = 1/25000
all.earnings = data.frame(state.code = cps.data$gestfips,
                          age=cps.data$a_age,
                          income=cps.data$htotval, 
                          residents=cps.data$h_numper,
                          education.code=cps.data$a_hga,
                          education=case_match(cps.data$a_hga,
                                           0  ~ 0,   # child
                                           31 ~ 0,   # < grade 1
                                           32 ~ 4,   #  grade 1-4
                                           33 ~ 6,   #  grade 5-6
                                           34 ~ 8,   #  grade 7-8 
                                           35 ~ 9,   #  grade 9
                                           36 ~ 10,  #  grade 10
                                           37 ~ 11,  #  grade 11
                                           38 ~ 11,  #  grade 12 no diploma
                                           39 ~ 12,  #  high school grad
                                           40 ~ 13,  #  some college
                                           41 ~ 14,  #  associate degree (vocational)
                                           42 ~ 14,  #  associate degree (academic)
                                           43 ~ 16,  #  bachelors degree
                                           44 ~ 18,  #  masters degree
                                           45 ~ 20,  #  professional school degree
                                           46 ~ 20), #  doctorate
                        metro.size=case_match(cps.data$gtcbsasz, # cps reports a metro-size category, we map that to an approximate metro size
                                           0~0,       # < 100k
                                           2~1e5,     # 100-250k
                                           3~2.5e5,   # 250-500k
                                           4~5e5,     # 500k-1 mil
                                           5~1e6,     # 1-2.5 mil
                                           6~2.5e6,   # 2.5-5 mil     
                                           7~5e6))    # >= 5 mil  

in.interval = function(X, interval) { interval[1] <= X & X <= interval[2] }
earnings = all.earnings[all.earnings$residents==1 &
                        all.earnings$state.code == california.state.code &
                        all.earnings$age %>% in.interval(c(25,35)) &
                        all.earnings$education >= 8, ]

earnings$degree = factor(earnings$education >= 16, 
                     levels=c(FALSE, TRUE), 
                     labels=c('no 4-year degree','4-year degree')) 
```

```{r}
# To give our survey recipients a bit more reality, we'll give them fake locations on a map
# And we'll make up a population of California residents that they're sampled from

# maps are based on code from https://eriqande.github.io/rep-res-eeb-2017/map-making-in-R.html
# additional county info from https://en.wikipedia.org/wiki/List_of_counties_in_California
# converted to csv using https://wikitable2csv.ggor.de

to.number = function(s) { as.numeric(str_trim(gsub(",", "", s))) }

counties = map_data("county")
ca_county = counties %>%
  filter(region == "california")
ca_county_centers = ca_county %>%
  group_by(subregion) %>%
  summarize(county.lat = mean(lat), county.long=mean(long))
ca_county_info = read.csv('california_county_info.csv') %>%
  mutate(subregion = tolower(sub(' County', '', County))) %>%
  mutate(pop = to.number(Population..2022.)) %>%
  mutate(area = to.number(sub('sq.*', '', Area))) 
ca_county_more = left_join(ca_county_centers, ca_county_info, by="subregion")
ca_all = left_join(ca_county, ca_county_more, by="subregion")


ca.plot = ggplot(data=ca_all, mapping=aes(x=long, y=lat, group=group)) + 
  coord_quickmap() +
  geom_polygon(fill='gray', color = "white") +
      scale_fill_gradient(trans="log10") + 
      theme_void()

ca.pop.plot = ggplot(data=ca_all, mapping=aes(x=long, y=lat, group=group)) + 
  coord_quickmap() +
  geom_polygon(aes(fill = pop/area), color = "white") +
      scale_fill_gradient(trans="log10") + 
      theme_void()

## we'll make up a population as follows
# 1. choose the number of people to be the number of people in our sample scaled up to account for the cps sampling ratio
# 2. for each, sample a county with probability proportional to its finite.population
# 3. for each, sample a latitude and longitude by taking a randomly weighted combination of the    
#    county's vertices. As long as the county is convex, these points will be inside the county.
# 4. we'll give them a degree too, chosen independently of all else from the empirical distribution on the sample
n = nrow(earnings) / cps.sampling.ratio 
finite.population.subregion = sample(ca_county_more$subregion, size=n, replace=TRUE, prob=ca_county_more$pop)
finite.population = unique(finite.population.subregion) %>% 
  map_dfr(function(subregion) { 
    vertices = ca_county[ca_county$subregion==subregion, c('lat','long')]
    U = sum(finite.population.subregion==subregion)
    V = nrow(vertices)
    weights = abs(rt(U*V, 1))
    dim(weights) = c(U,V)
    weights = (1/rowSums(weights)) * weights
    data.frame(subregion=subregion,
               lat  = weights %*% vertices$lat,
               long = weights %*% vertices$long)
  })
finite.population$group=1
finite.population$degree = sample(earnings$degree, nrow(finite.population), replace=TRUE)

# to go easy on ggplot, we won't really display everbody 
display.population = sample_n(finite.population, 100000) 
# and we'll force the degree-distribution in the sample to match our survey responses
our.sample = sample_n(finite.population, nrow(earnings))
our.sample$degree=earnings$degree

pop.plot = ca.plot + geom_point(aes(x=long, y=lat), data=display.population, color='red', alpha=.01) 
sample.plot = ca.plot + geom_point(aes(x=long, y=lat), data=our.sample, color='red',    alpha=1)
pop.plot.inf = ca.pop.plot + geom_point(aes(x=long, y=lat), data=display.population, color='red', alpha=.01)  
sample.plot.inf = ca.pop.plot + geom_point(aes(x=long, y=lat), data=our.sample, color='red',    alpha=1)
```

## Overview

::: {.hidden}
$$
% color defs aren't making it into MathJax
%\definecolor{twocolor-red}{RGB}{248, 118, 109}   % red   used by ggplot in plots w/ 2 colors 
%\definecolor{twocolor-green}{RGB}{0, 191, 196}   % green used by ggplot in plots w/ 2 colors
% so create commands for using those colors instead
\newcommand{\inred}[1]{\textcolor[RGB]{248,118,109}{#1}}
\newcommand{\ingreen}[1]{\textcolor[RGB]{0,191,196}{#1}}

\newcommand{\ind}[1]{1_{\{#1\}}}
\newcommand{\with}{\ingreen{\text{degree}}}
\newcommand{\without}{\inred{\text{no degree}}}

\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\Var}{Var}
$$ 

```{r}
twocolor.red   = 'rgb(248,118,109)'
twocolor.green = 'rgb(0,191,196)'
```
:::

- Today, we'll work with data from the 2022 Current Population Survey.
- We'll look at the relationship between income and education.
- We've already compared people  [with]{.twocolor-green} and [without]{.twocolor-red} 4-year degrees.
- That's a way of summarizing what we see in the data based on a *binary covariate*
$$ 
X_i = \begin{cases} 
  \with & \text{ if } \ \text{person}_i \ \text{has a 4-year degree} \\
  \without & \text{ otherwise }
  \end{cases}
$$
- Today, we'll talk about summaries based on a *non-binary covariate*
- We'll categorize people by the years of education they've had

# Review

## 

:::: columns

::: column
### The sample
We'll look at the survey recipients who

  - were California residents 
  - between 25 and 35
  - who lived alone

This sample includes `r nrow(earnings)` people.
:::

::: column
### The finite population
It's roughly correct to think of them as  

- being chosen randomly
- with equal probability
- from the set of all people like this

This includes about `r compact_number(nrow(earnings) / cps.sampling.ratio)` people.
:::
::::



```{r}
#| layout-ncol: 2
#| layout-valign: bottom
sample.plot
pop.plot
```

::: {.notes}
Thinking about where these people might live can help us build intuition about our sample and population. 

- Each dot on the left map represents a person in the sample.
- Each dot on right map represents one in the population.

These locations are, to be clear, totally made up.

- The survey doesn't give us much location data.
- These made-up locations are just to make the people we talk about feel real.

:::

## An infinite-population approximation

:::: {.r-stack}

::: {.fragment .fade-out data-fragment-index="0"}

We can think of the people in our finite population as a large number of independent draws

  - from an infinite population of people 
  - meeting those critera
  - that *could exist*. 

Our sample is a small number of these people, chosen randomly.

```{r}
#| layout-ncol: 2
#| layout-valign: bottom
sample.plot
pop.plot.inf
```
:::

::: {.fragment .current-visible data-fragment-index="0"}

This *implies* that our sample is a smaller number of draws from the same infinite population.

  - The infinite population isn't real; it's an abstraction we use to make some things simple.
  - We tend to estimate summaries of this (abstract) infinite population.
  - Then interpret them in relation to our (real) finite population.

Interpretation aside, we don't need the finite population at all.

```{r}
#| layout-ncol: 2
#| layout-valign: bottom
sample.plot.inf + theme(legend.position="none")
ca.pop.plot
```
:::

::::


::: {.notes}
Let's think about where the people in our infinite population tend to live.

- I've mapped the *probability density* of their locations.
- A person drawn from this infinite population, 
  i.e. one in our finite population,
  tends to live in areas of high density (SF,LA).
- A person in the sample *is* a person in the finite population;
  they'll live in the same places with the same probabilities.

:::

## Subsamples and subpopulations

:::: {.r-stack}

::: {.fragment .fade-out data-fragment-index="0"}
  - Our survey doesn't tell us much about location. We've made that up.
  - But it does tell us some things about education.
    - It tells us whether the respondent [has]{.twocolor-green} 
      or [doesn't have]{.twocolor-red} a four-year degree.
    - That's what we've been using so far.
  - This lets us break our sample into **subsamples** of people 
    with [with]{.twocolor-green} and [without]{.twocolor-red} degrees.
  - And think of our population in terms of corresponding **subpopulations**
    they're drawn from.
:::

::: {.fragment .current-visible data-fragment-index="0"}
  - The **subsample** of [green dots]{.twocolor-green} on the left are survey recipients [with 4-year degrees]{.twocolor-green}.
  - They're selected randomly from the **finite subpopulation** drawn as [green dots]{.twocolor-green} on the right.
    - i.e. the population of California residents, 25-35 and living alone, [with 4 year-degrees]{.twocolor-green}. 
  
  - And therefore drawn from an **infinite subpopulation** of people like this who could exist.
  - Same deal with the [red dots]{.twocolor-red} and recipients [without 4-year degrees]{.twocolor-red}.
:::

::::

```{r}
#| layout-ncol: 2
#| layout-valign: bottom
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=our.sample, alpha=1) + 
          theme(legend.position="none")
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=display.population, alpha=.01) + 
          theme(legend.position="none")
```

## What we really have

- As an illustration, I've been drawing these people on a map.
- But that's fake. We don't have their addresses. 
- From here on, we'll focus on information we actually do have.
- Start where we've been...acting as if binary...

Two ways of looking at everything 
- Table, Scatter plot
- Long; Compressed
:::

::::

```{r}
sample.plot
```

## A Tabular View

:::: columns

::: column

```{r}
earnings.tab.rows = sample(1:nrow(earnings))[1:12]
earnings.tab = earnings[earnings.tab.rows, c('income', 'degree')]
earnings.tab$income = sprintf('$%dk', round(earnings.tab$income / 1000))
rownames(earnings.tab) = NULL
styled.tab = earnings.tab %>% 
  kable() %>%
  kable_styling() %>%
  row_spec(which(earnings.tab$degree=='4-year degree'), background=twocolor.green) %>%
  row_spec(which(earnings.tab$degree!='4-year degree'), background=twocolor.red) 
styled.tab
```
::: 

::: column

- We can make a table including all `r nrow(earnings)`, but it'd be big.
  - It'd be hard to look at it all at once.
  - But it's easy to look at a small part of it.
- I'm showing `r nrow(earnings.tab)` randomly-chosen people on the left. 


What does this table tell us about who makes more?  \
Is it people people [with]{.twocolor-green} or [without]{.twocolor-red} four year degrees?

::: fragment
- Most [with]{.twocolor-green} 4-year degrees make more than most [without]{.twocolor-red}. 
- But not all of them. 


We can get an idea of what's going on by looking at the table, \
but it helps to use other visualizations that show all the data.
:::

:::

::::



## A Scatter Plot

:::: {.r-stack }

::: {.fragment .fade-out data-fragment-index="0"}
```{r}
dot.size = 5
earnings.scatter = ggplot(earnings) + 
  geom_point(aes(x=degree, y=income, color=degree), size=dot.size, alpha=.2) + 
  scale_y_continuous(labels=label_dollar()) + 
  theme(legend.position="none") 
earnings.scatter
```
- Here I've plotted one dot per person.
- It's hard to see much because almost almost everyone is clustered at the bottom.
- We can zoom in there, cutting off a few high-earners, to get a better look.
:::

::: {.fragment .current-visible data-fragment-index="0"}
```{r}
earnings.scatter = ggplot(earnings) + 
  geom_point(aes(x=degree, y=income, color=degree), size=dot.size, alpha=.2) + 
  scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) + 
  theme(legend.position="none") 
earnings.scatter
```
- This is better, but it's still hard to see much.
- There are only two values of $X_i$, so the dots are all lying on top of one another. 
- We can spread them out a little horizontally to get a better look. 
:::

::: {.fragment .current-visible data-fragment-index="1"}
```{r}
earnings.scatter = ggplot(earnings) + 
  geom_point(aes(x=degree, y=income, color=degree), size=dot.size, position=position_jitter(w=.2), alpha=.2) + 
  scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) + 
  theme(legend.position="none") 
earnings.scatter
```

- It's called *jittering.* There's no meaning to the horizontal spread. It's just to make space.
- Now it's easier to see where there are more dots and where there are less.
- That can give you a rough sense of how much most people make.
:::

::::

## Descriptive Summarization
  - Mean [plot on scatter plot]
  - Standard Deviation [plot sd on scatter plot]
  - Histogram

## A Histogram

```{r}
earnings.histogram = ggplot(earnings[earnings$income<3e5, ]) + 
  geom_histogram(aes(x=income, y=after_stat(density)), alpha=.5, bins=15) + 
  coord_flip() + facet_wrap(~degree, strip.position='bottom') + aes(fill=degree) + 
  scale_x_continuous(labels=label_dollar()) + 
  theme(legend.position="none", axis.text.x = element_blank()) 
earnings.histogram
```

- So far this semester, we've used histograms instead of jittered scatter plots.
- In this histogram, the length of the bar at each income level  
  shows the fraction of people that make roughly that much.
- Compared to the scatter plot, we've traded dot density for bar length.
  - You can see what most people make by looking for the peak. That's easier. 
- But we'll be using scatter plots more going forward. 
  - They work better when we have more values of the covariate $X_i$.


## Statistical Summarization
 - Point estimates
 - Interval Estimates [last week]

 ## Point Estimate Properties
 - Unbiased (not in itself interesting, but ...)
 - Convergent: sd $\propto 1/\sqrt{n}$ 
 - Approx normal 
 - Review in Detail Later

## Interval Estimate Properties 
  - Coverage
  - More next time

## 






## Beyond Binary Comparisons

- Using two categories is throwing away a lot of information about education. \
- We can, for example, categorize people by the **years of education** they've had. 

```{r}
years = unique(earnings$education)
ggplot(earnings) + geom_point(aes(x=education, y=income, color=degree), 
                              size=dot.size, position=position_jitter(w=.2), alpha=.2) +
                   scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) + 
                   scale_x_continuous(breaks=years, labels=sprintf('%d years', years)) + 
                   theme(legend.position="none")
```

:::: {.r-stack}

::: {.fragment .fade-out data-fragment-index="0"}
### We haven't lost anything.

- You have a four year degree if, and only if, you have $12+4=6$ years of education.
- Thinking visually, we can get the plot we had before by shuffling the red and green dots.
- This means we can still see how much people with and without 4-year degrees tend to make.
- But not shuffling lets us see some new stuff, too. 

### What else is there to see?
:::

::: {.fragment .current-visible data-fragment-index="0"}

### What do you see, in purely visual terms?
$$
$$

### What does it mean?
$$
$$

:::

::::

::: {.notes}

### One thing I see
The centers of our subsamples are going up as we move to the right.

### Meaning
Even among people with (or without) a 4-year degree, 
people with more education tend to make more money.

:::




## Including Years of Education
- This does, of course, require that our survey asks about years of education. 
- And it does, more or less, so we can make a table like this.

```{r}
earnings.tab$education = sprintf('%d years', earnings$education[earnings.tab.rows])
earnings.tab %>% 
  kable() %>%
  kable_styling() %>%
  row_spec(which(earnings.tab$degree=='4-year degree'), background=twocolor.green) %>%
  row_spec(which(earnings.tab$degree!='4-year degree'), background=twocolor.red)
```
- The survey doesn't ask exactly this *how many years* question.
- It takes a little pre-processing to get this.

## Pre-processing

```{r}
#| echo: true
education=case_match(cps.data$a_hga,
                          0  ~ 0,   # child
                          31 ~ 0,   # < grade 1
                          32 ~ 4,   #  grade 1-4
                          33 ~ 6,   #  grade 5-6
                          34 ~ 8,   #  grade 7-8 
                          35 ~ 9,   #  grade 9
                          36 ~ 10,  #  grade 10
                          37 ~ 11,  #  grade 11
                          38 ~ 11,  #  grade 12 no diploma
                          39 ~ 12,  #  high school grad
                          40 ~ 13,  #  some college
                          41 ~ 14,  #  associate degree (vocational)
                          42 ~ 14,  #  associate degree (academic)
                          43 ~ 16,  #  bachelors degree
                          44 ~ 18,  #  masters degree
                          45 ~ 20,  #  professional school degree
                          46 ~ 20)  #  doctorate
```         

- We've had to make a few things up in translation.
  - There's no differentiation between a 1 or 2-year masters; we assume 2.
  - We're assuming professional and doctoral degrees take 4 years; that's approximate.
  - Nobody has 17 or 19 years in our plots; that's why.
- That's part of data analysis. 
- Ultimately, you'll want to be careful about quirks introduced by translation.
  - But it's important to keep your job manageable. 
  - It can help to ignore the quirks when you're getting started.


## Subsamples and Subpopulations

```{r}
years = unique(earnings$education)
ggplot(earnings) + geom_point(aes(x=education, y=income, color=degree), 
                              size=dot.size, position=position_jitter(w=.2), alpha=.2) +
                   scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) + 
                   scale_x_continuous(breaks=years, labels=sprintf('%d years', years)) + 
                   theme(legend.position="none")

```


:::: {.r-stack}

::: {.fragment .fade-out data-fragment-index="0"}  
- When we thought of education as binary, we only had two subsamples to talk about.
  - People with 4-year degrees --- the [green dots]{.twocolor-green}.
  - People without them --- the [red dots]{.twocolor-red}.
- Now that we're taking a more granular view, we can talk about *many*.
:::

::: {.fragment .current-visible data-fragment-index="0"}
- People with $x$ years of education for $x=10,11,\ldots$.
     - This is the *finest* way to break down our sample in terms of $X$.
- People with and without 4-year degrees, as before.
     - This is a coarser breakdown. They're *aggregates*.
     - To get these, we combine our finer subsamples.
- People whose highest degree is graduate, 4-year,  2-year, high-school, or none.
      - That's in between. It's a finer aggregate.

:::

::: {.fragment .current-visible data-fragment-index="1"}
- Naturally, it's the same for subpopulations.
- For each subsample, there's a corresponding subpopulation.

:::

::::


## New Questions

Taking this more granular view of education allows us to pose new questions.

1. What's the typical difference in income between people with *exactly* 16 years of education
   and *exactly* 12?
    - This is a different question than the one we addressed in review, but it has a similar structure.
    - We want a summary of income differences in two groups. Those can be subsamples or subpopulations.
    - We can calculate and relate these summaries exactly like we did in review.
2. What's the typical difference in income between people with 4-year degrees and people without them?
    - That's exactly the question we talked about in review.
    - We'll talk about how to calculate it a different way that'll generalize better to other questions.
    - We can check our work. Obviously our answers should be exactly the same as we got earlier.
3. How much change would we expect in a similar population in which people tend to get an additional year of education?
    - This is a question unlike the ones we talked about earlier.
    - We'll need to be a bit careful about what we mean.
    - But once we've gotten there, we can base our calculations on the ones we did in the last question.
4. Homework
    - Grad vs Ugrad

# Descriptive analysis

## Q1

```{r}
earnings$highest.degree = case_match(earnings$education,
                            0:11 ~ 'none',
                            12:13 ~ 'high school',
                            14  ~   'associates',
                            16  ~   'bachelors',
                            18:20 ~ 'graduate')

fine = earnings %>% group_by(education) %>% summarize(mean.income = mean(income))
coarse = earnings %>% group_by(degree) %>% summarize(mean.income = mean(income), lb=min(education), ub=max(education)) 
intermediate = earnings %>% group_by(highest.degree) %>% summarize(mean.income = mean(income), lb=min(education), ub=max(education))

base.plot = ggplot(earnings) + geom_point(aes(x=education, y=income, color=degree), 
                              size=dot.size, position=position_jitter(w=.2), alpha=.2) +
                   scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) + 
                   scale_x_continuous(breaks=years, labels=sprintf('%d years', years)) + 
                   theme(legend.position="none")

base.plot + geom_segment(aes(x=education-.2, xend=education+.2, y=mean.income, yend=mean.income), data=fine)
```



## Q2
```{r}
base.plot + geom_segment(aes(x=lb-.2, xend=ub+.2, y=mean.income, yend=mean.income), data=coarse) 
```

  - Means of Aggregates are Aggregates of Means
  - Weighted Aggregates
  - [do somethoing with fine picture combo]

## Q3

- some kind of picture showing jumps

# Population analyis: Review

## Unbiasedness Subsample and subpopulation means

- A reasonable starting point for describing income in a subsample is its mean income.
  - We'll think about the subsample with $X_i=x$ for some value $x$, 
  - e.g. $X_i=\with$ for people with 4-year degrees.
- We can think of our subsample's  mean as an estimate of the mean of the corresponding finite subpopulation.
- But we'll focus on interpreting it as an estimate of the mean of the *infinite subpopulation* from which they're both drawn.
$$
\begin{aligned}
\hat\mu(x) &= \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) Y_i \quad \text{ for } \quad N_x = \sum_{i=1}^n \ind{x}(X_i) && \text{subsample mean} \\
\mu(x) &= \E[Y_i \mid X_i=x] && \text{infinite subpopulation mean} \\
\end{aligned}
$$

- The subsample mean is an *unbiased estimate* of the subpopulation mean. 
- That means its expectation *is* the subpopulation mean: $\E \hat\mu(x) = \mu(x)$ 
- As a warm up exercise, we'll prove it
- We'll start by reviewing a few things from probability.

##  Properties of Population Means: Conditioning

### The law of iterated expectations
$$
\E\qty{ \E\qty( Y \mid X ) } \quad \text{ for any random variables $X, Y$} \\
$$
To calculate the mean of $Y$, we can average within subpopulations, then across subpopulations.

### Irrelevance of independent conditioning variables
$$ \E \qty( Y \mid X, X' ) = \E \qty( Y \mid X ) \quad \text{ when $X'$ is independent of $X,Y$ } $$
If $X'$ is unrelated to $X$ and $Y$, holding it constant doesn't affect the relationship between them.

## Properties of Population Means: Linearity

### The linearity of expectations
$$ 
\begin{aligned}
\E\qty( a Y + b Z ) 
&= \E\qty(aY) + \E\qty(bZ) \\
&= a\E(Y) + b\E(Z)  \quad \text{ for random variables $Y, Z$ and numbers $a,b$ }
\end{aligned}
$$
  There are two things going on here.

  - To average a sum of two things, we can take two averages and sum.
  - To average a constant times a random variable, we can multiply the random variable's average by the constant.
  - In other words, we can *distribute* expectations and can pull constants out of them.

### The linearity of *conditional* expectations
$$ 
\begin{aligned}
\E\qty{ a(X) Y + b(X) Z \mid X } 
&= \E\qty{a(X)Y \mid X} + \E\qty{ b(X)Z \mid X} \\
&= a(X)\E\qty(Y \mid X) + b(X)\E\qty(Z \mid X)
\end{aligned}
$$
   
This is like linearity of expectations, but with a twist.

   - When we condition on $X$, we're working with subpopulations in which $X$ is constant.
   - This means we can we can act as if functions of $X$ *are* constants.
   - So we can pull them out of expectations that are conditional on $X$.

## Thinking clearly about conditioning

- It's important to distinguish between two things
  1.  The *conditional expectation function* $\mu(x)=\E[Y \mid X=x]$.
    - $\mu$ is a function; evaluated at $x$, it's a number. It's not random.
    - It's the mean of the infinite subpopulation of people with $X=x$.
    - This is the thing we're trying to estimate with $\hat\mu(x)$.
  2.  The conditional expectation $\mu(X)=E[Y \mid X]$.
    - $\mu(X)$ is the mean of a *random* infinite subpopulation of people.
    - It's the conditional expectation function *evaluated at* the random variable $X$.
    - This is the sort of thing that shows up when we use the law of iterated expectations.
- Let's do an easy calculation to try to make this clear. 
- We'll calculate the mean income in our population using the law of iterated expectations. 

## Example: Conditioning
  
  - Let's think about a random person drawn from this population
    - $Y$ is their income 
    - $X$ is an indicator for having a four-year degree. 
  - Suppose the subpopulation means are [100k]{.twocolor-green} for people with degrees 
    and [50k]{.twocolor-red} for people without.
  - And that 7/10 of our population has degrees.
  
  What is $\E(Y \mid X)$?

::: fragment

  - It's a random variable that is *either* 100k or 50k. A *random value* of the function $\mu$.
    - It's [$\mu(\with)=100k$]{.twocolor-green} with probability $7/10$,   when $X=\with$. 
    - And it's [$\mu(\without)=50k$]{.twocolor-red} with probability $3/10$, when $X=\without$.
:::

What is $\E(Y)$?

::: fragment

  - It's the mean of this random variable. A number.
$$
\begin{aligned} 
E\qty{ E\qty( Y \mid X ) } 
&= E(Y \mid X=\with)P(X=\with) + E(Y \mid X=\without)P(X=\without) \\
&= 100k \cdot 7/10 + 50k \cdot 3/10 = 70k + 15k = 85k.
\end{aligned}
$$

:::


## Exercise: Unbiasedness of the sample mean
$$
\begin{aligned}
\text{Prove } \ \E \hat\mu(x) &= \mu(x) \qquad \text{ where } \\
\hat\mu(x) &= \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) Y_i \quad \text{ for } \quad N_x = \sum_{i=1}^n \ind{x}(X_i) \\
\mu(x) &= \E[Y_i \mid X_i=x] && 
\end{aligned}
$$


### Hint: The Indicator Trick

- Using the properties we've reviewed, we can write $\E \hat \mu(x)$ in terms of the random variable $\mu(X_i)$.
  - That's a *random value* of the conditional expectation function $\mu$.
  - We want an expression in terms of the value $\mu(x)$ it takes on when $X_i=x$. 
- In this context, it seems like they should be the same. Everyone in our subsample has $X_i=x$.
- To formalize this, we observe that $\mu(X_i)$ appears multiplied by the indicator $\ind{x}(X_i)$.
- And when that indicator is nonzero, $\mu(X_i)$ and $\mu(x)$ coincide.
$$
\mu(X_i)\ind{x}(X_i) = 
\begin{cases} 
\mu(X_i) & \text{ when } X_i=x \\
0 &      \text{ otherwise}
\end{cases} 
=
\begin{cases}
\mu(x)  & \text{ when } X_i=x \\
0       & \text{ otherwise }
\end{cases} 
= \mu(x)\ind{x}(X_i)
$$




::: {.notes}
$$
{\tiny
\begin{aligned}
\E \hat\mu(x) =& \E\qty{\frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) Y_i}  \\
\underset{iterated}{=}& \E\qty[ \E\qty{ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) Y_i \mid X_1 \ldots X_n} ] \\
\underset{linearity}{=}& \E\qty[ \frac{1}{N_x} \E\qty{ \sum_{i=1}^n \ind{x}(X_i) Y_i \mid X_1 \ldots X_n} ]  \\ 
\underset{linearity}{=}& \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \E\qty{ \ind{x}(X_i) Y_i \mid X_1 \ldots X_n} ]  \\ 
\underset{linearity}{=}& \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) \E\qty{ Y_i \mid X_1 \ldots X_n} ]  \\
\underset{irrelevance}{=}& \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) \E\qty{Y_i \mid X_i} ] \\
\underset{definition}{=}& \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) \mu(X_i) ]  \\
\underset{hint}{=}& \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) \mu(x) ] \\   
\underset{linearity}{=}& \mu(x) \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) ]   \\
\underset{definition}{=}& \mu(x) \E\qty[ \frac{1}{N_x} N_{x}    ] = \mu(x)\E[ 1 ] = \mu(x)   
\end{aligned}
}
$$



## The Indicator Trick in General

- It's useful to generalize the *indicator trick* we've just used.
- We can expand a function of a random variable $X$ as a sum of 
  - the value it takes at every distinct value of $X$
  - times an indicator that $X$ takes on that value
$$
f(X) = f(a) \ind{a}(X) + f(b) \ind{b}(X) + f(c) \ind{c}(X) + \ldots \qfor a\neq b \neq c \ldots.
$$

- In our exercise, we essentially used this for the function $\mu$.
- Multiplying by the indicator $1_{x}(X_i)$ *picked out* one term in the expansion.
$$ 
\begin{aligned}
\ind{x}(X_i) \mu(X_i) &= \ind{x}(X_i) \sum_{x'} \ind{x'}(X_i) \mu(x') \\
                      &= \sum_{x'} \ind{x}(X_i)\ind{x'}(X_i) \mu(x') = \mu(x) 
\end{aligned}
$$
- We'll be using this later on. 
- Sometimes we'll multiply nu an indicator that picks out multiple terms.


::: notes
It's easy to check this equivalence. 
  - When $X=a$, the expansion's first indicator 'hits' and the others don't. 
  - That means the expanded version is $f(a)$. That's right.
  - It's the same for $b,c,\ldots$.
:::

# Q1

# Q2

# Q3




# Junk after this point; maybe can be cut into something

# Question 1

##
```{r}
earnings.scatter + scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) 
```

**Q**. Which group tends to make more money? 

::: {.fragment}
People with 4 year degrees tend to make more.
:::

# Question 2

##
```{r}
earnings.histogram
```

**Q**. Roughly how much more?

::: {.fragment}
 Typically \$50k more, comparing the two peaks in the histogram.
:::

# Question 3

##

```{r}
earnings.scatter + scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) 
```

How would we calculate the difference in **mean income** between people in our sample with and without 4-year degrees?

### Outline
::: {.fragment}
1. We calculate the mean income among the *subsample* [with]{.twocolor-green} 4-year degrees.
2. We do the same for the subsample [without]{.twocolor-red} degrees. 
3. We subtract.
:::

## Step 1

```{r}
earnings.scatter + scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) 
```

We calculate the mean income among the *subsample* [with]{.twocolor-green} 4-year degrees.

:::: {.r-stack}

::: {.fragment .current-visible data-frame-index="1"}

We can write it out in mathematical notation.

$$
\hat\mu(\text{with}) = \frac{\sum_{i : X_i=\text{with}} Y_i }{ \sum_{i : X_i=\text{with}} 1 }
$$
:::

::: {.fragment .current-visible data-frame-index="2"}
```{r}
Y = earnings$income
X = earnings$degree
```

Writing it in *R* actually gives us an answer.

```{r}
#| echo: true
mu.hat.with = mean(Y[X == '4-year degree'])
```

What we get is roughly \$`{r} as.integer(round(mu.hat.with, digits=-3))`.
:::

:::: 

## Step 2 
```{r}
earnings.scatter + scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) 
```

We do the same for the subsample [without]{.twocolor-red} degrees. 

:::: {.r-stack}

::: {.fragment .current-visible data-frame-index="1"}

We can write it out in mathematical notation.

$$
\hat\mu(\text{with}) = \frac{\sum_{i : X_i=\without} Y_i }{ \sum_{i : X_i=\without} 1 }
$$
:::

::: {.fragment .current-visible data-frame-index="2"}

Writing it in *R* actually gives us an answer.

```{r}
#| echo: true
mu.hat.without = mean(Y[X != '4-year degree'])
```

What we get is roughly \$`{r} as.integer(round(mu.hat.without, digits=-3))`.
:::

:::: 


## Step 3
```{r}
earnings.scatter + scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) 
```

We subtract.

:::: {.r-stack}

::: {.fragment .current-visible data-frame-index="1"}
Here's the math.
$$
\hat \Delta = \hat\mu(\text{with}) - \hat\mu(\text{without})  \vphantom{ \frac{\sum_{i : X_i=\text{without}} Y_i} { \sum_{i : X_i=\text{without}} 1 } }
$$
:::

::: {.fragment .current-visible data-frame-index="2"}
Here's the code.
```{r}
#| echo: true
difference = mu.hat.with - mu.hat.without
```

What we get is roughly \$`{r} as.integer(round(difference, digits=-3))`.
:::

::::

# Question 4

## 
```{r}
earnings.histogram
```
What *Summary of the Population* does this estimate?

::: fragment

- The difference in mean income among *subpopulations* with and without 4-year degrees. \
- That's a difference in conditional expectations.
$$
\Delta = \mu(\with)-\mu(\text{without}) \quad \text{ where } \quad \mu(x)=\E[Y_i \mid X_i=x]
$$
:::

# Question 5

##

$$\text{What is our estimator's bias,} \ \E \hat\Delta - \Delta ? $$

::: {.fragment }
#### It's zero. Why?
:::

$$
$$

:::: {.r-stack}

::: {.fragment .current-visible data-frame-index="2"}
#### Part of the story: subsample means are **unbiased** estimators of subpopulation means.

  - In particular, $\hat\mu(x)$ is an unbiased estiate of $\mu(x)$.
  $$
  \E\hat\mu(x)=\mu(x)
  $$
  - We proved this in Homework ? using the law of iterated expectations.
  
$$
TODO: Proof
$$

:::

::: {.fragment .current-visible data-frame-index="3"}
#### The rest: Expectation is **linear**. 

  - The expectation of a difference is the difference in expectations.
  - Subtracting unbiased estimates of our subpopulation means gives us \
    an unbiased estimate of the **difference in those subpopulation means**.
$$ 
\begin{aligned}
\E \hat\Delta 
&= \E \left\{ \hat\mu(\text{with}) - \hat\mu(\text{without}) \right\} && \text{definition} \\
&= \E \hat\mu(\text{with}) - \E \hat \mu(\text{without}) && \text{linearity} \\
&= \mu(\text{with}) - \mu(\text{without})  && \text{unbiasedness} \\
&= \Delta && \text{definition} 
\end{aligned}
$$
:::

::::

# Question 6

##

How can we approximate the *sampling distribution* of this estimator?

$$
$$

TODO: Picture of sampling dist vs approximation

### Outline 

::: fragment
1. We find normal approximations to the sampling distributions of each of our subsample means.
2. We characterize their difference as another normal distribution. 
3. We calculate its mean.
4. We calculate its variance.
:::

## Step 1

We observe that our subsample means are *approximately independent and normal* 
$$
\hat\mu(x) \approx Z_x \quad \text{ for } \quad Z_x \sim N\left( \E \hat\mu(x), \ \Var \hat\mu(x) \right)
$$
and
$$
Z_{x} \quad \text{ is independent of } \quad Z_{x'}  \quad  \text{ when } \quad x \neq x'. 
$$

They're approximately independent because our subsamples are *disjoint*, i.e., nobody is in both.


## Step 2

We'll approximate the difference in our subsample means by 
the corresponding difference of *independent normals*.
$$
\hat\mu(\text{with}) - \hat\mu(\text{without}) \approx Z_{\text{with}} - Z_{\text{without}}
$$
Then use the rule for subtracting independent normals: subtract the means; add the variances.
$$
\begin{aligned}
Z_{\text{with}} - Z_{\text{without}} &\sim  N\left( \mu_{\Delta}, \sigma^2_\Delta \right) \\
\text{ where } \quad    \mu_{\Delta} &= \E \hat\mu(\text{with})  - \E \hat\mu(\text{without}) \\
\text{ and  } \ \ \quad \sigma^2_\Delta &= \Var \hat\mu(\text{with}) + \Var \hat\mu(\text{without}) 
\end{aligned}
$$ 

## Step 3
We subtract the means, i.e., we calculate this difference. 
$$
\mu_{\Delta} = \E \hat\mu(\text{with})  - \E \hat\mu(\text{without}). 
$$
We know, from our last question, that this is the population summary we want.
$$\mu_{\Delta} = \hat\mu(\text{with})  - \hat\mu(\text{without}) = \Delta$$

## Step 4
We add the variances, i.e., we calculate this sum.
$$ 
\sigma^2_\Delta = \Var \hat\mu(\text{with}) + \Var \hat\mu(\text{without})
$$
The **variance** of each subsample mean $\hat\mu(x)$ is determined by 

1. The conditional variance of the terms in it
2. The size of the subsample.

$$ 
\sigma^2_\Delta = \frac{\Var(Y_i \mid X_i = \text{with})}{n_\text{with}} + \frac{\Var(Y_i \mid X_i = \text{without})}{n_{\text{without}}} 
$$

TODO: Do we define $n_\text{with}$ and $n_\text{with}$ properly as expectations of inverses of proportions?

## Summary.

We can approximate the sampling distribution of our estimator
by that of a *normal random variable* with the mean we want.
$$ 
\begin{aligned}
\hat\mu(\text{with}) - \hat\mu(\text{without}) &\approx Z_{\Delta} \quad \text{ for } \quad Z_{\Delta} \sim N(\Delta, \sigma^2_\Delta) \\
\text{ where } \quad \sigma^2_\Delta &= \frac{\Var(Y_i \mid X_i = \text{with})}{n_\text{with}} + \frac{\Var(Y_i \mid X_i = \text{without})}{n_{\text{without}}}
\end{aligned}
$$


# Question 7

##
How can we use this to calculate a 95\% confidence interval for this summary $\Delta$?

$$ $$

### Outline
::: fragment

1. We transform our estimate into a *pivot*: a random variable with an known sampling distribution.
2. We choose an interval that contains the pivot with 95\% probability.
3. We invert our transformation to get an interval for our estimate.

TODO: Fix up to deal with approximate pivots.
:::


## Step 1
t-stat calculation + picture of std normal reference distn

## Step 2
formula + picture of interval in reference distribution

## Step 3
formula + picture of transformed reference distributon

## Summary
something


## Most of the Remainder
We'll answer the questions

## Hopefully this fits at the end
- We'll explain why we might need an unsaturated model and how that results in bias.
- We'll imagine a reality in which we know whether there are 1 or 2 year masters degrees
  and think of the bias associated with grouping them for Question 3.
