---
title: "Working with Non-Binary Covariates"
subtitle: "QTM 220 Lecture 5"
format: 
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    embed-resources: true
    slide-number: true
    show-notes: separate-page
    pdf-separate-fragments: true
    fig-format: svg
    fig-align: left
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---

```{r load-libraries}
library(purrr)
library(tibble)
library(dplyr)
library(magrittr)
library(stringr)
library(ggplot2)
library(htmltools)
library(kableExtra)
library(scales)
library(cpsR)
library(mapdata)

source('../../shared-code/plotting.R')
theme_set(simplified.theme)
set.seed(1)

compact_number = label_number(accuracy=.1, scale_cut=cut_short_scale())
```

```{r get-data}
#| cache: true
#| 
# for cps variable interpretation see https://www2.census.gov/programs-surveys/cps/datasets/2022/march/asec2022_ddl_pub_full.pdf
census.api.key = '8084f81e4f71e4d32638dac15f188b424f72d063'

california.state.code = 6
year = 2022
cps.data = get_asec(year=year, vars=c('pearnval', 'a_age', 'a_hga', 'gestfips', 'gtco', 'gtcbsasz'), key=census.api.key)
```

```{r transform-data}
#| cache: true
#| 
cps.sampling.ratio = 1/25000
all.earnings = data.frame(state.code = cps.data$gestfips,
                          age=cps.data$a_age,
                          income=cps.data$pearnval, 
                          education.code=cps.data$a_hga,
                          education=case_match(cps.data$a_hga,
                                           0  ~ 0,   # child
                                           31 ~ 0,   # < grade 1
                                           32 ~ 4,   #  grade 1-4
                                           33 ~ 6,   #  grade 5-6
                                           34 ~ 8,   #  grade 7-8 
                                           35 ~ 9,   #  grade 9
                                           36 ~ 10,  #  grade 10
                                           37 ~ 11,  #  grade 11
                                           38 ~ 11,  #  grade 12 no diploma
                                           39 ~ 12,  #  high school grad
                                           40 ~ 13,  #  some college
                                           41 ~ 14,  #  associate degree (vocational)
                                           42 ~ 14,  #  associate degree (academic)
                                           43 ~ 16,  #  bachelors degree
                                           44 ~ 18,  #  masters degree
                                           45 ~ 20,  #  professional school degree
                                           46 ~ 20), #  doctorate
                        metro.size=case_match(cps.data$gtcbsasz, # cps reports a metro-size category, we map that to an approximate metro size
                                           0~0,       # < 100k
                                           2~1e5,     # 100-250k
                                           3~2.5e5,   # 250-500k
                                           4~5e5,     # 500k-1 mil
                                           5~1e6,     # 1-2.5 mil
                                           6~2.5e6,   # 2.5-5 mil     
                                           7~5e6))    # >= 5 mil  

in.interval = function(X, interval) { interval[1] <= X & X <= interval[2] }
earnings = all.earnings[all.earnings$state.code == california.state.code &
                        all.earnings$age %>% in.interval(c(25,35)) &
                        all.earnings$education >= 8, ]

earnings$degree = factor(earnings$education >= 16, 
                     levels=c(FALSE, TRUE), 
                     labels=c('no 4-year degree','4-year degree')) 
```

```{r make-maps}
#| cache: true
#| 

# To give our survey recipients a bit more reality, we'll give them fake locations on a map
# And we'll make up a population of California residents that they're sampled from

# maps are based on code from https://eriqande.github.io/rep-res-eeb-2017/map-making-in-R.html
# additional county info from https://en.wikipedia.org/wiki/List_of_counties_in_California
# converted to csv using https://wikitable2csv.ggor.de

to.number = function(s) { as.numeric(str_trim(gsub(",", "", s))) }

counties = map_data("county")
ca_county = counties %>%
  filter(region == "california")
ca_county_centers = ca_county %>%
  group_by(subregion) %>%
  summarize(county.lat = mean(lat), county.long=mean(long))
ca_county_info = read.csv('california_county_info.csv') %>%
  mutate(subregion = tolower(sub(' County', '', County))) %>%
  mutate(pop = to.number(Population..2022.)) %>%
  mutate(area = to.number(sub('sq.*', '', Area))) 
ca_county_more = left_join(ca_county_centers, ca_county_info, by="subregion")
ca_all = left_join(ca_county, ca_county_more, by="subregion")


ca.plot = ggplot(data=ca_all, mapping=aes(x=long, y=lat, group=group)) + 
  coord_quickmap() +
  geom_polygon(fill='gray', color = "white") +
      scale_fill_gradient(trans="log10") + 
      theme_void()

ca.pop.plot = ggplot(data=ca_all, mapping=aes(x=long, y=lat, group=group)) + 
  coord_quickmap() +
  geom_polygon(aes(fill = pop/area), color = "white") +
      scale_fill_gradient(trans="log10") + 
      theme_void()

## we'll make up a population as follows
# 1. choose the number of people to be the number of people in our sample scaled up to account for the cps sampling ratio
# 2. for each, sample a county with probability proportional to its finite.population
# 3. for each, sample a latitude and longitude by taking a randomly weighted combination of the    
#    county's vertices. As long as the county is convex, these points will be inside the county.
# 4. we'll give them a degree too, chosen independently of all else from the empirical distribution on the sample
n = nrow(earnings) / cps.sampling.ratio 
finite.population.subregion = sample(ca_county_more$subregion, size=n, replace=TRUE, prob=ca_county_more$pop)
finite.population = unique(finite.population.subregion) %>% 
  map_dfr(function(subregion) { 
    vertices = ca_county[ca_county$subregion==subregion, c('lat','long')]
    U = sum(finite.population.subregion==subregion)
    V = nrow(vertices)
    weights = abs(rt(U*V, 1))
    dim(weights) = c(U,V)
    weights = (1/rowSums(weights)) * weights
    data.frame(subregion=subregion,
               lat  = weights %*% vertices$lat,
               long = weights %*% vertices$long)
  })
finite.population$group=1
finite.population$degree = sample(earnings$degree, nrow(finite.population), replace=TRUE)

# to go easy on ggplot, we won't really display everbody 
display.population = sample_n(finite.population, min(100000, nrow(finite.population)))
```

```{r make-sample}
# and we'll force the degree-distribution in the sample to match our survey responses
our.sample = sample_n(finite.population, nrow(earnings))
our.sample$degree=earnings$degree

pop.plot = ca.plot + geom_point(aes(x=long, y=lat), data=display.population, color='red', alpha=.01) 
sample.plot = ca.plot + geom_point(aes(x=long, y=lat), data=our.sample, color='red',    alpha=1)
pop.plot.inf = ca.pop.plot + geom_point(aes(x=long, y=lat), data=display.population, color='red', alpha=.01)  
sample.plot.inf = ca.pop.plot + geom_point(aes(x=long, y=lat), data=our.sample, color='red',    alpha=1)
```

## Summary

::: {.hidden}
$$
% color defs aren't making it into MathJax
%\definecolor{twocolor-red}{RGB}{248, 118, 109}   % red   used by ggplot in plots w/ 2 colors 
%\definecolor{twocolor-green}{RGB}{0, 191, 196}   % green used by ggplot in plots w/ 2 colors
% so create commands for using those colors instead
\newcommand{\inred}[1]{\textcolor[RGB]{248,118,109}{#1}}
\newcommand{\ingreen}[1]{\textcolor[RGB]{0,191,196}{#1}}
\newcommand{\inblue}[1]{\textcolor{blue}{#1}}

\newcommand{\ind}[1]{1_{\{#1\}}}
\newcommand{\with}{\ingreen{\text{degree}}}
\newcommand{\without}{\inred{\text{no degree}}}

\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\Var}{Var}
$$ 

```{r define-colors}
twocolor.red   = 'rgb(248,118,109)'
twocolor.green = 'rgb(0,191,196)'
```
:::

- Today, we'll work with data from the 2022 Current Population Survey.
- We'll look at the relationship between income and education.
- We've already compared people  [with]{.twocolor-green} and [without]{.twocolor-red} 4-year degrees.
- That's a way of summarizing what we see in the data based on a *binary covariate*.
$$ 
X_i = \begin{cases} 
  \with & \text{ if } \ \text{person}_i \ \text{has a 4-year degree} \\
  \without & \text{ otherwise }
  \end{cases}
$$
- Today, we'll talk about summaries based on a *many-valued covariate*.
- We'll categorize people by the years of education they've had.


<<<<<<< HEAD
<<<<<<< HEAD
- A quick review of the binary case including
  - *descriptive statistics*: summaries of the data.
  - *inferential statistics*: summaries of a related population.
- Descriptive statistics in many-valued case.
- Inferential statistics in the many-valued case.
    - A more calculation-focused review of the binary case.
    - Generalizations for the many-valued case 
=======
1. A quick review of the binary case including
  - *descriptive statistics*: summaries of the data.
  - *inferential statistics*: summaries of a related population.
2. Descriptive statistics in many-valued case.
3. Inferential statistics in the many-valued case.
    a. A more calculation-focused review of the binary case.
    b. Generalizations for the many-valued case 
>>>>>>> 02ffdda (minor changes to lec 5 + added compiled html version (lecture5.html))
      - We'll do this as a set of in-class exercises.
      - I'll introduce techniques that'll make this straightforward.
      

# A Quick Review

## 
=======
## Our sample
>>>>>>> 11a3a80 (simplified lec 5; still incomplete)

:::: columns

::: {.column}
We'll look at the survey respondents who

  - were California residents 
  - between 25 and 35
  - who lived alone

This sample includes `r nrow(earnings)` people. 
You can think of them as living

```{r sample-map}
ca.plot + geom_point(aes(x=long, y=lat), color='red', data=our.sample, alpha=1)  + guides(color=FALSE)
```
:::

::: {.column}
```{r sample-table}
style.table = function(tab) {
  tab$income = sprintf('$%dk', round(tab$income / 1000))
  rownames(tab) = NULL
  kable(tab) %>%
    kable_styling() %>%
    row_spec(which(tab$degree=='4-year degree'), background=twocolor.green) %>%
    row_spec(which(tab$degree!='4-year degree'), background=twocolor.red) 
}

style.table( earnings[1:12, c('income', 'degree', 'education')] )
```
:::
::::

## A binary look 

## In a Scatter Plot

```{r binary-scatter-in-columns}
#| out-size: 500px
dot.size = 5
earnings.scatter = ggplot(earnings) + 
  geom_point(aes(x=degree, y=income, color=degree), size=dot.size, alpha=.2) + 
  scale_y_continuous(labels=label_dollar()) + guides(color=FALSE) + class.theme + xlab('') + class.theme
earnings.scatter
```


## Making Tables Usable

:::: columns

::: column

```{r short-binary-table}
random.rows = sample(1:nrow(earnings), 12)
style.table( earnings[random.rows, c('income', 'degree')] )
```

:::

::: column

- The major downside of tables is that they're usually big. 
  - It's hard to look at all `r nrow(earnings)` people at once.
  - But it's easy to look at a small part of it.
- I'm showing `r length(random.rows)` randomly-chosen people on the left. 
  - This is less overwhelming so we can see something.
  - But it is only a fraction of the data.

:::

::::

## Answering Questions by Looking at Tables


:::: columns

::: column

```{r short-binary-table-2}
style.table( earnings[random.rows, c('income', 'degree')] )
```

:::

::: column

- What does this table tell us about who makes more?  
    - Is it people [with]{.twocolor-green} or [without]{.twocolor-red} four year degrees?
    - Roughly how much more do they tend to make?

:::

::::

::: notes
- Of the randomly-chosen people we're looking at, 
  most [with 4-year degrees]{.twocolor-green} make more than most [without]{.twocolor-red}
- But the incomes we're seeing are all over the place, so
  - it's not clear that this holds up in the sample 
  - it's hard to quantify the difference
:::


## Making Scatter Plots Usable

:::: {.r-stack }

::: {.fragment .fade-out data-fragment-index="0"}
```{r binary-scatter-no-scaling}
#| out-height: 300px
dot.size = 5
earnings.scatter = ggplot(earnings) + 
  geom_point(aes(x=degree, y=income, color=degree), size=dot.size, alpha=.2) + 
  scale_y_continuous(labels=label_dollar()) + 
  guides(color=FALSE) + xlab('') + class.theme
earnings.scatter
```

- The major downside of scatter plots is that they can be dense.
  - I've plotted one dot for each person. 
  - They overlap to the point we can't see individual dots.
- Part of problem is wasted space. Everyone is clustered at the bottom.
- We can zoom in there, cutting off a few high-earners, to get a better look.

:::

::: {.fragment .current-visible data-fragment-index="0"}
```{r binary-scatter-scaled}
#| out-height: 300px
earnings.scatter + scale_y_continuous(labels=label_dollar(), limits=c(0,3e5))  
```

- Now I've cut off people who make more than \$300k
  - This is better, but the dots are still overlapping too much.
- That's still a problem of wasted space, but now the problem is mostly horizontal.
  - There are only two values of $X_i$, so we've got two vertical lines and a lot of blank space.
  - We can spread them out a little horizontally to get a better look. 
:::

::: {.fragment .current-visible data-fragment-index="1"}
```{r binary-jittered}
#| out-height: 300px
earnings.scatter.jittered =  ggplot(earnings) + 
  geom_point(aes(x=degree, y=income, color=degree), 
             position=position_jitter(w=.2), size=dot.size, alpha=.2) + 
  guides(color=FALSE) + xlab('') + class.theme
earnings.scatter.jittered + scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) 
```

- To use the horizontal space, we use a technique called *jittering*. 
  - We just give the points some arbitrary horizontal spread.
  - There's no meaning to it. It's just to use the space better.
- Now it's easier to see where there are more dots and where there are less.
- That can give you a rough sense of how much most people make.
:::

::::

## Answering Questions using Scatter Plots

```{r binary-scatter-jitter-2}
#| out-height: 300px
earnings.scatter.jittered + scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) 
```


- What does this plot tell us about who makes more?  
  - Is it people [with]{.twocolor-green} or [without]{.twocolor-red} four year degrees?
  - And roughly how much more do they make?

::: notes
- People [with 4-year degrees]{.twocolor-green} in the plot tend to make more than most [without]{.twocolor-red}. 
  - Now we're looking at all but a couple people.
  - So that's true about our sample as a whole.
- And it looks like they make roughly \$50k more.
  - Shifting up the dense cluster of [red dots]{.twocolor-red} by \$50k will
    roughly align it with the dense cluster of [green ones]{.twocolor-green}.
:::


## Summary Statistics

```{r binary-mean-sd}
summaries = earnings %>% 
          group_by(degree) %>% 
          summarize(mean=mean(income), sd=sd(income))


difference.in.means = as.numeric(summaries[summaries$degree=='4-year degree', 'mean'] - 
                                 summaries[summaries$degree=='no 4-year degree', 'mean'])
```

- Looking at the raw data helps us get a sense of what's going on.
- It's a good way to get rough answers to vague questions.
- But when we want to describe a sample or subsample precisely, we use *summary statistics*.
    - e.g., the mean and standard deviation
    - e.g., a histogram
- One way to pose the 'how much more' question precisely to ask for the difference in subsample means
- It's \$`r compact_number(difference.in.means)`. That's close to our visual estimate of \$50k.

## Mean and Standard Deviation

```{r binary-mean-sd-plot}
#| out-height: 250px
#| fig-asp: .3 
earnings.scatter.jittered + 
  geom_pointrange(aes(x=degree, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries) +
  ylim(-1e5, 4e5)
```

:::: {.r-stack}

::: {.fragment .fade-out data-fragment-index="0"}
### Means
- The subsample means are location summaries. 
  - They tell us where our subsamples are centered.
$$
\hat\mu(x) = \frac{1}{N_x} \sum_{i:X_i=x} Y_i \qfor N_x = \sum_{i:X_i=x} 1
$$
  - The black dots in our plot are these means.
  - We've got one for $x=\with$ and another for $x=\without$.

:::

::: {.fragment .current-visible data-fragment-index="0"}
### Standard Deviations
- The subsample standard deviations are spread summaries. 
  - They tell us how spread out our subsamples are.
$$
\hat\sigma(x) = \sqrt{ \frac{1}{N_x} \sum_{i: X_i=x}  \qty{ Y_i - \hat\mu(x) }^2 }  \qfor N_x = \sum_{i : X_i=x} 1
$$
  - It's the square root of the *mean squared distance* between the incomes in our subsample and their mean
  - The black bars in our plot show one standard deviation above and below the mean
  - We expect most of the incomes in our subsample to be within these bars.
:::

::::


## Issues

:::: columns

::: column
```{r binary-mean-sd-col}
earnings.scatter.jittered + 
  geom_pointrange(aes(x=degree, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries) +
  ylim(-1e5, 4e5)
```
:::

::: column
```{r binary-mean-sd-col-tall}
earnings.scatter.jittered + 
  geom_pointrange(aes(x=degree, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries) 
```
:::

::::

- Means and standard deviations are strongly affected by extreme observations
  - That's why the mean for [people with degrees]{.twocolor-green} is a bit
    bigger than the typical value we estimated by eye.
  - It's also why their standard deviation is so big.
- There are many alternatives, but we'll mostly focus on one this semester: histograms.

## Histograms

```{r}
#| out-height: 300px
#| fig-asp: .5 
earnings.histogram = ggplot(earnings) + 
  geom_histogram(aes(x=income, y=after_stat(count/sum(count))), alpha=.5, binwidth=25000) + 
  coord_flip() + facet_wrap(~degree, strip.position='bottom') + aes(fill=degree) +
  guides(fill=FALSE) + scale_y_continuous(labels=label_percent()) + ylab('') + class.theme
```

:::: {.r-stack}

::: {.fragment .fade-out data-fragment-index="0"}
```{r}
#| out-height: 300px
#| fig-asp: .5 
earnings.histogram + scale_x_continuous(labels=label_dollar(), limits=c(0,3e5)) 
```

- In these histograms, we show the fraction of people
  that make roughly a given amount of money.
- To be precise, we show the *subsample means*
  of **indicators** for different income ranges.
- The ranges we're using are $[0 \ldots 25k)$, $[25k \ldots 50k)$, $[50k \ldots 75k)$, etc.

:::

::: {.fragment .current-visible data-fragment-index="0"}

```{r}
#| out-height: 300px
#| fig-asp: .5 
earnings.histogram + scale_x_continuous(labels=label_dollar())
```

- What we get in one income range doesn't affect the others.
- That means our summaries of incomes in most ranges aren't
  affected by how much our high-income folks make.
- We just weren't showing the high-income folks in our last plot.

:::

::::

$$
\begin{aligned}
\text{frac}_{[a \ldots b)} &= \frac{1}{N_x}\sum_{i:X_i=x} 1_{[a \ldots b)}(Y_i) \qfor  1_{[a,b)}(y) = \begin{cases} 1 & \text{ if } a \leq y < b \\
                             0 & \text{ otherwise } \end{cases}
\end{aligned}
$$


## Limitations of what we've done
  - We've done a fairly complete story of the relationship between income and education
    - We've talked about descriptive statistics, i.e., sample summaries
    - We've talked about inferential statistics, i.e., population summaries
  - But we've based all of this on very coarse idea of education
    - We've just asked whether people have a 4-year degree or not
  - This means there are really only a couple questions to answer
    - How much people with and without them usually make?
    - How different that is between the two groups.
    - How much income varies, overall or within these two groups.
  - We can say much more if we think of education as being many-valued.
    - e.g. 0 years, 1 year, 2 years, ... 20 years of formal education.
    - There are many more questions we can ask when we do that.
  - That's what we'll talk about now.

# Beyond Binary Comparisons: Descriptive Analysis

## Making Education Multi-Valued 

- Using two categories is throwing away a lot of information about education. 
- We can categorize people by the **years of education** they've had. 

```{r}
#| out-height: 250px
years = unique(earnings$education)
ggplot(earnings) + geom_point(aes(x=education, y=income, color=degree), 
                              size=dot.size, position=position_jitter(w=.2), alpha=.2) +
                   scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) + 
                   scale_x_continuous(breaks=years, labels=sprintf('%d years', years)) + 
                   guides(color=FALSE)
```

:::: {.r-stack}

::: {.fragment .fade-out data-fragment-index="0"}
### We haven't lost anything.

- You have a four year degree if, and only if, you have $12+4=6$ years of education.
- Thinking visually, we can get the plot we had before by shuffling the red and green dots.
- This means we can still see how much people with and without 4-year degrees tend to make.
- But not shuffling lets us see some new stuff, too. 

### What else is there to see?
:::

::: {.fragment .current-visible data-fragment-index="0"}

### What do you see, in purely visual terms?
$$
$$

### What does it mean?
$$
$$

:::

::::

::: {.notes}

### One thing I see
The centers of our subsamples are going up as we move to the right.

### Meaning
Even among people with (or without) a 4-year degree, 
people with more education tend to make more money.

:::




## Including Years of Education
- This does, of course, require that our survey asks about years of education. 
- And it does, more or less, so we can make a table like this.

```{r}
earnings.tab = earnings
earnings.tab$education = sprintf('%d years', earnings$education)
style.table( earnings.tab[random.rows,  c('income', 'degree', 'education')] )
```

- The survey doesn't ask exactly this *how many years* question.
- It takes a little pre-processing to get this.

## Pre-processing

```{r}
#| echo: true
education=case_match(cps.data$a_hga,
                          0  ~ 0,   # child
                          31 ~ 0,   # < grade 1
                          32 ~ 4,   #  grade 1-4
                          33 ~ 6,   #  grade 5-6
                          34 ~ 8,   #  grade 7-8 
                          35 ~ 9,   #  grade 9
                          36 ~ 10,  #  grade 10
                          37 ~ 11,  #  grade 11
                          38 ~ 11,  #  grade 12 no diploma
                          39 ~ 12,  #  high school grad
                          40 ~ 13,  #  some college
                          41 ~ 14,  #  associate degree (vocational)
                          42 ~ 14,  #  associate degree (academic)
                          43 ~ 16,  #  bachelors degree
                          44 ~ 18,  #  masters degree
                          45 ~ 20,  #  professional school degree
                          46 ~ 20)  #  doctorate
```         

- We've had to make a few things up in translation.
  - There's no differentiation between a 1 or 2-year masters; we assume 2.
  - We're assuming professional and doctoral degrees take 4 years; that's approximate.
  - Nobody has 17 or 19 years in our plots; that's why.
- That's part of data analysis. 
- Ultimately, you'll want to be careful about quirks introduced by translation.
  - But it's important to keep your job manageable. 
  - It can help to ignore the quirks when you're getting started.


## Subsamples 

```{r}
#| out-height: 250px
years = unique(earnings$education)
ggplot(earnings) + geom_point(aes(x=education, y=income, color=degree), 
                              size=dot.size, position=position_jitter(w=.2), alpha=.2) +
                   scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) + 
                   scale_x_continuous(breaks=years, labels=sprintf('%d years', years)) + 
                   guides(color=FALSE)

```


:::: {.r-stack}

::: {.fragment .fade-out data-fragment-index="0"}  
- When we thought of education as binary, we only had two subsamples to talk about.
  - People with 4-year degrees --- the [green dots]{.twocolor-green}.
  - People without them --- the [red dots]{.twocolor-red}.
- Now that we're taking a more granular view, we can talk about *many*.
:::

::: {.fragment .current-visible data-fragment-index="0"}
- People with $x$ years of education for $x=10,11,\ldots$.
     - This is the *finest* way to break down our sample in terms of $X$.
- People with and without 4-year degrees, as before.
     - This is a coarser breakdown. They're *aggregates*.
     - To get these, we combine our finer subsamples.
- People whose highest degree is graduate, 4-year,  2-year, high-school, or none.
     - That's in between. It's a finer aggregate.

:::

::::


## New Questions

Taking this more granular view of education allows us to pose new questions.

1. What's the typical difference in income between people with *exactly* 16 years of education
   and *exactly* 12?
    - This is a different question than the one we addressed in review, but it has a similar structure.
    - We want a summary of income differences in two groups. Those can be subsamples or subpopulations.
    - We can calculate and relate these summaries exactly like we did in review.
2. What's the typical difference in income between people with 4-year degrees and people without them?
    - That's exactly the question we talked about in review.
    - We'll talk about how to calculate it a different way that'll generalize better to other questions.
    - We can check our work. Obviously our answers should be exactly the same as we got earlier in the semester.
3. How much change would we expect in a similar population in which people tend to get an additional year of education?
    - This is a question unlike the ones we talked about earlier.
    - We'll need to be a bit careful about what we mean.
    - But once we've gotten there, we can base our calculations on ones we've done already.

More for HW.

## Q1

```{r}
#| out-height: 250px
earnings$highest.degree = case_match(earnings$education,
                            0:11 ~ 'none',
                            12:13 ~ 'high school',
                            14  ~   'associates',
                            16  ~   'bachelors',
                            18:20 ~ 'graduate')

fine = earnings %>% group_by(education) %>% summarize(mean.income = mean(income))
coarse = earnings %>% group_by(degree) %>% summarize(mean.income = mean(income), lb=min(education), ub=max(education)) 
intermediate = earnings %>% group_by(highest.degree) %>% summarize(mean.income = mean(income), lb=min(education), ub=max(education))

base.plot = ggplot(earnings) + geom_point(aes(x=education, y=income, color=degree), 
                              size=dot.size, position=position_jitter(w=.2), alpha=.2) +
                   scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) + 
                   scale_x_continuous(breaks=years, labels=sprintf('%d years', years)) + 
                   guides(color=FALSE)

base.plot + geom_segment(aes(x=education-.2, xend=education+.2, y=mean.income, yend=mean.income), data=fine)
```



## Q2
```{r}
base.plot + geom_segment(aes(x=lb-.2, xend=ub+.2, y=mean.income, yend=mean.income), data=coarse) 
```

  - Means of Aggregates are Aggregates of Means
  - Weighted Aggregates
  - [do somethoing with fine picture combo]

## Q3

- some kind of picture showing jumps


## Population Summaries
 
 - Often, what we want isn't a summary of our sample.  We want to summarize our population instead.
    - e.g., we don't want the mean income of the [people with 4-year degrees]{.twocolor-green} on the left map---the ones we've actually surveyed. 
    - We want to know the now the mean income of [people with 4-year degrees]{.twocolor-green} on the right map---who we mostly haven't.
$$
$$

```{r map-subsamples-again}
#| layout-ncol: 2
#| layout-valign: bottom
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=our.sample, alpha=1)  + guides(color=FALSE)
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=display.population, alpha=.01) + guides(color=FALSE)
```

 - We can't do this perfectly because we're trying to summarize data we don't have.
 - But we can **estimate** these population summaries using our sample.


## Point and Interval Estimation 

- We've covered two types of estimate in this class.
  - A **point estimate** is a single number that we think is close to the population summary.
  - An **interval estimate** is a range of numbers that we think contains the population summary.
- We'll focus on point estimates today and interval estimates next class.


## Subsample and Subpopulation Means 

- We can think of *subsample means* as point estimates of the corresponding *subpopulation means*.
- When we do this, we're thinking of our sample as a random draw from our population.
- This means that summary statistics, like the subsample mean $\hat\mu(x)$, are random variables. 
  - Why? It means the observations in our sample are random variables.
  - And anything we calculate in terms of random variables is itself a random variable.

:::: columns

::: column
### Subsample Mean
$$
\hat\mu(x) = \frac{1}{N_x} \sum_{i:X_i=x} Y_i \qfor N_x = \sum_{i:X_i=x} 1
$$
:::

::: column
### Subpopulation Mean
$$
\mu(x) = \E[ Y_i \mid X_i = x]
$$

:::

::::

## Sampling Distributions

- A summary statistic, like all random variables, has a distribution.
- We call this its **sampling distribution**.
- That's the distribution the summary would have if we were to redraw the sample, over and over, from the population.
  - i.e., if we were to send out the Current Population Survey many times
  - each time randomly selecting the people who receive it survey the same way.
- In real life, we usually can't do that. 
  - That means we see only one draw from the sampling distribution.
  - We can, however, try to *estimate the sampling distribution* using only our sample.
- When we're simulating data, we can redraw our sample as many times as we like.
  - That means we can see what the sampling distribution looks like.
  - This allows us to check, to some degree, whether our estimates work.
  

## Sampling Distributions: A Picture
### TODO 
Sequence of histograms of some summary statistic
  - Maybe show K samples as scatter plots w/ mean plotted
  - And a histogram the K subsample means
  - e.g. for K=9, 9^2 [we can do 3x3 and 9x9 displays of the sample]


### Expectations

- Because a summary statistics is a random variable, it has a mean, standard deviation, etc.
- These are the mean and standard deviation of a draw from the summary's sampling distribution.
-

### TODO: illustrate this as an overlay on one of the sampling distribution pictures


## Subsample Means as Point Estimators
  - They are unbiased, i.e., their expectation is what we're trying to estimate.
  $$
  \E \hat\mu(x) = \mu(x)
  $$
  - They are convergent, i.e., they get closer to that expectation as the sample gets bigger.
    - In particular, their standard deviations are proportional to the inverse square root of the sample size.
  $$
  \begin{aligned}
  \sqrt{\Var \hat\mu(x)} = \E\qty{ \frac{ \sigma^2(x) }{\sqrt{N_x}} } \qfor \sigma^2(x) &= \Var(Y_i \mid X_i=x) %\\ &= \E \qty[ \qty{ Y_i - \mu(x) }^2 \mid X_i=x] 
  \end{aligned}
 $$ 
 - They are approximately normal, i.e., their sampling distribution is close to normal for large samples.
   - This is a consequence of the central limit theorem.
   - It's a good thing because it lets us base **interval estimates** on these normal approximations.

---
# Maybe use this stuff in the next class
# 
### Interval Estimates
#
#  - Estimate of where the pop summary is that accounts for uncertainty
#  - Based on the sampling distribution of our estimator
#  - Intention: in 95/100 replications, the interval estimate contains the pop summary
#  - Coverage: the fraction of replications where the interval estimate actually does contain
#  - We'll focus on point estimation today and interval estimation next class
---

## Comparing Means
- If we want to compare means for two subpopulations, we can do the same thing.
- The difference between the corresponding subsample means is an unbiased estimate.
$$ \E\qty{ \hat \mu(\with) - \hat \mu(\without) } = \mu(\with) - \mu(\without)  $$
- It's convergent and approximately normal, too.
- We'll review all this later today.


# Preparation for Inferential Analysis

## Review of Calculation Techniques

- All of our questions are comparisons, so the closest we've come is the comparison we did in the binary case.
- We'll review what happens when we estimate the difference in population means by the difference in sample means.
- In particular, we'll calculate an approximation that estimator's sampling distribution.
  - We'll start by showing that it's unbiased.
  - Then we'll calculate its variance.
  - We'll wave our hands about normality.

## Unbiasedness of subsample means
- We'll start by showing that a subsample's mean is an unbiased estimate of the corresponding subpopulation mean.
  - We'll think about the subsample and subpopulation with $X_i=x$ for some value $x$ 
  - e.g. $X_i=\with$ for people with 4-year degrees.
- That means its expectation *is* the subpopulation mean. 
- We can think of our subsample mean as an estimate of the mean of the corresponding *finite subpopulation*.
- But we'll focus on interpreting it as an estimate of the mean of the *infinite subpopulation* from which they're both drawn.
$$
\begin{aligned}
\hat\mu(x) &= \frac{1}{N_x} \sum_{i:X_i=x} Y_i \quad \text{ for } \quad N_x = \sum_{i:X_i=x} 1  && \text{subsample mean} \\
\mu(x) &= \E[Y_i \mid X_i=x] && \text{infinite subpopulation mean} \\
\end{aligned}
$$

## Proof Prep: Notation and Speaking Conventions
  
  - We'll use capitals for random variables and lower case for nonrandom ones
    - e.g. we'll use $X$ for the random variable and $x$ for a value it can take on.
    - and  we'll say **big x** and **little x** to distinguish between the two.
    - same deal for $Y$ and $y$, \ $Z$ and $z$, etc.
  - When describing a random sample, we'll use $X_i$ for the $i$th person's value of $X$.
    - and we'll say **big x i** if we need to be precise. 
    - but we'll usually just say **big x** and assume you know what we mean.
   - This gives us a two ways to pronounce $X_i = x$ 
      - **big x i equals little x** is the precise one.
      - **big x equals little x** is the usual one.
   

## Proof Prep: Thinking clearly about conditioning

- It's important to distinguish between two things
  1.  The *conditional expectation function* $\mu(x)=\E[Y \mid X=x]$.
    - $\mu$ is a function; evaluated at $x$, it's a number. It's not random.
    - It's the mean of the infinite subpopulation of people with $X=x$.
    - This is the thing we're trying to estimate with $\hat\mu(x)$.
  2.  The conditional expectation $\mu(X)=E[Y \mid X]$.
    - $\mu(X)$ is the mean of a *random* infinite subpopulation of people.
    - It's the conditional expectation function *evaluated at* the random variable $X$.
    - This is the sort of thing that shows up when we use the law of iterated expectations.
- Let's do an easy calculation to try to make this clear. 
- We'll calculate the mean income in our population using the law of iterated expectations. 

## Example: Conditioning
  
  - Let's think about a random person drawn from this population
    - $Y$ is their income 
    - $X$ is an indicator for having a four-year degree. 
  - Suppose the subpopulation means are [100k]{.twocolor-green} for people with degrees 
    and [50k]{.twocolor-red} for people without.
  - And that 7/10 of our population has degrees.
  
  What is $\E(Y \mid X)$?

::: fragment

  - It's a random variable that is *either* 100k or 50k. A *random value* of the function $\mu$.
    - It's [$\mu(\with)=100k$]{.twocolor-green} with probability $7/10$,   when $X=\with$. 
    - And it's [$\mu(\without)=50k$]{.twocolor-red} with probability $3/10$, when $X=\without$.
:::

What is $\E(Y)$?

::: fragment

  - It's the mean of this random variable. A number.
$$
\begin{aligned} 
E\qty{ E\qty( Y \mid X ) } 
&= E(Y \mid X=\with)P(X=\with) + E(Y \mid X=\without)P(X=\without) \\
&= 100k \cdot 7/10 + 50k \cdot 3/10 = 70k + 15k = 85k.
\end{aligned}
$$

:::

## Unbiasedness Proof
### Step 1

- Rewrite so we're summing over the whole sample.
- To do this, we use subsample indicators $1_x(X_i)$.
- These are random variables that are 
  - 1 if the person is in the subsample 
  - 0 otherwise.
$$
\begin{aligned}
\hat\mu(x) 
&= \frac{1}{N_x} \sum_{i:X_i=x} Y_i               &&\text{ for } \quad N_x = \sum_{i:X_i=x} 1  \\
&= \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i)\ Y_i   &&\text{ for } \quad N_x = \sum_{i=1}^n \ind{x}(X_i) 
\end{aligned}
$$


### Indicators
:::: {.r-stack}

::: {.fragment .fade-out data-fragment-index="0"}
$$
\ind{x}(X_i) = \begin{cases} 1 & \text{ if }  X_i = x \\
                             0 & \text{ otherwise } \end{cases}
$$

:::

::::

## Unbiasedness Proof 
### Step 2

- Rewrite in terms of an expectation that's conditional on our covariates.
- The law of iterated expectations tells us we can.
$$
\begin{aligned}
\E \hat\mu(x) 
&= \E\qty[\frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i)\ Y_i] \quad &&\text{ for } \quad N_x = \sum_{i=1}^n \ind{x}(X_i) \\
&= \E\qty[ \inblue{\E\qty{ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i)\ Y_i  \mid X_1 \ldots X_n}} ] 
\end{aligned}
$$

### The law of iterated expectations
$$
\E Y = \E\qty[ \E\qty{ Y \mid X  } ] 
$$
To calculate the mean of $Y$, we can 
  - first average within subpopulations
  - then across subpopulations.

## Unbiasedness Proof
### Step 3

- Move the conditional expectation inward. It's linear, so we can.
- We can go all the way in to $Y_i$. Everything else is a function of the covariates.
$$
\begin{aligned}
\E \hat\mu(x) 
&= \E\qty[ \inblue{\E \qty{ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i)\ Y_i  \mid X_1 \ldots X_n}}  ]  &&\text{ for } \quad N_x = \sum_{i=1}^n \ind{x}(X_i) \\
&= \E\qty[ \frac{1}{N_x} \inblue{\E\qty{ \sum_{i=1}^n \ind{x}(X_i)\ Y_i \mid X_1 \ldots X_n }} ]  \\
&= \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \inblue{\E\qty{ \ind{x}(X_i)\ Y_i \mid X_1 \ldots X_n }} ]  \\
&= \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i)\ \inblue{\E\qty{ Y_i \mid X_1 \ldots X_n }} ]  
\end{aligned}
$$

:::: columns
::: column
### Linearity of expectations
$$ 
\begin{aligned}
\E\qty( a Y + b Z ) 
&= \E\qty(aY) + \E\qty(bZ) \\
&= a\E(Y) + b\E(Z)  
\end{aligned}
$$

:::

::: column

### Linearity of *conditional* expectations
$$ 
\begin{aligned}
\E\qty{ a(X) Y + b(X) Z \mid X } 
&= \E\qty{a(X)Y \mid X} + \E\qty{ b(X)Z \mid X} \\
&= a(X)\E\qty(Y \mid X) + b(X)\E\qty(Z \mid X)
\end{aligned}
$$

:::

::::

::: notes

  - Linearity tells us two things.
    - To average a sum of two things, we can take two averages and sum.
    - To average a constant times a random variable, we can multiply the random variable's average by the constant.
    - In other words, we can *distribute* expectations and can pull constants out of them.
  - Conditional linearity tells us the same thing, but with a twist.
    - When we condition on $X$, we're working with subpopulations in which $X$ is constant.
    - This means we can we can act as if functions of $X$ *are* constants.
    - So we can pull them out of expectations that are conditional on $X$.
    
:::


## Unbiasedness Proof
### Step 4

- Stop conditioning one person's outcome on *other people's covariates*. 
- They're independent, so we can. And the result is familiar. 
- We get each person's conditional mean outcome $\mu(X_i)$
$$
\begin{aligned}
\E \hat\mu(x) 
&= \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i)\ \inblue{\E\qty{ Y_i \mid X_1 \ldots X_n }} ]  &&\text{ for } \quad N_x = \sum_{i=1}^n \ind{x}(X_i)  \\
&= \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i)\ \inblue{\E\qty{ Y_i \mid X_i}} ] \\ 
&= \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i)\ \inblue{\mu(X_i)} ]  
\end{aligned}
$$

### Irrelevance of independent conditioning variables
$$ \E \qty( Y \mid X, X' ) = \E \qty( Y \mid X ) \quad \text{ when $X'$ is independent of $X,Y$ } $$
If $X'$ is unrelated to $X$ and $Y$, holding it constant doesn't affect the relationship between them.


## Unbiasedness Proof
### Step 5

- It's the *random function value* $\mu(X_i)$.
- We now replace with the fixed function value $\mu(x)$
- We can because it's multiplied by an *indicator* that's nonzero only when $X_i=x$.
$$
\begin{aligned}
\E \hat\mu(x) 
&= \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i)\ \inblue{\mu(X_i)} ]  &&\text{ for } \quad N_x = \sum_{i=1}^n \ind{x}(X_i)  \\ 
&= \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i)\ \inblue{\mu(x)} ]  
\end{aligned}
$$

### The Indicator Trick
$$
\begin{aligned}
\ind{x}(X_i)\ \mu(X_i) & \ \ = \ \ 
\begin{cases} 
\mu(X_i) & \text{ when } X_i=x \\
0 &      \text{ otherwise}
\end{cases} \\
& \ \ = \ \ 
\begin{cases}
\mu(x)  & \ \ \text{ when } X_i=x \\
0       & \ \ \text{ otherwise }
\end{cases} 
\ \ = \ \ \ind{x}(X_i) \ \mu(x)
\end{aligned}
$$



## Unbiasedness Proof
### Step 6

- We conclude by pulling the constant $\mu(x)$ out of the expectation.
- We can because expectations are linear.
- What's left inside is $1/N_x$ times $N_x$, i.e., 1.

$$
\begin{aligned}
\E \hat\mu(x) 
&= \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) \ \inblue{\mu(x)} ]   &&\text{ for } \quad N_x = \sum_{i=1}^n \ind{x}(X_i) \\
&= \inblue{\mu(x)} \ \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) ]  \\
&= \inblue{\mu(x)} \ \E[1] = \inblue{\mu(x)}. 
\end{aligned}
$$

### Linearity of Expectations
$$ 
\begin{aligned}
\E\qty( a Y + b Z ) 
&= \E\qty(aY) + \E\qty(bZ) \\
&= a\E(Y) + b\E(Z)  
\end{aligned}
$$


## Proof
### Summary
$$
{\tiny
\begin{aligned}
\E \hat\mu(x) =& \E\qty{\frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) Y_i}  \\
\underset{iterated}{=}& \E\qty[ \E\qty{ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) Y_i \mid X_1 \ldots X_n} ] \\
\underset{linearity}{=}& \E\qty[ \frac{1}{N_x} \E\qty{ \sum_{i=1}^n \ind{x}(X_i) Y_i \mid X_1 \ldots X_n} ]  \\ 
\underset{linearity}{=}& \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \E\qty{ \ind{x}(X_i) Y_i \mid X_1 \ldots X_n} ]  \\ 
\underset{linearity}{=}& \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) \E\qty{ Y_i \mid X_1 \ldots X_n} ]  \\
\underset{irrelevance}{=}& \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) \E\qty{Y_i \mid X_i} ] \\
\underset{definition}{=}& \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) \mu(X_i) ]  \\
\underset{indicator}{=}& \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) \mu(x) ] \\   
\underset{linearity}{=}& \mu(x) \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) ]   \\
\underset{definition}{=}& \mu(x) \E\qty[ \frac{1}{N_x} N_{x}    ] = \mu(x)\E[ 1 ] = \mu(x)   
\end{aligned}
}
$$

## Remark: Indicator Notation

- In this proof we used an *indicator* that $X_i$ was equal to $x$.
- This is a special case of a more general concept: *indicators* that $X_i$ is *in a set*.
  - In this special case, our set $\{x\}$ contained one element: $x$.
  - We've used another one, in which our set was an interval $[a \ldots b)$, when we talked about histograms.
- We can think of these as *indicator functions* evaluated at the random variable $X_i$.
$$
\begin{aligned}
1_{A}(X) &= \begin{cases} 1 & \text{ if } \ X \in A \\
                             0 & \text{ otherwise } \end{cases} && \text{general case} \\
\ind{x}(X) &= \begin{cases} 1 & \text{ if }\  X = x \\
                             0 & \text{ otherwise } \end{cases} && \text{one-element case} \\
1_{[a \ldots b)}(X) &= \begin{cases} 1 & \text{ if }\  a \le X < b \\
                             0 & \text{ otherwise } \end{cases} && \text{interval case} 
\end{aligned}
$$



## Remark: Indicator Expansion

- It's useful to generalize the *indicator trick* we've used in our proof.
- We can expand a function of a random variable $X$ as a sum of 
  - the value it takes at every distinct value of $X$
  - times an indicator that $X$ takes on that value
$$
f(X) = f(a) \ \ind{a}(X) + f(b) \ \ind{b}(X) + f(c) \ \ind{c}(X) + \ldots \qfor a\neq b \neq c \ldots.
$$

- In our proof, we used this for the function $\mu$.
- The subsample indicator $1_{x}(X_i)$ *picked out* one term in the expansion.
$$ 
\begin{aligned}
\ind{x}(X_i) \mu(X_i) &= \inblue{\ind{x}(X_i)} \sum_{x'} \inblue{\ind{x'}(X_i)} \ \mu(x') \\
                      &= \sum_{x'} \inblue{\ind{x}(X_i)\ind{x'}(X_i) }\ \mu(x') \\
                      &= \sum_{x'} \mu(x') \inblue{\begin{cases} 1 & \text{ if } X_i=x' \text{ and } X_i=x \\
                                                                 0 & \text{ otherwise } \end{cases}} \\
                      &= \mu(x)  
\end{aligned}
$$


::: notes
It's easy to check this equivalence. 

  - When $X=a$, the expansion's first indicator 'hits' and the others don't. 
  - That means the expanded version is $f(a)$. That's right.
  - It's the same for $b,c,\ldots$.
:::


# Below here is old

##  Properties of Population Means: Conditioning

### The law of iterated expectations
$$
\E\qty{ \E\qty( Y \mid X ) } \quad \text{ for any random variables $X, Y$} \\
$$
To calculate the mean of $Y$, we can average within subpopulations, then across subpopulations.

### Irrelevance of independent conditioning variables
$$ \E \qty( Y \mid X, X' ) = \E \qty( Y \mid X ) \quad \text{ when $X'$ is independent of $X,Y$ } $$
If $X'$ is unrelated to $X$ and $Y$, holding it constant doesn't affect the relationship between them.

## Properties of Population Means: Linearity

### The linearity of expectations
$$ 
\begin{aligned}
\E\qty( a Y + b Z ) 
&= \E\qty(aY) + \E\qty(bZ) \\
&= a\E(Y) + b\E(Z)  \quad \text{ for random variables $Y, Z$ and numbers $a,b$ }
\end{aligned}
$$
  There are two things going on here.

  - To average a sum of two things, we can take two averages and sum.
  - To average a constant times a random variable, we can multiply the random variable's average by the constant.
  - In other words, we can *distribute* expectations and can pull constants out of them.

### The linearity of *conditional* expectations
$$ 
\begin{aligned}
\E\qty{ a(X) Y + b(X) Z \mid X } 
&= \E\qty{a(X)Y \mid X} + \E\qty{ b(X)Z \mid X} \\
&= a(X)\E\qty(Y \mid X) + b(X)\E\qty(Z \mid X)
\end{aligned}
$$
   
This is like linearity of expectations, but with a twist.

   - When we condition on $X$, we're working with subpopulations in which $X$ is constant.
   - This means we can we can act as if functions of $X$ *are* constants.
   - So we can pull them out of expectations that are conditional on $X$.

## Thinking clearly about conditioning

- It's important to distinguish between two things
  1.  The *conditional expectation function* $\mu(x)=\E[Y \mid X=x]$.
    - $\mu$ is a function; evaluated at $x$, it's a number. It's not random.
    - It's the mean of the infinite subpopulation of people with $X=x$.
    - This is the thing we're trying to estimate with $\hat\mu(x)$.
  2.  The conditional expectation $\mu(X)=E[Y \mid X]$.
    - $\mu(X)$ is the mean of a *random* infinite subpopulation of people.
    - It's the conditional expectation function *evaluated at* the random variable $X$.
    - This is the sort of thing that shows up when we use the law of iterated expectations.
- Let's do an easy calculation to try to make this clear. 
- We'll calculate the mean income in our population using the law of iterated expectations. 

## Example: Conditioning
  
  - Let's think about a random person drawn from this population
    - $Y$ is their income 
    - $X$ is an indicator for having a four-year degree. 
  - Suppose the subpopulation means are [100k]{.twocolor-green} for people with degrees 
    and [50k]{.twocolor-red} for people without.
  - And that 7/10 of our population has degrees.
  
  What is $\E(Y \mid X)$?

::: fragment

  - It's a random variable that is *either* 100k or 50k. A *random value* of the function $\mu$.
    - It's [$\mu(\with)=100k$]{.twocolor-green} with probability $7/10$,   when $X=\with$. 
    - And it's [$\mu(\without)=50k$]{.twocolor-red} with probability $3/10$, when $X=\without$.
:::

What is $\E(Y)$?

::: fragment

  - It's the mean of this random variable. A number.
$$
\begin{aligned} 
E\qty{ E\qty( Y \mid X ) } 
&= E(Y \mid X=\with)P(X=\with) + E(Y \mid X=\without)P(X=\without) \\
&= 100k \cdot 7/10 + 50k \cdot 3/10 = 70k + 15k = 85k.
\end{aligned}
$$

:::


Unbiasedness of the sample mean
$$
\begin{aligned}
\text{Prove } \ \E \hat\mu(x) &= \mu(x) \qquad \text{ where } \\
\hat\mu(x) &= \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) Y_i \quad \text{ for } \quad N_x = \sum_{i=1}^n \ind{x}(X_i) \\
\mu(x) &= \E[Y_i \mid X_i=x] && 
\end{aligned}
$$


### Hint: The Indicator Trick

- Using the properties we've reviewed, we can write $\E \hat \mu(x)$ in terms of the random variable $\mu(X_i)$.
  - That's a *random value* of the conditional expectation function $\mu$.
  - We want an expression in terms of the value $\mu(x)$ it takes on when $X_i=x$. 
- In this context, it seems like they should be the same. Everyone in our subsample has $X_i=x$.
- To formalize this, we observe that $\mu(X_i)$ appears multiplied by the indicator $\ind{x}(X_i)$.
- And when that indicator is nonzero, $\mu(X_i)$ and $\mu(x)$ coincide.
$$
\mu(X_i)\ind{x}(X_i) = 
\begin{cases} 
\mu(X_i) & \text{ when } X_i=x \\
0 &      \text{ otherwise}
\end{cases} 
=
\begin{cases}
\mu(x)  & \text{ when } X_i=x \\
0       & \text{ otherwise }
\end{cases} 
= \mu(x)\ind{x}(X_i)
$$





## Variance ...


# Q1

# Q2

# Q3




# Junk after this point; maybe can be cut into something

# Question 1

##
```{r}
earnings.scatter + scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) 
```

**Q**. Which group tends to make more money? 

::: {.fragment}
People with 4 year degrees tend to make more.
:::

# Question 2

##
```{r}
earnings.histogram
```

**Q**. Roughly how much more?

::: {.fragment}
 Typically \$50k more, comparing the two peaks in the histogram.
:::

# Question 3

##

```{r}
earnings.scatter + scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) 
```

How would we calculate the difference in **mean income** between people in our sample with and without 4-year degrees?

### Outline
::: {.fragment}
1. We calculate the mean income among the *subsample* [with]{.twocolor-green} 4-year degrees.
2. We do the same for the subsample [without]{.twocolor-red} degrees. 
3. We subtract.
:::

## Step 1

```{r}
earnings.scatter + scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) 
```

We calculate the mean income among the *subsample* [with]{.twocolor-green} 4-year degrees.

:::: {.r-stack}

::: {.fragment .current-visible data-frame-index="1"}

We can write it out in mathematical notation.

$$
\hat\mu(\text{with}) = \frac{\sum_{i : X_i=\text{with}} Y_i }{ \sum_{i : X_i=\text{with}} 1 }
$$
:::

::: {.fragment .current-visible data-frame-index="2"}
```{r}
Y = earnings$income
X = earnings$degree
```

Writing it in *R* actually gives us an answer.

```{r}
#| echo: true
mu.hat.with = mean(Y[X == '4-year degree'])
```

What we get is roughly \$`{r} as.integer(round(mu.hat.with, digits=-3))`.
:::

:::: 

## Step 2 
```{r}
earnings.scatter + scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) 
```

We do the same for the subsample [without]{.twocolor-red} degrees. 

:::: {.r-stack}

::: {.fragment .current-visible data-frame-index="1"}

We can write it out in mathematical notation.

$$
\hat\mu(\text{with}) = \frac{\sum_{i : X_i=\without} Y_i }{ \sum_{i : X_i=\without} 1 }
$$
:::

::: {.fragment .current-visible data-frame-index="2"}

Writing it in *R* actually gives us an answer.

```{r}
#| echo: true
mu.hat.without = mean(Y[X != '4-year degree'])
```

What we get is roughly \$`{r} as.integer(round(mu.hat.without, digits=-3))`.
:::

:::: 


## Step 3
```{r}
earnings.scatter + scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) 
```

We subtract.

:::: {.r-stack}

::: {.fragment .current-visible data-frame-index="1"}
Here's the math.
$$
\hat \Delta = \hat\mu(\text{with}) - \hat\mu(\text{without})  \vphantom{ \frac{\sum_{i : X_i=\text{without}} Y_i} { \sum_{i : X_i=\text{without}} 1 } }
$$
:::

::: {.fragment .current-visible data-frame-index="2"}
Here's the code.
```{r}
#| echo: true
difference = mu.hat.with - mu.hat.without
```

What we get is roughly \$`{r} as.integer(round(difference, digits=-3))`.
:::

::::

# Question 4

## 
```{r}
earnings.histogram
```
What *Summary of the Population* does this estimate?

::: fragment

- The difference in mean income among *subpopulations* with and without 4-year degrees. \
- That's a difference in conditional expectations.
$$
\Delta = \mu(\with)-\mu(\text{without}) \quad \text{ where } \quad \mu(x)=\E[Y_i \mid X_i=x]
$$
:::

# Question 5

##

$$\text{What is our estimator's bias,} \ \E \hat\Delta - \Delta ? $$

::: {.fragment }
#### It's zero. Why?
:::

$$
$$

:::: {.r-stack}

::: {.fragment .current-visible data-frame-index="2"}
#### Part of the story: subsample means are **unbiased** estimators of subpopulation means.

  - In particular, $\hat\mu(x)$ is an unbiased estiate of $\mu(x)$.
  $$
  \E\hat\mu(x)=\mu(x)
  $$
  - We proved this in Homework ? using the law of iterated expectations.
  
$$
TODO: Proof
$$

:::

::: {.fragment .current-visible data-frame-index="3"}
#### The rest: Expectation is **linear**. 

  - The expectation of a difference is the difference in expectations.
  - Subtracting unbiased estimates of our subpopulation means gives us \
    an unbiased estimate of the **difference in those subpopulation means**.
$$ 
\begin{aligned}
\E \hat\Delta 
&= \E \left\{ \hat\mu(\text{with}) - \hat\mu(\text{without}) \right\} && \text{definition} \\
&= \E \hat\mu(\text{with}) - \E \hat \mu(\text{without}) && \text{linearity} \\
&= \mu(\text{with}) - \mu(\text{without})  && \text{unbiasedness} \\
&= \Delta && \text{definition} 
\end{aligned}
$$
:::

::::

# Question 6

##

How can we approximate the *sampling distribution* of this estimator?

$$
$$

TODO: Picture of sampling dist vs approximation

### Outline 

::: fragment
1. We find normal approximations to the sampling distributions of each of our subsample means.
2. We characterize their difference as another normal distribution. 
3. We calculate its mean.
4. We calculate its variance.
:::

## Step 1

We observe that our subsample means are *approximately independent and normal* 
$$
\hat\mu(x) \approx Z_x \quad \text{ for } \quad Z_x \sim N\left( \E \hat\mu(x), \ \Var \hat\mu(x) \right)
$$
and
$$
Z_{x} \quad \text{ is independent of } \quad Z_{x'}  \quad  \text{ when } \quad x \neq x'. 
$$

They're approximately independent because our subsamples are *disjoint*, i.e., nobody is in both.


## Step 2

We'll approximate the difference in our subsample means by 
the corresponding difference of *independent normals*.
$$
\hat\mu(\text{with}) - \hat\mu(\text{without}) \approx Z_{\text{with}} - Z_{\text{without}}
$$
Then use the rule for subtracting independent normals: subtract the means; add the variances.
$$
\begin{aligned}
Z_{\text{with}} - Z_{\text{without}} &\sim  N\left( \mu_{\Delta}, \sigma^2_\Delta \right) \\
\text{ where } \quad    \mu_{\Delta} &= \E \hat\mu(\text{with})  - \E \hat\mu(\text{without}) \\
\text{ and  } \ \ \quad \sigma^2_\Delta &= \Var \hat\mu(\text{with}) + \Var \hat\mu(\text{without}) 
\end{aligned}
$$ 

## Step 3
We subtract the means, i.e., we calculate this difference. 
$$
\mu_{\Delta} = \E \hat\mu(\text{with})  - \E \hat\mu(\text{without}). 
$$
We know, from our last question, that this is the population summary we want.
$$\mu_{\Delta} = \hat\mu(\text{with})  - \hat\mu(\text{without}) = \Delta$$

## Step 4
We add the variances, i.e., we calculate this sum.
$$ 
\sigma^2_\Delta = \Var \hat\mu(\text{with}) + \Var \hat\mu(\text{without})
$$
The **variance** of each subsample mean $\hat\mu(x)$ is determined by 

1. The conditional variance of the terms in it
2. The size of the subsample.

$$ 
\sigma^2_\Delta = \frac{\Var(Y_i \mid X_i = \text{with})}{n_\text{with}} + \frac{\Var(Y_i \mid X_i = \text{without})}{n_{\text{without}}} 
$$

TODO: Do we define $n_\text{with}$ and $n_\text{with}$ properly as expectations of inverses of proportions?

## Summary.

We can approximate the sampling distribution of our estimator
by that of a *normal random variable* with the mean we want.
$$ 
\begin{aligned}
\hat\mu(\text{with}) - \hat\mu(\text{without}) &\approx Z_{\Delta} \quad \text{ for } \quad Z_{\Delta} \sim N(\Delta, \sigma^2_\Delta) \\
\text{ where } \quad \sigma^2_\Delta &= \frac{\Var(Y_i \mid X_i = \text{with})}{n_\text{with}} + \frac{\Var(Y_i \mid X_i = \text{without})}{n_{\text{without}}}
\end{aligned}
$$


# Question 7

##
How can we use this to calculate a 95\% confidence interval for this summary $\Delta$?

$$ $$

### Outline
::: fragment

1. We transform our estimate into a *pivot*: a random variable with an known sampling distribution.
2. We choose an interval that contains the pivot with 95\% probability.
3. We invert our transformation to get an interval for our estimate.

TODO: Fix up to deal with approximate pivots.
:::


## Step 1
t-stat calculation + picture of std normal reference distn

## Step 2
formula + picture of interval in reference distribution

## Step 3
formula + picture of transformed reference distributon

## Summary
something


## Most of the Remainder
We'll answer the questions

## Hopefully this fits at the end
- We'll explain why we might need an unsaturated model and how that results in bias.
- We'll imagine a reality in which we know whether there are 1 or 2 year masters degrees
  and think of the bias associated with grouping them for Question 3.
