---
title: "Working with Non-Binary Covariates"
subtitle: "QTM 220 Lecture 5"
format: 
  revealjs:
    theme: class.scss
    show-notes: separate-page
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    fig.width: 10
    fig.asp: 2
knitr:
  opts_chunk:
    fig.path: generated-figures/
execute: 
  cache: true
---



```{r setup}
library(purrr)
library(ggplot2)
library(htmltools)

source('../../shared-code/plotting.R')
theme_set(simplified.theme)
knitr::opts_chunk$set(dev.args=list(bg="transparent"))
```

```{r data-setup}
# for cps variable interpretation see https://www2.census.gov/programs-surveys/cps/datasets/2022/march/asec2022_ddl_pub_full.pdf
census.api.key = '8084f81e4f71e4d32638dac15f188b424f72d063'
library(cpsR)
library(dplyr)
library(magrittr)
library(kableExtra)
library(scales)

cps.data = cpsR::get_asec(year=2022, vars=c('h_numper', 'htotval', 'a_age', 'a_hga'), key=census.api.key)
all.earnings = data.frame(age=cps.data$a_age,
                          income=cps.data$htotval, 
                          education=case_match(cps.data$a_hga,
                                           0  ~ 0,   # child
                                           31 ~ 0,   # < grade 1
                                           32 ~ 4,   #  grade 1-4
                                           33 ~ 6,   #  grade 5-6
                                           34 ~ 8,   #  grade 7-8 
                                           35 ~ 9,   #  grade 9
                                           36 ~ 10,  #  grade 10
                                           37 ~ 11,  #  grade 11
                                           38 ~ 11,  #  grade 12 no diploma
                                           39 ~ 12,  #  high school grad
                                           40 ~ 13,  #  some college
                                           41 ~ 14,  #  associate degree (vocational)
                                           42 ~ 14,  #  associate degree (academic)
                                           43 ~ 16,  #  bachelors degree
                                           44 ~ 18,  #  masters degree
                                           45 ~ 20,  #  professional school degree
                                           46 ~ 20)) #  doctorate

in.interval = function(X, interval) { interval[1] <= X & X <= interval[2] }
earnings = all.earnings[all.earnings$age %>% in.interval(c(25,35)) & 
                        all.earnings$education >= 11, ]

earnings$raw.income = earnings$income
earnings$degree = factor(earnings$education >= 16, 
                     levels=c(FALSE, TRUE), 
                     labels=c('no 4-year degree','4-year degree')) 
rownames(earnings)=NULL
```

# Intuition

## CPS Income Data

::: {.hidden}
$$
\require{color}
\definecolor{twocolor-red}{HTML}{F8766D}    % red   used by ggplot in plots w/ 2 colors 
\definecolor{twocolor-green}{HTML}{00BFC4}  % green used by ggplot in plots w/ 2 colors
$$ 
:::

So far, we've been looking at our income data as if people fall into two groups. \
Our groups are people [with]{.twocolor-green} and [without]{.twocolor-red} 4-year degrees. Here's some of the data.
```{r}
earnings.tab = earnings[1:12, c('income', 'degree')]
earnings.tab$income = sprintf('$%dk', round(earnings.tab$income / 1000))
earnings.tab %>% 
  kable() %>%
  kable_styling() %>%
  row_spec(which(earnings.tab$degree=='4-year degree'), background='#00BFC4') %>%
  row_spec(which(earnings.tab$degree!='4-year degree'), background='#F8766D') 
```

In what's shown here, most people with 4-year degrees make more than most without. 
But not all of them. And this is just a small fraction of the data. Let's look at all of it.

## A Scatter Plot

:::: {.r-stack }

::: {.fragment .fade-out data-fragment-index="0"}
```{r}
p = ggplot(earnings) + geom_point(aes(x=degree, y=income, color=degree), position=position_jitter(w=.4), alpha=.2) + 
                       scale_y_continuous(labels=label_dollar()) + theme(legend.position="none") 
p
```

- Here I've plotted one dot per person. 
- There's lot of people, so the dots are all lying on top of one another. 
- You can see what most people make by looking for a dark band.
:::

::: {.fragment .current-visible data-fragment-index="0"}
```{r}
p +  scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) 
```

- But you really have to look carefully to compare them.
- It's hard to see much because so little space is given to what most people make.
- We need to do that to include people who make *a lot* of money.
::: 

::: {.fragment .current-visible}
```{r}
p +  scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) 
```

- We can see things a bit better when we zoom in on people who make $\le \$300$k. 
- This leaves out a few millionaires, but that's mostly fine. We can do that.
- And if that's still hard to work with, we can use another visualization.
:::

::::

## A Histogram

```{r}
ggplot(earnings[earnings$income<3e5, ]) + 
                   geom_histogram(aes(x=income, y=after_stat(density)), alpha=.5, bins=40) + 
                   coord_flip() + facet_wrap(~degree, strip.position='bottom') + aes(fill=degree) + 
                   scale_x_continuous(labels=label_dollar()) + 
                   theme(legend.position="none", axis.text.x = element_blank()) 
```

:::: {.r-stack}

::: {.fragment .fade-out data-fragment-index="0"}

- The bar length at each income level indicates the fraction 
      of people that make roughly that amount of money.
- You can see what most people make by looking for the peak.

:::

::: {.fragment .current-visible data-fragment-index="0"}
**Q**. Which group tends to make more money? And how much more? 

:::

::: {.fragment .fade-in data.fragment-index="1"}

- People with 4 year degrees tend to make more.
- Roughly \$50k more, comparing the two peaks.
- Recall confidence statement about the difference in means from last week.

:::

::::


## Taking a Closer Look
Using two categories is throwing away a lot of information about education. \
We can, for example, categorize people by the **years of education** they've had. 
:::: {.r-stack}

::: {.fragment}
```{r}
years = unique(earnings$education)
ggplot(earnings) + geom_point(aes(x=education, y=income, color=degree), position=position_jitter(w=.4), alpha=.2) +
                   scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) + 
                   scale_x_continuous(breaks=years, labels=sprintf('%d years', years)) + 
                   theme(legend.position="none")
```

This includes same information we were working with before. 
$$ \text{ 16 years education } \quad \implies \quad  \text{ 4-year degree } $$
If we shuffled all the red and green dots, we'd get the plot we had before. 

This means we can still see that people with 4 year degrees make more. \
But we can see some new stuff too. What else can we see? 

A thing: Even among people with (and without) a 4-year degree, \
people with more education tend to make more money.
:::

::: {.fragment}
 ```{r}
ggplot(earnings[earnings$income<3e5, ]) + 
                   geom_histogram(aes(x=income, y=after_stat(density)), alpha=.5, bins=40) + aes(fill=degree) + 
                   coord_flip() + facet_wrap(~education, strip.position='bottom', ncol=length(unique(earnings$education))) + 
                   scale_x_continuous(labels=label_dollar()) + 
                   theme(legend.position="none", axis.text.x = element_blank()) 
```
:::

::::




