---
title: "Working with Non-Binary Covariates"
subtitle: "QTM 220 Lecture 5"
format: 
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---

```{r load-libraries}
library(purrr)
library(tibble)
library(dplyr)
library(magrittr)
library(stringr)
library(ggplot2)
library(htmltools)
library(kableExtra)
library(scales)
library(cpsR)
library(mapdata)

source('../../shared-code/plotting.R')
theme_set(simplified.theme)
set.seed(1)

compact_number = label_number(accuracy=.1, scale_cut=cut_short_scale())
```

```{r get-data}
#| cache: true
#| 
# for cps variable interpretation see https://www2.census.gov/programs-surveys/cps/datasets/2022/march/asec2022_ddl_pub_full.pdf
census.api.key = '8084f81e4f71e4d32638dac15f188b424f72d063'

california.state.code = 6
years = 2022
cps.data = map_dfr(years, function(year) {
  get_asec(year=year, vars=c('pearnval', 'a_age', 'a_hga', 'gestfips', 'gtco', 'gtcbsasz'), key=census.api.key)
})
```

```{r transform-data}
#| cache: true
#| 
cps.sampling.ratio = 1/2500
all.earnings = data.frame(state.code = cps.data$gestfips,
                          county.code = cps.data$gtco,
                          age=cps.data$a_age,
                          income=cps.data$pearnval,
                          education.code=cps.data$a_hga,
                          education=case_match(cps.data$a_hga,
                                           0  ~ 0,   # child
                                           31 ~ 0,   # < grade 1
                                           32 ~ 4,   #  grade 1-4
                                           33 ~ 6,   #  grade 5-6
                                           34 ~ 8,   #  grade 7-8 
                                           35 ~ 9,   #  grade 9
                                           36 ~ 10,  #  grade 10
                                           37 ~ 11,  #  grade 11
                                           38 ~ 11,  #  grade 12 no diploma
                                           39 ~ 12,  #  high school grad
                                           40 ~ 13,  #  some college
                                           41 ~ 14,  #  associate degree (vocational)
                                           42 ~ 14,  #  associate degree (academic)
                                           43 ~ 16,  #  bachelors degree
                                           44 ~ 18,  #  masters degree
                                           45 ~ 20,  #  professional school degree
                                           46 ~ 20), #  doctorate
                        metro.size=case_match(cps.data$gtcbsasz, # cps reports a metro-size category, we map that to an approximate metro size
                                           0~0,       # < 100k
                                           2~1e5,     # 100-250k
                                           3~2.5e5,   # 250-500k
                                           4~5e5,     # 500k-1 mil
                                           5~1e6,     # 1-2.5 mil
                                           6~2.5e6,   # 2.5-5 mil     
                                           7~5e6))    # >= 5 mil  

in.interval = function(X, interval) { interval[1] <= X & X <= interval[2] }
earnings = all.earnings[all.earnings$state.code == california.state.code &
                        all.earnings$age %>% in.interval(c(25,35)), ]

earnings$degree = factor(earnings$education >= 16, 
                     levels=c(FALSE, TRUE), 
                     labels=c('no 4-year degree','4-year degree')) 
```

```{r make-maps}
#| cache: true
#| 

# To give our survey recipients a bit more reality, we'll give them fake locations on a map
# And we'll make up a population of California residents that they're sampled from

# maps are based on code from https://eriqande.github.io/rep-res-eeb-2017/map-making-in-R.html
# additional county info from https://en.wikipedia.org/wiki/List_of_counties_in_California
# converted to csv using https://wikitable2csv.ggor.de

to.number = function(s) { as.numeric(str_trim(gsub(",", "", s))) }

counties = map_data("county")
ca_county = counties %>%
  filter(region == "california")
ca_county_centers = ca_county %>%
  group_by(subregion) %>%
  summarize(county.lat = mean(lat), county.long=mean(long))
ca_county_info = read.csv('california_county_info.csv') %>%
  mutate(subregion = tolower(sub(' County', '', County))) %>%
  mutate(pop = to.number(Population..2022.)) %>%
  mutate(area = to.number(sub('sq.*', '', Area))) 
ca_county_more = left_join(ca_county_centers, ca_county_info, by="subregion")
ca_all = left_join(ca_county, ca_county_more, by="subregion")


ca.plot = ggplot(data=ca_all, mapping=aes(x=long, y=lat, group=group)) + 
  coord_quickmap() +
  geom_polygon(fill='gray', color = "white", alpha=.4) +
      scale_fill_gradient(trans="log10") + 
      theme_void()

ca.pop.plot = ggplot(data=ca_all, mapping=aes(x=long, y=lat, group=group)) + 
  coord_quickmap() +
  geom_polygon(aes(fill = pop/area), color = "white") +
      scale_fill_gradient(trans="log10") + 
      theme_void()

## we'll make up a population as follows
# 1. choose the number of people to be the number of people in our sample scaled up to account for the cps sampling ratio
# 2. for each, sample a county with probability proportional to its finite.population
# 3. for each, sample a latitude and longitude by taking a randomly weighted combination of the    
#    county's vertices. As long as the county is convex, these points will be inside the county.
# 4. we'll give them a degree too, chosen independently of all else from the empirical distribution on the sample
n = nrow(earnings) / cps.sampling.ratio

# takes a list of subregions, possibly with duplicates
# for each element, chooses random lat/long coordinates within the sample.subregion
# by taking a random convex combination of the subregion's vertices
sample.within.subregion = function(subregions) {
  output = data.frame(subregion=subregions, lat=NA, long=NA)
  for(subregion in unique(subregions)) { 
      vertices = ca_county[ca_county$subregion==subregion, c('lat','long')]
      U = sum(subregions==subregion)
      V = nrow(vertices)
      weights = abs(rt(U*V, 1))
      dim(weights) = c(U,V)
      weights = (1/rowSums(weights)) * weights
      output[subregions==subregion, c('lat', 'long')] = 
                 cbind(weights %*% vertices$lat, weights %*% vertices$long)
    }
    output
}

finite.population.subregion = sample(ca_county_more$subregion, size=n, replace=TRUE, prob=ca_county_more$pop)
finite.population = sample.within.subregion(finite.population.subregion)
finite.population$degree = sample(earnings$degree, nrow(finite.population), replace=TRUE)
finite.population$group=1 # for plotting

display.population = sample_n(finite.population, min(100000, nrow(finite.population)))
```


```{r make-sample}
# and we'll force the degree-distribution in the sample to match our survey responses
our.sample = sample_n(finite.population, nrow(earnings))
our.sample$degree=earnings$degree

pop.plot = ca.plot + geom_point(aes(x=long, y=lat), data=display.population, size=.02, alpha=.1) 
sample.plot = ca.plot + geom_point(aes(x=long, y=lat), data=our.sample, size=.2, alpha=.5)
pop.plot.inf = ca.pop.plot + geom_point(aes(x=long, y=lat), data=display.population, size=.01, alpha=.1)  
sample.plot.inf = ca.pop.plot + geom_point(aes(x=long, y=lat), data=our.sample, size=.2, alpha=.5)
```

## Summary

::: {.hidden}
$$
% color defs aren't making it into MathJax
%\definecolor{twocolor-red}{RGB}{248, 118, 109}   % red   used by ggplot in plots w/ 2 colors 
%\definecolor{twocolor-green}{RGB}{0, 191, 196}   % green used by ggplot in plots w/ 2 colors
% so create commands for using those colors instead
\newcommand{\inred}[1]{\textcolor[RGB]{248,118,109}{#1}}
\newcommand{\ingreen}[1]{\textcolor[RGB]{0,191,196}{#1}}
\newcommand{\inblue}[1]{\textcolor{blue}{#1}}

\newcommand{\ind}[1]{1_{\{#1\}}}
\newcommand{\with}{\ingreen{\text{degree}}}
\newcommand{\without}{\inred{\text{no degree}}}

\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\Var}{Var}
$$ 

```{r define-colors}
twocolor.red   = 'rgb(248,118,109)'
twocolor.green = 'rgb(0,191,196)'
```
:::

- Today, we'll work with data from the 2022 Current Population Survey.
- We'll look at the relationship between income and education.
- We've already compared people  [with]{.twocolor-green} and [without]{.twocolor-red} 4-year degrees.
- That's a way of summarizing what we see in the data based on a *binary covariate*
$$ 
X_i = \begin{cases} 
  \with & \text{ if } \ \text{person}_i \ \text{has a 4-year degree} \\
  \without & \text{ otherwise }
  \end{cases}
$$
- Today, we'll talk about summaries based on a *many-valued covariate*
- We'll categorize people by the years of education they've had

## The Sample

- We'll look at surveyed California residents between the ages of 25 and 35.
- This sample includes `r nrow(earnings)` people. To fit our table on the page, we're showing a few chosen at random. 

$$
$$

:::: columns

::: column
```{r map-sample-no-color}
#| fig-width: 5 
sample.plot
```
:::

::: column

```{r table-no-color}
earnings.county = left_join(earnings, ca_county_info[,c('FIPS.code', 'subregion')], by=c('county.code' = 'FIPS.code'))
earnings$county = earnings.county$subregion
random.rows = sample(nrow(earnings), 12)

style.table.no.color = function(tab) {
  tab$income = sprintf('$%dk', round(tab$income / 1000))
  rownames(tab) = NULL
  kable(tab) %>%
    kable_styling()
}
style.table.no.color( earnings[random.rows, c('income', 'education', 'age', 'county')] )
```
:::



::::

::: aside
The locations shown on the map are made up. They're not the actual locations of the people in the sample. \
The survey includes some location information for some people, but as you can see in the table, not for everyone.
:::

## Our Interpretation so Far

- So far, we've been thinking of education in binary terms.
- We've been comparing two subsamples: people [with]{.twocolor-green} and [without]{.twocolor-red} 4-year degrees.

$$
$$

:::: columns

::: column
```{r map-sample-with-color}
#| fig-width: 5 
ca.plot + geom_point(aes(x=long, y=lat, color=degree), data=our.sample, size=.1, alpha=.5) + guides(color=FALSE)
```
:::

::: column
```{r table-with-color}
style.table = function(tab) {
  tab$income = sprintf('$%dk', round(tab$income / 1000))
  rownames(tab) = NULL
  kable(tab) %>%
    kable_styling() %>%
    row_spec(which(tab$education >= 16), background=twocolor.green) %>%
    row_spec(which(tab$education < 16), background=twocolor.red) 
}
style.table( earnings[random.rows, c('income', 'education', 'age', 'county')] )
```
:::

::::

::: aside
The locations shown on the map are made up. They're not the actual locations of the people in the sample. \
The survey includes some location information for some people, but as you can see in the table, not for everyone.
:::


## What we Lost


:::: {.r-stack}

```{r}
dot.size = 2
years = unique(earnings$education)
```

::: {.fragment .fade-out data-fragment-index="0"}
We were throwing away a lot of information thinking of education as binary.
```{r scatter-plot-degree}
#| out-height: 600px
ggplot(earnings) + geom_point(aes(x=degree, y=income, color=degree), 
                              size=dot.size, position=position_jitter(w=.3), alpha=.1) +
                   scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) + 
                   guides(color=FALSE)
```
:::

::: {.fragment .current-visible data-fragment-index="0"}
Today, we'll categorize people by the **years of education** they've had. 
```{r scatter-plot-years}
#| out-height: 600px
#| 
earnings.jitter = ggplot(earnings) + geom_point(aes(x=education, y=income, color=degree), 
                              size=dot.size, position=position_jitter(w=.3), alpha=.1) +
                   scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) + 
                   scale_x_continuous(breaks=years, labels=sprintf('%dy', years), limits=c(-1,21)) + 
                   guides(color=FALSE) 
earnings.jitter
```
:::

::: {.fragment .current-visible data-fragment-index="1"}
By using the binary version, we were in effect **shuffling** the years of education within each group. 
```{r scatter-plot-years-shuffled}
#| out-height: 600px
earnings.shuffle = earnings
earnings.shuffle$education[earnings$education >= 16] = sample(earnings$education[earnings$education >= 16])
earnings.shuffle$education[earnings$education <  16] = sample(earnings$education[earnings$education <  16])
earnings.jitter.shuffled = ggplot(earnings.shuffle) + geom_point(aes(x=education, y=income, color=degree), 
                              size=dot.size, position=position_jitter(w=.3), alpha=.1) +
                   scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) + 
                   scale_x_continuous(breaks=years, labels=sprintf('%dy', years), limits=c(-1,21)) + 
                   guides(color=FALSE) 
earnings.jitter.shuffled
```
:::

::: {.fragment .current-visible data-fragment-index="2"}
This is easier to see looking at summary statistics. Here's the mean income for each group in black.
```{r scatter-plot-years-mean}
#| out-height: 600px
summaries = earnings %>% 
          group_by(education) %>% 
          summarize(mean=mean(income), sd=sd(income))
earnings.jitter + geom_point(aes(x=education, y=mean), size=dot.size, data=summaries)
```
:::

::: {.fragment .current-visible data-fragment-index="3"}
Here's the same thing, with the years of education shuffled within each group.
```{r scatter-plot-years-mean-shuffled}
#| out-height: 600px
summaries.shuffle = earnings.shuffle %>% 
          group_by(education) %>% 
          summarize(mean=mean(income), sd=sd(income))
earnings.jitter.shuffled + geom_point(aes(x=education, y=mean), data=summaries.shuffle, size=dot.size)
```
:::

::::

## dfsd

:::: {.r-stack}

::: {.fragment .fade-out data-fragment-index="0"}
### We haven't lost anything.

- You have a four year degree if, and only if, you have $12+4=6$ years of education.
- Thinking visually, we can get the plot we had before by shuffling the red and green dots.
- This means we can still see how much people with and without 4-year degrees tend to make.
- But not shuffling lets us see some new stuff, too. 

### What else is there to see?
:::

::: {.fragment .current-visible data-fragment-index="0"}

### What do you see, in purely visual terms?
$$
$$

### What does it mean?
$$
$$

:::

::::

::: {.notes}

### One thing I see
The centers of our subsamples are going up as we move to the right.

### Meaning
Even among people with (or without) a 4-year degree, 
people with more education tend to make more money.

:::


## Pre-processing

```{r}
#| echo: true
education=case_match(cps.data$a_hga,
                          0  ~ 0,   # child
                          31 ~ 0,   # < grade 1
                          32 ~ 4,   #  grade 1-4
                          33 ~ 6,   #  grade 5-6
                          34 ~ 8,   #  grade 7-8 
                          35 ~ 9,   #  grade 9
                          36 ~ 10,  #  grade 10
                          37 ~ 11,  #  grade 11
                          38 ~ 11,  #  grade 12 no diploma
                          39 ~ 12,  #  high school grad
                          40 ~ 13,  #  some college
                          41 ~ 14,  #  associate degree (vocational)
                          42 ~ 14,  #  associate degree (academic)
                          43 ~ 16,  #  bachelors degree
                          44 ~ 18,  #  masters degree
                          45 ~ 20,  #  professional school degree
                          46 ~ 20)  #  doctorate
```         

- We've had to make a few things up in translation.
  - There's no differentiation between a 1 or 2-year masters; we assume 2.
  - We're assuming professional and doctoral degrees take 4 years; that's approximate.
  - Nobody has 17 or 19 years in our plots; that's why.
- That's part of data analysis. 
- Ultimately, you'll want to be careful about quirks introduced by translation.
  - But it's important to keep your job manageable. 
  - It can help to ignore the quirks when you're getting started.


## Subsamples 

```{r}
#| out-height: 250px
years = unique(earnings$education)
ggplot(earnings) + geom_point(aes(x=education, y=income, color=degree), 
                              size=dot.size, position=position_jitter(w=.2), alpha=.2) +
                   scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) + 
                   scale_x_continuous(breaks=years, labels=sprintf('%d years', years)) + 
                   guides(color=FALSE)

```


:::: {.r-stack}

::: {.fragment .fade-out data-fragment-index="0"}  
- When we thought of education as binary, we only had two subsamples to talk about.
  - People with 4-year degrees --- the [green dots]{.twocolor-green}.
  - People without them --- the [red dots]{.twocolor-red}.
- Now that we're taking a more granular view, we can talk about *many*.
:::

::: {.fragment .current-visible data-fragment-index="0"}
- People with $x$ years of education for $x=10,11,\ldots$.
     - This is the *finest* way to break down our sample in terms of $X$.
- People with and without 4-year degrees, as before.
     - This is a coarser breakdown. They're *aggregates*.
     - To get these, we combine our finer subsamples.
- People whose highest degree is graduate, 4-year,  2-year, high-school, or none.
     - That's in between. It's a finer aggregate.

:::

::::


## New Questions

Taking this more granular view of education allows us to pose new questions.

1. What's the typical difference in income between people with *exactly* 16 years of education
   and *exactly* 12?
    - This is a different question than the one we addressed in review, but it has a similar structure.
    - We want a summary of income differences in two groups. Those can be subsamples or subpopulations.
    - We can calculate and relate these summaries exactly like we did in review.
2. What's the typical difference in income between people with 4-year degrees and people without them?
    - That's exactly the question we talked about in review.
    - We'll talk about how to calculate it a different way that'll generalize better to other questions.
    - We can check our work. Obviously our answers should be exactly the same as we got earlier in the semester.
3. How much change would we expect in a similar population in which people tend to get an additional year of education?
    - This is a question unlike the ones we talked about earlier.
    - We'll need to be a bit careful about what we mean.
    - But once we've gotten there, we can base our calculations on ones we've done already.

More for HW.

## Q1

```{r}
#| out-height: 250px
earnings$highest.degree = case_match(earnings$education,
                            0:11 ~ 'none',
                            12:13 ~ 'high school',
                            14  ~   'associates',
                            16  ~   'bachelors',
                            18:20 ~ 'graduate')

fine = earnings %>% group_by(education) %>% summarize(mean.income = mean(income))
coarse = earnings %>% group_by(degree) %>% summarize(mean.income = mean(income), lb=min(education), ub=max(education)) 
intermediate = earnings %>% group_by(highest.degree) %>% summarize(mean.income = mean(income), lb=min(education), ub=max(education))

base.plot = ggplot(earnings) + geom_point(aes(x=education, y=income, color=degree), 
                              size=dot.size, position=position_jitter(w=.2), alpha=.2) +
                   scale_y_continuous(labels=label_dollar(), limits=c(0,3e5)) + 
                   scale_x_continuous(breaks=years, labels=sprintf('%d years', years)) + 
                   guides(color=FALSE)

base.plot + geom_segment(aes(x=education-.2, xend=education+.2, y=mean.income, yend=mean.income), data=fine)
```



## Q2
```{r}
base.plot + geom_segment(aes(x=lb-.2, xend=ub+.2, y=mean.income, yend=mean.income), data=coarse) 
```

  - Means of Aggregates are Aggregates of Means
  - Weighted Aggregates
  - [do somethoing with fine picture combo]

## Q3

- some kind of picture showing jumps



# Preparation for Inferential Analysis

## Review of Calculation Techniques

- All of our questions are comparisons, so the closest we've come is the comparison we did in the binary case.
- We'll review what happens when we estimate the difference in population means by the difference in sample means.
- In particular, we'll calculate an approximation that estimator's sampling distribution.
  - We'll start by showing that it's unbiased.
  - Then we'll calculate its variance.
  - We'll wave our hands about normality.

## Unbiasedness of subsample means
- We'll start by showing that a subsample's mean is an unbiased estimate of the corresponding subpopulation mean.
  - We'll think about the subsample and subpopulation with $X_i=x$ for some value $x$ 
  - e.g. $X_i=\with$ for people with 4-year degrees.
- That means its expectation *is* the subpopulation mean. 
- We can think of our subsample mean as an estimate of the mean of the corresponding *finite subpopulation*.
- But we'll focus on interpreting it as an estimate of the mean of the *infinite subpopulation* from which they're both drawn.
$$
\begin{aligned}
\hat\mu(x) &= \frac{1}{N_x} \sum_{i:X_i=x} Y_i \quad \text{ for } \quad N_x = \sum_{i:X_i=x} 1  && \text{subsample mean} \\
\mu(x) &= \E[Y_i \mid X_i=x] && \text{infinite subpopulation mean} \\
\end{aligned}
$$

## Proof Prep: Notation and Speaking Conventions
  
  - We'll use capitals for random variables and lower case for nonrandom ones
    - e.g. we'll use $X$ for the random variable and $x$ for a value it can take on.
    - and  we'll say **big x** and **little x** to distinguish between the two.
    - same deal for $Y$ and $y$, \ $Z$ and $z$, etc.
  - When describing a random sample, we'll use $X_i$ for the $i$th person's value of $X$.
    - and we'll say **big x i** if we need to be precise. 
    - but we'll usually just say **big x** and assume you know what we mean.
   - This gives us a two ways to pronounce $X_i = x$ 
      - **big x i equals little x** is the precise one.
      - **big x equals little x** is the usual one.
   

## Proof Prep: Thinking clearly about conditioning

- It's important to distinguish between two things
  1.  The *conditional expectation function* $\mu(x)=\E[Y \mid X=x]$.
    - $\mu$ is a function; evaluated at $x$, it's a number. It's not random.
    - It's the mean of the infinite subpopulation of people with $X=x$.
    - This is the thing we're trying to estimate with $\hat\mu(x)$.
  2.  The conditional expectation $\mu(X)=E[Y \mid X]$.
    - $\mu(X)$ is the mean of a *random* infinite subpopulation of people.
    - It's the conditional expectation function *evaluated at* the random variable $X$.
    - This is the sort of thing that shows up when we use the law of iterated expectations.
- Let's do an easy calculation to try to make this clear. 
- We'll calculate the mean income in our population using the law of iterated expectations. 

## Example: Conditioning
  
  - Let's think about a random person drawn from this population
    - $Y$ is their income 
    - $X$ is an indicator for having a four-year degree. 
  - Suppose the subpopulation means are [100k]{.twocolor-green} for people with degrees 
    and [50k]{.twocolor-red} for people without.
  - And that 7/10 of our population has degrees.
  
  What is $\E(Y \mid X)$?

::: fragment

  - It's a random variable that is *either* 100k or 50k. A *random value* of the function $\mu$.
    - It's [$\mu(\with)=100k$]{.twocolor-green} with probability $7/10$,   when $X=\with$. 
    - And it's [$\mu(\without)=50k$]{.twocolor-red} with probability $3/10$, when $X=\without$.
:::

What is $\E(Y)$?

::: fragment

  - It's the mean of this random variable. A number.
$$
\begin{aligned} 
E\qty{ E\qty( Y \mid X ) } 
&= E(Y \mid X=\with)P(X=\with) + E(Y \mid X=\without)P(X=\without) \\
&= 100k \cdot 7/10 + 50k \cdot 3/10 = 70k + 15k = 85k.
\end{aligned}
$$

:::

## Unbiasedness Proof
### Step 1

- Rewrite so we're summing over the whole sample.
- To do this, we use subsample indicators $1_x(X_i)$.
- These are random variables that are 
  - 1 if the person is in the subsample 
  - 0 otherwise.
$$
\begin{aligned}
\hat\mu(x) 
&= \frac{1}{N_x} \sum_{i:X_i=x} Y_i               &&\text{ for } \quad N_x = \sum_{i:X_i=x} 1  \\
&= \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i)\ Y_i   &&\text{ for } \quad N_x = \sum_{i=1}^n \ind{x}(X_i) 
\end{aligned}
$$


### Indicators
:::: {.r-stack}

::: {.fragment .fade-out data-fragment-index="0"}
$$
\ind{x}(X_i) = \begin{cases} 1 & \text{ if }  X_i = x \\
                             0 & \text{ otherwise } \end{cases}
$$

:::

::::

## Unbiasedness Proof 
### Step 2

- Rewrite in terms of an expectation that's conditional on our covariates.
- The law of iterated expectations tells us we can.
$$
\begin{aligned}
\E \hat\mu(x) 
&= \E\qty[\frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i)\ Y_i] \quad &&\text{ for } \quad N_x = \sum_{i=1}^n \ind{x}(X_i) \\
&= \E\qty[ \inblue{\E\qty{ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i)\ Y_i  \mid X_1 \ldots X_n}} ] 
\end{aligned}
$$

### The law of iterated expectations
$$
\E Y = \E\qty[ \E\qty{ Y \mid X  } ] 
$$
To calculate the mean of $Y$, we can 
  - first average within subpopulations
  - then across subpopulations.

## Unbiasedness Proof
### Step 3

- Move the conditional expectation inward. It's linear, so we can.
- We can go all the way in to $Y_i$. Everything else is a function of the covariates.
$$
\begin{aligned}
\E \hat\mu(x) 
&= \E\qty[ \inblue{\E \qty{ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i)\ Y_i  \mid X_1 \ldots X_n}}  ]  &&\text{ for } \quad N_x = \sum_{i=1}^n \ind{x}(X_i) \\
&= \E\qty[ \frac{1}{N_x} \inblue{\E\qty{ \sum_{i=1}^n \ind{x}(X_i)\ Y_i \mid X_1 \ldots X_n }} ]  \\
&= \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \inblue{\E\qty{ \ind{x}(X_i)\ Y_i \mid X_1 \ldots X_n }} ]  \\
&= \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i)\ \inblue{\E\qty{ Y_i \mid X_1 \ldots X_n }} ]  
\end{aligned}
$$

:::: columns
::: column
### Linearity of expectations
$$ 
\begin{aligned}
\E\qty( a Y + b Z ) 
&= \E\qty(aY) + \E\qty(bZ) \\
&= a\E(Y) + b\E(Z)  
\end{aligned}
$$

:::

::: column

### Linearity of *conditional* expectations
$$ 
\begin{aligned}
\E\qty{ a(X) Y + b(X) Z \mid X } 
&= \E\qty{a(X)Y \mid X} + \E\qty{ b(X)Z \mid X} \\
&= a(X)\E\qty(Y \mid X) + b(X)\E\qty(Z \mid X)
\end{aligned}
$$

:::

::::

::: notes

  - Linearity tells us two things.
    - To average a sum of two things, we can take two averages and sum.
    - To average a constant times a random variable, we can multiply the random variable's average by the constant.
    - In other words, we can *distribute* expectations and can pull constants out of them.
  - Conditional linearity tells us the same thing, but with a twist.
    - When we condition on $X$, we're working with subpopulations in which $X$ is constant.
    - This means we can we can act as if functions of $X$ *are* constants.
    - So we can pull them out of expectations that are conditional on $X$.
    
:::


## Unbiasedness Proof
### Step 4

- Stop conditioning one person's outcome on *other people's covariates*. 
- They're independent, so we can. And the result is familiar. 
- We get each person's conditional mean outcome $\mu(X_i)$
$$
\begin{aligned}
\E \hat\mu(x) 
&= \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i)\ \inblue{\E\qty{ Y_i \mid X_1 \ldots X_n }} ]  &&\text{ for } \quad N_x = \sum_{i=1}^n \ind{x}(X_i)  \\
&= \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i)\ \inblue{\E\qty{ Y_i \mid X_i}} ] \\ 
&= \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i)\ \inblue{\mu(X_i)} ]  
\end{aligned}
$$

### Irrelevance of independent conditioning variables
$$ \E \qty( Y \mid X, X' ) = \E \qty( Y \mid X ) \quad \text{ when $X'$ is independent of $X,Y$ } $$
If $X'$ is unrelated to $X$ and $Y$, holding it constant doesn't affect the relationship between them.


## Unbiasedness Proof
### Step 5

- It's the *random function value* $\mu(X_i)$.
- We now replace with the fixed function value $\mu(x)$
- We can because it's multiplied by an *indicator* that's nonzero only when $X_i=x$.
$$
\begin{aligned}
\E \hat\mu(x) 
&= \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i)\ \inblue{\mu(X_i)} ]  &&\text{ for } \quad N_x = \sum_{i=1}^n \ind{x}(X_i)  \\ 
&= \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i)\ \inblue{\mu(x)} ]  
\end{aligned}
$$

### The Indicator Trick
$$
\begin{aligned}
\ind{x}(X_i)\ \mu(X_i) & \ \ = \ \ 
\begin{cases} 
\mu(X_i) & \text{ when } X_i=x \\
0 &      \text{ otherwise}
\end{cases} \\
& \ \ = \ \ 
\begin{cases}
\mu(x)  & \ \ \text{ when } X_i=x \\
0       & \ \ \text{ otherwise }
\end{cases} 
\ \ = \ \ \ind{x}(X_i) \ \mu(x)
\end{aligned}
$$



## Unbiasedness Proof
### Step 6

- We conclude by pulling the constant $\mu(x)$ out of the expectation.
- We can because expectations are linear.
- What's left inside is $1/N_x$ times $N_x$, i.e., 1.

$$
\begin{aligned}
\E \hat\mu(x) 
&= \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) \ \inblue{\mu(x)} ]   &&\text{ for } \quad N_x = \sum_{i=1}^n \ind{x}(X_i) \\
&= \inblue{\mu(x)} \ \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) ]  \\
&= \inblue{\mu(x)} \ \E[1] = \inblue{\mu(x)}. 
\end{aligned}
$$

### Linearity of Expectations
$$ 
\begin{aligned}
\E\qty( a Y + b Z ) 
&= \E\qty(aY) + \E\qty(bZ) \\
&= a\E(Y) + b\E(Z)  
\end{aligned}
$$


## Proof
### Summary
$$
{\tiny
\begin{aligned}
\E \hat\mu(x) =& \E\qty{\frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) Y_i}  \\
\underset{iterated}{=}& \E\qty[ \E\qty{ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) Y_i \mid X_1 \ldots X_n} ] \\
\underset{linearity}{=}& \E\qty[ \frac{1}{N_x} \E\qty{ \sum_{i=1}^n \ind{x}(X_i) Y_i \mid X_1 \ldots X_n} ]  \\ 
\underset{linearity}{=}& \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \E\qty{ \ind{x}(X_i) Y_i \mid X_1 \ldots X_n} ]  \\ 
\underset{linearity}{=}& \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) \E\qty{ Y_i \mid X_1 \ldots X_n} ]  \\
\underset{irrelevance}{=}& \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) \E\qty{Y_i \mid X_i} ] \\
\underset{definition}{=}& \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) \mu(X_i) ]  \\
\underset{indicator}{=}& \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) \mu(x) ] \\   
\underset{linearity}{=}& \mu(x) \E\qty[ \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) ]   \\
\underset{definition}{=}& \mu(x) \E\qty[ \frac{1}{N_x} N_{x}    ] = \mu(x)\E[ 1 ] = \mu(x)   
\end{aligned}
}
$$

## Remark: Indicator Notation

- In this proof we used an *indicator* that $X_i$ was equal to $x$.
- This is a special case of a more general concept: *indicators* that $X_i$ is *in a set*.
  - In this special case, our set $\{x\}$ contained one element: $x$.
  - We've used another one, in which our set was an interval $[a \ldots b)$, when we talked about histograms.
- We can think of these as *indicator functions* evaluated at the random variable $X_i$.
$$
\begin{aligned}
1_{A}(X) &= \begin{cases} 1 & \text{ if } \ X \in A \\
                             0 & \text{ otherwise } \end{cases} && \text{general case} \\
\ind{x}(X) &= \begin{cases} 1 & \text{ if }\  X = x \\
                             0 & \text{ otherwise } \end{cases} && \text{one-element case} \\
1_{[a \ldots b)}(X) &= \begin{cases} 1 & \text{ if }\  a \le X < b \\
                             0 & \text{ otherwise } \end{cases} && \text{interval case} 
\end{aligned}
$$



## Remark: Indicator Expansion

- It's useful to generalize the *indicator trick* we've used in our proof.
- We can expand a function of a random variable $X$ as a sum of 
  - the value it takes at every distinct value of $X$
  - times an indicator that $X$ takes on that value
$$
f(X) = f(a) \ \ind{a}(X) + f(b) \ \ind{b}(X) + f(c) \ \ind{c}(X) + \ldots \qfor a\neq b \neq c \ldots.
$$

- In our proof, we used this for the function $\mu$.
- The subsample indicator $1_{x}(X_i)$ *picked out* one term in the expansion.
$$ 
\begin{aligned}
\ind{x}(X_i) \mu(X_i) &= \inblue{\ind{x}(X_i)} \sum_{x'} \inblue{\ind{x'}(X_i)} \ \mu(x') \\
                      &= \sum_{x'} \inblue{\ind{x}(X_i)\ind{x'}(X_i) }\ \mu(x') \\
                      &= \sum_{x'} \mu(x') \inblue{\begin{cases} 1 & \text{ if } X_i=x' \text{ and } X_i=x \\
                                                                 0 & \text{ otherwise } \end{cases}} \\
                      &= \mu(x)  
\end{aligned}
$$


::: notes
It's easy to check this equivalence. 
  - When $X=a$, the expansion's first indicator 'hits' and the others don't. 
  - That means the expanded version is $f(a)$. That's right.
  - It's the same for $b,c,\ldots$.
:::


# Below here is old

##  Properties of Population Means: Conditioning

### The law of iterated expectations
$$
\E\qty{ \E\qty( Y \mid X ) } \quad \text{ for any random variables $X, Y$} \\
$$
To calculate the mean of $Y$, we can average within subpopulations, then across subpopulations.

### Irrelevance of independent conditioning variables
$$ \E \qty( Y \mid X, X' ) = \E \qty( Y \mid X ) \quad \text{ when $X'$ is independent of $X,Y$ } $$
If $X'$ is unrelated to $X$ and $Y$, holding it constant doesn't affect the relationship between them.

## Properties of Population Means: Linearity

### The linearity of expectations
$$ 
\begin{aligned}
\E\qty( a Y + b Z ) 
&= \E\qty(aY) + \E\qty(bZ) \\
&= a\E(Y) + b\E(Z)  \quad \text{ for random variables $Y, Z$ and numbers $a,b$ }
\end{aligned}
$$
  There are two things going on here.

  - To average a sum of two things, we can take two averages and sum.
  - To average a constant times a random variable, we can multiply the random variable's average by the constant.
  - In other words, we can *distribute* expectations and can pull constants out of them.

### The linearity of *conditional* expectations
$$ 
\begin{aligned}
\E\qty{ a(X) Y + b(X) Z \mid X } 
&= \E\qty{a(X)Y \mid X} + \E\qty{ b(X)Z \mid X} \\
&= a(X)\E\qty(Y \mid X) + b(X)\E\qty(Z \mid X)
\end{aligned}
$$
   
This is like linearity of expectations, but with a twist.

   - When we condition on $X$, we're working with subpopulations in which $X$ is constant.
   - This means we can we can act as if functions of $X$ *are* constants.
   - So we can pull them out of expectations that are conditional on $X$.

## Thinking clearly about conditioning

- It's important to distinguish between two things
  1.  The *conditional expectation function* $\mu(x)=\E[Y \mid X=x]$.
    - $\mu$ is a function; evaluated at $x$, it's a number. It's not random.
    - It's the mean of the infinite subpopulation of people with $X=x$.
    - This is the thing we're trying to estimate with $\hat\mu(x)$.
  2.  The conditional expectation $\mu(X)=E[Y \mid X]$.
    - $\mu(X)$ is the mean of a *random* infinite subpopulation of people.
    - It's the conditional expectation function *evaluated at* the random variable $X$.
    - This is the sort of thing that shows up when we use the law of iterated expectations.
- Let's do an easy calculation to try to make this clear. 
- We'll calculate the mean income in our population using the law of iterated expectations. 

## Example: Conditioning
  
  - Let's think about a random person drawn from this population
    - $Y$ is their income 
    - $X$ is an indicator for having a four-year degree. 
  - Suppose the subpopulation means are [100k]{.twocolor-green} for people with degrees 
    and [50k]{.twocolor-red} for people without.
  - And that 7/10 of our population has degrees.
  
  What is $\E(Y \mid X)$?

::: fragment

  - It's a random variable that is *either* 100k or 50k. A *random value* of the function $\mu$.
    - It's [$\mu(\with)=100k$]{.twocolor-green} with probability $7/10$,   when $X=\with$. 
    - And it's [$\mu(\without)=50k$]{.twocolor-red} with probability $3/10$, when $X=\without$.
:::

What is $\E(Y)$?

::: fragment

  - It's the mean of this random variable. A number.
$$
\begin{aligned} 
E\qty{ E\qty( Y \mid X ) } 
&= E(Y \mid X=\with)P(X=\with) + E(Y \mid X=\without)P(X=\without) \\
&= 100k \cdot 7/10 + 50k \cdot 3/10 = 70k + 15k = 85k.
\end{aligned}
$$

:::


Unbiasedness of the sample mean
$$
\begin{aligned}
\text{Prove } \ \E \hat\mu(x) &= \mu(x) \qquad \text{ where } \\
\hat\mu(x) &= \frac{1}{N_x} \sum_{i=1}^n \ind{x}(X_i) Y_i \quad \text{ for } \quad N_x = \sum_{i=1}^n \ind{x}(X_i) \\
\mu(x) &= \E[Y_i \mid X_i=x] && 
\end{aligned}
$$


### Hint: The Indicator Trick

- Using the properties we've reviewed, we can write $\E \hat \mu(x)$ in terms of the random variable $\mu(X_i)$.
  - That's a *random value* of the conditional expectation function $\mu$.
  - We want an expression in terms of the value $\mu(x)$ it takes on when $X_i=x$. 
- In this context, it seems like they should be the same. Everyone in our subsample has $X_i=x$.
- To formalize this, we observe that $\mu(X_i)$ appears multiplied by the indicator $\ind{x}(X_i)$.
- And when that indicator is nonzero, $\mu(X_i)$ and $\mu(x)$ coincide.
$$
\mu(X_i)\ind{x}(X_i) = 
\begin{cases} 
\mu(X_i) & \text{ when } X_i=x \\
0 &      \text{ otherwise}
\end{cases} 
=
\begin{cases}
\mu(x)  & \text{ when } X_i=x \\
0       & \text{ otherwise }
\end{cases} 
= \mu(x)\ind{x}(X_i)
$$





## Variance ...


# Q1

# Q2

# Q3


