---
title: "QTM 220 Final"
format: 
  pdf:
    include-in-header:
      text: |
        \usepackage{physics}
        \usepackage{mathtools}
        \DeclareMathOperator{\argmin}{argmin}
        \DeclareMathOperator{\E}{E}
        \newcommand{\model}{\mathcal{M}}
        \newcommand{\R}{\mathbb{R}}
execute:
  echo: false
  warning: false
---

```{r}
library(purrr)
library(tibble)
library(dplyr)
library(magrittr)
library(ggplot2)
theme_set(theme_bw())
```


\section*{Problem 1 }

Suppose we have a sample of $Y_1 \ldots Y_n$ and binary $X_1 \ldots X_n$ stored in an \textbf{R} dataframe \`df' with $n$ rows.  SAY POPULATION... We run this code.

```{r}
#| echo: false
library(ggplot2)
n <- 500
x <- c(rep(0,n/2),rep(1,n/2))
set.seed(220220)
y <- 2.6*x +  rnorm(n) + runif(n,-1,1)


df <- data.frame(y,x)



q2<- ggplot(data = df, aes(x = x, y = y)) +
  geom_jitter(width=.25) +
  geom_hline(yintercept=mean(df$y[df$x==0]), linetype="dashed")+ 
   geom_hline(yintercept=mean(df$y[df$x==1]), linetype="dashed")+ 
  xlab("x") +
  ylab("y") +
  theme_bw() 

```

```{r}
#| echo: true
B=10000
theta<- rep(NA,B)
for( b in 1:B ){
bootind = sample(1:n,size=n,replace=T)
dfboot = df[bootind,]
theta[b]<- (mean(dfboot$y[dfboot$x==1])-mean(dfboot$y[dfboot$x==0]))
}
```

```{r}
#| out-height: 3.5in
#| fig-ncol: 2
#| fig-subcap:
#|  -  Histogram of the vector Y's elements
#|  -  Jittered Scatterplot of X and Y
ggplot() + geom_histogram(aes(x=y,     y=after_stat(density)), alpha=.3) + 
           geom_vline(aes(xintercept=mean(y)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(y)+c(-1,1)*sd(y)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(y), xmin=mean(y)-1.96*sd(y), xmax=mean(y)+1.96*sd(y), y=.025)) + theme_bw() + xlab('') + ylab('')

q2 
```

```{r}
#| out-height: 3.5in
#| fig-ncol: 2
#| fig-subcap:
#|  -  Histogram of the theta's elements

ggplot() + geom_histogram(aes(x=theta, y=after_stat(density)), alpha=.3) + 
           geom_vline(aes(xintercept=mean(theta)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(theta)+c(-1,1)*sd(theta)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(theta), xmin=mean(theta)-1.96*sd(theta), xmax=mean(theta)+1.96*sd(theta), y=.25)) + theme_bw() + xlab('') + ylab('')

```

\subsection*{Part a}

What does this code calculate? State it in words and mark it with a 'a', and an arrow or bracket if necessary for clarity, on the appropriate plot above.

```{r}
#| echo: true
#| eval: false
sd(theta)
```

\color{red}

The bootstrap sampling distribution's standard deviation for the difference in the mean of y between x=1 and x=0. Estimated standard error. It's the distance between the solid vertical and either dashed vertical line on the theta sampling distribution histogram. \color{black}

\subsection*{Part b}

What does this code calculate? State it in words and mark it with a 'b', and an arrow or bracket if necessary for clarity, on the appropriate plot above.

```{r}
#| echo: true
#| eval: false
(mean(y[x==1])-mean(y[x==0])) + 1.96*(sd(y[x==1])/sqrt(n1) + sd(y[x==0])/sqrt(n0))
(mean(y[x==1])-mean(y[x==0])) - 1.96*(sd(y[x==1])/sqrt(n1) + sd(y[x==0])/sqrt(n0))

```

\color{red}

A 95% confidence interval --- based on normal approximation --- for the difference in means for the population from which this sample was drawn. This is indicated by the horizontal segment (with a dot in the middle) on the sampling distribution histogram (theta). \color{black}

\subsection*{Part c}

How could you calculate roughly the same thing as in part b using 'sd(theta)'

\color{red}

We can use the standard deviation of the bootstrap sampling distribution as our estimate of standard error.

```{=tex}
\begin{verbatim}
(mean(y[x==1])-mean(y[x==0]))  + 1.96 * sd(theta)
(mean(y[x==1])-mean(y[x==0]))  - 1.96 * sd(theta)
\end{verbatim}
```
\color{black}

\subsection*{Part d}

Why does part c not include $n_1$ or $n_0$?

\color{red}

We are using the bootstrap estimated sampling distribution to directly estimate standard error. The estimated sampling distribution already takes the sample size (and the subsample of treated and untreated) into account.

\color{black}
\section*{Problem 2}

Below, I've plotted two samples. Each is drawn with replacement from some population.
On top of them, I've drawn lines fit to them via least squares as follows.
$$ \hat \mu = \argmin_{m \in \model} \sum_{i=1}^n \qty{m(X_i) - Y_i}^2 \qfor \model = \qty{m(x) = b_0 + b_1 x \ : \ x \in \R^2} 
$$

Below these, in some order, I've plotted the sampling distributions of the predictions $\hat\mu(0)$ based on those two models. As usual, I've indicated the mean of each sampling distribution with a solid blue line, one standard error in either direction with a dashed blue line, and two  standard errors in either direction with a dotted blue line. The value of $\mu(0)$, the mean of $y$ among individuals in the population with $x=0$, is indicated by a black line. 

\subsection*{Part A}
True or False. These plots are in correspondence, i.e., the sampling distribution on the left corresponds to the line fit to the sample on the left.

\subsection*{Part B}
Suppose I run the following code to calculate the bootstrap sampling distribution of each estimator...
Suppose that, for each estimator $\hat\mu(0)$ 

\subsection*{Part C}

we've used the bootstrap to give us estimates $\hat\text{se}$ of the standard error of each estimator $\hat\mu()

\subsection*{Part D}
Approximate the coverage of 95\% confidence intervals of the form $\hat\mu(0) \pm 2 \hat\text{se}$ based on these two


 sampling distributions shown below.

The sampling distribution plotted on the left corresponds to the line
fit to the sample 

drawn the sampling distribution of $\hat\mu(0)$ for the line shown on the left is drawn below it and similarly for the line shown on the right.


 Explain why or why not.
The sampling distribution of the least squares fit to the sample on the left is centered at the true value of $\mu(0)$.
Are these plots in correspondence or flipped?  Circle one. Here, by in correspondence, I mean that 
i.e., is the sampling distribution based on each
fit drawn below it, and is the black line drawn below the blue line)? Explain why or why not.

```{r}
ft the one on the left)? Explain why or why not.

```{r}
#| layout-ncol: 2
#| layout-nrow: 2

set.seed(1)
mu = function(x){.2 + x + .5*sinpi(x)}
m = 1000000
X = round(3*runif(m))/3
Y = mu(X) + .5*rnorm(m)
pop = data.frame(X=X, Y=Y)

n=50

samp = pop[sample(1:nrow(pop), n, replace=TRUE), ]
model1 = lm(Y~X, data=samp)
model2 = lm(Y~X, data=samp[samp$X %in% c(0,1), ])
grid = expand.grid(X=seq(0,1,by=.01))
ggplot(samp) + geom_point(aes(x=X, y=Y), color='red', alpha=.3) + 
  geom_line(aes(x=X, y=predict(model1, newdata=data.frame(X=X))), data=grid, color='blue') +
  coord_cartesian(ylim=c(-3,3), xlim=c(0,1))

ggplot(samp[samp$X %in% c(0,1), ]) +
 geom_point(aes(x=X, y=Y), color='red', alpha=.3) + 
  geom_line(aes(x=X, y=predict(model2, newdata=data.frame(X=X))), data=grid, color='blue') +
  coord_cartesian(ylim=c(-3,3), xlim=c(0,1)) 

sampdist = purrr::map(1:1000, function(rep) {
  samp = pop[sample(1:nrow(pop), n, replace=TRUE), ]
  model1 = lm(Y~X, data=samp)
  model2 = lm(Y~X, data=samp[samp$X %in% c(0,1), ])
  data.frame(val=c(coef(model1)[1], coef(model2)[1]), model=c('model1', 'model2'))
}) %>% bind_rows()

ggplot(sampdist[sampdist$model=='model2', ])  +
  geom_histogram(aes(x=val, y=after_stat(density)), bins=50, alpha=.2) +
  geom_vline(aes(xintercept=mu(0)), color='black') +
  geom_vline(aes(xintercept=mean(val)), color='blue') +
  geom_vline(aes(xintercept=mean(val)-sd(val)), color='blue', linetype='dashed') + 
  geom_vline(aes(xintercept=mean(val)+sd(val)), color='blue', linetype='dashed') + 
  geom_vline(aes(xintercept=mean(val)-2*sd(val)), color='blue', linetype='dotted') +
  geom_vline(aes(xintercept=mean(val)+2*sd(val)), color='blue', linetype='dotted') 

ggplot(sampdist[sampdist$model=='model1', ])  +
  geom_histogram(aes(x=val, y=after_stat(density)), bins=50, alpha=.2) +
  geom_vline(aes(xintercept=mu(0)), color='black') +
  geom_vline(aes(xintercept=mean(val)), color='blue') +
  geom_vline(aes(xintercept=mean(val)-sd(val)), color='blue', linetype='dashed') + 
  geom_vline(aes(xintercept=mean(val)+sd(val)), color='blue', linetype='dashed') + 
  geom_vline(aes(xintercept=mean(val)-2*sd(val)), color='blue', linetype='dotted') +
  geom_vline(aes(xintercept=mean(val)+2*sd(val)), color='blue', linetype='dotted') 
```

\newpage


\subsection*{Problem 1}

You have a sample $Y_1,Y_2...,Y_n$ drawn with replacement from a population with mean $\mu$ and standard deviation $\sigma$.
The usual estimate for $\mu$ is the sample mean $\bar{Y} = \frac{1}{n}\sum_{i=1}^n Y_i$. But one of your coworkers has a clever idea to reduce variance---instead of dividing by $n$, they divide by $n+10$. This is their estimator: $\tilde{Y}_{10} = \frac{1}{n+10}\sum_{i=1}^n Y_i$. You try to convince them it's a bad idea by being a bit hyperbolic---you argue that if that's a good idea, then why not divide by $n+100$? They, unfortunately, \emph{love} this idea.  You now have to explain why this is a bad idea.

\subsubsection*{Part A}

Calculate the bias and standard error of the three estimators $\bar{Y}$, $\tilde Y_{2}$, and $\tilde Y_{100} = \frac{1}{n+100}\sum_{i=1}^n Y_i$.  

\subsection*{Part B}

For $\mu=0$ and $\sigma=1$, draw a picture of the sampling distribution of $\bar{Y}$, $\tilde Y_{2}$, and $\tilde Y_{100}$ at 
sample sizes $n=100$ and $n=10,000$. Clearly label (i) $\mu$, (ii) the center of your sampling distribution,
and (iii) the standard error of your sampling distribution. Repeat for $\mu=1$ and $\sigma=1$.

\subsection*{Part C}

Using your drawings from part B, explain why both your coworker's proposal and your hyperbolic proposal to use $\tilde Y_{100}$ to estimate $\mu$ is a bad idea. 

\subsection*{Part D}

This coworker isn't going anywhere. Next time, it might be better to bite your tongue and  save time. If your sample size was $n=10,000$, maybe your coworker's idea to use $\tilde Y_{10}$ really wasn't creating much of a problem. What would you need to know about $\mu$ and $\sigma$ to justify this? The pictures you drew in part B might not really illustrate the potential problem. You may want to draw a new one.


\subsection*{Problem 2}

```{r}
library(ggplot2)
mu = function(d, x){.2 + (1-d)*sinpi(x/2) + d*x}
pi = function(x){ 1 - (.1 + .8*abs(cospi(x))) }

set.seed(1)
n = 1000
X = sample(seq(0,1,by=.333), n, replace=T)
D = rbinom(n, 1, pi(X))
Y = mu(D,X) + .02*rnorm(n)
data = data.frame(X=X, D=D, Y=Y)
summaries = data %>% group_by(D,X) %>% summarize(mu=mean(Y), n = n()) %>%
                     group_by(D) %>% mutate(f=n/sum(n))

scatter = ggplot(data) + 
  geom_point(aes(x=X+.01*(2*D-1), y=Y, color=factor(D)), 
             position=position_jitter(w=.1,h=0), alpha=.2) + 
  scale_x_continuous(breaks=seq(0,1,by=.1)) + xlab('') + ylab('') + 
  geom_col(aes(x=X, y=f, fill=factor(D), color=factor(D)), data=summaries, position='identity', alpha=.2, width=.25) +
  guides(color='none', fill='none') + theme_bw()
scatter 
```

We have a sample of $(D_1, X_1, Y_1) \ldots (D_n,X_n,Y_n)$ stored in an \textbf{R} dataframe \`data' with $n$ rows. I've plotted it above. The $X$ variable is on the horizontal axis, the $Y$ variable is on the vertical axis, and the $D$ variable, which is binary, is indicated by color---green for $D=1$, red for $D=0$. The sample is from a conditionally randomized study, in which $D$ is randomized conditional on $X$ with probability $\pi(x)=P(D=1 \mid X=x)$. The proportion of times that $X$ takes on each value $x$ when $D=0$ and when $D=1$, $\hat f_{0}(x)$ and $\hat f_1(x)$ respectively, are plotted as histograms in matching colors. Keep in mind that the grayish part of the histogram is where the red and green one overlap. $\hat f_0$ and $\hat f_1$ are also given, as numbers, in the table below.

```{r}
f0 = summaries %>% filter(D==0)
f0 = f0[order(f0$X),]
f1 = summaries %>% filter(D==1)
f1 = f1[order(f1$X),]
options(digits=2)
```

$$
\begin{array}{lll}
x & \hat f_0(x) & \hat f_1(x) \\
\hline
0 & `r f0$f[1]` & `r f1$f[1]` \\
\frac13 & `r f0$f[2]` & `r f1$f[2]` \\
\frac23 & `r f0$f[3]` & `r f1$f[3]` \\
1 & `r f0$f[4]` & `r f1$f[4]` \\
\end{array}
$$


\subsubsection*{Part A}

Suppose your coworker runs the following code.

```{r}
#| echo: true
additive.model = lm(Y~D+X, data=data) 
summary(additive.model)
```

They claim that the coefficient on $D$, `r coef(additive.model)[2]`, is a pretty good estimate of the ATT. 
You're not sure, so you plot their fitted model. Here's what you see.

```{r}
muhat.additive = function(d,x) { predict(additive.model, newdata=data.frame(X=x,D=d)) }
grid = expand.grid(X=seq(0,1,by=.01), D=0:1)
scatter + geom_line(aes(x=X, y=muhat.additive(D,X), color=factor(D)), data=grid)
```

Do you think their estimate is biased up or down? Referring to the plot, explain why. 

\subsection*{Part B}

Hearing your concerns, your coworker decides to try a different model. They fit the following model.

```{r}
#| echo: true

quadratic.formula  = Y ~ D*(1+X+I(X^2))
quadratic.model = lm(quadratic.formula, data=data) 
summary(quadratic.model)
```

And they show it does a pretty good job of fitting the data.
```{r}
muhat.quadratic = function(d,x) { predict(quadratic.model, newdata=data.frame(X=x,D=d)) }
grid = expand.grid(X=seq(0,1,by=.01), D=0:1)
scatter + geom_line(aes(x=X, y=muhat.quadratic(D,X), color=factor(D)), data=grid)
```


They're confused. It looks to them like the model is fitting the data. But the coefficient on $D$ in this model, `r coef(quadratic.model)[2]`, doesn't seem to be a good estimate of the ATT. How would you propose to use their model to estimate the ATT? Give a formula for you estimator and then, using the information above, calculate an actual numerical value. I haven't tried to make the numbers nice, so you may want a calculator for this one.

\subsubsection*{Part C}

```{r}
estimate.ATT = function(model, data) {
  X = data$X
  D = data$D
  muhat = function(d,x) { predict(model, newdata=data.frame(X=x,D=d)) }
  mean(muhat(1,X[D==1]) - muhat(0, X[D==1]))
}
```

You've managed to convince your coworker that looking at the coefficient on $D$ is not the right way to estimate the ATT. You've got them estimating it correctly. For the sake of not spoiling Part B, let's say we've got a function 'estimate.ATT(model,data)' that does that.
Your coworker wants use the bootstrap to estimate their estimator's standard error. They run the following code.

```{r}
set.seed(1)
```

```{r}
#| echo: true  
#| cache: true 
B=1000
ATT.boot = rep(NA,B)
for( b in 1:B ){
  bootind = sample(1:nrow(data),size=nrow(data),replace=TRUE)
  data.boot = data[bootind,]
  model.boot = lm(quadratic.formula, data=data.boot) 
  ATT.boot[b] = estimate.ATT(model.boot, data.boot)
}
```

This is their estimate of the standard error.
```{r}
#| echo: true
sd(ATT.boot)/sqrt(n)
```

Is this a good estimate of the standard error of the ATT? Explain why or why not. To justify your answer, you may want to refer to the following plot of the bootstrap distribution they calculated.

```{r}
ggplot() + geom_histogram(aes(x=ATT.boot, y=after_stat(density)), alpha=.3, fill='orange') 
```

\subsubsection*{Part D}
Your boss asks you to come up with your own estimate of the ATT. To keep things simple, you decide to use the additive bivariate model, but fit with inverse probability weights. Here's your code.

```{r}
ATTweights = function(d,x) { d + (1-d)*pi(x)/(1-pi(x)) }
data$ATTweights = ATTweights(data$D, data$X)
additive.model.ipw = lm(Y~D+X, data=data, weights=ATTweights) 
summary(additive.model.ipw)
```

You report as your estimate of the coefficient on $D$, `r coef(additive.model.ipw)[2]`. Your coworker is confused. Reporting this coefficient is exactly what you told them not to do. And you're using a model that doesn't seem to fit the data any better than their did. Explain why it works, intuitively, with reference to the plot on the following page. In it, the dashed lines are the fitted values from your model, and the solid lines are the fitted values from your coworker's additive model from Part A. 

```{r}
#| layout-nrow: 2
muhat.ipw.additive = function(d,x) { predict(additive.model.ipw, newdata=data.frame(X=x,D=d)) }
scatter + 
  geom_line(aes(x=X, y=muhat.additive(D, X), color=factor(D)), data=grid) +
  geom_line(aes(x=X, y=muhat.ipw.additive(D,X), color=factor(D)), data=grid, linetype='dashed')
```


\subsubsection*{Part D}
Your supervisor, seeing three estimates, needs convincing that they should use yours. You plot the bootstrap sampling distributions of three estimators of the ATT: one based on your coworker's unweighted fit of the additive model (orange); another on your coworker's unweighted fit of the quadratic model (magenta); and the last on your inverse probability weighted fit of the additive model (blue). You also plot the confidence intervals you'd report for each of these estimators. Explain what this plot tells you about the actual value of the ATT. 

```{r}
#| cache: true
set.seed(1) 
B=5000
ATT.boot = data.frame(additive=rep(NA,B), additive.ipw=rep(NA,B), quadratic=rep(NA,B))
for( b in 1:B ){
  bootind = sample(1:nrow(data),size=nrow(data),replace=TRUE)
  data.boot = data[bootind,]
  boot.model.additive.ipw = lm(Y~D+X, data=data.boot, weights=ATTweights)
  boot.model.quadratic = lm(quadratic.formula, data=data.boot)
  boot.model.additive = lm(Y~D+X, data=data.boot)
  ATT.boot$additive[b] = estimate.ATT(boot.model.additive, data.boot)
  ATT.boot$additive.ipw[b] = estimate.ATT(boot.model.additive.ipw, data.boot)
  ATT.boot$quadratic[b] = estimate.ATT(boot.model.quadratic, data.boot) 
}
```

```{r}
ATT.point = data.frame(additive.ipw = estimate.ATT(additive.model.ipw, data),
                       additive = estimate.ATT(additive.model, data),
                       quadratic = estimate.ATT(quadratic.model, data))
```

```{r}
ggplot(ATT.boot) + 
  geom_histogram(aes(x=additive, y=after_stat(density)), alpha=.3, fill='orange', bins=100) +
  geom_histogram(aes(x=quadratic, y=after_stat(density)), alpha=.3,  fill='magenta', bins=100) + 
  geom_histogram(aes(x=additive.ipw, y=after_stat(density)), alpha=.3,  fill='blue', bins=100) +
  geom_pointrange(aes(y=20, x=ATT.point$additive, xmin=ATT.point$additive-2*sd(ATT.boot$additive), xmax=ATT.point$additive+2*sd(ATT.boot$additive)), color='orange') + 
  geom_pointrange(aes(y=30, x=ATT.point$quadratic, xmin=ATT.point$quadratic-2*sd(ATT.boot$quadratic), xmax=ATT.point$quadratic+2*sd(ATT.boot$quadratic)), color='magenta') +
  geom_pointrange(aes(y=40, x=ATT.point$additive.ipw, xmin=ATT.point$additive.ipw-2*sd(ATT.boot$additive.ipw), xmax=ATT.point$additive.ipw+2*sd(ATT.boot$additive.ipw)), color='blue') + xlab('') + ylab('')
```


\subsection*{Problem 3}

\subsubsection*{Part A}
```{r}
set.seed(5)

mu = function(d, x){.2 + x + (1-d) }
pi = function(x){ 1 - .01 -.98*abs(cospi(x/2)) }

m = 10000
X = sample(seq(0,1,by=.3333), m, replace=T)
D = rbinom(m, 1, pi(X))
Y = mu(D,X) + .02*rnorm(m)
pop = data.frame(X=X, D=D, Y=Y)

pop.scatter = ggplot(pop) + 
  geom_point(aes(x=X+.01*(2*D-1), y=Y, color=factor(D)), 
             position=position_jitter(w=.1,h=0), alpha=.3, size=1) + 
  scale_x_continuous(breaks=seq(0,1,by=.1)) + xlab('') + ylab('') + 
  geom_histogram(aes(x=X, y=.4*after_stat(density), fill=factor(D), color=factor(D)), 
    position='identity', bins=10, alpha=.2) +
  guides(color='none', fill='none') + theme_bw()

n = 1000
set.seed(7)
data = pop[sample(1:nrow(pop), n, replace=TRUE), ]
summaries = data %>% group_by(D,X) %>% summarize(mu=mean(Y), n = n()) %>%
                     group_by(D) %>% mutate(f=n/sum(n))

grid = expand.grid(X=seq(0,1,by=.3333), D=0:1)
sample.scatter = ggplot(data) + 
  geom_point(aes(x=X+.01*(2*D-1), y=Y, color=factor(D)), 
             position=position_jitter(w=.1,h=0), alpha=.2, size=2) + 
  scale_x_continuous(breaks=seq(0,1,by=.1)) + xlab('') + ylab('') + 
  geom_col(aes(x=X, y=f, fill=factor(D), color=factor(D)), data=summaries, position='identity', alpha=.2, width=.25) +
  geom_line(aes(x=X, y=pi(X)), data=grid, size=1, color='purple') +
  geom_point(aes(x=X, y=pi(X)), data=grid, color='purple') +
  guides(color='none', fill='none') + theme_bw()

#pop.scatter 
sample.scatter
```

Some people don't like inverse probability weighting. What they say is that often, some observations get way more weight --- and therefore influence on the least squares fit --- than others. Suppose we're estimating the ATT. Above, I've plotted a sample in which this will happen. The probability of treatment, $\pi(x)=\E[D_i \mid X_i=x]$, is drawn in purple. 
\begin{enumerate} 
\item Indicate, in the plot above, which observations would ---if sampled---get huge weights.
\item State a formula for the inverse probability weights. They should have the property that, if you were to fit the additive bivariate model and used it to estimate the ATT, it'd be an unbiased estimate. Prove it. The formula you give should be a function of $D$ and $X$ that you can calculate, at every $D$ and $X$, using information in the plot. But you might need a magnifying glass and a ruler to actually do that. 
\item Suppose you instead fit the additive bivariate model via unweighted least squares and used it to estimate the ATT. 
In terms of your estimator's limit $\tilde \mu(d,x)$ and the subpopulation means $\mu(d,x)$, what is the bias of your estimator?
\end{enumerate}


\subsubsection*{Part B}

To argue against inverse probability weighting, a colleague of yours has produced the following plot. In it, we see the sampling distribution of ATT estimators based on the additive bivariate model, fit with (in blue) and without (in orange) inverse probability weights. And, as you can see, the two estimators have the same mean, but the inverse-probability-weighted version has a larger variance. To produce it, they've drawn `r B` samples of size n=`r n` with replacement from some population they've made up. The sample we looked at in Part A was, in fact, one of these samples. Many of your colleagues are convinced by this plot. Are you? Explain why or why not. Hint: Think about your answer to Problem 1 Part B.

```{r}
B = 5000
estimates = data.frame(additive=rep(NA,B), additive.ipw=rep(NA,B)) 
set.seed(7)
for( b in 1:B ){
  data = pop[sample(1:nrow(pop), n, replace=TRUE), ]
  data$ATTweights = data$D + (1-data$D)*pi(data$X)/(1-pi(data$X)) 
  additive.model.ipw = lm(Y~D+X, data=data, weights=ATTweights)
  additive.model     = lm(Y~D+X, data=data)
  estimates$additive[b]     = estimate.ATT(additive.model, data)
  estimates$additive.ipw[b] = estimate.ATT(additive.model.ipw, data)
}
```

```{r}
ggplot(estimates) + 
  geom_histogram(aes(x=additive, y=after_stat(density)), alpha=.3, fill='orange', bins=50) +
  geom_histogram(aes(x=additive.ipw, y=after_stat(density)), alpha=.3, fill='blue', bins=50)
```

\subsection*{Problem 4}

\subsubsection*{Part A}
```{r}
set.seed(7)
data = pop[sample(1:nrow(pop), n, replace=TRUE), ]
lm(Y~D, data=data)
```


One of your coworkers, confused by something you said about inverse probability weighting, has become convinced that they can fit any model with group-specific intercepts and get an unbiased estimate of the ATT. They didn't hear the bit about inverse probability weighting. So they've run the following simple regression.

$$
\hat\mu_{\text{simple}}(d,x) = \hat \beta_0 + \hat \beta_1 d \qfor \hat\beta = \argmin_{b \in \R^2} \sum_{i=1}^n \qty{b_0  + b_1 D_i  - Y_i}^2 
$$

The data they were working with was, fortunately, the fake data from Problem 3 (Part A). No real harm done. But they're confused by their effect estimate of `r coef(lm(Y~D, data=data))[2]`. It's nothing like the estimates they're seeing in the histograms from Problem 3 (Part B). Tell them how far off their estimate of the ATT is --- and in what direction --- from one based on the least squares estimate of the additive model using the same data.

$$
\hat\mu_{\text{add}}(d,x) = \hat\gamma_0 + \hat \gamma_1 d + \hat \gamma_2 d x \qfor \hat\gamma = \argmin_{b \in \R^3} \sum_{i=1}^n \qty{b_0 + b_1 D_i + b_2 X_i - Y_i}^2
$$

Explain how to use the plot from Problem 3 (Part A) to determine the direction of the difference. To determine the size, you'll probably want to use the balance table and regression summary below.


$$
\begin{array}{llll}
\bar{X} & \bar{D} & \frac{\sum_{i}D_i X_i}{\sum_i D_i} & \frac{\sum_{i}(1-D_i) X_i}{\sum_i (1-D_i)} \\
\hline
`r mean(data$X)` & `r mean(data$D)` & `r mean(data$X[data$D==1])` & `r mean(data$X[data$D==0])` \\
\end{array}
$$

```{r}
additive.model = lm(Y~D+X, data=data)
summary(additive.model)
```

\subsubsection*{Part B}
Your coworker got the message about inverse probability weighting. They now use inverse probability weighting when they fit their simple univariate regression model. They're getting the same estimate of the ATT they expected.

```{r}
data$weights = data$D + (1-data$D)*pi(data$X)/(1-pi(data$X))
univariate.model.ipw = lm(Y~D, data=data, weights=weights)
summary(univariate.model.ipw)
```

But something confusing has come up. They know that they're supposed to average the CATE $\hat\mu(1,x) - \hat\mu(0,x)$ over the distribution of $X$ in the treated group to get the ATT and over the distribution of $X$ in the whole sample to get the ATE. But $\hat\mu(1,x)-\hat\mu(0,x)=\hat\beta_1$ is a constant---the average is the same in both cases. 
How can they estimate the ATE by fitting this simple univariate model?  Give them a formula, but also explain the intuition behind it using the plot below of the model they've just fit to the data when they were estimating the ATT. The solid lines show $\hat\mu(d,x)$ for $d=0,1$ in the colors you'd expect.

```{r}
muhat.univariate.ipw = function(d,x) { predict(univariate.model.ipw, newdata=data.frame(X=x,D=d)) }
sample.scatter + geom_line(aes(x=X, y=muhat.univariate.ipw(D,X), color=factor(D)), data=grid) + 
                 geom_point(aes(x=X, y=muhat.univariate.ipw(D,X), color=factor(D)), data=grid)  
```

To simplify this one, I'm going to make this multiple choice. The curve they'll want is
$$
\hat \mu_{\text{univariate}}(d,x) = \hat \beta_0 + \hat \beta_1 d \qfor
\hat \beta = \argmin_{b \in \R^2} \sum_{i=1}^n \text{weights}(D_i,X_i) \ \qty{b_0 + b_1 D_i - Y_i}^2
$$
for one of these three choices of $\text{weights}(d,x)$.
$$
\begin{aligned}
\text{weights}_a(D,X) &= (1-D) + \frac{D \qty{1-\pi(X)}}{\pi(X)} \\
\text{weights}_b(D,X) &= (1-D)\pi(X) + D \{ 1-\pi(X) \} \\
\text{weights}_c(D,X) &= \frac{1-D}{1-\pi(X)} + \frac{D}{\pi(X)} 
\end{aligned}
$$

Which one is it?  Prove that you've chosen the right one by showing that the resulting estimate of the ATE is unbiased.


Here's a long hint for those who missed our review of PS6 on Friday. You can show, by looking at the partial derivatives of  the minimization problem that $\tilde \mu$ solves, that $\E[\text{weights}(D,X) \qty{\tilde\mu(D,X) - \mu(D,X)} m(D,X)] = 0$ for every curve $m(D,X)$ in the model we've just fit. By choosing $m(D,X)=D$ and $m(D,X)=1-D$, both of which are in our model, you can show things about $\text{weights}(d,X)\qty{\tilde\mu(d,X)-\mu(d,X)}$ for $d=0$ and $d=1$ using the indicator trick. That is, using the following equivalences.
$$
\begin{aligned}
f(D,X)D &=\begin{cases} 
  f(1,X)D & \text{if } D=1 \\
  0 & \text{if } D=0  
\end{cases} = f(1,X)D  && \text{ and analogously } \\
f(1-D,X)(1-D) &= f(0,X)(1-D) 
\end{aligned}
$$

For one of the options above, a little additional work will reveal that your estimator is unbiased. In particular, it'll show that $\E[\qty{\tilde\mu(1,X) - \mu(1,X)}]=0$ and $\E[\qty{\tilde\mu(0,X) - \mu(0,X)}]=0$.  You'll need the usual computational tools: the law of iterated expectations and linearity of conditional expectations. Note that, by definition, $\E[D \mid X]=\pi(X)$ and consequently --- via linearity --- $\E[1-D \mid X] = 1-\pi(X)$. 

Write your proof in full, so it can be understood without reading this hint. That said, go ahead and use the property that $\E[\text{weights}(D,X) \qty{\tilde\mu(D,X) - \mu(D,X)} m(D,X)] = 0$ for all $m$ in the model without proving that.

\subsubsection*{Part C}
This coworker, troubled, revisits an estimates of the ATE and the ATT that they'd gotten without inverse probability weighting. 
They'd just compared subsample means and then taken the natural-seeming averages. Nobody had complained about that. Here are their estimates.

$$
\begin{aligned}
\widehat{ATE} &= \frac{1}{n}\sum_{i=1}^n \qty{\hat\mu(1,X_i) - \hat\mu(0,X_i)}  && \qqtext{ and } \\
\widehat{ATT} &= \frac{1}{\sum_{i=1}^n D_i}\sum_{i:D_i=1} \qty{\hat\mu(1,X_i) - \hat\mu(0,X_i)} && \qfor \hat \mu(d,x) = \frac{\sum_{i:D_i=d,X_i=x}Y_i}{\sum_{i:D_i=d,X_i=x} 1}
\end{aligned}
$$

Explain why there isn't a problem with either of them.


\subsubsection*{Part D}
This is a bonus question. Choose either of the weights that weren't your answer to Part B.
Explain what you wind up estimating. You may want to draw something on top of the plot from Part B to illustrate your answer. Extra points if you do \emph{both} of the weights that weren't your answer to Part B.