---
title: "QTM 220 Final Practice Problems"
format: pdf
execute:
  echo: false
  warning: false
editor: visual
---

\subsection*{Problem 1}

Suppose you have a sample $Y_1,Y_2...,Y_n$ drawn with replacement from a population with mean $\mu$ and standard deviation $\sigma$.

Consider the following estimator for $\mu$: $\tilde{Y} = \frac{1}{n+2}\sum_{i=1}^n Y_i$

\subsubsection*{Part A}

What is $E[\tilde{Y}]$?

\subsubsection*{Part B}

How does $E[\tilde{Y}]$ compare to $E[\bar{Y}]$? (where $\bar{Y} = \frac{1}{n}\sum_{i=1}^n Y_i$)

\subsubsection*{Part C}

Showing your work, calculate $Var[\tilde{Y}]$.

\subsubsection*{Part D}

How does $Var[\tilde{Y}]$ compare to $Var[\bar{Y}]$?

\subsection*{Problem 2}

Suppose we have a sample of $Y_1 \ldots Y_n$ and $X_1 \ldots X_n$ stored in an \textbf{R} dataframe \`df' with $n$ rows. We run this code.

```{r}
#| echo: false
library(ggplot2)
p3n <- 200
z <- c(rep(0,p3n/2),rep(1,p3n/2))
set.seed(220220)
x <-   z + rnorm(p3n,mean=1,sd=1) + z*rnorm(p3n,mean=3,sd = 1)
y <- (-0.5)*x + (.8)*x*z + rnorm(p3n)


dfp3 <- data.frame(y,x,z)
df<-dfp3


q2<- ggplot(data = dfp3, aes(x = x, y = y)) +
  geom_point() +
  xlab("x") +
  ylab("y") +
  theme_bw() 
n<-p3n


modone<- lm(y~x,data=dfp3)

```

```{r}
#| echo: true
B=10000
beta_0<- rep(NA,B)
beta_1<- rep(NA,B)
for( b in 1:B ){
bootind = sample(1:n,size=n,replace=T)
dfboot = df[bootind,]
mod<- lm(y~x, data=dfboot)
beta_0[b] = mod$coef[1]
beta_1[b] = mod$coef[2]
}
```

```{r}
#| out-height: 3.5in
#| fig-ncol: 2
#| fig-subcap:
#|  -  Histogram of the vector Y's elements
#|  -  Scatterplot of X and Y
ggplot() + geom_histogram(aes(x=y,     y=after_stat(density)), alpha=.3) + 
           geom_vline(aes(xintercept=mean(y)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(y)+c(-1,1)*sd(y)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(y), xmin=mean(y)-1.96*sd(y), xmax=mean(y)+1.96*sd(y), y=.025)) + theme_bw() + xlab('') + ylab('')

q2 + geom_abline(slope=modone$coef[2], intercept=modone$coef[1])
```

```{r}
#| out-height: 3.5in
#| fig-ncol: 2
#| fig-subcap:
#|  -  Histogram of the beta_0's elements
#|  -  Histogram of the beta_1's elements
ggplot() + geom_histogram(aes(x=beta_0, y=after_stat(density)), alpha=.3) + 
           geom_vline(aes(xintercept=mean(beta_0)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(beta_0)+c(-1,1)*sd(beta_0)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(beta_0), xmin=mean(beta_0)-1.96*sd(beta_0), xmax=mean(beta_0)+1.96*sd(beta_0), y=.25)) + theme_bw() + xlab('') + ylab('')

ggplot() + geom_histogram(aes(x=beta_1, y=after_stat(density)), alpha=.3) + 
           geom_vline(aes(xintercept=mean(beta_1)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(beta_1)+c(-1,1)*sd(beta_1)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(beta_1), xmin=mean(beta_1)-1.96*sd(beta_1), xmax=mean(beta_1)+1.96*sd(beta_1), y=1)) + theme_bw() + xlab('') + ylab('')

```

\subsection*{Part a}

What does this code calculate? State it in words and mark it with an 'a', and an arrow or bracket if necessary for clarity, on the appropriate plot above.

```{r}
#| echo: true
#| eval: false
mean(Y)
```

\subsection*{Part b}

What does this code calculate? State it in words and mark it with a 'b', and an arrow or bracket if necessary for clarity, on the appropriate plot above.

```{r}
#| echo: true
#| eval: false
sd(Y)
```

\subsection*{Part c}

What does this code calculate? State it in words and mark it with a 'c', and an arrow or bracket if necessary for clarity, on the appropriate plot above.

```{r}
#| echo: true
#| eval: false
mod<- lm(y~x, data=df)
mod$coef[1]
```

\subsection*{Part d}

What does this code calculate? State it in words and mark it with a 'd', and an arrow or bracket if necessary for clarity, on the appropriate plot above.

```{r}
#| echo: true
#| eval: false
sd(beta_1)
```

\subsection*{Part e}

What does this code calculate? State it in words and mark it with a 'e', and an arrow or bracket if necessary for clarity, on the appropriate plot above.

```{r}
#| echo: true
#| eval: false
mod$coef[2] + sqrt(diag(solve(t(X)%*%X)%*%t(X)%*%diag(resid^2)%*%X%*%solve(t(X)%*%X)))[2]

mod$coef[2] - sqrt(diag(solve(t(X)%*%X)%*%t(X)%*%diag(resid^2)%*%X%*%solve(t(X)%*%X)))[2]
```

\subsection*{Part f}

How could you calculate roughly the same thing as in part e using 'sd(beta_1)'
