---
title: "Final Exam"
subtitle: "QTM 220"
---


::: hidden

$$
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\CATE}{CATE}
\DeclareMathOperator{\Var}{V}
\DeclareMathOperator{\hVar}{\widehat{V}}
\DeclareMathOperator{\bias}{bias}
\DeclareMathOperator*{\argmin}{argmin}
\newcommand{\model}{\mathcal{M}}
$$

:::

```{r}
library(tidyverse)
theme_set(theme_minimal())
```

# Problem 1

```{r}
#| echo: false
library(ggplot2)
n <- 500
X <- c(rep(0,n/2),rep(1,n/2))
set.seed(220220)
Y <- 2.6*X +  rnorm(n) + runif(n,-1,1)
N1 = sum(X==1)
N0 = sum(X==0)
```

Suppose we have a sample of $(X_1,Y_1) \ldots (X_n,Y_n)$ drawn with replacement from a population
$(x_1, y_1) \ldots (x_m, y_m)$. The variable $X$ is binary. We have our sample stored in two **R** vectors $X$ and $Y$ of length $n$. We run this code.

```{r}
#| echo: true
B=10000
theta.star = rep(NA,B)
for( b in 1:B ){
  J.star = sample(1:n,size=n,replace=TRUE)
  X.star = X[J.star]
  Y.star = Y[J.star]
  theta.star[b] = mean(Y.star[X.star==1]) - mean(Y.star[X.star==0])
}
```

```{r}
#| out-height: 3.5in
#| fig-ncol: 2
#| fig-cap:
#|  -  Histogram of the vector Y's elements
#|  -  Jittered Scatterplot of X and Y
ggplot(data.frame(X=X,Y=Y)) + 
    geom_histogram(aes(x=Y, y=after_stat(density)), alpha=.3, bins=30) + 
    geom_vline(aes(xintercept=mean(Y)), alpha=.5) + 
    geom_vline(aes(xintercept=mean(Y)-sd(Y)), linetype='dashed', alpha=.5) + 
    geom_vline(aes(xintercept=mean(Y)+sd(Y)), linetype='dashed', alpha=.5) + 
    geom_pointrange(aes(x=mean(Y), xmin=mean(Y)-1.96*sd(Y), xmax=mean(Y)+1.96*sd(Y), y=.025)) + 
    xlab('') + ylab('')

mean_sd = function(y){ data.frame(y=mean(y), ymin=mean(y)-sd(y), ymax = mean(y) + sd(y)) }
ggplot(data.frame(X=X,Y=Y)) + 
    geom_point(aes(x=X, y=Y), position = position_jitter(w=.2), alpha=.2) + 
    stat_summary(aes(x=X, y=Y), fun.data=mean_sd, geom='pointrange') + 
    scale_x_continuous(breaks=c(0,1)) 
```

```{r}
#| out-height: 3.5in
#| fig-ncol: 2
#| fig-cap:
#|  -  Histogram of theta.star's elements

ggplot() + geom_histogram(aes(x=theta.star, y=after_stat(density)), alpha=.3, bins=50) + 
           geom_vline(aes(xintercept=mean(theta.star)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(theta.star)+c(-1,1)*sd(theta.star)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(theta.star), 
                               xmin=mean(theta.star)-1.96*sd(theta.star), 
                               xmax=mean(theta.star)+1.96*sd(theta.star), y=.25)) 

```

::: {.callout-exercise}
## Part A
What does this code calculate?  State your answer in words in terms of the sample and/or population, not the variables in the code.
And mark it with a 'a', and an arrow or bracket if necessary for clarity, on the appropriate plot above.

```{r}
#| echo: true
#| eval: false
sd(theta)
```

:::
::: {.callout-solution}

The bootstrap sampling distribution's standard deviation for the difference in the mean of y between x=1 and x=0. Estimated standard error. It's the distance between the solid vertical and either dashed vertical line on the theta sampling distribution histogram. 
:::


::: {style="height:2in"}
:::

::: {.callout-exercise}
### Part B

What does this code calculate? State your answer in words in terms of the sample and/or population, not the variables in the code. And mark it with a 'b', and an arrow or bracket if necessary for clarity, on the appropriate plot above.

```{r}
#| echo: true
#| eval: true
(mean(Y[X==1])-mean(Y[X==0])) + 1.96 * (sd(Y[X==1])/sqrt(N1) + sd(Y[X==0])/sqrt(N0))
(mean(Y[X==1])-mean(Y[X==0])) - 1.96 * (sd(Y[X==1])/sqrt(N1) + sd(Y[X==0])/sqrt(N0))
```

:::

::: {style="height:2in"}
:::

::: {.callout-solution}
A 95\% confidence interval --- based on normal approximation --- for the difference in means for the population from which this sample was drawn. This is indicated by the horizontal segment (with a dot in the middle) on the sampling distribution histogram (theta). 
:::

::: {.callout-exercise}
### Part C
Which of the following would calculate roughly the same thing as in Part B? Explain why.

```{r}
#| echo: true
#| eval: false
mean(Y[X==1])-mean(Y[X==0]) + 1.96 * sd(theta.star)/sqrt(B)
mean(Y[X==1])-mean(Y[X==0]) - 1.96 * sd(theta.star)/sqrt(B)
```

```{r}
#| echo: true
#| eval: false
mean(Y[X==1])-mean(Y[X==0]) + 1.96 * sd(theta.star)/sqrt(n)
mean(Y[X==1])-mean(Y[X==0]) - 1.96 * sd(theta.star)/sqrt(n)
```

```{r}
#| echo: true
#| eval: false
mean(Y[X==1])-mean(Y[X==0]) + 1.96 * sd(theta.star)
mean(Y[X==1])-mean(Y[X==0]) - 1.96 * sd(theta.star)
```

:::

::: {style="height:2in"}
:::

::: {.callout-solution}
The third one. The bootstrap sampling distribution is an approximation of the sampling distribution of our
point estimate `mean(Y[X==1])-mean(Y[X==0])`, so we can use the bootstrap standard deviation (i.e. the spread of the bootstrap sampling distribution)
as an estimator the standard error of our point estimate (i.e. the spread of our point estimate's actual sampling distribution).
:::

# Problem 2

Below, I've plotted samples drawn, with replacement, from two populations.
On top of them, I've drawn lines fit to them via least squares as follows.
$$ 
\hat \mu = \argmin_{m \in \model} \sum_{i=1}^n \qty{Y_i - m(X_i)}^2 \qfor \model = \qty{m(x) = a + b x  \ \text{ for numbers $a$, $b$} }
$$

Below these, in some order, I've plotted the sampling distributions of the predictions $\hat\mu(0)$ I get when I fit this model to each sample.
As usual, I've indicated the mean of each sampling distribution with a solid blue line, one standard error in either direction with a dashed blue line, and two  standard errors in either direction with a dotted blue line. The value of $\mu(0)$, the mean of $y$ among individuals in the population with $x=0$, is indicated by a green line. 


```{r}
#| layout-ncol: 2
#| layout-nrow: 2

set.seed(1)
mu = function(x){.2 + .7*x + .8*sinpi(x)}
m = 1000000
X = round(3*runif(m))/3
Y = mu(X) + .5*rnorm(m)
pop = data.frame(X=X, Y=Y)

n=30

samp = pop[sample(1:nrow(pop), n, replace=TRUE), ]
model1 = lm(Y~X, data=samp)
model2 = lm(Y~X, data=samp[samp$X %in% c(0,1), ])
grid = expand.grid(X=seq(0,1,by=.01))
ggplot(samp) + geom_point(aes(x=X, y=Y), color='red', alpha=.3) + 
  geom_line(aes(x=X, y=predict(model1, newdata=data.frame(X=X))), data=grid, color='blue') +
  scale_x_continuous(breaks=round(seq(0,1,by=1/3),2)) +
  coord_cartesian(ylim=c(-3,3), xlim=c(-.1,1.1))

ggplot(samp[samp$X %in% c(0,1), ]) +
 geom_point(aes(x=X, y=Y), color='red', alpha=.3) + 
  geom_line(aes(x=X, y=predict(model2, newdata=data.frame(X=X))), data=grid, color='blue') +
  scale_x_continuous(breaks=round(seq(0,1))) +
  coord_cartesian(ylim=c(-3,3), xlim=c(-.1,1.1)) 

sampdist = purrr::map(1:1000, function(rep) {
  samp = pop[sample(1:nrow(pop), n, replace=TRUE), ]
  model1 = lm(Y~X, data=samp)
  model2 = lm(Y~X, data=samp[samp$X %in% c(0,1), ])
  data.frame(val=c(coef(model1)[1], coef(model2)[1]), model=c('model1', 'model2'))
}) |> bind_rows()

ggplot(sampdist[sampdist$model=='model2', ])  +
  geom_histogram(aes(x=val, y=after_stat(density)), bins=50, alpha=.2) +
  geom_vline(aes(xintercept=mu(0)), color='green', linewidth=1.5) +
  geom_vline(aes(xintercept=mean(val)), color='blue') +
  geom_vline(aes(xintercept=mean(val)-sd(val)), color='blue', linetype='dashed') + 
  geom_vline(aes(xintercept=mean(val)+sd(val)), color='blue', linetype='dashed') + 
  geom_vline(aes(xintercept=mean(val)-2*sd(val)), color='blue', linetype='dotted') +
  geom_vline(aes(xintercept=mean(val)+2*sd(val)), color='blue', linetype='dotted') 

ggplot(sampdist[sampdist$model=='model1', ])  +
  geom_histogram(aes(x=val, y=after_stat(density)), bins=50, alpha=.2) +
  geom_vline(aes(xintercept=mu(0)), color='green', linewidth=1.5) +
  geom_vline(aes(xintercept=mean(val)), color='blue') +
  geom_vline(aes(xintercept=mean(val)-sd(val)), color='blue', linetype='dashed') + 
  geom_vline(aes(xintercept=mean(val)+sd(val)), color='blue', linetype='dashed') + 
  geom_vline(aes(xintercept=mean(val)-2*sd(val)), color='blue', linetype='dotted') +
  geom_vline(aes(xintercept=mean(val)+2*sd(val)), color='blue', linetype='dotted') 
```

::: {.callout-exercise}
True or False. These plots are in correspondence, i.e., the sampling distribution on the left corresponds to the line fit to the sample on the left.
:::

::: {.callout-exercise}
Suppose I use the bootstrap to get a good estimate $\widehat{\text{se}}$ of the standard error of each estimator $\hat{\mu}(0)$.
For each of these estimators, approximate the coverage of the 95\% confidence interval $\hat\mu(0) \pm 2 \widehat{\text{se}}$.
Is it closer to 95\%, 50\%, or 0\%?  So that the correctness of this one isn't tied to the previous one, please report this for...

- the estimator whose sampling distribution is on the left.
- the estimator whose sampling distribution is on the right.

:::

::: {style="height:2in"}
:::


# Problem 3

In this problem, we'll work with a sample $(W_1, X_1, Y_1) \ldots (W_n,X_n,Y_n)$ from some population $(w_1, x_1, y_1) \ldots (w_m, x_m, y_m)$. 
We'll plot $X_i$ on the horizontal axis, $Y_i$ is on the vertical axis, and $W_i$ variable, which is binary, is indicated by color---green for $W_i=1$, red for $W_i=0$. 
The sample is from a conditionally randomized study: for all members of the population with the same value of $x_j$, $w_j$ is chosen by flipping the same weighted coin,
so there is some function $\pi(x)$ for which it comes up heads for people with $x_j=x$ with probability $\pi(x)=P(W_i=1 \mid X_i=x)$.

We'll be estimating the the Average Treatment Effect on the Treated (ATT).
$$
\begin{aligned}
\text{ATT} &= \frac{1}{m_1} \sum_{j:w_j=1} \qty{\mu(1,x_j) - \mu(0,x_j)} \qfor m_w = \sum_{j:w_j=w} 1 \qand \mu(w,x) = \frac{\sum_{j:w_j=w \ \text{ and } x_j = x} y_j}{\sum_{j:w_j=w \ \text{ and } x_j = x} 1}
\end{aligned}
$$

As usual, we'll do this by using least squares regression to get an estimate $\hat\mu$ of $\mu$, then plugging it into a version of the formula above
where we replace population averages with sample averages. That is, we'll use this formula in terms of $\hat\mu$.
$$
\begin{aligned}
\widehat{ATT} &= \frac{1}{N_1} \sum_{i:W_i=1} \qty{\hat\mu(1,X_i) - \hat\mu(0,X_i)} \qfor N_w = \sum_{i:W_i=w} 1.  
\end{aligned}
$$


## A Small Study
We start by working with the small sample plotted below. I've fit the *two lines* model to the data using unweighted least squares, i.e., $\hat \mu$ is chosen as follows.
$$
\begin{aligned}
\hat \mu &= \argmin_{m \in \model} \sum_{i=1}^n \qty{Y_i - m(W_i,X_i)}^2  \\
& \qfor \model = \qty{m(w,x) = a(w) + b(w)x  \ \text{ for any functions $a$, $b$ } }
\end{aligned}
$$

In the plot, the solid lines show $\hat\mu$ with $\hat\mu(1,x)$ and $\hat\mu(0,x)$ in green and red respectively as usual.

```{r}
X <- c(0,2,4,0,0,2,4,4)
Y <- c(2,4,6,12.25,8,5.25,3.25,1)
D <- c(1,1,1,0,0,0,0,0)

data <- data.frame(X=X,Y=Y,D=D)
model = lm(Y~D*X, data=data)
grid = expand.grid(X=seq(0,4,by=.01), D=0:1)
grid$muhat = predict(model, newdata=grid)
plot<- ggplot(data = data, aes(x = X, y = Y)) + 
  geom_point(aes(color=factor(D)),size=2) + 
  scale_y_continuous(breaks = seq(0, 13, by = 1)) + 
  geom_line(aes(x=X, y=muhat, color=factor(D)), data=grid) 
    
plot + guides(color='none')
```


::: {.callout-exercise}
### Part A
What is the ATT estimate $\widehat{ATT}$ based on this model and data? Show your calculations.
:::

::: {style="height:5in"}
:::

::: {.callout-solution}
$\frac{1}{3}[(2-10)+(4-6)+(6-2)] = -2$ 
:::

::: {.callout-exercise}
### Part B
Calculate the *difference in means* of the treated and control groups, i.e.,
$$
\hat\Delta = \frac{1}{N_1} \sum_{i:W_i=1} Y_i - \frac{1}{N_0} \sum_{i:W_i=0} Y_i \quad \text{ for } \quad N_w = \sum_{i:W_i=w} 1.
$$

Compare this to the ATT. Is it bigger or smaller? Explain why.
:::

::: {.callout-tip}
Does it matter that we're averaging $Y_i$ instead of $\hat\mu(W_i,X_i)$ when we calculate $\hat\Delta$?
No need to write an answer to this question. It's just a prompt to help you think about Part B.
:::

:::{style="height:5in"}
:::


## A Bigger Version
In this question, we'll repeat Exercise A using a larger sample then do a bit more.
The set-up is exactly the same as before. We have a sample of $(W_1, X_1, Y_1) \ldots (W_n,X_n,Y_n)$ drawn with replacement from some population $(w_1, x_1, y_1) \ldots (w_m, x_m, y_m)$. 
I've plotted it below. The $X$ variable is on the horizontal axis, the $Y$ variable is on the vertical axis, and the $W$ variable, which is binary, is indicated by color---green for $W=1$, red for $W=0$. I've used some jitter when plotting to make things easier to see---$X$ really takes on the four values $0$,$1/3$,$2/3$,and $1$.

The sample is from a conditionally randomized study; for all members of the population with the same value of $x_j$, $w_j$ is chosen by flipping the same weighted coin,
so there is some function $\pi(x)$ for which it comes up heads for people with $x_j=x$ with probability $\pi(x)=P(W_i=1 \mid X_i=x)$. The population we're using here is
not necessarily the same as the one from which we drew the small sample in the previous question, so don't expect to see exactly the same relationships in the data.

As before, I've fit the *two lines*  model to the data using least squares. The solid lines show the fitted model $\hat\mu$ with $\hat\mu(1,x)$ and $\hat\mu(0,x)$ in green and red respectively.
I've also plotted $\bar Y_{w,x}$, the average of $Y$ among individuals with $W=w$ and $X=x$, as well as arms spanning $\pm 2$ standard errors, for all pairs of $w$ and $x$.

```{r scatter}
mu = function(w, x){.5 + (1-w)*sinpi(x/2) + w*x}
pi = function(x){ .5 + .4*cospi(2*x) }

set.seed(1)
n = 400
X = sample(seq(0,1,by=.333), n, replace=T)
W = rbinom(n, 1, pi(X))
Y = mu(W,X) + .1*rnorm(n)
data = data.frame(X=X, W=W, Y=Y)
summaries = data %>% group_by(W,X) %>% summarize(mu=mean(Y), n = n(), se=sd(Y)/sqrt(n)) %>%
                     group_by(W) %>% mutate(f=n/sum(n))

model = lm(Y~W*X, data=data)
summaries$muhat = predict(model, newdata=summaries)

summaries0 = summaries %>% filter(W==0)
summaries1 = summaries %>% filter(W==1)
summaries0 = summaries0[order(summaries0$W),]
summaries1 = summaries1[order(summaries1$W),]
f = mean(W) * summaries0$f + (1-mean(W)) * summaries1$f

scatter = ggplot(data) + 
  geom_point(aes(x=X+.025*(2*W-1), y=Y, color=factor(W)), 
             position=position_jitter(w=.025,h=0), alpha=.2, size=1) + 
  geom_line(aes(x=X, y=muhat, color=factor(W)), data=summaries, size=.5) +
  scale_x_continuous(breaks=round(seq(0,1,by=1/3),2)) + xlab('') + ylab('') + 
  geom_col(aes(x=X+.025*(2*W-1), y=f, fill=factor(W), color=factor(W)), data=summaries, position='identity', alpha=.2, width=.1) +
  geom_col(aes(x=X, y=f), fill='purple', color='purple', 
    data=data.frame(f=f, X=summaries0$X), position='identity', alpha=.2, width=.05) +
  geom_pointrange(aes(x=X+.025*(2*W-1), y=mu, ymin=mu-2*se, ymax=mu+2*se, color=factor(W)), 
                  data=summaries, alpha=.8, size=.2) +
  guides(color='none', fill='none') + theme_bw()
scatter 
```

This time, you probably won't want to count dots. Instead, you'll work with summaries of the data's distribution.
The proportion of times that $X$ takes on each value $x$ when $W=0$ and when $W=1$, $\hat p(0,x)$ and $\hat p(1,x)$ respectively, are plotted as histograms in matching colors. In purple, I've plotted $\hat p(x)$, the proportion of times $X$ takes on each value $x$ among all observations---I've made that one a bit thinner so it's easier to see the other two.

```{r}
options(digits=2)
```

$$
\begin{aligned}
\hat p(0,x)  &= \frac{1}{N_0}\sum_{i:W_i=0 \text{ and } X_i=x} 1  \quad \text{for} \quad N_0 = \sum_{i:W_i=0} 1 \\
\hat p(1, x) &= \frac{1}{N_1}\sum_{i:W_i=1 \text{ and } X_i=x} 1  \quad \text{for} \quad N_1 = \sum_{i:W_i=1} 1 \\
\hat p(x) &= \frac{1}{n}\sum_{i:X_i=x} 1 \\
\end{aligned}
$$

Here's a table of the summaries you see in the plot.

$$
\begin{array}{llllllll}
x & \hat p(0,x) & \hat p(1,x) & \hat p(x) & \hat\mu(0,x) & \hat \mu(1,x) & \bar Y_{0,x} & \bar Y_{1,x} \\
\hline
0 & `r summaries0$f[1]` & `r summaries1$f[1]` & `r f[1]` & `r summaries0$muhat[1]` & `r summaries1$muhat[1]` & `r summaries0$mu[1]` & `r summaries1$mu[1]` \\
\frac13 & `r summaries0$f[2]` & `r summaries1$f[2]` & `r f[2]` & `r summaries0$muhat[2]` & `r summaries1$muhat[2]` & `r summaries0$mu[2]` & `r summaries1$mu[2]` \\
\frac23 & `r summaries0$f[3]` & `r summaries1$f[3]` & `r f[3]` & `r summaries0$muhat[3]` & `r summaries1$muhat[3]` & `r summaries0$mu[3]` & `r summaries1$mu[3]` \\
1 & `r summaries0$f[4]` & `r summaries1$f[4]` & `r f[4]` & `r summaries0$muhat[4]` & `r summaries1$muhat[4]` & `r summaries0$mu[4]` & `r summaries1$mu[4]` \\
\end{array}
$$


::: {.callout-exercise}
### Part C
What is your ATT estimate $\widehat{ATT}$ based on this model and data? Show your calculations.
:::
::: {style="height:5in"}
:::

::: {.callout-exercise}
### Part D
Is this estimate's bias positive, negative, or roughly zero? With reference to the plot above explain your answer and explain why this is happening. 
:::
::: {style="height:5in"}
:::

::: {.callout-exercise}
### Part E
In terms of quantities available in the plot and table, write an *unbiased* estimate of the ATT. 
Write this first as a formula involving quantities like $\hat p(0,x)$, $\hat p(1,x)$, etc. 
Then calculate its value using the numbers in the table. 
:::
::: {style="height:5in"}
:::

## Extra Credit

::: {.callout-exercise}
### Extra Credit

Repeat Exercises A-E for the Average Treatment Effect on the whole population (ATE) instead of the ATT.
$$
ATE = \frac{1}{m} \sum_{j=1}^m \qty{ \mu(1,x_j) - \mu(0,x_j) }.
$$

:::
