---
title: "QTM 220 Final"
format: 
  pdf:
    include-in-header:
      text: |
        \usepackage{physics}
        \usepackage{mathtools}
        \DeclareMathOperator{\argmin}{argmin}
        \DeclareMathOperator{\E}{E}
        \newcommand{\model}{\mathcal{M}}
        \newcommand{\R}{\mathbb{R}}
        \newenvironment{answer}{\ \\ \color{white}}{\color{black}}
        % replace 'red' with 'white' above  to hide solutions (in printout)
execute:
  echo: false
  warning: false
editor: visual
---

```{r}
library(purrr)
library(tibble)
library(dplyr)
library(magrittr)
library(ggplot2)
theme_set(theme_bw())
```


\section*{Problem 1}
Suppose we have a sample of $(X_1,Y_1) \ldots (X_n,Y_n)$ drawn with replacement from a population. $X$ is binary. It's stored in an \textbf{R} dataframe `df' with $n$ rows. We run this code.

```{r}
#| echo: false
library(ggplot2)
n <- 500
x <- c(rep(0,n/2),rep(1,n/2))
set.seed(220220)
y <- 2.6*x +  rnorm(n) + runif(n,-1,1)


df <- data.frame(y,x)

q2<- ggplot(data = df) +
  geom_jitter(aes(x=x,y=y), width=.25, alpha=.2) +
  geom_pointrange(aes(x=x, y=mean(y), ymin=mean(y)-2*sd(y), ymax=mean(y)+2*sd(y)), data=df[df$x==0,]) + 
  geom_pointrange(aes(x=x, y=mean(y), ymin=mean(y)-2*sd(y), ymax=mean(y)+2*sd(y)), data=df[df$x==1,]) + 
  xlab("x") +
  ylab("y") +
  theme_bw() 
```

```{r}
#| echo: true
B=10000
theta = rep(NA,B)
for( b in 1:B ){
  bootind = sample(1:n,size=n,replace=TRUE)
  dfboot = df[bootind,]
  theta[b] = mean(dfboot$y[dfboot$x==1]) - mean(dfboot$y[dfboot$x==0])
}
```

```{r}
#| out-height: 3.5in
#| fig-ncol: 2
#| fig-subcap:
#|  -  Histogram of the vector Y's elements
#|  -  Jittered Scatterplot of X and Y
ggplot() + geom_histogram(aes(x=y,     y=after_stat(density)), alpha=.3) + 
           geom_vline(aes(xintercept=mean(y)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(y)+c(-1,1)*sd(y)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(y), xmin=mean(y)-1.96*sd(y), xmax=mean(y)+1.96*sd(y), y=.025)) + theme_bw() + xlab('') + ylab('')

q2 
```

```{r}
#| out-height: 3.5in
#| fig-ncol: 2
#| fig-subcap:
#|  -  Histogram of the theta's elements

ggplot() + geom_histogram(aes(x=theta, y=after_stat(density)), alpha=.3) + 
           geom_vline(aes(xintercept=mean(theta)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(theta)+c(-1,1)*sd(theta)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(theta), xmin=mean(theta)-1.96*sd(theta), xmax=mean(theta)+1.96*sd(theta), y=.25)) + theme_bw() + xlab('') + ylab('')

```

\subsection*{Part a}

What does this code calculate? State it in words and mark it with a 'a', and an arrow or bracket if necessary for clarity, on the appropriate plot above.

```{r}
#| echo: true
#| eval: false
sd(theta)
```

\begin{answer}

The bootstrap sampling distribution's standard deviation for the difference in the mean of y between x=1 and x=0. Estimated standard error. It's the distance between the solid vertical and either dashed vertical line on the theta sampling distribution histogram. 
\end{answer}

\subsection*{Part b}

What does this code calculate? State it in words and mark it with a 'b', and an arrow or bracket if necessary for clarity, on the appropriate plot above.

```{r}
#| echo: true
#| eval: false
(mean(y[x==1])-mean(y[x==0])) + 1.96*(sd(y[x==1])/sqrt(n1) + sd(y[x==0])/sqrt(n0))
(mean(y[x==1])-mean(y[x==0])) - 1.96*(sd(y[x==1])/sqrt(n1) + sd(y[x==0])/sqrt(n0))

```

\begin{answer}
A 95\% confidence interval --- based on normal approximation --- for the difference in means for the population from which this sample was drawn. This is indicated by the horizontal segment (with a dot in the middle) on the sampling distribution histogram (theta). 
\end{answer}

\subsection*{Part c}

How could you calculate roughly the same thing as in part b using 'sd(theta)'

\begin{answer}
We can use the standard deviation of the bootstrap sampling distribution as our estimate of standard error.

\begin{verbatim}
(mean(y[x==1])-mean(y[x==0]))  + 1.96 * sd(theta)
(mean(y[x==1])-mean(y[x==0]))  - 1.96 * sd(theta)
\end{verbatim}

\end{answer}

\subsection*{Part d}

Why does part c not include $n_1$ or $n_0$?

\begin{answer}
We are using the bootstrap estimated sampling distribution to directly estimate standard error. The estimated sampling distribution already takes the sample size (and the subsample of treated and untreated) into account.

\end{answer}

\newpage


\section*{Problem 2}

Below, I've plotted two samples. Each is drawn with replacement from some population.
On top of them, I've drawn lines fit to them via least squares as follows.
$$ \hat \mu = \argmin_{m \in \model} \sum_{i=1}^n \qty{m(X_i) - Y_i}^2 \qfor \model = \qty{m(x) = b_0 + b_1 x \ : \ x \in \R^2} 
$$

Below these, in some order, I've plotted the sampling distributions of the predictions $\hat\mu(0)$ based on those two models. As usual, I've indicated the mean of each sampling distribution with a solid blue line, one standard error in either direction with a dashed blue line, and two  standard errors in either direction with a dotted blue line. The value of $\mu(0)$, the mean of $y$ among individuals in the population with $x=0$, is indicated by a black line. 



```{r}
#| layout-ncol: 2
#| layout-nrow: 2

set.seed(1)
mu = function(x){.2 + x + .5*sinpi(x)}
m = 1000000
X = round(3*runif(m))/3
Y = mu(X) + .5*rnorm(m)
pop = data.frame(X=X, Y=Y)

n=50

samp = pop[sample(1:nrow(pop), n, replace=TRUE), ]
model1 = lm(Y~X, data=samp)
model2 = lm(Y~X, data=samp[samp$X %in% c(0,1), ])
grid = expand.grid(X=seq(0,1,by=.01))
ggplot(samp) + geom_point(aes(x=X, y=Y), color='red', alpha=.3) + 
  geom_line(aes(x=X, y=predict(model1, newdata=data.frame(X=X))), data=grid, color='blue') +
  coord_cartesian(ylim=c(-3,3), xlim=c(0,1))

ggplot(samp[samp$X %in% c(0,1), ]) +
 geom_point(aes(x=X, y=Y), color='red', alpha=.3) + 
  geom_line(aes(x=X, y=predict(model2, newdata=data.frame(X=X))), data=grid, color='blue') +
  coord_cartesian(ylim=c(-3,3), xlim=c(0,1)) 

sampdist = purrr::map(1:1000, function(rep) {
  samp = pop[sample(1:nrow(pop), n, replace=TRUE), ]
  model1 = lm(Y~X, data=samp)
  model2 = lm(Y~X, data=samp[samp$X %in% c(0,1), ])
  data.frame(val=c(coef(model1)[1], coef(model2)[1]), model=c('model1', 'model2'))
}) %>% bind_rows()

ggplot(sampdist[sampdist$model=='model2', ])  +
  geom_histogram(aes(x=val, y=after_stat(density)), bins=50, alpha=.2) +
  geom_vline(aes(xintercept=mu(0)), color='black') +
  geom_vline(aes(xintercept=mean(val)), color='blue') +
  geom_vline(aes(xintercept=mean(val)-sd(val)), color='blue', linetype='dashed') + 
  geom_vline(aes(xintercept=mean(val)+sd(val)), color='blue', linetype='dashed') + 
  geom_vline(aes(xintercept=mean(val)-2*sd(val)), color='blue', linetype='dotted') +
  geom_vline(aes(xintercept=mean(val)+2*sd(val)), color='blue', linetype='dotted') 

ggplot(sampdist[sampdist$model=='model1', ])  +
  geom_histogram(aes(x=val, y=after_stat(density)), bins=50, alpha=.2) +
  geom_vline(aes(xintercept=mu(0)), color='black') +
  geom_vline(aes(xintercept=mean(val)), color='blue') +
  geom_vline(aes(xintercept=mean(val)-sd(val)), color='blue', linetype='dashed') + 
  geom_vline(aes(xintercept=mean(val)+sd(val)), color='blue', linetype='dashed') + 
  geom_vline(aes(xintercept=mean(val)-2*sd(val)), color='blue', linetype='dotted') +
  geom_vline(aes(xintercept=mean(val)+2*sd(val)), color='blue', linetype='dotted') 
```

\subsection*{Part A}
True or False. These plots are in correspondence, i.e., the sampling distribution on the left corresponds to the line fit to the sample on the left.

\subsection*{Part B}
Suppose I use the bootstrap to get a good estimate $\hat\text{se}$ of the standard error of each estimator $\hat\mu(0)$.
For each of these estimators, approximate the coverage of the 95\% confidence interval $\hat\mu(0) \pm 2 \hat\text{se}$.
Is it closer to 95\%, 50\%, or 0\%?  So that the correctness of this one isn't tied to the previous one, please report this for...
\begin{itemize}
\item the estimator whose sampling distribution is on the left.
\item the estimator whose sampling distribution is on the right.
\end{itemize}


\newpage

\subsection*{Problem 3}

In this problem, we'll work with a sample $(D_1, X_1, Y_1) \ldots (D_n,X_n,Y_n)$ from some population. I've plotted it below. The $X$ variable is on the horizontal axis, the $Y$ variable is on the vertical axis, and the $D$ variable, which is binary, is indicated by color---green for $D=1$, red for $D=0$. The sample is from a conditionally randomized study, in which $D$ is randomized conditional on $X$ with probability $\pi(x)=P(D=1 \mid X=x)$. 
I've fit an interacted bivariate model to the data using unweighted least squares. 
The solid lines show the fitted model $\hat\mu$ with $\hat\mu(1,x)$
and $\hat\mu(0,x)$ in green and red respectively.


```{r}
X <- c(0,2,4,0,0,2,4,4)
Y <- c(2,4,6,12.25,8,5.25,3.25,1)
D <- c(1,1,1,0,0,0,0,0)

data <- data.frame(X=X,Y=Y,D=D)
model = lm(Y~D*X, data=data)
grid = expand.grid(X=seq(0,4,by=.01), D=0:1)
grid$muhat = predict(model, newdata=grid)
plot<- ggplot(data = data, aes(x = X, y = Y)) + 
  geom_point(aes(color=factor(D)),size=2) + 
  scale_y_continuous(breaks = seq(0, 13, by = 1)) + 
  geom_line(aes(x=X, y=muhat, color=factor(D)), data=grid) 
    
plot + theme(legend.position="none")
```




You'll be estimating the Average Treatment Effect (ATE)  and the Average Treatment Effect on the Treated (ATT).
$$
\begin{aligned}
\text{ATE} &= \E[ \mu(1,X) - \mu(0,X) ] \\
\text{ATT} &= \E[ \mu(1,X) - \mu(0,X) \mid D=1] 
\end{aligned}
$$

As your estimates, you'll use the following.
$$
\begin{aligned}
\widehat{ATE} &= \frac{1}{n} \sum_{i=1}^n \qty{\hat\mu(1,X_i) - \hat\mu(0,X_i)} \\
\widehat{ATT} &= \frac{1}{N_1} \sum_{i:D_i=1} \qty{\hat\mu(1,X_i) - \hat\mu(0,X_i)} \qfor N_1 = \sum_{i=1}^n D_i 
\end{aligned}
$$



\subsection*{Part a}

What is your ATE estimate $\widehat{ATE}$ based on this model and data? Show your calculations.

\begin{answer}

$\frac{1}{8}[(2-10)+(2-10)+(2-10)+(4-6)+(4-6)+(6-2)+(6-2)+(6-2)] = -2$

\end{answer}

\subsection*{Part b}

What is the $\widehat{ATT}$ based on this model and data? Show your calculations.

\begin{answer}

$\frac{1}{3}[(2-10)+(4-6)+(6-2)] = -2$ 

\end{answer}

\newpage

\subsection*{Problem 4}
In this question, we'll repeat our exercise from Problem 3 using a larger sample, then go a bit further.
The set-up is exactly the same as before. We have a sample of $(D_1, X_1, Y_1) \ldots (D_n,X_n,Y_n)$ from some population. I've plotted it below. The $X$ variable is on the horizontal axis, the $Y$ variable is on the vertical axis, and the $D$ variable, which is binary, is indicated by color---green for $D=1$, red for $D=0$. I've used some jitter when plotting to make things easier to see---$X$ really takes on the four values $0$,$1/3$,$2/3$,and $1$.


The sample is from a conditionally randomized study, in which $D$ is randomized conditional on $X$ with probability $\pi(x)=P(D=1 \mid X=x)$. I've fit an interacted bivariate model to the data using unweighted least squares. The solid lines show the fitted model $\hat\mu$ with $\hat\mu(1,x)$ and $\hat\mu(0,x)$ in green and red respectively.
I've also plotted $\bar Y_{x,d}$, the average of $Y$ among individuals with $X=x$ and $D=d$, as well as arms spannding $\pm 2$ standard errors, for all $x$ and $d$.

```{r scatter}
mu = function(d, x){.5 + (1-d)*sinpi(x/2) + d*x}
pi = function(x){ .1 + .6*abs(cospi(x)) }

set.seed(1)
n = 100
X = sample(seq(0,1,by=.333), n, replace=T)
D = rbinom(n, 1, pi(X))
Y = mu(D,X) + .1*rnorm(n)
data = data.frame(X=X, D=D, Y=Y)
summaries = data %>% group_by(D,X) %>% summarize(mu=mean(Y), n = n(), se=sd(Y)/sqrt(n)) %>%
                     group_by(D) %>% mutate(f=n/sum(n))

model = lm(Y~D*X, data=data)
summaries$muhat = predict(model, newdata=summaries)

summaries0 = summaries %>% filter(D==0)
summaries1 = summaries %>% filter(D==1)
summaries0 = summaries0[order(summaries0$X),]
summaries1 = summaries1[order(summaries1$X),]
f = mean(D) * summaries0$f + (1-mean(D)) * summaries1$f

scatter = ggplot(data) + 
  geom_point(aes(x=X+.025*(2*D-1), y=Y, color=factor(D)), 
             position=position_jitter(w=.025,h=0), alpha=.2, size=.5) + 
  geom_line(aes(x=X, y=muhat, color=factor(D)), data=summaries, size=.5) +
  scale_x_continuous(breaks=round(seq(0,1,by=1/3),2)) + xlab('') + ylab('') + 
  geom_col(aes(x=X+.025*(2*D-1), y=f, fill=factor(D), color=factor(D)), data=summaries, position='identity', alpha=.2, width=.1) +
  geom_col(aes(x=X, y=f), fill='purple', color='purple', 
    data=data.frame(f=f, X=summaries0$X), position='identity', alpha=.2, width=.05) +
  geom_pointrange(aes(x=X+.025*(2*D-1), y=mu, ymin=mu-2*se, ymax=mu+2*se, color=factor(D)), 
                  data=summaries, alpha=.8, size=.2) +
  guides(color='none', fill='none') + theme_bw()
scatter 
```

This time, you probably won't want to count dots. Instead, you'll work with summaries of the data's distribution.
The proportion of times that $X$ takes on each value $x$ when $D=0$ and when $D=1$, $\hat f_{0}(x)$ and $\hat f_1(x)$ respectively, are plotted as histograms in matching colors. In purple, I've plotted $\hat f$, the proportion of times $X$ takes on each value $x$ among all observations---I've made that one a bit thinner so it's easier to see the other two.

```{r}
options(digits=2)
```
$$
\begin{aligned}
\hat f_0(x) &= \frac{1}{N_0}\sum_{i:D_i=0} 1(X_i=x)  \qfor N_0 = \sum_{i=1}^n (1-D_i) \\
\hat f_1(x) &= \frac{1}{N_1}\sum_{i:D_i=1} 1(X_i=x)  \qfor N_1 = \sum_{i=1}^n D_i \\
\hat f(x)   &= \frac{1}{n}\sum_{i=1}^n 1(X_i=x)
\end{aligned}
$$

Here's a table of the summaries you see in the plot.

$$
\begin{array}{llllllll}
x & \hat f_0(x) & \hat f_1(x) & \hat f(x) & \hat\mu(0,x) & \hat \mu(1,x) & \bar Y_{0,x} & \bar Y_{1,x} \\
\hline
0 & `r summaries0$f[1]` & `r summaries1$f[1]` & `r f[1]` & `r summaries0$muhat[1]` & `r summaries1$muhat[1]` & `r summaries0$mu[1]` & `r summaries1$mu[1]` \\
\frac13 & `r summaries0$f[2]` & `r summaries1$f[2]` & `r f[2]` & `r summaries0$muhat[2]` & `r summaries1$muhat[2]` & `r summaries0$mu[2]` & `r summaries1$mu[2]` \\
\frac23 & `r summaries0$f[3]` & `r summaries1$f[3]` & `r f[3]` & `r summaries0$muhat[3]` & `r summaries1$muhat[3]` & `r summaries0$mu[3]` & `r summaries1$mu[3]` \\
1 & `r summaries0$f[4]` & `r summaries1$f[4]` & `r f[4]` & `r summaries0$muhat[4]` & `r summaries1$muhat[4]` & `r summaries0$mu[4]` & `r summaries1$mu[4]` \\
\end{array}
$$


\subsection*{Part a}

What is your ATE estimate $\widehat{ATE}$ based on this model and data? Show your calculations.

\begin{answer}

\vspace{1in}

\end{answer}

\subsection*{Part b}

What is your ATT estimate $\widehat{ATT}$ based on this model and data? Show your calculations.

\begin{answer}

\vspace{2in}

\end{answer}

\subsection*{Part c}

For each of your two estimates: is its bias positive, negative, or roughly zero? With reference to the plot above explain your answer and explain why this is happening. If it's hard to see what's going on in that plot, it may help to take a look at the plot below, in which I've done exactly the same thing with a larger sample.

\paragraph{Bonus.} Explain why the bias we get at this larger sample size tells us something about the bias we get at the smaller sample size we worked with above.

```{r scatter-big}
mu = function(d, x){.5 + (1-d)*sinpi(x/2) + d*x}
pi = function(x){ .1 + .6*abs(cospi(x)) }

set.seed(1)
n = 1000
X = sample(seq(0,1,by=.333), n, replace=T)
D = rbinom(n, 1, pi(X))
Y = mu(D,X) + .1*rnorm(n)
data = data.frame(X=X, D=D, Y=Y)
summaries = data %>% group_by(D,X) %>% summarize(mu=mean(Y), n = n(), se=sd(Y)/sqrt(n)) %>%
                     group_by(D) %>% mutate(f=n/sum(n))

model = lm(Y~D*X, data=data)
summaries$muhat = predict(model, newdata=summaries)

summaries0 = summaries %>% filter(D==0)
summaries1 = summaries %>% filter(D==1)
summaries0 = summaries0[order(summaries0$X),]
summaries1 = summaries1[order(summaries1$X),]
f = mean(D) * summaries0$f + (1-mean(D)) * summaries1$f

scatter = ggplot(data) + 
  geom_point(aes(x=X+.025*(2*D-1), y=Y, color=factor(D)), 
             position=position_jitter(w=.025,h=0), alpha=.2, size=.5) + 
  geom_line(aes(x=X, y=muhat, color=factor(D)), data=summaries, size=.5) +
  scale_x_continuous(breaks=round(seq(0,1,by=1/3),2)) + xlab('') + ylab('') + 
  geom_col(aes(x=X+.025*(2*D-1), y=f, fill=factor(D), color=factor(D)), data=summaries, position='identity', alpha=.2, width=.1) +
  geom_col(aes(x=X, y=f), fill='purple', color='purple', 
    data=data.frame(f=f, X=summaries0$X), position='identity', alpha=.2, width=.05) +
  geom_pointrange(aes(x=X+.025*(2*D-1), y=mu, ymin=mu-2*se, ymax=mu+2*se, color=factor(D)), 
                  data=summaries, alpha=.8, size=.2) +
  guides(color='none', fill='none') + theme_bw()
scatter 
```




\begin{answer}

\vspace{2in}
\end{answer}

\subsection*{Part d}

In terms of quantities available in the plot and table, write \emph{unbiased} estimates of the ATE and ATT. Write these first as a formula involving quantities like $\hat f_0(x)$, $\hat f_1(x)$, etc. then calculate their value using the numbers in the table. 

\begin{answer}

\vspace{2in}
\end{answer}

\newpage

\section*{Problem 5}
```{r}
#| layout-ncol: 2
set.seed(5)

mu = function(d, x){.2 + x + (1-d) }
pi = function(x){ 1 - .01 -.98*abs(cospi(x/2)) }

m = 10000
X = sample(seq(0,1,by=.3333), m, replace=T)
D = rbinom(m, 1, pi(X))
Y = mu(D,X) + .02*rnorm(m)
pop = data.frame(X=X, D=D, Y=Y)

pop.scatter = ggplot(pop) + 
  geom_point(aes(x=X+.01*(2*D-1), y=Y, color=factor(D)), 
             position=position_jitter(w=.1,h=0), alpha=.3, size=1) + 
  scale_x_continuous(breaks=seq(0,1,by=.1)) + xlab('') + ylab('') + 
  geom_histogram(aes(x=X, y=.4*after_stat(density), fill=factor(D), color=factor(D)), 
    position='identity', bins=10, alpha=.2) +
  guides(color='none', fill='none') + theme_bw()

n = 1000
set.seed(7)
data = pop[sample(1:nrow(pop), n, replace=TRUE), ]
summaries = data %>% group_by(D,X) %>% summarize(mu=mean(Y), n = n()) %>%
                     group_by(D) %>% mutate(f=n/sum(n))

grid = expand.grid(X=seq(0,1,by=.3333), D=0:1)
sample.scatter = ggplot(data) + 
  geom_point(aes(x=X+.01*(2*D-1), y=Y, color=factor(D)), 
             position=position_jitter(w=.1,h=0), alpha=.2, size=2) + 
  geom_line(aes(x=X, y=mu, color=factor(D)), data=summaries) +
  scale_x_continuous(breaks=seq(0,1,by=.1)) + xlab('') + ylab('') + 
  geom_col(aes(x=X, y=f, fill=factor(D), color=factor(D)), data=summaries, position='identity', alpha=.2, width=.25) +
  geom_line(aes(x=X, y=pi(X)), data=grid, size=1, color='purple') +
  geom_point(aes(x=X, y=pi(X)), data=grid, color='purple') +
  guides(color='none', fill='none') + theme_bw()

#pop.scatter 
sample.scatter

estimate.ATT = function(model, data) {
  X = data$X
  D = data$D
  muhat = function(d,x) { predict(model, newdata=data.frame(X=x,D=d)) }
  mean(muhat(1,X[D==1]) - muhat(0, X[D==1]))
}
B = 5000
estimates = data.frame(additive=rep(NA,B), additive.ipw=rep(NA,B)) 
set.seed(7)
for( b in 1:B ){
  data = pop[sample(1:nrow(pop), n, replace=TRUE), ]
  data$ATTweights = data$D + (1-data$D)*pi(data$X)/(1-pi(data$X)) 
  additive.model.ipw = lm(Y~D*X, data=data, weights=ATTweights)
  additive.model     = lm(Y~D*X, data=data)
  estimates$additive[b]     = estimate.ATT(additive.model, data)
  estimates$additive.ipw[b] = estimate.ATT(additive.model.ipw, data)
}
ggplot(estimates) + 
  geom_histogram(aes(x=additive, y=after_stat(density)), alpha=.3, fill='orange', bins=50) +
  geom_histogram(aes(x=additive.ipw, y=after_stat(density)), alpha=.3, fill='blue', bins=50)
```

Some people don't like inverse probability weighting. What they say is that often, some observations get way more weight --- and therefore influence on the least squares fit --- than others. Suppose we're estimating the ATT. Above, I've plotted a sample in which this will happen. The probability of treatment, $\pi(x)=\E[D_i \mid X_i=x]$, is drawn in purple. And the subpopulations means, $\mu(d,x)$, are drawn in as solid lines in red and green as usual. 
You know these because this is a sample drawn from a population your coworker made up.
Your coworker, who wrote the code that generated the fake data above, fits the additive bivariate model via unweighted least squares. They then use that model to estimate the ATT. Here's their estimate, which we'll call `the orange estimate'.

$$
\begin{aligned}
\textcolor{orange}{\widehat{ATT}} &= \frac{1}{N_1} \sum_{i:D_i=1} \qty{\hat\mu(1,X_i) - \hat\mu(0,X_i)} \qfor N_1 = \sum_{i=1}^n D_i \qqtext{where} \\
\hat\mu &= \argmin_{m \in \model} \sum_{i=1}^n \qty{m(D_i, X_i) - Y_i}^2 \qfor \\
\model &= \qty{m(d, x) = b_0 + b_1 x + b_2 d + b_3 dx \ : \ x \in \R^4}
\end{aligned}
$$

You suggest a small tweak: that they use inverse probability weighted least squares when they estimate $\mu(x)$. Here's your estimate, which we'll call `the blue estimate'.

$$
\begin{aligned}
\textcolor{blue}{\widehat{ATT}} &= \frac{1}{N_1} \sum_{i:D_i=1} \qty{\hat\mu(1,X_i) - \hat\mu(0,X_i)} \qfor N_1 = \sum_{i=1}^n D_i \qqtext{where} \\
\hat\mu &= \argmin_{m \in \model} \sum_{i=1}^n \qty{D_i+\frac{(1-D_i)\pi(X_i)}{1-\pi(X_i)}} \qty{m(D_i, X_i) - Y_i}^2 \qfor \\
\model &= \qty{m(d, x) = b_0 + b_1 x + b_2 d + b_3 dx \ : \ x \in \R^4}
\end{aligned}
$$

They argue that their estimator is better by producing a plot of the sampling distributions of the two estimators. In it, we see the sampling distribution of ATT estimators based on the interacted bivariate model, fit with (in blue) and without (in orange) inverse probability weights. And, as you can see, the two estimators have the same mean, but the inverse-probability-weighted version has a larger variance. To produce it, they've drawn `r B` samples of size n=`r n` with replacement from their population and calculated the two estimators. You know that their estimator only works because the regression model they're using is not misspecified --- $\mu(0,x)$ and $\mu(1,x)$ really are lines. To demonstrate their error, you want to make a plot like theirs, using the same estimators, but a slightly different population. 

\subsection*{Part A}
You try two things.
\begin{enumerate}
\item Changing their population so that $\mu(1,x)=x^2$, leaving $\mu(0,x)$ as a line. 
\item Changing their population so that $\mu(0,x)=1+x^2$, leaving $\mu(1,x)$ as a line.
\end{enumerate}
Here, in some order, are the plots you get. Only one of these is useful in arguing their point. 
Which one is it? Which of the two changes did you make to the population to get it? Explain how you know.

```{r}
#| layout-ncol: 2
set.seed(5)

mu = function(d, x){.2 + d*x^2 + (1-d)*(1+x) }
pi = function(x){ 1 - .01 -.98*abs(cospi(x/2)) }

m = 10000
X = sample(seq(0,1,by=.3333), m, replace=T)
D = rbinom(m, 1, pi(X))
Y = mu(D,X) + .02*rnorm(m)
pop = data.frame(X=X, D=D, Y=Y)

estimate.ATT = function(model, data) {
  X = data$X
  D = data$D
  muhat = function(d,x) { predict(model, newdata=data.frame(X=x,D=d)) }
  mean(muhat(1,X[D==1]) - muhat(0, X[D==1]))
}

B = 5000
estimates = data.frame(additive=rep(NA,B), additive.ipw=rep(NA,B)) 
set.seed(7)
for( b in 1:B ){
  data = pop[sample(1:nrow(pop), n, replace=TRUE), ]
  data$ATTweights = data$D + (1-data$D)*pi(data$X)/(1-pi(data$X)) 
  additive.model.ipw = lm(Y~D*X, data=data, weights=ATTweights)
  additive.model     = lm(Y~D*X, data=data)
  estimates$additive[b]     = estimate.ATT(additive.model, data)
  estimates$additive.ipw[b] = estimate.ATT(additive.model.ipw, data)
}
ggplot(estimates) + 
  geom_histogram(aes(x=additive, y=after_stat(density)), alpha=.3, fill='orange', bins=50) +
  geom_histogram(aes(x=additive.ipw, y=after_stat(density)), alpha=.3, fill='blue', bins=50)

set.seed(5)

mu = function(d, x){.2 + d*x + (1-d)*(1+x^2) }
pi = function(x){ 1 - .01 -.98*abs(cospi(x/2)) }

m = 10000
X = sample(seq(0,1,by=.3333), m, replace=T)
D = rbinom(m, 1, pi(X))
Y = mu(D,X) + .02*rnorm(m)
pop = data.frame(X=X, D=D, Y=Y)
data = pop[sample(1:nrow(pop), n, replace=TRUE), ]
summaries = data %>% group_by(D,X) %>% summarize(mu=mean(Y), n = n()) %>%
                     group_by(D) %>% mutate(f=n/sum(n))

B = 5000
estimates = data.frame(additive=rep(NA,B), additive.ipw=rep(NA,B)) 
set.seed(7)
for( b in 1:B ){
  data = pop[sample(1:nrow(pop), n, replace=TRUE), ]
  data$ATTweights = data$D + (1-data$D)*pi(data$X)/(1-pi(data$X)) 
  additive.model.ipw = lm(Y~D*X, data=data, weights=ATTweights)
  additive.model     = lm(Y~D*X, data=data)
  estimates$additive[b]     = estimate.ATT(additive.model, data)
  estimates$additive.ipw[b] = estimate.ATT(additive.model.ipw, data)
}
ggplot(estimates) + 
  geom_histogram(aes(x=additive, y=after_stat(density)), alpha=.3, fill='orange', bins=50) +
  geom_histogram(aes(x=additive.ipw, y=after_stat(density)), alpha=.3, fill='blue', bins=50)
```

\newpage

\subsection*{Part B}

Suppose you'd been estimating the ATE instead of the ATT and you'd used appropriate weights ---
$D_i/\pi(X_i) + (1-D_i)/(1-\pi(X_i))$ --- so that your estimator was unbiased. Would you see a problem with the unweighted approach using your coworker's population? What about with the two populations you tried in part A? Explain your answer. 