---
title: "Final"
subtitle: "QTM 285-1"
crossref: 
  chapters: true
---


::: {.hidden}
{{< include shared-code.qmd >}}
{{< include more-shared-code.qmd >}}

```{r}
theme_set(whitebg.theme())
```

:::

# Problem 1 

```{r}

set.seed(14)
m = 1000
y = rnorm(m); y=round(.3 + ((y-mean(y)) / sd(y)))
theta = mean(y) 
n = 20

set.seed(3) # use seed 3 for 93% coverage
J = sample(1:m, n, replace=FALSE)
Y = y[J]
set.seed(0)
```

In this problem, we'll be working with a sample $Y_1 \ldots Y_{`r n`}$ drawn with replacement from a population $y_1 \ldots y_{`r m`}$. 
The responses $y_j$ are integers between -3 and 3. I've plotted the distribution of responses in the population and the sample below.
We're interested in estimating the population mean $\mu = \frac{1}{m} \sum_{j=1}^m y_j$.


```{r}
#| layout-ncol: 2
#| fig-cap: 
#|  - "The distribution of responses in the population"
#|  - "The distribution of responses in the sample"

pop.scales = list(scale_x_continuous(breaks=seq(-5,5,by=1), limits=c(-3,3)), 
              scale_y_continuous(breaks=seq(0,1,by=.1), limits=c(0,.6)))
ggplot() + geom_bar(aes(x=y,y=after_stat(prop))) + pop.scales
ggplot() + geom_bar(aes(x=Y,y=after_stat(prop))) + pop.scales
```

::: {#exr-1a .callout-exercise}
Below, I've plotted the sampling distributions of four estimators of [$\mu$]{.target}.

$$
\begin{aligned}
\hat\mu_1 &= \bar Y = \frac{1}{`r n`} \sum_{i=1}^{`r n`} Y_i && \text{sample mean}  \\
\hat\mu_2 &= \frac{1}{`r n+20`} \sum_{i=1}^{`r n`} Y_i && \text{shrinkage} \\
\hat\mu_3 &= \begin{cases} 0 & \text{if } \abs{\bar Y} \leq .125 \\ \bar Y & \text{otherwise} \end{cases} && \text{keep-or-kill} \\
\hat\mu_4 &= Y_1 && \text{one observation}
\end{aligned}
$$
Which is which? 

Your answer should be a list of four number/letter pairs, e.g. 1:A, 2:B, 3:C, 4:D. 
:::

```{r}
#| layout-ncol: 2
#| layout-nrow: 2
#| fig-width: 5 
#| fig-cap: 
#|   - "Sampling Distribution A"
#|   - "Sampling Distribution B"
#|   - "Sampling Distribution C"
#|   - "Sampling Distribution D"

set.seed(15)
samples = 1:10000 |> map( function(.) { 
  J = sample(1:m, n, replace=TRUE)  
  Y = y[J]                          
})
samples[1]=Y

mean.samples = samples |> map_vec(mean)
scales = list(scale_x_continuous(breaks=seq(-.6,1.2,by=.2), 
                labels=function(x)sprintf("%1.1f", x)), 
              scale_y_continuous(breaks=seq(0,5,by=.25)),
              coord_cartesian(ylim=c(0,5), xlim=c(-.6,1.2)))

binwidth = .05
centers = seq(-1.5,1.5,by=binwidth) 
ggplot(data.frame(x=mean.samples)) + 
  geom_histogram(aes(x=x, y=after_stat(density)), alpha=.2, center=0, binwidth=binwidth, fill=pink, color='black', linewidth=.05) +  
  with_annotations(mean.samples, dots=0, height=2, dot.size=1, estimand=mean(y)) +
  scales

kok = function(Y, kill.threshold=binwidth*2.5) { 
  if(abs(mean(Y)) <= kill.threshold) { 0 } else { mean(Y)}
}

kok.samples = samples |> map_vec(kok)
ggplot(data.frame(x=kok.samples)) + 
  geom_histogram(aes(x=x, y=after_stat(density)), alpha=.2, center=0, binwidth=binwidth, fill=pink, color='black', linewidth=.05) +  
  with_annotations(kok.samples, dots=1, height=2, dot.size=1, estimand=mean(y)) +
  scales

shrinkage.samples = samples |> map_vec(\(x)sum(x)/(20+length(x)))
ggplot(data.frame(x=shrinkage.samples)) + 
  geom_histogram(aes(x=x, y=after_stat(density)), alpha=.2, center=0, binwidth=binwidth, fill=pink, color='black', linewidth=.05) + 
  with_annotations(shrinkage.samples, dots=0, height=2, dot.size=1, estimand=mean(y)) +
  scales
  
ggplot() + geom_bar(aes(x=Y,y=after_stat(prop)), alpha=.2, 
  fill=pink, color='black', linewidth=.05) + 
  with_annotations(Y, dots=0, height=1.5, dot.size=1, estimand=mean(y)) +
  pop.scales
```


::: {#exr-1b .callout-exercise}
Suppose you want to know the coverage of the interval estimator $\hat\mu_2 \pm 1.96 \ \text{sd}(\hat\mu_2^\star)$ 
where $\text{sd}(\hat\mu_2^\star)$ is the standard deviation of draws from the bootstrap sampling distribution of $\hat\mu_2$.
You have the sample $Y_1 \ldots Y_{`r n`}$ stored in an **R** vector `Y` of length `r n` and the population $y_1 \ldots y_{`r m`}$ stored in an **R** vector `y`. Is the code below correct? If not, explain how to fix it.
:::

```{r}
#| echo: true 

estimator = function(Y) { sum(Y) / (n + 20) }

bootstrap.sampling.distribution = function(Y) { 
  1:1000 |> map_vec( function(.) {  
    J.star = sample(1:n, n, replace=TRUE)
    Y.star = Y[J.star]
    estimator(Y.star)
  })
}

se = sd(bootstrap.sampling.distribution(Y))
covers = 1:1000 |> map_vec( function(.) { 
  J = sample(1:m, n, replace=TRUE)  
  Y = y[J]                          
  estimator(Y) - 1.96*se <= theta & theta <= estimator(Y) + 1.96*se
})
mean(covers)
```



# Problem 2

Below, I've plotted samples drawn with replacement from two populations.
On top of them, I've drawn lines fit to them via least squares as follows.
$$ 
\hat \mu = \argmin_{m \in \model} \sum_{i=1}^n \qty{Y_i - m(X_i)}^2 \qfor \model = \qty{m(x) = a + b x  \ \text{ for numbers $a$, $b$} }
$$

Below these, in some order, I've plotted the sampling distributions of the predictions $\hat\mu(0)$ I get when I fit this model to each sample. As usual, I've indicated the mean of each sampling distribution with a solid blue line, one standard deviation in either direction with a dashed blue line, and two standard deviations in either direction with a dotted blue line. The value of $\mu(0)$, the mean of $y$ among individuals in the population with $x=0$, is indicated by a green line. 

```{r}
#| layout-ncol: 2
#| layout-nrow: 2

set.seed(1)
mu = function(x){.2 + .7*x + .8*sinpi(x)}
m = 1000000
X = round(3*runif(m))/3
Y = mu(X) + .5*rnorm(m)
pop = data.frame(X=X, Y=Y)

n=30

samp = pop[sample(1:nrow(pop), n, replace=TRUE), ]
model1 = lm(Y~X, data=samp)
model2 = lm(Y~X, data=samp[samp$X %in% c(0,1), ])
grid = expand.grid(X=seq(0,1,by=.01))
ggplot(samp) + geom_point(aes(x=X, y=Y), color='red', alpha=.3) + 
  geom_line(aes(x=X, y=predict(model1, newdata=data.frame(X=X))), data=grid, color='blue', linetype='dashed') +
  geom_point(aes(x=X, y=predict(model1, newdata=data.frame(X=X))), data=samp, color='blue', shape=23) +
  scale_x_continuous(breaks=round(seq(0,1,by=1/3),2)) +
  coord_cartesian(ylim=c(-3,3), xlim=c(-.1,1.1))

ggplot(samp[samp$X %in% c(0,1), ]) +
 geom_point(aes(x=X, y=Y), color='red', alpha=.3) + 
  geom_line(aes(x=X, y=predict(model2, newdata=data.frame(X=X))), data=grid, color='blue', linetype='dashed') +
  geom_point(aes(x=X, y=predict(model2, newdata=data.frame(X=X))), data=samp, color='blue', shape=23) +
  scale_x_continuous(breaks=round(seq(0,1))) +
  coord_cartesian(ylim=c(-3,3), xlim=c(-.1,1.1)) 

sampdist = purrr::map(1:1000, function(rep) {
  samp = pop[sample(1:nrow(pop), n, replace=TRUE), ]
  model1 = lm(Y~X, data=samp)
  model2 = lm(Y~X, data=samp[samp$X %in% c(0,1), ])
  data.frame(val=c(coef(model1)[1], coef(model2)[1]), model=c('model1', 'model2'))
}) |> bind_rows()

ggplot(sampdist[sampdist$model=='model2', ])  +
  geom_histogram(aes(x=val, y=after_stat(density)), bins=50, alpha=.2) +
  geom_vline(aes(xintercept=mu(0)), color='green', linewidth=1.5) +
  geom_vline(aes(xintercept=mean(val)), color='blue') +
  geom_vline(aes(xintercept=mean(val)-sd(val)), color='blue', linetype='dashed') + 
  geom_vline(aes(xintercept=mean(val)+sd(val)), color='blue', linetype='dashed') + 
  geom_vline(aes(xintercept=mean(val)-2*sd(val)), color='blue', linetype='dotted') +
  geom_vline(aes(xintercept=mean(val)+2*sd(val)), color='blue', linetype='dotted') 

ggplot(sampdist[sampdist$model=='model1', ])  +
  geom_histogram(aes(x=val, y=after_stat(density)), bins=50, alpha=.2) +
  geom_vline(aes(xintercept=mu(0)), color='green', linewidth=1.5) +
  geom_vline(aes(xintercept=mean(val)), color='blue') +
  geom_vline(aes(xintercept=mean(val)-sd(val)), color='blue', linetype='dashed') + 
  geom_vline(aes(xintercept=mean(val)+sd(val)), color='blue', linetype='dashed') + 
  geom_vline(aes(xintercept=mean(val)-2*sd(val)), color='blue', linetype='dotted') +
  geom_vline(aes(xintercept=mean(val)+2*sd(val)), color='blue', linetype='dotted') 
```

::: {#exr-correspondence-tf .callout-exercise}
True or False. These plots are in correspondence, i.e., the sampling distribution on the left corresponds to the line fit to the sample on the left.
:::

::: {#exr-coverage-tf .callout-exercise}
Suppose I use the bootstrap to get a good estimate $\widehat{\text{se}}$ of the standard deviation of each estimator $\hat{\mu}(0)$.
For each of these estimators, approximate the coverage of the 95\% confidence interval $\hat\mu(0) \pm 1.96\ \widehat{\text{se}}$.
Is it closer to 95\%, 50\%, or 0\%?  So that the correctness of this one isn't tied to the previous one, please report this for...

1. The estimator whose sampling distribution is on the left.
2. The estimator whose sampling distribution is on the right.

:::


# Problem 3

You've run a turnout poll like the ones we discussed earlier in the semester.
This gives you a sample $Y_1 \ldots Y_n$ drawn with replacement from a population 
of binary responses $y_1 \ldots y_n$ with mean $\mu$. But instead of using the 
sample mean as your estimator of $\mu$, you've used this shrinkage estimator.
$$
\hat\mu = \argmin_{\text{numbers}\ m} \sum_{i=1}^n \qty{ Y_i - m }^2 + \lambda \qty(m-\frac{1}{2})^2
$$

::: {#exr-ridge-bias-se .callout-exercise}
What is this estimator's bias and standard deviation?
Your answers should be formulas involving $n$, $\lambda$, and $\mu$.
:::

::: {#exr-ridge-coverage .callout-exercise}
For this estimator, the ratio of its bias and its standard deviation is
$$
\frac{\text{bias}}{\text{se}} = \frac{\frac{1}{2} - \mu}{\sqrt{\mu(1-\mu)}} \times \frac{\lambda}{\sqrt{n}}.
$$
Suppose the population mean is $\mu=.7$, our sample size is $n=100$, and we use 'shrinkage level' $\lambda=10$ and consider the interval estimator $\hat\mu \pm 1.96 \ \text{se}$ where $\text{se}$ is the standard deviation of $\hat\mu$.

Approximately what is this estimator's coverage: 90%, 50%, or 10%?
:::


::: {#exr-ridge-ec .callout-exercise}
*Extra Credit*.

The plots below show four sampling distributions. For each, we're using a sample $Y_1 \ldots Y_n$ drawn with replacement from a population of binary responses $y_1 \ldots y_n$ with mean [$\mu=.7$]{.target}.

1. The sampling distribution of $\hat\mu$ for $\lambda=10$ when $n=100$.
2. The sampling distribution of $\hat\mu$ for $\lambda=100$ when $n=100$.
3. The sampling distribution of $\hat\mu$ for $\lambda=10$ when $n=400$. 
4. The sampling distribution of a *randomized estimator* that is, depending on a weighted coin-flip, either $1/2$ or the sample mean $\bar Y$ when $n=100$.

Which is which? 

Your answer should be a list of four number/letter pairs, e.g. 1:A, 2:B, 3:C, 4:D. 
:::


```{r}
#| layout-nrow: 2
#| layout-ncol: 2
#| fig-width: 5
#| fig-cap: 
#|   - "Sampling Distribution A"
#|   - "Sampling Distribution B"
#|   - "Sampling Distribution C"
#|   - "Sampling Distribution D" 

n = 100
mu = .7

coords = list(
  scale_x_continuous(breaks = seq(0,1,by=.1)),
  scale_y_continuous(breaks = seq(0,20,by=2)),
  coord_cartesian(xlim=c(.4,.9), ylim=c(0,20)))

plot.sampling.dist = function(n, mu, lambda, mixture=FALSE) { 

  x = seq(0,1,by=1/n)
  mean.pmf  = dbinom(n*x, n, mu)
  alpha = lambda/(n+lambda)
  se = (n / (n+lambda)) * sqrt(mu*(1-mu)/n) 
  
  histogram = if(mixture) {
     other.pmf = 0+(x==1/2)
     geom_col(aes(x=x,y=n*((1-alpha)*mean.pmf + alpha*other.pmf)), alpha=.2) 
  } else {
    coverage = sum(mean.pmf[ abs((x+alpha/2)*(1-alpha) - mu) <= 1.96*se ])
    geom_col(aes(x=(1-alpha)*x + alpha*(1/2), y=n*mean.pmf), alpha=.2)  
  }

  ggplot() + 
      geom_vline(xintercept=mu, color='green', alpha=.9) +
      geom_vline(xintercept=(1-alpha)*mu + alpha*(1/2), color='blue', alpha=.9) + 
      histogram 
     # annotate('rect', xmin=mu-1.96*se, xmax=mu+1.96*se, ymin=0, ymax=Inf, fill='yellow', alpha=.2) + 
}

plot.sampling.dist(n,mu,10,mixture=TRUE) + coords
plot.sampling.dist(n,mu,100) + coords
plot.sampling.dist(n,mu,10) + coords
plot.sampling.dist(4*n,mu,10) + coords
```

::: {.callout-tip}
### Tip.
Use information from different exercises to check your answers. 

1. Use the formula given in @exr-ridge-coverage to check your answer to @exr-ridge-bias-se.
2. Identify the plot of the sampling distribution of $\hat\mu$ for $\lambda=10$ when $n=100$ in @exr-ridge-ec, then use that to check your answer to @exr-ridge-coverage.

:::


# Problem 4

```{r}
m0x.mild     = rep(c(10,12,14,16,18), rev(c(80, 90, 100, 110, 120)))
m0x.moderate = rep(c(10,12,14,16,18), rev(c(50, 75, 100, 125, 150)))
m0x.extreme  = rep(c(10,12,14,16,18), rev(c(2,49, 100, 149, 198)))
m0x.flat = rep(c(10,12,14,16,18), c(100,100,100, 100,100))
m1x.mild     = rep(c(10,12,14,16,18), c(80, 90, 100, 110, 120))
m1x.moderate = rep(c(10,12,14,16,18), c(50, 75, 100, 125, 150))
m1x.extreme  = rep(c(10,12,14,16,18), c(2,49, 100, 149, 198))
m1x.flat = rep(c(10,12,14,16,18), c(100,100,100, 100,100))

exercise.pop = function(m0x, m1x, mu, sigma=\(w,x)mu(w,x)/4 ) {
  eps0 = rnorm(length(m0x))*sigma(0,m0x)
  eps1 = rnorm(length(m1x))*sigma(1,m1x)
  for(uu in unique(m1x)) { 
    eps0[m0x==uu] = eps0[m0x==uu] - mean(eps0[m0x==uu])
    eps1[m1x==uu] = eps1[m1x==uu] - mean(eps1[m1x==uu])
  }
  rbind(
    data.frame(w=0, x=m0x, y=mu(0,m0x)+eps0),
    data.frame(w=1, x=m1x, y=mu(1,m1x)+eps1))
}
```


```{r}
default.shift = .25
proportion.scale = 25
plotX = function(w,x, shift=default.shift) { ifelse(w==0, x-shift, x+shift) }
interleaved.bars = list(
  geom_bar(aes(x=plotX(w,x), y=proportion.scale*after_stat(prop), 
               fill=factor(w)), 
           position='identity', alpha=.1, linewidth=.1, color='black'))
purple.bars = geom_bar(aes(x=x, y=proportion.scale*after_stat(prop)), 
                       alpha=.1, linewidth=.1, fill='purple', color='black', width=.4)
scales = list(
  scale_y_continuous(
    breaks=seq(0,20,by=2.5),
    labels=function(x) sprintf("%1.1f", x),
    sec.axis=sec_axis(~./proportion.scale, name='proportion')), 
  coord_cartesian(ylim=c(0,20)),
  guides(color='none', fill='none', size='none') 
)
```

```{r}
lsq.plot = function(sam, formula=NULL, shift=default.shift, line.alpha=.7, size.scale=.1) {
  sam$weights = rep(1, nrow(sam))
  max.weight = max(sam$weights)
  max.size = size.scale * max.weight 
  
  base.plot = ggplot(sam) + 
    geom_point(aes(x=plotX(w,x), y=y,color=factor(w), size=I(weights)), 
      position=position_jitter(seed=0, width=shift),
      alpha=.2) + 
    stat_summary(aes(x=plotX(w,x), y=y, fill=factor(w)), 
      fun.data=mean_sd, geom='pointrange', 
      shape=21, alpha=.5, color='black') +
    interleaved.bars + purple.bars + scales
    
    if(is.null(formula)) { base.plot } else { 
        model = lm(formula, data=sam, weights=weights)
        gridx = seq(-10, 50, length.out=1000)
        muhat = function(w,x) predict(model, newdata=data.frame(w=w,x=x))   
        
        the.plot = base.plot 
        if(line.alpha > 0) { 
            the.plot = base.plot + 
            geom_line(aes(x=plotX(w,x), y=muhat(w,x), color=factor(w)), 
              data = expand_grid(w=0:1, x=gridx), alpha=line.alpha, linetype='dashed') 
        }
        the.plot + 
          geom_point(aes(x=plotX(w,x), y=muhat(w,x), fill=factor(w)),  
            color='black', alpha=.5, shape=23)
    }
}
```


```{r}
Delta1.target = function(pop=NULL) {
  function(mu, sam) { 
    if(is.null(pop)) { pop = sam }
    mean(mu(1,pop$x[pop$w==1]) - mu(0,pop$x[pop$w==1]))
  }
}
with.constant.weights = function(sam) { 
  sam$weights = 1  
  sam
}
with.ipw = function(sam, from=sam) { 
  summaries = from |> group_by(w,x) |> summarize(Nwx=n(), .groups='drop')
  Nwx = summary.lookup('Nwx', summaries)
  sam$weights = Nwx(1,sam$x) / Nwx(sam$w,sam$x)
  sam 
} 

plot.sampling.distribution = function(pop, formula, n = nrow(sam), 
target=Delta1.target(), pop.version=FALSE, replications=5000,
with.weights=with.constant.weights, seed=1) { 
    estimates = with_seed(function() { 
      1:replications |> map_vec(function(rep) { 
        J = sample(1:nrow(pop), size=n, replace=TRUE)
        sam = pop[J,] |> with.weights() 

        fitted.model = lm(formula, data=sam, weights=weights) 
        muhat = function(w,x) { predict(fitted.model, newdata=data.frame(w=w, x=x)) } 
        target(muhat, sam)
    })}, seed=seed)
    
    mu.model = lm(y~w*factor(x), data=pop)
    mu = function(w,x) { predict(mu.model, newdata=data.frame(w=w, x=x)) }
    the.target = target(mu, pop)
    

    mutilde.model = lm(formula, data=pop |> with.weights(), weights=weights)
    mutilde = function(w,x) { predict(mutilde.model, newdata=data.frame(w=w, x=x)) }
    the.pop.version = target(mutilde, pop)
    
    ggplot() +
     with_annotations(estimates, estimand=the.target, dots=0) +  
     geom_histogram(aes(x=estimates, y=after_stat(density)), bins=50, alpha=.2)
}
```

```{r}
#| fig-width: 5
#| layout-ncol: 2
#| fig-cap:
#|  - Population A  
#|  - Population B
set.seed(0)
mu = function(w,x) 3 + (x/(8-2*w))^2
population.A = exercise.pop(m1x.moderate, m1x.moderate, mu)
population.B = exercise.pop(m0x.flat, m1x.moderate, mu)

lsq.plot(population.A)
lsq.plot(population.B)
```

The questions below refer to the following estimation targets:
$$
\begin{aligned}
\Delta_{\text{raw}} &= \frac{1}{m_1}\sum_{j:w_j=1} y_j - \frac{1}{m_0}\sum_{j:w_j=0} y_j \qfor m_w = \sum_{j:w_j=w} 1 \\
\Delta_{0}   &= \frac{1}{m_0}\sum_{j:w_j=0} \lcb{ \mu(1,x_j) - \mu(0,x_j) }  \\
\Delta_{1}   &= \frac{1}{m_1}\sum_{j:w_j=1} \lcb{ \mu(1,x_j) - \mu(0,x_j) } \\ 
\Delta_{\text{all}}   &= \frac{1}{m}\sum_{j=1}^m \lcb{ \mu(1,x_j) - \mu(0,x_j) }  
\end{aligned}
$$

And the following estimator of $\Delta_1$ based on the least squares predictor in some linear model $\model$.

$$
\begin{aligned}
\hat\Delta_1 &= \frac{1}{N_1}\sum_{i:W_i=1} \lcb{ \hat\mu(1,X_i) - \hat\mu(0,X_i) } \qfor N_w = \sum_{i:W_i=w} 1 \\
& \qqtext{ where } \hat\mu = \argmin_{m \in \model} \sum_{i=1}^n \lcb{ Y_i - m(W_i, X_i) }^2 
\end{aligned}
$$ {#eq-Deltahat-1}


::: {#exr-4a .callout-exercise}
For the the two populations plotted above (A and B), answer the following questions.

1. Is $\Delta_1 > \Delta_\text{raw}$? 
2. Is $\Delta_1 > \Delta_0$?
3. Is $\Delta_1 > \Delta_\text{all}$?

I'm not asking you to compare the magnitudes of these estimation targets. If $\Delta_1$ is positive and $\Delta_\text{raw}$ is negative, then no matter what the magnitudes are, $\Delta_1 > \Delta_\text{raw}$.
:::



::: {#exr-4b .callout-exercise}
Consider the estimator $\hat\Delta_1$ we get when we use the *parallel lines model* in @eq-Deltahat-1.

$$
\model=\qty{ m(w,x) = a(w) + bx \ \text{ for all functions $a(w)$ and numbers $b$} \ }
$$

Below, in some order, I've plotted the sampling distribution of $\hat\Delta_1$ for samples of size $n=500$ drawn with replacement from Populations A and from Population B. 

1. True or False: the plot on the *left* shows the sampling distribution of $\Delta_1$ for Population A. 
2. Explain how you know. 

:::

```{r}
#| fig-width: 5
#| layout-ncol: 2

scales = list(
  scale_y_continuous(breaks=seq(0,2,    by=.5),   labels=function(x) sprintf("%1.1f", x)),
  scale_x_continuous(breaks=seq(1.5,3.5,by=.5),   labels=function(x) sprintf("%1.1f", x)),
  coord_cartesian(ylim=c(0,2.2), xlim=c(1.4,3.6)))

population.B |>
  plot.sampling.distribution(y~w+x, n=500) + scales
population.A |>
  plot.sampling.distribution(y~w+x, n=500) + scales
```



::: {#exr-4c .callout-exercise}
*Extra Credit*.


Consider the estimators $\hat\Delta_1$ we get when we use the *lines model* and *additive model* in @eq-Deltahat-1.
$$
\begin{aligned}
\model &= \{ m(w,x) = a(w) + b(w)x  \ \text{ for all functions $a(w)$ and $b(w)$ } \} && \text{lines model} \\
\model &= \{ m(w,x) = a(w) + b(x) \ \text{ for all functions $a(w)$ and $b(x)$} \} && \text{additive model}
\end{aligned}
$$

Below, in some order, I've plotted the sampling distribution of $\hat\Delta_1$ for these two estimators in
samples of size $n=500$ drawn with replacement from Population B. 

1. True or False: the plot on the *left* shows the sampling distribution we get when we use the *additive model*. 
2. Explain how you know.

:::

```{r}
#| fig-width: 5
#| layout-ncol: 2
population.B |>
  plot.sampling.distribution(y~w*x, n=500) + scales
population.B |>
  plot.sampling.distribution(y~w+factor(x), n=500) + scales
```

