---
title: "Midterm 1"
subtitle: "QTM 220"
---


```{r}
#| echo: false
library(tidyverse) 
library(scales)

theme_set(theme_minimal())
options(scipen = 1, digits = 2)
```

::: hidden
$$
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\Var}{V}
\DeclareMathOperator{\hVar}{\widehat{V}}
\DeclareMathOperator{\bias}{bias}
\DeclareMathOperator*{argmin}{argmin}
\newcommand{\model}{\mathcal{M}}
$$
:::

::: {.callout-warning}
I got a lot of answers on Problem 1 that explain what the code calculates in terms of the vector $Y$, not in terms of the sample or the population it's drawn from. In later exams, esp. the final, I tried to fix that. Let's try to do that here. 

And maybe let's give the estimator $\tilde{Y}$ from Problem 3 a bit more story behind it, e.g. dividing by $n+2$ with the intention of 
reducing variance instead of $n-2$.
:::

# Problem 1
Suppose we have a sample $Y_1 \ldots Y_n$ stored on an \textbf{R} vector $Y$ of length $n$.
We run this code.

```{r}
set.seed(123)
n = 1000
Y = rnorm(n)
```

```{r}
#| echo: true
B=10000
Y.bar = rep(NA,B)
for(b in 1:B) {
  Y.bar[b]= mean(sample(Y,size=n,replace=TRUE))
}
```

```{r}
#| out-height: 6in
#| fig-ncol: 2
#| fig-subcap:
#|  -  Histogram of the vector Y's elements
#|  -  Histogram of the Y.bar's elements
ggplot() + geom_histogram(aes(x=Y,     y=after_stat(density)), alpha=.3) + 
           geom_vline(aes(xintercept=mean(Y)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(Y)+c(-1,1)*sd(Y)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(Y), xmin=mean(Y)-1.96*sd(Y), xmax=mean(Y)+1.96*sd(Y), y=.15)) + theme_bw() + xlab('') + ylab('')
ggplot() + geom_histogram(aes(x=Y.bar, y=after_stat(density)), alpha=.3) + 
           geom_vline(aes(xintercept=mean(Y.bar)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(Y.bar)+c(-1,1)*sd(Y.bar)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(Y.bar), xmin=mean(Y.bar)-1.96*sd(Y)/sqrt(n), xmax=mean(Y.bar)+1.96*sd(Y)/sqrt(n), y=5)) + theme_bw() + xlab('') + ylab('')
```

::: {.callout-exercise}
### Part A
What does this code calculate? State it in words and mark it with an 'a', and an arrow or bracket
if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
mean(Y)
```

:::
::: {.callout-solution}
The sample mean. It's indicated by the solid vertical line on the left histogram.
:::

::: {.callout-exercise}
### Part B
What does this code calculate? State it in words and mark it with a 'b', and an arrow or bracket if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
sd(Y)
```

:::
::: {.callout-solution}
The sample standard deviation. It's the distance between the solid vertical and either dashed vertical line on the left histogram.
:::

::: {.callout-exercise}
### Part C
What does this code calculate? State it in words and mark it with an 'c', and an arrow or bracket if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
sd(Y.bar)
```

:::
::: {.callout-solution}
The bootstrap sampling distribution's standard deviation. 
It's the distance between the solid vertical and either dashed vertical line on the left histogram.
:::

::: {.callout-exercise}
### Part D
What does this code calculate? State it in words and mark it with an 'd', and an arrow or bracket if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
mean(Y) - 1.96* sd(Y)/sqrt(n)
mean(Y) + 1.96* sd(Y)/sqrt(n)
```

:::
::: {.callout-solution}
A 95\% confidence interval --- based on normal approximation --- for the mean of the population our sample $Y_1 \ldots Y_n$ is drawn from. This is indicated by the horizontal segment (with a dot in the middle) on the right histogram.
:::

::: {.callout-exercise}
### Part E
How could you calculate roughly the same thing as in part d using `sd(Y.bar)` and not `sd(Y)`?
:::
::: {.callout-solution}
We can use the standard deviation of the bootstrap sampling distribution, rather than `sd(Y)/sqrt(n)`, as our estimate of standard error.
```{r}
#| echo: true
#| eval: false
mean(Y) - 1.96* sd(Y.bar)
mean(Y) + 1.96* sd(Y.bar)
```

:::

# Problem 2
In 2015, Berkeley implemented a penny-per-ounce tax on soda. Advocates claimed the tax would reduce the rate of diabetes in the city's population. Now we want to check this claim---in a sense. What we want to know is the 
difference between the rate of diabetes among residents of Berkeley --- one city in the San Francisco Bay Area --- to the rate among residents of the Bay Area as a whole. 

What we can use to do this is a sample, drawn with replacement, from the population of Bay Area Residents. To be precise, suppose the Bay Area has $m$ residents, each with an id $j \in 1 \ldots m$. Our population is $(x_1,y_1) \ldots (x_m, y_m)$ 
where 
$$
\begin{aligned}
x_j &= \begin{cases} 1 & \ \text{ if Bay Area resident $j$ lives in Berkeley} \\  0 & \ \text{ otherwise} \end{cases} \\
y_j &= \begin{cases} 1 & \ \text{ if Bay Area resident $j$ has diabetes} \\ 0 & \ \text{ otherwise} \end{cases}
\end{aligned}
$$
We have a sample $(X_1, Y_1) \ldots (X_n, Y_n)$ drawn with replacement from this population.


::: {.callout-exercise}
### Part A
Express your estimand --- the population summary you want to estimate--- in terms of $(x_1, y_1) \ldots (x_m,y_m)$.
:::
::: {.callout-solution}
$$
\frac{\sum_{j=1}^m x_j y_j}{\sum_{j=1}^m x_j} - \frac{\sum_{j=1}^m y_j}{m}
$$
:::

::: {.callout-exercise}
### Part B
Express that estimand, using conditional and unconditional expectations, in terms of a single pair of random variables $(X_i,Y_i)$ drawn from this population.
:::
::: {.callout-solution}
$$ 
\E[Y_i \mid X_i=1] - \E[Y_i] 
$$
:::

::: {.callout-exercise}
### Part C
In terms of $\overline{X}=\frac1n\sum_{i=1}^n X_i$, $\overline{Y}=\frac1n\sum_{i=1}^n Y_i$, and $\overline{XY}=\frac1n\sum_{i=1}^n X_i Y_i$, state an *unbiased point estimate*.
:::
::: {.callout-solution}
$$
\frac{\sum_{i=1}^n X_i Y_i}{\sum_{i=1}^n X_i} - \frac{\sum_{i=1}^n Y_i}{n} = \overline{XY}/\overline{X} - \overline{Y}
$$
:::


# Problem 3

 Suppose you have a sample $Y_1, Y_2, \ldots, Y_n$ drawn with replacement from a population with mean $\mu$ and standard deviation $\sigma$. 
 
 
 Consider the following estimator for $\mu$: $\tilde{Y} = \frac{1}{n-2} \sum_{i=1}^n Y_i$.

::: {.callout-exercise}
### Part A
 What is $E[\tilde{Y}]$?
:::
::: {.callout-solution}
$$
\begin{aligned}
 \E\qty[\tilde{Y}] &= E\qty[ \frac{1}{n-2} \sum_{i=1}^n Y_i] \\
 &= \frac{1}{n-2} E\qty[ \sum_{i=1}^n Y_i]\\
 &= \frac{1}{n-2} \qty(\sum_{i=1}^n E[Y_i])\\
 &=\frac{1}{n-2} \qty(\sum_{i=1}^n \mu )\\
 &=\frac{1}{n-2} n\mu \\
 &= \frac{n}{n-2}\mu
\end{aligned}
$$
:::

::: {.callout-exercise}
### Part B
What is the estimator's bias, $E[\tilde{Y}] - \mu$?
:::
::: {.callout-solution}
$$
\begin{aligned}
 \E[\tilde{Y} - \mu] &= \frac{n}{n-2}\mu - \mu \\
 &= \qty(\frac{n}{n-2} - 1)\mu \\
 &= \qty(\frac{n}{n-2} - \frac{n-2}{n-2})\mu \\
 &= \frac{2}{n-2}\mu
\end{aligned}
$$
:::

::: {.callout-exercise}
### Part C
Showing your work, calculate $Var[\tilde{Y}]$.
:::
::: {.callout-solution}
 $$
 \begin{aligned}
 \Var\qty[\tilde{Y}] &= \Var\qty[ \frac{1}{n-2} \displaystyle \sum_{i=1}^n Y_i] \\
 &= \frac{1}{(n-2)^2} \Var[ \sum_{i=1}^n Y_i]\\
 &= \frac{1}{(n-2)^2} \qty(\sum_{i=1}^n V[Y_i])\\
 &=\frac{1}{(n-2)^2} \qty(\sum_{i=1}^n \sigma^2 )\\
 &=\frac{1}{(n-2)^2} n\sigma^2 \\
 &= \frac{n\sigma^2}{(n-2)^2}
\end{aligned}
$$
:::

::: {.callout-exercise}
### Part D
The plot below shows the sampling distributions of three estimators at three sample sizes. 
The vertical line indicates the mean of the sampling distribution. Which of these estimators is
$\tilde{Y}$? Is it a, b, or c?

```{r}
#| echo: false
#| out-width: 6.5in 
#| fig-asp: 1
set.seed(1234)
mu=4
sigma = 1

error = 2*rnorm(100000)
bias = list(a=function(n) 0, b=function(n) 2*mu/(n-2), c=function(n) 2*mu/(40-2))
se = function(n) { sqrt(n*sigma^2 / (n-2)^2) }

plotdata = expand_grid(estimator=names(bias), n=c(10,20,40)) %>%
  pmap(function(estimator, n) {
    thebias = bias[[estimator]](n)
    data.frame(estimator=estimator, n=n, muhat=thebias + error * se(n))
    }) %>% bind_rows() 
summaries = plotdata %>% group_by(n, estimator) %>% summarize(mean=mean(muhat))

breaks = seq(-4, 4, by=2)/2 
signs  = case_when(
  breaks < 0 ~ '-', 
  breaks == 0 ~ '',
  breaks > 0 ~ '+')
offsets = case_when(
  breaks != 0 ~ sprintf('%d', abs(breaks)),
  breaks == 0 ~ '')
labels.abstract = sprintf('μ %s %s', signs, offsets)

ggplot(plotdata) + geom_histogram(aes(x=muhat, y=after_stat(density)), bins=100, alpha=.4) + 
                   geom_vline(aes(xintercept=mean), data=summaries) + 
                   scale_x_continuous(breaks=breaks, labels=labels.abstract, limits=range(breaks)) + 
                   facet_grid(n ~ estimator, labeller=labeller(n=label_both, estimator=label_both)) + 
                   theme_bw() + theme(axis.text.x = element_text(angle = -90, vjust = 0.5, hjust=1)) + 
                   xlab('') + ylab('')
```

:::
::: {.callout-solution}
It's b. $a$ is unbiased and $c$'s bias doesn't decrease with $n$.
:::

# Problem 4

Suppose a friend has polled 100 of Georgia's registered voters
about their intention to vote in the 2024 election. The interval
estimate they report for turnout is [.59,.61]. 

::: {.callout-exercise}
Could this interval estimate have 95\% coverage?
Explain your answer. \emph{Hint. If $Y$ is a binary random variable
that takes on the value 1 with probability $p$ (and 0 with probability $1-p$), $\Var[Y]=p(1-p)$.}
:::

::: {.callout-solution}
No. The standard error of the mean of $100$ binary random variables
 $Y_i$ with $\E Y_i=p$ is $\sqrt{p(1-p)/100}$. Because we know that $p\approx 6/10$, this is roughly $(6/10)(4/10)/100=\sqrt{24}/100 \approx 5/100$. 
 As a result, a 95\% confidence interval should have a width of roughly
 $2 \times 1.96 \times 5/100 \approx 20/100=.2$. The width of their interval is 
 $1/10$ of this.
:::
