---
title: "Midterm 1"
subtitle: "QTM 285-1"
---


```{r}
#| echo: false
library(tidyverse) 
library(scales)

theme_set(theme_minimal())
options(scipen = 1, digits = 2)
```

::: hidden
$$
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\Var}{V}
\DeclareMathOperator{\hVar}{\widehat{V}}
\DeclareMathOperator{\bias}{bias}
\DeclareMathOperator*{argmin}{argmin}
\newcommand{\model}{\mathcal{M}}
$$
:::


# Problem 1
Suppose we've drawn a sample $Y_1 \ldots Y_n$ with replacement from a population $y_1 \ldots y_m$ with mean $\theta=\frac{1}{m}\sum_{j=1}^m y_j$. 
And we've stored it in an **R** vector $Y$ of length $n$. What I'm calling $Y_i$ in mathematical notation is  `Y[i]` in **R**. We run this code.

```{r}
set.seed(123)
m = 10000000 
y = rnorm(m) + (6*rbinom(m, size=1, prob=.5)-3)

n = 10000
J = sample(1:m, size=n, replace=TRUE)
Y = y[J]
```

```{r}
#| echo: true
#| cache: false
theta.hat.star = 1:10000 |> map_vec(function(.) { 
  I = sample(1:n, size=n, replace=TRUE)
  Y.star = Y[I]
  mean(Y.star)
}) 
```


```{r} 
#| fig-ncol: 2
#| fig-width: 5.5
#| fig-subcap:
#|  -  Histogram of the vector Y's elements
#|  -  Histogram of the vector theta.hat.star's elements
ggplot() + geom_histogram(aes(x=Y, y=after_stat(density)), 
bins=50, alpha=.3) + 
           geom_vline(aes(xintercept=mean(Y)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(Y)+c(-1,1)*sd(Y)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(Y), xmin=mean(Y)-1.96*sd(Y), xmax=mean(Y)+1.96*sd(Y), y=.05)) + theme_bw() + xlab('') + ylab('') +
           scale_x_continuous(breaks=seq(-8,8,by=2)) +
           scale_y_continuous(breaks=seq(0,.20,by=.05)) + 
           coord_cartesian(xlim=c(-8,8), ylim=c(0,.20))

ggplot() + geom_histogram(aes(x=theta.hat.star, y=after_stat(density)), 
                           bins=50, alpha=.3) + 
           geom_vline(aes(xintercept=mean(theta.hat.star)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(theta.hat.star)+c(-1,1)*sd(theta.hat.star)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(theta.hat.star), xmin=mean(theta.hat.star)-1.96*sd(theta.hat.star), xmax=mean(theta.hat.star)+1.96*sd(theta.hat.star), y=5)) + theme_bw() + xlab('') + ylab('') + 
           scale_x_continuous(breaks=seq(-.08,.12,by=.02)) +
           scale_y_continuous(breaks=seq(0,20,by=5)) + 
           coord_cartesian(xlim=c(-.08,.12), ylim=c(0,20))
```

::: {.callout-note}
### What Your Answers Should Look Like
When you answer the 'What does this code calculate?' questions below, 
please phrase your answer in statistical terms. Say things like 
'the sample _'or 'the _ sampling distribution' or 'a _ based on _ approximation'
rather than 'the _ of the vector `Y`'.
:::

::: {.callout-exercise}
### Part A
What does this code calculate? State it in words and mark it with an 'A', and an arrow or bracket
if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
mean(Y)
```

:::
::: {.callout-solution}
The sample mean. It's indicated by the solid vertical line on the left histogram.
:::

::: {.callout-exercise}
### Part B
What does this code calculate? State it in words and mark it with a 'B', and an arrow or bracket if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
sd(Y)
```

:::
::: {.callout-solution}
The sample standard deviation. It's the distance between the solid vertical and either dashed vertical line on the left histogram.
:::

::: {.callout-exercise}
### Part C
What does this code calculate? State it in words and mark it with an 'C', and an arrow or bracket if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
sd(theta.hat.star)
```

:::
::: {.callout-solution}
The bootstrap sampling distribution's standard deviation. 
It's the distance between the solid vertical and either dashed vertical line on the left histogram.
:::

::: {.callout-exercise}
### Part D
What does this code calculate? State it in words and mark it with an 'D', and an arrow or bracket if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
mean(Y) - 1.96* sd(Y)/sqrt(n)
mean(Y) + 1.96* sd(Y)/sqrt(n)
```

:::
::: {.callout-solution}
A 95\% confidence interval --- based on normal approximation --- for the mean of the population our sample $Y_1 \ldots Y_n$ is drawn from. This is indicated by the horizontal segment (with a dot in the middle) on the right histogram.
:::

::: {.callout-exercise}
### Part E
How could you calculate roughly the same thing as in Part D, if you had `mean(Y)` and `theta.hat.star` but not `Y` itself?
:::
::: {.callout-solution}
We can use the standard deviation of the bootstrap sampling distribution, rather than `sd(Y)/sqrt(n)`, as our estimate of standard error.
```{r}
#| echo: true
#| eval: false
mean(Y) - 1.96* sd(theta.hat.star)
mean(Y) + 1.96* sd(theta.hat.star)
```

:::

# Problem 2 

Suppose that you've drawn a sample $Y_1 \ldots Y_{n}$ of size $n=998$ with replacement from a binary population $y_1 \ldots y_{m}$
of size $m=1332$. You want a 95% confidence interval for $\theta$, the proportion of 1s in the population, but you don't want to do the 
calculations yourself, so you ask a classmate to do it. You tell them the sample proportion $\hat\theta$ is $1/3$ and the 
sample and population sizes, but you forget to tell them you sampled with replacement, so they do the calculation as if you sampled *without replacement*. 

Here's a plot that's pretty informative about what's going on. On the left is the Binomial sampling distribution estimate you'd have used for calibration if we'd done it yourself. On the right is the Hypergeometric estimate your friend used for calibration. I've drawn the interval your friend gave you on top of both. I've also drawn dotted vertical lines at the endpoints of the interval.

```{r}
#| echo: false
#| fig-width: 5.5
#| layout-ncol: 2
#| fig-cap: 
#|  - The Binomial distribution for n trials with success probability 1/3   
#|  - The Hypergeometric distribution for n trials with success probability 1/3 in a population of size m

n = 998
m = 1332
theta.hat = 1/3 
x = (0:998)/998
xx = seq(0,1,by=.001)

scales = list(scale_x_continuous(breaks=seq(0,1,by=1/10)),
              scale_y_continuous(breaks=seq(0,1,by=1/100)), 
              coord_cartesian(xlim=c(.25,.4), ylim=c(0,.06)),
              labs(x='', y=''))
sigma.hat.with = sqrt(theta.hat*(1-theta.hat)/n)
sigma.hat.without = sqrt(((m-n)/(m-1))*theta.hat*(1-theta.hat)/n)

ggplot() + geom_col(aes(x=x, y=dbinom(x*n, size=n, prob=theta.hat)), alpha=.5) + 
           geom_pointrange(aes(x=theta.hat, xmin=theta.hat-2*sigma.hat.without, xmax=theta.hat+2*sigma.hat.without, y=.01)) +
           geom_vline(xintercept=c(theta.hat-2*sigma.hat.without, theta.hat+2*sigma.hat.without), linetype='dotted') + 
           geom_vline(xintercept=c(theta.hat), alpha=.2) + 
           scales
ggplot() + geom_col(aes(x=x, y=dhyper(x*n, m*theta.hat, m*(1-theta.hat), n)), alpha=.5) + 
           geom_pointrange(aes(x=theta.hat, xmin=theta.hat-2*sigma.hat.without, xmax=theta.hat+2*sigma.hat.without, y=.01)) +
           geom_vline(xintercept=c(theta.hat-2*sigma.hat.without, theta.hat+2*sigma.hat.without), linetype='dotted') + 
           geom_vline(xintercept=c(theta.hat), alpha=.2) + 
           scales

```

This tells you a lot. But calibration isn't really about a single interval estimate. It's about an *interval estimator*: a function of the sample that gives you an interval estimate. So let's define the one we're using. Let $w_H(x)$ be the width your friend would have used if you'd told them the sample proportion was $x$---the width of the middle 95% of the sample-and-population-size-appropriate Hypergeometric distribution when $x$ is the proportion of 1s in the population. Your interval estimator is $\hat\theta \pm w_H(\hat\theta)/2$. 

If you want to calculate the coverage probability of this interval estimator *exactly*, you could. Or you could if you had a computer. The last extra credit problem on the calibration homework was about this. But to start with, let's make this simpler.

::: {.callout-exercise}
### Part A
Suppose the sampling distribution estimate I drew above on the left is *exactly the same shape* as the actual sampling distribution of the sample proportion $\hat\theta$, like I just picked the actual sampling distribution up from where it was centered---at the population proportion $\theta$---and set it down centered at $\hat\theta$. And suppose that our friend wasn't really using the estimate we gave them to calculate the width of the interval. No matter what we told them about $\hat \theta$, they'd have given us an interval $\hat\theta \pm w_H/2$ where $w_H$ is the interval width we see in the plots above. If all that were true, roughly what is the coverage probability of the interval estimator $\hat\theta \pm w_H/2$: 5%, 33%, 50%, 66%, or 95%?

*Hint.* If you're not sure because you're having trouble 'measuring the picture' accurately, here's a tip that may help. If you want to compare the area in the middle of a histogram to the area at the edges, think about the edges as two identical triangles. The combined area of the two triangles is the area of *a rectangle* with the same width and height as one of the triangles---the rectangle we get by rotating one of the triangles 180 degrees and setting it on top of the other. Compare the area of that rectangle to the area of the middle of the histogram. You've got 95% coverage if the ratio is 5/95=1/19, so you can fit 19 copies of the rectangle in the middle of the histogram. You've got 66% coverage if it's $1/2$, so you can fit 2 copies of the rectangle in the middle of the histogram. And you've got 50% coverage if you can fit in one copy of the rectangle.
:::

Now let's get back to reality. Your interval estimate really is $\hat\theta \pm w_H(\hat\theta)/2$ where $w_H(x)$ is as it is described above Part A.
And what's drawn above on the left really is the Binomial distribution for $n$ trials with success probability $\hat\theta$, not a picked-up-and-moved version of the Binomial for success probability $\theta$. To help make sense of what these differences mean, I've provided you with a new version of a familiar plot. It's the plot we used, at the end of our first lecture on real-world calibration, to argue that we could get away with calibrating an interval estimate around $\hat\theta$ as if its sampling distribution where Binomial with success probability $\hat\theta$ instead of $\theta$ itself.

Here is the plot. It shows the Binomial distribution for success probability $x$ and the corresponding version of our interval estimate, $x \pm w_H(x)/2$, at 3 values of $x$: $x=\hat \theta$ and $x=\hat\theta \pm 1.96\sqrt{.5^2/n}$. On the left, we see those things where they are. On the right, they're shifted so they're centered at zero.
```{r}
#| echo: false
#| layout-ncol: 2
#| fig-width: 5.5

alpha = .003
K = 3
binom.width = function(p, n, alpha=alpha) {
  2*qnorm(1-alpha/2)*sqrt(p*(1-p)/n)
}
hyper.width = function(p, n, m, alpha=alpha) {
  2*qnorm(1-alpha/2)*sqrt(p*(1-p)*(m-n)/(m-1)/n)
}
height = (1/2)*max(dhyper(0:n, m*theta.hat, m*(1-theta.hat), n))

thetas = c(theta.hat-1.96*sqrt(.25/n), theta.hat, theta.hat+1.96*sqrt(.25/n))
density.stability.plot = ggplot() + 
           geom_line(aes(x=x/n, y=dbinom(x,n,p), color=factor(p)), data=expand.grid(p=thetas, x=0:n), alpha=.5) +
           geom_pointrange(aes(x=p, xmin=p-w/2, xmax=p+w/2, y=y, color=factor(p)), data=data.frame(p=thetas, w=hyper.width(thetas, n, m,alpha=.05), y=c(.3,.375,.45)*height)) +
           scales + guides(color=FALSE) + labs(x='',y='')

centered.scales = list(
  scale_x_continuous(breaks=seq(-1,1,by=1/10)),
  scale_y_continuous(breaks=seq(0,1,by=1/100)), 
  coord_cartesian(xlim=c(-.075,.075), ylim=c(0,.06)),
  labs(x='', y=''))
centered.density.stability.plot = ggplot() + 
           geom_line(aes(x=x/n-p, y=dbinom(x,n,p), color=factor(p)), data=expand.grid(p=thetas, x=0:n), alpha=.5) +
           geom_pointrange(aes(x=0, xmin=-w/2, xmax=w/2, y=y, color=factor(p)), data=data.frame(p=thetas, w=hyper.width(thetas, n, m, alpha=.05), y=c(.3,.375,.45)*height)) + 
           centered.scales + guides(color=FALSE) + labs(x='',y='')
           
density.stability.plot
centered.density.stability.plot
```

::: {.callout-exercise}
Is your answer to Part A is a good approximation to the actual coverage of your interval estimate?  
In a sentence or two, referring to the plot above, explain why or why not. You may take as a given that $\theta \in \hat\theta \pm 1.96\sqrt{.5^2/n}$. 

*Extra credit*. In a sentence, explain why it's pretty likely that 
$\theta \in \hat\theta \pm 1.96\sqrt{.5^2/n}$. 
:::


# Problem 3

 Suppose you have a sample $Y_1 \ldots Y_n$ drawn with replacement from a population $y_1 \ldots y_m$ with mean $\theta$ and standard deviation $\sigma$. You're going to answer a few questions about a very simple estimator for $\theta$: $\hat\theta=Y_1$. 

::: {.callout-exercise}
### Part A
What is this estimator's expected value, $\E[\hat\theta]$?
:::

::: {.callout-exercise}
### Part B
What is this estimator's bias, $\E[\hat\theta]-\theta$?
:::

::: {.callout-exercise}
### Part C
What is this estimator's variance, $\Var[\hat\theta]$? 
:::

::: {.callout-exercise}
### Part D
The plot below shows the sampling distributions of three estimators at three sample sizes. 
The vertical line indicates the mean of the sampling distribution. Which of these is $\hat\theta$: estimator a, b, or c?

```{r}
#| echo: false
#| out-width: 6.5in 
#| fig-asp: 1
set.seed(1234)
mu = 4
sigma = 1

error = rnorm(100000)
bias = list(a=function(n) 0, b=function(n) 2*mu/(n-2), c=function(n)0)
se = list(a=function(n) sigma, b=function(n) { sqrt(n*sigma^2 / (n-2)^2) }, c=function(n) { sigma/sqrt(n) })

plotdata = expand_grid(estimator=names(bias), n=c(10,20,40)) %>%
  pmap(function(estimator, n) {
    thebias = bias[[estimator]](n)
    these = se[[estimator]](n)
    data.frame(estimator=estimator, n=n, muhat=thebias + error * these)
    }) %>% bind_rows() 
summaries = plotdata %>% group_by(n, estimator) %>% summarize(mean=mean(muhat))

breaks = seq(-4, 4, by=2)/2 
signs  = case_when(
  breaks < 0 ~ '-', 
  breaks == 0 ~ '',
  breaks > 0 ~ '+')
offsets = case_when(
  breaks != 0 ~ sprintf('%d', abs(breaks)),
  breaks == 0 ~ '')
labels.abstract = sprintf('θ %s %s', signs, offsets)

ggplot(plotdata) + geom_histogram(aes(x=muhat, y=after_stat(density)), bins=100, alpha=.4) + 
                   geom_vline(aes(xintercept=mean), data=summaries) + 
                   scale_x_continuous(breaks=breaks, labels=labels.abstract, limits=range(breaks)) + 
                   facet_grid(n ~ estimator, labeller=labeller(n=label_both, estimator=label_both)) + 
                   theme_bw() + theme(axis.text.x = element_text(angle = -90, vjust = 0.5, hjust=1)) + 
                   xlab('') + ylab('')
```

:::

::: {.callout-exercise}
### Part E 

Suppose $Y_1 \ldots Y_n$ is the sample from Problem 1. 
On one of the grids below, sketch the bootstrap sampling distribution of $\hat\theta$. 
Then indicate the mean and the mean &plusmn; 1 standard deviation on your sketch. 

I've given you two grids so that, if you've attempted a drawing in pen or have hit 
the limit of what your eraser can handle, you can cross out one grid and work on the other.
On the grids, I've left off the actual numbers labeling the 'ticks' on the x and y axes.
That way, you can choose the scale to make it as easy as possible to draw. When you 
have chosen a scale, please do label those ticks.

*Hint*. In more formal notation, the estimator we're talking about is the result of applying a function of $n$ variables that always gives you the first one, $\hat\theta(a_1 \ldots a_n) = a_1$,
to the sample $Y_1 \ldots Y_n$: $\hat\theta(Y_1\ldots Y_n) = Y_1$. Its bootstrap sampling distribution is the distribution of estimates we get when we apply this function to bootstrap samples, i.e., to
samples $Y_1^* \ldots Y_n^*$ drawn with replacement from the sample $Y_1 \ldots Y_n$.
:::


```{r} 
#| fig-ncol: 2
#| fig-width: 5.5
 
ggplot() + 
  geom_blank(aes(x=1, y=1)) + 
  scale_x_continuous(breaks=seq(-8,8,by=2)) +
  scale_y_continuous(breaks=seq(0,.20,by=.05)) + 
  coord_cartesian(xlim=c(-8,8), ylim=c(0,.20)) +
  theme(axis.text.x = element_blank(), axis.text.y = element_blank()) + 
  xlab('') + ylab('')

ggplot() + 
  geom_blank(aes(x=1, y=1)) +
  scale_x_continuous(breaks=seq(-.08,.12,by=.02)) +
  scale_y_continuous(breaks=seq(0,20,by=5)) + 
  coord_cartesian(xlim=c(-.08,.12), ylim=c(0,20)) +
  theme(axis.text.x = element_blank(), axis.text.y = element_blank()) + 
  xlab('') + ylab('')
```


# Problem 4

Suppose a friend has polled 100 of Georgia's registered voters
about their intention to vote in the 2024 election. The interval
estimate they report for turnout is [.58,.62]. 

::: {.callout-exercise}
### Part A
Could this interval estimate be (correctly) calibrated to have 95\% coverage? Explain your answer. 
:::

::: {.callout-solution}
No. The standard error of the mean of $100$ binary random variables
$Y_i$ with $\E[Y_i]=\theta$ is $\sqrt{p(1-p)/100}$. Because we know that $p\approx 6/10$, this is roughly $(6/10)(4/10)/100=\sqrt{24}/100 \approx 5/100$. 
As a result, a 95\% confidence interval should have a width of roughly
$2 \times 1.96 \times 5/100 \approx 20/100=.2$. The width of their interval is 
$1/10$ of this.
:::

::: {.callout-exercise}
### Part B
If they wanted an interval estimate that was accurate to within 1%, 
i.e. an interval of the form $\hat\theta \pm .01$, roughly how many voters should they have polled? 
Choose the closest number from this list: 100, 1000, 10000, 100000.
::: 

# Problem 5 (Extra Credit)

In this problem, we're going to be thinking about the bootstrap
in more formal terms than usual. Suppose we have an estimator $\hat\theta$ that
we can think of as the result of applying a function of $n$ variables, $\hat\theta(a_1 \ldots a_n)$,
to the sample $Y_1 \ldots Y_n$: $\hat\theta = \hat\theta(Y_1 \ldots Y_n)$. 
We usually calculate its bootstrap sampling distribution by histogramming 
bootstrap estimates  $\hat\theta^\star=\hat\theta(Y_1^\star \ldots Y_n^\star)$ 
that we get by applying the function $\hat\theta$ to bootstrap samples 
$Y_1^\star \ldots Y_n^\star$ drawn, with replacement, from our original sample $Y_1 \ldots Y_n$.
The thing we see when we do this is a *conditional distribution*. 
It's the distribution of  $\hat\theta^\star$ conditional on the original sample $Y_1 \ldots Y_n$. 

Really, what we see is a histogram approximation to it, but we can draw as many bootstrap
samples as we want, so we can draw a near-perfect histogram approximation if we want to.
Let's forget about the histogram stuff---I mention it just for the sake of familiarity--- 
and think about the conditional distribution we're using the histogram to visualize.

When we talk about the mean and variance of the bootstrap sampling distribution, 
what we're talking about are conditional expectations. They're 
'second stage expected values' if you think of a multi-stage sampling process 
where you first draw $Y_1 \ldots Y_n$ and then draw $Y_1^\star \ldots Y_n^\star$
which is, after all, what we're doing.

$$
\begin{aligned}
&\E[\hat\theta^\star \mid Y_1 \ldots Y_n]   && \qqtext{  is the bootstrap sampling distribution's mean.} \\
&\Var[\hat\theta^\star \mid Y_1 \ldots Y_n] && \qqtext{  is the bootstrap sampling distribution's variance.}
\end{aligned}
$$

Here we're going to going to talk about the bootstrap sampling distribution of the sample mean $\hat\theta=\frac1n \sum_{i=1}^n Y_i$.

::: {.callout-exercise}
### Part A
Whenever we draw the bootstrap sampling distribution of a sample mean $\hat\theta$, its mean $\E[\hat\theta^\star \mid Y_1 \ldots Y_n]$ always seems to be the sample mean $\hat\theta$ itself. 

Using what you know about the conditional distribution of 
$Y_1^\star \ldots Y_n^\star$ given $Y_1 \ldots Y_n$, calculate 
this conditional expectation and show that it's the sample mean $\hat\theta=\bar Y$.
:::

::: {.callout-exercise}
### Part B 
In our practice exam, we used the bootstrap sampling distribution's 
standard deviation as an estimate of the standard deviation of our 
actual sampling distribution. It seemed to work. 

Find a formula,
in terms of $Y_1 \ldots Y_n$, for the bootstrap sampling distribution's variance. Then explain why using the bootstrap variance is a reasonable alternative to using the usual formula: $\hat\sigma^2/n$ where $\hat\sigma^2=\frac{1}{n}\sum_{i=1}^n (Y_i-\bar Y)^2$ is the sample variance.
:::


::: {.callout-exercise}
### Part C 
Repeat Part A for the 'mean with prior observations', 
$\hat\theta=\frac{1}{n_0+n}\left(n_0\theta_0 + \sum_{i=1}^n Y_i\right)$. 
To keep your calculations compact, it may be helpful to write this 
as a simple weighted average of the sample mean and the constant $\theta_0$:
it's $a\theta_0 + b\bar Y$  where $a$, $b$, and $\theta_0$ are constants and $a+b=1$. 

If we think of our 'bootstrap estimation target' as $\theta^\star = \bar Y$, the mean in the 'population' our bootstrap sample is drawn from, what is this estimator's *conditional bias* 
$\E[\hat\theta^\star \mid Y_1 \ldots Y_n] - \theta^\star$? Suppose that, thinking of this as an estimate of your actual estimator's bias $\E[\hat\theta] -\theta$, you define a *bias-corrected* version of your estimator by subtracting this estimate from your estimator.

$$
 \hat\theta_{bc} = \hat\theta - \text{bias}^\star 
 \qfor
 \bias^\star = \E[\hat\theta^\star \mid Y_1 \ldots Y_n] - \theta^\star
$$

What's the bias of your bias-corrected estimator?
What about its variance? Is it larger or smaller than the variance of
your original estimator $\hat\theta$?

:::