---
title: "Practice Final"
subtitle: "QTM 285-1"
crossref: 
  chapters: true
---

{{< include shared-code.qmd >}}

```{r}
library(tidyverse)
theme_set(whitebg.theme())
```

### Summary
Whether you're in academia or industry, you'll run into people who are analyzing data badly. 
Often, what they're doing is both the established common practice and obviously wrong. In this practice exam,
I'm going to run you through some examples in which this is happening. I'm going to ask you how to do better;
explain what's going wrong; and, in some cases, deal with what happens when you've convinced someone to do better
but they haven't entirely understood you. Obviously, in real life not everything that happens can be explained
using the concepts we've learned in this class, and often the problems are a bit better-hidden than they are here. 
But it's surprising---at least to me---how often these things come up in a relatively obvious way. 

Aside from giving you some practice recognizing and dealing with this stuff, what I hope to do here is give you some examples to help you 
explain what's going on when these things happen. Often, if someone's making a mistake in a subtler form or a more complex problem, you can 
point out that the situation in analogous to one where the problem is much more obvious, as it is in many of these examples. 

I hope you find this useful.  For the exam, expect similar exercises with less text around them. Much less text, in the case of Problem 5 here.   

This is not graded, so exercises marked *Extra Credit* are not actually worth extra credit. 
But they're similar in character to what you might see as an extra credit exercise on the exam.


# Problem 1

```{r}
#| fig-width: 10
set.seed(1)
mu = function(x){.2 + x + .37*sinpi(x)}
n = c(10, 20, 40, 1000)
X = round(runif(max(n)), 1)
Y = mu(X) + .4*rnorm(max(n))
data = data.frame(X=X, Y=Y)

ggplot(data) + geom_point(aes(x=X, y=Y), position=position_jitter(w=.025,h=0), alpha=.3)  
```

Suppose we observe a sample of $(X_1, Y_1), \ldots, (X_n, Y_n)$ drawn with replacement from the population $(x_1,y_1) \ldots (x_m,y_m)$ plotted above and we're interested in estimating the subpopulation mean $\mu(0) = \sum_{j:x_j=0} y_j / \sum_{j:x_j=0} 1$.

To do this, we use least squares to fit a line to the data and then let its value $\hat\mu(0)$ at $x=0$ be our estimate. Below is a plot of its sampling distribution at sample size $n=13$. As usual, I've plotted its mean, mean plus or minus one standard deviation, and mean plus or minus two standard deviations as solid, dashed, and dotted blue lines respectively. I've also plotted the subpopulation mean $\mu(0)$ itself as a green line. 

You may assume that the standard deviation of the estimator at sample size $n$ is $\text{se}=\sigma/\sqrt{n}$ for some number $\sigma$ *that does not vary with sample size*, just like it would be if $\hat\mu(0)$ were the mean of a sample of size $n$.  



```{r} 
#| fig-width: 10
sampdist = replicate(10000, {
  samp = data[sample(1:nrow(data), 13, replace=TRUE), ]
  model = lm(Y~X, data=samp)
  c(coef(model)[1])
})
ggplot() + geom_histogram(aes(x=X, y=after_stat(density)), 
  data=data.frame(X=sampdist), bins=50, alpha=.3) +
  geom_vline(aes(xintercept=mu(0)), color='green', linewidth=1.5) +
  geom_vline(aes(xintercept=mean(sampdist)), color='blue') +
  geom_vline(aes(xintercept=mean(sampdist)+c(-1,1)*sd(sampdist)), color='blue', linetype='dashed') + 
  geom_vline(aes(xintercept=mean(sampdist)+c(-2,2)*sd(sampdist)), color='blue', linetype='dotted') 
```

::: {#exr-1a .callout-exercise} 
At which of these three samples sizes --- 25, 50, or 100 --- would you expect that interval $\hat\mu(0) \pm 1.96\ \text{se}$ to have a width of roughly $1/2$? 
:::

::: {.callout-solution}
50. 

Here's why. If standard error is $\sigma/\sqrt{n}$, increasing sample size by a factor of $s$ will decrease it by a factor of
$(\sigma/\sqrt{sn}) / (\sigma/\sqrt{n}) = \sqrt{n/(sn)} = 1/\sqrt{s}$. So, if we quadruple the sample size, the standard error will be halved. 
In this instance, we can see from the plot that the standard error at sample size $13$ is roughly $1$, so quadrupling that to roughly $50$ will give us a standard error of roughly $1/2$.
:::

::: {#exr-1b .callout-exercise}
At which of these three samples sizes --- 25, 50, or 100 --- would you expect the coverage of the usual 95\% confidence interval $\hat\mu(0) \pm 1.96\ \text{se}$ to be closest to 50\%?
:::
::: {.callout-solution}
50. 

Here's why. The bias of a least squares estimator is the error it makes when fit to the population and therefore does not depend on sample size. 
Our bias is roughly 1 standard error at sample size $n=13$ and therefore will be roughly 2 standard errors at sample size $n=50$. And we get roughly 50% coverage when the bias is roughly 2 standard errors.

You can see this by drawing the integral you'd use to do the coverage calculation. You do not, of course, actually have the sampling distribution for $n=50$ so you can't draw that exactly. But you know it should be centered in the same place as the one you do have, but about half as wide. I've drawn that below in red. For what it's worth, I've also drawn the actual sampling distribution for $n=50$ in blue. You'd integrate this sampling distribution from two standard errors left of the estimation target to two standard deviations right of it---i.e., over the region I've shaded yellow. And because that yellow region extends right to the center of the (roughly symmetric) sampling distribution and left far enough to include essentially all of its probability mass, this integral is roughly $.5$---half of the sampling distribution's total mass. 
```{r} 
sampdist.50 = replicate(1000, {
  samp = data[sample(1:nrow(data), 50, replace=TRUE), ]
  model = lm(Y~X, data=samp)
  c(coef(model)[1])
})
ggplot() + 
  geom_histogram(aes(x=X, y=after_stat(density)), 
    data=data.frame(X=sampdist.50), bins=50, alpha=.3, fill='red') +
  geom_histogram(aes(x=X, y=after_stat(density)), 
    data=data.frame(X=(1/2)*sampdist + (1/2)*mean(sampdist)), bins=50, alpha=.3, fill='blue') +
  geom_vline(aes(xintercept=mu(0)), color='black') +
  geom_vline(aes(xintercept=mean(sampdist.50)), color='blue') +
  geom_vline(aes(xintercept=mean(sampdist.50)+c(-1,1)*sd(sampdist.50)), color='blue', linetype='dashed') + 
  geom_vline(aes(xintercept=mean(sampdist.50)+c(-2,2)*sd(sampdist.50)), color='blue', linetype='dotted') +
  geom_area(aes(y=5, x=mu(0)+c(-2,2)*sd(sampdist.50)), fill='yellow', color='yellow', alpha=.3)
```

:::

::: {.callout-tip collapse="true"}
### Tip. Draw 
If the answers aren't obvious to you, it might be helpful to sketch the sampling distribution of the estimator at the different sample sizes
you're considering. How does their width vary as you increase sample size? What about their center?
:::

# Problem 2

You have a sample $Y_1,Y_2...,Y_n$ drawn with replacement from a population with mean $\mu$ and standard deviation $\sigma$.
The usual estimate for $\mu$ is the sample mean $\bar{Y}_0 = \frac{1}{n}\sum_{i=1}^n Y_i$. But one of your coworkers has a clever idea to reduce variance---instead of dividing by $n$, they divide by $n+10$. This is their estimator: $\bar{Y}_{10} = \frac{1}{n+10}\sum_{i=1}^n Y_i$. You try to convince them it's a bad idea by being a bit hyperbolic---you argue that if that's a good idea, then why not divide by $n+100$? They, unfortunately, *love* this idea.  You now have to explain why this is a bad idea. **There are some plots in Appendix @sec-appendix-plots that may help.**

Your coworker has calculated formulas for the bias and standard deviation of the 'divide by $n+k$ estimator' $\bar{Y}_k=\frac{1}{n+k}\sum_{i=1}^n Y_i$.
$$
\begin{aligned}
\E[\bar{Y}_k] - \mu &= -\frac{k}{n+k}\mu \\
\sd[\bar{Y}_k] &= \frac{\sigma}{\sqrt{n}} \times \frac{n}{n+k}.
\end{aligned}
$$

::: {.callout-note collapse="true"}
### Formula Derivation. Optional Reading.
$$
\begin{aligned}
 E[\bar{Y}_k] - \mu 
 &= \E[\frac{n}{n+k}\bar Y_0] - \mu \\
 &= \frac{n}{n+k}\E[\bar Y_0] - \mu \\
 &=\frac{n}{n+k}\mu - \mu \\
 &= \qty(\frac{n}{n+k} - 1)\mu \\
 &= \qty(\frac{n}{n+k} - \frac{n+k}{n+k})\mu \\
 &= -\frac{k}{n+k}\mu \\
 & \\
\Var[\tilde{Y}_k] 
 &= \Var[\frac{n}{n+k} \bar {Y}_0] \\
 &= \qty(\frac{n}{n+k})^2 V[\bar{Y}_0] \\
 &= \frac{n^2}{(n+k)^2} \frac{\sigma^2}{n} \\
 &= \frac{ n\sigma^2}{(n+k)^2}=\frac{\sigma^2}{n} \times \frac{n^2}{(n+k)^2}
 \end{aligned}
$$
:::

And they've drawn the sampling distribution of [$\bar{Y}_0$]{.pollb}, [$\bar{Y}_{10}$]{.pollc}, and [$\bar{Y}_{100}$]{.polld} (top, middle, and bottom rows respectively)
for sample sizes $n=100$ and $n=10,000$ (left and right columns respectively) *for $\mu=0$ and $\sigma=1$*. 
The solid line is the estimator's expected value $\E[\bar{Y}_k]$, 
the dashed lines are $\E[\bar Y_k] \pm \sd[\bar Y_k]$, and the dotted lines are $\E[\bar Y_k] \pm 2 \sd[\bar Y_k]$.
```{r manynormals}
#| layout-nrow: 3
#| layout-ncol: 2
#| fig-width: 6
 
x = seq(-2,2,by=.001)
se = function(n,k){1/sqrt(n) * n/(n+k)}

color = function(k) { 
  if(k==0) pollb else if(k==10) pollc else polld 
}
plot.normals = function(mu, n, k) {
  plots = map(k, function(k) {
    the.se   = se(n,k)
    the.bias = -(mu*k/sqrt(n+k)) * the.se
    list(
      geom_area(aes(x=x, y=dnorm(x, mean=mu+the.bias, sd=the.se)), fill=color(k), color=color(k), alpha=.1), 
      geom_vline(aes(xintercept=mu+the.bias + c(-2,2)*the.se), color=color(k), linetype='dotted', alpha=.7),
      geom_vline(aes(xintercept=mu+the.bias + c(-1,1)*the.se),  color=color(k), linetype='dashed', alpha=.7),
      geom_vline(aes(xintercept=mu+the.bias), color=color(k), linetype='solid', alpha=.7)
    )
  }) |> list_c()
  ggplot() + plots + xlab('') + ylab('') + guides(fill='none', color='none')
}

mu = 0
scale.100 = scale_x_continuous(breaks=mu+seq(-.5, .5, by=1/10), limits=mu+c(-.5,.5)) 
scale.10000 = scale_x_continuous(breaks=mu+seq(-5/100, 5/100, by=1/100), limits=mu+c(-5/100,5/100)) 
plot.normals(mu,n=100,k=0)  + scale.100
plot.normals(mu,n=10000,k=0) + scale.10000
plot.normals(mu,n=100,k=10) + scale.100
plot.normals(mu,n=10000,k=10) + scale.10000
plot.normals(mu,n=100,k=100) + scale.100
plot.normals(mu,n=10000,k=100) + scale.10000
```

On the basis of those drawings, your coworker proposes using $\bar{Y}_{100}$ to estimate $\mu$.

::: {#exr-2a .callout-exercise}
Suppose your sample size is $n=100$. Explain why your coworker's original proposal to use $\tilde Y_{10}$ and
your hyperbolic suggestion to use $\tilde Y_{100}$ are *appears*, from the drawings, to be a way to increase precision without increasing bias. 
To explain why this is misleading, describe what happens in the case that $\mu=11/10$ and $\sigma=1$ by doing the following.

- Modifying the drawings to show the sampling distributions in this new case.
- Stating an approximation to the coverage of each of the confidence intervals $\bar{Y}_{k} \pm 1.96\ \text{se}$ for $k=0$, $k=10$, and $k=100$.
  
*Tip*. You can 'change the location' of the distributions in the drawing above by changing the labels on the x-axis.
:::

::: {#exr-2b .callout-exercise}
This coworker isn't going anywhere. Next time, it might be better to bite your tongue and  save time. If your sample size was $n=10,000$, maybe your coworker's idea to use $\tilde Y_{10}$ really wasn't creating much of a problem. What would you need to know about $\mu$ and $\sigma$ to guarantee at least 85% coverage?  
:::

::: {.callout-solution}
What you really need to know is that $\text{bias}/\text{se}=\mu k/ \sigma \sqrt{n}$ is small. For $k=10$ and $n=10,000$,
that's $\mu/ 10\sigma$. We don't see a problem with $Y_{10}$ in our drawings because this is either $0$ (in the plots on the left) or $1/10$ (in the plots on the right). But if $\mu$ were very large or $\sigma$ very small this could be a problem. But, if you like, you can estimate both, see whether it's a problem, and humor your coworker if it's not.
:::

::: {#exr-2c .callout-exercise}
Suppose the context for all this is that you are running a poll of 100 registered voters 
to find out whether they intend to vote for the candidate from [Party 0]{.groupa} or [Party 1]{.groupb}.
Elections are always pretty close, so you can assume that the proportion registered voters who intend to vote for Party 1 
is between .4 and .6. Approximately how much coverage does using the interval estimator $\bar{Y}_{k} \pm 1.96\ \text{se}$
guarantee you for $k=0$, $k=10$, and $k=100$?
:::

# Problem 3


```{r}
m0x.mild     = rep(c(10,12,14,16,18), rev(c(80, 90, 100, 110, 120)))
m0x.moderate = rep(c(10,12,14,16,18), rev(c(50, 75, 100, 125, 150)))
m0x.extreme  = rep(c(10,12,14,16,18), rev(c(2,49, 100, 149, 198)))
m0x.flat = rep(c(10,12,14,16,18), c(100,100,100, 100,100))
m1x.mild     = rep(c(10,12,14,16,18), c(80, 90, 100, 110, 120))
m1x.moderate = rep(c(10,12,14,16,18), c(50, 75, 100, 125, 150))
m1x.extreme  = rep(c(10,12,14,16,18), c(2,49, 100, 149, 198))
m1x.flat = rep(c(10,12,14,16,18), c(100,100,100, 100,100))

exercise.pop = function(m0x, m1x, mu, sigma=\(w,x)mu(w,x)/4 ) {
  eps0 = rnorm(length(m0x))*sigma(0,m0x)
  eps1 = rnorm(length(m1x))*sigma(1,m1x)
  for(uu in unique(m1x)) { 
    eps0[m0x==uu] = eps0[m0x==uu] - mean(eps0[m0x==uu])
    eps1[m1x==uu] = eps1[m1x==uu] - mean(eps1[m1x==uu])
  }
  rbind(
    data.frame(w=0, x=m0x, y=mu(0,m0x)+eps0),
    data.frame(w=1, x=m1x, y=mu(1,m1x)+eps1))
}
```


```{r}
default.shift = .25
proportion.scale = 20
plotX = function(w,x, shift=default.shift) { ifelse(w==0, x-shift, x+shift) }
interleaved.bars = list(
  geom_bar(aes(x=plotX(w,x), y=proportion.scale*after_stat(prop), 
               fill=factor(w)), 
           position='identity', alpha=.1, linewidth=.05, color='black'))
purple.bars = geom_bar(aes(x=x, y=proportion.scale*after_stat(prop)), 
                       alpha=.1, linewidth=.05, fill='purple', color='black', width=.4)
scales = list(
  scale_y_continuous(sec.axis=sec_axis(~./proportion.scale, name='proportion')), 
  coord_cartesian(ylim=c(0,30)),
  guides(color='none', fill='none', size='none') 
)
```

```{r}
lsq.plot = function(sam, formula=NULL, shift=default.shift, line.alpha=.7, size.scale=.1) {
  sam$weights = rep(1, nrow(sam))
  max.weight = max(sam$weights)
  max.size = size.scale * max.weight 
  
  base.plot = ggplot(sam) + 
    geom_point(aes(x=plotX(w,x), y=y,color=factor(w), size=I(weights)), 
      position=position_jitter(seed=0, width=shift),
      alpha=.3) + 
    stat_summary(aes(x=plotX(w,x), y=y, fill=factor(w)), 
      fun.data=mean_sd, geom='pointrange', 
      shape=21, alpha=.5, color='black') +
    interleaved.bars + purple.bars + scales
    
    if(is.null(formula)) { base.plot } else { 
        model = lm(formula, data=sam, weights=weights)
        gridx = seq(-10, 50, length.out=1000)
        muhat = function(w,x) predict(model, newdata=data.frame(w=w,x=x))   
        
        the.plot = base.plot 
        if(line.alpha > 0) { 
            the.plot = base.plot + 
            geom_line(aes(x=plotX(w,x), y=muhat(w,x), color=factor(w)), 
              data = expand_grid(w=0:1, x=gridx), alpha=line.alpha, linetype='dashed') 
        }
        the.plot + 
          geom_point(aes(x=plotX(w,x), y=muhat(w,x), fill=factor(w)),  
            color='black', alpha=.5, shape=23)
    }
}
```

::: {.panel-tabset}
### Example A
```{r}
#| fig-width: 10
set.seed(0)
exercise.pop(m0x.moderate, m1x.moderate, \(w,x) x) |> 
  lsq.plot()
```

### Example B
```{r}
#| fig-width: 10
set.seed(0)
exercise.pop(m0x.flat, m0x.moderate, \(w,x) 3*sqrt(x/2+w*x^2/16)) |> 
  lsq.plot()
```

### Example C

```{r}
#| fig-width: 10
set.seed(0)
exercise.pop(m1x.moderate, m1x.extreme, \(w,x) 20*exp(-w*x/20)) |> 
  lsq.plot()
```

:::

Consider the following estimation targets.
$$
\begin{aligned}
\Delta_{\text{raw}} &= \frac{1}{m_1}\sum_{j:w_j=1} y_j - \frac{1}{m_0}\sum_{j:w_j=0} y_j \qfor m_w = \sum_{j:w_j=w} 1 \\
\Delta_{0}   &= \frac{1}{m_0}\sum_{j:w_j=0} \lcb{ \mu(1,x_j) - \mu(0,x_j) }  \\
\Delta_{1}   &= \frac{1}{m_1}\sum_{j:w_j=1} \lcb{ \mu(1,x_j) - \mu(0,x_j) } \\ 
\Delta_{\text{all}}   &= \frac{1}{m}\sum_{j=1}^m \lcb{ \mu(1,x_j) - \mu(0,x_j) }  
\end{aligned}
$$
          

::: {#exr-3 .callout-exercise}
For each of the examples above, answer the following questions.

1. Is $\Delta_1 > \Delta_\text{raw}$? 
2. Is $\Delta_1 > \Delta_0$?
3. Is $\Delta_1 > \Delta_\text{all}$?

I'm not asking you to compare the magnitudes of these estimation targets. If $\Delta_1$ is positive and $\Delta_\text{raw}$ is negative, then no matter what the magnitudes are, $\Delta_1 > \Delta_\text{raw}$.
:::

# Problem 4

Your friend tells you about a new estimator for the mean $\theta=\frac1m\sum_{j=1}^m y_j$ of a population $y_1 \ldots y_m$. They like to use it when they think the population mean $\theta$ is close to zero.

::: {.panel-tabset}
### In Math
$$
\hat \theta = \begin{cases}
  0 & \text{if } \abs{\bar Y} < .1 \\
  \bar Y & \text{otherwise}
\end{cases}
$$

### In Code 
```{r}
#| echo: true
estimator = function(x, kill.below=.1) {
  if(abs(mean(x)) < kill.below) { 0 } else { mean(x) }
}
```
:::

To show you how it works, they've make up a population of $y_1 \ldots y_{1000}$ random variables with
mean $\theta=.02$ and standard deviation 1. They show you the sampling distribution of this estimator
when you draw samples $Y_1 \ldots Y_n$ of size $n=25$ with replacement from this population.

```{r}
#| fig-width: 10
set.seed(11)

m = 1000
y = rnorm(m); y=.02 + ((y-mean(y)) / sd(y))
theta = mean(y) 
n = 25

samples = 1:10000 |> map(function(.) {
  J = sample(1:m, n, replace=TRUE)
  Y = y[J]
})
estimator.samples = samples |> map_vec(estimator)
mean.samples = samples |> map_vec(mean)

estimator.rmse = sqrt(mean((estimator.samples-theta)^2))
mean.rmse = sqrt(mean((mean.samples-theta)^2))

ggplot() + 
  geom_histogram(aes(x=estimator.samples, y=after_stat(density)), 
    bins=50, fill=teal, alpha=.1, color=teal, linewidth=.05) + 
geom_histogram(aes(x=mean.samples |> map_vec(mean), y=after_stat(density)),
    bins=50, fill=pink, alpha=.1, color=pink, linewidth=.05) + 
    geom_vline(xintercept=theta, color='green', alpha=.5) + 
    simplified.theme 
  
```

They're pretty pleased with themselves. Their estimator is ...
  
  - almost never further from $\theta$ than the sample mean is. It's further in `r percent(mean(estimator.rmse >= mean.rmse))` of samples. 
  - often closer to $\theta$ than the sample mean is. It's closer in `r percent(mean(abs(estimator.samples-theta) < abs(mean.samples-theta)))` of samples. 

This is a weird estimator, so you ask if you can still use the bootstrap to 
calibrate confidence intervals around it. Your friend says ...

> I checked this morning. The coverage probability is $.93$. That's not bad when you're getting a
> more accurate estimate `r percent(mean(abs(estimator.samples-theta) < abs(mean.samples-theta)))` of the time.

They pull up the code below to show you how they checked, but when they run it, it doesn't say $.93$. It says $.84$.

```{r}
set.seed(7) # use seed 3 for 93% coverage
J = sample(1:m, n, replace=TRUE)
Y = y[J]
set.seed(0)
```

```{r}
#| echo: true
bootstrap.sampling.distribution = function(Y) { 
  1:1000 |> map_vec( function(.) {  
    J.star = sample(1:n, n, replace=TRUE)
    Y.star = Y[J.star]
    estimator(Y.star)
  })
}

se = sd(bootstrap.sampling.distribution(Y))
covers = 1:1000 |> map_vec( function(.) { 
  J = sample(1:m, n, replace=TRUE)  
  Y = y[J]                          
  estimator(Y) - 1.96*se <= theta & theta <= estimator(Y) + 1.96*se
})
mean(covers)
```

::: {#exr-4a .callout-exercise}
Tell your friend what's wrong with their code and how to fix it. A sentence or two should do. 
:::

::: {.callout-tip collapse="true"}
### Hint. 
It doesn't take me more than 5 seconds to fix and I don't type very fast.
:::

::: {.callout-solution}
```{r}
set.seed(0)
covers = 1:1000 |> map_vec( function(Y) { 
  J = sample(1:m, n, replace=TRUE)
  Y = y[J]
  se = sd(bootstrap.sampling.distribution(Y))
  estimator(Y) - 1.96*se <= theta & theta <= estimator(Y) + 1.96*se
})
mean(covers)
```
:::

::: {#exr-4b .callout-exercise}
*Extra Credit*. Suppose that somewhere in the code they used to draw those plots, 
they were actually drawing a sample `Y` from the population `y`. They were running this code.
```{r}
#| echo: true
#| eval: false
J = sample(1:m, n, replace=TRUE)  
Y = y[J]    
```
They ran it before they calculated the coverage probability this morning and they ran it again before they calculated it just now. Why do you think the two coverage probabilities they calculated are so different? 
:::

::: {.callout-tip collapse="true"}
### Hint.
Take a look at the plots below. Each one shows the bootstrap sampling distributions 
of [this new estimator in teal]{.teal} and [the sample mean in pink]{.pink} for a random sample $Y_1 \ldots Y_n$ drawn from the population $y_1 \ldots y_m$. Each estimator's bootstrap mean &plusmn; 2 bootstrap standard deviations is shown by a pair of vertical 
in the same colors. 

```{r}
#| fig-width: 10
#| layout-nrow: 5
#| layout-ncol: 5
#| fig-asp: 1

bootstrap.sampling.distribution = function(Y, estimator=mean) { 
  1:1000 |> map_vec( function(.) {  
    J.star = sample(1:n, n, replace=TRUE)
    Y.star = Y[J.star]
    estimator(Y.star)
  })
}

plot.sampling.dist = function(.) { 
  J = sample(1:m, n, replace=TRUE)  
  Y = y[J] 
  estimator.samples = bootstrap.sampling.distribution(Y, estimator=estimator)
  meanY.samples = bootstrap.sampling.distribution(Y, estimator=mean)
  ggplot() + 
    geom_histogram(aes(x=estimator.samples, y=after_stat(density)), 
      bins=50, fill=teal, alpha=.1, color=teal, linewidth=.05) + 
    geom_histogram(aes(x=meanY.samples, y=after_stat(density)), 
      bins=50, fill=pink, alpha=.1, color=pink, linewidth=.05) + 
    geom_vline(xintercept=theta, color='green', alpha=.5) +
    geom_vline(xintercept=mean(estimator.samples)+c(-2,2)*sd(estimator.samples), 
      color=teal, alpha=.3) +
    geom_vline(xintercept=mean(meanY.samples)+c(-2,2)*sd(meanY.samples), 
      color=pink, alpha=.3) +
    scale_x_continuous(breaks=seq(-1.5,1.5,by=.5), limits=c(-1.5,1.5)) + 
    coord_cartesian(xlim=c(-1,1), ylim=c(0,10))
} 

plot.sampling.dist() 
plot.sampling.dist() 
plot.sampling.dist() 
plot.sampling.dist() 
plot.sampling.dist() 

plot.sampling.dist() 
plot.sampling.dist() 
plot.sampling.dist() 
plot.sampling.dist() 
plot.sampling.dist() 

plot.sampling.dist() 
plot.sampling.dist() 
plot.sampling.dist() 
plot.sampling.dist() 
plot.sampling.dist() 

plot.sampling.dist() 
plot.sampling.dist() 
plot.sampling.dist() 
plot.sampling.dist() 
plot.sampling.dist() 

plot.sampling.dist() 
plot.sampling.dist() 
plot.sampling.dist() 
plot.sampling.dist() 
plot.sampling.dist() 
```
:::

# Problem 5

```{r scatter}
#| fig-width: 10
mu = function(w, x){1 + (8/3)*((1-2*w)*sinpi(x/8) + 2*w^2*(x/4)^2)}
pi = function(x){ 
  case_match(x, 
             0 ~ .2, 
             1 ~ .3,
             2 ~ .9,
             3 ~ .7,
             4 ~ .5)
}
  

trim = function(x,a,b) { pmin(pmax(x, a), b) }
set.seed(1)
m = 10000
x = sample(c(rep(seq(0,3),8),4),  m, replace=T)
w = rbinom(m, 1, pi(x)) 
y = (mu(w,x) + .75*rnorm(m)) |> trim(0,4)
pop = data.frame(x=x, w=w, y=y)

n = 1000
J = sample(1:m, n, replace=T)
sam = pop[J,]
X = sam$x
W = sam$w
Y = sam$y
summaries = sam |> group_by(w,x) |> summarize(mu=mean(y), n = n(), .groups='drop') |>
                     group_by(w) |> mutate(f=n/sum(n))
plotX = function(w,x, shift=.2){ x + shift*(2*w-1) }

plot.summaries =  list(
  scale_x_continuous(breaks=seq(0,4,by=1)),
  scale_y_continuous(breaks=seq(0,4,by=1), sec.axis=sec_axis(~./4)),
  geom_bar(aes(x=plotX(w,x), y=4*after_stat(prop), fill=factor(w)), 
      position='identity', width=.4, 
      alpha=.1, color='gray', linewidth=.05),
  comparison_styles, simplified.theme)
  
scatter = ggplot(sam) + 
  geom_point(aes(x=plotX(w,x), y=y, color=factor(w)), 
             position=position_jitter(w=.18,h=0, seed=0), alpha=.3, size=.5) + 
             plot.summaries
scatter 
```

You're working at a company that makes a fancy alarm clock that watches you sleep. You don't tell it when to wake you up.
You tell it when you need to be up by; it wakes you earlier if, based on your sleep patterns, it thinks that's better for you.
You can, of course, turn that feature off so it wakes you when you ask like a regular alarm clock. You're on a team that's 
trying to figure out how well this feature is working. 

You have a population $w_1, x_1, y_1 \ldots w_m, x_m, y_m$ of people who have been using the clock.

  - $w_j$ indicates whether they have the feature turned [on]{.groupb} or [off]{.groupa}. 
  - $x_j$ is a sleep quality score they reported when they bought the clock (0 is worst, 4 is best).
  - $y_j$ is the average of the sleep quality scores they've logged since they started using the clock.
    
Privacy is a big deal when you're selling a device that watches people sleep, so your app doesn't report 
sleep quality scores directly to your company. You actually have to run a survey. The result is a sample 
$(W_1, X_1, Y_1) \ldots (W_n,X_n,Y_n)$ that's drawn with replacement from the population. You have it  
in an **R** dataframe `sam` with `n` rows. I've plotted it above. As usual, $X$ is on the horizontal axis,
$Y$ variable is on the vertical axis, and $W$ variable, which is binary, is indicated by color: green for $W=1$, red for $W=0$. 
The proportion of times that $X$ takes on each value $x$ when $W=0$ and when $W=1$, 
$\ingroupa{P_{x\mid 0}}$ and $\ingroupb{P_{x\mid 1}}$ respectively, are plotted as histograms in matching colors. 
They're also shown in the table below.

```{r summaries}
f0 = summaries |> filter(w==0)
f0 = f0[order(f0$x),]
f1 = summaries |> filter(w==1)
f1 = f1[order(f1$x),]
options(digits=2)
```

$$
\color{gray}
\begin{array}{lll}
x & \ingroupa{P_{x \mid 0}} & \ingroupb{P_{x \mid 1}}  \\
\hline
0 & \ingroupa{`r f0$f[1]`} & \ingroupb{`r f1$f[1]`} \\
1 & \ingroupa{`r f0$f[2]`} & \ingroupb{`r f1$f[2]`} \\
2 & \ingroupa{`r f0$f[3]`} & \ingroupb{`r f1$f[3]`} \\
3 & \ingroupa{`r f0$f[4]`} & \ingroupb{`r f1$f[4]`} \\
4 & \ingroupa{`r f0$f[5]`} & \ingroupb{`r f1$f[5]`} \\
\end{array}
$$


## Part 1 {#sec-5a}

Suppose your coworker fits  the *additive model* by least squares. 
That is, they choose $\hat\mu(w,x)$ as their estimate of $\mu(w,x)$ like this.
$$
\begin{aligned}
\hat \mu &= \argmin_{m \in \model} \sum_{i=1}^n \qty{ Y_i - m(W_i,X_i) }^2 \qquad \\
&\qqtext{ where } \model = \qty{m(w,x) = a(w) + b(x)  \ \text{ for all functions $a(w)$ and $b(x)$ }}
\end{aligned}
$$

```{r}
#| echo: false
additive.predictor = lm(y~w+factor(x), data=sam) 
muhat.additive = function(w,x) { predict(additive.predictor, newdata=data.frame(w=w, x=x)) }
```

They claim that the difference in the green and red lines' values at $x=0$, $\hmub{0} - \hmua{0} = `{r} muhat.additive(1,0)-muhat.additive(0,0)`$, is a good estimate of the adjusted difference in means $\Delta_1$. ^[They don't *say* 'the difference $\hmub{0} - \hmua{0}$'. We'll get to that in the next section.] ^[They describe what they're estimating as the *average treatment effect on the treated* (ATT), but the treatment $W_i$ isn't randomized; the people in our population chose it themselves. People, myself included, do this pretty often. We say *average treatment effect on the treated* (ATT), *average treatment effect on the controls* (ATC), and *average treatment effect* (ATE) to refer to $\Delta_1$, $\Delta_0$, and $\Delta_\text{all}$ respectively even outside the context of randomized experiments. Sometimes that's because they really want to estimate those treatment effects and they're doing their best despite lack of randomization.  That's what's going on in Section 2.1 of [Kim et al. (2023)](https://arxiv.org/pdf/2311.00568). But sometimes it's just something we do because we don't have a good alternative: there isn't standard terminology that differentiates between $\Delta_1$, $\Delta_0$, and $\Delta_\text{all}$ that isn't about causality. Sometimes it's something in between, like when people talk about estimating the 'effect of race' or 'effect of gender'; they want to be making a causal claim but what they're calling a treatment is a social construct and what exactly our potential outcomes mean is unclear. If you're interested in questions like that, take a look at Issa Kohler-Hausmann's paper [Eddie Murphy and the Danger of Counterfactual Causal Thinking About Detecting Racial Discrimination](https://scholarlycommons.law.northwestern.edu/nulr/vol113/iss5/6/).]

$$
\Delta_1 = \ingroupb{\frac{1}{m_1}\sum_{j:w_j=1}} \lcb{ \mub{X_i} - \mua{X_i} } \qfor m_1 = \sum_{j:w_j=1} 1
$$

You're not sure, so you plot their fitted model. Here's what you see. As usual, $\mub{x}$ is the green line and $\mua{x}$ is the red one.

```{r}
#| fig-width: 10
grid = expand.grid(x=unique(sam$x), w=0:1)
scatter + 
  geom_line(aes(x=x, y=muhat.additive(w,x), color=factor(w)), data=grid, linetype='dashed') +
  geom_point(aes(x=x, y=muhat.additive(w,x), fill=factor(w)), shape=filled$square, color='black', data=sam) + 
  guides(fill='none', shape='none', color='none')
```

::: {.callout-exercise #exr-additive-bias}
Do you think their estimate is biased up (too large) or down (too small)? Referring to the plot, explain why. 
:::

::: {.callout-solution}
It's biased up. Looking at the distribution of the treated people --- count the green dots or look at the green histogram, where it's been done for you --- we see that the ATT is essentially an average of the difference between the green and red lines in the middle of the plot. And when you look there the green line fits the data but the red line is an underestimate: 
$\text{good} - \text{too small} = \text{too big}$.
:::

## Part 2 {#sec-5b}

When they calculate their estimator, they're probably not *explicitly* making a comparison at $x=0$, e.g. by writing code that says something like `muhat(1,0)-muhat(0,0)`. They're probably just looking at the *coefficient on $w$* in the output of the `lm` function.
```{r}
#| echo: true 
additive.predictor = lm(y~w+factor(x), data=sam) 
additive.predictor    
```
That's the number in front of $w$ when you write out your least squares predictor out in terms of a certain basis: it's $\hat a_1$ for $\hat\mu(w,x) = \hat a_0 + \hat a_1 w + \hat a_2 1_{=2}(x) + \ldots + \hat a_5 1_{=5}(x)$. When they hear what you said in the @exr-additive-bias, they change their model to one they've seen you use before. This is their new least squares predictor.
$$
\hat\mu = \argmin_{m \in \model} \sum_{i=1}^n \qty{ Y_i - m(W_i,X_i)}^2 \qquad \text{ where } \model = \qty{\text{all functions} \ m(w,x)  }
$$

```{r}
all.functions.predictor = lm(y~w*factor(x), data=sam) 
```

And they do exactly what they did before. They report the coefficient on $w$ that `lm` gives them: `r sprintf("%.4f", coef(all.functions.predictor)['w'])`.
```{r}
#| echo: true
all.functions.predictor = lm(y~w*factor(x), data=sam) 
all.functions.predictor
```
That's $\hat a_1$ when we write out the least squares predictor like this.
$$
\hat\mu(w,x) = \hat a_0 + \hat a_1 w + \hat a_2 1_{=2}(x) + \ldots + \hat a_5 1_{=5}(x) + \hat a_6 w 1_{=2}(x) + \ldots + \hat a_{9} w 1_{=5}(x)
$$ {#eq-all-functions-mu}

That number doesn't sound right to them, so they plot their new predictor using the code you just wrote. Here's what they see.

```{r}
#| fig-width: 10
muhat.all.functions = function(w,x) { predict(all.functions.predictor, newdata=data.frame(w=w, x=x)) }
scatter + 
  geom_line(aes(x=x, y=muhat.all.functions(w,x), color=factor(w)), data=grid, linetype='dashed') +
  geom_point(aes(x=X, y=muhat.all.functions(w,x), fill=factor(W)), shape=filled$square, color='black', data=sam) + 
  guides(fill='none', shape='none', color='none')
```

They're confused. The plot tells basically the same story as the one for the additive model. There's a [green curve]{.groupb} below a [red one]{.groupa}. It seems like they should be getting a negative number, not zero.  

::: {#exr-att-calc .callout-exercise}

1. Using the output of `lm` above and other information you have above, estimate $\Delta_1$. 
Or, if you don't feel like getting out your calculator, explain how to do it very concretely,
so your confused coworker handle it from there. Don't expect too much from them. Say something
like 'take this number from this place, multiply it by this number from this other place, etc.'

2. Explain why what they were doing before worked for the additive model but not for the all functions model.
   
:::


::: {.callout-tip collapse="true"}
### If You Don't Want to Interpret the Output of `lm` 
Interpreting the output of `lm` isn't something we've spent much time on in this class. And it's not something you'll have to do on the actual final exam. 

If you want to try, go ahead. Reading the next tip might help. But if you don't, you're welcome to use this table describing the values of $\hat\mu(w,x)$. 
And if you're writing instructions for your coworker instead of doing the calculation yourself, you're welcome to refer to this table in your instructions.

```{r}
library(gt)
grid = expand.grid(w=0:1, x=unique(sam$x))
grid$muhat = predict(all.functions.predictor, newdata=grid)
grid |> gt()
```

:::

::: {.callout-tip collapse="true"}
### Interpreting the Output of `lm` 
To interpret the coefficients in the output of `lm`, you need to understand how it describes the basis it's thinking in. This is one of those skills that's not all that necessary when you're working on your own---you can just use `predict` to calculate $\hat\mu(w,x)$ and not worry about how **R** is actually doing it under the hood. But it's a good skill to have around other people, because a lot of them are in the habit of thinking about *coefficients* instead of *predictions*, and when that results in mistakes like the one 'your coworker' made here, you'll need to do a little translation between their way of thinking and yours so you can say something more helpful than 'Forget everything you know. Just do it my way.'

In the output of `lm(y~w*factor(x), data=data)`, they're thinking of $\hat\mu$ like I've written it in @eq-all-functions-mu.

 - What I've called $\hat a_0$, **R** writes under '(Intercept)'. It's the value of $\hat\mu(0,0)$.
 - What I've called $\hat a_1$, **R** writes under 'w'. It's what's multiplying $w$ in
@eq-all-functions-mu. 
 - What I've called $\hat a_2 \ldots a_5$, **R** writes under 'factor(x)1' ... 'factor(x)4'. What it means by 'factor(x)*k*' is the *indicator* function $1_{=k}(x)$.
 - What I've called $\hat a_6 \ldots \hat a_10$, **R** writes under 'w:factor(x)1' ... 'w:factor(x)4'. What it means by 'w:factor(x)*k*' is the product of $w$ and the *indicator* function $1_{=k}(x)$.

*Why it says factor(x)*. 
**R** says `factor(x)1`, etc., because I've converted $x$ to a 'factor'--- something where we want an indicator for each possible value in our model instead of a single slope --- directly in the formula `y~w*factor(x)`. If $x$ were already a factor, you'd see 'x1' ... 'x4' instead of 'factor(x)1' ... 'factor(x)4' and 'w:x1' ... 'w:x4' instead of 'w:factor(x)1' ... 'w:factor(x)4' in the output as you do below.  
```{r}
#| echo: true
another.all.functions.predictor = lm(y~w*x, data=sam |> mutate(x=factor(x)))
another.all.functions.predictor
```
The diffence between converting it in the formula and before is what `predict` expects. 
If you say `factor(x)` in the formula, you can pass in data in which $x$ is not a factor. It'll convert for you.
```{r}
#| echo: true
predict(all.functions.predictor, data.frame(w=1, x=1:4))
```
If you convert it ahead of time when you call `lm`, you have to convert it when you call predict, too.
```{r}
#| echo: true
predict(another.all.functions.predictor, data.frame(w=1, x=factor(1:4)))
```
If you don't, it gives you an error.
```{r}
#| error: true
predict(another.all.functions.predictor, data.frame(w=1, x=1:4))
```

:::

::: {.callout-solution}

To estimate the ATT, we want to average the difference $\hat\mu(1,x)-\hat\mu(0,x)$ between the green and red curves over the distribution of the green points, i.e., using the weights $\hat p(1, x)$ shown in the green histogram. Abstractly, here's our estimate.
$$ 
\widehat{ATT} = \sum_x \hat p(1,x) \qty{ \hat\mu(1,x) - \hat\mu(0,x) }
$$

We've got a table of $\hat p(1, x)$ above, so what we need to actually calculate this is add a row of the difference 
$\hat\mu(1,x) - \hat\mu(0,x)$. And for this model, FINISH THIS.

:::

 
## Part 3 {#sec-5c}

```{r}
estimate.Delta.1 = function(sam, formula=y~w*factor(x)) {
  if(!('weights' %in% colnames(sam))) { 
    sam$weights = rep(1, nrow(sam)) 
  }
  predictor = lm(formula, data=sam, weights=weights)
  muhat = function(w,x) { predict(predictor, newdata=data.frame(w=w, x=x)) }
  mean(muhat(1,sam$x[sam$w==1]) - muhat(0, sam$x[sam$w==1]))
}
```

Now you've managed to get your coworker estimating $\Delta_1$ by using `predict` instead
of just looking at coefficients. Congratulations! That means no matter what model they're 
using, if $\hat\mu(w,x)$ is a reasonable estimate of $\mu(w,x)$, they'll get a reasonable 
estimate of $\Delta_1$. And you've got them packaging everything up into a function `estimate.Delta.1` 
that takes in a sample, does all the work, and gives you back an estimate of the $\Delta_1$.
```{r}
#| echo: true
Delta1.hat = estimate.Delta.1(sam)
```
Their point estimate of $\Delta_1$ is `{r} round(Delta1.hat,3)`.

Now they want a confidence interval, so they bootstrap their estimator. 
To do it, they first write a function that calculates the sampling distribution 
of *any* estimator that's written as a function of a sample when you sample with replacement from a population. 
Something like this.
```{r}
#| echo: true
sampling.distribution = function(pop, estimator, draws=1000) { 
  map_vec(1:draws, function(b) {
    J = sample(1:nrow(pop), nrow(pop), replace=TRUE)
    sam = pop[J,]
    estimator(sam)
  }) 
}
```
Then, to get the bootstrap sampling distribution of their estimator, they use it 
as if their sample were the population.
```{r}
set.seed(1)
```

```{r}
#| echo: true
Delta1.hat.star = sampling.distribution(sam, estimate.Delta.1) 
```

They calculate a 95% interval like this. 
```{r}
#| echo: true
se = sd(Delta1.hat.star)
interval = c(Delta1.hat - 1.96*se, Delta1.hat + 1.96*se)
```
They get this interval $[`r sprintf('%.2f', min(interval))` \ , \  `r sprintf('%.2f', max(interval))` ]$. 
And they've done everything right. They're using an interval estimator with 95% coverage.
But your boss, who has a habit of correcting employees without really 
spending the time to understand what they're doing, glances at the code and says this:

> You forgot to divide by $\sqrt{n}$. Rookie mistake.

This isn't great for your coworker's confidence. To help explain the $\sqrt{n}$ thing, 
you draw a few things. You draw the bootstrap sampling distribution of $\Delta_1$, 
the bootstrap sampling distribution of $\bar{Y}$, and the distribution of $Y$ in the sample.

```{r}
#| fig-width: 10
#| layout-nrow: 2 
#| layout-ncol: 2
#| fig-cap: 
#|  - "" 
#|  - "The bootstrap sampling distribution of Delta.hat.1."
#|  - "The distribution of Y in the sample." 
#|  - "The bootstrap sampling distribution of the sample mean." 
ggplot()
ggplot() + 
  geom_histogram(aes(x=Delta1.hat.star, y=after_stat(density)), alpha=.3, fill='red', bins=50) +
  with_annotations(Delta1.hat.star, highlight.dots=0) 

ggplot() + 
  geom_histogram(aes(x=y, y=after_stat(density)), alpha=.3, fill='red', bins=50, data=sam)  +
  with_annotations(sam$y, highlight.dots=0) 

twodigits = function(x) { sprintf('%.2f', x) } 
meanY.star = sampling.distribution(sam, function(sam) mean(sam$y)) 
ggplot() + 
  geom_histogram(aes(x=meanY.star, y=after_stat(density)), alpha=.3, fill='red', bins=50) + 
  with_annotations(meanY.star, highlight.dots=0) +
  scale_x_continuous(breaks=seq(1.3,1.6,by=.05), labels=twodigits)
```

Here's what you say.

> Suppose we were just using the sample mean $\bar Y$ to estimate the mean outcome in the population, $\frac{1}{m}\sum_{j=1}^m y_j$.
> One way of calculating the standard deviation of $\bar Y$ is to take the standard deviation of the outcomes $Y_1 \ldots Y_n$ and divide by $\sqrt{n}$.
> But when we bootstrap $\bar Y$, we get a distribution that looks like the distribution of *our estimator*, not our outcomes. 
> It's already narrower by a factor of $1/\sqrt{n}$. 
You even calculate the standard deviation of $Y_1 \ldots Y_n$---it's `r sd(Y)`---divide it by $\sqrt{n}\approx `r round(sqrt(n))`$, and compare it to the bootstrap standard deviation of $\bar Y$, which is `r sprintf('%1.3f', sd(meanY.star))`.
But you really want to end your explanation with a flourish. You say this.
> If you want to think of the standard deviation of $Y_1 \ldots Y_n$---the one you'd divide by $\sqrt{n}$---in bootstrap terms, it's the bootstrap 
> standard deviation of an estimator you'd never actually use. It's the standard deviation of ...

::: {.callout-exercise #exr-5c}
Finish the sentence. What simple estimator has bootstrap standard deviation $\sigma=\sqrt{\frac{1}{n}\sum\{Y_i - \bar Y\}^2}$,
the same standard deviation as the sample $Y_1 \ldots Y_n$? 

*Tip*. There's more than one. Any one you can think of is fine.
:::

::: {.callout-solution}
The estimator $\hat\theta = Y_1$. Or $\hat\theta=Y_2$, $\hat\theta=Y_3$, etc.
:::

## Part 4 {#sec-5d}

```{r}
#| cache: true 
B=4000

difmeans.estimator = function(sam) { 
  mean(sam$y[sam$w==1]) - mean(sam$y[sam$w==0]) 
}
difmeans = difmeans.estimator(sam)
bootstrap.difmeans = sampling.distribution(sam, difmeans.estimator)
se.difmeans = sd(bootstrap.difmeans)
```

When you and your coworker bring a summary of your findings to your boss, they're annoyed. They say this.

> Why are you doing all this complicated nonsense? 

They pull up the data, subtract the mean of the green points from the mean of the red points. They estimate the standard deviation of their estimator $\hat\Delta_{\text{raw}}$ 
to be `{r} se.difmeans`, which gives them a 95% confidence interval of $\hat\Delta_{\text{raw}} \pm `r 1.96*se.difmeans`$.

> Much better! Simple. And more precise. 
> Your interval was $\pm `r 1.96*se`$. 
> Mine is $\pm `r 1.96*se.difmeans`$.

But width isn't everything. Their interval isn't just narrower. It's nowhere near yours. 

::: {.callout-exercise #exr-5d}
Is their point estimate $\hat\Delta_{\text{raw}}$ bigger or smaller than your point estimate $\hat\Delta_1$?  Explain why. 
:::

## Part 5 

Your coworker decides to do a little simulation to test out all the new stuff they've learned. 

```{r}
#| fig-width: 10
trim = function(x,a,b) { pmin(pmax(x, a), b) }
set.seed(1)
m = 10000
pop = expand.grid(w=c(0,1), x=rep(seq(0,4), m/10)) |>
  mutate(y=(mu(w,x) + .5*rnorm(m)) |> trim(0,4))

scatter = ggplot(pop) + 
  geom_point(aes(x=plotX(w,x), y=y, color=factor(w)), 
             position=position_jitter(w=.18,h=0, seed=0), alpha=.15, size=.1) + 
             plot.summaries

scatter 
```

They make up a population of 10,000 people. The population they've made up is plotted above.
They draw samples of size `r n` repeatedly to compare the sampling distributions of four different estimators of $\Delta_1$.

1. The estimator they were using at the beginning, in @sec-5a, where they used $\hat\mu$ is the least squares predictor in the *additive model*.
2. The estimator you calculated in @exr-att-calc. You used the least squares predictor in the *all functions model*. 
3. An estimator like the one they used in @sec-5a, but based on the *inverse probability weighted* least squares predictor in the *additive model*.
4. Your boss's estimator $\Delta_{\text{raw}}$ from @sec-5d.

From what they see, none of these estimators is biased. 
They sampling distributions of all four, plotted below, 
are centered at the estimation target $\Delta_1$. 

```{r}
#| cache: true
sim.draws = 1000
mu.model = lm(y~w*factor(x), data=pop)
mu = function(w,x) { predict(mu.model, newdata=data.frame(w=w, x=x)) }
att = mean(mu(1,pop$x[pop$w==1]) - mu(0,pop$x[pop$w==1]))

pop.summaries = pop |> group_by(w,x) |> summarize(mu=mean(y), n=n(), .groups='drop') |>
                     group_by(w) |> mutate(f=n/sum(n)) 
pi = function(x) { 1/2 }

with.weights = function(sam) sam |> mutate(weights=ifelse(w==1, 1, pi(x)/(1-pi(x))))
estimators = list(
  additive = function(sam) { sam |> estimate.Delta.1(formula=y~w+factor(x)) },
  all.functions = function(sam) { sam |> estimate.Delta.1(formula=y~w*factor(x)) },
  ipw.additive = function(sam) { sam |> with.weights() |> estimate.Delta.1(formula=y~w+factor(x)) },
  raw = function(sam) { difmeans.estimator(sam) })

estimates = estimators |> map(\(estimator) data.frame(estimate=pop |> 
                  sampling.distribution(estimator, draws=sim.draws))) |>
  list_rbind(names_to='estimator')
```

```{r}
#| fig-width: 10
ggplot(estimates) + 
  geom_vline(xintercept=att, linewidth=2, color='green', alpha=.5) +
  geom_histogram(aes(x=estimate, y=after_stat(density), group=estimator), alpha=.3, bins=50) + 
  stat_summary(aes(x=att, y=estimate, group=estimator), geom='vline',
    fun.data=\(x) data.frame(xintercept=mean(x)), alpha=.5, color='blue') +
    stat_summary(aes(x=att, y=estimate, group=estimator), geom='vline',
    fun.data=\(x) data.frame(xintercept=c(mean(x)-sd(x), mean(x)+sd(x))), 
    alpha=.3, linetype='dashed', color='blue') +
  stat_summary(aes(x=att, y=estimate, group=estimator), geom='vline',
    fun.data=\(x) data.frame(xintercept=c(mean(x)-1.96*sd(x), mean(x)+1.96*sd(x))), 
    alpha=.3, linetype='dotted', color='blue') +
    scale_x_continuous(breaks=seq(-2,2,by=.05), labels=twodigits) + guides(fill='none') +
   coord_cartesian(xlim=c(att-.1, att+.1), ylim=c(0,35)) +
  facet_wrap(~estimator, nrow=2)  
```

::: {.callout-exercise #exr-5e}
What is it about this population that makes this happen? 

No need for a long explanation. A sentence should be enough.
:::

## Part 6 

Your coworker has lost confidence in their ability to make up fake populations 
that they can use to test estimators. To try to test their four estimators 
again, they get a fake population from an old paper published in *Statistical Science*.

```{r}
set.seed(1)
m = 10000
x = rep(seq(0,4), m/5)
pi = function(x) { 1-(.2+.6*x/4) }
w = rbinom(m, 1, pi(x))
pop = data.frame(w=w, x=x) |>
  mutate(y=3+w-x/4 + (rnorm(n) |> trim(-1,1)))
  
scatter = ggplot(pop) + 
  geom_point(aes(x=plotX(w,x), y=y, color=factor(w)), 
             position=position_jitter(w=.18,h=0, seed=0), alpha=.15, size=.1) + 
             plot.summaries

scatter 
```

```{r}
#| cache: true
mu.model = lm(y~w*factor(x), data=pop)
mu = function(w,x) { predict(mu.model, newdata=data.frame(w=w, x=x)) }
att = mean(mu(1,pop$x[pop$w==1]) - mu(0,pop$x[pop$w==1]))

with.weights = function(sam) sam |> mutate(weights=ifelse(w==1, 1, pi(x)/(1-pi(x))))
estimators = list(
  additive = function(sam) { sam |> estimate.Delta.1(formula=y~w+factor(x)) },
  all.functions = function(sam) { sam |> estimate.Delta.1(formula=y~w*factor(x)) },
  ipw.additive = function(sam) { sam |> with.weights() |> estimate.Delta.1(formula=y~w+factor(x)) },
  raw = function(sam) { difmeans.estimator(sam) })

estimates = estimators |> map(\(estimator) data.frame(estimate=pop |> 
                  sampling.distribution(estimator, draws=1000))) |>
  list_rbind(names_to='estimator')
```

```{r}
ggplot(estimates) + 
  geom_vline(xintercept=att, linewidth=2, color='green', alpha=.5) +
  geom_histogram(aes(x=estimate, y=after_stat(density), group=estimator), alpha=.3, bins=50) + 
  stat_summary(aes(x=att, y=estimate, group=estimator), geom='vline',
    fun.data=\(x) data.frame(xintercept=mean(x)), alpha=.5, color='blue') +
    stat_summary(aes(x=att, y=estimate, group=estimator), geom='vline',
    fun.data=\(x) data.frame(xintercept=c(mean(x)-sd(x), mean(x)+sd(x))), 
    alpha=.3, linetype='dashed', color='blue') +
  stat_summary(aes(x=att, y=estimate, group=estimator), geom='vline',
    fun.data=\(x) data.frame(xintercept=c(mean(x)-1.96*sd(x), mean(x)+1.96*sd(x))), 
    alpha=.3, linetype='dotted', color='blue') +
  facet_wrap(~estimator, nrow=2)  + simplified.theme + guides(fill='none') + 
  coord_cartesian(xlim=c(att-.1, att+.1)) 
```

Now they're angry. The problem they had in their last simuation is gone---your boss's estimator, at least, is biased. But they're still not seeing any problem with their estimator from @sec-5a. All the changes they've considered have increased their estimator's variance without meaningfully reducing bias.

::: {.callout-exercise #exr-5f}
What's going on? What is it about this population that makes this happen? 

No need for a long explanation. A sentence should be enough.
:::


::: {.callout-exercise #exr-5f-p2}
When they use the same axes for all the plots, the sampling distribution of your boss's estimator $\hat\Delta_{\text{raw}}$ just isn't there. That estimator is so biased that its sampling distribution
is off the plot to one side or the other. Which side is it on? Is the bias of $\hat\Delta_{\text{raw}}$ positive or negative?
:::



# Appendix {#sec-appendix-plots}

::: {#fig-coverage}
```{r}
coverage = function(bias.over.se){ pnorm(1.96+bias.over.se + 1.96) - pnorm(bias.over.se - 1.96) }
x=seq(-3,3,by=.01)
ggplot() + geom_line(aes(x=x, y=coverage(x)), color='blue') + xlab('bias/sd') + ylab('coverage')
```
The coverage of a 95% confidence interval $\hat\theta \pm 1.96\ \sd[\hat\theta]$ when $\hat\theta$ has a normal sampling distribution.
:::

::: {#fig-thetaratio}

```{r}
x=seq(0,1,by=.01)
ggplot() + 
  geom_line(aes(x=x, y=x/sqrt(x * (1-x)))) + scale_y_continuous(limits=c(0,10), breaks=seq(0,10,by=1)) + labs(x='', y='') 
```
This is a plot of $f(x) = x/\sqrt{x(1-x)}$.
:::
