---
title: "Midterm 2"
subtitle: "QTM 285-1"
---

{{< include shared-code.qmd >}}
{{< include ca-income.qmd >}}

```{r}
#| echo: false
theme_set(theme_minimal())
options(scipen = 1, digits = 2)
```

```{r}
#| echo: false
#| warning: false
library(patchwork)
```

# Problem 1

Suppose we have a sample of $(X_1,Y_1) \ldots (X_n,Y_n)$ drawn with replacement from a population. It's stored in vectors $X$ and $Y$ of length $n$. $X_i$ can take the values of 1,2,3,4 or 5. 
```{r}
#| echo: false
set.seed(123)
n = 1000
X = c(rep(1,350), rep(2,300),rep(3,200),rep(4,100),rep(5,50))
Y = X + rnorm(n,0,2)
df<- data.frame(x=X,y=Y)
```

Here's what the data looks like.

```{r}
#| out-height: 3.5in
#| fig-ncol: 2
#| fig-subcap:
#|  -  Histogram of the vector Y's elements
#|  -  Jittered Scatterplot of X and Y
ggplot() + geom_histogram(aes(x=Y,     y=after_stat(density)), alpha=.2, bins=50) + 
           geom_vline(aes(xintercept=mean(Y)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(Y)+c(-1,1)*sd(Y)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(Y), xmin=mean(Y)-1.96*sd(Y), xmax=mean(Y)+1.96*sd(Y), y=.025)) + theme_bw() + xlab('') + ylab('')

ggplot() + geom_point(aes(x=X, y=Y), alpha=.1, position=position_jitter(w=.2, h=0)) + xlab('') + ylab('') +
           stat_summary(aes(x=X, y=Y), geom='pointrange', 
              fun=mean, 
              fun.min=function(x) mean(x)-sd(x),
              fun.max=function(x) mean(x)+sd(x))  
```


We run this code.

```{r}
#| echo: true
B=1000

thetastar =  rep(NA,B)
for( b in 1:B ){
  Jstar = sample(1:n,size=n,replace=TRUE)
  Xstar = X[Jstar]
  Ystar = Y[Jstar]
  thetastar[b] = (1/4)*(mean(Ystar[Xstar==5]) - mean(Ystar[Xstar==1]))
}
```

```{r}
#| out-height: 3.5in
#| fig-cap: Histogram of the thetastar's elements
ggplot() + geom_histogram(aes(x=thetastar, y=after_stat(density)), alpha=.2) + 
           geom_vline(aes(xintercept=mean(thetastar)), alpha=.5) + 
           geom_vline(aes(xintercept=mean(thetastar)+c(-1,1)*sd(thetastar)), linetype='dashed', alpha=.5) + 
           geom_pointrange(aes(x=mean(thetastar), xmin=mean(thetastar)-1.96*sd(thetastar), xmax=mean(thetastar)+1.96*sd(thetastar), y=1)) + xlab('') + ylab('')

```


## Part A
What does this code calculate? State it in words and mark it with an 'a', and an arrow or bracket
if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
mean(Y)
```

::: {.callout-solution}
The mean of the sample $Y_1 \ldots Y_n$.
:::

## Part B
What does this code calculate? State it in words and mark it with a 'b', and an arrow or bracket if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
sd(Y)
```

::: {.callout-solution}
The standard deviation of the sample $Y_1 \ldots Y_n$.
:::


## Part C
What does this code calculate? State it in words and mark it with a 'c', and an arrow or bracket
if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
mean(Y[X==1])
```

::: {.callout-solution}
The mean of the outcomes $Y_1 \ldots Y_n$  that are in the subsample (or column) where $X_i=1$.  
:::

## Part D   
What does this code calculate? State it in words and mark it with a 'd', and an arrow or bracket if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
sd(thetastar)
```

::: {.callout-solution}
The bootstrap standard deviation of an estimator you could reasonably describe two ways.

1. The mean of the four increments $\hat\mu(x)-\hat\mu(x-1)$ for $x=2,3,4,5$. 
2. A fourth of the difference between the mean outcome in the subsamples where $X_i=5$ and $X_i=1$.
:::


## Part E
What does this code calculate? State it in words and mark it with a 'e', and an arrow or bracket if necessary for clarity, on the appropriate plot above.
```{r}
#| echo: true
#| eval: false
(1/4)(mean(Y[X=5]) - mean(Y[X=1])) 
- 1.96* sqrt((1/4)^2*var(Y[X=5])/length(Y[X=5]) + (-1/4)^2*var(Y[X=1])/length(Y[X=1]))

(1/4)(mean(Y[X=5]) - mean(Y[X=1]))
+ 1.96* sqrt((1/4)^2*var(Y[X=5])/length(Y[X=5]) + (-1/4)^2*var(Y[X=1])/length(Y[X=1]))
```

::: {.callout-solution}
A 95% interval based on normal approximation centered on the same estimator as in Part D. 
:::


## Part F
How could you calculate roughly the same thing as in Part E if you knew `sd(thetastar)` 
but not `Var(Y[X=5])` and `Var(Y[X=1])`?

::: {.callout-solution}
Replace the formula-based standard deviation estimate---the thing that comes after the 1.96---with the bootstrap standard deviation `sd(thetastar)`.
:::

# Problem 2


```{r}
ca = earnings 
sam = data.frame(
  w = 0 + (ca$sex == 'female'),
  x = ca$education,
  y = ca$income)
```

```{r}
summaries = sam |> 
    group_by(w,x) |> 
    summarize(muhat = mean(y), sigmahat = sd(y), Nwx = n(), .groups='drop')
  
summary.lookup = function(column.name, summaries, default=NA) { #<1>
  function(w,x) { 
    out = left_join(data.frame(w=w,x=x),                              #<2>
                    summaries[, c('w','x',column.name)],              #<2>
                    by=c('w','x')) |>                                 #<2>
      pull(column.name)                                               #<2>
    out[is.na(out)] = default                                         #<3>
    out
  }
}

muhat    = summary.lookup('muhat',    summaries)
sigmahat = summary.lookup('sigmahat', summaries)
Nwx      = summary.lookup('Nwx',      summaries)

W = sam$w
X = sam$x
Y = sam$y
x = unique(X)
n = length(Y)
```

```{r}
proportion.scale = 350000
plotX = function(w,x, shift=default.shift) { ifelse(w==0, x-shift, x+shift) }
interleaved.bars = list(
  geom_bar(aes(x=plotX(w,x), y=proportion.scale*after_stat(prop), 
               fill=factor(w), width=w), 
           position='identity', alpha=.1, linewidth=.05, color='black'))
purple.bars = geom_bar(aes(x=x, y=proportion.scale*after_stat(prop)), 
                       alpha=.1, linewidth=.05, fill='purple', color='black', width=.2)
scales = list(
  scale_y(sec.axis=sec_axis(~./proportion.scale, name='proportion')), 
  scale_x(),
  coord_cartesian(ylim=c(0,1.5e5)),
  guides(color='none', fill='none', size='none') 
)
```

```{r}
m0x.mild     = rep(c(10,12,14,16,18), rev(c(80, 90, 100, 110, 120)))
m0x.moderate = rep(c(10,12,14,16,18), rev(c(50, 75, 100, 125, 150)))
m0x.extreme  = rep(c(10,12,14,16,18), rev(c(2,49, 100, 149, 198)))
m0x.flat = rep(c(10,12,14,16,18), c(100,100,100, 100,100))
m1x.mild     = rep(c(10,12,14,16,18), c(80, 90, 100, 110, 120))
m1x.moderate = rep(c(10,12,14,16,18), c(50, 75, 100, 125, 150))
m1x.extreme  = rep(c(10,12,14,16,18), c(2,49, 100, 149, 198))
m1x.flat = rep(c(10,12,14,16,18), c(100,100,100, 100,100))

exercise.pop = function(m0x, m1x, mu=NULL) {
  if(is.null(mu)) { 
    fit = lm(y ~ w*factor(x), data=sam)
    mu = function(w,x) { predict(fit, newdata=data.frame(w=w, x=x)) }
  }
  eps0 = rnorm(length(m0x))*sigmahat(0,m0x)/3
  eps1 = rnorm(length(m1x))*sigmahat(1,m1x)/3
  for(uu in unique(m1x)) { 
    eps0[m0x==uu] = eps0[m0x==uu] - mean(eps0[m0x==uu])
    eps1[m1x==uu] = eps1[m1x==uu] - mean(eps1[m1x==uu])
  }
  rbind(
    data.frame(w=0, x=m0x, y=mu(0,m0x)+eps0),
    data.frame(w=1, x=m1x, y=mu(1,m1x)+eps1))
}

exercise.pop.mild = exercise.pop(m0x.mild, m1x.mild)
exercise.pop.moderate = exercise.pop(m0x.moderate, m1x.moderate)
exercise.pop.extreme = exercise.pop(m0x.extreme, m1x.extreme)

set.seed(0)
J.sam = sample(1:nrow(exercise.pop.moderate), 500, replace=TRUE)
exercise.sam = exercise.pop.moderate[J.sam,]
n = nrow(exercise.sam)
```

```{r}
Delta0.target = function(pop=NULL) {
  function(mu, sam) { 
    if(is.null(pop)) { pop = sam }
    mean(mu(1,pop$x[pop$w==0]) - mu(0,pop$x[pop$w==0]))
  }
}
with.constant.weights = function(sam) { 
  sam$weights = 1  
  sam
}
with.ipw = function(sam, from=sam) { 
  summaries = from |> group_by(w,x) |> summarize(Nwx=n(), .groups='drop')
  Nwx = summary.lookup('Nwx', summaries)
  sam$weights = Nwx(0,sam$x) / Nwx(sam$w,sam$x)
  sam 
} 

plot.sampling.distribution = function(pop, formula, n = nrow(sam), 
target=Delta0.target(), pop.version=FALSE, replications=5000,
with.weights=with.constant.weights, seed=1) { 
    estimates = with_seed(function() { 
      1:replications |> map_vec(function(rep) { 
        J = sample(1:nrow(pop), size=n, replace=TRUE)
        sam = pop[J,] |> with.weights() 

        fitted.model = lm(formula, data=sam, weights=weights) 
        muhat = function(w,x) { predict(fitted.model, newdata=data.frame(w=w, x=x)) } 
        target(muhat, sam)
    })}, seed=seed)
    
    mu.model = lm(y~w*factor(x), data=pop)
    mu = function(w,x) { predict(mu.model, newdata=data.frame(w=w, x=x)) }
    the.target = target(mu, pop)
    

    mutilde.model = lm(formula, data=pop |> with.weights(), weights=weights)
    mutilde = function(w,x) { predict(mutilde.model, newdata=data.frame(w=w, x=x)) }
    the.pop.version = target(mutilde, pop)
    
    p = ggplot() + 
     geom_vline(xintercept = the.target, color='green', linewidth=2, alpha=.5) 
    if(pop.version) { p = p + geom_vline(xintercept = the.pop.version, color=pink, linewidth=2, alpha=.5) }
    p +
     geom_vline(xintercept = mean(estimates), color='blue', linewidth=.75, alpha=.9) +
     geom_histogram(aes(x=estimates, y=after_stat(density)), bins=50, alpha=.2)
}
```

```{r}
default.shift = .15
biggest.max.weight = exercise.pop.extreme |> with.ipw() |> pull(weights) |> max()
lsq.plot = function(sam, formula=NULL, shift=default.shift, line.alpha=.7, size.scale=.5) {
  if(is.null(sam$weights)) { 
    sam$weights = rep(1, nrow(sam))
  }
  max.weight = max(sam$weights)
  max.size = size.scale * max.weight 
  
  base.plot = ggplot(sam) + 
    geom_point(aes(x=plotX(w,x), y=y,color=factor(w), size=I(weights)), 
      position=position_jitter(seed=0, width=shift),
      alpha=.3) + 
    stat_summary(aes(x=plotX(w,x), y=y, fill=factor(w)), 
      fun.data=mean_sd, geom='pointrange', 
      shape=21, alpha=.5, color='black') +
    interleaved.bars + purple.bars + scales
    
    if(is.null(formula)) { base.plot } else { 
        model = lm(formula, data=sam, weights=weights)
        gridx = seq(-10, 50, length.out=1000)
        muhat = function(w,x) predict(model, newdata=data.frame(w=w,x=x))   
        
        the.plot = base.plot 
        if(line.alpha > 0) { 
            the.plot = base.plot + 
            geom_line(aes(x=plotX(w,x), y=muhat(w,x), color=factor(w)), 
              data = expand_grid(w=0:1, x=gridx), alpha=line.alpha, linetype='dashed') 
        }
        the.plot + 
          geom_point(aes(x=plotX(w,x), y=muhat(w,x), fill=factor(w)),  
            color='black', alpha=.5, shape=23)
    }
}
```

```{r}
#| out-width: 100%
exercise.pop(m1x.moderate, m0x.moderate) |>  lsq.plot(y ~ 1+w)
```
                                                     

Consider the following population summaries.
$$
\begin{aligned}
\Delta_{\text{raw}} &= \frac{1}{m_1}\sum_{j:w_j=1} y_j - \frac{1}{m_0}\sum_{j:w_j=0} y_j \qfor m_w = \sum_{j:w_j=w} 1 \\
\Delta_{0}   &= \frac{1}{m_0}\sum_{j:w_j=0} \lcb{ \mu(1,x_j) - \mu(0,x_j) }  \\
\Delta_{1}   &= \frac{1}{m_1}\sum_{j:w_j=1} \lcb{ \mu(1,x_j) - \mu(0,x_j) } \\ 
\Delta_{\text{all}}   &= \frac{1}{m}\sum_{j=1}^m \lcb{ \mu(1,x_j) - \mu(0,x_j) }  
\end{aligned}
$$
                                       
::: {.callout-exercise}
### Part A
In the population shown above, with annotations interpreted as usual, which is bigger, $\Delta_{\text{raw}}$ or $\Delta_{\text{1}}$?
:::

::: {.callout-solution}
$\Delta_{\text{1}}$ is bigger. 

*Explanations*.

1. A visual one. The raw difference---the height of the green horizontal line minus that of the red one --- is smaller (i.e. further below zero)
than any of the within-column differences in means that we average to get $\Delta_{\text{1}}$.

2. A math one. The raw difference in means $\Delta_\text{raw}$
is equal the sum of $\Delta_{\text{1}}$ and a covariate shift term $\sum_x \ingroupb{p_{x \mid 1}}\  \ingroupa{\mu(0,x)} - \sum_x \ingroupa{p_{x \mid 0} \  \mu(0,x)}$
The function $\ingroupa{\mu(0,x)}$ is increasing from left to right and our covariate shift from $\ingroupa{p_{x \mid 0}}$ to $\ingroupb{p_{x \mid 1}}$ is from right to left, from where $\ingroupa{\mu(0,x)}$ is big to where it's small, so the covariate shift term is negative. $\Delta_{\text{raw}} = \Delta_1 + \text{negative covariate shift term}$,
so $\Delta_\text{raw} < \Delta_1$. See [Lecture 9](`r lecture(9)`) for reference. 

:::

::: {.callout-exercise}
### Part B
In the population shown above, with annotations interpreted as usual, which is bigger, $\Delta_{\text{1}}$ or $\Delta_{\text{0}}$?
:::

::: {.callout-solution}
$\Delta_{1}$ is bigger. 

*Explanation*.

The within-column differences $\ingroupb{\mu(1,x)} - \ingroupa{\mu(0,x)}$ we 
average to get both $\Delta_1$ and $\Delta_0$ decrease from left to right.
$\Delta_0$ puts more weight on the the differences on the right
and $\Delta_1$ puts more weight on the differences on the left,
and the ones on the left are bigger, so $\Delta_1$ is bigger.
:::

::: {.callout-exercise}
### Part C
In the population shown above, with annotations interpreted as usual, which is bigger, $\Delta_{\text{1}}$ or $\Delta_{\text{all}}$?
:::

::: {.callout-solution}
$\Delta_1$ is bigger. 

*Explanation*.

$\Delta_{\text{all}}$ is a weighted average of $\Delta_{\text{1}}$ and $\Delta_{\text{0}}$,
so it's between them (see the [Week 7 Homework](`r homework(7, 'exr-shift-between-adjusted-all')`)). 
Since $\Delta_1$ is bigger than $\Delta_0$, it follows that $\Delta_1 > \Delta_{\text{all}} > \Delta_0$.
:::

::: {.callout-exercise}
### Part D
Suppose I'm talking about an entirely new population in which $\Delta_{\text{1}} < \Delta_{\text{0}}$.
Which is bigger, $\Delta_{\text{1}}$ or $\Delta_{\text{all}}$? If you can't say for sure, explain why.
:::

::: {.callout-solution}
$\Delta_{\text{all}}$ is bigger. 

*Explanation*.

It's the reasoning as in Part C, but now with $\Delta_1 < \Delta_0$ and therefore $\Delta_1 < \Delta_{\text{all}} < \Delta_0$.
:::

# Problem 3

Consider the following code.

```{r}
#| echo: true
#| eval: false 

sampling.dist = function(pop) { 
  map(1:10000, function(.) { 
     J = sample(1:m, size=n, replace=TRUE)                                      #<1>
     sam = pop[J,]                                                              #<1>
     model.fit = lm(y ~ w*factor(x), data=sam)                                  #<2>
     muhat = function(w,x) { predict(model.fit, newdata=data.frame(w=w, x=x)) } #<2>
     thetahat = mean(muhat(1,sam$x) - muhat(0,sam$x))
     thetahat
  }
}

estimator.samples  = sampling.dist(pop) 
bootstrap.samples  = sampling.dist(sam)
bootstrap.width = 2*1.96*sd(bootstrap.samples) 

coverage.probability = 
  mean(estimator.samples - bootstrap.width/2 <= theta & 
       theta <= estimator.samples + bootstrap.width/2)
```
1. Here we're drawing a sample of size `r n` with replacement from our population. This is exactly how we got our sample `sam` from the population `pop`.
2. We're using the 'all functions' model here, so this is just the mean of the observations with $W_i=w$ and $X_i=x$.

Suppose you're interested in checking out the interval estimator $\hat\theta \pm 1.96\text{sd}(\hat\theta^\star)$ 
where
$$ 
\begin{aligned}
\hat\theta = \frac{1}{n}\sum_{i=1}^n \qty{\mu(1,X_i) - \mu(0,X_i)} 
&&\qfor \hat\mu(w,x)=\frac{1}{N_{wx}}\sum_{i:W_i=w,X_i=x} Y_i \\
&&\qand N_{wx} = \sum_{i:W_i=w,X_i=x} 1
\end{aligned}
$$
and $\text{sd}(\hat\theta^\star)$ is the standard deviation of draws from the bootstrap sampling distribution of $\hat\theta$.
To do this, you're seeing if it works on data sampled with replacement from the made-up population `pop`.
After defining `theta` to be $\frac{1}{m}\sum_{i=1}^m \qty{\mu(1,x_j) - \mu(0,x_j)}$
and drawing a sample `sam` with replacement from your population `pop`, you run the code above.


::: {.callout-exercise}
### Part A
Explain what you'd do to get an accurate estimate of this interval estimator's coverage probability. 
'Just report `coverage.probability`' is technically an answer, but I hope it's not yours. You don't
have to write code, but you should be able to describe what you'd do in a way that a good programmer 
who didn't know anything about statistics could write the code.
:::

::: {.callout-solution}
The problem with this code is that we're calculating width based on the one sample `sam`
that exists outside the `map` call. [We should be calculating width exactly the same way,
but inside the `map` call, using the same sample used to calculate the point estimate 
we're going to center our interval at.]{.target} 

To fix this, copy the code for the function `sampling.dist` then,
after the line that defines `theta.hat`, paste in the lines that define `bootstrap.samples` and `bootstrap.width`
and then check if `theta` is in the interval `theta.hat - bootstrap.width/2` to `theta.hat + bootstrap.width/2`. 
The result will be a function that takes a population and returns 10000 TRUE/FALSE values indicating whether each of 
the 10000 interval estimators we calculated covers the true parameter. Average them to get the coverage probability.

Hopefully, if you tell someone to do that, you'll get this.
```{r}
#| eval: false
coverage.samples = function(pop) { 
  map(1:10000, function(.) { 
     J = sample(1:m, size=n, replace=TRUE)                                      
     sam = pop[J,]                                                              
     model.fit = lm(y ~ w*factor(x), data=sam)                                  
     muhat = function(w,x) { predict(model.fit, newdata=data.frame(w=w, x=x)) } 
     thetahat = mean(muhat(1,sam$x) - muhat(0,sam$x))
     bootstrap.samples = sampling.dist(sam)
     bootstrap.width = 2*1.96*sd(bootstrap.samples)
     interval.covers = thetahat - bootstrap.width/2 <= theta & theta <= thetahat + bootstrap.width/2
     interval.covers
  }
}
coverage.probability = mean(coverage.samples(pop))
```

*What I was looking for.*
The first paragraph, the second, or some approximation of the code above would each be a good answer on their own in my book.
What I was looking for was that you'd either written something like the second sentence of the first paragraph---the one I 
[highlighted like this]{.target}---or some implementation of that idea.
:::

::: {.callout-exercise}
### Part B
Suppose that the code above is your friend's, and when they see your answer to Part A, 
they say "That seems like a lot of work. We both calculated the coverage probability to be 0.95. I think mine's good enough."
Explain what's wrong with their approach. 


*Tip.* Try to provide a simple example in which their approach will often (i.e. in most samples `sam` drawn with replacement from the population) arrive at the correct coverage probability, but won't doesn't always. Your estimator and target don't have to be this one. Perhaps it's easier to explain if you're just estimating the population's mean. If you like, go ahead and assume the outcome is binary
so that mean is a proportion.
:::

::: {.callout-solution}
What's wrong is that the width is calculated based on a different sample than the one used to calculate the point estimate.
The conceptual issue is that when we're simulating the process of constructing an interval estimator to check coverage, 
our sample is the one inside the `map` call---and our width has the be based on that sample. 

One reason the code above might calculate about the same thing as the correct code is that every one of these widths, the ones our fixed-up
code calculates inside the `map` call and the one the incorrect code calculates outside it, is an estimate of the same thing:
the width `2*1.96*sd(estimator.samples)` of our estimator's actual sampling distribution. If those estimates are all close to
the width they're estimating, then it won't matter much whether we do it inside the `map` call or outside it. This tends 
to happen when the sample is large. We talked about this in the [Week 5 Homework](`r homework(5)`).

But, especially in smaller samples, this approach can get us in trouble. Suppose we're using the incorrect approach above 
to calculate the coverage probability of an interval estimator of a population proportion $\theta$ centered on the sample proportion $\hat\theta$.
If $\theta=.2$ and the sample proportion $\hat\theta$ is $.3$, then our estimator's standard deviation is $\sqrt{.2 \times .8/n}$ and we've
estimated it to be $\sqrt{.3 \times .7/n}$, roughly $\sqrt{.3 \times .7 / .2 \times .8} \approx `r sqrt(.3*.7/(.2*.8))`$ bigger 
than our estimator's actual standard deviation. Using intervals based on this one overestimate of our estimator's standard deviation
around each of our 10000 point estimates, rather than intervals that are sometimes wider and sometimes narrower than ones based on our estimator's actual standard deviation
as we would if we calculated width within the `map` call, will tell us our interval estimator's coverage probability is larger than it actually is.
:::


# Problem 4 

In the plot below, I'm showing some data from a (fake) survey of people in your city who've recently finished with school and begun working.
For whatever reason, nobody who attends school in this city goes to college, so everyone finishes with between 8 and 12 years of education.
There are 10,000 of these people, and we mailed out (and received back) 2000 surveys, sampling from this population of 10,000 with replacement.
The x-axis shows the survey recipient's years of education, the y-axis shows their income, the points with arms are within-column means and standard deviations, and the heights of the bars are proportional to the proportion of recipients with each level of education.

```{r}
#| out-width: 100%
set.seed(1)
xs = 8:12
p  = \(x) (x-6)/sum(xs-6)
e  = \(x) (x-6)^2/sum((xs-6)^2)
mu = \(w, x) exp(x/2)/max(exp(xs/2)) * 50000

m = 10000
pop.x =  sample(xs, size=m, replace=TRUE, prob=p(xs))
pop.w =  0 + (runif(m) <= e(pop.x))
eps = runif(m, min=-1, max=1)*mu(pop.w, pop.x)
pop.y =  mu(pop.w, pop.x) + eps
pop = tibble(x=pop.x, w=pop.w, y=pop.y)

n = 2000
sam = pop[sample(1:m, n, replace=TRUE), ]

mean_sd = \(x, mult=1) {
  data.frame(y=mean(x)*mult, ymin=mean(x)-sd(x)*mult, ymax=mean(x)+sd(x)*mult)
}

ggplot(sam) + 
  geom_point(aes(x=x, y=y), alpha=.1, size=1, position=position_jitter(w=.2,h=0)) + 
  stat_summary(aes(x=x, y=y), fun.data=mean_sd, geom='pointrange') +
  geom_bar(aes(x=x, y=max(pop$y)*after_stat(prop)), alpha=.1, color='black', linewidth=.05) + 
  scale_y_continuous(labels=label_dollar()) +
  xlab('') + ylab('')  
```

::: {.callout-exercise}
### Part A
Consider two summaries of the incremental value of a year of high school in this sample.

$$
\begin{aligned}
\hat\theta_{\text{years}}  &= \frac{\sum_{x \in 9 \ldots 12}     \qty{\hat\mu(x) - \hat\mu(x-1)}}{4} \\
\hat\theta_{\text{people}} &= \frac{\sum_{i:X_i \in 9 \ldots 12} \qty{\hat\mu(X_i) - \hat\mu(X_i-1)}}{\sum_{i:X_i \in 9\ldots 12} 1} \\
\end{aligned}
$$

Which is larger? Explain how you know.
:::

::: {.callout-solution}
$\hat\theta_{\text{people}}$ is larger. The increments $\hat\mu(x) - \hat\mu(x-1)$ get larger as $x$ increases,
and so does the proportion of individuals with that level of education (i.e. the number of dots in the $\text{years}=x$ column). In $\hat\theta_{\text{people}}$, these larger increments get more weight because there are more with these lager values of $x$; in $\hat\theta_{\text{years}}$ all increments get equal weight.
:::


These survey recipients went to two schools: [School 0]{.groupa} and [School 1]{.groupb}
```{r}
#| out-width: 100%
plotX = \(w,x) x + .15*(2*w-1)

phat = (\(w,x) { sum(sam$x==x & sam$w==w)  / sum(sam$w==w) }) |> Vectorize()

ggplot(sam) + 
  geom_point(aes(x=plotX(w,x), y=y, color=factor(w)), alpha=.1, size=1, position=position_jitter(w=.075,h=0)) + 
  stat_summary(aes(x=plotX(w,x), y=y, color=factor(w)), fun.data=mean_sd, geom='pointrange') +
  geom_col(aes(x=plotX(w,x), y=max(pop$y)*phat(w,x)), alpha=.1, color='black', linewidth=.05, 
           data=expand_grid(w=c(0,1), x=xs)) + 
  scale_y_continuous(labels=label_dollar()) +
  guides(color='none') +
  xlab('') + ylab('')  

```

Children in your district go to School 0.  Your cousin, who lives in a district that goes to School 1, encourages you to pretend that you and your family live with them so that your child can go to School 1. This is a thing that happens, at least on the TV show 'Friday Night Lights.' Their argument is that School B must be better because the average income of students who have attended it is clearly higher. 
Check out this confidence interval for the difference in mean incomes of students who have attended the two schools, they say.

```{r}
#| echo: false
Y = sam$y
W = sam$w
X = sam$x
n = nrow(sam)

width = function(samples, center=mean(samples), alpha=.05) {
  covg.dif = function(w) { 
    actual = mean( center - w/2 <= samples & samples <= center + w/2)
    nominal = 1-alpha 
    actual - nominal  
  }
  uniroot(covg.dif, interval=c(0, 2*max(abs(samples-center))))$root
}
```

```{r}
#| echo: true

Delta.hat = mean(Y[W==1]) - mean(Y[W==0])

Delta.hat.star = replicate(1000, {
  J.star = sample(1:n, size=n, replace=TRUE)
  W.star = W[J.star]
  X.star = X[J.star]
  Y.star = Y[J.star]
  
  mean(Y.star[W.star==1]) - mean(Y.star[W.star==0])
})

interval.width = width(Delta.hat.star)
confidence.interval = c(Delta.hat - interval.width/2, 
                        Delta.hat + interval.width/2)
confidence.interval
```

They even based their calculation on a lecture from the class website, so it's got to be legit. School 1 is at least \$`{r} round(min(confidence.interval))` better. 

You, on the other hand, look at the data and claim that a year at School 0 is worth just as much as a year at School 1. And you're right. You calculate school-specific versions of $\hat\theta_{\text{years}}$, bootstrap the difference between them, 
and find convincing evidence that the incremental value of a year at each school is the roughly the same.

```{r}
#| echo: true

your.summary = function(W,X,Y) {
  model.fit = lm(y ~ w*factor(x), data=data.frame(w=W, x=X, y=Y))            #<1>
  muhat = function(w,x) { predict(model.fit, newdata=data.frame(w=w, x=x)) } #<1>
  (mean(muhat(1,9:12)-muhat(1,8:11))) - (mean(muhat(0,9:12)-muhat(0,8:11)))
}

theta.hat.years.difference = your.summary(W,X,Y)

theta.hat.years.difference.star = replicate(100, {
  J.star = sample(1:n, size=n, replace=TRUE)
  your.summary(W[J.star], X[J.star], Y[J.star])
})
 

interval.width = width(theta.hat.years.difference.star)
confidence.interval = c(theta.hat.years.difference - interval.width/2, 
                        theta.hat.years.difference + interval.width/2)
confidence.interval
```
1. We're using the 'all functions' model here, so this is just the mean of the observations with $W_i=w$ and $X_i=x$.

::: {.callout-exercise}
### Part B
Explain, in terms your cousin can understand, how it's possible that the average income of students who have attended School 1 is higher than the average income of students who have attended School 0, but the average incremental value of a year at both schools is the same.
:::

::: {.callout-solution}
Students at School 1 tend to stay in school longer than students at school 0, so even though
the incremental value of each year is the same at the two schools, the students at School 1 tend to 
*actually get* more of this value because they stay in school for more of these years.
:::
