---
title: "QTM 220: Lectures 17"
subtitle: "Least Squares Regression in Linear Models: Part 2"
format: 
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    default-image-extension: svg
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---


```{r load-libraries}
library(purrr)
library(tibble)
library(dplyr)
library(magrittr)
library(stringr)
library(ggplot2)
library(htmltools)
library(kableExtra)
library(scales)
library(cpsR)
library(maps)

options(knitr.kable.NA = '')

census.api.key = '8084f81e4f71e4d32638dac15f188b424f72d063'
# for cps variable interpretation see https://www2.census.gov/programs-surveys/cps/datasets/2022/march/asec2022_ddl_pub_full.pdf
percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }
```

```{r ggplot-themes}
fontsize = 16
textcolor = "#888888"

class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
        panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x  = element_text(colour = textcolor, size=fontsize),
        axis.text.y  = element_text(colour = textcolor, size=fontsize),
        axis.title.x  = element_text(colour = textcolor),
        axis.title.y  = element_text(colour = textcolor, angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
              axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
           panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(class.theme)
```


```{r table-styling}
ellipsis <- function(df, nhead=5L, ntail=2) {
    rownames(df) = as.character(1:nrow(df))
    stopifnot("data.frame" %in% class(df))
    els <- rep(NA, ncol(df)) %>% 
        matrix(nrow=1, dimnames=list(NULL, names(df))) %>% 
        data.frame(stringsAsFactors=FALSE) %>% 
        tibble::as_tibble()
    rownames(els) = "⋮"
    dplyr::bind_rows(head(df, nhead), els, tail(df, ntail))
}
```

```{r get-data}
#| cache: true
#| 
# for cps variable interpretation see https://www2.census.gov/programs-surveys/cps/datasets/2022/march/asec2022_ddl_pub_full.pdf
census.api.key = '8084f81e4f71e4d32638dac15f188b424f72d063'
cps.sampling.ratio = 1/25 # Not true. 1/2500 households are surveyed. But for the sake of plotting, it's ok.

years = c(2019,2020,2021,2022)
cps.data = purrr::map(years, function(year) { 
  data=get_asec(year=year, vars=c('pearnval', 'a_age', 'a_hga', 'gestfips', 'gtco', 'gtcbsasz'), key=census.api.key)
  data$year=year
  data
}) %>% bind_rows()
```

```{r transform-data}
#| cache: false
#| 
all.earnings = data.frame(state.code = cps.data$gestfips,
                          county.code = cps.data$gtco,
                          year=cps.data$year,
                          age=cps.data$a_age,
                          income=cps.data$pearnval,
                          education.code=cps.data$a_hga,
                          education=case_match(cps.data$a_hga,
                                           0  ~ 0,   # child
                                           31 ~ 0,   # < grade 1
                                           32 ~ 4,   #  grade 1-4
                                           33 ~ 6,   #  grade 5-6
                                           34 ~ 8,   #  grade 7-8 
                                           35 ~ 9,   #  grade 9
                                           36 ~ 10,  #  grade 10
                                           37 ~ 11,  #  grade 11
                                           38 ~ 11,  #  grade 12 no diploma
                                           39 ~ 12,  #  high school grad
                                           40 ~ 13,  #  some college
                                           41 ~ 14,  #  associate degree (vocational)
                                           42 ~ 14,  #  associate degree (academic)
                                           43 ~ 16,  #  bachelors degree
                                           44 ~ 18,  #  masters degree
                                           45 ~ 20,  #  professional school degree
                                           46 ~ 20)) #  doctorate

all.earnings$degree = factor(all.earnings$education >= 16, 
                     levels=c(FALSE, TRUE), 
                     labels=c('no 4-year degree','4-year degree')) 

coarsen = function(x, group, seed=0) {
  current.seed = .Random.seed
  set.seed(seed)
  x.coarsened = rep(NA, length(x))
  for(gg in unique(group)) {
    x.coarsened[group==gg] = runif(sum(group==gg), min(x[group==gg]), max(x[group==gg]))
  }
  set.seed(current.seed)
  x.coarsened
} 
```

# Linear Regression Models 

::: {.hidden}
$$
\newcommand{\alert}[1]{\textcolor{red}{#1}}
\newcommand{\model}{\mathcal{M}}
\newcommand{\R}{\mathbb{R}}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator{\V}{V}
\DeclareMathOperator{\Var}{\V}
\newcommand{\inred}[1]{\textcolor{red}{#1}}
\newcommand{\inblue}[1]{\textcolor{blue}{#1}}
\newcommand{X}{\mathbf{X}}
\newcommand{Y}{\mathbf{Y}}
$$ 
:::


## Least Squares Regression with Linear Models {.smaller}

$$
\hat \mu = \argmin_{m \in \model} \frac{1}{n}\sum_{i=1}^n \qty{ m(\textbf{X}_i) - Y_i}^2 \qqtext{ for } \model = \qty{ m(x) = \sum_k b_k \phi_k(x) : b \in \R^{K+1} }
$$

- In this class, we focus on fitting **linear models** via least squares.
- You can think of linear models, generally speaking, as the set of **all linear combinations** of a set of **features**.
- We've been doing this for a while, but it was only last class that we talked about it in general.
- Let's go through a few familiar examples.


## The Saturated Model
```{r}
#| out-height: 400px
 
earnings = all.earnings %>% filter(education >= 8 & state.code==36 & county.code==61)
group.summaries = earnings %>% group_by(year, degree) %>% summarize(mean=mean(income), sd=sd(income), n=n())
summaries = earnings %>% group_by(year) %>% summarize(mean=mean(income), sd=sd(income), n=n())

saturated.model.plot = ggplot(earnings) + 
  geom_point(aes(x=year, y=income, color=degree), alpha=.4, size=.5, position=position_jitter(width=.1)) +
  geom_pointrange(aes(x=year, y=mean, ymin=mean-sd/sqrt(n), ymax=mean+sd/sqrt(n), color=degree), data=group.summaries) +
  scale_y_continuous(labels=dollar_format(), limits=c(0, 200000)) 
saturated.model.plot
```

- This is the model we've worked with most.
- In this model, we have a feature for every level of $x$.
- That feature is just a binary indicator that $X_i$ takes on that value.
- The least squares curve in this model is just the **subsample mean** of $Y_i$ at each level of $x$

$$
\small{
\hat \mu(x) = \frac{1}{N_x}\sum_{i:X_i=x} Y_i \qqtext{ for } \model = \qty{ m(x) = \sum_j b_k 1_{x_k}(x) : b_k \in \R } 
}
$$    


## Lines and Polynomials {.smaller}

```{r}
#| out-height: 400px
 
line.fit = lm(income ~ year, data=earnings)
quad.fit = lm(income ~ poly(year,2), data=earnings)
cube.fit = lm(income ~ poly(year,3), data=earnings)
x = seq(min(earnings$year), max(earnings$year), by=.01)
newdata = data.frame(year = x)
newdata$muhat.line  =  predict(line.fit, newdata)
newdata$muhat.quad  =  predict(quad.fit, newdata)
newdata$muhat.cube  =  predict(cube.fit, newdata)

saturated.model.plot +
 geom_pointrange(aes(x=year, y=mean, ymin=mean-sd/sqrt(n), ymax=mean+sd/sqrt(n)), data=summaries) +
 geom_line(aes(x=year, y=muhat.line), data=newdata, alpha=.6, color='blue') +
 geom_line(aes(x=year, y=muhat.quad), data=newdata, alpha=.8, color='red') + 
 geom_line(aes(x=year, y=muhat.cube), data=newdata, alpha=.6, color='magenta') 
```

- We've also worked with the set of **lines**.
- But with many-valued $x$, we started having to worry about **misspecification**.
- To reduce misspecification, people often use higher-order **polynomials** instead.

$$\small{
\begin{aligned}
\model &= \qty{ m(x) = b_0 + b_1 x : b \in \R^2 } && \textcolor{blue}{\text{lines}} \\
\model &= \qty{ m(x) = b_0 + b_1 x + b_2 x^2: b \in \R^3 } && \textcolor{red}{\text{quadratic polynomials}} \\
\model &= \qty{ m(x) = b_0 + b_1 x + b_2 x^2 + b_3 x^3 : b \in \R^4 } && \textcolor{magenta}{\text{cubic polynomials}}
\end{aligned}
}
$$

## Bivariate Line-Based Models Models {.smaller}

```{r}
#| out-height: 400px
 
additive.line.fit   = lm(income ~ degree+poly(year,1), data=earnings)
interacted.line.fit = lm(income ~ degree*poly(year,1), data=earnings)
newdata = expand.grid(year = x, degree=unique(earnings$degree))
newdata$muhat.additive    =  predict(additive.line.fit, newdata)
newdata$muhat.interacted  =  predict(interacted.line.fit, newdata)

saturated.model.plot +
 geom_pointrange(aes(x=year, y=mean, ymin=mean-sd/sqrt(n), ymax=mean+sd/sqrt(n)), data=summaries) +
 geom_line(aes(x=year, y=muhat.additive, color=degree), data=newdata, size=.5, alpha=.6, linetype='solid') +
 geom_line(aes(x=year, y=muhat.interacted, color=degree), data=newdata, size=.5, alpha=.6, linetype='dashed')
```

- In the last two classes, we've talked about two *bivariate models*.
- In additive models, we get *parallel lines* for each level of $z$.
- In interacted models, we get *totally different lines* for each level of $z$.
- In this case, they look very similar.

$$\small{
\begin{aligned}
\model &= \qty{ m(x,z) = b_0 + b_1x + b_2 z : b \in \R^3 }                 &&\text{additive bivariate lines} \\
\model &= \qty{ m(x,z) = b_0 + b_1x + b_2 z + b_3 x z: b \in \R^4 }      &&\text{interacted bivariate lines} \\
\end{aligned}
}
$$



## Bivariate Polynomial Models {.smaller}

```{r}
#| out-height: 400px
 
additive.line.fit   = lm(income ~ degree+poly(year,3), data=earnings)
interacted.line.fit = lm(income ~ degree*poly(year,3), data=earnings)
newdata = expand.grid(year = x, degree=unique(earnings$degree))
newdata$muhat.additive    =  predict(additive.line.fit, newdata)
newdata$muhat.interacted  =  predict(interacted.line.fit, newdata)

saturated.model.plot +
 geom_pointrange(aes(x=year, y=mean, ymin=mean-sd/sqrt(n), ymax=mean+sd/sqrt(n)), data=summaries) +
 geom_line(aes(x=year, y=muhat.additive, color=degree), data=newdata, size=.5, alpha=.6, linetype='solid') +
 geom_line(aes(x=year, y=muhat.interacted, color=degree), data=newdata, size=.5, alpha=.6, linetype='dashed')
```


- We can also do this with *polynomials*. Here we're doing cubics.
- In additive models, we get *parallel polynomials* for each level of $z$. They differ by a constant.
- In interacted models, we get *totally different polynomials* for each level of $z$.
- These are different. With different cubics, we can go through the subsample means. But with parallel ones, we can't.

$$\small{
\begin{aligned}
\model &= \qty{ m(x,z) = b_0 + b_1x + b_2 x^2 + b_3 x^3 + b_4 z : b \in \R^5 }                 &&\text{additive bivariate cubics} \\
\model &= \qty{ m(x,z) = b_0 + b_1x + b_2 x^2 + b_3 x^3 + b_4 z + b_5 x^2z + b_6 x^2 z + b_7 x^3 z: b \in \R^8 } &&\text{interacted bivariate cubics} \\
\end{aligned}
}
$$


## Saturated Models Can Be Too Big {.smaller}

```{r}
#| out-height: 400px
 
earnings = all.earnings %>% filter(education >= 8 & state.code==36 & county.code==61)
group.summaries = earnings %>% group_by(education, year) %>% summarize(mean=mean(income), sd=sd(income), n=n())
summaries = earnings %>% group_by(education) %>% summarize(mean=mean(income), sd=sd(income), n=n())
mean.year = mean(unique(earnings$year))


additive.line.fit   = lm(income ~ factor(year)+poly(education,8), data=earnings)
interacted.line.fit = lm(income ~ factor(year)*poly(education,8), data=earnings)
small.interacted.fit = lm(income ~ factor(year)*poly(education,3), data=earnings)
piecewise.fit = lm(income ~ factor(year)*(I(education > 12) + I(education > 16)), data=earnings)
newdata = expand.grid(year = factor(unique(earnings$year)), education = seq(7.5,20.5, by=.01))
newdata$muhat.additive    =  predict(additive.line.fit, newdata)
newdata$muhat.interacted  =  predict(interacted.line.fit, newdata)
newdata$muhat.smallinteracted  =  predict(small.interacted.fit, newdata)
newdata$muhat.piecewise   = predict(piecewise.fit, newdata)

edandyear.plot = ggplot(earnings) + 
  geom_point(aes(x=education, y=income, color=factor(year)), alpha=.4, size=.5, position=position_jitter(width=.3)) +
  geom_pointrange(aes(x=education+.1*(2*(year-mean.year)-1), y=mean, ymin=mean-sd/sqrt(n), ymax=mean+sd/sqrt(n), color=factor(year)), data=group.summaries) +
  scale_y_continuous(labels=dollar_format(), limits=c(-1500, 300000)) 
edandyear.plot
```

- So far, we've had relatively few levels of the covariates.
  - This meant we could usually get away with using saturated models. 
  - And it seemed like a good idea because the consequences of misspeccification are pretty bad.
- But when we use more covariates---or covariates with more levels---we get noisy estimates.
- Here we're plotting income (y) against education level (x) and year (z) for surveyed Manhattan residents.
  

## Saturated Models Can Be Too Big {.smaller}
```{r}
#| out-height: 400px

edandyear.plot
``` 
- There are 10 levels of education and 4 years, so we have 40 levels of the covariates---40 subsamples.
  - Some of these are very small. As a result, who happens to get surveyed has a big effect on the subsample mean.
  - And we see that they jump around a lot. That's probably not what's going on in the population.
- It's even possible that we get 'subsamples' of zero people. 
  - i.e., possible levels of the covariates that we don't get in the sample.
- To get around this, *we have to use models with fewer features*. 


## Smaller Models {.smaller}

```{r}
#| out-height: 400px
 
edandyear.plot + 
geom_line(aes(x=education, y=muhat.interacted, color=year), data=newdata, size=.5, alpha=.6)
```

- Here's an interacted model with 8th-order polynomials.
  - This has $4 \times 9 = 36$ features.
  - That's almost as big as the saturated model.
- And we see it's similarly messy.

## Smaller Models {.smaller}

```{r}
#| out-height: 400px
 
edandyear.plot + 
geom_line(aes(x=education, y=muhat.smallinteracted, color=year), data=newdata, size=.5, alpha=.6)
```

- Here's an interacted model with 3rd-order polynomials.
  - This has $4 \times 4 = 16$ features.
  - That's much smaller. And it's less noisy.

## Smaller Models {.smaller}

```{r}
#| out-height: 400px
 
edandyear.plot + 
geom_line(aes(x=education, y=muhat.additive, color=year), data=newdata, size=.5, alpha=.6)
```

- Here's an additive model with 8th-order polynomials.
  - This has $4 + 8 = 12$ features.
  - That's even smaller. And it's not too noisy either.
- Note that it looks ok *at observed levels of the covariate* but does weird stuff *in between.*
- High-order polynomials do that. Be careful if you want predictions at unobserved levels of the covariate.

## Smaller Models {.smaller}

```{r}
#| out-height: 400px

edandyear.plot + 
geom_line(aes(x=education, y=muhat.piecewise, color=year), data=newdata, size=.5, alpha=.6)
```

- An alternative to polynomials is *piecewise-constant models*.
- These are sets of curves that are constant between specified levels of the covariate---where they jump.
- We have seen this already.
  - We spoke for a while about using 4-year degree status instead of education.
  - This is a piecewise-constant model with a jump at $x=16$.

## Summary

```{r}
edandyear.plot + 
geom_line(aes(x=education, y=muhat.piecewise, color=year), data=newdata, size=.5, alpha=.6)
```

- When you have many covariates or many-valued covariates, you may not want a saturated model.
- There are lots of models you can choose from.
- We'll talk about *how to choose* in a few weeks.
- But before we get there, we'll talk about the *consequences of our choice*.

# Least Squares with Linear Models


## Orthogonality of Residuals to Features

- The residuals $\hat\varepsilon_i = Y_i - \hat\mu(X_i)$ are **orthogonal** to the every feature $\phi_k(X_i)$.
- That is, the dot product between the [**vector of residuals**]{.fg style="color:orange"} and the [**vector of feature observations**]{.fg style="color:blue"} is zero.
$$
\small{
\begin{aligned}
\hat\varepsilon^T \phi_k(X) 
= \begin{pmatrix} \hat\varepsilon_1 & \hat\varepsilon_2 & \ldots & \hat\varepsilon_n \end{pmatrix} 
\begin{pmatrix} \phi_k(X_1) \\ \phi_k(X_2) \\ \vdots \\ \phi_k(X_n) \end{pmatrix} 
= \sum_{i=1}^n \hat\varepsilon_i \phi_k(X_i) = 0 \qfor k=0,1,\ldots,K 
\end{aligned}
}
$$

- This is what that looks like in 2D, i.e., when we have two observations. 
  - We plot the feature and residual for one observation on the $x$-axis and the other on the $y$-axis.
  - We've used only one feature in each regression. 
  - If we used more, our plot would be less interesting. Why?

```{r}
#| out-height: 400px
#| layout-ncol: 2
 
Y = c(3/4, 3/2)
X = c(1,1)
residuals = lm(Y~X-1)$residuals
ggplot() + 
  geom_segment(aes(x=0, xend=residuals[1], y=0, yend=residuals[2]), color='orange') +
  geom_point(aes(x=residuals[1], y=residuals[2]), color='orange') +
  geom_segment(aes(x=0, xend=X[1], y=0, yend=X[2]), color='blue')  +
  geom_point(aes(x=X[1], y=X[2]), color='blue') + 
  geom_point(aes(x=0, y=0), color='gray') +
  coord_fixed() + xlab('i=1') + ylab('i=2') 

Y = c(1,  1)
X = c(1, 0)
residuals = lm(Y~X-1)$residuals
ggplot() + 
  geom_segment(aes(x=0, xend=residuals[1], y=0, yend=residuals[2]), color='orange') +
  geom_point(aes(x=residuals[1], y=residuals[2]), color='orange') +
  geom_segment(aes(x=0, xend=X[1], y=0, yend=X[2]), color='blue')  +
  geom_point(aes(x=X[1], y=X[2]), color='blue') + 
  geom_point(aes(x=0, y=0), color='gray') + 
  coord_fixed() + xlab('i=1') + ylab('i=2')
```

## Orthogonality of Residuals to Curves

- This means they're **orthogonal** to every curve $m(x)$ in the model.
$$
\small{
\begin{aligned}
\hat\varepsilon^T \phi_k(X) 
= \begin{pmatrix} \hat\varepsilon_1 & \hat\varepsilon_2 & \ldots & \hat\varepsilon_n \end{pmatrix} 
\begin{pmatrix} m(X_1) \\ m(X_2) \\ \vdots \\ m(X_n) \end{pmatrix} 
= \sum_{i=1}^n \hat\varepsilon_i m(X_i) = 0 \qfor m \in \model.
\end{aligned}
}
$$

- That's because curves in the models are linear combinations of features.
- And linear combinations of zeros are zero.

$$
\small{
\begin{aligned}
\sum_{i=1}^n \hat\varepsilon_i m(X_i) 
&= \sum_{i=1}^n \hat\varepsilon_i \qty{ \sum_k b_k \phi_k(X_i) } && \text{ for some $b_0,b_1,\ldots$ for any $m \in \model$ } \\
&= \sum_{i=1}^n \sum_k b_k \hat\varepsilon_i \phi_k(X_i) && \text{ distributing } \\
&= \sum_{k} b_k \sum_{i=1}^n \hat\varepsilon_i \phi_k(X_i) && \text{ re-ordering sums and pulling out constants } \\
&= \sum_{k} b_k \times 0 = 0                               && \text{ because residuals are orthogonal to features}. 
\end{aligned}
}
$$

## Uncorrelatedness of Residuals

- If the constant $1$ is in the model, this means they're **uncorrelated** with every curve in the model. Why?

::: fragment
- Because the *covariance* of the residuals and $m$ is the dot product of the residuals and another function $m-\bar m$.
$$
\sum_{i=1}^n \qty{m(X_i) - \bar m} \hat\varepsilon_i = \sum_{i=1}^n \tilde m(X_i) \hat\varepsilon_i \qfor \tilde m(x) = m(x)-\bar m.
$$

- And that function is in the model too, as it's a linear combination of two curves in the model.
$$
\small{
m \in \model \qqtext{ and } 1 \in \model \implies m - \bar m \times 1 \in \model \qqtext{ for the `coefficient' } \bar m = \frac{1}{n}\sum_{i=1}^n m(X_i)
}
$$

:::


## Residual Plots
[ TODO: incoporate old plots]

- When we have more than two observations, we can't plot one observation on each axis.
- Instead, we plot residuals on the $y$-axis and features on the $x$.
- It's a bit harder to see orthogonality/uncorrelatedness, as we don't actually get a right angle in the plot.
- This is what it looks like.


## Orthogonality Proof

$$
\small{
\qqtext{Let's prove that } \sum_{i=1}^n \hat\varepsilon_i \phi_k(X_i) = 0 \qfor k=0,1,\ldots,K
}
$$

::: fragment
$$
\begin{aligned}
0 &= \frac{\partial}{\partial b_k}\mid_{b=\hat\beta}  \frac{1}{n}\sum_{i=1}^n \qty{ \sum_{j=0}^K b_j \phi_j(X_i) - Y_i}^2 \\
&= \frac{1}{n}\sum_{i=1}^n 2 \qty{ \sum_{k=0}^K \beta_k \phi_k(X_i) - Y_i } \phi_k(X_i) \\
&= \frac{1}{n}\sum_{i=1}^n 2 \qty{ \hat\mu(X_i) - Y_i } \phi_k(X_i)  \\
&= \frac{1}{n}\sum_{i=1}^n 2 \qty{ -\hat\varepsilon_i } \phi_k(X_i)  \\ 
&= -\frac{2}{n}\sum_{i=1}^n \hat\varepsilon_i \phi_k(X_i)  
\end{aligned}
$$

- Versions of this proof come up *a lot*. It's an important one to remember.
- You can work out almost everything there is to know about least squares from it.

:::


## Limits

-  Suppose our observations $(X_1,Y_1) \ldots (X_n,Y_n)$ are sampled with replacement from a population $(x_1, y_1) \ldots (x_m,y_m)$.
- The **law of large numbers** tells you that, for any curve $m$, ... 
  - Mean squared error in the sample approximates its expected value.
  - And that expected value is mean squared error in the population.
$$
\frac{1}{n}\sum_{i=1}^n \qty{ m(X_i) - Y_i }^2 \approx \E\qty{ m(X_i) - Y_i}^2 = \frac{1}{m}\sum_{j=1}^m \qty{ m(x_j) - y_j }^2 
$$

- That suggests that if we minimize these two errors over curves in the same model, we should get similar curves.
- i.e., our least squares estimator approximates the population least squares curve.
$$
\hat\mu = \argmin_{m \in \model} \frac{1}{n}\sum_{i=1}^n \qty{ m(X_i) - Y_i }^2
\qqtext{approximates}
\tilde\mu = \argmin_{m \in \model} \frac{1}{m}\sum_{j=1}^m \qty{ m(x_j) - y_j }^2.
$$ 

- This is --- roughly speaking --- what actually happens.
- For the models we'll use in this class, this approximation is very good.
- There are some subtleties here that come up when you use very large models. 
- You can learn about them, e.g. in my Machine Learning Theory class.

## Population Least Squares

- We can say all this residual orthogonality stuff for **least squares in the whole population**.
- All curves in the model are **orthogonal** to the **population residuals**.

$$
\begin{aligned}
\tilde \mu &= \argmin_{m \in \model} \frac{1}{m}\sum_{j=1}^m \qty{ m(x_j) - y_j }^2 
&& \text{the population least squares curve} \\
\tilde \varepsilon_j &= y_j - \tilde \mu(x_j) 
&& \text{the population residuals} \\
0 &= \sum_{j=1}^m  \tilde \varepsilon_j m(x_j) 
&& \text{ for all $m \in \model$ }
\end{aligned}
$$

- We can see something interesting when we write this in terms of random variables and expectations.
- The **errors** our population least squares estimator makes---as an estimator of the conditional mean $\mu(x)$---are orthogonal to all curves in the model.

::: fragment

$$
\small{
\begin{aligned}
\tilde \mu = \argmin_{m \in \model} \E[ \qty{ m(X_i) - Y_i }^2 ]
\qqtext{ satisfies }
0 &= \E\qty[ \qty{Y_i - \tilde \mu(X_i)} m(X_i) ] && \text{for all} \quad m \in \model \\
  &= \E\qty[ \E\qty[ \qty{Y_i - \tilde \mu(X_i)} m(X_i) \mid X_i ]]  &&\text{via iterated expectations} \\
  &= \E\qty[ \qty{\E\qty[Y_i \mid X_i] - \tilde\mu(X_i)} m(X_i)  ]  &&\text{via linearity of cond. expectations} \\
  &= \E\qty[ \qty{\mu(X_i) - \tilde\mu(X_i)} m(X_i)  ]              &&\text{by definition of $\mu$}
\end{aligned}
}
$$

:::

- This suggests that the population least squares estimator *doesn't depend on variation* $Y_i-\mu(X_i)$ around the subsample means.
- It's the curve in the model that *best fits* the subpopulation means. We can see that more directly, too.

## Proof: Population Least Squares as **Noiseless** 

$$
\text{Our claim: }
\tilde\mu 
= \underset{\text{\color{gray}'noisy' population least squares}}{\argmin_{m \in \model} \E[ \qty{ m(X_i) - Y_i }^2 ]} 
= \underset{\text{\color{gray}'noiseless' population least squares}}{\argmin_{m \in \model} \E[ \qty{ m(X_i) - \mu(X_i) }^2]}
$$


::: fragment
$$
\small{
\begin{aligned}
&\E\qty[ \qty( m(X_i) - Y_i )^2 ] \\ 
&= \E\qty[ \qty( \qty{m(X_i) - \mu(X_i)} + \qty{\mu(X_i) - Y_i} )^2 ] 
&& \text{ adding and subtracting }\\
&= \E\qty[ \qty{m(X_i) - \mu(X_i)}^2 + 2\qty{m(X_i) - \mu(X_i)} \qty{\mu(X_i) - Y_i} + \qty{\mu(X_i) - Y_i}^2 ] 
&& \text{ via FOIL } \\
&= \E \qty{m(X_i) - \mu(X_i)}^2 + 2\textcolor{blue}{\E\qty{m(X_i) - \mu(X_i)} \qty{\mu(X_i) - Y_i}} + \textcolor{red}{\E\qty{\mu(X_i) - Y_i}^2}  
&& \text{ via linearity of expectations } 
\end{aligned}
}
$$


- Here the [third term]{.fg style="color:red"} doesn't depend on $m$, so it doesn't change the minimizer. 
- And the [second term]{.fg style="color:blue"} is zero.
$$
\small{
\begin{aligned}
&\textcolor{blue}{\E \qty{m(X_i) - \mu(X_i)} \qty{\mu(X_i) - Y_i}} \\
&= \E\qty[ \E\qty[ \qty{m(X_i) - \mu(X_i)} \qty{\mu(X_i) - Y_i}  \mid X_i]] && \text{ via iterated expectations} \\
&=  \E\qty[ \qty{m(X_i) - \mu(X_i)} \qty{\mu(X_i) - \E[Y_i \mid X_i]}] && \text{ via linearity } \\
&= \E \qty{m(X_i) - \mu(X_i)} \E\qty{\mu(X_i) - \mu(X_i)} = 0 &&.
\end{aligned}
}
$$

:::

## Example



```{r}
#| cache: true
#| out-height: 400px
#| layout-ncol: 2
 
scale_y = function(limits=c(-.3e5, 3e5)) { 
  scale_y_continuous(labels=label_dollar(), limits=limits)
}

years = 8:20
scale_x = function(limits=c(min(years)-.5, max(years)+.5)) { 
  scale_x_continuous(breaks=years, labels=sprintf('%dy', years), limits=limits)
}
default.dot.size = 2


# make a fake population by sampling with replacement from the sample then adding fuzz to incomes
# the actual sampled proportion in CPS is about 1/2500, so by doing 25x more dots,
# we're letting each dot  stand in for 100 people in the population
earnings.pop = earnings[sample(1:nrow(earnings), nrow(earnings)*25, replace=TRUE),]
earnings.pop$income = earnings.pop$income + .3*earnings.pop$income*runif(nrow(earnings.pop), min=-1,max=1)
earnings.pop = earnings.pop[earnings.pop$education <= 12,]

n = 1000
theme_set(simplified.theme)
set.seed(2)
earnings.samp = earnings.pop[sample(1:n, n, replace=TRUE), ]

plot.earnings = function(earnings, dot.size=default.dot.size, formula=NULL) {
  summaries = earnings %>% group_by(education) %>% summarize(mean=mean(income), sd=sd(income), n=n())
  p = ggplot(earnings) + 
    geom_point(aes(x=education, y=income), color='#F7766D',
               size=dot.size, position=position_jitter(seed=0,w=.3), alpha=.15) +
    geom_pointrange(aes(x=education, y=mean, ymin=mean-2*sd/sqrt(n), ymax=mean+2*sd/sqrt(n)), data=summaries) +
    scale_x(c(7.5, 12.5)) + scale_y(c(-.3e5, 1.5e5)) + guides(color=FALSE)

  if(!is.null(formula)) {
    model = lm(formula, data=earnings)
    x = seq(min(earnings$education), max(earnings$education), by=.01)
    linedata = data.frame(education=x, income=predict(model, data.frame(education=x)))
    p = p + geom_line(aes(x=education, y=income), data=linedata, color='blue', alpha=.5)
  }

  p
}

plot.earnings(earnings.pop, dot.size=default.dot.size/10)
plot.earnings(earnings.samp[1:1000,])
```



- Let's look at this stuff in the context of some data. 
- We'll use fake data so we have the population. That's what we're showing here.
- We'll base it on the California CPS data we've used to talk about education.
- We'll focus on residents with no more than 12 years of education.
- And we'll look --- starting on the next slide---at samples of three sizes: 50, 200, and 1000.

## Our Data
```{r}
#| out-height: 150px
#| layout-ncol: 2
#| layout-nrow: 2

formula = NULL
plot.earnings(earnings.pop, dot.size=default.dot.size/10, formula=formula)
plot.earnings(earnings.samp[1:1000,], formula=formula)
plot.earnings(earnings.samp[1:50,], formula=formula)
plot.earnings(earnings.samp[1:200,], formula=formula)
```

- We're plotting the subsample means +- 2 standard errors in black.

## Fitting Lines
```{r}
#| out-height: 150px
#| layout-ncol: 2
#| layout-nrow: 2

formula = income~education
plot.earnings(earnings.pop, dot.size=default.dot.size/10, formula=formula)
plot.earnings(earnings.samp[1:1000,], formula=formula)
plot.earnings(earnings.samp[1:50,], formula=formula)
plot.earnings(earnings.samp[1:200,], formula=formula)
```

- As sample size increases, our least squares lines get close to the population least squares line.
- But the population least squares line is not all that close to some of the subpopulation means.

## Fitting A Piecewise-Constant Model

```{r}
#| out-height: 150px
#| layout-ncol: 2
#| layout-nrow: 2

formula = income~I(education>=11) 
plot.earnings(earnings.pop, dot.size=default.dot.size/10, formula=formula)
plot.earnings(earnings.samp[1:1000,], formula=formula)
plot.earnings(earnings.samp[1:50,], formula=formula)
plot.earnings(earnings.samp[1:200,], formula=formula)
```

- This is a piecewise-constant model with a jump at $x=11$.
- It's a bit better than the line in that it never predicts *negative income*.

## Fitting A Cubic Polynomial

```{r}
#| out-height: 150px
#| layout-ncol: 2
#| layout-nrow: 2

formula = income~poly(education,3)
plot.earnings(earnings.pop, dot.size=default.dot.size/10, formula=formula)
plot.earnings(earnings.samp[1:1000,], formula=formula)
plot.earnings(earnings.samp[1:50,], formula=formula)
plot.earnings(earnings.samp[1:200,], formula=formula)
```

- This gets close to the subsample means. 
- But it has some weird wiggles in our smallest sample

## Fitting A Quartic Polynomial

```{r}
#| out-height: 150px
#| layout-ncol: 2
#| layout-nrow: 2

formula = income~poly(education,4)
plot.earnings(earnings.pop, dot.size=default.dot.size/10, formula=formula)
plot.earnings(earnings.samp[1:1000,], formula=formula)
plot.earnings(earnings.samp[1:50,], formula=formula)
plot.earnings(earnings.samp[1:200,], formula=formula)
```

- This *goes through* the subsample means.
- There's a 4th-order polynomial through any 5 points.

## Bias

```{r}
#| out-height: 400px
#| layout-ncol: 2
formula = income~education
plot.earnings(earnings.pop, dot.size=default.dot.size/10, formula=formula)
plot.earnings(earnings.samp[1:1000,], formula=formula)
```

- As sample size increases, the least squares line gets increasingly close to the population least squares line.
- But the population least squares line is not close to all the subpopulation means.
- As a result, no matter what sample size we use, we'll get bias.
- It's useful to think of our error at having two pieces.
  - *estimation error*. This is random. It depends on the sample.
  - *modeling error*.  This isn't. It depends only on the model and the population.
- Estimation error goes away in large samples. Modeling error doesn't.

$$
\hat\mu(x) - \mu(x) = \underset{\text{estimation error}}{\qty{ \hat\mu(x) - \tilde\mu(x) }} + 
  \underset{\text{modeling error}}{\qty{ \tilde\mu(x) - \mu(x) }}
$$

- All this passes through to summaries. What are the two parts of the average-over-years summary?


## Inference

$$
\hat\mu(x) - \mu(x) = \underset{\text{estimation error}}{\qty{ \hat\mu(x) - \tilde\mu(x) }} + 
  \underset{\text{modeling error}}{\qty{ \tilde\mu(x) - \mu(x) }}
$$

- The estimation error has mean zero and is approximately normal with variance proportional to $1/\sqrt{n}$.
- And the modeling error does not depend on the sample at all---it's constant. It's the bias.
- This is exactly the scenario we looked at on the last question on Midterm 2. 
- That, fundamentally, is what happens when you do least squares with linear models.
- In homework, we'll look into ... 
  - how to get a sense of what modeling error will be.
  - the tradeoff between modeling error and estimation error and its impact on inference.


