---
title: "Lab Nov 17"
format: pdf
editor: visual
---

## Matrices in R 

Let's use our WHI experiment data 

```{r}
whi<- read.csv("whidata.csv")
whi$age50s<- ifelse(whi$age=="50-59", 1, 0)
whi$age60s<- ifelse(whi$age=="60-69", 1, 0)
whi$age70s<- ifelse(whi$age=="70-79",1,0)

```

Start with simple additive model 

```{r}
whimod1<- lm(breastcancer ~ treat + age60s + age70s, data=whi)

summary(whimod1)
```


How can we get the same thing using matrices? We want $(X'X)^{-1}X'Y$, so we need a matrix $X$ that has our covariates in it and a column of 1s so we can estimate an intercept. 

```{r}
df.mat<- data.frame(constant=rep(1,length(whi$breastcancer)), treat=whi$treat, age60s=whi$age60s, age70s=whi$age70s)

X<- data.matrix(df.mat)

dim(X)

```

To test our matrix functions, I want to use something smaller to be able to check it. I will use the matrix from PS 3: $C = \begin{bmatrix} 3 \\ 2 \\ 5 \\ 7 \end{bmatrix}$ and $A = \begin{bmatrix} 1 & 0\\ 1 & 1 \\ 1 & 1\\ 1 & 2 \end{bmatrix}$

```{r}
A<- matrix(c(1,1,1,1,0,1,1,2),ncol=2)
dim(A)
C<- matrix(c(3,2,5,7),ncol=1)
dim(C)
#What is the transpose of A?
t(A)

t(A)%*%A
#What is the inverse of A'A?
solve(t(A)%*%A)

solve(t(A)%*%A)%*%t(A)%*%C
``` 

Let's go back to our WHI data 

```{r}
betas<- solve(t(X)%*%X)%*%t(X)%*%whi$breastcancer
dim(betas)
betas
```

We got the same answer as lm! 

optional: make a new version of $X$ that includes all three age variables. What happens? 

```{r}
df.mat2<- data.frame(constant=rep(1,length(whi$breastcancer)), treat=whi$treat, age50s=whi$age50s ,age60s=whi$age60s, age70s=whi$age70s)

X2<- data.matrix(df.mat2)

dim(X2)

#solve(t(X2)%*%X2)
```


How do we use matrices to estimate standard errors for our betas? What is the sandwich estimator? 

$$\hat V\left[\hat{ \beta}\right]  = \left[X^T X \right]^{-1}  X^T
     {\text{diag}} \left[ (Y - X\hat{\beta})^2 \right]
      X \left[X^T X \right]^{-1}.$$

We know we can deal with $X$ now, but what is the thing in the middle of the sandwich? A diagonal matrix of the residuals squared. 


```{r}
resid<- residuals(whimod1)
```

so we have our residuals and we can square them, but how do we make a diagonal matrix? 

```{r}
test<- c(4,5,6,7,8)

diag(test)

```

So we can make a diagonal matrix of our squared residuals the same way. What should the dimensions of the matrix be? 

```{r}
D<- diag(resid^2)

dim(D)

```

Let's use our sandwich estimator 

```{r}
betavar<- solve(t(X)%*%X)%*%t(X)%*%D%*%X%*%solve(t(X)%*%X)
```


Are we done? The diagonal is the variance of our betas and we need to take the squareroot to get the standard error. 

```{r}
sqrt(diag(betavar))
```


That was complicated. Big matrices. Is there an easier way to do it? The sandwich package 

```{r}
library(sandwich)
sandwich(whimod1)

sqrt(diag(sandwich(whimod1)))

```

