---
title: "Problem Set 1"
subtitle: "QTM 220"
format: pdf
editor: visual
---

```{r}
#| echo: false
library(ggplot2)
theme_set(theme_bw())
```

# Mean and Standard Deviation Review

## Calculations

- Each of the following lists has a mean of 50. For which is the standard deviation biggest? Smallest? 

  1. 0, 20, 40, 50, 60, 80, 1002.
  2. 0, 48, 49, 50, 51, 52, 100 
  3. 0, 1, 2, 50, 98, 99, 100  


- For the two lists below, calculate the mean and standard deviation of the numbers. Then compare the your answers for the two lists. Think about this comparison. How is the first list related to the second? How does this relationship carry over to the mean? The standard deviation? 

 
  1. 1, 3, 4, 5, 7 
  2. 6, 8, 9, 10, 12 

-  Repeat this exercise for a new pair of lists. 

  1. 1, 3, 4, 5, 7 
  2. 3, 9, 12, 15, 21

- And again. 

  1. 5, -4, 3, -1, 7
  2. -5, 4, -3, 1, -7 

## Properties

- Can a standard deviation ever be negative? Explain.

- For set of positive numbers, can the standard deviation ever be larger than the average? Explain. 

## Visualization

```{r}
#| echo: false
set.seed(220220)
sample.1 = rbeta(10000, 5, 2)
sample.2 = rbeta(10000, 2, 5)
sample.3 = rbeta(10000, 5, 5)
```

Consider the three histograms below.


- The means of the samples we've histogrammed are approximately `r round(mean(sample.2),1)`,
`r round(mean(sample.1),1)`, `r round(mean(sample.3),1)`. Which histogram corresponds to which mean?
- True or false: the standard deviation of the sample summarized by Histogram 1 is a lot smaller than the one summarized by Histogram 2. Explain. 

```{r}
#| echo: false
#| layout-ncol: 3
#| fig-cap:
#|  - "Histogram 1"
#|  - "Histogram 2"
#|  - "Histogram 3" 

ggplot() + geom_histogram(aes(x=sample.1, y=after_stat(density)), bins=100, alpha=.3, color='black') + xlim(0,1) 
ggplot() + geom_histogram(aes(x=sample.2, y=after_stat(density)), bins=100, alpha=.3, color='black') + xlim(0,1)
ggplot() + geom_histogram(aes(x=sample.3, y=after_stat(density)), bins=100, alpha=.3, color='black') + xlim(0,1)
```

# Supreme Court Justices

## The Data
Load the provided dataset "EMdata.csv". To do this, start an R session in the same folder as the csv file, then say 

```{r}
#| echo: true
EMdata = read.csv("EMdata.csv")
```

- This data is on 27 justices from the Warren ('53 - '69), Burger ('69 - '86), and Rehnquist ('86 - '05) courts
- The data can be interpreted as a census of justices for the 1953 - 2005 era.  Each row is a justice and each column is a variable. The column 'justice' is the name of the justice. We'll be looking at a few other variables.
   - CLlib: The percentage of votes in liberal direction for each justice in civil liberties cases
   - party: the political party that nominated the justice (Republican =0, Democrat=1)
   - ur:   the justice is a member of an under-represented group, such as a racial or gender minority (under-represented group=1, not in under-represented group=0)
- To get you started, I'm going to plot a histogram of the percentage of liberal votes, identifying the mean with a vertical line. You may want to edit this code to answer future questions.

```{r}
#| echo: true

CL.histogram = ggplot(EMdata) + 
           geom_histogram(aes(x=CLlib, y=after_stat(density)), 
                          bins=10, alpha=.3, color='black') +  
           geom_vline(aes(xintercept=mean(CLlib)), 
                          color="blue") +
           xlab("% Support for Liberal Position on Civil Liberties Cases (CLlib)")

CL.histogram
```

## Exercise 1 

- Write your own function to calculate the standard deviation of CLlib (i.e. not using "sd") and report it. Use 'sd' to check your answer.^[You may be slightly off. In particular, you may be off by a factor of `r n=nrow(EMdata); n-1`/`r n`. That's
ok. There are two conventions for calculating the sample standard deviation---one involves division by $n$ and the other $n-1$.]
Add, to the plot above, vertical lines indicating the mean plus and minus 1 and 2 standard deviations. 
identification of a standard deviation on each side of the mean. What is the substantive interpretation of this mean?



## Exercise 2

Replicate the plot above, i.e. a histogram with lines for the mean plus or minus two standard deviations, for the variable 'ur'. What is this mean? And what is its substantive interpretation?


## Exercise 3

Draw two histograms of CLlib, one for Republican-nominated justices and one for Democratic-nominated justices. Calculate the mean of CLlib in each group. Which is larger, the mean among Republican justices or Democratic justices? Give a substantive interpretation of this difference.



## Exercise 4

Below, I've drawn a scatter plot of CLlib with the nominating party as the x-axis. I've used `stat_summary` to draw in the mean plus and minus two standard deviations for each group. Calculate these means and standard deviations and report the mean and endpoints of the intervals that have been drawn in. Check your answer by visually for agreement with the plot. Can you give a substantive interpretation of these intervals?

```{r}
#| echo: true

ggplot(EMdata) +  
  geom_point(aes(x=party, y=CLlib), 
             position=position_jitter(w=.1, h=0), alpha=.4) + 
  stat_summary(aes(x=party, y=CLlib), 
              fun=mean, 
              fun.min = function(x) mean(x) - 2*sd(x), 
              fun.max = function(x) mean(x) + 2*sd(x), 
              geom="pointrange") + 
  xlab("Party of Appointing President") + 
  ylab("%Support for Liberal Position on CL Cases")
```


## Exercise 5
Repeat the exercise above, using the variable 'ur' instead of 'CLlib'. Draw your own plot.


# NBA Players

You and a friend each assemble some data describing the performance of NBA players in the 2023 season. 
From the population of all players that played in the league that year, you each draw a sample of 100 players
with replacement. Then you record some information about each player in your sample. 

```{r}
#| echo: true
sample.1 = read.csv("nba_sample_1.csv") 
sample.2 = read.csv("nba_sample_2.csv") 
```

Here's a list of variables.

- Player - The player's name.
- Team: The team whose this player is playing for this season in abbreviated term.
- Age:  The Age of the player.
- GP: The number of games that the player has played in this season.
- W: The number of games won in which the player has played.
- L: The numer of games lost in which the player has played.
- Min: The minutes the player has played for this season.
- PTS: The number of points made by the player.


## Exercise 1

Let's look the variable 'PTS'---the number of points players scored over the course of the season. For sample 1, calculate the mean and standard deviation. Then, draw a histogram with the mean and mean plus or minus two standard deviations indicated by vertical lines. Repeat this for sample 2.


## Exercise 2 

Using sample 1, calculate a 95% confidence interval for the population mean. To calibrate it, use the bootstrap.
Then draw the interval on top of your histogram from Exercise 1. And draw it on top of a similarly annotated histogram of the sample mean's bootstrap sampling distribution, too---one with lines indicating its mean plus or minus one and two standard deviations.  If you've got a ggplot called 'histogram' and have calculated the center ('center') and width ('w') of your interval, you can add the interval like this.

```{r}
#| echo: true
#| eval: false

histogram + geom_pointrange(y=0, x=center, xmin=center-w/2, xmax=center+w/2)
```


To get you started, I'll give you a function that'll calculate your interval's width. What you'll need to do is draw samples from the bootstrap sampling distribution and pass them in.


```{r width-code}
width = function(samples, center=mean(samples), alpha=.05) {
  covg.dif = function(w) { 
    actual = mean( center - w/2 <= samples & samples <= center + w/2)
    nominal = 1-alpha 
    actual - nominal  
  }
  uniroot(covg.dif, interval=c(0, 2*max(abs(samples-center))))$root
}
```

Do this using at least 10,000 bootstrap samples. Then repeat this exercise using sample 2.


**About the code.** In essence, it tries different widths $w$ until it finds one for which 95% of the samples fall in the interval $\text{center} \pm \text{w}/2$. It's easy to find one because the number of samples included is *increasing* with the width of the interval. You know to increase it if you're below 95% and decrease it if you're above. That's what the `R` built-in function `uniroot` does---it searches for $w$ for which the difference between the actual proportion of samples included and 95% is zero. I'll explain how this works a little more precisely later on in class. Note that it's the argument `alpha` that specifies that we want a 95% interval---`alpha` is the proportion of samples we want to *exclude* from the interval. If the proportion of samples we want to include is $.95$, the proportion we want to exclude is $1-.95=.05$. 



## Exercise 3

Let's compare our 95% interval to differently-calibrated intervals. We'll use sample 1 for this exercise.
Calculate, then draw into your plot of the bootstrap sampling distribution from the previous exercise, 90% and 99% intervals in addition to the 95% interval you already have. How do your three intervals compare? It'll help to read the 'about the code' paragraph above if you haven't already. 

**Plotting Tip.**
You'll want to specify different 'y' values in `geom_pointrange` so your intervals aren't drawn on top of each other. 
Choose whatever makes sense to you visually. 


# Comparing Sample and Population

Noting your interest in NBA stats, a benevolent data analyst sends you data for the full population. That is, they send you a table like the one you've been working with, but with a row for every player in the league the 2023 season. 

```{r}
population = read.csv("nba_population.csv")
```

We're going to use this to check our work. We can draw samples of size `r n`, 
just like the two we actually have, from this population.
And we can use them to calculate the actual sampling distribution of the mean number of points players scored. This is just like what we did for our turnout poll using our post-election knowledge of turnout.


## Exercise 1

What is the mean number of points scored for players in our population? How does it compare to our two sample means? Is it in our 95% confidence intervals? What about our 90% and 99% intervals?

## Exercise 2

By drawing samples of size $n=`r nrow(sample.1)`$ with replacement from the population, 
you can simulate the surveys you and your friend ran earlier. Do this 10,000 times, calculating the sample mean each time, and use the results to draw a histogram showing the sampling distribution of your sample mean.
Then add the following to your plot.

- The sampling distribution's mean and mean plus or minus one and two standard deviations, just like you did for the bootstrap sampling distribution. 
- The population mean as a vertical line. Make it green like I have in lecture. 
- Your two 95\% confidence intervals based on samples 1 and 2. Use different 'y' values so they don't overlap. 

If either interval doesn't include the population mean, explain what's gone wrong. Is it an unusually bad point estimate? A too-narrow interval? Or both? Refer to the plot.