---
title: "Problem Set 2"
subtitle: "QTM 220"
execute:
  echo: false
  warning: false
format: 
  pdf:
    include-in-header: 
      text: |
        \usepackage{amsmath}
        \usepackage{setspace}
        \onehalfspacing
        \usepackage{amssymb}
        \usepackage{physics}
        \usepackage{color}
        \usepackage{fullpage}
        \usepackage{graphicx}
        \usepackage{cancel}
        \usepackage{physics}
        \usepackage{fullpage}
        \DeclareMathOperator{\E}{E}
        \DeclareMathOperator{\Var}{V}
        \DeclareMathOperator{\bias}{bias}
        \usepackage{environ}
---

```{r setup} 
library(ggplot2)
library(tidyverse)
if (knitr::is_latex_output())
  knitr::opts_chunk$set(dev = "cairo_pdf")
```


# Consistency

We call an estimator *consistent* if, when our sample size $n$ is large enough, it converges to the thing we intend to estimate. There are multiple ways to talk about the
convergence of random variables, but we'll start with once called *convergence in mean square*. An estimator $\hat\mu$ converges in mean-square to $\mu$ if $\E\qty[(\hat \mu - \mu)^2] \to 0$ as $n \to \infty$.
What does that mean? Well, let's recall our `excess' formula for the variance. 
$$ 
\Var\qty[\hat\mu - \mu] = \E\qty[(\hat\mu - \mu)^2] - \qty(\E[\hat\mu - \mu])^2 \qqtext{ and equivalently } \underset{\text{mean squared error}}{\E\qty[(\hat\mu - \mu)^2]} =  \underset{\text{bias}^2}{\qty(\E[\hat\mu - \mu])^2} + \underset{\text{variance}}{\Var\qty[\hat\mu - \mu]}.
$$
The rearranged form of this, which decomposes the estimator's mean squared error as the sum of its squared bias and its variance, is called the \emph{bias/variance decomposition}. 
Since both $\text{bias}^2$ and \text{variance} are positive, this tells us that an estimator is consistent if and only if both its bias and its variance go to zero. That is,
if, by using a large enough sample size, we can make two things happen.

1. We can make the center of its sampling distribution arbitrarily close to the estimation target $\mu$. That's the bias part.
2. We can make the spread of the sampling distribution arbitrarily small. That's the variance part.

One example of a consistent estimator, when we observe a sample with replacement $Y_1 \ldots Y_n$ from a population with mean $\mu$
and standard deviation $\sigma$, is the sample mean $\hat \mu = \frac{1}{n}\sum_{i=1}^n Y_i$.

::: exercise
When we have a sample of size $n$, what is the bias of $\hat\mu?$ What is its standard deviation?
:::

Now lets's try to get a sense of what consistency does and doesn't look like. The plot below shows the sampling distributions of three estimators at three sample sizes. 
The vertical line indicates the mean of the sampling distribution. 

::: exercise
Which of these estimators are consistent? Circle zero or more: a, b, c.
:::

```{r}
#| echo: false
#| out-height: 4in 
set.seed(1234)
error = 2*rnorm(100000)
bias = list(a=function(n) 0, b=function(n) sqrt(10)/sqrt(n), c=function(n) 1)

plotdata = expand_grid(estimator=names(bias), n=c(10,40,160)) %>%
  pmap(function(estimator, n) {
    thebias = bias[[estimator]](n)
    data.frame(estimator=estimator, n=n, muhat=thebias + error / sqrt(n))
    }) |> bind_rows() 
summaries = plotdata |> group_by(n, estimator) |> summarize(mean=mean(muhat))

breaks = seq(-4, 4, by=2)/2 
signs  = case_when(
  breaks < 0 ~ '-', 
  breaks == 0 ~ '',
  breaks > 0 ~ '+')
offsets = case_when(
  breaks != 0 ~ sprintf('%d', abs(breaks)),
  breaks == 0 ~ '')
labels.abstract = sprintf('μ %s %s', signs, offsets)

ggplot(plotdata) + 
  geom_histogram(aes(x=muhat, y=after_stat(density)), bins=100, alpha=.4) + 
  geom_vline(aes(xintercept=mean), data=summaries) + 
  scale_x_continuous(breaks=breaks, labels=labels.abstract, limits=range(breaks)) + 
  theme_bw() + theme(axis.text.x = element_text(angle = -90, vjust = 0.5, hjust=1)) + 
  xlab('') + ylab('') + 
  facet_grid(n ~ estimator, labeller=labeller(n=label_both, estimator=label_both))  
```


# Convergence in Probability
Now let's think about another notion of convergence. If an estimator's sampling distribution
ends up in the right location ($\text{bias} \to 0$) with arbitrarily little spread ($\text{variance} \to 0$),
then it makes sense that \emph{any draw from it} will be close to the estimation target $\mu$. 

Recall that you can visualize your estimator as a single dot drawn
from its sampling distribution. Like the black one below. And if you want to think of what happens if you were to repeat your survey 100 times, you can visualize the results by plotting 100 dots. Like the purple ones.
```{r}
K = 100
plotdata$height = runif(nrow(plotdata), 0, 1.2)
dotplotdata = plotdata |> filter(n==40 & estimator=='b')
dotplotsummaries = summaries |> filter(n==40 & estimator=='b')

ggplot(dotplotdata) + 
  geom_histogram(aes(x=muhat, y=after_stat(density)), bins=100, alpha=.4) + 
  geom_vline(aes(xintercept=mean), data=dotplotsummaries) + 
  geom_point(aes(x=muhat, y=height), alpha=.2,
             color='purple', data=dotplotdata[2:K,]) +
  geom_point(aes(x=muhat, y=height), alpha=1,
            color='black', data=dotplotdata[1,]) + 
  scale_x_continuous(breaks=breaks, labels=labels.abstract, limits=range(breaks)) + 
  theme_bw() + theme(axis.text.x = element_text(angle = -90, vjust = 0.5, hjust=1)) + 
  xlab('') + ylab('') 

```


We formalize this idea by talking about \emph{convergence in probability}. We say an estimator $\hat\mu$ converges in probability to $\mu$ if, for anyone's idea of `sufficiently close', the probability that $\hat\mu$ is sufficiently close to $\mu$ goes to one. 
$$
P(\lvert\hat\mu - \mu\rvert > \epsilon) \to 0 \qqtext{ as } n \to \infty
$$

Convergence in mean square *implies* convergence in probability. Let's see why.
Markov's inequality says that the probability the absolute value of a random variable $X$ 
exceeds some threshold $t$ is bounded by the ratio of its expectated absolute value and $t$, i.e.,
$$ P(X \geq t) \leq \frac{\E[X]}{t}$$
Often, instead of using this to bound the random variable we're interested in directly,
e.g. $X=\lvert\hat\mu - \mu\rvert$, we use it to bound its square. $X \ge t$ if and only if $X^2 \ge t^2$, 
so applying Markov's inequality to the random variable $X^2$ gives us a bound in terms of $X$'s mean square.
$$
P(X \ge t) = P(X^2 \ge t^2) \leq \frac{\E[X^2]}{t^2} \qqtext{ e.g. } 
P(\lvert\hat\mu - \mu\rvert \ge \epsilon) = P((\hat\mu - \mu)^2 \ge \epsilon^2) \leq \frac{\E[(\hat\mu - \mu)^2]}{\epsilon^2}
$$

This tells us, in terms of the mean squared error of $\hat\mu$, how likely it is that $\hat\mu$ is far from $\mu$. And, in particular, it tells us that if the mean squared error goes to zero, then the probability that $\hat\mu$ is any distance $\epsilon$ away from $\mu$ goes to zero. 


Now we're going to do two things. First, we're going to prove Markov's inequality. It's good practice working with expectations and probabilities. Second, we're going to look at what Markov's inequality tells us about
how close $\hat\mu$ is to $\mu$, i.e., we're going to use it to calculate the coverage of an interval estimate.


## Proving Markov's Inequality

::: exercise
Prove Markov's inequality. 
:::

To help you out, I'll outline the proof. You'll need to fill in the details. But treat this like an 'essay question', not a 'short-answer' thing. Write out the whole argument, so you can read it without the outline.

Here's the outline. We'll be thinking about a binary random variable $1_{\ge t}(X)$. This is the random variable we get by plugging the random variable $X$ into the *indicator function* for exceeding $t$.
$$
1_{\ge t}(x) = \begin{cases} 1 & \text{ if } x \ge t \\ 0 & \text{ if } x < t \end{cases}
$$

1. The expectation of $1_{\ge t}(X)$ is the probability that $X$ exceeds $t$, $P(X \ge t)$. Explain why.
2. If we have some function of $u(x)$ that's always larger than $1_{\ge t}(x)$, i.e., 
   one satisfying $u(x) \ge 1_{\ge t}(x)$ for all $x$, we know that $\E[u(X)] \ge \E 1_{\ge t}(X)$.^[You can take this as given, but it's easy to prove: use linearity of expectations to show that $\E[u(X)] - \E 1_{\ge t}(X) \ge 0$ when $u(x) - 1_{\ge t}(x) \ge 0$ for all $x$.] Argue that the function $u(x)=x/t$ 
   does the trick by considering the two cases: $x \ge t$ and $x < t$.
3. What is $\E u(X)$. Are you done?

## Markov's Inequality and Interval Estimation

So far, when we've calculated the coverage of an interval estimate in terms of its mean and variance, we've relied on normal approximation. That is, we've been using a different formula for 
$P(\lvert\hat\mu - \mu\rvert \ge \epsilon)$ based on the idea that $\hat\mu-\mu$ is has approximately the same distribution as a normally-distributed random variable $Z$ with mean zero and variance $\sigma^2=\Var[\hat\mu-\mu]$. We can use this approximation, in combination with our knowlege of the probability density of the normal approximation, to get a bound on the probability that $\lvert\hat\mu - \mu\rvert \ge \epsilon$.

$$
P(\abs{\hat\mu - \mu} \ge \epsilon) \approx P(\abs{Z} \ge \epsilon) = 1-P(\abs{Z} < \epsilon) = 1- \int_{-\epsilon}^{\epsilon} f_{\sigma}(z) dz \qfor f_{\sigma}(z) = \frac{1}{\sqrt{2\pi}\sigma} e^{-z^2/2\sigma^2}
$$

The reason we're always talking about intervals of the form $\hat\mu \pm 2\sigma$ is that, using this approximation, that yields 95% coverage. That is, $P(\abs{Z} < 2\sigma) \approx .95$. To see why we 
use normal approximation instead of Markov's inequality, let's see what Markov's inequality tells us about the coverage of this interval.

We'll start with an abstract mathematical question.
::: exercise
Suppose $Z$ is a random variable, not necessarily normally distributed, 
with mean zero and variance $\sigma^2$. What does Markov's inequality tell 
us about the probability that $P(\abs{Z} \ge t)$ for $t=2\sigma$?
:::

Now let's translate that into practical terms.
::: exercise
In a previous exercise, you calculated the mean and variance of the sample mean $\hat\mu$ when we have a sample of size $n$ drawn with replacement from a population with mean $\mu$ and variance $\sigma^2$. What does Markov's inequality tell us about the coverage probability of the interval $\hat\mu \pm 2\sigma$, i.e., the probability that $\abs{\hat\mu - \mu} < 2\sigma$? And how big do we need to make $t$ for it to tell us that $\hat\mu \pm t\sigma$ has coverage probability .95?
:::

Finally, let's draw it in the context of a real dataset. We'll be estimating the mean number of points NBA players scored over the course of a season. As our estimate, we'll us the corresponding mean of a sample of these players---the one in 'nba_sample_1.csv' from Problem Set 1.

::: exercise
As in Problem Set 1, draw 10,000 samples from the bootstrap sampling distribution of the sample mean $\hat\mu$.
Make a histogram showing the bootstrap sampling distribution of the sample mean. Then plot three 95\% confidence intervals. The first should be, as in Problem Set 1, based on the width of the bootstrap sampling distribution;
the second, an interval $\hat\mu \pm 2\hat\sigma$ based on normal approximation where $\hat\sigma$ is the standard deviation of your bootstrap samples; and the third, an interval $\hat\mu \pm t\sigma$ for the appropriate threshold based on Markov's inequality. And give a short, e.g. one-sentence, answer to this. **Q.** Why do we usually prefer the intervals based on the bootstrap and normal approximation to the one based on Markov's inequality? 
:::

## Power Calculation

Let's think back to our discussion of the NSW experiment/job-training program. 
Our estimate of the treatment effect was `r 
We
were a bit dissapointed with the results of the experiment because,
when we looked at our interval estimate for the treatment effect, it was so wide
that we couldn't tell if the job-training program

so wide 
that it included zero. That is, we couldn't reject the null hypothesis that the treatment had no effect.


