---
title: "Problem Set 6 Solutions"
subtitle: "QTM 220"
format: pdf
editor: visual
---

```{r}
#| echo: false
library(ggplot2)
library(sandwich)
```

Due 1 December at 2:00pm ET on Canvas. Questions about submission requirements should be directed to your TA.

\subsection{California CPS Revisited}

(a) Using the California sample of 25-35 year olds, regress income on education and age (in an additive model). Report your results in a table, including the standard errors that R automatically calculates.

```{r}
cps<- read.csv("California_CPS_degrees.csv")

mod1<- lm(income~ education + age,data=cps)

summary(mod1)

```

(b) Create a matrix of your $X$ information for your model (education and age) and include a column of 1s to estimate an intercept. Using the following matrix functions in base R, calculate $(X'X)^{-1}X'Y$ and confirm your resulting $\hat{\beta}s$ are the same as your $lm$ results:

Let $X$ be your matrix in R. Use $\%*\%$ to multiply matrices, i.e. $X\%*\%X$ will yield $XX$

Use $t(X)$ to take the transpose of matrix $X$

Use solve($X$) to take the inverse of matrix $X$

```{r}
df.mat<- data.frame(constant=rep(1,length(cps$income)), edu=cps$education, age=cps$age)

X<- data.matrix(df.mat)

solve(t(X)%*%X)%*%t(X)%*%(cps$income)

```

(c) Extract the predicted values and residuals from your model using predict(model name) and residuals(model name), then create a scatter plot of your predicted values and residuals. What do you see? Would you guess the errors are homoskedastic or heteroskedastic? Why?

```{r}

pred<- predict(mod1)
resid<- residuals(mod1)

df<- data.frame(pred=pred,resid=resid)


rplot<- ggplot(df, aes(x=pred,y=resid)) + geom_point(alpha=.2)

rplot
```

(d) install the sandwich package and get heteroskdasticity consistent standard errors for your model

```{r}
sandwich(mod1)

sqrt(diag(sandwich(mod1)))

```



(e) do matrix algebra to get the same thing

```{r}
D<- diag(resid^2)

dim(D)

betavar<- solve(t(X)%*%X)%*%t(X)%*%D%*%X%*%solve(t(X)%*%X)

sqrt(diag(betavar))

```


(f) Bootstrap! 


(g) Compare the standard errors you got from R in part (a) to the robust (sandwich) you calculated in parts (e) and (f). Why are they different? How do these compare to what you estimated from the bootstrap? 

\subsection{Potential outcomes review???}

potential outcomes table-- use conditional randomization, show that ATE among groups is the same as interactive model, show wrong model gives wrong answer, show APD=ATE 

\subsection{Neto and Cox}

Potential outcomes table ?

Basic APD 

Propensity score matching

make a balance table

\subsection{Menopausal Hormone Therapy}

compare RCT and observational studies

read some stuff

reflect
