---
title: "220 Quiz 1"
format: pdf
editor: visual
---
## Question 1 
A company wants to assess customer satisfaction after implementing a new online support system. They conduct a survey among their customers, asking them to rate their experience on a scale from 1 (very dissatisfied) to 5 (very satisfied). The company collects responses from a random sample of 500 customers. They find that the average satisfaction score among the 500 customers is 4.2. 

\begin{enumerate}
\item What is the estimand? 
\vspace{3cm}

\item What is the estimator? 

\vspace{3cm}

\item What is the estimate? 


\end{enumerate}

\newpage

## Question 2 

```{r}
#| echo: false
library(truncnorm)
pop<- rtruncnorm(3000, a=1, b=5, mean=4.2, sd=.75)
sampdisty<- NA
for(i in 1:10000) {
    Y.star = pop[sample(1:length(pop), length(pop), replace=TRUE)]
    sampdisty[i] = mean(Y.star) 
}

y.a<- sampdisty
#hist(y.a)

y.b<- sample(pop, 500, replace=T)
#hist(y.b)
```

Below are two histograms, labeled "y.a" and "y.b" Indicate which histogram visualizes the distribution of the sampled respondents' satisfaction ratings and which histogram visualizes the sampling distribution of the sample mean of respondents' satisfaction for a sample size of 500. In one to two sentences, describe how you determined which is which. 

```{r}
#| echo: false
#| out-height: 400px 
#| layout-ncol: 2 
hist(y.a)
hist(y.b)
```