---
title: "QTM 220: Lecture 18"
subtitle: "Inverse Probability Weighting"
format: 
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: png
    default-image-extension: png
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---


```{r load-libraries}
library(purrr)
library(tibble)
library(dplyr)
library(magrittr)
library(stringr)
library(ggplot2)
library(htmltools)
library(kableExtra)
library(scales)
library(cpsR)
library(maps)

options(knitr.kable.NA = '')

census.api.key = '8084f81e4f71e4d32638dac15f188b424f72d063'
# for cps variable interpretation see https://www2.census.gov/programs-surveys/cps/datasets/2022/march/asec2022_ddl_pub_full.pdf
percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }
```

```{r ggplot-themes}
fontsize = 16
textcolor = "#888888"

class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
        panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x  = element_text(colour = textcolor, size=fontsize),
        axis.text.y  = element_text(colour = textcolor, size=fontsize),
        axis.title.x  = element_text(colour = textcolor),
        axis.title.y  = element_text(colour = textcolor, angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
              axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
           panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(simplified.theme)
```


```{r table-styling}
ellipsis <- function(df, nhead=5L, ntail=2) {
    rownames(df) = as.character(1:nrow(df))
    stopifnot("data.frame" %in% class(df))
    els <- rep(NA, ncol(df)) %>% 
        matrix(nrow=1, dimnames=list(NULL, names(df))) %>% 
        data.frame(stringsAsFactors=FALSE) %>% 
        tibble::as_tibble()
    rownames(els) = "⋮"
    dplyr::bind_rows(head(df, nhead), els, tail(df, ntail))
}
```


```{r population}
set.seed(1) 
m = 100000
pi = function(x) { 1/(1+exp(-sinpi((x-60)/60))) }
mu = function(w, x) { 2 - 2*((x-50)/30)^3/3 + w*((x-50)/30)^2 }  

X = 50+round(30*runif(m))
W = rbinom(m, 1, pi(X))
eps = .5*rnorm(m)
for(x in unique(X)) {
  for(w in c(0,1)) {
    eps[X==x & W==w] = eps[X==x & W==w] - mean(eps[X==x & W==w])
  }
}
Y = mu(W,X) + eps

pop = data.frame(X=X, W=W, Y=Y, Y.counterfactual=Y-mu(W,X)+mu(1-W,X))
```

```{r estimates}
#| cache: true
ns = 100*2^(0:10)
grid = expand.grid(n=ns, rep=1:200, type=c('unweighted', 'weighted', 'weighted-estimates-linear', 'weighted-estimates-cubic')) 

estimates = purrr::pmap(grid, function(n, rep, type) {
    set.seed(rep)
    data = pop[sample(1:nrow(pop), n, replace=TRUE), ]
    if(type=='unweighted') { 
      model = lm(Y ~ W*X, data=data) 
    } else if(type=='weighted') { 
      model = lm(Y ~ W*X, data=data, 
        weights=data$W + (1-data$W)*pi(data$X)/(1-pi(data$X))) 
    } else if(type == 'weighted-estimates-linear') {
      propensity.model = glm(W~X, family='binomial', data=data)
      pihat = function(x) { predict(propensity.model, newdata=data.frame(X=x), type='response') }
      model = lm(Y ~ W*X, data=data, weights=data$W + (1-data$W)*pihat(data$X)/(1-pihat(data$X))) 
    } else {
      propensity.model = glm(W~poly(X,3), family='binomial', data=data)
      pihat = function(x) { predict(propensity.model, newdata=data.frame(X=x), type='response') }
      model = lm(Y ~ W*X, data=data, weights=data$W + (1-data$W)*pihat(data$X)/(1-pihat(data$X))) 
    }
    cate75 = predict(model, newdata=data.frame(X=75, W=1)) - 
             predict(model, newdata=data.frame(X=75, W=0))
    att1 = mean(predict(model, newdata=data.frame(X=data$X[data$W==1], W=1)))
    att0 = mean(predict(model, newdata=data.frame(X=data$X[data$W==1], W=0)))
    data.frame(n=n, rep=rep, type=type, cate75=cate75, att=att1-att0, att1=att1, att0=att0)
}) %>% bind_rows()

estimands = data.frame(cate75 = mu(1, 75) - mu(0, 75),
                       att = mean(mu(1, pop$X[pop$W==1]) - mu(0, pop$X[pop$W==1])),
                       att1 = mean(mu(1, pop$X[pop$W==1])),
                       att0 = mean(mu(0, pop$X[pop$W==1])))

pop.model = lm(Y ~ W*X, data=pop)
mutilde = function(w,x) { predict(pop.model, newdata=data.frame(W=w,X=x)) }
pop.estimates = data.frame(cate75 = mutilde(1,75) - mutilde(0,75),
                           att    =  mean(mutilde(1, pop$X[pop$W==1]) - mutilde(0, pop$X[pop$W==1])),
                           att1 = mean(mutilde(1, pop$X[pop$W==1])),
                           att0 = mean(mutilde(0, pop$X[pop$W==1]))) 

cate.summaries = estimates %>% group_by(n, type) %>% summarize(bias=mean(cate75)-estimands$cate75, se=sd(cate75))
att.summaries = estimates %>% group_by(n, type) %>% summarize(bias=mean(att)-estimands$att, se=sd(att))
att1.summaries = estimates %>% group_by(n, type) %>% summarize(bias=mean(att1)-estimands$att1, se=sd(att1))
att0.summaries = estimates %>% group_by(n, type) %>% summarize(bias=mean(att0)-estimands$att0, se=sd(att0))
```


# Today

## An Example
::: {.hidden}
$$
\newcommand{\alert}[1]{\textcolor{red}{#1}}
\newcommand{\model}{\mathcal{M}}
\newcommand{\R}{\mathbb{R}}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator{\V}{V}
\DeclareMathOperator{\Var}{\V}
\newcommand{\inblue}[1]{\textcolor{blue}{#1}}
\newcommand{\inred}[1]{\textcolor[RGB]{248,118,109}{#1}}
\newcommand{\ingreen}[1]{\textcolor[RGB]{0,191,196}{#1}}
\newcommand{X}{\mathbf{X}}
\newcommand{Y}{\mathbf{Y}}
$$ 
:::

```{r}
#| out-height: 400px
#| fig-asp: .4 

set.seed(1)

n = 1600
data = pop[sample(nrow(pop), n, replace=TRUE),]

plotX = function(X,W) { X + .15*(2*W-1) }
scaleY = scale_y_continuous(limits=c(0,4), breaks=seq(0,4,by=1))

pop.scatter = ggplot(pop) + 
  geom_point(aes(x=plotX(X,W), y=Y, color=factor(W)), alpha=.035, position=position_jitter(w=.15, h=0), size=.1) + 
  guides(color=FALSE) + scaleY
counterfactual.pop.scatter = ggplot(pop) +
  geom_point(aes(x=plotX(X,!W), y=Y.counterfactual, color=factor(W)), alpha=.035, position=position_jitter(w=.15, h=0), size=.1) + 
  guides(color=FALSE) + scaleY

scatter = ggplot(data) + 
  geom_point(aes(x=plotX(X,W), y=Y, color=factor(W)), alpha=.1, position=position_jitter(w=.15, h=0)) + 
  guides(color=FALSE) + scaleY
pop.scatter
```



- Researchers hypothesize that MHT increases bone density. 
- We'll think about how to look into this. We'll be working with fake data.
- Above, we've got a population of `r m` people eligible for some form of MHT.
  - Their age at the time the study begins is on the $x$-axis.
  - Bone density, measured 2 years later, is on the $y$-axis.
  - Color indicates whether they [did]{.twocolor-green} or [didn't]{.twocolor-red} do MHT.
- What we want to know is how well MHT works, on average, for the people who take it.
- That means comparing green folks---people who take MHT---to versions of themselves who didn't.

::: aside
To make things easier to look at, I'm drawing people who did and didn't receive treatment side-by-side. 
This shift isn't meaningful. Age is a round number: 60, 61, ...
:::

## Counterfactuals

```{r}
#| out-height: 400px
#| layout-ncol: 2 
#| fig-subcap:
#|   - "Real Population: Green People Did MHT, Red People Didn't"
#|   - "Counterfactual Population: Green People Didn't Do MHT, Red People Did" 
pop.scatter
counterfactual.pop.scatter
```

- We can't observe these *counterfactual* versions of people.
- But we can think about what the data would look like if we did.
  - On the left, we've got our real population.
  - On the right, we've got a hypothetical one where everyone who did MHT didn't and vice versa.
  - Note we're keeping the colors the same---counterfactual green people are still green.
- We can think of the average difference as the *average effect of MHT for those who took it*.

$$
\begin{aligned}
\frac{\sum_{j:d_j=1} y_j(1) - y_j(0)}{\sum_{j:d_j=1} 1} = 
\frac{\sum_{j:d_j=1} y_j}{\sum_{j:d_j=1} 1} - \textcolor{blue}{\frac{\sum_{j:d_j=1} y_j(0)}{\sum_{j:d_j=1} 1}}
\end{aligned}
$$

- We call this the *average treatment effect on the treated* (ATT). It's a difference of two means. 
- One is observable (if we have the population). [The other isn't]{.fg style="color:blue"}.

## Summarizing these Populations

```{r}
#| out-height: 400px
#| layout-ncol: 2 
#| fig-asp: .4 
#| fig-subcap:
#|   - "Real Population: Green People Did MHT, Red People Didn't"
#|   - "Counterfactual Population: Green People Didn't Do MHT, Red People Did" 
#| 
pop.summary = pop %>% group_by(X, W) %>% summarize(mean=mean(Y), counterfactual.mean = mean(Y.counterfactual), n=n())
summarized.pop = pop.scatter + 
  geom_point(aes(x=plotX(X,W), y=mean, color=factor(W)), data=pop.summary) + 
  geom_line(aes(x=plotX(X,W), y=mean, color=factor(W)), data=pop.summary) + 
  geom_bar(aes(x=X, y=40*after_stat(prop), fill=factor(W)), 
    position='identity', alpha=.2, data=pop) + 
  guides(fill=FALSE, color=FALSE)

summarized.counterfactual.pop = counterfactual.pop.scatter + 
  geom_point(aes(x=plotX(X,!W), y=counterfactual.mean, color=factor(W)), data=pop.summary) + 
  geom_line(aes(x=plotX(X,!W),  y=counterfactual.mean, color=factor(W)),  data=pop.summary) + 
  geom_bar(aes(x=X, y=40*after_stat(prop), fill=factor(W)), 
    position='identity', alpha=.2, data=pop) + 
  guides(fill=FALSE, color=FALSE)

summarized.pop
summarized.counterfactual.pop
```

- The scatter plot is hard to read. To help, I've overlaid some summaries.
  - Mean bone density for each age and treatment group.
  - Histograms of age for each treatment group.
- We can think of what we're doing in terms of these summaries. Here's the observable mean.

$$
\begin{aligned}
\text{observable mean} 
= \frac{\sum_{j:d_j=1} y_j}{\sum_{j:d_j=1} 1} 
= \sum_x \frac{\sum_{j:d_j=1, x_j=x} 1}{\sum_{j:d_j=1} 1} \frac{\sum_{j:d_j=1, x_j=x} y_j}{\sum_{j:d_j=1, x_j=x} 1} 
= \sum_x \ingreen{p(1,x)}\ingreen{\mu(1,x)}
\end{aligned}
$$

- It's the average of the green curve over the distribution of the green points---i.e., the green histogram.
- The counterfactual mean is the same thing, but in the counterfactual plot.

## Working With Summaries

```{r}
#| out-height: 400px
#| layout-ncol: 2 
#| fig-asp: .4 
#| fig-subcap:
#|   - "Real Population: Green People Did MHT, Red People Didn't"
#|   - "Counterfactual Population: Green People Didn't Do MHT, Red People Did" 
summarized.pop
summarized.counterfactual.pop
```


- This seems like it'd be a problem, since we don't actually see the counterfactual plot. 
- But, comparing our plots, we see something helpful: the green curve on the left looks like the red curve on the left.
- In other words, we can use *what actually happened to the red people* to estimate *what would have happened to the green people*.

$$
\begin{aligned}
\text{observable mean} &= \sum_x \ingreen{p(1,x)}\ingreen{\mu(1,x)} \\
\text{counterfactual mean} &= \ingreen{\sum_x p(1,x)}\inred{\mu(0,x)}
\end{aligned}
$$

- In real data, there's no counterfactual plot, so we can't do this comparison visually.
- But *we've proven* that these curves look the same *in conditionally randomized experiments*.

## Working With Summaries
```{r}
#| out-height: 400px
#| layout-ncol: 2 
#| fig-asp: .4 
#| fig-subcap:
#|   - "Real Population: Green People Did MHT, Red People Didn't"
#|   - "Counterfactual Population: Green People Didn't Do MHT, Red People Did" 
summarized.pop
summarized.counterfactual.pop
```

### Summary

- To calculate the ATT in a conditionally randomized experiment, we ...
  - average the green curve over the distribution of the green points
  - average the red curve over the distribution of the green distribution
  - and subtract 

$$
ATT = \sum_x \ingreen{p(1,x)}\ingreen{\mu(1,x)} - \sum_x \ingreen{p(1,x)}\inred{\mu(0,x)}
$$

- That's what we'd do if we had the population.
- But we have a sample from it, so we'll do the same with *estimates*.

## Working With A Sample

```{r}
#| out-height: 400px
#| layout-ncol: 2 
#| fig-asp: .4 
#| fig-subcap:
#|   - "A Sample from the Population" 
#|   - "A Population: Green People Did MHT, Red People Didn't"
summary = data %>% group_by(X, W) %>% summarize(mean=mean(Y), sd=sd(Y), n=n())
summarized.scatter = scatter + 
  geom_pointrange(aes(x=plotX(X,W), y=mean, ymin=mean-2*sd/sqrt(n), ymax=mean+2*sd/sqrt(n), color=factor(W)), data=summary) + 
  geom_bar(aes(x=X, y=40*after_stat(prop), fill=factor(W)), 
    position='identity', alpha=.2, data=data) + 
  guides(fill=FALSE, color=FALSE) 
summarized.scatter
summarized.pop
```

- This is what we actually have. It's a sample of `r n` people. 
- To give you a sense of the uncertainty, I've added 95% confidence intervals for each subpopulation mean.

$$
\widehat{ATT} = \sum_x \ingreen{\hat p(1,x)}\ingreen{\hat\mu(1,x)} - \sum_x \ingreen{\hat p(1,x)}\inred{\hat\mu(0,x)}
$$

- The problematic part is the second term---averaging the red means over the green distribution. 
    - The green distribution weights the means on the right --- older folks --- heavily.
    - And that's where our red subsample means have large standard errors.
- To reduce variance, we often use other estimates of the subpopulation means $\inred{\mu(0,x)}$.

## Working With A Sample

```{r}
#| out-height: 400px
#| fig-asp: .4  

sample.model = lm(Y ~ W*X, data=data)
pop.model = lm(Y ~ W*X, data=pop)
grid = expand.grid(X=seq(50,80,by=.1), W=c(0,1))
grid$muhat = predict(sample.model, newdata=grid)
grid$mutilde = predict(pop.model, newdata=grid)
grid$mu = mu(grid$W, grid$X)

summarized.scatter.annotated = summarized.scatter + 
  geom_line(aes(x=X, y=muhat, color=factor(W)), data=grid, linetype='dashed') + 
  geom_line(aes(x=X, y=mutilde, group=factor(W)), data=grid, linetype='twodash', color='black', alpha=.4) +
  geom_line(aes(x=X, y=mu, color=factor(W)), data=grid, linetype='solid', alpha=.5)
summarized.scatter.annotated
```

- Here we've fit separately to each group, i.e., the interacted bivariate model.
    - The red/green solid lines are, as before, the subpopulation means
    - The red/green dashed lines are what we'd use to estimate them---the model fit on the sample.
    - The gray dashed lines are their large-sample limits---the model fit on the population.

$$ 
\begin{aligned}
\hat \mu &= \argmin_{m \in \model} \frac{1}{n}\sum_{i=1}^n \qty{ m(D_i, X_i) - Y_i }^2 \\
&\approx \\
\tilde \mu &= \argmin_{m \in \model} \E \qty{ m(D_i, X_i) - Y_i }^2 
\end{aligned}
$$

- I'm showing the estimate and its limit because it's useful to think of our error in two parts.


## Bias 

```{r}
#| out-height: 400px
#| fig-asp: .4  

summarized.scatter.annotated
```

$$
\hat\mu(d,x) - \mu(d,x) = \underset{\text{estimation error}}{\qty{ \hat\mu(d,x) - \tilde\mu(d,x) }} + 
  \underset{\text{modeling error}}{\qty{ \tilde\mu(d,x) - \mu(d,x) }}
$$

- The first part is the difference between our predictions and those we'd get in the large-sample limit.
    - It's the difference between the colored and gray dashed lines. 
    - This has mean zero and we have techniques to understand it---confidence intervals.
- The second part is the difference between the predictions we'd get in the large-sample limit and the ones we want.
    - It's the difference between the gray dashed lines and the solid colored ones. 
    - This is the bias of our prediction. It depends on the population and the model, but not the sample.
    - This is what we're going to focus on today.
- All this passes through to summaries like the CATE and ATT


## Bias for the CATE

```{r}
sample.dist.plot = function(estimate.samples, pop.estimate, estimand) { 
  data = data.frame(estimates=estimate.samples)
  p = ggplot(data) + geom_histogram(aes(x=estimates, y=after_stat(density)), fill='black', bins=30, alpha=.2) + 
           geom_vline(aes(xintercept=mean(estimates)), color='blue') +
           geom_vline(aes(xintercept=mean(estimates) - 2*sd(estimates)), color='blue', linetype='dotted') +
           geom_vline(aes(xintercept=mean(estimates) + 2*sd(estimates)), color='blue', linetype='dotted') +
           geom_vline(aes(xintercept=estimand), color='black', alpha=.5, linewidth=2)
  if(!is.na(pop.estimate)) {
    p = p + geom_vline(aes(xintercept=pop.estimate), color='red', alpha=.5, linewidth=2)
  }
  p
}
```

```{r}
#| out-height: 400px
#| fig-asp: .4 
summarized.scatter.annotated
```

- We can see that our CATE estimate should be fairly biased for older folks (large $x$).
- That's because our fitteed line $\inred{\hat\mu(0,x)}$ gives a poor estimate of our subpopulation means there.

$$
\begin{aligned}
\widehat{CATE}(x) - CATE(x) &=  \qty{\hat\mu(1, x) - \tilde\mu(1, x)} - \qty{\hat\mu(0,x) - \tilde\mu(0,x)} && \text{estimation error} \\
                            &+  \qty{\tilde\mu(1, x) - \mu(1, x)} - \qty{\tilde\mu(0,x) - \mu(0,x)}  && \text{modeling error} \\ 
\end{aligned}
$$

## Implications for Inference

```{r}
#| out-height: 400px
#| fig-asp: .4 
#| layout-ncol: 2 
sample.dist.plot(subset(estimates, n==1600 & type=='unweighted')$cate75, 
                 pop.estimates$cate75, estimands$cate75) + xlim(.5, .8)    
sample.dist.plot(subset(estimates, n==6400 & type=='unweighted')$cate75, 
                 pop.estimates$cate75, estimands$cate75) + xlim(.5, .8)
```

- Above, we're looking at sampling distributions for our CATE estimate at age $x=75$. 
  - On the left, we have sample size $n=1600$.
  - On the right, we have sample size $n=6400$.
- We have vertical lines indicating ...
  - its mean (in blue) +- 2 standard errors (dotted)
  - an estimate based on the population regression (in red) 
  - the CATE itself (in black)
- At $n=1600$, our bias is about one standard error. At $n=6400$, it's twice that.
- Why twice? And what are the inferential implications?


## Bias for the ATT

```{r}
#| out-height: 400px
#| fig-asp: .4 
summarized.scatter.annotated
```

- Our ATT estimate should also be fairly biased because it's an average of the CATE for older folks.

$$
\begin{aligned}
\widehat{ATT} - ATT &=   \sum_x p(1,x) \qty{\hat\mu(1, x) - \tilde\mu(1, x)} - \sum_x p(1,x) \qty{\hat\mu(0,x) - \tilde\mu(0,x)} &&\text{estimation error}  \\
                    &+ \sum_x p(1,x) \qty{\tilde\mu(1, x) - \mu(1, x)} - \sum_x p(1,x) \qty{\tilde\mu(0,x) - \mu(0,x)} &&\text{modeling error}
\end{aligned}
$$

## Implications for Inference

```{r}
#| out-height: 400px
#| fig-asp: .4 
#| layout-ncol: 2 
sample.dist.plot(subset(estimates, n==1600 & type=='unweighted')$att, 
                 pop.estimates$att, estimands$att) + xlim(.3,.5) 
sample.dist.plot(subset(estimates, n==6400 & type=='unweighted')$att, 
                 pop.estimates$att, estimands$att) + xlim(.3,.5)
```

- Above, we're looking at sampling distributions for our ATT estimate.
  - On the left, we have sample size $n=1600$.
  - On the right, we have sample size $n=6400$.
- We have vertical lines indicating ...
  - its mean (in blue) +- 2 standard errors (dotted)
  - an estimate based on the population regression (in red) 
  - the CATE itself (in black)
- At $n=1600$, bias is small relative to standard error. At $n=6400$, it's still not bad.
- Why are we doing better for this estimand?

## Understanding Bias 

```{r}
#| out-height: 400px
#| fig-asp: .4  

summarized.pop.annotated = summarized.pop + 
  geom_line(aes(x=X, y=mutilde, group=factor(W)), 
  data=grid, linetype='twodash', color='black', alpha=.4)
summarized.pop.annotated
```

$$
\begin{aligned}
\text{modeling error} &= \sum_x \ingreen{p(1,x)} \qty{\ingreen{\tilde\mu(1, x) - \mu(1, x)}} - \sum_x \ingreen{p(1,x)} \qty{\inred{\tilde\mu(0,x) - \mu(0,x)}} \\
\qfor \tilde \mu &= \argmin_{m \in \model} \E \qty{ m(D_i, X_i) - Y_i }^2 
\end{aligned}
$$

- Let's break this bias down as a sum of two terms.
 - One associated with our estimate of the [mean outcome with treatment]{.twocolor-green} 
 - One associated with our estimate of the [mean outcome without]{.twocolor-red}
- The first looks like less of a problem. It looks like a line and we've fit a line.
- The second curve $\inred{\mu(0,x)}$ we want to estimate is curved---it's not a line.
- And the line we fit to it is a decent approximation *on the left side of the plot*.
    - That's what least squares is telling it to do. It's where most of our red points are.
    - But it's not what we want. We're averaging it over the distribution of the green points, which are mostly on the right.

## The First Term Is Not A Problem

```{r}
#| out-height: 400px
#| fig-asp: .4  

summarized.scatter.annotated
```

$$
\text{modeling error} = \sum_x \ingreen{p(1,x)} \qty{\ingreen{\tilde\mu(1, x) - \mu(1, x)}} - \textcolor{lightgray}{\sum_x p(1,x) \qty{\tilde\mu(0,x) - \mu(0,x)}}
$$

- And it's not just because we're fitting a line to data with a linear trend.
- In fact, this bias term will zero *for any model with group-specific intercepts*.
- We can prove it by thinking about the population residuals $\tilde \varepsilon = Y - \tilde\mu(D,X)$.
    - We'll use the property that the population residuals are orthogonal to every curve in the model.
    - And the property that an indicator for the group---the curve $m(d,x)=d$---is in the model.

$$
\begin{aligned}
0 &= \E\qty[ D \times (Y-\tilde\mu(D, X)) ]     && \text{orthogonality} \\
  &= \E\qty[ D \times (\mu(D, X) - \tilde\mu(D, X)) ] && \text{iterated expectations} \\
  &= \sum_x 1 \times \qty{\ingreen{\mu(1,x) - \tilde\mu(1,x)}} \ingreen{p(1,x)} && \text{calculating expectations} \\
  &+ \sum_x 0 \times \textcolor{gray}{\qty{\mu(0,x) - \tilde\mu(0,x)} p(0,x)}  && 
\end{aligned}
$$

## Sampling Distributions: 3 Estimators, 3 Sample Sizes
```{r}
#| cache: true
formulas = list(interacted=Y~X*W, additive=Y~X+W, intercepts=Y~W)
term.grid = expand.grid(n=c(1000, 4000, 16000), rep=1:500, formula=names(formulas))

term.estimates = rep(list(), nrow(term.grid))
for(ii in 1:nrow(term.grid)) { 
    n = term.grid$n[ii]
    rep = term.grid$rep[ii]
    formula = term.grid$formula[ii]

    set.seed(rep)
    data = pop[sample(1:nrow(pop), n, replace=TRUE), ]

    model = lm(formula=formulas[[formula]], data=data) 
    att1 = mean(predict(model, newdata=data.frame(X=data$X[data$W==1], W=1)))
    att0 = mean(predict(model, newdata=data.frame(X=data$X[data$W==1], W=0)))

    weights = data$W + (1-data$W)*pi(data$X)/(1-pi(data$X))
    wmodel = lm(formula=formulas[[formula]], data=data, weights = weights) 
    watt1 = mean(predict(wmodel, newdata=data.frame(X=data$X[data$W==1], W=1)))
    watt0 = mean(predict(wmodel, newdata=data.frame(X=data$X[data$W==1], W=0)))
    term.estimates[[ii]] = data.frame(n=n, rep=rep, formula=formula, att1=att1, att0=att0, watt1=watt1, watt0=watt0)
} 
term.estimates = term.estimates %>% bind_rows() 
```

```{r}
#| out-height: 400px
#| fig-asp: .4 
#| layout-nrow: 3 
#| layout-ncol: 3
sample.dist.plot(subset(term.estimates, n==1000 & formula=='interacted')$att1,
                 NA, estimands$att1) + xlim(2.1, 2.3)
sample.dist.plot(subset(term.estimates, n==4000 & formula=='interacted')$att1, 
                 NA, estimands$att1) + xlim(2.1, 2.3) 
sample.dist.plot(subset(term.estimates, n==16000 & formula=='interacted')$att1, 
                 NA, estimands$att1) + xlim(2.1, 2.3)
sample.dist.plot(subset(term.estimates, n==1000 & formula=='additive')$att1,
                 NA, estimands$att1) + xlim(2.1, 2.3)
sample.dist.plot(subset(term.estimates, n==4000 & formula=='additive')$att1, 
                 NA, estimands$att1) + xlim(2.1, 2.3)
sample.dist.plot(subset(term.estimates, n==16000 & formula=='additive')$att1, 
                 NA, estimands$att1) + xlim(2.1, 2.3)
sample.dist.plot(subset(term.estimates, n==1000 & formula=='intercepts')$att1,
                 NA, estimands$att1) + xlim(2.1, 2.3)
sample.dist.plot(subset(term.estimates, n==4000 & formula=='intercepts')$att1, 
                 NA, estimands$att1) + xlim(2.1, 2.3)
sample.dist.plot(subset(term.estimates, n==16000 & formula=='intercepts')$att1, 
                 NA, estimands$att1) + xlim(2.1, 2.3)
```

- Top to Bottom: estimates based on the interacted bivariate, additive bivariate, and two-intercepts ($\hat\mu(d,x) = \hat\beta_0 + \hat\beta_1 d$) models.
- Left to Right: sampling distributions at sample sizes $n=1000, 4000, 16000$.

## Our 3 Estimators

```{r}
#| out-height: 400px
#| fig-asp: .4 
#| layout-ncol: 3
sample.model = lm(Y ~ W*X, data=data)
grid$muhat = predict(sample.model, newdata=grid)
summarized.scatter + geom_line(aes(x=X, y=muhat, color=factor(W)), data=grid, linetype='dashed') 

sample.model = lm(Y ~ W+X, data=data)
grid$muhat = predict(sample.model, newdata=grid)
summarized.scatter + geom_line(aes(x=X, y=muhat, color=factor(W)), data=grid, linetype='dashed') 

sample.model = lm(Y ~ W, data=data)
grid$muhat = predict(sample.model, newdata=grid)
summarized.scatter + geom_line(aes(x=X, y=muhat, color=factor(W)), data=grid, linetype='dashed') 
```

- You can see that we're focusing on the wrong thing when we estimate $\ingreen{\mu(0,x)}$.
- We're estimating it well on the left, but we're using it on the right.

## The Second Term 

```{r}
#| out-height: 400px
#| fig-asp: .4  

summarized.pop + geom_line(aes(x=X, y=mutilde, color=factor(W)), data=grid, linetype='dashed') 
```

- The second term is more troublesome.
- If we repeat this argument as-is, replacing our indicator $D$ with $1-D$, we won't get what we want.

$$
\begin{aligned}
0 &= \E\qty[ (1-D) \times (Y-\tilde\mu(D, X)) ]     && \text{orthogonality} \\
  &= \E\qty[ (1-D) \times (\mu(D, X) - \tilde\mu(D, X)) ] && \text{iterated expectations} \\
  &= \sum_x 1 \times \qty{\inred{\mu(0,x) - \tilde\mu(0,x)}} \inred{p(0,x)} && \text{calculating expectations} \\
  &+ \sum_x 0 \times \textcolor{gray}{\qty{\mu(1,x) - \tilde\mu(1,x)} p(1,x)}  && 
\end{aligned}
$$

- That says the red curve's modeling error would be zero if we were averaging it over the red distribution.
- But we're not. We're averaging it over the green distribution. And we can see this is a problem.

## Sampling Distribution: 3 Estimators, 3 Sample Sizes

```{r}
#| out-height: 400px
#| fig-asp: .4 
#| layout-nrow: 3 
#| layout-ncol: 3

scale_x = coord_cartesian(xlim=c(1.7,2))
sample.dist.plot(subset(term.estimates, n==1000 & formula=='interacted')$att0,
                 NA, estimands$att0)  + scale_x
sample.dist.plot(subset(term.estimates, n==4000 & formula=='interacted')$att0, 
                 NA, estimands$att0) + scale_x 
sample.dist.plot(subset(term.estimates, n==16000 & formula=='interacted')$att0, 
                 NA, estimands$att0) + scale_x 
sample.dist.plot(subset(term.estimates, n==1000 & formula=='additive')$att0,
                 NA, estimands$att0) + scale_x
sample.dist.plot(subset(term.estimates, n==4000 & formula=='additive')$att0, 
                 NA, estimands$att0) + scale_x 
sample.dist.plot(subset(term.estimates, n==16000 & formula=='additive')$att0, 
                 NA, estimands$att0) + scale_x
sample.dist.plot(subset(term.estimates, n==1000 & formula=='intercepts')$att0,
                 NA, estimands$att0) + scale_x
sample.dist.plot(subset(term.estimates, n==4000 & formula=='intercepts')$att0, 
                 NA, estimands$att0) + scale_x
sample.dist.plot(subset(term.estimates, n==16000 & formula=='intercepts')$att0, 
                 NA, estimands$att0) + scale_x
```


## Weighted Least Squares

```{r}
#| out-height: 400px
#| fig-asp: .4  

summarized.pop + geom_line(aes(x=X, y=mutilde, color=factor(W)), data=grid, linetype='dotted') 
```

$$
\text{modeling error} = \textcolor{lightgray}{\sum_x p(1,x) \qty{\tilde\mu(1, x) - \mu(1, x)}} - \sum_x \ingreen{p(1,x)} \qty{\inred{\tilde\mu(0,x) - \mu(0,x)}}
$$

- What will work is *weighted least squares regression*.
- The residuals from this have an analogous orthogonality property in terms of a *weighted* average.
- To get the what we want, we'll use weights $w(x) = \ingreen{p(1,x)}/\inred{p(0,x)}$.
$$
\begin{aligned}
0 &= \E[ (1-D)(Y-\tilde\mu(X)) w(X)] && \text{weighted orthogonality} \\
  &= \E[ (1-D)(\mu(X) - \tilde\mu(X)) w(X)] && \text{iterated expectations} \\
  &= \sum_x 1 \times \qty{\mu(0,x) - \tilde\mu(0,x)} w(x) p(0,x) && \text{calculating expectations} \\
  &+ \sum_x 0 \times \textcolor{gray}{\qty{\mu(1,x) - \tilde\mu(1,x)} w(x) p(0,x)}  && 
  &= \sum_x \qty{\inred{\mu(0,x) - \tilde\mu(0,x)}} \ingreen{p(1,x)}/\inred{p(0,x)} p(0,x) &&
\end{aligned}
$$


## This Works

```{r}
#| out-height: 400px
#| fig-asp: .4 
#| layout-nrow: 3 
#| layout-ncol: 3

scale_x = coord_cartesian(xlim=c(1.7,2))
sample.dist.plot(subset(term.estimates, n==1000 & formula=='interacted')$watt0,
                 NA, estimands$att0)  + scale_x
sample.dist.plot(subset(term.estimates, n==4000 & formula=='interacted')$watt0, 
                 NA, estimands$att0) + scale_x 
sample.dist.plot(subset(term.estimates, n==16000 & formula=='interacted')$watt0, 
                 NA, estimands$att0) + scale_x 
sample.dist.plot(subset(term.estimates, n==1000 & formula=='additive')$watt0,
                 NA, estimands$att0) + scale_x
sample.dist.plot(subset(term.estimates, n==4000 & formula=='additive')$watt0, 
                 NA, estimands$att0) + scale_x 
sample.dist.plot(subset(term.estimates, n==16000 & formula=='additive')$watt0, 
                 NA, estimands$att0) + scale_x
sample.dist.plot(subset(term.estimates, n==1000 & formula=='intercepts')$watt0,
                 NA, estimands$att0) + scale_x
sample.dist.plot(subset(term.estimates, n==4000 & formula=='intercepts')$watt0, 
                 NA, estimands$att0) + scale_x
sample.dist.plot(subset(term.estimates, n==16000 & formula=='intercepts')$watt0, 
                 NA, estimands$att0) + scale_x
```


## Getting Both Terms at Once

$$
\text{modeling error} = \sum_x p(1,x) \qty{\tilde\mu(1, x) - \mu(1, x)} - \sum_x \ingreen{p(1,x)} \qty{\inred{\tilde\mu(0,x) - \mu(0,x)}}
$$

- To be precise, we use this weighted regression. This uses weight ...
  - $1$ for the $D=1$ observations, so our no-bias argument for the first term works.
  - $\ingreen{p(1,x)}/\inred{p(0,x)}$ for the $D=0$ observations, so the one for the second term works, too.

$$
\hat\mu = \argmin_{m \in \model} \frac{1}{n} \sum_{i=1}^n w(D_i, X_i) \qty{ m(D_i, X_i) - Y_i }^2 \qfor w(D,X) = D + (1-D)\frac{\ingreen{p(1,X)}}{\inred{p(0,X)}}
$$

- It may be inobvious that we can do this because we don't know these densities $p(1,x)$ and $p(0,x)$.
- But --- in a conditionally randomized experiment --- we do know their ratio. 
- It's proportional to a function of the within-group assignment probability $\pi(x)$.
- And we don't need to know the constant of proportionality to do weighted least squares.

$$
\begin{aligned}
\frac{p(1,x)}{p(0,x)} 
&= \frac{P(X=x \mid D=1)}{P(X=x \mid D=0)} 
= \frac{P(X=x, D=1) / P(D=1)}{ P(X=x, D=0) / P(D=0)} \\
&= \frac{P(D=1 \mid X=x) P(X=x) / P(D=1)}{ P(D=0 \mid X=x) P(X=x) / P(D=0)}
= \frac{\pi(x)}{1-\pi(x)} \cdot \frac{P(D=1)}{P(D=0)}
\end{aligned}
$$

- This is equivalent, but implementable.
$$
\hat\mu = \argmin_{m \in \model} \frac{1}{n} \sum_{i=1}^n w(D_i, X_i) \qty{ m(D_i, X_i) - Y_i }^2 \qfor w(D,X) = D + (1-D)\frac{\pi(X)}{1-\pi(X)}
$$

## Our 3 Estimators

```{r}
#| out-height: 400px
#| fig-asp: .4 
#| layout-ncol: 3

weights = data$W + (1-data$W)*pi(data$X)/(1-pi(data$X)) 
sample.model = lm(Y ~ W*X, weights = weights, data=data)
grid$muhat = predict(sample.model, newdata=grid)
summarized.scatter + geom_line(aes(x=X, y=muhat, color=factor(W)), data=grid, linetype='dashed') 

sample.model = lm(Y ~ W+X, weights = weights, data=data)
grid$muhat = predict(sample.model, newdata=grid)
summarized.scatter + geom_line(aes(x=X, y=muhat, color=factor(W)), data=grid, linetype='dashed') 

sample.model = lm(Y ~ W,   weights = weights, data=data)
grid$muhat = predict(sample.model, newdata=grid)
summarized.scatter + geom_line(aes(x=X, y=muhat, color=factor(W)), data=grid, linetype='dashed') 
```

- You can see we're getting better fits on the right, where it matters.

## IPW and Point Estimation

```{r}
#| out-height: 600px
rmse.plot = ggplot(att.summaries[att.summaries$type %in% c('unweighted', 'weighted'), ]) + 
  geom_line(aes(x=n, y=sqrt(bias^2 + se^2), color=type )) +
  geom_point(aes(x=n, y=sqrt(bias^2 + se^2), color=type )) + 
  geom_hline(aes(yintercept=0), alpha=.3, linewidth=2) +
  scale_x_sqrt(breaks=ns)
rmse.plot
```

- Here's a simulation study comparing ATT estimators based on [weighted]{.twocolor-green} and [unweighted]{.twocolor-red] least squares fits of the interacted bivariate model.
- I'm plotting the square root of mean squared error (RMSE) at a range of sample sizes.
- We can see that, except in very small samples, the weighted one performs better.
- And the error curve of the [unweighted]{.twocolor-red] one is flattening out.
- It's not converging to zero. What is it converging to?

## IPW and Point Estimation

```{r}
#| out-height: 600px
rmse.plot + geom_hline(aes(yintercept=abs(pop.estimates$att-estimands$att)), color='red', alpha=.8, linetype='dashed')
```

- It's converging to the error we get using the **population least squares fit**.
- I've plotted that as a red dashed horizontal line.

## IPW in Observational Studies 

```{r}
#| out-height: 600px
rmse.plot + geom_hline(aes(yintercept=abs(pop.estimates$att-estimands$att)), color='red', alpha=.8, linetype='dashed')
```

- Often, we analyze observational studies as if they were conditionally randomized experiments.
  - This works if, among groups with identical covariates, people select treatment with the same probability.
  - This is usually an imperfect approximation, but often it's the best we can do,
- What this means, implementation-wise, is that we don't know the randomization probabilities.
- And we have to make do with estimates. How well do you think this works?

$$
\hat\mu = \argmin_{m \in \model} \frac{1}{n} \sum_{i=1}^n w(D_i, X_i) \qty{ m(D_i, X_i) - Y_i }^2 \qfor \hat w(D,X) = D + (1-D)\frac{\hat\pi(X)}{1-\hat\pi(X)}
$$


## It Usually Works Pretty Well

```{r}
#| out-height: 600px
rmse.plot.all = ggplot(att.summaries) + 
  geom_line(aes(x=n, y=sqrt(bias^2 + se^2), color=type )) +
  geom_point(aes(x=n, y=sqrt(bias^2 + se^2), color=type )) + 
  geom_hline(aes(yintercept=0), alpha=.3, linewidth=2) +
  scale_x_sqrt(breaks=ns) 
rmse.plot.all + geom_hline(aes(yintercept=abs(pop.estimates$att-estimands$att)), color='red', alpha=.8, linetype='dashed')
```



