---
title: "QTM 220: Lecture 18"
subtitle: "Causal Inference"
format: 
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    default-image-extension: png
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---


```{r load-libraries}
library(purrr)
library(tibble)
library(dplyr)
library(magrittr)
library(stringr)
library(ggplot2)
library(htmltools)
library(kableExtra)
library(scales)
library(cpsR)
library(maps)

options(knitr.kable.NA = '')

census.api.key = '8084f81e4f71e4d32638dac15f188b424f72d063'
# for cps variable interpretation see https://www2.census.gov/programs-surveys/cps/datasets/2022/march/asec2022_ddl_pub_full.pdf
percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }
```

```{r ggplot-themes}
fontsize = 16
textcolor = "#888888"

class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
        panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x  = element_text(colour = textcolor, size=fontsize),
        axis.text.y  = element_text(colour = textcolor, size=fontsize),
        axis.title.x  = element_text(colour = textcolor),
        axis.title.y  = element_text(colour = textcolor, angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
              axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
           panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(simplified.theme)
```


```{r table-styling}
ellipsis <- function(df, nhead=5L, ntail=2) {
    rownames(df) = as.character(1:nrow(df))
    stopifnot("data.frame" %in% class(df))
    els <- rep(NA, ncol(df)) %>% 
        matrix(nrow=1, dimnames=list(NULL, names(df))) %>% 
        data.frame(stringsAsFactors=FALSE) %>% 
        tibble::as_tibble()
    rownames(els) = "⋮"
    dplyr::bind_rows(head(df, nhead), els, tail(df, ntail))
}
```

# Today

## An Example
::: {.hidden}
$$
\newcommand{\alert}[1]{\textcolor{red}{#1}}
\newcommand{\model}{\mathcal{M}}
\newcommand{\R}{\mathbb{R}}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator{\V}{V}
\DeclareMathOperator{\Var}{\V}
\newcommand{\inblue}[1]{\textcolor{blue}{#1}}
\newcommand{\inred}[1]{\textcolor[RGB]{248,118,109}{#1}}
\newcommand{\ingreen}[1]{\textcolor[RGB]{0,191,196}{#1}}
\newcommand{X}{\mathbf{X}}
\newcommand{Y}{\mathbf{Y}}
$$ 
:::

```{r}
#| out-height: 400px
#| fig-asp: .4 
 
set.seed(1) 
m = 2e4
pi = function(x) { 1/(1+exp(-(x-60)/8 )) }
mu = function(w, x) { 1 - (x-50)/30 + w*((x-50)/30)^2/2 }  

X = 50+round(30*runif(m))
W = 0 + (runif(m) <= pi(X))
Y = mu(W,X) + rnorm(m)
pop = data.frame(X=X, W=W, Y=Y, Y.counterfactual=Y-mu(W,X)+mu(1-W,X))

n = 5000
data = pop[sample(nrow(pop), n, replace=TRUE),]

plotX = function(X,W) { X + .15*(2*W-1) }
pop.scatter = ggplot(pop) + 
  geom_point(aes(x=plotX(X,W), y=Y, color=factor(W)), alpha=.1, position=position_jitter(w=.15, h=0), size=.2) + 
  guides(color=FALSE)
counterfactual.pop.scatter = ggplot(pop) +
  geom_point(aes(x=plotX(X,!W), y=Y.counterfactual, color=factor(W)), alpha=.1, position=position_jitter(w=.15, h=0), size=.2) + 
  guides(color=FALSE)

scatter = ggplot(data) + 
  geom_point(aes(x=plotX(X,W), y=Y, color=factor(W)), alpha=.1, position=position_jitter(w=.15, h=0)) + 
  guides(color=FALSE)
pop.scatter
```


- Researchers hypothesize that MHT increases bone density. 
- We'll think about how to look into this.
- Above, we've got a population of `r approx(m)` people eligible for some form of MHT.
  - Their age at the time the study begins is on the $x$-axis.
  - Bone density, measured 2 years later, is on the $y$-axis.
  - Color indicates whether they [did]{.twocolor-green} or [didn't]{.twocolor-red} do MHT.
- What we want to know is how well MHT works, on average, for the people who take it.
- That means comparing green folks---people who take MHT---to versions of themselves who didn't.

::: aside
To make things easier to look at, I'm drawing people who did and didn't receive treatment side-by-side. 
This shift isn't meaningful. Age is a round number: 60, 61, ...
:::

## Counterfactuals

```{r}
#| out-height: 400px
#| layout-ncol: 2 
#| fig-subcap:
#|   - "Real Population: Green People Did MHT, Red People Didn't"
#|   - "Counterfactual Population: Green People Didn't Do MHT, Red People Did" 
pop.scatter
counterfactual.pop.scatter
```

- We can't observe these *counterfactual* versions of people.
- But we can think about what the data would look like if we did.
  - On the left, we've got our real population.
  - On the right, we've got a hypothetical one where everyone who did MHT didn't and vice versa.
  - Note we're keeping the colors the same---counterfactual green people are still green.
- We can think of the average difference as the *average effect of MHT for those who took it*.

$$
\begin{aligned}
\frac{\sum_{j:d_j=1} y_j(1) - y_j(0)}{\sum_{j:d_j=1} 1} = 
\frac{\sum_{j:d_j=1} y_j}{\sum_{j:d_j=1} 1} - \textcolor{blue}{\frac{\sum_{j:d_j=1} y_j(0)}{\sum_{j:d_j=1} 1}}
\end{aligned}
$$

- We call this the *average treatment effect on the treated* (ATT). It's a difference of two means. 
- One is observable (if we have the population). [The other isn't]{.fg style="color:blue"}.

## Summarizing these Populations

```{r}
#| out-height: 400px
#| layout-ncol: 2 
#| fig-asp: .4 
#| fig-subcap:
#|   - "Real Population: Green People Did MHT, Red People Didn't"
#|   - "Counterfactual Population: Green People Didn't Do MHT, Red People Did" 
#| 
mean.shape=21
pop.summary = pop %>% group_by(X, W) %>% summarize(mean=mean(Y), counterfactual.mean = mean(Y.counterfactual), n=n())
summarized.pop = pop.scatter + 
  geom_point(aes(x=plotX(X,W), y=mean, color=factor(W)), data=pop.summary, shape=mean.shape, color='black') + 
  geom_line(aes(x=plotX(X,W), y=mean, color=factor(W)), data=pop.summary) + 
  geom_histogram(aes(x=X, y=30*after_stat(density), fill=factor(W)), 
    position='identity', alpha=.2, data=pop, bins=length(unique(pop$X))) + 
  guides(fill=FALSE, color=FALSE)

summarized.counterfactual.pop = counterfactual.pop.scatter + 
  geom_point(aes(x=plotX(X,!W), y=counterfactual.mean, color=factor(W)), data=pop.summary, shape=mean.shape, color='black') + 
  geom_line(aes(x=plotX(X,!W),  y=counterfactual.mean, color=factor(W)),  data=pop.summary) + 
  geom_histogram(aes(x=X, y=30*after_stat(density), fill=factor(W)), 
    position='identity', alpha=.2, data=pop, bins=length(unique(pop$X))) + 
  guides(fill=FALSE, color=FALSE)

summarized.pop
summarized.counterfactual.pop
```

- The scatter plot is hard to read. To help, I've overlaid some summaries.
  - Mean bone density for each age and treatment group.
  - Histograms of age for each treatment group.
- We can think of what we're doing in terms of these summaries. Here's the observable mean.

$$
\begin{aligned}
\text{observable mean} 
= \frac{\sum_{j:d_j=1} y_j}{\sum_{j:d_j=1} 1} 
= \sum_x \frac{\sum_{j:d_j=1, x_j=x} 1}{\sum_{j:d_j=1} 1} \frac{\sum_{j:d_j=1, x_j=x} y_j}{\sum_{j:d_j=1, x_j=x} 1} 
= \sum_x \ingreen{p(1,x)}\ingreen{\mu(1,x)}
\end{aligned}
$$

- It's the average of the green curve over the distribution of the green points---i.e., the green histogram.
- The counterfactual mean is the same thing, but in the counterfactual plot.

## Working With Summaries

```{r}
#| out-height: 400px
#| layout-ncol: 2 
#| fig-asp: .4 
#| fig-subcap:
#|   - "Real Population: Green People Did MHT, Red People Didn't"
#|   - "Counterfactual Population: Green People Didn't Do MHT, Red People Did" 
summarized.pop
summarized.counterfactual.pop
```


- This seems like it'd be a problem, since we don't actually see the counterfactual plot. 
- But, comparing our plots, we see something helpful: the green curve on the left looks like the red curve on the left.
- In other words, we can use *what actually happened to the red people* to estimate *what would have happened to the green people*.

$$
\begin{aligned}
\text{observable mean} &= \sum_x \ingreen{p(1,x)}\ingreen{\mu(1,x)} \\
\text{counterfactual mean} &= \ingreen{\sum_x p(1,x)}\inred{\mu(0,x)}
\end{aligned}
$$

- In real data, there's no counterfactual plot, so we can't do this comparison visually.
- But *we've proven* that these curves look the same *in conditionally randomized experiments*.

## Sampling

```{r}
#| out-height: 400px
#| layout-ncol: 2 
#| fig-asp: .4 
summary = data %>% group_by(X, W) %>% summarize(mean=mean(Y), sd=sd(Y), n=n())
summarized.scatter = scatter + 
  geom_pointrange(aes(x=X+.2*(2*W-1), y=mean, ymin=mean-2*sd/sqrt(n), ymax=mean+2*sd/sqrt(n), color=factor(W)), data=summary) + 
  geom_histogram(aes(x=X, y=120*after_stat(density), fill=factor(W)), 
    position='identity', alpha=.2, data=data, bins=length(unique(data$X))) + 
  guides(fill=FALSE, color=FALSE)
summarized.scatter
summarized.pop
```

- The other part of our problem is the **statistical element**.
  - In reality, we don't have the whole population. 
  - We just have a sample of, say, `r approx(n)` people. 
- So we have to ask whether what we see in the sample is true of the population.

## Plan
  
- We're going to talk about how to answer these questions.
- You've seen the main ideas before in simpler contexts.
- All we've got to do is generalize them and put them together.

# Conditionally Randomized Experiments

## The Idea

```{r}
#| out-height: 500px
#| fig-asp: .4 
twolevel.data = data[data$X %in% c(60,70), ]
twolevel.scatter = ggplot(twolevel.data) + 
  geom_point(aes(x=X, y=Y, color=factor(W)), alpha=.5, size=3, position=position_jitter(w=3,h=0)) + 
  guides(color=FALSE) + scale_x_continuous(breaks=c(60,05))
twolevel.scatter
```

- To make things simple, let's start by looking at a subset of the data.
  - We'll look at participants who were either 60 or 70 when the study began.
  - We'll ignore the others. This gives us two age groups---a binary covariate.
- Suppose that, for each group, a coin was flipped to determine whether they do MHT.
  - Different --- and differently weighted --- coins. 
  - That's why different proportions of people do MHT in the two groups.
- We call this a **conditionally randomized experiment**.
  - a.k.a. a **blocked** or **stratified** randomized experiment.

## Essentially, We Have Two Experiments

$$
\begin{array}{cccc|ccc}
i & X_i & D_i & Y_i & Y_i(1) & Y_i(0) & Y_i(1)-Y_i(0) \\
\hline
1 & \textcolor{blue}{60} &1 & 70 & \textcolor{blue}{70} & \alert{70} & \alert{0} \\
2 & \textcolor{magenta}{70} &1 & 80 & \textcolor{magenta}{80} & \alert{70} & \alert{10} \\
3 & \textcolor{blue}{60} & 0 & 60 & \alert{60} & \textcolor{blue}{60} & \alert{0}\\
4 & \textcolor{magenta}{70} &0& 70 & \alert{80} & \textcolor{magenta}{70} & \alert{10} \\
5 & \textcolor{blue}{60} & 0&70 & \alert{70} & \textcolor{blue}{70} & \alert{0} \\
6 & \textcolor{blue}{60}& 0 & 80 & \alert{80} & \textcolor{blue}{80} & \alert{0} \\
\end{array}
$$

- For each age group, we have a randomized experiment. 
- Among those with $\textcolor{blue}{X=60}$ and $\textcolor{magenta}{X=70}$ ...
  - Everything we've said before abiout randomized experiments is true.
  - We can think of the people who do and don't do MHT as being drawn from the same population.
  - That means that --- if we ignore $D_i$ and $Y_i$ --- they have the same distribution.
  - We have to ignore this stuff to get idenitical distribution because it's *downstream of randomization*.
- This allows us to get at the *distribution of our missing data*. 
- [Remember red numbers are not observed]{.fg style="color:red"}.

## Filling in Missing Data

$$
\begin{array}{cccc|ccc}
i & X_i & D_i & Y_i & Y_i(1) & Y_i(0) & Y_i(1)-Y_i(0) \\
\hline
1 & \textcolor{blue}{60} &1 & 70 & \textcolor{blue}{70} & \alert{70} & \alert{0} \\
2 & \textcolor{magenta}{70} &1 & 80 & \textcolor{magenta}{80} & \alert{70} & \alert{10} \\
3 & \textcolor{blue}{60} & 0 & 60 & \alert{60} & \textcolor{blue}{60} & \alert{0}\\
4 & \textcolor{magenta}{70} &0& 70 & \alert{80} & \textcolor{magenta}{70} & \alert{10} \\
5 & \textcolor{blue}{60} & 0&70 & \alert{70} & \textcolor{blue}{70} & \alert{0} \\
6 & \textcolor{blue}{60}& 0 & 80 & \alert{80} & \textcolor{blue}{80} & \alert{0} \\
\end{array}
$$

What is the average $Y_i(1)$ among the participants with $X=60$?

::: fragment

$$
\frac{\textcolor{blue}{70} +\alert{60} +\alert{70} + \alert{80}}{\textcolor{blue}{1} + \alert{3}}  = 70
$$

:::


What is the average $Y_i(1)$ among the *treated* ($D=1$) participants with $X=60$?

::: fragment
$$
\frac{\textcolor{blue}{70} }{\textcolor{blue}{1}}  = 70
$$
:::

What is the average $Y_i(0)$ among the participants with $X=60$?

::: fragment
$$
\frac{\textcolor{red}{70} +\textcolor{blue}{60} +\textcolor{blue}{70}+\textcolor{blue}{80}}{4}  = 70
$$
:::


What is the average $Y_i(0)$ among the *untreated* ($D=0$) participants with $X=60$?

::: fragment
$$
\frac{\textcolor{blue}{60} +\textcolor{blue}{70}+\textcolor{blue}{80}}{3}  = 70
$$

:::

::: aside
This is a stylized example. Randomization implies equivalance *in distribution*, but not in any given sample.
:::


## Conditional ATE for $\textcolor{blue}{X=60}$

$$
\begin{array}{cccc|ccc}
i & X_i & D_i & Y_i & Y_i(1) & Y_i(0) & Y_i(1)-Y_i(0) \\
\hline
1 & \textcolor{blue}{60} &1 & 70 & \textcolor{blue}{70} & \alert{70} & \alert{0} \\
2 & \textcolor{magenta}{70} &1 & 80 & \textcolor{magenta}{80} & \alert{70} & \alert{10} \\
3 & \textcolor{blue}{60} & 0 & 60 & \alert{60} & \textcolor{blue}{60} & \alert{0}\\
4 & \textcolor{magenta}{70} &0& 70 & \alert{80} & \textcolor{magenta}{70} & \alert{10} \\
5 & \textcolor{blue}{60} & 0&70 & \alert{70} & \textcolor{blue}{70} & \alert{0} \\
6 & \textcolor{blue}{60}& 0 & 80 & \alert{80} & \textcolor{blue}{80} & \alert{0} \\
\end{array}
$$

- Because we have a randomized experiment within the  $\textcolor{blue}{X=60}$ group ...
- The difference in *subpopulation means* is equal to the average effect of treatment within that group.
- We call this the **conditional average treatment effect** ($CATE(60)$).
$$
CATE(60) = \E[Y_i(1) - Y_i(0) \mid X_i=0] = \E[Y_i \mid D_i=1, X_i=60] - \E[Y_i \mid D_i=0, X_i=60]
$$
- The difference in subsample means $\textcolor{blue}{X=60}$ is an unbiased estimator of 
this quantity.

$$
\begin{aligned}
\widehat{CATE}(60) &=\frac{\textcolor{blue}{70} }{1}  - \frac{\textcolor{blue}{60} +\textcolor{blue}{70}+\textcolor{blue}{80}}{3} \\
&= \frac{\textcolor{blue}{70} +\alert{60} +\alert{70} + \alert{80}}{4} - \frac{\textcolor{red}{70} +\textcolor{blue}{60} +\textcolor{blue}{70}+\textcolor{blue}{80}}{4} \\
&= \frac{\alert{0} +\alert{0} +\alert{0} + \alert{0}}{4}
\end{aligned}
$$



## Conditional ATE for $\textcolor{magenta}{X=70}$

$$
\begin{array}{cccc|ccc}
i & X_i & D_i & Y_i & Y_i(1) & Y_i(0) & Y_i(1)-Y_i(0) \\
\hline
1 & \textcolor{blue}{60} &1 & 70 & \textcolor{blue}{70} & \alert{70} & \alert{0} \\
2 & \textcolor{magenta}{70} &1 & 80 & \textcolor{magenta}{80} & \alert{70} & \alert{10} \\
3 & \textcolor{blue}{60} & 0 & 60 & \alert{60} & \textcolor{blue}{60} & \alert{0}\\
4 & \textcolor{magenta}{70} &0& 70 & \alert{80} & \textcolor{magenta}{70} & \alert{10} \\
5 & \textcolor{blue}{60} & 0&70 & \alert{70} & \textcolor{blue}{70} & \alert{0} \\
6 & \textcolor{blue}{60}& 0 & 80 & \alert{80} & \textcolor{blue}{80} & \alert{0} \\
\end{array}
$$

- Because we have a randomized experiment within the  $\textcolor{magenta}{X=70}$ group ...
- The difference in *subpopulation means* is equal to the average effect of treatment within that group.
- We call this the **conditional average treatment effect** ($CATE(70)$).
$$
CATE(70) = \E[Y_i(1) - Y_i(0) \mid X_i=0] = \E[Y_i \mid D_i=1, X_i=70] - \E[Y_i \mid D_i=0, X_i=70]
$$
- The difference in subsample means $\textcolor{magenta}{X=70}$ is an unbiased estimator of 
this quantity.

$$
\begin{aligned}
\widehat{CATE}(70) &= \frac{\textcolor{magenta}{80} }{1}  - \frac{\textcolor{magenta}{70} }{1} \\
&=\frac{\textcolor{magenta}{80} +\alert{80} }{2} - \frac{\alert{70}  + \textcolor{magenta}{70}}{2} \\
&= \frac{\alert{10} +\alert{10}}{2}
\end{aligned}
$$

## The ATE

$$
\begin{array}{cccc|ccc}
i & X_i & D_i & Y_i & Y_i(1) & Y_i(0) & Y_i(1)-Y_i(0) \\
\hline
1 & \textcolor{blue}{60} &1 & 70 & \textcolor{blue}{70} & \alert{70} & \alert{0} \\
2 & \textcolor{magenta}{70} &1 & 80 & \textcolor{magenta}{80} & \alert{70} & \alert{10} \\
3 & \textcolor{blue}{60} & 0 & 60 & \alert{60} & \textcolor{blue}{60} & \alert{0}\\
4 & \textcolor{magenta}{70} &0& 70 & \alert{80} & \textcolor{magenta}{70} & \alert{10} \\
5 & \textcolor{blue}{60} & 0&70 & \alert{70} & \textcolor{blue}{70} & \alert{0} \\
6 & \textcolor{blue}{60}& 0 & 80 & \alert{80} & \textcolor{blue}{80} & \alert{0} \\
\end{array}
$$

- To get an *aggregate summary* of the effect of MHT, we can average the two within-subpopulation effects.
$$
\begin{aligned}
\text{ATE} 
&= \E\qty[ \E\qty{ Y_i(1) - Y_i(0) \mid X_i } ] \\
&= \E\qty[ \E\qty{ Y_i \mid D_i=1, X_i } - \E\qty{ Y_i \mid D_i=0, X_i } ] \\
&= \E\qty[ CATE(X_i) ] 
\end{aligned}
$$

- We can estimate this with a weighted average of our two estimates.
$$
\begin{aligned}
\widehat{ATE} &= \textcolor{blue}{\frac{4}{6}} \widehat{CATE}(60) + \textcolor{magenta}{\frac{2}{6}} \widehat{CATE}(70) \\
&=\textcolor{blue}{\frac{4}{6}} \cdot \left\{\frac{70}{1} -  \frac{60+70+80}{3}\right\} + \textcolor{magenta}{\frac{2}{6}} \cdot \left\{\frac{80}{1} -  \frac{70}{1}\right\} \\
&= \frac{20}{6}
\end{aligned}
$$


## The ATT

$$
\begin{array}{cccc|ccc}
i & X_i & D_i & Y_i & Y_i(1) & Y_i(0) & Y_i(1)-Y_i(0) \\
\hline
1 & \textcolor{blue}{60} &1 & 70 & \textcolor{blue}{70} & \alert{70} & \alert{0} \\
2 & \textcolor{magenta}{70} &1 & 80 & \textcolor{magenta}{80} & \alert{70} & \alert{10} \\
3 & \textcolor{blue}{60} & 0 & 60 & \alert{60} & \textcolor{blue}{60} & \alert{0}\\
4 & \textcolor{magenta}{70} &0& 70 & \alert{80} & \textcolor{magenta}{70} & \alert{10} \\
5 & \textcolor{blue}{60} & 0&70 & \alert{70} & \textcolor{blue}{70} & \alert{0} \\
6 & \textcolor{blue}{60}& 0 & 80 & \alert{80} & \textcolor{blue}{80} & \alert{0} \\
\end{array}
$$

- We can, if we prefer, average the CATE over *the distribution of treated units*. 
- This changes our weights. And gives us a different summary. It's the ATT.
$$
\begin{aligned}
\text{ATT} 
&= \E\qty[ \E\qty{ Y_i(1) - Y_i(0) \mid X_i } \mid D_i = 1] \\
&= \E\qty[ \E\qty{ Y_i \mid D_i=1, X_i } - \E\qty{ Y_i \mid D_i=0, X_i } \mid D_i=1] \\
&= \E\qty[ CATE(X_i) \mid D_i=1] 
\end{aligned}
$$

- We can estimate this with a weighted average of our two estimates.
$$
\begin{aligned}
\widehat{ATT} &= \textcolor{blue}{\frac{1}{2}} \widehat{CATE}(60) + \textcolor{magenta}{\frac{1}{2}} \widehat{CATE}(70) \\
&=\textcolor{blue}{\frac{1}{2}} \cdot \left\{\frac{70}{1} -  \frac{60+70+80}{1}\right\} + \textcolor{magenta}{\frac{1}{2}} \cdot \left\{\frac{80}{1} -  \frac{70}{1}\right\} \\
&= \frac{10}{2}
\end{aligned}
$$

## Working With All the Data
```{r}
#| out-height: 400px
#| layout-ncol: 2 
#| fig-asp: .4 
summarized.pop
summarized.counterfactual.pop
```

- This formalizes what we were talking about earlier.
- It tells us we can use subpopulation means *that we can estimate* in place of *counterfactual means* we (otherwise) cannot.

$$
\begin{aligned}
\E[Y(1) \mid D_i=1] &=  \E[ \E[Y_i \mid D_i=1, X_i]] = \sum_x p(1,x)\mu(1,x) \\
\E[Y(0) \mid D_i=1] &=  \E[ \E[Y_i \mid D_i=0, X_i]] = \sum_x p(1,x)\mu(0,x).
\end{aligned}
$$

## The Challenge

```{r}
#| out-height: 700px
#| layout-ncol: 2 
#| fig-asp: .4  

sample.model = lm(Y ~ W*X, data=data)
pop.model = lm(Y ~ W*X, data=pop)
grid = expand.grid(X=seq(50,80,by=.1), W=c(0,1))
grid$muhat = predict(sample.model, newdata=grid)
grid$mutilde = predict(pop.model, newdata=grid)

summarized.scatter + geom_line(aes(x=X, y=muhat, color=factor(W)), data=grid, linetype='dashed')
summarized.counterfactual.pop + geom_line(aes(x=X, y=mutilde, color=factor(W)), data=grid, linetype='dashed')
```

- It's not always so easy to estimate these subpopulation means.
- Often, the corresponding subsamples are pretty small.
- This means we often wind up using estimates that are *biased*.


# Bias and Inverse Probability Weighting

## Bias

```{r}
#| out-height: 400px
#| fig-asp: .4 
model = lm(Y ~ W*X, data=data)
grid = expand.grid(X=seq(50,80,by=.1), W=c(0,1))
grid$muhat = predict(model, newdata=grid)
summarized.scatter + geom_line(aes(x=X, y=muhat, color=factor(W)), data=grid)
```

- As sample size increases, the random variation becomes less impactful.
- The predictions we get from our *estimated regression model* will get close to what we'd get from regression in the population.
$$ \hat \mu = \argmin_{m \in \model} \frac{1}{n}\sum_{i=1}^n \qty{ m(D_i, X_i) - Y_i }^2 \quad \approx \quad
   \tilde \mu = \argmin_{m \in \model} \E \qty{ m(D_i, X_i) - Y_i }^2 
$$
- But the predictions of the population regression model are not necessarily close to the subpopulation means.
- As a result, no matter what sample size we use, we'll get bias.
- It's useful to think of our error at having two pieces.
  - *estimation error*. This is random. It depends on the sample.
  - *modeling error*.  This isn't. It depends only on the model and the population.
- Estimation error goes away in large samples. Modeling error doesn't.

$$
\hat\mu(x) - \mu(x) = \underset{\text{estimation error}}{\qty{ \hat\mu(x) - \tilde\mu(x) }} + 
  \underset{\text{modeling error}}{\qty{ \tilde\mu(x) - \mu(x) }}
$$

## Bias and Summaries

```{r}
#| out-height: 400px
#| fig-asp: .4 
model = lm(Y ~ W + X, data=data)
grid = expand.grid(X=seq(50,80,by=.1), W=c(0,1))
grid$muhat = predict(model, newdata=grid)
summarized.scatter + geom_line(aes(x=X, y=muhat, color=factor(W)), data=grid)
```

- All this passes through to summaries like the CATE, ATE, and ATT

$$
\begin{aligned}
\widehat{CATE}(x) - CATE(x) &= \underset{\text{estimation error}}{\qty{ \hat\mu(1, x) - \tilde\mu(0, x) }} + 
  \underset{\text{modeling error}}{\qty{ \tilde\mu(1, x) - \mu(0, x) }} \\
\widehat{ATE} - ATE &= \underset{\text{estimation error}}{ \sum_x P_x \qty{ \hat\mu(1, x) - \tilde\mu(0, x) }} + 
  \underset{\text{modeling error}}{\sum_x P_x \qty{ \hat\mu(1, x) - \tilde\mu(0, x) }}
\end{aligned}
$$

## How Do We Make This Modeling Error Go Away?

```{r}
#| out-height: 500px
#| fig-asp: .4 
summary = data %>% group_by(X, W) %>% summarize(mean=mean(Y), sd=sd(Y), n=n())
summarized.scatter = scatter + 
  geom_pointrange(aes(x=X+.2*(2*W-1), y=mean, ymin=mean-2*sd/sqrt(n), ymax=mean+2*sd/sqrt(n), color=factor(W)), data=summary) + 
  geom_histogram(aes(x=X, y=120*after_stat(density), fill=factor(W)), 
    position='identity', alpha=.2, data=data, bins=length(unique(data$X))) + 
  guides(fill=FALSE, color=FALSE)
summarized.scatter
```
