---
format:
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---

```{r setup}
library(usmap)
library(sf)

library(ggplot2)

library(scales)
library(knitr)

percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }
```

```{r ggplot-themes}
class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
		    panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
		    axis.ticks.x = element_blank(),
		    axis.ticks.y = element_blank(),
		    axis.text.x  = element_text(colour = "#aaaaaa"),
		    axis.text.y  = element_text(colour = "#aaaaaa"),
		    axis.title.x  = element_text(colour = "#aaaaaa"),
		    axis.title.y  = element_text(colour = "#aaaaaa", angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
   	         axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
	         panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(simplified.theme)
```

```{r population-and-sample}
#| echo: false 
our.seed = 220
set.seed(our.seed)
N = 7233584
n = 625

b0 = floor(.7*N)
b1 = ceiling(.3*N)
y0 = rbinom(n=b0,size=1,prob=.68)
y1 = rbinom(n=b1,size=1,prob=.74)
dfpop = data.frame(y=c(y0,y1), x=c(rep(0,b0),rep(1,b1)))
dfpop = dfpop[sample(1:N),] # shuffle the rows
x = dfpop$x
y = dfpop$y


# Generate Sample Data for us (J1) and three friends (J2,J3,JM)
J1 = sample(1:N,size=n,replace=T)
J2 = sample(1:N,size=n,replace=T)
J3 = sample(1:N,size=n,replace=T)
JM = sample(1:N,size=n,replace=T)

J = J1
dfsam = dfpop[J,]
Y = dfsam$y
X = dfsam$x

# Generate Sampling Distribution
S = 10000
subsamMean = rep(NA,S)
diffMean = rep(NA,S)
for( s in 1:S ){
	ind = sample(1:N,size=625,replace=T)
	dfind = dfpop[ind,]
	subsamMean[s] = mean(dfind$y[dfind$x==1])
	diffMean[s] = mean(dfind$y[dfind$x==1])-mean(dfind$y[dfind$x==0])	
}

B = 10000
bootsubsamMean = rep(NA,B)
bootdiffsamMean = rep(NA,B)
for( b in 1:B ){
bootind = sample(1:625,size=625,replace=T)
dfboot = dfsam[bootind,]
bootsubsamMean[b] = mean(dfboot$y[dfboot$x==1])
bootdiffsamMean[b] = mean(dfboot$y[dfboot$x==1]) - mean(dfboot$y[dfboot$x==0])
}
```

```{r mapping}
# Create fake locations for the people in our population
# by taking random convex combinations of state vertices
set.seed(0)
ga.shape = us_map(include = c("GA"))

# To keep all this from being overwhelming, we'll have 10000 fake locations
# The jth person is at the (j %% M + 1)st location
M = 10000 
V = nrow(ga.shape)
weights = abs(rt(M*V,1))
dim(weights) = c(M,V)
weights = (1/rowSums(weights)) * weights
locations = data.frame(x=weights %*% ga.shape$x, y=weights %*% ga.shape$y, 
                       black=factor(x[1:M], levels=c(0,1), labels=c('non-Black', 'Black')), 
                       black.numeric = x[1:M])

ga = ggplot() + geom_polygon(data=ga.shape, aes(x = x, y = y), fill="white", color="black") + coord_fixed() + blank.theme
```

# Review

## Point and Interval Estimation

```{r}
#| out-height: 400px

meanY.samples = rbinom(1000000, size=n, prob=mean(y)) / n
width = function(Y, alpha=.05) {
  covg.dif = function(w) { 
    actual = pbinom(n*(mean(Y)+w/2), size=n, prob=mean(Y)) - pbinom(n*(mean(Y)-w/2), size=n, prob=mean(Y)) 
    nominal = 1-alpha 
    actual - nominal  
  }
  uniroot(covg.dif, interval=c(0,1))$root
}

annotated.histogram = ggplot() + geom_histogram(aes(x=meanY.samples, y=after_stat(density)), binwidth=.001, alpha=.1) +
           geom_vline(aes(xintercept=mean(y) + c(-1,1)*width(y, alpha=.33)/2), color="blue", linetype="dashed") + 
           geom_vline(aes(xintercept=mean(y) + c(-1,1)*width(y, alpha=.05)/2), color="blue", linetype="dotted") + 
           geom_vline(aes(xintercept=mean(y)), color='green', size=1, alpha=.7) +
           scale_x_continuous(breaks=seq(.6,.8, by=.025), limits=c(.6, .8)) + 
           scale_y_continuous(breaks=seq(0,40,  by=10), limits=c(0, 40)) 

Y2 = y[J2]
Y3 = y[J3]
K = 97
widths = sapply(1:K, function(k) { width(meanY.samples[k], alpha=.05) })

interval.plot = annotated.histogram + geom_pointrange(aes(y=30*(1:K)/K, x=meanY.samples[1:K], 
                                                        xmin=meanY.samples[1:K]-widths/2, 
                                                        xmax=meanY.samples[1:K]+widths/2), color='purple', alpha=.1) +
                      geom_pointrange(aes(y=5, x=mean(Y),   xmin=mean(Y)-width(Y)/2,   xmax=mean(Y)+width(Y)/2)) + 
                      geom_pointrange(aes(y=15, x=mean(Y2), xmin=mean(Y2)-width(Y2)/2, xmax=mean(Y2)+width(Y2)/2), color='magenta') +   
                      geom_pointrange(aes(y=25, x=mean(Y3), xmin=mean(Y3)-width(Y3)/2, xmax=mean(Y3)+width(Y3)/2), color='cyan') 
interval.plot      
```

- In previous lectures, we discussed learning about population proportions and means.
- Our estimates were proportions/means of a random sample (with replacement) from the population. 
- We saw that for large enough sample sizes, these estimates were likely to be accurate.
- We demonstrated that we could predict how accurate we'd be using confidence intervals. 
  - To get these intervals, we add 'arms' of equal length to our estimate.
  - We choose the arm length so that, approximately 95% of the time, the population value is within them.

## Calibration of Interval Estimates
```{r}
#| out-height: 400px
interval.plot
```

- If we had access to the sampling distribution, we could choose so we'd be right exactly 95% of the time.
  - We'd draw arms from the mean of the sampling distribution that span 95% of point estimates.
  - We can see that this is the right length by a shift of perspective.
    - A point estimate can touch the mean with its arms iff^[if and only if] the mean can touch it with equally long arms.
    - So 95% of intervals cover if and only if 95% of point estimates are within the mean's arms.
- We can't do this in practice because we don't know the sampling distribution.
- What we can do is use an *estimate* of the sampling distribution in its place. For example ...
  1. The Bootstrap Sampling Distribution
  2. A 'Plug-in' Normal Approximation using an Estimated Standard Error

## Implementation

```{r}
#| echo: true
width = function(samples, alpha=.05) {
  covg.dif = function(w) { 
    actual = mean( mean(samples) - w/2 <= samples & samples <= mean(samples) + w/2)
    nominal = 1-alpha 
    actual - nominal  
  }
  uniroot(covg.dif, interval=c(0, 2*max(abs(samples-mean(samples)))))$root
}
```

- It's easy to find this width if we have samples from (an approximation to) the sampling distribution.
- We can just search for the arm length that covers the right proportion of the samples.
  - As implemented above, that proportion is $1-\alpha$.
  - The default, $\alpha=.05$, gives us a 95% confidence interval.
- We use **R**'s built-in root-finder **uniroot** to do this search.
  - It finds a point $w$ where a function $f(w)=0$.
  - The function we want to zero is the difference between ...
    - Actual coverage---the proportion of samples within the mean's arms---at the arm length $w/2$.
    - Nominal coverage---the proportion of samples we want within the arms.

## An Approximate Implementation

- If our sampling distribution estimate is symmetric, this is a bit easier.
  - 95% of samples lie between its 2.5th and 97.5th percentiles. That doesn't require symmetry.
  - But with symmetry those percentiles re equidistant from its mean.
  - This distance should be the length of our interval's arms.
- There's an even simpler interpretation of this interval.
- It *is* the interval between the 2.5th and 97.5th percentiles of our sampling distribution estimate.
  - Assuming our point estimate is the mean of our sampling distribution estimate.
  - Because adding those arms to its mean takes us back to those percentiles.
- Absent symmetry, these 3 approaches give us different intervals. Usually ... 
  - The first (last slide) works.
  - The second (percentile width) works less well.
  - The third (percentile range) works even less well.
- But it hasn't really come up in a big way.
   - The sampling distributions we've seeen have been approximately symmetric. They're approximately normal.
   - That's why we've gotten away with being inconsistent about how, exactly, we've constructed our interval.

::: aside
Don't worry about being tested on this. You won't be.
I just thought that, after some time being vague and/or inconsistent 
from Lecture to Lecture to Homework, it'd be good to offer clarification.
:::

# Today

## Summary

```{r}
#| out-height: 300px
#| layout-ncol: 2

sample.locations = locations[J %% M + 1, ]
sample.locations$label = 1:nrow(sample.locations)
# Fix blackness indicator at sampled locations to align with sample
# We can't actually use the location's because multiple people 'live' at a location 
# and the blackness of a location is inherited from an arbitrarily-chosen one of them
sample.locations$black = factor(x[J], levels=c(0,1), labels=c('non-Black', 'Black'))
sample.locations$black.numeric = x[J]

ga + geom_text(aes(x = x, y = y, color=black, label=label), alpha=.5, data=sample.locations) + guides(color=FALSE)
ga + geom_point(aes(x=x, y=y, color=black), alpha=.2, size=.1, data=locations) + guides(color=FALSE)
```

- We'll expand the scope of what we can estimate.
- We'll discuss learning subpopulation proportions and differences between them.
- As before, we'll work with a random sample (with replacement) from the population. 
- We'll revisit turnout in the 2020 Georgia presidential election.
- We'll consider the estimation of two quantities.
  1. The turnout rate of [Black voters]{.twocolor-green}.
  2. The difference in turnout between [Black]{.twocolor-green} and [non-Black voters]{.twocolor-red}.
- This is a little more complicated than what we've done before.
  - The number of [Black]{.twocolor-green} and [non-Black voters]{.twocolor-red} varies from sample to sample.
  - To deal with this, we'll need some new calculational tools. 
  - But we'll be able to defer that stuff to next lecture.

## Context: 2020 Presidential Election Turnout in Georgia

- Recall that ... 
  - Based on a poll of `r n` registered voters a week before the election, we estimated a turnout rate of `r approx(mean(Y))` 
  - After the election, we found that the turnout rate among all registered voters was `r approx(mean(y))`
- Let's suppose we asked for a little more new information. 
- We know from the voter file that `r percent(mean(x==1))` of registered voters are Black.
- In our poll, we found that...
  - `r sum(X==1)` of the people we called (`r percent(mean(X==1))`) were Black with a turnout rate of `r approx(mean(Y[X==1]))`
  - `r sum(X==0)` of the people we called (`r percent(mean(X==0))`) were non-Black with a turnout rate of `r approx(mean(Y[X==0]))`   
- After the election, we found that ...
  - Turnout among Black registered voters was `r approx(mean(y[x==1]))` 
  - Turnout among non-Black registered voters was `r approx(mean(y[x==0]))`.

## Our Sampling Scheme

:::: {.r-stack}
::: {.fragment .fade-out data-fragment-index="0"}

```{r}
#| out-height: 300px
#| layout-ncol: 2
ga + geom_text(aes(x = x, y = y, color=black, label=label), alpha=.5, data=sample.locations[c(1,2,n),]) + guides(color=FALSE)
ga + geom_point(aes(x=x, y=y, color=black), alpha=.2, size=.1, data=locations) + guides(color=FALSE)
```
:::

::: {.fragment .current-visible data-fragment-index="0"}

```{r}
#| out-height: 300px
#| layout-ncol: 2
ga + geom_text(aes(x = x, y = y, color=black, label=label), alpha=.5, data=sample.locations[1:10,]) + guides(color=FALSE)
ga + geom_point(aes(x=x, y=y, color=black), alpha=.2, size=.1, data=locations) + guides(color=FALSE)
```
:::

::: {.fragment .current-visible data-fragment-index="1"}

```{r}
#| out-height: 300px
#| layout-ncol: 2
ga + geom_text(aes(x = x, y = y, color=black, label=label), alpha=.5, data=sample.locations) + guides(color=FALSE)
ga + geom_point(aes(x=x, y=y, color=black), alpha=.2, size=.1, data=locations) + guides(color=FALSE)
```
:::

::::

$$
\small{
\begin{array}{r|rrr|rrr|r|rrr|rr}
\text{call} &  1 & & & 2 & & & \dots & `r n` & & & & & \\
\text{question}  & J_1 & X_1 & Y_1 & J_2 & X_2 & Y_2 & \dots & J_{`r n`} & X_{`r n`} & Y_{`r n`} & \overline{X}_{`r n`} & \overline{Y}_{`r n`} \\
\text{outcome}    & `r J[1]` & \underset{\textcolor{gray}{x_{`r J[1]`}}}{`r X[1]`} & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{x_{`r J[2]`}}}{`r X[2]`} & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`}
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{x_{`r J[n]`}}}{`r X[n]`} & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`}
                  & `r approx(mean(X))` & `r approx(mean(Y))` \\
\end{array}
}
$$


- For each call $i$, we randomly select a voter with an id we'll call $J_i$.
- And we record two things.
    - Our outcome $Y_i=y_{J_i}$, an indicator for turnout: 1 if they actually vote; 0 if they don't.
    - Our covariate $X_i=x_{J_i}$, an indicator for Blackness: 1 if they identify as Black; 0 if they don't.
    - In our plots, we use color to indicate [Black]{.twocolor-green} and [non-Black]{.twocolor-red} voters.
- On each call, each registered voter has the same probability of being called: $1/`r approx(N)`$.
    - As a result, the sample is roughly representative of the population.
    - In particular, the sample is roughly `r percent(mean(x))` Black, just like the population of Georgia's registered voters.


## 4+1 Questions 

1. Our estimate of the Black turnout rate, `r approx(mean(Y[X==1]), accuracy=.001)`, is close to the actual rate of `r approx(mean(y[x==1]), accuracy=.001)`. Did we get lucky?  
2. Could we have predicted how close we would get before the election?  
3. The survey turnout gap of `r approx(mean(Y[X==1])-mean(Y[X==0]))` is close to the population turnout gap of `r approx(mean(y[x==1]) - mean(y[x==0]))`. Did we get lucky? 
4. Could we have predicted how close we would get before the election? 
 
+1: Can we do better using a different sampling scheme?

# Alternative Sampling Schemes

## Reinterpreting Sampling with Replacement

- We'll start by reinterpreting the scheme we've been using.
- Our sampling scheme is equivalent to a two-step process.
    1. We flip a weighted coin to decide whether to call a [Black]{.twocolor-green} or [non-Black voter]{.twocolor-red}.
    2. We choose a voter uniformly at random from the group we've selected.
- Our coin is weighted so we select a Black voter with probability equal to the proportion in our population.
- That makes sampling representative: each registered voter has the same probability of being selected.
  
$$
P(J_i=j) = 
\begin{cases}
   \frac{N_{black}}{N} \ \ \ \ \ \times \frac{1}{N_{black}} \ \ \ \ \ =\ \frac{1}{N} & \text{if the $j$th voter in our population is black ($x_j=1$) } \\
    \frac{N-N_{black}}{N} \times \frac{1}{N-N_{black}} \ = \ \frac{1}{N} & \text{otherwise}
\end{cases}
$$

```{r}
#| out-height: 400px
ga + geom_text(aes(x = x, y = y, color=black, label=label), alpha=.5, data=sample.locations[1:10,]) + guides(color=FALSE)
```

**Check your understanding**. Which of our first 10 coin flips were 'heads'---i.e., selected a black voter?

## Visualizing the Two-Step Process

```{r}
#| out-height: 300px
#| layout-ncol: 2
#| layout-nrow: 2

ga + geom_text(aes(x = x, y = y, color=black, label=label, alpha=black.numeric), data=sample.locations[1:10,]) + 
    scale_alpha(range=.5*c(0,1)) + guides(color=FALSE, alpha=FALSE)
ga + geom_point(aes(x=x, y=y, color=black, alpha=black.numeric), size=.1, data=locations) + 
    scale_alpha(range=.2*c(0,1)) + guides(color=FALSE, alpha=FALSE)

ga + geom_text(aes(x = x, y = y, color=black, label=label, alpha=1-black.numeric), data=sample.locations[1:10,]) +
    scale_alpha(range=.5*c(0,1)) + guides(color=FALSE, alpha=FALSE)
ga + geom_point(aes(x=x, y=y, color=black, alpha=1-black.numeric), size=.1, data=locations) +
    scale_alpha(range=.2*c(0,1)) + guides(color=FALSE, alpha=FALSE)
```

- `r sum(X[1:10])` of the voters we spoke to in our first 10 calls were [Black]{.twocolor-green}.
- `r percent(mean(X))` of the voters we spoke to in our `r n` calls were [Black]{.twocolor-green}.

## Repeating it: a Friend's Poll

```{r}
#| out-height: 300px
#| layout-ncol: 2
#| layout-nrow: 2


sample.locations = locations[J2 %% M + 1, ]
sample.locations$label = 1:nrow(sample.locations)
sample.locations$black = factor(x[J2], levels=c(0,1), labels=c('non-Black', 'Black'))
sample.locations$black.numeric = x[J2]
X2 = x[J2]

ga + geom_text(aes(x = x, y = y, color=black, label=label, alpha=black.numeric), data=sample.locations[1:10,]) + 
    scale_alpha(range=.5*c(0,1)) + guides(color=FALSE, alpha=FALSE)
ga + geom_point(aes(x=x, y=y, color=black, alpha=black.numeric), size=.1, data=locations) + 
    scale_alpha(range=.2*c(0,1)) + guides(color=FALSE, alpha=FALSE)

ga + geom_text(aes(x = x, y = y, color=black, label=label, alpha=1-black.numeric), data=sample.locations[1:10,]) +
    scale_alpha(range=.5*c(0,1)) + guides(color=FALSE, alpha=FALSE)
ga + geom_point(aes(x=x, y=y, color=black, alpha=1-black.numeric), size=.1, data=locations) +
    scale_alpha(range=.2*c(0,1)) + guides(color=FALSE, alpha=FALSE)
```

- `r sum(X2[1:10])` of the voters they spoke to in their first 10 calls were [Black]{.twocolor-green}.
- `r percent(mean(X2))` of the voters they spoke to in their `r n` calls were [Black]{.twocolor-green}.

## Non-Representative Sampling with Replacement


## Nonrepresentative Sampling with Replacement

- We can use a differently-weighted coin to get a nonrepresentative sample.
- e.g., a sample in which approximately half of our call recipients are Black.
- In this sample, we select voters independently with *unequal* probability.

$$
P(J_i=j) = 
\begin{cases}
   \frac{1}{2} \times \frac{1}{N_{Black}} \ \ \ \ \ =\ \frac{1}{2N_{Black}} & \text{if the $j$th voter in our population is black ($x_j=1$) } \\
   \frac{1}{2} \times \frac{1}{N-N_{Black}} \ = \ \frac{1}{2(N-N_{Black})} & \text{otherwise}
\end{cases}
$$

## Sampling Voters --- Our Nonrepresentative Poll

```{r}
#| out-height: 300px
#| layout-ncol: 2
#| layout-nrow: 2

set.seed(our.seed)
Jnr = sample(N, n, replace=TRUE, prob=ifelse(x==1, 1/(2*sum(x)), 1/(2*sum(1-x))))
Xnr = x[Jnr]
sample.locations = locations[Jnr %% M + 1, ]
sample.locations$label = 1:nrow(sample.locations)
sample.locations$black = factor(x[Jnr], levels=c(0,1), labels=c('non-Black', 'Black'))
sample.locations$black.numeric = x[Jnr]

ga + geom_text(aes(x = x, y = y, color=black, label=label, alpha=black.numeric), data=sample.locations[1:10,]) + 
    scale_alpha(range=.5*c(0,1)) + guides(color=FALSE, alpha=FALSE)
ga + geom_point(aes(x=x, y=y, color=black, alpha=black.numeric), size=.1, data=locations) + 
    scale_alpha(range=.2*c(0,1)) + guides(color=FALSE, alpha=FALSE)

ga + geom_text(aes(x = x, y = y, color=black, label=label, alpha=1-black.numeric), data=sample.locations[1:10,]) +
    scale_alpha(range=.5*c(0,1)) + guides(color=FALSE, alpha=FALSE)
ga + geom_point(aes(x=x, y=y, color=black, alpha=1-black.numeric), size=.1, data=locations) +
    scale_alpha(range=.2*c(0,1)) + guides(color=FALSE, alpha=FALSE)
```

- `r sum(Xnr[1:10])` of the voters we spoke to in our first 10 calls were [Black]{.twocolor-green}.
- `r percent(mean(Xnr))` of the voters we spoke to in our `r n` calls were [Black]{.twocolor-green}.

## Sampling Voters --- A Friend's Nonrepresentative Poll

```{r}
#| out-height: 300px
#| layout-ncol: 2
#| layout-nrow: 2

Jnr = sample(N, n, replace=TRUE, prob=ifelse(x==1, 1/(2*sum(x)), 1/(2*sum(1-x))))
Xnr = x[Jnr]
sample.locations = locations[Jnr %% M + 1, ]
sample.locations$label = 1:nrow(sample.locations)
sample.locations$black = factor(x[Jnr], levels=c(0,1), labels=c('non-Black', 'Black'))
sample.locations$black.numeric = x[Jnr]


ga + geom_text(aes(x = x, y = y, color=black, label=label, alpha=black.numeric), data=sample.locations[1:10,]) + 
    scale_alpha(range=.5*c(0,1)) + guides(color=FALSE, alpha=FALSE)
ga + geom_point(aes(x=x, y=y, color=black, alpha=black.numeric), size=.1, data=locations) + 
    scale_alpha(range=.2*c(0,1)) + guides(color=FALSE, alpha=FALSE)

ga + geom_text(aes(x = x, y = y, color=black, label=label, alpha=1-black.numeric), data=sample.locations[1:10,]) +
    scale_alpha(range=.5*c(0,1)) + guides(color=FALSE, alpha=FALSE)
ga + geom_point(aes(x=x, y=y, color=black, alpha=1-black.numeric), size=.1, data=locations) +
    scale_alpha(range=.2*c(0,1)) + guides(color=FALSE, alpha=FALSE)
```

- `r sum(Xnr[1:10])` of the voters they spoke to in their first 10 calls were [Black]{.twocolor-green}.
- `r percent(mean(Xnr))` of the voters they spoke to in their `r n` calls were [Black]{.twocolor-green}.

## Stratified Sampling

- Nonrepresentative sampling with replacement gives us roughly the proportion of Black voters we specify.
- But we can also dispense with the coin flips to get the *exact proportion* we choose.
    - e.g. 50%
    - e.g. `r percent(mean(x))`, the exact proportion in our population.
    - e.g. 100%
- We can simply sample $n_{Black}$ Black voters and $n_{non-Black}$ non-Black voters with replacement from those groups.
    - This gives us two independent samples, one of each group.
    - We can then combine them to get a single sample of size $n=n_{Black}+n_{non-Black}$.
- When we do this, observations are not identically distributed, which makes calculations a bit more complicated.
  - It's not a big challenge. The complexity is more a matter of notation than anything.
  - But to keep things simple, we won't talk more about stratified sampling in this class.


# Estimating the Turnout Rate among Black Registered Voters

## The Sampling Distribution Table and Black Turnout
$$
\small{
\begin{array}{r|rrr|rrr|r|rrr|rrr}
\text{call} &  1 & & & 2 & & & \dots & `r n` & & & & & & \\
\text{pollster} & J_1 & X_1 & Y_1 & J_2 & X_2 & Y_2   & \dots & J_{`r n`} & X_{`r n`} & Y_{`r n`}  & \overline{X} & \overline{Y}  & \overline{XY} \\
\hline
\color{black}   1   
    & \color{black}   `r J=J1; Y=y[J]; X=x[J]; J[1]`    & \color{black}  `r X[1]`      & \color{black} `r Y[1]` 
    & \color{black}   `r J[2]`                          & \color{black}  `r X[2]`      & \color{black} `r Y[2]`      
                                                        & \color{black}   \dots 
    & \color{black}   `r J[n]`                          & \color{black}   `r X[n]`     & \color{black} `r Y[n]`       
    & \color{black}   `r approx(mean(X))`   & \color{black} `r approx(mean(Y))`   &  \color{black}  `r approx(mean(X*Y))` \\
\color{magenta}   2   
    & \color{magenta}   `r J=J2; Y=y[J]; X=x[J]; J[1]`  & \color{magenta}  `r X[1]`      & \color{magenta} `r Y[1]` 
    & \color{magenta}   `r J[2]`                        & \color{magenta}  `r X[2]`      & \color{magenta} `r Y[2]`      
                                                        & \color{magenta}   \dots 
    & \color{magenta}   `r J[n]`                        & \color{magenta}   `r X[n]`     & \color{magenta} `r Y[n]`       
    & \color{magenta}   `r approx(mean(X))` & \color{magenta} `r approx(mean(Y))` &  \color{magenta} `r approx(mean(X*Y))` \\
\color{cyan}    3   
    & \color{cyan}    `r J=J3; Y=y[J]; X=x[J]; J[1]`    & \color{cyan}   `r X[1]`      & \color{cyan}  `r Y[1]` 
    & \color{cyan}    `r J[2]`                          & \color{cyan}   `r X[2]`      & \color{cyan}  `r Y[2]`      
                                                        & \color{cyan}    \dots 
    & \color{cyan}    `r J[n]`                          & \color{cyan}    `r X[n]`     & \color{cyan}  `r Y[n]`       
    & \color{cyan}    `r approx(mean(X))`   & \color{cyan}  `r approx(mean(Y))`   &  \color{cyan}    `r approx(mean(X*Y))` \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{gray}    1M   
    & \color{gray}    `r J=JM; Y=y[J]; X=x[J]; J[1]`    & \color{gray}   `r X[1]`      & \color{gray}  `r Y[1]` 
    & \color{gray}    `r J[2]`                          & \color{gray}   `r X[2]`      & \color{gray}  `r Y[2]`      
                                                        & \color{gray}    \dots 
    & \color{gray}    `r J[n]`                          & \color{gray}    `r X[n]`     & \color{gray}  `r Y[n]`       
    & \color{gray}    `r approx(mean(X))`   & \color{gray}  `r approx(mean(Y))`   &  \color{gray}    `r approx(mean(X*Y))` \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\end{array}
}
$$

:::: r-stack

::: {.fragment .fade-out data-fragment-index="0"}
$$
\text{Let's think about} \ \ \sum_{i=1}^{`r n`}X_i Y_i=n\overline{XY}
$$

- That's the number of polled Black voters who turn out. 
  - It's a sum of `r n` independent and identically distributed Bernoulli random variables with mean $p=`r approx(mean(x*y))`$.
  - That's the proportion of registered voters who both identify as Black and turn out.
  - As a result, it's Binomial: $n\overline{XY} \sim Bin(n,p)$  and $\overline{XY}=Bin(n,p)/n$.
- This proportion is something we could do inference for exactly as we did for the overall turnout rate.
- But it's not the quantity we're interested in. We're interested in the proportion of Black voters who turn out. 
- Using the quantities in our table, how do we calculate that?

:::
::: {.fragment .current-visible data-fragment-index="0"}

$$ 
\text{It's} \ \
\frac{\sum_{i=1}^{`r n`} X_i Y_i}{\sum_{i=1}^{`r n`}} = \frac{n\overline{XY}}{n\overline{X}} = \frac{\overline{XY}}{\overline{X}}.
$$ 

- It's the number of polled Black voters who turn out divided by the number of polled Black voters.
  - That ratio *doesn't* have a Binomial distribution---or a scaled Binomial distribution.
  - The problem is that the denominator is not a constant. It's a random variable.
- To simulate its sampling distribution, we'll repeatedly sample with replacement from the population and compute this quantity $\overline{XY}/\overline{X}$. 
- To estimate its sampling distribution, we'll repeatedly sample with replacement from the sample and do the same. That is, we'll bootstrap.

:::
::::

## Simulating the Sampling Distribution of Black Turnout

```{r}
plot.samples = function(samples) { 
	ggplot() + geom_histogram(aes(x=samples, y=after_stat(density)), binwidth=.001, alpha=.2) +
           geom_vline(aes(xintercept=mean(samples)), color="blue") +
           geom_vline(aes(xintercept=mean(samples)+c(-1,1)*width(samples, alpha=.33)/2), color="blue", linetype='dashed') +
           geom_vline(aes(xintercept=mean(samples)+c(-1,1)*width(samples, alpha=.05)/2), color="blue", linetype='dotted') +
           geom_vline(aes(xintercept=mean(y[x==1])), color='green', size=3, alpha=.7) +
           scale_x_continuous(breaks=seq(.5,.9, by=.025), limits=c(.5, .9)) +
           scale_y_continuous(breaks=seq(0,40,  by=10), limits=c(0, 40))
}
```

```{r}
set.seed(111)
```

```{r}
#| echo: true
turnout.samples = replicate(10000, {
  Js = sample(1:N,size=n,replace=TRUE)
  Xs = x[Js]
  Ys = y[Js]
  mean(Xs*Ys)/mean(Xs)
})
```

```{r}
#| out-height: 400px
plot.samples(turnout.samples)
```

- There are a few features we may want to highlight.
  - The *mean of the sampling distribution* is the solid blue line.
  - The middle 2/3 of the sampling distribution lies between the dashed blue lines.
  - The middle 95\% of the sampling distribution lies between the dotted blue lines. 
- I've also drawn the *subpopulation proportion* as a wide green line.
  - We see that it appears as a kind of highlight on the solid blue line. 
  - It's in exactly the same place. What does this mean? 


## Estimating the Sampling Distribution of Black Turnout

```{r}
#| echo: true
turnout.bootstrap.samples = replicate(10000, {
  Jb = sample(1:n,size=n,replace=TRUE)
  Xb = X[Jb]
  Yb = Y[Jb]
  mean(Xb*Yb)/mean(Xb)
})
```

```{r}
#| out-height: 400px
plot.samples(turnout.bootstrap.samples)
```

- There are a few features we may want to highlight.
  - The *mean of the bootstrap sampling distribution* is the solid blue line.
  - The middle 2/3 of the sampling distribution lies between the dashed blue lines.
  - The middle 95\% of the sampling distribution lies between the dotted blue lines. 
- I've also drawn the *subpopulation proportion* as a wide green line.
  - We see that it's not quite equal to the bootstrap sampling distribution's mean.
  - Why is this?


## A Comparison
```{r}
#| fig-nrow: 2
#| fig-asp: .5
#| out-height: 400px
plot.samples(turnout.samples)
plot.samples(turnout.bootstrap.samples)
```

## A Post-Centering Comparison
```{r}
#| fig-nrow: 2
#| fig-asp: .5
#| out-height: 400px
plot.samples(turnout.samples)
plot.samples(turnout.bootstrap.samples + mean(turnout.samples)-mean(turnout.bootstrap.samples))
```

## A Bootstrap Confidence Interval

```{r}
#| echo: true
w = width(turnout.bootstrap.samples, alpha=.05)
c(mean(X*Y)/mean(X) - w/2, mean(X*Y)/mean(X) + w/2)
```

```{r}
interval.plot = plot.samples(turnout.samples) + geom_pointrange(aes(y=10, x=mean(X*Y)/mean(X), xmin=mean(X*Y)/mean(X)-w/2, xmax=mean(X*Y)/mean(X)+w/2)) 
interval.plot
```

## Coverage

```{r}
#| echo: true
#| cache: true
S = 100
intervals = array(dim=c(S, 2))
for(ss in 1:S) {
  J = sample(1:N,size=n,replace=TRUE)
  Xs = x[J]
  Ys = y[J]
  
  bootstrap.samples = replicate(1000, {
    Jb = sample(1:n,size=n,replace=TRUE)
    Xb = Xs[Jb]
    Yb = Ys[Jb]
    mean(Xb*Yb)/mean(Xb)
  })

  w = width(bootstrap.samples, alpha=.05)
  intervals[ss,] = c(mean(Xs*Ys)/mean(Xs) - w/2, mean(Xs*Ys)/mean(Xs) + w/2)
}
coverage = mean( intervals[,1] <= mean(x*y)/mean(x) & mean(x*y)/mean(x) <= intervals[,2] )
width.mean = mean(intervals[,2]-intervals[,1])
```

```{r}
#| out-height: 400px
K = 99
interval.plot + geom_pointrange(aes(y=20*(1:K)/K, xmin=intervals[1:K,1], xmax=intervals[1:K,2], x=(intervals[1:K,1]+intervals[1:K,2])/2), alpha=.1, color='purple')
```

- `r S*coverage`/`r S`  intervals (`r percent(coverage)`) cover the true proportion of Black voters who turn out.
  - Pretty close to the nominal 95% coverage rate.
- Their mean width is 2 x `r approx(width.mean/2)`.
  - We talk about widths as twice something because our interval has two arms.
  - The width is twice the length of an arm---twice the distance from point-estimate to endpoint.


## Questions 1-2  
```{r}
as.good = mean(abs(turnout.samples - mean(y[x==1])) <=abs(mean(Y[X==1]) - mean(y[x==1])))
width.66 = width(turnout.samples, alpha=1/3)
width.95 = width(turnout.samples, alpha=.05)
```

1. Our estimate of the Black turnout rate, `r approx(mean(X*Y)/mean(X), accuracy=.001)` is very close to the actual rate of `r approx(mean(x*y)/mean(x), accuracy=.001)`. Did we get lucky?  
    - Maybe a litte.  We come this close in only `r percent(as.good)` of polls.
    - But in 2/3 of polls (i.e. draws from the sampling distribution), we're within `r approx(width.66/2)` of the true proportion.
    - And in 95% of polls (i.e. draws from the sampling distribution), we're within `r approx(width.95/2)` of the true proportion.
  
2. Could we have predicted how close we would get before the election? 
    - Luck aside, yes. We can use the bootstrap to construct at 95% confidence interval.
    - This interval has `r percent(coverage)` coverage and average width of 2 x `r approx(width.mean/2)`.

+1: Can we do better using a different sampling scheme?

# +1. Redo with New Sampling Scheme

## How and Why

- Polling non-Black voters isn't useful for estimating Black turnout.
- So it makes sense to reduce the proportion of non-Black voters in our sample.
- We can use non-respresentative sampling with replacement to do this.
- How should we weight our coin?

::: fragment

- We should weight it so it *always* lands heads, i.e., always selects a black voter.
- By doing this, we are in effect sampling with replacement from the subpopulation of Black voters.

:::


## Relative Performance: Accuracy

```{r}
#| echo: true
j.black = which(x==1) 
nonrep.turnout.samples = replicate(10000, {
  Js = sample(j.black,size=n,replace=TRUE)
  Xs = x[Js]
  Ys = y[Js]
  mean(Xs*Ys)/mean(Xs)
})
```

```{r}
#| fig-nrow: 2
#| fig-asp: .5
#| out-height: 300px
#| fig-subcap: 
#|   - "Sampling Distribution with Nonrepresentative Sampling with Replacement"
#|   - "Sampling Distribution with Representative Sampling with Replacement"
plot.samples(nonrep.turnout.samples)
plot.samples(turnout.samples)
```


## Relative Performance: Intervals


```{r}
#| cache: true
S = 100
nonrep.intervals = array(dim=c(S, 2))
for(ss in 1:S) { 
  J = sample(j.black, size=n, replace=TRUE)
  Xs = x[J]
  Ys = y[J]
  
  bootstrap.samples = replicate(1000, {
    Jb = sample(1:n,size=n,replace=TRUE)
    Xb = Xs[Jb]
    Yb = Ys[Jb]
    mean(Xb*Yb)/mean(Xb)
  })

  w = width(bootstrap.samples, alpha=.05)
  nonrep.intervals[ss,] = c(mean(Xs*Ys)/mean(Xs) - w/2, mean(Xs*Ys)/mean(Xs) + w/2)
}
nonrep.coverage = mean( nonrep.intervals[,1] <= mean(x*y)/mean(x) & mean(x*y)/mean(x) <= nonrep.intervals[,2] )
```

```{r}
#| fig-nrow: 2
#| fig-asp: .5
#| out-height: 300px
#| fig-subcap: 
#|   - "Sampling Distribution with Nonrepresentative Sampling with Replacement"
#|   - "Sampling Distribution with Representative Sampling with Replacement"
K = 100
plot.samples(nonrep.turnout.samples) + 
   geom_pointrange(aes(y=40*(1:K)/K, xmin=nonrep.intervals[1:K,1], xmax=nonrep.intervals[1:K,2], x=(nonrep.intervals[1:K,1]+nonrep.intervals[1:K,2])/2), alpha=.1, color='purple')
plot.samples(turnout.samples) + geom_pointrange(aes(y=40*(1:K)/K, xmin=intervals[1:K,1], xmax=intervals[1:K,2], x=(intervals[1:K,1]+intervals[1:K,2])/2), alpha=.1, color='purple') 
```


- When we use this new approach to sampling ...
  - `r percent(nonrep.coverage)` of intervals cover the true proportion.
  - The average width of our intervals is 2 x `r approx(mean(nonrep.intervals[,2]-nonrep.intervals[,1])/2)`.
- This compares favorably to what we got with representative sampling.
  - `r percent(nonrep.coverage)` of intervals covered the true proportion.
  - The average width of our intervals was 2 x `r approx(mean(intervals[,2]-intervals[,1])/2)`.


# Estimating the Difference in Turnout Rates

## The Sampling Distribution Table and the Difference in Turnout

$$
\small{
\begin{array}{r|rrr|rrr|r|rrr|rrr}
\text{call} &  1 & & & 2 & & & \dots & `r n` & & & & & & \\
\text{pollster} & J_1 & X_1 & Y_1 & J_2 & X_2 & Y_2   & \dots & J_{`r n`} & X_{`r n`} & Y_{`r n`}  & \overline{X} & \overline{Y}  & \overline{XY} \\
\hline
\color{black}   1   
    & \color{black}   `r J=J1; Y=y[J]; X=x[J]; J[1]`    & \color{black}  `r X[1]`      & \color{black} `r Y[1]` 
    & \color{black}   `r J[2]`                          & \color{black}  `r X[2]`      & \color{black} `r Y[2]`      
                                                        & \color{black}   \dots 
    & \color{black}   `r J[n]`                          & \color{black}   `r X[n]`     & \color{black} `r Y[n]`       
    & \color{black}   `r approx(mean(X))`   & \color{black} `r approx(mean(Y))`   &  \color{black}  `r approx(mean(X*Y))` \\
\color{magenta}   2   
    & \color{magenta}   `r J=J2; Y=y[J]; X=x[J]; J[1]`  & \color{magenta}  `r X[1]`      & \color{magenta} `r Y[1]` 
    & \color{magenta}   `r J[2]`                        & \color{magenta}  `r X[2]`      & \color{magenta} `r Y[2]`      
                                                        & \color{magenta}   \dots 
    & \color{magenta}   `r J[n]`                        & \color{magenta}   `r X[n]`     & \color{magenta} `r Y[n]`       
    & \color{magenta}   `r approx(mean(X))` & \color{magenta} `r approx(mean(Y))` &  \color{magenta} `r approx(mean(X*Y))` \\
\color{cyan}    3   
    & \color{cyan}    `r J=J3; Y=y[J]; X=x[J]; J[1]`    & \color{cyan}   `r X[1]`      & \color{cyan}  `r Y[1]` 
    & \color{cyan}    `r J[2]`                          & \color{cyan}   `r X[2]`      & \color{cyan}  `r Y[2]`      
                                                        & \color{cyan}    \dots 
    & \color{cyan}    `r J[n]`                          & \color{cyan}    `r X[n]`     & \color{cyan}  `r Y[n]`       
    & \color{cyan}    `r approx(mean(X))`   & \color{cyan}  `r approx(mean(Y))`   &  \color{cyan}    `r approx(mean(X*Y))` \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\color{gray}    1M   
    & \color{gray}    `r J=JM; Y=y[J]; X=x[J]; J[1]`    & \color{gray}   `r X[1]`      & \color{gray}  `r Y[1]` 
    & \color{gray}    `r J[2]`                          & \color{gray}   `r X[2]`      & \color{gray}  `r Y[2]`      
                                                        & \color{gray}    \dots 
    & \color{gray}    `r J[n]`                          & \color{gray}    `r X[n]`     & \color{gray}  `r Y[n]`       
    & \color{gray}    `r approx(mean(X))`   & \color{gray}  `r approx(mean(Y))`   &  \color{gray}    `r approx(mean(X*Y))` \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\end{array}
}
$$

- We know how to calculate Black turnout. 

$$ 
\frac{\sum_{i=1}^{`r n`} X_i Y_i}{\sum_{i=1}^{`r n`} X_i} = 
\frac{n\overline{XY}}{n\overline{X}} = 
\frac{\overline{XY}}{\overline{X}}
$$

- What about non-Black turnout? 

::: fragment

$$
\frac{\sum_{i=1}^{`r n`} (1-X_i) Y_i}{\sum_{i=1}^{`r n`} (1-X_i)} = 
\frac{\sum_{i=1}^{`r n`} Y_i- \sum_{i=1}^{`r n`} X_i Y_i}{\sum_{i=1}^{`r n`} 1 - \sum_{i=1}^{`r n`} X_i} =
\frac{n\overline{Y} - n\overline{XY}}{n - n\overline{X}} = 
\frac{\overline{Y}-\overline{XY}}{1-\overline{X}}
$$

:::

- The difference doesn't have a simple distribution, so we simulate/bootstrap.

## Sampling Dist: Black Turnout Minus non-Black Turnout

$$ 
\begin{aligned}
\text{Black Turnout}     &= \frac{\overline{XY}}{\overline{X}} \\
\text{non-Black Turnout} &= \frac{\overline{Y}-\overline{XY}}{1-\overline{X}} 
\end{aligned}
$$

Here's our code for the sampling distribution of black turnout.

```{r}
#| echo: true
turnout.samples = replicate(10000, {
  Js = sample(1:N,size=n,replace=TRUE)
  Xs = x[Js]
  Ys = y[Js]
  mean(Xs*Ys)/mean(Xs)
})
```

How do we modify it to get the sampling distribution of the difference?

::: fragment

```{r}
#| echo: true
turnout.dif.samples = replicate(10000, {
  Js = sample(1:N,size=n,replace=TRUE)
  Xs = x[Js]
  Ys = y[Js]
  mean(Xs*Ys)/mean(Xs) - mean(Ys*(1-Xs))/(1-mean(Xs))
})
```

```{r}
plot.samples = function(samples) { 
	ggplot() + geom_histogram(aes(x=samples, y=after_stat(density)), binwidth=.001, alpha=.2) +
           geom_vline(aes(xintercept=mean(samples)), color="blue") +
           geom_vline(aes(xintercept=mean(samples)+c(-1,1)*width(samples, alpha=.33)/2), color="blue", linetype='dashed') +
           geom_vline(aes(xintercept=mean(samples)+c(-1,1)*width(samples, alpha=.05)/2), color="blue", linetype='dotted') +
           geom_vline(aes(xintercept=mean(y[x==1])-mean(y[x==0])), color='green', size=3, alpha=.7) +
           scale_x_continuous(breaks=seq(-.2,.2, by=.025), limits=c(-.2, .2)) +
           scale_y_continuous(breaks=seq(0,40,  by=10), limits=c(0, 40))
}
plot.samples(turnout.dif.samples)
```

:::


## Bootstrapping the Turnout Difference

 Here's our code for sampling from the sampling distribution.

```{r}
#| echo: true
turnout.dif.samples = replicate(10000, {
  Js = sample(1:N,size=n,replace=TRUE)
  Xs = x[Js]
  Ys = y[Js]
  mean(Xs*Ys)/mean(Xs) - mean(Ys*(1-Xs))/(1-mean(Xs))
})
```

How do we modify it to give us a sample from the bootstrap sampling distribution?

::: fragment

```{r}
#| echo: true
turnout.dif.bootstrap.samples = replicate(10000, {
  Jb = sample(1:n,size=n,replace=TRUE)
  Xb = X[Jb]
  Yb = Y[Jb]
  mean(Xb*Yb)/mean(Xb) - mean(Yb*(1-Xb))/(1-mean(Xb))
})
```

```{r}
plot.samples(turnout.dif.bootstrap.samples)
```

:::

## A Post-Centering Comparison

```{r}
#| fig-nrow: 2
#| out-height: 400px
plot.samples(turnout.dif.samples)
plot.samples(turnout.dif.bootstrap.samples + mean(turnout.dif.samples)-mean(turnout.dif.bootstrap.samples))
```


## A Bootstrap Confidence Interval

```{r}
#| echo: true
w = width(turnout.dif.bootstrap.samples, alpha=.05)
c(mean(X*Y)/mean(X) - w/2, mean(X*Y)/mean(X) + w/2)
```


```{r}
estimated.dif = mean(X*Y)/mean(X) - mean(Y*(1-X))/(1-mean(X))
interval.plot = plot.samples(turnout.dif.samples) + 
  geom_pointrange(aes(y=10, x=estimated.dif, xmin=estimated.dif-w/2, xmax=estimated.dif + w/2))
interval.plot
```

## Coverage

```{r}
#| echo: true
#| cache: true
S = 100
intervals = array(dim=c(S, 2))
for(ss in 1:S) {
  J = sample(1:N,size=n,replace=TRUE)
  Xs = x[J]
  Ys = y[J]
  
  bootstrap.samples = replicate(1000, {
    Jb = sample(1:n,size=n,replace=TRUE)
    Xb = Xs[Jb]
    Yb = Ys[Jb]
    mean(Xb*Yb)/mean(Xb) - mean(Yb*(1-Xb))/(1-mean(Xb))
  })

  w = width(bootstrap.samples, alpha=.05)
  estimated.dif.s = mean(Xs*Ys)/mean(Xs) - mean(Ys*(1-Xs))/(1-mean(Xs))
  intervals[ss,] = c(estimated.dif.s - w/2, estimated.dif.s + w/2)
}
pop.dif = mean(x*y)/mean(x) - mean((1-x)*y)/mean(1-x)
coverage = mean( intervals[,1] <= pop.dif & pop.dif <= intervals[,2] )
mean.width = mean( intervals[,2] - intervals[,1] )
```

```{r}
#| out-height: 400px
K = 99
interval.plot + geom_pointrange(aes(y=20*(1:K)/K, xmin=intervals[1:K,1], xmax=intervals[1:K,2], x=(intervals[1:K,1]+intervals[1:K,2])/2), alpha=.1, color='purple')
```

- `r S*coverage`/`r S`  intervals (`r percent(coverage)`) cover the true difference in turnout.
  - Pretty close to the intended 95% coverage rate.
- Their mean width is 2 x `r approx(mean.width/2)`.
  - Remember: we use this phrasing because widths are twice the distance from point-estimate to endpoint.


## Questions 3-4  
```{r}
as.good = mean(abs(turnout.dif.bootstrap.samples - pop.dif) <= abs(estimated.dif - pop.dif))
width.66 = width(turnout.samples, alpha=1/3)
width.95 = width(turnout.samples, alpha=.05)
```

3. Our estimate of the turnout rate difference, `r approx(estimated.dif)`, is close to the actual rate of `r approx(pop.dif)`. Did we get lucky?  
    - Maybe a little. We come this close in only `r percent(as.good)` of polls.
    - But in 2/3 of polls (i.e. draws from the sampling distribution), we're within `r approx(width.66/2)` of the population difference.
    - And in 95% of polls (i.e. draws from the sampling distribution), we're within `r approx(width.95/2)` of the population difference.
  
4. Could we have predicted how close we would get before the election? 
    - Luck aside, yes. 
    - Bootstrap confidence intervals have `r percent(coverage)` coverage and average width of 2 x `r approx(width.mean/2)`.


+1: Can we do better using a different sampling scheme? 

# +1: Redo with New Sampling Scheme

## What should it be?

- We'll use non-representative sampling with replacement.
- Your poll will decide the probability that, on each call, we select a Black voter. 
- Which will it be?
    - 30% 
    - 50%
    - 80%
    - 100%

