---
title: "Lecture 4+"
subtitle: "Nonrepresentative Sampling"
format:
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---

```{r setup}
library(usmap)
library(sf)

library(ggplot2)

library(scales)
library(knitr)

percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }
```

```{r ggplot-themes}
class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
		    panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
		    axis.ticks.x = element_blank(),
		    axis.ticks.y = element_blank(),
		    axis.text.x  = element_text(colour = "#aaaaaa"),
		    axis.text.y  = element_text(colour = "#aaaaaa"),
		    axis.title.x  = element_text(colour = "#aaaaaa"),
		    axis.title.y  = element_text(colour = "#aaaaaa", angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
   	         axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
	         panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(simplified.theme)
```

```{r population}
our.seed = 0
set.seed(our.seed)
N=7233584

# TODO: Fix proportions
Nblack = round(0.3*N)
Nvote.black    = round(.7*Nblack)
Nvote.nonblack = round(.6*(N-Nblack))

x= c(rep(1, Nblack), rep(0, N-Nblack))
y= c(rep(1,Nvote.black),rep(0,Nblack-Nvote.black),
     rep(1,Nvote.nonblack),rep(0,N-Nblack-Nvote.nonblack))

# shuffle the order 
o = sample(1:N)
x=x[o]
y=y[o]
```

```{r sample}
set.seed(our.seed)

n = 625
J = sample(N, n, replace=TRUE)
Y = y[J]
X = x[J]
```

##  

```{r mapping}
# Create fake locations for the people in our population
# by taking random convex combinations of state vertices
set.seed(0)
ga.shape = us_map(include = c("GA"))

# To keep all this from being overwhelming, we'll have 10000 fake locations
# The jth person is at the (j %% M + 1)st location
M = 10000 
V = nrow(ga.shape)
weights = abs(rt(M*V,1))
dim(weights) = c(M,V)
weights = (1/rowSums(weights)) * weights
locations = data.frame(x=weights %*% ga.shape$x, y=weights %*% ga.shape$y, black=x[1:M])

ga = ggplot() + geom_polygon(data=ga.shape, aes(x = x, y = y), fill="white", color="black") + coord_fixed() + blank.theme
```

:::: {.r-stack}
::: {.fragment .fade-out data-fragment-index="0"}

```{r}
#| out-height: 300px
#| layout-ncol: 2
sample.locations = locations[J %% M + 1, ]
sample.locations$label = 1:nrow(sample.locations)
ga + geom_text(aes(x = x, y = y, color=factor(black), label=label), alpha=.5, data=sample.locations[c(1,2,n),]) + guides(color=FALSE)
ga + geom_point(aes(x=x, y=y, color=factor(black)), alpha=.2, size=.1, data=locations) + guides(color=FALSE)
```
:::

::: {.fragment .current-visible data-fragment-index="0"}

```{r}
#| out-height: 300px
#| layout-ncol: 2
ga + geom_text(aes(x = x, y = y, color=factor(black), label=label), alpha=.5, data=sample.locations[1:10,]) + guides(color=FALSE)
ga + geom_point(aes(x=x, y=y, color=factor(black)), alpha=.2, size=.1, data=locations) + guides(color=FALSE)
```
:::

::: {.fragment .current-visible data-fragment-index="1"}

```{r}
#| out-height: 300px
#| layout-ncol: 2
ga + geom_text(aes(x = x, y = y, color=factor(black), label=label), alpha=.5, data=sample.locations) + guides(color=FALSE)
ga + geom_point(aes(x=x, y=y, color=factor(black)), alpha=.2, size=.1, data=locations) + guides(color=FALSE)
```
:::

::::

$$
\small{
\begin{array}{r|rrr|rrr|r|rrr|rr}
\text{call} &  1 & & & 2 & & & \dots & `r n` & & & & & \\
\text{question}  & J_1 & X_1 & Y_1 & J_2 & X_2 & Y_2 & \dots & J_{`r n`} & X_{`r n`} & Y_{`r n`} & \overline{X}_{`r n`} & \overline{Y}_{`r n`} \\
\text{outcome}    & `r J[1]` & \underset{\textcolor{gray}{x_{`r J[1]`}}}{`r X[1]`} & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{x_{`r J[2]`}}}{`r X[2]`} & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`}
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{x_{`r J[n]`}}}{`r X[n]`} & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`}
                  & `r approx(mean(X))` & `r approx(mean(Y))` \\
\end{array}
}
$$


- For each call $i$, we randomly select a voter with an id we'll call $J_i$.
- And we record two things.
    - Our outcome,   an indicator for turnout: $Y_i=y_{J_i}$.
    - Our covariate, an indicator for blackness: $X_i=x_{J_i}$
    - In our plots, we use color to indicate [black]{.twocolor-green} and [nonblack]{.twocolor-red} voters.
- On each call, each registered voter has the same probability of being called: $1/`r approx(N)`$.
    - As a result, the sample is roughly representative of the population.
    - In particular, the sample is roughly `r percent(mean(x))` black, just like the population of Georgia's registered voters.


## Our Poll

```{r}
# Now we'll repeat our sampling so we have polls for a few friends
set.seed(our.seed)
J1 = sample(N, n, replace=TRUE)
J2 = sample(N, n, replace=TRUE)
J3 = sample(N, n, replace=TRUE)
JM = sample(N, n, replace=TRUE)
```



```{r}
J = J1
Y = y[J1]
X = x[J1]

sample.locations = locations[J %% M + 1, ]
sample.locations$label = 1:nrow(sample.locations)
ga + geom_text(aes(x = x, y = y, label=label, color=factor(black)), alpha=.5, data=sample.locations[1:10, ]) + guides(color=FALSE)
```

- `r sum(X[1:10])` of the voters we spoke to in our first 10 calls were [black]{.twocolor-green}.
- `r percent(mean(X))` of the voters we spoke to in our `r n` calls were [black]{.twocolor-green}.

$$
\small{
\begin{array}{r|rrr|rrr|r|rrr|rr}
\text{call} &  1 & & & 2 & & & \dots & `r n` & & & & & \\
\text{question}  & J_1 & X_1 & Y_1 & J_2 & X_2 & Y_2 & \dots & J_{`r n`} & X_{`r n`} & Y_{`r n`} & \overline{X}_{`r n`} & \overline{Y}_{`r n`} \\
\text{outcome}    & `r J[1]` & \underset{\textcolor{gray}{x_{`r J[1]`}}}{`r X[1]`} & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{x_{`r J[2]`}}}{`r X[2]`} & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`}
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{x_{`r J[n]`}}}{`r X[n]`} & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`}
                  & `r approx(mean(X))` & `r approx(mean(Y))` \\
\end{array}
}
$$


## Our First Friend's Poll

```{r}
J = J2
Y = y[J2]
X = x[J2]

sample.locations = locations[J %% M + 1, ]
sample.locations$label = 1:nrow(sample.locations)
ga + geom_text(aes(x = x, y = y, label=label, color=factor(black)), alpha=.5, data=sample.locations[1:10, ]) + guides(color=FALSE)
```

- `r sum(X[1:10])` of the voters they spoke to in their first 10 calls were [black]{.twocolor-green}.
- `r percent(mean(X))` of the voters they spoke to in their `r n` calls were [black]{.twocolor-green}.

$$
\small{
\begin{array}{r|rrr|rrr|r|rrr|rr}
\text{call} &  1 & & & 2 & & & \dots & `r n` & & & & & \\
\text{question}  & J_1 & X_1 & Y_1 & J_2 & X_2 & Y_2 & \dots & J_{`r n`} & X_{`r n`} & Y_{`r n`} & \overline{X}_{`r n`} & \overline{Y}_{`r n`} \\
\text{outcome}    & `r J[1]` & \underset{\textcolor{gray}{x_{`r J[1]`}}}{`r X[1]`} & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{x_{`r J[2]`}}}{`r X[2]`} & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`}
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{x_{`r J[n]`}}}{`r X[n]`} & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`}
                  & `r approx(mean(X))` & `r approx(mean(Y))` \\
\end{array}
}
$$

## Our Second Friend's Poll

```{r}
J = J3
Y = y[J3]

sample.locations = locations[J %% M + 1, ]
sample.locations$label = 1:nrow(sample.locations)
ga + geom_text(aes(x = x, y = y, label=label, color=factor(black)), alpha=.5, data=sample.locations[1:10, ]) + guides(color=FALSE)
```

- `r sum(X[1:10])` of the voters they spoke to in their first 10 calls were [black]{.twocolor-green}.
- `r percent(mean(X))` of the voters they spoke to in their `r n` calls were [black]{.twocolor-green}.

$$
\small{
\begin{array}{r|rrr|rrr|r|rrr|rr}
\text{call} &  1 & & & 2 & & & \dots & `r n` & & & & & \\
\text{question}  & J_1 & X_1 & Y_1 & J_2 & X_2 & Y_2 & \dots & J_{`r n`} & X_{`r n`} & Y_{`r n`} & \overline{X}_{`r n`} & \overline{Y}_{`r n`} \\
\text{outcome}    & `r J[1]` & \underset{\textcolor{gray}{x_{`r J[1]`}}}{`r X[1]`} & \underset{\textcolor{gray}{y_{`r J[1]`}}}{`r Y[1]`}
                  & `r J[2]` & \underset{\textcolor{gray}{x_{`r J[2]`}}}{`r X[2]`} & \underset{\textcolor{gray}{y_{`r J[2]`}}}{`r Y[2]`}
                  & \dots 
                  & `r J[n]` & \underset{\textcolor{gray}{x_{`r J[n]`}}}{`r X[n]`} & \underset{\textcolor{gray}{y_{`r J[n]`}}}{`r Y[n]`}
                  & `r approx(mean(X))` & `r approx(mean(Y))` \\
\end{array}
}
$$

## A Reinterpretation


- Our sampling scheme is equivalent to a two-step process.
    1. We flip a weighted coin to decide whether to call a black or nonblack voter.
    2. We choose a voter uniformly at random from the group we've selected.
- Our coin is weighted so we select a black voter with probability equal to the proportion in our population.
- That's what makes sampling representative. 
- With this scheme, each registered voter has the same probability of being selected.

$$
P(J_i=j) = 
\begin{cases}
   \frac{N_{black}}{N} \ \ \ \ \ \times \frac{1}{N_{black}} \ \ \ \ \ =\ \frac{1}{N} & \text{if the $j$th voter in our population is black ($x_j=1$) } \\
    \frac{N-N_{black}}{N} \times \frac{1}{N-N_{black}} \ = \ \frac{1}{N} & \text{otherwise}
\end{cases}
$$

## Sampling Our First Ten Voters


```{r}
#| out-height: 300px
#| layout-ncol: 2
#| layout-nrow: 2
#| 
sample.locations = locations[J %% M + 1, ]
sample.locations$label = 1:nrow(sample.locations)
ga + geom_text(aes(x = x, y = y, color=factor(black), label=label, alpha=black), data=sample.locations[1:10,]) + 
    scale_alpha(range=.5*c(0,1)) + guides(color=FALSE, alpha=FALSE)
ga + geom_point(aes(x=x, y=y, color=factor(black), alpha=black), size=.1, data=locations) + 
    scale_alpha(range=.2*c(0,1)) + guides(color=FALSE, alpha=FALSE)
ga + geom_text(aes(x = x, y = y, color=factor(black), label=label, alpha=1-black), data=sample.locations[1:10,]) +
    scale_alpha(range=.5*c(0,1)) + guides(color=FALSE, alpha=FALSE)
ga + geom_point(aes(x=x, y=y, color=factor(black), alpha=1-black), size=.1, data=locations) +
    scale_alpha(range=.2*c(0,1)) + guides(color=FALSE, alpha=FALSE)

```

Which of our coin flips were 'heads'---i.e., selected a black voter?


## Nonrepresentative Sampling with Replacement

- We can use a differently-weighted coin to get a nonrepresentative sample.
- e.g., a sample in which approximately half of our call recipients are black.
- In this sample, we select voters independently with *unequal* probability.

$$
P(J_i=j) = 
\begin{cases}
   \frac{1}{2} \times \frac{1}{N_{black}} \ \ \ \ \ =\ \frac{1}{2N_{black}} & \text{if the $j$th voter in our population is black ($x_j=1$) } \\
   \frac{1}{2} \times \frac{1}{N-N_{black}} \ = \ \frac{1}{2(N-N_{black})} & \text{otherwise}
\end{cases}
$$

## Sampling Voters --- Our Nonrepresentative Poll

```{r}
#| out-height: 300px
#| layout-ncol: 2
#| layout-nrow: 2

set.seed(our.seed)
Jnr = sample(N, n, replace=TRUE, prob=ifelse(x==1, 1/(2*sum(x)), 1/(2*sum(1-x))))
X = x[Jnr]
sample.locations = locations[Jnr %% M + 1, ]
sample.locations$label = 1:nrow(sample.locations)

ga + geom_text(aes(x = x, y = y, color=factor(black), label=label, alpha=black), data=sample.locations[1:10,]) + 
    scale_alpha(range=.5*c(0,1)) + guides(color=FALSE, alpha=FALSE)
ga + geom_point(aes(x=x, y=y, color=factor(black), alpha=black), size=.1, data=locations) + 
    scale_alpha(range=.2*c(0,1)) + guides(color=FALSE, alpha=FALSE)

ga + geom_text(aes(x = x, y = y, color=factor(black), label=label, alpha=1-black), data=sample.locations[1:10,]) +
    scale_alpha(range=.5*c(0,1)) + guides(color=FALSE, alpha=FALSE)
ga + geom_point(aes(x=x, y=y, color=factor(black), alpha=1-black), size=.1, data=locations) +
    scale_alpha(range=.2*c(0,1)) + guides(color=FALSE, alpha=FALSE)
```

- `r sum(X[1:10])` of the voters we spoke to in our first 10 calls were [black]{.twocolor-green}.
- `r percent(mean(X))` of the voters we spoke to in our `r n` calls were [black]{.twocolor-green}.

## Sampling Voters --- Our First Friend's

```{r}
#| out-height: 300px
#| layout-ncol: 2
#| layout-nrow: 2

Jnr = sample(N, n, replace=TRUE, prob=ifelse(x==1, 1/(2*sum(x)), 1/(2*sum(1-x))))
X = x[Jnr]
sample.locations = locations[Jnr %% M + 1, ]
sample.locations$label = 1:nrow(sample.locations)

ga + geom_text(aes(x = x, y = y, color=factor(black), label=label, alpha=black), data=sample.locations[1:10,]) + 
    scale_alpha(range=.5*c(0,1)) + guides(color=FALSE, alpha=FALSE)
ga + geom_point(aes(x=x, y=y, color=factor(black), alpha=black), size=.1, data=locations) + 
    scale_alpha(range=.2*c(0,1)) + guides(color=FALSE, alpha=FALSE)

ga + geom_text(aes(x = x, y = y, color=factor(black), label=label, alpha=1-black), data=sample.locations[1:10,]) +
    scale_alpha(range=.5*c(0,1)) + guides(color=FALSE, alpha=FALSE)
ga + geom_point(aes(x=x, y=y, color=factor(black), alpha=1-black), size=.1, data=locations) +
    scale_alpha(range=.2*c(0,1)) + guides(color=FALSE, alpha=FALSE)
```

- `r sum(X[1:10])` of the voters they spoke to in their first 10 calls were [black]{.twocolor-green}.
- `r percent(mean(X))` of the voters they spoke to in their `r n` calls were [black]{.twocolor-green}.

## Sampling Voters --- Our Second Friend's

```{r}
#| out-height: 300px
#| layout-ncol: 2
#| layout-nrow: 2

Jnr = sample(N, n, replace=TRUE, prob=ifelse(x==1, 1/(2*sum(x)), 1/(2*sum(1-x))))
X = x[Jnr]
sample.locations = locations[Jnr %% M + 1, ]
sample.locations$label = 1:nrow(sample.locations)

ga + geom_text(aes(x = x, y = y, color=factor(black), label=label, alpha=black), data=sample.locations[1:10,]) + 
    scale_alpha(range=.5*c(0,1)) + guides(color=FALSE, alpha=FALSE)
ga + geom_point(aes(x=x, y=y, color=factor(black), alpha=black), size=.1, data=locations) + 
    scale_alpha(range=.2*c(0,1)) + guides(color=FALSE, alpha=FALSE)

ga + geom_text(aes(x = x, y = y, color=factor(black), label=label, alpha=1-black), data=sample.locations[1:10,]) +
    scale_alpha(range=.5*c(0,1)) + guides(color=FALSE, alpha=FALSE)
ga + geom_point(aes(x=x, y=y, color=factor(black), alpha=1-black), size=.1, data=locations) +
    scale_alpha(range=.2*c(0,1)) + guides(color=FALSE, alpha=FALSE)
```

- `r sum(X[1:10])` of the voters they spoke to in their first 10 calls were [black]{.twocolor-green}.
- `r percent(mean(X))` of the voters they spoke to in their `r n` calls were [black]{.twocolor-green}.

## Stratified Sampling

- Nonrepresentative sampling with replacement gives us roughly the proportion of black voters we specify.
- But we can also dispense with the coin flips to get the *exact proportion* we choose.
    - e.g. 50%
    - e.g. `r percent(mean(x))`, the exact proportion in our population.
- We can simply sample $n_{black}$ black voters and $n_{nonblack}$ nonblack voters with replacement from those groups.
    - This gives us two independent samples, one of each group.
    - We can then combine them to get a single sample of size $n=n_{black}+n_{nonblack}$.
- When we do this, observations are not identically distributed, which makes calculations a bit more complicated.
- For this reason, we won't talk more about stratified sampling in this class.

## Why Use Nonrepresentative Sampling?


##

```{r}
#| cache: true 
S = 100

diffMean    = rep(NA,S)
diffMean.nr = rep(NA,S)
for( s in 1:S ){
    J = sample(N,size=n,replace=TRUE)
    X = x[J]
    Y = y[J]
    diffMean[s] = mean(Y[X==1])-mean(Y[X==0])

    Jnr = sample(N, n, replace=TRUE, prob=ifelse(x==1, 1/(2*sum(x)), 1/(2*sum(1-x))))
    X = x[Jnr]
    Y = y[Jnr]
    diffMean.nr[s] = mean(Y[X==1])-mean(Y[X==0])
}
```

:::: columns
::: column
Representative Sampling 
```{r}
ggplot() + geom_histogram(aes(x=diffMean, y=after_stat(density)))  + 
           geom_vline(aes(xintercept=mean(y[x==1])-mean(y[x==0])), color='green') +
           xlim(-.1,.3) + ylim(0, 20) + coord_flip()
```

:::
::: column

Nonrepresentative Sampling -- 50% Black
```{r}
ggplot() + geom_histogram(aes(x=diffMean.nr, y=after_stat(density)))  + 
           geom_vline(aes(xintercept=mean(y[x==1])-mean(y[x==0])), color='green') +
           xlim(-.1,.3) + ylim(0,20) + coord_flip()
```

:::
::::

- Here are the sampling distributions for the difference in turnout using these two sampling schemes.
- Why does one look better than the other?

  