---
title: "QTM 220: Lecture 19"
subtitle: "Inverse Probability Weighting"
format: 
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: png
    default-image-extension: png
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---


```{r load-libraries}
library(purrr)
library(tibble)
library(dplyr)
library(magrittr)
library(stringr)
library(ggplot2)
library(htmltools)
library(kableExtra)
library(scales)
library(cpsR)
library(maps)

options(knitr.kable.NA = '')

census.api.key = '8084f81e4f71e4d32638dac15f188b424f72d063'
# for cps variable interpretation see https://www2.census.gov/programs-surveys/cps/datasets/2022/march/asec2022_ddl_pub_full.pdf
percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }
```

```{r ggplot-themes}
fontsize = 16
textcolor = "#888888"

class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
        panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x  = element_text(colour = textcolor, size=fontsize),
        axis.text.y  = element_text(colour = textcolor, size=fontsize),
        axis.title.x  = element_text(colour = textcolor),
        axis.title.y  = element_text(colour = textcolor, angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
              axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
           panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(simplified.theme)
```


```{r table-styling}
ellipsis <- function(df, nhead=5L, ntail=2) {
    rownames(df) = as.character(1:nrow(df))
    stopifnot("data.frame" %in% class(df))
    els <- rep(NA, ncol(df)) %>% 
        matrix(nrow=1, dimnames=list(NULL, names(df))) %>% 
        data.frame(stringsAsFactors=FALSE) %>% 
        tibble::as_tibble()
    rownames(els) = "⋮"
    dplyr::bind_rows(head(df, nhead), els, tail(df, ntail))
}
```


```{r population}
set.seed(1) 
m = 100000
pi = function(x) { 1/(1+exp(-sinpi((x-60)/60))) }
mu = function(w, x) { 2 - 2*((x-50)/30)^3/3 + w*((x-50)/30)^2 }  

X = 50+round(30*runif(m))
W = rbinom(m, 1, pi(X))
eps = .5*rnorm(m)
for(x in unique(X)) {
  for(w in c(0,1)) {
    eps[X==x & W==w] = eps[X==x & W==w] - mean(eps[X==x & W==w])
  }
}
Y = mu(W,X) + eps

pop = data.frame(X=X, W=W, Y=Y, Y.counterfactual=Y-mu(W,X)+mu(1-W,X))
```

```{r estimands}
#| cache: true

calculate.summaries = function(muhat, data) { 
  data.frame(cate78 = muhat(1,78) - muhat(0,78),
             att    =  mean(muhat(1, data$X[data$W==1]) - muhat(0, data$X[data$W==1])),
             att1 = mean(muhat(1, data$X[data$W==1])),
             att0 = mean(muhat(0, data$X[data$W==1]))) 
}

mu.formulas = c('Y~W*X', 'Y~W+X', 'Y~W')
pi.formulas = c('known', 'W~X', 'W~1')
formula.grid = expand.grid(mu.formula=mu.formulas, pi.formula=pi.formulas, stringsAsFactors = FALSE)

formula.summaries = function(mu.formula, pi.formula, data) {
    pihat = if (pi.formula=='known') { pi }  else { 
      pi.model = glm(pi.formula, family='binomial', data=data)
      function(x) { predict(pi.model, newdata=data.frame(X=x), type='response') }
    }
    data$weights = data$W + (1-data$W)*pihat(data$X)/(1-pihat(data$X))
    mu.model = lm(formula=mu.formula, weights=weights, data=data)
    muhat = function(w,x) { predict(mu.model, newdata=data.frame(W=w, X=x)) }
    
    summaries = calculate.summaries(muhat, data)
    summaries$mu.formula = mu.formula
    summaries$pi.formula = pi.formula
    summaries
}

estimands = calculate.summaries(mu, pop)
pop.estimates =  purrr::pmap(formula.grid, function(mu.formula, pi.formula) {
    formula.summaries(mu.formula, pi.formula, pop)
}) %>% bind_rows()
```

```{r estimates}
#| cache: true

# This takes ~5 minutes to run

ns = c(100, 200, 500, 1000, 2000, 5000, 10000, 20000, 50000, 100000)  
grid = expand.grid(n=ns, rep=1:200)

estimates = purrr::pmap(grid, function(n, rep) {
    set.seed(rep)
    data = pop[sample(1:nrow(pop), n, replace=TRUE), ]
    summaries = purrr::pmap(formula.grid, function(mu.formula, pi.formula) {
      formula.summaries(mu.formula, pi.formula, data)
    }) %>% bind_rows()
    summaries$n = n
    summaries$rep = rep
    summaries
})  %>% bind_rows()

cate.summaries = estimates %>% group_by(n,  mu.formula, pi.formula) %>% summarize(bias=mean(cate78)-estimands$cate78, se=sd(cate78))
att.summaries  = estimates %>% group_by(n,  mu.formula, pi.formula) %>% summarize(bias=mean(att)-estimands$att, se=sd(att))
att1.summaries = estimates %>% group_by(n,  mu.formula, pi.formula) %>% summarize(bias=mean(att1)-estimands$att1, se=sd(att1))
att0.summaries = estimates %>% group_by(n,  mu.formula, pi.formula) %>% summarize(bias=mean(att0)-estimands$att0, se=sd(att0))
```

# Statistical Inference for Summaries

## An Example
::: {.hidden}
$$
\newcommand{\alert}[1]{\textcolor{red}{#1}}
\newcommand{\model}{\mathcal{M}}
\newcommand{\R}{\mathbb{R}}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator{\V}{V}
\DeclareMathOperator{\Var}{\V}
\newcommand{\inblue}[1]{\textcolor{blue}{#1}}
\newcommand{\inred}[1]{\textcolor[RGB]{248,118,109}{#1}}
\newcommand{\ingreen}[1]{\textcolor[RGB]{0,191,196}{#1}}
\newcommand{\ingray}[1]{\textcolor[RGB]{150,150,150}{#1}}
\newcommand{X}{\mathbf{X}}
\newcommand{Y}{\mathbf{Y}}
$$ 
:::

```{r}
#| out-height: 400px
#| fig-asp: .4 

set.seed(1)

n = 1000
data = pop[sample(nrow(pop), n, replace=TRUE),]

plotX = function(X,W) { X + .15*(2*W-1) }
scaleY = scale_y_continuous(limits=c(0,4), breaks=seq(0,4,by=1))

pop.scatter = ggplot(pop) + 
  geom_point(aes(x=plotX(X,W), y=Y, color=factor(W)), alpha=.035, position=position_jitter(w=.15, h=0), size=.1) + 
  guides(color=FALSE) + scaleY
counterfactual.pop.scatter = ggplot(pop) +
  geom_point(aes(x=plotX(X,!W), y=Y.counterfactual, color=factor(W)), alpha=.035, position=position_jitter(w=.15, h=0), size=.1) + 
  guides(color=FALSE) + scaleY

scatter = ggplot(data) + 
  geom_point(aes(x=plotX(X,W), y=Y, color=factor(W)), alpha=.1, position=position_jitter(w=.15, h=0)) + 
  guides(color=FALSE) + scaleY
pop.scatter
```



- Researchers hypothesize that MHT increases bone density. 
- We'll think about how to look into this. We'll be working with fake data.
- Above, we've got a population of `r format(m,scientific=FALSE)` people eligible for some form of MHT.
  - Their age at the time the study begins is on the $x$-axis.
  - Bone density, measured 2 years later, is on the $y$-axis.
  - Color indicates whether they [did]{.twocolor-green} or [didn't]{.twocolor-red} do MHT.
- What we want to know is how well MHT works, on average, for the people who take it.
- That means comparing green folks---people who take MHT---to versions of themselves who didn't.

::: aside
To make things easier to look at, I'm drawing people who did and didn't receive treatment side-by-side. 
This shift isn't meaningful. Age is a round number: 60, 61, ...
:::

## Counterfactuals

```{r}
#| out-height: 400px
#| layout-ncol: 2 
#| fig-subcap:
#|   - "Real Population: Green People Did MHT, Red People Didn't"
#|   - "Counterfactual Population: Green People Didn't Do MHT, Red People Did" 
pop.scatter
counterfactual.pop.scatter
```

- We can't observe these *counterfactual* versions of people.
- But we can think about what the data would look like if we did.
  - On the left, we've got our real population.
  - On the right, we've got a hypothetical one where everyone who did MHT didn't and vice versa.
  - Note we're keeping the colors the same---counterfactual green people are still green.
- We can think of the average difference as the *average effect of MHT for those who took it*.

$$
\begin{aligned}
\frac{\sum_{j:d_j=1} y_j(1) - y_j(0)}{\sum_{j:d_j=1} 1} = 
\frac{\sum_{j:d_j=1} y_j(1)}{\sum_{j:d_j=1} 1} - \textcolor{blue}{\frac{\sum_{j:d_j=1} y_j(0)}{\sum_{j:d_j=1} 1}}
\end{aligned}
$$

- We call this the *average treatment effect on the treated* (ATT). It's a difference of two means. 
- One is observable (if we have the population). [The other isn't]{.fg style="color:blue"}.

## Summarizing these Populations

```{r}
#| out-height: 400px
#| layout-ncol: 2 
#| fig-asp: .4 
#| fig-subcap:
#|   - "Real Population: Green People Did MHT, Red People Didn't"
#|   - "Counterfactual Population: Green People Didn't Do MHT, Red People Did" 
#| 
pop.summary = pop %>% group_by(X, W) %>% summarize(mean=mean(Y), counterfactual.mean = mean(Y.counterfactual), n=n())
summarized.pop = pop.scatter + 
  geom_point(aes(x=plotX(X,W), y=mean, color=factor(W)), data=pop.summary) + 
  geom_line(aes(x=plotX(X,W), y=mean, color=factor(W)), data=pop.summary) + 
  geom_bar(aes(x=X, y=30*after_stat(prop), fill=factor(W)), 
    position='identity', alpha=.15, data=pop) + 
  guides(fill=FALSE, color=FALSE)

summarized.counterfactual.pop = counterfactual.pop.scatter + 
  geom_point(aes(x=plotX(X,!W), y=counterfactual.mean, color=factor(W)), data=pop.summary) + 
  geom_line(aes(x=plotX(X,!W),  y=counterfactual.mean, color=factor(W)),  data=pop.summary) + 
  geom_bar(aes(x=X, y=30*after_stat(prop), fill=factor(W)), 
    position='identity', alpha=.15, data=pop) + 
  guides(fill=FALSE, color=FALSE)

summarized.pop
summarized.counterfactual.pop
```

- The scatter plot is hard to read. To help, I've overlaid some summaries.
  - Mean bone density for each age and treatment group: connected dots $\ingreen{\mu(1,x)}$ and  $\inred{\mu(0,x)}$.
  - The frequency of each age within each treatment group: histograms $\ingreen{f_1(x)}$
  and $\inred{f_0(x)}$.
  - Note that where the histograms overlap, the colors mix into gray.
- We can think of what we're doing in terms of these summaries. 
- The observable mean is the average of the green curve over the distribution of the green points.
  - i.e., the green histogram.

$$
\begin{aligned}
\text{observable mean} 
= \frac{\sum_{j:d_j=1} y_j}{\sum_{j:d_j=1} 1} 
= \sum_x \frac{\sum_{j:d_j=1, x_j=x} 1}{\sum_{j:d_j=1} 1} \frac{\sum_{j:d_j=1, x_j=x} y_j}{\sum_{j:d_j=1, x_j=x} 1} 
= \sum_x \ingreen{f_1(x) \ \mu(1,x)}
\end{aligned}
$$


- The counterfactual mean is the same thing, but in the counterfactual plot.

## Working With Summaries

```{r}
#| out-height: 400px
#| layout-ncol: 2 
#| fig-asp: .4 
#| fig-subcap:
#|   - "Real Population: Green People Did MHT, Red People Didn't"
#|   - "Counterfactual Population: Green People Didn't Do MHT, Red People Did" 
summarized.pop
summarized.counterfactual.pop
```


- This seems like it'd be a problem, since we don't actually see the counterfactual plot. 
- But, comparing our plots, we see something helpful: the green curve on the left looks like the red curve on the left.
- In other words, we can use *what actually happened to the red people* to estimate *what would have happened to the green people*.

$$
\begin{aligned}
\text{observable mean} &= \sum_x \ingreen{f_1(x)}\ingreen{\mu(1,x)} \\
\text{counterfactual mean} &= \sum_x \ingreen{f_1(x)}\inred{\mu(0,x)}
\end{aligned}
$$

- In real data, there's no counterfactual plot, so we can't do this comparison visually.
- But *we've proven* that these curves look the same *in conditionally randomized experiments*.

## Working With Summaries
```{r}
#| out-height: 400px
#| layout-ncol: 2 
#| fig-asp: .4 
#| fig-subcap:
#|   - "Real Population: Green People Did MHT, Red People Didn't"
#|   - "Counterfactual Population: Green People Didn't Do MHT, Red People Did" 
summarized.pop
summarized.counterfactual.pop
```

### In short

- To calculate the ATT in a conditionally randomized experiment, we ...
  - average the green curve over the distribution of the green points
  - average the red curve over the distribution of the green distribution
  - subtract 

$$
ATT = \sum_x \ingreen{f_{1}(x)}\ingreen{\mu(1,x)} - \sum_x \ingreen{f_{1}(x)}\inred{\mu(0,x)}
$$

- That's what we'd do if we had the population.
- But we have a sample from it, so we'll do the same with *estimates*.

## Working With A Sample

```{r}
#| out-height: 400px
#| layout-ncol: 2 
#| fig-asp: .4 
#| fig-subcap:
#|   - "A Sample from the Population" 
#|   - "A Population: Green People Did MHT, Red People Didn't"
summary = data %>% group_by(X, W) %>% summarize(mean=mean(Y), sd=sd(Y), n=n())
summarized.scatter = scatter + 
  geom_pointrange(aes(x=plotX(X,W), y=mean, ymin=mean-2*sd/sqrt(n), ymax=mean+2*sd/sqrt(n), 
  color=factor(W)), data=summary, alpha=.5) + 
  geom_bar(aes(x=X, y=30*after_stat(prop), fill=factor(W)), 
    position='identity', alpha=.15, data=data) + 
  guides(fill=FALSE, color=FALSE) 
summarized.scatter
summarized.pop
```

- This is what we actually have. It's a sample of `r n` people. 
- To give you a sense of the uncertainty, I've drawn 95% confidence intervals for each subpopulation mean
around the subsample means.

$$
\widehat{ATT} = \sum_x \ingreen{\hat f_1(x)}\ingreen{\hat\mu(1,x)} - \sum_x \ingreen{\hat f_1(x)}\inred{\hat\mu(0,x)}
$$

- To reduce variance, we often use other estimates of the subpopulation means $\inred{\mu(0,x)}$.
- These estimates are often biased and, as a result, our ATT estimator usually will be too.

## Working With A Sample

```{r}
sample.model = lm(Y ~ W*X, data=data)
pop.model = lm(Y ~ W*X, data=pop)
grid = expand.grid(X=seq(50,80,by=.1), W=c(0,1))
grid$muhat = predict(sample.model, newdata=grid)
grid$mutilde = predict(pop.model, newdata=grid)
grid$mu = mu(grid$W, grid$X)

summarized.scatter.annotated.nopop = summarized.scatter + 
  geom_line(aes(x=X, y=muhat, color=factor(W)), data=grid, linetype='dashed') + 
  geom_line(aes(x=X, y=mu, color=factor(W)), data=grid, linetype='solid', alpha=.7) 
summarized.scatter.annotated = summarized.scatter.annotated.nopop +
  geom_line(aes(x=X, y=mutilde, group=factor(W)), data=grid, linetype='solid', color='black', alpha=.4)
```

:::: r-stack
::: {.fragment .fade-out data-fragment-index="0"}

```{r}
#| out-height: 400px
#| fig-asp: .4  
summarized.scatter.annotated.nopop
```

:::
::: {.fragment .current-visible data-fragment-index="0"}
```{r}
#| out-height: 400px
#| fig-asp: .4  
summarized.scatter.annotated
```

:::
::::

- Here we've fit separately to each group, i.e., the interacted bivariate model.
    - The solid curves are, as before, the subpopulation means
    - The dashed lines are what we'd use to estimate them: the model fit on the sample.
- In large samples, these estimates approach a limit: the model fit on the population. 
  - We call this the $\ingray{\text{population regression } \tilde\mu(d,x)}$. 
  - We'll plot it in gray. You'll have to infer $d$---the treatment status---from context. 

$$ 
\begin{aligned}
\hat \mu &= \argmin_{m \in \model} \frac{1}{n}\sum_{i=1}^n \qty{ m(D_i, X_i) - Y_i }^2 \\
&\approx \\
\tilde \mu &= \argmin_{m \in \model} \E \qty{ m(D_i, X_i) - Y_i }^2 
\end{aligned}
$$

## Types of Error

```{r}
#| out-height: 400px
#| fig-asp: .4  

summarized.scatter.annotated
```

$$
\hat\mu(d,x) - \mu(d,x) = \underset{\text{estimation error}}{\qty{ \hat\mu(d,x) - \tilde\mu(d,x) }} + 
  \underset{\text{modeling error}}{\qty{ \tilde\mu(d,x) - \mu(d,x) }}
$$

- It's helpful to decompose error into two parts: one that depends on the sample and one that doesn't.
- The first part is the difference between our predictions and those we'd get in the large-sample limit.
    - It's the difference between the colored and gray dashed lines. 
    - This has mean zero and we have techniques to understand it---confidence intervals.
- The second part is the difference between the predictions we'd get in the large-sample limit and the ones we want.
    - It's the difference between the gray dashed lines and the solid colored ones. 
    - This is the bias of our prediction. It depends on the population and the model, but not the sample.
    - This is what we're going to focus on today.
- All this passes through to summaries like the CATE and ATT


## Bias for the CATE

```{r}
sample.dist.plot = function(estimate.samples, pop.estimate, estimand) { 
  data = data.frame(estimates=estimate.samples)
  p = ggplot(data) + geom_histogram(aes(x=estimates, y=after_stat(density)), fill='black', bins=30, alpha=.2) + 
           geom_vline(aes(xintercept=mean(estimates)), color='blue') +
           geom_vline(aes(xintercept=mean(estimates) - 2*sd(estimates)), color='blue', linetype='dotted') +
           geom_vline(aes(xintercept=mean(estimates) + 2*sd(estimates)), color='blue', linetype='dotted') +
            geom_vline(aes(xintercept=mean(estimates) - sd(estimates)), color='blue', linetype='dashed') +
           geom_vline(aes(xintercept=mean(estimates)  + sd(estimates)), color='blue', linetype='dashed') +
           geom_vline(aes(xintercept=estimand), color='black', alpha=.5, linewidth=2)
  if(!is.na(pop.estimate)) {
    p = p + geom_vline(aes(xintercept=pop.estimate), color='red', alpha=.5, linewidth=2)
  }
  p
}
plot.sample.dist = function(outcome, mu.formula, pi.formula, n) {
  estimate.samples = estimates[estimates$mu.formula == mu.formula & 
                               estimates$pi.formula == pi.formula & 
                               estimates$n == n, outcome]
  pop.estimate = pop.estimates[pop.estimates$mu.formula == mu.formula & 
                               pop.estimates$pi.formula == pi.formula, outcome]
  sample.dist.plot(estimate.samples, pop.estimate, estimands[[outcome]])
}
```

```{r}
#| out-height: 400px
#| fig-asp: .4 
summarized.scatter.annotated
```

- We can see that our CATE estimate should be fairly biased for older folks (large $x$).
- That's because our fitted line $\inred{\hat\mu(0,x)}$ gives a poor estimate of our subpopulation means there.

$$
\begin{aligned}
\widehat{CATE}(x) - CATE(x) = & \ingreen{\qty{\hat\mu(1, x) - \ingray{\tilde\mu(1, x)}}} - \inred{\qty{\hat\mu(0,x) - \ingray{\tilde\mu(0,x)}}} && \text{estimation error} \\
                            +& \ingreen{\qty{\ingray{\tilde\mu(1, x)} - \mu(1, x)}} - \inred{\qty{\ingray{\tilde\mu(0,x)} - \mu(0,x)}}  && \text{modeling error} \\ 
\end{aligned}
$$

## Implications for Inference

```{r}
#| out-height: 400px
#| fig-asp: .8 
#| layout-ncol: 3 

mu.formula = 'Y~W*X'
pi.formula = 'W~1'
outcome = 'cate78'

scale_x = coord_cartesian(xlim = c(.5, 1)) 
plot.sample.dist(outcome, mu.formula, pi.formula, n=500) + scale_x
plot.sample.dist(outcome, mu.formula, pi.formula, n=1000) + scale_x
plot.sample.dist(outcome, mu.formula, pi.formula, n=2000) + scale_x
```

- Above, we're looking at sampling distributions for our CATE estimate at age $x=78$. 
  - From left to right, we have sample sizes $n=500,1000,2000$
- We have vertical lines indicating ...
  - its mean (in blue) +- 1 and 2 standard errors (dashed and dotted respectively)
  - an estimate based on the population regression (in red) 
  - the CATE itself (in black)
- At $n=500$, our bias is about 1.5 standard errors. At $n=2000$, it's twice that.
- Why twice? And what are the inferential implications?


## Bias for the ATT

```{r}
#| out-height: 400px
#| fig-asp: .4 
summarized.scatter.annotated
```

- Our ATT estimate should also be fairly biased because it's an average of the CATE for older folks.

$$
\begin{aligned}
\widehat{ATT} - ATT &=   \sum_x \ingreen{f_1(x)} \ingreen{\qty{\hat\mu(1, x) - \ingray{\tilde\mu(1, x)}}} - \sum_x \ingreen{f_1(x)} \inred{\qty{\hat\mu(0,x) - \ingray{\tilde\mu(0,x)}}} &&\text{estimation error}  \\
                    &+ \sum_x \ingreen{f_1(x)} \ingreen{\qty{\ingray{\tilde\mu(1, x)} - \mu(1, x)}} - \sum_x \ingreen{f_1(x)} \inred{\qty{\ingray{\tilde\mu(0,x)} - \mu(0,x)}} &&\text{modeling error}
\end{aligned}
$$

## Implications for Inference

```{r}
#| out-height: 400px
#| fig-asp: .8 
#| layout-ncol: 3 

mu.formula = 'Y~W*X'
pi.formula = 'W~1'
outcome = 'att'


scale_x = coord_cartesian(xlim=c(.2, .5)) 
plot.sample.dist(outcome, mu.formula, pi.formula, n=500) + scale_x
plot.sample.dist(outcome, mu.formula, pi.formula, n=1000) + scale_x
plot.sample.dist(outcome, mu.formula, pi.formula, n=2000) + scale_x
```

- Above, we're looking at sampling distributions for our ATT estimate.
  - On the left, we have sample size $n=500$.
  - On the right, we have sample size $n=2000$.
- We have vertical lines indicating ...
  - its mean (in blue) +- 2 standard errors (dotted)
  - an estimate based on the population regression (in red) 
  - the ATT itself (in black)
- At $n=500$, bias is very small relative to standard error. At $n=2000$, it's still not bad. 

## Things Are Worse If You Use The Additive Model
```{r}
#| out-height: 400px
#| fig-asp: .8 
#| layout-ncol: 3 
#| layout-nrow: 2 

mu.formula = 'Y~W*X'
pi.formula = 'W~1'
outcome = 'att'

scale_x = coord_cartesian(xlim=c(.2, .5)) 
plot.sample.dist(outcome, mu.formula, pi.formula, n=500) + scale_x
plot.sample.dist(outcome, mu.formula, pi.formula, n=1000) + scale_x
plot.sample.dist(outcome, mu.formula, pi.formula, n=2000) + scale_x

mu.formula = 'Y~W+X'
plot.sample.dist(outcome, mu.formula, pi.formula, n=500) + scale_x
plot.sample.dist(outcome, mu.formula, pi.formula, n=1000) + scale_x
plot.sample.dist(outcome, mu.formula, pi.formula, n=2000) + scale_x
```

Sampling distributions for ATT estimates based on the interacted (top) and additive (bottom) bivariate models.

## It Doesn't Fit Very Well
```{r}
sample.model.additive = lm(Y ~ W+X, data=data)
pop.model.additive = lm(Y~W+X, data=pop)
grid$muhat.additive = predict(sample.model.additive, newdata=grid)
grid$mutilde.additive = predict(pop.model.additive, newdata=grid)

summarized.scatter.annotated.nopop  +
  geom_line(aes(x=X, y=muhat.additive, color=factor(W)), data=grid, linetype='dotted') 
```

- Interacted (dashed) and additive (dotted) bivariate models fit to the sample. 
- Solid lines are the population means.


## Conclusion

- When you do least squares regression ...
  - your predictions $\hat\mu(d,x)$ will be *biased estimates* of the corresponding subpopulation means $\mu(d,x)$.
  - unless you use a saturated model, i.e., unless your predictions are the correspondig subsample means.
- And if you use them to estimate the ATT---or any other summary---that estimate will be biased too.
- That means that, in large enough samples, you'll be claiming coverage much better than what you actually have.
- In the next section, we'll talk about how to fix this. 
  - To prepare, we'll need to review some ideas from probability.
  - Before we do that, let's review our notation and pictures.

## Review Exercise 1

```{r}
#| out-height: 600px
 
sample.model.additive = lm(Y ~ W+X, data=data)
pop.model.additive = lm(Y ~ W+X, data=pop)
grid$muhat.additive = predict(sample.model.additive, newdata=grid)
grid$mutilde.additive = predict(pop.model.additive, newdata=grid)
summarized.scatter + 
  geom_line(aes(x=X, y=muhat.additive, color=factor(W)), linetype='dashed', data=grid) +
  geom_line(aes(x=X, y=mutilde.additive, group=factor(W)), data=grid, linetype='dashed', color='black', alpha=.7) 
```

- In the plot above, we're working with the additive bivariate model.
  - Label $\hat\mu(d, x)$, $\tilde\mu(d, x)$, $\hat\mu(d, x)$ for $d=0$ and $d=1$
  - Illustrate $CATE(x)$, $\widehat{CATE}(x)$, and the bias of $\widehat{CATE}(x)$.
- Describe all of those things in words.
- Explain, in terms of the the plot, how to calculate $\widehat{ATT}$.

## Review Exercise 2

```{r}
#| out-height: 600px

plot.sample.dist(outcome, mu.formula, pi.formula, n=500) + scale_x 
```

- This is a plot of the sampling distribution of $\widehat{ATT}$.
  - Label $\widehat{ATT}$, $ATT$, and the bias of $\widehat{ATT}$.
  - Illustrate a calculation of the coverage of the interval $\widehat{ATT} \pm 2\text{se}$.
  - Based on your illustration, give a rough approximation to this coverage.

# Probability Review

## Orthogonality of Population Regression Errors

- The population residuals, $\tilde \varepsilon = Y - \tilde \mu(X)$, are orthogonal to the features 
$\phi_k(X)$ used in the regression.
- As a result, they're orthogonal to any curve in the model.
$$
\begin{aligned}
0&=\sum_j \tilde \varepsilon_j \phi_k(x_j =\E\qty[ \tilde \varepsilon_i \phi_k(X_i) ] && \qfor k=0,1,2,\ldots \\
\implies 
0&=\sum_j \tilde \varepsilon_j m(x_j) =\E\qty[ \tilde \varepsilon_i m(X_i) ] && \qqtext{ for any } m \in \model 
\end{aligned}
$$

- This implies the *population regression errors*, $\mu(X)-\tilde\mu(X)$ are orthogonal, too.
$$
\begin{aligned}
0 &= \E\qty[ \tilde \varepsilon_i m(X_i) ] && \text{from above} \\
  &= \E\qty[ \qty{Y_i - \tilde \mu(X_i)} m(X_i) ] && \text{definitionally} \\
  &= \E\qty[ \E(\qty{Y_i - \tilde \mu(X_i)} m(X_i) \mid X_i) ] && \text{law of iterated expectations} \\
  &= \E\qty[ \qty{\E\qty(Y_i \mid X_i) - \tilde \mu(X_i)} m(X_i) ] && \text{linearity of conditional expectations} \\
  &= \E\qty[ \qty{\mu(X_i) - \tilde \mu(X_i)} m(X_i) ] && \text{definitionally} 
\end{aligned}
$$

- This is when we think of $X$ as everything we condition on, including the treatment indicator $D$. 
- Usually, we'd write it this way.

$$
\begin{aligned}
0 &= \E\qty[ \qty{\mu(D_i, X_i) - \tilde \mu(D_i, X_i)} m(D_i, X_i) ] \qqtext{ for any curve $m \in \model$ } \\
 &= \E\qty[ g(D,X) ] \qfor g(D,X) = \qty{\mu(D, X) - \tilde \mu(D, X)} m(D, X)
\end{aligned}
$$

- To think about what this is, let's think of it as an average of an arbitrary function $g(D,X)$.

## Calculating Expectations With Binary Treatment
```{r}
#| out-height: 400px
#| fig-asp: .4  
pop.model = lm(Y ~ W*X, data=pop) 
mutilde = predict(pop.model, newdata=grid)
summarized.pop.annotated = summarized.pop + 
  geom_line(aes(x=X, y=mutilde, group=factor(W)), 
  data=grid, color='black', alpha=.4)
summarized.pop.annotated
```


- When $D$ is binary, there's a convenient way to think about expectations like this.
  (a) We use the law of iterated expectations, conditioning on $D$.
  (b) This makes it an average of a random variable taking on two values.
  (c) And these two values are the averages of $g(D,X)$ over the populations with $D=0$ and $D=1$.

$$
\begin{aligned}
\E\qty[ g(D,X) ] 
&\overset{(a)}{=} \E\qty[ \E\qty{ g(D,X) \mid D } ] \\
&\overset{(b)}{=} P(D=0) \E\qty[ g(D,X) \mid D=0] \quad + \quad P(D=1)\E\qty[ g(D,X) \mid D=1]  \\
&\overset{(c)}{=} P(D=0) \sum_x \inred{f_0(x)} g(0,x)  \quad + \quad P(D=1) \sum_x \ingreen{f_1(x)} g(1,x) \\
\end{aligned}
$$

# Bias and Inverse Probability Weighting

## Understanding Bias 

```{r}
#| out-height: 400px
#| fig-asp: .4  
summarized.pop.annotated
```

$$
\begin{aligned}
\text{modeling error} &= \sum_x \ingreen{f_1(x)} \qty{\ingreen{\tilde\mu(1, x) - \mu(1, x)}} - \sum_x \ingreen{f_1(x)} \qty{\inred{\tilde\mu(0,x) - \mu(0,x)}} \\
\qfor \tilde \mu &= \argmin_{m \in \model} \E \qty{ m(D_i, X_i) - Y_i }^2 
\end{aligned}
$$

- Let's break this bias down as a sum of two terms.
 - One associated with our estimate of the [mean outcome with treatment]{.twocolor-green} 
 - One associated with our estimate of the [mean outcome without]{.twocolor-red}
- The first looks like less of a problem. The curve $\ingreen{\mu(1,x)}$ looks like a line and we've fit a line.
- The second curve $\inred{\mu(0,x)}$ we want to estimate is curved---it's not a line.
- And the line we fit to it is a decent approximation *on the left side of the plot*.
    - That's what least squares is telling it to do. It's where most of our red points are.
    - But it's not what we want. We're averaging it over the distribution of the green points, which are mostly on the right.

## The First Term Is Not A Problem

```{r}
#| out-height: 400px
#| fig-asp: .4  

summarized.scatter.annotated
```

$$
\text{modeling error} = \sum_x \ingreen{f_1(x)} \ingreen{\qty{\ingray{\tilde\mu(1, x)} - \mu(1, x)}} - \textcolor{lightgray}{\sum_x f_1(x) \qty{\tilde\mu(0,x) - \mu(0,x)}}
$$

:::: r-stack
::: {.fragment .fade-out data-fragment-index="0"}

- And it's not just because we're fitting a line to data with a linear trend.
- In fact, this bias term will zero *for any model with group-specific intercepts*.
- We can prove it by thinking about the population regression errors $\mu(D,X) - \tilde\mu(D,X)$.
    - We'll use the property that they're orthogonal to every curve in the model.
    - And the property that an indicator for the group---the curve $m(d,x)=d$---is in the model.

:::
::: {.fragment .current-visible data-fragment-index="0"}

$$
\begin{aligned}
0 &= \E\qty[ D \ \qty{\mu(D,X)-\tilde\mu(D, X)} ]    
&& \text{orthogonality of errors} \\
&= P(D=0) \E\qty[D \ \qty{\mu(D,X)-\tilde\mu(D, X)} \mid D=0] 
&& \\
&+ P(D=1) \E\qty[D \ \qty{\mu(D,X)-\tilde\mu(D, X)} \mid D=1] 
&& \text{iterated expectations} \\
&= P(D=0) \sum_x  \inred{f_0(x)}   \cdot  0 \cdot \textcolor{lightgray}{\qty{\mu(0,x) - \tilde\mu(0,x)} }  
&& \\
&+ P(D=1) \sum_x \ingreen{f_1(x)}   \cdot 1 \cdot \qty{\ingreen{\mu(1,x) - \ingray{\tilde\mu(1,x)}}}  
&& \text{calculating expectations} \\
\end{aligned}
$$

:::
::::

## Sampling Distributions: Interacted Bilinear Model
```{r}
#| cache: true
formulas = list(interacted=Y~X*W, additive=Y~X+W, intercepts=Y~W)
term.grid = expand.grid(n=c(500, 1000, 2000), rep=1:500, formula=names(formulas))

term.estimates = rep(list(), nrow(term.grid))
for(ii in 1:nrow(term.grid)) { 
    n = term.grid$n[ii]
    rep = term.grid$rep[ii]
    formula = term.grid$formula[ii]

    set.seed(rep)
    data = pop[sample(1:nrow(pop), n, replace=TRUE), ]
    data$weights = data$W + (1-data$W)*pi(data$X)/(1-pi(data$X))

    model = lm(formula=formulas[[formula]], data=data) 
    att1 = mean(predict(model, newdata=data.frame(X=data$X[data$W==1], W=1)))
    att0 = mean(predict(model, newdata=data.frame(X=data$X[data$W==1], W=0)))

    wmodel = lm(formula=formulas[[formula]], data=data, weights = weights) 
    watt1 = mean(predict(wmodel, newdata=data.frame(X=data$X[data$W==1], W=1)))
    watt0 = mean(predict(wmodel, newdata=data.frame(X=data$X[data$W==1], W=0)))
    term.estimates[[ii]] = data.frame(n=n, rep=rep, formula=formula, att1=att1, att0=att0, watt1=watt1, watt0=watt0)
} 
term.estimates = term.estimates %>% bind_rows() 
```

```{r}
#| out-height: 400px
#| fig-asp: .4 
#| layout-nrow: 2 
#| layout-ncol: 2
sample.model = lm(Y ~ W*X, data=data)
grid$muhat = predict(sample.model, newdata=grid)

summarized.scatter + 
  geom_line(aes(x=X, y=muhat, color=factor(W)), data=grid, linetype='dashed') +
  geom_line(aes(x=X, y=mu, color=factor(W)), data=grid, linetype='solid', alpha=.7)  

scale_x = coord_cartesian(xlim=c(2,2.3))
sample.dist.plot(subset(term.estimates, n==500 & formula=='interacted')$att1,
                 NA, estimands$att1)  + scale_x
sample.dist.plot(subset(term.estimates, n==1000 & formula=='interacted')$att1, 
                 NA, estimands$att1)  + scale_x
sample.dist.plot(subset(term.estimates, n==2000 & formula=='interacted')$att1, 
                 NA, estimands$att1)  + scale_x 
```

- Here we're seeing the interacted model fit at sample size $n=`r n`$
- And sampling distributions at sample sizes 500, 1000, and 2000


## Sampling Distributions: Additive Bilinear Model

```{r}
#| out-height: 400px
#| fig-asp: .4 
#| layout-nrow: 2 
#| layout-ncol: 2
sample.model = lm(Y ~ W+X, data=data)
grid$muhat = predict(sample.model, newdata=grid)

summarized.scatter + 
  geom_line(aes(x=X, y=muhat, color=factor(W)), data=grid, linetype='dashed') +
  geom_line(aes(x=X, y=mu, color=factor(W)), data=grid, linetype='solid', alpha=.7)  

scale_x = coord_cartesian(xlim=c(2,2.3))
sample.dist.plot(subset(term.estimates, n==500 & formula=='additive')$att1,
                 NA, estimands$att1)  + scale_x
sample.dist.plot(subset(term.estimates, n==1000 & formula=='additive')$att1, 
                 NA, estimands$att1)  + scale_x
sample.dist.plot(subset(term.estimates, n==2000 & formula=='additive')$att1, 
                 NA, estimands$att1)  + scale_x 
```

## Sampling Distributions: Two-Intercepts Model

```{r}
#| out-height: 400px
#| fig-asp: .4 
#| layout-nrow: 2 
#| layout-ncol: 2
sample.model = lm(Y ~ W, data=data)
grid$muhat = predict(sample.model, newdata=grid)

summarized.scatter + 
  geom_line(aes(x=X, y=muhat, color=factor(W)), data=grid, linetype='dashed') +
  geom_line(aes(x=X, y=mu, color=factor(W)), data=grid, linetype='solid', alpha=.7)  

scale_x = coord_cartesian(xlim=c(2,2.3))
sample.dist.plot(subset(term.estimates, n==500 & formula=='intercepts')$att1,
                 NA, estimands$att1)  + scale_x
sample.dist.plot(subset(term.estimates, n==1000 & formula=='intercepts')$att1, 
                 NA, estimands$att1)  + scale_x
sample.dist.plot(subset(term.estimates, n==2000 & formula=='intercepts')$att1, 
                 NA, estimands$att1)  + scale_x 
```


## The Second Term 

```{r}
#| out-height: 400px
#| fig-asp: .4  

summarized.pop + geom_line(aes(x=X, y=mutilde, color=factor(W)), data=grid, linetype='dashed') 
```

- The second term is more troublesome.
- If we repeat this argument as-is, replacing our indicator $D$ with $1-D$, we won't get what we want.

$$
\begin{aligned}
0 &= \E\qty[ (1-D) \ \qty{\mu(D,X)-\tilde\mu(D, X)} ]    
&& \text{orthogonality of errors} \\
&= P(D=0) \E\qty[(1-D) \ \qty{\mu(D,X)-\tilde\mu(D, X)} \mid D=0] 
&& \\
&+ P(D=1) \E\qty[(1-D) \ \qty{\mu(D,X)-\tilde\mu(D, X)} \mid D=1] 
&& \text{iterated expectations} \\
&= P(D=0) \sum_x   \inred{f_0(x)} \cdot 1 \cdot \inred{\qty{\mu(0,x) - \ingray{\tilde\mu(0,x)}} f_0(x)}  
&& \\
&+  P(D=1) \sum_x  \ingreen{f_1(x)} \cdot 0 \cdot \textcolor{lightgray}{\qty{\mu(1,x) - \tilde\mu(1,x)}} 
&& \text{calculating expectations} \\
\end{aligned}
$$

- That says the red curve's modeling error would be zero if we were averaging it over the red distribution.
- But we're not. We're averaging it over the green distribution. And we can see this is a problem.


## The Problem: Interacted Bilinear Model

```{r}
#| out-height: 400px
#| fig-asp: .4 
#| layout-nrow: 2 
#| layout-ncol: 2
sample.model = lm(Y ~ W*X, data=data)
grid$muhat = predict(sample.model, newdata=grid)

summarized.scatter + 
  geom_line(aes(x=X, y=muhat, color=factor(W)), data=grid, linetype='dashed') +
  geom_line(aes(x=X, y=mu, color=factor(W)), data=grid, linetype='solid', alpha=.7)  

scale_x = coord_cartesian(xlim=c(1.65,1.95))
sample.dist.plot(subset(term.estimates, n==500 & formula=='interacted')$att0,
                 NA, estimands$att0)  + scale_x
sample.dist.plot(subset(term.estimates, n==1000 & formula=='interacted')$att0, 
                 NA, estimands$att0)  + scale_x
sample.dist.plot(subset(term.estimates, n==2000 & formula=='interacted')$att0, 
                 NA, estimands$att0)  + scale_x 
```

- Here we're seeing the interacted model fit at sample size $n=`r n`$
- And sampling distributions at sample sizes 500, 1000, and 2000
- This isn't all that bad because our bias is small relative to standard error.
  - The line doesn't fit terribly.
  - Although you can imagine that, at larger sample sizes, it becomes a bigger deal.


## The Problem: Additive Bilinear Model

```{r}
#| out-height: 400px
#| fig-asp: .4 
#| layout-nrow: 2 
#| layout-ncol: 2
sample.model = lm(Y ~ W+X, data=data)
grid$muhat = predict(sample.model, newdata=grid)

summarized.scatter + 
  geom_line(aes(x=X, y=muhat, color=factor(W)), data=grid, linetype='dashed') +
  geom_line(aes(x=X, y=mu, color=factor(W)), data=grid, linetype='solid', alpha=.7)  

scale_x = coord_cartesian(xlim=c(1.65,1.95))
sample.dist.plot(subset(term.estimates, n==500 & formula=='additive')$att0,
                 NA, estimands$att0)  + scale_x
sample.dist.plot(subset(term.estimates, n==1000 & formula=='additive')$att0, 
                 NA, estimands$att0)  + scale_x
sample.dist.plot(subset(term.estimates, n==2000 & formula=='additive')$att0, 
                 NA, estimands$att0)  + scale_x 
```

- Here we're seeing the additive model fit at sample size $n=`r n`$
- And sampling distributions at sample sizes 500, 1000, and 2000
- This is bad.
  - The [red line]{.twocolor-red} fits terribly.
  - And bias is large relative to standard error.



## The Problem: Two-Intercepts Model

```{r}
#| out-height: 400px
#| fig-asp: .4 
#| layout-nrow: 2 
#| layout-ncol: 2
sample.model = lm(Y ~ W, data=data)
grid$muhat = predict(sample.model, newdata=grid)

summarized.scatter + 
  geom_line(aes(x=X, y=muhat, color=factor(W)), data=grid, linetype='dashed') +
  geom_line(aes(x=X, y=mu, color=factor(W)), data=grid, linetype='solid', alpha=.7)  

scale_x = coord_cartesian(xlim=c(1.65,1.95))
sample.dist.plot(subset(term.estimates, n==500 & formula=='intercepts')$att0,
                 NA, estimands$att0)  + scale_x
sample.dist.plot(subset(term.estimates, n==1000 & formula=='intercepts')$att0, 
                 NA, estimands$att0)  + scale_x
sample.dist.plot(subset(term.estimates, n==2000 & formula=='intercepts')$att0, 
                 NA, estimands$att0)  + scale_x 
```

- Here we're seeing the two-intercepts model fit at sample size $n=`r n`$
- And sampling distributions at sample sizes 500, 1000, and 2000
- This is worse.
  - The [red line]{.twocolor-red} fits terribly.
  - And bias is large relative to standard error.

## How Do We Fix This?

```{r}
#| out-height: 300px
#| fig-asp: .3 
sample.model = lm(Y ~ W*X, data=data)
grid$muhat = predict(sample.model, newdata=grid)

summarized.scatter + 
  geom_line(aes(x=X, y=muhat, color=factor(W)), data=grid, linetype='dashed') +
  geom_line(aes(x=X, y=mu, color=factor(W)), data=grid, linetype='solid', alpha=.7)  
```

- We're fitting the red curve in the wrong place for our summary.
- What can we do to fit it in the right place?


## Weighted Least Squares

```{r}
#| out-height: 300px
#| fig-asp: .3  

summarized.pop + geom_line(aes(x=X, y=mutilde, color=factor(W)), data=grid, linetype='dotted') 
```

$$
\text{modeling error} = \textcolor{lightgray}{\sum_x f_1(x) \qty{\tilde\mu(1, x) - \mu(1, x)}} - \sum_x \ingreen{f_1(x)} \inred{\qty{\ingray{\tilde\mu(0,x)} - \mu(0,x)}}
$$

- What will work is *weighted least squares regression*.
- The population regression errors from this have an orthogonality property in terms of a *weighted* average.
- To get the what we want, we'll use weights $w(x) = \ingreen{f_1(x)}/\inred{f_0(x)}$.

$$
\begin{aligned}
0 &= \E\qty[ w(X) \ (1-D) \ \qty{\mu(D,X)-\tilde\mu(D, X)} ]    
&& \text{orthogonality of errors} \\
&= P(D=0) \E\qty[ w(X) \ (1-D) \ \qty{\mu(D,X)-\tilde\mu(D, X)} \mid D=0] 
&& \\
&+ P(D=1) \E\qty[ w(X) \ (1-D) \ \qty{\mu(D,X)-\tilde\mu(D, X)} \mid D=1] 
&& \text{iterated expectations} \\
&=  P(D=0) \sum_x   \inred{f_0(x)} \frac{\ingreen{f_1(x)}}{\inred{f_0(x)}} \cdot 1 \cdot \inred{\qty{\mu(0,x) - \ingray{\tilde\mu(0,x)}}}  
&& \\
&+  P(D=1) \sum_x \ingreen{f_1(x)} \frac{\ingreen{f_1(x)}}{\inred{f_0(x)}} \cdot 0 \cdot \textcolor{lightgray}{\qty{\mu(1,x) - \tilde\mu(1,x)} } 
&& \text{calculating expectations} 
\end{aligned}
$$


## This Works: Interacted Bilinear Model

```{r}
#| out-height: 400px
#| fig-asp: .4 
#| layout-nrow: 2 
#| layout-ncol: 2

formula = 'interacted'

weights = pi(data$X)/ (1-pi(data$X)) 
sample.model = lm(formulas[[formula]], weights=weights, data=data)
grid$muhat = predict(sample.model, newdata=grid)

unweighted.model = lm(formulas[[formula]],  data=data)
grid$muhat.unweighted = predict(unweighted.model, newdata=grid)

summarized.scatter + 
  geom_line(aes(x=X, y=muhat, color=factor(W)), data=grid, linetype='dashed') +
  geom_line(aes(x=X, y=muhat.unweighted, color=factor(W)), data=grid, linetype='dotted') +
  geom_line(aes(x=X, y=mu, color=factor(W)), data=grid, linetype='solid', alpha=.7)  

scale_x = coord_cartesian(xlim=c(1.65, 1.95))
sample.dist.plot(subset(term.estimates, n==500 & formula==formula)$watt0,
                 NA, estimands$att0)  + scale_x
sample.dist.plot(subset(term.estimates, n==1000 & formula==formula)$watt0, 
                 NA, estimands$att0)  + scale_x
sample.dist.plot(subset(term.estimates, n==2000 & formula==formula)$watt0, 
                 NA, estimands$att0)  + scale_x 
```

- Here we're seeing the interacted model fit at sample size $n=`r n`$ with weights (dashed) and without (dotted)
- And $\widehat{ATT}$ sampling distributions based on our weighted regression at sample sizes 500, 1000, and 2000
- The impact of the weights is subtle visually, but it'll be huge inferentially at larger sample sizes.

## This Works: Additive Bilinear Model

```{r}
#| out-height: 400px
#| fig-asp: .4 
#| layout-nrow: 2 
#| layout-ncol: 2

formula = 'additive'

weights = pi(data$X)/ (1-pi(data$X)) 
sample.model = lm(formulas[[formula]], weights=weights, data=data)
grid$muhat = predict(sample.model, newdata=grid)

unweighted.model = lm(formulas[[formula]],  data=data)
grid$muhat.unweighted = predict(unweighted.model, newdata=grid)

summarized.scatter + 
  geom_line(aes(x=X, y=muhat, color=factor(W)), data=grid, linetype='dashed') +
  geom_line(aes(x=X, y=muhat.unweighted, color=factor(W)), data=grid, linetype='dotted') +
  geom_line(aes(x=X, y=mu, color=factor(W)), data=grid, linetype='solid', alpha=.7)  

scale_x = coord_cartesian(xlim=c(1.65, 1.95))
sample.dist.plot(subset(term.estimates, n==500 & formula==formula)$watt0,
                 NA, estimands$att0)  + scale_x
sample.dist.plot(subset(term.estimates, n==1000 & formula==formula)$watt0, 
                 NA, estimands$att0)  + scale_x
sample.dist.plot(subset(term.estimates, n==2000 & formula==formula)$watt0, 
                 NA, estimands$att0)  + scale_x 
```

## This Works: Two-Intercepts Model

```{r}
#| out-height: 400px
#| fig-asp: .4 
#| layout-nrow: 2 
#| layout-ncol: 2

formula = 'intercepts'

weights = pi(data$X)/ (1-pi(data$X)) 
sample.model = lm(formulas[[formula]], weights=weights, data=data)
grid$muhat = predict(sample.model, newdata=grid)

unweighted.model = lm(formulas[[formula]],  data=data)
grid$muhat.unweighted = predict(unweighted.model, newdata=grid)

summarized.scatter + 
  geom_line(aes(x=X, y=muhat, color=factor(W)), data=grid, linetype='dashed') +
  geom_line(aes(x=X, y=muhat.unweighted, color=factor(W)), data=grid, linetype='dotted') +
  geom_line(aes(x=X, y=mu, color=factor(W)), data=grid, linetype='solid', alpha=.7)  

scale_x = coord_cartesian(xlim=c(1.65, 1.95))
sample.dist.plot(subset(term.estimates, n==500 & formula==formula)$watt0,
                 NA, estimands$att0)  + scale_x
sample.dist.plot(subset(term.estimates, n==1000 & formula==formula)$watt0, 
                 NA, estimands$att0)  + scale_x
sample.dist.plot(subset(term.estimates, n==2000 & formula==formula)$watt0, 
                 NA, estimands$att0)  + scale_x 
```


## Getting Both Terms at Once

$$
\text{modeling error} = \sum_x \ingreen{f_1(x)} \ingreen{\qty{\ingray{\tilde\mu(1, x)} - \mu(1, x)}} - \sum_x \ingreen{f_1(x)} \inred{\qty{\ingray{\tilde\mu(0,x)} - \mu(0,x)}}
$$

- To be precise, we use this weighted regression. This uses weight ...
  - $1$ for the $D=1$ observations, so our no-bias argument for the first term works.
  - $\ingreen{f_1(x)}/\inred{f_0(x)}$ for the $D=0$ observations, so the one for the second term works, too.

$$
\hat\mu = \argmin_{m \in \model} \frac{1}{n} \sum_{i=1}^n w(D_i, X_i) \qty{ m(D_i, X_i) - Y_i }^2 \qfor w(D,X) = D + (1-D)\frac{\ingreen{f_1(X)}}{\inred{f_0(X)}}
$$

- It may be inobvious that we can do this because we don't know these densities $\ingreen{f_1(x)}$ and $\inred{f_0(x)}$.
- But --- in a conditionally randomized experiment --- we do know their ratio. 
- We can calculate it from the within-group assignment probability $\pi(x)$ using *Bayes' rule*.
- Up to constant factors. And that's all we need---a constant here doesn't change the solution $\hat\mu$.

## The Equivalent, Implementable Form

$$
\begin{aligned}
\hat\mu = \argmin_{m \in \model} \frac{1}{n} \sum_{i=1}^n w(D_i, X_i) \qty{ m(D_i, X_i) - Y_i }^2 
&\qfor             w(D,X) &= D + (1-D)\frac{\ingreen{f_1(X)}}{\inred{f_0(X)}} \\
&\qqtext{or  } \   w(D,X) &= D + (1-D) \frac{\pi(X)}{1-\pi(X)} 
\end{aligned}
$$

- To show these weights are equivalent up to constant factors, we use Bayes' rule.
$$
\begin{aligned}
P(A|B) &= \frac{P(B|A)P(A)}{P(B)} && \qqtext{ is Bayes' rule} \\
\frac{P(A|B)}{P(A|B')} &= \frac{P(B|A)}{P(B'|A)} \times \frac{P(B')}{P(B)} && \qqtext{ follows from it.}
\end{aligned}
$$

- In particular, we take ...
  - $A$ to be the event that $X=x$ 
  - $B$ to be the event that $D=1$
  - $B'$ to be the event that $D=0$
- Plugging these in ...

$$
\begin{aligned}
\frac{f_1(x)}{f_0(x)} 
&= \frac{P(X=x \mid D=1)}{P(X=x \mid D=0)} && \text{ is just notation } \\ 
&= \frac{P(D=1 \mid X=x)}{P(D=0 \mid X=x)} \times \frac{P(D=0)}{P(D=1)} && \text{ is Bayes' Rule} \\
&= \frac{\pi(x)}{1-\pi(x)} \ \frac{P(D=1)}{P(D=0)} && \text{ is just notation again } \\
\end{aligned}
$$


## IPW and Point Estimation

```{r}
#| out-height: 600px

att.summaries$weighted = att.summaries$pi.formula %in% c('W~X', 'known')
att.summaries$estimated.weights = att.summaries$pi.formula %in% c('W~X')

mu.formula = 'Y~W*X' 
pop.est = pop.estimates[pop.estimates$pi.formula == 'W~1' &
                        pop.estimates$mu.formula == mu.formula, 'att'] 

rmse.plot = ggplot(att.summaries[att.summaries$pi.formula %in% c('W~1', 'known') & 
                                 att.summaries$mu.formula == mu.formula, ]) + 
  geom_line(aes(x=n, y=sqrt(bias^2 + se^2), color=factor(weighted)), alpha=.5) +
  geom_point(aes(x=n, y=sqrt(bias^2 + se^2), color=factor(weighted)), alpha=.5) + 
  geom_hline(aes(yintercept=0), alpha=.3, linewidth=2) +
  geom_hline(aes(yintercept=abs(pop.est-estimands$att)), color='black', alpha=.2) +
  scale_x_sqrt(breaks=ns, labels=label_comma()) + 
  theme_update(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  guides(color=FALSE)
rmse.plot
```

- Here's a simulation study comparing ATT estimators based on [unweighted]{.twocolor-red} and [weighted]{.twocolor-green} least squares fits of the interacted bivariate model.
- I'm plotting the square root of mean squared error (RMSE) at a range of sample sizes.
- We can see that, except in very small samples, the weighted one performs better.
- And the error curve of the [unweighted]{.twocolor-green} one is flattening out.
- It's not converging to zero. What is it converging to?

## IPW and Point Estimation

```{r}
#| out-height: 600px
rmse.plot 
```

- It's converging to the error we get using the **population least squares fit**.
- I've plotted that as a [thin gray horizontal line]{.fg style="color:gray"}

## IPW in Observational Studies 

```{r}
#| out-height: 600px
rmse.plot 
```

- Often, we analyze observational studies as if they were conditionally randomized experiments.
  - This works if, among groups with identical covariates, people select treatment with the same probability.
  - This is usually an imperfect approximation, but often it's the best we can do,
- What this means, implementation-wise, is that we don't know the randomization probabilities.
- And we have to make do with estimates. How well do you think this works?

$$
\hat\mu = \argmin_{m \in \model} \frac{1}{n} \sum_{i=1}^n w(D_i, X_i) \qty{ m(D_i, X_i) - Y_i }^2 \qfor \hat w(D,X) = D + (1-D)\frac{\hat\pi(X)}{1-\hat\pi(X)}
$$


## It Usually Works Pretty Well

```{r}
#| out-height: 600px

mu.formula = 'Y~W*X' 
pop.est = pop.estimates[pop.estimates$pi.formula == 'W~1' &
                        pop.estimates$mu.formula == mu.formula, 'att'] 

rmse.plot.all = ggplot(att.summaries[att.summaries$mu.formula == mu.formula, ]) + 
  geom_line(aes(x=n, y=sqrt(bias^2 + se^2), color=factor(weighted), linetype=factor(estimated.weights)), alpha=.5) +
  geom_point(aes(x=n, y=sqrt(bias^2 + se^2), color=factor(weighted), shape=factor(estimated.weights)), alpha=.5) + 
  geom_hline(aes(yintercept=0), alpha=.3, linewidth=2) +
  geom_hline(aes(yintercept=abs(pop.est-estimands$att)), color='black', alpha=.2) +
  scale_x_sqrt(breaks=ns, labels=label_comma()) + 
  theme_update(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  guides(color=FALSE, shape=FALSE, linetype=FALSE) 
rmse.plot.all
```

- In this instance it works very well. I've drawn the fit with estimated weights using a [dashed green line]{.twocolor-green}.
- At this scale, we can't see the difference between what we get with the actual or estimated weights.
- Even though we've used a misspecified logit-linear model to estimate the assignment probabilities $\pi(x)$.

## It Usually Works Pretty Well
```{r}
#| out-height: 600px
rmse.plot.all + coord_cartesian(xlim=c(20000-10000, 100000+1000), ylim=c(0, .015)) 
```

- Zooming in, we see that we're doing a bit *better* with the estimated weights.


## Estimated Treatment Probabilities and Weights

```{r}
#| out-height: 600px
#| layout-ncol: 2
#| fig-asp: .6 
 
x=seq(50,80,by=.01)
pimodel = glm(W~X, family=binomial(), data) 
pihat = function(x) { predict(pimodel, newdata=data.frame(X=x), type='response') }
ggplot() + geom_line(aes(x=x, y=pi(x))) + geom_line(aes(x=x, y=pihat(x)), color='blue')
ggplot() + geom_line(aes(x=x, y=pi(x)/(1-pi(x)))) + geom_line(aes(x=x, y=pihat(x)/(1-pihat(x))), color='blue')
```

- On the left, we're showing the probabilities $\pi(x)$ vs. estimates [$\hat\pi(x)$]{.fg style="color:blue"}
- On the right, the corresponding weights $\pi(x)/\{1-\pi(x)\}$ and [$\hat\pi(x)/\{1-\hat\pi(x)\}$]{.fg style="color:blue"}.
- We do pretty well, but we're far from perfect. Like we do when we fit of our interacted model to the outcomes.
- You can think of IPW as combining two regressions in a way that *amplifies* their accuracy.

## What Happens With An Additive Model
```{r}
#| out-height: 600px

mu.formula = 'Y~W+X' 
pop.est = pop.estimates[pop.estimates$pi.formula == 'W~1' &
                        pop.estimates$mu.formula == mu.formula, 'att'] 

rmse.plot.all = ggplot(att.summaries[att.summaries$mu.formula == mu.formula, ]) + 
  geom_line(aes(x=n, y=sqrt(bias^2 + se^2), color=factor(weighted), linetype=factor(estimated.weights)), alpha=.5) +
  geom_point(aes(x=n, y=sqrt(bias^2 + se^2), color=factor(weighted), shape=factor(estimated.weights)), alpha=.5) + 
  geom_hline(aes(yintercept=0), alpha=.3, linewidth=2) +
  geom_hline(aes(yintercept=abs(pop.est-estimands$att)), color='black', alpha=.2) +
  scale_x_sqrt(breaks=ns, labels=label_comma()) + 
  theme_update(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  guides(color=FALSE, shape=FALSE, linetype=FALSE) 

rmse.plot.all
```

## What Happens With An Additive Model --- Zoomed In
```{r}
#| out-height: 600px
rmse.plot.all + coord_cartesian(xlim=c(20000-10000, 100000+1000), ylim=c(0, .015)) 
```

## A Comparison
```{r}
#| out-height: 600px

mu.formula = c('Y~W+X', 'Y~W*X') 
pi.formula = c('W~1', 'known')
rmse.plot.all = ggplot(att.summaries[(att.summaries$mu.formula %in% mu.formula) &
                                     (att.summaries$pi.formula %in% pi.formula), ]) +
  geom_line(aes(x=n, y=sqrt(bias^2 + se^2),  color=factor(pi.formula), linetype=factor(mu.formula)), alpha=.5) +
  geom_point(aes(x=n, y=sqrt(bias^2 + se^2), color=factor(pi.formula), shape=factor(mu.formula)), alpha=.5) + 
  geom_hline(aes(yintercept=0), alpha=.3, linewidth=2) +
  scale_x_sqrt(breaks=ns, labels=label_comma()) + 
  theme_update(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  guides(color=FALSE, shape=FALSE, linetype=FALSE) 

rmse.plot.all
```

- As before, we're showing weighted (green) and unweighted (red) least squares fits.
- But now we're comparing the interacted model (solid) to the additive model (dashed).
  - Without weights, the additive model is terrible.
  - With weights, it's worse than the interacted one, but not bad.
- To see much of a difference, we need to zoom in.

## A Comparison -- Zoomed In

```{r}
#| out-height: 600px
rmse.plot.all + coord_cartesian(xlim=c(20000-10000, 100000+1000), ylim=c(0, .015)) 
```

## Advice

```{r}
#| out-height: 400px
#| fig-asp: .4 
#| layout-ncol: 2

formula = 'interacted'

weights = pi(data$X)/ (1-pi(data$X)) 
sample.model = lm(formulas[[formula]], weights=weights, data=data)
grid$muhat = predict(sample.model, newdata=grid)

unweighted.model = lm(formulas[[formula]],  data=data)
grid$muhat.unweighted = predict(unweighted.model, newdata=grid)

summarized.scatter + 
  geom_line(aes(x=X, y=muhat, color=factor(W)), data=grid, linetype='dashed') +
  geom_line(aes(x=X, y=muhat.unweighted, color=factor(W)), data=grid, linetype='dotted') +
  geom_line(aes(x=X, y=mu, color=factor(W)), data=grid, linetype='solid', alpha=.7)  

formula = 'additive'

weights = pi(data$X)/ (1-pi(data$X)) 
sample.model = lm(formulas[[formula]], weights=weights, data=data)
grid$muhat = predict(sample.model, newdata=grid)

unweighted.model = lm(formulas[[formula]],  data=data)
grid$muhat.unweighted = predict(unweighted.model, newdata=grid)

summarized.scatter + 
  geom_line(aes(x=X, y=muhat, color=factor(W)), data=grid, linetype='dashed') +
  geom_line(aes(x=X, y=muhat.unweighted, color=factor(W)), data=grid, linetype='dotted') +
  geom_line(aes(x=X, y=mu, color=factor(W)), data=grid, linetype='solid', alpha=.7)  
```

- Use inverse probability weighted regression, estimating $\pi$ if necessary.
- Use a good regression model, too. One that can fit the data.

$$
\begin{aligned}
\hat\mu = \argmin_{m \in \model} \frac{1}{n} \sum_{i=1}^n w(D_i, X_i) \qty{ m(D_i, X_i) - Y_i }^2 
\qfor  &w(D,X) = D + (1-D) \frac{\pi(X)}{1-\pi(X)} \\
\qqtext{ or } &w(D,X) = D + (1-D) \frac{\hat\pi(X)}{1-\hat\pi(X)}
\end{aligned}
$$

- You need different inverse probability weights to estimate different summaries: ATT, ATC, ATE, etc.

