---
title: "QTM 220 Lecture 10"
subtitle: "Inferential Statistics with Non-Binary Covariates"
format: 
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---

```{r load-libraries}
library(purrr)
library(tibble)
library(dplyr)
library(magrittr)
library(stringr)
library(ggplot2)
library(htmltools)
library(kableExtra)
library(scales)
library(cpsR)
library(mapdata)

options(knitr.kable.NA = '')

compact_number = label_number(accuracy=.1, scale_cut=cut_short_scale())
ellipsis <- function(df, nhead=5L, ntail=2) {
    rownames(df) = as.character(1:nrow(df))
    stopifnot("data.frame" %in% class(df))
    els <- rep(NA, ncol(df)) %>% 
        matrix(nrow=1, dimnames=list(NULL, names(df))) %>% 
        data.frame(stringsAsFactors=FALSE) %>% 
        tibble::as_tibble()
    rownames(els) = "⋮"
    dplyr::bind_rows(head(df, nhead), els, tail(df, ntail))
}

percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }
```

```{r ggplot-themes}
fontsize = 12
textcolor = "#888888"

class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
        panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x  = element_text(colour = textcolor, size=fontsize),
        axis.text.y  = element_text(colour = textcolor, size=fontsize),
        axis.title.x  = element_text(colour = textcolor),
        axis.title.y  = element_text(colour = textcolor, angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
              axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
           panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(simplified.theme)
```




```{r get-data}
#| cache: true
#| 
# for cps variable interpretation see https://www2.census.gov/programs-surveys/cps/datasets/2022/march/asec2022_ddl_pub_full.pdf
census.api.key = '8084f81e4f71e4d32638dac15f188b424f72d063'
cps.sampling.ratio = 1/25 # Not true. 1/2500 households are surveyed. But for the sake of plotting, it's ok.

california.state.code = 6
year = 2020
cps.data = results=get_asec(year=year, vars=c('pearnval', 'a_age', 'a_hga', 'gestfips', 'gtco', 'gtcbsasz'), key=census.api.key)
```

```{r transform-data}
#| cache: false
#| 
all.earnings = data.frame(state.code = cps.data$gestfips,
                          county.code = cps.data$gtco,
                          age=cps.data$a_age,
                          income=cps.data$pearnval,
                          education.code=cps.data$a_hga,
                          education=case_match(cps.data$a_hga,
                                           0  ~ 0,   # child
                                           31 ~ 0,   # < grade 1
                                           32 ~ 4,   #  grade 1-4
                                           33 ~ 6,   #  grade 5-6
                                           34 ~ 8,   #  grade 7-8 
                                           35 ~ 9,   #  grade 9
                                           36 ~ 10,  #  grade 10
                                           37 ~ 11,  #  grade 11
                                           38 ~ 11,  #  grade 12 no diploma
                                           39 ~ 12,  #  high school grad
                                           40 ~ 13,  #  some college
                                           41 ~ 14,  #  associate degree (vocational)
                                           42 ~ 14,  #  associate degree (academic)
                                           43 ~ 16,  #  bachelors degree
                                           44 ~ 18,  #  masters degree
                                           45 ~ 20,  #  professional school degree
                                           46 ~ 20)) #  doctorate

all.earnings$degree = factor(all.earnings$education >= 16, 
                     levels=c(FALSE, TRUE), 
                     labels=c('no 4-year degree','4-year degree')) 

coarsen = function(x, group, seed=0) {
  current.seed = .Random.seed
  set.seed(seed)
  x.coarsened = rep(NA, length(x))
  for(gg in unique(group)) {
    x.coarsened[group==gg] = runif(sum(group==gg), min(x[group==gg]), max(x[group==gg]))
  }
  set.seed(current.seed)
  x.coarsened
} 

in.interval = function(X, interval) { interval[1] <= X & X <= interval[2] }
earnings = filter(all.earnings, education >= 8 & 
                                state.code == california.state.code & 
                                age %>% in.interval(c(25,35)))
Y = earnings$income
X = earnings$education
```


```{r get-county-info}
# county info from https://en.wikipedia.org/wiki/List_of_counties_in_California
# converted to csv using https://wikitable2csv.ggor.de

to.number = function(s) { as.numeric(str_trim(gsub(",", "", s))) }
ca_county_info = read.csv('california_county_info.csv') %>%
  mutate(subregion = tolower(sub(' County', '', County))) %>%
  mutate(pop = to.number(Population..2022.)) %>%
  mutate(area = to.number(sub('sq.*', '', Area))) 

earnings.county = left_join(earnings, ca_county_info[,c('FIPS.code', 'subregion')], by=c('county.code' = 'FIPS.code'))
earnings$county = earnings.county$subregion
earnings$county[is.na(earnings$county)]='unknown'
```


```{r make-maps}
#| cache: true
#| 

# To give our survey recipients a bit more reality, we'll give them fake locations on a map
# And we'll make up a population of California residents that they're sampled from
# maps are based on code from https://eriqande.github.io/rep-res-eeb-2017/map-making-in-R.html

counties = map_data("county")
ca_county = counties %>%
  filter(region == "california")
ca_county_centers = ca_county %>%
  group_by(subregion) %>%
  summarize(county.lat = mean(lat), county.long=mean(long))

ca_county_more = left_join(ca_county_centers, ca_county_info, by="subregion")
ca_all = left_join(ca_county, ca_county_more, by="subregion")


ca.plot = ggplot(data=ca_all, mapping=aes(x=long, y=lat, group=group)) + 
  coord_quickmap() +
  geom_polygon(fill='gray', color = "white", alpha=.4) +
      scale_fill_gradient(trans="log10") + 
      theme_void()

ca.pop.plot = ggplot(data=ca_all, mapping=aes(x=long, y=lat, group=group)) + 
  coord_quickmap() +
  geom_polygon(aes(fill = pop/area), color = "white") +
      scale_fill_gradient(trans="log10") + 
      theme_void()

## we'll make up a population as follows
# 1. choose the number of people to be the number of people in our sample scaled up to account for the cps sampling ratio
# 2. for each, sample a county with probability proportional to its finite.population
# 3. for each, sample a latitude and longitude by taking a randomly weighted combination of the    
#    county's vertices. As long as the county is convex, these points will be inside the county.
# 4. we'll give them a degree too, chosen independently of all else from the empirical distribution on the sample
n = nrow(earnings) / cps.sampling.ratio

# takes a list of subregions, possibly with duplicates
# for each element, chooses random lat/long coordinates within the sample.subregion
# by taking a random convex combination of the subregion's vertices
sample.within.subregion = function(subregions) {
  output = data.frame(subregion=subregions, lat=NA, long=NA)
  for(subregion in unique(subregions)) { 
      vertices = ca_county[ca_county$subregion==subregion, c('lat','long')]
      U = sum(subregions==subregion)
      V = nrow(vertices)
      weights = abs(rt(U*V, 1))
      dim(weights) = c(U,V)
      weights = (1/rowSums(weights)) * weights
      output[subregions==subregion, c('lat', 'long')] = 
                 cbind(weights %*% vertices$lat, weights %*% vertices$long)
    }
    output
}

finite.population.subregion = sample(ca_county_more$subregion, size=n, replace=TRUE, prob=ca_county_more$pop)
finite.population = sample.within.subregion(finite.population.subregion)
finite.population$degree = sample(earnings$degree, nrow(finite.population), replace=TRUE)
finite.population$group=1 # for plotting

display.population = sample_n(finite.population, min(100000, nrow(finite.population)))
```


```{r make-sample}
# and we'll force the degree-distribution in the sample to match our survey responses
our.sample = sample_n(finite.population, nrow(earnings))
our.sample$degree=earnings$degree

pop.plot = ca.plot + geom_point(aes(x=long, y=lat), data=display.population, size=.02, alpha=.1) 
sample.plot = ca.plot + geom_point(aes(x=long, y=lat), data=our.sample, size=.2, alpha=.5)
pop.plot.inf = ca.pop.plot + geom_point(aes(x=long, y=lat), data=display.population, size=.01, alpha=.1)  
sample.plot.inf = ca.pop.plot + geom_point(aes(x=long, y=lat), data=our.sample, size=.2, alpha=.5)
```

```{r visualization-defaults}
scale_y = function(limits=c(-.3e5, 3e5)) { 
  scale_y_continuous(labels=label_dollar(), limits=limits)
}

years = 8:20
scale_x = function(limits=c(min(years)-.5, max(years)+.5)) { 
  scale_x_continuous(breaks=years, labels=sprintf('%dy', years), limits=limits)
}

dot.size = 2
```


```{r define-colors}
twocolor.red   = 'rgb(248,118,109)'
twocolor.green = 'rgb(0,191,196)'
```

## Last Time

::: {.hidden}
$$
% color defs aren't making it into MathJax
%\definecolor{twocolor-red}{RGB}{248, 118, 109}   % red   used by ggplot in plots w/ 2 colors 
%\definecolor{twocolor-green}{RGB}{0, 191, 196}   % green used by ggplot in plots w/ 2 colors
% so create commands for using those colors instead
\newcommand{\inred}[1]{\textcolor[RGB]{248,118,109}{#1}}
\newcommand{\ingreen}[1]{\textcolor[RGB]{0,191,196}{#1}}
\newcommand{\inblue}[1]{\textcolor{blue}{#1}}

\newcommand{\ind}[1]{1_{\{#1\}}}
\newcommand{\with}{\ingreen{\text{degree}}}
\newcommand{\without}{\inred{\text{no degree}}}

\DeclareMathOperator{\E}{E}degredegre
\DeclareMathOperator{\Var}{Var}
$$ 

:::

```{r scatter-plots}

earnings.jitter = ggplot(earnings) + geom_point(aes(x=education, y=income), color='#F7766D',
                              size=dot.size, position=position_jitter(seed=0,w=.3), alpha=.1) +
                   scale_x() + scale_y() + guides(color=FALSE)

summaries = earnings %>% 
          group_by(education) %>% 
          summarize(mean=mean(income),
                    sd=sd(income))
                    
```


```{r}
#| out-height: 300px
#| fig-asp: .3

earnings.jitter
```

- We talked summaries of the relationship between income and education *in a sample*.
- We talked about a number of types of summaries. 
- What you'd use depends on what you're interested in.
- Let's step through them again.


## Fine Comparisons

```{r}
#| out-height: 300px
#| fig-asp: .3
      
earnings.jitter + geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries)
```

- One simple option is to look at groups where the covariate takes on a single value. 
- e.g. at the income mean and standard deviation for people with 8,9,10, etc. years of schooling.

$$
\begin{aligned}
\hat \mu(x) &= \frac{1}{N_x}\ \sum_{i:X_i=x} Y_i \qfor N_x = \sum_{i:X_i=x} 1 \\
\hat \sigma(x) &= \sqrt{ \frac{1}{N_x}\sum_{i:X_i=x} \qty{Y_i - \hat \mu(x)}^2 }
\end{aligned}
$$

## Fine Comparisons

```{r}
#| out-height: 300px
#| fig-asp: .3
      
earnings.jitter + geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries)
```

- Or compare any pair of those groups. It's natural to look at groups that are in some sense adjacent.
    - $\hat\mu(18) - \hat\mu(16)$: the average value of a masters' degree (only) vs. a 4-year one (only).
    - $\hat\mu(16) - \hat\mu(14)$: the value of a 4-year degree (only) vs. a 2-year degree (only).
    - $\hat\mu(14) - \hat\mu(12)$: the value of a 2-year degree (only) vs. a high school degree (only).
- Sometimes we're interested in exactly one of these things and this is what you want.
- But it's not so great if we're a bit less focused.
  - That'd mean we might have to look at and report a lot of different numbers.
  - It can get to be too much. We're not summarizing enough.


## Coarsening

```{r}
#| out-height: 300px
#| fig-asp: .3

groups = factor(1:4, 
               levels=1:4, 
               labels=c('no college degree', '2-year degree', '4-year degree', 'graduate degree'))

group = function(x) { 
  case_match(x,
    8:13  ~ groups[1],
    14    ~ groups[2],
    16    ~ groups[3],
    18:20 ~ groups[4])
}


coarsened.summaries = data.frame(X=X, Y=Y, G=group(X)) %>% 
  group_by(G) %>% 
  summarize(mean=mean(Y), sd=sd(Y), mean.X = mean(coarsen(X, G)))


ggplot() + geom_point(aes(x=coarsen(X,group(X)), y=Y), color='#F7766D', position=position_jitter(w=.3, seed=0), alpha=.3) +
           geom_pointrange(aes(x=mean.X, y=mean, ymin=mean-sd, ymax=mean+sd), data=coarsened.summaries) + scale_x() + scale_y()
```

- One remedy is to work with coarser *aggregate groups*, i.e. groups where the covariate takes on several values.
- Here, for example, is a breakdown in 4 groups.
    - &lt; 14 years: no college degree. $X_i \in \{ 8 \ldots 13\}$.
    - =14 years: 2-year degree.         $X_i \in \{14\}$.         
    - =16 years: 4-year degree.         $X_i \in \{ 16 \}$ 
    - &gt; 16 years: graduate degree.   $X_i \in \{ 18 \ldots 20 \}$.

## Coarsening
  
```{r}
#| out-height: 300px
#| fig-asp: .3

ggplot() + geom_point(aes(x=coarsen(X,group(X)), y=Y), color='#F7766D', position=position_jitter(w=.3, seed=0), alpha=.3) +
           geom_pointrange(aes(x=mean.X, y=mean, ymin=mean-sd, ymax=mean+sd), data=coarsened.summaries) + scale_x() + scale_y()
```


- We can express the means within these groups as *weighted averages* of the fine-group means.
$$ 
\begin{aligned}
\hat \mu(\{8...13\})    
&= \frac{\sum_{i : X_i \in \{8 \ldots 13\}} Y_i}{\sum_{i : X_i \in \{8 \ldots 13\}} 1} 
= \frac{\sum_{x \in \{8 \ldots 13\}}\ \sum_{i : X_i = x} Y_i}{\sum_{x \in \{8 \ldots 13\}}\ \sum_{i : X_i = x} 1} \\
&= \frac{\sum_{x \in \{8 \ldots 13\}}\  N_x \ \hat \mu(x)}{\sum_{x \in \{8 \ldots 13\}} N_x} 
= \sum_{x \in \{8 \ldots 13\}} P_x \hat \mu(x) \quad \text{ for } \quad P_x = \frac{N_x}{\sum_{x \in \{8 \ldots 13\}} N_x}
\end{aligned}
$$
- The weights $P_x$ are the *proportions* of people in each fine group.

## Coarsening

```{r}
#| out-height: 300px
#| fig-asp: .3

ggplot() + geom_point(aes(x=coarsen(X,group(X)), y=Y), color='#F7766D', position=position_jitter(w=.3, seed=0), alpha=.3) +
           geom_pointrange(aes(x=mean.X, y=mean, ymin=mean-sd, ymax=mean+sd), data=coarsened.summaries) + scale_x() + scale_y()
```


- Again, sometimes we're interested in exactly one of these things and this is what you want.
- But it's also a bit less overwhelming to think about/report all of it than what we had without coarsening.


## Aggregating Comparisons

```{r}
#| out-height: 300px
#| fig-asp: .3

earnings.jitter.hs = ggplot(earnings) + geom_point(aes(x=education, y=income, alpha=education <= 12), color='#F7766D',
                              size=dot.size, position=position_jitter(seed=0,w=.3)) +
                   scale_x() + scale_y() + scale_alpha_manual(breaks=c(FALSE, TRUE), values=c(.05,.3)) + guides(color=FALSE, alpha=FALSE)

aggregate.comparison.plot = earnings.jitter.hs +
  geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries[summaries$education <= 12,]) +
  geom_line(aes(x=education, y=mean), data=summaries[summaries$education<=12,], color='black', linetype='dashed') +
  scale_y(c(0, .7e5))

aggregate.comparison.plot
```

- Another option---rather than comparing aggregates---is aggregating comparisons.
- We can summarize the 'value' of each year of high school by the associated increment in mean income.
- Visually, that's the slope of the dashed black line leading up to it.
- To reduce these to a single number, we can *aggregate* these increments. 
- There are a few ways to do that.
  

## Aggregating Comparisons {.smaller}

```{r}
#| out-height: 300px
#| fig-asp: .3
aggregate.comparison.plot
```


1. We can average over the four years of high school 
    $$
    \small{
    \frac14 \sum_{x=9}^{12} \qty{ \hat\mu(x) - \hat\mu(x-1) } = \hat \mu(12) - \hat\mu(8)
    }
    $$
2. Or average over the people with &ge;1 year of high school.
    $$
    \small{
    \frac{1}{N_{\{9\ldots 12\}}}\sum_{i:X_i \in 9 \ldots 12} \qty{ \hat\mu(X_i) - \hat\mu(X_i-1) } 
    = P_{12}\hat\mu(12) + \qty(P_{11} - P_{12}) \hat\mu(11)  + \qty(P_{10} - P_{11})\hat\mu(10)  + \qty(P_{9} - P_{10}) \hat\mu(9) - P_9 \hat\mu(8)
    }
    $$
3. Or we can *fit a line* to these people we're comparing---those with 8-12 years of education---and take its slope.
    $$
    \small{
    \frac{\sum_{i : X_i \in 8 \ldots 12} (X_i - \bar X) Y_i}{\sum_{i : X_i \in 8 \ldots 12} (X_i - \bar X)^2} 
    = \sum_{x \in 8 \ldots 12} w (x) \ \hat\mu(x)  \qfor 
    \bar X = \frac{1}{N_{\{8 \ldots 12\}}} \sum_{i: X_i \in \{8 \ldots 12\}} X_i \qand
    w(x) = \frac{N_x (x-\bar X)}{\sum_{i:X_i \in 8 \ldots 12} (X_i-\bar X)^2} 
    }
    $$


## These Can Be Very Different

```{r}

mu = map_dbl(1:12, function(x) { mean(earnings$income[earnings$education==x]) })
epsilon = rnorm(nrow(earnings)) * .1*sd(earnings$income[earnings$education %in% 8:12])

earnings1 = earnings
earnings1$income = mu[ifelse(earnings$education==12, 12, 8)] + epsilon

earnings2 = earnings
earnings2$income = mu[ifelse(earnings$education >= 10, 12, 8)] + epsilon

mu3 = NA+mu
mu3[8]=mu[8]
mu3[9]=mu[10]
mu3[10]=(mu[11]+mu[12])/2
mu3[11]=mu[12]
mu3[12]=mu[12]
earnings3 = earnings
earnings3$income = mu3[earnings$education] + epsilon


over.people = function(earnings) {
  hsed = earnings$education[earnings$education %in% 9:12]
  mu = map_dbl(1:12, function(x) { mean(earnings$income[earnings$education==x]) })
  mean(mu[hsed] - mu[hsed-1])
}
over.years = function(earnings) {
  mu = map_dbl(1:12, function(x) { mean(earnings$income[earnings$education==x]) })
  mean(mu[9:12] - mu[8:11])
}
lsq.slope = function(earnings) { 
  model = lm(income ~ education, data=earnings[earnings$education <= 12,])
  coef(model)[2]
}

plot.aggregate.comparison = function(earnings2) {
  summaries2 = earnings2 %>% group_by(education) %>% summarize(mean=mean(income), sd=sd(income), mean.X = mean(education))
  aggregate.comparison.plot= ggplot(earnings2) + geom_point(aes(x=education, y=income), color='#F7766D', alpha=.3,
                              size=dot.size, position=position_jitter(seed=0,w=.3)) +
    geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries2[summaries2$education <= 12,]) +
    geom_line(aes(x=education, y=mean), data=summaries2[summaries2$education<=12,], color='black', linetype='dashed') 

  model = lm(income ~ education, data=earnings2[earnings2$education <= 12,])
  lm.predictions = data.frame(education=seq(8,12,by=.01))
  lm.predictions$muhat = predict(model, newdata=lm.predictions)
  aggregate.comparison.plot + scale_x(c(7.5,12.5)) + scale_y(c(0, .7e5)) +
      geom_line(aes(x=education, y=muhat), data=lm.predictions, color='red', alpha=.4)
}

```


:::: columns

::: column

  1. All that matters is the diploma.
  2. All that matters is 10th grade algebra.
  3. They teach the most important stuff first. 
:::

::: column


$$
\small{
\begin{array}{r|rrr}
\text{Story} & 1 & 2 & 3 \\
\hline
\text{Least Squares Slope}  & `r approx(lsq.slope(earnings1))`   &  `r approx(lsq.slope(earnings2))` & `r approx(lsq.slope(earnings3))`  \\
\text{Average over People}  & `r approx(over.people(earnings1))` &  `r approx(over.people(earnings2))` & `r approx(over.people(earnings3))`  \\
\text{Average over Years}   & `r approx(over.years(earnings1))`  &  `r approx(over.years(earnings2))` & `r approx(over.years(earnings3))`  \\
\end{array}
}
$$

:::
::::

```{r}
#| layout-ncol: 3
#| out-height: 400px
#| fig-asp: 1
#| fig-subcap:
#|  - Story 1
#|  - Story 2
#|  - Story 3 

plot.aggregate.comparison(earnings1)
plot.aggregate.comparison(earnings2)
plot.aggregate.comparison(earnings3)
```


## What All These Things Have in Common


- Each of these summaries is a *linear combination* of the subsamples means. Except the standard deviations $\hat\sigma(x)$.
$$
\text{summary} = \sum_x \hat \alpha_x \hat\mu(x) 
$$

- In some cases, the weights $\hat\alpha_x$ don't depend on the covariates.
  - $\hat \mu(18) - \hat\mu(16)$
  - $\frac14 \sum_{x=9}^{12} \qty{ \hat\mu(x) - \hat\mu(x-1) } = \frac{\hat\mu(12) - \hat\mu(8)}{4}$ 
- In others, they do.
  - $\hat \mu(14) - \hat\mu(\{8...12\}) = \hat\mu(14) - \sum_{x=8}^{12} P_x \hat\mu(x) \qfor P_x = \frac{N_x}{\sum_{x \in \{8 \ldots 12\}} N_x}$
  - $\frac{1}{N_{\{9\ldots 12\}}}\sum_{i:X_i \in 9 \ldots 12} \qty{ \hat\mu(X_i) - \hat\mu(X_i-1) } = P_{12}\hat\mu(12) + \qty(P_{11} - P_{12}) \hat\mu(11)  + \ldots$
  - $\frac{\sum_{i : X_i \in 8 \ldots 12} (X_i - \bar X) Y_i}{\sum_{i : X_i \in 8 \ldots 12} (X_i - \bar X)^2} = \sum_{x \in 8 \ldots 12} w (x) \ \hat\mu(x) \qfor w(x) = \frac{N_x (x-\bar X)}{\sum_{i:X_i \in 8 \ldots 12} (X_i-\bar X)^2}$
- But we'll see that this'll allow us to talk about statistical inference in a unified way.


## Today

```{r}
#| out-height: 700px
#| fig-ncol: 2
sample.plot
pop.plot
```


- We'll talk about *statistical inference* based on these sample summaries.
- That is, we'll talk about what they tell us about the population our sample was drawn from.
- As before, we'll work with data from the 2022 Current Population Survey.
- We'll look at California residents age 25-35 with at least an 8th-grade education.
  - This sample includes `r nrow(earnings)` people. 
  - We'll think of them as having been sampled with replacement from a population.
  - That is, from the population of all California residents meeting these criteria.
  - That's an approximation---the CPS is a complex survey---but it's not too bad.


## What We're Estimating

- We've talked about a lot of sample summaries. And we can summarize the entire population in the same ways.
- It's natural to think of the sample summaries as *estimates* of the population summaries.

:::: columns

::: column
### Sample
$$
\begin{array}{ccccccc}
X_1 & Y_1 & X_2 & Y_2 &  \ldots & X_n & Y_n \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\end{array}
$$
:::
::: column
### Population
$$
\begin{array}{ccccccc}
x_1 & y_1 & x_2 & y_2 &  \ldots & x_m & y_m \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\end{array}
$$
:::
::::

::: fragment
$$
\small{
\begin{aligned}
&\hat \mu(x) = \frac{1}{N_x}\ \sum_{i:X_i=x} Y_i \qfor N_x = \sum_{i:X_i=x} 1 && \text{ estimates } \quad \mu(x) = \frac{1}{m_x}\ \sum_{j:x_j=x} y_j \qfor m_x = \sum_{j:x_j=x} 1 \\
&\frac{\sum_{i:X_i \in 8 \ldots 12} Y_i}{\sum_{i:X_i \in 8 \ldots 12} 1} && \text{ estimates } \quad \frac{\sum_{j:x_j \in 8 \ldots 12} y_j}{\sum_{j:x_j \in 8 \ldots 12} 1} \\
&\frac{1}{4}\sum_{x=9}^{12} \qty{ \hat \mu(x) - \hat\mu(x-1) }                       && \text{ estimates } \quad \frac{1}{4}\sum_{x=9}^{12} \qty{ \mu(x) - \mu(x-1) }   \\
&\frac{\sum_{i:X_i \in 9 \ldots 12} \qty{ \hat\mu(X_i) - \hat\mu(X_i-1) }}{\sum_{i:X_i \in 9 \ldots 12} 1} && \text{ estimates }  \quad  \frac{\sum_{j:x_j \in 9 \ldots 12} \qty{ \mu(x_j) - \mu(x_j-1) }}{\sum_{j:x_j \in 9 \ldots 12} 1} \\ 
&\frac{\sum_{i : X_i \in 8 \ldots 12} (X_i - \bar X) Y_i}{\sum_{i : X_i \in 8 \ldots 12} (X_i - \bar X)^2} && \text{ estimates } \quad \frac{\sum_{j : x_j \in 8 \ldots 12} (x_j - \bar x) y_j}{\sum_{j : x_j \in 8 \ldots 12} (x_j - \bar x)^2}
\end{aligned}
}
$$
:::


## Are We Successful?

- Let's characterize the *error* of our estimates.
- That is, let's characterize how far off they are from the population summaries.
- We'll calculate their bias and variance.


# Bias

## Fine-Group Subsample Means are Unbiased

- This is the same proof we did last time, but with fancier notation.
- We'll show something stronger than unbiasedness: unbiasedness *conditional on the covariates*.

::: fragment
$$
\begin{aligned}
\E\qty[\hat\mu(x) \mid X_1 \ldots X_n] 
&= \E\qty[\frac{\sum_{i:X_i=x} Y_i}{\sum_{i:X_i=x} 1} \mid X_1 \ldots X_n]  && \\
&= \E\qty[\frac{\sum_{i=1}^n 1_x(X_i) Y_i}{\sum_{i=1}^n 1_x(X_i)} \mid X_1 \ldots X_n] && \qfor 1_x(X) = \begin{cases} 1 & \text{ if } x = X  \\ 0 & \text{ otherwise} \end{cases} \\
&= \frac{\sum_{i=1}^n 1_x(X_i) \E[ Y_i \mid X_1 \ldots X_n]}{\sum_{i=1}^n 1_x(X_i)} && \text{ by linearity of conditional expectations } \\
&= \frac{\sum_{i=1}^n 1_x(X_i) \E[ Y_i \mid X_i]}{\sum_{i=1}^n 1_x(X_i)}          && \text{  by irrelevance of independent conditioning variables } \\
&= \frac{\sum_{i=1}^n 1_x(X_i) \mu(X_i)}{\sum_{i=1}^n 1_x(X_i)}                     && \text{  is a matter of notation } \\
&= \frac{\sum_{i=1}^n 1_x(X_i) \mu(x)}{\sum_{i=1}^n 1_x(X_i)}                       && \text{ by the \emph{indicator trick} } 1_x(X_i)\mu(X_i) = 1_x(X_i)\mu(x) \\
&= \mu(x) \frac{\sum_{i=1}^n 1_x(X_i)}{\sum_{i=1}^n 1_x(X_i)} = \mu(x) 
\end{aligned}
$$
:::

Why does this imply (unconditional) unbiasedness?

::: fragment

The law of iterated expectations.

$$ \E[\hat\mu(x)]  = \E\qty[ \E\qty[\hat\mu(x) \mid X_1 \ldots X_n]  ] = \E\qty[ \mu(x) ] = \mu(x) $$

:::


## Linear Combinations **With Non-Random Coefficients** are Unbiased

$$ \E\qty[\sum_x \hat\alpha_x \hat\mu(x)] = \sum_x \alpha_x \mu(x) \quad \text{if} \hat\alpha_x = \alpha_x $$

- Examples
  - $\hat \mu(18) - \hat\mu(16)$ is unbiased for $\mu(18) - \mu(16)$.
  - $\frac14 \sum_{x=9}^{12} \qty{ \hat\mu(x) - \hat\mu(x-1) }$ is unbiased for $\frac14 \sum_{x=9}^{12} \qty{ \mu(x) - \mu(x-1) }$.
- Why?


::: {.fragment}

Linearity of Expectations and unbiasedness of the subsample means.

$$
\E\qty[\sum_x \alpha_x \hat\mu(x)] = \sum_x \alpha_x \E\qty[\hat\mu(x)] = \sum_x \alpha_x \mu(x)
$$

:::

## Linear Combinations **With Random Coefficients** are Unbiased ...

- ... if the sample coefficients $\hat\alpha(x)$ depend only on the covariates $X_1 \ldots X_n$ ...
- and are unbiased estimators of the population coeffients $\alpha_x$.

$$ \E\qty[\sum_x \hat\alpha_x \hat\mu(x)] = \sum_x \alpha_x \mu(x) \quad \text{if} \E[\hat\alpha_x] = \alpha_x $$

- Examples
  - Proportions $P_x$ are unbiased for the population proportions $p_x$.
  - Differences in Proportions $P_x - P_{x-1}$ are unbiased for the population differences $p_x - p_{x-1}$.
- And consequently 
  - Aggregate-group subsample means are unbiased.
  - And so is our averaged-over-people value of a year of high school.
- The proof is *almost* the same as before. 
  - But now we use linearity of *conditional expectations*.
  - And *conditional unbiasedness* of the subsample means.


::: {.fragment}
$$
\begin{aligned}
\E\qty[\sum_x \alpha_x \hat\mu(x)]   
&= \E\qty[ \E\qty{ \sum_x \alpha_x \hat\mu(x) \mid X_1 \ldots X_n } ] && \text{via law of iterated expectations} \\
&= \E\qty[ \sum_x \alpha_x \E\qty{ \hat\mu(x) \mid X_1 \ldots X_n } ] && \text{via linearity of conditional expectations} \\
&= \E\qty[ \sum_x \alpha_x \mu(x) ]                                   && \text{via conditional unbiasedness of the subsample means} \\
&= \sum_x \E[\hat\alpha_x] \mu(x) = \sum_x \alpha_x \mu(x)            && \text{via linearity of expectations}  
\end{aligned} 
$$
:::


## Summary

- We've seen that our population summaries have natural unbiased estimators: the corresponding summaries of the sample.
  - With the possible exception of the least-squares slope, which we haven't talked about. We'll get to that later in the semester.
- That means that our ability to estimate those population summaries comes down to the *variance* of those sample summaries.
- That's what we'll talk about next.


# Variance

## Variance Calculation Step 1: Using the Law of Total Variance


- Let's talk about the variance of a linear combination of the sample means.
- We'll use the law of total variance to break the variance of our estimator into two parts.
- When the coefficients in our combination are non-random, one of these --- the $\Var\E[\ldots]$ part --- is zero.
  - This happens because the conditional expectation of our subsample means are constants
  - We've seen a special case of this earlier, when discussing differences of subsample means.
- More generally, we'll have to deal with both terms.

$$
\begin{aligned}
\Var\qty[ \sum_x \hat\alpha_x \hat \mu(x) ] 
&= \E\qty[ \Var\qty{ \sum_x \hat\alpha_x \hat \mu(x) \mid X_1 \ldots X_n }] + \Var\qty[ \E{\sum_x \hat\alpha_x \hat \mu(x) \mid X_1 \ldots X_n} ] \\
&= \E\qty[ \Var\qty{ \sum_x \hat\alpha_x \hat \mu(x) \mid X_1 \ldots X_n }] + \Var\qty[ \sum_x \hat \alpha_x \E{ \hat \mu(x) \mid X_1 \ldots X_n} ] \\
&= \E\qty[ \Var\qty{ \sum_x \hat\alpha_x \hat \mu(x) \mid X_1 \ldots X_n }] + \Var\qty[ \sum_x \hat \alpha_x \mu(x) ] \\
\end{aligned}
$$


## Variance Calculation Step 2: The $\E[\Var[\ldot]]$ Term.

- What we get for the $\E[\Var[\ldots]]$ part is familiar.
  - We *sum* the conditional variances of the individual terms, then take the expectation of the result.
  - This is what we did when we calculated the variance of the difference of subsample means.

$$\small{
\begin{aligned}
\E\qty[ \Var\qty{ \sum_x \hat\alpha_x \hat \mu(x) \mid X_1 \ldots X_n }] 
&=\E\qty[ \E\qty{ \qty(\sum_x \hat\alpha_x \hat\mu(x) - \E[\sum_x \hat\alpha_x \hat\mu(x) \mid X_1 \ldots X_n])^2 \mid X_1 \ldots X_n }] \\
&=\E\qty[ \E\qty{ \qty(\sum_x \hat\alpha_x \hat\mu(x) - \hat\alpha_x \E\qty[\hat\mu(x) \mid X_1 \ldots X_n])^2 \mid X_1 \ldots X_n }] \\
&=\E\qty[ \E\qty{ \qty(\sum_x \hat\alpha_x \qty{\hat\mu(x) - \E[\hat\mu(x) \mid X_1 \ldots X_n]})^2 \mid X_1 \ldots X_n }] \\
&=\E\qty[ \E\qty{ \sum_x\sum_x' \hat\alpha_x \hat\alpha x' \qty{\hat\mu(x) - \E[\hat\mu(x) \mid X_1 \ldots X_n]}  \qty{\hat\mu(x) - \E[\hat\mu(x) \mid X_1 \ldots X_n]} \mid X_1 \ldots X_n }] \\ 
&=\E\qty[ \sum_x\sum_x' \hat\alpha_x \hat\alpha x' \E\qty{\qty{\hat\mu(x) - \E[\hat\mu(x) \mid X_1 \ldots X_n]}  \qty{\hat\mu(x) - \E[\hat\mu(x) \mid X_1 \ldots X_n]} \mid X_1 \ldots X_n }] \\ 
&=\E\qty[ \sum_x\sum_x' \hat\alpha_x \hat\alpha x' \begin{cases} \Var\qty{\hat\mu(x) \mid X_1 \ldots X_n} = \frac{\sigma^2(x)}{N_x} & \text{ if } x=x' \\ 0 & \text{ otherwise} \end{cases} ] \\
&=\E\qty[ \sum_x \hat \alpha_x^2 \frac{\sigma^2(x)}{N_x} ] = \sum_x \sigma^2(x) \times \E\qty[ \frac{\hat\alpha_x^2}{N_x} ] \\
\end{aligned}
}
$$

## Variance Calculation Step 3: The $\Var[\E[\ldot]]$ Term.


## Summary

$$
\Var[ \sum_x \alpha_x \hat\mu(x) ] = \sum_x \sigma^2(x) \times \E\qty[ \frac{\hat\alpha_x^2}{N_x} ] +  \Var\qty[ \sum_x \hat \alpha_x \mu(x) ]
$$


# Implications

## Some things are *easier to estimate* than others.

TODO: A picture of corresponding interval estimates

- Compare variance of the average over years vs average over population things.


