---
title: "QTM 220 Lecture 10"
subtitle: "Inferential Statistics with Non-Binary Covariates: Part 1"
format: 
  clean-revealjs:
    height: 1000
    width: 1290
    theme: [default, "class.scss"]
    smaller: true
    slide-number: true
    pdf-separate-fragments: true
    fig-format: svg
    html-math-method: mathjax
    include-in-header: 
      text: |
        <script>
          window.MathJax = {
          loader: {load: ['[tex]/physics']},
          tex: {packages: {'[+]': ['physics']}}
          }
    
        </script>
knitr:
  opts_chunk:
    fig.path: generated-figures/
    dev.args:
      bg: transparent
---

```{r load-libraries}
library(purrr)
library(tibble)
library(dplyr)
library(magrittr)
library(stringr)
library(ggplot2)
library(htmltools)
library(kableExtra)
library(scales)
library(cpsR)
library(mapdata)

options(knitr.kable.NA = '')

compact_number = label_number(accuracy=.1, scale_cut=cut_short_scale())
ellipsis <- function(df, nhead=5L, ntail=2) {
    rownames(df) = as.character(1:nrow(df))
    stopifnot("data.frame" %in% class(df))
    els <- rep(NA, ncol(df)) %>% 
        matrix(nrow=1, dimnames=list(NULL, names(df))) %>% 
        data.frame(stringsAsFactors=FALSE) %>% 
        tibble::as_tibble()
    rownames(els) = "⋮"
    dplyr::bind_rows(head(df, nhead), els, tail(df, ntail))
}

percent = label_percent(accuracy=1)
approx  = function(x,accuracy=.01) { label_number(accuracy=accuracy, scale_cut=cut_short_scale())(x) }
```

```{r ggplot-themes}
fontsize = 16
textcolor = "#888888"

class.theme = theme(plot.background = element_rect(fill = "transparent", colour = NA),
        panel.background = element_rect(fill = "transparent", colour = NA),
                    legend.background = element_rect(fill="transparent", colour = NA),
                    legend.box.background = element_rect(fill="transparent", colour = NA),
                    legend.key = element_rect(fill="transparent", colour = NA),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x  = element_text(colour = textcolor, size=fontsize),
        axis.text.y  = element_text(colour = textcolor, size=fontsize),
        axis.title.x  = element_text(colour = textcolor),
        axis.title.y  = element_text(colour = textcolor, angle=90)
)

theme_set(class.theme)
theme_update(axis.title.x  = element_blank(),
             axis.title.y  = element_blank())
simplified.theme = theme_get()

theme_update(axis.text.x   = element_blank(), 
              axis.text.y   = element_blank())
blank.theme = theme_get()

theme_update(panel.grid.major=element_line(color=rgb(1,0,0,.1,  maxColorValue=1)),
           panel.grid.minor=element_line(color=rgb(0,1,0,.1, maxColorValue=1)))
grid.theme = theme_get()

theme_set(simplified.theme)
```




```{r get-data}
#| cache: true
#| 
# for cps variable interpretation see https://www2.census.gov/programs-surveys/cps/datasets/2022/march/asec2022_ddl_pub_full.pdf
census.api.key = '8084f81e4f71e4d32638dac15f188b424f72d063'
cps.sampling.ratio = 1/25 # Not true. 1/2500 households are surveyed. But for the sake of plotting, it's ok.

california.state.code = 6
year = 2020
cps.data = results=get_asec(year=year, vars=c('pearnval', 'a_age', 'a_hga', 'gestfips', 'gtco', 'gtcbsasz'), key=census.api.key)
```

```{r transform-data}
#| cache: false
#| 
all.earnings = data.frame(state.code = cps.data$gestfips,
                          county.code = cps.data$gtco,
                          age=cps.data$a_age,
                          income=cps.data$pearnval,
                          education.code=cps.data$a_hga,
                          education=case_match(cps.data$a_hga,
                                           0  ~ 0,   # child
                                           31 ~ 0,   # < grade 1
                                           32 ~ 4,   #  grade 1-4
                                           33 ~ 6,   #  grade 5-6
                                           34 ~ 8,   #  grade 7-8 
                                           35 ~ 9,   #  grade 9
                                           36 ~ 10,  #  grade 10
                                           37 ~ 11,  #  grade 11
                                           38 ~ 11,  #  grade 12 no diploma
                                           39 ~ 12,  #  high school grad
                                           40 ~ 13,  #  some college
                                           41 ~ 14,  #  associate degree (vocational)
                                           42 ~ 14,  #  associate degree (academic)
                                           43 ~ 16,  #  bachelors degree
                                           44 ~ 18,  #  masters degree
                                           45 ~ 20,  #  professional school degree
                                           46 ~ 20)) #  doctorate

all.earnings$degree = factor(all.earnings$education >= 16, 
                     levels=c(FALSE, TRUE), 
                     labels=c('no 4-year degree','4-year degree')) 

coarsen = function(x, group, seed=0) {
  current.seed = .Random.seed
  set.seed(seed)
  x.coarsened = rep(NA, length(x))
  for(gg in unique(group)) {
    x.coarsened[group==gg] = runif(sum(group==gg), min(x[group==gg]), max(x[group==gg]))
  }
  set.seed(current.seed)
  x.coarsened
} 

in.interval = function(X, interval) { interval[1] <= X & X <= interval[2] }
earnings = filter(all.earnings, education >= 8 & 
                                state.code == california.state.code & 
                                age %>% in.interval(c(25,35)))
Y = earnings$income
X = earnings$education
```


```{r get-county-info}
# county info from https://en.wikipedia.org/wiki/List_of_counties_in_California
# converted to csv using https://wikitable2csv.ggor.de

to.number = function(s) { as.numeric(str_trim(gsub(",", "", s))) }
ca_county_info = read.csv('california_county_info.csv') %>%
  mutate(subregion = tolower(sub(' County', '', County))) %>%
  mutate(pop = to.number(Population..2022.)) %>%
  mutate(area = to.number(sub('sq.*', '', Area))) 

earnings.county = left_join(earnings, ca_county_info[,c('FIPS.code', 'subregion')], by=c('county.code' = 'FIPS.code'))
earnings$county = earnings.county$subregion
earnings$county[is.na(earnings$county)]='unknown'
```


```{r make-maps}
#| cache: true
#| 

# To give our survey recipients a bit more reality, we'll give them fake locations on a map
# And we'll make up a population of California residents that they're sampled from
# maps are based on code from https://eriqande.github.io/rep-res-eeb-2017/map-making-in-R.html

counties = map_data("county")
ca_county = counties %>%
  filter(region == "california")
ca_county_centers = ca_county %>%
  group_by(subregion) %>%
  summarize(county.lat = mean(lat), county.long=mean(long))

ca_county_more = left_join(ca_county_centers, ca_county_info, by="subregion")
ca_all = left_join(ca_county, ca_county_more, by="subregion")


ca.plot = ggplot(data=ca_all, mapping=aes(x=long, y=lat, group=group)) + 
  coord_quickmap() +
  geom_polygon(fill='gray', color = "white", alpha=.4) +
      scale_fill_gradient(trans="log10") + 
      theme_void()

ca.pop.plot = ggplot(data=ca_all, mapping=aes(x=long, y=lat, group=group)) + 
  coord_quickmap() +
  geom_polygon(aes(fill = pop/area), color = "white") +
      scale_fill_gradient(trans="log10") + 
      theme_void()

## we'll make up a population as follows
# 1. choose the number of people to be the number of people in our sample scaled up to account for the cps sampling ratio 1/2500
#    Because this would wind up being computationally and visually problematic, we'll scale up by 25 instead of 2500;
#    each population dot in the population stands for 100 people in the population.
# 2. for each, sample a county with probability proportional to its finite.population
# 3. for each, sample a latitude and longitude by taking a randomly weighted combination of the    
#    county's vertices. As long as the county is convex, these points will be inside the county.
# 4. we'll give them a degree too, chosen independently of all else from the empirical distribution on the sample
m = nrow(earnings) * 25 

# takes a list of subregions, possibly with duplicates
# for each element, chooses random lat/long coordinates within the sample.subregion
# by taking a random convex combination of the subregion's vertices
sample.within.subregion = function(subregions) {
  output = data.frame(subregion=subregions, lat=NA, long=NA)
  for(subregion in unique(subregions)) { 
      vertices = ca_county[ca_county$subregion==subregion, c('lat','long')]
      U = sum(subregions==subregion)
      V = nrow(vertices)
      weights = abs(rt(U*V, 1))
      dim(weights) = c(U,V)
      weights = (1/rowSums(weights)) * weights
      output[subregions==subregion, c('lat', 'long')] = 
                 cbind(weights %*% vertices$lat, weights %*% vertices$long)
    }
    output
}

finite.population.subregion = sample(ca_county_more$subregion, size=m, replace=TRUE, prob=ca_county_more$pop)
finite.population = sample.within.subregion(finite.population.subregion)
finite.population$degree = sample(earnings$degree, nrow(finite.population), replace=TRUE)
finite.population$group=1 # for plotting

display.population = finite.population
```


```{r make-sample}
# and we'll force the degree-distribution in the sample to match our survey responses
our.sample = sample_n(finite.population, nrow(earnings))
our.sample$degree=earnings$degree

pop.plot = ca.plot + geom_point(aes(x=long, y=lat), data=display.population, size=.02, alpha=.1) 
sample.plot = ca.plot + geom_point(aes(x=long, y=lat), data=our.sample, size=.2, alpha=.5)
pop.plot.inf = ca.pop.plot + geom_point(aes(x=long, y=lat), data=display.population, size=.01, alpha=.1)  
sample.plot.inf = ca.pop.plot + geom_point(aes(x=long, y=lat), data=our.sample, size=.2, alpha=.5)
```

```{r visualization-defaults}
scale_y = function(limits=c(-.3e5, 3e5)) { 
  scale_y_continuous(labels=label_dollar(), limits=limits)
}

years = 8:20
scale_x = function(limits=c(min(years)-.5, max(years)+.5)) { 
  scale_x_continuous(breaks=years, labels=sprintf('%dy', years), limits=limits)
}

dot.size = 2
```


```{r define-colors}
twocolor.red   = 'rgb(248,118,109)'
twocolor.green = 'rgb(0,191,196)'
```

## Last Time

::: {.hidden}
$$
% color defs aren't making it into MathJax
%\definecolor{twocolor-red}{RGB}{248, 118, 109}   % red   used by ggplot in plots w/ 2 colors 
%\definecolor{twocolor-green}{RGB}{0, 191, 196}   % green used by ggplot in plots w/ 2 colors
% so create commands for using those colors instead
\newcommand{\inred}[1]{\textcolor[RGB]{248,118,109}{#1}}
\newcommand{\ingreen}[1]{\textcolor[RGB]{0,191,196}{#1}}
\newcommand{\inblue}[1]{\textcolor{blue}{#1}}

\newcommand{\ind}[1]{1_{\{#1\}}}
\newcommand{\with}{\ingreen{\text{degree}}}
\newcommand{\without}{\inred{\text{no degree}}}

\DeclareMathOperator{\E}{E}degredegre
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\Cov}{Cov}
$$ 

:::

```{r scatter-plots}

earnings.jitter = ggplot(earnings) + geom_point(aes(x=education, y=income), color='#F7766D',
                              size=dot.size, position=position_jitter(seed=0,w=.3), alpha=.1) +
                   scale_x() + scale_y() + guides(color=FALSE)

summaries = earnings %>% 
          group_by(education) %>% 
          summarize(mean=mean(income),
                    sd=sd(income))
                    
```


```{r}
#| out-height: 300px
#| fig-asp: .3

earnings.jitter
```

- We talked summaries of the relationship between income and education *in a sample*.
- We talked about a number of types of summaries. 
- What you'd use depends on what you're interested in.


## Three Types

```{r}
#| layout-ncol: 3
#| out-height: 400px
#| fig-asp: 1
#| fig-subcap:
#|  - A Fine Group Comparison $$\hat\mu(16) - \hat\mu(12)$$
#|  - An Aggregate Group Comparison $$\hat\mu(16) - \hat\mu(\le 12)$$
#|  - An Aggregated Comparison $$\frac{1}{4}\sum_{x= 9}^{12} \qty{ \hat\mu(x) - \hat\mu(x-1) } $$



ggplot(earnings) + geom_point(aes(x=education, y=income, alpha=education %in% c(16,12), color=education > 12), 
                                  size=dot.size, position=position_jitter(seed=0,w=.3)) + 
                   geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries[summaries$education %in% c(16,12), ]) +
                   scale_x() + scale_y() + scale_alpha_manual(breaks=c(FALSE, TRUE), values=c(.01,.3)) + guides(color=FALSE, alpha=FALSE) 

coarsened.summaries = earnings[earnings$education %in% c(16,8:12),] %>% 
  group_by(education <= 12) %>% 
  summarize(mean=mean(income),
            sd=sd(income),
            mean.X = mean(education))

ggplot(earnings) + geom_point(aes(x=education, y=income, alpha=education %in% c(16,8:12), color=education > 12), 
                                  size=dot.size, position=position_jitter(seed=0,w=.3)) + 
                   geom_pointrange(aes(x=mean.X, y=mean, ymin=mean-sd, ymax=mean+sd), data=coarsened.summaries) +
                   scale_x() + scale_y() + scale_alpha_manual(breaks=c(FALSE, TRUE), values=c(.01,.3)) + guides(color=FALSE, alpha=FALSE) 

earnings.jitter.hs = ggplot(earnings) + geom_point(aes(x=education, y=income, alpha=education <= 12), color='#F7766D',
                              size=dot.size, position=position_jitter(seed=0,w=.3)) +
                   scale_x() + scale_y() + scale_alpha_manual(breaks=c(FALSE, TRUE), values=c(.01,.3)) + guides(color=FALSE, alpha=FALSE)

aggregate.comparison.plot = earnings.jitter.hs +
  geom_pointrange(aes(x=education, y=mean, ymin=mean-sd, ymax=mean+sd), data=summaries[summaries$education <= 12,]) +
  geom_line(aes(x=education, y=mean), data=summaries[summaries$education<=12,], color='black', linetype='dashed') +
  scale_y(c(0, .7e5))

aggregate.comparison.plot
```



## Today

```{r}
#| out-height: 400px
#| fig-asp: .3

education.to.summary.row = rep(NA, max(summaries$education))
education.to.summary.row[summaries$education] = 1:nrow(summaries)
muhat = function(x) { 
  summaries$mean[education.to.summary.row[x]]
}
increments = muhat(9:12)-muhat(8:11)
average.increment = mean(increments)

hs.subsample = earnings$education[earnings$education %in% 9:12]
sample.average.increment = mean(muhat(hs.subsample) - muhat(hs.subsample-1))
sai.data = data.frame(education=c(8,12), mean=c(muhat(12)-sample.average.increment*4, muhat(12)))

aggregate.comparison.plot + xlim(c(7.5,12.5))
```


- We'll talk about *statistical inference* based on these sample summaries.
- That is, we'll talk about what they tell us about the population our sample was drawn from.
- To keep us focused, we'll talk mostly about two different *aggregated comparisons*.
  - Both are summaries of the value of a year of high school. 
  - They're different averages of the year-on-year increments in mean income.


## Today

```{r}
#| out-height: 400px
#| fig-asp: .3
aggregate.comparison.plot + xlim(c(7.5,12.5))
```

1. The average, over the four years of high school, of the increments.
$$
\frac{1}{4}\sum_{x=9}^{12} \qty{ \hat\mu(x) - \hat\mu(x-1) } = \frac{\hat\mu(12) - \hat\mu(8)}{4}
$$

2. The average, over people with &ge;1 year of high school, of the increments *affecting them* in their last year. 
$$
\frac{\sum_{i:X_i \in 9 \ldots 12} \qty{ \hat\mu(X_i) - \hat\mu(X_i-1) }}{\sum_{i:X_i \in 9 \ldots 12} 1} 
=  \sum_{x \in 9 \ldots 12} P_x \ \qty{ \hat\mu(x) - \hat\mu(x-1) } \quad \text{ for } \quad P_x = \frac{N_x}{\sum_{x \in 9 \ldots 12}N_x}.
$$




## Why This is Enough

- These aren't the only summaries we might be interested in. We talked about many more last time. 
- And we saw that all were linear combinations --- coefficient weighted sums --- of the subsample means.
- Some, like the average over years, involved coefficients that didn't depend on the covariates.

$$
\small{
\frac14\sum_{x=9}^{12} \qty{ \hat\mu(x) - \hat\mu(x-1)} = \sum_x \alpha(x) \hat \mu(x)  \qfor \alpha(x) = \begin{cases} \frac14 & \text{ if } x = 12 \\ -\frac14 & \text{ if } x = 8 \\ 0 & \text{ otherwise} \end{cases}
}
$$

- Others, like the average over people with &ge;1 year of high school, involved coefficients that did.

$$
\small{
\sum_{x=9}^{12} P_x \qty{ \hat\mu(x) - \hat\mu(x-1)} = \sum_x \hat \alpha(x) \hat \mu(x) \qfor
\hat\alpha(x) = \begin{cases}
  P_{12} & \text{ if } x = 12 \\
  P_{x} - P_{x+1} & \text{ if } x \in \{9 \ldots 11\} \\
  -P_9 & \text{ if } x = 8 
\end{cases}
}
$$

- We'll tackle these two general cases, using our two examples as illustrations.


# What We're Doing

## Our Data

```{r}
#| out-height: 700px
#| fig-ncol: 2
sample.plot
pop.plot
```



- As before, we'll work with data from the 2022 Current Population Survey.
- We'll look at California residents age 25-35 with at least an 8th-grade education.
  - This sample includes `r nrow(earnings)` people. 
  - We'll think of them as having been sampled with replacement from a population.
  - That is, from the population of all California residents meeting these criteria.
  - That's an approximation---the CPS is a complex survey---but it's not too bad.

::: aside
As discussed earlier in the course, these locations are fictitious. The maps are just an illustration of the relationship between sample and population.
:::

## What We're Estimating {.smaller}

```{r}
#| out-height: 250px
#| layout-ncol: 2
#| cache: true


earnings.jitter = ggplot(earnings) + geom_point(aes(x=education, y=income), color='#F7766D',
                              size=dot.size, position=position_jitter(seed=0,w=.3), alpha=.1) +
                   scale_x() + scale_y() + guides(color=FALSE)


# make a fake population by sampling with replacement from the sample then adding fuzz to incomes
# the actual sampled proportion in CPS is about 1/2500, so by doing 25x more dots,
# we're letting each dot  stand in for 100 people in the population
earnings.pop = earnings[sample(1:nrow(earnings), nrow(earnings)*25, replace=TRUE),]
earnings.pop$income = earnings.pop$income + rnorm(nrow(earnings.pop), sd=5000)
earnings.jitter.pop = ggplot(earnings.pop) + geom_point(aes(x=education, y=income), color='#F7766D',
                              size=dot.size/10, position=position_jitter(seed=0,w=.3), alpha=.05) +
                   scale_x() + scale_y() + guides(color=FALSE)

earnings.jitter
earnings.jitter.pop
```


:::: columns

::: column
### Sample
$$
\small{
\begin{array}{ccccccc}
X_1 & Y_1 & X_2 & Y_2 &  \ldots & X_n & Y_n \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\end{array}
}
$$

:::
::: column
### Population
$$
\small{
\begin{array}{ccccccc}
x_1 & y_1 & x_2 & y_2 &  \ldots & x_m & y_m \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\end{array}
}
$$

:::
::::

- Qualitatively, the sample and the population are the same kind of thing. 
- They're both collections of people with incomes and education levels. 
- That means we can summarize them in the same ways.  
- It's natural to think of sample summaries as *estimates* of the corresponding population summaries.

$$
\small{
\begin{aligned}
&\hat \mu(x) = \frac{1}{N_x}\ \sum_{i:X_i=x} Y_i \qfor N_x = \sum_{i:X_i=x} 1 
&& \text{ estimates } \quad \mu(x) = \frac{1}{m_x}\ \sum_{j:x_j=x} y_j \qfor m_x = \sum_{j:x_j=x} 1 \\
&\frac{1}{4}\sum_{x=9}^{12} \qty{ \hat \mu(x) - \hat\mu(x-1) }                       
&& \text{ estimates } \quad \frac{1}{4}\sum_{x=9}^{12} \qty{ \mu(x) - \mu(x-1) }   \\
&\sum_{x=9}^{12} P_x \qty{ \hat \mu(x) - \hat\mu(x-1) } \qfor P_x = \frac{N_x}{\sum_{x=9}^{12} N_x}                       
&& \text{ estimates } \quad \sum_{x=9}^{12} p_x \qty{ \mu(x) - \mu(x-1) }   \qfor p_x = \frac{m_x}{\sum_{x=9}^{12} m_x} \\
\end{aligned}
}
$$


## What We Want To Know

```{r}
#| cache: true

average.over.people = function(summaries) {
  muhat = function(x) { summaries$mean[summaries$education==x] }
  P = function(x) { summaries$N[summaries$education==x] / sum(summaries$N) }
  alphahat = function(x) { if (x==12) { P(12) } else if (x==8) { -P(9) } else { P(x) - P(x+1) } }  
  sapply(8:12, function(x) { alphahat(x) * muhat(x) }) %>% sum
}
run.boot = function(earnings, B=1000) { purrr::map(1:B, function(i) {
  earnings.boot = earnings[sample(1:nrow(earnings), replace=TRUE),]
  summaries = earnings.boot[earnings.boot$education <= 12, ] %>% group_by(education) %>% summarize(mean=mean(income), N=length(income))
  model = lm(income ~ education, data=earnings.boot[earnings.boot$education <= 12,])
  data.frame(people=average.over.people(summaries), 
             years=(summaries$mean[summaries$education==12] - summaries$mean[summaries$education==8])/4,
             line = coef(model)[2])
}) %>% bind_rows()
}
boots = run.boot(earnings)
summaries = earnings[earnings$education <= 12, ] %>% group_by(education) %>% summarize(mean=mean(income), N=length(income))
years.estimate = (summaries$mean[summaries$education==12] - summaries$mean[summaries$education==8])/4
```

```{r}
#| layout-ncol: 2
#| out-height: 300px
#| fig-asp: .3 
#| fig-subcap:
#|  - Interval Estimate for the Average over Years
#|  - Interval Estimate for the Average over People 

width = function(samples, center=mean(samples), alpha=.05) {
  covg.dif = function(w) { 
    actual = mean( center - w/2 <= samples & samples <= center + w/2)
    nominal = 1-alpha 
    actual - nominal  
  }
  uniroot(covg.dif, interval=c(0, 2*max(abs(samples-center))))$root
}

distribution.plot = function(boot.samples, point.estimate, w=width(boot.samples), interval.replications=0, point.replications=0) { 
  drange = c(0, dnorm(0, mean=0, sd=sd(boot.samples)))
  p = ggplot(data.frame(samples=boot.samples)) + 
    geom_histogram(aes(x=samples, y=after_stat(density)), alpha=.3, bins=50)  +
    geom_pointrange(aes(y=dnorm(2*sd(samples), mean=0, sd=sd(samples)), 
      x=point.estimate, xmin=point.estimate - w/2, xmax=point.estimate + w/2)) + 
      geom_vline(xintercept=mean(boot.samples), color='blue') + 
      geom_vline(xintercept=quantile(boot.samples, c(.025,.975)), color='blue', linetype='dotted') 
  if(interval.replications > 0) { 
    K = interval.replications
    p = p + geom_pointrange(aes(y=y, x=x, xmin=xmin, xmax=xmax), 
          data=data.frame(y=seq(min(drange), max(drange), length.out=K), 
                  x=boot.samples[1:K], xmin=boot.samples[1:K] - w/2, xmax=boot.samples[1:K] + w/2), color='purple', alpha=.2)
  }
  if(point.replications > 0) { 
    K = point.replications
    p = p + geom_point(aes(y=y, x=x), 
          data=data.frame(y=seq(min(drange), max(drange), length.out=K), 
                  x=boot.samples[1:K], xmin=boot.samples[1:K] - w/2, xmax=boot.samples[1:K] + w/2), color='purple', alpha=.2)
  }
  p
}


distribution.plot(boots$years, years.estimate) + xlim(-5000,10000)
distribution.plot(boots$people, average.over.people(summaries)) + xlim(-5000,10000)
```

1. How accurate our estimates tend to be. 
2. How to characterize their accuracy, e.g. with a confidence interval. 
 
\ \

- We do this exactly the same way we've done it before. 
  1. We estimate the sampling distribution of our estimator.
  2. We draw arms, legs, and a head on it.

::: fragment
  - That was what Github Copilot --- GPT4 --- suggested after I typed 'We'll draw arms'.
  - I didn't want to  keep that one to myself.
:::

## What We Want To Know
```{r}
#| layout-ncol: 2
#| out-height: 300px
#| fig-asp: .3 
#| fig-subcap:
#|  - Interval Estimate for the Average over Years
#|  - Interval Estimate for the Average over People 

distribution.plot(boots$years, years.estimate) + xlim(-5000,10000)
distribution.plot(boots$people, average.over.people(summaries)) + xlim(-5000,10000)
```


1. How accurate our estimates tend to be. 
2. How to characterize their accuracy, e.g. with a confidence interval. 
 
\ \

- We do this exactly the same way we've done it before. 
  1. We estimate the sampling distribution of our estimator.
  2. We draw arms on our point estimate wide enough to give it 95% coverage.
- And we see something interesting. 
- Our estimate of the average over years is much more precise. 
  - Its arms are much shorter.


## Review: How Interval Estimation Works.
```{r}
#| out-height: 300px
#| fig-asp: .3 

distribution.plot(boots$years, years.estimate, interval.replications=100) + xlim(-2000,8000)
```

- These arms should be wide enough hat ...
  - They'd cover 95% of draws from the sampling distribution
  - if drawn from the sampling distribution's mean.
- If the mean's arms reach the point estimate, then the point estimate's arms reach the mean.

## We'll Talk About Two Ways of Estimating the Sampling Distribution

```{r}
#| out-height: 300px
#| fig-asp: .3 

x = seq(min(boots$years), max(boots$years), length.out=1000)
distribution.plot(boots$years, years.estimate) + 
    geom_area(aes(x=x, y=dnorm(x, mean=mean(boots$years), sd=sd(boots$years))), color='purple', fill='purple', alpha=.08) +
    geom_vline(aes(xintercept=mean(boots$years)), color='purple', linetype='solid') +
    geom_vline(aes(xintercept=xi), data=data.frame(xi=qnorm(c(.025,.975), mean=mean(boots$years), sd=sd(boots$years))), color='purple', linetype='dotted') +
    xlim(-2000,8000)
```


- The bootstrap.
  - This is easy if you have a computer. 
  - There's a procedure---you just do it.  

- Normal approximation
  - This is, in a way, harder.
    - You have to work out formulas for the estimator's mean and variance.
    - Then estimate the stuff in them, e.g. subpopulation variances.
  - We'll do it because having formula will give us a better sense of what's going on. 
    - It'll tell us why one of our population summaries is easier to estimate than the other.
    - i.e., why we get a narrower interval around the corresponding sample summary.


## The Bootstrap Approach

```{r}
#| out-height: 300px
#| fig-asp: .3

distribution.plot(boots$years, years.estimate, point.replications=1000) + xlim(-2000,8000) 
```

- Calculate your summary of the sample. That's your point estimate of the population summary.
- To estimate its sampling distribution ...
  - Draw $B$ samples with replacement from the sample.
  - Calculate your summary in each of these $B$ *bootstrap samples*.
  - Histogramming these gives you an estimate of the sampling distribution.
- Use it to calculate a 95% interval around your point estimate.
  - The 'arms' of the interval should be wide enough to extend ...
  - from your estimate to cover 95% of your bootstrap samples.

:::: columns

::: column
### Bootstrap Sample

$$
\small{
\begin{array}{ccccccc}
X_1^* & Y_1^* & X_2^* & Y_2^* &  \ldots & X_n^* & Y_n^* \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\end{array}
}
$$

:::

::: column

### Sample
$$
\small{
\begin{array}{ccccccc}
X_1 & Y_1 & X_2 & Y_2 &  \ldots & X_n & Y_n \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
\end{array}
}
$$

:::
::::

## The Normal Approximation Approach
```{r}
#| out-height: 300px
#| fig-asp: .3

point.estimate = years.estimate
w = 2*1.96*sd(boots$years) 
ggplot() + 
    geom_area(aes(x=x, y=dnorm(x, mean=mean(boots$years), sd=sd(boots$years))), color='purple', fill='purple', alpha=.08) +
    geom_vline(aes(xintercept=mean(boots$years)), color='purple', linetype='solid') +
    geom_vline(aes(xintercept=xi), data=data.frame(xi=qnorm(c(.025,.975), mean=mean(boots$years), sd=sd(boots$years))), color='purple', linetype='dotted') +
    geom_pointrange(aes(y=dnorm(2*sd(boots$years), mean=0, sd=sd(boots$years)), x=point.estimate, xmin=point.estimate - w/2, xmax=point.estimate + w/2)) + 
    xlim(-2000,8000)
```

1. We'll show that our estimator is unbiased

$$ 
\E[\hat\theta] = \theta
$$

- This implies our sampling distribution is *in the right place*.

2. We'll calculate --- and then estimate --- their *standard errors*.
$$
\hat \sigma_{\hat\theta} \approx \sqrt{ \Var[\hat\theta] }
$$

3.  And we'll use normal approximation to construct *confidence intervals*.
$$
\theta \in \qty[ \hat \theta - 1.96\hat\sigma_{\hat\theta}, \ \hat \theta + 1.96\hat\sigma_{\hat\theta}] 
$$


## What's New

- Not that much. This is just like what we did earlier in the semester. 
- What's new is that we'll be doing it for linear combinations of the subsample means.
- This means we'll need new unbiasedness proofs and variance calculations.
  - Today, we'll do the the simpler case: linear combinations with non-random coefficients.
  - Next time, we'll generalize to linear combinations with random coefficients.


# Inference for Linear Combinations of Subsample Means: Non-Random Coefficients

## Unbiasedness Step 1: Subsample Means

- We'll start with something we've essentially shown before --- that $\hat\mu(x)$ is an unbiased estimator of $\mu(x)$. 
  - $\E[\hat\mu(x)] = \mu(x)$
- We'll show something stronger than that. We'll show that it's *covariate-conditional mean* is $\mu(x)$.
  - $\E[\hat\mu(x) \mid X_1 \ldots X_n] = \mu(x)$
- Unbiasedness follows from this by the law of iterated expectations.
  - $\E[\hat\mu(x)] = \E[\E[\hat\mu(x) \mid X_1 \ldots X_n]] = \E[\mu(x)] = \mu(x)$


::: fragment
$$
\small{
\begin{aligned}
&\E\qty[\hat\mu(x) \mid X_1 \ldots X_n] \\
&= \E\qty[\frac{\sum_{i:X_i=x} Y_i}{\sum_{i:X_i=x} 1} \mid X_1 \ldots X_n]  && \\
&= \E\qty[\frac{\sum_{i=1}^n 1_x(X_i) Y_i}{\sum_{i=1}^n 1_x(X_i)} \mid X_1 \ldots X_n] && \qfor 1_x(X) = \begin{cases} 1 & \text{ if } x = X  \\ 0 & \text{ otherwise} \end{cases} \\
&= \frac{\sum_{i=1}^n 1_x(X_i) \E[ Y_i \mid X_1 \ldots X_n]}{\sum_{i=1}^n 1_x(X_i)} && \text{ by linearity of conditional expectations } \\
&= \frac{\sum_{i=1}^n 1_x(X_i) \E[ Y_i \mid X_i]}{\sum_{i=1}^n 1_x(X_i)}          && \text{  by irrelevance of independent conditioning variables } \\
&= \frac{\sum_{i=1}^n 1_x(X_i) \mu(X_i)}{\sum_{i=1}^n 1_x(X_i)}                     && \text{  is a matter of notation } \\
&= \frac{\sum_{i=1}^n 1_x(X_i) \mu(x)}{\sum_{i=1}^n 1_x(X_i)}                       && \text{ by the \emph{indicator trick} } 1_x(X_i)\mu(X_i) = 1_x(X_i)\mu(x) \\
&= \mu(x) \frac{\sum_{i=1}^n 1_x(X_i)}{\sum_{i=1}^n 1_x(X_i)} = \mu(x) 
\end{aligned}
}
$$
:::


## Unbiasedness Step 2: Linear Combinations 

$$ \E\qty[\sum_x \alpha(x) \hat\mu(x)] = \sum_x \alpha(x) \mu(x) $$

- Examples
$$
\small{
\begin{aligned}
&\hat \mu(18) - \hat\mu(16) && \qqtext{ is unbiased for } \mu(18) - \mu(16) \\
&\frac14 \sum_{x=9}^{12} \qty{ \hat\mu(x) - \hat\mu(x-1) } &&\qqtext{ is unbiased for } \frac14 \sum_{x=9}^{12} \qty{ \mu(x) - \mu(x-1) }
\end{aligned}
}
$$

- Why is this true?

::: fragment
- Expectation is linear and subsample means are unbiased.

$$ \E\qty[\sum_x \alpha(x) \hat\mu(x)] =  \sum_x \alpha(x) \E\qty[ \hat\mu(x)] = \sum_x \alpha(x) \mu(x) $$

:::

## Variance Calculation Step 1: Total Variance


- Let's talk about the variance of a linear combination of the sample means.
- We'll use the law of total variance to break the variance of our estimator into two parts.
- We'll see that one of these parts --- the $\Var\E[\ldots]$ part --- is zero.
  - This happens because the conditional expectation of our subsample means are constants
  - We saw this before when we discussed differences of subsample means. 
- What we'll get for the remaining $\E[\Var[\ldots]]$ part is familiar.
  - We *sum* the conditional variances of the individual terms, then take the expectation of the result.
  - This is what we did when we calculated the variance of the difference of subsample means.

$$
\small{
\begin{aligned}
&\Var\qty[ \sum_x \alpha(x) \hat \mu(x) ] \\
&= \E\qty[ \Var\qty{ \sum_x \alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] + \Var\qty[ \E\qty{\sum_x \alpha(x) \hat \mu(x) \mid X_1 \ldots X_n} ] \\
&= \E\qty[ \Var\qty{ \sum_x \alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] + \Var\qty[ \sum_x  \alpha(x) \E\qty{ \hat \mu(x) \mid X_1 \ldots X_n} ] \\
&= \E\qty[ \Var\qty{ \sum_x \alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] + \Var\qty[ \sum_x \alpha(x) \mu(x) ] \\
&= \textcolor{blue}{ \sum_x \sigma^2(x) \alpha^2(x) \times \E\qty[ \frac{1}{N_x} ]} + 0
\end{aligned}
}
$$


## Variance Calculation Step 2: $\E[\Var[\ldots]]$ 

$$\small{
\begin{aligned}
&\Var\qty[ \sum_x \alpha(x) \hat \mu(x) ] \\ 
&=\E\qty[ \Var\qty{ \sum_x \alpha(x) \hat \mu(x) \mid X_1 \ldots X_n }] \\
&=\E\qty[ \E\qty{ \qty(\sum_x \alpha(x) \hat\mu(x) - \E\qty[\sum_x \alpha(x) \hat\mu(x) \mid X_1 \ldots X_n])^2 \mid X_1 \ldots X_n }] \\
&=\E\qty[ \E\qty{ \qty(\sum_x \alpha(x) \hat\mu(x) - \alpha(x) \E\qty[\hat\mu(x) \mid X_1 \ldots X_n])^2 \mid X_1 \ldots X_n }] \\
&=\E\qty[ \E\qty{ \qty(\sum_x \alpha(x) \qty{\hat\mu(x) - \E\qty[\hat\mu(x) \mid X_1 \ldots X_n]})^2 \mid X_1 \ldots X_n }] \\
&=\E\qty[ \E\qty{ \qty(\sum_x \alpha(x) Z_x)^2  \mid X_1 \ldots X_n }] \quad \text{ for } \quad  Z_x=\hat\mu(x) - \E\qty[\hat\mu(x) \mid X_1 \ldots X_n] \quad \text{ with } \quad \E[Z_x^2 \mid X_1 \ldots X_n] = \Var[\hat\mu(x) \mid X_1 \ldots X_n] \\
&=\E\qty[ \E\qty{ \sum_x\sum_x' \alpha(x) \alpha(x') Z_x Z_{x'} \mid X_1 \ldots X_n }] \\ 
&=\E\qty[ \sum_x\sum_x' \alpha(x) \alpha(x') \E\qty{ Z_x Z_{x'} \mid X_1 \ldots X_n}] \\ 
&=\E\qty[ \sum_x\sum_x' \alpha(x) \alpha(x') \begin{cases} \Var\qty{\hat\mu(x) \mid X_1 \ldots X_n} = \frac{\sigma^2(x)}{N_x} & \text{ if } x=x' \\ 0 & \text{ otherwise} \end{cases} ] \\
&=\E\qty[ \sum_x \alpha(x)^2 \frac{\sigma^2(x)}{N_x} ] = \sum_x \sigma^2(x) \alpha^2(x) \times \E\qty[ \frac{1}{N_x} ] \\
\end{aligned}
}
$$

## Variance Estimation

$$
\Var[\sum_x \alpha(x) \hat \mu(x)] = \sum_x \sigma^2(x) \alpha^2(x) \times \E\qty[ \frac{1}{N_x} ] \quad \text{ where } \quad \sigma^2(x) = \Var[Y_i \mid X_i=x]
$$

- This has two quantities we don't know.
  - The subpopulation variances $\sigma^2(x)$.
  - The expected inverse subsample size $1 / N_x$.
- But we can estimate both easily.
  - The subsample variances are good esimates of the subpopulation variances
$$
\sigma^2(x) \approx \hat\sigma^2(x) := \frac{1}{N_x} \sum_{i:X_i=x} \qty{Y_i - \hat\mu(x)}^2
$$

  - The inverse subsample size is a good estimate of its expectation.

$$ 1/N_x \approx \E[1/N_x] $$

- Plugging them in gives us an estimate of our summary's variance.

$$
\hat\sigma_{\hat\theta}^2 = \sum_x \hat\sigma^2(x) \alpha^2(x) \times \frac{1}{N_x}
$$



## Summary

- If what we're estimating is a linear combination of the *subpopulation means*, with non-random coefficients
$$ 
\small{
\theta = \sum_x \alpha(x) \mu(x)
}
$$
- Then the same linear combination of *subsample means* is unbiased.
$$
\small{
\hat \theta = \sum_x \alpha(x) \hat\mu(x) \quad \text{ satisfies } \E[\hat\theta] = \theta
}
$$
- Its variance is determined by the variance in the subpopulation, the coefficients, and the subsample sizes. And we can estimate it.
$$
\small{
\Var[ \hat\theta ] = \sum_x \sigma^2(x) \alpha^2(x) \times \E\qty[ \frac{1}{N_x} ] \approx  \sum_x \hat\sigma^2(x)\alpha^2(x) \times \frac{1}{N_x}  =: \hat\sigma_{\hat\theta}^2
}
$$
- We can use this to construct a confidence interval based on normal approximation.
$$
\small{
\theta \in \qty[ \hat \theta - 1.96\hat\sigma_{\hat\theta}, \ \hat \theta + 1.96\hat\sigma_{\hat\theta}] \quad \text{ with probability } \approx .95 \qfor \hat\sigma_{\hat\theta}^2 = \sum_x \hat\sigma^2(x)\alpha^2(x) \times \frac{1}{N_x}  
}
$$

- Fundamentally, we're going to get an imprecise estimate *if a small subsample has a large weight*. 
- By imprecise, we mean a large standard error and a wide confidence interval.


## Application {.smaller}

```{r}
#| out-height: 150px
#| fig-asp: .2

summaries = earnings %>% group_by(education) %>% 
          summarize(mean=mean(income), sd=sd(income))
estimate = (summaries$mean[summaries$education==12] - summaries$mean[summaries$education==8])/4
se = sqrt( summaries$sd[summaries$education==12]^2 / (16 * sum(earnings$education==12)) + summaries$sd[summaries$education==8]^2 / (16 * sum(earnings$education==8)) )

blueline = data.frame(x=c(8,12), y=c(summaries$mean[summaries$education==8], summaries$mean[summaries$education==12]))
aggregate.comparison.plot + geom_line(aes(x=x, y=y), data=blueline, color='blue', alpha=.4) + xlim(7.7,12.5)
```

- Let's use this to estimate the value of a year of high school in our population.
- The one we've addressed is the average, over the four years of high school, of the increments.
$$
\small{
\theta = \sum_{x=9}^{12} \frac14 \qty{ \mu(x) - \mu(x-1) } = \frac14 \qty{ \mu(12) - \mu(8) }
}
$$

- That's the slope of the secant---blue line---in an analogous plot for the population.
- We're estimating it by the slope of the secant in the plot of our sample above.
$$
\small{
\hat\theta = \sum_{x=9}^{12} \frac14 \qty{ \hat\mu(x) - \hat\mu(x-1) } = \frac14 \qty{ \hat\mu(12) - \hat\mu(8) } = `r approx(estimate)`
}
$$

- This is our estimate of its standard error, i.e., the square root of our estimate of its variance.

$$
\small{
\hat\sigma_{\hat\theta}^2 
= \frac{\hat\sigma^2(12) \times \qty(\frac{1}{4})^2}{ N_{12} } + \frac{\hat\sigma^2(8) \times \qty(\frac{1}{4})^2}{ N_{8} } 
= \frac{`r approx(summaries$sd[summaries$education==12])`^2}{16 \times `r sum(earnings$education==12)`} 
+ \frac{`r approx(summaries$sd[summaries$education==8])`^2}{16 \times `r sum(earnings$education==8)`} = `r approx(se^2)` \approx `r approx(se)`^2 
}
$$

## This is the resulting interval.

$$
\small{
\qty[ \hat \theta - 1.96\hat\sigma_{\hat\theta}, \ \hat \theta + 1.96\hat\sigma_{\hat\theta}] = \qty[`r approx(estimate - 1.96*se)` , `r approx(estimate + 1.96*se)`]
}
$$

